### 2015-02-04 ACMTransCL/HK_TOCL.tex
Initial commit

\documentclass[acmtocl]{acmsmall}

\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usepackage{amssymb}
\usepackage{amsfonts,amsmath,latexsym,stmaryrd}
\usepackage{listings}
\usepackage{cite}
\input{macro}
\usepackage{alltt}
\usepackage[all]{xy}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{pdflscape}
\usepackage{booktabs}
\usetikzlibrary{shapes,snakes}
% 
% \usepackage{theorem}
% 
% \theorembodyfont{\rmfamily}
% \newtheorem{Thm}{Theorem}
% \newtheorem{Defn}[Thm]{Definition}
% \newtheorem{Ex}[Thm]{Example}
% \newtheorem{Lem}[Thm]{Lemma}
% \newtheorem{Prop}[Thm]{Proposition}
\newtheorem{PS}[theorem]{Proof Strategy}
\newtheorem{PEM}[theorem]{Proof Exploration Method}
% \newtheorem{GR}{General Remark}
% \newtheorem{Alg}{Algorithm}

\begin{document}

\markboth{J. Heras and E. Komendantskaya}{Proof-Pattern Search in Coq/SSReflect}

\title{Proof-Pattern Search in Coq/SSReflect}

%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{J\'onathan Heras
\affil{Department of Mathematics and Computer Science, University of La Rioja, Spain}
Ekaterina Komendantskaya
\affil{School of Computing, University of Dundee, UK}
}
%

\begin{abstract}
ML4PG is an extension of the Proof General interface, allowing
the user to invoke machine-learning algorithms and find proof similarities
in Coq/SSReflect libraries. In this paper, we present four new improvements to
%techniques that enhance the quality of the results obtained by 
ML4PG.
First, a new method of \lq\lq{}recurrent clustering" is introduced to collect statistical features from Coq terms.
Now, the user can receive suggestions about similar definitions,
types and lemma statements, in addition to proof strategies. Second, Coq
proofs are split into patches to capture proof strategies that could arise
at different stages of a proof. Third, we present a method to automatically generate Coq proofs based on ML4PG's output. Finally,
we introduce several visualisation tools that help to interpret the families of similar theorems/terms generated by ML4PG.
\end{abstract}

\category{F.4.1}{Mathematical Logic and Formal Languages}{Mathematical Logic -- Mechanical theorem proving}
\category{D.2.4}{Software Engineering}{Software/Program Verification -- Formal methods.}

\terms{}

\keywords{Coq/SSReflect, Proof Patterns, Recurrent Clustering, Pattern Recognition, Feature Extraction.}

\acmformat{J\'onathan Heras and Ekaterina Komendantskaya, 2014.
Proof-Pattern Search in Coq/SSReflect.}

\begin{bottomstuff}
The work was supported by EPSRC grants EP/J014222/1 and EP/K031864/1.

Author's addresses: J\'onathan Heras and Ekaterina Komendantskaya,
School of Computing, University of Dundee, UK
\end{bottomstuff}

\maketitle

\input{introduction}

\input{lemmaclustering}

\input{rec}

\input{patches}

\input{evaluation}

\input{visualisation}

\input{relatedwork}

\input{conclusions}

%\section{Acknowledgements}

\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{biblio}

\pagebreak

\appendix

\input{gallina}

\input{coqtactics}

\input{automaton}

\end{document}

### 2015-02-04 ACMTransCL/conceptualisation.tex
Initial commit

\section{Visualisation tools in ML4PG}\label{sec:visualisation}

In the previous section, we have seen how ML4PG can be used to find families of proofs following 
a common proof pattern. However, the output provided by clustering algorithms is just a set
of similar patches with no hints of why these proof-patches are deemed similar. 
% In addition, the user needs
% to replay the proof to infer the pattern. %Therefore, the conceptualisation of proof-patterns can be
% %a challenge. 
In this section, we present our approach to facilitate the understanding of proof patterns.

The first problem that we address is the discovery of the key features that were taken
into account during the cluster formation. This is a well-known problem in machine-learning 
known as \emph{feature selection}~\cite{Weka}. From a given set of features, feature selection algorithms 
generate different subsets of features and create a ranking of feature subsets.
ML4PG uses the \emph{correlation-based feature subset selection} algorithm implemented in Weka~\cite{Weka}
-- the machine-learning toolbox employed by ML4PG to extract the relevant features.

\begin{example}\label{ex:correlation-features}
In the case study presented at the end of the previous section, the relevant features were:

$-$ The tactics applied in the first and second steps of the proof-patches (\lstinline?case? and \lstinline?rewrite? respectively).

$-$ The type of the argument of the tactic applied in the first step of the proof-patches (\lstinline?nat?).

$-$ The second top symbol of the goal in the second step of the proof-patches (the $\sum$ symbol).

$-$ The first and second auxiliary lemmas applied in the the third step of the proof-patches. The first auxiliary lemma (\lstinline?sumrB?) is used to expand the summations and is 
common to all the proofs. There are two different \lq\lq{}second\rq\rq{} auxiliary lemmas that occur in the proofs of Lemma~\ref{lem:fundamental} and~\ref{lem:nilpotent2} (\lstinline?big_nat_recr? and
\lstinline?big_nat_recl? -- both extract elements of a summation); however, the recurrent feature extraction process have assigned similar values to them.
The ``automaton'' of Example~\ref{ex:correlation-features} is depicted in Figure~\ref{fig:automata} and Appendix~\ref{sec:automaton}.
\end{example}

ML4PG uses this information and produces an automaton-shape representation for discovered % to help in the visualisation of 
proof-patterns and the correlated features that determined the patterns. 
Our graphical representation is simpler than other works where automata are used to represent models that are inferred from proof traces~\cite{GWR14}.  
% This feature selection mechanism helps to explain why the proof-patches are similar since it indicates the 
% features that have been relevant for the construction of the cluster. To show the selected features and their relation with 
% the proof-patches, we introduce an automaton-shape representation. In~\cite{GWR14}, a tool to infer models 
% from proof traces was presented. Those models help to understand the behaviour exhibited by a collection
% of proofs and is represented by automata.

Generally, given a cluster of proof patches ${\cal C}$, we have an automaton $A$ with $5$ consecutive states. The $i$th state of $A$ is labelled
with the list of $i$th goals in the proof-patches contained in ${\cal C}$. The transitions between the $i$th and $i+1$th states are given by 
the $i$th tactic of each proof-patch of ${\cal C}$. If two or more tactics belong to the same group (see Figure~\ref{tab:tactics}),
they are merged in a unique transition; otherwise, the tactics will be shown as different transitions. In addition, each state is annotated with 
features 
whose correlation determined the cluster.
%for clustering that are related to that 
%state.

% $-$ there are several transitions from the $j$th to the $j+1$th state labelled by the $j$th tactic of each proof-patch of $P$ -- if either the $j$th tactic of two or more
% proof-patches is the same or belong to the same group (see Table~\ref{tab:tactics}), the transitions are merged in a unique transition labelled with the set of those tactics,
% 
% $-$ if a proof ends after the $j$th tactic of a proof-patch, the $j+1$ state is represented as final.

%\begin{example}

\begin{figure}[t]
\centering 
\includegraphics[scale=.19]{itp6.png}
\caption{\scriptsize{\emph{Fragment of the automaton 
 corresponding to the cluster of four lemma fragments described in the case study of Section~\ref{sec:recurrent}} (the whole automaton can be seen in Appendix~\ref{sec:automaton}). 
It shows correspondence between certain proof steps of lemmas % the following lemmas were discovered:
\texttt{telescope} (for the lemma of Example~\ref{example0}), \texttt{nilpotent\_inverse} (for Lemma~\ref{lem:nilpotent2}), and \texttt{persistent\_homology} 
(for Lemma~\ref{lem:fundamental}). Square boxes denote feature correlation where it exists. 
To be compact, we  hide full lemma statements, tactic and auxiliary lemma names, symbols, etc., but they can be shown by ML4PG. 
%For example, the full statements of the following cluster lemmas will be shown:
%The names of the lemmas of that cluster are: 
}}\label{fig:automata}
\end{figure}

%\end{example}

### 2015-02-04 ACMTransCL/conclusions.tex
Initial commit

\section{Conclusions}\label{sec:conclusions}

We have presented several enhancements of the ML4PG system. \emph{Term clustering} adds a new 
functionality to ML4PG: the user can receive suggestions about families of similar definitions, types and lemma shapes (in fact, any Coq terms). 
The \emph{proof-patch method} is employed to analyse the properties of the patches that constitute a proof. The whole syntax of Coq libraries is now subject to \emph{recurrent clustering}, which %combination
%of term clustering and proof-patches
 increases the number and accuracy of families of similar proofs suggested by ML4PG. In addition, this information is taken into account in an automatic proof-generation method. Finally, the different visualisation tools implemented in ML4PG facilitates the interpretation of clusters of similar proof-patches and terms.

Further improvements in \emph{accuracy} (e.g. including proof contexts into the analysis) and \emph{conceptualisation} for clustered terms and proofs are planned. The families of similar proofs and terms can be the basis to
apply symbolic techniques to, for instance, infer models from proof traces~\cite{GNR14}, or generate auxiliary results using
mutation of lemmas~\cite{lpar13}. We expect that the incorporation of these techniques help in the goal pursued by ML4PG: make the 
proof development easier.

%  \textbf{Related Work}
%  on using statistical machine-learning in ITPs was focused on speeding up proof automation in different ITPs~\cite{lpar-urban,mash,UrbanSPV08}. 
%  Comparing to these tools, we use unsupervised, rather than supervised, learning; and we do not use sparse  machine-learning methods.
%  (See also \cite{KHG13} for a detailed comparison of different machine learning tools applied in various theorem provers.)
%  We do not have a quantitative target when it comes to improving \emph{interactive} proof building experience: 
%  no longer speed up in automated proof search or the number of automatically proven theorems are the main criteria of success.
%  Instead, the user experience is the main parameter we target.
%  We generally follow the ``qualitative'' intuition that ML4PG, being an interactive hint generator, must provide interesting and non-trivial hints on user's demand,
%  and should be flexible and fast enough to do so in real  time,  at any stage of the proof, and relative to any chosen proof library.

### 2015-02-04 ACMTransCL/evaluation.tex
Initial commit

\section{Evaluation}\label{sec:evaluation}

In the previous sections, we have shown how ML4PG has been designed to provide \emph{interesting} and \emph{non-trivial} hints on user's demand, 
and to be flexible enough to do so at any stage of the proof, and relative to any chosen proof library. 
However, it is difficult to measure how ML4PG improves the interactive proof building experience, since
the usability of a hint varies from user to user. In this section, we present a quantitative method to evaluate
how useful ML4PG can be during the proof development.

Machine-learning techniques have been previously used for the automatically generation of proofs in 
ITPs, see~\cite{Duncan02,DixonF03,GNR14}. ML4PG was not initially designed with this aim;
but we can use the information obtained from clustering to automatically generate proof attempts for a given theorem.

\begin{PEM}\label{pem}
Given the statement of a theorem $T$ and a set of lemmas $L=\{L_i\}_i$, we can use ML4PG to find a proof for $T$ as follows:
\begin{enumerate}
 \item Using term-clustering and $T\cup L$ as dataset, obtain the cluster $C$ that contains the theorem $T$ (i.e. $C=T\cup \{L_j\}_j$ where $\{L_j\}_j$ is the set of lemmas similar to $T$).
 \item Obtain the sequence of tactics $\{T_1^j,\ldots,T_{n_i}^j\}_{j}$ used to prove each lemma $L_j$ in $C$.
 \item For each $j$, try to prove $T$ using $T_1^j,\ldots,T_{n_j}^j$.
 \item If no sequence of tactics prove $T$, then for each tactic use ML4PG to infer the argument for each tactic $T_k^j$:
    \begin{itemize}
     \item If the argument of $T_k^j$ is an internal hypothesis from the context of a proof, try all the internal hypothesis from the context of the current proof. 
     \item If the argument of $T_k^j$ is an external lemma $E$, use term-clustering and $L$ as dataset to compute the lemmas in the same cluster as $E$ and try all those lemmas.
     \item[***] This can be naturally extended to tactics with several arguments, just trying all the possible combinations. 
  \end{itemize}    
\end{enumerate}
\end{PEM}

\begin{example}\label{ex:maxnACA}
Let $T$ be the lemma \lstinline?maxnACA? that states the inner commutativity of the maximum of two natural numbers in the SSReflect library:

\begin{lstlisting}
Lemma maxnACA : interchange maxn maxn. 
\end{lstlisting}

\noindent and $L$ be the \lstinline?ssrnat? library of SSReflect. Using Proof Exploration Method~\ref{pem}, we can construct a 
proof of \lstinline?maxnACA? as follows.

\begin{enumerate}
 \item $T$ belongs to the cluster containing the lemmas $\{$\lstinline?addnACA?, \lstinline?minnACA?, \lstinline?mulnACA?$\}$ --- these
 3 lemmas state the inner commutativity of addition, multiplication and the minimum of two naturals respectively. 
 \item For this example, we will only consider the proof of the lemma \lstinline?addnACA? that is proven using the sequence of tactics \lstinline?by move=> m n p q; rewrite -!addnA (addnCA n).?
 \item The proof of \lstinline?maxnACA? fails using the sequence of tactics from \lstinline?addnACA?.
 \item The proof of \lstinline?addnACA? uses the lemmas \lstinline?addnA? and \lstinline?addnCA?. If we cluster these lemmas with the
 rest of the lemmas of the ssrnat library we find the cluster $\{$\lstinline?minnA?, \lstinline?mulnA?, \lstinline?maxnA?$\}$ for 
 \lstinline?addnA?, and the cluster $\{$\lstinline?minnAC?, \lstinline?mulnAC?, \lstinline?maxnAC?$\}$ for 
 \lstinline?addnAC?. 
 \item Using the lemmas \lstinline?maxnA? and \lstinline?maxnAC?, we construct the sequence of tactics \lstinline?by move=> m n p q; rewrite -!maxnA (maxnCA n).? that proves the lemma \lstinline?maxnACA?.
\end{enumerate}

\end{example}

Using 5 Coq theories of varied sizes, we perform an empirical evaluation of our proof exploration method. Our test data consists of the Basic infrastructure of SSReflect library~\cite{SSReflect}, the formalisation about Java-like bytecode presented in~\cite{HK14}, the formalisation of persistent homology~\cite{HCMS12}, the paths library of the HoTT development~\cite{hottbook}, and two formalisations of Nash Equilibrium~\cite{Ves06,nash}. Using Proof exploration method~\ref{pem}, we try to prove every lemma of these libraries with a fully-honest approach (following the terminology from~\cite{KaliszykU14}): clustering is only performed against the lemmas that have previously
proven in the given library. In addition, we study the impact of changing the granularity value.

The results of our experiments can be seen in Table~\ref{tab:reproven}. The success rate of the proof exploration method 
depends on how similar are the proofs of theorems in a given library. This explains the high success rate in the \lstinline?Paths?
library, where most of the lemmas are proven using almost the same sequence of tactics, and the low rate in the Persistent Homology 
library, where just a few lemmas are similarly proven. The granularity value does not have a big impact in the performance of 
our experiments, and almost the same amount of lemmas are proven with the different values. In some cases, like in the Nash equilibrium
library, a small granularity value generates bigger clusters that increase the exploration space allowing to prove more lemmas. 
However, reducing the granularity value can also have a negative impact; for instance, in the JVM library the number of clusters
is reduced and this means a reduction of the proven theorems.

The evaluation method presented in this section shows an estimation of how ML4PG can help during the proof development. 
This evaluation is a bit artificial since it heavily relies on having a well-developed background theory. For instance, 
in Example~\ref{ex:maxnACA}, the proof exploration method is able to automatically generate a proof for \lstinline?maxnACA? since
the lemmas \lstinline?maxnA? and \lstinline?maxnAC? were already in the library, but this is not a common scenario when the user
is creating his library. However, ML4PG can be useful for the user even if those two lemmas are not available: ML4PG can suggest
that \lstinline?maxnACA? is similar to  \lstinline?addnACA?, and from the proof of the latter theorem, the user can infer that he needs
to state the lemmas \lstinline?maxnA? and \lstinline?maxnAC? that were missing in the library. The automatic generation of lemmas based
on clustering information was studied for the ACL2 system in~\cite{lpar13} --- the extrapolation of the technique presented in~\cite{lpar13} to Coq is not trivial since ACL2 works with first-order logic and Coq works with higher-order logic.

% 
% \begin{itemize}
%  \item Basic infrastructure of SSReflect library~\cite{SSReflect}: 1389 theorems formalised using SSReflect.
%  \item Formalisation about Java-like bytecode~\cite{HK14}: 49 theorems formalised using SSReflect.
%  \item Formalisation of Persistent Homology~\cite{HCMS12}: 306 theorems formalised using SSReflect.
%  \item Library Paths of HoTT development~\cite{hottbook}: 80 theorems formalised using plain Coq.
%  \item Formalisation of Nash Equilibrium~\cite{Ves06,nash}: 145 theorems formalised using plain Coq.
% \end{itemize}

\begin{table}
\tbl{\scriptsize{\emph{Percentage of automatically re-proven theorems.}}\label{tab:reproven}}{
\centering
\begin{tabular}{cccccc}
 \hline
 Library & Language & Granularity 1 & Granularity 3 & Granularity 5 & Theorems \\
 \hline

 SSReflect library  & SSReflect & $36\%$ & $35\%$ & $28\%$ & 1389\\

  JVM & SSReflect  & $56\%$ & $58\%$ & $65\%$ & 49\\

 Persistent Homology & SSReflect  & $0\%$ & $10\%$ & $12\%$ & 306\\

 Paths (HoTT) & Coq & $92\%$ & $91\%$ & $94\%$ &80\\

 Nash Equilibrium & Coq & $40\%$ & $37\%$ & $36\%$ & 145\\
   \hline
\end{tabular}}

\end{table}

### 2015-02-04 ACMTransCL/gallina.tex
Initial commit

\section{Formula for Gallina tokens}\label{sec:gallinasyntax}

We split Gallina tokens into the following groups.

\begin{itemize}
 \item Group 1: \lstinline?forall?, \lstinline?->?.
 \item Group 2: \lstinline?fun?,
 \item Group 3: \lstinline?let?, \lstinline?let fix?, \lstinline?let cofix?.
 \item Group 4: \lstinline?fix?, \lstinline?cofix?.
 \item Group 5: \lstinline?@?,
 \item Group 6: \lstinline?match?, \lstinline?if?.
 \item Group 7: \lstinline?:=?, \lstinline?=>?, \lstinline?is?. 
 \item Group 8: \lstinline?Inductive?, \lstinline?CoInductive?.
 \item Group 9: \lstinline?exists?, \lstinline?exists2?.
 \item Group 10: \lstinline?:?, \lstinline?:>?, \lstinline?<:?, \lstinline?%?.

\end{itemize}

The formula for the $j$th Gallina token of the $n$th group is given by the formula $$- (n + \sum_{i=0}^j \frac{1}{10\times 2^{i-1}})$$

### 2015-02-04 ACMTransCL/introduction.tex
Initial commit

\section{Introduction}\label{sec:introduction}

Development of Interactive Theorem Provers (ITPs) has led to the creation of big libraries and varied infrastructures for formal
mathematical proofs. 
These frameworks usually involve thousands of definitions and theorems (for instance, there are approximately
4200 definitions and 15000 theorems in the formalisation of the Feit-Thompson theorem~\cite{FTT}). % Definitions,
%theorems and proof-strategies used in
Parts of  those libraries can often be re-applied in new domains; however, 
it is a challenge for expert and non-expert users alike to trace them, and find re-usable concepts and proof ideas.

% 
% \begin{figure}
%  \centering
%  \begin{tikzpicture}[scale=.5]
%  
% \draw[fill=gray,draw=gray] (-1.95,.45) rectangle (2.05,-.55);  
% \draw[fill=white] (-2,.5) rectangle (2,-.5); 
% \node (0,0) {{\scriptsize Proof General} };
%  
% 
% \draw[fill=gray,draw=gray] (7.95,.45) rectangle (12.15,-.55);  
% \draw[fill=white] (7.9,.5) rectangle (12.1,-.5); 
% \draw (10,0) node{{\scriptsize MATLAB/Weka}};
% 
% 
% \draw (5,0) node{\textbf{ML4PG}};
% 
% 
% \draw[latex-,shorten <=2pt,shorten >=2pt,dashed] (7.9,.4) .. controls (6,1) and (4,1) .. (2,.4); 
% \draw (5,1.3) node[anchor=north,fill=white]{\emph{\tiny{feature extraction}}};
%  
% \draw[latex-,shorten <=2pt,shorten >=2pt,dashed] (2,-.4) .. controls (4,-1) and (6,-1) .. (7.9,-.4); 
% \draw (5,-1.3) node[anchor=south,fill=white]{\emph{\tiny{proof families}}}; 
%  
%  
% \draw[fill=white,rounded corners] (-2,-2) rectangle (2,-3); 
% \draw (0,-2.3) node{{\tiny Interactive Prover:}} ;
% \draw (0,-2.7) node{{\tiny Coq, SSReflect}} ;
% 
% 
% \draw[dotted,shorten <=2pt,shorten >=2pt,-latex] (0,-.5) -- (0,-2);
% \draw[dotted,shorten <=2pt,shorten >=2pt,-latex] (10,-.5) -- (10,-2); 
%  
% \draw[fill=white,rounded corners] (7.7,-2) rectangle (12.3,-3); 
% \draw (10,-2.3) node{{\tiny Clustering:}} ;
% \draw (10,-2.7) node{{\tiny K-means, Gaussian, \ldots}} ;
%  
%  
% \end{tikzpicture}
% \caption{\emph{ML4PG architecture.}}\label{fig:ml4pg-architecture}
% \end{figure}

% 
% Coq/SSReflect already provides comprehensive search mechanisms to search the corpus of results available in different libraries.
% There are several search commands in Coq:  \lstinline?Search?~, \lstinline?SearchAbout?, \lstinline?SearchPattern?
% and \lstinline?SearchRewrite?~\cite{Coq}. In addition, SSReflect implements its own version of the \lstinline?Search?
% command \cite{SSReflect} -- SSReflect's \lstinline?Search? gathers the functionality of the 4 Coq's search commands.
% The Whelp platform~\cite{AspertiGCTZ04} is a web search engine for mathematical knowledge formalised in Coq, which features 3 functionalities:
% \lstinline?Match? (similar to Coq's \lstinline?Search? command), \lstinline?Hint? (that finds all the theorems which can
% be applied to derive the current goal) and \lstinline?Elim? (that retrieves all the eliminators of a given type).
% %In this section, we compare ML4PG with these tools.  
% 
% %Let us explain first when these tools are usually invoked. 
% The existing search mechanisms can be used in two different scenarios. If the user knows how to continue the proof, but
% he does not remember (or know) the concrete name of the desired auxiliary lemma, it suffices to provide a search pattern to Coq searching engines, e.g.
% using commands of the form
% ``\lstinline?Search "distr"? \lstinline?in bigop?'' or   
% ``\lstinline?Search _ (_ * (\big[_/_]_(_ <- _| _) _))?'',
%  where \texttt{bigop} is a library, \texttt{"distr"} is a pattern in a lemma name, and  \lstinline? _ (_ * (\big[_/_]_(_ <- _| _) _))?
% is a pattern for search.
%  
% %\begin{example}\label{ex:distr}
% % In the proof of Lemma~\ref{lem:nilpotent}, we may want to introduce the term $(1-M)$ inside the sum $\sum_{i=0}^{n-1} M^i$, but 
% % if we have no experience with the bigop library of SSReflect, it is likely that we do not know the name of the concrete lemma. 
% % The lemma which carries out this task can be find using, for instance, the commands \lstinline?Search "distr" in bigop.? or  
% %\lstinline?Search _ (_ * (\big[_/_]_(_ <- _| _) _))?. 
% %\end{example}
% In the second scenario, the user needs help in the middle of the proof. Searching mechanisms can be also useful in this 
% case; for instance, using the \lstinline?Search? command the user can search all the lemmas associated with a concrete term
% of the current goal. The \lstinline?Hint? mechanism of Whelp can find all the theorems which can be applied to derive the current goal.
% 
% %\begin{example}\label{ex:fact}
% % In the proof of Lemma~\ref{lem:factorial}, we can find all the lemmas related to the factorial function using the command
% % \lstinline?SearchAbout factorial?. 
% %\end{example}
% 
% %ML4PG is helpful in the second scenario, since it provides families of similar proofs. Inspecting that family of proofs, 
% %the user can find a pattern to finish the proof that she is formalising. 
% 
% %As we have seen in Examples~\ref{ex:distr} and~\ref{ex:fact}, 
% The above search mechanisms are goal directed and deterministic. That is, the user
% searches  the chosen  libraries for lemmas related to a concrete type, term or pattern. 
% If the patterns defined by the user are present in the given library, then the user is guaranteed to see the relevant lemmas on the screen.
% 
% The situation is more complicated if the user does not know the right pattern to search for. Imagine, for example, being in the middle of constructing a proof, 
% and wishing to get some higher-level hint on how to proceed, wishing there was an ITP expert near, who would suggest a further proof strategy based on his previous experience.

The ML4PG (``Machine-Learning for Proof-General'') tool~\cite{KHG13,CICM13,HK14} was created to help the user in such a situation for the particular case of the Coq system~\cite{Coq} and its SSReflect extension~\cite{SSReflect} developments. ML4PG uses statistical machine-learning algorithms to discover re-usable proof-patterns based on ``previous experience'' acquired from previous developments. A proof-pattern was defined as a correlation between the tactics and the types/shapes of subgoals resulting from the tactic applications within a few proof steps. The resulting tool could indeed find some interesting --- unexpected and yet relevant --- proof-patterns,
across different notation, types, and libraries. Our experiments spanned several subjects: basic mathematical infrastructures, Computer Algebra, Game Theory, and the certification of Java-like bytecode. The results are best summarised in~\cite{CICM13,HK14}.

% 
% In comparison to the more traditional \emph{search engines}, ML4PG is a goal-independent (unsupervised) tool, i.e. the user does not have to know the required pattern in advance.
% But, as many 
% statistical machine-learning applications, it is also non-deterministic. That is, the tool failing to suggest a proof pattern does not mean there is no ``interesting'' pattern to be found.
% Actually, the notion of a proof pattern being ``interesting'' is left to the user's judgement, as well.

That initial approach had two inherent limitations. First, the essence of a Coq/SSReflect proof is not fully expressible by a sequence of applied tactics. The definitions, types, and shapes of auxiliary lemmas used in a proof can be much more sophisticated and conceptual than a proof script calling them. Therefore, although ML4PG could find interesting, and often useful, sequences of tactics; it could not go further to recognise, for instance, similar definitions. Second, the notion of a proof-pattern being ``interesting'' or useful is left to the user's judgement.

In this paper, we present the most recent extensions to ML4PG  involving all kinds of Coq terms --- type declarations, definitions and lemma statements --- into the process of proof-pattern search. This required additional algorithms of feature extraction that reflect the mutual dependency of various proof objects in Coq\rq{}s dependently-typed setting; see Sections~\ref{sec:lemmaclustering} and~\ref{sec:reclemmaclustering}. 
This major step in ML4PG development prompted other improvements. The initial ML4PG was considering features arising from the first 5 proof-steps in a proof, whereas now we treat every proof as a collection of
proof patches, each potentially
representing an interesting proof strategy. Moreover, if say 15th-20th step in one proof resembles a 115th-120th step in another, the tool is now able to detect such patterns deep down.  
The feature extraction algorithms for proof features have been further refined to include the data collected from Coq terms, and now the 
whole syntax of the chosen proof libraries is subject
to \emph{recurrent clustering} --- a novel technique for ML4PG. %, but see~\cite{lpar13}. 
All these extensions to proof-feature extraction are explained in Section~\ref{sec:recurrent}.

ML4PG used to show, in response to the user's call, a set of similar proofs,
with no hints of why these proofs are deemed similar. We now introduce two extensions that improve ML4PG's user-experience. 
First, we have designed a method to automatically proof theorems based on ML4PG clustering methods (cf. Section~\ref{sec:evaluation}). Additionally, several visualisation tools (using automaton-shape or tree-shape representations) have been developed to show the proof-features that correlate, see Section~\ref{sec:visualisation}. This partially addresses the drawback of the subjective approach to 
the pattern's ``interestingness'' --- now 
the tool clearly declares correlation of which proof features defined the suggested proof pattern. 
Finally, in Section~\ref{sec:relatedwork}, we compare ML4PG with other machine-learning and searching approaches available in the literature, and conclude the paper in Section~\ref{sec:conclusions}.

Examples we use throughout the paper come from several Coq and SSReflect libraries: the basic infrastructure of SSReflect~\cite{SSReflect}, a matrix library~\cite{GarillotEtAl09}, a formalisation of persistent homology~\cite{HCMS12}, the HoTT library~\cite{hottbook}, two formalisation related to Nash equilibrium~\cite{Ves06,nash}, and the formalisation about Java-like bytecode presented in~\cite{HK14}. ML4PG is a part of standard Proof General distribution; the novel features we present here are available at~\cite{HK12}. % and can be used in any Coq or SSReflect development.

%On the contrary, the proof patterns found by
%ML4PG are unknown in advance. In this way, ML4PG can find some unexpected results but which can still be useful for the user.

%\begin{example}\label{ex:cancel}
 % The command \lstinline?Search _ (\sum_(_ <- _) (_ - _))?  can find Lemmas~\ref{lem:lemma3} and~\ref{lem:fundamental}; this can 
 % be useful in the proof of Lemma~\ref{lem:nilpotent} since the three lemmas follow the same pattern. However, Lemma~\ref{lem:nilpotent2}
 % whose proof is closer to the proof of Lemma~\ref{lem:nilpotent} (see Table~\ref{tab:compare}) cannot be found using the previous search command
 % or other commands involving the terms of Lemma~\ref{lem:nilpotent}: \lstinline?Search _ (\sum_(_ <- _) _ *m _)?, 
 % \lstinline?Search _ (_ *m (_ - _))? or \lstinline?Search _ (_ *m _)? (this last command finds Lemma~\ref{lem:nilpotent2}, but there are 
 % approximately 120 additional suggestions).
%\end{example}

%Finally, we can consider the relevance of suggestions provided by search mechanisms and ML4PG. In both cases, some of the results returned by these tools
%can be useless; however, they can be refined modifying the input of the search mechanisms or changing some parameters in ML4PG.

%\begin{example}\label{ex:relevance}
% In Benchmark 1, a search of all the lemmas involving big operations produce more than $300$ lemmas. If we refine the search to lemmas involving $\sum$,
% we can find $250$ lemmas; and the command \lstinline?Search _ (\sum_(_ <- _) (_ - _))? produces $4$ lemmas. In the same way, we have seen in Table~\ref{tab:compare}
% that the modification of parameters such as the granularity or the algorithm produce better results. %
%\end{example}

%The main contributions of this paper are:

%\begin{itemize}
% \item a new feature extraction method based on the shape of Coq terms, see Section~\ref{sec:lemmaclustering}.
% \item an extension of ML4PG's feature extraction method~\cite{KHG13} to capture information from complete 
% proofs of varied sizes using patches, see Section~\ref{sec:patches}.
% \item an evaluation of these techniques in one of the most relevant developments in theorem proving: 
% the Feit-Thompson theorem~\cite{FTT}, see Section~\ref{sec:evaluation}.

%\end{itemize}

### 2015-02-04 ACMTransCL/lemmaclustering.tex
Initial commit

\section{Feature Extraction for Coq terms}\label{sec:lemmaclustering}

ML4PG uses (unsupervised) \emph{clustering algorithms}~\cite{Bishop} to find patterns in Coq syntax and proofs.
Clustering algorithms % is a set of machine-learning techniques that 
divide data into $n$ groups of similar objects (called \emph{clusters}), where the value of $n$
is a parameter provided by the user. In ML4PG, the value of $n$ is automatically computed depending on the number 
of objects to cluster, and using the formula provided in~\cite{KHG13,lpar13}. 
A detailed exposition of machine-learning algorithms involved in ML4PG can be found in \cite{KHG13}. In this paper, our first goal consists in improving ML4PG's feature extraction algorithm.
%In this section, we present a method to collect features from Coq terms.

\emph{Feature extraction}~\cite{Bishop} is a research area developing methods for 
discovery of statistically significant features in data. % in machine-learning, known as . 
We adopt the following standard terminology. % concerning feature extraction. 
We assume there is a training data set, containing some samples (or objects). 
\emph{Features} are given by a set of statistical parameters chosen to represent all objects in the given data set.
If $n$ features are chosen, one says that object classification is 
conducted in an $n$-dimensional space. 
For this reason, most pattern-recognition tools will require that %the features have numeric values; and 
the number of selected features is limited and fixed (sparse methods, like the ones applied in
e.g.~\cite{lpar-urban,K13,UrbanSPV08} are the exception to this rule). 
\emph{Feature values} are rational numbers used to instantiate the features for every given object.  
If an object is characterised by $n$ feature values, these $n$ values together form a \emph{feature vector} for this object. 
%Irrespective of the particular algorithm
%used, 
A function that assigns, to every object of the data set, a feature vector is called a \emph{feature extraction function}.
Normally, feature extraction is a data pre-processing stage, and it is separated from the actual pattern-recognition process. %ML4PG relies on unsupervised clustering methods~\cite{KHG13}.

Feature extraction from terms or \emph{term trees} is common to most feature extraction algorithms implemented
in theorem provers~\cite{lpar13,lpar-urban,K13,UrbanSPV08}. In~\cite{lpar13}, we introduced a feature extraction mechanism
for ACL2 first-order terms. Here, that ACL2 method is substantially re-defined to capture the higher-order dependently-typed language of Coq.
%However, due to the different nature of ACL2 and Coq -- ACL2 being a first-order untyped language, and Coq being a higher-order
%dependently-typed language -- the algorithm presented in~\cite{lpar13} cannot be extrapolated to Coq, and a different method is necessary.
%Therefore, we introduce here a new feature-extraction method for Coq terms.

The underlying formal language of Coq is known as the \emph{Predicative Calculus of (Co)Inductive Constructions} (pCIC)~\cite{Coq,CoquandH88,CoPa89}. % -- a 
%weaker calculus of the \emph{Calculus of (Co)Inductive Constructions}. 
The terms of pCIC are built from
the following rules:

\begin{definition}[pCIC term]\label{def:coqterms}

$-$ The sorts \lstinline?Set?, \lstinline?Prop?, \lstinline?Type(i)? (\lstinline?i?$\in \mathbb{N}$) are terms.

$-$ The global names of the environment are terms.

$-$ Variables are terms.

$-$ If \lstinline?x? is a variable and \lstinline?T, U? are terms, then \lstinline?forall x:T,U? 
 is a term. If \lstinline?x? does not occur in \lstinline?U?, then \lstinline?forall x:T,U? will be written as \lstinline?T -> U?. A term of the 
 form \lstinline?forall x1:T1, forall x2:T2, ..., forall xn:Tn, U? will be written as \lstinline?forall (x1:T1) (x2:T2) ...(xn:Tn), U?.

$-$ If \lstinline?x? is a variable and \lstinline?T, U? are terms, then \lstinline?fun x:T => U? 
 is a term. A term of the form \lstinline?fun x1:T1 => fun x2:T2 => ... => fun xn:Tn => U? will be written as \lstinline?fun (x1:T1) (x2:T2) ...(xn:Tn) => U?.

 $-$ If \lstinline?T? and \lstinline?U? are terms, then \lstinline?(T U)? is a term -- we use an uncurried notation (\lstinline?(T U1 U2 ... Un)?) 
 for nested applications (\lstinline?(((T U1) U2) ... Un)?).

 $-$ If \lstinline?x? is a variable, and \lstinline?T, U? are terms, then (\lstinline?let x:=T in U?) is a term.

\end{definition}

The syntax of Coq terms~\cite{Coq} includes some terms that do not appear in Definition~\ref{def:coqterms}; e.g. given 
 a variable \lstinline?x?, and  terms \lstinline?T? and \lstinline?U?, \lstinline?fix name (x:T) := U? is a Coq term used to declare a recursive definition.
The notion of a term in Coq covers a very general syntactic
category in the Gallina specification language~\cite{Coq} and corresponds to the intuitive notion of well-formed expression. 
%Our feature extraction
%method works for all well-formed expressions that can be defined in Coq; 
However, for the purpose of concise exposition, % purposes,
we will restrict our notion of a term to Definition~\ref{def:coqterms},
giving the full treatment of the whole Coq syntax in the actual ML4PG implementation.
 % %definitions to the syntax of Coq terms coming from pCIC terms, thereby remembering that full Coq syntax is handled in the actual ML4PG implementation.

\begin{definition}[ML4PG term tree]\label{def:ml4pgtermtree}
Given a Coq term \lstinline?C?, we define its associated term tree as follows:

$-$ If \lstinline?C? is one of the sorts \lstinline?Set?, \lstinline?Prop? or \lstinline?Type(i)?, then the term tree of 
 \lstinline?C? consists of one single node, labelled respectively by \lstinline?Set:Type(0)?, \lstinline?Prop:Type(0)? or \lstinline?Type(i):Type(i+1)?.

$-$ If \lstinline?C? is a name or a variable, then 
 the term tree of \lstinline?C? consists of one single node, labelled by the name or the variable itself together with its type.

$-$ If \lstinline?C? is a term of the form \lstinline?forall (x1:T1) (x2:T2) ...(xn:Tn), U? (analogously for \lstinline?fun (x1:T1) (x2:T2) ...(xn:Tn) => U?); 
then, the term tree of \lstinline?C? is the tree with the root node labelled by \lstinline?forall? (respectively \lstinline?fun?) 
and its immediate subtrees given by the trees representing \lstinline?x1:T1?, \lstinline?x2:T2?, \lstinline?xn:Tn? and \lstinline?U?.

$-$ If \lstinline?C? is a term of the form \lstinline?let x:=T in U?, then the term tree of \lstinline?C? 
 is the tree with the root node labelled by \lstinline?let?, having three subtrees given by the trees corresponding to \lstinline?x?, \lstinline?T? and \lstinline?U?.

$-$ If \lstinline?C? is a term of the form  \lstinline?T -> U?, then  the term tree of \lstinline?C?  is represented by the tree with the root node labelled by 
\lstinline?->?, and its immediate subtrees given by the trees representing  \lstinline?T? and \lstinline?U?.

$-$ If \lstinline?C? is a term of the form  \lstinline?(T U1 ... Un)?, then we have two cases.
If \lstinline?T? is a name, the term tree of \lstinline?C?  is represented by 
the tree with the root node labelled by \lstinline?T? together with its type, and its immediate subtrees given by the trees
representing \lstinline?U1?,\ldots, \lstinline?Un?. If \lstinline?T? is not a  name, the term tree of \lstinline?C?
is the tree with the root node labelled by \lstinline?@?, and its immediate subtrees given by the trees 
representing \lstinline?T?, \lstinline?U1,...,Un?.

\end{definition}

Note that ML4PG term trees consist of two kinds of nodes: \emph{Gallina} and \emph{term-type} nodes. The Gallina nodes are 
labelled by Gallina-keywords and special tokens such as \lstinline?forall?, \lstinline?fun?, \lstinline?let? or \lstinline?->? (from now on, we will call them Gallina tokens); 
%to
%Gallina keywords and special tokens); 
and the term-type nodes are labelled by expressions of the form \lstinline?t1:t2? where \lstinline?t1? is a sort, a variable or 
a name, and  \lstinline?t2? is the type of \lstinline?t1?.

\begin{figure}[t]
\centering
\begin{tikzpicture}[level 1/.style={sibling distance=40mm},
 level 2/.style={sibling distance=70mm},
 level 3/.style={sibling distance=30mm},
 level 4/.style={sibling distance=30mm},scale=.8,font=\footnotesize]

   \node (root) {\lstinline?forall?} [level distance=10mm]
             child { node {\lstinline?n : nat?}}
             child { node {\lstinline?H : even n?}}
             child { node {\lstinline?odd : nat -> Prop?} 
                      child { node {\lstinline?+ : nat -> nat -> nat?}
                           child { node {\lstinline?n : nat?}}
                           child { node {\lstinline?1 : nat?}}
                           }
                   }

    ;
 \end{tikzpicture}
\caption{\scriptsize{\emph{ML4PG term tree for the term \texttt{forall (n : nat) (H : even n), odd (n + 1).}}}}\label{fig:termtree}
\end{figure}

\begin{example}\label{ex1}
 Given the term \lstinline?forall (n : nat) (H : even n), odd (n + 1)?, its ML4PG term tree is depicted in Figure~\ref{fig:termtree}.
 \end{example}

We represent ML4PG term trees as feature matrices, further to be flattened into feature vectors for clustering. %they in turn will be the input of machine-learning algorithms.
A variety of methods exists to represent trees as matrices, for instance using
adjacency or incidence matrices. The adjacency matrix and the various previous methods of  feature
extraction (e.g.~\cite{lpar-urban,K13,UrbanSPV08}) share the following common properties: different %tree
%nodes 
library symbols are represented by distinct features, and the %matrix entries
feature values are binary. %; and the size of the matrix depends on the tree size. 
For
big libraries and growing syntax, feature vectors grow very large (up to $10^6$ in some experiments)
and at the same time very sparse, which implies the use of sparse machine-learning methods.

%In~\cite{lpar13}, we represent ACL2 term trees as matrices with the following properties: the features are 
%given by the arity of functions and the term tree depth -- two properties common to every term tree -- and the 
%feature values are rational numbers dynamically computed depending on the definition of the functions involved in the term tree. The size of 
%those matrices is fixed to $49$; hence, allowing the application of data-mining techniques to both big and small 
%data sets of ACL2 theorems -- large and sparse feature vectors require to use big data sets, 
%where the size of the example data set is comparable to the number of features; on the contrary, compact feature extraction methods work equally well 
%with both big and small datasets. The method presented in~\cite{lpar13}
%can be applied not only to ACL2 but also to any first-order theory; however, it is not sufficient to capture all the information 
%from the higher-order dependently-typed language of Coq.

We develop a new compact method that tracks a large (potentially unlimited) number of Coq terms by a finite number of features and an unlimited number of feature-values.
In our method, the features are given by two properties common to all possible term trees: the term tree depth and the level index of nodes.
The most important information about the term is then encoded by improving precision of feature values %means of a more precise specification of feature values, 
using rational-valued feature-extraction functions. Taking just $300$ features, the new feature-extraction method recursively adjusts the feature values, 
adapting to the growing language syntax. %; thus producing a machine-learning analogue of how an unlimited number of Coq terms can be generated from the limited syntactic constructs.  
The resulting feature vectors have an average density ratio of 60\%.

Given a Coq expression, we can differentiate its term and type components; the feature values capture information from
these components and also the structure of the tree. In particular, each tree node is encoded by distinct feature values
given by a triple of rational numbers to represent the term component, the type component, and the level index of the parent node in the term tree,
cf. Table~\ref{ml4pgtermtable}. Our feature extraction method is formalised in the following definitions.

\begin{table}[t]
\tbl{\scriptsize{\emph{ML4PG term tree matrix for \texttt{forall (n : nat) (H : even n), odd (n+1)}.}}\label{ml4pgtermtable}}{
\centering
{\scriptsize 
\begin{tabular}{|c||c|c|c|}
\hline
 &  level index 0 & level index 1 & level index 2 \\
 \hline
td0 & ($[\texttt{forall}]_{Gallina}$,-1,-1) & (0,0,0) & (0,0,0)\\
\hline
td1 & ($[\texttt{n}]_{term}$,$[\texttt{nat}]_{type}$,0) & ($[\texttt{H}]_{term}$,$[\texttt{even n}]_{type}$,0)& ($[\texttt{odd}]_{term}$,$[\texttt{nat-> Prop}]_{type}$,0) \\
\hline
td2 & ($[\texttt{+}]_{term}$,$[\texttt{nat -> nat -> nat}]_{type}$,2) & (0,0,0)& (0,0,0)\\
\hline
td3 & ($[\texttt{n}]_{term}$,$[\texttt{nat}]_{type}$,0) & ($[\texttt{1}]_{term}$,$[\texttt{nat}]_{type}$,0)  & (0,0,0)\\
\hline
\end{tabular}}}
\end{table}

\begin{definition}[Term tree depth level and level index]\label{def:termtreelevel}
Given  a term tree $T$, the \emph{depth} of the node $t$ in $T$, denoted by \emph{depth(t)}, is defined as follows:

$-$ $depth(t) = 0$, if $t$ is a root node;

$-$ $depth(t) = n+1$, where $n$ is the depth of the parent node of $t$.

The \emph{$n$th level} of $T$ is the ordered sequence of nodes of depth $n$ --- using the classical representation for trees, the order of the sequence is 
given by visiting the nodes of depth $n$ from left to right. The \emph{level index} of a node with depth $n$ is the position of the node in the $n$th level of $T$.
We denote by $T(i,j)$ the node of $T$ with depth $i$ and index level $j$.
\end{definition}

	%%% Introduce common notation for all three; may be call it more like feature exatraction... template f.e.? pre-fe.e?

% \begin{definition}[ML4PG term tree matrix]\label{df:matrix}
% Given a term tree $T$ for a Coq term, and three injective functions $[.]_{Gallina}: Gallina~tokens \rightarrow \mathbb{Q}^-$, $[.]_{term}: Coq~Terms \rightarrow \mathbb{Q}^+$, $[.]_{type}: Coq~Terms \rightarrow \mathbb{Q}^+$, 
% the ML4PG term tree matrix $M_M$ of $T$ is a $10 \times 10$ matrix such that the $(i,j)$-th entry of $M_M$ captures information from the node of depth $i$ and
% level index $j$ of $T$ as follows:
% 
% $-$ if the node of depth $i$ and level index $j$ of $T$ is a Gallina node $g$, then the $(i,j)$-th entry of $M_M$ is a triple $([g]_{Gallina},-1,p)$ 
% where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$. 
% 
% $-$ if the node of depth $i$ and level index $j$ of $T$ is a term-type node \lstinline?t1:t2?, then the $(i,j)$-th entry of $M_M$ is a triple $([t1]_{term},[t2]_{type},p)$
% where $p$ is the level index of the direct parent of the node.
% 
% \end{definition}

We use the notation $M[\mathbb{Q}]_{n\times m}$ to denote the set of matrices of size  $n\times m$ with rational coefficients.

\begin{definition}[ML4PG term tree feature extraction]\label{df:matrix}
Given a term \lstinline?t?, its corresponding term tree $T_{\texttt{t}}$, and three injective functions 
$[.]_{term}: Coq~terms \rightarrow \mathbb{Q}^+$, $[.]_{type}: Coq~terms \rightarrow \mathbb{Q}^+$
and $[.]_{Gallina}: Gallina~tokens \rightarrow \mathbb{Q}^-$, then the feature extraction function 
$[.]_M=\langle[.]_{term}, [.]_{type},[.]_{Gallina}\rangle : Coq~terms \rightarrow M[\mathbb{Q}]_{10\times 10}$
builds the  \emph{term tree matrix of \texttt{t}}, $[\texttt{t}]_M$,
where the $(i,j)$-th entry of $[\texttt{t}]_M$ captures information from the node $T_{\texttt{t}}(i,j)$ as follows:

$-$ if $T_{\texttt{t}}(i,j)$ is a Gallina node $g$, then the $(i,j)$th entry of $[\texttt{t}]_M$ is a triple $([g]_{Gallina},-1,p)$ 
where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$.

$-$ if $T_{\texttt{t}}(i,j)$ is a term-type node \lstinline?t1:t2?, then the $(i,j)$th entry of $[\texttt{t}]_M$ is a triple $([t1]_{term},[t2]_{type},p)$
where $p$ is the level index of the parent of the node.

\end{definition}

In the above definition, we fix  the maximum depth and  maximum level index of a node to $10$; this makes the feature extraction mechanism uniform
across all Coq terms appearing in the libraries. We may lose some information if  pruning is needed, but the chosen size works well for most terms appearing in Coq libraries. 
%The above method works for any Coq term, irrespective of the size and complexity. 
If a term tree does not fit into $10 \times 10$ term tree dimensions, its $10 \times 10$
subtree is still considered by ML4PG.
%For the machine-learning algorithms, 
The term tree matrix is flattened into a feature vector, and each triple will be split 
into three components of the vector, giving a feature vector size of $300$, %-- bigger than the size of vectors in~\cite{lpar13}, 
still smaller than in sparse approaches~\cite{lpar-urban,K13,UrbanSPV08}.

In Definition~\ref{df:matrix}, we deliberately specify the functions $[.]_{Gallina}, [.]_{term}$ and $[.]_{type}$ just by their signature. 
The function $[.]_{Gallina}$ is a predefined function.
The number of Gallina tokens (\lstinline?forall?, \lstinline?fun?, \lstinline?->? and so on) is fixed and cannot be 
expanded by the Coq user. Therefore, we know in advance all the Gallina tokens that can appear in a development, and we can
assign a concrete value to each of them. The function $[.]_{Gallina}: Gallina~tokens  \rightarrow \mathbb{Q}^-$ is an injective 
function carefully defined to assign close values to similar Gallina tokens %(for instance, 
%$[$\lstinline?forall?$]_{Gallina}=-1$ and $[$\lstinline?->?$]_{Gallina}=-1.1$);
and more distant numbers to unrelated tokens 
%(e.g. $[$\lstinline?let?$]_{Gallina}=-3$)
--- see Appendix~\ref{sec:gallinasyntax} for the exact encoding.
 %; thus distinguishing the 
%different kinds of Gallina tokens unambiguously.

%On the contrary to the function $[.]_{Gallina}$, 
The functions $[.]_{term}$ and $[.]_{type}$ are dynamically re-defined for every 
library and every given proof stage, to adapt to the changing syntax.
 %, as we are going to describe in 
%Definition~\ref{def:funterm}. 
In practice, there will be  %set of functions computed in 
%every session: 
new $[.]_{term}$ and $[.]_{type}$ functions computed whenever ML4PG is called.
This
brings the element of the ``acquired knowledge/experience'' to the machine-learning cycle, as will be formalised in the next section.

% In the previous section, we have presented a method to extract features from 
% Coq terms; however, the functions $[.]_{Gallina}$, $[.]_{term}$ and $[.]_{type}$
% are still undefined. 
% % The definition of these functions influences the results that 
% % are obtained by machine-learning algorithms.
% In this section, we explain how these functions are defined and how this has led to several improvements in ML4PG.

%We finish this section with the following property about ML4PG term tree matrices.

% \begin{lemma}
% Let ${\cal T}$ be the set of Coq terms whose nodes have maximum depth $10$ and level index $10$. 
% Then, the function $[.]_{T}$ restricted to ${\cal T}$ is a one-to-one function. 
% %  Given a term tree matrix $M$ associated with a Coq term $C$. If the maximum depth of the term tree of $C$ is 10 and the maximum index level of the 
% %  nodes of the tree is $10$, then the term $C$ can be reconstructed from $M$. 
% \end{lemma}

### 2015-02-04 ACMTransCL/patches.tex
Initial commit

\section{Recurrent Proof Clustering}\label{sec:recurrent}

The method presented in the previous section can cluster similar statements of all Coq terms, including definitions and theorems. 
However, this method does not capture the interactive nature of Coq proofs.
In this section, we involve proofs into the recurrent clustering of Coq libraries.

 In~\cite{KHG13}, 
we introduced a feature extraction method for Coq proofs capturing
the user's interaction through the applied tactics. That method 
%tracks simple, 
traced 
low-level properties present in proof's subgoals, e.g. ``the
top symbol'' or ``the argument type''. Further, these features were taken
in relation to the statistics of user actions on every subgoal: how many and what
kind of tactics he applied, and what kind of arguments he provided to the
tactics. Finally, a few proof-steps were taken in relation to each other.
This method  had two drawbacks.\\ 
% (1) It only captures information 
% about the top symbol of the goal; this means that information is mainly collected 
% from symbols such as \lstinline?forall? or \lstinline?equal? (common to most proofs)
% and features from other symbols (that are likely more relevant for the proof) are never captured.
(1) It was  focused on the first
five proof-steps of a proof; therefore, some information was lost. We address this issue by implementing automatic split of each proof into proof-patches; thus, allowing ML4PG to analyse a proof by the properties
of the patches that constitute the proof. \\
(2) The method assigned most feature-values blindly; thus, being insensitive to many important parameters, such as e.g. the structure of lemmas and hypotheses used as 
tactic arguments within a proof. 
The previous section gave us the way of involving all Coq objects into recurrent feature re-evaluation.
%, and, as we have explained in the previous section, machine-learning algorithms are sensitive to 
%the assignment of feature values. These drawbacks have been overcome combining a new method called the 
%\emph{proof-patch method} and the term feature extraction method presented in the previous section.

%We start with introducing \emph{proof-patches} to address the first drawback. The underlying idea is that

%We start defining the notions of Coq proof and proof-patch matrices.

\begin{definition}[Coq proof]
 Given a statement $S$ in Coq, a \emph{Coq proof} of $S$ is given by a sequence of triples $((\Gamma_i,G_i,T_i))_{0\leq i\leq n}$ where $\Gamma_i$ is a context,
 $G_i$ is a goal and $T_i$ is a sequence of tactics satisfying:

  - $G_0=S$, and for all $i$, $\Gamma_i$ is the context of the goal $G_i$,

  - for all $i$ with $0<i\leq n$, $\Gamma_i,G_i$ are respectively the context and goal obtained after applying $T_{i-1}$, and 
    the application of $T_n$ completes the proof.
\end{definition}

In this paper, we focus on the goals and tactics of Coq proofs; thus, we do not consider the contexts and denote the 
Coq proof $((\Gamma_i,G_i,T_i))_{0\leq i\leq n}$ by $((G_i,T_i))_{0\leq i\leq n}$. Involving contexts into proof-pattern search may be a subject for future work.

\begin{table}[t]
\tbl{\scriptsize{\emph{Proof for the lemma of Example~\ref{example0} in SSReflect.}}\label{tab:sumfirstn}}{
 	\centering
 	\tiny{
 		\begin{tabular}{|l|l|}
 		\hline
 	Goals and Subgoals & Applied Tactics \\
 		\hline
 		\hline
 	{\scriptsize $G_0) \forall~n,\sum\limits_{i=0}^{n} (g(i+1) - g(i)) = g(n+1) - g(0)$} & \\
 			& $T_0)$ {\scriptsize \lstinline?case : n => [|n _].?} \\
        $G_1) \sum\limits_{i=0}^{0} (g(i+1) - g(i)) = g(1) - g(0)$ & \\
        & $T_1)$  {\scriptsize\lstinline?by rewrite big_nat1.?}\\
        $G_2)  \sum\limits_{i=0}^{n+1} (g(i+1) - g(i)) = g(n+2) - g(0)$ &\\

        & $T_2)$ {\scriptsize\lstinline?rewrite sumrB big_nat_recr big_nat_recl ?}\\
          & ~~~~~~~~{\scriptsize\lstinline?     addrC addrC -subr_sub -!addrA addrA.?} \\
        $G_3) g(n+2) + \sum\limits_{i=0}^{n} g(i+1) -  \sum\limits_{i=0}^{n} g(i+1) - g(0) =$ &\\
        $  g(n+2) - g(0)$ &\\
        &$T_3)$  {\scriptsize\lstinline?move : eq_refl.?}\\
        $G_4) \sum\limits_{i=0}^{n} g(i+1) == \sum\limits_{i=0}^{n} g(i+1) \rightarrow  $ &\\
        $g(n+2) + \sum\limits_{i=0}^{n} g(i+1) -  \sum\limits_{i=0}^{n} g(i+1) - g(0) =$ &\\
        $  g(n+2) - g(0)$ &\\
        &$T_4)$  {\scriptsize\lstinline?rewrite -subr_eq0.?}\\
        $G_5) \sum\limits_{i=0}^{n} g(i+1) - \sum\limits_{i=0}^{n} g(i+1) == 0\rightarrow  $ &\\
        $g(n+2) + \sum\limits_{i=0}^{n} g(i+1) -  \sum\limits_{i=0}^{n} g(i+1) - g(0) =$ &\\
        $  g(n+2) - g(0)$ &\\
        &$T_5)$  {\scriptsize\lstinline?move/eqP => ->.?}\\
         $G_6)  g(n+2) + 0 - g(0) = g(n+2) - g(0)$ &\\
        &$T_6)$  {\scriptsize\lstinline?by rewrite sub0r.?}\\
 		$\Box$ & \\
 		& {\scriptsize\lstinline?Qed.?}\\
 		\hline
 		\end{tabular}

 		}}
 \end{table}

\begin{example}\label{example0}
Table~\ref{tab:sumfirstn} shows the Coq proof of the following statement:
 $$\forall g:\mathbb{N} \rightarrow \mathbb{Z}\implies \sum_{0\leq i \leq n} (g(i+1) - g(i)) = g(n+1) - g(0)$$
\end{example}

One small proof may potentially 
resemble a fragment of a bigger proof; also, various small ``patches'' of different big proofs may resemble.

\begin{definition}[Proof-patch]
 Given a \emph{Coq proof} $C=((G_i,T_i))_{0\leq i\leq n}$, a \emph{proof-patch} of $C$ is a subsequence of at most $5$ consecutive 
 pairs of $C$.
\end{definition}

From proof-patches, we can construct their feature matrices. %that will be provided to the machine-learning algorithms. 
We will shortly define the feature extraction function $[.]_P=\langle[.]_M,[.]_{tac}\rangle : proof~patches \rightarrow M[\mathbb{Q}]_{5\times 6}$,
where $[.]_{tac}$ is an injective function that has been introduced to assign values to tactics.
%To be more concrete, 
We have defined two versions of  $[.]_{tac}$: one for Coq tactics and another for SSReflect tactics. 
% -- although SSReflect is an extension of Coq, this package implements a set of proof tactics designed to support the extensive
% use of small-scale reflection in formal proofs~\cite{SSReflect}. 
In the SSReflect case, we divide the tactics into 7 groups and assign similar
values to each tactic in the group, see Table~\ref{tab:tactics}. 
% If several tactics are concatenated using ``\lstinline?;?'', the value returned by
% $[.]_{tac}$ is the concatenation of the values of the individual tactics. 
Analogously for Coq tactics, cf. Appendix~\ref{sec:coqtactics}.

%These clusters in turn are used to define a new function $[.]_{term}$. The function $[.]_{term}$ assigns numbers using the 
%formula $[l]_{term}=2\times j + p$ where $C_j$ is a cluster such that $l\in C_j$ and $p$ is the proximity value of $l$ in $C_j$. This process assigns 
%closer values to statements in the same cluster, and more distant numbers across groups. This function will be used in the following section.

\begin{definition}[Proof-patch matrix]\label{def:ptm}
Given a Coq proof $C=((G_i,T_i))_{0\leq i\leq n}$, and a proof patch $p=((G_{i_0},T_{i_0}),\ldots,(G_{i_4},T_{i_4}))$ of $C$, 
the feature extraction function $[.]_P: proof~patches \rightarrow M[\mathbb{Q}]_{5\times 6}$ constructs the \emph{proof-patch matrix} $[p]_P$ 
as follows:

\begin{itemize}
 \item the $(j,0)$-th entry of $[p]_P$ is a 4-tuple $([T_{i_j}^1]_{tac}, [T_{i_j}^2]_{tac},[T_{i_j}^3]_{tac},[T_{i_j}^r]_{tac})$ 
 where $T_{i_j}^1, T_{i_j}^2$ and $T_{i_j}^3$ are the three first tactics of $T_{i_j}$, and $T_{i_j}^r$ is the list of the rest of tactics of $T_{i_j}$,
 \item the $(j,1)$-th entry of $[p]_P$ is the number of tactics appearing in $T_{i_j}$,
 \item the $(j,2)$-th entry of $[p]_P$ is a 4-tuple $([t_1]_{type},[t_2]_{type},[t_3]_{type},[t_{i_j}]_{type})$ where 
 $t_1, t_2$ and $t_3$ are the three first argument-types of $T_{i_j}$, and $t_{i_j}$ is the set of the rest of 
 the argument-types of $T_{i_j}$ (insensitive to order or repetition),
 \item the $(j,3)$-th entry of $[p]_P$ is a 4-tuple  $([l_{i_{j_1}}]_{term},[l_{i_{j_2}}]_{term},[l_{i_{j_3}}]_{term},[l_{i_j}]_{term})$
 where $l_{i_{j_1}}$, $l_{i_{j_2}}$ and $l_{i_{j_3}}$ are the three first lemmas applied in $T_{i_j}$ and $l_{i_j}$ is the list of the rest of lemmas 
used in $T_{i_j}$ (sensitive to order and repetition),
 \item the $(j,4)$-th entry of $[p]_P$ is a triple  $([s_1]_{term},[s_2]_{term},[s_3]_{term})$ where $s_1,s_2$ and $s_3$ are respectively the top, second, and third
 symbol of $G_{i_j}$,
 \item the $(j,5)$-th entry of $[p]_P$ is the number of subgoals after applying $T_{i_j}$ to $G_{i_j}$.
\end{itemize} 
\end{definition}

\begin{table}[t]
\tbl{\scriptsize{\emph{Formulas computing the value of SSReflect tactics.}
they serve to assign closer values to the tactics within each of the seven groups, and more distant numbers 
across the groups.  If a new tactic is defined, ML4PG automatically assigns a new number to it, using the next available natural number $n$ 
in the formula $n+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.}\label{tab:tactics}}{
\centering
{\scriptsize
\begin{tabular}{ll}
\hline
$\ast$ Bookkeeping ($b=\{$\lstinline?move:, move => ?$\}$): &$[b_i]_{tac}=1+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$ (where $b_i$ is the $i$th element of $b$).\\
$\ast$ Case and Induction ($c=\{$\lstinline?case, elim?$\}$):& $[c_i]_{tac}=2+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.\\
$\ast$ Discharge ($d=\{$\lstinline?apply, exact, congr?$\}$):& $[d_i]_{tac}=3+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.\\
$\ast$ Simplification ($s=\{$\lstinline?//, /=, //=?$\}$):& $[s_i]_{tac}=4+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.\\
$\ast$ Rewrite:& $[$\lstinline?rewrite?$]_{tac} = 5$. \\
$\ast$ Forward Chaining  ($f=\{$\lstinline?have, suff, wlog?$\}$):& $[f_i]_{tac}=6+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.\\
$\ast$ Views and reflection  ($v=\{$\lstinline?move/, apply/, elim/, case/?$\}$):& $[v_i]_{tac}=7+\sum_{j=1}^i \frac{1}{10\times 2^{j-1}}$.\\
\hline
\end{tabular}}}
\end{table}

\begin{example}\label{example1}
Given the proof of Example~\ref{example0} and the proof-patch $((G_{i},T_{i}))_{0\leq i \leq 4}$, 
the top table of Table~\ref{tab:patches} shows their proof-patch matrix.
\end{example}

The \emph{proof-patch method} considers several proof-patches to collect information from a concrete proof. 
In particular, given a Coq proof $C=((G_i,T_i))_{0\leq i\leq n}$ , the proof $C$ can be split into patches $C_0,\ldots,C_m$ where 
$m=\lceil\frac{n}{5}\rceil+1$. The patches are defined as follows: $C_j=((G_j,T_j),\ldots,(G_{j+4},T_{j+4}))$ 
for $0\leq j < m$ (some patches might contain less than $5$ proof-steps), and $C_m=((G_{n-4},T_{n-4}),\ldots,(G_{n},T_{n}))$ --- the last patch captures 
the last five proof-steps.

\begin{example}
Using the proof-patch method, we  can split the proof presented in Example~\ref{example1} into three proof-patches $((G_i,T_i))_{0\leq i \leq 4}$, 
$((G_5,T_5),(G_6,T_6))$ and $((G_i,T_i))_{2\leq i \leq 6}$; the corresponding proof-patch matrices are given in Table~\ref{tab:patches}. 
\end{example}

The proof-patch method together with the feature function $[.]_P$ solve the two drawbacks of the old method~\cite{KHG13}: 
the new method captures information about the whole proof and the feature values are dynamically computed to assign close
values to similar terms, types, tactics and lemma statements used as tactic arguments.

We finish this section with a case study that illustrates 
the use of the proof-patch method, and shows the differences with the results obtained with the old method~\cite{KHG13}. This case study concerns the discovery of proof patterns in mathematical proofs
across formalisations of apparently disjoint mathematical theories: Linear Algebra, Combinatorics and Persistent Homology  (across 758 lemmas and 5~libraries).
In this scenario, we use statistically discovered proof patterns to advance the proof of a given ``problematic'' lemma. 
Namely,  a few initial steps in its proof are clustered against several mathematical libraries. 
We deliberately take lemmas belonging to very different SSReflect libraries. The lemma introduced in Example~\ref{example0} is a basic fact about
summations. Lemma~\ref{lem:nilpotent} states a result about \emph{nilpotent} matrices %~\cite{BR91} 
(a square matrix $M$ is \emph{nilpotent}
if there exists an $n$ such that $M^n=0$). Finally, Lemma~\ref{lem:fundamental} is a generalisation of the \emph{fundamental lemma of 
Persistent Homology}~\cite{HCMS12}.

\begin{lemma}\label{lem:nilpotent}
 Let $M$ be a square matrix and $n$ be a natural number such that $M^n=0$, then $(1-M)\times \sum\limits_{i=0}^{n-1} M^i = 1$.
\end{lemma}

\begin{lemma}\label{lem:fundamental}
Let $\beta_n^{k,l}:\mathbb{N} \times \mathbb{N} \times \mathbb{N} \rightarrow \mathbb{Z}$, then
$$\sum_{0\leq i \leq k} \sum_{l<j\leq m} (\beta_n^{i,j-1} - \beta_n^{i,j}) - (\beta_n^{i-1,j-1} - \beta_n^{i-1,j}) = \beta_n^{k,l} - \beta_n^{k,m}.$$
\end{lemma}

\begin{table}
\tbl{\scriptsize{\emph{Proof-patch matrices for the proof of Example~\ref{example1}.} \textbf{Top.} Proof-patch matrix of the patch $((G_i,T_i))_{0\leq i\leq 4}$. 
 \textbf{Centre.} Proof-patch matrix of the patch $((G_i,T_i))_{5\leq i\leq 6}$ (rows that are not included in the table are filled with zeroes).
 \textbf{Bottom.} Proof-patch matrix of the patch $((G_i,T_i))_{2\leq i\leq 6}$.  
Where we use notation $EL$, ML4PG gathers the lemma names: (\lstinline?addrC?, \lstinline?addrC?~, \lstinline?subr_sub?, \ldots).}\label{tab:patches}}{
\centering
\tiny{
\begin{tabular}{|l||l|l|l|l|l|l|}
\hline
 & \emph{tactics} & \emph{n} & \emph{arg type} & \emph{arg} & \emph{symbols} & \emph{goals} \\
\hline
\hline
\emph{g1}& $([case]_{tac},0,0,0)$ & $1$  & $([nat]_{type},0,0,0)$  & $([Hyp]_{term},0,0,0)$ & $([\forall]_{term},[=]_{term},[sum]_{term})$ & $2$ \\
 \hline
  \emph{g2} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$  & $([big\_nat1]_{term},0,0,0)$ & $([=]_{term},[\sum]_{term},[-]_{term})$& $0$ \\
 \hline
  \emph{g3} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},$ & $([surB]_{term},$  & $([=]_{term},[+]_{term},[-]_{term})$& $1$ \\
 & & &$[Prop]_{type},$ & $[big\_nat\_recr]_{term}$ & & \\
 & & &$[Prop]_{type},$ & $[big\_nat\_recl]_{term}$ & & \\
 & & &$[Prop]_{type})$ & $[EL]_{term})$ & & \\
 \hline
   \emph{g4} & $([move:]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$ &  $([eq\_refl]_{term},0,0,0)$  & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline 
  \emph{g5} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$  & $([subr\_eq0]_{term},0,0,0)$ & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline

 \multicolumn{7}{c}{}\\
% Second table 
\hline
 & \emph{tactics} & \emph{n} & \emph{arg type} & \emph{arg} & \emph{symbols} &  \emph{goals} \\
\hline
\hline
\emph{g1} & $([move/]_{tac},[\texttt{->}]_{tac},0,0)$ & $2$  & $([Prop]_{type},0,0,0)$ &  $([eq\_refl]_{term},0,0,0)$  & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline
  \emph{g2} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$  & $([subr\_eq0]_{term},0,0,0)$ & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline
%   \emph{g3} & - & -  & - & -  & - & -  \\
%  \hline
%   \emph{g4} & - & -  & - & -  & - & -  \\
%  \hline 
%   \emph{g5} & - & -  & - & -  & - & - \\
%  \hline

 \multicolumn{7}{c}{}\\
%Third table
 \hline
 & \emph{tactics} & \emph{n} & \emph{arg type} & \emph{arg} & \emph{symbols} & \emph{goals} \\
\hline
\hline
\emph{g1}&  $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},$ & $([surB]_{term},$  & $([=]_{term},[+]_{term},[-]_{term})$& $1$ \\
 & & & $[Prop]_{type},$ & $[big\_nat\_recr]_{term}$ & & \\
 & & & $[Prop]_{type},$& $[big\_nat\_recl]_{term}$ & & \\
 & & & $[Prop]_{type})$& $[EL]_{term})$ & & \\
 \hline
  \emph{g2} & $([move:]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$ &  $([eq\_refl]_{term},0,0,0)$  & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline
  \emph{g3} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$  & $([subr\_eq0]_{term},0,0,0)$ & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline
\emph{g4}& $([move/]_{tac},[\texttt{->}]_{tac},0,0)$ & $2$  & $([Prop]_{type},0,0,0)$  & $([eqP]_{term},0,0,0)$ & $([=]_{term},[+]_{term},[-]_{term})$ & $1$ \\
 \hline
  \emph{g5} & $([rewrite]_{tac},0,0,0)$ & $1$  & $([Prop]_{type},0,0,0)$  & $([sub0r]_{term},0,0,0)$ & $([=]_{term},[+]_{term},[-]_{term})$ & $0$ \\
 \hline
  \end{tabular} 
  }}
 \end{table}

When proving Lemma~\ref{lem:nilpotent}, the user may call ML4PG after completing a few standard proof steps: apply induction and solve the base case using rewriting.
At this point it is difficult, even for an expert user, % 60 B?
to get the intuition that he could reuse the proofs from Example~\ref{example0} and Lemma~\ref{lem:fundamental}. There are several reasons for this.
First of all, the formal proofs of these lemmas are in different libraries; then, it is difficult to establish a conceptual connection among them. Moreover,
although the three lemmas involve summations,
the types of the terms of those summations are different. Therefore, search based on types or keywords would not help. Even search  % *a* search removed -- because 59B??? 
of all the lemmas involving summations does not provide a clear suggestion, since there are more than $250$ lemmas (a considerable number for handling them manually) --- a clever search-pattern will considerably reduced the number of suggested lemmas, but such a pattern is not trivial at all.

% \begin{table}
%  	\centering
%  	\tiny{
%  		\begin{tabular}{|l|l|}
%  		\hline
%  	Goals and Subgoals & Proof-Steps (Tactics) \\
%  		\hline
%  		\hline
%  	{\scriptsize $\forall~(M:M_n) (m:nat), M^m=0 \implies (1-M)\times \sum\limits_{i=0}^{m-1} M^i = 1$} & \\
%  			& {\scriptsize \lstinline?move => M m nilpotent.?} \\
%          {\scriptsize $(1-M)\times \sum\limits_{i=0}^{m-1} M^i = 1$} & \\
%         &   {\scriptsize\lstinline?rewrite big_distrr mulmxBr mul1mx.?}\\
%         {\scriptsize$   \sum\limits_{i=0}^{m-1} M^i - M^{i+1}$} &\\
%         
%         & {\scriptsize\lstinline?case : n.?}\\
%         {\scriptsize $ \forall~(M:M_0) (m:nat),M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
%         &  {\scriptsize\lstinline?by rewrite !thinmx0.?}\\
%          {\scriptsize$ \forall~(M:M_{n+1}) (m:nat), M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $}&\\
%  		\hline
%  		\end{tabular}
%  		
%  		}
%  	\caption{\textbf{First steps of the proof of Lemma~\ref{lem:nilpotent},} \emph{where the user stops and invokes ML4PG for a hint.}}
%  	\label{tab:sumfirstn}
%  \end{table}

However, if only the lemmas from Example~\ref{example0} and Lemma~\ref{lem:fundamental} are suggested when proving Lemma~\ref{lem:nilpotent}, the user would be able to
spot the following common proof pattern.

%\begin{IP}\label{ip:math}
%This proof pattern is followed by Lemmas~\ref{lem:nilpotent},~\ref{lem:fundamental} and~\ref{lem:lemma3}.

\begin{PS}\label{ps:math}

\emph{
%\begin{enumerate}
  %\item 
	Apply case on $n$.
  \begin{enumerate}
   \item Prove the base case (a simple task).
   \item Prove the case $0<n$:
     \begin{enumerate}
     \item expand the summation,
     \item cancel the terms pairwise,
     \item the terms remaining after the cancellation are the first and the last one. 
     \end{enumerate}
  \end{enumerate}
% \end{enumerate}
}

\end{PS}

% \begin{table}
%  	\centering
%  	\footnotesize{
%  		\begin{tabular}{|l|l|}
%  		\hline
%  	Goals and Subgoals & Proof-Steps (Tactics) \\
%  		\hline
%  		\hline
%  	{\scriptsize $\forall~(M:M_n) (m:nat), M^m=0 \implies \exists N, N \times (1-M)= 1$} & \\
%  			& {\scriptsize \lstinline?move => M m nilpotent.?} \\
%          {\scriptsize $\exists N, N \times (1-M)= 1$} & \\
%         &   {\scriptsize\lstinline?exists?}\\
%         &   {\scriptsize~~\lstinline?\sum_(0<=i<m.+1) (pot_matrix M i).?}\\
%         {\scriptsize$   (\sum\limits_{i=0}^{m-1} M^i) \times (1 - M) $} &\\
%         
%         & {\scriptsize\lstinline?rewrite big_distrl mulmxrB mulmx1.?}\\
%         {\scriptsize $ \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
%                 & {\scriptsize\lstinline?case : n.?}\\
%         {\scriptsize $ \forall~(M:M_0) (m:nat),M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
%         &  {\scriptsize\lstinline?by rewrite !thinmx0.?}\\
%          {\scriptsize$ \forall~(M:M_{n+1}) (m:nat), M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $}&\\
%  		\hline
%  		\end{tabular}
%  		
%  		}
%  	\caption{\textbf{First steps of the proof of Lemma~\ref{lem:nilpotent2}.} \emph{Note the correlation between this proof and the proof of Lemma~\ref{lem:nilpotent} in Table~\ref{tab:sumfirstn}.}}
%  	\label{tab:nilpotent2}
%  \end{table}

Using the method presented in~\cite{KHG13}, if ML4PG was invoked during the proof of Lemma~\ref{lem:nilpotent}, it would suggest the lemmas from Example~\ref{example0} and 
Lemma~\ref{lem:fundamental}. However, 5 irrelevant lemmas 
 about summations would also be suggested (irrelevant in the sense that they do not follow Proof Strategy~\ref{ps:math}). The cluster containing just the two desired lemmas could be obtained after
increasing the \emph{granularity} value~\cite{KHG13} --- a statistical ML4PG parameter that can be adjusted by the user to obtain more precise clusters. The new version of ML4PG suggests four proof fragments, all following Strategy~\ref{ps:math} without needing to adjust granularity.

%There are two enhancements when we use the proof-patch method instead of the method presented in~\cite{KHG13}: 
The new method brings two improvements:
(1) the number of suggestions is increased, but, at the same time, (2) 
the clusters are more accurate. The proof-patch method considers fragments of proofs that are deep in the proof and were not considered before; therefore, it can find lemmas (more precisely patches of lemmas) that were not
included previously in the clusters. In our case study, ML4PG suggests two additional interesting proof fragments. The first one is an intermediate patch of the proof of 
Lemma~\ref{lem:fundamental}; then, two patches are suggested from this 
lemma: the proof-patch of the inner sum, and the proof-patch of the outer sum (both of them following Proof Strategy~\ref{ps:math}). The following lemma is also 
suggested.

\begin{lemma}\label{lem:nilpotent2}
Let $M$ be a nilpotent matrix, then there exists a matrix $N$ such that $N \times (1-M)=1$. 
\end{lemma}

At first sight, the proof of this lemma is an unlikely candidate to follow  Proof Strategy~\ref{ps:math}, since the statement of the lemma
does not involve summations. However, inspecting its proof, we can see that it uses $\sum_{i=0}^{n-1} M^i$ as witness for $N$ and 
then follows Proof Strategy~\ref{ps:math}. In this case, ML4PG suggests the patch from the last five proof-steps that correspond to 
the application of Proof Strategy~\ref{ps:math}.

%The proof-patch method increases the number of suggestions; this could be a problem if more irrelevant lemmas were suggested. However, 
The new numbering of features produces more accurate clusters removing the irrelevant lemmas. In particular, using the default settings, 
ML4PG only suggests the lemma from Example~\ref{example0}, the two patches from Lemma~\ref{lem:fundamental} and the last patch from 
Lemma~\ref{lem:nilpotent2}. If the granularity is increased, the last patch from 
Lemma~\ref{lem:nilpotent2} is the only suggestion --- note that this is the closest lemma to Lemma~\ref{lem:nilpotent}.

%The discovery of proof-patterns can be a challenging task, but increasing the number and precision of suggestions simplifies this problem. 
%With the same goal (facilitate the discovery of proof-pattens), we introduce a graphical representation in the next section.

% the four lemmas given above follow the Strategy \ref{ps:math} is our next benchmarking task for ML4PG.
% Table~\ref{tab:compare0} shows the results that ML4PG will obtain, if the user varies the clustering algorithms and the size of clusters when 
% calling ML4PG within the proof environment; we also highlight the exact solution to our benchmark task, given by calling K-means algorithm with granularity $4$.

% \begin{table}
% \centering
% {\scriptsize 
%    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
%      \hline
%     & $g=1$& $g=2$& $g=3$& $g=4$ & $g=5$\\
%       Algorithm: & ($n=84$) & ($n=94$) & ($n=108$) & ($n=126$) & ($n=151$)\\     \hline
%      K-means &76$^{a-d}$ &51$^{a-d}$  &16$^{a-d}$ &\textbf{4}$^{a-d}$&2$^{b,d}$  \\
%      \hline
%      E.M. &26$^{a-d}$  &51$^{a,b,d}$ &41$^{a,b,d}$  &11$^{b,c,d}$ &3$^{a,b,d}$ \\
%      \hline
%      FarthestFirst &81$^{a-d}$ &48$^{a-d}$ &31$^{a-d}$  &25$^{a-d}$  &21$^{a-d}$  \\
%      \hline
%     \end{tabular}      }
%           
%    \caption{\textbf{Clustering experiments discovering Proof Strategy~\ref{ps:math}.} \emph{When  $g$ is chosen by the user, 
%    ML4PG dynamically calculates the number of clusters $n$. Choosing Lemma~\ref{lem:nilpotent} for pattern-search,  
%   the table shows the size of the single cluster that ML4PG displays to the user for every choice of $g$.
%    The table shows sizes of clusters containing: $a)$ Lemma~\ref{lem:nilpotent}, $b)$ Lemma~\ref{lem:lemma3},  $c)$ Lemma~\ref{lem:fundamental} and
%    $d)$ Lemma~\ref{lem:nilpotent2}. Note that with all variations of the learning algorithms and parameters, ML4PG is consistently grouping the
% correct lemmas into clusters, albeit with varied degree of precision. Note also the effect of the granularity parameter, the smallest granularity value $g=1$ produces big 
% clusters that are difficult to inspect; on the contrary, the biggest value $g=5$ produces small clusters that might not contain enough elements to spot a pattern. 
% The most accurate result for this example is obtained using the K-means algorithm and 4 as granularity parameter.}}\label{tab:compare0}
%   \end{table}
%  
% 
% Table~\ref{tab:compare0} gives an insight on how granularity parameter $g$ can be used in pattern-recognition within ML4PG.
%  One can use a ``bottom-up approach'' to detect patterns starting with the granularity value of 1 and increasing that value in successive calls.
% Using the default values $g=3$ and K-means algorithm, ML4PG obtains 15 suggestions related to lemmas about summations including Lemmas~\ref{lem:lemma3} 
% and~\ref{lem:fundamental}. Increasing the granularity level to 4, ML4PG discovers that Lemma  \ref{lem:nilpotent} is similar to 
% Lemmas~\ref{lem:lemma3}, \ref{lem:fundamental} and \ref{lem:nilpotent2}. Finally, increasing the granularity level to the maximum level of $5$, 
% ML4PG just discovers that Lemmas \ref{lem:nilpotent} and  \ref{lem:nilpotent2} are similar. %and also the following lemma;

% When the proof analogy is found (as all four proof patches taken by ML4PG follow steps $1$ and $2(a)$ of Proof Strategy \ref{ps:math}), the user can use the remaining proof steps for Lemma~\ref{lem:nilpotent2} as a template for reconstructing the remaining proof for Lemma~\ref{lem:nilpotent} (by analogy with the steps $2(b)$ and $2(c)$ in Lemma~\ref{lem:nilpotent2}). % in case of the four lemmas mentioned here, the remaining steps all follow the Strategy \ref{ps:math}. Thus, ML4PG finds patterns arising from 
% The heterogeneous cluster obtained in this scenario belongs to the category of clusters consisting of theorems that share the same proof structure and as well as an 
% auxiliary lemma (e.g. the lemma used to expand the summation).

### 2015-02-04 ACMTransCL/relatedwork.tex
Initial commit

\section{Related Work}\label{sec:relatedwork}

In this section, we compare ML4PG with tools available in Coq and other ITPs, an overview of this comparison is provided in
Table~\ref{tab:comparison}.

\begin{table}
\tbl{\scriptsize{\emph{Comparison of ML4PG with premise-selection methods, Coq's search mechanisms and dependency graphs.}}\label{tab:comparison}}{
\centering
{\scriptsize
 \begin{tabular}{ p{2cm}  p{1.75cm}  p{1.75cm}  p{1.75cm}   p{1.75cm}   p{1.75cm} }
   \toprule
    & Premise Selection & Search & Dependency graphs & Model Inference & ML4PG  \\
		%\hline
		%\hline
  	%	Fib example & No & No &   \alert{Yes} \\
		%	\hline
\midrule
\midrule
\textbf{Method} & Statistical Proof-Premise Selection & Search & Parsing & Model Inference &  Statistical Pattern-recognition \\

\midrule
   \textbf{Goal-oriented?} & Yes  & Yes &  Not necessarily & No & Not necessarily  \\
\midrule
   \textbf{Deterministic?} & No & Yes &Yes  & No & No   \\
\midrule
	\textbf{Visual representation} & No & No  & Yes  & Yes   & Yes \\
\midrule
	\textbf{Covers proofs or terms?} & Both &Terms/types & Both & Proofs  &  Both  \\
\midrule	
			\textbf{Takes into consideration dependencies?} & Yes & No & Yes & Yes  & Yes \\
\midrule
		\textbf{Finds structural similarities beyond concrete syntax?} & No & No & No  & No & Yes  \\
\midrule
	\textbf{Finds structurally similar patterns in proofs?} & No & No & No & No  & Yes  \\
\midrule
  \textbf{Good for proof automation?} & Depend on background theory & No & No & Depend on background theory & Depend on background theory \\
   \bottomrule
		\end{tabular} }}
\end{table}

\paragraph{Statistical Proof-Premise Selection in ITPs} Proof automation in ITPs has been enhanced with the connection of this kind of 
systems with Automated Theorem Provers (ATPs). The workflow of tools like Sledgehammer~\cite{Paulson_threeyears} or HolyHammer~\cite{holyhammer} consists of the following steps: (1) translation of the statement of the theorem (from the ITP format) to a
first order format suitable for ATPs; (2) selection of the lemmas (or premises) that could be relevant to prove a theorem; 
(3) invocation of several ATPs to prove the result; and (4) if an ATP is successful in the proof, reconstruction of the proof in
the ITP from the output generated by the ATP.

An important issue in the above procedure is the premise selection mechanism, especially when dealing with big libraries, since proofs of some results can be infeasible for the ATPs if they receive too many premises. Statistical machine-learning methods have been 
successfully used to tackle this problem in~\cite{UrbanSPV08,K13,holyhammer}. In particular, a classifier is constructed for every 
lemma of the library. Given two lemmas $A$ and $B$, if $B$ was used in the proof of $A$, a feature vector $|A|$  is extracted 
from the statement of $A$, and is sent as a positive example to the classifier $<B>$ constructed for $B$; otherwise, $|A|$ is
used as a negative example to train $<B>$.
Note that, $|A|$  captures statistics of $A$'s syntactic form relative to
every symbol in the library; and the resulting feature vector is a sparse (including up to $10^6$ features).
 After such training, the classifier $<B>$ can predict whether a new lemma $C$ requires 
the lemma $B$ in its proof, by testing $<B>$ with the input vector $|C|$. On the basis of such predictions for all lemmas in the library,  this tool constructs a hierarchy of the premises that are most likely to be used in the proof
of $C$.

There are several differences between the statistical premise-selection tools and ML4PG:

\begin{itemize}
 \item \emph{Aim:} premise selection tools help the prover, trying to increase the number of goals automatically proved by ATPs, and ML4PG
 assists the user, providing suggestions as proof hints.
 \item \emph{Features:} premise selection tools extract features from first-order formulas translated from the original syntax of the
 theorem prover; on the contrary, ML4PG directly works with Coq's higher-order formulas and proofs. 
 \item \emph{Dependencies:} premise selection tools capture the dependencies of lemmas used in the proof of theorems; ML4PG not only 
 captures these dependencies but also information about, for instance, tactics involved in a proof.
 \item \emph{Machine-learning methods:} Premise selection tools use supervised learning methods, and ML4PG uses unsupervised techniques.
 \item \emph{Proof Automation:} the success of tools like Sledgehammer depends on several aspects: good methods for premise selection,
 availability of a useful background theory, and success of ATPs and the proof reconstruction tool. In the case of ML4PG, the 
 availability of a background theory is also a key aspect for success proof automation, but it also depends on the existence of 
 theorems that required similar proofs. Hence, both tools are only as good as the background theory that was previously developed. \end{itemize}

% As in the case of proof-automation in ML4PG (cf. Section~\ref{sec:evaluation}), tools like Sledgehammer~\cite{Paulson_threeyears} or HolyHammer~\cite{holyhammer} are only as good as the background theory that was previously developed. Therefore, premise selection tools
% can improve the performance of those tools as long as the necessary lemmas are available.

\paragraph{ML4PG vs Coq's Searching Tools} Coq/SSReflect already provides comprehensive search mechanisms to inspect the corpus of results available in different libraries.
There are several search commands in Coq:  \lstinline?Search?~, \lstinline?SearchAbout?, \lstinline?SearchPattern?
and \lstinline?SearchRewrite?~\cite{Coq}. In addition, SSReflect implements its own version of the \lstinline?Search?
command \cite{SSReflect} --- SSReflect's \lstinline?Search? gathers the functionality of the 4 Coq's search commands.
The Whelp platform~\cite{AspertiGCTZ04} is a web search engine for mathematical knowledge formalised in Coq, which features 3 functionalities:
\lstinline?Match? (similar to Coq's \lstinline?Search? command), \lstinline?Hint? (that finds all the theorems which can
be applied to derive the current goal) and \lstinline?Elim? (that retrieves all the eliminators of a given type).
%In this section, we compare ML4PG with these tools.

%Let us explain first when these tools are usually invoked. 
The existing search mechanisms can be used in two different scenarios. If the user knows how to continue the proof, but
he does not remember (or know) the concrete name of the desired auxiliary lemma, it suffices to provide a search pattern to Coq searching engines, e.g.
using commands of the form
``\lstinline?Search "distr"? \lstinline?in bigop?'' or   
``\lstinline?Search _ (_ * (\big[_/_]_(_ <- _| _) _))?'',
 where \texttt{bigop} is a library, \texttt{"distr"} is a pattern in a lemma name, and  \lstinline? _ (_ * (\big[_/_]_(_ <- _| _) _))?
is a pattern for search.

%\begin{example}\label{ex:distr}
% In the proof of Lemma~\ref{lem:nilpotent}, we may want to introduce the term $(1-M)$ inside the sum $\sum_{i=0}^{n-1} M^i$, but 
% if we have no experience with the bigop library of SSReflect, it is likely that we do not know the name of the concrete lemma. 
% The lemma which carries out this task can be find using, for instance, the commands \lstinline?Search "distr" in bigop.? or  
%\lstinline?Search _ (_ * (\big[_/_]_(_ <- _| _) _))?. 
%\end{example}
In the second scenario, the user needs help in the middle of the proof. Searching mechanisms can be also useful in this 
case; for instance, using the \lstinline?Search? command the user can search all the lemmas associated with a concrete term
of the current goal. The \lstinline?Hint? mechanism of Whelp can find all the theorems which can be applied to derive the current goal.

%\begin{example}\label{ex:fact}
% In the proof of Lemma~\ref{lem:factorial}, we can find all the lemmas related to the factorial function using the command
% \lstinline?SearchAbout factorial?. 
%\end{example}

%ML4PG is helpful in the second scenario, since it provides families of similar proofs. Inspecting that family of proofs, 
%the user can find a pattern to finish the proof that she is formalising.

%As we have seen in Examples~\ref{ex:distr} and~\ref{ex:fact}, 
The above search mechanisms are goal directed and deterministic. That is, the user
searches  the chosen  libraries for lemmas related to a concrete type, term or pattern. 
If the patterns defined by the user are present in the given library, then the user is guaranteed to see the relevant lemmas on the screen.

The situation is more complicated if the user does not know the right pattern to search for. Imagine, for example, being in the middle of constructing a proof, 
and wishing to get some higher-level hint on how to proceed, wishing there was an ITP expert near, who would suggest a further proof strategy based on his previous experience. ML4PG was created to emulate such intelligent help automatically, using statistical machine-learning algorithms to use the information arising from the ``previous experience''.

In comparison to the more traditional \emph{search engines}, ML4PG is a goal-independent (unsupervised) tool, i.e. the user does not have to know the required pattern in advance. But, as many statistical machine-learning applications, it is also non-deterministic. That is, the tool failing to suggest a proof pattern does not mean there is no ``interesting'' pattern to be found. Actually, the notion of a proof pattern being ``interesting'' is left to the user's judgement, as well.

\paragraph{ML4PG vs Coq's Visualisation Tools} Several visualisation tools have been developed to facilitate the understanding
of the dependencies that arise in Coq developments. Namely, Coq provides tools to visualise dependencies among files, dependencies 
between theorems and definitions, proof-trees, and proof-models.

Large Coq developments involve hundreds of files (e.g. 125 files in the proof of the Feit-Thompson theorem); therefore, the 
computation and visualisation of the dependencies among those files is an important issue to maintain those developments. The 
Coq distribution provides the \lstinline?coqdep? tool to compute file dependencies. The output generated by \lstinline?coqdep? 
can be fed into the tool developed in~\cite{dependtohtml} to visualise file dependencies (see Figure~\ref{fig:depend-files}).

\begin{figure}[t]
\centering 
\includegraphics[scale=.4]{HoTTCorefragment.png}
\caption{\scriptsize{\emph{Fragment of the dependency graph of the HoTT library.} The complete dependency graph is available at the HoTT wiki\protect\cite{hottbook}.}}\label{fig:depend-files}
\end{figure}

A different set of tools~\cite{dpdgraph,BPP00} has been developed to compute and visualise dependencies between theorems and definitions, see Figure~\ref{fig:dgraph1}. An analysis of proof terms makes it possible to know whether some theorem is used when proving another result or when defining a new concept. Visualising the complete graph of dependence is useful to detect the results that are not used, and to foresee the impact of modifying a theorem or definition.

These dependency graphs are designed to, mainly, work with proofs rather than object definitions and do not show the term structure per se,
but only the dependencies between the auxiliary lemmas/constructors used. Moreover, they give a complete information of all the results that were necessary to prove a given lemma, making hard to read them, due to the presence of this complete information. Hence,  statistical machine learning may be useful for data-mining the above information in order to discriminate unimportant features of the
lemma, and highlight those that are important. ML4PG essentially does this. ML4PG's recurrent clustering  captures term and lemma dependencies recurrently, via the feature extraction which itself involves recurrent clustering of all previously
defined terms.

ML4PG's proof clustering in particular is very close to dependency graphs for lemmas, in that both are capturing mutual or inductive dependencies of all lemmas needed to prove the given statement. As a result, ML4PG can be seen as a post-processing tool for dependency
graphs. Given a big graph as shown in Figure~\ref{fig:dgraph1}, what is the right way to discriminate its unimportant features? How to decide which of them are important or not? One of the ways to do this, is to determine that relatively to other lemmas in the library. This is achieved by applying clustering in ML4PG. Taking all of the above features into account, ML4PG can associate various object definitions
and lemma statements and lemma proofs.

\begin{figure}
\centering
\includegraphics[scale=.15]{graph_dpath_path_lr.png}
\caption{\scriptsize{\emph{Dependency graph for the theorem \texttt{dpath\_path\_lr} included in the \texttt{Paths} library of HoTT.}}}\label{fig:dgraph1}
\end{figure}

The tools presented in~\cite{dpdgraph,BPP00} can also be used to visualise dependency graphs for terms, see Figure~\ref{fig:dgraph2}. 
ML4PG’s term-structure clustering goes beyond these dependency graphs for terms, and carefully captures the tree term shape, structure,
and dependency on other term- and type- structures present in the library (see Figure~\ref{fig:treevisualisation}). ML4PG's approach makes sure that dependency of auxiliary terms/types on other terms/types defined in Coq libraries is taken into account recurrently.

\begin{figure}
\centering
\includegraphics[scale=.35]{graphpaths.png}
\caption{\scriptsize{\emph{Dependency graph for the definition of \texttt{transport\_paths\_r} included in the \texttt{Paths} library of HoTT.}}}\label{fig:dgraph2}
\end{figure}

Coq also supports the visualisation of the proof-tree during an interactive proof development using the Prooftree tool~\cite{prooftree}. This tool was aimed to visualise the different subgoals that arise during the development of a proof, see Figure~\ref{fig:prooftree}. ML4PG automaton representation of proofs (cf. Figure~\ref{fig:automata}) not only shows the applied tactics, but also the generated-subgoals in each step, and compare the proof flow of the current proof with the steps followed in other theorems.

\begin{figure}
\centering
\includegraphics[scale=.5]{prooftree.png}
\caption{\scriptsize{\emph{Proof tree for the theorem \texttt{dpath\_path\_lr} included in the \texttt{Paths} library of HoTT.}}}\label{fig:prooftree}
\end{figure}

Finally, the tool presented in~\cite{GNR14} can infer models from Coq developments. Those models are represented as automata that are latter used to generate proof attempts. Since the models were designed with the aim of automatically generate proofs, they are not human-readable. 
ML4PG's graphical representation of automata is simpler than the work presented in~\cite{GNR14}.

### 2015-02-04 ACMTransCL/visualisation.tex
Initial commit

\section{Visualisation tools in ML4PG}\label{sec:visualisation}

\begin{figure}

$$
\xy0;/r.12pc/: 
*[o]=<80pt,30pt>\hbox{\txt{Coq libraries:\\one or many?}}="a"*\frm<8pt>{-,},
(40,35)*[o]=<80pt,30pt>\hbox{\txt{General\\pattern-search}}="r"*\frm<8pt>{-},
(40,-35)*[o]=<80pt,30pt>\hbox{\txt{Relative to\\ Coq object}}="b"*\frm<8pt>{-},
(80,55)*[o]=<80pt,15pt>\hbox{\txt{Terms}}="c"*\frm<8pt>{-},
(80,15)*[o]=<80pt,15pt>\hbox{\txt{Proofs}}="d"*\frm<8pt>{-},
(80,-15)*[o]=<80pt,15pt>\hbox{\txt{Terms}}="e"*\frm<8pt>{-},
(80,-55)*[o]=<80pt,15pt>\hbox{\txt{Proofs}}="f"*\frm<8pt>{-},
(160,55)*[o]=<120pt,15pt>\hbox{\txt{Visualise term clusters?}}="g"*\frm<8pt>{-,},
(160,15)*[o]=<120pt,15pt>\hbox{\txt{Visualise proof families?}}="h"*\frm<8pt>{-,},
(160,-15)*[o]=<120pt,15pt>\hbox{\txt{Visualise the term tree?}}="i"*\frm<8pt>{-,},
(160,-55)*[o]=<120pt,15pt>\hbox{\txt{Visualise the proof flow?}}="j"*\frm<8pt>{-,},
"a";"b" **\dir{-} ?>*\dir{>},
"a";"r" **\dir{-} ?>*\dir{>},
"r";"c" **\dir{-} ?>*\dir{>},
"r";"d" **\dir{-} ?>*\dir{>},
"b";"e" **\dir{-} ?>*\dir{>},
"b";"f" **\dir{-} ?>*\dir{>},
"c";"g" **\dir{-} ?>*\dir{>},
"d";"h" **\dir{-} ?>*\dir{>},
"e";"i" **\dir{-} ?>*\dir{>},
"f";"j" **\dir{-} ?>*\dir{>},
%"r";"a" **\dir{-} ?>*\dir{>},
%"c";"d" **\dir{-} ?>*\dir{>},
%"d";"h" **\dir{-} ?>*\dir{>},
%"c";"e" **\dir{-} ?>*\dir{>},
%"e";"c" **\dir{-} ?>*\dir{>}
%+<0em,0em>@+ 
\endxy
$$
\caption{Visualisation techniques included in ML4PG.}\label{fig:visualisation}
\end{figure}

In the previous sections, we have seen how ML4PG can be used to find families of proofs following 
a common pattern, and use those families to automatically prove theorems. However, when the proof-generation
method fails, the output provided by ML4PG algorithms is just a set of similar patches with no hints of why 
these proof-patches are deemed similar. 
% In addition, the user needs
% to replay the proof to infer the pattern. %Therefore, the conceptualisation of proof-patterns can be
% %a challenge. 
In this section, we present our approach to facilitate the understanding of proof patterns using different visualisation tools
that are summarised in Figure~\ref{fig:visualisation}.

\paragraph{An automaton-like representation of the proof-flow}
The first problem that we address is the discovery of the key features that were taken
into account during the cluster formation. This is a well-known problem in machine-learning 
known as \emph{feature selection}~\cite{Bishop}. From a given set of features, feature selection algorithms 
generate different subsets of features and create a ranking of feature subsets.
ML4PG uses the \emph{correlation-based feature subset selection} algorithm implemented in Weka~\cite{Weka}
--- the machine-learning toolbox employed by ML4PG to extract the relevant features.

\begin{example}\label{ex:correlation-features}
In the case study presented at the end of Section~\ref{sec:recurrent}, the relevant features were:

$-$ The tactics applied in the first and second steps of the proof-patches (\lstinline?case? and \lstinline?rewrite? respectively).

$-$ The type of the argument of the tactic applied in the first step of the proof-patches (\lstinline?nat?).

$-$ The second top symbol of the goal in the second step of the proof-patches (the $\sum$ symbol).

$-$ The first and second auxiliary lemmas applied in the the third step of the proof-patches. The first auxiliary lemma (\lstinline?sumrB?) is used to expand the summations and is 
common to all the proofs. There are two different \lq\lq{}second\rq\rq{} auxiliary lemmas that occur in the proofs of Lemma~\ref{lem:fundamental} and~\ref{lem:nilpotent2} (\lstinline?big_nat_recr? and
\lstinline?big_nat_recl? --- both extract elements of a summation); however, the recurrent feature extraction process have assigned similar values to them.

ML4PG uses this information and produces an automaton-shape representation for discovered % to help in the visualisation of 
proof-patterns and the correlated features that determined the patterns. In our example, the associated ``automaton'' is depicted in Figure~\ref{fig:automata}. 
\end{example}

% This feature selection mechanism helps to explain why the proof-patches are similar since it indicates the 
% features that have been relevant for the construction of the cluster. To show the selected features and their relation with 
% the proof-patches, we introduce an automaton-shape representation. In~\cite{GWR14}, a tool to infer models 
% from proof traces was presented. Those models help to understand the behaviour exhibited by a collection
% of proofs and is represented by automata.

Generally, given a cluster of proof patches ${\cal C}$, we have an automaton $A$ with $5$ consecutive states. The $i$th state of $A$ is labelled
with the list of $i$th goals in the proof-patches contained in ${\cal C}$. The transitions between the $i$th and $i+1$th states are given by 
the $i$th tactic of each proof-patch of ${\cal C}$. If two or more tactics belong to the same group (see Table~\ref{tab:tactics}),
they are merged in a unique transition; otherwise, the tactics will be shown as different transitions. In addition, each state is annotated with 
features 
whose correlation determined the cluster.

\begin{figure}[t]
\centering 
\includegraphics[scale=.19]{itp6.png}
\caption{\scriptsize{\emph{Fragment of the automaton 
 corresponding to the cluster of four lemma fragments described in the case study of Section~\ref{sec:recurrent}} (the whole automaton can be seen in Appendix~\ref{sec:automaton}). 
It shows correspondence between certain proof steps of lemmas % the following lemmas were discovered:
\texttt{telescope} (for the lemma of Example~\ref{example0}), \texttt{nilpotent\_inverse} (for Lemma~\ref{lem:nilpotent2}), and \texttt{persistent\_homology} 
(for Lemma~\ref{lem:fundamental}). Square boxes denote feature correlation where it exists. 
To be compact, we  hide full lemma statements, tactic and auxiliary lemma names, symbols, etc., but they can be shown by ML4PG. 
%For example, the full statements of the following cluster lemmas will be shown:
%The names of the lemmas of that cluster are: 
}}\label{fig:automata}
\end{figure}

\paragraph{A tree representation for terms}

As we have previously explained, ML4PG can cluster terms using a recurrent clustering process (see Section~\ref{sec:reclemmaclustering}). When using this tool, the output generated by ML4PG is a list of names of similar terms, and the user should inspect the libraries to compare the different terms. In order to simplify this process, we have created a visualisation tool that generates the ML4PG term-tree (cf. Definition~\ref{def:ml4pgtermtree}) given the name associated with a term.

\begin{example}
 In Example~\ref{ex:maxnACA}, we have shown how the lemma \lstinline?maxnACA? is automatically proven thanks to its similarity with lemmas \lstinline?addnACA?, \lstinline?minnACA?, \lstinline?mulnACA?. From Figure~\ref{fig:treevisualisation}, it is trivial why these 4 lemmas are similar.

\begin{figure}
\centering
\begin{tikzpicture}
\draw (0,0) node{\includegraphics[scale=.2]{maxnACA.png}} ;
\draw (7,0) node{\includegraphics[scale=.2]{minnACA.png}};
\draw (0,-1.5) node{\includegraphics[scale=.2]{addnACA.png}} ;
\draw (7,-1.5) node{\includegraphics[scale=.2]{mulnACA.png}}  ;
\end{tikzpicture}
\caption{Term trees associated with \texttt{maxnACA} (top left), \texttt{minnACA} (top right), \texttt{addnACA} (bottom left), \texttt{mulnACA} (bottom right).}\label{fig:treevisualisation} 
\end{figure}

 \end{example}

\paragraph{Cluster visualisation}

Finally, we have also improved the visualisation of clusters. Initially, ML4PG's output is a list of similar lemmas (or in general terms), and the user can adjust the accuracy of those families varying the granularity value (see the case studied included at the end of Section~\ref{sec:recurrent}). However, this means that he has to run the clustering process several times, and remember the families that were generated in each iteration. These problems have been tackled thanks to a new visualisation tool that captures the relations between the families of similar proofs/terms obtained using different granularity values, see Figure~\ref{fig:clustervisualisation}.

\begin{figure}
\centering
\includegraphics[scale=.3]{clustervisualisation.png}
\caption{Cluster visualisation for the case study included at the end of Section~\ref{sec:recurrent} regarding the families of similar lemmas to Lemma~\ref{lem:nilpotent}: \texttt{telescope} (for the lemma of Example~\ref{example0}), \texttt{nilpotent\_inverse} (for Lemma~\ref{lem:nilpotent2}), and \texttt{persistent\_homology} 
(for Lemma~\ref{lem:fundamental}). The outer, middle and inner square contain, respectively, the proof-patches obtained with granularities 1, 3 and 5. It shows correspondence between certain proof steps of lemmas % the following lemmas were discovered:
}\label{fig:clustervisualisation} 
\end{figure}

%for clustering that are related to that 
%state.

% $-$ there are several transitions from the $j$th to the $j+1$th state labelled by the $j$th tactic of each proof-patch of $P$ -- if either the $j$th tactic of two or more
% proof-patches is the same or belong to the same group (see Table~\ref{tab:tactics}), the transitions are merged in a unique transition labelled with the set of those tactics,
% 
% $-$ if a proof ends after the $j$th tactic of a proof-patch, the $j+1$ state is represented as final.

%\begin{example}

%\end{example}

### 2015-02-04 ACMTransCL/conceptualisation.tex
Remove commented-out parts

@@
 of similar patches with no hints of why these proof-patches are deemed similar.
-% In addition, the user needs
-% to replay the proof to infer the pattern. %Therefore, the conceptualisation of proof-patterns can be
-% %a challenge. 
 In this section, we present our approach to facilitate the understanding of proof patterns.
@@
 
-ML4PG uses this information and produces an automaton-shape representation for discovered % to help in the visualisation of 
+ML4PG uses this information and produces an automaton-shape representation for discovered
 proof-patterns and the correlated features that determined the patterns.
 Our graphical representation is simpler than other works where automata are used to represent models that are inferred from proof traces~\cite{GWR14}.
-% This feature selection mechanism helps to explain why the proof-patches are similar since it indicates the 
-% features that have been relevant for the construction of the cluster. To show the selected features and their relation with 
-% the proof-patches, we introduce an automaton-shape representation. In~\cite{GWR14}, a tool to infer models 
-% from proof traces was presented. Those models help to understand the behaviour exhibited by a collection
-% of proofs and is represented by automata.
 
@@
 whose correlation determined the cluster.
-%for clustering that are related to that 
-%state.
-
-% $-$ there are several transitions from the $j$th to the $j+1$th state labelled by the $j$th tactic of each proof-patch of $P$ -- if either the $j$th tactic of two or more
-% proof-patches is the same or belong to the same group (see Table~\ref{tab:tactics}), the transitions are merged in a unique transition labelled with the set of those tactics,
-% 
-% $-$ if a proof ends after the $j$th tactic of a proof-patch, the $j+1$ state is represented as final.
-
-%\begin{example}
 
@@
  corresponding to the cluster of four lemma fragments described in the case study of Section~\ref{sec:recurrent}} (the whole automaton can be seen in Appendix~\ref{sec:automaton}).
-It shows correspondence between certain proof steps of lemmas % the following lemmas were discovered:
+It shows correspondence between certain proof steps of lemmas
 \texttt{telescope} (for the lemma of Example~\ref{example0}), \texttt{nilpotent\_inverse} (for Lemma~\ref{lem:nilpotent2}), and \texttt{persistent\_homology}
@@
 To be compact, we  hide full lemma statements, tactic and auxiliary lemma names, symbols, etc., but they can be shown by ML4PG.
-%For example, the full statements of the following cluster lemmas will be shown:
-%The names of the lemmas of that cluster are: 
+
 }}\label{fig:automata}
 \end{figure}
-
-%\end{example}

### 2015-02-04 ACMTransCL/conclusions.tex
Remove commented-out parts

@@
 functionality to ML4PG: the user can receive suggestions about families of similar definitions, types and lemma shapes (in fact, any Coq terms).
-The \emph{proof-patch method} is employed to analyse the properties of the patches that constitute a proof.
-The whole syntax of Coq libraries is now subject to \emph{recurrent clustering}, which %combination
-%of term clustering and proof-patches
+The \emph{proof-patch method} is employed to analyse the properties of the patches that constitute a proof. The whole syntax of Coq libraries is now subject to \emph{recurrent clustering}, which
  increases the number and accuracy of families of similar proofs suggested by ML4PG.
@@
 proof development easier.
-
-%  \textbf{Related Work}
-%  on using statistical machine-learning in ITPs was focused on speeding up proof automation in different ITPs~\cite{lpar-urban,mash,UrbanSPV08}. 
-%  Comparing to these tools, we use unsupervised, rather than supervised, learning; and we do not use sparse  machine-learning methods.
-%  (See also \cite{KHG13} for a detailed comparison of different machine learning tools applied in various theorem provers.)
-%  We do not have a quantitative target when it comes to improving \emph{interactive} proof building experience: 
-%  no longer speed up in automated proof search or the number of automatically proven theorems are the main criteria of success.
-%  Instead, the user experience is the main parameter we target.
-%  We generally follow the ``qualitative'' intuition that ML4PG, being an interactive hint generator, must provide interesting and non-trivial hints on user's demand,
-%  and should be flexible and fast enough to do so in real  time,  at any stage of the proof, and relative to any chosen proof library.

### 2015-02-04 ACMTransCL/patches.tex
Remove commented-out parts

@@
 the user's interaction through the applied tactics. That method
-%tracks simple, 
 traced
@@
 This method  had two drawbacks.\\
-% (1) It only captures information 
-% about the top symbol of the goal; this means that information is mainly collected 
-% from symbols such as \lstinline?forall? or \lstinline?equal? (common to most proofs)
-% and features from other symbols (that are likely more relevant for the proof) are never captured.
 (1) It was  focused on the first
@@
 The previous section gave us the way of involving all Coq objects into recurrent feature re-evaluation.
-%, and, as we have explained in the previous section, machine-learning algorithms are sensitive to 
-%the assignment of feature values. These drawbacks have been overcome combining a new method called the 
-%\emph{proof-patch method} and the term feature extraction method presented in the previous section.
-
-%We start with introducing \emph{proof-patches} to address the first drawback. The underlying idea is that
-
-%We start defining the notions of Coq proof and proof-patch matrices.
 
@@
 
-From proof-patches, we can construct their feature matrices. %that will be provided to the machine-learning algorithms. 
+From proof-patches, we can construct their feature matrices.
 We will shortly define the feature extraction function $[.]_P=\langle[.]_M,[.]_{tac}\rangle : proof~patches \rightarrow M[\mathbb{Q}]_{5\times 6}$,
 where $[.]_{tac}$ is an injective function that has been introduced to assign values to tactics.
-%To be more concrete, 
 We have defined two versions of  $[.]_{tac}$: one for Coq tactics and another for SSReflect tactics.
-% -- although SSReflect is an extension of Coq, this package implements a set of proof tactics designed to support the extensive
-% use of small-scale reflection in formal proofs~\cite{SSReflect}. 
 In the SSReflect case, we divide the tactics into 7 groups and assign similar
 values to each tactic in the group, see Table~\ref{tab:tactics}.
-% If several tactics are concatenated using ``\lstinline?;?'', the value returned by
-% $[.]_{tac}$ is the concatenation of the values of the individual tactics. 
 Analogously for Coq tactics, cf. Appendix~\ref{sec:coqtactics}.
-
-%These clusters in turn are used to define a new function $[.]_{term}$. The function $[.]_{term}$ assigns numbers using the 
-%formula $[l]_{term}=2\times j + p$ where $C_j$ is a cluster such that $l\in C_j$ and $p$ is the proximity value of $l$ in $C_j$. This process assigns 
-%closer values to statements in the same cluster, and more distant numbers across groups. This function will be used in the following section.
 
@@
 We deliberately take lemmas belonging to very different SSReflect libraries. The lemma introduced in Example~\ref{example0} is a basic fact about
-summations. Lemma~\ref{lem:nilpotent} states a result about \emph{nilpotent} matrices %~\cite{BR91} 
+summations. Lemma~\ref{lem:nilpotent} states a result about \emph{nilpotent} matrices
 (a square matrix $M$ is \emph{nilpotent}
@@
  \hline
-%   \emph{g3} & - & -  & - & -  & - & -  \\
-%  \hline
-%   \emph{g4} & - & -  & - & -  & - & -  \\
-%  \hline 
-%   \emph{g5} & - & -  & - & -  & - & - \\
-%  \hline
 
@@
 When proving Lemma~\ref{lem:nilpotent}, the user may call ML4PG after completing a few standard proof steps: apply induction and solve the base case using rewriting.
-At this point it is difficult, even for an expert user, % 60 B?
+At this point it is difficult, even for an expert user,
 to get the intuition that he could reuse the proofs from Example~\ref{example0} and Lemma~\ref{lem:fundamental}. There are several reasons for this.
@@
 although the three lemmas involve summations,
-the types of the terms of those summations are different. Therefore, search based on types or keywords would not help. Even search  % *a* search removed -- because 59B??? 
+the types of the terms of those summations are different. Therefore, search based on types or keywords would not help. Even search
 of all the lemmas involving summations does not provide a clear suggestion, since there are more than $250$ lemmas (a considerable number for handling them manually) --- a clever search-pattern will considerably reduced the number of suggested lemmas, but such a pattern is not trivial at all.
-
-% \begin{table}
-%  	\centering
-%  	\tiny{
-%  		\begin{tabular}{|l|l|}
-%  		\hline
-%  	Goals and Subgoals & Proof-Steps (Tactics) \\
-%  		\hline
-%  		\hline
-%  	{\scriptsize $\forall~(M:M_n) (m:nat), M^m=0 \implies (1-M)\times \sum\limits_{i=0}^{m-1} M^i = 1$} & \\
-%  			& {\scriptsize \lstinline?move => M m nilpotent.?} \\
-%          {\scriptsize $(1-M)\times \sum\limits_{i=0}^{m-1} M^i = 1$} & \\
-%         &   {\scriptsize\lstinline?rewrite big_distrr mulmxBr mul1mx.?}\\
-%         {\scriptsize$   \sum\limits_{i=0}^{m-1} M^i - M^{i+1}$} &\\
-%         
-%         & {\scriptsize\lstinline?case : n.?}\\
-%         {\scriptsize $ \forall~(M:M_0) (m:nat),M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
-%         &  {\scriptsize\lstinline?by rewrite !thinmx0.?}\\
-%          {\scriptsize$ \forall~(M:M_{n+1}) (m:nat), M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $}&\\
-%  		\hline
-%  		\end{tabular}
-%  		
-%  		}
-%  	\caption{\textbf{First steps of the proof of Lemma~\ref{lem:nilpotent},} \emph{where the user stops and invokes ML4PG for a hint.}}
-%  	\label{tab:sumfirstn}
-%  \end{table}
 
@@
 
-%\begin{IP}\label{ip:math}
-%This proof pattern is followed by Lemmas~\ref{lem:nilpotent},~\ref{lem:fundamental} and~\ref{lem:lemma3}.
-
 \begin{PS}\label{ps:math}
@@
 \emph{
-%\begin{enumerate}
-  %\item 
 	Apply case on $n$.
@@
   \end{enumerate}
-% \end{enumerate}
 }
@@
 \end{PS}
-
-% \begin{table}
-%  	\centering
-%  	\footnotesize{
-%  		\begin{tabular}{|l|l|}
-%  		\hline
-%  	Goals and Subgoals & Proof-Steps (Tactics) \\
-%  		\hline
-%  		\hline
-%  	{\scriptsize $\forall~(M:M_n) (m:nat), M^m=0 \implies \exists N, N \times (1-M)= 1$} & \\
-%  			& {\scriptsize \lstinline?move => M m nilpotent.?} \\
-%          {\scriptsize $\exists N, N \times (1-M)= 1$} & \\
-%         &   {\scriptsize\lstinline?exists?}\\
-%         &   {\scriptsize~~\lstinline?\sum_(0<=i<m.+1) (pot_matrix M i).?}\\
-%         {\scriptsize$   (\sum\limits_{i=0}^{m-1} M^i) \times (1 - M) $} &\\
-%         
-%         & {\scriptsize\lstinline?rewrite big_distrl mulmxrB mulmx1.?}\\
-%         {\scriptsize $ \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
-%                 & {\scriptsize\lstinline?case : n.?}\\
-%         {\scriptsize $ \forall~(M:M_0) (m:nat),M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $ }&\\
-%         &  {\scriptsize\lstinline?by rewrite !thinmx0.?}\\
-%          {\scriptsize$ \forall~(M:M_{n+1}) (m:nat), M^m=0 \implies \sum\limits_{i=0}^{m-1} M^i - M^{i+1} $}&\\
-%  		\hline
-%  		\end{tabular}
-%  		
-%  		}
-%  	\caption{\textbf{First steps of the proof of Lemma~\ref{lem:nilpotent2}.} \emph{Note the correlation between this proof and the proof of Lemma~\ref{lem:nilpotent} in Table~\ref{tab:sumfirstn}.}}
-%  	\label{tab:nilpotent2}
-%  \end{table}
 
@@
 
-%There are two enhancements when we use the proof-patch method instead of the method presented in~\cite{KHG13}: 
 The new method brings two improvements:
@@
 
-%The proof-patch method increases the number of suggestions; this could be a problem if more irrelevant lemmas were suggested. However, 
 The new numbering of features produces more accurate clusters removing the irrelevant lemmas. In particular, using the default settings,
@@
 Lemma~\ref{lem:nilpotent2} is the only suggestion --- note that this is the closest lemma to Lemma~\ref{lem:nilpotent}.
-
-%The discovery of proof-patterns can be a challenging task, but increasing the number and precision of suggestions simplifies this problem. 
-%With the same goal (facilitate the discovery of proof-pattens), we introduce a graphical representation in the next section.
-
-% the four lemmas given above follow the Strategy \ref{ps:math} is our next benchmarking task for ML4PG.
-% Table~\ref{tab:compare0} shows the results that ML4PG will obtain, if the user varies the clustering algorithms and the size of clusters when 
-% calling ML4PG within the proof environment; we also highlight the exact solution to our benchmark task, given by calling K-means algorithm with granularity $4$.
-
-% \begin{table}
-% \centering
-% {\scriptsize 
-%    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-%      \hline
-%     & $g=1$& $g=2$& $g=3$& $g=4$ & $g=5$\\
-%       Algorithm: & ($n=84$) & ($n=94$) & ($n=108$) & ($n=126$) & ($n=151$)\\     \hline
-%      K-means &76$^{a-d}$ &51$^{a-d}$  &16$^{a-d}$ &\textbf{4}$^{a-d}$&2$^{b,d}$  \\
-%      \hline
-%      E.M. &26$^{a-d}$  &51$^{a,b,d}$ &41$^{a,b,d}$  &11$^{b,c,d}$ &3$^{a,b,d}$ \\
-%      \hline
-%      FarthestFirst &81$^{a-d}$ &48$^{a-d}$ &31$^{a-d}$  &25$^{a-d}$  &21$^{a-d}$  \\
-%      \hline
-%     \end{tabular}      }
-%           
-%    \caption{\textbf{Clustering experiments discovering Proof Strategy~\ref{ps:math}.} \emph{When  $g$ is chosen by the user, 
-%    ML4PG dynamically calculates the number of clusters $n$. Choosing Lemma~\ref{lem:nilpotent} for pattern-search,  
-%   the table shows the size of the single cluster that ML4PG displays to the user for every choice of $g$.
-%    The table shows sizes of clusters containing: $a)$ Lemma~\ref{lem:nilpotent}, $b)$ Lemma~\ref{lem:lemma3},  $c)$ Lemma~\ref{lem:fundamental} and
-%    $d)$ Lemma~\ref{lem:nilpotent2}. Note that with all variations of the learning algorithms and parameters, ML4PG is consistently grouping the
-% correct lemmas into clusters, albeit with varied degree of precision. Note also the effect of the granularity parameter, the smallest granularity value $g=1$ produces big 
-% clusters that are difficult to inspect; on the contrary, the biggest value $g=5$ produces small clusters that might not contain enough elements to spot a pattern. 
-% The most accurate result for this example is obtained using the K-means algorithm and 4 as granularity parameter.}}\label{tab:compare0}
-%   \end{table}
-%  
-% 
-% Table~\ref{tab:compare0} gives an insight on how granularity parameter $g$ can be used in pattern-recognition within ML4PG.
-%  One can use a ``bottom-up approach'' to detect patterns starting with the granularity value of 1 and increasing that value in successive calls.
-% Using the default values $g=3$ and K-means algorithm, ML4PG obtains 15 suggestions related to lemmas about summations including Lemmas~\ref{lem:lemma3} 
-% and~\ref{lem:fundamental}. Increasing the granularity level to 4, ML4PG discovers that Lemma  \ref{lem:nilpotent} is similar to 
-% Lemmas~\ref{lem:lemma3}, \ref{lem:fundamental} and \ref{lem:nilpotent2}. Finally, increasing the granularity level to the maximum level of $5$, 
-% ML4PG just discovers that Lemmas \ref{lem:nilpotent} and  \ref{lem:nilpotent2} are similar. %and also the following lemma;
-
-% When the proof analogy is found (as all four proof patches taken by ML4PG follow steps $1$ and $2(a)$ of Proof Strategy \ref{ps:math}), the user can use the remaining proof steps for Lemma~\ref{lem:nilpotent2} as a template for reconstructing the remaining proof for Lemma~\ref{lem:nilpotent} (by analogy with the steps $2(b)$ and $2(c)$ in Lemma~\ref{lem:nilpotent2}).
-% in case of the four lemmas mentioned here, the remaining steps all follow the Strategy \ref{ps:math}.
-Thus, ML4PG finds patterns arising from
-% The heterogeneous cluster obtained in this scenario belongs to the category of clusters consisting of theorems that share the same proof structure and as well as an 
-% auxiliary lemma (e.g. the lemma used to expand the summation).

### 2015-02-04 ACMTransCL/rec.tex
Remove commented-out parts

\section{Recurrent Term Clustering}\label{sec:reclemmaclustering}

The previous section introduced a method of defining statistically significant features.
It remains to define the functions
 $[.]_{term}$ and $[.]_{type}$ that will determine the feature values.
These functions must be sensitive to the structure of terms, assigning close values
to similar terms, and more distant values to unrelated terms.

A term \lstinline?t? is represented by the $300$ feature values of $[\texttt{t}]_M$.
The values of $[.]_{M}$
for  variables  and pre-defined sorts in \lstinline?t? are fixed, but the values of user-defined terms (and types!) contained in \texttt{t}
have to be computed recursively, based
on the structures of their definitions, and
using clustering to compute their feature vectors, and their representative values for $[\texttt{t}]_M$.
It is the nature of functional languages to have terms depending on other terms, and feature extraction/clustering cycle is
repeated recursively to reflect complex mutual term dependencies as feature values.
 We call this method \emph{recurrent clustering}: the function $[.]_M$ automatically (and recurrently) adapts to the given libraries and the current proof stage.
This differs from the standard machine-learning approach (and the old version of ML4PG), where the process of feature extraction is separated from running pattern-recognition algorithms. Here, one is a crucial part of another.

When Coq objects are divided into clusters, a unique integer number is assigned to each cluster. Clustering algorithms compute
a \emph{proximity value} (ranging from $0$ to $1$) to every object in a cluster to indicate the certainty of the given example belonging
to the cluster. The cluster numbers and the proximity values are used in the definitions of $[.]_{term}$ and $[.]_{type}$ below.

\begin{definition}\label{def:funterm}
Given a term  \lstinline?t? of a Coq library, the functions $[.]_{term}$ and $[.]_{type}$ are defined respectively for the term component \lstinline?t1?
and the type component \lstinline?t2? of every term-type node in the ML4PG term tree of \lstinline?t? as follows:

$-$ $[\texttt{t1}]_{term/type}=i$, if \lstinline?t1? is the $i$th distinct variable in \lstinline?t?.

$-$ $[\texttt{t1}]_{term/type}=100+\sum_{j=1}^i\frac{1}{10\times 2^{j-1}}$, if \lstinline?t1? is the $i$th element of the set\\ $\{\texttt{Set},\texttt{Prop},\texttt{Type(0)},
\texttt{Type(1)}, \texttt{Type(2)},\ldots\}$.

$-$ $[\texttt{t1}]_{term}=200+2\times j + p$, where $j$ is a number of a cluster $C_j$ computed by the latest run of term clustering,
such that $p$ is the proximity value of $\texttt{t1}$ in $C_j$.

$-$ $[\texttt{t2}]_{type}=200+2\times j + p$, where $j$ is a number of a cluster $C_j$ computed by the latest run of type clustering (i.e. term clustering restricted to types),
such that $p$ is the proximity value of $\texttt{t2}$ in $C_j$.
\end{definition}

Note the recurrent nature of the functions $[.]_{term}$ and $[.]_{type}$ where numbering of components of \lstinline?t? depends on the term definitions
and types included in the library, assuming those values are computed by iterating the process back to the basic definitions.  In addition, the function $[.]_{term}$ internally uses the function $[.]_{type}$ in the recurrent clustering
process and \emph{vice versa}.

In the above definition,
the variable encoding reflects the number and order of unique variables appearing in the term, note its similarity to the de Bruijn indexes.
In the formula for sorts, $\sum_{j=1}^i\frac{1}{10\times 2^{j-1}}$ reflects the close relation among sorts, and
$100$ is used to  distinguish sorts from variables and names.

Finally, the formula $200+2\times j + p$ assigns $[\texttt{t1}]$ (or $[\texttt{t2}]$) a value within $[200+2\times j,200+2\times j+1]$ depending on the
statistical proximity of \lstinline?t1? (or \lstinline?t2?) in cluster $j$. Thus, elements of the same cluster have closer values comparing to the values
assigned to elements of other clusters, sorts, and variables. The formula is the same for the functions $[.]_{term}$ and $[.]_{type}$, but it is computed with different
clusters and the values
occur in different cells of the term-tree matrices (cf. Definition~\ref{df:matrix}); thus, clustering algorithms distinguish terms and types on the level of features rather than feature values.

We can now state the main property of the ML4PG feature extraction.

\begin{proposition}
Let ${\cal T}$ be the set of Coq terms  whose trees have maximum depth $10$ and level index $10$.
Then, the function $[.]_{M}$ restricted to ${\cal T}$ is a one-to-one function.
\end{proposition}

Once the feature values of ML4PG term tree matrices have been computed, we can cluster these matrices and
obtain groups of similar terms. In particular, ML4PG can be used to cluster definitions, types and lemma
statements.
We finish this section with some clusters discovered among the 457 definitions of the basic infrastructure of the SSReflect
library~\cite{SSReflect}.

\begin{example}
We include here 3 of the 91 clusters discovered by ML4PG automatically in the SSReflect library of 457 terms (across 12 standard files), within 5--10 seconds.
Note that this example of cluster-search is not goal-oriented, ML4PG discovers patterns without any user guidance, and offers the user to consider term similarities of which he may
not be aware.

\begin{itemize}
 \item Cluster 1:
{\scriptsize \begin{lstlisting}
 Fixpoint eqn (m n : nat) :=
   match m, n with
   | 0, 0 => true | m'.+1, n'.+1 => eqn m' n'
   | _, _ => false end.
 Fixpoint eqseq (s1 s2 : seq T)  :=
   match s1, s2 with
   | [::], [::] => true | x1 :: s1', x2 :: s2' => (x1 == x2) && eqseq s1' s2'
   | _, _ => false end.
\end{lstlisting}}

\item  Cluster 2:
{\scriptsize
\begin{lstlisting}
 Fixpoint drop n s := match s, n with | _ :: s', n'.+1 => drop n' s' | _, _ => s end.
 Fixpoint take n s := match s, n with | x :: s', n'.+1 => x :: take n' s' | _, _ => [::] end.
\end{lstlisting}}

\item Cluster 3:
{\scriptsize
\begin{lstlisting}
 Definition flatten := foldr cat (Nil T).
 Definition sumn := foldr addn 0.
\end{lstlisting}}
\end{itemize}

\end{example}

The first cluster contains the definitions of equality for natural numbers and lists --- showing that
ML4PG can spot similarities across libraries. The second cluster discovers the relation between \lstinline?take? (takes the first $n$ elements of
a list) and \lstinline?drop? (drops the first $n$ elements of a list).
The last pattern is less trivial of the three, as it depends on
other definitions, like \lstinline?foldr?, \lstinline?cat? (concatenation of lists) and \lstinline?addn? (sum
of natural numbers).
Recurrent term clustering handles such dependencies well: it assigns close values to \lstinline?cat? and \lstinline?addn?,  since they have been discovered to belong to the same cluster.
Note the precision of ML4PG clustering. Among $457$ terms it considered, $15$ used \lstinline?foldr?, however,  Cluster 3 contained only $2$ definitions, excluding e.g.
\lstinline?Definition allpairs s t:=foldr (fun x => cat (map (f x) t)) [::] s? ; \lstinline?Definition divisors n:=foldr add_divisors [:: 1] (prime_decomp n)?  or \lstinline?Definition Poly:=foldr cons_poly 0.? --- this is due to the recurrent clustering process since functions like \lstinline?add_divisors? or \lstinline?cons_poly? are not clustered together with functions \lstinline?cat? and \lstinline?addn?.

To summarise, there are three main properties that distinguish ML4PG pattern search from standard Coq search commands:
\begin{itemize}
	\item the user does not have to know and provide any search pattern;
	\item the discovered clusters do not have to follow a \lq\lq{}pattern\rq\rq{} in a strict sense (e.g. neither exact symbol names nor their order make a pattern), but ML4PG considers structures and background information found in the library; and,
	\item working with potentially huge sets of Coq objects, ML4PG makes its own intelligent discrimination of more significant and less significant patterns,
	as example with \lstinline?foldr? has shown. This is opposed to the classic search for \lstinline?foldr? pattern that would present the user with a set of $15$ definitions.
\end{itemize}

ML4PG can also work in a goal-directed mode,
and discover only clusters of terms that are similar to the given term \lstinline?t?.
This can speed-up the proof development in two different ways.
In addition, clustering will provide definitions of terms similar to \lstinline?t?;
hence, the proofs of the theorems
involving those terms may follow similar patterns.
Clustering can also discover that a newly defined term \lstinline?t? was previously defined (perhaps in a different notation, as ML4PG works with structures across notations);
in that case, the user can use the existing library definition and all its background theory instead of defining it from scratch.

### 2015-02-04 ACMTransCL/visualisation.tex
Remove commented-out parts

@@
 "f";"j" **\dir{-} ?>*\dir{>},
-%"r";"a" **\dir{-} ?>*\dir{>},
-%"c";"d" **\dir{-} ?>*\dir{>},
-%"d";"h" **\dir{-} ?>*\dir{>},
-%"c";"e" **\dir{-} ?>*\dir{>},
-%"e";"c" **\dir{-} ?>*\dir{>}
-%+<0em,0em>@+ 
 \endxy
@@
 these proof-patches are deemed similar.
-% In addition, the user needs
-% to replay the proof to infer the pattern. %Therefore, the conceptualisation of proof-patterns can be
-% %a challenge. 
 In this section, we present our approach to facilitate the understanding of proof patterns using different visualisation tools
@@
 
-ML4PG uses this information and produces an automaton-shape representation for discovered % to help in the visualisation of 
+ML4PG uses this information and produces an automaton-shape representation for discovered
 proof-patterns and the correlated features that determined the patterns. In our example, the associated ``automaton'' is depicted in Figure~\ref{fig:automata}.
 \end{example}
-
-% This feature selection mechanism helps to explain why the proof-patches are similar since it indicates the 
-% features that have been relevant for the construction of the cluster. To show the selected features and their relation with 
-% the proof-patches, we introduce an automaton-shape representation. In~\cite{GWR14}, a tool to infer models 
-% from proof traces was presented. Those models help to understand the behaviour exhibited by a collection
-% of proofs and is represented by automata.
 
@@
 To be compact, we  hide full lemma statements, tactic and auxiliary lemma names, symbols, etc., but they can be shown by ML4PG.
-%For example, the full statements of the following cluster lemmas will be shown:
-%The names of the lemmas of that cluster are: 
 }}\label{fig:automata}
@@
 The outer, middle and inner square contain, respectively, the proof-patches obtained with granularities 1, 3 and 5.
-It shows correspondence between certain proof steps of lemmas % the following lemmas were discovered:
+It shows correspondence between certain proof steps of lemmas
 }\label{fig:clustervisualisation}
 \end{figure}
-
-%for clustering that are related to that 
-%state.
-
-% $-$ there are several transitions from the $j$th to the $j+1$th state labelled by the $j$th tactic of each proof-patch of $P$ -- if either the $j$th tactic of two or more
-% proof-patches is the same or belong to the same group (see Table~\ref{tab:tactics}), the transitions are merged in a unique transition labelled with the set of those tactics,
-% 
-% $-$ if a proof ends after the $j$th tactic of a proof-patch, the $j+1$ state is represented as final.
-
-%\begin{example}
-
-%\end{example}

### 2015-02-04 CICM2015/automaton.tex
Renamed

\newpage

\section{Automaton}\label{sec:automaton}

\begin{figure}[h]
    \centering
\includegraphics[scale=0.28]{itp.png} 
\caption{\scriptsize{\emph{Automaton corresponding to the proof cluster of four lemma fragments described in the case study of Section~\ref{sec:recurrent}.} The automaton shows the five proof steps, of which the first three are shown to influence the cluster formation.
It uses Lemma names: \texttt{telescope} for the lemma of Example~\ref{example0}, \texttt{nilpotent\_inverse} for Lemma~\ref{lem:nilpotent2}, and \texttt{persistent\_homology} 
for Lemma~\ref{lem:fundamental}.
  In addition to Lemma names, ML4PG can show lemma statements, and it can provide details of the \lq\lq{}Tactic applied\rq\rq{}, \lq\lq{}Type tactic argument\rq\rq{}, \lq\lq{}Second top symbol\rq\rq{}, \lq\lq{}First/Second lemma applied\rq\rq{} fields.}}
\end{figure}

### 2015-02-04 CICM2015/coqtactics.tex
Renamed

\newpage

\section{Groups of Coq tactics and number assignment}\label{sec:coqtactics}

% We have divided the Coq tactics in the groups presented in Table~\ref{tab:coqtactics}.

Table~\ref{tab:coqtactics} splits the Coq tactics into different groups. The formula used in the function $[.]_{tactic}$ 
to compute the value of a Coq tactic is given by 
$i+\sum_{j=0}^k\frac{1}{10 \times 2^{j-1}}$ where $i$ is the group of the tactic and $k$ is the position 
of the tactic in that group (cf. right side of Table~\ref{tab:coqtactics}).

\begin{table}[h]
\tbl{{\scriptsize \emph{Groups of Coq tactics}}\label{tab:coqtactics}}{
\centering
\begin{tabular}{l||l}

Group & Tactics of the group\\

\hline
\hline
Group 1:           &  \lstinline?exact, eexact, assumption, eassumption, ?\\
Applying theorems &  \lstinline?refine, apply, eapply, simple apply, lapply, ?\\
\hline
Group 2:           &  \lstinline?constructor, split, exists, left, right, ?\\
Managing inductive constructors & \lstinline?econstructor, esplit, eexists, eleft, eright?\\
\hline
Group 3:           &  \lstinline?intro, intros, clear, rever, move, rename,  ?\\
Managing local context & \lstinline?set, remember, pose, decompose?\\
\hline
Group 4:            &  \lstinline?assert, cut, pose, specialize, generalize, ?\\
Controlling proof flow & \lstinline?evar, instantiate, admit, absurd, ?\\
& \lstinline?contradition, contradict, exfalso?\\
\hline
Group 5:            & \lstinline?destruct, case, ecase, simple destruct,?\\
Case analysis and induction & \lstinline?induction, einduction, elim, eelim, ?\\
 & \lstinline?simple induction, double induction, ?\\
     & \lstinline?dependent induction, functional induction,?\\
  & \lstinline? discriminate, injection, fix, cofix, ?\\
    & \lstinline? case_eq, elimtype?\\
\hline
Group 6:          &  \lstinline?rewrite, erewrite, cutrewrite, replace, ?\\
Rewriting expressions &  \lstinline?reflexivity, symmetry, transitivity, subst, ?\\
&  \lstinline?stepl, change?\\
\hline
Group 7:          &  \lstinline?cbv, compute, vm_compute, red, hnf, simpl, ?\\
Performing computations  &  \lstinline?unfold, fold, pattern, conv_tactic?\\
\hline
Group 8:           &  \lstinline?auto, trivial, eauto, autounfold, ?\\
Automation       &  \lstinline?autorewrite?\\
\hline
Group 9:           &  \lstinline?tauto, intuition, rtauto, firstorder, ?\\
Decision procedures & \lstinline?congruence?\\
\hline
Group 10 &  \lstinline?decide equality, compare, simplify_eq,?\\
Equality & \lstinline?esimplify_eq?\\
\hline
Group 11 &  \lstinline?inversion, dependent inversion,?\\
Inversion & \lstinline?functional inversion, quote?\\
\hline
Group 12 & \lstinline?classical_left, classical_right?\\
Classical tactics & \lstinline??\\
\hline
Group 13 & \lstinline?omega, ring, field, fourier?\\
Automatizing & \\
\hline
Group 14 & Rest of Coq tactics\\
\hline
\end{tabular}}
\end{table}

### 2015-02-04 CICM2015/defManSSR.tex
Renamed

%=======================================================================
% Défini le style ssr Gallina pour les listings (Assia Mahboubi 2007)

\lstdefinelanguage{SSR} {

% Anything betweeen $ becomes LaTeX math mode
mathescape=true,						
% Comments may or not include Latex commands
texcl=false,

% Vernacular commands
morekeywords=[1]{
Section, Module, End, Require, Import, Export, Coercions,
Variable, Variables, Parameter, Parameters, Axiom, Hypothesis, Hypotheses,
Notation, Local, Tactic, Reserved, Scope, Open, Close, Bind, Delimit,
Definition, Let, Ltac, Fixpoint, CoFixpoint, Add, Morphism, Relation,
Implicit, Arguments, Set, Unset, Contextual, Strict, Prenex, Implicits,
Inductive, CoInductive, Record, Structure, Canonical, Coercion,
Theorem, Lemma, Corollary, Proposition, Fact, Remark, Example,
Proof, Goal, Save, Qed, Defined, Hint, Resolve, Rewrite, View,
Search, Check, Show, Print, Printing, All, Graph, Projections, inside,
outside, Locate, Maximal, Types, Type, SearchAbout, SearchPattern, SearchRewrite,
Match, Elim},

% Gallina
morekeywords=[2]{forall, exists, exists2, fun, fix, cofix, struct,
      match, with, end, as, ' in ', return, let, if, is, then, else,
      for, of, nosimpl},

% Sorts
morekeywords=[3]{Set, Prop},

% Various tactics, some are std Coq subsumed by ssr, for the manual purpose
morekeywords=[4]{
         pose, set, move, case, elim, apply, clear,
            hnf, intro, intros, generalize, rename, pattern, after,
	    destruct, induction, using, refine, inversion, injection,
         rewrite, congr, unlock, compute, ring, field,
            replace, fold, unfold, change, cutrewrite, simpl,
         have, suff, wlog, suffices, without, loss, nat_norm,
            assert, cut, trivial, revert, bool_congr, nat_congr,
	 symmetry, transitivity, auto, split, left, right, autorewrite},

% Terminators
morekeywords=[5]{
         by, done, exact, reflexivity, tauto, romega, omega,
         assumption, solve, contradiction, discriminate},

% Control
morekeywords=[6]{do, last, first, try, idtac, repeat},

% Various symbols
% For the ssr manual we turn off the prettyprint of formulas
% literate=	
%	{->}{{$\rightarrow\,$}}2		
% 	{->}{{\tt ->}}3
%	{<-}{{$\leftarrow\,$}}2	
% 	{<-}{{\tt <-}}2			
% 	{>->}{{$\mapsto$}}3		
% 	{<=}{{$\leq$}}1			
% 	{>=}{{$\geq$}}1		
% 	{<>}{{$\neq$}}1
% 	{/\\}{{$\wedge$}}2	
% 	{\\/}{{$\vee$}}2 		
% 	{<->}{{$\leftrightarrow\;$}}3
% 	{<=>}{{$\Leftrightarrow\;$}}3
% 	{:nat}{{$~\in\mathbb{N}$}}3	
%	{fforall\ }{{$\forall_f\,$}}1		
%	{forall\ }{{$\forall\,$}}1		
%	{exists\ }{{$\exists\,$}}1
% 	{negb}{{$\neg$}}1
% 	{spp}{{:*:\,}}1
% 	{~}{{$\sim$}}1
% 	{\\in}{{$\in\;$}}1
% 	{/\\}{$\land\,$}1
% 	{:*:}{{$*$}}2
%	{=>}{{$\,\Rightarrow\ $}}1
% 	{=>}{{\tt =>}}2
% 	{:=}{{{\tt:=}\,\,}}2
% 	{==}{{$\equiv$}\,}2
% 	{!=}{{$\neq$}\,}2
% 	{^-1}{{$^{-1}$}}1
% 	{elt'}{elt'}1
%       {=}{{\tt=}\,\,}2
%       {+}{{\tt+}\,\,}2,

% Comments delimiters, we do turn this off for the manual
%comment=[s]{(*}{*)},

% Spaces are not displayed as a special character
showstringspaces=false,

% String delimiters
morestring=[b]",
morestring=[d]’,

% Size of tabulations
tabsize=3,

% Enables ASCII chars 128 to 255
extendedchars=true,

% Case sensitivity
sensitive=true,

% Automatic breaking of long lines
breaklines=true,

% Default style fors listings
basicstyle=\ttfamily,

% Position of captions is bottom
captionpos=b,

% Full flexible columns 
columns=[l]fullflexible,

% Style for (listings') identifiers
identifierstyle={\ttfamily\color{black}},
% Note : highlighting of Coq identifiers is done through a new
% delimiter definition through an lstset at the begining of the
% document. Don't know how to do better.

% Style for declaration keywords
keywordstyle=[1]{\ttfamily\color{dkviolet}},

% Style for gallina keywords
keywordstyle=[2]{\ttfamily\color{dkgreen}},

% Style for sorts keywords
keywordstyle=[3]{\ttfamily\color{lightblue}},

% Style for tactics keywords
keywordstyle=[4]{\ttfamily\color{dkblue}},

% Style for terminators keywords
keywordstyle=[5]{\ttfamily\color{red}},

%Style for iterators
keywordstyle=[6]{\ttfamily\color{dkviolet}},

% Style for strings
stringstyle=\ttfamily,

% Style for comments
commentstyle=\rmfamily,

}

### 2015-02-04 CICM2015/lemmaclustering.tex
Renamed

--- 2015-02-04 ACMTransCL/lemmaclustering.tex
@@
 ML4PG uses (unsupervised) \emph{clustering algorithms}~\cite{Bishop} to find patterns in Coq syntax and proofs.
-Clustering algorithms % is a set of machine-learning techniques that 
+Clustering algorithms
 divide data into $n$ groups of similar objects (called \emph{clusters}), where the value of $n$
@@
 A detailed exposition of machine-learning algorithms involved in ML4PG can be found in \cite{KHG13}. In this paper, our first goal consists in improving ML4PG's feature extraction algorithm.
-%In this section, we present a method to collect features from Coq terms.
 
 \emph{Feature extraction}~\cite{Bishop} is a research area developing methods for
-discovery of statistically significant features in data. % in machine-learning, known as . 
-We adopt the following standard terminology. % concerning feature extraction. 
+discovery of statistically significant features in data.
+We adopt the following standard terminology.
 We assume there is a training data set, containing some samples (or objects).
@@
 conducted in an $n$-dimensional space.
-For this reason, most pattern-recognition tools will require that %the features have numeric values; and 
+For this reason, most pattern-recognition tools will require that
 the number of selected features is limited and fixed (sparse methods, like the ones applied in
@@
 If an object is characterised by $n$ feature values, these $n$ values together form a \emph{feature vector} for this object.
-%Irrespective of the particular algorithm
-%used, 
 A function that assigns, to every object of the data set, a feature vector is called a \emph{feature extraction function}.
-Normally, feature extraction is a data pre-processing stage, and it is separated from the actual pattern-recognition process. %ML4PG relies on unsupervised clustering methods~\cite{KHG13}.
+Normally, feature extraction is a data pre-processing stage, and it is separated from the actual pattern-recognition process.
 
@@
 for ACL2 first-order terms. Here, that ACL2 method is substantially re-defined to capture the higher-order dependently-typed language of Coq.
-%However, due to the different nature of ACL2 and Coq -- ACL2 being a first-order untyped language, and Coq being a higher-order
-%dependently-typed language -- the algorithm presented in~\cite{lpar13} cannot be extrapolated to Coq, and a different method is necessary.
-%Therefore, we introduce here a new feature-extraction method for Coq terms.
 
-The underlying formal language of Coq is known as the \emph{Predicative Calculus of (Co)Inductive Constructions} (pCIC)~\cite{Coq,CoquandH88,CoPa89}. % -- a 
-%weaker calculus of the \emph{Calculus of (Co)Inductive Constructions}. 
+The underlying formal language of Coq is known as the \emph{Predicative Calculus of (Co)Inductive Constructions} (pCIC)~\cite{Coq,CoquandH88,CoPa89}.
 The terms of pCIC are built from
@@
 category in the Gallina specification language~\cite{Coq} and corresponds to the intuitive notion of well-formed expression.
-%Our feature extraction
-%method works for all well-formed expressions that can be defined in Coq; 
-However, for the purpose of concise exposition, % purposes,
+However, for the purpose of concise exposition,
 we will restrict our notion of a term to Definition~\ref{def:coqterms},
 giving the full treatment of the whole Coq syntax in the actual ML4PG implementation.
- % %definitions to the syntax of Coq terms coming from pCIC terms, thereby remembering that full Coq syntax is handled in the actual ML4PG implementation.
 
@@
 labelled by Gallina-keywords and special tokens such as \lstinline?forall?, \lstinline?fun?, \lstinline?let? or \lstinline?->? (from now on, we will call them Gallina tokens);
-%to
-%Gallina keywords and special tokens); 
 and the term-type nodes are labelled by expressions of the form \lstinline?t1:t2? where \lstinline?t1? is a sort, a variable or
@@
 
-We represent ML4PG term trees as feature matrices, further to be flattened into feature vectors for clustering. %they in turn will be the input of machine-learning algorithms.
+We represent ML4PG term trees as feature matrices, further to be flattened into feature vectors for clustering.
 A variety of methods exists to represent trees as matrices, for instance using
 adjacency or incidence matrices. The adjacency matrix and the various previous methods of  feature
-extraction (e.g.~\cite{lpar-urban,K13,UrbanSPV08}) share the following common properties: different %tree
-%nodes 
-library symbols are represented by distinct features, and the %matrix entries
-feature values are binary. %; and the size of the matrix depends on the tree size. 
+extraction (e.g.~\cite{lpar-urban,K13,UrbanSPV08}) share the following common properties: different
+library symbols are represented by distinct features, and the
+feature values are binary.
 For
@@
 
-%In~\cite{lpar13}, we represent ACL2 term trees as matrices with the following properties: the features are 
-%given by the arity of functions and the term tree depth -- two properties common to every term tree -- and the 
-%feature values are rational numbers dynamically computed depending on the definition of the functions involved in the term tree. The size of 
-%those matrices is fixed to $49$; hence, allowing the application of data-mining techniques to both big and small 
-%data sets of ACL2 theorems -- large and sparse feature vectors require to use big data sets, 
-%where the size of the example data set is comparable to the number of features; on the contrary, compact feature extraction methods work equally well 
-%with both big and small datasets. The method presented in~\cite{lpar13}
-%can be applied not only to ACL2 but also to any first-order theory; however, it is not sufficient to capture all the information 
-%from the higher-order dependently-typed language of Coq.
-
 We develop a new compact method that tracks a large (potentially unlimited) number of Coq terms by a finite number of features and an unlimited number of feature-values.
 In our method, the features are given by two properties common to all possible term trees: the term tree depth and the level index of nodes.
-The most important information about the term is then encoded by improving precision of feature values %means of a more precise specification of feature values, 
+The most important information about the term is then encoded by improving precision of feature values
 using rational-valued feature-extraction functions. Taking just $300$ features, the new feature-extraction method recursively adjusts the feature values,
-adapting to the growing language syntax. %; thus producing a machine-learning analogue of how an unlimited number of Coq terms can be generated from the limited syntactic constructs.  
+adapting to the growing language syntax.
 The resulting feature vectors have an average density ratio of 60\%.
@@
 
-	%%% Introduce common notation for all three; may be call it more like feature exatraction... template f.e.? pre-fe.e?
-
-% \begin{definition}[ML4PG term tree matrix]\label{df:matrix}
-% Given a term tree $T$ for a Coq term, and three injective functions $[.]_{Gallina}: Gallina~tokens \rightarrow \mathbb{Q}^-$, $[.]_{term}: Coq~Terms \rightarrow \mathbb{Q}^+$, $[.]_{type}: Coq~Terms \rightarrow \mathbb{Q}^+$,
-% the ML4PG term tree matrix $M_M$ of $T$ is a $10 \times 10$ matrix such that the $(i,j)$-th entry of $M_M$ captures information from the node of depth $i$ and
-% level index $j$ of $T$ as follows:
-% 
-% $-$ if the node of depth $i$ and level index $j$ of $T$ is a Gallina node $g$, then the $(i,j)$-th entry of $M_M$ is a triple $([g]_{Gallina},-1,p)$ 
-% where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$. 
-% 
-% $-$ if the node of depth $i$ and level index $j$ of $T$ is a term-type node \lstinline?t1:t2?, then the $(i,j)$-th entry of $M_M$ is a triple $([t1]_{term},[t2]_{type},p)$
-% where $p$ is the level index of the direct parent of the node.
-% 
-% \end{definition}
-
 We use the notation $M[\mathbb{Q}]_{n\times m}$ to denote the set of matrices of size  $n\times m$ with rational coefficients.
@@
 $-$ if $T_{\texttt{t}}(i,j)$ is a Gallina node $g$, then the $(i,j)$th entry of $[\texttt{t}]_M$ is a triple $([g]_{Gallina},-1,p)$
-where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$.
+where $p$ is the level index of the parent of $g$.
 
@@
 across all Coq terms appearing in the libraries. We may lose some information if  pruning is needed, but the chosen size works well for most terms appearing in Coq libraries.
-%The above method works for any Coq term, irrespective of the size and complexity. 
 If a term tree does not fit into $10 \times 10$ term tree dimensions, its $10 \times 10$
 subtree is still considered by ML4PG.
-%For the machine-learning algorithms, 
 The term tree matrix is flattened into a feature vector, and each triple will be split
-into three components of the vector, giving a feature vector size of $300$, %-- bigger than the size of vectors in~\cite{lpar13}, 
+into three components of the vector, giving a feature vector size of $300$,
 still smaller than in sparse approaches~\cite{lpar-urban,K13,UrbanSPV08}.
@@
 assign a concrete value to each of them. The function $[.]_{Gallina}: Gallina~tokens  \rightarrow \mathbb{Q}^-$ is an injective
-function carefully defined to assign close values to similar Gallina tokens %(for instance, 
-%$[$\lstinline?forall?$]_{Gallina}=-1$ and $[$\lstinline?->?$]_{Gallina}=-1.1$);
+function carefully defined to assign close values to similar Gallina tokens
 and more distant numbers to unrelated tokens
-%(e.g. $[$\lstinline?let?$]_{Gallina}=-3$)
 --- see Appendix~\ref{sec:gallinasyntax} for the exact encoding.
- %; thus distinguishing the 
-%different kinds of Gallina tokens unambiguously.
 
-%On the contrary to the function $[.]_{Gallina}$, 
 The functions $[.]_{term}$ and $[.]_{type}$ are dynamically re-defined for every
 library and every given proof stage, to adapt to the changing syntax.
- %, as we are going to describe in 
-%Definition~\ref{def:funterm}. 
-In practice, there will be  %set of functions computed in 
-%every session: 
+In practice, there will be
 new $[.]_{term}$ and $[.]_{type}$ functions computed whenever ML4PG is called.
@@
 brings the element of the ``acquired knowledge/experience'' to the machine-learning cycle, as will be formalised in the next section.
-
-% In the previous section, we have presented a method to extract features from 
-% Coq terms; however, the functions $[.]_{Gallina}$, $[.]_{term}$ and $[.]_{type}$
-% are still undefined. 
-% % The definition of these functions influences the results that 
-% % are obtained by machine-learning algorithms.
-% In this section, we explain how these functions are defined and how this has led to several improvements in ML4PG.
-
-%We finish this section with the following property about ML4PG term tree matrices.
-
-% \begin{lemma}
-% Let ${\cal T}$ be the set of Coq terms whose nodes have maximum depth $10$ and level index $10$. 
-% Then, the function $[.]_{T}$ restricted to ${\cal T}$ is a one-to-one function. 
-% %  Given a term tree matrix $M$ associated with a Coq term $C$. If the maximum depth of the term tree of $C$ is 10 and the maximum index level of the 
-% %  nodes of the tree is $10$, then the term $C$ can be reconstructed from $M$. 
-% \end{lemma}

### 2015-08-06 AI4FM/render.sh
Added render

@@ -1,9 +1,25 @@
 #! /usr/bin/env nix-shell
 #! nix-shell -i bash -p bash -p md2pdf
 
-function go {
+function renderMd {
     ARGS="--filter pandoc-citeproc" md2pdf
 }
+
+function mdToLatex {
+    pandoc -s --filter pandoc-citeproc -o abstract.tex abstract.md
+}
+
+function renderLatex {
+    latex abstract
+    latex abstract
+    pdflatex abstract
+}
+
+function go {
+    renderLatex
+}
+
+# Run once (default) or forever (if given any argument)
 
 go
 

### 2015-08-06 AI4FM/abstract.tex
Render markdown to TeX

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Chris Warburton},
            pdftitle={Scaling Automated Theory Exploration},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

\title{Scaling Automated Theory Exploration}
\author{Chris Warburton}
\date{}
\usepackage[T1]{fontenc}
\usepackage{upquote}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\section{Abstract}\label{abstract}

In this paper we investigate the \textbf{theory exploration} (TE)
paradigm for computer-assisted Mathematics and identify limitations and
improvements of current approaches. Unlike the theorem-proving paradigm,
which requires user-provided conjectures, TE performs an open-ended
search for theorems satisfying given criteria. We see promise in TE for
identifying new abstractions and connections in libraries of software
and proofs, but realising this potential requires more scalable
algorithms than presently used.

\section{Introduction}\label{introduction}

The \emph{theory exploration} (TE) paradigm provides software support
for traditional Mathematical workflows; namely, deriving ``interesting''
consequences from formal definitions (Buchberger 1999). Early
implementations like \textsc{Theorema} (Buchberger 2000) emphasised
interactivity, in a similar way to computer algebra systems like
\textsc{Mathematica} (in which \textsc{Theorema} is implemented) or
interactive theorem provers. Subsequent systems have investigated
\emph{automated} theory exploration, for tasks such as lemma discovery
(Johansson et al. 2014).

By removing user interaction, automated TE systems require an algorithm
for deciding whether a theorem is ``interesting''. In existing systems,
this is intimately connected to the choice of search algorithm, to
ensure tractability. For example, \textsc{IsaCoSy} (Johansson, Dixon,
and Bundy 2009) discovers equations, which are defined as
``interesting'' if they cannot be simplified using previously discovered
equations. The intuition for such criteria is to avoid special cases of
known theorems, such as \(0 + 0 = 0\), \(0 + 1 = 1\), etc. when we
already know \(\forall x. 0 + x = x\). To work effectively, this
requires general forms to be found \emph{before} special cases.

TE is a ``bottom-up'' process: new theorems are derived from given
knowledge and previous theorems. In particular, there is no special
``goal'' theorem(s) as found in interactive theorem proving; all we have
is our (proxy) ``interesting'' criterion, and the definitions of our
theory. This makes theory exploration a \emph{combinatorial
optimisation} problem, which is well studied in Artificial Intelligence
and Machine Learning. Existing TE systems rely on complete, brute-force
search techniques which do not scale to theories of a significant size;
these could be replaced by a variety of faster, approximate algorithms,
eg. those surveyed in (Blum et al. 2011).

Whilst efficient on a small scale, where we expect the chosen inputs to
be related, we cannot expect an effective output criterion to coincide
so readily with search algorithm and output criterion to be effective
for large inputs containing many irrelevant connections. umbers of also
consider data mining approaches, which infer general rules \emph{after}
enumerating some number of special cases. A variety of
``interestingness'' measures are used in domains such as data mining
(Geng and Hamilton 2006), which may be adapted to a TE setting. These
include objective measures, eg. based on information theory, and
subjective measures which ask for the user's opinion. We can avoid
asking users directly, by treating existing proof libraries and test
suites as representative examples of what is interesting, and hence
construct an oracle; such oracles have already been used for
precision-recall experiments to evaluate the performance of
\textsc{HipSpec} (Claessen et al. 2013).

\section{Tackling Complexity}\label{tackling-complexity}

Even with a slow search algorithm, we can use a divide and conquor
approach to limit the number of allowed combinations, either by using
stricter types to prevent composition, or by partitioning the theory
into small independently-searched sub-theories. Of course, such
restrictions should strike a balance between the efficiency gained and
the potential to forbid some interesting theorems.

\section{Existing Work}\label{existing-work}

Automated theory exploration has been applied to libraries in Isabelle
and Haskell, although we focus on the latter as its implementations are
the most mature (\textsc{Hipster} actually explores Isabelle by
translating code to Haskell first!). Haskell is interesting to target,
since its functional purity and algebraic structure make equational
properties common; recursion and higher-order functions make automation
non-trivial; and since these properties can't be expressed in Haskell
itself (without difficulty (Lindley and McBride 2014), at least), less
effort is spent discovering and proving these properties compared to
proof-oriented systems like Isabelle.

Due to Haskell's relative popularity, there are large code repositories
such as \textsc{Hackage} available to explore, with the potential to
benefit existing library authors and users in comprehending and
maintaining their code (Claessen, Smallbone, and Hughes 2010).

Currently, the most powerful TE system for Haskell is \textsc{HipSpec}.
This uses \textsc{QuickSpec} to search through \emph{expressions}
(combinations of the Haskell terms given by the theory), rather than
searching through the space of equations or proofs directly. Expressions
are grouped into equivalence classes, such that the \textsc{QuickCheck}
counterexample finder cannot distinguish between the elements; equations
relating the members of these classes are then conjectured, and sent to
existing automated theorem provers to try and prove (Rosén 2012). This
approach works well as a lemma generation system, making
\textsc{HipSpec} a capable inductive theorem prover as well as a theory
exploration system (Claessen et al. 2013).

\section{Going Forward}\label{going-forward}

Given this state of the art, we identify the following as potential
areas for improvement:

\begin{itemize}
\tightlist
\item
  Expression enumeration is brute-force; this could scale to larger
  terms and theories using a heuristic algorithm.
\item
  ``Interestingness'' is a fixed part of the algorithm: an equation is
  interesting if it cannot be derived from previous equations. As we
  increase the size of our theory, this becomes unsatisfying in two
  ways:

  \begin{itemize}
  \tightlist
  \item
    The number of irreducible equations grows, making it desirable to
    impose extra conditions of a more subjective nature.
  \item
    Surprising, insightful equations may be discarded, if they are
    actually reducible in some complex, non-obvious way. A more
    subjective interestingness measure could be used to veto such
    rejections.
  \end{itemize}
\item
  The system does not propose candidate equations by data mining
  previous results; generalisation methods like anti-unification could
  do this, and at the same time remove the requirement that general
  forms must be enumerated early.
\item
  All type-safe combinations of the given expressions are tried, whilst
  it may be discernable a priori that some combinations are not worth
  considering (either because they are never related, or because their
  relations are never interesting). A pre-processor could make large
  theories more tractable by selecting combinations which are likely to
  be related, similar to premise selection in automated theorem proving
  (Kühlwein et al. 2012).
\end{itemize}

We are implementing a system called \textsc{ML4HS} to investigate these
ideas. Our initial hypothesis that expressions with similar definitions
are more likely to be related by equational properties than those
without, and hence a similarity-based clustering method such as that of
\textsc{ML4PG} (Heras and Komendantskaya 2013) can be used to implement
a divide and conquor pre-processor.

Since our aim is to scale up theory exploration, we treat entire
\textsc{Hackage} packages as our theories. \textsc{ML4HS} will manage
downloading, compiling, managing dependencies, etc. automatically. This
both eliminates the need to define theories manually, and may be useful
in its own right as a mechanism to execute arbitrary Haskell code from
arbitrary modules in arbitrary packages.

\section{Acknowlegements}\label{acknowlegements}

I am grateful for those who have helped formulate these ideas through
conversation, especially the HipSpec team at Chalmers University (Moa
Johansson, Koen Claessen, Nick Smallbone and Dan Rosén) and my
supervisor Katya Komendantskaya. I also wish to thank the implementors
of the systems we are building on, including HipSpec, QuickSpec and
QuickCheck on the theory exploration side, as well as GHC, Cabal and Nix
on the infrastructure side.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hyperdef{}{ref-blum2011hybrid}{\label{ref-blum2011hybrid}}
Blum, Christian, Jakob Puchinger, Günther R Raidl, and Andrea Roli.
2011. ``Hybrid Metaheuristics in Combinatorial Optimization: A Survey.''
\emph{Applied Soft Computing} 11 (6). Elsevier: 4135--51.

\hyperdef{}{ref-RISC1482}{\label{ref-RISC1482}}
Buchberger, Bruno. 1999. ``Theory Exploration Versus Theorem Proving.''
RISC Report Series 99-46. Schloss Hagenberg, 4232 Hagenberg, Austria:
Research Institute for Symbolic Computation (RISC), Johannes Kepler
University Linz.

\hyperdef{}{ref-buchberger2000theory}{\label{ref-buchberger2000theory}}
---------. 2000. ``Theory Exploration with Theorema.'' \emph{Analele
Universitatii Din Timisoara, Ser. Matematica-Informatica} 38 (2).
Citeseer: 9--32.

\hyperdef{}{ref-claessen2013automating}{\label{ref-claessen2013automating}}
Claessen, Koen, Moa Johansson, Dan Rosén, and Nicholas Smallbone. 2013.
``Automating Inductive Proofs Using Theory Exploration.'' In
\emph{Automated Deduction--CADE-24}, 392--406. Springer.

\hyperdef{}{ref-QuickSpec}{\label{ref-QuickSpec}}
Claessen, Koen, Nicholas Smallbone, and John Hughes. 2010. ``QuickSpec:
Guessing Formal Specifications Using Testing.'' In \emph{Tests and
Proofs}, edited by Gordon Fraser and Angelo Gargantini, 6143:6--21.
Lecture Notes in Computer Science. Springer Berlin Heidelberg.
\href{http://doi.org/10.1007/978-3-642-13977-2_3}{doi:10.1007/978-3-642-13977-2\_3}.

\hyperdef{}{ref-geng2006interestingness}{\label{ref-geng2006interestingness}}
Geng, Liqiang, and Howard J Hamilton. 2006. ``Interestingness Measures
for Data Mining: A Survey.'' \emph{ACM Computing Surveys (CSUR)} 38 (3).
ACM: 9.

\hyperdef{}{ref-journalsux2fcorrux2fabs-1302-6421}{\label{ref-journalsux2fcorrux2fabs-1302-6421}}
Heras, Jónathan, and Ekaterina Komendantskaya. 2013. ``ML4PG:
proof-mining in Coq.'' \emph{CoRR} abs/1302.6421.
\url{http://dblp.uni-trier.de/db/journals/corr/corr1302.html\#abs-1302-6421}.

\hyperdef{}{ref-johansson2009isacosy}{\label{ref-johansson2009isacosy}}
Johansson, Moa, Lucas Dixon, and Alan Bundy. 2009. ``Isacosy: Synthesis
of Inductive Theorems.'' In \emph{Workshop on Automated Mathematical
Theory Exploration (Automatheo)}.

\hyperdef{}{ref-Hipster}{\label{ref-Hipster}}
Johansson, Moa, Dan Rosén, Nicholas Smallbone, and Koen Claessen. 2014.
``Hipster: Integrating Theory Exploration in a Proof Assistant.'' In
\emph{Intelligent Computer Mathematics}, edited by StephenM. Watt,
JamesH. Davenport, AlanP. Sexton, Petr Sojka, and Josef Urban,
8543:108--22. Lecture Notes in Computer Science. Springer International
Publishing.
\href{http://doi.org/10.1007/978-3-319-08434-3_9}{doi:10.1007/978-3-319-08434-3\_9}.

\hyperdef{}{ref-kuhlwein2012overview}{\label{ref-kuhlwein2012overview}}
Kühlwein, Daniel, Twan van Laarhoven, Evgeni Tsivtsivadze, Josef Urban,
and Tom Heskes. 2012. ``Overview and Evaluation of Premise Selection
Techniques for Large Theory Mathematics.'' In \emph{Automated
Reasoning}, 378--92. Springer.

\hyperdef{}{ref-lindley2014hasochism}{\label{ref-lindley2014hasochism}}
Lindley, Sam, and Conor McBride. 2014. ``Hasochism: The Pleasure and
Pain of Dependently Typed Haskell Programming.'' \emph{ACM SIGPLAN
Notices} 48 (12). ACM: 81--92.

\hyperdef{}{ref-rosen2012proving}{\label{ref-rosen2012proving}}
Rosén, Dan. 2012. ``Proving Equational Haskell Properties Using
Automated Theorem Provers.'' PhD thesis, Master's thesis, University of
Gothenburg, Sweden.

\end{document}

### 2015-08-06 AI4FM/abstract.tex
Switch refs

@@
 
-\section{Abstract}\label{abstract}
+\begin{abstract}
 
@@
 
+\end{abstract}
+
 \section{Introduction}\label{introduction}
@@
 for traditional Mathematical workflows; namely, deriving ``interesting''
-consequences from formal definitions (Buchberger 1999). Early
-implementations like \textsc{Theorema} (Buchberger 2000) emphasised
+consequences from formal definitions \cite{RISC1482}. Early
+implementations like \textsc{Theorema} \cite{buchberger2000theory} emphasised
 interactivity, in a similar way to computer algebra systems like
@@
 \emph{automated} theory exploration, for tasks such as lemma discovery
-(Johansson et al. 2014).
+\cite{Hipster}.
 
@@
 this is intimately connected to the choice of search algorithm, to
-ensure tractability. For example, \textsc{IsaCoSy} (Johansson, Dixon,
-and Bundy 2009) discovers equations, which are defined as
+ensure tractability. For example, \textsc{IsaCoSy} \cite{johansson2009isacosy}
+discovers equations, which are defined as
 ``interesting'' if they cannot be simplified using previously discovered
@@
 these could be replaced by a variety of faster, approximate algorithms,
-eg. those surveyed in (Blum et al. 2011).
+eg. those surveyed in \cite{blum2011hybrid}.
 
@@
 ``interestingness'' measures are used in domains such as data mining
-(Geng and Hamilton 2006), which may be adapted to a TE setting. These
+\cite{geng2006interestingness}, which may be adapted to a TE setting. These
 include objective measures, eg. based on information theory, and
@@
 precision-recall experiments to evaluate the performance of
-\textsc{HipSpec} (Claessen et al. 2013).
+\textsc{HipSpec} \cite{claessen2013automating}.
 
@@
 non-trivial; and since these properties can't be expressed in Haskell
-itself (without difficulty (Lindley and McBride 2014), at least), less
+itself (without difficulty \cite{lindley2014hasochism}, at least), less
 effort is spent discovering and proving these properties compared to
@@
 benefit existing library authors and users in comprehending and
-maintaining their code (Claessen, Smallbone, and Hughes 2010).
+maintaining their code \cite{QuickSpec}.
 
@@
 relating the members of these classes are then conjectured, and sent to
-existing automated theorem provers to try and prove (Rosén 2012). This
+existing automated theorem provers to try and prove \cite{rosen2012proving}. This
 approach works well as a lemma generation system, making
 \textsc{HipSpec} a capable inductive theorem prover as well as a theory
-exploration system (Claessen et al. 2013).
+exploration system \cite{claessen2013automating}.
 
@@
   be related, similar to premise selection in automated theorem proving
-  (Kühlwein et al. 2012).
+  \cite{kuhlwein2012overview}.
 \end{itemize}
@@
 without, and hence a similarity-based clustering method such as that of
-\textsc{ML4PG} (Heras and Komendantskaya 2013) can be used to implement
+\textsc{ML4PG} \cite{journals/corr/abs-1302-6421} can be used to implement
 a divide and conquor pre-processor.
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\hyperdef{}{ref-blum2011hybrid}{\label{ref-blum2011hybrid}}
-Blum, Christian, Jakob Puchinger, Günther R Raidl, and Andrea Roli.
-2011. ``Hybrid Metaheuristics in Combinatorial Optimization: A Survey.''
-\emph{Applied Soft Computing} 11 (6). Elsevier: 4135--51.
-
-\hyperdef{}{ref-RISC1482}{\label{ref-RISC1482}}
-Buchberger, Bruno. 1999. ``Theory Exploration Versus Theorem Proving.''
-RISC Report Series 99-46. Schloss Hagenberg, 4232 Hagenberg, Austria:
-Research Institute for Symbolic Computation (RISC), Johannes Kepler
-University Linz.
-
-\hyperdef{}{ref-buchberger2000theory}{\label{ref-buchberger2000theory}}
----------. 2000. ``Theory Exploration with Theorema.'' \emph{Analele
-Universitatii Din Timisoara, Ser. Matematica-Informatica} 38 (2).
-Citeseer: 9--32.
-
-\hyperdef{}{ref-claessen2013automating}{\label{ref-claessen2013automating}}
-Claessen, Koen, Moa Johansson, Dan Rosén, and Nicholas Smallbone. 2013.
-``Automating Inductive Proofs Using Theory Exploration.'' In
-\emph{Automated Deduction--CADE-24}, 392--406. Springer.
-
-\hyperdef{}{ref-QuickSpec}{\label{ref-QuickSpec}}
-Claessen, Koen, Nicholas Smallbone, and John Hughes. 2010. ``QuickSpec:
-Guessing Formal Specifications Using Testing.'' In \emph{Tests and
-Proofs}, edited by Gordon Fraser and Angelo Gargantini, 6143:6--21.
-Lecture Notes in Computer Science. Springer Berlin Heidelberg.
-\href{http://doi.org/10.1007/978-3-642-13977-2_3}{doi:10.1007/978-3-642-13977-2\_3}.
-
-\hyperdef{}{ref-geng2006interestingness}{\label{ref-geng2006interestingness}}
-Geng, Liqiang, and Howard J Hamilton. 2006. ``Interestingness Measures
-for Data Mining: A Survey.'' \emph{ACM Computing Surveys (CSUR)} 38 (3).
-ACM: 9.
-
-\hyperdef{}{ref-journalsux2fcorrux2fabs-1302-6421}{\label{ref-journalsux2fcorrux2fabs-1302-6421}}
-Heras, Jónathan, and Ekaterina Komendantskaya. 2013. ``ML4PG:
-proof-mining in Coq.'' \emph{CoRR} abs/1302.6421.
-\url{http://dblp.uni-trier.de/db/journals/corr/corr1302.html\#abs-1302-6421}.
-
-\hyperdef{}{ref-johansson2009isacosy}{\label{ref-johansson2009isacosy}}
-Johansson, Moa, Lucas Dixon, and Alan Bundy. 2009. ``Isacosy: Synthesis
-of Inductive Theorems.'' In \emph{Workshop on Automated Mathematical
-Theory Exploration (Automatheo)}.
-
-\hyperdef{}{ref-Hipster}{\label{ref-Hipster}}
-Johansson, Moa, Dan Rosén, Nicholas Smallbone, and Koen Claessen. 2014.
-``Hipster: Integrating Theory Exploration in a Proof Assistant.'' In
-\emph{Intelligent Computer Mathematics}, edited by StephenM. Watt,
-JamesH. Davenport, AlanP. Sexton, Petr Sojka, and Josef Urban,
-8543:108--22. Lecture Notes in Computer Science. Springer International
-Publishing.
-\href{http://doi.org/10.1007/978-3-319-08434-3_9}{doi:10.1007/978-3-319-08434-3\_9}.
-
-\hyperdef{}{ref-kuhlwein2012overview}{\label{ref-kuhlwein2012overview}}
-Kühlwein, Daniel, Twan van Laarhoven, Evgeni Tsivtsivadze, Josef Urban,
-and Tom Heskes. 2012. ``Overview and Evaluation of Premise Selection
-Techniques for Large Theory Mathematics.'' In \emph{Automated
-Reasoning}, 378--92. Springer.
-
-\hyperdef{}{ref-lindley2014hasochism}{\label{ref-lindley2014hasochism}}
-Lindley, Sam, and Conor McBride. 2014. ``Hasochism: The Pleasure and
-Pain of Dependently Typed Haskell Programming.'' \emph{ACM SIGPLAN
-Notices} 48 (12). ACM: 81--92.
-
-\hyperdef{}{ref-rosen2012proving}{\label{ref-rosen2012proving}}
-Rosén, Dan. 2012. ``Proving Equational Haskell Properties Using
-Automated Theorem Provers.'' PhD thesis, Master's thesis, University of
-Gothenburg, Sweden.
+\bibliographystyle{plain} \bibliography{/home/chris/Documents/ArchivedPapers/Bibtex}
 

### 2015-08-06 AI4FM/render.sh

@@ -11,6 +11,7 @@ function mdToLatex {
 
 function renderLatex {
     latex abstract
+    bibtex abstract
     latex abstract
     pdflatex abstract
 }

### 2015-08-06 AI4FM/abstract.tex
Added hardlink to Bibtex.bib

@@
 and Haskell, although we focus on the latter as its implementations are
-the most mature (\textsc{Hipster} actually explores Isabelle by
-translating code to Haskell first!). Haskell is interesting to target,
-since its functional purity and algebraic structure make equational
-properties common; recursion and higher-order functions make automation
-non-trivial; and since these properties can't be expressed in Haskell
-itself (without difficulty \cite{lindley2014hasochism}, at least), less
+the most mature (demonstrated by \textsc{Hipster}, which explores Isabelle by
+translating code to Haskell first). Haskell is interesting to target,
+since its emphasis on pure functions and algebraic datatypes make it common for
+functions to obey equational properties; although Haskell is not a theorem prover,
+and hence cannot represent such equations . Also, the widespread use of
+recursion and higher-order functions makes automated deduction non-trivial
 effort is spent discovering and proving these properties compared to
@@
 
-\bibliographystyle{plain} \bibliography{/home/chris/Documents/ArchivedPapers/Bibtex}
+\bibliographystyle{plain} \bibliography{../Bibtex}
 

### 2015-08-06 AI4FM/render.sh

@@ -16,6 +16,16 @@ function renderLatex {
     pdflatex abstract
 }
 
+function clean {
+    rm abstract.aux
+    rm abstract.blg
+    rm abstract.log
+    rm abstract.pdf
+    rm abstract.bbl
+    rm abstract.dvi
+    rm abstract.out
+}
+
 function go {
     renderLatex
 }

### 2015-08-06 AI4FM/abstract.tex
Edits

@@
-\documentclass[]{article}
+\documentclass[]{eceasst}
 \usepackage{lmodern}
@@
 }{}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
+% \ifxetex
+%   \usepackage[setpagesize=false, % page size defined by xetex
+%               unicode=false, % unicode breaks when used with xetex
+%               xetex]{hyperref}
+% \else
+%   \usepackage[unicode=true]{hyperref}
+% \fi
 \hypersetup{breaklinks=true,
@@
 By removing user interaction, automated TE systems require an algorithm
-for deciding whether a theorem is ``interesting''. In existing systems,
-this is intimately connected to the choice of search algorithm, to
-ensure tractability. For example, \textsc{IsaCoSy} \cite{johansson2009isacosy}
-discovers equations, which are defined as
-``interesting'' if they cannot be simplified using previously discovered
-equations. The intuition for such criteria is to avoid special cases of
-known theorems, such as \(0 + 0 = 0\), \(0 + 1 = 1\), etc. when we
-already know \(\forall x. 0 + x = x\). To work effectively, this
+for deciding whether to inform the user of a particular theorem or not; we refer
+to this search criterion as the theorem being ``interesting''.
+
+In existing systems, this criterion is intimately connected to the
+search algorithm; this coupling improves efficiency, even for brute-force
+search. As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
+equations, which are defined as ``interesting'' if they cannot be simplified
+using previously discovered equations. The intuition for such criteria is to
+avoid special cases of known theorems, such as \(0 + 0 = 0\), \(0 + 1 = 1\),
+etc. when we already know \(\forall x. 0 + x = x\). To work effectively, this
 requires general forms to be found \emph{before} special cases.
 
-TE is a ``bottom-up'' process: new theorems are derived from given
-knowledge and previous theorems. In particular, there is no special
-``goal'' theorem(s) as found in interactive theorem proving; all we have
-is our (proxy) ``interesting'' criterion, and the definitions of our
-theory. This makes theory exploration a \emph{combinatorial
-optimisation} problem, which is well studied in Artificial Intelligence
-and Machine Learning. Existing TE systems rely on complete, brute-force
-search techniques which do not scale to theories of a significant size;
-these could be replaced by a variety of faster, approximate algorithms,
-eg. those surveyed in \cite{blum2011hybrid}.
-
-Whilst efficient on a small scale, where we expect the chosen inputs to
-be related, we cannot expect an effective output criterion to coincide
-so readily with search algorithm and output criterion to be effective
-for large inputs containing many irrelevant connections. umbers of also
-consider data mining approaches, which infer general rules \emph{after}
-enumerating some number of special cases. A variety of
-``interestingness'' measures are used in domains such as data mining
-\cite{geng2006interestingness}, which may be adapted to a TE setting. These
-include objective measures, eg. based on information theory, and
-subjective measures which ask for the user's opinion. We can avoid
-asking users directly, by treating existing proof libraries and test
-suites as representative examples of what is interesting, and hence
-construct an oracle; such oracles have already been used for
-precision-recall experiments to evaluate the performance of
-\textsc{HipSpec} \cite{claessen2013automating}.
+More generally, automated TE is an \emph{optimisation problem}; an area which
+has been studied extensively in the fields of Artificial Intelligence and
+Machine Learning. For a more thorough treatment, \cite{geng2006interestingness}
+surveys recent work on search algorithms and \cite{blum2011hybrid} surveys
+interestingness criteria in the context of data mining.
 
@@
 translating code to Haskell first). Haskell is interesting to target,
-since its emphasis on pure functions and algebraic datatypes make it common for
-functions to obey equational properties; although Haskell is not a theorem prover,
-and hence cannot represent such equations . Also, the widespread use of
+since its emphasis on pure functions and algebraic datatypes make equational
+properties common; however, Haskell is not a theorem prover, and hence it cannot
+represent such as universally quantified equations. Also, the widespread use of
 recursion and higher-order functions makes automated deduction non-trivial

### 2015-08-06 AI4FM/abstract.tex
Use template

@@
-\documentclass[]{eceasst}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-% \ifxetex
-%   \usepackage[setpagesize=false, % page size defined by xetex
-%               unicode=false, % unicode breaks when used with xetex
-%               xetex]{hyperref}
-% \else
-%   \usepackage[unicode=true]{hyperref}
-% \fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={Chris Warburton},
-            pdftitle={Scaling Automated Theory Exploration},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\documentclass{eceasst}
+% This is an empty ECEASST article that can be used as a template
+% by authors.
+% Just uncomment the appropriate frontmatter commands and provide
+% the parameters.
+
+% Required packages
+% =================
+% Your \usepackage commands go here.
+
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
 
-\title{Scaling Automated Theory Exploration}
-\author{Chris Warburton}
-\date{}
-\usepackage[T1]{fontenc}
-\usepackage{upquote}
+% Volume frontmatter
+% ==================
+\input{frontmatter}
 
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+% Article frontmatter
+% ===================
+\title{Scaling Automated Theory Exploration} % Title of the article
+%\short{} % Short title of the article (optional)
+\author{Chris Warburton\sponsor{}} % Authors and references to addresses
+\institute{ % Institutes with labels
+\email{cmwarburton@dundee.ac.uk}, \url{http://www.computing.dundee.ac.uk/about/staff/124}\\
+School of Computing\\
+University of Dundee}
+\abstract{
+We investigate the \textbf{theory exploration} (TE)
+paradigm for computer-assisted Mathematics and identify limitations and
+improvements for current approaches. Unlike the theorem-proving paradigm,
+which requires user-provided conjectures, TE performs an open-ended
+search for theorems satisfying given criteria. We see promise in TE for
+identifying new abstractions and connections in libraries of software
+and proofs, but realising this potential requires more scalable
+algorithms than presently used.} % Abstract of the article
+%\keywords{} % Keywords for the article
 
@@
 
-\begin{abstract}
-
-In this paper we investigate the \textbf{theory exploration} (TE)
-paradigm for computer-assisted Mathematics and identify limitations and
-improvements of current approaches. Unlike the theorem-proving paradigm,
-which requires user-provided conjectures, TE performs an open-ended
-search for theorems satisfying given criteria. We see promise in TE for
-identifying new abstractions and connections in libraries of software
-and proofs, but realising this potential requires more scalable
-algorithms than presently used.
-
-\end{abstract}
-
-\section{Introduction}\label{introduction}
+% Main part of your article
+% =========================
+\section{Introduction}
 
@@
 
-\section{Tackling Complexity}\label{tackling-complexity}
+\section{Tackling Complexity}
 
@@
 
-\section{Existing Work}\label{existing-work}
+\section{Existing Work}
 
@@
 
-\section{Going Forward}\label{going-forward}
+\section{Going Forward}
 
@@
 
-\section{Acknowlegements}\label{acknowlegements}
-
+% Acknowledgements for colleagues, referees, ...
+% ==============================================
+\begin{acknowledge}
 I am grateful for those who have helped formulate these ideas through
 conversation, especially the HipSpec team at Chalmers University (Moa
-Johansson, Koen Claessen, Nick Smallbone and Dan Rosén) and my
-supervisor Katya Komendantskaya. I also wish to thank the implementors
+Johansson, Koen Claessen, Nick Smallbone, Dan Rosén and Irene Lobo Valbuena)
+and my supervisor Katya Komendantskaya. I also wish to thank the implementors
 of the systems we are building on, including HipSpec, QuickSpec and
@@
 on the infrastructure side.
+\end{acknowledge}
 
-\bibliographystyle{plain} \bibliography{../Bibtex}
+% Bibliography with BibTeX
+% ========================
+\bibliographystyle{eceasst}
+\bibliography{../Bibtex}
 

### 2015-08-06 AI4FM/article.tex
Fixed underscores

--- 2015-08-06 AI4FM/abstract.tex
@@
 % Your \usepackage commands go here.
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \institute{ % Institutes with labels
-\email{cmwarburton@dundee.ac.uk}, \url{http://www.computing.dundee.ac.uk/about/staff/124}\\
+\email{cmwarburton@dundee.ac.uk}\\
 School of Computing\\
@@
 using previously discovered equations. The intuition for such criteria is to
-avoid special cases of known theorems, such as \(0 + 0 = 0\), \(0 + 1 = 1\),
-etc. when we already know \(\forall x. 0 + x = x\). To work effectively, this
+avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$,
+etc. when we already know $\forall x. 0 + x = x$. To work effectively, this
 requires general forms to be found \emph{before} special cases.
@@
 
-\section{Tackling Complexity}
+In the following, we give an overview of the state of the art in automated
+theory exploration; identify potential improvements; and present our initial
+attempts to implement some of these ideas.
 
-Even with a slow search algorithm, we can use a divide and conquor
-approach to limit the number of allowed combinations, either by using
-stricter types to prevent composition, or by partitioning the theory
-into small independently-searched sub-theories. Of course, such
-restrictions should strike a balance between the efficiency gained and
-the potential to forbid some interesting theorems.
-
-\section{Existing Work}
+\section{Theory Exploration in Haskell}
 
@@
 
-\section{Going Forward}
+\section{Identifying Improvements}
+
+Even with a slow search algorithm, we can use a divide and conquor
+approach to limit the number of allowed combinations, either by using
+stricter types to prevent composition, or by partitioning the theory
+into small independently-searched sub-theories. Of course, such
+restrictions should strike a balance between the efficiency gained and
+the potential to forbid some interesting theorems.
 
@@
 \begin{itemize}
-\tightlist
 \item
@@
   \begin{itemize}
-  \tightlist
   \item
@@
 
+\section{The \textsc{ML4HS} Framework}
+
 We are implementing a system called \textsc{ML4HS} to investigate these
@@
 conversation, especially the HipSpec team at Chalmers University (Moa
-Johansson, Koen Claessen, Nick Smallbone, Dan Rosén and Irene Lobo Valbuena)
+Johansson, Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena)
 and my supervisor Katya Komendantskaya. I also wish to thank the implementors

### 2015-08-06 AI4FM/render.sh

@@ -6,24 +6,24 @@ function renderMd {
 }
 
 function mdToLatex {
-    pandoc -s --filter pandoc-citeproc -o abstract.tex abstract.md
+    pandoc -s --filter pandoc-citeproc -o article.tex article.md
 }
 
 function renderLatex {
-    latex abstract
-    bibtex abstract
-    latex abstract
-    pdflatex abstract
+    latex article
+    bibtex article
+    latex article
+    pdflatex article
 }
 
 function clean {
-    rm abstract.aux
-    rm abstract.blg
-    rm abstract.log
-    rm abstract.pdf
-    rm abstract.bbl
-    rm abstract.dvi
-    rm abstract.out
+    rm article.aux
+    rm article.blg
+    rm article.log
+    rm article.pdf
+    rm article.bbl
+    rm article.dvi
+    rm article.out
 }
 
 function go {

### 2015-08-06 AI4FM/article.tex
Improvements

@@
 
-More generally, automated TE is an \emph{optimisation problem}; an area which
-has been studied extensively in the fields of Artificial Intelligence and
-Machine Learning. For a more thorough treatment, \cite{geng2006interestingness}
-surveys recent work on search algorithms and \cite{blum2011hybrid} surveys
-interestingness criteria in the context of data mining.
-
 In the following, we give an overview of the state of the art in automated
@@
 and Haskell, although we focus on the latter as its implementations are
-the most mature (demonstrated by \textsc{Hipster}, which explores Isabelle by
-translating code to Haskell first). Haskell is interesting to target,
-since its emphasis on pure functions and algebraic datatypes make equational
-properties common; however, Haskell is not a theorem prover, and hence it cannot
-represent such as universally quantified equations. Also, the widespread use of
-recursion and higher-order functions makes automated deduction non-trivial
-effort is spent discovering and proving these properties compared to
-proof-oriented systems like Isabelle.
+the most mature (demonstrated by the fact that \textsc{Hipster} explores
+Isabelle by first translating it to Haskell). Haskell is interesting to target,
+since its use of pure functions and algebraic datatypes causes many functions to
+follow algebraic laws; however, since Haskell's type system cannot encode such
+laws (without difficulty \cite{lindley2014hasochism}, at least), less
+effort is given to finding and stating them; compared to a theorem proving
+setting like Isabelle, where these laws are often the main focus.
 
-Due to Haskell's relative popularity, there are large code repositories
-such as \textsc{Hackage} available to explore, with the potential to
-benefit existing library authors and users in comprehending and
-maintaining their code \cite{QuickSpec}.
+Due to Haskell's relative popularity, there are also large code repositories
+such as \textsc{Hackage} available to explore. This gives several potential
+benefits: in the short term, existing library authors and users may be assisted
+in comprehending and maintaining their code \cite{QuickSpec}; more notably, TE
+could be used to data mine new abstractions, such as typeclasses, with relevance
+to real-world code.
 
@@
 
-\begin{itemize}
-\item
-  Expression enumeration is brute-force; this could scale to larger
-  terms and theories using a heuristic algorithm.
-\item
-  ``Interestingness'' is a fixed part of the algorithm: an equation is
-  interesting if it cannot be derived from previous equations. As we
-  increase the size of our theory, this becomes unsatisfying in two
-  ways:
+For a more thorough treatment,
+surveys recent work on search algorithms and  surveys
+interestingness criteria in the context of data mining.
 
@@
 \item
-    The number of irreducible equations grows, making it desirable to
+  Enumerating expressions is a brute-force approach. Identifying more scalable
+  alternatives to brute-force search is a well-studied area of Artificial
+  Intelligence and Machine Learning. In particular, there is potential in
+  applying heuristic algorithms like those surveyed in \cite{blum2011hybrid}.
+\item
+  Many ``interestingness'' criteria are surveyed in
+  \cite{geng2006interestingness}, but existing systems are tightly coupled to
+  one particular notion. Whilst pre- and post-processors could provide
+  additional filtering, there is merit in building a more loosely-coupled
+  alternative, for two reasons:
+
+  \begin{itemize}
+  \item
+    As the number of irreducible equations grows, it may be desirable to
     impose extra conditions of a more subjective nature.
   \item
-    Surprising, insightful equations may be discarded, if they are
-    actually reducible in some complex, non-obvious way. A more
+    Surprising, insightful equations may be discarded by current systems, if
+    they are actually reducible in some complex, non-obvious way. A more
     subjective interestingness measure could be used to veto such
@@
 \item
-  The system does not propose candidate equations by data mining
+  HipSpec does not propose candidate equations by data mining
   previous results; generalisation methods like anti-unification could
@@
   considering (either because they are never related, or because their
-  relations are never interesting). A pre-processor could make large
-  theories more tractable by selecting combinations which are likely to
-  be related, similar to premise selection in automated theorem proving
-  \cite{kuhlwein2012overview}.
+  relations are never interesting).
 \end{itemize}
+
+We consider all of these worth investigating, although in this paper we
+focus on the last item.
 
@@
 
-We are implementing a system called \textsc{ML4HS} to investigate these
-ideas. Our initial hypothesis that expressions with similar definitions
-are more likely to be related by equational properties than those
-without, and hence a similarity-based clustering method such as that of
-\textsc{ML4PG} \cite{journals/corr/abs-1302-6421} can be used to implement
-a divide and conquor pre-processor.
+We are implementing a system called \textsc{ML4HS} to investigate scalable
+automated theory exploration for Haskell. At this stage, we provide a
+pre-processor for \textsc{QuickSpec} which attempts to make exploration of large
+theories more tractable by splitting them into a collection of smaller theories.
 
-Since our aim is to scale up theory exploration, we treat entire
-\textsc{Hackage} packages as our theories. \textsc{ML4HS} will manage
-downloading, compiling, managing dependencies, etc. automatically. This
-both eliminates the need to define theories manually, and may be useful
-in its own right as a mechanism to execute arbitrary Haskell code from
-arbitrary modules in arbitrary packages.
+Inspired by premise selection in automated theorem proving
+\cite{kuhlwein2012overview}, and similarity-based clustering method such as that
+of \textsc{ML4PG} \cite{journals/corr/abs-1302-6421}, we are investigating the
+hypothesis that similar expressions are more likely to be related by equational
+properties than dissimilar expressions. This has lead to a divide and conquor
+approach, shown in figure 1.
+
+Since our aim is to scale up theory exploration, our pre-processor also
+increases the level of automation compared to existing systems. We consider
+\textsc{Hackage} packages to be theories, and treat package management,
+downloading, compiling, dependency resolution, etc. as necessary tasks for a TE
+system such as \textsc{ML4HS} to implement.
+
+Not only does this eliminate the need to define theories manually, as a sub-set
+of available expressions, but may also be useful in its own right as a mechanism
+to execute arbitrary Haskell code from arbitrary modules in arbitrary packages.
 

### 2015-08-06 AI4FM/article.tex
Added diagram

@@
 for deciding whether to inform the user of a particular theorem or not; we refer
-to this search criterion as the theorem being ``interesting''.
+to this search criterion as the theorem being ``interesting''. In existing
+systems, this criterion is intimately connected to the
+search algorithm; such coupling improves efficiency, even for brute-force
+search, but makes it difficult to try alternative measures.
 
-In existing systems, this criterion is intimately connected to the
-search algorithm; this coupling improves efficiency, even for brute-force
-search. As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
+As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
 equations, which are defined as ``interesting'' if they cannot be simplified
@@
 avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$,
-etc. when we already know $\forall x. 0 + x = x$. To work effectively, this
-requires general forms to be found \emph{before} special cases.
+etc. when we already know $\forall x. 0 + x = x$. Whilst elegantly implemented
+as a term rewriting system, to work effectively this requires general forms to
+be found \emph{before} special cases.
 
 In the following, we give an overview of the state of the art in automated
-theory exploration; identify potential improvements; and present our initial
-attempts to implement some of these ideas.
+theory exploration (\autoref{haskell}). We identify potential improvements to
+these systems (\autoref{improve}) and present our initial implementation
+attempts in the form of the \textsc{ML4HS} system (\autoref{ml4hs}).
 
-\section{Theory Exploration in Haskell}
+\section{Theory Exploration in Haskell}\label{haskell}
 
@@
 since its use of pure functions and algebraic datatypes causes many functions to
-follow algebraic laws; however, since Haskell's type system cannot encode such
-laws (without difficulty \cite{lindley2014hasochism}, at least), less
-effort is given to finding and stating them; compared to a theorem proving
-setting like Isabelle, where these laws are often the main focus.
+follow algebraic laws. However, since Haskell's type system cannot encode such
+laws (without difficulty \cite{lindley2014hasochism}, at least); less
+effort is given to finding and stating them, compared to a theorem proving
+setting like Isabelle. Hence, Haskell provides much low-hanging fruit for a
+broad but shallow exploration to discover.
 
@@
 such as \textsc{Hackage} available to explore. This gives several potential
-benefits: in the short term, existing library authors and users may be assisted
-in comprehending and maintaining their code \cite{QuickSpec}; more notably, TE
-could be used to data mine new abstractions, such as typeclasses, with relevance
-to real-world code.
+benefits: in the short term, applying TE to such repositories may assist
+existing library authors and users in comprehending and maintaining their code
+\cite{QuickSpec}; more notably, TE could be used to data mine new abstractions,
+such as typeclasses, with relevance to real-world applications.
 
 Currently, the most powerful TE system for Haskell is \textsc{HipSpec}.
-This uses \textsc{QuickSpec} to search through \emph{expressions}
-(combinations of the Haskell terms given by the theory), rather than
-searching through the space of equations or proofs directly. Expressions
-are grouped into equivalence classes, such that the \textsc{QuickCheck}
-counterexample finder cannot distinguish between the elements; equations
-relating the members of these classes are then conjectured, and sent to
-existing automated theorem provers to try and prove \cite{rosen2012proving}. This
-approach works well as a lemma generation system, making
-\textsc{HipSpec} a capable inductive theorem prover as well as a theory
+This uses the earlier \textsc{QuickSpec} system to search through
+\emph{expressions} (combinations of the Haskell terms given by the theory),
+rather than searching through the space of equations or proofs directly.
+Expressions are grouped into equivalence classes, such that the
+\textsc{QuickCheck} counterexample finder cannot distinguish between the
+elements; equations relating the members of these classes are then conjectured,
+and sent to existing automated theorem provers to try and prove
+\cite{rosen2012proving}. This approach works well as a lemma generation system,
+making \textsc{HipSpec} a capable inductive theorem prover as well as a theory
 exploration system \cite{claessen2013automating}.
 
-\section{Identifying Improvements}
-
-Even with a slow search algorithm, we can use a divide and conquor
-approach to limit the number of allowed combinations, either by using
-stricter types to prevent composition, or by partitioning the theory
-into small independently-searched sub-theories. Of course, such
-restrictions should strike a balance between the efficiency gained and
-the potential to forbid some interesting theorems.
+\section{Identifying Improvements}\label{improve}
 
 Given this state of the art, we identify the following as potential
-areas for improvement:
-
-For a more thorough treatment,
-surveys recent work on search algorithms and  surveys
-interestingness criteria in the context of data mining.
+areas for investigation:
 
@@
 \item
-  Enumerating expressions is a brute-force approach. Identifying more scalable
-  alternatives to brute-force search is a well-studied area of Artificial
-  Intelligence and Machine Learning. In particular, there is potential in
-  applying heuristic algorithms like those surveyed in \cite{blum2011hybrid}.
+  More scalable alternatives to brute-force enumeration. This is a well-studied
+  area of Artificial Intelligence and Machine Learning. In particular, there is
+  potential in applying heuristic algorithms like those surveyed in
+  \cite{blum2011hybrid}.
 \item
-  Many ``interestingness'' criteria are surveyed in
-  \cite{geng2006interestingness}, but existing systems are tightly coupled to
-  one particular notion. Whilst pre- and post-processors could provide
-  additional filtering, there is merit in building a more loosely-coupled
-  alternative, for two reasons:
+  Using alternative ``interestingness'' criteria, like those surveyed in
+  \cite{geng2006interestingness}. In particular, we envision two problems with
+  large-scale exploration which may require more sophisticated criteria:
 
@@
     As the number of irreducible equations grows, it may be desirable to
-    impose extra conditions of a more subjective nature.
+    impose extra conditions of a more subjective nature; e.g. to distinguish
+    useful relationships from incidental coincidences.
   \item
@@
 \item
-  HipSpec does not propose candidate equations by data mining
-  previous results; generalisation methods like anti-unification could
-  do this, and at the same time remove the requirement that general
-  forms must be enumerated early.
+  Data mining previous results; for example, using generalisation methods like
+  anti-unification \cite{bulychev2010anti}. This would make more efficient use
+  of data, and also remove the requirement that general forms be enumerated
+  early.
 \item
-  All type-safe combinations of the given expressions are tried, whilst
-  it may be discernable a priori that some combinations are not worth
-  considering (either because they are never related, or because their
-  relations are never interesting).
+  Divide and conquer approaches. Rather than try all type-safe combinations of
+  the given expressions, it may be discernable a priori that some combinations
+  are not worth considering (either because they are never related, or because
+  their relations are never interesting). In other words, automating the theory
+  selection process currently performed by the user.
 \end{itemize}
@@
 
-\section{The \textsc{ML4HS} Framework}
+\section{The \textsc{ML4HS} Framework}\label{ml4hs}
+
+\begin{figure}
+  \centering
+  \includegraphics[]{ml4hs}
+  \caption{An overview of the \textsc{ML4HS} system}
+  \label{ml4hsdia}
+\end{figure}
 
 We are implementing a system called \textsc{ML4HS} to investigate scalable
-automated theory exploration for Haskell. At this stage, we provide a
-pre-processor for \textsc{QuickSpec} which attempts to make exploration of large
-theories more tractable by splitting them into a collection of smaller theories.
+automated theory exploration for Haskell. The current form of \textsc{ML4HS} is
+a pre-processor for \textsc{QuickSpec} which attempts to make exploration of
+large theories more tractable by splitting them into a collection of smaller
+theories: the divide and conquer approach mentioned above. This is
+similar to reducing the search space in automated theorem proving via premise
+selection \cite{kuhlwein2012overview}. An overview of \textsc{ML4HS} is given in
+\autoref{ml4hsdia}. We have integrated our pre-processor with the \textsc{Cabal} and
+\textsc{Nix} package managers so that \textsc{ML4HS} can use regular Haskell
+packages as theories.
 
-Inspired by premise selection in automated theorem proving
-\cite{kuhlwein2012overview}, and similarity-based clustering method such as that
-of \textsc{ML4PG} \cite{journals/corr/abs-1302-6421}, we are investigating the
-hypothesis that similar expressions are more likely to be related by equational
-properties than dissimilar expressions. This has lead to a divide and conquor
-approach, shown in figure 1.
-
-Since our aim is to scale up theory exploration, our pre-processor also
-increases the level of automation compared to existing systems. We consider
-\textsc{Hackage} packages to be theories, and treat package management,
-downloading, compiling, dependency resolution, etc. as necessary tasks for a TE
-system such as \textsc{ML4HS} to implement.
-
-Not only does this eliminate the need to define theories manually, as a sub-set
-of available expressions, but may also be useful in its own right as a mechanism
-to execute arbitrary Haskell code from arbitrary modules in arbitrary packages.
+When selecting the components of each sub-theory, we want to keep together those
+expressions which combine to form interesting theorems, and separate those which
+combine in uninteresting ways. Our hypothesis is that similarity-based
+clustering of expressions, as found in \textsc{ML4PG}
+\cite{journals/corr/abs-1302-6421}, is an effective method for performing this
+division. Future experiments will test this by measuring throughput rates of
+\textsc{ML4HS} compared to \textsc{QuickSpec} and \textsc{HipSpec}.
 
@@
 \begin{acknowledge}
-I am grateful for those who have helped formulate these ideas through
-conversation, especially the HipSpec team at Chalmers University (Moa
-Johansson, Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena)
-and my supervisor Katya Komendantskaya. I also wish to thank the implementors
-of the systems we are building on, including HipSpec, QuickSpec and
-QuickCheck on the theory exploration side, as well as GHC, Cabal and Nix
-on the infrastructure side.
+Thank you to my supervisor Katya Komendantskaya and members of the
+\textsc{HipSpec} team at Chalmers University (Moa Johansson, Koen Claessen, Nick
+Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful discussions of these
+ideas.
 \end{acknowledge}

### 2015-08-06 AI4FM/article.tex
Reworded

@@
 Isabelle by first translating it to Haskell). Haskell is interesting to target,
-since its use of pure functions and algebraic datatypes causes many functions to
-follow algebraic laws. However, since Haskell's type system cannot encode such
-laws (without difficulty \cite{lindley2014hasochism}, at least); less
-effort is given to finding and stating them, compared to a theorem proving
-setting like Isabelle. Hence, Haskell provides much low-hanging fruit for a
-broad but shallow exploration to discover.
+since its use of pure functions and algebraic datatypes causes many programs to
+follow algebraic laws. However, since Haskell's type system cannot easily
+encode such laws, less effort is given to finding and stating them; compared to
+a theorem proving setting like Isabelle.
 

### 2015-08-06 AI4FM/article.tex
Tweaks

@@
 consequences from formal definitions \cite{RISC1482}. Early
-implementations like \textsc{Theorema} \cite{buchberger2000theory} emphasised
-interactivity, in a similar way to computer algebra systems like
+implementations like \textsc{Theorema} \cite{buchberger2000theory} provided
+interactive environments, similar to computer algebra systems like
 \textsc{Mathematica} (in which \textsc{Theorema} is implemented) or
@@
 similar to reducing the search space in automated theorem proving via premise
-selection \cite{kuhlwein2012overview}. An overview of \textsc{ML4HS} is given in
-\autoref{ml4hsdia}. We have integrated our pre-processor with the \textsc{Cabal} and
-\textsc{Nix} package managers so that \textsc{ML4HS} can use regular Haskell
-packages as theories.
+selection \cite{kuhlwein2012overview}. The basic idea of \textsc{ML4HS} is shown
+in \autoref{ml4hsdia}: theories, in the form of Haskell packages, are split up
+and sent to concurrent instances of QuickSpec, and the results collated.
 

### 2015-08-07 AI4FM/article.tex
Preparing to change

@@
 \begin{acknowledge}
-Thank you to my supervisor Katya Komendantskaya and members of the
-\textsc{HipSpec} team at Chalmers University (Moa Johansson, Koen Claessen, Nick
-Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful discussions of these
-ideas.
+Thank you to the \textsc{HipSpec} team at Chalmers University (Moa Johansson,
+Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful
+discussions of these ideas.
 \end{acknowledge}

### 2015-08-07 AI4FM/article.tex
Use LLNCS style

@@
-\documentclass{eceasst}
-% This is an empty ECEASST article that can be used as a template
-% by authors.
-% Just uncomment the appropriate frontmatter commands and provide
-% the parameters.
+% This is LLNCS.DEM the demonstration file of
+% the LaTeX macro package from Springer-Verlag
+% for Lecture Notes in Computer Science,
+% version 2.4 for LaTeX2e as of 16. April 2010
+%
+\documentclass{llncs}
+%
+\usepackage{makeidx}  % allows for indexgeneration
+\usepackage{amsmath,hyperref}
+\usepackage{graphicx}
+\hypersetup{colorlinks=true}
+%
+\begin{document}
+%
+%\frontmatter          % for the preliminaries
+%
+\pagestyle{headings}  % switches on printing of running heads
+%\addtocmark{Hamiltonian Mechanics} % additional mark in the TOC
+%
 
-% Required packages
-% =================
-% Your \usepackage commands go here.
+% %
+% This textbook is intended for use by students of physics, physical
+% chemistry, and theoretical chemistry. The reader is presumed to have a
+% basic knowledge of atomic and quantum physics at the level provided, for
+% example, by the first few chapters in our book {\it The Physics of Atoms
+% and Quanta}. The student of physics will find here material which should
+% be included in the basic education of every physicist. This book should
+% furthermore allow students to acquire an appreciation of the breadth and
+% variety within the field of molecular physics and its future as a
+% fascinating area of research.
 
-% Volume frontmatter
-% ==================
-\input{frontmatter}
+% For the student of chemistry, the concepts introduced in this book will
+% provide a theoretical framework for that entire field of study. With the
+% help of these concepts, it is at least in principle possible to reduce
+% the enormous body of empirical chemical knowledge to a few basic
+% principles: those of quantum mechanics. In addition, modern physical
+% methods whose fundamentals are introduced here are becoming increasingly
+% important in chemistry and now represent indispensable tools for the
+% chemist. As examples, we might mention the structural analysis of
+% complex organic compounds, spectroscopic investigation of very rapid
+% reaction processes or, as a practical application, the remote detection
+% of pollutants in the air.
 
-% Article frontmatter
-% ===================
-\title{Scaling Automated Theory Exploration} % Title of the article
-%\short{} % Short title of the article (optional)
-\author{Chris Warburton\sponsor{}} % Authors and references to addresses
-\institute{ % Institutes with labels
-\email{cmwarburton@dundee.ac.uk}\\
-School of Computing\\
-University of Dundee}
-\abstract{
+% \vspace{1cm}
+% \begin{flushright}\noindent
+% April 1995\hfill Walter Olthoff\\
+% Program Chair\\
+% ECOOP'95
+% \end{flushright}
+% %
+% \chapter*{Organization}
+% ECOOP'95 is organized by the department of Computer Science, Univeristy
+% of \AA rhus and AITO (association Internationa pour les Technologie
+% Object) in cooperation with ACM/SIGPLAN.
+% %
+% \section*{Executive Commitee}
+% \begin{tabular}{@{}p{5cm}@{}p{7.2cm}@{}}
+% Conference Chair:&Ole Lehrmann Madsen (\AA rhus University, DK)\\
+% Program Chair:   &Walter Olthoff (DFKI GmbH, Germany)\\
+% Organizing Chair:&J\o rgen Lindskov Knudsen (\AA rhus University, DK)\\
+% Tutorials:&Birger M\o ller-Pedersen\hfil\break
+% (Norwegian Computing Center, Norway)\\
+% Workshops:&Eric Jul (University of Kopenhagen, Denmark)\\
+% Panels:&Boris Magnusson (Lund University, Sweden)\\
+% Exhibition:&Elmer Sandvad (\AA rhus University, DK)\\
+% Demonstrations:&Kurt N\o rdmark (\AA rhus University, DK)
+% \end{tabular}
+% %
+% \section*{Program Commitee}
+% \begin{tabular}{@{}p{5cm}@{}p{7.2cm}@{}}
+% Conference Chair:&Ole Lehrmann Madsen (\AA rhus University, DK)\\
+% Program Chair:   &Walter Olthoff (DFKI GmbH, Germany)\\
+% Organizing Chair:&J\o rgen Lindskov Knudsen (\AA rhus University, DK)\\
+% Tutorials:&Birger M\o ller-Pedersen\hfil\break
+% (Norwegian Computing Center, Norway)\\
+% Workshops:&Eric Jul (University of Kopenhagen, Denmark)\\
+% Panels:&Boris Magnusson (Lund University, Sweden)\\
+% Exhibition:&Elmer Sandvad (\AA rhus University, DK)\\
+% Demonstrations:&Kurt N\o rdmark (\AA rhus University, DK)
+% \end{tabular}
+% %
+% \begin{multicols}{3}[\section*{Referees}]
+% V.~Andreev\\
+% B\"arwolff\\
+% E.~Barrelet\\
+% H.P.~Beck\\
+% G.~Bernardi\\
+% E.~Binder\\
+% P.C.~Bosetti\\
+% Braunschweig\\
+% F.W.~B\"usser\\
+% T.~Carli\\
+% A.B.~Clegg\\
+% G.~Cozzika\\
+% S.~Dagoret\\
+% Del~Buono\\
+% P.~Dingus\\
+% H.~Duhm\\
+% J.~Ebert\\
+% S.~Eichenberger\\
+% R.J.~Ellison\\
+% Feltesse\\
+% W.~Flauger\\
+% A.~Fomenko\\
+% G.~Franke\\
+% J.~Garvey\\
+% M.~Gennis\\
+% L.~Goerlich\\
+% P.~Goritchev\\
+% H.~Greif\\
+% E.M.~Hanlon\\
+% R.~Haydar\\
+% R.C.W.~Henderso\\
+% P.~Hill\\
+% H.~Hufnagel\\
+% A.~Jacholkowska\\
+% Johannsen\\
+% S.~Kasarian\\
+% I.R.~Kenyon\\
+% C.~Kleinwort\\
+% T.~K\"ohler\\
+% S.D.~Kolya\\
+% P.~Kostka\\
+% U.~Kr\"uger\\
+% J.~Kurzh\"ofer\\
+% M.P.J.~Landon\\
+% A.~Lebedev\\
+% Ch.~Ley\\
+% F.~Linsel\\
+% H.~Lohmand\\
+% Martin\\
+% S.~Masson\\
+% K.~Meier\\
+% C.A.~Meyer\\
+% S.~Mikocki\\
+% J.V.~Morris\\
+% B.~Naroska\\
+% Nguyen\\
+% U.~Obrock\\
+% G.D.~Patel\\
+% Ch.~Pichler\\
+% S.~Prell\\
+% F.~Raupach\\
+% V.~Riech\\
+% P.~Robmann\\
+% N.~Sahlmann\\
+% P.~Schleper\\
+% Sch\"oning\\
+% B.~Schwab\\
+% A.~Semenov\\
+% G.~Siegmon\\
+% J.R.~Smith\\
+% M.~Steenbock\\
+% U.~Straumann\\
+% C.~Thiebaux\\
+% P.~Van~Esch\\
+% from Yerevan Ph\\
+% L.R.~West\\
+% G.-G.~Winter\\
+% T.P.~Yiou\\
+% M.~Zimmer\end{multicols}
+% %
+% \section*{Sponsoring Institutions}
+% %
+% Bernauer-Budiman Inc., Reading, Mass.\\
+% The Hofmann-International Company, San Louis Obispo, Cal.\\
+% Kramer Industries, Heidelberg, Germany
+% %
+% \tableofcontents
+% %
+\mainmatter              % start of the contributions
+%
+\title{Scaling Automated Theory Exploration}
+%
+%\titlerunning{Hamiltonian Mechanics}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
+\author{Chris Warburton \and Ekaterina Komendantskaya}
+%
+%\authorrunning{Ivar Ekeland et al.} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+% \tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
+% Craig Chambers, Kim B. Bruce, and Elisa Bertino}
+%
+\institute{University of Dundee,\\
+\texttt{http://tocai.computing.dundee.ac.uk}}
+
+\maketitle              % typeset the title of the contribution
+
+\begin{abstract}
 We investigate the \textbf{theory exploration} (TE)
@@
 and proofs, but realising this potential requires more scalable
-algorithms than presently used.} % Abstract of the article
-%\keywords{} % Keywords for the article
-
-\begin{document}
-\maketitle
-
-% Main part of your article
-% =========================
+algorithms than presently used.
+\keywords{theory exploration, Haskell}
+\end{abstract}
+%
 \section{Introduction}
@@
 
-% Acknowledgements for colleagues, referees, ...
-% ==============================================
 \begin{acknowledge}
@@
 
-% Bibliography with BibTeX
-% ========================
-\bibliographystyle{eceasst}
+\bibliographystyle{plain}
 \bibliography{../Bibtex}
 
+%
+% ---- Bibliography ----
+%
+\begin{thebibliography}{5}
+%
+\bibitem {clar:eke}
+Clarke, F., Ekeland, I.:
+Nonlinear oscillations and
+boundary-value problems for Hamiltonian systems.
+Arch. Rat. Mech. Anal. 78, 315--333 (1982)
+
+\bibitem {clar:eke:2}
+Clarke, F., Ekeland, I.:
+Solutions p\'{e}riodiques, du
+p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
+Note CRAS Paris 287, 1013--1015 (1978)
+
+\bibitem {mich:tar}
+Michalek, R., Tarantello, G.:
+Subharmonic solutions with prescribed minimal
+period for nonautonomous Hamiltonian systems.
+J. Diff. Eq. 72, 28--55 (1988)
+
+\bibitem {tar}
+Tarantello, G.:
+Subharmonic solutions for Hamiltonian
+systems via a $\bbbz_{p}$ pseudoindex theory.
+Annali di Matematica Pura (to appear)
+
+\bibitem {rab}
+Rabinowitz, P.:
+On subharmonic solutions of a Hamiltonian system.
+Comm. Pure Appl. Math. 33, 609--633 (1980)
+
+\end{thebibliography}
+
 \end{document}

### 2015-08-07 AI4FM/article.tex
More changes

--- 2015-08-07 AI4FM/article.tex
@@
-\documentclass{eceasst}
-% This is an empty ECEASST article that can be used as a template
-% by authors.
-% Just uncomment the appropriate frontmatter commands and provide
-% the parameters.
+% This is LLNCS.DEM the demonstration file of
+% the LaTeX macro package from Springer-Verlag
+% for Lecture Notes in Computer Science,
+% version 2.4 for LaTeX2e as of 16. April 2010
+%
+\documentclass{llncs}
+%
+\usepackage{makeidx}  % allows for indexgeneration
+\usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage{mathtools}
+\usepackage{amssymb}
+\usepackage{enumitem}
+%\hypersetup{colorlinks=true}
+%
+\begin{document}
+%
+%\frontmatter          % for the preliminaries
+%
+\pagestyle{headings}  % switches on printing of running heads
 
-% Required packages
-% =================
-% Your \usepackage commands go here.
+\mainmatter              % start of the contributions
+%
+\title{Scaling Automated Theory Exploration}
+%
+%\titlerunning{Hamiltonian Mechanics}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
+\author{Chris Warburton \and Ekaterina Komendantskaya}
+%
+%\authorrunning{Ivar Ekeland et al.} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+% \tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
+% Craig Chambers, Kim B. Bruce, and Elisa Bertino}
+%
+\institute{University of Dundee,\\
+\texttt{http://tocai.computing.dundee.ac.uk}}
 
-% Volume frontmatter
-% ==================
-\input{frontmatter}
+\maketitle              % typeset the title of the contribution
 
-% Article frontmatter
-% ===================
-\title{Scaling Automated Theory Exploration} % Title of the article
-%\short{} % Short title of the article (optional)
-\author{Chris Warburton\sponsor{}} % Authors and references to addresses
-\institute{ % Institutes with labels
-\email{cmwarburton@dundee.ac.uk}\\
-School of Computing\\
-University of Dundee}
-\abstract{
+\begin{abstract}
 We investigate the \textbf{theory exploration} (TE)
@@
 and proofs, but realising this potential requires more scalable
-algorithms than presently used.} % Abstract of the article
-%\keywords{} % Keywords for the article
-
-\begin{document}
-\maketitle
-
-% Main part of your article
-% =========================
+algorithms than presently used.
+\keywords{theory exploration, Haskell}
+\end{abstract}
+%
 \section{Introduction}
 
-The \emph{theory exploration} (TE) paradigm provides software support
-for traditional Mathematical workflows; namely, deriving ``interesting''
-consequences from formal definitions \cite{RISC1482}. Early
-implementations like \textsc{Theorema} \cite{buchberger2000theory} provided
-interactive environments, similar to computer algebra systems like
-\textsc{Mathematica} (in which \textsc{Theorema} is implemented) or
-interactive theorem provers. Subsequent systems have investigated
-\emph{automated} theory exploration, for tasks such as lemma discovery
-\cite{Hipster}.
+Given a signature and a set of variables (collectively a \emph{theory}), then
+\emph{theory exploration} (TE) describes any process
+$(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing
+terms of the theory which are well-formed, provable and satisfy some criterion
+referred to as ``interesting''. These conditions give rise to the following
+questions, which we use to characterise TE systems:
+\begin{description}
+\item [Q1] \label{Q1} How do we form terms?
+\item [Q2] \label{Q2} How do we guarantee well-formedness?
+\item [Q3] \label{Q3} How do we prove terms?
+\item [Q4] \label{Q4} What is considered ``interesting''?
+\end{description}
 
-By removing user interaction, automated TE systems require an algorithm
-for deciding whether to inform the user of a particular theorem or not; we refer
-to this search criterion as the theorem being ``interesting''. In existing
-systems, this criterion is intimately connected to the
-search algorithm; such coupling improves efficiency, even for brute-force
-search, but makes it difficult to try alternative measures.
+Early implementations like \textsc{Theorema} \cite{buchberger2000theory}
+provided interactive environments, similar to computer algebra systems and
+interactive theorem provers, to assist the user in finding theorems. In this
+setting, terms are formed by the user in whichever way they find interesting,
+whilst the software provides support for \textbf{Q2} and \textbf{Q3}.
 
-As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
-equations, which are defined as ``interesting'' if they cannot be simplified
+Subsequent systems have investigated \emph{automated} theory exploration, for
+tasks such as lemma discovery \cite{Hipster}. By removing user interaction,
+\textbf{Q1} and \textbf{Q4} must be solved by algorithms. In existing
+systems these are tightly coupled to improve efficiency, which makes it
+difficult to try different approaches independently.
+
+As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about
+Haskell code, which are defined as ``interesting'' if they cannot be simplified
 using previously discovered equations. The intuition for such criteria is to
-avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$,
-etc. when we already know $\forall x. 0 + x = x$. Whilst elegantly implemented
-as a term rewriting system, to work effectively this requires general forms to
-be found \emph{before} special cases.
+avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
+when we already know $0 + x = x$. Whilst \textsc{Q4} is elegantly implemented
+with a congruence closure relation (version 1) and a term rewriting system
+(version 2), the term generation for \textsc{Q1} is performed by brute-force.
 
 In the following, we give an overview of the state of the art in automated
-theory exploration (\autoref{haskell}). We identify potential improvements to
-these systems (\autoref{improve}) and present our initial implementation
-attempts in the form of the \textsc{ML4HS} system (\autoref{ml4hs}).
+theory exploration, then present potential improvements and our initial attempts
+at implementation.
 
@@
 encode such laws, less effort is given to finding and stating them; compared to
-a theorem proving setting like Isabelle.
+full theorem provers like Isabelle. Hence we imagine even a shallow exploration
+of code repositories such as \textsc{Hackage} could find many interesting
+theorems.
 
-Due to Haskell's relative popularity, there are also large code repositories
-such as \textsc{Hackage} available to explore. This gives several potential
-benefits: in the short term, applying TE to such repositories may assist
-existing library authors and users in comprehending and maintaining their code
-\cite{QuickSpec}; more notably, TE could be used to data mine new abstractions,
-such as typeclasses, with relevance to real-world applications.
-
-Currently, the most powerful TE system for Haskell is \textsc{HipSpec}.
-This uses the earlier \textsc{QuickSpec} system to search through
-\emph{expressions} (combinations of the Haskell terms given by the theory),
-rather than searching through the space of equations or proofs directly.
-Expressions are grouped into equivalence classes, such that the
-\textsc{QuickCheck} counterexample finder cannot distinguish between the
-elements; equations relating the members of these classes are then conjectured,
-and sent to existing automated theorem provers to try and prove
-\cite{rosen2012proving}. This approach works well as a lemma generation system,
-making \textsc{HipSpec} a capable inductive theorem prover as well as a theory
-exploration system \cite{claessen2013automating}.
-
-\section{Identifying Improvements}\label{improve}
-
-Given this state of the art, we identify the following as potential
-areas for investigation:
-
-\begin{itemize}
-\item
-  More scalable alternatives to brute-force enumeration. This is a well-studied
-  area of Artificial Intelligence and Machine Learning. In particular, there is
-  potential in applying heuristic algorithms like those surveyed in
-  \cite{blum2011hybrid}.
-\item
-  Using alternative ``interestingness'' criteria, like those surveyed in
-  \cite{geng2006interestingness}. In particular, we envision two problems with
-  large-scale exploration which may require more sophisticated criteria:
-
-  \begin{itemize}
-  \item
-    As the number of irreducible equations grows, it may be desirable to
-    impose extra conditions of a more subjective nature; e.g. to distinguish
-    useful relationships from incidental coincidences.
-  \item
-    Surprising, insightful equations may be discarded by current systems, if
-    they are actually reducible in some complex, non-obvious way. A more
-    subjective interestingness measure could be used to veto such
-    rejections.
-  \end{itemize}
-\item
-  Data mining previous results; for example, using generalisation methods like
-  anti-unification \cite{bulychev2010anti}. This would make more efficient use
-  of data, and also remove the requirement that general forms be enumerated
-  early.
-\item
-  Divide and conquer approaches. Rather than try all type-safe combinations of
-  the given expressions, it may be discernable a priori that some combinations
-  are not worth considering (either because they are never related, or because
-  their relations are never interesting). In other words, automating the theory
-  selection process currently performed by the user.
-\end{itemize}
-
-We consider all of these worth investigating, although in this paper we
-focus on the last item.
+Currently, the most powerful TE system for Haskell is \textsc{HipSpec}. This
+uses the earlier \textsc{QuickSpec} system to enumerate all type-correct
+combinations of the terms in the theory, group them into equivalence
+classes using the \textsc{QuickCheck} counterexample finder, then conjecture
+equations relating the members of these classes. Off-the-shelf automated theorem
+provers (ATPs) are used to prove these conjectures \cite{rosen2012proving}. This
+approach works well as a lemma generation system, making \textsc{HipSpec} a
+capable inductive theorem prover as well as a theory exploration system
+\cite{claessen2013automating}.
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[]{ml4hs}
-  \caption{An overview of the \textsc{ML4HS} system}
-  \label{ml4hsdia}
-\end{figure}
+We consider the use of type systems and ATPs as satisfactory solutions to
+\textbf{Q2} and \textbf{Q3}, respectively. For the other aspect, we identify
+the following potential improvements:
 
-We are implementing a system called \textsc{ML4HS} to investigate scalable
-automated theory exploration for Haskell. The current form of \textsc{ML4HS} is
-a pre-processor for \textsc{QuickSpec} which attempts to make exploration of
-large theories more tractable by splitting them into a collection of smaller
-theories: the divide and conquer approach mentioned above. This is
-similar to reducing the search space in automated theorem proving via premise
-selection \cite{kuhlwein2012overview}. The basic idea of \textsc{ML4HS} is shown
-in \autoref{ml4hsdia}: theories, in the form of Haskell packages, are split up
-and sent to concurrent instances of QuickSpec, and the results collated.
+\begin{description}
+\item [Q1]
+  Enumerating all type-correct terms is a brute-force solution to this question.
+  Scalable alternatives to brute-force algorithms are a well-studied area of
+  Artificial Intelligence and Machine Learning. In particular, there is
+  potential in applying heuristic algorithms like those surveyed in
+  \cite{blum2011hybrid}. We could also data-mine previous results using
+  generalisation methods like anti-unification \cite{bulychev2010anti}, and
+  use statistical methods to prioritise which elements of the theory we
+  investigate.
+\item [Q4]
+  Various ``interestingness'' criteria have been proposed, for example those
+  surveyed in \cite{geng2006interestingness}. Augmenting or replacing the
+  criteria may be useful, for example to distinguish useful relationships from
+  incidental coincidences; or to prevent surprising, insightful equations from
+  being discarded simply because they are reducible.
+\end{description}
 
-When selecting the components of each sub-theory, we want to keep together those
-expressions which combine to form interesting theorems, and separate those which
-combine in uninteresting ways. Our hypothesis is that similarity-based
-clustering of expressions, as found in \textsc{ML4PG}
-\cite{journals/corr/abs-1302-6421}, is an effective method for performing this
-division. Future experiments will test this by measuring throughput rates of
-\textsc{ML4HS} compared to \textsc{QuickSpec} and \textsc{HipSpec}.
+We are implementing a system called \textsc{ML4HS} to investigate the use of
+these ideas for a scalable automated theory exploration system for Haskell. In
+its current form, we are investigating the prioritisation of theory elements by
+augmenting \textsc{QuickSpec} with a pre-processor. Inspired by the use of
+premise selection \cite{kuhlwein2012overview} to reduce the search space in ATP,
+we select sub-sets of the given theory to explore, chosen to try and keep
+together those expressions which combine in interesting ways, and to separate
+those which combine in uninteresting ways.
 
-% Acknowledgements for colleagues, referees, ...
-% ==============================================
-\begin{acknowledge}
+We hypothesise that similarity-based clustering of expressions, inspired by that
+of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618}, is an effective method for
+performing this separation. Future experiments will test this by measuring
+throughput rates of \textsc{QuickSpec} with and without the \textsc{ML4HS}
+pre-processor, to see how the performance changes with the size of the theory,
+the elements of the theory (especially their types), and how much reduction we
+performed with our pre-processor.
+
+\section*{Acknowledgements}
+
 Thank you to the \textsc{HipSpec} team at Chalmers University (Moa Johansson,
@@
 discussions of these ideas.
-\end{acknowledge}
 
-% Bibliography with BibTeX
-% ========================
-\bibliographystyle{eceasst}
+\bibliographystyle{plain}
 \bibliography{../Bibtex}

### 2015-08-07 AI4FM/render.sh

@@ -10,9 +10,9 @@ function mdToLatex {
 }
 
 function renderLatex {
-    latex article
-    bibtex article
-    latex article
+    latex    article &&
+    bibtex   article &&
+    latex    article &&
     pdflatex article
 }
 

### 2015-08-07 AI4FM/article.tex
Final?

@@
 algorithms than presently used.
-\keywords{theory exploration, Haskell}
+%\keywords{theory exploration, Haskell}
 \end{abstract}
@@
 
-Given a signature and a set of variables (collectively a \emph{theory}), then
-\emph{theory exploration} (TE) describes any process
-$(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing
-terms of the theory which are well-formed, provable and satisfy some criterion
-referred to as ``interesting''. These conditions give rise to the following
-questions, which we use to characterise TE systems:
+Given a signature $\Sigma$ and a set of variables $V$, we call the pair
+$(\Sigma, V)$ a \emph{theory} and use \emph{theory exploration} (TE) to refer
+to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$
+for producing terms of the theory which are well-formed, provable and satisfy
+some criterion referred to as ``interesting''. These conditions give rise to the
+following questions, which we use to characterise TE systems:
 \begin{description}
-\item [Q1] \label{Q1} How do we form terms?
+\item [Q1] \label{Q1} How do we generate terms?
 \item [Q2] \label{Q2} How do we guarantee well-formedness?
@@
 avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$. Whilst \textsc{Q4} is elegantly implemented
+when we already know $0 + x = x$. Whilst \textbf{Q4} is elegantly implemented
 with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the term generation for \textsc{Q1} is performed by brute-force.
+(version 2), the term generation for \textbf{Q1} is performed by brute-force.
+
+Although \textsc{QuickSpec} only \emph{tests} its equations rather than
+proving them, it is still used as the exploration component of more rigorous
+systems like \textsc{HipSpec} and \textsc{Hipster}.
 
@@
 
-Currently, the most powerful TE system for Haskell is \textsc{HipSpec}. This
-uses the earlier \textsc{QuickSpec} system to enumerate all type-correct
-combinations of the terms in the theory, group them into equivalence
-classes using the \textsc{QuickCheck} counterexample finder, then conjecture
-equations relating the members of these classes. Off-the-shelf automated theorem
-provers (ATPs) are used to prove these conjectures \cite{rosen2012proving}. This
-approach works well as a lemma generation system, making \textsc{HipSpec} a
-capable inductive theorem prover as well as a theory exploration system
-\cite{claessen2013automating}.
+Currently, the most powerful TE system for Haskell is \textsc{HipSpec}, which
+uses off-the-shelf automated theorem provers (ATPs) to verify the conjectures of
+\textsc{QuickSpec}. \textsc{QuickSpec}, in turn, enumerates all type-correct combinations
+of the terms in the theory up to some depth, groups them into equivalence
+classes using the \textsc{QuickCheck} counterexample finder, then conjectures
+equations relating the members of these classes. This approach works well as a
+lemma generation system, making \textsc{HipSpec} a capable inductive theorem
+prover as well as a theory exploration system \cite{claessen2013automating}.
 
@@
 
-We consider the use of type systems and ATPs as satisfactory solutions to
-\textbf{Q2} and \textbf{Q3}, respectively. For the other aspect, we identify
-the following potential improvements:
+We consider \textbf{Q2} and \textbf{Q3} to be adequately solved by the existing
+use of type systems and ATPs, respectively. We identify the following potential
+improvements for the other questions:
 
@@
   Scalable alternatives to brute-force algorithms are a well-studied area of
-  Artificial Intelligence and Machine Learning. In particular, there is
-  potential in applying heuristic algorithms like those surveyed in
-  \cite{blum2011hybrid}. We could also data-mine previous results using
-  generalisation methods like anti-unification \cite{bulychev2010anti}, and
-  use statistical methods to prioritise which elements of the theory we
-  investigate.
+  Artificial Intelligence and Machine Learning. In particular, heuristic
+  search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
+  We could also use Machine Learning methods to identify some sub-set of a given
+  theory, to prioritise over the rest.
 \item [Q4]
-  Various ``interestingness'' criteria have been proposed, for example those
-  surveyed in \cite{geng2006interestingness}. Augmenting or replacing the
-  criteria may be useful, for example to distinguish useful relationships from
-  incidental coincidences; or to prevent surprising, insightful equations from
-  being discarded simply because they are reducible.
+  Various alternative ``interestingness'' criteria have been proposed, for
+  example those surveyed in \cite{geng2006interestingness}. Augmenting or
+  replacing the criteria may be useful, for example to distinguish useful
+  relationships from incidental coincidences; or to prevent surprising,
+  insightful equations from being discarded because they can be simplified.
 \end{description}
 
-We are implementing a system called \textsc{ML4HS} to investigate the use of
-these ideas for a scalable automated theory exploration system for Haskell. In
-its current form, we are investigating the prioritisation of theory elements by
-augmenting \textsc{QuickSpec} with a pre-processor. Inspired by the use of
-premise selection \cite{kuhlwein2012overview} to reduce the search space in ATP,
+We are implementing a system called \textsc{ML4HS} to investigate these ideas.
+Its current form is a pre-processor for \textsc{QuickSpec} for prioritising
+theory elements. Inspired by the use of premise selection
+\cite{kuhlwein2012overview} to reduce the search space in ATP,
 we select sub-sets of the given theory to explore, chosen to try and keep
@@
 We hypothesise that similarity-based clustering of expressions, inspired by that
-of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618}, is an effective method for
-performing this separation. Future experiments will test this by measuring
-throughput rates of \textsc{QuickSpec} with and without the \textsc{ML4HS}
-pre-processor, to see how the performance changes with the size of the theory,
-the elements of the theory (especially their types), and how much reduction we
-performed with our pre-processor.
+of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and related work in ACL2
+\cite{heras2013proof}, is an effective method for performing this separation.
+Future experiments will test this by comparing the throughput of
+\textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
 

### 2015-08-17 AI4FM/render.sh
Use make

@@ -1,34 +1,0 @@
-#! /usr/bin/env nix-shell
-#! nix-shell -i bash -p bash
-
-function renderLatex {
-    latex    article &&
-    bibtex   article &&
-    latex    article &&
-    pdflatex article
-}
-
-function clean {
-    { rm article.aux;
-      rm article.blg;
-      rm article.log;
-      rm article.pdf;
-      rm article.bbl;
-      rm article.dvi;
-      rm article.out; } 2> /dev/null
-    return 0
-}
-
-function go {
-    renderLatex
-}
-
-# Run once (default) or forever (if given any argument)
-
-go
-
-while [ "x$1" = "xloop" ]
-do
-    sleep 2
-    go
-done

### 2015-09-24 TransferReport/report.tex
Added transfer report

--- 2015-08-07 AI4FM/article.tex
@@
-% This is LLNCS.DEM the demonstration file of
-% the LaTeX macro package from Springer-Verlag
-% for Lecture Notes in Computer Science,
-% version 2.4 for LaTeX2e as of 16. April 2010
-%
-\documentclass{llncs}
-%
-\usepackage{makeidx}  % allows for indexgeneration
+\documentclass[]{article}
+
 \usepackage{hyperref}
@@
 \usepackage{enumitem}
-%\hypersetup{colorlinks=true}
-%
+\usepackage{paralist}
+\usepackage{csquotes}
+\usepackage[affil-it]{authblk}
+
 \begin{document}
-%
-%\frontmatter          % for the preliminaries
-%
+
 \pagestyle{headings}  % switches on printing of running heads
 
-\mainmatter              % start of the contributions
-%
-\title{Scaling Automated Theory Exploration}
-%
-%\titlerunning{Hamiltonian Mechanics}  % abbreviated title (for running head)
-%                                     also used for the TOC unless
-%                                     \toctitle is used
-%
-\author{Chris Warburton \and Ekaterina Komendantskaya}
-%
-%\authorrunning{Ivar Ekeland et al.} % abbreviated author list (for running head)
-%
-%%%% list of authors for the TOC (use if author list has to be modified)
-% \tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
-% Craig Chambers, Kim B. Bruce, and Elisa Bertino}
-%
-\institute{University of Dundee,\\
+\title{Machine Learning Methods in Functional Programming and Interactive Theorem Proving}
+
+\author{Chris Warburton}
+
+\affil{University of Dundee,\\
 \texttt{http://tocai.computing.dundee.ac.uk}}
@@
 \maketitle              % typeset the title of the contribution
+
+\begin{abstract}
+Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
+Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
+\end{abstract}
+
+\section{Mechanised Proof}
+
+When computers are used to aid the development of formal proofs, it is usually within either the \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP) framework.
+Such theorem proving approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them.
+In addition to these traditional approaches, we also consider the recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.
+
+\subsection{Interactive Theorem Proving}
+
+ITP systems are built around an automated \emph{proof checker}.
+This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic TODO: Reference some example, eg.
+proof of false in Coq.
+A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}: implementing a very simple ``core'' checker, and having all other features be translated into a form checkable by this core \cite[\S~3]{barendregt2005challenge} TODO Maybe Chapter 18.
+Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers).
+instead?.
+This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.
+
+In the ITP setting, proving a conjecture is equivalent to producing an input to the checker which \begin{inparaenum}[a)]
+  \item is accepted and
+  \item has that conjecture as its conclusion.
+\end{inparaenum}
+
+TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
+
+Of course, \emph{producing} such proofs is much more difficult than checking them; as the name suggests, ITP leaves this task up to the user.
+To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format.
+Besides theoretical elegance, due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming.
+Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process.
+Secondly, the process of \emph{software verification}, a common task for formal proof, can be achieved in a very straightforward way: the software can be implemented inside the proof language, certified correct by the proof checker, then \emph{extracted} into an executable form.
+
+Despite these conveniences, ITP still requires enormous effort for non-trivial proof or verification tasks (e.g.
+see \cite{hales2015formal}).
+The most obvious way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible.
+The most striking example of this approach is the \emph{sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.
+
+\subsection{Automated Theorem Proving}
+
+ATP is the autonomous search for proofs in some logic.
+Every ATP system is built around a complete or partial decision procedure such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294}, usually augmented by heuristics and often adjustable using parameters.
+Most standalone ATP systems use a first-order logic \cite[\S~10]{Russell:2003:AIM:773294}, since many tasks in higher-order logics are undecidable (e.g.
+\cite{huet1973undecidability}).
+
+ATP is a classic field of research in artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
+Even by that time Newell and Simon has developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
+
+In a similar way to ITP, the central problem of ATP is simple to state but unfeasible to execute: search through the space of provable statements until the given conjecture is found; the path from conjecture to the initial axioms consitutes the proof.
+However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+
+\subsection{Theory Exploration}
+
+Our main subject of investigation is the use of computers for augmenting and automating the process of \emph{theory exploration} (TE).
+This is the discovery of ``interesting'' patterns, relationships and structure in a formal system.
+
+Buchberger identifies TE as \textquote{building up (large) mathematical knowledge bases in an efficient, reliable, well-structured, re-usable, and flexible way} \cite{buchberger2004algorithm}, and considers it \textquote{the natural paradigm for using automated or computer-supported proving systems} \cite{buchberger2000theory}.
+In other words, the ITP and ATP approaches described above are fundamental components of any TE system, but the primary focus is on \emph{generating} conjectures.
+
+The most striking difference between TE and theorem proving is the lack of an explicit goal (e.g.
+a user-provided conjecture to prove).
+Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
+Of course, a key question to ask is what do we mean by ``interesting''?
+There is no single answer, although many approximate measures have been proposed.
+The choice of what counts as ``interesting'' has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
+
+In the case of Buchberger's \texttt{Theorema}, the first self-confessed TE system, only the questions of well-formedness and proof were tackled; specifically by the use of types and ATP algorithms, respectively.
+Similar to the approach of ITP, the more difficult tasks (term generation and the determination of their interest) is left up to the user.
+
+\iffalse
+
+  task of theory exploration, and its existing software implementations, by Buchberger as complementary to  takes place after concrete definitions have been established (eg.
+by the process of \emph{theory formation})  identified by .
+Theory exploration is similar to \emph{experimental mathematics} process of identifying \emph{interesting} consequences of a given theory.
+
+ - Relation to Science
+  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
+ - Relation to AI tasks: exploring surroundings, etc.
+
+Theory exploration must take place after \emph{theory formation}
+
+Major fields include computer algebra and theorem proving.
+
+- Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+
+Computer algebra systems, such as Mathematica, are
+There arithmetic, has a Initially used for ``computation'', ie. evaluation of expressions,
+especially in arithmetic numerical The application to evaluation
+
+Theory formation: Alison? Others.
+Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
+Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
+Communication: Latex, Wikis, APIs, communicating with aliens
+
+\subsection{Exploration in Artificial Intelligence and Machine Learning}
+
+The theme of goals and exploration which occur in theorem proving and TE have analogues in the domains of artificial intelligence (AI) and machine learning (ML).
+In  in an commonly appearing in the form of the \emph{exploration vs exploitation problem} in goal-driven systems.
+
+Machine Learning
+
+Symbolic AI: ATP, theory exploration?
+
+Integrating statistical and symbolic
+
+Machine learning over structured data:
+1D is common: parsing natural language
+2D is common; images
+Trees are fractal
+Backpropagation through structure
+LSTM with recursive structure
+Most work tries to identify structure; we already have it
+
+One way to apply fixed-size machine learning algorithms to arbitrarily-sized recursive structures is to use a \emph{distributed representation}.
+These mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
+
+CURRENT PROGRESS
+
+Divide and conquor for theory exploration
+Does it help?
+
+Clustering and feature extraction
+
+FUTURE WORK
+
+QuickSpec: extend or extinguish?
+
+Improve and find other use cases/scenarios for clustering and feature extraction
+
+Other directions for Theory Exploration?
+
+How about systems based on term rewriting, logic programming, etc.?
 
@@
 algorithms than presently used.
-%\keywords{theory exploration, Haskell}
 \end{abstract}
-%
-\section{Introduction}
 
@@
 
+\fi
+
 \bibliographystyle{plain}

### 2015-09-24 TransferReport/render.sh

@@ -1,0 +1,15 @@
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p ditaa bash
+
+NAME=report
+
+run() {
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape "$1"
+}
+
+for EXT in aux bbl blg log out pdf
+do
+    rm -f "${NAME}.${EXT}"
+done
+
+latex "$NAME" && bibtex "$NAME" && latex "$NAME" && pdflatex "$NAME"

### 2015-10-08 STP/talks/2015/10/render.sh
Added accumulated changes

@@ -1,0 +1,3 @@
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p haskellPackages.pandoc-citeproc panpipe panhandle ditaa
+pandoc -w dzslides --standalone --self-contained --mathml -o slides.html --filter=pandoc-citeproc --filter=panpipe --filter=panhandle slides.md

### 2015-10-08 STP/talks/render.sh

@@ -1,0 +1,3 @@
+#!/usr/bin/env bash
+cd 2015/10
+./render.sh

### 2015-10-08 TransferReport/report.tex
More changes

@@
 \usepackage{amssymb}
-\usepackage{enumitem}
+%\usepackage{enumitem}
 \usepackage{paralist}
@@
 
-\section{Mechanised Proof}
+\section{Introduction}
+
+The automation of ``routine'' mathematical reasoning by computers is increasingly important for both mathematics and computer science.
+What has previously been seen as a pure research field of academic interest TODO: better wording [CITE] is becoming an indispensable tool at the cutting edge of these fields.
+The controversy surrounding Hales' TODO proof of the Kepler conjecture, and its subsequent formalisation [CITE], is perhaps the most famous example of mathematical results becoming too large and intricate to be feasibly verified, let alone discovered, without computer assistance.
+
+At the same time, the modern increasing reliance of a proof which relies on computers to such an extent that it is in cannot be understood  requirement
+ - Maths, verification, etc. are important
+ - Theorem proving: ATP and ITP
+
+Any theorem proving approach is fundamentally limited by its reliance on the user to provide the conjectures to be (dis)proved. We focus This
+
+ - (Automated) theory exploration
+ - Theory of ``interestingness'' is missing
+
+TODO: Include work trying to assign probability to formal sentences
+
+In \S\ref{background}, we formalise the notions of \emph{theory}, \emph{theory exploration} and \emph{interestingness}, along with neccessary background such as \emph{automated theorem proving}.
+In \S\ref{related} we review related work, including the application of statistical methods to formal systems and the use of pure exploration in artificial intelligence domains.
+\S\ref{current} summarises our recent work undertaken in this area, and \S\ref{future} presents future directions this can be taken.
+
+\section{Background}
+\label{background}
+
+\iffalse
+
+Before formalising To frame the concepts we will be building on, we first present an informal, simplistic picture.
+followed by individual define our terms in turn.
+Take a typical Gentzen-style judgement, e.g.
+in type theory:
+
+\[\Gamma \vdash p : S\]
+
+We can interpret this judgement as ``From context $\Gamma$, the proof $p$ of statement $S$ is derivable''.
+We can hang the components of a proof framework from the symbols of this judgement, in the following way:
+
+\begin{itemize}
+  \item We can use $\Gamma$ to represent a \emph{theory}; the extra-logical definitions whose consequences are to be explored.
+The field of \emph{theory formation}, also known as \emph{concept formation}, studies the discovery or invention of formal theories from informal sources, e.g.
+modelling scientific observations.
+  \item $S$ can represent our \emph{conjectures}
+  \item The witness $p$ represents a \emph{proof}. Discovering proofs \emph{theorem proving}
+  \item $\vdash$, the provability relation, can represent \emph{proof search}.
+  \item $:$ is a typing judgement, denoting that the term on its left-hand-side is an element of the type on its right-hand-side, or equivalently (via \emph{propositions as types}) that the proof object on its left is a witness of the proposition on its right.
+In our picture, this represents the process of \emph{proof checking} (or, equivalently, of type checking).
+
+The context for our investigation can be Constructing  Informally, we Roughly speaking,  this work, we focus on the conjecture-generation aspect of theory exploration.
+We define a theory exploration system as
+\end{itemize}
+
+TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to
+
+TODO: Background section
+
+TODO: Reword this, theory exploration is our main point.
+It is a research direction/area, etc.
+Don't define it negatively (``without goals''), define it positively.
+Maybe stop trying to encompass everyone's different demarcations, and have a ``for our purposes, ....''
 
 When computers are used to aid the development of formal proofs, it is usually within either the \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP) framework.
-Such theorem proving approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them.
-In addition to these traditional approaches, we also consider the recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.
+These approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them.
+In this work we focus on the automation of the conjecture generation process itself, a task we refer to as \emph{theory exploration}.
+We can characterise theory
+
+This reliance on goals In addition to these traditional approaches, we also consider the related recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.
 
@@
 
-ITP systems are built around an automated \emph{proof checker}.
-This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic TODO: Reference some example, eg.
+The simplest approach to proof automation automated ITP systems are built around an automated \emph{proof checker}.
+This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic (for example, TODO: Reference some example, eg.
 proof of false in Coq.
-A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}: implementing a very simple ``core'' checker, and having all other features be translated into a form checkable by this core \cite[\S~3]{barendregt2005challenge} TODO Maybe Chapter 18.
+A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  TODO Maybe Chapter 18.
 Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers).
@@
 TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
+      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?
 
@@
 To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format.
-Besides theoretical elegance, due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming.
+This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming.
 Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process.
@@
 The most obvious way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible.
-The most striking example of this approach is the \emph{sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.
+The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.
 
@@
 
+TODO: Traditionally, ATP describes.... We care about the algorithms (not UIs, languages, etc.)... We use ATP to refer to...
 ATP is the autonomous search for proofs in some logic.
-Every ATP system is built around a complete or partial decision procedure such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294}, usually augmented by heuristics and often adjustable using parameters.
-Most standalone ATP systems use a first-order logic \cite[\S~10]{Russell:2003:AIM:773294}, since many tasks in higher-order logics are undecidable (e.g.
+Every ATP system is built around a complete, partial or semi-decision procedure such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294}, usually augmented by heuristics and often adjustable using parameters.
+Most standalone TODO: EXPLAIN ATP systems use a sub-set of first-order logic \cite[\S~10]{Russell:2003:AIM:773294} to ensure (semi-)decidability, since many tasks in higher-order logics are undecidable (e.g.
 \cite{huet1973undecidability}).
 
+TODO: ATP isn't *contained in* AI; it overlaps with (eg. some ATP people may not like being called AI researchers)
 ATP is a classic field of research in artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
@@
 This is the discovery of ``interesting'' patterns, relationships and structure in a formal system.
+TODO: ``discovery'' in which sense?
+``patterns'': is ML4PG a system?
+
+TODO: Definition of TE, rather than description. Maybe ``for our purposes''. Maybe just focus on the conjecture generation, rather than including theorem proving under the same umbrella.
 
 Buchberger identifies TE as \textquote{building up (large) mathematical knowledge bases in an efficient, reliable, well-structured, re-usable, and flexible way} \cite{buchberger2004algorithm}, and considers it \textquote{the natural paradigm for using automated or computer-supported proving systems} \cite{buchberger2000theory}.
-In other words, the ITP and ATP approaches described above are fundamental components of any TE system, but the primary focus is on \emph{generating} conjectures.
+In this sense, the TE paradigm encompasses both the ITP and ATP approaches described above, but \emph{also} includes the necessary prior step of \emph{generating} (interactively or automatically) the conjectures to be proved (interactively or automatically).
 
-The most striking difference between TE and theorem proving is the lack of an explicit goal (e.g.
+TODO: If we define TE and ATP/ITP more concretely, this is more of a comment; the difference is obvious
+The most striking difference between TE and theorem proving is the former's lack of an explicit goal (e.g.
 a user-provided conjecture to prove).
@@
 
-\iffalse
+TODO: expansion, eg. interactive vs automated.
 
@@
 
-\subsection{Exploration in Artificial Intelligence and Machine Learning}
+TODO: Motivating theme/narrative. For example, interestingess measures for theory exploration
 
-The theme of goals and exploration which occur in theorem proving and TE have analogues in the domains of artificial intelligence (AI) and machine learning (ML).
+TODO: Related work section
+\section{Related Work}
+\label{related}
+
+\subsection{Conjecture Generation}
+
+The task of \emph{conjecture generation} lies at the heart of theory exploration, and
+
+\subsubsection{Lemma generation}
+
+\subsection{Exploration in Artificial Intelligence}
+
+TODO: https en.wikipedia.org/wiki/Discovery system
+
+Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
+Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
+
+\subsubsection{Evolutionary Computation}
+
+\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
+Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications in a domain like theory exploration, where recursive structures of unbounded depth arise.
+Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a viable approach to the search component of theory exploration.
+
+Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
+Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
+
+The evolutionary analogue of theory exploration is to remove the (explicit) fitness function, just like we remove explicit goals to turn the theorem proving task into the theory exploration task.
+There are two main reasons for studying such purely-exploratory evolutionary systems: \emph{artificial life} and \emph{deceptive problems}.
+The former attempts to gain insight into the nature of life and biology through competition over limited resources.
+Whilst this may have utility in resource allocation, e.g.
+efficient scheduling of a portfolio of ATP programs, there is no direct connection to the search aspect of theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
+\cite{baum2000evolution}).
+
+On the other hand, work on deceptive problems is highly relevant, as it has lead to studying \emph{exploration in general}, and hence provides insight for our narrower domain of theory exploration.
+Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (eg.
+by hill climbing), but few global optima which are hard to find.
+Many approaches to avoiding deception, such as \emph{niching methods} \cite{sareni1998fitness}, alter the output of the fitness function to promote \emph{diversity} and \emph{novelty}.
+For example, \emph{fitness sharing} promotes diversity by dividing up the fitness of identical or similar solutions, e.g.
+the result of a fitness function $f$ may be shared by two identical solutions $s_1$ and $s_2$, such that the augmented fitness $f'(s_1) = \frac{f(s_1)}{2}$ and $f'(s_2) = \frac{f(s_2)}{2}$.
+As always, there are many variations on this theme, such as how to judge similarity (e.g.
+``genetically'' by comparing the solutions themselves, or ``phenotypically'' by comparing their fitness), but all strive to penalise duplication and hence promote diversity.
+
+Once we augment fitness functions, to a greater or lesser degree, an obvious question arises: what if the augmented function contains nothing of the original fitness?
+This is the pure exploration scenario we are interested in, which has been found to provide useful ``stepping stones'', even in objective-driven domains \cite{lehman2011abandoning}.
+
+TODO: Expand
+
+\subsubsection{Artificial Curiosity}
+
+\emph{Artificial curiosity} (AC) is the drive to seek out input or data which is in some sense ``interesting'' \cite{schmidhuber2006developmental}.
+As an unsupervised learning task, AC has no access to labels or meanings associated with its data; the only features it can learn are the structure and relationships between observations and actions.
+The aim of using an ``interestingness'' measure is to force the AC system away from inputs which are not amenable to learning; either because they are simple enough to be completely characterised already, or because they are so complex as to be unintelligible.
+This property is succinctly captured by the \emph{Wundt curve} TODO: Cite.
+
+One common approach to striking this balance is to measure some information-theoretic property of the system or data, and reward the learner accordingly.
+A survey of interestingness measures in artificial curiosity is given in \cite{oudeyer2007intrinsic}.
+Of particular interest is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
+Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
+This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, new theorems and we select for useful generalisations which allow old special-case lemmas to be discarded.
+
+TODO: Coevolution
+
+Approaches to artificial curiosity mostly follow two themes: the definition of an explicit ``interestingness'' measure, from which rewards are calculated; or else the use of \emph{co-evolution} to \emph{emergent}.
+
+Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
+This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
+For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
+However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
+In this sense, the
+
+The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
+in a robot).
+Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
+Embodiment emerged as a response to the symbolic techniques of so-called \emph{Good Old Fashioned AI} (GOFAI), of which theorem proving is a classic example.
+In this sense, the fields of theory exploration and developmental robotics seem rather disconnected.
+Nevertheless, TE still exhibits a form of embodiment in two ways.
+Firstly, the abstract, mathematical domain being explored is not a model of some external, physical environment; the domain \emph{is} the environment; in other words, TE does not suffer from the same grounding problem as GOFAI.
+In particular, \emph{resource usage} is a critical factor in any search problem; if it weren't, then brute force would be a viable solution.
+
+\iffalse
+
+\subsubsection{Universal Drives}
+
+PhysRevLett.110.168702.pdf
+Omohundro? Too physical.
+\emph{Universal drives} are those
+
+ takes the heavily automated field of theorem proving   mbodies The attempt to automate earlier transition from goal-driven problem solving to goal-less exploration and problem invention is not unique to theorem proving; it has also been investigated occurred in  and exploration which occur in theorem proving and TE have analogues in the domains of artificial intelligence (AI) and machine learning (ML).
 In  in an commonly appearing in the form of the \emph{exploration vs exploitation problem} in goal-driven systems.
 
-Machine Learning
+\subsection{Statistics of Formal Systems}
 
-Symbolic AI: ATP, theory exploration?
+\subsubsection{Relevance}
 
-Integrating statistical and symbolic
+ML4PG
+ACL2(ml)
+Josef Urban
+Sledgehammer
+Naive Bayes
+
+\subsubsection{Concept Formation}
+
+Alan Bundy et al
+
+Eurisko, AM, etc.?
+
+Genetic programming? GP in finite algebras
+
+\subsubsection{Probability of Sentences}
+
+\subsubsection{Learning on Structured Data}
 
@@
 
+Recurrent neural networks
+Backpropagation through structure
+
 One way to apply fixed-size machine learning algorithms to arbitrarily-sized recursive structures is to use a \emph{distributed representation}.
@@
 
-CURRENT PROGRESS
+\section{CURRENT PROGRESS}
+\label{current}
 
-Divide and conquor for theory exploration
+Divide and conquer for theory exploration
 Does it help?
@@
 
-FUTURE WORK
+\section{FUTURE WORK}
+\label{future}
 
@@
 discussions of these ideas.
-
 \fi

### 2015-10-08 TheoryExploration/render.sh

@@ -1,5 +1,9 @@
 #!/bin/sh
-nix-shell --pure --show-trace --command "sh" <<EOF
-  echo "Works"
-  #ARGS="--filter panpipe --filter panhandle" PATTERN="notes" renderWatch md2pdf
+nix-shell --pure --show-trace --command "sh" <<'EOF'
+  for src in *.md
+  do
+    doc=$(basename "$src" .md)
+    echo "Rendering $doc"
+    pandoc --filter panpipe --filter panhandle -o "$doc.pdf" "$doc.md"
+  done
 EOF

### 2015-10-08 TransferReport/render.sh

@@ -12,4 +12,4 @@ do
     rm -f "${NAME}.${EXT}"
 done
 
-latex "$NAME" && bibtex "$NAME" && latex "$NAME" && pdflatex "$NAME"
+run "$NAME" && bibtex "$NAME" && run "$NAME" && run "$NAME"

### 2015-10-20 TransferReport/report.tex
More report

@@
 \usepackage[affil-it]{authblk}
+\usepackage{tikz}
+\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
 
@@
 
-The automation of ``routine'' mathematical reasoning by computers is increasingly important for both mathematics and computer science.
-What has previously been seen as a pure research field of academic interest TODO: better wording [CITE] is becoming an indispensable tool at the cutting edge of these fields.
-The controversy surrounding Hales' TODO proof of the Kepler conjecture, and its subsequent formalisation [CITE], is perhaps the most famous example of mathematical results becoming too large and intricate to be feasibly verified, let alone discovered, without computer assistance.
+As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
+Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
+By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
+However, the question still remains: on what should we focus that analysis?
 
-At the same time, the modern increasing reliance of a proof which relies on computers to such an extent that it is in cannot be understood  requirement
- - Maths, verification, etc. are important
- - Theorem proving: ATP and ITP
+In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems, especially software, as a way to make productive use of resources in an often intractable domain.
+Appeals to interestingness arise when more direct measures, such as utility, are not available; as an extreme example, we could this is analogous to for example, we could characterise the distinction between pure and applied mathematics.
+much work in pure mathematics could be considered For example, the inclusion of particular statements in a program or proof development can be easily justified by whether they contribute to the solution; however, in a \emph{library} there is no particular problem to be solved, in which case statements may instead be judged on how interesting they are.
+Appeals to interestingness abound in the history of computer-assisted reasoning; for example Lenat's 1977 description of the AM system using information to \textquote{concentrate upon the most interesting [concepts]} \cite{lenat1977automated}.
+Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
 
-Any theorem proving approach is fundamentally limited by its reliance on the user to provide the conjectures to be (dis)proved. We focus This
-
- - (Automated) theory exploration
- - Theory of ``interestingness'' is missing
-
-TODO: Include work trying to assign probability to formal sentences
-
-In \S\ref{background}, we formalise the notions of \emph{theory}, \emph{theory exploration} and \emph{interestingness}, along with neccessary background such as \emph{automated theorem proving}.
-In \S\ref{related} we review related work, including the application of statistical methods to formal systems and the use of pure exploration in artificial intelligence domains.
-\S\ref{current} summarises our recent work undertaken in this area, and \S\ref{future} presents future directions this can be taken.
+We begin in \S \ref{background} by defining a formal framework for our investigation, and how it relates to the existing theorem proving landscape.
+In \S \ref{conceptformation} we consider those heuristics proposed in the literature to date, whilst \S \ref{related} casts a wider net to encompass pure exploration and interestingness in other domains of machine learning and artificial intelligence.
+We consider a selection of results from the theorem proving community in \S \ref{examples}, both to evaluate these heuristics and to identify common features.
+We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
 
@@
 
+We will focus our investigation on inductive theorems regarding programs in the Haskell language, since this is an area with existing work \cite{Claessen_hipspec:automating} against which we can compare our methods.
+
+\begin{figure}
+  \centering
+  \tikzstyle{decision} = [diamond, draw, fill=blue!20, text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
+  \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text centered, rounded corners, minimum height=4em]
+  \tikzstyle{line} = [draw, -latex']
+  \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm, minimum height=2em]
+
+  \begin{tikzpicture}[node distance = 2cm, auto]
+    % Place nodes
+    \node [block] (cf) {Concept Formation};
+    \node [block, below of=cf] (cg) {Conjecture Generation};
+    \node [block, below of=cg] (tp) {Theorem Proving};
+
+    % Draw edges
+    \path [line] (cf) -- (cg);
+    \path [line] (cg) -- (tp);
+
+    % Draw braces
+    \draw [decorate,decoration={brace,amplitude=10pt, mirror}]
+    (cf.north west) -- (tp.south west) node [black,midway,xshift=-1.5cm]
+    {\footnotesize ITP};
+
+    \draw [decorate,decoration={brace,amplitude=10pt}]
+    (tp.north east) -- (tp.south east) node [black,midway,xshift=0.3cm]
+    {\footnotesize ATP};
+
+    \draw [decorate,decoration={brace,amplitude=10pt}, xshift=5cm]
+    ($ (cg.north east) + (1,0) $) -- ($ (tp.south east) + (1,0) $)
+    node [black,midway, xshift=0.6cm] {\footnotesize Theory Exploration};
+  \end{tikzpicture}
+  \caption{Core stages of computer-assisted proof \label{proofprocess}}
+\end{figure}
+
 \iffalse
 
-Before formalising To frame the concepts we will be building on, we first present an informal, simplistic picture.
+Before defining formalising To frame the concepts we will be building on, we first present an informal, simplistic picture.
 followed by individual define our terms in turn.
@@
 
-TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to
+\fi
 
-TODO: Background section
+% TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to
 
-TODO: Reword this, theory exploration is our main point.
+% TODO: Background section
+
+% TODO: Reword this, theory exploration is our main point.
 It is a research direction/area, etc.
@@
 
-When computers are used to aid the development of formal proofs, it is usually within either the \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP) framework.
-These approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them.
-In this work we focus on the automation of the conjecture generation process itself, a task we refer to as \emph{theory exploration}.
-We can characterise theory
-
-This reliance on goals In addition to these traditional approaches, we also consider the related recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.
+To set the scene for the formal definitions which follow, we first give an overview of our key terms and concepts, which are also depicted in figure \ref{proofprocess}.
+Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
+Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characteris both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
+In this work, we push the focus back to the prior steps of \emph{concept formation} (\S \ref{conceptformation}) and \emph{conjecture generation} (\S \ref{conjecturegeneration}), which are characterised as being \emph{open-ended} and \emph{creative}, in that solutions must be judged (as interesting or not) rather than merely constructed.
 
 \subsection{Interactive Theorem Proving}
+\label{itp}
 
-The simplest approach to proof automation automated ITP systems are built around an automated \emph{proof checker}.
-This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic (for example, TODO: Reference some example, eg.
-proof of false in Coq.
-A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  TODO Maybe Chapter 18.
+The ITP process relies on a \emph{proof checker} $C$, which is a decision procedure for determining if a given proof object $P$ witnesses a given statement $S$:
+
+$$ C : (P \times S) \rightarrow Boolean $$
+
+The ITP process then consists of finding an appropriate $P$ for the given \emph{goal} $S$, which is accepted by $C$:
+
+$$ ITP : (S \times C) \rightarrow \{ P | C(P, S) = True \} $$
+
+Automating a proof checker is relatively simple \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic \iffalse (for example, TODO: Reference some example, eg.
+proof of false in Coq).
+\fi A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  \iffalse TODO Maybe Chapter 18.
 Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers).
 instead?.
-This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.
+\fi This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.
 
-In the ITP setting, proving a conjecture is equivalent to producing an input to the checker which \begin{inparaenum}[a)]
-  \item is accepted and
-  \item has that conjecture as its conclusion.
-\end{inparaenum}
-
-TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
-      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?
+% TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
+%      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?
 
@@
 To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format.
-This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming.
+This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, and there are also practical benefits to turning theorem proving into programming.
 Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process.
@@
 see \cite{hales2015formal}).
-The most obvious way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible.
-The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.
+One common way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible \iffalse Adam Chlipala \fi.
+The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
 
 \subsection{Automated Theorem Proving}
+\label{itp}
 
@@
 
-TODO: Related work section
+\section{Examples of Interesting Objects}
+\label{examples}
+
+Before exploring abstract definitions of interestingness, we can first consider some examples of theorems, lemmas and properties which occur in existing developments, to see what real users considered interesting enough to formalise.
+This will give an empirical basis for our intuitions, which we can compare with the definitions to follow.
+
+\subsection{Generalisations}
+
+When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
+Proofs by induction often require generalisation, since the specific obligations which arise during the proof may be unsuitable for the inductive hypotheses which are available.
+However, we cannot blindly generalise all obligations we encounter, since \emph{over-generalising} results in obligations which are too strong to prove at all.
+We must therefore rely on heuristics to choose when and how to generalise, which can in part consider how interesting the generalisations are.
+
+As an example, we can take Boyer and Moore's proof of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
+
+$$(x * y) * z = x * (y * z)$$
+
+During the course of the proof, the following obligation arises:
+
+\begin{equation}
+  \tag{conc3}
+  (y + (x * y)) * z = (y * z) + ((x * y) * z)
+  \label{eq:conc3}
+\end{equation}
+
+ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
+
+\begin{equation}
+  \tag{conc4}
+  (y + w) * z = (y * z) + (w * z)
+  \label{eq:conc4}
+\end{equation}
+
+This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
+
+\subsection{Auxilliary Lemmas}
+
+Many of the above generalisations, such as  do not rely on context
+
+\subsection{QuickCheck Properties}
+
 \section{Related Work}
@@
 \subsubsection{Artificial Curiosity}
+\label{curiosity}
 
@@
 
-\iffalse
-
 \subsubsection{Universal Drives}
@@
 
+The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
+This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
+
 \subsubsection{Relevance}
+
+The combinatorial nature of formal systems leads many algorithms in this domain, such as proof search methods, to have exponential time complexityCITE; hence even a modest size increase can turn a trivial problem into an intractable one.
+It is currently infeasible to improve the scaling behaviour of many such algorithms; for example, problems like TODO are in the infamous NP-complete complexity classCITE.
+On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
+We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
+This leads to the idea of \emph{relevance filtering}.
+
+Relevance filtering simplifies a proof search problem by removing from consideration those axioms, definitions, lemmas, etc.
+which are deemed \emph{irrelevant}.
+The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only those lemmas which are deemed ``relevant'' (required for proving the statement) are included.
+This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a lemma is relevant: how do we know what will be required, before we have the proof?
+
+Sledgehammer's approach is to use \emph{supervised learning} to approximate the relevance relation: a naive bayes classifier is trained on TODO
 
@@
 
-\subsubsection{Concept Formation}
+\subsubsection{Interestingness in Concept Formation}
+\label{conceptformation}
 
-Alan Bundy et al
+% Alan Bundy et al
 
-Eurisko, AM, etc.?
+% Eurisko, AM, etc.?
 
-Genetic programming? GP in finite algebras
+% Genetic programming? GP in finite algebras
 
@@
 
-\section{CURRENT PROGRESS}
+\section{Contributions}
 \label{current}
 
-Divide and conquer for theory exploration
-Does it help?
+\subsection{The \textsc{ML4HS} Framework}\label{ml4hs}
 
-Clustering and feature extraction
+We consider \textbf{Q2} and \textbf{Q3} to be adequately solved by the existing
+use of type systems and ATPs, respectively. We identify the following potential
+improvements for the other questions:
 
-\section{FUTURE WORK}
+\begin{description}
+\item [Q1]
+  Enumerating all type-correct terms is a brute-force solution to this question.
+  Scalable alternatives to brute-force algorithms are a well-studied area of
+  Artificial Intelligence and Machine Learning. In particular, heuristic
+  search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
+  We could also use Machine Learning methods to identify some sub-set of a given
+  theory, to prioritise over the rest.
+\item [Q4]
+  Various alternative ``interestingness'' criteria have been proposed, for
+  example those surveyed in \cite{geng2006interestingness}. Augmenting or
+  replacing the criteria may be useful, for example to distinguish useful
+  relationships from incidental coincidences; or to prevent surprising,
+  insightful equations from being discarded because they can be simplified.
+\end{description}
+
+We are implementing a system called \textsc{ML4HS} to investigate these ideas.
+Its current form is a pre-processor for \textsc{QuickSpec} for prioritising
+theory elements. Inspired by the use of premise selection
+\cite{kuhlwein2012overview} to reduce the search space in ATP,
+we select sub-sets of the given theory to explore, chosen to try and keep
+together those expressions which combine in interesting ways, and to separate
+those which combine in uninteresting ways.
+
+We hypothesise that similarity-based clustering of expressions, inspired by that
+of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and related work in ACL2
+\cite{heras2013proof}, is an effective method for performing this separation.
+Future experiments will test this by comparing the throughput of
+\textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
+
+% Divide and conquer for theory exploration
+% Does it help?
+
+% Clustering and feature extraction
+
+\section{Future Work}
 \label{future}
+
+\iffalse
 
@@
 How about systems based on term rewriting, logic programming, etc.?
+\fi
+
+\section{Conclusion}
+\label{conclusion}
+
+\iffalse
 
@@
 
-\section{The \textsc{ML4HS} Framework}\label{ml4hs}
-
-We consider \textbf{Q2} and \textbf{Q3} to be adequately solved by the existing
-use of type systems and ATPs, respectively. We identify the following potential
-improvements for the other questions:
-
-\begin{description}
-\item [Q1]
-  Enumerating all type-correct terms is a brute-force solution to this question.
-  Scalable alternatives to brute-force algorithms are a well-studied area of
-  Artificial Intelligence and Machine Learning. In particular, heuristic
-  search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
-  We could also use Machine Learning methods to identify some sub-set of a given
-  theory, to prioritise over the rest.
-\item [Q4]
-  Various alternative ``interestingness'' criteria have been proposed, for
-  example those surveyed in \cite{geng2006interestingness}. Augmenting or
-  replacing the criteria may be useful, for example to distinguish useful
-  relationships from incidental coincidences; or to prevent surprising,
-  insightful equations from being discarded because they can be simplified.
-\end{description}
-
-We are implementing a system called \textsc{ML4HS} to investigate these ideas.
-Its current form is a pre-processor for \textsc{QuickSpec} for prioritising
-theory elements. Inspired by the use of premise selection
-\cite{kuhlwein2012overview} to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep
-together those expressions which combine in interesting ways, and to separate
-those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by that
-of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and related work in ACL2
-\cite{heras2013proof}, is an effective method for performing this separation.
-Future experiments will test this by comparing the throughput of
-\textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
 \section*{Acknowledgements}
@@
 discussions of these ideas.
+
 \fi

### 2015-10-27 TransferReport/report.tex
Big updates

@@
 \usepackage[affil-it]{authblk}
+\usepackage{listings}
 \usepackage{tikz}
@@
 
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems, especially software, as a way to make productive use of resources in an often intractable domain.
-Appeals to interestingness arise when more direct measures, such as utility, are not available; as an extreme example, we could this is analogous to for example, we could characterise the distinction between pure and applied mathematics.
-much work in pure mathematics could be considered For example, the inclusion of particular statements in a program or proof development can be easily justified by whether they contribute to the solution; however, in a \emph{library} there is no particular problem to be solved, in which case statements may instead be judged on how interesting they are.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example Lenat's 1977 description of the AM system using information to \textquote{concentrate upon the most interesting [concepts]} \cite{lenat1977automated}.
+In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
+To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
+Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
+
+Appeals to interestingness arise when more direct measures, such as utility, are not available.
+For example, the inclusion of particular statements in a program or proof development can be easily justified based on its contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
+Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
 Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
 
-We begin in \S \ref{background} by defining a formal framework for our investigation, and how it relates to the existing theorem proving landscape.
-In \S \ref{conceptformation} we consider those heuristics proposed in the literature to date, whilst \S \ref{related} casts a wider net to encompass pure exploration and interestingness in other domains of machine learning and artificial intelligence.
-We consider a selection of results from the theorem proving community in \S \ref{examples}, both to evaluate these heuristics and to identify common features.
+We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
+We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
+A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{relatedwork}.
+We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are also experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
+Recent efforts in this area have lead to principled theories to emerge, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
+
 We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
@@
 
-We will focus our investigation on inductive theorems regarding programs in the Haskell language, since this is an area with existing work \cite{Claessen_hipspec:automating} against which we can compare our methods.
+Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
+Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
 
-\begin{figure}
-  \centering
-  \tikzstyle{decision} = [diamond, draw, fill=blue!20, text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
-  \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text centered, rounded corners, minimum height=4em]
-  \tikzstyle{line} = [draw, -latex']
-  \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm, minimum height=2em]
+In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
+Since many theories will admit an infinite number of trivial theorems (e.g.
+$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
+This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
 
-  \begin{tikzpicture}[node distance = 2cm, auto]
-    % Place nodes
-    \node [block] (cf) {Concept Formation};
-    \node [block, below of=cf] (cg) {Conjecture Generation};
-    \node [block, below of=cg] (tp) {Theorem Proving};
+Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
+This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
+We decided to focus on Haskell, as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \cite{QuickSpec} and \textsc{HipSpec} \cite{claessen2013automating}).
+This is evident even to users of Isabelle/HOL, which boasts its own \textsc{IsaCoSy} \cite{johansson2009isacosy} and \textsc{Hipster} \cite{Hipster} theory exploration systems, since \textsc{Hipster} is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
 
-    % Draw edges
-    \path [line] (cf) -- (cg);
-    \path [line] (cg) -- (tp);
+\subsection{Haskell}\label{haskell}
 
-    % Draw braces
-    \draw [decorate,decoration={brace,amplitude=10pt, mirror}]
-    (cf.north west) -- (tp.south west) node [black,midway,xshift=-1.5cm]
-    {\footnotesize ITP};
+\lstset{
+  frame=none,
+  xleftmargin=2pt,
+  stepnumber=1,
+  numbers=none,
+  numbersep=5pt,
+  numberstyle=\ttfamily\tiny\color[gray]{0.3},
+  belowcaptionskip=\bigskipamount,
+  captionpos=b,
+  escapeinside={*'}{'*},
+  language=haskell,
+  tabsize=2,
+  emphstyle={\bf},
+  commentstyle=\it,
+  stringstyle=\mdseries\rmfamily,
+  showspaces=false,
+  keywordstyle=\bfseries\rmfamily,
+  columns=flexible,
+  basicstyle=\small\sffamily,
+  showstringspaces=false,
+  morecomment=[l]\%,
+}
 
-    \draw [decorate,decoration={brace,amplitude=10pt}]
-    (tp.north east) -- (tp.south east) node [black,midway,xshift=0.3cm]
-    {\footnotesize ATP};
+\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
 
-    \draw [decorate,decoration={brace,amplitude=10pt}, xshift=5cm]
-    ($ (cg.north east) + (1,0) $) -- ($ (tp.south east) + (1,0) $)
-    node [black,midway, xshift=0.6cm] {\footnotesize Theory Exploration};
-  \end{tikzpicture}
-  \caption{Core stages of computer-assisted proof \label{proofprocess}}
-\end{figure}
+Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
+Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to a \emph{GHC Core}).
+For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \cite{marlow2010haskell} \footnote{Version 2010 was current as this was written}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
 
-\iffalse
+\begin{description}
 
-Before defining formalising To frame the concepts we will be building on, we first present an informal, simplistic picture.
-followed by individual define our terms in turn.
-Take a typical Gentzen-style judgement, e.g.
-in type theory:
+\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
 
-\[\Gamma \vdash p : S\]
+\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
 
-We can interpret this judgement as ``From context $\Gamma$, the proof $p$ of statement $S$ is derivable''.
-We can hang the components of a proof framework from the symbols of this judgement, in the following way:
+\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
+
+\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
+performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
+This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
+For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
+Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
+
+\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
+For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
+an ADT's declaration specifies all of the normal forms for that type.
+This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
+
+\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
+This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
+t -> t}.
+\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
+\hs{id x = x}.
+This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \cite{wadler1989theorems}.
+
+\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
+Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
+As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+op :: forall t. Semigroup t => t -> t -> t
+\end{lstlisting}
+
+$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
+
+The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
+\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
+As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
+Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
+\hs{Integer} with \hs{op} performing addition.
+Many more examples can be found in the \emph{typeclassopedia} \cite{yorgey2009typeclassopedia}.
+This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
+
+\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
+This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
+Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
+
+\end{description}
+
+Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
+However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \cite{lindley2014hasochism}).
+This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
+in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
+We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell in thus in a unique position with regards to the discovery of interesting theorems.
+Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
+The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
+
+Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
 
 \begin{itemize}
-  \item We can use $\Gamma$ to represent a \emph{theory}; the extra-logical definitions whose consequences are to be explored.
-The field of \emph{theory formation}, also known as \emph{concept formation}, studies the discovery or invention of formal theories from informal sources, e.g.
-modelling scientific observations.
-  \item $S$ can represent our \emph{conjectures}
-  \item The witness $p$ represents a \emph{proof}. Discovering proofs \emph{theorem proving}
-  \item $\vdash$, the provability relation, can represent \emph{proof search}.
-  \item $:$ is a typing judgement, denoting that the term on its left-hand-side is an element of the type on its right-hand-side, or equivalently (via \emph{propositions as types}) that the proof object on its left is a witness of the proposition on its right.
-In our picture, this represents the process of \emph{proof checking} (or, equivalently, of type checking).
 
-The context for our investigation can be Constructing  Informally, we Roughly speaking,  this work, we focus on the conjecture-generation aspect of theory exploration.
-We define a theory exploration system as
+\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
+
+\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
+
+\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
+
 \end{itemize}
-
-\fi
-
-% TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to
-
-% TODO: Background section
-
-% TODO: Reword this, theory exploration is our main point.
-It is a research direction/area, etc.
-Don't define it negatively (``without goals''), define it positively.
-Maybe stop trying to encompass everyone's different demarcations, and have a ``for our purposes, ....''
-
-To set the scene for the formal definitions which follow, we first give an overview of our key terms and concepts, which are also depicted in figure \ref{proofprocess}.
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characteris both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-In this work, we push the focus back to the prior steps of \emph{concept formation} (\S \ref{conceptformation}) and \emph{conjecture generation} (\S \ref{conjecturegeneration}), which are characterised as being \emph{open-ended} and \emph{creative}, in that solutions must be judged (as interesting or not) rather than merely constructed.
 
@@
 
-The ITP process relies on a \emph{proof checker} $C$, which is a decision procedure for determining if a given proof object $P$ witnesses a given statement $S$:
+ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
 
-$$ C : (P \times S) \rightarrow Boolean $$
+$$ C \colon (P \times S) \rightarrow Boolean $$
 
-The ITP process then consists of finding an appropriate $P$ for the given \emph{goal} $S$, which is accepted by $C$:
+The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
 
-$$ ITP : (S \times C) \rightarrow \{ P | C(P, S) = True \} $$
+$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
 
-Automating a proof checker is relatively simple \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic \iffalse (for example, TODO: Reference some example, eg.
-proof of false in Coq).
-\fi A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  \iffalse TODO Maybe Chapter 18.
-Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers).
-instead?.
-\fi This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.
+It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
+In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
+There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
 
-% TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
-%      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?
+This coincidence is due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, which identifies programming languages with systems of logic.
+Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
+With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
 
-Of course, \emph{producing} such proofs is much more difficult than checking them; as the name suggests, ITP leaves this task up to the user.
-To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format.
-This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, and there are also practical benefits to turning theorem proving into programming.
-Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process.
-Secondly, the process of \emph{software verification}, a common task for formal proof, can be achieved in a very straightforward way: the software can be implemented inside the proof language, certified correct by the proof checker, then \emph{extracted} into an executable form.
+Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
+In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
+For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
+Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
+This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof} \cite{barendregt2001proof}.
+One nice consequence of this approach, which is followed for example by Isabelle \cite{nipkow2002isabelle} and Coq \cite{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of our confidence in the results.
 
-Despite these conveniences, ITP still requires enormous effort for non-trivial proof or verification tasks (e.g.
+Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
 see \cite{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible \iffalse Adam Chlipala \fi.
-The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
+One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
+The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
 
 \subsection{Automated Theorem Proving}
-\label{itp}
+\label{atp}
 
-TODO: Traditionally, ATP describes.... We care about the algorithms (not UIs, languages, etc.)... We use ATP to refer to...
-ATP is the autonomous search for proofs in some logic.
-Every ATP system is built around a complete, partial or semi-decision procedure such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294}, usually augmented by heuristics and often adjustable using parameters.
-Most standalone TODO: EXPLAIN ATP systems use a sub-set of first-order logic \cite[\S~10]{Russell:2003:AIM:773294} to ensure (semi-)decidability, since many tasks in higher-order logics are undecidable (e.g.
-\cite{huet1973undecidability}).
+ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
+By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \cite{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
+In particular, most ATP systems (such as E \cite{schulz2013system} and Vampire \cite{riazanov2003implementing}) operate in classical first-order logic.
+This is the most striking difference from ITP systems (e.g.
+Coq, Agda and Isabelle) which operate in higher-order logics.
 
-TODO: ATP isn't *contained in* AI; it overlaps with (eg. some ATP people may not like being called AI researchers)
-ATP is a classic field of research in artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
+In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
 Even by that time Newell and Simon has developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
+The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a form amenable to first-order reasoning.
 
-In a similar way to ITP, the central problem of ATP is simple to state but unfeasible to execute: search through the space of provable statements until the given conjecture is found; the path from conjecture to the initial axioms consitutes the proof.
-However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
 
 \subsection{Theory Exploration}
+\label{theoryexploration}
 
-Our main subject of investigation is the use of computers for augmenting and automating the process of \emph{theory exploration} (TE).
-This is the discovery of ``interesting'' patterns, relationships and structure in a formal system.
-TODO: ``discovery'' in which sense?
-``patterns'': is ML4PG a system?
-
-TODO: Definition of TE, rather than description. Maybe ``for our purposes''. Maybe just focus on the conjecture generation, rather than including theorem proving under the same umbrella.
-
-Buchberger identifies TE as \textquote{building up (large) mathematical knowledge bases in an efficient, reliable, well-structured, re-usable, and flexible way} \cite{buchberger2004algorithm}, and considers it \textquote{the natural paradigm for using automated or computer-supported proving systems} \cite{buchberger2000theory}.
-In this sense, the TE paradigm encompasses both the ITP and ATP approaches described above, but \emph{also} includes the necessary prior step of \emph{generating} (interactively or automatically) the conjectures to be proved (interactively or automatically).
-
-TODO: If we define TE and ATP/ITP more concretely, this is more of a comment; the difference is obvious
-The most striking difference between TE and theorem proving is the former's lack of an explicit goal (e.g.
+The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
 a user-provided conjecture to prove).
@@
 
-In the case of Buchberger's \texttt{Theorema}, the first self-confessed TE system, only the questions of well-formedness and proof were tackled; specifically by the use of types and ATP algorithms, respectively.
-Similar to the approach of ITP, the more difficult tasks (term generation and the determination of their interest) is left up to the user.
+Following \cite{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
+\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
 
-TODO: expansion, eg. interactive vs automated.
+The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
+Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
 
-  task of theory exploration, and its existing software implementations, by Buchberger as complementary to  takes place after concrete definitions have been established (eg.
-by the process of \emph{theory formation})  identified by .
-Theory exploration is similar to \emph{experimental mathematics} process of identifying \emph{interesting} consequences of a given theory.
+Early implementations like \textsc{Theorema} \cite{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
+Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
 
- - Relation to Science
-  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
- - Relation to AI tasks: exploring surroundings, etc.
+Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \cite{Hipster}.
+By removing user interaction, these properties must be implemented by algorithms.
+In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
 
-Theory exploration must take place after \emph{theory formation}
+As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
+The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
+when we already know $0 + x = x$.
+Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
+(version 2), the conjecture generation is performed in a brute-force way.
 
-Major fields include computer algebra and theorem proving.
+Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
 
-- Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
+It uses off-the-shelf ATP to verify the conjectures of \textsc{QuickSpec}.
+\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
+This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}, useful alongside automated test suites.
 
-Computer algebra systems, such as Mathematica, are
-There arithmetic, has a Initially used for ``computation'', ie. evaluation of expressions,
-especially in arithmetic numerical The application to evaluation
+In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems to prove, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
+There are likely lessons that each can learn from the other's approach to term generation.
 
-Theory formation: Alison? Others.
-Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-Communication: Latex, Wikis, APIs, communicating with aliens
+% Theory exploration is similar to \emph{experimental mathematics}
 
-TODO: Motivating theme/narrative. For example, interestingess measures for theory exploration
+% - Relation to Science
+%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
+% - Relation to AI tasks: exploring surroundings, etc.
 
-\section{Examples of Interesting Objects}
+% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+
+% Theory formation: Alison? Others.
+% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
+% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
+% Communication: Latex, Wikis, APIs, communicating with aliens
+
+\section{Exploration in Theorem Proving}
 \label{examples}
 
-Before exploring abstract definitions of interestingness, we can first consider some examples of theorems, lemmas and properties which occur in existing developments, to see what real users considered interesting enough to formalise.
-This will give an empirical basis for our intuitions, which we can compare with the definitions to follow.
+Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
+An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform out later exploration of the literature.
 
-\subsection{Generalisations}
+\subsection{Generalisation}
 
@@
 However, we cannot blindly generalise all obligations we encounter, since \emph{over-generalising} results in obligations which are too strong to prove at all.
-We must therefore rely on heuristics to choose when and how to generalise, which can in part consider how interesting the generalisations are.
+We must therefore rely on heuristics to guide the generation of generalised conjectures, which we can view as a simple form of interestingness.
 
-As an example, we can take Boyer and Moore's proof of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
+A classic example of generalisation occurs when we try to reason about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
+Consider the following addition functions, defined for the Peano naturals \hs{Z :: Nat} and \hs{S :: Nat -> Nat}:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+plus :: Nat -> Nat -> Nat
+plus n  Z    = n
+plus n (S m) = S (plus n m)
+
+plus_tail :: Nat -> Nat -> Nat
+plus_tail n  Z    = n
+plus_tail n (S m) = plus_tail (S n) m
+\end{lstlisting}
+
+Both of these functions implement addition, but the \hs{plus_tail} variant is tail recursive: it does not modify the return value obtained through recursion (in the second clause), and hence we can \emph{replace} the caller's stack frame with the callee's, giving constant space usage.
+rather than pushing a new frame on top.
+Compare this to the use of recursion in \hs{plus}, where the value gets passed to the \hs{S} constructor; this manipulation requires the original stack frame to be available after the recursive call returns, and hence we must push a new frame for each iteration.
+ch recursive call, so that we can pop back to and hence we require the stack frame.
+In practical terms, the recursive call in \hs{plus_tail} can be implemented in constant space, since the we can discard the stack frames
+\fi
+
+As a concrete example, we can take Boyer and Moore's proof of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
 
@@
 This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
+It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
 
-\subsection{Auxilliary Lemmas}
+The ACL2 heuristics which lead to this result are described as
 
-Many of the above generalisations, such as  do not rely on context
+\subsection{Intention}
+
+One consideration when generating conjectures is the difference (if any) between theorems, lemmas, corollaries, etc.
+From a logical point of view, these are all equivalent; it Most proof assistants do not distinguish between these, although some (like Coq) may allow users to annotate their statements with label.
 
@@
 
+TODO: \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
+TODO: \cite{Piantadosi.Tenenbaum.Goodman:2012}
+TODO: \cite{Wille:2005}
+TODO: \cite{colton1999automatic}
+TODO: \cite{colton2000agent}
+TODO: \cite{colton2012automated}
+TODO: \cite{lenat1977automated}
+TODO: \cite{mullerunderstanding}
+TODO: \cite{zzzzz57269a963d945d8964c69b117f24e24c}
+TODO: \cite{Bundy.Cavallo.Dixon.ea:2015}
+TODO: \cite{johansson2009isacosy}
+TODO: \cite{spector2008genetic}
+
+However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+
 % Alan Bundy et al
@@
 
-\subsubsection{Learning on Structured Data}
+The most important property of a logical formula is its truth value.
+Although we may be able to determine some truth values exactly, eg.
+using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
+One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
+This provides a   may be able to determine this truth value, it may not always be even an interesting formula
+TODO: \cite{Hutter.Lloyd.Ng.ea:2013}
+
+\subsubsection{Learning From Structured Data}
+
+One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
+In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
+Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
+Here we review a selection of solutions to this problem, and compare their distinguishing properties.
+
+The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, and it is these models which are
+operate alloe If we for example, a statement like may from the nested form of logical  example, logical terms may be nested to arbitrary depths a large amount of information is represented   to a domain like formal mathematics is the question of how to deal with structure.
+
+TODO: \cite{Gartner2003}
+TODO: \cite{Oveisi.Oveisi.Erfanian.ea:2012}
+TODO: \cite{bakir2007predicting}
+TODO: \cite{conf/ijcai/Plate91}
+TODO: \cite{goller1996learning}
+TODO: \cite{kwasny1995tail}
+TODO: \cite{pollack1990recursive}
+TODO: \cite{zanzotto2012distributed}
+TODO: \cite{zzzzzba6c2484534f9d1105400e6d693a1d3b}
 
@@
 \iffalse
-
 QuickSpec: extend or extinguish?
@@
 
-\iffalse
-
-\begin{abstract}
-We investigate the \textbf{theory exploration} (TE)
-paradigm for computer-assisted Mathematics and identify limitations and
-improvements for current approaches. Unlike the theorem-proving paradigm,
-which requires user-provided conjectures, TE performs an open-ended
-search for theorems satisfying given criteria. We see promise in TE for
-identifying new abstractions and connections in libraries of software
-and proofs, but realising this potential requires more scalable
-algorithms than presently used.
-\end{abstract}
-
-Given a signature $\Sigma$ and a set of variables $V$, we call the pair
-$(\Sigma, V)$ a \emph{theory} and use \emph{theory exploration} (TE) to refer
-to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$
-for producing terms of the theory which are well-formed, provable and satisfy
-some criterion referred to as ``interesting''. These conditions give rise to the
-following questions, which we use to characterise TE systems:
-\begin{description}
-\item [Q1] \label{Q1} How do we generate terms?
-\item [Q2] \label{Q2} How do we guarantee well-formedness?
-\item [Q3] \label{Q3} How do we prove terms?
-\item [Q4] \label{Q4} What is considered ``interesting''?
-\end{description}
-
-Early implementations like \textsc{Theorema} \cite{buchberger2000theory}
-provided interactive environments, similar to computer algebra systems and
-interactive theorem provers, to assist the user in finding theorems. In this
-setting, terms are formed by the user in whichever way they find interesting,
-whilst the software provides support for \textbf{Q2} and \textbf{Q3}.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for
-tasks such as lemma discovery \cite{Hipster}. By removing user interaction,
-\textbf{Q1} and \textbf{Q4} must be solved by algorithms. In existing
-systems these are tightly coupled to improve efficiency, which makes it
-difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about
-Haskell code, which are defined as ``interesting'' if they cannot be simplified
-using previously discovered equations. The intuition for such criteria is to
-avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$. Whilst \textbf{Q4} is elegantly implemented
-with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the term generation for \textbf{Q1} is performed by brute-force.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than
-proving them, it is still used as the exploration component of more rigorous
-systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-In the following, we give an overview of the state of the art in automated
-theory exploration, then present potential improvements and our initial attempts
-at implementation.
-
-\section{Theory Exploration in Haskell}\label{haskell}
-
-Automated theory exploration has been applied to libraries in Isabelle
-and Haskell, although we focus on the latter as its implementations are
-the most mature (demonstrated by the fact that \textsc{Hipster} explores
-Isabelle by first translating it to Haskell). Haskell is interesting to target,
-since its use of pure functions and algebraic datatypes causes many programs to
-follow algebraic laws. However, since Haskell's type system cannot easily
-encode such laws, less effort is given to finding and stating them; compared to
-full theorem provers like Isabelle. Hence we imagine even a shallow exploration
-of code repositories such as \textsc{Hackage} could find many interesting
-theorems.
-
-Currently, the most powerful TE system for Haskell is \textsc{HipSpec}, which
-uses off-the-shelf automated theorem provers (ATPs) to verify the conjectures of
-\textsc{QuickSpec}. \textsc{QuickSpec}, in turn, enumerates all type-correct combinations
-of the terms in the theory up to some depth, groups them into equivalence
-classes using the \textsc{QuickCheck} counterexample finder, then conjectures
-equations relating the members of these classes. This approach works well as a
-lemma generation system, making \textsc{HipSpec} a capable inductive theorem
-prover as well as a theory exploration system \cite{claessen2013automating}.
-
-\section*{Acknowledgements}
-
-Thank you to the \textsc{HipSpec} team at Chalmers University (Moa Johansson,
-Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful
-discussions of these ideas.
-
-\fi
-
 \bibliographystyle{plain}

### 2015-10-27 TransferReport/render.sh

@@ -4,12 +4,17 @@
 NAME=report
 
 run() {
+    echo "RUNNING pdflatex"
     pdflatex -interaction=nonstopmode -halt-on-error --shell-escape "$1"
 }
 
-for EXT in aux bbl blg log out pdf
+for EXT in aux bbl blg dvi log out pdf
 do
     rm -f "${NAME}.${EXT}"
 done
 
-run "$NAME" && bibtex "$NAME" && run "$NAME" && run "$NAME"
+run "$NAME"           &&
+echo "RUNNING bibtex" &&
+bibtex "$NAME"        &&
+run "$NAME"           &&
+run "$NAME"

### 2015-10-28 TransferReport/report.tex
More work

@@
 \textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP to verify the conjectures of \textsc{QuickSpec}.
+It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
 \textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}, useful alongside automated test suites.
+This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}.
+\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
 
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems to prove, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
+In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
 There are likely lessons that each can learn from the other's approach to term generation.
@@
 Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform out later exploration of the literature.
+An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
 
@@
 
+\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+
 When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Proofs by induction often require generalisation, since the specific obligations which arise during the proof may be unsuitable for the inductive hypotheses which are available.
-However, we cannot blindly generalise all obligations we encounter, since \emph{over-generalising} results in obligations which are too strong to prove at all.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, which we can view as a simple form of interestingness.
+Although counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
+This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
+
+However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
+We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
+
+\subsubsection{Tail-recursion}
 
 A classic example of generalisation occurs when we try to reason about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
-Consider the following addition functions, defined for the Peano naturals \hs{Z :: Nat} and \hs{S :: Nat -> Nat}:
+Consider the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Inductive Nat : Set := Z : Nat
+                     | S : Nat -> Nat.
+
+Fixpoint plus      (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => S (plus n' m)
+                                  end.
+
+Fixpoint plus_tail (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => plus_tail n' (S m)
+                                  end.
+\end{lstlisting}
+
+\iffalse
+Haskell equivalent:
+
 plus :: Nat -> Nat -> Nat
@@
 plus_tail n (S m) = plus_tail (S n) m
+\fi
+
+Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive; i.e.
+it can be implemented in constant space using a loop.
+However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+(* Solve equalities by beta-normalising both sides *)
+Ltac triv := try (simpl; reflexivity).
+
+(* Prove equivalence of plus and plus_tail *)
+Theorem equiv : forall n m, plus n m = plus_tail n m.
+  induction n; triv. (* Base case is trivial *)
+
+  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
+  intro m.
+
+  (* Beta-reduce the right-hand-side (justification is trivial) *)
+  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
+
+  (* Use induction hypothesis to replace plus_tail with plus *)
+  rewrite <- (IHn (S m)).
 \end{lstlisting}
 
-Both of these functions implement addition, but the \hs{plus_tail} variant is tail recursive: it does not modify the return value obtained through recursion (in the second clause), and hence we can \emph{replace} the caller's stack frame with the callee's, giving constant space usage.
-rather than pushing a new frame on top.
-Compare this to the use of recursion in \hs{plus}, where the value gets passed to the \hs{S} constructor; this manipulation requires the original stack frame to be available after the recursive call returns, and hence we must push a new frame for each iteration.
-ch recursive call, so that we can pop back to and hence we require the stack frame.
-In practical terms, the recursive call in \hs{plus_tail} can be implemented in constant space, since the we can discard the stack frames
-\fi
+At this point we must prove \coq{plus (S n) m = plus n (S m)}, which seems reasonable.
+However, if we try to proceed by another induction on \coq{n}, we get stuck with incompatible induction hypotheses.
+Specifically, the \emph{conclusion} of our new induction hypothesis is exactly the equation we need:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
+\end{lstlisting}
+
+Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn : forall x, plus (S n') x = plus_tail (S n') x
+\end{lstlisting}
+
+We are forced to abandon the proof, despite such a reasonable-looking intermediate goal. In fact, we \emph{can} prove that goal in isolation:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Lemma gen n m : plus (S n) m = plus n (S m).
+  induction n; triv. (* Base case is trivial *)
+
+  (* Move all S constructors outside *)
+  simpl. rewrite <- IHn. simpl.
+
+  (* Trivial *)
+  reflexivity.
+Defined.
+\end{lstlisting}
+
+We can now resume our proof of \coq{equiv}, which is easily discharged:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  rewrite (gen n m).
+  reflexivity.
+Defined.
+\end{lstlisting}
+
+\subsubsection{Associativity}
 
@@
 
-TODO: https en.wikipedia.org/wiki/Discovery system
-
 Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
@@
 
-TODO: Expand
+% TODO: Expand
 
@@
 The aim of using an ``interestingness'' measure is to force the AC system away from inputs which are not amenable to learning; either because they are simple enough to be completely characterised already, or because they are so complex as to be unintelligible.
-This property is succinctly captured by the \emph{Wundt curve} TODO: Cite.
+This property is succinctly captured by the \emph{Wundt curve} (see figure )
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[scale=0.75]
+    %\node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{/home/chris/DELETEME/screenshot.png}};
+    \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
+    \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
+    \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
+
+    \draw[black,ultra thick] (3, 4)
+         to[out=0,   in=240] (5.05, 6)
+         to[out=60,  in=180] (6.5,  8.2)
+         to[out=0,   in=105] (8.1,  6)
+         to[out=-75, in=120] (8.8,  4)
+         to[out=-60, in=150] (11.4, 0.8);
+
+    % This version is closer, but a little jagged
+    \iffalse
+    \draw[black,ultra thick] (3, 4)
+         to[out=0,   in=240] (5.05, 6)
+         to[out=60,  in=225] (6,    8)
+         to[out=45,  in=180] (6.5,  8.2)
+         to[out=0,   in=135] (7.2,  8)
+         to[out=-45, in=105] (8.1,  6)
+         to[out=-75, in=120] (8.8,  4)
+         to[out=-60, in=150] (11.4, 0.8);
+    \fi
+\end{tikzpicture}
+
+\caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
+The author describes ``hedonic value'' and ``arousal potential'' as terms meant to cover \textquote{both reward value\dots and preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
+
+\label{wundt}
+\end{figure}
+
+% TODO: Cite.
 
@@
 
+\subsubsection{Probability of Sentences}
+
+The most important property of a logical formula is its truth value.
+Although we may be able to determine some truth values exactly, eg.
+using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
+One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
+This provides a   may be able to determine this truth value, it may not always be even an interesting formula
+TODO: \cite{Hutter.Lloyd.Ng.ea:2013}
+
 \subsubsection{Interestingness in Concept Formation}
@@
 TODO: \cite{mullerunderstanding}
-TODO: \cite{zzzzz57269a963d945d8964c69b117f24e24c}
 TODO: \cite{Bundy.Cavallo.Dixon.ea:2015}
@@
 TODO: \cite{spector2008genetic}
+TODO: https en.wikipedia.org/wiki/Discovery system
 
@@
 % Genetic programming? GP in finite algebras
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, eg.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-This provides a   may be able to determine this truth value, it may not always be even an interesting formula
-TODO: \cite{Hutter.Lloyd.Ng.ea:2013}
 
@@
 TODO: \cite{zanzotto2012distributed}
-TODO: \cite{zzzzzba6c2484534f9d1105400e6d693a1d3b}
 
@@
 \bibliographystyle{plain}
-\bibliography{../Bibtex}
+\bibliography{/home/chris/Documents/ArchivedPapers/Bibtex}
 

### 2015-10-29 PhDMeetings/2015-09-16/summer.tex
Additions

\documentclass{article}

\begin{filecontents*}{\jobname.64}
     /-------\
     |Hackage|
     \-------/
         |
         V
   +----------+
   |AST Plugin|
   +----------+
         |    /----\
         |    |    |
         V    V    |
   +------------+  |
   |AST database|  | annotate
   +------------+  |
         |    |    |
         V    \----/
   +------------+
   |Monomorphise|
   +------------+
         |
         V
 +-----------------+
 |Theory definition|
 +-----------------+
         |
         V
+------------------+
|Theory exploration|
+------------------+
\end{filecontents*}

\usepackage{graphicx}
\begin{document}
\title{Summer 2015 Report}

\section{Theory Exploration System}\label{theory-exploration-system}

A lot of my time over summer was spent on building/improving/fixing the wrapper
around QuickSpec which I've dubbed \texttt{ML4HS}. Whilst theoretically
straightforward, peculiarities of Haskell and its infrastructure (GHC,
Cabal, etc.) has lead to several difficulties when trying to build
QuickSpec signatures. The common theme is that it would be useful to
have a function like
\texttt{eval :: [PackageName] -> [ModuleName] -> Expr -> IO (Maybe Expr)}
to (attempt to) evaluate Haskell code in the context of other packages
and modules. This way, we could simply ignore values which lead to
errors.

Until someone implements such a thing, our QuickSpec integration
is instead rather fragile: we must identify and remove any potentially
troublesome values long before they reach QuickSpec, since most of the
errors we encounter are unrecoverable:

\begin{itemize}
\item
  Syntax errors. These can occur when extracting functions from Core,
  since implicit values like type parameters and typeclass dictionaries
  are reified to values with invalid Haskell names (eg.
  \texttt{\$c\textless{}} for an implementation of
  \texttt{<})
\item
  Type errors. These can occur when we try to add functions to a
  signature which do not implement the required typeclasses
  (\texttt{Ord} and \texttt{Typeable}).
\item
  Scope errors. These can occur if we try to use values which are not
  exported.
\item
  GHC panics. These can occur due to mishandling of GHC's API.
\item
  All of the above, at the Template Haskell level (which we use for
  monomorphising)
\end{itemize}

An extensive test suite for every compoenent has helped pin-down and eliminate
most of the sources of errors. The one known bug is that we cannot look up the
names of some packages during compilation: we have their \emph{package key}, but
this is a hash referring to a particular compiled copy on disk; we would rather
have the Hackage name, so we can identify dependencies across packages and
across compilations. The function we currently use to look up names causes a GHC
panic for some packages, although this hasn't been investigated yet.

Conceptually, the project is quite straightforward:

\immediate\write18{ditaa \jobname.64 \jobname-tmp.png}
\fbox{\includegraphics[width=6cm]{\jobname-tmp.png}}

We use a custom GHC plugin to render Core ASTs for top-level values as they are
encountered. These ASTs are labelled with the value's name, its module name and
its package name. This triple acts as a unique ID within the system. Inside the
AST, any references to variables, data constructors or type constructors are
rendered as s-expressions denoting the ID being referenced.

These labelled ASTs are then fed through a series of \emph{annotation} scripts.
Each one appends extra labels to the ASTs, including:

\begin{itemize}
  \item
    The value's type
  \item
    The value's arity (between 0 and 5)
  \item
    The value's \emph{dependencies}, ie. the set of IDs it references.
  \item
    The value's \emph{feature vector}, as determined by one of several implementations.
  \item
    The value's \emph{cluster}, as determined by Weka based on the feature vectors
\end{itemize}

The dependency annotation was implemented by Ouanis, who has also made a simple
script for topologically sorting the ASTs based on their dependencies.

Currently, the \texttt{ML4HSFE} feature extraction script performs a similar
AST-to-matrix transformation as ML4PG, replacing IDs by their cluster as given
by the AST database. If an ID is not found, a default value (eg. ``0'') is used.

To implement recurrent feature extraction, we simply need to perform Ouanis's
sort operation on the database before performing feature extraction, and add extra
calls to the clusterer after each simply-connected-component (ie. mutually-recursive set of functions) has been processed.

\subsection{Going Forward}

The \texttt{ML4HSFE} feature extractor is meant to emulate that of ML4PG (representing the AST nodes in a matrix). We can also use a feature extractor similar to the ACL2 approach (classifying AST nodes based on arities, etc. and using that as the matrix), or similar to Sledgehammer and the ATP work of Josef Urban (counting occurrences of each symbol; I already have a prototype of this).

Moa is also interested in using the output of these tools, so I think it makes sense to release these, eg. to Hackage, sooner rather than later (they are already available from \texttt{chriswarbo.net/git} but may be difficult to install, as Ouanis found).

\section{Reading}\label{reading}

Over the summer I've read many papers, mostly in machine learning, to
better understand the landscape of approaches, their relationships,
history and the current state of the art.

I've also started following Andrew Ng's machine learning lectures from
Stanford.

I've particularly focused on k-means and associated algorithms,
specifically to understand the following:

\begin{itemize}
\item
  How repeatable are its results? \cite{bottou1994convergence}
\item
  How to choose parameters, eg. the number of clusters
  \cite{pelleg2000x} and their initial locations \cite{arthur2007k}?
\item
  How to ensure scalability/tractability? \cite{bahmani2012scalable}
\end{itemize}

I've also kept a few questions in mind regarding recurrent clustering
(RC):

\begin{itemize}
\item
  Can we guarantee convergence of RC?
\item
  How repeatable/replicable is RC (eg. do results vary widely based on
  initial conditions)?
\item
  How does RC compare with in-lining of definitions?
\end{itemize}

As well as AI/ML methods, I've also looked into their application to
Mathematics. From early theorem provers like the Geometry Theorem Prover (which uses counterexamples to eliminate obvious non-theorems before trying to prove them), to exploratory
applications in universal algebra \cite{spector2008genetic}. Other
reading has been ad-hoc, along the lines of either following citation
chains or proceedings (eg. I've been working my way through the
Journal of Machine Learning Research).

\bibliographystyle{plain}
\bibliography{Bibtex}

\end{document}

### 2015-10-29 PhDMeetings/2015-09-16/test.tex
Additions

\documentclass{article}
\begin{filecontents*}{\jobname.64}
JVBERi0xLjUKJbXtrvsKMyAwIG9iago8PCAvTGVuZ3RoIDQgMCBSCiAgIC9GaWx0ZXIgL0ZsYXRl
RGVjb2RlCj4+CnN0cmVhbQp4nJWTO1LEMAyG+5xCNYXQy69j7BFoWIpsAdx/Bjlksx4nZE0VWSP/
3y9Z+ZwIY46SAjRBThRyhq8rvL4RXL8nAlFDTQw3jxTZvAxmkCKohSEKclzOaCF5SV6KZ/iY3l8c
IZSyKTSBFmGvdASBkqFZce0H50BrTdTqsNBW8cKapNpeA66qXfFts6rEGKP9w7zfKJGhCaJokNKb
vw9GyStcaD5kHvfgQ3HzsAW/PXSCz2bfw1btIMH9Pb7JCsmA+YbVvsvZ1DmonzMYJeRUx86p+F4Z
GGfMXHXvGRXBEkPNrLdUGP0zwzSwNuxGPTK/k6nu5ZpQUTSRI9ne3dAL7zkVoGCaUIo2DTUt/gU6
+9U6/xtIJSzkocmdrVNnfAyw6+3ZUg3Ma4ftX25hXKYfzwLt9wplbmRzdHJlYW0KZW5kb2JqCjQg
MCBvYmoKICAgMzA1CmVuZG9iagoyIDAgb2JqCjw8CiAgIC9FeHRHU3RhdGUgPDwKICAgICAgL2Ew
IDw8IC9DQSAxIC9jYSAxID4+CiAgID4+Cj4+CmVuZG9iago1IDAgb2JqCjw8IC9UeXBlIC9QYWdl
CiAgIC9QYXJlbnQgMSAwIFIKICAgL01lZGlhQm94IFsgMCAwIDI5Mi4zODk5MjMgNDM3LjI5MzI0
MyBdCiAgIC9Db250ZW50cyAzIDAgUgogICAvR3JvdXAgPDwKICAgICAgL1R5cGUgL0dyb3VwCiAg
ICAgIC9TIC9UcmFuc3BhcmVuY3kKICAgICAgL0NTIC9EZXZpY2VSR0IKICAgPj4KICAgL1Jlc291
cmNlcyAyIDAgUgo+PgplbmRvYmoKMSAwIG9iago8PCAvVHlwZSAvUGFnZXMKICAgL0tpZHMgWyA1
IDAgUiBdCiAgIC9Db3VudCAxCj4+CmVuZG9iago2IDAgb2JqCjw8IC9DcmVhdG9yIChjYWlybyAx
LjExLjIgKGh0dHA6Ly9jYWlyb2dyYXBoaWNzLm9yZykpCiAgIC9Qcm9kdWNlciAoY2Fpcm8gMS4x
MS4yIChodHRwOi8vY2Fpcm9ncmFwaGljcy5vcmcpKQo+PgplbmRvYmoKNyAwIG9iago8PCAvVHlw
ZSAvQ2F0YWxvZwogICAvUGFnZXMgMSAwIFIKPj4KZW5kb2JqCnhyZWYKMCA4CjAwMDAwMDAwMDAg
NjU1MzUgZiAKMDAwMDAwMDcwNSAwMDAwMCBuIAowMDAwMDAwNDE5IDAwMDAwIG4gCjAwMDAwMDAw
MTUgMDAwMDAgbiAKMDAwMDAwMDM5NyAwMDAwMCBuIAowMDAwMDAwNDkxIDAwMDAwIG4gCjAwMDAw
MDA3NzAgMDAwMDAgbiAKMDAwMDAwMDg5NyAwMDAwMCBuIAp0cmFpbGVyCjw8IC9TaXplIDgKICAg
L1Jvb3QgNyAwIFIKICAgL0luZm8gNiAwIFIKPj4Kc3RhcnR4cmVmCjk0OQolJUVPRgo=
\end{filecontents*}
\usepackage{graphicx}

\begin{document}

\immediate\write18{base64 -d \jobname.64 > \jobname-tmp.pdf}

picture is

\fbox{\includegraphics[width=3cm]{\jobname-tmp.pdf}}

\end{document}

### 2015-10-29 TransferReport/report.tex
Additions

@@
 
-\emph{Artificial curiosity} (AC) is the drive to seek out input or data which is in some sense ``interesting'' \cite{schmidhuber2006developmental}.
-As an unsupervised learning task, AC has no access to labels or meanings associated with its data; the only features it can learn are the structure and relationships between observations and actions.
-The aim of using an ``interestingness'' measure is to force the AC system away from inputs which are not amenable to learning; either because they are simple enough to be completely characterised already, or because they are so complex as to be unintelligible.
-This property is succinctly captured by the \emph{Wundt curve} (see figure )
-
 \begin{figure}
@@
   \begin{tikzpicture}[scale=0.75]
-    %\node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{/home/chris/DELETEME/screenshot.png}};
+      % The image, for reference
+      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
+
+      % Axes
       \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
@@
 
+      % Curve. The numbers come from tracing over wundt.png
       \draw[black,ultra thick] (3, 4)
@@
   \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
-The author describes ``hedonic value'' and ``arousal potential'' as terms meant to cover \textquote{both reward value\dots and preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
+The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
 
@@
 
-% TODO: Cite.
+\emph{Artificial curiosity} (AC) describes learning systems which are rewarded based on the interesting of the input or data they discover \cite{schmidhuber2006developmental}, and hence provides a general mechanism for converting goal-driven algorithms into exploration systems.
+As an unsupervised learning task, AC has no access to labels or meanings associated with this input; the only features it can learn are the structure and relationships inherent in the data.
+The aim of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{(In fact, many measures avoid negative values for simplicity, in which case we can just replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
 
-One common approach to striking this balance is to measure some information-theoretic property of the system or data, and reward the learner accordingly.
-A survey of interestingness measures in artificial curiosity is given in \cite{oudeyer2007intrinsic}.
-Of particular interest is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
+We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm, such that the behaviour follows a Wundt curve.
+
+A framework encompassing many interestingness measures, referred to as \emph{implicit rewards} in the reinforcement learning context, is given in \cite{oudeyer2007intrinsic}.
+
+One particularly general measure is \emph{compression progress}, investigated in  given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
 Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
@@
 
-TODO: Coevolution
+% TODO: Examples
+% TODO: \cite{Schmidhuber1999} Bet
+
+Two sources of intrinsic reward are proposed in \cite{Hester.Stone:2012} for \emph{random forests}.
+A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
+The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply add up the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
+
+\begin{equation}
+  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
+\end{equation}
+
+$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
+For parts of the state space which have been fully learned, the models will agree on accurate predictions.
+For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
+Hence it is the states which are amenable to learning which produce the largest disagreement.
+
+The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to $R_{max}$.
+This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
+
+A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
+we can see which paths are taken through each decision tree).
+% TODO: The accuracy of these random forest models are compared
+
+% TODO: \cite{Kaplan2006}
+% TODO: \cite{Lipson2007}
+% TODO: \cite{Luciw2011}
+% TODO: \cite{Macedo2000}
+% TODO: \cite{Ramik.Sabourin.Madani:2013}
+% TODO: \cite{Roa.Kruijff.Jacobsson:2009}
+% TODO: \cite{Schaul.Sun.Wierstra.ea:2011}
+% TODO: \cite{Schmidhuber1999}
+% TODO: \cite{Schmidhuber:1991}
+% TODO: \cite{Scott1989}
+% TODO: \cite{Steunebrink.Koutnik.Thorisson.ea:2013}
+% TODO: \cite{maher2008achieving}
+% TODO: \cite{meyer1991possibility}
+% TODO: \cite{oudeyer2004intelligent}
+% TODO: \cite{oudeyer2014evolution}
+% TODO: \cite{schmidhuber2006developmental}
+
+\subsubsection{Implicit Artificial Curiosity}
+
+We refer to an AC system as being \emph{implicit} when the \emph{outputs} of multiple learning algorithms are combined, to produce Wundt-like behaviour in the overall system. % TODO: Examples
+
+% TODO: Coevolution
 

### 2015-10-29 Balmoral2015/render.sh

@@ -1,4 +1,5 @@
-#!/bin/sh
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p bash haskellPackages.Agda
 # --self-contained
 HTML_ARGS="-w dzslides --standalone --filter panpipe --filter panhandle"
 

### 2015-10-29 Chalmers2015-03-25/render.sh

@@ -1,3 +1,4 @@
-#!/bin/sh
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p pandoc panpipe panhandle
 
-nix-shell --pure --command "md2pdf beamer"
+pandoc -t beamer -o slides.pdf slides.md

### 2015-10-29 PhDMeetings/2015-09-16/render.sh

@@ -1,0 +1,5 @@
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p ditaa bash
+pdflatex --shell-escape summer
+bibtex summer
+pdflatex --shell-escape summer

### 2015-10-29 TransferReport/report.tex
Rename overlapping key

@@
 
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm, such that the behaviour follows a Wundt curve.
+We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system still follows a Wundt curve as an emergent property.
 
@@
 \bibliographystyle{plain}
-\bibliography{/home/chris/Documents/ArchivedPapers/Bibtex}
+\bibliography{../Bibtex}
 

### 2015-10-30 TransferReport/reverse.sh
Added reverse.sh

@@ -1,0 +1,39 @@
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p xidel bash
+
+# Find the most popular Haskell packages which use property testing
+# We find the reverse dependencies of some property checkers, find the most
+# downloaded Haskell packages, and take the intersection
+
+function reverse_deps {
+    xidel --extract '//table//td[not(@class="version")]' "http://packdeps.haskellers.com/reverse/$1"
+}
+
+function check_deps {
+    for PKG in QuickCheck smallcheck lazysmallcheck sparsecheck
+    do
+        reverse_deps "$PKG"
+    done
+}
+
+function top_packages {
+    xidel --extract '//table//tr/td/a/text()' "http://hackage.haskell.org/packages/top" 2> /dev/null
+}
+
+function top_tests {
+    DEPS=$(check_deps)
+    while read -r PKG
+    do
+        [[ -z "$PKG" ]] && continue
+        while read -r DEP
+        do
+            [[ "x$PKG" = "x$DEP" ]] && {
+                echo "$PKG"
+                break
+            }
+        done < <(echo "$DEPS")
+    done < <(top_packages)
+}
+
+# Take the top 10
+head -n10 < <(top_tests)

### 2015-10-30 TransferReport/report.tex
Imported feature extraction writeup

@@
 \usepackage{listings}
+
+% Treat paragraph as subsubsubsection
+\usepackage{titlesec}
+\setcounter{secnumdepth}{4}
+
 \usepackage{tikz}
 \usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
+
+\DeclareMathOperator{\md5}{MD5}
 
@@
 Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to a \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \cite{marlow2010haskell} \footnote{Version 2010 was current as this was written}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
+Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
+For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \cite{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \cite{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
 
@@
 
-\subsection{Intention}
+\subsection{Lemmas}
 
-One consideration when generating conjectures is the difference (if any) between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent; it Most proof assistants do not distinguish between these, although some (like Coq) may allow users to annotate their statements with label.
+One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
+From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
+However, their \emph{intention} may be different, and hence some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
+despite their internal representations being the same.
 
-\subsection{QuickCheck Properties}
+One striking observation is that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}
+
+\subsection{Tests}
+
+Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
+For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
+Whilst many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
+One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
+Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \cite{claessen2011quickcheck} and \textsc{SmartCheck} \cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \cite{reich2013advances}) and logic programming (\textsc{SparseCheck} \cite{sparsecheck}).
+
+Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
+
+Here we consider 10 most-downloaded packages from hackage.haskell.org (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
 
@@
 % TODO: Examples
-% TODO: \cite{Schmidhuber1999} Bet
+\cite{Schmidhuber1999}
 
@@
 
-% TODO: \cite{Kaplan2006}
-% TODO: \cite{Lipson2007}
-% TODO: \cite{Luciw2011}
-% TODO: \cite{Macedo2000}
-% TODO: \cite{Ramik.Sabourin.Madani:2013}
-% TODO: \cite{Roa.Kruijff.Jacobsson:2009}
-% TODO: \cite{Schaul.Sun.Wierstra.ea:2011}
-% TODO: \cite{Schmidhuber1999}
-% TODO: \cite{Schmidhuber:1991}
-% TODO: \cite{Scott1989}
-% TODO: \cite{Steunebrink.Koutnik.Thorisson.ea:2013}
-% TODO: \cite{maher2008achieving}
-% TODO: \cite{meyer1991possibility}
-% TODO: \cite{oudeyer2004intelligent}
-% TODO: \cite{oudeyer2014evolution}
-% TODO: \cite{schmidhuber2006developmental}
+% TODO
+\cite{Kaplan2006}
+\cite{Lipson2007}
+\cite{Luciw2011}
+\cite{Macedo2000}
+\cite{Ramik.Sabourin.Madani:2013}
+\cite{Roa.Kruijff.Jacobsson:2009}
+\cite{Schaul.Sun.Wierstra.ea:2011}
+\cite{Schmidhuber1999}
+\cite{Schmidhuber:1991}
+\cite{Scott1989}
+\cite{Steunebrink.Koutnik.Thorisson.ea:2013}
+\cite{maher2008achieving}
+\cite{meyer1991possibility}
+\cite{oudeyer2004intelligent}
+\cite{oudeyer2014evolution}
+\cite{schmidhuber2006developmental}
 
@@
 
-The combinatorial nature of formal systems leads many algorithms in this domain, such as proof search methods, to have exponential time complexityCITE; hence even a modest size increase can turn a trivial problem into an intractable one.
-It is currently infeasible to improve the scaling behaviour of many such algorithms; for example, problems like TODO are in the infamous NP-complete complexity classCITE.
+The combinatorial nature of formal systems causes many proof search methods such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
+Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
+determining satisfiability) or co-NP-complete (e.g.
+determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
 On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
@@
 
-Relevance filtering simplifies a proof search problem by removing from consideration those axioms, definitions, lemmas, etc.
-which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only those lemmas which are deemed ``relevant'' (required for proving the statement) are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a lemma is relevant: how do we know what will be required, before we have the proof?
+Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
+The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
+This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
 
-Sledgehammer's approach is to use \emph{supervised learning} to approximate the relevance relation: a naive bayes classifier is trained on TODO
+The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
+There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
+given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
+To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
 
-ML4PG
-ACL2(ml)
-Josef Urban
-Sledgehammer
-Naive Bayes
+Recently, a variety of alternative algorithms have also been investigated, including:
+
+\begin{description}
+
+  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \cite{kuhlwein2013mash}.
+The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
+To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
+
+  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathamatical Library for use by the Vampire ATP system \cite{alama2014premise}.
+\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
+
+  % TODO:
+  \item{\textsc{SInE}}
+  \item{\textsc{BliStr}}
+  \item{\textsc{HOLyHammer}}
+  \item{\textsc{MoMM}}
+  \item{\textsc{SNoW}}
+  \item{\textsc{MPTP 0.2}}
+  \item{\textsc{MaLARea}}
+  \item{\textsc{MaLARea SG1}}
+
+\end{description}
+
+\subsubsection{Clustering}
+% TODO: ML4PG
+% TODO: ACL2(ml)
 
@@
 One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-This provides a   may be able to determine this truth value, it may not always be even an interesting formula
-TODO: \cite{Hutter.Lloyd.Ng.ea:2013}
+
+% TODO
+\cite{Hutter.Lloyd.Ng.ea:2013}
 
@@
 
-TODO: \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
-TODO: \cite{Piantadosi.Tenenbaum.Goodman:2012}
-TODO: \cite{Wille:2005}
-TODO: \cite{colton1999automatic}
-TODO: \cite{colton2000agent}
-TODO: \cite{colton2012automated}
-TODO: \cite{lenat1977automated}
-TODO: \cite{mullerunderstanding}
-TODO: \cite{Bundy.Cavallo.Dixon.ea:2015}
-TODO: \cite{johansson2009isacosy}
-TODO: \cite{spector2008genetic}
-TODO: https en.wikipedia.org/wiki/Discovery system
+% TODO:
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}
+\cite{Piantadosi.Tenenbaum.Goodman:2012}
+\cite{Wille:2005}
+\cite{colton1999automatic}
+\cite{colton2000agent}
+\cite{colton2012automated}
+\cite{lenat1977automated}
+\cite{mullerunderstanding}
+\cite{Bundy.Cavallo.Dixon.ea:2015}
+\cite{johansson2009isacosy}
+\cite{spector2008genetic}
+\cite{colton2012automated}
+% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
 
-However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
 
@@
 % Eurisko, AM, etc.?
-
-% Genetic programming? GP in finite algebras
 
@@
 
-The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, and it is these models which are
-operate alloe If we for example, a statement like may from the nested form of logical  example, logical terms may be nested to arbitrary depths a large amount of information is represented   to a domain like formal mathematics is the question of how to deal with structure.
+\paragraph{Feature Extraction}
 
-TODO: \cite{Gartner2003}
-TODO: \cite{Oveisi.Oveisi.Erfanian.ea:2012}
-TODO: \cite{bakir2007predicting}
-TODO: \cite{conf/ijcai/Plate91}
-TODO: \cite{goller1996learning}
-TODO: \cite{kwasny1995tail}
-TODO: \cite{pollack1990recursive}
-TODO: \cite{zanzotto2012distributed}
+\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
+Rather than feeding "raw" data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
+This has two benefits:
+
+\begin{itemize}
+  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
+This reduces the size of the ML problem, improving efficiency (eg.
+running time).
+  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
+\end{itemize}
+
+Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
+Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (eg.
+\emph{artificial neural networks} with fixed connections between nodes).
+This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
+
+For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
+Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
+In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
+
+\paragraph{Significance}
+
+As an example, say we want to learn relationships between the following program fragments:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+data Maybe a = Nothing | Just a
+
+data Either a b = Left a | Right b
+\end{lstlisting}
+
+We might hope our algorithm discovers relationships like:
+
+\begin{itemize}
+  \item Both are valid Haskell code.
+  \item Both describe datatypes.
+  \item Both datatypes have two constructors.
+  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
+  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
+  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
+  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
+\end{itemize}
+
+However, this is too optimistic.
+Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
+Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
+
+\paragraph{Practical Considerations}
+
+In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
+Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
+
+There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
+Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
+\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
+If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
+
+Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
+This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
+
+Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
+This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
+
+\paragraph{Truncation and Padding}
+
+The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
+This is the approach taken by ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
+
+Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
+Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
+The less these assumptions hold, the less we can truncate.
+Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
+
+\paragraph{Dimension Reduction}
+
+A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
+
+Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
+More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
+
+There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
+
+\begin{itemize}
+  \item They operate *offline*, requiring all input points up-front
+  \item All input points must have the same dimensionality
+\end{itemize}
+
+In particular, the second constraint is precisely what we're trying to avoid.
+Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
+
+The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features.
+Such systems are possible, but out of scope for this small initial implementation.
+
+\paragraph{Structured Data}
+
+The task of dimension reduction changes when we consider \emph{structured} data.
+Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
+For data mining context-free languages (eg.
+those of programming and theorem-proving systems), we will mainly be concerned with tree structures of the following form (where \hs{FV} denotes a feature vector):
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+data Tree = Leaf FV
+          | Branch FV Tree Tree
+\end{lstlisting}
+
+\paragraph{Truncation}
+
+Just like the non-recursive case, the simplest way to reduce our dimensionality is to truncate. To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceD 1 (Branch fv l r) = Leaf   fv
+reduceD n (Leaf   fv)     = Leaf   fv
+reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
+                                      (reduceD (n-1) r)
+\end{lstlisting}
+
+This is simple, but suffers two problems:
+
+\begin{itemize}
+ \item We must choose a conservatively large \hs{d}, since we're throwing away information
+ \item The number of feature vectors grows exponentially as \hs{d} increases
+\end{itemize}
+
+The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
+One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}.
+We can do this with \hs{tabulate w d t}:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+tabulate w d t = take d (map (take w) (tabulate' t))
+
+tabulate' (Leaf   fv)     = [fv] : []
+tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
+
+merge    []     ys  = ys
+merge    xs     []  = xs
+merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
+\end{lstlisting}
+
+\paragraph{Extracting Features from Haskell}
+
+The Glasgow Haskell Compiler provides a library interface to its operation, which we can use to parse Haskell code into tree structures.
+It also provides a \emph{renaming} transformation, which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+
+\paragraph{Linear Combinations}
+
+Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
+Following \cite{zanzotto2012distributed} we can use \emph{circular convolution} (\hs{cc}) as our combining function.
+Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceC (Leaf fv)       = fv
+reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
+\end{lstlisting}
+
+Circular convolution has the following desirable properties:
+
+\begin{description}
+
+  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
+
+  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
+
+\end{description}
+
+Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
+This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
+In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \cite{plate1991holographic}.
+
+\paragraph{Extracting Features from XML}
+
+This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
+The application parses arbitrary trees of XML and uses binary features, ie.
+feature vectors are bit vectors.
+
+To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
+
+\begin{equation}
+feature(n) = 2^{\md5(s) \pmod{L}}
+\end{equation}
+
+Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
+The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
+Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
+
+In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
+
+\paragraph{Application to Coq}
+
+Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
+We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
+
+Coq will also interpret the standard output of the command, to generate terms and tactics.
+This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing for powerful ML approaches like \emph{reinforcement learning}.
+
+\paragraph{Sequencing}
+
+Any investigation of variable-size input would be incomplete without mentioning the \emph{sequencing}.
+This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
+The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
+This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
+
+Not all ML algorithms can be adapted to accept sequences.
+One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
+Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
+
+The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
+If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
+However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
+The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \cite{hochreiter1997long}).
+
+The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
+
+% TODO
+\cite{Gartner2003}
+\cite{Oveisi.Oveisi.Erfanian.ea:2012}
+\cite{bakir2007predicting}
+\cite{conf/ijcai/Plate91}
+\cite{goller1996learning}
+\cite{kwasny1995tail}
+\cite{pollack1990recursive}
+\cite{zanzotto2012distributed}
+
+\iffalse
 
@@
 These mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
+\fi
 

### 2015-10-30 TransferReport/report.tex
Made text readable

@@
 Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on its contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
+For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
 Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
@@
 We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{relatedwork}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are also experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to principled theories to emerge, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
+A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
+We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
+Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
 
@@
 This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell, as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \cite{QuickSpec} and \textsc{HipSpec} \cite{claessen2013automating}).
-This is evident even to users of Isabelle/HOL, which boasts its own \textsc{IsaCoSy} \cite{johansson2009isacosy} and \textsc{Hipster} \cite{Hipster} theory exploration systems, since \textsc{Hipster} is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
+We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \cite{QuickSpec} and \textsc{HipSpec} \cite{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
 
@@
 in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell in thus in a unique position with regards to the discovery of interesting theorems.
+We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
 Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
@@
 Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof} \cite{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \cite{nipkow2002isabelle} and Coq \cite{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of our confidence in the results.
+This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
+One nice consequence of this approach, which is followed for example by Isabelle \cite{nipkow2002isabelle} and Coq \cite{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
 
@@
 In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon has developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a form amenable to first-order reasoning.
+Even by that time Newell and Simon had developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
+The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
 
@@
 There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as ``interesting'' has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
+The choice of what counts as interesting has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
 
@@
 When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
+Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
 This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
@@
 Consider the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
+
+TODO: This is more of an auxiliary lemma than a generalisation
 
@@
 it can be implemented in constant space using a loop.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties:
+However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
+In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
+
+\iffalse
 
@@
 
-At this point we must prove \coq{plus (S n) m = plus n (S m)}, which seems reasonable.
-However, if we try to proceed by another induction on \coq{n}, we get stuck with incompatible induction hypotheses.
-Specifically, the \emph{conclusion} of our new induction hypothesis is exactly the equation we need:
+\fi
+
+Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
 
@@
 
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal. In fact, we \emph{can} prove that goal in isolation:
+We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
+
+In fact, if we attempt to prove that goal \emph{in isolation}, we can make the proof go through; even though it is stronger due to the absence of the \coq{IHn} assumption.
+We can then specialise this lemma to make our original proof of equivalence go through.
+
+\iffalse
 
@@
 \end{lstlisting}
-
-We can now resume our proof of \coq{equiv}, which is easily discharged:
 
@@
 
+\fi
+
 \subsubsection{Associativity}
 
-As a concrete example, we can take Boyer and Moore's proof of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
+An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
 
@@
 
-The ACL2 heuristics which lead to this result are described as
-
-\subsection{Lemmas}
+% TODO: Describe the ACL2 heuristics
+
+\subsection{Auxiliary Lemmas}
 
@@
 
-One striking observation is that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}
+One striking observation is that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
+
+% TODO: Analyse them
+
+% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
 
@@
 For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Whilst many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
+Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
 One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
@@
 
-Here we consider 10 most-downloaded packages from hackage.haskell.org (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
+Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
+
+% TODO: Analysis
 
@@
 
+\iffalse
+
 \subsection{Conjecture Generation}
@@
 \subsubsection{Lemma generation}
+
+\fi
 
@@
 On the other hand, work on deceptive problems is highly relevant, as it has lead to studying \emph{exploration in general}, and hence provides insight for our narrower domain of theory exploration.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (eg.
+Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
 by hill climbing), but few global optima which are hard to find.
@@
 Once we augment fitness functions, to a greater or lesser degree, an obvious question arises: what if the augmented function contains nothing of the original fitness?
-This is the pure exploration scenario we are interested in, which has been found to provide useful ``stepping stones'', even in objective-driven domains \cite{lehman2011abandoning}.
+This is the pure exploration scenario we are interested in, which has been found to provide useful ``stepping stones'', even in objective-driven domains \cite{lehman2011abandoning} \cite{hutter2002fitness}.
 
@@
 
-\emph{Artificial curiosity} (AC) describes learning systems which are rewarded based on the interesting of the input or data they discover \cite{schmidhuber2006developmental}, and hence provides a general mechanism for converting goal-driven algorithms into exploration systems.
+\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}, and hence provides a general mechanism for converting goal-driven algorithms into exploration systems.
 As an unsupervised learning task, AC has no access to labels or meanings associated with this input; the only features it can learn are the structure and relationships inherent in the data.
 The aim of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{(In fact, many measures avoid negative values for simplicity, in which case we can just replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system still follows a Wundt curve as an emergent property.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
+
+We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
 
@@
 
-One particularly general measure is \emph{compression progress}, investigated in  given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
+One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
 Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, new theorems and we select for useful generalisations which allow old special-case lemmas to be discarded.
+This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
 
@@
 A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply add up the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
+The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
 
@@
 
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to $R_{max}$.
+The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
 This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
@@
 
-\subsubsection{Implicit Artificial Curiosity}
-
-We refer to an AC system as being \emph{implicit} when the \emph{outputs} of multiple learning algorithms are combined, to produce Wundt-like behaviour in the overall system. % TODO: Examples
-
 % TODO: Coevolution
-
-Approaches to artificial curiosity mostly follow two themes: the definition of an explicit ``interestingness'' measure, from which rewards are calculated; or else the use of \emph{co-evolution} to \emph{emergent}.
 
@@
 However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-In this sense, the
 
@@
 Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of so-called \emph{Good Old Fashioned AI} (GOFAI), of which theorem proving is a classic example.
-In this sense, the fields of theory exploration and developmental robotics seem rather disconnected.
-Nevertheless, TE still exhibits a form of embodiment in two ways.
-Firstly, the abstract, mathematical domain being explored is not a model of some external, physical environment; the domain \emph{is} the environment; in other words, TE does not suffer from the same grounding problem as GOFAI.
-In particular, \emph{resource usage} is a critical factor in any search problem; if it weren't, then brute force would be a viable solution.
+Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
+Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
+
+\begin{itemize}
+
+  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
+
+  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
+If it weren't, then brute force enumeration of proofs would be a viable solution.
+In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
+
+\end{itemize}
+
+\iffalse
 
@@
 
- takes the heavily automated field of theorem proving   mbodies The attempt to automate earlier transition from goal-driven problem solving to goal-less exploration and problem invention is not unique to theorem proving; it has also been investigated occurred in  and exploration which occur in theorem proving and TE have analogues in the domains of artificial intelligence (AI) and machine learning (ML).
-In  in an commonly appearing in the form of the \emph{exploration vs exploitation problem} in goal-driven systems.
+\fi
 
@@
 
-The combinatorial nature of formal systems causes many proof search methods such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
+The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
 Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
@@
 
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathamatical Library for use by the Vampire ATP system \cite{alama2014premise}.
+  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
 \iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
@@
 The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, eg.
+Although we may be able to determine some truth values exactly, e.g.
 using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
@@
 \emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding "raw" data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
+Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
 This has two benefits:
@@
   \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (eg.
+This reduces the size of the ML problem, improving efficiency (e.g.
 running time).
@@
 Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (eg.
+Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
 \emph{artificial neural networks} with fixed connections between nodes).
@@
 In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-\paragraph{Significance}
 
@@
 
-\paragraph{Practical Considerations}
-
 In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
@@
 \begin{itemize}
-  \item They operate *offline*, requiring all input points up-front
+  \item They operate \emph{offline}, requiring all input points up-front
   \item All input points must have the same dimensionality
@@
 
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features.
-Such systems are possible, but out of scope for this small initial implementation.
-
-\paragraph{Structured Data}
+The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
+
+\paragraph{Sequencing}
 
@@
 Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (eg.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-\paragraph{Truncation}
-
-Just like the non-recursive case, the simplest way to reduce our dimensionality is to truncate. To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}.
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\paragraph{Extracting Features from Haskell}
-
-The Glasgow Haskell Compiler provides a library interface to its operation, which we can use to parse Haskell code into tree structures.
-It also provides a \emph{renaming} transformation, which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-
-\paragraph{Linear Combinations}
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \cite{zanzotto2012distributed} we can use \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \cite{plate1991holographic}.
-
-\paragraph{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\paragraph{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing for powerful ML approaches like \emph{reinforcement learning}.
-
-\paragraph{Sequencing}
-
-Any investigation of variable-size input would be incomplete without mentioning the \emph{sequencing}.
+For data mining context-free languages (e.g.
+those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
+
+Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
 This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
@@
 
-One way to apply fixed-size machine learning algorithms to arbitrarily-sized recursive structures is to use a \emph{distributed representation}.
-These mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
 \fi
 
-\section{Contributions}
-\label{current}
-
-\subsection{The \textsc{ML4HS} Framework}\label{ml4hs}
-
-We consider \textbf{Q2} and \textbf{Q3} to be adequately solved by the existing
-use of type systems and ATPs, respectively. We identify the following potential
-improvements for the other questions:
+\section{Contributions}\label{current}
+
+\subsection{Circular Convolution}\label{circularconvolution}
+
+We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+data Tree = Leaf FV
+          | Branch FV Tree Tree
+\end{lstlisting}
+
+Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
+To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceD 1 (Branch fv l r) = Leaf   fv
+reduceD n (Leaf   fv)     = Leaf   fv
+reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
+                                      (reduceD (n-1) r)
+\end{lstlisting}
+
+This is simple, but suffers two problems:
+
+\begin{itemize}
+ \item We must choose a conservatively large \hs{d}, since we're throwing away information
+ \item The number of feature vectors grows exponentially as \hs{d} increases
+\end{itemize}
+
+The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
+One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
+
+\iffalse
+
+We can do this with \hs{tabulate w d t}:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+tabulate w d t = take d (map (take w) (tabulate' t))
+
+tabulate' (Leaf   fv)     = [fv] : []
+tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
+
+merge    []     ys  = ys
+merge    xs     []  = xs
+merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
+\end{lstlisting}
+
+\fi
+
+Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
+Following \cite{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
+Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceC (Leaf fv)       = fv
+reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
+\end{lstlisting}
+
+Circular convolution has the following desirable properties:
 
 \begin{description}
-\item [Q1]
+
+  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
+
+  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
+
+\end{description}
+
+Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
+This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
+In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \cite{conf/ijcai/Plate91}.
+
+Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
+
+\subsubsection{Extracting Features from XML}
+
+This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
+The application parses arbitrary trees of XML and uses binary features, ie.
+feature vectors are bit vectors.
+
+To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
+
+\begin{equation}
+feature(n) = 2^{\md5(s) \pmod{L}}
+\end{equation}
+
+Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
+The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
+Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
+
+In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
+
+\subsubsection{Application to Coq}
+
+Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
+We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
+
+Coq will also interpret the standard output of the command, to generate terms and tactics.
+This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
+
+\subsection{\texttt{nix-eval}}\label{nixeval}
+
+\subsection{\textsc{ASTPlugin}}\label{astplugin}
+
+The Glasgow Haskell Compiler provides a library interface to its operation, which we can use to parse Haskell code into tree structures.
+It also provides a \emph{renaming} transformation, which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+
+\subsection{\textsc{MLSpec}}\label{mlspec}
+
+\subsection{\textsc{ML4HS}}\label{ml4hs}
+
+We identify the following potential improvements for existing TE systems such as \textsc{HipSpec}:
+
+\begin{description}
+\item{Term generation}:
   Enumerating all type-correct terms is a brute-force solution to this question.
@@
   theory, to prioritise over the rest.
-\item [Q4]
-  Various alternative ``interestingness'' criteria have been proposed, for
+\item{Interestingness}
+  Various alternative interestingness criteria have been proposed, for
   example those surveyed in \cite{geng2006interestingness}. Augmenting or

### 2015-10-30 TransferReport/report.tex
Expanded examples

@@
 
-\title{Machine Learning Methods in Functional Programming and Interactive Theorem Proving}
+\title{Interestingness Measures for Theory Exploration}
 
@@
 
-\subsubsection{Tail-recursion}
-
-A classic example of generalisation occurs when we try to reason about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
-Consider the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-TODO: This is more of an auxiliary lemma than a generalisation
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive; i.e.
-it can be implemented in constant space using a loop.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{in isolation}, we can make the proof go through; even though it is stronger due to the absence of the \coq{IHn} assumption.
-We can then specialise this lemma to make our original proof of equivalence go through.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsubsection{Associativity}
-
 An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
@@
 
-\subsection{Auxiliary Lemmas}
+Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
+\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
+See \S \ref{auxiliarylemmas} for example definitions of each type.}
+
+\subsection{Analogy}
+
+One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
+By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
+
+Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
+If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
+
+The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \cite{Heras.Komendantskaya.Johansson.ea:2013}.
+These clusters are used in two ways:
+
+\begin{itemize}
+
+  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
+
+  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
+Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
+
+\end{itemize}
+
+The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
+
+\begin{itemize}
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
+
+  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
+
+\end{itemize}
+
+\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
 
@@
 From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different, and hence some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
+However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
+
+Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
 despite their internal representations being the same.
-
-One striking observation is that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
+This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
 
@@
 % TODO: Theory exploration as lemma generation; give example from a HipSpec paper
+
+We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
+Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Inductive Nat : Set := Z : Nat
+                     | S : Nat -> Nat.
+
+Fixpoint plus      (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => S (plus n' m)
+                                  end.
+
+Fixpoint plus_tail (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => plus_tail n' (S m)
+                                  end.
+\end{lstlisting}
+
+\iffalse
+Haskell equivalent:
+
+plus :: Nat -> Nat -> Nat
+plus      n  Z    = n
+plus      n (S m) = S (plus n m)
+
+plus_tail :: Nat -> Nat -> Nat
+plus_tail n  Z    = n
+plus_tail n (S m) = plus_tail (S n) m
+\fi
+
+Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
+However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
+In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
+
+\iffalse
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+(* Solve equalities by beta-normalising both sides *)
+Ltac triv := try (simpl; reflexivity).
+
+(* Prove equivalence of plus and plus_tail *)
+Theorem equiv : forall n m, plus n m = plus_tail n m.
+  induction n; triv. (* Base case is trivial *)
+
+  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
+  intro m.
+
+  (* Beta-reduce the right-hand-side (justification is trivial) *)
+  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
+
+  (* Use induction hypothesis to replace plus_tail with plus *)
+  rewrite <- (IHn (S m)).
+\end{lstlisting}
+
+\fi
+
+Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
+\end{lstlisting}
+
+Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn : forall x, plus (S n') x = plus_tail (S n') x
+\end{lstlisting}
+
+We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
+
+In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
+Using this separate result as a lemma, the pointwise equality is proven easily.
+
+\iffalse
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Lemma gen n m : plus (S n) m = plus n (S m).
+  induction n; triv. (* Base case is trivial *)
+
+  (* Move all S constructors outside *)
+  simpl. rewrite <- IHn. simpl.
+
+  (* Trivial *)
+  reflexivity.
+Defined.
+\end{lstlisting}
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  rewrite (gen n m).
+  reflexivity.
+Defined.
+\end{lstlisting}
+
+\fi
 

### 2015-11-02 TransferReport/report.tex
Current work

@@
 \usepackage{amssymb}
+\usepackage[T1]{fontenc}
 %\usepackage{enumitem}
@@
 \usepackage{listings}
+\usepackage{fixltx2e}
 
@@
 
+\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
+When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
+This mechanism allows modules to provide more guarantees than are available just in their types.
+For example, a module may represent email addresses in the following way:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
+module Email (Email(), at, render) where
+
+data Email = E String String
+
+render :: Email -> String
+render (E u h) = u ++ "@" ++ h
+
+at :: String -> String -> Maybe Email
+at "" h  = Nothing
+at u  "" = Nothing
+at u  h  = Just (E u h)
+\end{lstlisting}
+
+The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
+We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
+By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
+
 \end{description}
@@
 
-\subsection{Tests}
+\subsection{Tests} \label{tests}
 
@@
 
-A framework encompassing many interestingness measures, referred to as \emph{implicit rewards} in the reinforcement learning context, is given in \cite{oudeyer2007intrinsic}.
+A framework encompassing many explicit interestingness measures is given in \cite{oudeyer2007intrinsic}, as \emph{implicit rewards} in the context of reinforcement learning.
 
@@
 
-\paragraph{Feature Extraction}
+\paragraph{Feature Extraction}\label{featureextraction}
 
@@
 
+Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
+
 \subsection{Circular Convolution}\label{circularconvolution}
@@
 
+To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
+This is difficult, as many terms are functions, which Haskell does not let us inspect.
+This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
+To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
+
+\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
+In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
+
+On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
+Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
+Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
+
+We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
+These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
+
+\texttt{nix-eval} is build on top of the Nix package manager and GHC.
+Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
+
+The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
+When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
+GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
+The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
+The expression, therefore, must have type \hs{String}; although this is not enforced statically.
+
+The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
+The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
+The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
+It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
+using JSON.
+
 \subsection{\textsc{ASTPlugin}}\label{astplugin}
 
-The Glasgow Haskell Compiler provides a library interface to its operation, which we can use to parse Haskell code into tree structures.
-It also provides a \emph{renaming} transformation, which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
+
+Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
+Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
+However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
+
+This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
+All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
+
+Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
+All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
+This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
+
+GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
+We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
+
+There are several complications in the implementation of \textsc{AstPlugin}.
+We are working with Core, which as a variant of System F\textsubscript{C} \cite{sulzmann2007system} has slightly different syntax and semantics to Haskell.
+This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
+Three areas which make Core more difficult to handle are:
+
+\begin{description}
+  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
+In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
+Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
+This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
+
+  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
+Since Core does not make these distinctions, names may become ambiguous.
+For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
+To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
+This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
+
+  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
+When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
+Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
+\end{description}
+
+In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
+Those which fail, due to the reasons above and others, are discarded.
+
+The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
+Definitions are indexed by a combination of their package name, module name and binding name.
+The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
+Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
+
+Such data is useful for multiple purposes:
+
+\begin{itemize}
+  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
+those (package, module, name) combinations which it references.
+Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
+
+  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
+As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
+
+  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
+
+  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
+
+  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
+Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for all of the types in our theory; yet existing approaches to deriving such generators (e.g.
+as found in the \texttt{derive} package) rely on choose between constructors, recursively.
+As demonstrated in \S \ref{haskell}, a type's constructors may not be exported, in which case these methods fail; not to mention the exponential complexity of such recursive branching.
+It seems that a database of inspectable Haskell functions may be useful for solving this problem, although we have not yet performed experiments in this area.
+\end{itemize}
 
 \subsection{\textsc{MLSpec}}\label{mlspec}
+
+When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
+One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
+
+Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
+Features of \textsc{MLSpec} include:
+
+\begin{itemize}
+  \item{Monomorphising}: Given values of polymorphic type, e.g.
+\hs{safeHead :: forall t.
+[t] -> Maybe t} and \hs{[] :: forall t.
+[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
+Maybe t} without instantiating the variable \hs{t} to a specific type.
+Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
+We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
+\hs{forall c.
+c Bool -> c Bool}) or incompatible class constraints (e.g.
+\hs{forall t.
+IsString t => t}).
+
+  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
+There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
+
+  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
+
+  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
+
+\end{itemize}
 

### 2015-11-02 TransferReport/report.tex
Updates

@@
 
+We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
+
+\begin{description}
+\item{Term generation}:
+  Enumerating all type-correct terms is a brute-force solution to this question.
+Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
+In particular, heuristic search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
+  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
+
+\item{Interestingness}
+  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
+Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
+\end{description}
+
 % Theory exploration is similar to \emph{experimental mathematics}
@@
 
-\subsubsection{Relevance}
+\subsubsection{Relevance Filtering}
+\label{relevance}
+
+% TODO
+\cite{kuhlwein2012overview}
 
@@
 \subsubsection{Clustering}
+\label{clustering}
+
 % TODO: ML4PG
-% TODO: ACL2(ml)
+\cite{journals/corr/abs-1212-3618}
+% TODO: ACL2(ml) (also examples section)
+\cite{heras2013proof}
 
@@
 \cite{colton2012automated}
+ \cite{geng2006interestingness}
 % TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
@@
   \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for all of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choose between constructors, recursively.
-As demonstrated in \S \ref{haskell}, a type's constructors may not be exported, in which case these methods fail; not to mention the exponential complexity of such recursive branching.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem, although we have not yet performed experiments in this area.
+Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
+as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
+As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
+Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
+It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
 \end{itemize}
@@
 
-We identify the following potential improvements for existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-  Scalable alternatives to brute-force algorithms are a well-studied area of
-  Artificial Intelligence and Machine Learning. In particular, heuristic
-  search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
-  We could also use Machine Learning methods to identify some sub-set of a given
-  theory, to prioritise over the rest.
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, for
-  example those surveyed in \cite{geng2006interestingness}. Augmenting or
-  replacing the criteria may be useful, for example to distinguish useful
-  relationships from incidental coincidences; or to prevent surprising,
-  insightful equations from being discarded because they can be simplified.
-\end{description}
-
-We are implementing a system called \textsc{ML4HS} to investigate these ideas.
-Its current form is a pre-processor for \textsc{QuickSpec} for prioritising
-theory elements. Inspired by the use of premise selection
-\cite{kuhlwein2012overview} to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep
-together those expressions which combine in interesting ways, and to separate
-those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by that
-of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and related work in ACL2
-\cite{heras2013proof}, is an effective method for performing this separation.
-Future experiments will test this by comparing the throughput of
-\textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
+\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
+
+Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
+we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
+
+We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
+
+% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
 

### 2015-11-03 TransferReport/report.tex
Expanded evolutionary computation

@@
 \emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications in a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a viable approach to the search component of theory exploration.
+Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
+Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
 
@@
 Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-
-The evolutionary analogue of theory exploration is to remove the (explicit) fitness function, just like we remove explicit goals to turn the theorem proving task into the theory exploration task.
-There are two main reasons for studying such purely-exploratory evolutionary systems: \emph{artificial life} and \emph{deceptive problems}.
+If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
+
+Pure exploration (i.e.
+for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
 The former attempts to gain insight into the nature of life and biology through competition over limited resources.
 Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to the search aspect of theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
+efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
 \cite{baum2000evolution}).
 
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying \emph{exploration in general}, and hence provides insight for our narrower domain of theory exploration.
+On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
 Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
 by hill climbing), but few global optima which are hard to find.
-Many approaches to avoiding deception, such as \emph{niching methods} \cite{sareni1998fitness}, alter the output of the fitness function to promote \emph{diversity} and \emph{novelty}.
-For example, \emph{fitness sharing} promotes diversity by dividing up the fitness of identical or similar solutions, e.g.
-the result of a fitness function $f$ may be shared by two identical solutions $s_1$ and $s_2$, such that the augmented fitness $f'(s_1) = \frac{f(s_1)}{2}$ and $f'(s_2) = \frac{f(s_2)}{2}$.
-As always, there are many variations on this theme, such as how to judge similarity (e.g.
-``genetically'' by comparing the solutions themselves, or ``phenotypically'' by comparing their fitness), but all strive to penalise duplication and hence promote diversity.
-
-Once we augment fitness functions, to a greater or lesser degree, an obvious question arises: what if the augmented function contains nothing of the original fitness?
-This is the pure exploration scenario we are interested in, which has been found to provide useful ``stepping stones'', even in objective-driven domains \cite{lehman2011abandoning} \cite{hutter2002fitness}.
-
-% TODO: Expand
+Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \cite{sareni1998fitness}.
+
+One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
+Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
+In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
+We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
+if we allocate fitness uniformly between identical solutions we get:
+
+$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
+
+Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
+In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
+By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
+
+There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
+Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
+This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
+
+In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
+The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
+via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
+via mutation and crossover, to obtain new solutions.
+
+\newcommand{\argmin}{\operatornamewithlimits{argmin}}
+
+Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
+Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
+For example, the fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
+An individual $s$ is then selected with fitness closest to $f_t$, i.e.
+$\argmin{s}{\lvert f(s) - f_t \rvert}$.
+
+In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
+This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
+As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
+
+Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
+This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \cite{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
+Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
 

### 2015-11-06 Balmoral2015/render.sh
Fiddled render.sh scripts

@@ -1,5 +1,6 @@
 #! /usr/bin/env nix-shell
-#! nix-shell -i bash -p bash haskellPackages.Agda
+#! nix-shell -i bash -p bash haskellPackages.Agda graphviz
+
 # --self-contained
 HTML_ARGS="-w dzslides --standalone --filter panpipe --filter panhandle"
 

### 2015-11-06 STP/talks/render.sh

@@ -1,3 +1,0 @@
-#!/usr/bin/env bash
-cd 2015/10
-./render.sh

### 2015-11-06 TransferReport/interesting.tex
Pulled out lots of interestingess stuff into a separate file

--- 2015-11-03 TransferReport/report.tex
@@
+% Since writing this up, it was decided to leave as future work. Still, worth
+% having around for editing into such future work.
+
 \documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
 \begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
 
@@
 
-\section{Background}
-\label{background}
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \cite{QuickSpec} and \textsc{HipSpec} \cite{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \cite{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \cite{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
 \begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \cite{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \cite{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \cite{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \cite{nipkow2002isabelle} and Coq \cite{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \cite{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \cite{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \cite{schulz2013system} and Vampire \cite{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \cite{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \cite{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \cite{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
 \item{Interestingness}
@@
 An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$\argmin{s}{\lvert f(s) - f_t \rvert}$.
+$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
 
@@
 
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}, and hence provides a general mechanism for converting goal-driven algorithms into exploration systems.
-As an unsupervised learning task, AC has no access to labels or meanings associated with this input; the only features it can learn are the structure and relationships inherent in the data.
-The aim of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
+\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}.
+Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
+
+As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
+The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
 The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
@@
 
-A framework encompassing many explicit interestingness measures is given in \cite{oudeyer2007intrinsic}, as \emph{implicit rewards} in the context of reinforcement learning.
+In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
+A framework encompassing many examples is given in \cite{oudeyer2007intrinsic} in the context of reinforcement learning.
 
@@
 
-\section{Contributions}\label{current}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \cite{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \cite{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \cite{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
 \bibliographystyle{plain}

### 2015-11-09 TransferReport/report.tex
Fix argmin
Added syntax rules for Core

@@
 \usepackage{fixltx2e}
+\usepackage[numbers]{natbib}
+\usepackage{mathpartir}
+\usepackage{mmm}
 
@@
 
+\iffalse
+TODO
+Intro + motivation
+
+About 2 pages
+
+Functional programming is important (motivation)
+
+introduce running example
+
+Haskell is a mainstream FP language (motivation)
+
+Important to test (example)
+
+Tests are not sufficient (example)
+
+QuickCheck (example)
+
+How to come up with the properties
+
+Mention theory exploration
+\fi
+
 As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
@@
 For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
+Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
 Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
@@
 
+\iffalse
+TODO
+
+Background
+
+2.1 Introduce language signature
+ a) subset of Haskell
+ b) stuff from QuickCheck/Spec
+
+2.2? Machine learning and feature extraction
+~1 page
+
+2.3: QuickSpec, HipSpec
+\fi
+
+\subsection{Haskell}
+
+\newcommand{\fc}{System F\textsubscript{C}}
+
+\begin{figure}
+  \begin{equation*}
+    \begin{split}
+      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
+                         |\ & \texttt{Lit}\ literal                                  \\
+                         |\ & \texttt{App}\ expr\ expr                               \\
+                         |\ & \texttt{Lam}\ id\ expr                                 \\
+                         |\ & \texttt{Let}\ bind\ expr                               \\
+                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
+      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
+                         |\ & \texttt{Global}\ \mathcal{G} \\
+      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
+                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
+      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
+      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
+                         |\ & \texttt{LitAlt}\ literal                               \\
+                         |\ & \texttt{Default}                                       \\
+      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
+                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
+    \end{split}
+  \end{equation*}
+  where:
+  \begin{tabular}[t]{l @{ $=$ } l}
+    $\mathcal{S}$ & string literals    \\
+    $\mathcal{N}$ & numeric literals   \\
+    $\mathcal{L}$ & local identifiers  \\
+    $\mathcal{G}$ & global identifiers
+  \end{tabular}
+
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
+  \label{coresyntax}
+\end{figure}
+
+The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
+For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
+The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
+
+We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
+
 Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
@@
 This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \cite{QuickSpec} and \textsc{HipSpec} \cite{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
+We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
 
@@
 
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
+Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
 Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \cite{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \cite{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
+For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
 
@@
 \hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \cite{wadler1989theorems}.
+This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
 
@@
 \hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \cite{yorgey2009typeclassopedia}.
+Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
 This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
@@
 Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \cite{lindley2014hasochism}).
+However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
 This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
@@
 
-This coincidence is due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, which identifies programming languages with systems of logic.
+This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
 Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
@@
 This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \cite{nipkow2002isabelle} and Coq \cite{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
+One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
 
 Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \cite{hales2015formal}).
+see \citep{hales2015formal}).
 One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
+The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
 
@@
 ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \cite{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \cite{schulz2013system} and Vampire \cite{riazanov2003implementing}) operate in classical first-order logic.
+By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
+In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
 This is the most striking difference from ITP systems (e.g.
@@
 In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.
+Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
 The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
@@
 There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \cite{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
+The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
+
+Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
 \emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
@@
 
-Early implementations like \textsc{Theorema} \cite{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
+Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
 Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
 
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \cite{Hipster}.
+Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
 By removing user interaction, these properties must be implemented by algorithms.
@@
 
-As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
+As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
 The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
@@
 \textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}.
+This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
 \textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
@@
 Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
+In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
   We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
@@
 
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
+An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
 
@@
 
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
+Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
 \footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
@@
 
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \cite{Heras.Komendantskaya.Johansson.ea:2013}.
+The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
 These clusters are used in two ways:
@@
 One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \cite{claessen2011quickcheck} and \textsc{SmartCheck} \cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \cite{reich2013advances}) and logic programming (\textsc{SparseCheck} \cite{sparsecheck}).
+Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
 
@@
 % TODO: Analysis
+
+\section{Contributions}\label{current}
+
+\iffalse
+TODO
+
+3 Recurrent clustering
+
+Good analysis of differences with ML4PG/ACL2ML
+\fi
+
+\section{Implementation}
+
+Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
+
+\subsection{Circular Convolution}\label{circularconvolution}
+
+We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+data Tree = Leaf FV
+          | Branch FV Tree Tree
+\end{lstlisting}
+
+Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
+To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceD 1 (Branch fv l r) = Leaf   fv
+reduceD n (Leaf   fv)     = Leaf   fv
+reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
+                                      (reduceD (n-1) r)
+\end{lstlisting}
+
+This is simple, but suffers two problems:
+
+\begin{itemize}
+ \item We must choose a conservatively large \hs{d}, since we're throwing away information
+ \item The number of feature vectors grows exponentially as \hs{d} increases
+\end{itemize}
+
+The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
+One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
+
+\iffalse
+
+We can do this with \hs{tabulate w d t}:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+tabulate w d t = take d (map (take w) (tabulate' t))
+
+tabulate' (Leaf   fv)     = [fv] : []
+tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
+
+merge    []     ys  = ys
+merge    xs     []  = xs
+merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
+\end{lstlisting}
+
+\fi
+
+Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
+Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
+Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceC (Leaf fv)       = fv
+reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
+\end{lstlisting}
+
+Circular convolution has the following desirable properties:
+
+\begin{description}
+
+  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
+
+  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
+
+\end{description}
+
+Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
+This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
+In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
+
+Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
+
+\subsubsection{Extracting Features from XML}
+
+This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
+The application parses arbitrary trees of XML and uses binary features, ie.
+feature vectors are bit vectors.
+
+To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
+
+\begin{equation}
+feature(n) = 2^{\md5(s) \pmod{L}}
+\end{equation}
+
+Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
+The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
+Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
+
+In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
+
+\subsubsection{Application to Coq}
+
+Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
+We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
+
+Coq will also interpret the standard output of the command, to generate terms and tactics.
+This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
+
+\subsection{\texttt{nix-eval}}\label{nixeval}
+
+To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
+This is difficult, as many terms are functions, which Haskell does not let us inspect.
+This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
+To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
+
+\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
+In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
+
+On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
+Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
+Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
+
+We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
+These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
+
+\texttt{nix-eval} is build on top of the Nix package manager and GHC.
+Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
+
+The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
+When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
+GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
+The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
+The expression, therefore, must have type \hs{String}; although this is not enforced statically.
+
+The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
+The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
+The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
+It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
+using JSON.
+
+\subsection{\textsc{ASTPlugin}}\label{astplugin}
+
+The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
+
+Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
+Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
+However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
+
+This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
+All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
+
+Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
+All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
+This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
+
+GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
+We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
+
+There are several complications in the implementation of \textsc{AstPlugin}.
+We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
+This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
+Three areas which make Core more difficult to handle are:
+
+\begin{description}
+  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
+In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
+Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
+This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
+
+  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
+Since Core does not make these distinctions, names may become ambiguous.
+For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
+To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
+This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
+
+  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
+When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
+Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
+\end{description}
+
+In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
+Those which fail, due to the reasons above and others, are discarded.
+
+The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
+Definitions are indexed by a combination of their package name, module name and binding name.
+The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
+Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
+
+Such data is useful for multiple purposes:
+
+\begin{itemize}
+  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
+those (package, module, name) combinations which it references.
+Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
+
+  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
+As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
+
+  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
+
+  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
+
+  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
+Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
+as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
+As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
+Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
+It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
+\end{itemize}
+
+\subsection{\textsc{MLSpec}}\label{mlspec}
+
+When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
+One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
+
+Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
+Features of \textsc{MLSpec} include:
+
+\begin{itemize}
+  \item{Monomorphising}: Given values of polymorphic type, e.g.
+\hs{safeHead :: forall t.
+[t] -> Maybe t} and \hs{[] :: forall t.
+[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
+Maybe t} without instantiating the variable \hs{t} to a specific type.
+Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
+We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
+\hs{forall c.
+c Bool -> c Bool}) or incompatible class constraints (e.g.
+\hs{forall t.
+IsString t => t}).
+
+  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
+There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
+
+  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
+
+  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
+
+\end{itemize}
+
+\subsection{\textsc{ML4HS}}\label{ml4hs}
+
+\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
+
+Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
+we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
+
+We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
+
+% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
+
+% Divide and conquer for theory exploration
+% Does it help?
+
+% Clustering and feature extraction
+
+\section{First Results}
+
+\iffalse
+TODO
+
+Future work?
+Hypothetical use cases?
+\fi
 
@@
 
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
+\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
 Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
+Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
 
@@
 efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\cite{baum2000evolution}).
+\citep{baum2000evolution}).
 
 On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
+Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
 by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \cite{sareni1998fitness}.
+Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
 
@@
 Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
+For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
 An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$\argmin{s}{\lvert f(s) - f_t \rvert}$.
+$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
 
@@
 Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \cite{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
+This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
 Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
@@
 
-  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
+  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
 The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
@@
 
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}, and hence provides a general mechanism for converting goal-driven algorithms into exploration systems.
-As an unsupervised learning task, AC has no access to labels or meanings associated with this input; the only features it can learn are the structure and relationships inherent in the data.
-The aim of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
+\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
+Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
+
+As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
+The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
 
@@
 
-A framework encompassing many explicit interestingness measures is given in \cite{oudeyer2007intrinsic}, as \emph{implicit rewards} in the context of reinforcement learning.
+In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
+A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
 
@@
 % TODO: Examples
-\cite{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \cite{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
+\citep{Schmidhuber1999}
+
+Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
+A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
 The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
@@
 % TODO
-\cite{Kaplan2006}
-\cite{Lipson2007}
-\cite{Luciw2011}
-\cite{Macedo2000}
-\cite{Ramik.Sabourin.Madani:2013}
-\cite{Roa.Kruijff.Jacobsson:2009}
-\cite{Schaul.Sun.Wierstra.ea:2011}
-\cite{Schmidhuber1999}
-\cite{Schmidhuber:1991}
-\cite{Scott1989}
-\cite{Steunebrink.Koutnik.Thorisson.ea:2013}
-\cite{maher2008achieving}
-\cite{meyer1991possibility}
-\cite{oudeyer2004intelligent}
-\cite{oudeyer2014evolution}
-\cite{schmidhuber2006developmental}
+\citep{Kaplan2006}
+\citep{Lipson2007}
+\citep{Luciw2011}
+\citep{Macedo2000}
+\citep{Ramik.Sabourin.Madani:2013}
+\citep{Roa.Kruijff.Jacobsson:2009}
+\citep{Schaul.Sun.Wierstra.ea:2011}
+\citep{Schmidhuber1999}
+\citep{Schmidhuber:1991}
+\citep{Scott1989}
+\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
+\citep{maher2008achieving}
+\citep{meyer1991possibility}
+\citep{oudeyer2004intelligent}
+\citep{oudeyer2014evolution}
+\citep{schmidhuber2006developmental}
 
@@
 in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
+Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
 Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
@@
 % TODO
-\cite{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
+\citep{kuhlwein2012overview}
+
+The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
 Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
@@
 
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
 Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
@@
 
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \cite{kuhlwein2013mash}.
+  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
 The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
@@
 
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
+  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
 \iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
@@
 % TODO: ML4PG
-\cite{journals/corr/abs-1212-3618}
+\citep{journals/corr/abs-1212-3618}
 % TODO: ACL2(ml) (also examples section)
-\cite{heras2013proof}
+\citep{heras2013proof}
 
@@
 % TODO
-\cite{Hutter.Lloyd.Ng.ea:2013}
+\citep{Hutter.Lloyd.Ng.ea:2013}
 
@@
 % TODO:
-\cite{Montano-Rivas.McCasland.Dixon.ea:2012}
-\cite{Piantadosi.Tenenbaum.Goodman:2012}
-\cite{Wille:2005}
-\cite{colton1999automatic}
-\cite{colton2000agent}
-\cite{colton2012automated}
-\cite{lenat1977automated}
-\cite{mullerunderstanding}
-\cite{Bundy.Cavallo.Dixon.ea:2015}
-\cite{johansson2009isacosy}
-\cite{spector2008genetic}
-\cite{colton2012automated}
- \cite{geng2006interestingness}
+\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
+\citep{Piantadosi.Tenenbaum.Goodman:2012}
+\citep{Wille:2005}
+\citep{colton1999automatic}
+\citep{colton2000agent}
+\citep{colton2012automated}
+\citep{lenat1977automated}
+\citep{mullerunderstanding}
+\citep{Bundy.Cavallo.Dixon.ea:2015}
+\citep{johansson2009isacosy}
+\citep{spector2008genetic}
+\citep{colton2012automated}
+ \citep{geng2006interestingness}
 % TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
@@
 Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
+Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
 \emph{artificial neural networks} with fixed connections between nodes).
@@
 The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
+This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
 
@@
 The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
+If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
 However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \cite{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
+The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
+
+The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
 
 % TODO
-\cite{Gartner2003}
-\cite{Oveisi.Oveisi.Erfanian.ea:2012}
-\cite{bakir2007predicting}
-\cite{conf/ijcai/Plate91}
-\cite{goller1996learning}
-\cite{kwasny1995tail}
-\cite{pollack1990recursive}
-\cite{zanzotto2012distributed}
+\citep{Gartner2003}
+\citep{Oveisi.Oveisi.Erfanian.ea:2012}
+\citep{bakir2007predicting}
+\citep{conf/ijcai/Plate91}
+\citep{goller1996learning}
+\citep{kwasny1995tail}
+\citep{pollack1990recursive}
+\citep{zanzotto2012distributed}
 
@@
 
-\section{Contributions}\label{current}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \cite{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \cite{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \cite{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
 \section{Future Work}
@@
 \iffalse
+TODO
+
+interestingness
+\fi
+
+\iffalse
+
 QuickSpec: extend or extinguish?

### 2015-11-10 TransferReport/report.tex
Intro to Haskell/testing

@@
 
+\lstset{
+  frame=none,
+  xleftmargin=2pt,
+  stepnumber=1,
+  numbers=none,
+  numbersep=5pt,
+  numberstyle=\ttfamily\tiny\color[gray]{0.3},
+  belowcaptionskip=\bigskipamount,
+  captionpos=b,
+  escapeinside={*'}{'*},
+  language=haskell,
+  tabsize=2,
+  emphstyle={\bf},
+  commentstyle=\it,
+  stringstyle=\mdseries\rmfamily,
+  showspaces=false,
+  keywordstyle=\bfseries\rmfamily,
+  columns=flexible,
+  basicstyle=\small\sffamily,
+  showstringspaces=false,
+  morecomment=[l]\%,
+}
+
+\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
+
+\newenvironment{haskell}
+  {
+    \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  }
+  {
+    \end{lstlisting}
+  }
+
+\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+
 \begin{document}
@@
 
-Functional programming is important (motivation)
-
-introduce running example
-
 Haskell is a mainstream FP language (motivation)
 
-Important to test (example)
-
-Tests are not sufficient (example)
-
 QuickCheck (example)
@@
 As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
+The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
+For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
+
+Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
+These ideas are thus spreading to mainstream software engineering, in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
+
+Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g.
+life-) critical systems.
+To see why, consider the following Haskell function, along with some unit tests (see \ref{haskell} for more details on Haskell):
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  factorial 0 = 1
+  factorial n = n * factorial (n-1)
+
+  fact_base      = factorial 0 == factorial 1
+  fact_increases = factorial 4 >  factorial 3
+  fact_div       = div (factorial 5) (factorial 4) == 5
+\end{lstlisting}
+
+The intent of the function is to map an input $n$ to an output $n!$.
+The tests check a few properties of the implementation, including the base case, that the function is monotonically increasing, and a relationship between adjacent outputs.
+However, these tests will \emph{not} expose a serious problem with the implementation: it diverges on half of its possible inputs!
+
+All of Haskell's built-in numeric types allow negative numbers, which this implementation doesn't take into account.
+Whilst this is a rather trivial example, it highlights a common problem: unit tests are insufficient to expose incorrect assumptions.
+In this case, our assumption that numbers are positive has caused a bug in the implementation \emph{and} limited the tests we've written.
+
+If we do manage spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+factorial 0 = 1
+factorial n = let nPos = abs n
+               in nPos * factorial (nPos - 1)
+
+fact_neg = factorial (-1) == factorial 1
+\end{lstlisting}
+
+However, this is \emph{still} not enough: since this function only uses generic numeric operations, it will be polymorphic; allowing all numeric types\footnote{Its type will be of the form \hs{forall t.
+Num t => t -> t}, where \hs{Num} constrains the type variable \hs{t} to be numeric.
+There will also be extra contraints like \hs{Eq} due to the use of \hs{==} in the unit tests.}.
+If we provide a fractional value, the function will again diverge.
+Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
+
+Haskell offers a partial solution to this problem in the form of \emph{property checking}.
+Tools such as \textsc{QuickCheck} separate tests into three components: a \emph{property} to check, a source of values to check and a stopping criterion.
+Here is how we might restate our unit tests as properties:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+fact_base = factorial 0 == factorial 1 -- Unchanged; unit tests are valid (albeit limited) properties
+
+fact_increases n = factorial (n+1) >= factorial n  -- >= since factorial 1 == factorial 0
+
+fact_div n = factorial (n+1) `div` factorial n == n+1
+
+fact_neg n = factorial (n * -1) == factorial n
+\end{lstlisting}
+
+Turning particular values (like \hs{5} and \hs{4}) into parameters tells \textsc{QuickCheck} to \emph{universally quantify} them, i.e.
+a unit test stating $factorial(-1) = factorial(1)$ becomes a property stating $\forall n, factorial(n \times -1) = factorial(n)$.
+
+%TODO example, eg. int -> int -> int requires 2^64 * 2^64 cases for full coverage
+%     NO! we want random tests first; show something with edge-cases; what about assuming Ints are positive? what about %assuming Int when it's actually Num
+
+\iffalse
+
 Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
@@
 
+\fi
+
 \section{Background}
@@
 2.1 Introduce language signature
- a) subset of Haskell
+ a) subset of Haskell DONE
  b) stuff from QuickCheck/Spec
@@
 \subsection{Haskell}
+\label{haskell}
 
@@
 For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
+The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} we erase types and use a custom representation of names.
+There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
+
+\iffalse
 
@@
 
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
+\fi
 
@@
 
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
 When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
@@
 \iffalse
+
 Haskell equivalent:
@@
 plus_tail n (S m) = plus_tail (S n) m
+
 \fi
@@
 \label{related}
+
+\subsection{Haskell}
+\label{haskelldesc}
+
+Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
+This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
+We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
+
+Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
+Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
+For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
+
+\begin{description}
+
+\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
+
+\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
+
+\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
+
+\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
+performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
+This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
+For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
+Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
+
+\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
+For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
+an ADT's declaration specifies all of the normal forms for that type.
+This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
+
+\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
+This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
+t -> t}.
+\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
+\hs{id x = x}.
+This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
+
+\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
+Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
+As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+op :: forall t. Semigroup t => t -> t -> t
+\end{lstlisting}
+
+$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
+
+The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
+\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
+As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
+Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
+\hs{Integer} with \hs{op} performing addition.
+Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
+This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
+
+\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
+This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
+Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
+
+\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
+When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
+This mechanism allows modules to provide more guarantees than are available just in their types.
+For example, a module may represent email addresses in the following way:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
+module Email (Email(), at, render) where
+
+data Email = E String String
+
+render :: Email -> String
+render (E u h) = u ++ "@" ++ h
+
+at :: String -> String -> Maybe Email
+at "" h  = Nothing
+at u  "" = Nothing
+at u  h  = Just (E u h)
+\end{lstlisting}
+
+The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
+We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
+By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
+
+\end{description}
+
+Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
+However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
+This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
+in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
+We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
+Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
+The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
+
+Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
+
+\begin{itemize}
+
+\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
+
+\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
+
+\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
+
+\end{itemize}
 

### 2015-11-10 TransferReport/haskell_example/src/Example.hs

@@ -1,0 +1,32 @@
+module Example where
+{-
+type Nat = [()]
+
+render :: Nat -> String
+render = show . length
+
+mult :: Nat -> Nat -> Nat
+mult x y = case x of
+                []     -> []
+                ():as  -> y
+                (a:as) -> y ++ mult as y
+
+mult_tail' :: Nat -> Nat -> Nat -> Nat
+mult_tail' x y acc = case x of
+                          []     -> acc
+                          (a:as) -> mult_tail' as y (y ++ acc)
+
+mult_tail :: Nat -> Nat -> Nat
+mult_tail x y = mult_tail' x y []
+-}
+factorial1 0 = 1
+factorial1 n = n * factorial1 (n-1)
+
+base      f = f 0    ==    f 1
+increases f = f 4    >     f 3
+divides   f = f 5    `div` f 4 == 5
+negative  f = f (-1) ==    f 1
+
+factorial2 0 = 1
+factorial2 n = let nPos = abs n
+                in nPos * factorial2 (nPos - 1)

### 2015-11-10 TransferReport/report.tex
Updates

@@
 
+\newcommand{\fc}{System F\textsubscript{C}}
+
+\newcommand{\qcheck}{\textsc{QuickCheck}}
+\newcommand{\qspec}{\textsc{QuickSpec}}
+\newcommand{\hspec}{\textsc{HipSpec}}
+
 \providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
@@
 Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
-These ideas are thus spreading to mainstream software engineering, in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
+Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
+
+Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive.
+Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
+
+\iffalse
+TODO
+Our contributions are:
+
+\begin{itemize}
+  \item A framework for applying theory exploration tools such as \qspec to Haskell's existing package system.
+  \item A
+\end{itemize}
+\fi
+
+\section{Background}
+\label{background}
+
+\iffalse
+TODO
+
+Background
+
+2.1 Introduce language signature
+ a) subset of Haskell DONE
+ b) stuff from QuickCheck/Spec
+
+2.2? Machine learning and feature extraction
+~1 page
+
+2.3: QuickSpec, HipSpec
+\fi
+
+\subsection{Testing}
 
@@
 fact_base      = factorial 0 == factorial 1
-  fact_increases = factorial 4 >  factorial 3
-  fact_div       = div (factorial 5) (factorial 4) == 5
+fact_increases = factorial 3 <= factorial 4
+fact_div       = factorial 4 == factorial 5 `div` 5
 \end{lstlisting}
@@
 
-fact_neg = factorial (-1) == factorial 1
+fact_neg = factorial 1 == factorial (-1)
 \end{lstlisting}
@@
 
-Haskell offers a partial solution to this problem in the form of \emph{property checking}.
-Tools such as \textsc{QuickCheck} separate tests into three components: a \emph{property} to check, a source of values to check and a stopping criterion.
+Haskell offers a partial solution to this problem in the form of \emph{property checking}\footnote{See section \ref{propertychecking} for more information}.
+Tools such as \qcheck separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
 Here is how we might restate our unit tests as properties:
@@
 \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-fact_base = factorial 0 == factorial 1 -- Unchanged; unit tests are valid (albeit limited) properties
-
-fact_increases n = factorial (n+1) >= factorial n  -- >= since factorial 1 == factorial 0
-
-fact_div n = factorial (n+1) `div` factorial n == n+1
-
-fact_neg n = factorial (n * -1) == factorial n
-\end{lstlisting}
-
-Turning particular values (like \hs{5} and \hs{4}) into parameters tells \textsc{QuickCheck} to \emph{universally quantify} them, i.e.
-a unit test stating $factorial(-1) = factorial(1)$ becomes a property stating $\forall n, factorial(n \times -1) = factorial(n)$.
-
-%TODO example, eg. int -> int -> int requires 2^64 * 2^64 cases for full coverage
-%     NO! we want random tests first; show something with edge-cases; what about assuming Ints are positive? what about %assuming Int when it's actually Num
+fact_base        = factorial 0 == factorial 1
+fact_increases n = factorial n <= factorial (n+1)
+fact_div       n = factorial n == factorial (n+1) `div` (n+1)
+fact_neg       n = factorial n == factorial (-n)
+\end{lstlisting}
+
+The free variables (all called \hs{n} in this case) are abstracted as function parameters; these parameters are implicitly \emph{universally quantified}, i.e.
+we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
+Notice that unit tests like \hs{fact_base} are valid properties; they just assert rather weak statements.
+
+To check these properties, \qcheck treats closed terms (like \hs{fact_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
+For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
+The default stopping criterion for \qcheck (for each test) is when a single generated test fails, or when 100 generated tests pass.
+
+The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with closed values. In the \hs{factorial} example, this manifests in two ways:
+
+\begin{itemize}
+  \item \qcheck cannot test polymorphic functions; they type must be \emph{monomorphised} first (instantiated to a particular concrete type).
+This is a technical limitation, since \qcheck must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
+
+  \item The generators used by \qcheck depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
+This will expose the problem with negative numbers, which we weren't expecting.
+\end{itemize}
+
+Property checking is certainly an improvement over unit testing, but the problem of tests being biased towards expected cases remains, since we are manually specifying the properties to be checked.
+
+We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec and \hspec.
+These programs \emph{discover} properties of a ``theory'' (e.g.
+a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec) automated theorem proving.\footnote{See section \ref{atp} for more information on automated theorem proving.}
 
@@
 
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell DONE
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
 \subsection{Haskell}
 \label{haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
 
@@
 
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
+As an example, \qspec \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
 The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
@@
 
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
+Although \qspec only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \hspec and \textsc{Hipster}.
+
+\hspec is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
+It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \qspec.
+\qspec, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \qcheck counterexample finder, then conjectures equations relating the members of these classes.
+This approach works well as a lemma generation system, making \hspec a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
+\hspec is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \hspec alongside more traditional QA tools like \qcheck, \textsc{HUnit} and \textsc{Criterion}.
+
+In fact, there are similarities between the way a TE system like \hspec can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \qcheck can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
 There are likely lessons that each can learn from the other's approach to term generation.
 
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
+We identify the following areas for researching potential improvements in existing TE systems such as \hspec:
 
@@
 One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
+Many Haskell property checkers exist, based on random testing (\qcheck \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
 
@@
 Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
+Indeed, \hspec already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
 However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
@@
 
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
+In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \qspec.
 Those which fail, due to the reasons above and others, are discarded.
@@
 
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec to explore.
+
+  \item{Reflection}: As mentioned above, \hspec uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
 Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
@@
 
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
+Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec, based on the databases constructed by \textsc{AstPlugin} described above.
 Features of \textsc{MLSpec} include:
@@
 [t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
+[t]}, a testing-based system like \qspec is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
 Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
+Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck by attempting to instantiate all type variables to \hs{Integer}.
 We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
@@
 
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
+  \item{Variable definition}: Once a \qspec theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec's exhaustive search).
 
@@
 
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
+% TODO: Future experiments will test this by comparing the throughput of \qspec with and without the \textsc{ML4HS} pre-processor.
 
@@
 This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
+We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\qspec \citep{QuickSpec} and \hspec \citep{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec.
 
@@
 
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
+\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec and \hspec, and counterexample finders like \qcheck, \textsc{SmallCheck} and \textsc{SmartCheck}.
 

### 2015-11-12 TransferReport/background.tex
Split up monolothic report file

--- 2015-11-10 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\newcommand{\qcheck}{\textsc{QuickCheck}}
-\newcommand{\qspec}{\textsc{QuickSpec}}
-\newcommand{\hspec}{\textsc{HipSpec}}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-\newenvironment{haskell}
-  {
-    \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  }
-  {
-    \end{lstlisting}
-  }
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Haskell is a mainstream FP language (motivation)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
-For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
-
-Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
-Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
-
-Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive.
-Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
-
-\iffalse
-TODO
-Our contributions are:
-
-\begin{itemize}
-  \item A framework for applying theory exploration tools such as \qspec to Haskell's existing package system.
-  \item A
-\end{itemize}
-\fi
-
 \section{Background}
@@
 
-If we do manage spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
+If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
 
@@
 However, this is \emph{still} not enough: since this function only uses generic numeric operations, it will be polymorphic; allowing all numeric types\footnote{Its type will be of the form \hs{forall t.
-Num t => t -> t}, where \hs{Num} constrains the type variable \hs{t} to be numeric.
+Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
 There will also be extra contraints like \hs{Eq} due to the use of \hs{==} in the unit tests.}.
@@
 Haskell offers a partial solution to this problem in the form of \emph{property checking}\footnote{See section \ref{propertychecking} for more information}.
-Tools such as \qcheck separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
+Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
 Here is how we might restate our unit tests as properties:
@@
 
-To check these properties, \qcheck treats closed terms (like \hs{fact_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
+To check these properties, \qcheck{} treats closed terms (like \hs{fact_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
 For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
-The default stopping criterion for \qcheck (for each test) is when a single generated test fails, or when 100 generated tests pass.
+The default stopping criterion for \qcheck{} (for each test) is when a single generated test fails, or when 100 generated tests pass.
 
@@
 \begin{itemize}
-  \item \qcheck cannot test polymorphic functions; they type must be \emph{monomorphised} first (instantiated to a particular concrete type).
-This is a technical limitation, since \qcheck must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
+  \item \qcheck{} cannot test polymorphic functions; they type must be \emph{monomorphised} first (instantiated to a particular concrete type).
+This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
 
-  \item The generators used by \qcheck depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
+  \item The generators used by \qcheck{} depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
 This will expose the problem with negative numbers, which we weren't expecting.
@@
 
-We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec and \hspec.
+We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}.
 These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec) automated theorem proving.\footnote{See section \ref{atp} for more information on automated theorem proving.}
+a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{atp} for more information on automated theorem proving.}
 
@@
 
-As an example, \qspec \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
+As an example, \qspec{} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
 The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
@@
 
-Although \qspec only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \hspec and \textsc{Hipster}.
+Although \qspec{} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \hspec{} and \textsc{Hipster}.
 
-\hspec is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \qspec.
-\qspec, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \qcheck counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \hspec a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\hspec is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \hspec alongside more traditional QA tools like \qcheck, \textsc{HUnit} and \textsc{Criterion}.
+\hspec{} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
+It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \qspec{}.
+\qspec{}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
+This approach works well as a lemma generation system, making \hspec{} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
+\hspec{} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \hspec{} alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
 
-In fact, there are similarities between the way a TE system like \hspec can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \qcheck can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
+In fact, there are similarities between the way a TE system like \hspec{} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
 There are likely lessons that each can learn from the other's approach to term generation.
 
-We identify the following areas for researching potential improvements in existing TE systems such as \hspec:
+We identify the following areas for researching potential improvements in existing TE systems such as \hspec{}:
 
@@
 % Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\qcheck \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \hspec already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \qspec.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec to explore.
-
-  \item{Reflection}: As mentioned above, \hspec uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \qspec is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \qspec theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec's exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \qspec with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
-
-\subsection{Haskell}
-\label{haskelldesc}
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\qspec \citep{QuickSpec} and \hspec \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec.
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec and \hspec, and counterexample finders like \qcheck, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/contributions.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Functional programming is important (motivation)
-
-introduce running example
-
-Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
 \section{Contributions}\label{current}
@@
 \fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/examples.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Functional programming is important (motivation)
-
-introduce running example
-
-Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
 \section{Exploration in Theorem Proving}
@@
 \subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
 
@@
 One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
+Many Haskell property checkers exist, based on random testing (\qcheck{} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
 
@@
 % TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/implementation.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Functional programming is important (motivation)
-
-introduce running example
-
-Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
 \section{Implementation}
@@
 
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
+  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec to explore.
 
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+  \item{Reflection}: As mentioned above, \hspec uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
 Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
@@
 % Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/intro.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
 \section{Introduction}
@@
 
-Functional programming is important (motivation)
-
-introduce running example
-
 Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
 
@@
 As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
+The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
+For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
 
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
+Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
+Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
 
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
+Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive.
+Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
 
@@
 TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
+Our contributions are:
 
 \begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
+  \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
+  \item A
 \end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
 \fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/related.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
-\documentclass[]{article}
-
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage[T1]{fontenc}
-%\usepackage{enumitem}
-\usepackage{paralist}
-\usepackage{csquotes}
-\usepackage[affil-it]{authblk}
-\usepackage{listings}
-\usepackage{fixltx2e}
-\usepackage[numbers]{natbib}
-\usepackage{mathpartir}
-\usepackage{mmm}
-
-% Treat paragraph as subsubsubsection
-\usepackage{titlesec}
-\setcounter{secnumdepth}{4}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}
-
-\DeclareMathOperator{\md5}{MD5}
-
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Functional programming is important (motivation)
-
-introduce running example
-
-Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
+\section{Related Work}
+\label{related}
 
 \subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
+\label{haskelldesc}
 
@@
 This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=none,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  captionpos=b,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
+We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
 
@@
 
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
+\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}, and counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}.
 
@@
 \end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
-
-\section{First Results}
-
-\iffalse
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
-\section{Related Work}
-\label{related}
 
@@
 \fi
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
-\section{Conclusion}
-\label{conclusion}
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2015-11-12 TransferReport/report.tex
Split up monolothic report file

--- 2015-11-09 TransferReport/report.tex
@@
 
-\begin{document}
-
-\pagestyle{headings}  % switches on printing of running heads
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Functional programming is important (motivation)
-
-introduce running example
-
-Haskell is a mainstream FP language (motivation)
-
-Important to test (example)
-
-Tests are not sufficient (example)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\section{Background}
-\label{background}
-
-\iffalse
-TODO
-
-Background
-
-2.1 Introduce language signature
- a) subset of Haskell
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\subsection{Haskell}
-
-\newcommand{\fc}{System F\textsubscript{C}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{The current version being present in GHC 7.10.2}, we erase types, use a custom  various forms of literal simplifications include perform type erasure
-
-We follow it is capable of expressing just as is capable of used as an intermediate which the Glasgow Haskell Compiler (GHC) target by
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\textsc{QuickSpec} \citep{QuickSpec} and \textsc{HipSpec} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \textsc{HipSpec}.
-
-\subsection{Haskell}\label{haskell}
-
 \lstset{
@@
 
+\newcommand{\fc}{System F\textsubscript{C}}
+
+\newcommand{\qcheck}{\textsc{QuickCheck}}
+\newcommand{\qspec}{\textsc{QuickSpec}}
+\newcommand{\hspec}{\textsc{HipSpec}}
+
 \providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
 
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
-
-\begin{description}
-
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
+\newenvironment{haskell}
+  {
     \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
+  }
+  {
     \end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \textsc{QuickSpec} and \textsc{HipSpec}, and counterexample finders like \textsc{QuickCheck}, \textsc{SmallCheck} and \textsc{SmartCheck}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
-\subsection{Theory Exploration}
-\label{theoryexploration}
-
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
-
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
-
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
-
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
-
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
-
-As an example, \textsc{QuickSpec} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
-
-Although \textsc{QuickSpec} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \textsc{HipSpec} and \textsc{Hipster}.
-
-\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}.
-\textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
-
-We identify the following areas for researching potential improvements in existing TE systems such as \textsc{HipSpec}:
-
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
-
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
+  }
 
@@
 
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
+\begin{document}
 
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
+\pagestyle{headings}  % switches on printing of running heads
 
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \citep{boyer1983proof}:
+\title{Interestingness Measures for Theory Exploration}
 
-$$(x * y) * z = x * (y * z)$$
+\author{Chris Warburton}
 
-During the course of the proof, the following obligation arises:
+\affil{University of Dundee,\\
+\texttt{http://tocai.computing.dundee.ac.uk}}
 
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
+\maketitle              % typeset the title of the contribution
 
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
+\begin{abstract}
+Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
+Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
+\end{abstract}
 
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \citep{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \citep{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \citep{claessen2011quickcheck} and \textsc{SmartCheck} \citep{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \citep{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \citep{reich2013advances}) and logic programming (\textsc{SparseCheck} \citep{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Contributions}\label{current}
-
-\iffalse
-TODO
-
-3 Recurrent clustering
-
-Good analysis of differences with ML4PG/ACL2ML
-\fi
-
-\section{Implementation}
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{Circular Convolution}\label{circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}\label{nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \textsc{HipSpec} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \textsc{QuickSpec}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \textsc{QuickSpec} to explore.
-
-  \item{Reflection}: As mentioned above, \textsc{HipSpec} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
-
-\subsection{\textsc{MLSpec}}\label{mlspec}
-
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \textsc{QuickSpec}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
-
-\begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \textsc{QuickSpec} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \textsc{QuickCheck} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
-
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
-
-  \item{Variable definition}: Once a \textsc{QuickSpec} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \textsc{QuickSpec}'s exhaustive search).
-
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
-
-\end{itemize}
-
-\subsection{\textsc{ML4HS}}\label{ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
-
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
-
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
-
-% TODO: Future experiments will test this by comparing the throughput of \textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
-
-% Divide and conquer for theory exploration
-% Does it help?
-
-% Clustering and feature extraction
+\input{intro.tex}
+\input{background.tex}
+\input{examples.tex}
+\input{contributions.tex}
+\input{implementation.tex}
 
@@
 
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\citep{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
+\input{related.tex}
 

### 2015-11-16 TransferReport/background.tex
Updates

@@
 \iffalse
-TODO
+TODO{2015-11-14}
 
@@
 2.1 Introduce language signature
- a) subset of Haskell DONE
  b) stuff from QuickCheck/Spec
@@
 
-\subsection{Testing}
+\iffalse
+
+Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
+By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
+However, the question still remains: on what should we focus that analysis?
+
+In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
+To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
+Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
+
+Appeals to interestingness arise when more direct measures, such as utility, are not available.
+For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
+Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
+Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
+
+We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
+We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
+A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
+We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
+Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
+
+We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
+
+\fi
+
+\subsection{Haskell}
+\label{haskell}
+
+\begin{figure}
+  \begin{equation*}
+    \begin{split}
+      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
+                         |\ & \texttt{Lit}\ literal                                  \\
+                         |\ & \texttt{App}\ expr\ expr                               \\
+                         |\ & \texttt{Lam}\ \mathcal{L}\ expr                        \\
+                         |\ & \texttt{Let}\ bind\ expr                               \\
+                         |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
+      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
+                         |\ & \texttt{Global}\ \mathcal{G}                           \\
+      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
+                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
+      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )                      \\
+      altcon\  \rightarrow\ & \texttt{DataAlt}\ \mathcal{G}                          \\
+                         |\ & \texttt{LitAlt}\ literal                               \\
+                         |\ & \texttt{Default}                                       \\
+      bind\    \rightarrow\ & \texttt{NonRec}\ \mathcal{L}\ expr                     \\
+                         |\ & \texttt{Rec}\ [ ( \mathcal{L},\ expr ) ]
+    \end{split}
+  \end{equation*}
+  where:
+  \begin{tabular}[t]{l @{ $=$ } l}
+    $\mathcal{S}$ & string literals    \\
+    $\mathcal{N}$ & numeric literals   \\
+    $\mathcal{L}$ & local identifiers  \\
+    $\mathcal{G}$ & global identifiers
+  \end{tabular}
+
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
+  \label{coresyntax}
+\end{figure}
+
+The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
+For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
+The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} we erase types and use a custom representation of names.
+There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
+
+\subsection{QuickCheck}
 
@@
 
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\fi
-
-\subsection{Haskell}
-\label{haskell}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ id\ expr                                 \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ id\ \left[ alt \right]            \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G} \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
-      alt\     \rightarrow\ & ( altcon,\ id,\ expr )                                 \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ id                                   \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ id\ expr                              \\
-                         |\ & \texttt{Rec}\ [ ( id,\ expr ) ]
-    \end{split}
-  \end{equation*}
-  where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
-\end{figure}
-
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} we erase types and use a custom representation of names.
-There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
-
-\iffalse
-
 Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
@@
 
+\begin{figure}
+  \begin{equation*}
+    \begin{split}
+      name\   =\ & N_1\ |\ \dots\ |\ N_n                               \\
+      type\   =\ & T_1\ |\ \dots\ |\ T_m\ |\ type \rightarrow type   \\
+      sig\    =\ & \{(n_i | (name, type); sig                                     \\
+      expr_{\mathcal{s}}\ =\ & v    : t\ \text{if}\ v : t \in s                        \\
+              |\ & f\ x : b\ \text{if}\ f : a \in expr_s \wedge x : a \in expr_s \\
+      conj_s\ =\ & \{ a =_t b \mid a : t \in expr_s \wedge b : t \in expr_s \}
+    \end{split}
+  \end{equation*}
+
+  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_i$ and types $T_i$.}
+  \label{coresyntax}
+\end{figure}
+
 The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
@@
   Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO: Related work reference \fi.
+Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO{2015-11-15}: Related work reference \fi.
 In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.

### 2015-11-16 TransferReport/contributions.tex
Updates

@@
-\section{Contributions}\label{current}
+\section{Contributions}
+\label{current}
 
@@
 TODO
-
-3 Recurrent clustering
-
 Good analysis of differences with ML4PG/ACL2ML
 \fi
+
+\subsection{Recurrent Clustering}
+\label{recurrentclustering}
+
+We take the \emph{recurrent clustering} approach found in ML4PG and ACL2(ml), and implement a variant in the context of Haskell theory exploration.
+Here we give a brief description of recurrent clustering, describe our algorithm, and compare its similarity and differences to ML4PG and ACL2(ml).
+
+\subsubsection{Overview}
+
+The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs.
+In our case, we want compare the abstract syntax trees of Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions.
+For example, consider the following expressions which are identical except for an identifier:\footnote{The \texttt{(\&\&)} function is boolean AND, \texttt{(++)} appends lists and \texttt{map} applies a function to the elements of a list.}
+
+\begin{equation*}
+  \begin{split}
+    X =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.(\&\&)"})) \\
+    Y =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.(++)"})) \\
+    Z =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.map"}))
+  \end{split}
+\end{equation*}
+
+Two simple methods for comparison are to ignore identifiers completely, or to compare them for equality.
+Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).
+
+Ideally we would like something in between these two extremes, which requires a more sophisticated notion of similarity for identifiers.
+In the above example, we may intuitively consider \texttt{(\&\&)} and \texttt{(++)} to be closer to each other than to \texttt{map}; both in terms of \emph{definition} (e.g.
+only the former can be specified by pattern-matching on their first argument), and by their \emph{properties} (e.g.
+only the former can form monoids).
+
+Recurrent clustering tackles this problem in a straightforward way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
+This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
+For this recursive process to be well-founded, we impose a topological ordering on expressions based on their dependencies (the expressions they reference).
+This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level.
+
+\subsubsection{Algorithm}
+
+\begin{figure}
+  \begin{haskell}
+    type Env = [Local]
+    type Db  = [(Global, Int)]
+
+    data RoseTree a = Rose a [RoseTree a]
+
+    exprTree :: Env -> Db -> Expr -> RoseTree Int
+    exprTree env db e = case e of
+      Var (Global i) -> Rose (fromMaybe sUNKNOWN (lookup db  i)) []
+      Var (Local  i) -> Rose (fromMaybe sUNKNOWN (lookup env i)) []
+      Lit (LitNum _) -> Rose (sLITNUM) []
+      Lit (LitStr _) -> Rose (sLITSTR) []
+      App  e1 e2     -> Rose (sAPP) [exprTree    env  db e1,
+                                     exprTree    env  db e2]
+      Lam  i  e      -> Rose (sLAM) [exprTree (i:env) db e1]
+      Let  bs e      -> Rose (sLET)
+      Case e i alts  -> Rose (sCASE) (exprTree env db e : map (altTree (i:env) db) alts)
+
+    altTree :: Env -> Db -> Alt -> Rose Int
+    altTree env db alt = case alt of
+      (DataAlt _, vs, e) -> exprTree (vs ++ env) db e
+      (LitAlt  _, _,  e) -> exprTree env db e
+      (Default,   _,  e) -> exprTree env db e
+
+    closedTree :: Db -> Core -> RoseTree Int
+    closedTree = exprTree []
+  \end{haskell}
+  \caption{Feature extraction for Haskell Core, using recurrent clustering an k-means.}
+  \label{featureextractionalgorithm}
+\end{figure}
+
+Our feature extraction algorithm is given in figure \ref{featureextractionalgorithm}. Here we highlight some key features:
+
+\begin{itemize}
+  \item Since they are comparatively rare, we ignore the particular value of each literal.
+  \item Since we are currently focused on expressions, we ignore particular data constructors and types.
+  \item If we cannot find a particular identifier, we use the sentinel value \hs{sUNKNOWN}.
+In practice, this only occurs for mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
+  \item
+\end{itemize}
+
+\subsubsection{Comparison}
+
+Our algorithm is most similar to that of ML4PG, as our transformation maps each element in a tree to a distinct cell in its matrix.
+In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.
+
+There are two major differences between our algorithm and that of ML4PG: mutual-recursion and types.
+
+The special handling required for mutual recursion is discussed above (namely, topological sorting of expressions and the \hs{sUNKNOWN} sentinel).
+Such handling is not present in ML4PG, since the Coq code it analyses must, by virtue of the language, be written in dependency order to begin with.
+Coq \emph{does} have limited support for mutually-recursive functions, of the following form:
+
+\begin{coqblock}
+Fixpoint even n := match n with
+                       | O   => true
+                       | S m => odd m
+                   end
+    with odd  n := match n with
+                       | O   => false
+                       | S m => even m
+                   end.
+\end{coqblock}
+
+However, this is relatively uncommon and unsupported by ML4PG.
+
+\iffalse
+TODO
+Replacing mutually-recursive calls (i.e. unknown global references) by the \hs{sUNKNOWN} sentinel, has the effect of denotes an has a similar effect to a consistently
+\fi
+
+The more interesting differences come from our handling (or lack thereof) for types.
+Coq and ACL2 are at opposite ends of the typing spectrum, with the former treating types as first class entities of the language whilst the latter is untyped (or \emph{unityped}).
+In both cases, we have a \emph{single} language to analyse, by ML4PG and ACL2(ml) respectively.\footnote{ML4PG can also analyse Coq's \textsc{Ltac} meta-language.
+Haskell has its own meta-language, Template Haskell, but here we only consider the regular Haskell which it generates.}
+
+The situation is different for Haskell, where the type level is distinct from the value level, and there are strict rules for how they can influence each other.
+In particular, Haskell values can depend on types (via the type class mechanism) but types cannot depend on values.
+
+In our initial approach, we restrict ourselves to the value level. This has several consequences:
+
+\begin{itemize}
+  \item We cannot distinguish between data constructors, other than using exact equality.
+  \item The arity of Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, there may be spurious after erasure rather than the Our Core expressions may contain type-level operation
+\end{itemize}
+
+effectively a collection contains  As a sub-set of Common Lisp, ACL2 is an untyped language; whilst Coq makes heavy use of types as first-class values.
+Haskell differs from both (the ACL2 sub-set of) Common Lisp and Coq (by ML4PG) analyses an untyped language, Haskell effectively lives between the two extremes of untyped programming (as analysed by ACL2(ml)) and dependently typed programming (Coq (which is dependently typed) a sense, Haskell ACL2(ml)

### 2015-11-16 TransferReport/implementation.tex
Updates

@@
 Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \hspec already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
+Indeed, \hspec{} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
 However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
@@
 
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \qspec.
+In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \qspec{}.
 Those which fail, due to the reasons above and others, are discarded.
@@
 
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec to explore.
+  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec{} to explore.
 
-  \item{Reflection}: As mentioned above, \hspec uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+  \item{Reflection}: As mentioned above, \hspec{} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
 Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
@@
 
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec, based on the databases constructed by \textsc{AstPlugin} described above.
+Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec{}, based on the databases constructed by \textsc{AstPlugin} described above.
 Features of \textsc{MLSpec} include:
@@
 [t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \qspec is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
+[t]}, a testing-based system like \qspec{} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
 Maybe t} without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck by attempting to instantiate all type variables to \hs{Integer}.
+Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
 We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
@@
 
-  \item{Variable definition}: Once a \qspec theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec's exhaustive search).
+  \item{Variable definition}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).
 
@@
 
-% TODO: Future experiments will test this by comparing the throughput of \qspec with and without the \textsc{ML4HS} pre-processor.
+% TODO: Future experiments will test this by comparing the throughput of \qspec{} with and without the \textsc{ML4HS} pre-processor.
 

### 2015-11-17 TransferReport/background.tex
Updates

@@
 \label{background}
-
-\iffalse
-TODO{2015-11-14}
-
-Background
-
-2.1 Introduce language signature
- b) stuff from QuickCheck/Spec
-
-2.2? Machine learning and feature extraction
-~1 page
-
-2.3: QuickSpec, HipSpec
-\fi
-
-\iffalse
-
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\fi
 
@@
   \end{equation*}
-  where:
+  Where:
   \begin{tabular}[t]{l @{ $=$ } l}
@@
 
-The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc.
-For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} we erase types and use a custom representation of names.
+We decided to focus our theory exploration research on Haskell as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
+
+Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
+Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, is based on a variant of the polymorphic $\lambda$-calculus known as \fc{}.
+For simplicity, we will focus on this Core language rather than the relatively large and complex syntax of Haskell proper.
+For a full treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
+
+The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names.
 There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
@@
 \subsection{QuickCheck}
+\label{quickcheck}
 
@@
 
-However, this is \emph{still} not enough: since this function only uses generic numeric operations, it will be polymorphic; allowing all numeric types\footnote{Its type will be of the form \hs{forall t.
+However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t.
 Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
-There will also be extra contraints like \hs{Eq} due to the use of \hs{==} in the unit tests.}.
-If we provide a fractional value, the function will again diverge.
+In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge.
 Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
 
-Haskell offers a partial solution to this problem in the form of \emph{property checking}\footnote{See section \ref{propertychecking} for more information}.
+Haskell offers a partial solution to this problem in the form of \emph{property checking}.
 Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
@@
 
-The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with closed values. In the \hs{factorial} example, this manifests in two ways:
+The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:
 
 \begin{itemize}
-  \item \qcheck{} cannot test polymorphic functions; they type must be \emph{monomorphised} first (instantiated to a particular concrete type).
+  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type).
 This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
@@
 
-\iffalse
-
-Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP).
-Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).
-
-In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems.
-Since many theories will admit an infinite number of trivial theorems (e.g.
-$\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements.
-This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.
-
-\fi
-
-\subsection{Interactive Theorem Proving}
-\label{itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
-
 \subsection{Theory Exploration}
@@
     \begin{split}
-      name\   =\ & N_1\ |\ \dots\ |\ N_n                               \\
-      type\   =\ & T_1\ |\ \dots\ |\ T_m\ |\ type \rightarrow type   \\
-      sig\    =\ & \{(n_i | (name, type); sig                                     \\
-      expr_{\mathcal{s}}\ =\ & v    : t\ \text{if}\ v : t \in s                        \\
-              |\ & f\ x : b\ \text{if}\ f : a \in expr_s \wedge x : a \in expr_s \\
-      conj_s\ =\ & \{ a =_t b \mid a : t \in expr_s \wedge b : t \in expr_s \}
+      names\       =\ & \{ N_1, \dots, N_n \}                                                                  \\
+      types\       =\ & \{ T_1, \dots, T_m \} \cup \{a \rightarrow b \mid a \in types \wedge b \in types \} \\
+      exprs\       =\ & names \cup \{ f\ x \mid f \in exprs \wedge x \in exprs \wedge f : a \rightarrow b \wedge x : a \} \\
+      classes\     =\ & \{c \mid \forall x y \in c, x \in exprs \wedge y \in exprs \wedge x =_{QC} y \} \\
+      conjectures\ =\ & \bigcup_{c \in classes}\{ \ulcorner a = b \urcorner \mid a \in c \wedge b \in c \}
     \end{split}
@@
 
-  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_i$ and types $T_i$.}
-  \label{coresyntax}
+  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_1$ to $N_n$ and types $T_1$ to $T_m$.
+We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
+  \label{qspecmodel}
 \end{figure}
 
-The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g.
-a user-provided conjecture to prove).
-Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises.
-Of course, a key question to ask is what do we mean by ``interesting''?
-There is no single answer, although many approximate measures have been proposed.
-The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.
+In this work we focus on the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
+In \citep{warburtonscaling} we identify the method of conjecture generation as a key characteristic of any theory exploration system, and in particular lament the existing reliance on brute force methods.
 
-Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}.
-\emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.
+We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code.
+These are found through the process modelled in figure \ref{qspecmodel}: given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
 
-The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers.
-Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.
+The resulting set of $conjectures$ can be used in several ways: it can be simplified for direct presentation to the user, sent to a more rigorous system like \hspec{} and \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
 
-Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems.
-Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.
+\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
 
-Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}.
-By removing user interaction, these properties must be implemented by algorithms.
-In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.
+In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
+One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
 
-As an example, \qspec{} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations.
-The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
-when we already know $0 + x = x$.
-Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
-(version 2), the conjecture generation is performed in a brute-force way.
+We attempt to mitigate the cost of running \qspec{}, by providing a machine learning layer analogous to that of premise selection in theorem proving.
+Of particular concern is the set of expressions $exprs$, which grows exponentially as the size of the expressions increases, and hence slows down the rest of the algorithm.
+Although complete, the current enumeration approach is wasteful: many combinations are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
+The key idea of premise selection is to make such choices automatically, by only including those expressions which are considered \emph{relevant} to the problem.
 
-Although \qspec{} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \hspec{} and \textsc{Hipster}.
+\subsection{Clustering and Feature Extraction}
+\label{featureextraction}
 
-\hspec{} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}.
-It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \qspec{}.
-\qspec{}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
-This approach works well as a lemma generation system, making \hspec{} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}.
-\hspec{} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \hspec{} alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
+Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources.
+The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
+We describe this approach in more details in \S \ref{relevance}.
+Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
 
-In fact, there are similarities between the way a TE system like \hspec{} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-There are likely lessons that each can learn from the other's approach to term generation.
+Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
+The reason is that theory exploration has no distinct \emph{goal} to compare expressions against.
+Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}.
+A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
+See \S \ref{clustering} for more information.
 
-We identify the following areas for researching potential improvements in existing TE systems such as \hspec{}:
+Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller sets more amenable to brute force enumeration.
 
-\begin{description}
-\item{Term generation}:
-  Enumerating all type-correct terms is a brute-force solution to this question.
-Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO{2015-11-15}: Related work reference \fi.
-In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
-  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.
+Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
+This has two benefits:
 
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
+\begin{itemize}
+  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
+This reduces the size of the machine learning problem, improving efficiency (e.g.
+running time).
+  \item We avoid learning irrelevant details, such as the text encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
+\end{itemize}
 
-% Theory exploration is similar to \emph{experimental mathematics}
+Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
+Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
 
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
+As an example, say we want to learn relationships between the following program fragments:
 
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+\begin{haskell}
+data Maybe a = Nothing | Just a
 
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
+data Either a b = Left a | Right b
+\end{haskell}
+
+We might hope our algorithm discovers relationships like:
+
+\begin{itemize}
+  \item Both are valid Haskell code.
+  \item Both describe datatypes.
+  \item Both datatypes have two constructors.
+  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
+  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
+  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
+  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
+\end{itemize}
+
+However, this is too optimistic.
+Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
+Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
+
+In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
+Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
+
+There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
+Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
+\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
+If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
+
+Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
+This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
+
+In our case, we will use a novel feature extraction algorithm, described in \S \ref{recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.

### 2015-11-17 TransferReport/contributions.tex
Updates

@@
 \section{Contributions}
-\label{current}
-
-\iffalse
-TODO
-Good analysis of differences with ML4PG/ACL2ML
-\fi
+\label{contributions}
 
@@
 \subsubsection{Algorithm}
+\label{algorithm}
 
 \begin{figure}
-  \begin{haskell}
-    type Env = [Local]
-    type Db  = [(Global, Int)]
+  \begin{lstlisting}[language=Haskell, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
+    data RoseTree a = Node a [RoseTree a]
 
-    data RoseTree a = Rose a [RoseTree a]
+    eRT :: [Local] -> [[Global]] -> Expr -> RoseTree Feature
+    eRT env db e = case e of
+      Var (Global i) -> Node (lookupG db  i) []
+      Var (Local  i) -> Node (lookupL env i) []
+      Lit (LitNum _) -> Node sLITNUM []
+      Lit (LitStr _) -> Node sLITSTR []
+      Lam  i  e      -> Node sLAM [eRT (i:env) db e]
+      App  e1 e2     -> Node sAPP [eRT    env  db e1,
+                                   eRT    env  db e2]
+      Let  bs e      -> Node sLET (map (bRT (ids bs:env) db) bs ++
+                                        eRT (ids bs:env) db  e
+      Case e i alts  -> Node sCASE (eRT    env  db  e :
+                               map (aRT (i:env) db) alts)
 
-    exprTree :: Env -> Db -> Expr -> RoseTree Int
-    exprTree env db e = case e of
-      Var (Global i) -> Rose (fromMaybe sUNKNOWN (lookup db  i)) []
-      Var (Local  i) -> Rose (fromMaybe sUNKNOWN (lookup env i)) []
-      Lit (LitNum _) -> Rose (sLITNUM) []
-      Lit (LitStr _) -> Rose (sLITSTR) []
-      App  e1 e2     -> Rose (sAPP) [exprTree    env  db e1,
-                                     exprTree    env  db e2]
-      Lam  i  e      -> Rose (sLAM) [exprTree (i:env) db e1]
-      Let  bs e      -> Rose (sLET)
-      Case e i alts  -> Rose (sCASE) (exprTree env db e : map (altTree (i:env) db) alts)
+    aRT :: [Local] -> [[Global]] -> Alt -> RoseTree Feature
+    aRT env db alt = case alt of
+      (DataAlt _, vs, e) -> eRT (vs ++ env) db e
+      (LitAlt  _, _,  e) -> eRT env db e
+      (Default,   _,  e) -> eRT env db e
 
-    altTree :: Env -> Db -> Alt -> Rose Int
-    altTree env db alt = case alt of
-      (DataAlt _, vs, e) -> exprTree (vs ++ env) db e
-      (LitAlt  _, _,  e) -> exprTree env db e
-      (Default,   _,  e) -> exprTree env db e
+    bRT :: [Local] -> [[Global]] -> Bind -> RoseTree Feature
+    bRT env db b = case b of
+      NonRec i e -> eRT env db e
+      Rec es     -> Node sREC (map (eRT env db . snd) es)
 
-    closedTree :: Db -> Core -> RoseTree Int
-    closedTree = exprTree []
-  \end{haskell}
-  \caption{Feature extraction for Haskell Core, using recurrent clustering an k-means.}
+    level :: Int -> RoseTree Feature -> [Feature]
+    level 0 (Node x xs) = [x]
+    level n (Node x xs) = concatMap (level (n-1)) xs
+
+    rt :: [[Global]] -> Expr -> [Feature]
+    rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
+
+    recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
+    recurrentCluster = go ([],[])
+      where go (fs, db) []       = db
+            go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
+                                       db' = kMeans fs'
+                                    in go (fs', db') ess
+            extract db (i, e) = (i, rt db e)
+  \end{lstlisting}
+  \caption{Feature extraction for Haskell Core.
+We send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive.
+Syntax trees are first converted into \hs{RoseTree}s of \hs{Feature}s, these trees are converted to matrices, then flattened into feature vectors, then \hs{kMeans} clusters the feature vectors.
+The result is a list of clusters, naming their elements.
+More details are given in \S \ref{algorithm}.}
   \label{featureextractionalgorithm}
@@
   \item Since we are currently focused on expressions, we ignore particular data constructors and types.
-  \item If we cannot find a particular identifier, we use the sentinel value \hs{sUNKNOWN}.
-In practice, this only occurs for mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
-  \item
+  \item The \hs{lookupL} and \hs{lookupG} functions return the index containing their argument, if found.
+In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors.
+For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features.
+  \item \hs{lookupL} and \hs{lookupG} return a sentinel value \hs{sUNKNOWN} if the given ID is not found.
+In practice, this only occurs for global, mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
+  \item The \hs{Feature} type is abstract, but in practice it will be \hs{Float}.
+  \item Our algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
+The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
 \end{itemize}
@@
 
-\iffalse
-TODO
-Replacing mutually-recursive calls (i.e. unknown global references) by the \hs{sUNKNOWN} sentinel, has the effect of denotes an has a similar effect to a consistently
-\fi
-
 The more interesting differences come from our handling (or lack thereof) for types.
@@
 \begin{itemize}
-  \item We cannot distinguish between data constructors, other than using exact equality.
-  \item The arity of Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, there may be spurious after erasure rather than the Our Core expressions may contain type-level operation
+  \item Although they are values, we cannot distinguish between data constructors, other than using exact equality.
+  \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot completely erase type-level parameters unless they are fully applied.
+This can cause a function's Core arity to be greater than its Haskell arity.
 \end{itemize}
 
-effectively a collection contains  As a sub-set of Common Lisp, ACL2 is an untyped language; whilst Coq makes heavy use of types as first-class values.
-Haskell differs from both (the ACL2 sub-set of) Common Lisp and Coq (by ML4PG) analyses an untyped language, Haskell effectively lives between the two extremes of untyped programming (as analysed by ACL2(ml)) and dependently typed programming (Coq (which is dependently typed) a sense, Haskell ACL2(ml)
+There is certainly promise in including types in our analysis, by pairing every term with its type as in ML4PG.
+This will allow fine-grained distinction of expressions which are otherwise identical, especially data constructors.

### 2015-11-17 TransferReport/intro.tex
Updates

@@
 
-Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive.
+Functional programming is also well suited to more radical practices, such as \emph{property checking} (as popularised by \qcheck{}) and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet can be prohibitively expensive.
 Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
 
-\iffalse
-TODO
 Our contributions are:
@@
   \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
-  \item A
+  \item A feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf machine learning algorithms.
+  \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
 \end{itemize}
-\fi
+
+We begin in \S \ref{background} by providing a formal context for analysing Haskell expressions (\S \ref{haskell}) and the results of theory exploration systems like \qspec{} (\S \ref{theoryexploration}).
+We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{quickcheck}), as well as the machine learning approaches we are building on (\S \ref{featureextraction}).
+We discuss our contributions in more depth in \S \ref{contributions}, and provide implementation details \S \ref{implementation}.
+A variety of related work is surveyed in \S \ref{related}, and we give several potential directions for future research in \S \ref{future} before concluding in \S \ref{conclusion}.

### 2015-11-17 TransferReport/related.tex
Updates

@@
 
-Firstly, however, we give an overview of the Haskell programming language in \S \ref{haskell}.
-This allows us to establish many concepts in a straightforward software context, which we can then re-use in the potentially less-familiar realm of interactive theorem proving.
-We decided to focus on Haskell as it has mature, state-of-the-art theory exploration implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
-
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact the \textsc{GHC} compiler uses (variants of) the polymorphic $\lambda$-calculus \emph{System F} as an intermediate representation (referred to as \emph{GHC Core}).
-For a detailed exposition of the language syntax and semantics, you can consult the Haskell Language Report \citep{marlow2010haskell} \footnote{Version 2010 was current as this was written} or a textbook such as \citep{thompson1999haskell}; here we will focus on those language features which make it particularly suited for our formal reasoning purposes:
+Haskell is a convenient choice for our purposes for several reasons. Here we discuss the relevant language features from a high-level:
 
@@
 
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \citep{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
 \paragraph{Truncation and Padding}
@@
 \fi
+
+\subsection{Interactive Theorem Proving}
+\label{itp}
+
+ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
+
+$$ C \colon (P \times S) \rightarrow Boolean $$
+
+The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
+
+$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
+
+It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
+In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
+There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
+
+This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
+Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
+With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
+
+Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
+In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
+For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
+Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
+This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
+One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
+
+Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
+see \citep{hales2015formal}).
+One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
+The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
+
+\subsection{Automated Theorem Proving}
+\label{atp}
+
+ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
+By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
+In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
+This is the most striking difference from ITP systems (e.g.
+Coq, Agda and Isabelle) which operate in higher-order logics.
+
+In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
+Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
+The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
+
+Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.

### 2015-11-17 TransferReport/report.tex
Updates

@@
 \usepackage{amssymb}
+\usepackage{algorithm2e}
 \usepackage[T1]{fontenc}
@@
   numberstyle=\ttfamily\tiny\color[gray]{0.3},
+  columns=flexible,
   belowcaptionskip=\bigskipamount,
@@
   commentstyle=\it,
-  stringstyle=\mdseries\rmfamily,
+  stringstyle=\mdseries\ttfamily,
   showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
+  keepspaces=true,
+  keywordstyle=\bfseries\ttfamily,
+  basicstyle=\small\ttfamily,
   showstringspaces=false,
@@
 
-\newenvironment{haskell}
-  {
-    \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  }
-  {
-    \end{lstlisting}
-  }
+\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
 
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+\lstnewenvironment{haskell}{%
+  \lstset{language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth}}{}
+
+\lstnewenvironment{coqblock}{%
+  \lstset{language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth}}{}
 
@@
 
-\title{Interestingness Measures for Theory Exploration}
+\title{Improving Haskell Theory Exploration}
 
@@
 \begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
+Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
+Current techniques are limited by their computational cost, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
 \end{abstract}
@@
 \input{background.tex}
-\input{examples.tex}
 \input{contributions.tex}
@@
 
+\iffalse
 \section{First Results}
 
-\iffalse
 TODO
@@
 
-\iffalse
-
-QuickSpec: extend or extinguish?
-
-Improve and find other use cases/scenarios for clustering and feature extraction
-
-Other directions for Theory Exploration?
-
-How about systems based on term rewriting, logic programming, etc.?
-\fi
-
 \section{Conclusion}

### 2015-11-23 TransferReport/background.tex
Added more feedback

@@
 \section{Background}
-\label{background}
+\label{sec:background}
 
 \subsection{Haskell}
-\label{haskell}
+\label{sec:haskell}
 
@@
   \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{coresyntax}
+  \label{fig:coresyntax}
 \end{figure}
@@
 
-The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names.
+The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names.
 There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
@@
 \subsection{QuickCheck}
-\label{quickcheck}
+\label{sec:quickcheck}
 
@@
 life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests (see \ref{haskell} for more details on Haskell):
+To see why, consider the following Haskell function, along with some unit tests (see \ref{sec:haskell} for more details on Haskell):
 
@@
 These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{atp} for more information on automated theorem proving.}
+a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{sec:atp} for more information on automated theorem proving.}
 
 \subsection{Theory Exploration}
-\label{theoryexploration}
+\label{sec:theoryexploration}
 
@@
 We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
-  \label{qspecmodel}
+  \label{fig:qspecmodel}
 \end{figure}
@@
 We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code.
-These are found through the process modelled in figure \ref{qspecmodel}: given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
+These are found through the process modelled in figure \ref{fig:qspecmodel}: \iffalse TODO: Figure just shows the syntax, it does not model any process \fi given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
 
@@
 
+\iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
+\iffalse TODO: This detailed comparison with premise selection is opaque without definiing/explaining what premise selection is \fi
+
 \subsection{Clustering and Feature Extraction}
-\label{featureextraction}
+\label{sec:featureextraction}
 
@@
 The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{relevance}.
+We describe this approach in more details in \S \ref{sec:relevance}.
 Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
@@
 Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}.
-A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-See \S \ref{clustering} for more information.
+\iffalse TODO: every term in the given signature?
+\fi A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
+See \S \ref{sec:clustering} for more information.
 
 Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller sets more amenable to brute force enumeration.
+
+\iffalse TODO: Signatures or terms? \fi
+
+\iffalse TODO: I would play up the clustering intelligence.
+Clustering is not just about breaking up, it is about discovering significant patterns in data.
+By forgetting this, you make motivation for your work sound too ``small'' \fi
 
@@
 
-In our case, we will use a novel feature extraction algorithm, described in \S \ref{recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.
+In our case, we will use a novel feature extraction algorithm, described in \S \ref{sec:recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[scale=0.75, level/.style={sibling distance=60mm/#1}]
+      % Tree
+      \node {\texttt{Lam}}
+        child {node {\texttt{"x"}}}
+        child {node {\texttt{App}}
+          child {node {\texttt{Var}}
+            child {node {\texttt{Global}}
+              child {node {\texttt{"base:Prelude.not"}}}}}
+          child {node {\texttt{Var}}
+            child {node {\texttt{Local}}
+              child {node {\texttt{"x"}}}}}};
+
+      % Arrow
+      %\node[single arrow,
+      %      draw=black,
+      %      fill=black!10,
+      %      minimum height=2cm,
+      %      shape border rotate=0] at (0,-1) {};
+      %\draw[-latex] (A.east) -- (B.west);
+
+      % Feature vector
+
+  \end{tikzpicture}
+
+  \caption{}
+
+  \label{fig:featureextractionpic}
+\end{figure}
+
+\iffalse
+\subsection{Clustering}
+% TODO
+\fi

### 2015-11-23 TransferReport/implementation.tex
Updates
Added more feedback

@@
 \section{Implementation}
+\label{sec:implementation}
 
@@
 
-\subsection{Circular Convolution}\label{circularconvolution}
+\subsection{Circular Convolution}
+\label{sec:circularconvolution}
 
@@
 
-\subsection{\texttt{nix-eval}}\label{nixeval}
+\subsection{\texttt{nix-eval}}
+\label{sec:nixeval}
 
@@
 
-\subsection{\textsc{ASTPlugin}}\label{astplugin}
+\subsection{\textsc{ASTPlugin}}
+\label{sec:astplugin}
 
@@
 \begin{description}
-  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
+  \item{Type variables}: In \S \ref{sec:haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
 In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
@@
 
-  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module.
+  \item{Violating encapsulation}: As discussed in \S \ref{sec:haskell}, Haskell allows names to be \emph{private} to a module.
 When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
@@
   \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
+As mentioned in \S \ref{sec:featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
 
@@
 
-  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
+  \item{Counterexample Generation}: As mentioned in \S \ref{sec:tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
 Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
 as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail.
+As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail.
 Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
@@
 
-\subsection{\textsc{MLSpec}}\label{mlspec}
+\subsection{\textsc{MLSpec}}
+\label{sec:mlspec}
 
@@
 
-\subsection{\textsc{ML4HS}}\label{ml4hs}
+\subsection{\textsc{ML4HS}}
+\label{sec:ml4hs}
 
@@
 
-Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
+Inspired by the use of premise selection (see \S \ref{sec:relevance}) to reduce the search space in ATP,
 we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
 
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.
+We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{sec:clustering}), is an effective method for performing this separation.
 

### 2015-11-23 TransferReport/intro.tex
Added more feedback

--- 2015-11-12 TransferReport/intro.tex
@@
 
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
+\iffalse TODO: sounds strange...
+what does it mean for software to be capable or uncapable?
+of what?
+Rephrase?
+\fi As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
 The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
-For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
+\iffalse TODO: I am not sure the example makes sense without explaining what effects are and why separating them helps.
+Rather than going there, perhaps just say about declarative style of programming and types expressing prperties of computations?
+\fi For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
 
@@
 
-Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive.
+Functional programming is also well suited to less-widespread practices, such as \emph{property checking} (as popularised by \qcheck{}) and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet can be prohibitively expensive.
 Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
 
-\iffalse
-TODO
 Our contributions are:
@@
   \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
-  \item A
+  \item A feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf machine learning algorithms. \iffalse TODO: say how this item fits in the TE agenda. \fi
+  \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
 \end{itemize}
-\fi
+
+We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and the results of theory exploration systems like \qspec{} (\S \ref{sec:theoryexploration}).
+We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}).
+We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details \S \ref{sec:implementation}.
+A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future} before concluding in \S \ref{sec:conclusion}.

### 2015-11-23 TransferReport/related.tex
Added more feedback

@@
 \section{Related Work}
-\label{related}
+\label{sec:related}
 
 \subsection{Haskell}
-\label{haskelldesc}
+\label{sec:haskelldesc}
 
@@
 This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
+in the style of interactive theorem proving (see \S \ref{sec:itp}), as these results may be invalidated by seemingly innocuous code changes.
+We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{sec:theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
 Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
@@
 \subsubsection{Artificial Curiosity}
-\label{curiosity}
+\label{sec:curiosity}
 
@@
 
-  \label{wundt}
+  \label{fig:wundt}
 \end{figure}
@@
 The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
 
@@
 \subsubsection{Relevance Filtering}
-\label{relevance}
+\label{sec:relevance}
 
@@
 \subsubsection{Clustering}
-\label{clustering}
+\label{sec:clustering}
 
@@
 \subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
+\label{sec:conceptformation}
 
@@
 \subsection{Interactive Theorem Proving}
-\label{itp}
+\label{sec:itp}
 
@@
 \subsection{Automated Theorem Proving}
-\label{atp}
+\label{sec:atp}
 
@@
 
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.
+Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{sec:related}.

### 2015-11-23 TransferReport/report.tex
Added more feedback

@@
 Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
-Current techniques are limited by their computational cost, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
+Current techniques \iffalse TODO: it sounds as if you are talking about techniques of "formal analysis" here rather than techniques of TE.
+Reformulate.
+\fi are limited by their computational cost \iffalse TODO: Unclear what that means in this context.
+Computations of what is costly?
+why?
+same for "problem size" below \fi, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
 \end{abstract}
@@
 \section{Future Work}
-\label{future}
+\label{sec:future}
 
@@
 \section{Conclusion}
-\label{conclusion}
+\label{sec:conclusion}
 

### 2015-11-23 TransferReport/background.tex
More TODOs

@@
 \iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
-\iffalse TODO: This detailed comparison with premise selection is opaque without definiing/explaining what premise selection is \fi
+\iffalse TODO: This detailed comparison with premise selection is opaque without defining/explaining what premise selection is \fi
 
@@
 
-Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
+Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) \iffalse TODO: software vs logical expr?
+\fi expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
 This has two benefits:
@@
 \begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
+  \item \emph{Feature vectors} (ordered lists of features) are chosen to be more compact \iffalse TODO: unclear \fi than the data they're extracted from: feature extraction is \emph{lossy compression}.
 This reduces the size of the machine learning problem, improving efficiency (e.g.
@@
 Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
+\iffalse TODO: It depends on your feature extraction, and why bits?
+\fi Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
+\iffalse TODO: I do not see this has to be this way\fi
 
@@
 Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
+\iffalse TODO: not sure that's the right argument for a more-or-less common observation that ML & FE are coming together \fi \footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
 If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
@@
 In our case, we will use a novel feature extraction algorithm, described in \S \ref{sec:recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.
+
+\iffalse TODO: I have had a look now at section 5.
+I do not think feature extraction and clustering is explained/defined there, either.
+Seeing your contribution is feature extraction algorithms, you need to define feature vectors and clusters *here* \fi
 

### 2015-11-23 TransferReport/contributions.tex
Added more feedback
More TODOs

@@
 \section{Contributions}
-\label{contributions}
+\label{sec:contributions}
 
 \subsection{Recurrent Clustering}
-\label{recurrentclustering}
+\label{sec:recurrentclustering}
 
@@
 
-Recurrent clustering tackles this problem in a straightforward way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
+Recurrent clustering tackles this problem in a straightforward way: \iffalse TODO: What would be a non-straightforward way?
+\fi we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
 This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
@@
 \subsubsection{Algorithm}
-\label{algorithm}
+\label{sec:algorithm}
+
+\iffalse TODO: Not sure if you want to give a more conceptual explanation of the algorithm than the code in this figure? \fi
+
+\iffalse TODO: Connect this to figure 1 in your explanations \fi
+
+\iffalse TODO:
+ 1: Tree representation
+ 2. Feature vectors vs feature values
+ 3. Recurrence
+ 4. k-means algorithm
+\fi
 
@@
 The result is a list of clusters, naming their elements.
-More details are given in \S \ref{algorithm}.}
-  \label{featureextractionalgorithm}
+More details are given in \S \ref{sec:algorithm}.}
+  \label{fig:featureextractionalgorithm}
+  \iffalse TODO: Split this up, get rid of float warning \fi
 \end{figure}
 
-Our feature extraction algorithm is given in figure \ref{featureextractionalgorithm}. Here we highlight some key features:
+Our feature extraction algorithm is given in figure \ref{fig:featureextractionalgorithm}. Here we highlight some key features:
 

### 2015-11-23 TransferReport/implementation.tex
More TODOs

@@
 
-  \item{Counterexample Generation}: As mentioned in \S \ref{sec:tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
+  \item{Counterexample Generation}: As mentioned in \S \ref{sec:quickcheck}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
 Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.

### 2015-11-23 TransferReport/background.tex
Reformulated QuickSpec description

@@
 
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      names\       =\ & \{ N_1, \dots, N_n \}                                                                  \\
-      types\       =\ & \{ T_1, \dots, T_m \} \cup \{a \rightarrow b \mid a \in types \wedge b \in types \} \\
-      exprs\       =\ & names \cup \{ f\ x \mid f \in exprs \wedge x \in exprs \wedge f : a \rightarrow b \wedge x : a \} \\
-      classes\     =\ & \{c \mid \forall x y \in c, x \in exprs \wedge y \in exprs \wedge x =_{QC} y \} \\
-      conjectures\ =\ & \bigcup_{c \in classes}\{ \ulcorner a = b \urcorner \mid a \in c \wedge b \in c \}
-    \end{split}
-  \end{equation*}
-
-  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_1$ to $N_n$ and types $T_1$ to $T_m$.
-We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
-  \label{fig:qspecmodel}
-\end{figure}
+\newcommand{\blank}{\cdot}
 
 In this work we focus on the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-In \citep{warburtonscaling} we identify the method of conjecture generation as a key characteristic of any theory exploration system, and in particular lament the existing reliance on brute force methods.
+The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
 
-We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code.
-These are found through the process modelled in figure \ref{fig:qspecmodel}: \iffalse TODO: Figure just shows the syntax, it does not model any process \fi given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.
+We focus on \qspec{} \citep{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
+These conjectures are arrived at through the following stages:
 
-The resulting set of $conjectures$ can be used in several ways: it can be simplified for direct presentation to the user, sent to a more rigorous system like \hspec{} and \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
+\begin{enumerate}
+  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$.
+To ensure the list is finite, function applications are only nested up to a specified depth (e.g.
+3).
+  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
+  \item Each variable is instantiated to particular value, generated randomly by \qcheck{}.
+  \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
+If so, the class is split up to separate such members.
+  \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
+  \item A set of equations are then conjectured, relating each class's members.
+\end{enumerate}
+
+Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which is a can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
+
+As an example, we can consider a simple signature for Peano numerals, addition and multiplication:
+
+\begin{align*}
+  \Sigma_{\mathbb{N}} = \{ & 0     && : \mathbb{N}, \\
+                        & succ  && : \mathbb{N} \rightarrow \mathbb{N}, \\
+                        & plus  && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}, \\
+                        & times && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}\}
+\end{align*}
+
+Together with a set of variables, say $V_{\mathbb{N}} = \{x, y, z\}$, \qspec{}'s enumeration will resemble the following:
+
+\begin{multline}
+  terms_{\mathbb{N}} = \{0,\ succ,\ plus,\ times,\ x,\ y,\ succ(0),\ succ(x),\ succ(y),\ succ(z), \\
+                    plus(0,\blank),\ plus(x, \blank),\ \dots \}
+\end{multline}
+
+Notice that our functions are curried, where $f(x,\blank)$ indicates \emph{partial application} of $f$ to $x$.
+This is required as the construction of $terms$ applies functions to one argument at a time.
+
+These terms will be grouped into three classes, for $\mathbb{N}$, $\mathbb{N} \rightarrow \mathbb{N}$ and $\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}$.
+As the variables $x$, $y$ and $z$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until equations such as the following are conjectured (after simplification):
+
+% To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
+\begin{align*}
+                      plus(x, y) &= plus(y, x)
+                      plus(x, 0) &= x
+             plus(x, plus(y, z)) &= plus(y, plus(x, z))
+                     times(x, y) &= times(y, x)
+                     times(x, 0) &= 0
+           times(x, times(y, z)) &= times(y, times(x, z))
+                plus(x, succ(y)) &= succ(plus(x, y))
+               times(x, succ(y)) &= plus(x, times(x, y))
+            times(x, plus(y, y)) &= times(y, plus(x, x))
+  plus(times(x, y), times(x, z)) &= times(x, plus(y, z))
+\end{align*}
 
@@
 The reason is that theory exploration has no distinct \emph{goal} to compare expressions against.
-Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}.
-\iffalse TODO: every term in the given signature?
-\fi A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
+Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
+A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
 See \S \ref{sec:clustering} for more information.

### 2015-11-23 TransferReport/haskell_example/src/Example.hs

@@ -30,3 +30,21 @@ negative  f = f (-1) ==    f 1
 factorial2 0 = 1
 factorial2 n = let nPos = abs n
                 in nPos * factorial2 (nPos - 1)
+
+f1_tail :: Int -> Int
+f1_tail = go 1
+  where go acc 0 = acc
+        go acc n = go (n*acc) (n-1)
+
+f2_tail :: Int -> Int
+f2_tail = go 1
+  where go acc 0 = acc
+        go acc n = let nPos = abs n in go (nPos*acc) (nPos-1)
+
+f1 :: Int -> Int
+f1 = factorial1
+
+f2 :: Int -> Int
+f2 = factorial2
+
+foo n = -n

### 2015-11-23 TransferReport/haskell_example/src/QuickSpecExample.hs

@@ -1,0 +1,37 @@
+module QuickSpecExample where
+
+import Numeric.Natural
+import Test.QuickSpec
+
+i2N :: Int -> Natural
+i2N = fromInteger . toInteger
+
+sigma = [blind0 "()"      (),
+         blind0 "Z"       ([] :: [()]),
+         blind1 "length"  ((i2N . length) :: [()] -> Natural),
+         blind2 "++"      ((++) :: [()] -> [()] -> [()]),
+         --blind1 "reverse" (reverse :: [()] -> [()]),
+         blind2 "s"       ((:) :: () -> [()] -> [()]),
+         blind0 "F"       False,
+         blind0 "T"       True,
+         blind0 "0"       (0 :: Natural),
+         blind1 "S"       ((+ 1) :: Natural -> Natural),
+         blind2 "+"       ((+) :: Natural -> Natural -> Natural),
+         blind2 "*"       ((*) :: Natural -> Natural -> Natural),
+
+         vars ["b1", "b2", "b3"] True,
+         vars ["l1", "l2", "l3"] [()],
+         vars ["n1", "n2", "n3"] (0 :: Natural),
+         vars ["u1"]             (),
+
+         observer1 ((i2N . length) :: [()] -> Natural),
+         observer1 (id :: () -> ()),
+         observer1 ((`mod` 10) :: Natural -> Natural),
+         observer1 (id :: Bool -> Bool)]
+
+nat = [fun0 "0"     (0 :: Natural),
+       fun1 "succ"  ((1+) :: Natural -> Natural),
+       fun2 "plus"  ((+) :: Natural -> Natural -> Natural),
+       fun2 "times" ((*) :: Natural -> Natural -> Natural),
+
+       vars ["x", "y", "z"] (0 :: Natural)]

### 2015-11-23 TheoryExploration/interesting.tex
Keep interesting.tex in git

--- 2015-11-06 TransferReport/interesting.tex
@@
 In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
+To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{sec:haskell}.
 Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
@@
 
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
+We begin our undertaking in \S \ref{sec:background} by introducing the Haskell language, as well as the relevant fields of verification for context.
 We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
+A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{sec:related}.
 We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
@@
 
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
+We discuss our present contributions in \S \ref{sec:current} and future research directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.
 
@@
 \item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
+  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{sec:relatedwork}.
 Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
@@
 \section{Exploration in Theorem Proving}
-\label{examples}
+\label{sec:examples}
 
@@
 \footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
+See \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
 
@@
 
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
+\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
 
@@
 
-\subsection{Tests} \label{tests}
+\subsection{Tests} \label{sec:tests}
 
@@
 \section{Related Work}
-\label{related}
+\label{sec:related}
 
@@
 The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
 
@@
 
-\paragraph{Feature Extraction}\label{featureextraction}
+\paragraph{Feature Extraction}\label{sec:featureextraction}
 

### 2015-11-23 TransferReport/background.tex
Removed old FE cruft

@@
 
-Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller sets more amenable to brute force enumeration.
-
-\iffalse TODO: Signatures or terms? \fi
+Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller ones more amenable to brute force enumeration.
 
@@
 
-Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) \iffalse TODO: software vs logical expr?
-\fi expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
+Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
 This has two benefits:
@@
 \begin{itemize}
-  \item \emph{Feature vectors} (ordered lists of features) are chosen to be more compact \iffalse TODO: unclear \fi than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the machine learning problem, improving efficiency (e.g.
+  \item \emph{Feature vectors} (ordered lists of features) are chosen to represent the relevant information in a more compressed form than the raw data (for example, replacing descriptive identifier names with sequential numbers).
+This reduces the input size of the machine learning problem, improving efficiency (e.g.
 running time).
@@
 Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{haskell}
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{haskell}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-\iffalse TODO: It depends on your feature extraction, and why bits?
-\fi Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-\iffalse TODO: I do not see this has to be this way\fi
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\iffalse TODO: not sure that's the right argument for a more-or-less common observation that ML & FE are coming together \fi \footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
 
@@
 
-\iffalse
 \subsection{Clustering}
+\label{sec:clustering}
 % TODO
-\fi

### 2015-11-23 TransferReport/future.tex
Added future

\section{Future Work}
\label{sec:future}

\subsection{Extensions}
\label{sec:preprocessing}

Our use of clustering to pre-process \qcheck{} signatures necessarily involves many decisions and tradeoffs to be made. There are many alternative approaches which are ripe for investigation:

\begin{itemize}
  \item Improvements to our feature extraction algorithm, in particular to handle types.
  \item Alternative clustering algorithms.
  \item \emph{Feature learning} uses machine learning algorithms in place of hand-coded feature extraction algorithms such as ours. A comparison of our hand-picked features against a selection of learned representations would be a useful indication of the importance that understanding the expression language may or may not have on identifying salient aspects of expressions.
  \item
\end{itemize}

\subsection{Interestingness}
\label{sec:interestingness}

It is important to consider the question: \emph{what if we are right?} What if a more efficient theory exploration system were possible, capable of reading huge amounts of code and producing an  abundance of theorems? How could the output be made manageable, by finding the needles we are interested in among the haystack of potential laws?

The criteria of what is considered \emph{interesting} is a key property of a theory exploration system. \qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is to discard equations which are direct consequences of others; those which remain are considered interesting.

A variety of such ad-hoc heuristics can be found in systems for concept formation

 conjecture generation systems, like those described in \S \ref{sec:conceptformation}, use a variety of ad-hoc heuristics to

### 2015-11-24 TransferReport/report.tex
More TODOs
Added future
Remove examples for now

--- 2015-11-17 TransferReport/report.tex
@@
 Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
-Current techniques are limited by their computational cost, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
+Current theory exploration techniques are limited by their use of exponential time algorithms, which we mitigate by intelligently reducing the size of large inputs, using machine learning to preserve the relationships needed for successful exploration.
 \end{abstract}
@@
 \input{implementation.tex}
-
-\iffalse
-\section{First Results}
-
-TODO
-
-Future work?
-Hypothetical use cases?
-\fi
-
 \input{related.tex}
-
-\section{Future Work}
-\label{future}
-
-\iffalse
-TODO
-
-interestingness
-\fi
-
-\section{Conclusion}
-\label{conclusion}
+\input{evaluation.tex}
+\input{future.tex}
+\iffalse TODO: Section that brings back ML discussion to TE is missing \fi
 

### 2015-11-24 TransferReport/background.tex
Updates

@@
                          |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
+                         |\ & \texttt{Type}                                          \\
       id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
@@
 
-We decided to focus our theory exploration research on Haskell as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
+We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
 This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
 
-Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, is based on a variant of the polymorphic $\lambda$-calculus known as \fc{}.
-For simplicity, we will focus on this Core language rather than the relatively large and complex syntax of Haskell proper.
-For a full treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
+Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
+Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
+For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
+Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
 
-The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names.
-There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.
+The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to ignore type hints (explicit casts are removed, whilst particular types/kinds/coercions are all represented by \texttt{Type}) and to use a custom representation of names.
+We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals.
 
@@
 life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests (see \ref{sec:haskell} for more details on Haskell):
+To see why, consider the following Haskell function, along with some unit tests:
 
@@
 These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{sec:atp} for more information on automated theorem proving.}
+a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.
 
@@
 
-In this work we focus on the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
+In this work we consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
 The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
@@
   \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$.
-To ensure the list is finite, function applications are only nested up to a specified depth (e.g.
-3).
+To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
   \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item Each variable is instantiated to particular value, generated randomly by \qcheck{}.
+  \item Each variable is instantiated to a particular value, generated randomly by \qcheck{}.
   \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
-If so, the class is split up to separate such members.
+If so, the class is split up to separate such distinguishable members.
   \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
@@
 
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which is a can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
+Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
 
@@
 
-\begin{multline}
-  terms_{\mathbb{N}} = \{0,\ succ,\ plus,\ times,\ x,\ y,\ succ(0),\ succ(x),\ succ(y),\ succ(z), \\
-                    plus(0,\blank),\ plus(x, \blank),\ \dots \}
-\end{multline}
+\begin{align*}
+  terms_{\mathbb{N}} = [& 0,\ succ,\ plus,\ times,\ x,\ y,\ z,\ succ(0),\ succ(x),\ succ(y), \\
+                     & succ(z),\ plus(0,\blank),\ plus(x, \blank),\ \dots ]
+\end{align*}
 
@@
 
-These terms will be grouped into three classes, for $\mathbb{N}$, $\mathbb{N} \rightarrow \mathbb{N}$ and $\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}$.
-As the variables $x$, $y$ and $z$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until equations such as the following are conjectured (after simplification):
-
+\begin{figure}
   % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
   \begin{align*}
-                      plus(x, y) &= plus(y, x)
-                      plus(x, 0) &= x
-             plus(x, plus(y, z)) &= plus(y, plus(x, z))
-                     times(x, y) &= times(y, x)
-                     times(x, 0) &= 0
-           times(x, times(y, z)) &= times(y, times(x, z))
-                plus(x, succ(y)) &= succ(plus(x, y))
-               times(x, succ(y)) &= plus(x, times(x, y))
-            times(x, plus(y, y)) &= times(y, plus(x, x))
+                        plus(x, y) &= plus(y, x)            \\
+                        plus(x, 0) &= x                     \\
+               plus(x, plus(y, z)) &= plus(y, plus(x, z))   \\
+                       times(x, y) &= times(y, x)           \\
+                       times(x, 0) &= 0                     \\
+             times(x, times(y, z)) &= times(y, times(x, z)) \\
+                  plus(x, succ(y)) &= succ(plus(x, y))      \\
+                 times(x, succ(y)) &= plus(x, times(x, y))  \\
+              times(x, plus(y, y)) &= times(y, plus(x, x))  \\
     plus(times(x, y), times(x, z)) &= times(x, plus(y, z))
   \end{align*}
+  \caption{Equations conjectured by \qspec{} for Peano numerals, addition and multiplication; after simplification.}
+  \label{fig:qspecresult}
+\end{figure}
 
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
+These terms will be grouped into three classes, one each for $\mathbb{N}$, $\mathbb{N} \rightarrow \mathbb{N}$ and $\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}$.
+As the variables $x$, $y$ and $z$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
+
+Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
+Here we know that addition and multiplication are closely related, and hence obey many algebraic laws.
+Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
+
+\qspec{} (and \hspec{}) are also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
 
@@
 One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-
-We attempt to mitigate the cost of running \qspec{}, by providing a machine learning layer analogous to that of premise selection in theorem proving.
-Of particular concern is the set of expressions $exprs$, which grows exponentially as the size of the expressions increases, and hence slows down the rest of the algorithm.
-Although complete, the current enumeration approach is wasteful: many combinations are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-The key idea of premise selection is to make such choices automatically, by only including those expressions which are considered \emph{relevant} to the problem.
-
-\iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
-\iffalse TODO: This detailed comparison with premise selection is opaque without defining/explaining what premise selection is \fi
 
@@
 
+\iffalse
+
 \begin{figure}
@@
 % TODO
+\fi

### 2015-11-24 TransferReport/contributions.tex
Updates

--- 2015-11-17 TransferReport/contributions.tex
@@
 
-Two simple methods for comparison are to ignore identifiers completely, or to compare them for equality.
+Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality.
 Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).
@@
 
-Recurrent clustering tackles this problem in a straightforward way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
+Recurrent clustering tackles this problem in a simple way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
 This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
@@
 \subsubsection{Algorithm}
-\label{algorithm}
+\label{sec:algorithm}
 
-\begin{figure}
-  \begin{lstlisting}[language=Haskell, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
-    data RoseTree a = Node a [RoseTree a]
+Our algorithm proceeds in several stages.
+At the top level, we send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive:
 
-    eRT :: [Local] -> [[Global]] -> Expr -> RoseTree Feature
-    eRT env db e = case e of
-      Var (Global i) -> Node (lookupG db  i) []
-      Var (Local  i) -> Node (lookupL env i) []
-      Lit (LitNum _) -> Node sLITNUM []
-      Lit (LitStr _) -> Node sLITSTR []
-      Lam  i  e      -> Node sLAM [eRT (i:env) db e]
-      App  e1 e2     -> Node sAPP [eRT    env  db e1,
-                                   eRT    env  db e2]
-      Let  bs e      -> Node sLET (map (bRT (ids bs:env) db) bs ++
-                                        eRT (ids bs:env) db  e
-      Case e i alts  -> Node sCASE (eRT    env  db  e :
-                               map (aRT (i:env) db) alts)
-
-    aRT :: [Local] -> [[Global]] -> Alt -> RoseTree Feature
-    aRT env db alt = case alt of
-      (DataAlt _, vs, e) -> eRT (vs ++ env) db e
-      (LitAlt  _, _,  e) -> eRT env db e
-      (Default,   _,  e) -> eRT env db e
-
-    bRT :: [Local] -> [[Global]] -> Bind -> RoseTree Feature
-    bRT env db b = case b of
-      NonRec i e -> eRT env db e
-      Rec es     -> Node sREC (map (eRT env db . snd) es)
-
-    level :: Int -> RoseTree Feature -> [Feature]
-    level 0 (Node x xs) = [x]
-    level n (Node x xs) = concatMap (level (n-1)) xs
-
-    rt :: [[Global]] -> Expr -> [Feature]
-    rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
-
+\begin{haskell}
 recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
@@
         extract db (i, e) = (i, rt db e)
-  \end{lstlisting}
-  \caption{Feature extraction for Haskell Core.
-We send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive.
-Syntax trees are first converted into \hs{RoseTree}s of \hs{Feature}s, these trees are converted to matrices, then flattened into feature vectors, then \hs{kMeans} clusters the feature vectors.
-The result is a list of clusters, naming their elements.
-More details are given in \S \ref{algorithm}.}
-  \label{featureextractionalgorithm}
-\end{figure}
+\end{haskell}
 
-Our feature extraction algorithm is given in figure \ref{featureextractionalgorithm}. Here we highlight some key features:
+The overall result of the \hs{recurrentCluster} function is a list of clusters, containing the IDs of their elements.
+These are obtained by interleaving clustering (the \hs{kMeans} function) and feature extraction (the \hs{rt} function).
 
-\begin{itemize}
-  \item Since they are comparatively rare, we ignore the particular value of each literal.
-  \item Since we are currently focused on expressions, we ignore particular data constructors and types.
-  \item The \hs{lookupL} and \hs{lookupG} functions return the index containing their argument, if found.
+The feature extraction itself contains two parts: first, syntax trees matching the grammar in figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
+
+\begin{haskell}
+data RoseTree = Node Feature [RoseTree]
+
+eRT :: [Local] -> [[Global]] -> Expr -> RoseTree
+eRT env db e = case e of
+  ...
+\end{haskell}
+
+The easiest branches to handle are literals and types, which we represent using particular feature values (\hs{sLITNUM}, \hs{sLITSTR} and \hs{sTYPE}):
+
+\begin{haskell}
+  Lit (LitNum _) -> Node sLITNUM []
+  Lit (LitStr _) -> Node sLITSTR []
+  Type           -> Node sTYPE   []
+\end{haskell}
+
+Function application simply recurses into both sub-expressions:
+
+\begin{haskell}
+  App e1 e2 -> Node sAPP [eRT env db e1,
+                          eRT env db e2]
+\end{haskell}
+
+To look up variables locally and globally, we use the \hs{lookupL} and \hs{lookupG} functions, respectively.
+These return the index containing their argument, if found.
 In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors.
-For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features.
-  \item \hs{lookupL} and \hs{lookupG} return a sentinel value \hs{sUNKNOWN} if the given ID is not found.
-In practice, this only occurs for global, mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
-  \item The \hs{Feature} type is abstract, but in practice it will be \hs{Float}.
-  \item Our algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
+For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features:
+
+\begin{haskell}
+  Var (Global i) -> Node (lookupG db  i) []
+  Var (Local  i) -> Node (lookupL env i) []
+\end{haskell}
+
+Finally, when we traverse binders we must extend the environment \hs{env}:
+
+\begin{haskell}
+  Lam  i  e     -> Node sLAM [eRT (i:env) db e]
+  Let  bs e     -> Node sLET (map (bRT (ids bs:env) db) bs ++
+                                   eRT (ids bs:env) db  e
+  Case e i alts -> Node sCASE (eRT    env  db  e :
+                          map (aRT (i:env) db) alts)
+\end{haskell}
+
+Patterns and bindings are handled in a similar way:
+
+\begin{haskell}
+aRT :: [Local] -> [[Global]] -> Alt -> RoseTree
+aRT env db alt = case alt of
+  (DataAlt _, vs, e) -> eRT (vs ++ env) db e
+  (LitAlt  _, _,  e) -> eRT env db e
+  (Default,   _,  e) -> eRT env db e
+
+bRT :: [Local] -> [[Global]] -> Bind -> RoseTree
+bRT env db b = case b of
+  NonRec i e -> eRT env db e
+  Rec es     -> Node sREC (map (eRT env db . snd) es)
+\end{haskell}
+
+The result of \hs{eRT} is a \hs{RoseTree} whose branching structure mimics that of our original expression.
+We next need to convert this to a \emph{matrix} of features, which we do by converting each level of the \hs{RoseTree} into a row of the matrix (using \hs{pad} to ensure a consistent size).
+Finally we turn the matrix into a \emph{feature vector} by concatenating the rows:
+
+\begin{haskell}
+level :: Int -> RoseTree -> [Feature]
+level 0 (Node x xs) = [x]
+level n (Node x xs) = concatMap (level (n-1)) xs
+
+rt :: [[Global]] -> Expr -> [Feature]
+rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
+\end{haskell}
+
+Notice that this algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
 The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
-\end{itemize}
 
@@
   \item Although they are values, we cannot distinguish between data constructors, other than using exact equality.
-  \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot completely erase type-level parameters unless they are fully applied.
+  \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot always distinguish between them.
 This can cause a function's Core arity to be greater than its Haskell arity.

### 2015-11-24 TransferReport/evaluation.tex
Updates

\iffalse
\section{First Results}

TODO

Future work?
Hypothetical use cases?
\fi

### 2015-11-24 TransferReport/future.tex
Updates

@@
 A comparison of our hand-picked features against a selection of learned representations would be a useful indication of the importance that understanding the expression language may or may not have on identifying salient aspects of expressions.
-  \item
+  \item There are several potential alternatives to \qcheck{} for instantiating variables, which may improve performance of \qspec{} by either generating expressions more efficiently (as in \textsc{SmallCheck}'s enumeration approach) or by observing differences with fewer tests (as found in \textsc{Lazy SmallCheck}'s demand-driven enumeration).
 \end{itemize}
@@
 
-It is important to consider the question: \emph{what if we are right?} What if a more efficient theory exploration system were possible, capable of reading huge amounts of code and producing an  abundance of theorems?
+It is important to consider the question: \emph{what if we succeed?} What if a more efficient theory exploration system were possible, capable of reading huge amounts of code and producing an abundance of theorems?
 How could the output be made manageable, by finding the needles we are interested in among the haystack of potential laws?
 
-The criteria of what is considered \emph{interesting} is a key property of a theory exploration system.
-\qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is to discard equations which are direct consequences of others; those which remain are considered interesting.
+This is governed by the ``interestingness'' criteria of the theory exploration system: what to keep and what to discard, and even what areas of the search space to prioritise.
+\qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is very simple: we discard equations which are direct consequences of others, and keep all the rest.
 
-A variety of such ad-hoc heuristics can be found in systems for concept formation
+\subsubsection{Interestingness in Concept Formation}
+\label{sec:conceptformation}
 
- conjecture generation systems, like those described in \S \ref{sec:conceptformation}, use a variety of ad-hoc heuristics to
+% TODO:
+\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
+\citep{Piantadosi.Tenenbaum.Goodman:2012}
+\citep{Wille:2005}
+\citep{colton1999automatic}
+\citep{colton2000agent}
+\citep{colton2012automated}
+\citep{lenat1977automated}
+\citep{mullerunderstanding}
+\citep{Bundy.Cavallo.Dixon.ea:2015}
+\citep{johansson2009isacosy}
+\citep{spector2008genetic}
+\citep{colton2012automated}
+ \citep{geng2006interestingness}
+% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
+
+% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+
+% Alan Bundy et al
+
+% Eurisko, AM, etc.?
+
+\subsubsection{Artificial Curiosity}
+\label{sec:curiosity}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[scale=0.75]
+      % The image, for reference
+      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
+
+      % Axes
+      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
+      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
+      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
+
+      % Curve. The numbers come from tracing over wundt.png
+      \draw[black,ultra thick] (3, 4)
+           to[out=0,   in=240] (5.05, 6)
+           to[out=60,  in=180] (6.5,  8.2)
+           to[out=0,   in=105] (8.1,  6)
+           to[out=-75, in=120] (8.8,  4)
+           to[out=-60, in=150] (11.4, 0.8);
+
+      % This version is closer, but a little jagged
+      \iffalse
+      \draw[black,ultra thick] (3, 4)
+           to[out=0,   in=240] (5.05, 6)
+           to[out=60,  in=225] (6,    8)
+           to[out=45,  in=180] (6.5,  8.2)
+           to[out=0,   in=135] (7.2,  8)
+           to[out=-45, in=105] (8.1,  6)
+           to[out=-75, in=120] (8.8,  4)
+           to[out=-60, in=150] (11.4, 0.8);
+      \fi
+  \end{tikzpicture}
+
+  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
+The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
+
+  \label{fig:wundt}
+\end{figure}
+
+\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
+Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
+
+As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
+The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
+
+We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
+
+In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
+A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
+
+One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
+Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
+This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
+
+% TODO: Examples
+\citep{Schmidhuber1999}
+
+Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
+A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
+The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
+
+\begin{equation}
+  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
+\end{equation}
+
+$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
+For parts of the state space which have been fully learned, the models will agree on accurate predictions.
+For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
+Hence it is the states which are amenable to learning which produce the largest disagreement.
+
+The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
+This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
+
+A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
+we can see which paths are taken through each decision tree).
+% TODO: The accuracy of these random forest models are compared
+
+% TODO
+\citep{Kaplan2006}
+\citep{Lipson2007}
+\citep{Luciw2011}
+\citep{Macedo2000}
+\citep{Ramik.Sabourin.Madani:2013}
+\citep{Roa.Kruijff.Jacobsson:2009}
+\citep{Schaul.Sun.Wierstra.ea:2011}
+\citep{Schmidhuber1999}
+\citep{Schmidhuber:1991}
+\citep{Scott1989}
+\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
+\citep{maher2008achieving}
+\citep{meyer1991possibility}
+\citep{oudeyer2004intelligent}
+\citep{oudeyer2014evolution}
+\citep{schmidhuber2006developmental}
+
+% TODO: Coevolution
+
+Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
+This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
+For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
+However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
+
+The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
+in a robot).
+Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
+Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
+Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
+
+\begin{itemize}
+
+  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
+
+  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
+If it weren't, then brute force enumeration of proofs would be a viable solution.
+In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
+
+\end{itemize}
+
+\iffalse
+
+\subsubsection{Universal Drives}
+
+PhysRevLett.110.168702.pdf
+Omohundro? Too physical.
+\emph{Universal drives} are those
+
+\fi
+
+\subsubsection{Interestingness in Evolutionary Computation}
+
+\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
+Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
+Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
+
+Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
+Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
+If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
+
+Pure exploration (i.e.
+for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
+The former attempts to gain insight into the nature of life and biology through competition over limited resources.
+Whilst this may have utility in resource allocation, e.g.
+efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
+\citep{baum2000evolution}).
+
+On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
+Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
+by hill climbing), but few global optima which are hard to find.
+Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
+
+One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
+Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
+In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
+We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
+if we allocate fitness uniformly between identical solutions we get:
+
+$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
+
+Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
+In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
+By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
+
+There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
+Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
+This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
+
+In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
+The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
+via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
+via mutation and crossover, to obtain new solutions.
+
+\newcommand{\argmin}{\operatornamewithlimits{argmin}}
+
+Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
+Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
+For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
+An individual $s$ is then selected with fitness closest to $f_t$, i.e.
+$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
+
+In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
+This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
+As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
+
+Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
+This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
+Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
+
+\iffalse
+The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
+This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
+\fi

### 2015-11-24 TransferReport/intro.tex
Updates

@@
 
-\iffalse
-TODO
-Intro + motivation
-
-About 2 pages
-
-Haskell is a mainstream FP language (motivation)
-
-QuickCheck (example)
-
-How to come up with the properties
-
-Mention theory exploration
-\fi
-
-\iffalse TODO: sounds strange...
-what does it mean for software to be capable or uncapable?
-of what?
-Rephrase?
-\fi As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
+As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
 The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
-\iffalse TODO: I am not sure the example makes sense without explaining what effects are and why separating them helps.
-Rather than going there, perhaps just say about declarative style of programming and types expressing prperties of computations?
-\fi For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).
+For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
+Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code.
+Similarly, pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results.
 
@@
   \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
-  \item A feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf machine learning algorithms. \iffalse TODO: say how this item fits in the TE agenda. \fi
+  \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of large signatures, which are more tractable to explore.
+  \item A novel feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf learning algorithms.
   \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
@@
 We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details \S \ref{sec:implementation}.
-A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future} before concluding in \S \ref{sec:conclusion}.
+A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future}.

### 2015-11-24 TransferReport/related.tex
Updates

@@
 This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving (see \S \ref{sec:itp}), as these results may be invalidated by seemingly innocuous code changes.
+in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
 We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{sec:theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
@@
 
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{sec:curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{fig:wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\citep{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
+\subsection{Relevance Filtering}
 \label{sec:relevance}
@@
 
-\subsubsection{Clustering}
+\subsection{Clustering Expressions}
 \label{sec:clustering}
@@
 \citep{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\citep{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{sec:conceptformation}
-
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
 
@@
 \fi
-
-\subsection{Interactive Theorem Proving}
-\label{sec:itp}
-
-ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:
-
-$$ C \colon (P \times S) \rightarrow Boolean $$
-
-The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:
-
-$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$
-
-It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell.
-In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker.
-There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.
-
-This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic.
-Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers.
-With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.
-
-Most differences between functional programming and ITP stem from different \emph{expectations} of the user.
-In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete.
-For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made.
-Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness).
-This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}.
-One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.
-
-Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g.
-see \citep{hales2015formal}).
-One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible.
-The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.
-
-\subsection{Automated Theorem Proving}
-\label{sec:atp}
-
-ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker.
-By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic.
-In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic.
-This is the most striking difference from ITP systems (e.g.
-Coq, Agda and Isabelle) which operate in higher-order logics.
-
-In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference.
-Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}.
-The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.
-
-Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{sec:related}.

### 2015-11-30 TransferReport/background.tex
Added Core translations

@@
 
+\newcommand{\CVar}{\texttt{Var}}
+\newcommand{\CLit}{\texttt{Lit}}
+\newcommand{\CApp}{\texttt{App}}
+\newcommand{\CLam}{\texttt{Lam}}
+\newcommand{\CLet}{\texttt{Let}}
+\newcommand{\CCase}{\texttt{Case}}
+\newcommand{\CType}{\texttt{Type}}
+\newcommand{\CLocal}{\texttt{Local}}
+\newcommand{\CGlobal}{\texttt{Global}}
+\newcommand{\CLitNum}{\texttt{LitNum}}
+\newcommand{\CLitStr}{\texttt{LitStr}}
+\newcommand{\CDataAlt}{\texttt{DataAlt}}
+\newcommand{\CLitAlt}{\texttt{LitAlt}}
+\newcommand{\CDefault}{\texttt{Default}}
+\newcommand{\CNonRec}{\texttt{NonRec}}
+\newcommand{\CRec}{\texttt{Rec}}
+
 \begin{figure}
@@
     \begin{split}
-      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
-                         |\ & \texttt{Lit}\ literal                                  \\
-                         |\ & \texttt{App}\ expr\ expr                               \\
-                         |\ & \texttt{Lam}\ \mathcal{L}\ expr                        \\
-                         |\ & \texttt{Let}\ bind\ expr                               \\
-                         |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
-                         |\ & \texttt{Type}                                          \\
-      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
-                         |\ & \texttt{Global}\ \mathcal{G}                           \\
-      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
-                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
+      expr\    \rightarrow\ & \CVar\ id                                       \\
+                         |\ & \CLit\ literal                                  \\
+                         |\ & \CApp\ expr\ expr                               \\
+                         |\ & \CLam\ \mathcal{L}\ expr                        \\
+                         |\ & \CLet\ bind\ expr                               \\
+                         |\ & \CCase\ expr\ \mathcal{L}\ \left[ alt \right]   \\
+                         |\ & \CType                                          \\
+      id\      \rightarrow\ & \CLocal\ \mathcal{L}                            \\
+                         |\ & \CGlobal\ \mathcal{G}                           \\
+      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}                           \\
+                         |\ & \CLitStr\ \mathcal{S}                           \\
       alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )               \\
-      altcon\  \rightarrow\ & \texttt{DataAlt}\ \mathcal{G}                          \\
-                         |\ & \texttt{LitAlt}\ literal                               \\
-                         |\ & \texttt{Default}                                       \\
-      bind\    \rightarrow\ & \texttt{NonRec}\ \mathcal{L}\ expr                     \\
-                         |\ & \texttt{Rec}\ [ ( \mathcal{L},\ expr ) ]
+      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{G}                          \\
+                         |\ & \CLitAlt\ literal                               \\
+                         |\ & \CDefault                                       \\
+      bind\    \rightarrow\ & \CNonRec\ \mathcal{L}\ expr                     \\
+                         |\ & \CRec\ [ ( \mathcal{L},\ expr ) ]
     \end{split}
@@
 
-The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to ignore type hints (explicit casts are removed, whilst particular types/kinds/coercions are all represented by \texttt{Type}) and to use a custom representation of names.
+The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
 We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals.
+We denote names and literals using quoted strings, e.g.
+\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}.
+
+\begin{figure}
+  \begin{haskell}
+    data Nat = Z
+             | S Nat
+
+    plus :: Nat -> Nat -> Nat
+    plus    Z  y = y
+    plus (S x) y = S (plus x y)
+
+    mult :: Nat -> Nat -> Nat
+    mult    Z  y = Z
+    mult (S x) y = plus y (mult x y)
+
+    factorial :: Nat -> Nat
+    factorial    Z  = S Z
+    factorial (S n) = mult (S n) (factorial n)
+
+    odd :: Nat -> Bool
+    odd    Z  = False
+    odd (S n) = even n
+
+    even :: Nat -> Bool
+    even    Z  = True
+    even (S n) = odd n
+  \end{haskell}
+  \caption{A Haskell datatype and functions for Peano arithmetic; in particular, the \hs{even} and \hs{odd} functions are mutually recursive.
+The \hs{Bool} type is Haskell's built in representation of Booleans, which can be regarded as \hs{data Bool = True \| False}.}
+  \label{fig:haskellexample}
+\end{figure}
+
+\begin{figure}
+  \begin{subfigure}[plus]{\textwidth}
+    \begin{small}
+      \underline{\texttt{plus}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       [(DataAlt "Z", [],    Var (Local "y")),
+                        (DataAlt "S", ["x"], App (Var (Global "S"))
+                                                 (App (App (Var (Global "plus"))
+                                                           (Var (Local  "x")))
+                                                      (Var (Local "y"))))]))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[mult]{\textwidth}
+    \begin{small}
+      \underline{\texttt{mult}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       [(DataAlt "Z", [],    Var (Global "Z")),
+                        (DataAlt "S", ["x"], App (App (Var (Global "plus"))
+                                                      (Var (Local  "y")))
+                                                 (App (App (Var (Global "mult"))
+                                                           (Var (Local  "x")))
+                                                      (Var (Local  "y"))))]))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[factorial]{\textwidth}
+    \begin{small}
+      \underline{\texttt{factorial}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              [(DataAlt "Z", [],    App (Var (Global "S"))
+                                        (Var (Global "Z"))),
+               (DataAlt "S", ["n"], App (App (Var (Global "mult"))
+                                             (App (Var (Global "S"))
+                                                  (Var (Local  "n"))))
+                                        (App (Var (Global "factorial"))
+                                             (Var (Local  "n"))))])
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[odd]{\textwidth}
+    \begin{small}
+      \underline{\texttt{odd}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              [(DataAlt "Z", [],    Var (Global "False")),
+               (DataAlt "S", ["n"], App (Var (Global "even"))
+                                        (Var (Local  "n")))])
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[even]{\textwidth}
+    \begin{small}
+      \underline{\texttt{even}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              [(DataAlt "Z", [],    Var (Global "True")),
+               (DataAlt "S", ["n"], App (Var (Global "odd"))
+                                        (Var (Local  "n")))])
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}.
+Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
+Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
+  \label{fig:coreexample}
+\end{figure}
+
+Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
+The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}.
+Although the Core is more verbose, we can see similar structure across the definitions (such as \texttt{(App (App (Var (Global $\blank$)) (Var (Local "x"))) (Var (Local "y")))}) which arise from similarities in the Haskell definitions (in this case, \hs{(plus x y)} and \hs{(mult x y)}).
+This is clearest in the definitions of \hs{odd} and \hs{even}, which are identical except for the global variables.
+
+Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}.
+GHC can represent these, but in this work we only consider reducible expressions (i.e.
+value-level bindings of the form \hs{f a b ...
+= ...}).
 
@@
 If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
+
+\iffalse TODO: Re-use these examples later on, e.g. for QuickSpec exploration \fi
 
@@
 \label{sec:theoryexploration}
-
-\newcommand{\blank}{\cdot}
 
@@
 
-\subsection{Clustering and Feature Extraction}
+\subsection{Feature Extraction}
 \label{sec:featureextraction}
+
+\iffalse TODO: include definitions \fi
+\iffalse TODO: make more relevant to our context \fi
 
@@
 We describe this approach in more details in \S \ref{sec:relevance}.
-Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+Premise selection is a practical tool which has seen use in practical software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
 
 Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
-The reason is that theory exploration has no distinct \emph{goal} to compare expressions against.
+The reason is that theory exploration has no distinct \emph{goal} \iffalse TODO: implies that premise selection has a goal; rephrase: premise selection is given a concrete formula (called the ``goal'') to compare terms against \fi to compare expressions against.
 Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
@@
 Seeing your contribution is feature extraction algorithms, you need to define feature vectors and clusters *here* \fi
-
-\iffalse
 
@@
 % TODO
-\fi
+
+Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric.
+There are many variations on this theme, but in our case we will fix the number of clusters at \iffalse TODO: use ceiling brackets \fi $k = \sqrt{n}$ the data points will be $d$-dimensional feature vectors of rational numbers, the similarity metric is euclidean distance, denoted $d$.
+and clustering will be performed by the $k-means$ algorithm.
+We choose this setup as it is well studied and supported by off-the-shelf tools, and is used in the ML4PG and ACL2(ml) systems which served as inspiration for this work.
+
+We will use $S$ to denote our set of $n$ data points and $k$ as the number of clusters.
+
+\subsubsection{k-Means}
+
+The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements, which we will denote as $m_1$ to $m_k$.
+Since k-means is iterative, we will use superscript notation to denote time steps, such as $m_i^t$ for the value of $m_i$ at step $t$.
+
+Before k-means starts, we must choose \emph{seed} values for $m_1^0$ to $m_k^0$.
+Many methods have been proposed for choosing these values, e.g.
+see CITE.
+For simplicity, we will assume they are chosen at random from our data set.
+Subsequent values, i.e.
+$m_i^t$ for $t > 0$, are simply the mean values of the corresponding cluster $C_i^t$:
+
+TODO: We need to handle the vectors element-wise, but we've already used subscript notation!
+
+\begin{equation}
+  m_i^t =
+\end{equation}
+
+At time step $t$, the elements of each cluster $C_i^t$ are the data points closest to $m_i^{t-1}$ from the previous time step:
+
+\begin{equation}
+  C_i^t = \{ p \in S \mid i = \argmin\limits_j d(p, m_j^{t-1}) \}
+\end{equation}
+
+The mean values o determine which cluster a point $p$ belongs to, we
+The as the name suggests, relies on the mean value of each cluster to determine  represents each cluster by uses a \emph{representative} t step $t$, the with cluster membership  We begin at iteration 0 by \emph{seeding} the algorithm with , each cluster is represented by the mean value of its elements:
+
+\begin{equation}
+  C_k =
+\end{equation}

### 2015-11-30 TransferReport/contributions.tex
Added Core translations
Added TODOs

@@
 \subsubsection{Overview}
+
+\iffalse TODO: Much of this discussion reads as a minor implementation solution rather than an overview of the whole method \fi
+
+\iffalse TODO: Don't show the ``pkg:Mod:name'' format here, it will just be confusing. Leave that to the implementation section. \fi
+
+\iffalse TODO: Focus the section more on the problem of identifiers, and how recurrent clustering uses clustering as part of feature extraction to solve this \fi
+
+\iffalse TODO: Maybe present the translation from trees to matrices, to vectors first; leaving the conversion of terms to features until afterwards; similar to how ML4PG uses [foo] to mean ``the feature corresponding to foo, whatever that might be'' \fi
+
+\iffalse TODO: Maybe present recurrent clustering in a ``naive'' way: define the conversion function ``[]'' recursively.
+Leave the dependency-ordering, etc.
+for the Algorithm section, as a way of ``building up to'' the result, which avoids the inefficiencies of recursing.
+\fi
 
@@
 \label{sec:algorithm}
+\iffalse TODO: the code should go in the implementation section rather than here \fi
+\iffalse TODO: give a more abstract presentation, using mathematical notation \fi
+\iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi
+\iffalse TODO: Define rose trees \fi
+\iffalse TODO: Show an example from earlier, e.g. factorial \fi
+\iffalse TODO: Split into three parts: expressions to rose trees of features; rose trees to matrices to vectors; k-means clustering \fi
+\iffalse TODO: highlight the recurrent nature of the algorithm \fi
+\iffalse TODO: I would probably just restructure:
+ - start with the syntax, get to show how, for every line of figure 1, extraction works
+ - then transformation to feature vectors
+ - then k-means
+obviously, they are mutually recursive...
+\fi
 

### 2015-11-30 TransferReport/implementation.tex
Added TODOs

@@
 \label{sec:implementation}
+
+\iffalse TODO: How does this section differ from Algorithm in Contributions? \fi
+\iffalse TODO: Make this a concrete presentation of the abstract ideas from the Algorithm section, e.g.
+this is how we get the dependencies (AST plugin, etc.); this is a case-by-case of how we perform feature extraction; this is how we perform k-means (Weka), etc.
+\fi
 

### 2015-11-30 TheoryExploration/render.sh
Don't run Hoogle; it's too huge

@@ -1,5 +1,5 @@
 #!/bin/sh
-nix-shell --pure --show-trace --command "sh" <<'EOF'
+nix-shell --show-trace --command "sh" <<'EOF'
   for src in *.md
   do
     doc=$(basename "$src" .md)

### 2015-11-30 ML4HS1/render.sh
Use pandoc directly

@@ -1,2 +1,9 @@
 #!/bin/sh
-nix-shell --pure --command "md2pdf"
+nix-shell --command "sh" <<'EOF'
+  for src in *.md
+  do
+    doc=$(basename "$src" .md)
+    echo "Rendering $doc"
+    pandoc --filter panpipe --filter panhandle -o "$doc.pdf" "$doc.md"
+  done
+EOF

### 2015-12-01 TransferReport/background.tex
Feature extraction

@@
 The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
-We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals.
-We denote names and literals using quoted strings, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}.
+For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
+We will use quoted strings to denote names and literals, e.g.
+\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
 
@@
 
-    factorial :: Nat -> Nat
-    factorial    Z  = S Z
-    factorial (S n) = mult (S n) (factorial n)
-
     odd :: Nat -> Bool
@@
   \end{haskell}
-  \caption{A Haskell datatype and functions for Peano arithmetic; in particular, the \hs{even} and \hs{odd} functions are mutually recursive.
-The \hs{Bool} type is Haskell's built in representation of Booleans, which can be regarded as \hs{data Bool = True \| False}.}
+  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
+\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool = True \| False}.}
   \label{fig:haskellexample}
@@
                                                       (Var (Local  "y"))))]))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[factorial]{\textwidth}
-    \begin{small}
-      \underline{\texttt{factorial}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              [(DataAlt "Z", [],    App (Var (Global "S"))
-                                        (Var (Global "Z"))),
-               (DataAlt "S", ["n"], App (App (Var (Global "mult"))
-                                             (App (Var (Global "S"))
-                                                  (Var (Local  "n"))))
-                                        (App (Var (Global "factorial"))
-                                             (Var (Local  "n"))))])
       \end{verbatim}
@@
 The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see similar structure across the definitions (such as \texttt{(App (App (Var (Global $\blank$)) (Var (Local "x"))) (Var (Local "y")))}) which arise from similarities in the Haskell definitions (in this case, \hs{(plus x y)} and \hs{(mult x y)}).
-This is clearest in the definitions of \hs{odd} and \hs{even}, which are identical except for the global variables.
+Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the global variables.
+This correspondence allows us to analyse
 
@@
 
-\qspec{} (and \hspec{}) are also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
+\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
 
@@
 
-\iffalse TODO: include definitions \fi
-\iffalse TODO: make more relevant to our context \fi
+One of our main contributions is a \emph{feature extraction} algorithm for Core expressions.
+Feature extraction is the conversion of ``raw'' input data, such as audio, images or (in our case) expressions, into a form more suited to machine learning algorithms.
+By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
+
+We use a standard representation of features known as a \emph{feature vector}: a sequence of numbers $(x_1, \dots, x_n)$ where $x_i \in \mathbb{R}$.
+\footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience.
+Practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has two important advantages over our raw expressions:
+
+\begin{itemize}
+  \item All of our feature vectors will be the same size, i.e.
+they will all have length (or \emph{dimension}) $n$.
+This makes element-wise comparison between arbitrary feature vectors $x$ and $y$ trivial, as each element $x_i$ will have a corresponding element $y_i$.
+For raw expressions this is not so straightforward, as their nesting may give rise to very different shapes.
+  \item Unlike our discrete expressions, we can continuously transform one feature vector into another.
+This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, taking the arithmetic mean.
+\end{itemize}
 

### 2015-12-01 TransferReport/contributions.tex
Use simple quoted IDs

@@
 
-\iffalse TODO: Don't show the ``pkg:Mod:name'' format here, it will just be confusing. Leave that to the implementation section. \fi
-
 \iffalse TODO: Focus the section more on the problem of identifiers, and how recurrent clustering uses clustering as part of feature extraction to solve this \fi
@@
 In our case, we want compare the abstract syntax trees of Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions.
-For example, consider the following expressions which are identical except for an identifier:\footnote{The \texttt{(\&\&)} function is boolean AND, \texttt{(++)} appends lists and \texttt{map} applies a function to the elements of a list.}
+For example, consider the following expressions which are identical except for an identifier:
 
@@
   \begin{split}
-    X =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.(\&\&)"})) \\
-    Y =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.(++)"})) \\
-    Z =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"base:Prelude.map"}))
+    X =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"odd"})) \\
+    Y =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"even"})) \\
+    Z =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"plus"}))
   \end{split}

### 2015-12-01 TransferReport/background.tex
Fleshed out feature extraction

@@
 
-As an example, we can consider a simple signature for Peano numerals, addition and multiplication:
+As an example, we can consider a simple signature containing the expressions from figure \ref{fig:haskellexample}:
 
 \begin{align*}
-  \Sigma_{\mathbb{N}} = \{ & 0     && : \mathbb{N}, \\
-                        & succ  && : \mathbb{N} \rightarrow \mathbb{N}, \\
-                        & plus  && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}, \\
-                        & times && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}\}
+  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
 \end{align*}
 
-Together with a set of variables, say $V_{\mathbb{N}} = \{x, y, z\}$, \qspec{}'s enumeration will resemble the following:
+Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
 
 \begin{align*}
-  terms_{\mathbb{N}} = [& 0,\ succ,\ plus,\ times,\ x,\ y,\ z,\ succ(0),\ succ(x),\ succ(y), \\
-                     & succ(z),\ plus(0,\blank),\ plus(x, \blank),\ \dots ]
+  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
+                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
 \end{align*}
 
-Notice that our functions are curried, where $f(x,\blank)$ indicates \emph{partial application} of $f$ to $x$.
+Notice that Haskell curries functions, so the binary functions \hs{plus} and \hs{mult} can be treated as unary functions which return unary functions.
 This is required as the construction of $terms$ applies functions to one argument at a time.
@@
   % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
-  \begin{align*}
-                        plus(x, y) &= plus(y, x)            \\
-                        plus(x, 0) &= x                     \\
-               plus(x, plus(y, z)) &= plus(y, plus(x, z))   \\
-                       times(x, y) &= times(y, x)           \\
-                       times(x, 0) &= 0                     \\
-             times(x, times(y, z)) &= times(y, times(x, z)) \\
-                  plus(x, succ(y)) &= succ(plus(x, y))      \\
-                 times(x, succ(y)) &= plus(x, times(x, y))  \\
-              times(x, plus(y, y)) &= times(y, plus(x, x))  \\
-    plus(times(x, y), times(x, z)) &= times(x, plus(y, z))
-  \end{align*}
-  \caption{Equations conjectured by \qspec{} for Peano numerals, addition and multiplication; after simplification.}
+  \begin{haskell}
+                  plus a b = plus b a
+                  plus a Z = a
+         plus a (plus b c) = plus b (plus a c)
+                  mult a b = mult b a
+                  mult a Z = Z
+         mult a (mult b c) = mult b (mult a c)
+              plus a (S b) = S (plus a b)
+              mult a (S b) = plus a (mult a b)
+         mult a (plus b b) = mult b (plus a a)
+                 odd (S a) = even a
+            odd (plus a a) = odd Z
+           odd (times a a) = odd a
+                even (S a) = odd a
+           even (plus a a) = even Z
+          even (times a a) = even a
+plus (mult a b) (mult a c) = mult a (plus b c)
+  \end{haskell}
+  \caption{Equations conjectured by \qspec{} for the functions in figure \ref{fig:haskellexample}; after simplification.}
   \label{fig:qspecresult}
@@
 
-These terms will be grouped into three classes, one each for $\mathbb{N}$, $\mathbb{N} \rightarrow \mathbb{N}$ and $\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}$.
-As the variables $x$, $y$ and $z$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
+These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat} and \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
+As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
 
@@
 
-\subsection{Feature Extraction}
+\subsection{Clustering}
+\label{sec:clustering}
+
+Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources.
+The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
+We describe this approach in more details in \S \ref{sec:relevance}.
+Premise selection is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+
+Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
+The reason is that premise selection requires a distinguished expression to compare everything against; in practice this is the current goal of a theorem prover.
+Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
+
+A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
+Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
+
+\subsubsection{Feature Extraction}
 \label{sec:featureextraction}
 
-One of our main contributions is a \emph{feature extraction} algorithm for Core expressions.
-Feature extraction is the conversion of ``raw'' input data, such as audio, images or (in our case) expressions, into a form more suited to machine learning algorithms.
+Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
+This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms.
 By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
 
-We use a standard representation of features known as a \emph{feature vector}: a sequence of numbers $(x_1, \dots, x_n)$ where $x_i \in \mathbb{R}$.
+We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_p)$ where $x_i \in \mathbb{R}$.
 \footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience.
@@
   \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $n$.
-This makes element-wise comparison between arbitrary feature vectors $x$ and $y$ trivial, as each element $x_i$ will have a corresponding element $y_i$.
-For raw expressions this is not so straightforward, as their nesting may give rise to very different shapes.
-  \item Unlike our discrete expressions, we can continuously transform one feature vector into another.
-This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, taking the arithmetic mean.
+they will all have length (or \emph{dimension}) $p$.
+This makes element-wise comparison between arbitrary feature vectors $\mathbf{x}$ and $\mathbf{y}$ trivial: compare each element $x_i$ with $y_i$ for $1 \leq i \leq p$.
+For expressions would not be so straightforward, as their nesting may give rise to very different shapes.
+  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
+This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
 \end{itemize}
 
-Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources.
-The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{sec:relevance}.
-Premise selection is a practical tool which has seen use in practical software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+\subsubsection{K-Means}
+\label{sec:kmeans}
 
-Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
-The reason is that theory exploration has no distinct \emph{goal} \iffalse TODO: implies that premise selection has a goal; rephrase: premise selection is given a concrete formula (called the ``goal'') to compare terms against \fi to compare expressions against.
-Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
-A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-See \S \ref{sec:clustering} for more information.
+Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:
 
-Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller ones more amenable to brute force enumeration.
+\begin{itemize}
+  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$.
+  \item Data points will be $p$-dimensional feature vectors, as defined above.
+  \item We will use euclidean distance (denoted $d$) as our similarity metric.
+  \item Clustering will be performed by the $k-means$ algorithm.
+\end{itemize}
 
-\iffalse TODO: I would play up the clustering intelligence.
-Clustering is not just about breaking up, it is about discovering significant patterns in data.
-By forgetting this, you make motivation for your work sound too ``small'' \fi
+This is a standard setup, supported by off-the-shelf tools.
+In particular, it is similar to that used by the ML4PG and ACL2(ml) systems which served as inspiration for this work, which makes our results more easily comparable.
+
+The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements.
+Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
+We denote the clusters as $C^1$ to $C^k$, their mean values as $\mathbf{m}^1$ to $\mathbf{m}^k$.
+Subscripts are vector elements, so $m^i_j(t)$ is the $i$th element of the $j$th mean, at time $t$.
+
+Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^1(0)$ to $\mathbf{m}^k(0)$.
+Many methods have been proposed for choosing these values, e.g.
+see \citep{arthur2007k}.
+For simplicity, we will assume they are chosen at random from our data set.
+Subsequent values, i.e.
+$\mathbf{m}^i(t)$ for $t > 0$, are simply the mean values of the corresponding cluster $C^i(t)$:
+
+\begin{equation}
+  m^i_j(t) = \frac{\sum_{x \in C^i(t)} x_j}{|C^i(t)|}
+\end{equation}
+
+At time step $t$, the elements of each cluster $C_i^t$ are the data points closest to $m_i^{t-1}$ from the previous time step:
+
+\begin{equation}
+  C_i^t = \{ p \in S \mid i = \argmin\limits_j d(p, m_j^{t-1}) \}
+\end{equation}
+
+The mean values o determine which cluster a point $p$ belongs to, we
+The as the name suggests, relies on the mean value of each cluster to determine  represents each cluster by uses a \emph{representative} t step $t$, the with cluster membership  We begin at iteration 0 by \emph{seeding} the algorithm with , each cluster is represented by the mean value of its elements:
+
+\begin{equation}
+  C_k =
+\end{equation}
 
@@
 \end{figure}
-
-\subsection{Clustering}
-\label{sec:clustering}
-% TODO
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric.
-There are many variations on this theme, but in our case we will fix the number of clusters at \iffalse TODO: use ceiling brackets \fi $k = \sqrt{n}$ the data points will be $d$-dimensional feature vectors of rational numbers, the similarity metric is euclidean distance, denoted $d$.
-and clustering will be performed by the $k-means$ algorithm.
-We choose this setup as it is well studied and supported by off-the-shelf tools, and is used in the ML4PG and ACL2(ml) systems which served as inspiration for this work.
-
-We will use $S$ to denote our set of $n$ data points and $k$ as the number of clusters.
-
-\subsubsection{k-Means}
-
-The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements, which we will denote as $m_1$ to $m_k$.
-Since k-means is iterative, we will use superscript notation to denote time steps, such as $m_i^t$ for the value of $m_i$ at step $t$.
-
-Before k-means starts, we must choose \emph{seed} values for $m_1^0$ to $m_k^0$.
-Many methods have been proposed for choosing these values, e.g.
-see CITE.
-For simplicity, we will assume they are chosen at random from our data set.
-Subsequent values, i.e.
-$m_i^t$ for $t > 0$, are simply the mean values of the corresponding cluster $C_i^t$:
-
-TODO: We need to handle the vectors element-wise, but we've already used subscript notation!
-
-\begin{equation}
-  m_i^t =
-\end{equation}
-
-At time step $t$, the elements of each cluster $C_i^t$ are the data points closest to $m_i^{t-1}$ from the previous time step:
-
-\begin{equation}
-  C_i^t = \{ p \in S \mid i = \argmin\limits_j d(p, m_j^{t-1}) \}
-\end{equation}
-
-The mean values o determine which cluster a point $p$ belongs to, we
-The as the name suggests, relies on the mean value of each cluster to determine  represents each cluster by uses a \emph{representative} t step $t$, the with cluster membership  We begin at iteration 0 by \emph{seeding} the algorithm with , each cluster is represented by the mean value of its elements:
-
-\begin{equation}
-  C_k =
-\end{equation}

### 2015-12-01 TransferReport/contributions.tex
Fleshed out feature extraction

@@
 The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs.
-In our case, we want compare the abstract syntax trees of Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions.
-For example, consider the following expressions which are identical except for an identifier:
+In our case, we want to compare Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions.
+For example, consider the following expressions which are identical except for the global identifiers (which refer to the expressions in figure \ref{fig:coreexample}):
 
@@
 
-Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality.
+We saw in section Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality.
 Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).

### 2015-12-01 TransferReport/haskell_example/src/QuickSpecExample.hs

@@ -29,9 +29,11 @@ sigma = [blind0 "()"      (),
          observer1 ((`mod` 10) :: Natural -> Natural),
          observer1 (id :: Bool -> Bool)]
 
-nat = [fun0 "0"     (0 :: Natural),
-       fun1 "succ"  ((1+) :: Natural -> Natural),
-       fun2 "plus"  ((+) :: Natural -> Natural -> Natural),
-       fun2 "times" ((*) :: Natural -> Natural -> Natural),
+nat = [fun0 "Z"     (0    :: Natural),
+       fun1 "S"     ((1+) :: Natural -> Natural),
+       fun2 "plus"  ((+)  :: Natural -> Natural -> Natural),
+       fun2 "times" ((*)  :: Natural -> Natural -> Natural),
+       fun1 "odd"   (odd  :: Natural -> Bool),
+       fun1 "even"  (even :: Natural -> Bool),
 
        vars ["x", "y", "z"] (0 :: Natural)]

### 2015-12-01 TransferReport/background.tex
Clustering

@@
 
-These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat} and \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
+These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
 As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
@@
 
-Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources.
+Our approach to scaling up \qspec{} takes inspiration from two sources.
 The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
@@
 
-We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_p)$ where $x_i \in \mathbb{R}$.
-\footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience.
-Practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has two important advantages over our raw expressions:
+We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$.
+\footnote{In fact, ractical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
 
+Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
 \begin{itemize}
   \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $p$.
-This makes element-wise comparison between arbitrary feature vectors $\mathbf{x}$ and $\mathbf{y}$ trivial: compare each element $x_i$ with $y_i$ for $1 \leq i \leq p$.
-For expressions would not be so straightforward, as their nesting may give rise to very different shapes.
+they will all have length (or \emph{dimension}) $d$.
+Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
+For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
   \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
 This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
+  \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
+This reduces the input size of the machine learning problem, improving efficiency.
 \end{itemize}
@@
   \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$.
-  \item Data points will be $p$-dimensional feature vectors, as defined above.
-  \item We will use euclidean distance (denoted $d$) as our similarity metric.
-  \item Clustering will be performed by the $k-means$ algorithm.
+  \item Data points will be $d$-dimensional feature vectors, as defined above.
+  \item We will use euclidean distance (denoted $e$) as our similarity metric.
+  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm.
 \end{itemize}
 
-This is a standard setup, supported by off-the-shelf tools.
-In particular, it is similar to that used by the ML4PG and ACL2(ml) systems which served as inspiration for this work, which makes our results more easily comparable.
+This is a standard setup, supported by off-the-shelf tools. In particular, it is similar to that used by the ML4PG and ACL2(ml), which makes our results more easily comparable.
 
-The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements.
+The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as:
+
+\begin{equation*}
+  \mean{X}_i = \frac{\sum_{x \in X} x_i}{|X|}
+\end{equation*}
+
 Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$, their mean values as $\mathbf{m}^1$ to $\mathbf{m}^k$.
-Subscripts are vector elements, so $m^i_j(t)$ is the $i$th element of the $j$th mean, at time $t$.
+We denote the clusters as $C^1$ to $C^k$ and their mean values as $\vect{m}^1$ to $\vect{m}^k$, hence:
 
-Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^1(0)$ to $\mathbf{m}^k(0)$.
-Many methods have been proposed for choosing these values, e.g.
-see \citep{arthur2007k}.
-For simplicity, we will assume they are chosen at random from our data set.
-Subsequent values, i.e.
-$\mathbf{m}^i(t)$ for $t > 0$, are simply the mean values of the corresponding cluster $C^i(t)$:
+\begin{equation*}
+  \vect{m}^i(t) = \mean{C}^i(t) \quad \text{for $t > 0$}
+\end{equation*}
 
-\begin{equation}
-  m^i_j(t) = \frac{\sum_{x \in C^i(t)} x_j}{|C^i(t)|}
-\end{equation}
+Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^i(0)$.
+Many methods have been proposed for choosing these values \citep{arthur2007k}.
+For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
 
-At time step $t$, the elements of each cluster $C_i^t$ are the data points closest to $m_i^{t-1}$ from the previous time step:
+The elements of each cluster $C^i(t)$ are those data points closest to the mean value at the previous time step:
 
-\begin{equation}
-  C_i^t = \{ p \in S \mid i = \argmin\limits_j d(p, m_j^{t-1}) \}
-\end{equation}
+\begin{equation*}
+  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad for t > 0
+\end{equation*}
 
-The mean values o determine which cluster a point $p$ belongs to, we
-The as the name suggests, relies on the mean value of each cluster to determine  represents each cluster by uses a \emph{representative} t step $t$, the with cluster membership  We begin at iteration 0 by \emph{seeding} the algorithm with , each cluster is represented by the mean value of its elements:
+As $t$ increases, the clusters $C^i$ move from their initial location around the ``seeds'', to converge on a local minimum of the ``within-cluster sum of squared error'' objective:
 
-\begin{equation}
-  C_k =
-\end{equation}
-
-Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered lists of features) are chosen to represent the relevant information in a more compressed form than the raw data (for example, replacing descriptive identifier names with sequential numbers).
-This reduces the input size of the machine learning problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details, such as the text encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-
-In our case, we will use a novel feature extraction algorithm, described in \S \ref{sec:recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.
-
-\iffalse TODO: I have had a look now at section 5.
-I do not think feature extraction and clustering is explained/defined there, either.
-Seeing your contribution is feature extraction algorithms, you need to define feature vectors and clusters *here* \fi
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75, level/.style={sibling distance=60mm/#1}]
-      % Tree
-      \node {\texttt{Lam}}
-        child {node {\texttt{"x"}}}
-        child {node {\texttt{App}}
-          child {node {\texttt{Var}}
-            child {node {\texttt{Global}}
-              child {node {\texttt{"base:Prelude.not"}}}}}
-          child {node {\texttt{Var}}
-            child {node {\texttt{Local}}
-              child {node {\texttt{"x"}}}}}};
-
-      % Arrow
-      %\node[single arrow,
-      %      draw=black,
-      %      fill=black!10,
-      %      minimum height=2cm,
-      %      shape border rotate=0] at (0,-1) {};
-      %\draw[-latex] (A.east) -- (B.west);
-
-      % Feature vector
-
-  \end{tikzpicture}
-
-  \caption{}
-
-  \label{fig:featureextractionpic}
-\end{figure}
+\begin{equation*}
+  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
+\end{equation*}

### 2015-12-01 TransferReport/background.tex
Updated algorithm

@@
 \newcommand{\CLitStr}{\texttt{LitStr}}
+\newcommand{\CAlt}{\texttt{Alt}}
 \newcommand{\CDataAlt}{\texttt{DataAlt}}
@@
 \newcommand{\CRec}{\texttt{Rec}}
+\newcommand{\CBind}{\texttt{Bind}}
 
@@
                          |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ \left[ alt \right]   \\
+                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                          |\ & \CType                             \\
@@
                          |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )               \\
+      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
       altcon\  \rightarrow\ & \CDataAlt\ \mathcal{G}             \\
@@
                          |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ \mathcal{L}\ expr                     \\
-                         |\ & \CRec\ [ ( \mathcal{L},\ expr ) ]
+      bind\    \rightarrow\ & \CNonRec\ binder                   \\
+                         |\ & \CRec\ [binder]                    \\
+      binder   \rightarrow\ & \CBind \mathcal{L}\ expr
     \end{split}
@@
   \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
-\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool = True \| False}.}
+\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
   \label{fig:haskellexample}
@@
                        "b"
-                       [(DataAlt "Z", [],    Var (Local "y")),
-                        (DataAlt "S", ["x"], App (Var (Global "S"))
+                       [Alt (DataAlt "Z") (Var (Local "y")) [],
+                        Alt (DataAlt "S") (App (Var (Global "S"))
                                                (App (App (Var (Global "plus"))
                                                          (Var (Local  "x")))
-                                                      (Var (Local "y"))))]))
+                                                    (Var (Local "y"))))
+                                          ["x"]]))
       \end{verbatim}
@@
                        "b"
-                       [(DataAlt "Z", [],    Var (Global "Z")),
-                        (DataAlt "S", ["x"], App (App (Var (Global "plus"))
+                       [Alt (DataAlt "Z") (Var (Global "Z")) [],
+                        Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                     (Var (Local  "y")))
@@
                                                          (Var (Local  "x")))
-                                                      (Var (Local  "y"))))]))
+                                                    (Var (Local  "y"))))
+                                          ["x"]]))
       \end{verbatim}
@@
               "b"
-              [(DataAlt "Z", [],    Var (Global "False")),
-               (DataAlt "S", ["n"], App (Var (Global "even"))
-                                        (Var (Local  "n")))])
+              [Alt (DataAlt "Z") (Var (Global "False")) [],
+               Alt (DataAlt "S") (App (Var (Global "even"))
+                                      (Var (Local  "n")))
+                                 ["n"]])
       \end{verbatim}
@@
               "b"
-              [(DataAlt "Z", [],    Var (Global "True")),
-               (DataAlt "S", ["n"], App (Var (Global "odd"))
-                                        (Var (Local  "n")))])
+              [Alt (DataAlt "Z") (Var (Global "True")) [],
+               Alt (DataAlt "S") (App (Var (Global "odd"))
+                                      (Var (Local  "n")))
+                                 ["n"]])
       \end{verbatim}
@@
 
-\iffalse TODO: Re-use these examples later on, e.g. for QuickSpec exploration \fi
-
 \begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
@@
 we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
-Notice that unit tests like \hs{fact_base} are valid properties; they just assert rather weak statements.
+Notice that unit tests like \hs{fact\_base} are valid properties; they just assert rather weak statements.
 
-To check these properties, \qcheck{} treats closed terms (like \hs{fact_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
+To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
 For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
@@
 
-\subsubsection{K-Means}
+\subsubsection{k-Means}
 \label{sec:kmeans}
@@
 
-The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as:
+k-means works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as:
 
@@
 \begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad for t > 0
+  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
 \end{equation*}

### 2015-12-01 TransferReport/contributions.tex
Updated algorithm

--- 2015-12-01 TransferReport/contributions.tex
@@
 We take the \emph{recurrent clustering} approach found in ML4PG and ACL2(ml), and implement a variant in the context of Haskell theory exploration.
-Here we give a brief description of recurrent clustering, describe our algorithm, and compare its similarity and differences to ML4PG and ACL2(ml).
-
-\subsubsection{Overview}
+Here we describe our recurrent clustering algorithm and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
@@
 The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs.
-In our case, we want compare the abstract syntax trees of Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions.
-For example, consider the following expressions which are identical except for an identifier:
+In our case, we want to compare Core expressions which, as described above, requires a feature extraction algorithm to create a suitable representation.
+
+We extract features in two stages: first we construct \emph{rose trees} of features (real numbers), which maintain the nested structure of expressions shown in figure \ref{fig:coresyntax}.
+We then flatten these rose trees into feature vectors.
+
+\subsubsection{Rose Trees}
+
+Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+$T_i$ are the \emph{children of} $T$ and $f$ is the \emph{feature at} $T$.
+$n_T$ may be different for each child; trees where $n_T = 0$ are \emph{leaves}.
+
+We turn Core expressions into rose trees by converting terminal symbols (such as \hs{Lam} or \hs{Rec}) into features; we represent this conversion using $\feature{}$.
+Non-terminal symbols (such as $expr$ and $id$) are converted recursively and become children.
+This is shown in figure \ref{fig:totree}.
+
+\begin{figure}
+  \begin{align*}
+    toTree(e) &=
+    \begin{cases}
+      (\feature{\CVar}, toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
+      (\feature{\CLit}, toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
+      (\feature{\CApp}, toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
+      (\feature{\CLam}, leaf(l_1), toTree(e_1))                      & \text{if $e = \CLam\ l_1\ e_1$} \\
+      (\feature{\CLet}, toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
+      (\feature{\CCase}, toTree(e_1), leaf(l_1), toTree(a_1), \dots) & \text{if $e = \CCase\ e_1\ l_1 [a_1, \dots]$} \\
+      (\feature{\CType})                                            & \text{if $e = \CType$} \\
+      (\feature{\CLocal}, leaf(l_1))                                & \text{if $e = \CLocal\ l_1$} \\
+      (\feature{\CGlobal}, leaf(g_1))                               & \text{if $e = \CGlobal\ g_1$} \\
+      (\feature{\CLitNum})                                          & \text{if $e = \CLitNum\ n_1$} \\
+      (\feature{\CLitStr})                                          & \text{if $e = \CLitStr\ s_1$} \\
+      (\feature{\CAlt}, toTree(e_1), toTree(e_2), leaf{l_1}, \dots)  & \text{if $e = \CAlt\ e_1 e_2 [l_1, \dots]$}  \\
+      (\feature{\CDataAlt}, leaf(g_1))                              & \text{if $e = \CDataAlt\ g_1$}  \\
+      (\feature{\CLitAlt}, toTree(e_1))                             & \text{if $e = \CLitAlt\ e_1$}  \\
+      (\feature{\CDefault})                                         & \text{if $e = \CDefault$}  \\
+      (\feature{\CNonRec}, toTree(e_1))                             & \text{if $e = \CNonRec\ e_1$}  \\
+      (\feature{\CRec}, toTree(e_1), \dots)                         & \text{if $e = \CRec\ [e_1, \dots]$} \\
+      (\feature{\CBind}, leaf(l_1), toTree(e_1))                    & \text{if $e = \CBind\ l_1 e_1$}
+    \end{cases} \\
+    leaf(x) &= (\feature{x})
+  \end{align*}
+  \caption{Recursively converting Core expressions of figure \ref{fig:coresyntax} to rose trees.
+List elements (indicated with $\dots$) are appended as children, rather than being wrapped in another tree.}
+  \label{fig:totree}
+\end{figure}
+
+To flatten a rose tree we fetch the features from each ``level'', concatenating those from adjacent children:
+
+\begin{equation*}
+  level(l, (x, T_1, \dots, T_{n_T})) =
+    \begin{cases}
+      (x) & \text{if $l = 1$} \\
+      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$ and $\vect{x}_i = level(l - 1, T_i)$}
+    \end{cases}
+\end{equation*}
+
+Given a rose tree $T$, we pad each level $level(1, T)$, $level(2, T)$, $\dots$, $level(r, T)$ with zeros and truncate them to length $c$.
+The resulting vectors are concatenated to form the feature vector for $T$.
+
+See see figure \ref{fig:rosetreeexample} for an example of this construction.
+
+\begin{figure}
+  \caption{}
+  \label{fig:rosetreeexample}
+\end{figure}
+
+\subsubsection{Feature Extraction}
+
+We now define the algorithm for $\feature{}$, which turns terminal symbols into features.
+In the case of known language features, such as $\feature{\hs{Lam}}$ and $\feature{\hs{Case}}$, we just enumerate the possibilities (while trying to make ``similar'' symbols have similar features).
+
+The difficulty comes from identifiers, which we cannot hard-code into our algorithm in advance.
+The only meaningful way to directly compare identifiers is by equality, but since the space of identifiers is large, giving distinct features to distinct identifiers would make matches very unlikely.
+Since identifiers make up a significant portion of expressions, we would like a more fine-grained notion of similarity between their features than exact matching.
+
+For example, consider the following feature vectors, which are identical except for the global identifiers:
 
@@
   \begin{split}
-    X =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"odd"})) \\
-    Y =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"even"})) \\
-    Z =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"plus"}))
+    \vect{x} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"odd"})) \\
+    \vect{y} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"even"})) \\
+    \vect{z} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"plus"}))
   \end{split}
@@
 
-Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality.
-Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).
+From their definitions in figure \ref{fig:coreexample} we might consider \hs{odd} to be more similar to \hs{even} than to \hs{plus}, and hence for $X$ to be more similar to $Y$ than to $Z$.
+In this sense, exact equality will under-estimate the similarity of $X$ and $Y$, whilst ignoring identifiers will over-estimate the similarity of $X$ and $Z$.
 
-Ideally we would like something in between these two extremes, which requires a more sophisticated notion of similarity for identifiers.
-In the above example, we may intuitively consider \texttt{(\&\&)} and \texttt{(++)} to be closer to each other than to \texttt{map}; both in terms of \emph{definition} (e.g.
-only the former can be specified by pattern-matching on their first argument), and by their \emph{properties} (e.g.
-only the former can form monoids).
-
-Recurrent clustering tackles this problem in a simple way: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity in place of the identifiers'.
-This gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
+Recurrent clustering tackles this problem by comparing identifiers \emph{indirectly}: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity instead.
+However, this is a circular definition: to compare the similarity of, say $X$ and $Y$, we would need to look up the global identifiers and compare the expressions \hs{odd} and \hs{even}; yet \hs{odd} and \hs{even} are mutually recursive!
+To  themselves gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
 For this recursive process to be well-founded, we impose a topological ordering on expressions based on their dependencies (the expressions they reference).
@@
 
-\subsubsection{Algorithm}
-\label{sec:algorithm}
 \iffalse TODO: the code should go in the implementation section rather than here \fi

### 2015-12-01 TransferReport/implementation.tex
Updated algorithm

@@
 
-\subsection{Circular Convolution}
-\label{sec:circularconvolution}
+\subsection{\textsc{ML4HS}}
+\label{sec:ml4hs}
 
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.
-
-\subsection{\texttt{nix-eval}}
-\label{sec:nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
+\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin} and \textsc{MLSpec}, described below, with an implementation of our recurrent clustering algorithm and the \textsc{Weka} machine learning library.
 
@@
 
-\subsection{\textsc{ML4HS}}
-\label{sec:ml4hs}
+\subsection{\texttt{nix-eval}}
+\label{sec:nixeval}
 
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin}, \textsc{TreeFeatures}, the \textsc{Weka} machine learning library and \textsc{MLSpec}.
+To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
+This is difficult, as many terms are functions, which Haskell does not let us inspect.
+This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
+To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
 
-Inspired by the use of premise selection (see \S \ref{sec:relevance}) to reduce the search space in ATP,
-we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.
+\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
+In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
 
-We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{sec:clustering}), is an effective method for performing this separation.
+On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
+Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
+Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
 
-% TODO: Future experiments will test this by comparing the throughput of \qspec{} with and without the \textsc{ML4HS} pre-processor.
+We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
+These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
 
-% Divide and conquer for theory exploration
-% Does it help?
+\texttt{nix-eval} is build on top of the Nix package manager and GHC.
+Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
 
-% Clustering and feature extraction
+The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
+When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
+GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
+The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
+The expression, therefore, must have type \hs{String}; although this is not enforced statically.
+
+The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
+The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
+The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
+It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
+using JSON.
+
+\subsection{Circular Convolution}
+\label{sec:circularconvolution}
+
+We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+data Tree = Leaf FV
+          | Branch FV Tree Tree
+\end{lstlisting}
+
+Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
+To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceD 1 (Branch fv l r) = Leaf   fv
+reduceD n (Leaf   fv)     = Leaf   fv
+reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
+                                      (reduceD (n-1) r)
+\end{lstlisting}
+
+This is simple, but suffers two problems:
+
+\begin{itemize}
+ \item We must choose a conservatively large \hs{d}, since we're throwing away information
+ \item The number of feature vectors grows exponentially as \hs{d} increases
+\end{itemize}
+
+The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
+One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
+
+\iffalse
+
+We can do this with \hs{tabulate w d t}:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+tabulate w d t = take d (map (take w) (tabulate' t))
+
+tabulate' (Leaf   fv)     = [fv] : []
+tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
+
+merge    []     ys  = ys
+merge    xs     []  = xs
+merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
+\end{lstlisting}
+
+\fi
+
+Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
+Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
+Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+reduceC (Leaf fv)       = fv
+reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
+\end{lstlisting}
+
+Circular convolution has the following desirable properties:
+
+\begin{description}
+
+  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
+
+  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
+
+\end{description}
+
+Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
+This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
+In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
+
+Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
+
+\subsubsection{Extracting Features from XML}
+
+This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
+The application parses arbitrary trees of XML and uses binary features, ie.
+feature vectors are bit vectors.
+
+To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
+
+\begin{equation}
+feature(n) = 2^{\md5(s) \pmod{L}}
+\end{equation}
+
+Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
+The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
+Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
+
+In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
+
+\subsubsection{Application to Coq}
+
+Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
+We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
+
+Coq will also interpret the standard output of the command, to generate terms and tactics.
+This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.

### 2015-12-04 TransferReport/contributions.tex
More on recurrent clustering

@@
 \label{sec:recurrentclustering}
-
-We take the \emph{recurrent clustering} approach found in ML4PG and ACL2(ml), and implement a variant in the context of Haskell theory exploration.
-Here we describe our recurrent clustering algorithm and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
@@
 
-The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs.
-In our case, we want to compare Core expressions which, as described above, requires a feature extraction algorithm to create a suitable representation.
+We take the \emph{recurrent clustering} approach found in ML4PG and ACL2(ml), and implement a variant in the context of Haskell theory exploration.
+As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
+Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm, to avoid problems with identifiers (described below).
+Here we describe our approach to recurrent clustering and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
-We extract features in two stages: first we construct \emph{rose trees} of features (real numbers), which maintain the nested structure of expressions shown in figure \ref{fig:coresyntax}.
+We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat feature vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the notation $\feature{}$.
+
+\subsubsection{Flattening Expressions}
+
+Our approach will make use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be orthogonal.
+Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
+Since expressions can be nested in different ways, simple traversals (breadth-first, depth-first, post-order, etc.) may cause features from equivalent sub-expressions to be mis-aligned.
+For example, if we read features in breadth-first order we might get the following feature vectors (padded to the same length with $0$):
+
+\begin{small}
+  \begin{equation*}
+    \begin{array}{r@{}l@{}l@{}}
+      X\ &=\ \CAlt\ (\CDataAlt\ \id{x})\ & (\vlocal{y}) \\
+      Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
+    \end{array}
+  \end{equation*}
+
+  \begin{equation*}
+    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
+      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
+      breadthFirst(Y)\ &=\ (\feature{\CAlt},\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &)
+    \end{array}
+  \end{equation*}
+\end{small}
+
+Here the common sub-expression $\vlocal{y}$ is misaligned, such that only $\frac{1}{6}$ features are guaranteed to match (others may match by coincidence, depending on $\feature{}$).
+These feature vectors might be deemed very dissimilar, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
+
+If we were to align these features instead, by padding the third column rather than the sixth, then $\frac{4}{6}$ of the features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
+
+There are many approaches we could take to increase alignment.
+Our method, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size.
+This doesn't guarantee alignment, but it does prevent misalignment from accumulating from one level to the next.
+In this example, our method (when padding each level to 2 features) would give the following feature vectors:
+
+\begin{small}
+  \begin{equation*}
+    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
+      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
+      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
+    \end{array}
+  \end{equation*}
+\end{small}
+
+For this reason, we cannot Core expressions  we will associate first associate features (real numbers) with the syntax of figure \ref{fig:coresyntax}.
+To preserve the in two stages: first we construct \emph{rose trees} of features (real numbers), which maintain the nested structure of expressions shown in figure \ref{fig:coresyntax}.
 We then flatten these rose trees into feature vectors.
-
-\subsubsection{Rose Trees}
 
@@
 \begin{figure}
+  \begin{small}
+  \begin{align*}
+    toTree(\hs{plus}) &= \\
+      (\feature{\CLam},
+       \begin{array}[t]{@{}l@{}}
+         (\feature{\id{a}}), \\
+         (\feature{\CLam},
+          \begin{array}[t]{@{}l@{}}
+            (\feature{\id{y}}), \\
+            (\feature{\CCase},
+             \begin{array}[t]{@{}l@{}}
+               (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
+               (\feature{\id{b}}), \\
+               (\feature{\CAlt},
+                \begin{array}[t]{@{}l@{}}
+                  (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
+                  (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}})))),
+                \end{array} \\
+               (\feature{\CAlt},
+                \begin{array}[t]{@{}l@{}}
+                  (\feature{\CDataAlt}, (\feature{\id{S}})), \\
+                  (\feature{\CApp},
+                   \begin{array}[t]{@{}l@{}}
+                     (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{S}}))), \\
+                     (\feature{\CApp},
+                      \begin{array}[t]{@{}l@{}}
+                        (\feature{\CApp},
+                         \begin{array}[t]{@{}l@{}}
+                           (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{plus}}))), \\
+                           (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{x}})))),
+                         \end{array} \\
+                      (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}}))))),
+                      \end{array}
+                   \end{array} \\
+                  (\feature{\id{x}})))))
+                \end{array}
+             \end{array}
+          \end{array}
+       \end{array} \\
+foo &= bar
+  \end{align*}
+  \end{small}
   \caption{}
@@
 We now define the algorithm for $\feature{}$, which turns terminal symbols into features.
-In the case of known language features, such as $\feature{\hs{Lam}}$ and $\feature{\hs{Case}}$, we just enumerate the possibilities (while trying to make ``similar'' symbols have similar features).
+In the case of known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we just enumerate the possibilities (while trying to make ``similar'' symbols have similar features).
 

### 2015-12-04 TransferReport/background.tex
Fix some notation

@@
                        "b"
-                       [Alt (DataAlt "Z") (Var (Local "y")) [],
-                        Alt (DataAlt "S") (App (Var (Global "S"))
+                       (Alt (DataAlt "Z") (Var (Local "y")))
+                       (Alt (DataAlt "S") (App (Var (Global "S"))
                                                (App (App (Var (Global "plus"))
@@
                                                     (Var (Local "y"))))
-                                          ["x"]]))
+                                          "x")))
       \end{verbatim}
@@
                        "b"
-                       [Alt (DataAlt "Z") (Var (Global "Z")) [],
-                        Alt (DataAlt "S") (App (App (Var (Global "plus"))
+                       (Alt (DataAlt "Z") (Var (Global "Z")))
+                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                     (Var (Local  "y")))
@@
                                                     (Var (Local  "y"))))
-                                          ["x"]]))
+                                          "x")))
       \end{verbatim}
@@
               "b"
-              [Alt (DataAlt "Z") (Var (Global "False")) [],
-               Alt (DataAlt "S") (App (Var (Global "even"))
+              (Alt (DataAlt "Z") (Var (Global "False")))
+              (Alt (DataAlt "S") (App (Var (Global "even"))
                                       (Var (Local  "n")))
-                                 ["n"]])
+                                 "n"))
       \end{verbatim}
@@
               "b"
-              [Alt (DataAlt "Z") (Var (Global "True")) [],
-               Alt (DataAlt "S") (App (Var (Global "odd"))
+              (Alt (DataAlt "Z") (Var (Global "True")))
+              (Alt (DataAlt "S") (App (Var (Global "odd"))
                                       (Var (Local  "n")))
-                                 ["n"]])
+                                 "n"))
       \end{verbatim}

### 2015-12-04 TransferReport/contributions.tex
Fix some notation

@@
 Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
-Since expressions can be nested in different ways, simple traversals (breadth-first, depth-first, post-order, etc.) may cause features from equivalent sub-expressions to be mis-aligned.
-For example, if we read features in breadth-first order we might get the following feature vectors (padded to the same length with $0$):
+Since different patterns of nesting can alter the ``shape'' of expressions, simple traversals (breadth-first, depth-first, post-order, etc.) may cause features from equivalent sub-expressions to be mis-aligned.
+For example, consider the following expressions, which represent pattern-match clauses with different patterns but the same body (\hs{\vlocal{y}}):
 
-\begin{small}
 \begin{equation*}
   \begin{array}{r@{}l@{}l@{}}
-      X\ &=\ \CAlt\ (\CDataAlt\ \id{x})\ & (\vlocal{y}) \\
+    X\ &=\ \CAlt\ (\CDataAlt\ \id{C})\ & (\vlocal{y}) \\
     Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
@@
 
+If we traverse these expressions in breadth-first order, converting each token to a feature using $\feature{}$ and padding to the same length with $0$, we would get the following feature vectors:
+
+\begin{small}
   \begin{equation*}
@@
 
-Here the common sub-expression $\vlocal{y}$ is misaligned, such that only $\frac{1}{6}$ features are guaranteed to match (others may match by coincidence, depending on $\feature{}$).
-These feature vectors might be deemed very dissimilar, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
+Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\feature{}$).
+These feature vectors might be deemed very dissimilar during clustering, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
 
-If we were to align these features instead, by padding the third column rather than the sixth, then $\frac{4}{6}$ of the features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
+If we were to align these feature optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
 
-There are many approaches we could take to increase alignment.
-Our method, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size.
-This doesn't guarantee alignment, but it does prevent misalignment from accumulating from one level to the next.
-In this example, our method (when padding each level to 2 features) would give the following feature vectors:
+Our feature extraction method (described below) uses a variation of breadth-first traversal which pads each level of nesting to a fixed size.
+This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
+For this example, our method (when padding each level to 2 features) would give the following feature vectors:
 
@@
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
-      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
+      featureVec_2(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
+      featureVec_2(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
     \end{array}
@@
 
-For this reason, we cannot Core expressions  we will associate first associate features (real numbers) with the syntax of figure \ref{fig:coresyntax}.
-To preserve the in two stages: first we construct \emph{rose trees} of features (real numbers), which maintain the nested structure of expressions shown in figure \ref{fig:coresyntax}.
-We then flatten these rose trees into feature vectors.
+Here $\frac{1}{2}$ of the original 6 features align, which is more than $breadthFirst$ but not optimal.
+Both vectors have also been padded by an extra 2 zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
 
+To achieve this we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}.
 Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
-$T_i$ are the \emph{children of} $T$ and $f$ is the \emph{feature at} $T$.
-$n_T$ may be different for each child; trees where $n_T = 0$ are \emph{leaves}.
-
-We turn Core expressions into rose trees by converting terminal symbols (such as \hs{Lam} or \hs{Rec}) into features; we represent this conversion using $\feature{}$.
-Non-terminal symbols (such as $expr$ and $id$) are converted recursively and become children.
-This is shown in figure \ref{fig:totree}.
+$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
+$n_T$ may differ for each (sub-) tree for each sub-tree; trees where $n_T = 0$ are \emph{leaves}.
+Rose trees for the Peano arithmetic functions of figure \ref{fig:coreexample} are given in figure \ref{fig:rosetreeexample}.
 
@@
       (\feature{\CLet}, toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase}, toTree(e_1), leaf(l_1), toTree(a_1), \dots) & \text{if $e = \CCase\ e_1\ l_1 [a_1, \dots]$} \\
+      (\feature{\CCase}, toTree(e_1), leaf(l_1), toTree(a_1), \dots) & \text{if $e = \CCase\ e_1\ l_1 a_1 \dots$} \\
       (\feature{\CType})                                            & \text{if $e = \CType$} \\
@@
       (\feature{\CLitStr})                                          & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt}, toTree(e_1), toTree(e_2), leaf{l_1}, \dots)  & \text{if $e = \CAlt\ e_1 e_2 [l_1, \dots]$}  \\
+      (\feature{\CAlt}, toTree(e_1), toTree(e_2), leaf{l_1}, \dots)  & \text{if $e = \CAlt\ e_1 e_2 l_1 \dots$}  \\
       (\feature{\CDataAlt}, leaf(g_1))                              & \text{if $e = \CDataAlt\ g_1$}  \\
@@
       (\feature{\CNonRec}, toTree(e_1))                             & \text{if $e = \CNonRec\ e_1$}  \\
-      (\feature{\CRec}, toTree(e_1), \dots)                         & \text{if $e = \CRec\ [e_1, \dots]$} \\
+      (\feature{\CRec}, toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1 \dots$} \\
       (\feature{\CBind}, leaf(l_1), toTree(e_1))                    & \text{if $e = \CBind\ l_1 e_1$}
@@
   \end{align*}
-  \caption{Recursively converting Core expressions of figure \ref{fig:coresyntax} to rose trees.
-List elements (indicated with $\dots$) are appended as children, rather than being wrapped in another tree.}
+  \caption{Transforming Core expressions of figure \ref{fig:coresyntax} to rose trees.
+The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
+for $\CAlt$ we use $leaf$ for each $l_i$.
+Also note that we ignore the value of literals, as it simplifies our later definition of $\feature{}$ and we conjecture that the effect on clustering real code is low.}
   \label{fig:totree}
@@
 
-To flatten a rose tree we fetch the features from each ``level'', concatenating those from adjacent children:
+To flatten a rose tree we fetch the features from each ``level'', concatenating those from adjacent sub-trees:
 
@@
        \end{array} \\
-foo &= bar
+\iffalse
+    toTree(\hs{mult}) &= \\
+      (Lam "a" (Lam "y" (Case (Var (Local "a"))
+                              "b"
+                              [Alt (DataAlt "Z") (Var (Global "Z")) [],
+                               Alt (DataAlt "S") (App (App (Var (Global "plus"))
+                                                           (Var (Local  "y")))
+                                                      (App (App (Var (Global "mult"))
+                                                                (Var (Local  "x")))
+                                                           (Var (Local  "y"))))
+                                                 ["x"]])))
+
+    toTree(\hs{odd}) &= \\
+      (\feature{\CLam},
+       \begin{array}[t]{@{}l@{}}
+         (\feature{\id{a}}), \\
+         (\feature{\CCase},
+          \begin{array}[t]{@{}l@{}}
+            (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
+            (\feature{\id{b}}), \\
+            (\feature{\CAlt},
+             \begin{array}[t]{@{}l@{}}
+               (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
+               (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{False}})))),
+             \end{array} \\
+            (\feature{\CAlt},
+             \begin{array}[t]{@{}l@{}}
+               (\feature{\CDataAlt}, (\feature{\id{S}})), \\
+               (\feature{\CApp},
+                \begin{array}[t]{@{}l@{}}
+                  (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{even}}))), \\
+                  (\feature{\CVar}, (\feature{\CLocal},  (\feature{\id{n}})))),
+                \end{array} \\
+               (\feature{\id{n}}))))
+             \end{array}
+          \end{array}
+       \end{array}
+\fi
+
+    toTree(\hs{odd}) &= \\
+      \begin{array}
+        (\feature{\CLam}, &(\feature{\id{a}}),&                    &                      &                      &                           &                         \\
+                          &(\feature{\CCase}, &(\feature{\CVar},   &(\feature{\CLocal},   &(\feature{\id{a}}))), &                           &                         \\
+                          &                   &(\feature{\id{b}}), &                      &                      &                           &                         \\
+                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{Z}})),  &                           &                         \\
+                          &                   &                    &(\feature{\CVar},     &(\feature{\CGlobal},  &(\feature{\id{False}})))), &                         \\
+                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{S}})),  &                           &                         \\
+                          &                   &                    &(\feature{\CApp},     &(\feature{\CVar},     &(\feature{\CGlobal},       &(\feature{\id{even}}))), \\
+                          &                   &                    &                      &(\feature{\CVar},     &(\feature{\CLocal},        &(\feature{\id{n}})))),   \\
+                          &                   &                    &(\feature{\id{n}})))) &                      &                           &
+
+       \end{array}
   \end{align*}

### 2015-12-04 TransferReport/contributions.tex
Fleshed out flattening algorithm

@@
 $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
-$n_T$ may differ for each (sub-) tree for each sub-tree; trees where $n_T = 0$ are \emph{leaves}.
-Rose trees for the Peano arithmetic functions of figure \ref{fig:coreexample} are given in figure \ref{fig:rosetreeexample}.
+$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
+The results are illustrated in figure \ref{fig:rosetreeexample}.
 
@@
 
-To flatten a rose tree we fetch the features from each ``level'', concatenating those from adjacent sub-trees:
+\begin{figure}
+  \centering
+  \begin{subfigure}{\textwidth}
+    \begin{small}
+      \Tree[ .$\feature{\CLam}$
+                $\feature{\id{a}}$
+                [ .$\feature{\CCase}$
+                     [ .$\feature{\CVar}$
+                          [ .$\feature{\CLocal}$
+                               $\feature{\id{a}}$ ]]
+                     $\feature{\id{b}}$
+                     [ .$\feature{\CAlt}$
+                          [ .$\feature{\CDataAlt}$
+                               $\feature{\id{Z}}$ ]
+                          [ .$\feature{\CVar}$
+                               [ .$\feature{\CGlobal}$
+                                    $\feature{\id{False}}$ ]]]
+                     [ .$\feature{\CAlt}$
+                          [ .$\feature{\CDataAlt}$
+                               $\feature{\id{S}}$ ]
+                          [ .$\feature{\CApp}$
+                               [ .$\feature{\CVar}$
+                                    [ .$\feature{\CGlobal}$
+                                         $\feature{\id{even}}$ ]]
+                               [ .$\feature{\CVar}$
+                                    [ .$\feature{\CLocal}$
+                                         $\feature{\id{n}}$ ]]]
+                          $\feature{\id{n}}$ ]]]
+    \end{small}
+    \caption{Rose tree for the expression \hs{odd} from \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
+    \label{fig:rosetreeexample}
+  \end{subfigure}
+  \vspace{1ex}
+  \begin{subfigure}{\textwidth}
+    \begin{equation*}
+      \begin{bmatrix}
+        \feature{\CLam}      & 0                   & 0                  & 0                   & 0               & 0                \\
+        \feature{\id{a}}     & \feature{\CCase}    & 0                  & 0                   & 0               & 0                \\
+        \feature{\CVar}      & \feature{\id{b}}    & \feature{\CAlt}    & \feature{\CAlt}     & 0               & 0                \\
+        \feature{\CLocal}    & \feature{\CDataAlt} & \feature{\CVar}    & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
+        \feature{\id{a}}     & \feature{\id{Z}}    & \feature{\CGlobal} & \feature{\id{S}}    & \feature{\CVar} & \feature{\CVar}  \\
+        \feature{\id{False}} & \feature{\CGlobal}  & \feature{\CLocal}  & 0                   & 0               & 0                \\
+        \feature{\id{even}}  & \feature{\id{n}}    & 0                  & 0                   & 0               & 0
+      \end{bmatrix}
+    \end{equation*}
+    \caption{Matrix generated from figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of sub-trees corresponds to a row in the matrix.}
+    \label{fig:matrixexample}
+  \end{subfigure}
+  \vspace{1ex}
+  \begin{subfigure}{\textwidth}
+    \begin{equation*}
+      (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
+     \end{equation*}
+     \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
+     \label{fig:vectorexample}
+  \end{subfigure}
+  \caption{Feature extraction applied to the expression \hs{odd} from figure \ref{fig:coreexample}.}
+\end{figure}
+
+We convert these rose trees into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting using the following function:
 
 \begin{equation*}
-  level(l, (x, T_1, \dots, T_{n_T})) =
+  level(l, (f, T_1, \dots, T_{n_T})) =
     \begin{cases}
-      (x) & \text{if $l = 1$} \\
-      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$ and $\vect{x}_i = level(l - 1, T_i)$}
+      (f) & \text{if $l = 1$} \\
+      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$, where $\vect{x}_i = level(l - 1, T_i)$}
     \end{cases}
@@
 
-Given a rose tree $T$, we pad each level $level(1, T)$, $level(2, T)$, $\dots$, $level(r, T)$ with zeros and truncate them to length $c$.
-The resulting vectors are concatenated to form the feature vector for $T$.
+Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t))$, where $pad$ appends zeros to ensure each row has the same length.
+To improve alignment across different expressions, we actually pad each matrix to the same dimensions $r \times c$, which may require appending rows of zeros or truncating rows/columns as necessary.
 
-See see figure \ref{fig:rosetreeexample} for an example of this construction.
-
-\begin{figure}
-  \begin{small}
-  \begin{align*}
-    toTree(\hs{plus}) &= \\
-      (\feature{\CLam},
-       \begin{array}[t]{@{}l@{}}
-         (\feature{\id{a}}), \\
-         (\feature{\CLam},
-          \begin{array}[t]{@{}l@{}}
-            (\feature{\id{y}}), \\
-            (\feature{\CCase},
-             \begin{array}[t]{@{}l@{}}
-               (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
-               (\feature{\id{b}}), \\
-               (\feature{\CAlt},
-                \begin{array}[t]{@{}l@{}}
-                  (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
-                  (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}})))),
-                \end{array} \\
-               (\feature{\CAlt},
-                \begin{array}[t]{@{}l@{}}
-                  (\feature{\CDataAlt}, (\feature{\id{S}})), \\
-                  (\feature{\CApp},
-                   \begin{array}[t]{@{}l@{}}
-                     (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{S}}))), \\
-                     (\feature{\CApp},
-                      \begin{array}[t]{@{}l@{}}
-                        (\feature{\CApp},
-                         \begin{array}[t]{@{}l@{}}
-                           (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{plus}}))), \\
-                           (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{x}})))),
-                         \end{array} \\
-                      (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}}))))),
-                      \end{array}
-                   \end{array} \\
-                  (\feature{\id{x}})))))
-                \end{array}
-             \end{array}
-          \end{array}
-       \end{array} \\
-\iffalse
-    toTree(\hs{mult}) &= \\
-      (Lam "a" (Lam "y" (Case (Var (Local "a"))
-                              "b"
-                              [Alt (DataAlt "Z") (Var (Global "Z")) [],
-                               Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                           (Var (Local  "y")))
-                                                      (App (App (Var (Global "mult"))
-                                                                (Var (Local  "x")))
-                                                           (Var (Local  "y"))))
-                                                 ["x"]])))
-
-    toTree(\hs{odd}) &= \\
-      (\feature{\CLam},
-       \begin{array}[t]{@{}l@{}}
-         (\feature{\id{a}}), \\
-         (\feature{\CCase},
-          \begin{array}[t]{@{}l@{}}
-            (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
-            (\feature{\id{b}}), \\
-            (\feature{\CAlt},
-             \begin{array}[t]{@{}l@{}}
-               (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
-               (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{False}})))),
-             \end{array} \\
-            (\feature{\CAlt},
-             \begin{array}[t]{@{}l@{}}
-               (\feature{\CDataAlt}, (\feature{\id{S}})), \\
-               (\feature{\CApp},
-                \begin{array}[t]{@{}l@{}}
-                  (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{even}}))), \\
-                  (\feature{\CVar}, (\feature{\CLocal},  (\feature{\id{n}})))),
-                \end{array} \\
-               (\feature{\id{n}}))))
-             \end{array}
-          \end{array}
-       \end{array}
-\fi
-
-    toTree(\hs{odd}) &= \\
-      \begin{array}
-        (\feature{\CLam}, &(\feature{\id{a}}),&                    &                      &                      &                           &                         \\
-                          &(\feature{\CCase}, &(\feature{\CVar},   &(\feature{\CLocal},   &(\feature{\id{a}}))), &                           &                         \\
-                          &                   &(\feature{\id{b}}), &                      &                      &                           &                         \\
-                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{Z}})),  &                           &                         \\
-                          &                   &                    &(\feature{\CVar},     &(\feature{\CGlobal},  &(\feature{\id{False}})))), &                         \\
-                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{S}})),  &                           &                         \\
-                          &                   &                    &(\feature{\CApp},     &(\feature{\CVar},     &(\feature{\CGlobal},       &(\feature{\id{even}}))), \\
-                          &                   &                    &                      &(\feature{\CVar},     &(\feature{\CLocal},        &(\feature{\id{n}})))),   \\
-                          &                   &                    &(\feature{\id{n}})))) &                      &                           &
-
-       \end{array}
-  \end{align*}
-  \end{small}
-  \caption{}
-  \label{fig:rosetreeexample}
-\end{figure}
+Feature vectors are then simply the concatenation of matrix rows $(\mathbf{M}_{1,1} \mathbf{M}_{1,2} \dots \mathbf{M}_{2,1} \mathbf{M}_{2,2} \dots)$, as shown in figure \ref{fig:vectorexample}.
 

### 2015-12-07 TransferReport/background.tex
More contributions

@@
                          |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind \mathcal{L}\ expr
+      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
     \end{split}

### 2015-12-07 TransferReport/contributions.tex
More contributions

@@
 
-\iffalse TODO: Much of this discussion reads as a minor implementation solution rather than an overview of the whole method \fi
-
 \iffalse TODO: Focus the section more on the problem of identifiers, and how recurrent clustering uses clustering as part of feature extraction to solve this \fi
-
-\iffalse TODO: Maybe present the translation from trees to matrices, to vectors first; leaving the conversion of terms to features until afterwards; similar to how ML4PG uses [foo] to mean ``the feature corresponding to foo, whatever that might be'' \fi
-
-\iffalse TODO: Maybe present recurrent clustering in a ``naive'' way: define the conversion function ``[]'' recursively.
-Leave the dependency-ordering, etc.
-for the Algorithm section, as a way of ``building up to'' the result, which avoids the inefficiencies of recursing.
-\fi
 
@@
 
-\subsubsection{Flattening Expressions}
+\subsubsection{Expressions to Vectors}
 
-Our approach will make use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be orthogonal.
+Our recurrent clustering algorithm makes use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be orthogonal.
 Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
@@
 
-Our feature extraction method (described below) uses a variation of breadth-first traversal which pads each level of nesting to a fixed size.
+The method we use to ``flatten'' expressions, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size $c$ (for \emph{columns}).
 This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
-For this example, our method (when padding each level to 2 features) would give the following feature vectors:
+For this example, our method would give the following feature vectors when $c = 2$:
 
 \begin{small}
-  \begin{equation*}
+  \begin{equation}\label{eq:flattened}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      featureVec_2(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
-      featureVec_2(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
+      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
+      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
     \end{array}
-  \end{equation*}
+  \end{equation}
 \end{small}
@@
 
-To achieve this we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}.
+To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}.
 Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
@@
       (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
-      (\feature{\CLam}, leaf(l_1), toTree(e_1))                      & \text{if $e = \CLam\ l_1\ e_1$} \\
+      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
       (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase}, toTree(e_1), leaf(l_1), toTree(a_1), \dots) & \text{if $e = \CCase\ e_1\ l_1 a_1 \dots$} \\
+      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1 a_1 \dots$} \\
       (\feature{\CType})                                                & \text{if $e = \CType$} \\
@@
       (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt}, toTree(e_1), toTree(e_2), leaf{l_1}, \dots)  & \text{if $e = \CAlt\ e_1 e_2 l_1 \dots$}  \\
+      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1 e_2 l_1 \dots$}  \\
       (\feature{\CDataAlt}, leaf(g_1))                                  & \text{if $e = \CDataAlt\ g_1$}  \\
@@
       (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1 \dots$} \\
-      (\feature{\CBind}, leaf(l_1), toTree(e_1))                    & \text{if $e = \CBind\ l_1 e_1$}
+      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1 e_1$}
     \end{cases} \\
@@
 The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
-for $\CAlt$ we use $leaf$ for each $l_i$.
+for $\CRec$ we apply $toTree$ to each $e_i$.
 Also note that we ignore the value of literals, as it simplifies our later definition of $\feature{}$ and we conjecture that the effect on clustering real code is low.}
@@
 
-We convert these rose trees into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting using the following function:
+These rose trees are then turned into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting using the following function:
 
@@
 
-Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t))$, where $pad$ appends zeros to ensure each row has the same length.
-To improve alignment across different expressions, we actually pad each matrix to the same dimensions $r \times c$, which may require appending rows of zeros or truncating rows/columns as necessary.
+Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t), c)$, where $pad$ either truncates or appends zeros, until the row has length $c$.
+We also truncate/pad the number of rows to match a fixed number $r$.
+This way, all expressions give rise to $r \times c$ matrices, where the left-most features at each each level are aligned.
 
@@
 
-\subsubsection{Feature Extraction}
+\subsubsection{Symbols to Features}
 
-We now define the algorithm for $\feature{}$, which turns terminal symbols into features.
-In the case of known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we just enumerate the possibilities (while trying to make ``similar'' symbols have similar features).
+\iffalse TODO: Maybe present recurrent clustering in a ``naive'' way: define the conversion function ``[]'' recursively.
+Leave the dependency-ordering, etc.
+for the Algorithm section, as a way of ``building up to'' the result, which avoids the inefficiencies of recursing.
+\fi
 
-The difficulty comes from identifiers, which we cannot hard-code into our algorithm in advance.
-The only meaningful way to directly compare identifiers is by equality, but since the space of identifiers is large, giving distinct features to distinct identifiers would make matches very unlikely.
-Since identifiers make up a significant portion of expressions, we would like a more fine-grained notion of similarity between their features than exact matching.
+We now define the algorithm for $\feature{}$, which turns terminal symbols of Core syntax into features (real numbers).
+For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14} in Coq.
+We use a constant $\alpha$ to separate these values from those of other tokens (e.g.
+identifiers), but the order is essentially arbitrary: \footnote{In \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
 
-For example, consider the following feature vectors, which are identical except for the global identifiers:
+\begin{align*}
+  \feature{\CAlt}     &= \alpha      &
+  \feature{\CDataAlt} &= \alpha + 1  &
+  \feature{\CLitAlt}  &= \alpha + 2  \\
+  \feature{\CDefault} &= \alpha + 3  &
+  \feature{\CNonRec}  &= \alpha + 4  &
+  \feature{\CRec}     &= \alpha + 5  \\
+  \feature{\CBind}    &= \alpha + 6  &
+  \feature{\CLet}     &= \alpha + 7  &
+  \feature{\CCase}    &= \alpha + 8  \\
+  \feature{\CLocal}   &= \alpha + 9  &
+  \feature{\CGlobal}  &= \alpha + 10 &
+  \feature{\CVar}     &= \alpha + 11 \\
+  \feature{\CLam}     &= \alpha + 12 &
+  \feature{\CApp}     &= \alpha + 13 &
+  \feature{\CType}    &= \alpha + 14 \\
+  \feature{\CLit}     &= \alpha + 15 &
+  \feature{\CLitNum}  &= \alpha + 16 &
+  \feature{\CLitStr}  &= \alpha + 17
+\end{align*}
 
-\begin{equation*}
-  \begin{split}
-    \vect{x} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"odd"})) \\
-    \vect{y} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"even"})) \\
-    \vect{z} =\ & \texttt{App}\ (\texttt{Var}\ (\texttt{Local}\ \texttt{"f"}))\ (\texttt{Var}\ (\texttt{Global}\ \texttt{"plus"}))
-  \end{split}
-\end{equation*}
+To encode \emph{local} identifiers $\mathcal{L}$ we use their $de Bruijn index$, since this is a numeric value suitable for use as a feature, and it gives equal values for $\alpha$-equivalent expressions.
+To calculate these indices the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from figure \ref{fig:totree} for clarity).
+The context is a list of the local identifiers which are in scope, beginning with an empty list for top-level Haskell declarations and prepending identifiers as they are introduced.
 
-From their definitions in figure \ref{fig:coreexample} we might consider \hs{odd} to be more similar to \hs{even} than to \hs{plus}, and hence for $X$ to be more similar to $Y$ than to $Z$.
-In this sense, exact equality will under-estimate the similarity of $X$ and $Y$, whilst ignoring identifiers will over-estimate the similarity of $X$ and $Z$.
+For example, when calculating $toTree(\CLam\ i\ e)$ in context $c$, we make the recursive call $toTree(e)$ in the context of \emph{$c$ prepended with $i$}.
+Similar extensions of the context are performed in the cases of \CBind, \CCase\footnote{The $\mathcal{L}$ value in a \CCase\ expression is bound to the expression being matched against; a technical detail to preserve sharing.}, \CAlt\ (which may introduce an arbitrary number of local identifiers) and \CLet\ (where identifiers are taken from occurrences of \CBind\ in the first expression).
 
-Recurrent clustering tackles this problem by comparing identifiers \emph{indirectly}: we look up the expressions \emph{referenced by} each identifier, and use \emph{their} similarity instead.
-However, this is a circular definition: to compare the similarity of, say $X$ and $Y$, we would need to look up the global identifiers and compare the expressions \hs{odd} and \hs{even}; yet \hs{odd} and \hs{even} are mutually recursive!
-To  themselves gives rise to a recursive process, where feature extraction is interleaved with multiple rounds of clustering.
-For this recursive process to be well-founded, we impose a topological ordering on expressions based on their dependencies (the expressions they reference).
-This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level.
+Since well-formed Haskell declarations do not contain free variables, if we apply $toTree$ to the corresponding Core expression we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$.
+\iffalse NOTE: Can we prove this, or would it just be an exercise in formalising Haskell?
+\fi In which case we define:
+
+$$\feature{l} = i + 2 \alpha$$
+
+Where $i$ is the de Bruijn index of $l$ (the index of the first occurrence of $l$ in the context), and we again use $\alpha$ to separate these features from those of other constructs.
+
+Since the $toTree$ function discards the particular values of numerals and strings, the only remaining case is global identifiers $\mathcal{G}$.
+Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers.
+We also cannot directly encode the form of the identifiers, e.g.
+using a scheme like G{\"o}del numbering, since this is essentially arbitrary and has no effect on their semantic meaning (referencing other expressions).
+
+Instead, we encode global identifiers \emph{indirectly}, by looking up the expressions which they \emph{reference}.
+For $g \in \mathcal{G}$ we define $\feature{g}$ as the \emph{index of the cluster which $g$ appears in}.
+This is where the recurrent nature of the algorithm appears, since we perform k-means clustering \emph{during} feature extraction, and that clustering step, in turn, requires that we perform feature extraction.
+
+For this recursive process to be well-founded, we impose a topological ordering on declarations based on their dependencies (the expressions they reference).
+This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level in the ordering.
+Rather than following such cyclic references forever, references which point to the same level in the ordering (including self-references) get a constant feature value $r$.
+
+By working through a list of declarations $d$ in dependency order, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in figure .
+Any global identifier which has not been encountered yet must be a mutual- or self-reference.
+
+\begin{algorithm}
+  \begin{algorithmic}[1]
+    \Require List $d$ of (identifier, expression) pairs, in dependency order
+    \Procedure{RecurrentCluster}{}
+      \State $C  \gets []$
+      \State $DB \gets \varnothing$
+      \ForAll{$(i, e)$ \textbf{in} $d$}
+        \State $DB \gets DB \cup \{(i, featureVec(e))\}$
+        \State $C  \gets kMeans(DB)$
+      \EndFor
+      \Return $C$
+    \EndProcedure
+  \end{algorithmic}
+  \caption{Recurrent clustering of Core expressions. $d$, $C$ and $DB$ are lists.}\label{alg:recurrent}
+\end{algorithm}
+
+For example, from their definitions in figure \ref{fig:coreexample} we might consider \hs{odd} to be more similar to \hs{even} than to \hs{plus}, and hence for $\feature{\id{odd}}$ to be more similar to $\feature{\id{even}}$ than to $\feature{\id{plus}}$.
+In this sense, exact equality will under-estimate the similarity of $\feature{\id{odd}}$ and $\feature{\id{even}}$, whilst ignoring all differences between identifiers will over-estimate the similarity of $\feature{\id{odd}}$ and $\feature{\id{plus}}$.
 
 \iffalse TODO: the code should go in the implementation section rather than here \fi
-\iffalse TODO: give a more abstract presentation, using mathematical notation \fi
 \iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi
-\iffalse TODO: Define rose trees \fi
-\iffalse TODO: Show an example from earlier, e.g. factorial \fi
 \iffalse TODO: Split into three parts: expressions to rose trees of features; rose trees to matrices to vectors; k-means clustering \fi
@@
 \iffalse TODO: I would probably just restructure:
- - start with the syntax, get to show how, for every line of figure 1, extraction works
- - then transformation to feature vectors
  - then k-means

### 2015-12-07 TransferReport/background.tex
Finishing off contributions

--- 2015-12-04 TransferReport/background.tex
@@
 \newcommand{\CGlobal}{\texttt{Global}}
+\newcommand{\CConstructor}{\texttt{Constructor}}
 \newcommand{\CLitNum}{\texttt{LitNum}}
@@
                          |\ & \CGlobal\      \mathcal{G}         \\
+                         |\ & \CConstructor\ \mathcal{D}         \\
       literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
@@
       alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{G}             \\
+      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                          |\ & \CLitAlt\ literal                  \\
@@
     $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers
+    $\mathcal{G}$ & global identifiers \\
+    $\mathcal{D}$ & constructor identifiers
   \end{tabular}
@@
 We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
+\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
 
@@
                        (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Global "S"))
+                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
                                                (App (App (Var (Global "plus"))
@@
                        "b"
-                       (Alt (DataAlt "Z") (Var (Global "Z")))
+                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
                        (Alt (DataAlt "S") (App (App (Var (Global "plus"))
@@
               "b"
-              (Alt (DataAlt "Z") (Var (Global "False")))
+              (Alt (DataAlt "Z") (Var (Constructor "False")))
               (Alt (DataAlt "S") (App (Var (Global "even"))
@@
               "b"
-              (Alt (DataAlt "Z") (Var (Global "True")))
+              (Alt (DataAlt "Z") (Var (Constructor "True")))
               (Alt (DataAlt "S") (App (Var (Global "odd"))

### 2015-12-07 TransferReport/contributions.tex
Finishing off contributions

@@
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
-Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm, to avoid problems with identifiers (described below).
+Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as algorithm \ref{alg:recurrent}), which goes beyond a simple syntactic characterisation, to allow the features of an expression to depend on those it references.
 Here we describe our approach to recurrent clustering and compare its similarity and differences to those of ML4PG and ACL2(ml).
@@
 
-Our recurrent clustering algorithm makes use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be orthogonal.
+Our recurrent clustering algorithm makes use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be \emph{orthogonal}.
 Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
@@
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
+      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
       breadthFirst(Y)\ &=\ (\feature{\CAlt},\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &)
@@
 This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
-For this example, our method would give the following feature vectors when $c = 2$:
+Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the $toTree$ function would ignore the constructor identifier $\id{C}$ and never produce the feature $\feature{\id{C}}$.
+However, this example is still accurate in terms of laying out the features as given.}
 
 \begin{small}
-  \begin{equation}\label{eq:flattened}
+  \begin{equation*}\label{eq:flattened}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{x}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
+      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
       featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
     \end{array}
-  \end{equation}
+  \end{equation*}
 \end{small}
@@
       (\feature{\CGlobal},  leaf(g_1))                                  & \text{if $e = \CGlobal\ g_1$} \\
+      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
       (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
@@
       (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1 e_2 l_1 \dots$}  \\
-      (\feature{\CDataAlt}, leaf(g_1))                                  & \text{if $e = \CDataAlt\ g_1$}  \\
+      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
       (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
@@
 for $\CRec$ we apply $toTree$ to each $e_i$.
-Also note that we ignore the value of literals, as it simplifies our later definition of $\feature{}$ and we conjecture that the effect on clustering real code is low.}
+We ignore values of $\mathcal{D}$ as constructors can only be distinguished by their type, which we do not currently support.
+We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\feature{}$, and we conjecture that the effect on clustering real code is low.}
   \label{fig:totree}
@@
                      [ .$\feature{\CAlt}$
-                          [ .$\feature{\CDataAlt}$
-                               $\feature{\id{Z}}$ ]
+                          $\feature{\CDataAlt}$
                           [ .$\feature{\CVar}$
-                               [ .$\feature{\CGlobal}$
-                                    $\feature{\id{False}}$ ]]]
+                               $\feature{\CConstructor}$ ]]
                      [ .$\feature{\CAlt}$
-                          [ .$\feature{\CDataAlt}$
-                               $\feature{\id{S}}$ ]
+                          $\feature{\CDataAlt}$
                           [ .$\feature{\CApp}$
@@
     \end{small}
-    \caption{Rose tree for the expression \hs{odd} from \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
+    \caption{Rose tree for the expression \hs{odd} from figure \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
     \label{fig:rosetreeexample}
@@
         \feature{\CLocal}    & \feature{\CDataAlt}     & \feature{\CVar}   & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
-        \feature{\id{a}}     & \feature{\id{Z}}    & \feature{\CGlobal} & \feature{\id{S}}    & \feature{\CVar} & \feature{\CVar}  \\
-        \feature{\id{False}} & \feature{\CGlobal}  & \feature{\CLocal}  & 0                   & 0               & 0                \\
+        \feature{\id{a}}     & \feature{\CConstructor} & \feature{\CVar}   & \feature{\CVar}     & 0               & 0                \\
+        \feature{\CGlobal}   & \feature{\CLocal}       & 0                 & 0                   & 0               & 0                \\
         \feature{\id{even}}  & \feature{\id{n}}        & 0                 & 0                   & 0               & 0
@@
     \end{equation*}
-    \caption{Matrix generated from figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of sub-trees corresponds to a row in the matrix.}
+    \caption{Matrix generated from figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
     \label{fig:matrixexample}
@@
 
-These rose trees are then turned into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting using the following function:
+These rose trees are then turned into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting, and concatenating them together (written as $\cat$):
 
@@
       (f)                                                    & \text{if $l = 1$} \\
-      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$, where $\vect{x}_i = level(l - 1, T_i)$}
+      level(l - 1, T_1) \cat \dots \cat level(l - 1, T_{n_T}) & \text{if $l > 1$}
     \end{cases}
@@
 
-Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t), c)$, where $pad$ either truncates or appends zeros, until the row has length $c$.
+Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t))$, where $pad$ either truncates or appends zeros, until the row has a fixed length $c$.
 We also truncate/pad the number of rows to match a fixed number $r$.
@@
 
-Feature vectors are then simply the concatenation of matrix rows $(\mathbf{M}_{1,1} \mathbf{M}_{1,2} \dots \mathbf{M}_{2,1} \mathbf{M}_{2,2} \dots)$, as shown in figure \ref{fig:vectorexample}.
+\begin{sloppypar}
+Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_{1,1} \cat \mathbf{M}_{1,2} \cat \dots \cat \mathbf{M}_{2,1} \cat \mathbf{M}_{2,2} \cat \dots$, as shown in figure \ref{fig:vectorexample}:
+\end{sloppypar}
+
+\begin{equation}
+  featureVec(e) = pad(level(1, toTree(e))) \cat \dots \cat pad(level(r, toTree(e)))
+\end{equation}
 
@@
   \feature{\CGlobal}       &= \alpha + 10 &
-  \feature{\CVar}     &= \alpha + 11 \\
-  \feature{\CLam}     &= \alpha + 12 &
-  \feature{\CApp}     &= \alpha + 13 &
-  \feature{\CType}    &= \alpha + 14 \\
-  \feature{\CLit}     &= \alpha + 15 &
-  \feature{\CLitNum}  &= \alpha + 16 &
-  \feature{\CLitStr}  &= \alpha + 17
+  \feature{\CConstructor}  &= \alpha + 11 \\
+  \feature{\CVar}          &= \alpha + 12 &
+  \feature{\CLam}          &= \alpha + 13 &
+  \feature{\CApp}          &= \alpha + 14 \\
+  \feature{\CType}         &= \alpha + 15 &
+  \feature{\CLit}          &= \alpha + 16 &
+  \feature{\CLitNum}       &= \alpha + 17 \\
+  \feature{\CLitStr}       &= \alpha + 18
 \end{align*}
 
-To encode \emph{local} identifiers $\mathcal{L}$ we use their $de Bruijn index$, since this is a numeric value suitable for use as a feature, and it gives equal values for $\alpha$-equivalent expressions.
+To encode \emph{local} identifiers $\mathcal{L}$ we use their \emph{de Bruijn index}, since this is a numeric value suitable for use as a feature, and it gives equal values for $\alpha$-equivalent expressions.
 To calculate these indices the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from figure \ref{fig:totree} for clarity).
-The context is a list of the local identifiers which are in scope, beginning with an empty list for top-level Haskell declarations and prepending identifiers as they are introduced.
+The context is a list of the local identifiers which are in scope, prepended as they are encountered.
+The context is initially empty, and only extended when $toTree$ calls itself recursively.
 
-For example, when calculating $toTree(\CLam\ i\ e)$ in context $c$, we make the recursive call $toTree(e)$ in the context of \emph{$c$ prepended with $i$}.
+For example, when calculating $toTree(\CLam\ i\ e)$ in context $x$, we make the recursive call $toTree(e)$ in the context of \emph{$x$ prepended with $i$} (i.e.
+$(i) \cat x$).
 Similar extensions of the context are performed in the cases of \CBind, \CCase\footnote{The $\mathcal{L}$ value in a \CCase\ expression is bound to the expression being matched against; a technical detail to preserve sharing.}, \CAlt\ (which may introduce an arbitrary number of local identifiers) and \CLet\ (where identifiers are taken from occurrences of \CBind\ in the first expression).
 
-Since well-formed Haskell declarations do not contain free variables, if we apply $toTree$ to the corresponding Core expression we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$.
+Well-formed Haskell declarations do not contain free variables (or, equivalently, free variables are encoded as global identifiers rather than local identifiers).
+Hence if we apply $toTree$ to (the Core expression corresponding to) such declarations we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$.
 \iffalse NOTE: Can we prove this, or would it just be an exercise in formalising Haskell?
@@
 
-$$\feature{l} = i + 2 \alpha$$
+\begin{equation}
+  \feature{l} = i + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
+\end{equation}
 
-Where $i$ is the de Bruijn index of $l$ (the index of the first occurrence of $l$ in the context), and we again use $\alpha$ to separate these features from those of other constructs.
+Where $i$ is the index of the first occurrence of $l$ in the context; this is the de Bruijn index of $l$. We again use $\alpha$ to separate these features from those of other constructs.
 
-Since the $toTree$ function discards the particular values of numerals and strings, the only remaining case is global identifiers $\mathcal{G}$.
+Since the $toTree$ function discards numerals, strings and constructor identifiers, the only remaining case is global identifiers $\mathcal{G}$.
 Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers.
@@
 For this recursive process to be well-founded, we impose a topological ordering on declarations based on their dependencies (the expressions they reference).
-This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and several mutually-recursive expressions may appear at the same level in the ordering.
-Rather than following such cyclic references forever, references which point to the same level in the ordering (including self-references) get a constant feature value $r$.
+This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
+To handle this, we perform the sort using an algorithm such as Tarjan's which produces a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.
+References which point to the same SCC as the expression they appear in (which includes direct recursion and mutual recursion) are given a constant feature value $f_{recursion}$.
 
-By working through a list of declarations $d$ in dependency order, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in figure .
-Any global identifier which has not been encountered yet must be a mutual- or self-reference.
+By working through the sorted list of SCCs, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in algorithm \ref{alg:recurrent}.
+This makes cyclic references particularly easy to handle: global identifiers are looked up in the current clusters $C$, and if not found the value $f_{recursion}$ is used.
 
@@
   \begin{algorithmic}[1]
-    \Require List $d$ of (identifier, expression) pairs, in dependency order
+    \Require List $d$ contains SCCs of (identifier, expression) pairs, in dependency order.
     \Procedure{RecurrentCluster}{}
@@
       \State $DB \gets \varnothing$
-      \ForAll{$(i, e)$ \textbf{in} $d$}
-        \State $DB \gets DB \cup \{(i, featureVec(e))\}$
+      \ForAll{$scc$ \textbf{in} $d$}
+        \State $DB \gets DB \cup \{(i, featureVec(e)) \mid (i, e) \in scc\}$
         \State $C  \gets kMeans(DB)$
@@
   \end{algorithmic}
-  \caption{Recurrent clustering of Core expressions. $d$, $C$ and $DB$ are lists.}\label{alg:recurrent}
+  \caption{Recurrent clustering of Core expressions.}
+  \label{alg:recurrent}
 \end{algorithm}
 
-For example, from their definitions in figure \ref{fig:coreexample} we might consider \hs{odd} to be more similar to \hs{even} than to \hs{plus}, and hence for $\feature{\id{odd}}$ to be more similar to $\feature{\id{even}}$ than to $\feature{\id{plus}}$.
-In this sense, exact equality will under-estimate the similarity of $\feature{\id{odd}}$ and $\feature{\id{even}}$, whilst ignoring all differences between identifiers will over-estimate the similarity of $\feature{\id{odd}}$ and $\feature{\id{plus}}$.
+As an example of this recurrent process, we can consider the Peano arithmetic functions from figure \ref{fig:coreexample}.
+A valid topological ordering would be (eliding Core expressions to save space) $d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]$:
 
-\iffalse TODO: the code should go in the implementation section rather than here \fi
-\iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi
-\iffalse TODO: Split into three parts: expressions to rose trees of features; rose trees to matrices to vectors; k-means clustering \fi
-\iffalse TODO: highlight the recurrent nature of the algorithm \fi
-\iffalse TODO: I would probably just restructure:
- - then k-means
-obviously, they are mutually recursive...
-\fi
+\begin{itemize}
+  \item The first iteration through \textsc{RecurrentCluster}'s loop will set $scc \gets \{(\hs{plus}, \dots)\}$.
+  \item With $i = \hs{plus}$ and $e$ as its Core expression, calculating $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$ will become $f_{recursion}$ (since \hs{plus} doesn't appear in $C$).
+  \item The call to $kMeans$ will produce $C \gets [\{\hs{plus}\}]$, i.e. a single cluster containing \hs{plus}.
+  \item The next iteration will set $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
+  \item With $i = \hs{odd}$ and $e$ as its Core expression, the call to \hs{even} will result in $f_{recursion}$.
+  \item Likewise for the call to $\hs{even}$ when $i = \hs{odd}$.
+  \item Since the feature vectors for \hs{odd} and \hs{even} will be identical, $kMeans$ will put them in the same cluster.
+The number of clusters $k = \lceil \sqrt{\lvert DB \rvert} \rceil = 2$, so the other cluster must contain \hs{plus}.
+Their order is arbitrary, so one possibility is $C = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
+  \item Finally \hs{mult} will be clustered. The recursive call will become $f_{recursion}$ whilst the call to \hs{plus} will become $2$, since $\hs{plus} \in C_2$.
+  \item Since $k = 2$, the resulting clusters will be $C \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus}, \hs{mult}\}]$.
+\end{itemize}
 
-Our algorithm proceeds in several stages.
-At the top level, we send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive:
-
-\begin{haskell}
-recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
-recurrentCluster = go ([],[])
-  where go (fs, db) []       = db
-        go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
-                                   db' = kMeans fs'
-                                in go (fs', db') ess
-        extract db (i, e) = (i, rt db e)
-\end{haskell}
-
-The overall result of the \hs{recurrentCluster} function is a list of clusters, containing the IDs of their elements.
-These are obtained by interleaving clustering (the \hs{kMeans} function) and feature extraction (the \hs{rt} function).
-
-The feature extraction itself contains two parts: first, syntax trees matching the grammar in figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
-
-\begin{haskell}
-data RoseTree = Node Feature [RoseTree]
-
-eRT :: [Local] -> [[Global]] -> Expr -> RoseTree
-eRT env db e = case e of
-  ...
-\end{haskell}
-
-The easiest branches to handle are literals and types, which we represent using particular feature values (\hs{sLITNUM}, \hs{sLITSTR} and \hs{sTYPE}):
-
-\begin{haskell}
-  Lit (LitNum _) -> Node sLITNUM []
-  Lit (LitStr _) -> Node sLITSTR []
-  Type           -> Node sTYPE   []
-\end{haskell}
-
-Function application simply recurses into both sub-expressions:
-
-\begin{haskell}
-  App e1 e2 -> Node sAPP [eRT env db e1,
-                          eRT env db e2]
-\end{haskell}
-
-To look up variables locally and globally, we use the \hs{lookupL} and \hs{lookupG} functions, respectively.
-These return the index containing their argument, if found.
-In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors.
-For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features:
-
-\begin{haskell}
-  Var (Global i) -> Node (lookupG db  i) []
-  Var (Local  i) -> Node (lookupL env i) []
-\end{haskell}
-
-Finally, when we traverse binders we must extend the environment \hs{env}:
-
-\begin{haskell}
-  Lam  i  e     -> Node sLAM [eRT (i:env) db e]
-  Let  bs e     -> Node sLET (map (bRT (ids bs:env) db) bs ++
-                                   eRT (ids bs:env) db  e
-  Case e i alts -> Node sCASE (eRT    env  db  e :
-                          map (aRT (i:env) db) alts)
-\end{haskell}
-
-Patterns and bindings are handled in a similar way:
-
-\begin{haskell}
-aRT :: [Local] -> [[Global]] -> Alt -> RoseTree
-aRT env db alt = case alt of
-  (DataAlt _, vs, e) -> eRT (vs ++ env) db e
-  (LitAlt  _, _,  e) -> eRT env db e
-  (Default,   _,  e) -> eRT env db e
-
-bRT :: [Local] -> [[Global]] -> Bind -> RoseTree
-bRT env db b = case b of
-  NonRec i e -> eRT env db e
-  Rec es     -> Node sREC (map (eRT env db . snd) es)
-\end{haskell}
-
-The result of \hs{eRT} is a \hs{RoseTree} whose branching structure mimics that of our original expression.
-We next need to convert this to a \emph{matrix} of features, which we do by converting each level of the \hs{RoseTree} into a row of the matrix (using \hs{pad} to ensure a consistent size).
-Finally we turn the matrix into a \emph{feature vector} by concatenating the rows:
-
-\begin{haskell}
-level :: Int -> RoseTree -> [Feature]
-level 0 (Node x xs) = [x]
-level n (Node x xs) = concatMap (level (n-1)) xs
-
-rt :: [[Global]] -> Expr -> [Feature]
-rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
-\end{haskell}
-
-Notice that this algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
-The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
+Even in this very simple example we can see a few features of our algorithm emerge.
+For example, \hs{odd} and \hs{even} will always appear in the same cluster, since they only differ in their choice of constructor names (which are discarded by $toTree$) and recursive calls (which are replaced by $f_{recursion}$).
+A more extensive investigation of these features requires a concrete implementation, in particular to pin down values for the parameters such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
 
@@
 
-Our algorithm is most similar to that of ML4PG, as our transformation maps each element in a tree to a distinct cell in its matrix.
+Our algorithm is most similar to that of ML4PG, as our transformation maps the elements of a syntax tree to distinct cells in a matrix.
 In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.
@@
 
-The special handling required for mutual recursion is discussed above (namely, topological sorting of expressions and the \hs{sUNKNOWN} sentinel).
+The special handling required for mutual recursion is discussed above (namely, topological sorting of expressions and the $f_{recursion}$ feature).
 Such handling is not present in ML4PG, since the Coq code it analyses must, by virtue of the language, be written in dependency order to begin with.
@@
 \begin{itemize}
-  \item Although they are values, we cannot distinguish between data constructors, other than using exact equality.
+  \item Although they are values, we cannot distinguish between data constructors, other than using exact equality. Hence they are discarded by $toTree$.
   \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot always distinguish between them.
@@
 This will allow fine-grained distinction of expressions which are otherwise identical, especially data constructors.
+Integrating types into our algorithm, and extracting them from Core expressions, is hence left as future work.

### 2015-12-07 TransferReport/implementation.tex
Finishing off contributions

@@
 \label{sec:implementation}
+
+We have implemented these ideas in a theory exploration framework called \textsc{ML4HS}. It contains several components.
+
+\subsection{Recurrent Clustering}
+
+\iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi
+
+Our implementation is a rather direct translation of the algorithm described in \S \ref{sec:contributions} into Haskell.
+At the top level, we send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive:
+
+\begin{haskell}
+recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
+recurrentCluster = go ([],[])
+  where go (fs, db) []       = db
+        go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
+                                   db' = kMeans fs'
+                                in go (fs', db') ess
+        extract db (i, e) = (i, rt db e)
+\end{haskell}
+
+The overall result of the \hs{recurrentCluster} function is a list of clusters, containing the IDs of their elements.
+These are obtained by interleaving clustering (the \hs{kMeans} function) and feature extraction (the \hs{rt} function).
+
+The feature extraction itself contains two parts: first, syntax trees matching the grammar in figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
+
+\begin{haskell}
+data RoseTree = Node Feature [RoseTree]
+
+eRT :: [Local] -> [[Global]] -> Expr -> RoseTree
+eRT env db e = case e of
+  ...
+\end{haskell}
+
+The easiest branches to handle are literals and types, which we represent using particular feature values (\hs{sLITNUM}, \hs{sLITSTR} and \hs{sTYPE}):
+
+\begin{haskell}
+  Lit (LitNum _) -> Node sLITNUM []
+  Lit (LitStr _) -> Node sLITSTR []
+  Type           -> Node sTYPE   []
+\end{haskell}
+
+Function application simply recurses into both sub-expressions:
+
+\begin{haskell}
+  App e1 e2 -> Node sAPP [eRT env db e1,
+                          eRT env db e2]
+\end{haskell}
+
+To look up variables locally and globally, we use the \hs{lookupL} and \hs{lookupG} functions, respectively.
+These return the index containing their argument, if found.
+In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors.
+For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features:
+
+\begin{haskell}
+  Var (Global i) -> Node (lookupG db  i) []
+  Var (Local  i) -> Node (lookupL env i) []
+\end{haskell}
+
+Finally, when we traverse binders we must extend the environment \hs{env}:
+
+\begin{haskell}
+  Lam  i  e     -> Node sLAM [eRT (i:env) db e]
+  Let  bs e     -> Node sLET (map (bRT (ids bs:env) db) bs ++
+                                   eRT (ids bs:env) db  e
+  Case e i alts -> Node sCASE (eRT    env  db  e :
+                          map (aRT (i:env) db) alts)
+\end{haskell}
+
+Patterns and bindings are handled in a similar way:
+
+\begin{haskell}
+aRT :: [Local] -> [[Global]] -> Alt -> RoseTree
+aRT env db alt = case alt of
+  (DataAlt _, vs, e) -> eRT (vs ++ env) db e
+  (LitAlt  _, _,  e) -> eRT env db e
+  (Default,   _,  e) -> eRT env db e
+
+bRT :: [Local] -> [[Global]] -> Bind -> RoseTree
+bRT env db b = case b of
+  NonRec i e -> eRT env db e
+  Rec es     -> Node sREC (map (eRT env db . snd) es)
+\end{haskell}
+
+The result of \hs{eRT} is a \hs{RoseTree} whose branching structure mimics that of our original expression.
+We next need to convert this to a \emph{matrix} of features, which we do by converting each level of the \hs{RoseTree} into a row of the matrix (using \hs{pad} to ensure a consistent size).
+Finally we turn the matrix into a \emph{feature vector} by concatenating the rows:
+
+\begin{haskell}
+level :: Int -> RoseTree -> [Feature]
+level 0 (Node x xs) = [x]
+level n (Node x xs) = concatMap (level (n-1)) xs
+
+rt :: [[Global]] -> Expr -> [Feature]
+rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
+\end{haskell}
+
+Notice that this algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
+The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
 

### 2015-12-08 TransferReport/background.tex
More TODOs

@@
 Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
+Like most \emph{functional} programming languages, Haskell is essentially a variant of $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
 For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
@@
 
+\iffalse TODO: Make this more formal?
+ V \in Var
+ F \in Fun
+ T \in Term
+ T ::= V | F | T1 T2
+
+ Term ::= VAR | Const | Fun (Term)
+or
+ Term t ::= x | f | t t'
+\fi
+
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$.
+  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
+\fi.
 To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
@@
 This is required as the construction of $terms$ applies functions to one argument at a time.
+\iffalse TODO: Maybe mention that not all definitions of a term would regard plus as a well-formed term on its own; maybe make it more explicit that yes, we do regard functions as first-class values, etc \fi
 
@@
 
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
 \begin{itemize}
@@
 
-k-means works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as:
+k-means works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as: \iffalse TODO: Define environment? \fi
 
@@
 \end{equation*}
+
+\iffalse TODO: subsection Available Software \fi

### 2015-12-08 TransferReport/contributions.tex
More TODOs

@@
 We now define the algorithm for $\feature{}$, which turns terminal symbols of Core syntax into features (real numbers).
-For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14} in Coq.
+For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \citep{DBLP:journals/corr/HerasK14} in Coq.
 We use a constant $\alpha$ to separate these values from those of other tokens (e.g.
-identifiers), but the order is essentially arbitrary: \footnote{In \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
+identifiers), but the order is essentially arbitrary: \footnote{In \citep{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
 
@@
 For this recursive process to be well-founded, we impose a topological ordering on declarations based on their dependencies (the expressions they reference).
-This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
+\iffalse TODO: Maybe explain a little more about topological sorting; i.e.
+directed graphs, etc.
+\fi This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
 To handle this, we perform the sort using an algorithm such as Tarjan's which produces a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.

### 2015-12-08 TransferReport/intro.tex
More TODOs

@@
 As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis.
-For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
+The \emph{functional programming} paradigm has been proposed for addressing these issues \citep{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis.
+\iffalse TODO: Types?
+\fi For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
 Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code.
@@
 Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
-Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
+Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \citep{gosling2015java} and C++ \citep{willcock2006lambda}.
 
@@
 
-\begin{itemize}
-  \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
+\begin{enumerate}
   \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of large signatures, which are more tractable to explore.
   \item A novel feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf learning algorithms.
+  \item An implementation of these feature extraction and theory exploration approaches.
   \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
-\end{itemize}
+\end{enumerate}
 
-We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and the results of theory exploration systems like \qspec{} (\S \ref{sec:theoryexploration}).
+We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and describe the \qspec{} theory exploration system (\S \ref{sec:theoryexploration}).
 We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}).

### 2015-12-09 TransferReport/background.tex
Recurrent clustering section

@@
 
-The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
+The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
 For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
@@
   \end{subfigure}
-  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}.
+  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}.
 Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
@@
 Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
-The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the global variables.
-This correspondence allows us to analyse
+The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
+Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
+It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
 
@@
 GHC can represent these, but in this work we only consider reducible expressions (i.e.
-value-level bindings of the form \hs{f a b ...
-= ...}).
+value-level bindings of the form \mbox{\hs{f a b ...
+= ...}}).
 
@@
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
+  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions, which are from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
 \fi.
@@
 
-As an example, we can consider a simple signature containing the expressions from figure \ref{fig:haskellexample}:
+As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
 
@@
   \end{haskell}
-  \caption{Equations conjectured by \qspec{} for the functions in figure \ref{fig:haskellexample}; after simplification.}
+  \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
   \label{fig:qspecresult}
@@
 These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
+As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
 
@@
 
+\iffalse TODO: Say bold letters will denote vectors \fi
+\iffalse TODO: Maybe stick more closely to vectors, rather than sequences? \fi
+
 Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
@@
 \begin{itemize}
-  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$.
+  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$. \iffalse TODO: Any reason for this choice? YES: It should be n/max! \fi
   \item Data points will be $d$-dimensional feature vectors, as defined above.
   \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm.
+  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm. \iffalse TODO:Citation \fi
 \end{itemize}

### 2015-12-09 TransferReport/contributions.tex
Recurrent clustering section

@@
 
-We take the \emph{recurrent clustering} approach found in ML4PG and ACL2(ml), and implement a variant in the context of Haskell theory exploration.
+We adapt the methodology of \emph{recurrent clustering} suggested in CITE, and suggest a new recurrent clustering and feature extraction algorithm for Haskell \iffalse TODO: Fix the end of this sentence \fi theory exploration.
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
@@
 
-To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}.
-Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in Figure \ref{fig:totree}.
+\iffalse TODO: Make a formal definition in an environment, i.e.
+not inline?
+\fi Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
 $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
 $n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
-The results are illustrated in figure \ref{fig:rosetreeexample}.
+The results are illustrated in Figure \ref{fig:rosetreeexample}.
 
@@
     \end{cases} \\
-    leaf(x) &= (\feature{x})
+    leaf(x) &= (\feature{x}) \iffalse TODO: This is completely unnecessary... \fi
   \end{align*}
-  \caption{Transforming Core expressions of figure \ref{fig:coresyntax} to rose trees.
+  \caption{Transforming Core expressions of Figure \ref{fig:coresyntax} to rose trees.
 The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
 for $\CRec$ we apply $toTree$ to each $e_i$.
-We ignore values of $\mathcal{D}$ as constructors can only be distinguished by their type, which we do not currently support.
+We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
 We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\feature{}$, and we conjecture that the effect on clustering real code is low.}
@@
     \end{small}
-    \caption{Rose tree for the expression \hs{odd} from figure \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
+    \caption{Rose tree for the expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
     \label{fig:rosetreeexample}
@@
     \end{equation*}
-    \caption{Matrix generated from figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
+    \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
     \label{fig:matrixexample}
@@
   \end{subfigure}
-  \caption{Feature extraction applied to the expression \hs{odd} from figure \ref{fig:coreexample}.}
+  \caption{Feature extraction applied to the expression \hs{odd} from Figure \ref{fig:coreexample}.}
 \end{figure}
 
-These rose trees are then turned into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting, and concatenating them together (written as $\cat$):
-
+These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting, and concatenating them together (written as $\cat$):
+\iffalse TODO: Maybe make this a bit more of a formal definition? \fi
 \begin{equation*}
@@
 \begin{sloppypar}
-Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_{1,1} \cat \mathbf{M}_{1,2} \cat \dots \cat \mathbf{M}_{2,1} \cat \mathbf{M}_{2,2} \cat \dots$, as shown in figure \ref{fig:vectorexample}:
+Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_{1,1} \cat \mathbf{M}_{1,2} \cat \dots \cat \mathbf{M}_{2,1} \cat \mathbf{M}_{2,2} \cat \dots$, as shown in Figure \ref{fig:vectorexample}:
 \end{sloppypar}
@@
 
-\iffalse TODO: Maybe present recurrent clustering in a ``naive'' way: define the conversion function ``[]'' recursively.
-Leave the dependency-ordering, etc.
-for the Algorithm section, as a way of ``building up to'' the result, which avoids the inefficiencies of recursing.
-\fi
+\iffalse TODO: quine quotes look so much like ceiling. Maybe find a name for it and define it formally in an environment? \fi
 
@@
 To encode \emph{local} identifiers $\mathcal{L}$ we use their \emph{de Bruijn index}, since this is a numeric value suitable for use as a feature, and it gives equal values for $\alpha$-equivalent expressions.
-To calculate these indices the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from figure \ref{fig:totree} for clarity).
+\iffalse TODO: needs to be defined within the text to make formal sense.
+Maybe don't mention it here, but just say, once it's defined, that it's known as the de Bruijn index?
+\fi To calculate these indices the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from Figure \ref{fig:totree} for clarity).
 The context is a list of the local identifiers which are in scope, prepended as they are encountered.
@@
 
+\iffalse TODO: this is a novel feature compared to ML4PG, ACL2(ml), etc., right? Maybe... Emphasize either the connection to earlier work, or the novelty. \fi
 Instead, we encode global identifiers \emph{indirectly}, by looking up the expressions which they \emph{reference}.
@@
 
-As an example of this recurrent process, we can consider the Peano arithmetic functions from figure \ref{fig:coreexample}.
+As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}.
 A valid topological ordering would be (eliding Core expressions to save space) $d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]$:

### 2015-12-09 TransferReport/evaluation.tex
Added TODOs
Recurrent clustering section

@@
+\section{Evaluation}
+
+We evaluate our recurrent clustering algorithm in two ways: by comparison to the clustering algorithms of ML4PG and ACL2(ml), and by comparison to \qspec{}'s theory exploration.
+
+\subsection{Clustering}
+
+\iffalse TODO \fi
+
+\subsection{Theory Exploration}
+
+The most direct way to use our recurrent clustering algorithm for theory exploration is to construct a separate signature for each cluster.
+As we vary $k$ from 1 (equivalent to no clustering at all) to $n$ (the number of elements in the signature), the performance will increase since each signature will produce fewer terms (up to the allowed level of nesting).
+However, the number of equations produced will decrease, as terms in separate clusters are no longer combined.
+
 \iffalse
-\section{First Results}
-
 TODO

### 2015-12-09 TransferReport/future.tex
Added TODOs
Recurrent clustering section

@@
 
-Our use of clustering to pre-process \qcheck{} signatures necessarily involves many decisions and tradeoffs to be made. There are many alternative approaches which are ripe for investigation:
+Our use of clustering to pre-process \qspec{} signatures necessarily involves many decisions and tradeoffs. There are many alternative approaches which are ripe for investigation:
 
@@
 The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
+The resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
 
@@
 
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
 Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
@@
 An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
+$s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$
 

### 2015-12-09 TransferReport/implementation.tex
More TODOs
Recurrent clustering section

@@
 \label{sec:implementation}
+
+\iffalse TODO:
+  A: Conceptually: - implementation follows the inductive/recursive structure of the previous definitions
+e.g. definition of the rose tree is naturally represented as Haskell datatype definition
+
+B Module-by-module road map
+  - Treatment of Haskell package system
+
+C - External software
+\fi
 
@@
 
-The feature extraction itself contains two parts: first, syntax trees matching the grammar in figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
+The feature extraction itself contains two parts: first, syntax trees matching the grammar in Figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
 

### 2015-12-09 TransferReport/related.tex
Feature extraction
Updated algorithm
Recurrent clustering section

@@
 \label{sec:related}
+
+\iffalse TODO: Talk about MuCheck \fi
 
@@
 \iffalse
-
+TODO
 \subsection{Conjecture Generation}
@@
 
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+The initial approach, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
 Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
@@
   \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
+  \item{\textsc{BliStr}}: The Blind Strategy Maker
   \item{\textsc{HOLyHammer}}
@@
 
-\subsection{Clustering Expressions}
-\label{sec:clustering}
+\subsection{Recurrent Clustering}
+\label{sec:clusteringexpressions}
 
-% TODO: ML4PG
-\citep{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\citep{heras2013proof}
+Our recurrent clustering approach takes inspiration from the ML4PG \citep{journals/corr/abs-1212-3618} and ACL2(ml) \citep{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
+Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
+In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
 
-\subsubsection{Learning From Structured Data}
+However, the way we \emph{use} our clusters to inform theory exploration is similar to that of ACL2(ml).
+In ACL2(ml), the clusters are used to reason by analogy: finding theorem statements which are similar to the current goal being proved.
+Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
+For example, if we are trying to prove a theorem \texttt{multiply_commutes}, we might find it appears in the same cluster as \texttt{plus_commutes} which was proved using the lemma \texttt{plus_identity} relating \texttt{plus} and \texttt{0}.
+In this case ACL2(ml) might try mutating \texttt{plus_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \textt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply_identity}) would be given to ACL2 to try and prove \textt{multiply_commutes}.
+
+Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.
+
+\subsection{Feature Extraction}
 

### 2015-12-09 TransferReport/report.tex
Added TODOs
Clustering
Updated algorithm
More on recurrent clustering
Fleshed out flattening algorithm
More contributions
Finishing off contributions
Use ++ for concatenation
Recurrent clustering section

@@
 \usepackage{amssymb}
-\usepackage{algorithm2e}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 \usepackage[T1]{fontenc}
@@
 \usepackage{mmm}
+\usepackage{subcaption}
+\usepackage{tikz-qtree}
 
@@
 
+\newcommand{\blank}{\cdot}
+\newcommand*\mean[1]{\overline{#1}}
+\newcommand*\vect[1]{\mathbf{#1}}
+\newcommand{\argmin}{\operatornamewithlimits{argmin}}
+\newcommand{\feature}[1]{\left\ulcorner #1 \right\urcorner}
+\newcommand{\id}[1]{\texttt{"#1"}}
+\newcommand{\vlocal}[1]{\CVar\ (\CLocal\ \id{#1})}
+\newcommand{\vglobal}[1]{\CVar\ (\CGlobal\ \id{#1})}
+\newcommand{\cat}{\mbox{\ensuremath{+\!\!+\,}}}
+
 \newcommand{\qcheck}{\textsc{QuickCheck}}
@@
 \newcommand{\hspec}{\textsc{HipSpec}}
-
-\providecommand{\hs}[1]{\lstinline[language=Haskell]|#1|}
+\newcommand{\equal}{=}
+\newcommand{\hs}[1]{\texttt{#1}}
 
@@
 \begin{abstract}
-Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
+Theory Exploration is a promising approach to improving the quality and understanding of software.
+It extends previously existing methods available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
 Current theory exploration techniques are limited by their use of exponential time algorithms, which we mitigate by intelligently reducing the size of large inputs, using machine learning to preserve the relationships needed for successful exploration.
+
+\iffalse TODO: I would emphasise more selective (intelligent) approach, and as a result - better precision of TE. Rather than saying that it is merely a size/complexity problem. \fi
 \end{abstract}

### 2015-12-09 TransferReport/related.tex
Fixed rendering issue

@@
 Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
-For example, if we are trying to prove a theorem \texttt{multiply_commutes}, we might find it appears in the same cluster as \texttt{plus_commutes} which was proved using the lemma \texttt{plus_identity} relating \texttt{plus} and \texttt{0}.
-In this case ACL2(ml) might try mutating \texttt{plus_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \textt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply_identity}) would be given to ACL2 to try and prove \textt{multiply_commutes}.
+For example, if we are trying to prove a theorem \texttt{multiply\_commutes}, we might find it appears in the same cluster as \texttt{plus\_commutes} which was proved using the lemma \texttt{plus\_identity} relating \texttt{plus} and \texttt{0}.
+In this case ACL2(ml) might try mutating \texttt{plus\_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \texttt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply\_identity}) would be given to ACL2 to try and prove \texttt{multiply\_commutes}.
 

### 2015-12-09 TransferReport/related.tex
More feature extraction

@@
 
+\iffalse
+We could expand this a bit, e.g. talking about how we both use Weka, etc.
+\fi
+
 \subsection{Feature Extraction}
@@
 In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
+Such nested structures may extend to arbitrary depth, which makes them difficult to process with many machine learning algorithms.
+The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}.
+However, our method is not the only possible way to encode recursive structures as fixed-size features.
 
-\paragraph{Truncation and Padding}
+The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger.
+We use this to make our matrices a uniform size, borrowing the idea from ML4PG.
+Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
+In our case, we transform the recursive structure of expressions into matrices in order to give meanings such as ``the $i$th token at the $j$th level of nesting''.
 
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \citep{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
+Truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
+The less these assumptions hold, the less we can truncate.
+These seem to hold in the case of Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, due to laziness, the lower levels may never even be evaluated.
 
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
+By modelling our inputs as points in high-dimensional spaces, we can consider feature extraction as projection into a lower-dimensional space (known as \emph{dimension reduction}).
+Truncation is a trivial dimension reduction technique; more sophisticated projection functions consider the \emph{distribution} of the input points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
 
-\paragraph{Dimension Reduction}
+Techniques such as \emph{principle component analysis} (PCA) can be used to find these hyperplanes, but unfortunately require their inputs to already have a fixed, integer number of dimensions.
+In the case of our recursive expressions (which we may consider to have \emph{fractal} dimension), we would need another pre-processing stage to satisfy this requirement.
 
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
+There are some machine learning algorithms which can handle variable-size input.
+In particular, \emph{recurrent artificial neural networks} (RNNs) can learn \emph{sequences}, such as audio or written sentences.
+For example, we could render our expressions into a linear sequence of tokens (e.g.
+s-expressions) and process these with an RNN.
+Unfortunately there are several problems with this approach: firstly, we must train RNNs in a \emph{supervised} way, unlike clustering which is \emph{unsupervised}; this requires us to define an error signal which the RNN should minimise.
+There are ways to achieve clustering this way, for example by having the RNN \emph{compress} each expression to a fixed size we get a dimension reduction technique.
+\iffalse and, in fact, clustering CITE Clustering by compression\fi In this case the error signal is the difference between the original input and the output after decompression.
+This is the basis of \emph{autoencoders}.
 
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \citep{hochreiter1997long}).
+Another problem with RNNs is the difficulty of training them.
+If we extend the standard backpropagation algorithm (used for non-recurrent neural networks) to handle the cycles present in RNNs, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
+However, this suffers a problem known as the \emph{vanishing} or \emph{exploding gradient}: error values grow or decay exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
+The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train, rather than backpropagation) and \emph{long short-term memory} (LSTM; introducing a few special, nodes to persist values for long time periods, unaffected by training \citep{hochreiter1997long}).
 

### 2015-12-09 TransferReport/related.tex
Finished feature extraction/machine learning

@@
 
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to process with many machine learning algorithms.
-The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}.
-However, our method is not the only possible way to encode recursive structures as fixed-size features.
+One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
+This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways.
+The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
 
@@
 Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
-In our case, we transform the recursive structure of expressions into matrices in order to give meanings such as ``the $i$th token at the $j$th level of nesting''.
+In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.
 
-Truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-These seem to hold in the case of Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, due to laziness, the lower levels may never even be evaluated.
+Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian'').
+This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness.
+This allows us to truncate more aggressively than if the leaves were most significant.
 
@@
 
-There are some machine learning algorithms which can handle variable-size input.
-In particular, \emph{recurrent artificial neural networks} (RNNs) can learn \emph{sequences}, such as audio or written sentences.
-For example, we could render our expressions into a linear sequence of tokens (e.g.
-s-expressions) and process these with an RNN.
-Unfortunately there are several problems with this approach: firstly, we must train RNNs in a \emph{supervised} way, unlike clustering which is \emph{unsupervised}; this requires us to define an error signal which the RNN should minimise.
-There are ways to achieve clustering this way, for example by having the RNN \emph{compress} each expression to a fixed size we get a dimension reduction technique.
-\iffalse and, in fact, clustering CITE Clustering by compression\fi In this case the error signal is the difference between the original input and the output after decompression.
-This is the basis of \emph{autoencoders}.
+There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise.
+Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.
 
-Another problem with RNNs is the difficulty of training them.
-If we extend the standard backpropagation algorithm (used for non-recurrent neural networks) to handle the cycles present in RNNs, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing} or \emph{exploding gradient}: error values grow or decay exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train, rather than backpropagation) and \emph{long short-term memory} (LSTM; introducing a few special, nodes to persist values for long time periods, unaffected by training \citep{hochreiter1997long}).
+One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs).
+These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations.
+In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.
 
-The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
+Unfortunately RNNs are difficult to train.
+The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
+However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
+The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
 
-% TODO
-\citep{Gartner2003}
-\citep{Oveisi.Oveisi.Erfanian.ea:2012}
-\citep{bakir2007predicting}
-\citep{conf/ijcai/Plate91}
-\citep{goller1996learning}
-\citep{kwasny1995tail}
-\citep{pollack1990recursive}
-\citep{zanzotto2012distributed}
+Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
+The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs.
+Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}.
+More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
 
-\iffalse
+A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
+Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
 
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
+\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
+Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.

### 2015-12-09 TransferReport/related.tex
Relevance filtering

@@
 
-% TODO
-\citep{kuhlwein2012overview}
-
 The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
@@
 We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
+This leads to the idea of \emph{relevance filtering} (or, \emph{premise selection}, when viewed as the \emph{addition} of relevant information to an initially-empty problem).
+This is the core idea behind our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time.
 
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
+Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
+The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
 This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
 
-The initial approach, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+The initial approach taken by Sledgehammer, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
 Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
@@
 
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \citep{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
+Recently, a variety of alternative algorithms have also been investigated, for example, the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
+Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
 To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
 
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}: The Blind Strategy Maker
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
+Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
+Many more approaches are described and evaluated in \citep{kuhlwein2012overview}.
 

### 2015-12-09 TransferReport/related.tex
Haskell

@@
 
-Haskell is a convenient choice for our purposes for several reasons. Here we discuss the relevant language features from a high-level:
+Whilst \S \label{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
+Here we discuss the relevant language features from a high-level, which motivated our choice:
 
@@
 
-\item{Functional}: All control flow is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
+\item{Functional}: All control flow in Haskell is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
 
@@
 in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{sec:theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems.
+This puts Haskell in a rather special position with regards to the discovery of interesting theorems.
 Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
@@
 
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}, and counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}.
+\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
 
@@
 
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
+The task of \emph{conjecture generation} lies at the heart of theory exploration, and whilst its main use has been for lemma generation (i.e. performing ) and even significant theorems.
 

### 2015-12-09 TransferReport/contributions.tex
Fixes

@@
 
-We adapt the methodology of \emph{recurrent clustering} suggested in CITE, and suggest a new recurrent clustering and feature extraction algorithm for Haskell \iffalse TODO: Fix the end of this sentence \fi theory exploration.
+We adapt the methodology of \emph{recurrent clustering} proposed in \citep{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell \iffalse TODO: Fix the end of this sentence \fi theory exploration.
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).

### 2015-12-09 TransferReport/related.tex
Fixes

@@
 \label{sec:related}
-
-\iffalse TODO: Talk about MuCheck \fi
 
@@
 
-Whilst \S \label{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
+Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
 Here we discuss the relevant language features from a high-level, which motivated our choice:
@@
 
-\iffalse
-TODO
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and whilst its main use has been for lemma generation (i.e. performing ) and even significant theorems.
-
-\fi
-
 \subsection{Relevance Filtering}

### 2015-12-10 TransferReport/contributions.tex
Updates

@@
 
-For this recursive process to be well-founded, we impose a topological ordering on declarations based on their dependencies (the expressions they reference).
-\iffalse TODO: Maybe explain a little more about topological sorting; i.e.
-directed graphs, etc.
-\fi This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
-To handle this, we perform the sort using an algorithm such as Tarjan's which produces a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.
+For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference).
+To do this, we construct a directed graph, where each declaration is a node and edges denote an ``is referenced by'' relation.
+An example dependency graph is given in Figure \ref{fig:dependencyexample}.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[node distance=2cm]
+      \tikzset{vertex/.style = {minimum size=1.5em}}
+      \tikzset{edge/.style = {->,> = latex'}}
+
+      \node[vertex]                at (0.5\textwidth, 3)   (z)     {\hs{Z}};
+      \node[vertex]                at (0.5\textwidth, 0)   (s)     {\hs{S}};
+      \node[vertex]                at (0,             1.5) (plus)  {\hs{plus}};
+      \node[vertex, anchor=east]   at (\textwidth,    1.5) (false) {\hs{False}};
+      \node[vertex, right of=plus]                         (mult)  {\hs{mult}};
+      \node[vertex, left of=false]                         (odd)   {\hs{odd}};
+      \node[vertex, left of=odd]                           (even)  {\hs{even}};
+      \node[vertex, left of=even]                          (true)  {\hs{True}};
+
+      \draw[edge, dashed]     (z)     to (plus);
+      \draw[edge, dashed]     (z)     to (mult);
+      \draw[edge, dashed]     (z)     to (odd);
+      \draw[edge, dashed]     (z)     to (even);
+      \draw[edge, dashed]     (s)     to (plus);
+      \draw[edge, dashed]     (s)     to (mult);
+      \draw[edge, dashed]     (s)     to (odd);
+      \draw[edge, dashed]     (s)     to (even);
+      \draw[edge, dashed]     (false) to (odd);
+      \draw[edge, dashed]     (true)  to (even);
+      \draw[edge]             (plus)  to (mult);
+      \draw[edge]             (even)  to (odd);
+      \draw[edge]             (odd)   to (even);
+      \draw[edge, loop above] (plus)  to (plus);
+      \draw[edge, loop above] (mult)  to (mult);
+    \end{tikzpicture}
+
+    \caption{Dependency graph for Figure \ref{fig:haskellexample}.
+Loops indicate recursive functions, double arrows indicate mutual recursion.
+Dashed lines show references to data constructors, which we do not consider in this work.}
+    \label{fig:dagexample}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \centering
+    \tikzstyle{block} = [rectangle, draw]
+    \tikzstyle{line}  = [draw, -latex']
+    \begin{tikzpicture}[node distance=2cm]
+      \node [block] (plusset) {
+        \begin{tikzpicture}[node distance = 2cm]
+          \node [block] (plus) {\hs{plus}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=plusset] (multset) {
+        \begin{tikzpicture}[node distance = 2cm]
+          \node [block] (mult) {\hs{mult}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=multset] (parityset) {
+        \begin{tikzpicture}[node distance = 1cm]
+          \node [block]               (odd)  {\hs{odd}};
+          \node [block, below of=odd] (even) {\hs{even}};
+        \end{tikzpicture}
+      };
+
+      \path [line] (plusset)     -- (multset);
+      \path [line] (multset)     -- (parityset);
+    \end{tikzpicture}
+    \caption{One possible topological sorting of simply connected components for Figure \ref{fig:dagexample} (ignoring constructors).
+\hs{even} and \hs{odd} are mutually recursive, no neither can appear before the other.}
+    \label{fig:sccexample}
+  \end{subfigure}
+  \label{fig:dependencyexample}
+  \caption{Sorting functions from Figure \ref{fig:haskellexample} into dependency order.}
+\end{figure}
+
+In a topological sort we say that $A < B$ for nodes $A$ and $B$ if we can reach $B$ by following edges from $A$ (if $A$ and $B$ cannot be reached from each other, their order is arbitrary).
+This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
+To handle this, we use an algorithm such as Tarjan's \citep{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations (as shown in Figure \ref{fig:sccexample}).
 References which point to the same SCC as the expression they appear in (which includes direct recursion and mutual recursion) are given a constant feature value $f_{recursion}$.

### 2015-12-10 TransferReport/future.tex
Fixes
Updates

@@
 
+\iffalse
+TODO
+\subsection{Conjecture Generation}
+
+The task of \emph{conjecture generation} lies at the heart of theory exploration, and whilst its main use has been for lemma generation (i.e. performing ) and even significant theorems.
+
+\fi
+
 \subsubsection{Interestingness in Evolutionary Computation}
@@
 \fi
+
+\iffalse
+TODO
+
+Other uses of AST Plugin data. Maybe mention this in the context of data generation (e.g. lazysmallcheck and friends)
+
+\begin{itemize}
+  \item{Reflection}: As mentioned above, \hspec{} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
+Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
+
+  \item{Counterexample Generation}: As mentioned in \S \ref{sec:quickcheck}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
+Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
+as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
+As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail.
+Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
+It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
+\end{itemize}
+\fi

### 2015-12-10 TransferReport/implementation.tex
Updates

@@
 
-We have implemented these ideas in a theory exploration framework called \textsc{ML4HS}. It contains several components.
+We provide an implementation of our recurrent clustering algorithm in a tool called \textsc{ML4HS}, which consists of a loose collection of components shown in Figure \ref{fig:ml4hs}.
+This arrangement makes it easy to swap out parts for experimentation.
+In the following, we describe the custom components in the order they appear in the diagram.
+
+\begin{figure}
+  \centering
+  \tikzstyle{block} = [rectangle, draw, rounded corners]
+  \tikzstyle{container} = [rectangle, draw, rounded corners]
+  \tikzstyle{line} = [draw, -latex']
+  \colorlet{shade}{rgb:black,1;white,4}
+
+  \begin{tikzpicture}[node distance=2cm]
+    \node [block] (hackage) {Hackage};
+
+    \node [block, below of=hackage] (cabal) {
+      \begin{tikzpicture}[node distance = 2cm]
+        \node (caballabel) {Cabal};
+        \node [block, anchor=north west] at (caballabel.south) (ghc) {
+          \begin{tikzpicture}[node distance = 2cm]
+            \node (ghclabel) {GHC};
+            \node [block, anchor=north west, fill=shade] at (ghclabel.south) (plugin) {AST Plugin};
+          \end{tikzpicture}
+        };
+      \end{tikzpicture}
+    };
+
+    \node [block, below of=cabal, fill=shade]  (sorting) {Sorting};
+
+    \node [block, below of=sorting, fill=shade] (clustering) {
+      \begin{tikzpicture}[node distance = 2cm, auto]
+        \node (clusterlabel) {Recurrent Clustering};
+        \node [block, anchor=north west, fill=shade] at (clusterlabel.south west) (fe) {Feature Extraction};
+        \node [block, anchor=west, fill=white] at ([xshift=2em]fe.east) (weka) {Weka};
+
+        \path [line] (fe)   -- (weka);
+        \path [line] (weka) -- (fe);
+      \end{tikzpicture}
+    };
+
+    \node [block, below of=clustering, fill=shade] (mlspec) {
+      \begin{tikzpicture}[node distance = 2cm, auto]
+        \node (mlspeclabel) {MLSpec};
+        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {QuickSpec};
+      \end{tikzpicture}
+    };
+
+    \node [block, below of=mlspec] (user) {User};
+
+    \path [line] (hackage)    -- (cabal);
+    \path [line] (cabal)      -- (sorting);
+    \path [line] (sorting)    -- (clustering);
+    \path [line] (clustering) -- (mlspec);
+    \path [line] (mlspec)     -- (user);
+  \end{tikzpicture}
+  \caption{Components of the ML4HS theory exploration system. Custom components are shaded, arrows indicate data flow.}
+  \label{fig:ml4hs}
+\end{figure}
+
+\subsection{\textsc{ASTPlugin}}
+\label{sec:astplugin}
+
+The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core.
+It also includes a \emph{renaming} transformation which makes all names unique (either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
+
+Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
+Indeed, \hspec{} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
+However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
+
+This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
+All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
+
+Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
+All of the above complexities will be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
+This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
+
+GHC provides a plugin mechanism for manipulating Core during a build, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
+We provide a plugin called \textsc{AstPlugin} which emits a serialised version of each Core definition to the console (to satisfy the type system, it also implements a dummy ``optimisation'' which returns the Core unchanged).
+
+Compared to Haskell, Core is a much simpler language and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
+Three areas which make Core difficult to handle are:
+
+\begin{description}
+  \item{Type variables}: Parametric polymorphism (described in more detail in \S \ref{sec:haskelldesc}) can be thought of as values being parameterised by type-level objects.
+In System F, this is represented explicitly by a special abstraction form $\Lambda$, distinct from the $\lambda$ used for values.
+Core only has one abstraction form, \CLam, for both types and values.
+This alters function properties like arity.
+
+  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
+Since Core does not make these distinctions, names may become ambiguous.
+For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
+To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core.
+This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
+
+  \item{Violating encapsulation}: Although Haskell allows names to be \emph{private} to a module, when compiling Core we have full access to private definitions, as well as references to private names from within other definitions.
+Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
+\end{description}
+
+In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to reference that name within the GHCi interpreter.
+Names with the above problems will cause an error, and are discarded.
+
+The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
+Definitions are indexed by a combination of their package name, module name and binding name.
+The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which allows dependencies to be discovered.
+Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
+
+\subsection{Toplogical Sorting}
+
+As described in \S \ref{} We have a tool which annotates each definition with its dependencies, i.e.
+those (package, module, name) combinations which it references.
+Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
 
@@
 \textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin} and \textsc{MLSpec}, described below, with an implementation of our recurrent clustering algorithm and the \textsc{Weka} machine learning library.
-
-\subsection{\textsc{ASTPlugin}}
-\label{sec:astplugin}
-
-The Glasgow Haskell Compiler provides mechanisms for parsing Haskell code into tree structures, and a \emph{renaming} transformation which makes all names unique; either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \hspec{} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities can be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating its Core representation, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which implements an optimisation pass which returns its argument unchanged, but which also emits a serialised version of the definition to the console.
-
-There are several complications in the implementation of \textsc{AstPlugin}.
-We are working with Core, which as a variant of System F\textsubscript{C} \citep{sulzmann2007system} has slightly different syntax and semantics to Haskell.
-This is mostly a benefit, since Core is a much simpler language than Haskell and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core more difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: In \S \ref{sec:haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters.
-Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values.
-This difference from Haskell, which only has explicit binding of values, alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder \texttt{$\lambda$ \$dOrd} in Core.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: As discussed in \S \ref{sec:haskell}, Haskell allows names to be \emph{private} to a module.
-When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to type-check a Haskell expression which imports that name and passes it to \qspec{}.
-Those which fail, due to the reasons above and others, are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name (AKA database keys).
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-Such data is useful for multiple purposes:
-
-\begin{itemize}
-  \item{Dependency tracking}: We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
-
-  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored.
-As mentioned in \S \ref{sec:featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.
-
-  \item{Theory exploration}: The \textsc{MLSpec} tool, described below, uses the output of \textsc{AstPlugin} to construct theories suitable for \qspec{} to explore.
-
-  \item{Reflection}: As mentioned above, \hspec{} uses the GHC API to obtain definitions of Haskell terms, which are then converted for use by theorem provers.
-Although we have not yet investigated this use, it seems likely that the output of \textsc{AstPlugin} would provide a more comprehensive and robust alternative to the GHC API approach.
-
-  \item{Counterexample Generation}: As mentioned in \S \ref{sec:quickcheck}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing.
-Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g.
-as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing.
-As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail.
-Also, the exponential complexity of such naive recursion makes these generators unusable in many cases.
-It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
-\end{itemize}
 

### 2015-12-10 TransferReport/related.tex
Updates

@@
 
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
+Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little explanation was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widely used language like Java.
 Here we discuss the relevant language features from a high-level, which motivated our choice:
@@
 This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
+Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for theory exploration.
 
@@
 in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-This puts Haskell in a rather special position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
+This puts Haskell in a rather special position with regards to the discovery of interesting theorems; namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
+The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence theory exploration would be unlikely to discover unexpected properties.
 
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
+Other empirical advantages to studying Haskell, compared to other programming languages or theorem proving systems, include:
 
@@
 
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
+\item The existence of theory exploration systems such as \hspec{}, and related tools which we may be able to re-use, including conjecture generators like \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
 
@@
 
-The initial approach taken by Sledgehammer, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
+The initial approach taken by Sledgehammer, known as \textsc{MePo} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+Initially, the relevant symbols are those which occur in the goal to be proved, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
 There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
@@
 
-Recently, a variety of alternative algorithms have also been investigated, for example, the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
+Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
 Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
@@
 Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}.
+Many more approaches are described and evaluated in \citep{kuhlwein2012overview}, some of which may be directly applicable in the context of theory exploration.
 
@@
 In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
+Whilst it may be interesting to contrast our current algorithm with an alternative based on that of ACL2(ml), it is unclear how such summaries could be extended to include types, which seems the next logical step for our approach.
+The ML4PG algorithm extends trivially, by using (term, type) pairs instead of just terms.
 
-However, the way we \emph{use} our clusters to inform theory exploration is similar to that of ACL2(ml).
-In ACL2(ml), the clusters are used to reason by analogy: finding theorem statements which are similar to the current goal being proved.
-Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
-For example, if we are trying to prove a theorem \texttt{multiply\_commutes}, we might find it appears in the same cluster as \texttt{plus\_commutes} which was proved using the lemma \texttt{plus\_identity} relating \texttt{plus} and \texttt{0}.
-In this case ACL2(ml) might try mutating \texttt{plus\_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \texttt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply\_identity}) would be given to ACL2 to try and prove \texttt{multiply\_commutes}.
+The way we \emph{use} our clusters to inform theory exploration is actually more similar to that of ACL2(ml) than ML4PG.
+ML4PG can either present clusters to the user for inspection, or produce automata for recreating proofs.
+In ACL2(ml), the clusters are used to restrict the search space of a proof search, much like we restrict the scope of theory exploration.
+
+ACL2(ml) reasons by analogy: finding theorem statements which are similar to the current goal, and attempting to prove the goal in a similar way.
+In particular, the lemmas used to prove a theorem are mutated by substituting symbols for those which appear in the same cluster.
+For example, if \texttt{plus} and \texttt{multiply} are clustered together, and we are trying to prove a goal involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem involving \texttt{plus}.
+The lemmas used to prove that theorem will be mutated, for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an attempt to prove the goal.
 
@@
 One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
-This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways.
-The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
+This can lead to nested expressions of arbitrary depth, which are difficult to compare in numerical ways.
+One solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
 
@@
 
-One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs).
-These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations.
-In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.
+One example of learning from variable-size input is to use \emph{recurrent neural networks} (RNNs).
+These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' NNs, allowing state to persist between observations.
+In this way, each data point can be divided into a sequence of arbitrary length, for example an s-expression, and fed into an RNN one token at a time for processing.
 
 Unfortunately RNNs are difficult to train.
-The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
+The standard way to train NNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
 However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
@@
 Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}.
+The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward NN to learn recursive distributed representations \citep{pollack1990recursive} which correspond to the recursive structure of the inputs.
+Such distributed representations can also be used for sequences, which we can use to encode sub-trees when the branching factor of nodes is not  uniform \citep{kwasny1995tail}.
 More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
@@
 \emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.
+Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}, which could be used to learn more subtle relations between expressions than recurrent clustering alone.
+
+\subsection{K-Means}
+
+We use the Weka tool to perform k-means clustering \citep{Holmes.Donkin.Witten:1994}, since we are more concerned with the application of feature extraction to Haskell and its use in theory exploration, rather than precise tuning of learning algorithms.
+Since k-means is a standard method, there are many other implementations available.
+More interestingly, there are many other clustering algorithms we could use, such as \emph{expectation maximisation} \footnote{In fact, k-means is very similar to expectation-maximisation, as it alternates between an \emph{expectation step} (finding the mean value of each cluster) and a \emph{maximisation step} (assigning points to the cluster they're most similar to; or alternatively, \emph{minimising} the distance of each point to the centre of its cluster, as per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have shown little difference in their results; in effect, the quality of our features is the bottleneck to learning, so there is no reason to avoid a fast algorithm like k-means.
+
+In any case there are many conservative improvements to the standard k-means algorithm, which could be applied to our setup.
+For example, a more efficient approach like \emph{yinyang k-means} \citep{conf/icml/DingZSMM15} could make larger input sizes more practical to cluster, especially since recurrent clustering invokes k-means many times.
+The \emph{k-means++} approach \citep{arthur2007k, bahmani2012scalable} can be used to more carefully select the ``seed'' values for the first timestep, and the \emph{x-means} algorithm \citep{pelleg2000x} is able to estimate how many clusters to use (our \emph{final} clusters should be tuned to maximise the performance of the subsequent theory exploration step, but x-means could still be useful in the recurrent clustering steps).

### 2015-12-15 TransferReport/background.tex
Relevance filtering
Updates
Added Gantt chart and a little extra to evaluation

@@
 Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most \emph{functional} programming languages, Haskell is essentially a variant of $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
+Like most \emph{functional} programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
 For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
@@
 
-The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
+The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major restriction is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
 For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
@@
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions, which are from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
+  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
 \fi.
@@
 
-Notice that Haskell curries functions, so the binary functions \hs{plus} and \hs{mult} can be treated as unary functions which return unary functions.
+Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments.
+In addition, Haskell curries functions, so these binary functions will be treated as unary functions which return unary functions.
 This is required as the construction of $terms$ applies functions to one argument at a time.
-\iffalse TODO: Maybe mention that not all definitions of a term would regard plus as a well-formed term on its own; maybe make it more explicit that yes, we do regard functions as first-class values, etc \fi
 
@@
 Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
+The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
 We describe this approach in more details in \S \ref{sec:relevance}.
-Premise selection is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+Relevance filtering is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
 
-Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
-The reason is that premise selection requires a distinguished expression to compare everything against; in practice this is the current goal of a theorem prover.
+Despite the idea's promise, we cannot simply invoke existing relevance filter algorithms in our theory exploration setting.
+The reason is that relevance filtering is a supervised learning method, i.e.
+it would require a distinguished expression to compare everything against.
 Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
@@
 
-\iffalse TODO: Say bold letters will denote vectors \fi
-\iffalse TODO: Maybe stick more closely to vectors, rather than sequences? \fi
-
 Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
@@
 
-We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$.
-\footnote{In fact, ractical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
+We use a standard representation of features as a \emph{feature vector} of numbers $\vect{x} = (x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$ (we use \textbf{bold face} to represent vectors, including feature vectors).
+\footnote{In fact, practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
 
@@
 they will all have length (or \emph{dimension}) $d$.
-Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
+Many machine learning algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
 For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
   \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
-This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
+This enables many powerful machine learning algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
   \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
@@
 
-\subsubsection{k-Means}
+\subsubsection{K-Means}
 \label{sec:kmeans}
@@
 \begin{itemize}
-  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$. \iffalse TODO: Any reason for this choice? YES: It should be n/max! \fi
+  \item For simplicity, we use the ``rule of thumb'' given in \citep[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
   \item Data points will be $d$-dimensional feature vectors, as defined above.
   \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm. \iffalse TODO:Citation \fi
+  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \citep{lloyd1982least}.
 \end{itemize}
 
-This is a standard setup, supported by off-the-shelf tools. In particular, it is similar to that used by the ML4PG and ACL2(ml), which makes our results more easily comparable.
+This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \citep{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
 
-k-means works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as: \iffalse TODO: Define environment? \fi
+Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
+We denote the clusters as $C^1$ to $C^k$.
+As the name suggests, k-means uses the mean value of each cluster, which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
 
 \begin{equation*}
-  \mean{X}_i = \frac{\sum_{x \in X} x_i}{|X|}
+  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
 \end{equation*}
 
-Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$ and their mean values as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation*}
-  \vect{m}^i(t) = \mean{C}^i(t) \quad \text{for $t > 0$}
-\end{equation*}
-
-Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^i(0)$.
+Before k-means starts, we must choose \emph{seed} values for $\vect{m}^i(0)$.
 Many methods have been proposed for choosing these values \citep{arthur2007k}.
@@
 
-\begin{equation*}
+\begin{equation} \label{eq:kmeansobjective}
   \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation*}
-
-\iffalse TODO: subsection Available Software \fi
+\end{equation}

### 2015-12-15 TransferReport/future.tex
Added Gantt chart and a little extra to evaluation

--- 2015-12-09 TransferReport/future.tex
@@
 
-\subsection{Extensions}
+\begin{sidewaysfigure}
+  \centering
+  \includegraphics[trim=3cm 2cm 3cm 2cm, width=1.00\textwidth]{Gantt.pdf}
+  \caption{Gantt chart showing research direction, sub-tasks and approximate timeframes. CICM is the \emph{Conference on Intelligent Computer Mathematics}.}
+  \label{fig:gantt}
+\end{sidewaysfigure}
+
+Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made.
+Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further.
+In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique.
+The Gantt chart in Figure \ref{fig:gantt} shows how these relate to the short- and long-term direction being taken by this research.
+Below, we elaborate on the details, background and motivation for these choices.
+
+\subsection{Clustering Extensions}
 \label{sec:preprocessing}
 
-Our use of clustering to pre-process \qspec{} signatures necessarily involves many decisions and tradeoffs. There are many alternative approaches which are ripe for investigation:
+The most glaring omission in our algorithm is its disregard for types.
+By ignoring types, not only are we losing valuable information about expressions, but we also lose the ability to distinguish between constructors.
+This is because a constructor, like \hs{True} or \hs{Just}, has no internal structure; it is just a token.
+The distinguishing features of constructors are their types, which not only tell us which data type they construct, but also their arity, the types of their arguments, etc.
 
-\begin{itemize}
-  \item Improvements to our feature extraction algorithm, in particular to handle types.
-  \item Alternative clustering algorithms.
-  \item \emph{Feature learning} uses machine learning algorithms in place of hand-coded feature extraction algorithms such as ours.
-A comparison of our hand-picked features against a selection of learned representations would be a useful indication of the importance that understanding the expression language may or may not have on identifying salient aspects of expressions.
-  \item There are several potential alternatives to \qcheck{} for instantiating variables, which may improve performance of \qspec{} by either generating expressions more efficiently (as in \textsc{SmallCheck}'s enumeration approach) or by observing differences with fewer tests (as found in \textsc{Lazy SmallCheck}'s demand-driven enumeration).
-\end{itemize}
+Our algorithm closely follows that of ML4PG, which \emph{does} support types.
+This is handled by populating matrix cells with tokens \emph{and} their types.
+Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).
 
-\subsection{Interestingness}
+One partial solution would be leave most Core expressions without types, but to include them for non-local identifiers (i.e.
+globals and constructors), which we can look up in a database.
+In fact, our ML4HS framework already includes such type information in its database, alongside the Core syntax trees.
+Integrating this information into our algorithm is the next logical step.
+
+We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \citep{bengio2013representation}.
+This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.
+
+With more expressive features, it may also be useful to experiment with more powerful learning algorithms.
+An interesting possibility is to add a feedback loop between the theory exploration phase and the clustering phase, to more directly base the similarity of expressions on whether they (are predicted to) occur together in equations.
+
+\subsection{Theory Exploration Extensions}
+
+Our current approach is a rather conservative change to the existing theory exploration approaches, as it is essentially a wrapper around \qspec{}.
+There is potential for more radical changes to be made, which alter the search process itself.
+
+\subsubsection{Variable Instantiation}
+
+\qcheck{} is certainly the most popular property checker for Haskell, which motivates its use in \qspec{} to instantiate variables to random values.
+However, this task of finding type inhabitants has also been solved in many other ways, which may be worth investigating in place of \qcheck{} (or perhaps even as part of an ensemble).
+
+The \textsc{SmallCheck} system \citep{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly.
+Whilst this does not make \textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage is that it may use much less memory, as the generated values are built up incrementally.
+In contrast, \qcheck{} may generate very large values; in particular, generating tree structures na\"{\i}vely can cause them to grow exponentially.
+For example, here is a potential generator for \hs{RoseTree}s:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+genRoseTree = do f        <- arbitrary
+                 subtrees <- listOf genRoseTree
+                 return (Node f subtrees)
+\end{lstlisting}
+
+The \hs{listOf genRoseTree} call will return a list of arbitrary length, where each element is generated by \hs{genRoseTree}.
+This allows an arbitrary number of recursive calls to be made for each invocation of \hs{genRoseTree}, which will quickly exhaust the resources of any machine.
+Whilst such problems may be anticipated, or quickly spotted, in a property checking setting, this can be more difficult for our automated approach.
+For example, if a type does not have a generator available, we cannot use a library like \hs{derive} to create one automatically, as it suffers from this na\"{\i}vity problem.
+
+A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \citep{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded.
+This may narrow down our search procedures greatly, especially when predicates are involved.
+\qcheck{} allows predicates to restrict the values it tests with, and hence allows \emph{conditional} equations to be discovered.
+However, its implementation uses a simple rejection sampling technique: values are generated just as if the predicate were not there, and afterwards are filtered to reject any which do not satisfy the predicate.
+This makes it difficult to use very specific predicates, as it is unlikely that many of our random samples will exactly match our criteria.
+On the other hand, \textsc{Lazy SmallCheck} will focus its search on exactly those parts of the datastructure which are checked by the predicate, as those are the parts being forced to evaluate.
+This makes it much more likely that we will find values which satisfy the predicate, allowing us to effectively explore more specific conditional properties.
+
+Other approaches to generating inhabitants include \textsc{Djinn} \citep{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator).
+\textsc{MuCheck} \citep{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g.
+changing the order of pattern match clauses); whilst not as exhaustive as the other approaches, mutating existing values in this way is claimed to yield values which correspond more closely to what a programmer might write.
+This is an interesting possibility for focusing theory exploration on to more ``realistic'' areas of the search space, and hence avoiding some of the more useless or bizarre expressions that random search and enumeration may produce.
+
+In fact, the database generated by our \textsc{AstPlugin} may prove helpful in generating values, since its type information can be fed to a tool like \textsc{Djinn} to discover chains of function applications for building values, which would be particularly useful in cases where constructors are private, like in our email example.
+This is similar to the \textsc{Hoogle} tool, but also offers the ability to use dependency information to avoid potentially infinite recursion.
+
+The Core syntax trees in our database could also be used to generate theories for automated theorem provers.
+\hspec{} currently uses the GHC API to transform Core within its own process, however that approach suffers from the problems described in \S \ref{sec:astplugin}.
+
+\subsubsection{Interestingness}
 \label{sec:interestingness}
 
-It is important to consider the question: \emph{what if we succeed?} What if a more efficient theory exploration system were possible, capable of reading huge amounts of code and producing an abundance of theorems?
-How could the output be made manageable, by finding the needles we are interested in among the haystack of potential laws?
+If we do succeed in producing a fast theory exploration system, which chooses productive combinations of terms and finds a large number of properties, we encounter the problem of managing the output: finding the needles we are interested in among the haystack of trivialities and coincidences.
 
@@
 \qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is very simple: we discard equations which are direct consequences of others, and keep all the rest.
+Different, and more sophisticated notions of interestingness have been widely studied in other fields, which may be applied in the context of theory exploration.
 
-\subsubsection{Interestingness in Concept Formation}
-\label{sec:conceptformation}
+\paragraph{Concept Formation} \label{sec:conceptformation} \leavevmode \newline
 
-% TODO:
-\citep{Montano-Rivas.McCasland.Dixon.ea:2012}
-\citep{Piantadosi.Tenenbaum.Goodman:2012}
-\citep{Wille:2005}
-\citep{colton1999automatic}
-\citep{colton2000agent}
-\citep{colton2012automated}
-\citep{lenat1977automated}
-\citep{mullerunderstanding}
-\citep{Bundy.Cavallo.Dixon.ea:2015}
-\citep{johansson2009isacosy}
-\citep{spector2008genetic}
-\citep{colton2012automated}
- \citep{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
+One directly applicable area to consider is \emph{concept formation}, which considers the (automatic) generation of new definitions and axioms.
+Unlike theory exploration, such systems are not constrained by the requirement that their output be provable, and hence interestingness is an important way to judge the quality of the results.
 
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+Approaches vary, from those which are directly related to theory exploration (such as the scheme-instantiating approach of \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory exploration system), to others which are more closely related to theories of human learning and discovery \citep{Piantadosi.Tenenbaum.Goodman:2012, mullerunderstanding}.
+Those based on finding patterns in data, such as \citep{Wille:2005}, may be useful in tandem with our expression database, and the results of value generators like \qcheck{}.
 
-% Alan Bundy et al
+Since tools like \hspec{} already call out to third-party automated theorem provers, and indeed our own ML4HS system uses the external Weka program, there may also be merit in using external concept formation or conjecture generation tools (or reimplementations of their ideas), in order to build up more structure on top of that provided by the code we are exploring.
+For example, the approaches taken by AM \citep{lenat1977automated, lenat1979automated}, Graffiti \citep{delavina2005some, delavina2005graffiti} and HR \citep{colton1999automatic, colton2000agent} could be used alongside those of \qspec{}.
 
-% Eurisko, AM, etc.?
-
-\subsubsection{Artificial Curiosity}
-\label{sec:curiosity}
+\paragraph{Artificial Curiosity} \label{sec:curiosity} \leavevmode \newline
 
@@
 \emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
+Although framed in the context of \emph{reinforcement learning}, this reliance on interest is clearly relevant to our theory exploration setting.
 
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
+As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, in a similar way to our recurrent clustering algorithm.
 The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
 The resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
+This same behaviour may be applicable to the theorems produced by a theory exploration system.
 
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
+We can divide AC approaches into two groups: those which make \emph{explicit} use of interestingness, learning from signals which follow a Wundt curve; whilst \emph{implicit} approaches modify the \emph{output} of their learning algorithm(s), to engineer the overall system behaviour to follow a Wundt curve as an emergent property.
 
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning.
+A framework encompassing many examples of the explicit approach is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning; for comparison, many similar measures are surveyed in a data mining context in \citep{geng2006interestingness}.
+Many more reinforcement learning examples can be found in \citep{Kaplan2006, Lipson2007, Luciw2011, Macedo2000, Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991, oudeyer2004intelligent}; whilst more general descriptions are given in \citep{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be more amenable to our theory exploration setting.
 
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
+Many of these reward signals are based on information theory, with a prominent example being \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
+Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new insights into the structure of past experience can provide a space saving when compressed together.
+This seems particularly relevant for our problem of identifying interesting theorems: those new theorems (``observations'') which shorten the proofs of previously discovered theorems may be more general, more powerful and therefore more \emph{interesting} and hence worth keeping.
+In fact this is very similar to \qspec{}'s interestingness criterion.
 
-% TODO: Examples
-\citep{Schmidhuber1999}
+Another example of explicit artificial curiosity is given in \citep{Hester.Stone:2012}, where world states which cause \emph{disagreement} among a population of decision trees (a \emph{random forest} \citep{randomforests}) are considered interesting.
+Since the models make stochastic predictions, the disagreement follows a Wundt curve as the complexity of state transitions increases: for parts of the state space which have been fully learned, the models will agree on accurate predictions; for parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the same average value.
+Whilst the latter predictions may not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the interestingness of states which are too complex.
 
-Two sources of intrinsic reward are proposed in \citep{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \citep{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
+Many examples of the implicit case are based on \emph{coevolution}: rewarding one part of the system for exploiting another part, and vice versa.
+In \citep{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on the outcome of actions, and the winner is rewarded at the expense of the loser.
+Due to the risk involved, each algorithm will only bet when it is confident in its prediction, and bets will only be actioned when each algorithm is confident in a \emph{different} outcome.
+The overall behaviour of this system is therefore similar to the explicit measure of disagreement used in the random forest example.
+In terms of theory exploration, such a scheme could be used to find theorems which are \emph{non-obvious}, and hence informative in some way to the user.
 
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
+Another case is the ``darwinian brain'' of Fernando et al.
+\citep{fernando2013design1, fernando2013design2}.
+This coevolves a population of problem generators and problem solvers, rewarding the solvers based on their speed, and rewarding the generators based on the \emph{variance} of the solvers' speed.
+This avoids trivial problems (which all solvers can quickly overcome) and complex problems (which no solver can manage), and focuses on those with the most possibility for learning.
+It is easy to imagine such a general architecture being populated by conjecture generators and theorem provers to form a theory exploration system.
 
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
+\paragraph{Evolutionary Computation} \leavevmode \newline
 
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\citep{Kaplan2006}
-\citep{Lipson2007}
-\citep{Luciw2011}
-\citep{Macedo2000}
-\citep{Ramik.Sabourin.Madani:2013}
-\citep{Roa.Kruijff.Jacobsson:2009}
-\citep{Schaul.Sun.Wierstra.ea:2011}
-\citep{Schmidhuber1999}
-\citep{Schmidhuber:1991}
-\citep{Scott1989}
-\citep{Steunebrink.Koutnik.Thorisson.ea:2013}
-\citep{maher2008achieving}
-\citep{meyer1991possibility}
-\citep{oudeyer2004intelligent}
-\citep{oudeyer2014evolution}
-\citep{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \citep{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsubsection{Interestingness in Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
+Coevolution is a form of \emph{evolutionary computation}; an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
 Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
+Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting (there are also precedents for using evolutionary computation in a theorem proving domain \citep{spector2008genetic}).
 
@@
 Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\iffalse
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-\fi

### 2015-12-15 TransferReport/implementation.tex
Added Gantt chart and a little extra to evaluation

@@
 \label{sec:implementation}
-
-\iffalse TODO:
-  A: Conceptually: - implementation follows the inductive/recursive structure of the previous definitions
-e.g. definition of the rose tree is naturally represented as Haskell datatype definition
-
-B Module-by-module road map
-  - Treatment of Haskell package system
-
-C - External software
-\fi
 
@@
 
-\subsection{\textsc{ASTPlugin}}
+\subsection{\textsc{AST Plugin}}
 \label{sec:astplugin}
@@
 The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core.
-It also includes a \emph{renaming} transformation which makes all names unique (either by prefixing them with the name of the module which defines them, or by suffixing the name with a number.
+It also includes a \emph{renaming} transformation, which resolves global identifiers into a canonical form.
 This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
@@
 For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core.
+To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core, which is guaranteed not to conflict since this name would be invalid in Haskell.
 This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
@@
 
-The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}.
+The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \citep{mitchell2008hoogle}.
 Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which allows dependencies to be discovered.
+The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which makes it trivial to look up references in the database.
 Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
@@
 
-As described in \S \ref{} We have a tool which annotates each definition with its dependencies, i.e.
-those (package, module, name) combinations which it references.
-Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.
+As described in \S \ref{sec:symbolstofeatures}, we must topologically sort the output of \textsc{AstPlugin} in order for our recurrent clustering to be well-founded.
+Since our database keys (containing the package, module and binding names, as described above) match our representation of non-local references, it is simple to walk each syntax tree to obtain the set of references it makes.
+In addition, the resulting set of (identifier, list-of-referenced-identifiers) pairs exactly matches the (vertex, list-of-successor-vertices) format used to represent directed graphs by the popular \hs{containers} library which ships with GHC.
+This provides an implementation of topological sort for strongly connected components, which we use as-is.
+A simple shell script loops through these SCCs, invoking the recurrent clustering component for each and appending the resulting features and clusters to the database.
 
-\subsection{Recurrent Clustering}
+\subsection{Feature Extraction}
 
-\iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi
-
-Our implementation is a rather direct translation of the algorithm described in \S \ref{sec:contributions} into Haskell.
-At the top level, we send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive:
-
-\begin{haskell}
-recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
-recurrentCluster = go ([],[])
-  where go (fs, db) []       = db
-        go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
-                                   db' = kMeans fs'
-                                in go (fs', db') ess
-        extract db (i, e) = (i, rt db e)
-\end{haskell}
-
-The overall result of the \hs{recurrentCluster} function is a list of clusters, containing the IDs of their elements.
-These are obtained by interleaving clustering (the \hs{kMeans} function) and feature extraction (the \hs{rt} function).
-
-The feature extraction itself contains two parts: first, syntax trees matching the grammar in Figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:
+The implementation of our feature extraction algorithm is a rather direct translation of the description given in \S \ref{sec:contributions} into Haskell.
+We parse the s-expressions generated by \textsc{AstPlugin} into algebraic data types which correspond directly to the definitions in Figure \ref{fig:coresyntax}; this is routine, so we omit the details for brevity.
+Similarly, we can represent rose trees with a datatype corresponding to the definition given in \S \ref{sec:expressionstovectors}:
 
@@
 data RoseTree = Node Feature [RoseTree]
-
-eRT :: [Local] -> [[Global]] -> Expr -> RoseTree
-eRT env db e = case e of
-  ...
 \end{haskell}
 
-The easiest branches to handle are literals and types, which we represent using particular feature values (\hs{sLITNUM}, \hs{sLITSTR} and \hs{sTYPE}):
+For simplicity we use the representation \hs{Feature = Int}, as we do not have fractional values.
+Since we represent the symbols $expr$, $id$, etc.
+from Figure \ref{fig:coresyntax} using different datatypes, we cannot write one big definition of $toTree$ or $\phi$ which works on all tokens.
+Each case shown in Figure \ref{fig:totree} and equations \ref{eq:feature}, \ref{eq:localfeature} and \ref{eq:globalfeature} appears in the implementation, although they are spread across several functions.
+
+To support looking up local identifiers, our implementation of $toTree$ takes a context as argument, extending it as required.
+As an example of the complexity this adds, here is the \CCase\ branch of $toTree$:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=0pt, xrightmargin=0pt]
+toTree :: Context -> Expr -> RoseTree
+toTree ctx x = case x of
+  ...
+  Case e l as -> Node fCase (toTree ctx e : map (toTreeAlt (l:ctx)) as)
+  ...
+\end{lstlisting}
+
+Breaking this down we can see \hs{fCase} representing the value of $\feature{\CCase}$, and a list of sub-trees defined in parentheses.
+The first subtree is a straightforward recursive call in an unmodified context: \mbox{\hs{toTree ctx e}}.
+The rest of the list is formed by applying the function \mbox{\hs{toTreeAlt (l:ctx)}} to each element of the list \hs{as} of \CAlt\ clauses.
+
+The \hs{toTreeAlt} function contains those cases of $toTree$ which handle symbols in $alt$.
+We prepend the identifier \hs{l} to the context, to get the extended context \hs{l:ctx}.
+This is because \hs{l} will be bound the value of \hs{e}, in order to avoid re-computing its value several times.
+
+The other clauses are handled in a similar way.
+The trickiest is the \CLet\ clause, since the local identifiers aren't directly available; we must extract them from their \CRec, \CNonRec\ and \CBind\ wrappers first, which we do using helper functions.
+
+As shown above, the values from equations \ref{eq:feature} are encoded directly in $toTree$.
+For $phi(l \in \mathcal{L})$ we use standard Haskell functions to look up the required indices in the context:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+phiL :: Context -> Local -> Feature
+phiL ctx x = case elemIndex x ctx of
+  Nothing -> error (concat ["Local '", show x,
+                            "' not in context '",
+                            show ctx, "'"])
+  Just i  -> (2 * alpha) + i
+\end{lstlisting}
+
+As explained in \S \ref{sec:symbolstofeatures}, local identifiers should always exist in the context. If this precondition doesn't hold, we abort the program with an error rather than continuing.
+
+Global identifiers are kept as-is until we have access to the clusters from the last iteration. This takes place outside Haskell, using the \hs{jq} data processing tool.
+
+Our implementation of $level$ exactly matches equation \ref{eq:level}:
 
 \begin{haskell}
-  Lit (LitNum _) -> Node sLITNUM []
-  Lit (LitStr _) -> Node sLITSTR []
-  Type           -> Node sTYPE   []
+level :: Int -> RoseTree -> [[Feature]]
+level 1 (Node f _)  = [f]
+level n (Node _ ts) = concatMap (level (n-1)) ts
 \end{haskell}
 
-Function application simply recurses into both sub-expressions:
+To produce feature vectors, we do not directly construct the matrix; instead we generate the rows and concatenate them together in one step, using the \hs{concatMap} function:
 
-\begin{haskell}
-  App e1 e2 -> Node sAPP [eRT env db e1,
-                          eRT env db e2]
-\end{haskell}
+\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+featureVec :: Expr -> [Feature]
+featureVec e = concatMap (\m -> pad (level m tree)) [1..r]
+  where tree   = toTree [] e
+        pad xs = take c (xs ++ repeat 0)
+\end{lstlisting}
 
-To look up variables locally and globally, we use the \hs{lookupL} and \hs{lookupG} functions, respectively.
-These return the index containing their argument, if found.
-In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors.
-For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features:
+By providing \hs{featureVec} with the latest set of clusters, read from the \textsc{AstPlugin} database, we turn Core expressions into feature vectors, which are appended to the database.
 
-\begin{haskell}
-  Var (Global i) -> Node (lookupG db  i) []
-  Var (Local  i) -> Node (lookupL env i) []
-\end{haskell}
-
-Finally, when we traverse binders we must extend the environment \hs{env}:
-
-\begin{haskell}
-  Lam  i  e     -> Node sLAM [eRT (i:env) db e]
-  Let  bs e     -> Node sLET (map (bRT (ids bs:env) db) bs ++
-                                   eRT (ids bs:env) db  e
-  Case e i alts -> Node sCASE (eRT    env  db  e :
-                          map (aRT (i:env) db) alts)
-\end{haskell}
-
-Patterns and bindings are handled in a similar way:
-
-\begin{haskell}
-aRT :: [Local] -> [[Global]] -> Alt -> RoseTree
-aRT env db alt = case alt of
-  (DataAlt _, vs, e) -> eRT (vs ++ env) db e
-  (LitAlt  _, _,  e) -> eRT env db e
-  (Default,   _,  e) -> eRT env db e
-
-bRT :: [Local] -> [[Global]] -> Bind -> RoseTree
-bRT env db b = case b of
-  NonRec i e -> eRT env db e
-  Rec es     -> Node sREC (map (eRT env db . snd) es)
-\end{haskell}
-
-The result of \hs{eRT} is a \hs{RoseTree} whose branching structure mimics that of our original expression.
-We next need to convert this to a \emph{matrix} of features, which we do by converting each level of the \hs{RoseTree} into a row of the matrix (using \hs{pad} to ensure a consistent size).
-Finally we turn the matrix into a \emph{feature vector} by concatenating the rows:
-
-\begin{haskell}
-level :: Int -> RoseTree -> [Feature]
-level 0 (Node x xs) = [x]
-level n (Node x xs) = concatMap (level (n-1)) xs
-
-rt :: [[Global]] -> Expr -> [Feature]
-rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
-\end{haskell}
-
-Notice that this algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30).
-The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
-
-\iffalse TODO: How does this section differ from Algorithm in Contributions? \fi
-\iffalse TODO: Make this a concrete presentation of the abstract ideas from the Algorithm section, e.g.
-this is how we get the dependencies (AST plugin, etc.); this is a case-by-case of how we perform feature extraction; this is how we perform k-means (Weka), etc.
-\fi
-
-Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.
-
-\subsection{\textsc{ML4HS}}
-\label{sec:ml4hs}
-
-\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin} and \textsc{MLSpec}, described below, with an implementation of our recurrent clustering algorithm and the \textsc{Weka} machine learning library.
+We use the Weka system to perform our k-means clustering, as it is widely used, including by ML4PG.
+We select all feature vectors from our database, and write them in CSV format for Weka to process.
+The Weka CLI command is invoked, which appends a cluster number to each of these feature vectors; we read these off and append them to the database.
+As long as more SCCs remain unprocessed, we keep looping this process, using the database to communicate between the feature extraction and clustering phases.
 
@@
 
-When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks.
-One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.
-
-Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec{}, based on the databases constructed by \textsc{AstPlugin} described above.
-Features of \textsc{MLSpec} include:
+We cannot supply these clusters as-is to \qspec{}, since it must be provided with a \emph{signature}.
+These are constructed by our \textsc{MLSpec} tool, using information from the \textsc{AstPlugin} database.
+Tasks performed by \textsc{MLSpec} include:
 
@@
 [t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \qspec{} is unable to evaluate expressions such as \hs{safeHead [] :: forall t.
-Maybe t} without instantiating the variable \hs{t} to a specific type.
+[t]}, a testing-based system like \qspec{} is unable to evaluate these expressions without instantiating the variable \hs{t} to a specific type.
 Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
@@
   \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.
+There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules.
+In such cases the exploration process simply aborts.
 
@@
 
-  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.
+  \item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is that all required packages and modules must be provided up-front, usually by specification in a Cabal file.
+Since \textsc{MLSpec} builds signatures \emph{dynamically}, depending on the cluster information it is given, we do not know what packages it may need.
+To work around this problem, \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have built called \texttt{nix-eval}.
+This provides an \texttt{eval} function, like those commonly found in dynamic languages such as Python and Javascript, for evaluating Haskell expressions.
+The key feature of \texttt{nix-eval} is that these Haskell expressions may reference packages that are not installed on the system.
+When such expressions are evaluated, these packages will be automatically downloaded and installed into a sandbox using the Nix package manager, and GHC will be invoked in this sandbox to perform the evaluation.
 
 \end{itemize}
-
-\subsection{\texttt{nix-eval}}
-\label{sec:nixeval}
-
-To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation.
-This is difficult, as many terms are functions, which Haskell does not let us inspect.
-This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves.
-To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.
-
-\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically.
-In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.
-
-On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier.
-Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system.
-Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.
-
-We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system.
-These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.
-
-\texttt{nix-eval} is build on top of the Nix package manager and GHC.
-Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.
-
-The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code.
-When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression.
-GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression.
-The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output.
-The expression, therefore, must have type \hs{String}; although this is not enforced statically.
-
-The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above.
-The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way.
-The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}.
-It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g.
-using JSON.
-
-\subsection{Circular Convolution}
-\label{sec:circularconvolution}
-
-We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Tree = Leaf FV
-          | Branch FV Tree Tree
-\end{lstlisting}
-
-Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate.
-To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceD 1 (Branch fv l r) = Leaf   fv
-reduceD n (Leaf   fv)     = Leaf   fv
-reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
-                                      (reduceD (n-1) r)
-\end{lstlisting}
-
-This is simple, but suffers two problems:
-
-\begin{itemize}
- \item We must choose a conservatively large \hs{d}, since we're throwing away information
- \item The number of feature vectors grows exponentially as \hs{d} increases
-\end{itemize}
-
-The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}.
-One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.
-
-\iffalse
-
-We can do this with \hs{tabulate w d t}:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-tabulate w d t = take d (map (take w) (tabulate' t))
-
-tabulate' (Leaf   fv)     = [fv] : []
-tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)
-
-merge    []     ys  = ys
-merge    xs     []  = xs
-merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
-\end{lstlisting}
-
-\fi
-
-Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth.
-Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function.
-Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-reduceC (Leaf fv)       = fv
-reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
-\end{lstlisting}
-
-Circular convolution has the following desirable properties:
-
-\begin{description}
-
-  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}
-
-  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.
-
-\end{description}
-
-Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing.
-This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information.
-In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.
-
-Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.
-
-\subsubsection{Extracting Features from XML}
-
-This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}.
-The application parses arbitrary trees of XML and uses binary features, ie.
-feature vectors are bit vectors.
-
-To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:
-
-\begin{equation}
-feature(n) = 2^{\md5(s) \pmod{L}}
-\end{equation}
-
-Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument.
-The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$.
-Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.
-
-In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.
-
-\subsubsection{Application to Coq}
-
-Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning.
-We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.
-
-Coq will also interpret the standard output of the command, to generate terms and tactics.
-This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.

### 2015-12-15 TransferReport/intro.tex
Recurrent clustering section
Added Gantt chart and a little extra to evaluation

@@
 The \emph{functional programming} paradigm has been proposed for addressing these issues \citep{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis.
-\iffalse TODO: Types?
-\fi For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
+For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
 Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code.
-Similarly, pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results.
+Similarly, the results of pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results.
+By making state implicit in this way, powerful type systems can be used to constrain the behaviour of programs, and to give a rich, composable structure to data.
 
@@
 \begin{enumerate}
-  \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of large signatures, which are more tractable to explore.
+  \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of Haskell libraries, which are more tractable to explore.
   \item A novel feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf learning algorithms.
@@
 We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}).
-We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details \S \ref{sec:implementation}.
-A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future}.
+We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details in \S \ref{sec:implementation}.
+A variety of related work is surveyed in \S \ref{sec:related}, we briefly evaluate our implementations in \S \ref{sec:evaluation} and give several potential directions for future research in \S \ref{sec:future}.

### 2015-12-15 TransferReport/related.tex
Added Gantt chart and a little extra to evaluation

@@
 
+\subsection{Theory Exploration}
+
+We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements.
+The idea was first introduced in the \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
+This provided an interactive environment, similar to computer algebra systems and interactive theorem provers.
+In this setting, many of our concerns such as the generation of values and deciding which properties to explore are simply delegated to the user; the software would check for correctness, store results and perform searches; again, similar to interactive theorem provers.
+
+Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery.
+By removing user interaction, these concerns about directing search must be solved by algorithms.
+As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \citep{Montano-Rivas.McCasland.Dixon.ea:2012, johansson2009isacosy, Hipster}.
+
+We have focused our attention on \qspec{}, although it does not actually \emph{prove} its results, and hence may not be considered a theory exploration system on its own.
+However, it does form a vital component of \hspec{}, which uses off-the-shelf automated theorem provers (ATPs) to verify \qspec{}'s conjectures, forming a complete theory exploration system as well as a capable inductive theorem prover (by exploiting theory exploration for lemma generation) \cite{claessen2013automating}.
+Due to \hspec{}'s use in \textsc{Hipster}, improvements to \qspec{} also benefit work being pursued in Isabelle.
+
 \subsection{Relevance Filtering}
@@
 Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
+To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a na\"{\i}ve Bayes algorithm.
 
@@
 The \emph{k-means++} approach \citep{arthur2007k, bahmani2012scalable} can be used to more carefully select the ``seed'' values for the first timestep, and the \emph{x-means} algorithm \citep{pelleg2000x} is able to estimate how many clusters to use (our \emph{final} clusters should be tuned to maximise the performance of the subsequent theory exploration step, but x-means could still be useful in the recurrent clustering steps).
+
+\iffalse
+TODO: Theory Exploration?
+\citep{johansson2009isacosy}
+\fi

### 2015-12-15 TransferReport/haskell_example/src/Example.hs
Tail-recursive clustering example

@@ -19,6 +19,7 @@ mult_tail' x y acc = case x of
 mult_tail :: Nat -> Nat -> Nat
 mult_tail x y = mult_tail' x y []
 -}
+{-
 factorial1 0 = 1
 factorial1 n = n * factorial1 (n-1)
 
@@ -48,3 +49,4 @@ f2 :: Int -> Int
 f2 = factorial2
 
 foo n = -n
+-}

### 2015-12-15 TransferReport/haskell_example/src/QuickSpecExample.hs

@@ -2,7 +2,7 @@ module QuickSpecExample where
 
 import Numeric.Natural
 import Test.QuickSpec
-
+{-
 i2N :: Int -> Natural
 i2N = fromInteger . toInteger
 
@@ -37,3 +37,4 @@ nat = [fun0 "Z"     (0    :: Natural),
        fun1 "even"  (even :: Natural -> Bool),
 
        vars ["x", "y", "z"] (0 :: Natural)]
+-}

### 2016-01-26 CICM2016/contributions.tex
Copied TransferReport to CICM2016

--- 2015-12-10 TransferReport/contributions.tex
@@
 
-\iffalse TODO: Focus the section more on the problem of identifiers, and how recurrent clustering uses clustering as part of feature extraction to solve this \fi
-
-We adapt the methodology of \emph{recurrent clustering} proposed in \citep{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell \iffalse TODO: Fix the end of this sentence \fi theory exploration.
+We adapt the methodology of \emph{recurrent clustering} proposed in \citep{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
-Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as algorithm \ref{alg:recurrent}), which goes beyond a simple syntactic characterisation, to allow the features of an expression to depend on those it references.
+Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as Algorithm \ref{alg:recurrent}), which goes beyond a simple syntactic characterisation, to allow the features of an expression to depend on those it references.
 Here we describe our approach to recurrent clustering and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
-We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat feature vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the notation $\feature{}$.
+We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat feature vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote as the function $\phi$.
 
 \subsubsection{Expressions to Vectors}
+\label{sec:expressionstovectors}
 
@@
 
-If we traverse these expressions in breadth-first order, converting each token to a feature using $\feature{}$ and padding to the same length with $0$, we would get the following feature vectors:
+If we traverse these expressions in breadth-first order, converting each token to a feature using $\phi$ and padding to the same length with $0$, we would get the following feature vectors:
 
@@
 
-Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\feature{}$).
+Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\phi$).
 These feature vectors might be deemed very dissimilar during clustering, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
 
-If we were to align these feature optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
+If we were to align these features optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
 
@@
 To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in Figure \ref{fig:totree}.
-\iffalse TODO: Make a formal definition in an environment, i.e.
-not inline?
-\fi Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+We follow the presentation in \citep{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
 $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
@@
       (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1 a_1 \dots$} \\
+      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
       (\feature{\CType})                                                & \text{if $e = \CType$} \\
-      (\feature{\CLocal},   leaf(l_1))                                  & \text{if $e = \CLocal\ l_1$} \\
-      (\feature{\CGlobal},  leaf(g_1))                                  & \text{if $e = \CGlobal\ g_1$} \\
+      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
+      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
       (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
@@
       (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1 e_2 l_1 \dots$}  \\
+      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
       (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
@@
       (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
-      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1 \dots$} \\
-      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1 e_1$}
-    \end{cases} \\
-    leaf(x) &= (\feature{x}) \iffalse TODO: This is completely unnecessary... \fi
+      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
+      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
+    \end{cases}
   \end{align*}
@@
 We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
-We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\feature{}$, and we conjecture that the effect on clustering real code is low.}
+We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.}
   \label{fig:totree}
@@
 These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting, and concatenating them together (written as $\cat$):
-\iffalse TODO: Maybe make this a bit more of a formal definition? \fi
-\begin{equation*}
+
+\begin{equation} \label{eq:level}
   level(l, (f, T_1, \dots, T_{n_T})) =
@@
     \end{cases}
-\end{equation*}
+\end{equation}
 
@@
 \begin{sloppypar}
-Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_{1,1} \cat \mathbf{M}_{1,2} \cat \dots \cat \mathbf{M}_{2,1} \cat \mathbf{M}_{2,2} \cat \dots$, as shown in Figure \ref{fig:vectorexample}:
+Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_1 \cat \mathbf{M}_2 \cat \dots \cat \mathbf{M}_r$, as shown in Figure \ref{fig:vectorexample}:
 \end{sloppypar}
@@
 \subsubsection{Symbols to Features}
+\label{sec:symbolstofeatures}
 
-\iffalse TODO: quine quotes look so much like ceiling. Maybe find a name for it and define it formally in an environment? \fi
-
-We now define the algorithm for $\feature{}$, which turns terminal symbols of Core syntax into features (real numbers).
+We now define the function $\phi$, which turns terminal symbols of Core syntax into features (real numbers).
 For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \citep{DBLP:journals/corr/HerasK14} in Coq.
@@
 
-\begin{align*}
+\begin{equation} \label{eq:feature}
+  \begin{aligned}
     \feature{\CAlt}          &= \alpha      &
@@
     \feature{\CLitStr}       &= \alpha + 18
-\end{align*}
+  \end{aligned}
+\end{equation}
 
-To encode \emph{local} identifiers $\mathcal{L}$ we use their \emph{de Bruijn index}, since this is a numeric value suitable for use as a feature, and it gives equal values for $\alpha$-equivalent expressions.
-\iffalse TODO: needs to be defined within the text to make formal sense.
-Maybe don't mention it here, but just say, once it's defined, that it's known as the de Bruijn index?
-\fi To calculate these indices the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from Figure \ref{fig:totree} for clarity).
+To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which gives equal values for $\alpha$-equivalent expressions (i.e.
+renaming an identifier shouldn't affect the feature).
+To do this, the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from Figure \ref{fig:totree} for clarity).
 The context is a list of the local identifiers which are in scope, prepended as they are encountered.
@@
 
-For example, when calculating $toTree(\CLam\ i\ e)$ in context $x$, we make the recursive call $toTree(e)$ in the context of \emph{$x$ prepended with $i$} (i.e.
-$(i) \cat x$).
+For example, when calculating $toTree(\CLam\ i\ e)$ in context $\vect{x}$, we make the recursive call $toTree(e)$ in the context of \emph{$\vect{x}$ prepended with $i$} (i.e.
+$(i) \cat \vect{x}$).
 Similar extensions of the context are performed in the cases of \CBind, \CCase\footnote{The $\mathcal{L}$ value in a \CCase\ expression is bound to the expression being matched against; a technical detail to preserve sharing.}, \CAlt\ (which may introduce an arbitrary number of local identifiers) and \CLet\ (where identifiers are taken from occurrences of \CBind\ in the first expression).
@@
 Hence if we apply $toTree$ to (the Core expression corresponding to) such declarations we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$.
-\iffalse NOTE: Can we prove this, or would it just be an exercise in formalising Haskell?
-\fi In which case we define:
+In which case we define:
 
-\begin{equation}
+\begin{equation} \label{eq:localfeature}
   \feature{l} = i + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
@@
 
-Where $i$ is the index of the first occurrence of $l$ in the context; this is the de Bruijn index of $l$. We again use $\alpha$ to separate these features from those of other constructs.
+Where $i$ is the index of the first occurrence of $l$ in the context.
+This tells us how many binders a variable is nested inside, relative to the point it was introduced.
+This is invariant under renaming, as we wanted, and is known as the \emph{de Bruijn index} of $l$.
+We again use $\alpha$ to separate these features from those of other constructs.
 
@@
 
-\iffalse TODO: this is a novel feature compared to ML4PG, ACL2(ml), etc., right? Maybe... Emphasize either the connection to earlier work, or the novelty. \fi
-Instead, we encode global identifiers \emph{indirectly}, by looking up the expressions which they \emph{reference}.
-For $g \in \mathcal{G}$ we define $\feature{g}$ as the \emph{index of the cluster which $g$ appears in}.
-This is where the recurrent nature of the algorithm appears, since we perform k-means clustering \emph{during} feature extraction, and that clustering step, in turn, requires that we perform feature extraction.
+Instead, we use the approach taken in the latest versions of ML4PG and encode global identifiers \emph{indirectly}, by looking up the expressions which they \emph{reference}:
 
-For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference).
-To do this, we construct a directed graph, where each declaration is a node and edges denote an ``is referenced by'' relation.
-An example dependency graph is given in Figure \ref{fig:dependencyexample}.
+\begin{equation} \label{eq:globalfeature}
+  \feature{g \in \mathcal{G}} =
+    \begin{cases}
+      i + 3 \alpha \quad & \text{if $g \in C_i$} \\
+      f_{recursion}         & \text{otherwise}
+    \end{cases}
+\end{equation}
+
+Where $\vect{C}$ are our clusters (in some arbitrary order).
+This is where the recurrent nature of the algorithm appears: to determine $\vect{C}$ we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.
 
@@
 
-    \caption{Dependency graph for Figure \ref{fig:haskellexample}.
+    \caption{Dependency graph for Figure \ref{fig:coreexample}.
 Loops indicate recursive functions, double arrows indicate mutual recursion.
@@
   \end{subfigure}
-
   \begin{subfigure}{\textwidth}
@@
       };
-      \node [block, right of=plusset] (multset) {
-        \begin{tikzpicture}[node distance = 2cm]
-          \node [block] (mult) {\hs{mult}};
-        \end{tikzpicture}
-      };
-      \node [block, right of=multset] (parityset) {
+      \node [block, right of=plusset] (parityset) {
         \begin{tikzpicture}[node distance = 1cm]
@@
       };
+      \node [block, right of=parityset] (multset) {
+        \begin{tikzpicture}[node distance = 2cm]
+          \node [block] (mult) {\hs{mult}};
+        \end{tikzpicture}
+      };
 
-      \path [line] (plusset)     -- (multset);
-      \path [line] (multset)     -- (parityset);
+      \path [line] (plusset)   -- (parityset);
+      \path [line] (parityset) -- (multset);
     \end{tikzpicture}
     \caption{One possible topological sorting of simply connected components for Figure \ref{fig:dagexample} (ignoring constructors).
-\hs{even} and \hs{odd} are mutually recursive, no neither can appear before the other.}
+\hs{even} and \hs{odd} are mutually recursive, neither can appear before the other, so they are grouped into one component and handled concurrently by Algorithm \ref{alg:recurrent}.}
     \label{fig:sccexample}
   \end{subfigure}
+  \caption{Sorting functions from Figure \ref{fig:coreexample} into dependency order.}
   \label{fig:dependencyexample}
-  \caption{Sorting functions from Figure \ref{fig:haskellexample} into dependency order.}
 \end{figure}
+
+For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference).
+In this way, we can avoid looking up expressions which haven't been clustered yet.
+To perform our sort, we construct a directed graph, where each declaration is a node and edges denote an ``is referenced by'' relation.
+An example dependency graph is given in Figure \ref{fig:dependencyexample}.
 
@@
 To handle this, we use an algorithm such as Tarjan's \citep{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations (as shown in Figure \ref{fig:sccexample}).
-References which point to the same SCC as the expression they appear in (which includes direct recursion and mutual recursion) are given a constant feature value $f_{recursion}$.
+If an identifier cannot be found in any cluster, it must appear in the same SCC as the expression we are processing; hence we use the constant feature value $f_{recursion}$.
 
-By working through the sorted list of SCCs, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in algorithm \ref{alg:recurrent}.
-This makes cyclic references particularly easy to handle: global identifiers are looked up in the current clusters $C$, and if not found the value $f_{recursion}$ is used.
+By working through the sorted list of SCCs, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in Algorithm \ref{alg:recurrent}.
 
@@
     \Procedure{RecurrentCluster}{}
-      \State $C  \gets []$
+      \State $\vect{C}  \gets []$
       \State $DB \gets \varnothing$
@@
         \State $DB \gets DB \cup \{(i, featureVec(e)) \mid (i, e) \in scc\}$
-        \State $C  \gets kMeans(DB)$
+        \State $\vect{C}  \gets kMeans(DB)$
       \EndFor
-      \Return $C$
+      \Return $\vect{C}$
     \EndProcedure
@@
 As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}.
-A valid topological ordering would be (eliding Core expressions to save space) $d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]$:
+A valid topological ordering is given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding Core expressions to save space):
+
+\begin{equation*}
+  d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]
+\end{equation*}
+
+We can then trace the execution of Algorithm \ref{alg:recurrent} as follows:
 
@@
   \item The first iteration through \textsc{RecurrentCluster}'s loop will set $scc \gets \{(\hs{plus}, \dots)\}$.
-  \item With $i = \hs{plus}$ and $e$ as its Core expression, calculating $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$ will become $f_{recursion}$ (since \hs{plus} doesn't appear in $C$).
-  \item The call to $kMeans$ will produce $C \gets [\{\hs{plus}\}]$, i.e. a single cluster containing \hs{plus}.
+  \item With $i = \hs{plus}$ and $e$ as its Core expression, calculating $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$ will become $f_{recursion}$ (since \hs{plus} doesn't appear in $\vect{C}$).
+  \item The call to $kMeans$ will produce $\vect{C} \gets [\{\hs{plus}\}]$, i.e. a single cluster containing \hs{plus}.
   \item The next iteration will set $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
@@
   \item Since the feature vectors for \hs{odd} and \hs{even} will be identical, $kMeans$ will put them in the same cluster.
-The number of clusters $k = \lceil \sqrt{\lvert DB \rvert} \rceil = 2$, so the other cluster must contain \hs{plus}.
-Their order is arbitrary, so one possibility is $C = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
-  \item Finally \hs{mult} will be clustered. The recursive call will become $f_{recursion}$ whilst the call to \hs{plus} will become $2$, since $\hs{plus} \in C_2$.
-  \item Since $k = 2$, the resulting clusters will be $C \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus}, \hs{mult}\}]$.
+To avoid the degenerate case of a single cluster, for this example we will assume that $k = 2$; in which case the other cluster must contain \hs{plus}.
+Their order is arbitrary, so one possibility is $\vect{C} = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
+  \item Finally \hs{mult} will be clustered. The recursive call will become $f_{recursion}$ whilst the call to \hs{plus} will become $2 + 3 \alpha$, since $\hs{plus} \in C_2$.
+  \item \begin{sloppypar}Again assuming that $k = 2$, the resulting clusters will be \mbox{$\vect{C} \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus}, \hs{mult}\}]$}.\end{sloppypar}
 \end{itemize}

### 2016-01-26 CICM2016/evaluation.tex
Copied TransferReport to CICM2016

\section{Evaluation}
\label{sec:evaluation}

We have applied our recurrent clustering algorithm to several scenarios, with mixed results. A major difficulty in evaluating these clusters is that we have no ``ground truth'', i.e. there is no objectively correct way to compare expressions. Instead, we provide a qualitative overview of the more interesting characteristics.

As a simple example, we clustered (Haskell equivalents of) the running examples used to present ACL2(ml) \citep{heras2013proof}, shown in Figure \ref{fig:lisp}. These include tail-recursive and non-tail-recursive implementations of several functions. We expect those with similar \emph{structure} to be clustered together, rather than those which implement the same function. The results are shown in Figure \ref{fig:haskellcluster}, where we can see the ``\hs{Tail}'' functions clearly distinguished, with little distinction between the tail recursive and na\"{\i}ve implementations.

Next we tested whether these same functions would be clustered together when mixed with seemingly-unrelated functions, in this case 207 functions from Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail} functions appeared together in a separate cluster from the rest. This was unexpected, with no obvious semantic connection between these two functions and the others in their cluster (although most are recursive, due to the nature of the \hs{text} library).

\begin{figure}
  \begin{haskell}
(defun fact (n)
  (if (zp n) 1 (* n (fact (- n 1)))))

(defun helper-fact (n a)
  (if (zp n) a (helper-fact (- n 1) (* a n))))

(defun fact-tail (n)
  (helper-fact n 1))

(defun power (n)
  (if (zp n) 1 (* 2 (power (- n 1)))))

(defun helper-power (n a)
  (if (zp n) a (helper-power (- n 1) (+ a a))))

(defun power-tail (n)
  (helper-power n 1))

(defun fib (n)
  (if (zp n)
      0
      (if (equal n 1)
          1
          (+ (fib (- n 1)) (fib (- n 2))))))

(defun helper-fib (n j k)
  (if (zp n)
      j
      (if (equal n 1)
          k
          (helper-fib (- n 1) k (+ j k)))))

(defun fib-tail (n)
  (helper-fib n 0 1))
  \end{haskell}
  \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
  \label{fig:lisp}
\end{figure}

\begin{figure}
  \begin{haskell}
fact n = if n == 0
            then 1
            else n * fact (n - 1)

helperFact n a = if n == 0
                    then a
                    else helperFact (n - 1) (a * n)

factTail n = helperFact n 1

power n = if n == 0
             then 1
             else 2 * power (n - 1)

helperPower n a = if n == 0
                     then a
                     else helperPower (n - 1) (a + a)

powerTail n = helperPower n 1

fib n = if n == 0
           then 0
           else if n == 1
                   then 1
                   else fib (n - 1) + fib (n - 2)

helperFib n j k = if n == 0
                     then j
                     else if n == 1
                          then k
                          else helperFib (n - 1) k (j + k)

fibTail n = helperFib n 0 1
  \end{haskell}
  \caption{Haskell equivalents of the Common Lisp functions in Figure \ref{fig:lisp}.}
  \label{fig:haskelllisp}
\end{figure}

\begin{figure}
  \centering
    \tikzstyle{block} = [rectangle, draw]
    \tikzstyle{line}  = [draw, -latex']
    \begin{tikzpicture}[node distance=3cm]
      \node [block] (c1) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                    (factTail)  {\hs{factTail}};
          \node [block, below of=factTail] (fibTail)   {\hs{fibTail}};
          \node [block, below of=fibTail]  (powerTail) {\hs{powerTail}};
        \end{tikzpicture}
      };
      \node [block, right of=c1] (c2) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                (fact)       {\hs{fact}};
          \node [block, below of=fact] (helperFact) {\hs{helperFact}};
        \end{tikzpicture}
      };
      \node [block, right of=c2] (c3) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                       (fib)         {\hs{fib}};
          \node [block, below of=fib]         (helperFib)   {\hs{helperFib}};
          \node [block, below of=helperFib]   (helperPower) {\hs{helperPower}};
          \node [block, below of=helperPower] (power)       {\hs{power}};
        \end{tikzpicture}
      };
  \end{tikzpicture}
  \caption{Typical clusters for the functions in Figure \ref{fig:haskelllisp}.}
  \label{fig:haskellcluster}
\end{figure}

We have also applied our recurrent clustering algorithm to a variety of the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including \texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we expected functions with a similar purpose to appear together, such as the various reader and writer functions of \hs{pandoc}, there were always a few exceptions which became separate for reasons which are still unclear.

When clustering the \hs{yesod} Web framework, the clustering did seem to match our intuitions, in particular since all 15 of Yesod's MIME type identifiers appeared in the same cluster.

Whilst recurrent clustering has produced results which merit further investigation, the application to theory exploration has yet to be tested empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type class to generate random values for instantiating variables. Whilst we can automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not all types have \hs{Arbitrary} instances; those without cannot be given any variables in our signature, which severely limits the possible combinations which can be explored. In many cases, no variables can be included at all, leaving just equations involving constants. This has so far prevented us from measuring the direct impact on \qspec{} performance, either directly by exploring the sub-sets identified through recurrent clustering, or indirectly by comparing the equations generated by a full brute-force search to our recurrent clusters: those equations relating terms from different clusters would not be discovered by our method. It is this ratio of equations found through brute force to those found after narrowing-down by clusters which is one of our key objectives to maximise at this stage; until we begin to pursue the \emph{interestingness} of the properties.

The following less-serious problems were also encountered while applying \textsc{ML4HS} to \textsc{Hackage} packages:

\begin{itemize}
  \item Some packages, such as \texttt{warp} and \texttt{conduits}, get no declarations to cluster. This is because they make all of their declarations privately, e.g. in ``internal'' modules, then use separate modules to export the public declarations. GHC's renaming phase makes all references to such exports canonical, by pointing them to the private declarations. This forces us to ignore such declarations, as \qspec{} will not be able to access them.

  \item Since we do not support type-level entities, we ignore type classes. Unfortunately, this also means ignoring any value-level bindings (AKA ``methods'') which occur in a type class instance. Instead of being clustered, these result in references getting $f_{recursion}$ features. This is especially noticable in libraries like \hs{scientific}, where only the functions for constructing and destructing numbers in scientific notation are clustered; all of the arithmetic is defined in type classes. One difficulty with supporting methods is that their namespace in Core is disjoint from that of regular Haskell identifiers: a transformation layer would be required, along with explicit type annotations to avoid ambiguity.
\end{itemize}

It seems like this recurrent clustering method has promise, although it will require a more thorough exploration of the parameters to obtain more intuitively reliable results. These clusters can then be used in several ways to perform theory exploration; the most na\"{\i}ve way being to explore each cluster as \qspec{} signature in its own right. \textsc{ML4HS} already provides this functionality, although the lack of test generators severely limits what can be discovered.

As alluded to previously, we also have the opportunity to reason by \emph{analogy}. Similar to the work on ACL2(ml) \citep{heras2013proof}, we could produce a general ``scheme'' from each equation we find (either through \qspec{} or by data mining test suites); like Isabelle approaches have shown \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be instantiated to a variety of similar values, in an attempt to find new theorems which are analogues of existing results from a different context. ``Mutating'' existing theorem statements in such a way would also increase the chance of any result being considered interesting; since it's likely that the unmodified statement was deemed interesting, and the new result would not in general follow as a simple logical consequence.

### 2016-01-26 CICM2016/Email.hs

@@ -1,0 +1,21 @@
+module Email (User, Host, Email(), at, render) where
+type User  = String
+type Host  = String
+data Email = E User Host
+
+instance Show Email where
+  show = render
+
+eUser :: Email -> User
+eUser (E u h) = u
+
+eHost :: Email -> Host
+eHost (E u h) = h
+
+render :: Email -> String
+render (E u h) = u ++ "@" ++ h
+
+at :: User -> Host -> Maybe Email
+at "" h  = Nothing
+at u  "" = Nothing
+at u  h  = Just (E u h)

### 2016-01-26 CICM2016/render.sh

@@ -1,0 +1,4 @@
+#! /usr/bin/env nix-shell
+#! nix-shell -i bash -p bash gnumake
+make clean
+make all

### 2016-02-29 CICM2016/future.tex
Remove gantt chart

--- 2015-12-15 TransferReport/future.tex
@@
 \label{sec:future}
-
-\begin{sidewaysfigure}
-  \centering
-  \includegraphics[trim=3cm 2cm 3cm 2cm, width=1.00\textwidth]{Gantt.pdf}
-  \caption{Gantt chart showing research direction, sub-tasks and approximate timeframes. CICM is the \emph{Conference on Intelligent Computer Mathematics}.}
-  \label{fig:gantt}
-\end{sidewaysfigure}
 
@@
 Different, and more sophisticated notions of interestingness have been widely studied in other fields, which may be applied in the context of theory exploration.
-
-\paragraph{Concept Formation} \label{sec:conceptformation} \leavevmode \newline
-
-One directly applicable area to consider is \emph{concept formation}, which considers the (automatic) generation of new definitions and axioms.
-Unlike theory exploration, such systems are not constrained by the requirement that their output be provable, and hence interestingness is an important way to judge the quality of the results.
-
-Approaches vary, from those which are directly related to theory exploration (such as the scheme-instantiating approach of \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory exploration system), to others which are more closely related to theories of human learning and discovery \citep{Piantadosi.Tenenbaum.Goodman:2012, mullerunderstanding}.
-Those based on finding patterns in data, such as \citep{Wille:2005}, may be useful in tandem with our expression database, and the results of value generators like \qcheck{}.
-
-Since tools like \hspec{} already call out to third-party automated theorem provers, and indeed our own ML4HS system uses the external Weka program, there may also be merit in using external concept formation or conjecture generation tools (or reimplementations of their ideas), in order to build up more structure on top of that provided by the code we are exploring.
-For example, the approaches taken by AM \citep{lenat1977automated, lenat1979automated}, Graffiti \citep{delavina2005some, delavina2005graffiti} and HR \citep{colton1999automatic, colton2000agent} could be used alongside those of \qspec{}.
-
-\paragraph{Artificial Curiosity} \label{sec:curiosity} \leavevmode \newline
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{fig:wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this reliance on interest is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, in a similar way to our recurrent clustering algorithm.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-This same behaviour may be applicable to the theorems produced by a theory exploration system.
-
-We can divide AC approaches into two groups: those which make \emph{explicit} use of interestingness, learning from signals which follow a Wundt curve; whilst \emph{implicit} approaches modify the \emph{output} of their learning algorithm(s), to engineer the overall system behaviour to follow a Wundt curve as an emergent property.
-
-A framework encompassing many examples of the explicit approach is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning; for comparison, many similar measures are surveyed in a data mining context in \citep{geng2006interestingness}.
-Many more reinforcement learning examples can be found in \citep{Kaplan2006, Lipson2007, Luciw2011, Macedo2000, Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991, oudeyer2004intelligent}; whilst more general descriptions are given in \citep{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be more amenable to our theory exploration setting.
-
-Many of these reward signals are based on information theory, with a prominent example being \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new insights into the structure of past experience can provide a space saving when compressed together.
-This seems particularly relevant for our problem of identifying interesting theorems: those new theorems (``observations'') which shorten the proofs of previously discovered theorems may be more general, more powerful and therefore more \emph{interesting} and hence worth keeping.
-In fact this is very similar to \qspec{}'s interestingness criterion.
-
-Another example of explicit artificial curiosity is given in \citep{Hester.Stone:2012}, where world states which cause \emph{disagreement} among a population of decision trees (a \emph{random forest} \citep{randomforests}) are considered interesting.
-Since the models make stochastic predictions, the disagreement follows a Wundt curve as the complexity of state transitions increases: for parts of the state space which have been fully learned, the models will agree on accurate predictions; for parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the same average value.
-Whilst the latter predictions may not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the interestingness of states which are too complex.
-
-Many examples of the implicit case are based on \emph{coevolution}: rewarding one part of the system for exploiting another part, and vice versa.
-In \citep{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on the outcome of actions, and the winner is rewarded at the expense of the loser.
-Due to the risk involved, each algorithm will only bet when it is confident in its prediction, and bets will only be actioned when each algorithm is confident in a \emph{different} outcome.
-The overall behaviour of this system is therefore similar to the explicit measure of disagreement used in the random forest example.
-In terms of theory exploration, such a scheme could be used to find theorems which are \emph{non-obvious}, and hence informative in some way to the user.
-
-Another case is the ``darwinian brain'' of Fernando et al.
-\citep{fernando2013design1, fernando2013design2}.
-This coevolves a population of problem generators and problem solvers, rewarding the solvers based on their speed, and rewarding the generators based on the \emph{variance} of the solvers' speed.
-This avoids trivial problems (which all solvers can quickly overcome) and complex problems (which no solver can manage), and focuses on those with the most possibility for learning.
-It is easy to imagine such a general architecture being populated by conjecture generators and theorem provers to form a theory exploration system.
-
-\paragraph{Evolutionary Computation} \leavevmode \newline
-
-Coevolution is a form of \emph{evolutionary computation}; an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting (there are also precedents for using evolutionary computation in a theorem proving domain \citep{spector2008genetic}).
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.

### 2016-02-29 CICM2016/check.sh
Change page limit

@@ -1,6 +1,5 @@
 #!/usr/bin/env bash
 
-cd /home/chris/Writing/TransferReport
 ERR=0
 
 TODOS=""
@@ -123,17 +122,13 @@ fi
 if [[ -e report.pdf ]]
 then
     PAGES=$(pdfinfo report.pdf | grep Pages | grep -o "[0-9]*$")
-    TOTAL=60
-    REMAINING=$(( TOTAL - PAGES ))
-    NOW=$(date +%s)
-    DEADLINE=$(date -d "2015-12-17" +%s)
-    DAYS=$(( (DEADLINE - NOW) / (60 * 60 * 24) ))
-    SHOULD=$(( TOTAL - DAYS ))
-    if [[ "$REMAINING" -gt "$DAYS" ]]
+    ALLOWED=15
+    if [[ "$PAGES" -gt "$ALLOWED" ]]
     then
-        echo "report.pdf should have $SHOULD pages by now, only has $PAGES" >> /dev/stderr
+        echo "report.pdf should have $ALLOWED pages, instead has $PAGES" >> /dev/stderr
+        ERR=1
     else
-        echo "Page target is $SHOULD, you have $PAGES!"
+        echo "Page target is $ALLOWED, you have $PAGES!"
     fi
 fi
 

### 2016-02-29 CICM2016/check.sh
Don't use citep

@@ -8,12 +8,19 @@ CITES=""
 MACROS=""
 LABELS=""
 FIGS=""
+CITEP=""
 for FILE in *.tex
 do
     if grep "TODO[^{]" < "$FILE" > /dev/null
     then
         NUM=$(grep -c TODO < "$FILE")
         TODOS=$(echo "$TODOS"; echo "Found $NUM undated TODOs in $FILE")
+        ERR=1
+    fi
+
+    if grep "citep" < "$FILE" > /dev/null
+    then
+        CITEP=$(echo "$CITEP"; echo "Found citep in '$FILE', should use cite")
         ERR=1
     fi
 
@@ -79,6 +86,7 @@ done
 [[ -z "$MACROS" ]] || echo "$MACROS" >> /dev/stderr
 [[ -z "$LABELS" ]] || echo "$LABELS" >> /dev/stderr
 [[ -z "$FIGS"   ]] || echo "$FIGS"   >> /dev/stderr
+[[ -z "$CITEP"  ]] || echo "$CITEP"  >> /dev/stderr
 
 echo "Rendering report.tex" >> /dev/stderr
 touch *.tex

### 2016-02-29 CICM2016/background.tex
Use cite not citep, and equation not equation*

--- 2015-12-15 TransferReport/background.tex
@@
 \begin{figure}
-  \begin{equation*}
+  \begin{equation}
     \begin{split}
@@
     \end{split}
-  \end{equation*}
+  \end{equation}
   Where:
@@
 
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
+We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and \hspec{} \cite{claessen2013automating}).
+This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
 
-Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
+Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
 Like most \emph{functional} programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
 For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
-Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
+Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \cite[Appendix C]{sulzmann2007system}.
 
@@
 
-We focus on \qspec{} \citep{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
+We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
 These conjectures are arrived at through the following stages:
@@
 
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
+Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
 
@@
 \begin{itemize}
-  \item For simplicity, we use the ``rule of thumb'' given in \citep[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
+  \item For simplicity, we use the ``rule of thumb'' given in \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
   \item Data points will be $d$-dimensional feature vectors, as defined above.
   \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \citep{lloyd1982least}.
+  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \cite{lloyd1982least}.
 \end{itemize}
 
-This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \citep{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
+This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \cite{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
 
@@
 
-\begin{equation*}
+\begin{equation}
   m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
-\end{equation*}
+\end{equation}
 
 Before k-means starts, we must choose \emph{seed} values for $\vect{m}^i(0)$.
-Many methods have been proposed for choosing these values \citep{arthur2007k}.
+Many methods have been proposed for choosing these values \cite{arthur2007k}.
 For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
@@
 
-\begin{equation*}
+\begin{equation}
   C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
-\end{equation*}
+\end{equation}
 

### 2016-02-29 CICM2016/contributions.tex
Use cite not citep, and equation not equation*

@@
 
-We adapt the methodology of \emph{recurrent clustering} proposed in \citep{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
+We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
@@
 
-\begin{equation*}
+\begin{equation}
   \begin{array}{r@{}l@{}l@{}}
@@
   \end{array}
-\end{equation*}
+\end{equation}
 
@@
 \begin{small}
-  \begin{equation*}
+  \begin{equation}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
@@
     \end{array}
-  \end{equation*}
+  \end{equation}
 \end{small}
@@
 \begin{small}
-  \begin{equation*}\label{eq:flattened}
+  \begin{equation}\label{eq:flattened}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
@@
 To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in Figure \ref{fig:totree}.
-We follow the presentation in \citep{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
 $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
@@
   \begin{subfigure}{\textwidth}
-    \begin{equation*}
+    \begin{equation}
       \begin{bmatrix}
@@
       \end{bmatrix}
-    \end{equation*}
+    \end{equation}
     \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
@@
   \begin{subfigure}{\textwidth}
-    \begin{equation*}
+    \begin{equation}
       (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
-     \end{equation*}
+     \end{equation}
      \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
@@
 We now define the function $\phi$, which turns terminal symbols of Core syntax into features (real numbers).
-For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \citep{DBLP:journals/corr/HerasK14} in Coq.
+For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14} in Coq.
 We use a constant $\alpha$ to separate these values from those of other tokens (e.g.
-identifiers), but the order is essentially arbitrary: \footnote{In \citep{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
+identifiers), but the order is essentially arbitrary: \footnote{In \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
 
@@
 This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
-To handle this, we use an algorithm such as Tarjan's \citep{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations (as shown in Figure \ref{fig:sccexample}).
+To handle this, we use an algorithm such as Tarjan's \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations (as shown in Figure \ref{fig:sccexample}).
 If an identifier cannot be found in any cluster, it must appear in the same SCC as the expression we are processing; hence we use the constant feature value $f_{recursion}$.
@@
 
-\begin{equation*}
+\begin{equation}
   d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]
-\end{equation*}
+\end{equation}
 

### 2016-02-29 CICM2016/evaluation.tex
Use cite not citep, and equation not equation*

@@
 
-As a simple example, we clustered (Haskell equivalents of) the running examples used to present ACL2(ml) \citep{heras2013proof}, shown in Figure \ref{fig:lisp}.
+As a simple example, we clustered (Haskell equivalents of) the running examples used to present ACL2(ml) \cite{heras2013proof}, shown in Figure \ref{fig:lisp}.
 These include tail-recursive and non-tail-recursive implementations of several functions.
@@
 As alluded to previously, we also have the opportunity to reason by \emph{analogy}.
-Similar to the work on ACL2(ml) \citep{heras2013proof}, we could produce a general ``scheme'' from each equation we find (either through \qspec{} or by data mining test suites); like Isabelle approaches have shown \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be instantiated to a variety of similar values, in an attempt to find new theorems which are analogues of existing results from a different context.
+Similar to the work on ACL2(ml) \cite{heras2013proof}, we could produce a general ``scheme'' from each equation we find (either through \qspec{} or by data mining test suites); like Isabelle approaches have shown \cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be instantiated to a variety of similar values, in an attempt to find new theorems which are analogues of existing results from a different context.
 ``Mutating'' existing theorem statements in such a way would also increase the chance of any result being considered interesting; since it's likely that the unmodified statement was deemed interesting, and the new result would not in general follow as a simple logical consequence.

### 2016-02-29 CICM2016/future.tex
Use cite not citep, and equation not equation*

@@
 
-We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \citep{bengio2013representation}.
+We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \cite{bengio2013representation}.
 This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.
@@
 
-The \textsc{SmallCheck} system \citep{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly.
+The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly.
 Whilst this does not make \textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage is that it may use much less memory, as the generated values are built up incrementally.
@@
 
-A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \citep{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded.
+A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \cite{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded.
 This may narrow down our search procedures greatly, especially when predicates are involved.
@@
 
-Other approaches to generating inhabitants include \textsc{Djinn} \citep{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator).
-\textsc{MuCheck} \citep{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g.
+Other approaches to generating inhabitants include \textsc{Djinn} \cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator).
+\textsc{MuCheck} \cite{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g.
 changing the order of pattern match clauses); whilst not as exhaustive as the other approaches, mutating existing values in this way is claimed to yield values which correspond more closely to what a programmer might write.

### 2016-02-29 CICM2016/implementation.tex
Use cite not citep, and equation not equation*

--- 2015-12-15 TransferReport/implementation.tex
@@
 
-The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \citep{mitchell2008hoogle}.
+The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \cite{mitchell2008hoogle}.
 Definitions are indexed by a combination of their package name, module name and binding name.

### 2016-02-29 CICM2016/intro.tex
Use cite not citep, and equation not equation*

--- 2015-12-15 TransferReport/intro.tex
@@
 As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-The \emph{functional programming} paradigm has been proposed for addressing these issues \citep{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis.
+The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis.
 For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
@@
 Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
-Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \citep{gosling2015java} and C++ \citep{willcock2006lambda}.
+Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
 

### 2016-02-29 CICM2016/related.tex
Use cite not citep, and equation not equation*

--- 2015-12-15 TransferReport/related.tex
@@
 \hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
+This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \cite{wadler1989theorems}.
 
@@
 \hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
+Many more examples can be found in the \emph{typeclassopedia} \cite{yorgey2009typeclassopedia}.
 This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
@@
 Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
+However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \cite{lindley2014hasochism}).
 This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
@@
 
-\item The existence of theory exploration systems such as \hspec{}, and related tools which we may be able to re-use, including conjecture generators like \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
+\item The existence of theory exploration systems such as \hspec{}, and related tools which we may be able to re-use, including conjecture generators like \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \cite{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \cite{le2014mucheck}, \textsc{MagicHaskeller} \cite{katayama2011magichaskeller} and \textsc{Djinn} \cite{augustsson2005djinn}.
 
@@
 By removing user interaction, these concerns about directing search must be solved by algorithms.
-As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \citep{Montano-Rivas.McCasland.Dixon.ea:2012, johansson2009isacosy, Hipster}.
+As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012, johansson2009isacosy, Hipster}.
 
@@
 
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
+The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
 Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
@@
 
-The initial approach taken by Sledgehammer, known as \textsc{MePo} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
+The initial approach taken by Sledgehammer, known as \textsc{MePo} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
 Initially, the relevant symbols are those which occur in the goal to be proved, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
@@
 
-Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
+Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \cite{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
 Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
@@
 
-Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}, some of which may be directly applicable in the context of theory exploration.
+Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
+Many more approaches are described and evaluated in \cite{kuhlwein2012overview}, some of which may be directly applicable in the context of theory exploration.
 
@@
 
-Our recurrent clustering approach takes inspiration from the ML4PG \citep{journals/corr/abs-1212-3618} and ACL2(ml) \citep{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
+Our recurrent clustering approach takes inspiration from the ML4PG \cite{journals/corr/abs-1212-3618} and ACL2(ml) \cite{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
 Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
@@
 Unfortunately RNNs are difficult to train.
-The standard way to train NNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
+The standard way to train NNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
 However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
+The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \cite{hochreiter1997long}).
 
 Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward NN to learn recursive distributed representations \citep{pollack1990recursive} which correspond to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which we can use to encode sub-trees when the branching factor of nodes is not  uniform \citep{kwasny1995tail}.
-More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
+The \emph{back-propagation through structure} approach \cite{goller1996learning} is a more direct solution to this problem, using a feed-forward NN to learn recursive distributed representations \cite{pollack1990recursive} which correspond to the recursive structure of the inputs.
+Such distributed representations can also be used for sequences, which we can use to encode sub-trees when the branching factor of nodes is not  uniform \cite{kwasny1995tail}.
+More recent work has investigated storing recursive structures inside LSTM cells \cite{zhu2015long}.
 
-A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
-Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
+A simpler alternative for generating recursive distributed representations is to use circular convolution \cite{conf/ijcai/Plate91}.
+Although promising results are shown for its use in \emph{distributed tree kernels} \cite{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
 
-\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}, which could be used to learn more subtle relations between expressions than recurrent clustering alone.
+\emph{Kernel methods} have also been applied to structured information, for example in \cite{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
+Many more applications of kernel methods to structured domains are given in \cite{bakir2007predicting}, which could be used to learn more subtle relations between expressions than recurrent clustering alone.
 
@@
 
-We use the Weka tool to perform k-means clustering \citep{Holmes.Donkin.Witten:1994}, since we are more concerned with the application of feature extraction to Haskell and its use in theory exploration, rather than precise tuning of learning algorithms.
+We use the Weka tool to perform k-means clustering \cite{Holmes.Donkin.Witten:1994}, since we are more concerned with the application of feature extraction to Haskell and its use in theory exploration, rather than precise tuning of learning algorithms.
 Since k-means is a standard method, there are many other implementations available.
@@
 In any case there are many conservative improvements to the standard k-means algorithm, which could be applied to our setup.
-For example, a more efficient approach like \emph{yinyang k-means} \citep{conf/icml/DingZSMM15} could make larger input sizes more practical to cluster, especially since recurrent clustering invokes k-means many times.
-The \emph{k-means++} approach \citep{arthur2007k, bahmani2012scalable} can be used to more carefully select the ``seed'' values for the first timestep, and the \emph{x-means} algorithm \citep{pelleg2000x} is able to estimate how many clusters to use (our \emph{final} clusters should be tuned to maximise the performance of the subsequent theory exploration step, but x-means could still be useful in the recurrent clustering steps).
+For example, a more efficient approach like \emph{yinyang k-means} \cite{conf/icml/DingZSMM15} could make larger input sizes more practical to cluster, especially since recurrent clustering invokes k-means many times.
+The \emph{k-means++} approach \cite{arthur2007k, bahmani2012scalable} can be used to more carefully select the ``seed'' values for the first timestep, and the \emph{x-means} algorithm \cite{pelleg2000x} is able to estimate how many clusters to use (our \emph{final} clusters should be tuned to maximise the performance of the subsequent theory exploration step, but x-means could still be useful in the recurrent clustering steps).
 
@@
 TODO: Theory Exploration?
-\citep{johansson2009isacosy}
+\cite{johansson2009isacosy}
 \fi

### 2016-03-01 CICM2016/background.tex
Cut out a load

@@
 
-\begin{figure}
-  \begin{equation}
-    \begin{split}
-      expr\    \rightarrow\ & \CVar\ id                          \\
-                         |\ & \CLit\ literal                     \\
-                         |\ & \CApp\ expr\ expr                  \\
-                         |\ & \CLam\ \mathcal{L}\ expr           \\
-                         |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                         |\ & \CType                             \\
-      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                         |\ & \CGlobal\      \mathcal{G}         \\
-                         |\ & \CConstructor\ \mathcal{D}         \\
-      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                         |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                         |\ & \CLitAlt\ literal                  \\
-                         |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                         |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-    \end{split}
-  \end{equation}
-  Where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers \\
-    $\mathcal{D}$ & constructor identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
-
 We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and \hspec{} \cite{claessen2013automating}).
@@
 
-Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \cite{marlow2010haskell}.
-Like most \emph{functional} programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
+Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
 For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
@@
 \begin{figure}
-  \begin{haskell}
+  \begin{haskell}\begin{verbatim}
 data Nat = Z
@@
 even    Z  = True
-    even (S n) = odd n
+even (S n) = odd n\end{verbatim}
   \end{haskell}
@@
 \begin{figure}
-  \begin{subfigure}[plus]{\textwidth}
   \begin{small}
@@
     \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[mult]{\textwidth}
-    \begin{small}
+
     \underline{\texttt{mult}}
@@
     \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[odd]{\textwidth}
-    \begin{small}
+
     \underline{\texttt{odd}}
@@
     \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[even]{\textwidth}
-    \begin{small}
+
     \underline{\texttt{even}}
@@
   \end{small}
-  \end{subfigure}
   \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}.
@@
 
-\subsection{QuickCheck}
-\label{sec:quickcheck}
-
-Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g.
-life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = n * factorial (n-1)
-
-fact_base      = factorial 0 == factorial 1
-fact_increases = factorial 3 <= factorial 4
-fact_div       = factorial 4 == factorial 5 `div` 5
-\end{lstlisting}
-
-The intent of the function is to map an input $n$ to an output $n!$.
-The tests check a few properties of the implementation, including the base case, that the function is monotonically increasing, and a relationship between adjacent outputs.
-However, these tests will \emph{not} expose a serious problem with the implementation: it diverges on half of its possible inputs!
-
-All of Haskell's built-in numeric types allow negative numbers, which this implementation doesn't take into account.
-Whilst this is a rather trivial example, it highlights a common problem: unit tests are insufficient to expose incorrect assumptions.
-In this case, our assumption that numbers are positive has caused a bug in the implementation \emph{and} limited the tests we've written.
-
-If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = let nPos = abs n
-               in nPos * factorial (nPos - 1)
-
-fact_neg = factorial 1 == factorial (-1)
-\end{lstlisting}
-
-However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t.
-Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
-In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge.
-Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
-
-Haskell offers a partial solution to this problem in the form of \emph{property checking}.
-Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
-Here is how we might restate our unit tests as properties:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-fact_base        = factorial 0 == factorial 1
-fact_increases n = factorial n <= factorial (n+1)
-fact_div       n = factorial n == factorial (n+1) `div` (n+1)
-fact_neg       n = factorial n == factorial (-n)
-\end{lstlisting}
-
-The free variables (all called \hs{n} in this case) are abstracted as function parameters; these parameters are implicitly \emph{universally quantified}, i.e.
-we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
-Notice that unit tests like \hs{fact\_base} are valid properties; they just assert rather weak statements.
-
-To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
-For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
-The default stopping criterion for \qcheck{} (for each test) is when a single generated test fails, or when 100 generated tests pass.
-
-The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:
-
-\begin{itemize}
-  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type).
-This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
-
-  \item The generators used by \qcheck{} depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
-This will expose the problem with negative numbers, which we weren't expecting.
-\end{itemize}
-
-Property checking is certainly an improvement over unit testing, but the problem of tests being biased towards expected cases remains, since we are manually specifying the properties to be checked.
-
-We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}.
-These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.
-
 \subsection{Theory Exploration}
@@
 
-In this work we consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
+We consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
 The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
 
-We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
+We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about Haskell code.
+These may be fed into another tool, such as \hspec{}, for subsequent proving.
 These conjectures are arrived at through the following stages:
@@
   \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item Each variable is instantiated to a particular value, generated randomly by \qcheck{}.
-  \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
-If so, the class is split up to separate such distinguishable members.
-  \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
-  \item A set of equations are then conjectured, relating each class's members.
+  \item The equivalence of terms in each class is tested using \qcheck{}: variables are instantiated to particular values, generated randomly, and the resulting closed expressions are evaluated and compared for equality.
+  \item If a class is found to have non-equal members, it is split up to separate those members.
+  \item The previous steps of testing and splitting are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
+  \item For each class, one member is selected and equations are conjectured that it is equal to each of the other members.
 \end{enumerate}
 
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
+Such conjectures can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
 
+\iffalse
 As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
@@
 Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments.
-In addition, Haskell curries functions, so these binary functions will be treated as unary functions which return unary functions.
-This is required as the construction of $terms$ applies functions to one argument at a time.
+In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.
 
@@
   % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
-  \begin{haskell}
+  \begin{verbatim}
                   plus a b = plus b a
@@
 plus (mult a b) (mult a c) = mult a (plus b c)
-  \end{haskell}
+  \end{verbatim}
   \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
@@
 As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
+\fi
 
 Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-Here we know that addition and multiplication are closely related, and hence obey many algebraic laws.
+For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws.
 Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
 
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
-
+\iffalse
+TODO
 In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
 One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
+\fi
 
@@
 it would require a distinguished expression to compare everything against.
-Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
+Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{all} terms in a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
 
@@
 
-\subsubsection{Feature Extraction}
-\label{sec:featureextraction}
-
-Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
-This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms.
-By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
-
-We use a standard representation of features as a \emph{feature vector} of numbers $\vect{x} = (x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$ (we use \textbf{bold face} to represent vectors, including feature vectors).
-\footnote{In fact, practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
-
-\begin{itemize}
-  \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $d$.
-Many machine learning algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
-  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
-This enables many powerful machine learning algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
-  \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
-This reduces the input size of the machine learning problem, improving efficiency.
-\end{itemize}
-
-\subsubsection{K-Means}
-\label{sec:kmeans}
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:
-
-\begin{itemize}
-  \item For simplicity, we use the ``rule of thumb'' given in \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
-  \item Data points will be $d$-dimensional feature vectors, as defined above.
-  \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \cite{lloyd1982least}.
-\end{itemize}
-
-This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \cite{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
-
-Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$.
-As the name suggests, k-means uses the mean value of each cluster, which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation}
-  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
-\end{equation}
-
-Before k-means starts, we must choose \emph{seed} values for $\vect{m}^i(0)$.
-Many methods have been proposed for choosing these values \cite{arthur2007k}.
-For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
-
-The elements of each cluster $C^i(t)$ are those data points closest to the mean value at the previous time step:
-
-\begin{equation}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
-\end{equation}
-
-As $t$ increases, the clusters $C^i$ move from their initial location around the ``seeds'', to converge on a local minimum of the ``within-cluster sum of squared error'' objective:
-
-\begin{equation} \label{eq:kmeansobjective}
-  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation}
+We use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, as this setup is used by ML4PG and ACL2(ml).
+For simplicity, we will cluster $n$ points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters, following the ``rule of thumb'' given in \cite[pp.
+365]{mardia1979multivariate}, and will use randomly-selected elements of our data set to determine the initial clusters (known as the Forgy method).

### 2016-03-01 CICM2016/contributions.tex
Cut out a load

@@
 
-We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14, heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
+We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
 As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
-Its distinguishing characteristic is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as Algorithm \ref{alg:recurrent}), which goes beyond a simple syntactic characterisation, to allow the features of an expression to depend on those it references.
+Rather than clustering expressions directly, we first transform them into numeric vectors in a process called \emph{feature extraction}.
+The distinguishing characteristic of recurrent clustering is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as Algorithm \ref{alg:recurrent}), which allows the feature vector of an expression to depend on those of the expressions it references.
 Here we describe our approach to recurrent clustering and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
-We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat feature vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote as the function $\phi$.
+We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the function $\phi$.
 
@@
 
-Our recurrent clustering algorithm makes use of the k-means algorithm described in \S \ref{sec:kmeans}, which considers the elements of a feature vector to be \emph{orthogonal}.
+Our recurrent clustering algorithm is based on k-means clustering, which considers the elements of a feature vector to be \emph{orthogonal}.
 Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
@@
 
-\begin{equation}
+\begin{equation}\label{eq:xy}
   \begin{array}{r@{}l@{}l@{}}
@@
 
-To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in Figure \ref{fig:totree}.
-We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
-$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
-$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
-The results are illustrated in Figure \ref{fig:rosetreeexample}.
-
-\begin{figure}
-  \begin{align*}
-    toTree(e) &=
-    \begin{cases}
-      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
-      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
-      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
-      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
-      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
-      (\feature{\CType})                                                & \text{if $e = \CType$} \\
-      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
-      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
-      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
-      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
-      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
-      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
-      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
-      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
-      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
-      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
-      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
-    \end{cases}
-  \end{align*}
-  \caption{Transforming Core expressions of Figure \ref{fig:coresyntax} to rose trees.
-The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
-for $\CRec$ we apply $toTree$ to each $e_i$.
-We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
-We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.}
-  \label{fig:totree}
-\end{figure}
+To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the function $toTree$ given in Figure \ref{fig:totree}.
+Intuitively, these rose trees are simply s-expression representations of the ASTs, illustrated for the \hs{odd} function in Figure \ref{fig:rosetreeexample}.
 
@@
   \centering
-  \begin{subfigure}{\textwidth}
-    \begin{small}
+  \begin{scriptsize}
       \Tree[ .$\feature{\CLam}$
@@
                           $\feature{\id{n}}$ ]]]
-    \end{small}
-    \caption{Rose tree for the expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
-    \label{fig:rosetreeexample}
-  \end{subfigure}
-  \vspace{1ex}
-  \begin{subfigure}{\textwidth}
-    \begin{equation}
+  \end{scriptsize}
+  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the expression \hs{odd} from Figure \ref{fig:coreexample}.
+Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
+\end{figure}
+
+\begin{figure}
+    \begin{equation*}
       \begin{bmatrix}
@@
       \end{bmatrix}
-    \end{equation}
+    \end{equation*}
     \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
     \label{fig:matrixexample}
-  \end{subfigure}
-  \vspace{1ex}
-  \begin{subfigure}{\textwidth}
-    \begin{equation}
+\end{figure}
+
+\begin{figure}
+    \begin{equation*}
       (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
-     \end{equation}
+     \end{equation*}
      \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
      \label{fig:vectorexample}
-  \end{subfigure}
-  \caption{Feature extraction applied to the expression \hs{odd} from Figure \ref{fig:coreexample}.}
 \end{figure}
 
-These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting, and concatenating them together (written as $\cat$):
-
-\begin{equation} \label{eq:level}
-  level(l, (f, T_1, \dots, T_{n_T})) =
-    \begin{cases}
-      (f)                                                    & \text{if $l = 1$} \\
-      level(l - 1, T_1) \cat \dots \cat level(l - 1, T_{n_T}) & \text{if $l > 1$}
-    \end{cases}
-\end{equation}
-
-Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t))$, where $pad$ either truncates or appends zeros, until the row has a fixed length $c$.
-We also truncate/pad the number of rows to match a fixed number $r$.
-This way, all expressions give rise to $r \times c$ matrices, where the left-most features at each each level are aligned.
+These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}.
+Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, with any required padding added on the right.
+These matrices are either padded with $0$ or truncated, to fit a fixed size $r \times c$.
 
@@
 
-\begin{figure}
-  \centering
-  \begin{subfigure}{\textwidth}
-    \centering
-    \begin{tikzpicture}[node distance=2cm]
-      \tikzset{vertex/.style = {minimum size=1.5em}}
-      \tikzset{edge/.style = {->,> = latex'}}
-
-      \node[vertex]                at (0.5\textwidth, 3)   (z)     {\hs{Z}};
-      \node[vertex]                at (0.5\textwidth, 0)   (s)     {\hs{S}};
-      \node[vertex]                at (0,             1.5) (plus)  {\hs{plus}};
-      \node[vertex, anchor=east]   at (\textwidth,    1.5) (false) {\hs{False}};
-      \node[vertex, right of=plus]                         (mult)  {\hs{mult}};
-      \node[vertex, left of=false]                         (odd)   {\hs{odd}};
-      \node[vertex, left of=odd]                           (even)  {\hs{even}};
-      \node[vertex, left of=even]                          (true)  {\hs{True}};
-
-      \draw[edge, dashed]     (z)     to (plus);
-      \draw[edge, dashed]     (z)     to (mult);
-      \draw[edge, dashed]     (z)     to (odd);
-      \draw[edge, dashed]     (z)     to (even);
-      \draw[edge, dashed]     (s)     to (plus);
-      \draw[edge, dashed]     (s)     to (mult);
-      \draw[edge, dashed]     (s)     to (odd);
-      \draw[edge, dashed]     (s)     to (even);
-      \draw[edge, dashed]     (false) to (odd);
-      \draw[edge, dashed]     (true)  to (even);
-      \draw[edge]             (plus)  to (mult);
-      \draw[edge]             (even)  to (odd);
-      \draw[edge]             (odd)   to (even);
-      \draw[edge, loop above] (plus)  to (plus);
-      \draw[edge, loop above] (mult)  to (mult);
-    \end{tikzpicture}
-
-    \caption{Dependency graph for Figure \ref{fig:coreexample}.
-Loops indicate recursive functions, double arrows indicate mutual recursion.
-Dashed lines show references to data constructors, which we do not consider in this work.}
-    \label{fig:dagexample}
-  \end{subfigure}
-  \begin{subfigure}{\textwidth}
-    \centering
-    \tikzstyle{block} = [rectangle, draw]
-    \tikzstyle{line}  = [draw, -latex']
-    \begin{tikzpicture}[node distance=2cm]
-      \node [block] (plusset) {
-        \begin{tikzpicture}[node distance = 2cm]
-          \node [block] (plus) {\hs{plus}};
-        \end{tikzpicture}
-      };
-      \node [block, right of=plusset] (parityset) {
-        \begin{tikzpicture}[node distance = 1cm]
-          \node [block]               (odd)  {\hs{odd}};
-          \node [block, below of=odd] (even) {\hs{even}};
-        \end{tikzpicture}
-      };
-      \node [block, right of=parityset] (multset) {
-        \begin{tikzpicture}[node distance = 2cm]
-          \node [block] (mult) {\hs{mult}};
-        \end{tikzpicture}
-      };
-
-      \path [line] (plusset)   -- (parityset);
-      \path [line] (parityset) -- (multset);
-    \end{tikzpicture}
-    \caption{One possible topological sorting of simply connected components for Figure \ref{fig:dagexample} (ignoring constructors).
-\hs{even} and \hs{odd} are mutually recursive, neither can appear before the other, so they are grouped into one component and handled concurrently by Algorithm \ref{alg:recurrent}.}
-    \label{fig:sccexample}
-  \end{subfigure}
-  \caption{Sorting functions from Figure \ref{fig:coreexample} into dependency order.}
-  \label{fig:dependencyexample}
-\end{figure}
-
 For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference).
 In this way, we can avoid looking up expressions which haven't been clustered yet.
-To perform our sort, we construct a directed graph, where each declaration is a node and edges denote an ``is referenced by'' relation.
-An example dependency graph is given in Figure \ref{fig:dependencyexample}.
-
-In a topological sort we say that $A < B$ for nodes $A$ and $B$ if we can reach $B$ by following edges from $A$ (if $A$ and $B$ cannot be reached from each other, their order is arbitrary).
-This is slightly complicated in Haskell (compared to Coq, for example), since general recursion is permitted and hence the dependency graph may contain cycles.
-To handle this, we use an algorithm such as Tarjan's \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations (as shown in Figure \ref{fig:sccexample}).
+To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.
 If an identifier cannot be found in any cluster, it must appear in the same SCC as the expression we are processing; hence we use the constant feature value $f_{recursion}$.
@@
 
+\iffalse
 As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}.
@@
 A more extensive investigation of these features requires a concrete implementation, in particular to pin down values for the parameters such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
+\fi
 
@@
 
-\begin{coqblock}
+\begin{verbatim}
 Fixpoint even n := match n with
@@
                    end.
-\end{coqblock}
+\end{verbatim}
 

### 2016-03-01 CICM2016/future.tex
Cut out a load

@@
 In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique.
-The Gantt chart in Figure \ref{fig:gantt} shows how these relate to the short- and long-term direction being taken by this research.
 Below, we elaborate on the details, background and motivation for these choices.
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+\begin{haskell}
 genRoseTree = do f        <- arbitrary
@@
                  return (Node f subtrees)
-\end{lstlisting}
+\end{haskell}
 

### 2016-03-01 CICM2016/implementation.tex
Cut out a load

@@
 
-We provide an implementation of our recurrent clustering algorithm in a tool called \textsc{ML4HS}, which consists of a loose collection of components shown in Figure \ref{fig:ml4hs}.
-This arrangement makes it easy to swap out parts for experimentation.
-In the following, we describe the custom components in the order they appear in the diagram.
+We provide an implementation of our recurrent clustering algorithm in a tool called \textsc{ML4HS}.
+We obtain Core ASTs from Haskell packages using a plugin for the GHC compiler, which emits a serialised version of each Core definition as it is being compiled.
+This approach is more robust than, for example, parsing source files, since it avoids the complications of preprocessors and language extensions altering the syntax.
 
-\begin{figure}
-  \centering
-  \tikzstyle{block} = [rectangle, draw, rounded corners]
-  \tikzstyle{container} = [rectangle, draw, rounded corners]
-  \tikzstyle{line} = [draw, -latex']
-  \colorlet{shade}{rgb:black,1;white,4}
+A post-processing stage determines which Core definitions can be used by \qspec{}, based on their type, visibility (whether they are encapsulated inside their module or visible to importers), etc.
+For each definition, we simply use the GHCi interpreter to check if calling \qspec{} with that argument is well-typed.
 
-  \begin{tikzpicture}[node distance=2cm]
-    \node [block] (hackage) {Hackage};
+These definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}.
+Since the features associated with each identifier may vary between iterations, we leave the raw identifiers in the vector so their features can be extracted in the correct context.
 
-    \node [block, below of=hackage] (cabal) {
-      \begin{tikzpicture}[node distance = 2cm]
-        \node (caballabel) {Cabal};
-        \node [block, anchor=north west] at (caballabel.south) (ghc) {
-          \begin{tikzpicture}[node distance = 2cm]
-            \node (ghclabel) {GHC};
-            \node [block, anchor=north west, fill=shade] at (ghclabel.south) (plugin) {AST Plugin};
-          \end{tikzpicture}
-        };
-      \end{tikzpicture}
-    };
+The iterative algorithm \ref{alg:recurrent} is a simple shell script, which calls the Weka machine learning system to perform k-means clustering and associate each Core definition with a cluster number.
+These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and so on until all SCCs have been processed.
 
-    \node [block, below of=cabal, fill=shade]  (sorting) {Sorting};
-
-    \node [block, below of=sorting, fill=shade] (clustering) {
-      \begin{tikzpicture}[node distance = 2cm, auto]
-        \node (clusterlabel) {Recurrent Clustering};
-        \node [block, anchor=north west, fill=shade] at (clusterlabel.south west) (fe) {Feature Extraction};
-        \node [block, anchor=west, fill=white] at ([xshift=2em]fe.east) (weka) {Weka};
-
-        \path [line] (fe)   -- (weka);
-        \path [line] (weka) -- (fe);
-      \end{tikzpicture}
-    };
-
-    \node [block, below of=clustering, fill=shade] (mlspec) {
-      \begin{tikzpicture}[node distance = 2cm, auto]
-        \node (mlspeclabel) {MLSpec};
-        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {QuickSpec};
-      \end{tikzpicture}
-    };
-
-    \node [block, below of=mlspec] (user) {User};
-
-    \path [line] (hackage)    -- (cabal);
-    \path [line] (cabal)      -- (sorting);
-    \path [line] (sorting)    -- (clustering);
-    \path [line] (clustering) -- (mlspec);
-    \path [line] (mlspec)     -- (user);
-  \end{tikzpicture}
-  \caption{Components of the ML4HS theory exploration system. Custom components are shaded, arrows indicate data flow.}
-  \label{fig:ml4hs}
-\end{figure}
-
-\subsection{\textsc{AST Plugin}}
-\label{sec:astplugin}
-
-The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core.
-It also includes a \emph{renaming} transformation, which resolves global identifiers into a canonical form.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
-
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \hspec{} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
-
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
-
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities will be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
-
-GHC provides a plugin mechanism for manipulating Core during a build, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which emits a serialised version of each Core definition to the console (to satisfy the type system, it also implements a dummy ``optimisation'' which returns the Core unchanged).
-
-Compared to Haskell, Core is a much simpler language and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core difficult to handle are:
-
-\begin{description}
-  \item{Type variables}: Parametric polymorphism (described in more detail in \S \ref{sec:haskelldesc}) can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special abstraction form $\Lambda$, distinct from the $\lambda$ used for values.
-Core only has one abstraction form, \CLam, for both types and values.
-This alters function properties like arity.
-
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core, which is guaranteed not to conflict since this name would be invalid in Haskell.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
-
-  \item{Violating encapsulation}: Although Haskell allows names to be \emph{private} to a module, when compiling Core we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
-\end{description}
-
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to reference that name within the GHCi interpreter.
-Names with the above problems will cause an error, and are discarded.
-
-The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \cite{mitchell2008hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which makes it trivial to look up references in the database.
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
-
-\subsection{Toplogical Sorting}
-
-As described in \S \ref{sec:symbolstofeatures}, we must topologically sort the output of \textsc{AstPlugin} in order for our recurrent clustering to be well-founded.
-Since our database keys (containing the package, module and binding names, as described above) match our representation of non-local references, it is simple to walk each syntax tree to obtain the set of references it makes.
-In addition, the resulting set of (identifier, list-of-referenced-identifiers) pairs exactly matches the (vertex, list-of-successor-vertices) format used to represent directed graphs by the popular \hs{containers} library which ships with GHC.
-This provides an implementation of topological sort for strongly connected components, which we use as-is.
-A simple shell script loops through these SCCs, invoking the recurrent clustering component for each and appending the resulting features and clusters to the database.
-
-\subsection{Feature Extraction}
-
-The implementation of our feature extraction algorithm is a rather direct translation of the description given in \S \ref{sec:contributions} into Haskell.
-We parse the s-expressions generated by \textsc{AstPlugin} into algebraic data types which correspond directly to the definitions in Figure \ref{fig:coresyntax}; this is routine, so we omit the details for brevity.
-Similarly, we can represent rose trees with a datatype corresponding to the definition given in \S \ref{sec:expressionstovectors}:
-
-\begin{haskell}
-data RoseTree = Node Feature [RoseTree]
-\end{haskell}
-
-For simplicity we use the representation \hs{Feature = Int}, as we do not have fractional values.
-Since we represent the symbols $expr$, $id$, etc.
-from Figure \ref{fig:coresyntax} using different datatypes, we cannot write one big definition of $toTree$ or $\phi$ which works on all tokens.
-Each case shown in Figure \ref{fig:totree} and equations \ref{eq:feature}, \ref{eq:localfeature} and \ref{eq:globalfeature} appears in the implementation, although they are spread across several functions.
-
-To support looking up local identifiers, our implementation of $toTree$ takes a context as argument, extending it as required.
-As an example of the complexity this adds, here is the \CCase\ branch of $toTree$:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=0pt, xrightmargin=0pt]
-toTree :: Context -> Expr -> RoseTree
-toTree ctx x = case x of
-  ...
-  Case e l as -> Node fCase (toTree ctx e : map (toTreeAlt (l:ctx)) as)
-  ...
-\end{lstlisting}
-
-Breaking this down we can see \hs{fCase} representing the value of $\feature{\CCase}$, and a list of sub-trees defined in parentheses.
-The first subtree is a straightforward recursive call in an unmodified context: \mbox{\hs{toTree ctx e}}.
-The rest of the list is formed by applying the function \mbox{\hs{toTreeAlt (l:ctx)}} to each element of the list \hs{as} of \CAlt\ clauses.
-
-The \hs{toTreeAlt} function contains those cases of $toTree$ which handle symbols in $alt$.
-We prepend the identifier \hs{l} to the context, to get the extended context \hs{l:ctx}.
-This is because \hs{l} will be bound the value of \hs{e}, in order to avoid re-computing its value several times.
-
-The other clauses are handled in a similar way.
-The trickiest is the \CLet\ clause, since the local identifiers aren't directly available; we must extract them from their \CRec, \CNonRec\ and \CBind\ wrappers first, which we do using helper functions.
-
-As shown above, the values from equations \ref{eq:feature} are encoded directly in $toTree$.
-For $phi(l \in \mathcal{L})$ we use standard Haskell functions to look up the required indices in the context:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
-phiL :: Context -> Local -> Feature
-phiL ctx x = case elemIndex x ctx of
-  Nothing -> error (concat ["Local '", show x,
-                            "' not in context '",
-                            show ctx, "'"])
-  Just i  -> (2 * alpha) + i
-\end{lstlisting}
-
-As explained in \S \ref{sec:symbolstofeatures}, local identifiers should always exist in the context. If this precondition doesn't hold, we abort the program with an error rather than continuing.
-
-Global identifiers are kept as-is until we have access to the clusters from the last iteration. This takes place outside Haskell, using the \hs{jq} data processing tool.
-
-Our implementation of $level$ exactly matches equation \ref{eq:level}:
-
-\begin{haskell}
-level :: Int -> RoseTree -> [[Feature]]
-level 1 (Node f _)  = [f]
-level n (Node _ ts) = concatMap (level (n-1)) ts
-\end{haskell}
-
-To produce feature vectors, we do not directly construct the matrix; instead we generate the rows and concatenate them together in one step, using the \hs{concatMap} function:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
-featureVec :: Expr -> [Feature]
-featureVec e = concatMap (\m -> pad (level m tree)) [1..r]
-  where tree   = toTree [] e
-        pad xs = take c (xs ++ repeat 0)
-\end{lstlisting}
-
-By providing \hs{featureVec} with the latest set of clusters, read from the \textsc{AstPlugin} database, we turn Core expressions into feature vectors, which are appended to the database.
-
-We use the Weka system to perform our k-means clustering, as it is widely used, including by ML4PG.
-We select all feature vectors from our database, and write them in CSV format for Weka to process.
-The Weka CLI command is invoked, which appends a cluster number to each of these feature vectors; we read these off and append them to the database.
-As long as more SCCs remain unprocessed, we keep looping this process, using the database to communicate between the feature extraction and clustering phases.
-
-\subsection{\textsc{MLSpec}}
-\label{sec:mlspec}
-
-We cannot supply these clusters as-is to \qspec{}, since it must be provided with a \emph{signature}.
-These are constructed by our \textsc{MLSpec} tool, using information from the \textsc{AstPlugin} database.
-Tasks performed by \textsc{MLSpec} include:
+Once the recurrent clustering is complete, we split the definitions based on their final clusters, and use these clusters to generate strings of Haskell code for invoking \qspec{}.
+To construct a \qspec{} signature from these definitions, we must:
 
 \begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \qspec{} is unable to evaluate these expressions without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
+  \item{Monomorphise}: If a value has polymorphic type, e.g.
+\hs{equal :: forall t.
+t -> t -> Bool}, we must choose a concrete representation to use (in this case for \hs{t}), in order to know which random generator to use.
+We take the approach used in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
+Any values where this is invalid, e.g.
+if there are incompatible class constraints (e.g.
 \hs{forall t.
-IsString t => t}).
+IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will have already been discarded in our post-processing stage after extracting the ASTs.
 
@@
 Since \textsc{MLSpec} builds signatures \emph{dynamically}, depending on the cluster information it is given, we do not know what packages it may need.
-To work around this problem, \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have built called \texttt{nix-eval}.
-This provides an \texttt{eval} function, like those commonly found in dynamic languages such as Python and Javascript, for evaluating Haskell expressions.
-The key feature of \texttt{nix-eval} is that these Haskell expressions may reference packages that are not installed on the system.
-When such expressions are evaluated, these packages will be automatically downloaded and installed into a sandbox using the Nix package manager, and GHC will be invoked in this sandbox to perform the evaluation.
+To work around this problem, we evaluate these strings of Haskell using a custom library called \texttt{nix-eval}.
+This uses the Nix package manager to obtain all of the required packages and make them available to GHC.
 

### 2016-03-01 CICM2016/intro.tex
Cut out a load

@@
 
-As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis.
-For example, in pure functional programming all values are \emph{immutable}: defined once and never changed.
-Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code.
-Similarly, the results of pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results.
-By making state implicit in this way, powerful type systems can be used to constrain the behaviour of programs, and to give a rich, composable structure to data.
+Functional programming claims to be more ``mathematically tractable'' than other ``conventional'' approaches, due to properties like \emph{referential transparency} and \emph{immutability}, which allow equational reasoning without many of the complications caused by side effects and strict evaluation order \cite{hughes1989functional}.
+Languages such as Haskell combine these advantages with static type systems to restrict the possible values of every expression, allowing powerful analysis tools such as property checkers \cite{claessen2011quickcheck} and theorem provers \cite{rosen2012proving} to be developed.
 
-Whilst use of pure functional programming languages, like Haskell and Idris, is relatively rare, their features are well suited to common software engineering practices like \emph{unit testing}; where tasks are broken down into small, easily-specified ``units'', and tested in isolation for a variety of use-cases.
-Functional ideas are thus spreading to mainstream software engineering in a more dilute form; seen, for example, in the recent inclusion of first-class functions in Java \cite{gosling2015java} and C++ \cite{willcock2006lambda}.
-
-Functional programming is also well suited to less-widespread practices, such as \emph{property checking} (as popularised by \qcheck{}) and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet can be prohibitively expensive.
-Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.
+Whilst such tools can support or refute user-provided properties, we are interested in the related problem of \emph{discovering} such properties automatically: the task of \emph{theory exploration}.
+Whilst clever algorithms are used to guide their search, existing theory exploration tools, including those for Haskell, ultimately rely on enumerating all (well-typed) expressions.
+This limits their usefulness, as it is only feasible to explore a small collection of functions at a time, relying on the user to select promising combinations.
+We investigate the use of machine learning methods to perform this selection automatically, removing the need for user intervention whilst maintaining the ability to discover desirable properties.
 
@@
 
-We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and describe the \qspec{} theory exploration system (\S \ref{sec:theoryexploration}).
-We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}).
+We describe our treatment of Haskell expressions in \S \ref{sec:haskell} and describe the \qspec{} theory exploration system we build upon in \S \ref{sec:theoryexploration}.
 We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details in \S \ref{sec:implementation}.

### 2016-03-01 CICM2016/related.tex
Cut out a load

--- 2015-12-10 TransferReport/related.tex
@@
 
-\subsection{Haskell}
-\label{sec:haskelldesc}
+\subsection{Theory Exploration}
 
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little explanation was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widely used language like Java.
-Here we discuss the relevant language features from a high-level, which motivated our choice:
+We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements.
+This general idea is quite old, with .
+The current generation of theory exploration tools trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
+In such a setting, most of the automation focuses on bookkeeping (such as type-checking, managing scope, etc.) and more traditional automated theorem proving.
 
-\begin{description}
-
-\item{Functional}: All control flow in Haskell is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for theory exploration.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-This puts Haskell in a rather special position with regards to the discovery of interesting theorems; namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence theory exploration would be unlikely to discover unexpected properties.
-
-Other empirical advantages to studying Haskell, compared to other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of theory exploration systems such as \hspec{}, and related tools which we may be able to re-use, including conjecture generators like \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
+Subsequent systems have increased the level of automation, and have shown promise in tasks such as lemma discovery.
+As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012} \cite{johansson2009isacosy} \cite{Hipster}.
+Since \qspec{} is used for conjecture generation in many of those systems, they can also benefit from our work with machine learning.
 
@@
 
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering} (or, \emph{premise selection}, when viewed as the \emph{addition} of relevant information to an initially-empty problem).
-This is the core idea behind our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time.
+The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}.
+The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the input size of a task by removing redundant or irrelevant information from the input.
+Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is the essentially the same idea being applied in a novel setting.
 
@@
 
-The initial approach taken by Sledgehammer, known as \textsc{MePo} (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal to be proved, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
-Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}, some of which may be directly applicable in the context of theory exploration.
+Various approaches have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}, though the original approach is based on the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics).
 
@@
 
-\iffalse
-We could expand this a bit, e.g. talking about how we both use Weka, etc.
-\fi
-
 \subsection{Feature Extraction}
@@
 This can lead to nested expressions of arbitrary depth, which are difficult to compare in numerical ways.
-One solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
+One common approach to this problem is to represent such structures as \emph{sequences}.
+\emph{Recurrent neural networks} (RNNs) are a popular choice for processing sequences, especially when combined with mechanisms such as \emph{long short-term memory} (LSTM) for preserving information across long sequences \cite{hochreiter1997long}.
+Such systems have been used, for example, to parse and execute computer programs \cite{zaremba2014learning}.
+However, learning to parse sequences seems inefficient considering that we already have correctly-parsed ASTs.
 
-The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger.
-We use this to make our matrices a uniform size, borrowing the idea from ML4PG.
-Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
-In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.
-
-Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian'').
-This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness.
-This allows us to truncate more aggressively than if the leaves were most significant.
-
-By modelling our inputs as points in high-dimensional spaces, we can consider feature extraction as projection into a lower-dimensional space (known as \emph{dimension reduction}).
-Truncation is a trivial dimension reduction technique; more sophisticated projection functions consider the \emph{distribution} of the input points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-Techniques such as \emph{principle component analysis} (PCA) can be used to find these hyperplanes, but unfortunately require their inputs to already have a fixed, integer number of dimensions.
-In the case of our recursive expressions (which we may consider to have \emph{fractal} dimension), we would need another pre-processing stage to satisfy this requirement.
-
-There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise.
-Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.
-
-One example of learning from variable-size input is to use \emph{recurrent neural networks} (RNNs).
-These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' NNs, allowing state to persist between observations.
-In this way, each data point can be divided into a sequence of arbitrary length, for example an s-expression, and fed into an RNN one token at a time for processing.
-
-Unfortunately RNNs are difficult to train.
-The standard way to train NNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
-
-Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward NN to learn recursive distributed representations \citep{pollack1990recursive} which correspond to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which we can use to encode sub-trees when the branching factor of nodes is not  uniform \citep{kwasny1995tail}.
-More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
-
-A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
-Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
-
-\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}, which could be used to learn more subtle relations between expressions than recurrent clustering alone.
-
-\subsection{K-Means}
-
-We use the Weka tool to perform k-means clustering \citep{Holmes.Donkin.Witten:1994}, since we are more concerned with the application of feature extraction to Haskell and its use in theory exploration, rather than precise tuning of learning algorithms.
-Since k-means is a standard method, there are many other implementations available.
-More interestingly, there are many other clustering algorithms we could use, such as \emph{expectation maximisation} \footnote{In fact, k-means is very similar to expectation-maximisation, as it alternates between an \emph{expectation step} (finding the mean value of each cluster) and a \emph{maximisation step} (assigning points to the cluster they're most similar to; or alternatively, \emph{minimising} the distance of each point to the centre of its cluster, as per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have shown little difference in their results; in effect, the quality of our features is the bottleneck to learning, so there is no reason to avoid a fast algorithm like k-means.
-
-In any case there are many conservative improvements to the standard k-means algorithm, which could be applied to our setup.
-For example, a more efficient approach like \emph{yinyang k-means} \citep{conf/icml/DingZSMM15} could make larger input sizes more practical to cluster, especially since recurrent clustering invokes k-means many times.
-The \emph{k-means++} approach \citep{arthur2007k, bahmani2012scalable} can be used to more carefully select the ``seed'' values for the first timestep, and the \emph{x-means} algorithm \citep{pelleg2000x} is able to estimate how many clusters to use (our \emph{final} clusters should be tuned to maximise the performance of the subsequent theory exploration step, but x-means could still be useful in the recurrent clustering steps).
+Whilst neural networks have been applied directly to recursive structures \cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more popular approach is to use \emph{kernel methods} \cite{bakir2007predicting}.
+These are promising as a more principled alternative to our current hand-crafted translation of ASTs to vectors.

### 2016-03-01 CICM2016/appendices.tex
Added appendices

\iffalse
%\iftrue
\section*{Appendix: Syntax and Translation}

\begin{figure}
  \begin{equation}
    \begin{split}
      expr\    \rightarrow\ & \CVar\ id                          \\
                         |\ & \CLit\ literal                     \\
                         |\ & \CApp\ expr\ expr                  \\
                         |\ & \CLam\ \mathcal{L}\ expr           \\
                         |\ & \CLet\ bind\ expr                  \\
                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                         |\ & \CType                             \\
      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                         |\ & \CGlobal\      \mathcal{G}         \\
                         |\ & \CConstructor\ \mathcal{D}         \\
      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                         |\ & \CLitStr\ \mathcal{S}              \\
      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                         |\ & \CLitAlt\ literal                  \\
                         |\ & \CDefault                          \\
      bind\    \rightarrow\ & \CNonRec\ binder                   \\
                         |\ & \CRec\ [binder]                    \\
      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
    \end{split}
  \end{equation}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers \\
    $\mathcal{D}$ & constructor identifiers
  \end{tabular}

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
  \label{fig:coresyntax}
\end{figure}

\begin{figure}
  \begin{align*}
    toTree(e) &=
    \begin{cases}
      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
      (\feature{\CType})                                                & \text{if $e = \CType$} \\
      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
    \end{cases}
  \end{align*}
  \caption{Transforming Core expressions of Figure \ref{fig:coresyntax} to rose trees. The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support. We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.}
  \label{fig:totree}
We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
\end{figure}

\fi

### 2016-03-02 CICM2016/appendices.tex
Updates

@@
-\iffalse
-%\iftrue
-\section*{Appendix: Syntax and Translation}
+%\iffalse  % WITHOUT APPENDICES
+\iftrue    % WITH APPENDICES
 
-\begin{figure}
-  \begin{equation}
+\appendix
+
+\section{Core Syntax}\label{sec:core}
+
+We use a simplified syntax for the GHC Core language, given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively).
+We mechanically translate the full GHC Core language into this sub-set, which does not contain types, prior to performing recurrent clustering.
+
+%\begin{figure}
+  \begin{equation}\label{fig:coresyntax}
     \begin{split}
@@
   \end{tabular}
+%\end{figure}
 
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
+\section{Rose Trees}\label{sec:rosetree}
 
-\begin{figure}
-  \begin{align*}
+The $toTree$ function shown below transforms Core expressions, described in Appendix \ref{sec:core} to rose trees.
+We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
+$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
+
+The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
+for $\CRec$ we apply $toTree$ to each $e_i$.
+We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
+We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.
+
+%\begin{figure}
+  \begin{align*}\label{fig:totree}
     toTree(e) &=
@@
   \end{align*}
-  \caption{Transforming Core expressions of Figure \ref{fig:coresyntax} to rose trees.
-The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
-for $\CRec$ we apply $toTree$ to each $e_i$.
-We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
-We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.}
-  \label{fig:totree}
-We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
-$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
-$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
-\end{figure}
+
+%\end{figure}
 

### 2016-03-02 CICM2016/background.tex
Updates

@@
 
-plus :: Nat -> Nat -> Nat
-plus    Z  y = y
-plus (S x) y = S (plus x y)
-
-mult :: Nat -> Nat -> Nat
-mult    Z  y = Z
-mult (S x) y = plus y (mult x y)
-
 odd :: Nat -> Bool
@@
   \begin{small}
-    \underline{\texttt{plus}}
-    \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
-                                               (App (App (Var (Global "plus"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local "y"))))
-                                          "x")))
-    \end{verbatim}
-
-    \underline{\texttt{mult}}
-    \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
-                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                    (Var (Local  "y")))
-                                               (App (App (Var (Global "mult"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local  "y"))))
-                                          "x")))
-    \end{verbatim}
-
     \underline{\texttt{odd}}
@@
 \begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
+  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{odd}, \texttt{even}\}
 \end{align*}
@@
 \begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
+  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
+                     & \texttt{S}\ c,\ \texttt{odd Z},\ \texttt{odd}\ a,\ \dots ]
 \end{align*}
 
-Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments.
+Notice that functions such as \hs{odd} and \hs{even} are valid terms, despite not being applied to any arguments.
 In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.
 
-\begin{figure}
-  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
-  \begin{verbatim}
-                  plus a b = plus b a
-                  plus a Z = a
-         plus a (plus b c) = plus b (plus a c)
-                  mult a b = mult b a
-                  mult a Z = Z
-         mult a (mult b c) = mult b (mult a c)
-              plus a (S b) = S (plus a b)
-              mult a (S b) = plus a (mult a b)
-         mult a (plus b b) = mult b (plus a a)
-                 odd (S a) = even a
-            odd (plus a a) = odd Z
-           odd (times a a) = odd a
-                even (S a) = odd a
-           even (plus a a) = even Z
-          even (times a a) = even a
-plus (mult a b) (mult a c) = mult a (plus b c)
-  \end{verbatim}
-  \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
-  \label{fig:qspecresult}
-\end{figure}
-
-These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
+These terms will be grouped into four classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
+As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations such as $odd (S a) = even a$ and $even (S a) = odd a$ are conjectured.
 \fi

### 2016-03-02 CICM2016/contributions.tex
Updates

@@
 
-\begin{figure}
-    \begin{equation*}
-      (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
-     \end{equation*}
-     \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
-     \label{fig:vectorexample}
-\end{figure}
-
 These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}.
-Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, with any required padding added on the right.
-These matrices are either padded with $0$ or truncated, to fit a fixed size $r \times c$.
+Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, read left-to-right, followed by any required padding.
+These matrices are then either truncated, or padded with rows (on the bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.
 
 \begin{sloppypar}
-Feature vectors are then simply the concatenation of matrix rows $\mathbf{M}_1 \cat \mathbf{M}_2 \cat \dots \cat \mathbf{M}_r$, as shown in Figure \ref{fig:vectorexample}:
+Finally, matrices are turned into vectors by simply concatenating the rows from top to bottom, hence the matrix from Figure \ref{fig:matrixexample} will begin $(\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, \dots$.
 \end{sloppypar}
-
-\begin{equation}
-  featureVec(e) = pad(level(1, toTree(e))) \cat \dots \cat pad(level(r, toTree(e)))
-\end{equation}
 
@@
 renaming an identifier shouldn't affect the feature).
-To do this, the $toTree$ function maintains a \emph{context} as it recurses through expressions (we elided this from Figure \ref{fig:totree} for clarity).
-The context is a list of the local identifiers which are in scope, prepended as they are encountered.
-The context is initially empty, and only extended when $toTree$ calls itself recursively.
-
-For example, when calculating $toTree(\CLam\ i\ e)$ in context $\vect{x}$, we make the recursive call $toTree(e)$ in the context of \emph{$\vect{x}$ prepended with $i$} (i.e.
-$(i) \cat \vect{x}$).
-Similar extensions of the context are performed in the cases of \CBind, \CCase\footnote{The $\mathcal{L}$ value in a \CCase\ expression is bound to the expression being matched against; a technical detail to preserve sharing.}, \CAlt\ (which may introduce an arbitrary number of local identifiers) and \CLet\ (where identifiers are taken from occurrences of \CBind\ in the first expression).
-
-Well-formed Haskell declarations do not contain free variables (or, equivalently, free variables are encoded as global identifiers rather than local identifiers).
-Hence if we apply $toTree$ to (the Core expression corresponding to) such declarations we are guaranteed that $l \in \mathcal{L}$ will appear in the context whenever we encounter $\feature{l}$.
-In which case we define:
+To do this, we use the \emph{de Bruijn index} of the identifier \cite{de1972lambda}, denoted $i_l$:
 
 \begin{equation} \label{eq:localfeature}
-  \feature{l} = i + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
+  \feature{l} = i_l + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
 \end{equation}
 
-Where $i$ is the index of the first occurrence of $l$ in the context.
-This tells us how many binders a variable is nested inside, relative to the point it was introduced.
-This is invariant under renaming, as we wanted, and is known as the \emph{de Bruijn index} of $l$.
 We again use $\alpha$ to separate these features from those of other constructs.
 
-Since the $toTree$ function discards numerals, strings and constructor identifiers, the only remaining case is global identifiers $\mathcal{G}$.
+Since numerals, strings and constructor identifiers are discarded during conversion to rose trees, the only remaining case is global identifiers $\mathcal{G}$.
 Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers.
@@
 
-Where $\vect{C}$ are our clusters (in some arbitrary order).
-This is where the recurrent nature of the algorithm appears: to determine $\vect{C}$ we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.
+Where $C_i$ are our clusters (in some arbitrary order).
+This is where the recurrent nature of the algorithm appears: to determine the contents of $C_i$, we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.
 
@@
 To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.
-If an identifier cannot be found in any cluster, it must appear in the same SCC as the expression we are processing; hence we use the constant feature value $f_{recursion}$.
+If an identifier cannot be found amongst those clustered so far, it must appear in the same SCC as the expression we are processing; hence we give that identifier the constant feature value $f_{recursion}$.
 

### 2016-03-02 CICM2016/future.tex
Updates

@@
 Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further.
-In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique.
-Below, we elaborate on the details, background and motivation for these choices.
 
-\subsection{Clustering Extensions}
-\label{sec:preprocessing}
-
-The most glaring omission in our algorithm is its disregard for types.
-By ignoring types, not only are we losing valuable information about expressions, but we also lose the ability to distinguish between constructors.
-This is because a constructor, like \hs{True} or \hs{Just}, has no internal structure; it is just a token.
-The distinguishing features of constructors are their types, which not only tell us which data type they construct, but also their arity, the types of their arguments, etc.
-
-Our algorithm closely follows that of ML4PG, which \emph{does} support types.
-This is handled by populating matrix cells with tokens \emph{and} their types.
+The most obvious next step for our algorithm is to incorporate types.
+Types contain valuable information about an expression, and allow us to distinguish between constructors.
+Our algorithm closely follows that of ML4PG, which \emph{does} support types, by populating matrix cells with tokens \emph{and} their types.
 Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).
-
-One partial solution would be leave most Core expressions without types, but to include them for non-local identifiers (i.e.
-globals and constructors), which we can look up in a database.
-In fact, our ML4HS framework already includes such type information in its database, alongside the Core syntax trees.
-Integrating this information into our algorithm is the next logical step.
 
@@
 This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.
-
-With more expressive features, it may also be useful to experiment with more powerful learning algorithms.
-An interesting possibility is to add a feedback loop between the theory exploration phase and the clustering phase, to more directly base the similarity of expressions on whether they (are predicted to) occur together in equations.
-
-\subsection{Theory Exploration Extensions}
-
-Our current approach is a rather conservative change to the existing theory exploration approaches, as it is essentially a wrapper around \qspec{}.
-There is potential for more radical changes to be made, which alter the search process itself.
-
-\subsubsection{Variable Instantiation}
-
-\qcheck{} is certainly the most popular property checker for Haskell, which motivates its use in \qspec{} to instantiate variables to random values.
-However, this task of finding type inhabitants has also been solved in many other ways, which may be worth investigating in place of \qcheck{} (or perhaps even as part of an ensemble).
-
-The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly.
-Whilst this does not make \textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage is that it may use much less memory, as the generated values are built up incrementally.
-In contrast, \qcheck{} may generate very large values; in particular, generating tree structures na\"{\i}vely can cause them to grow exponentially.
-For example, here is a potential generator for \hs{RoseTree}s:
-
-\begin{haskell}
-genRoseTree = do f        <- arbitrary
-                 subtrees <- listOf genRoseTree
-                 return (Node f subtrees)
-\end{haskell}
-
-The \hs{listOf genRoseTree} call will return a list of arbitrary length, where each element is generated by \hs{genRoseTree}.
-This allows an arbitrary number of recursive calls to be made for each invocation of \hs{genRoseTree}, which will quickly exhaust the resources of any machine.
-Whilst such problems may be anticipated, or quickly spotted, in a property checking setting, this can be more difficult for our automated approach.
-For example, if a type does not have a generator available, we cannot use a library like \hs{derive} to create one automatically, as it suffers from this na\"{\i}vity problem.
-
-A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \cite{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded.
-This may narrow down our search procedures greatly, especially when predicates are involved.
-\qcheck{} allows predicates to restrict the values it tests with, and hence allows \emph{conditional} equations to be discovered.
-However, its implementation uses a simple rejection sampling technique: values are generated just as if the predicate were not there, and afterwards are filtered to reject any which do not satisfy the predicate.
-This makes it difficult to use very specific predicates, as it is unlikely that many of our random samples will exactly match our criteria.
-On the other hand, \textsc{Lazy SmallCheck} will focus its search on exactly those parts of the datastructure which are checked by the predicate, as those are the parts being forced to evaluate.
-This makes it much more likely that we will find values which satisfy the predicate, allowing us to effectively explore more specific conditional properties.
-
-Other approaches to generating inhabitants include \textsc{Djinn} \cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator).
-\textsc{MuCheck} \cite{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g.
-changing the order of pattern match clauses); whilst not as exhaustive as the other approaches, mutating existing values in this way is claimed to yield values which correspond more closely to what a programmer might write.
-This is an interesting possibility for focusing theory exploration on to more ``realistic'' areas of the search space, and hence avoiding some of the more useless or bizarre expressions that random search and enumeration may produce.
-
-In fact, the database generated by our \textsc{AstPlugin} may prove helpful in generating values, since its type information can be fed to a tool like \textsc{Djinn} to discover chains of function applications for building values, which would be particularly useful in cases where constructors are private, like in our email example.
-This is similar to the \textsc{Hoogle} tool, but also offers the ability to use dependency information to avoid potentially infinite recursion.
-
-The Core syntax trees in our database could also be used to generate theories for automated theorem provers.
-\hspec{} currently uses the GHC API to transform Core within its own process, however that approach suffers from the problems described in \S \ref{sec:astplugin}.
-
-\subsubsection{Interestingness}
-\label{sec:interestingness}
-
-If we do succeed in producing a fast theory exploration system, which chooses productive combinations of terms and finds a large number of properties, we encounter the problem of managing the output: finding the needles we are interested in among the haystack of trivialities and coincidences.
-
-This is governed by the ``interestingness'' criteria of the theory exploration system: what to keep and what to discard, and even what areas of the search space to prioritise.
-\qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is very simple: we discard equations which are direct consequences of others, and keep all the rest.
-Different, and more sophisticated notions of interestingness have been widely studied in other fields, which may be applied in the context of theory exploration.

### 2016-03-02 CICM2016/check.sh
Modularise tests

@@ -1,29 +1,112 @@
 #!/usr/bin/env bash
 
-ERR=0
+ERR=0 # Indicates failure, will be our exit code
 
-TODOS=""
-REFS=""
-CITES=""
-MACROS=""
-LABELS=""
-FIGS=""
-CITEP=""
-for FILE in *.tex
-do
-    if grep "TODO[^{]" < "$FILE" > /dev/null
-    then
-        NUM=$(grep -c TODO < "$FILE")
-        TODOS=$(echo "$TODOS"; echo "Found $NUM undated TODOs in $FILE")
+# "Top-level" tests, invoked directly
+
+function checkLatex {
+    FAIL=0 # A checkLatex-specific version of ERR
+    for FUNC in checkLatexUndatedTodos checkLatexDatedTodos checkLatexRefs \
+                checkLatexCites checkLatexMacros checkLatexLabels          \
+                checkLatexFigs checkLatexCitep
+    do
+        mapFiles "$FUNC" || FAIL=1
+    done
+    return "$FAIL"
+}
+
+function checkRender {
+    render || {
         ERR=1
+        echo "Couldn't render PDF" >> /dev/stderr
+        return 1
+    }
+    [[ -e report.pdf ]] || {
+        ERR=1
+        echo "report.pdf wasn't created" >> /dev/stderr
+        return 1
+    }
+}
+
+function checkBibtex {
+    if [[ -z "$(render | getBibStart)" ]]
+    then
+        ERR=1
+        echo "Bibtex didn't run?" >> /dev/stderr
+        return 1
+    fi
+}
+
+function checkWarnings {
+    BIBSTART=$(render | getBibStart)
+
+    [[ -z "$BIBSTART" ]] && return 1
+
+    WARNINGS=$(render | tail "-n$BIBSTART" | grep -ai "warning")
+
+    # Citations take two passes, so double-check
+    CITES=$(echo "$WARNINGS" | grep 'Citation `'  | sed -e 's/.*`\(.*\)'"'.*/\1/g")
+    while read -r CITE
+    do
+        grep "^@[^{]*{$CITE,$" < ~/Writing/Bibtex.bib > /dev/null ||
+            {
+                ERR=1
+                echo "Citation '$CITE' missing" >> /dev/stderr
+            }
+    done < <(echo "$CITES")
+
+    # Check for non-citation warnings
+    NONCIT=$(echo "$WARNINGS" | grep -v 'Citation `'                      |
[…]

### 2016-03-02 CICM2016/haskell_example/src/Eval.hs
Added haskell_example

@@ -1,0 +1,35 @@
+module Eval where
+
+fact n = if n == 0
+            then 1
+            else n * fact (n - 1)
+
+helperFact n a = if n == 0
+                    then a
+                    else helperFact (n - 1) (a * n)
+
+factTail n = helperFact n 1
+
+power n = if n == 0
+             then 1
+             else 2 * power (n - 1)
+
+helperPower n a = if n == 0
+                     then a
+                     else helperPower (n - 1) (a + a)
+
+powerTail n = helperPower n 1
+
+fib n = if n == 0
+           then 0
+           else if n == 1
+                   then 1
+                   else fib (n - 1) + fib (n - 2)
+
+helperFib n j k = if n == 0
+                     then j
+                     else if n == 1
+                          then k
+                          else helperFib (n - 1) k (j + k)
+
+fibTail n = helperFib n 0 1

### 2016-03-02 CICM2016/haskell_example/src/Example.hs

@@ -1,0 +1,52 @@
+module Example where
+{-
+type Nat = [()]
+
+render :: Nat -> String
+render = show . length
+
+mult :: Nat -> Nat -> Nat
+mult x y = case x of
+                []     -> []
+                ():as  -> y
+                (a:as) -> y ++ mult as y
+
+mult_tail' :: Nat -> Nat -> Nat -> Nat
+mult_tail' x y acc = case x of
+                          []     -> acc
+                          (a:as) -> mult_tail' as y (y ++ acc)
+
+mult_tail :: Nat -> Nat -> Nat
+mult_tail x y = mult_tail' x y []
+-}
+{-
+factorial1 0 = 1
+factorial1 n = n * factorial1 (n-1)
+
+base      f = f 0    ==    f 1
+increases f = f 4    >     f 3
+divides   f = f 5    `div` f 4 == 5
+negative  f = f (-1) ==    f 1
+
+factorial2 0 = 1
+factorial2 n = let nPos = abs n
+                in nPos * factorial2 (nPos - 1)
+
+f1_tail :: Int -> Int
+f1_tail = go 1
+  where go acc 0 = acc
+        go acc n = go (n*acc) (n-1)
+
+f2_tail :: Int -> Int
+f2_tail = go 1
+  where go acc 0 = acc
+        go acc n = let nPos = abs n in go (nPos*acc) (nPos-1)
+
+f1 :: Int -> Int
+f1 = factorial1
+
+f2 :: Int -> Int
+f2 = factorial2
+
+foo n = -n
+-}

### 2016-03-02 CICM2016/haskell_example/src/Foo.hs

@@ -1,0 +1,8 @@
+module Foo where
+{-
+data Foo a = Foo a
+
+class Bar a
+
+instance Foo a => Bar (Foo a)
+-}

### 2016-03-02 CICM2016/haskell_example/src/QuickSpecExample.hs

@@ -1,0 +1,42 @@
+module QuickSpecExample where
+
+import Numeric.Natural
+import Test.QuickSpec
+{-
+i2N :: Int -> Natural
+i2N = fromInteger . toInteger
+
+sigma = [blind0 "()"      (),
+         blind0 "Z"       ([] :: [()]),
+         blind1 "length"  ((i2N . length) :: [()] -> Natural),
+         blind2 "++"      ((++) :: [()] -> [()] -> [()]),
+         --blind1 "reverse" (reverse :: [()] -> [()]),
+         blind2 "s"       ((:) :: () -> [()] -> [()]),
+         blind0 "F"       False,
+         blind0 "T"       True,
+         blind0 "0"       (0 :: Natural),
+         blind1 "S"       ((+ 1) :: Natural -> Natural),
+         blind2 "+"       ((+) :: Natural -> Natural -> Natural),
+         blind2 "*"       ((*) :: Natural -> Natural -> Natural),
+
+         vars ["b1", "b2", "b3"] True,
+         vars ["l1", "l2", "l3"] [()],
+         vars ["n1", "n2", "n3"] (0 :: Natural),
+         vars ["u1"]             (),
+
+         observer1 ((i2N . length) :: [()] -> Natural),
+         observer1 (id :: () -> ()),
+         observer1 ((`mod` 10) :: Natural -> Natural),
+         observer1 (id :: Bool -> Bool)]
+-}
+
+nat = [fun0 "True"  (True  :: Bool),
+       fun0 "False" (False :: Bool),
+       fun0 "Z"     (0     :: Natural),
+       fun1 "S"     ((1+)  :: Natural -> Natural),
+       fun1 "not"   (not   :: Bool    -> Bool),
+       fun1 "odd"   (odd   :: Natural -> Bool),
+       fun1 "even"  (even  :: Natural -> Bool),
+
+       vars ["x", "y", "z"] (0    :: Natural),
+       vars ["a", "b", "c"] (True :: Bool)]

### 2016-03-02 CICM2016/appendices.tex
More changes

@@
 
-We use a simplified syntax for the GHC Core language, given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively).
-We mechanically translate the full GHC Core language into this sub-set, which does not contain types, prior to performing recurrent clustering.
+The GHC Core language is based on \fc{}, and described in detail in \cite[Appendix C]{sulzmann2007system}.
+For our machine learning purposes we are mostly interested in the syntax of reducible expressions (representing Haskell of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
 
-%\begin{figure}
 \begin{equation}\label{fig:coresyntax}
@@
 \end{tabular}
-%\end{figure}
+
+The full Core language emitted by GHC (as of version 7.10.2, the latest at the time of writing) is translated automatically to this simplified form prior to recurrent clustering.
+Our major restriction is to ignore type-level entities (such as datatype definitions, explicit casts, and differences between types/kinds/coercions).
+Our implementation also handles several other forms of literal (machine words of various sizes, individual characters, etc.), but we omit them here for brevity as their treatment is similar to those of strings and numerals.
 
@@
 We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
-We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect on clustering real code is low.
+We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect of such ``magic values'' on clustering real code is low.
 

### 2016-03-02 CICM2016/background.tex
More changes

@@
 Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
-Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \cite[Appendix C]{sulzmann2007system}.
+To avoid the relatively large and complex syntax of Haskell proper, we will focus on an intermediate representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described in Appendix \ref{sec:core}.
 
-The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major restriction is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
-For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
 We will use quoted strings to denote names and literals, e.g.
@@
   \begin{haskell}\begin{verbatim}
-data Nat = Z
-         | S Nat
+data Bool = True | False
+
+data Nat = Z | S Nat
+
+not :: Bool -> Bool
+not True  = False
+not False = True
 
@@
   \end{haskell}
-  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
-\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
+  \caption{Haskell datatypes for booleans and natural numbers, followed by some simple function definitions. Note that \hs{odd} and \hs{even} are mutually recursive.}
   \label{fig:haskellexample}
@@
   \begin{small}
+    \underline{\texttt{not}}
+    \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              (Alt (DataAlt "True")  (Var (Constructor "False")))
+              (Alt (DataAlt "False") (Var (Constructor "True"))))
+    \end{verbatim}
+
     \underline{\texttt{odd}}
@@
   \end{small}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}.
+  \caption{Translations of functions in Figure \ref{fig:haskellexample} into our Core syntax, with fresh variables chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.
 Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
+The \texttt{"b"} variables are introduced to preserve sharing.}
   \label{fig:coreexample}
@@
 
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
+Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with common definitions for boolean and natural number datatypes (analogous to those in Haskell's standard library).
 The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
+Although Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
 It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
-
-Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}.
-GHC can represent these, but in this work we only consider reducible expressions (i.e.
-value-level bindings of the form \mbox{\hs{f a b ...
-= ...}}).
 
@@
 
-Such conjectures can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
+Such conjectures can be used in several ways: they can be presented to the user, sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
+As an example, giving the definitions from Figure \ref{fig:haskellexample} to \qspec{}.
+along with some suitable \hs{Nat} and \hs{Bool} variables, produces equations including \hs{not (not a) == a} and \hs{not (odd x) == even x}.
 
@@
 For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
+Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which can be explored quickly, yet have the potential to give rise to many equations.
 
@@
 Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{sec:relevance}.
-Relevance filtering is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by only considering clauses deemed ``relevant'' to the problem \cite{meng2009lightweight}.
 
-Despite the idea's promise, we cannot simply invoke existing relevance filter algorithms in our theory exploration setting.
-The reason is that relevance filtering is a supervised learning method, i.e.
-it would require a distinguished expression to compare everything against.
-Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{all} terms in a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
+Relevance is determined by comparing clauses to the target theorem, but theory exploration does not have such a distinguished term.
+Instead, we are interested in relationships between \emph{all} terms in a signature, and hence we need a different algorithm for considering the relevance of \emph{all terms} to \emph{all other terms}.
 
@@
 
-We use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, as this setup is used by ML4PG and ACL2(ml).
-For simplicity, we will cluster $n$ points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters, following the ``rule of thumb'' given in \cite[pp.
-365]{mardia1979multivariate}, and will use randomly-selected elements of our data set to determine the initial clusters (known as the Forgy method).
+Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, with randomly-selected input elements as the initial clusters.
+Rather than relying on the user to provide the number of clusters $k$, we use the ``rule of thumb'' given in \cite[pp.
+365]{mardia1979multivariate} of clustering $n$ data points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.

### 2016-03-02 CICM2016/future.tex
More changes

@@
 
-The most obvious next step for our algorithm is to incorporate types.
-Types contain valuable information about an expression, and allow us to distinguish between constructors.
-Our algorithm closely follows that of ML4PG, which \emph{does} support types, by populating matrix cells with tokens \emph{and} their types.
-Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).
+The most obvious next step is to incorporate types.
+Types contain valuable information about an expression, and would allow us to distinguish between constructors.
+Since our algorithm closely follows that of ML4PG, which \emph{does} support types, the only barrier is the practical issue of propagating annotations through all of the Core definitions.
 
-We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \cite{bengio2013representation}.
-This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.
+More speculative directions include the use of \emph{learned} representations, rather than our hand-crafted features \cite{bengio2013representation}.
+This would provide an interesting comparison, as well as being more robust in the face of language evolution.
+Another intriguing possibility is to extend the recurrent nature of our algorithm to make use of the discovered properties during clustering; for example, by treating the discovered equations as rewrite rules to reduce the ASTs prior to feature extraction.

### 2016-03-02 CICM2016/implementation.tex
More changes

@@
 For each definition, we simply use the GHCi interpreter to check if calling \qspec{} with that argument is well-typed.
+This information, along with type signatures, arity, etc.
+are stored alongside the Core definitions in JSON format.
 
-These definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}.
-Since the features associated with each identifier may vary between iterations, we leave the raw identifiers in the vector so their features can be extracted in the correct context.
+The definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}.
+Since the features associated with each identifier may vary between iterations (as the clusters change), we leave the raw identifiers in the vector so their features can be extracted in the correct context.
 
-The iterative algorithm \ref{alg:recurrent} is a simple shell script, which calls the Weka machine learning system to perform k-means clustering and associate each Core definition with a cluster number.
-These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and so on until all SCCs have been processed.
+We implement Algorithm \ref{alg:recurrent} as a simple shell script, which invokes Weka for clustering and associates each Core definition with a cluster number.
+These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and the process is repeated until all SCCs have been processed.
 
-Once the recurrent clustering is complete, we split the definitions based on their final clusters, and use these clusters to generate strings of Haskell code for invoking \qspec{}.
-To construct a \qspec{} signature from these definitions, we must:
+Once the recurrent clustering is complete, we generate a string of Haskell code for each of the final clusters, which will invoke \qspec{} with a suitable signature. This involves:
 
 \begin{itemize}
-  \item{Monomorphise}: If a value has polymorphic type, e.g.
+  \item{Monomorphising}: If a value has polymorphic type, e.g.
 \hs{equal :: forall t.
@@
 We take the approach used in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
-Any values where this is invalid, e.g.
-if there are incompatible class constraints (e.g.
+Any values where this is invalid, such as those with incompatible class constraints (e.g.
 \hs{forall t.
-IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will have already been discarded in our post-processing stage after extracting the ASTs.
+IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will not be included in the signature (this is checked at the AST post-processing stage).
 
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
+  \item{Qualifying}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
 There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules.
@@
 
-  \item{Variable definition}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).
+  \item{Appending variables}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).
 
@@
 \end{itemize}
+
+The equations resulting from evaluating these strings are collected and outputted as JSON values, to ease further processing (e.g. displaying in some form, sending to a theorem prover, etc.).

### 2016-03-02 CICM2016/intro.tex
More changes

@@
 Functional programming claims to be more ``mathematically tractable'' than other ``conventional'' approaches, due to properties like \emph{referential transparency} and \emph{immutability}, which allow equational reasoning without many of the complications caused by side effects and strict evaluation order \cite{hughes1989functional}.
-Languages such as Haskell combine these advantages with static type systems to restrict the possible values of every expression, allowing powerful analysis tools such as property checkers \cite{claessen2011quickcheck} and theorem provers \cite{rosen2012proving} to be developed.
+Languages such as Haskell combine these advantages with static type systems to restrict the possible values of every expression, making programs amenable to powerful analysis tools such as property checking \cite{claessen2011quickcheck} and theorem proving \cite{rosen2012proving}.
 

### 2016-03-02 CICM2016/related.tex
More changes

@@
 We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements.
-This general idea is quite old, with .
-The current generation of theory exploration tools trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
-In such a setting, most of the automation focuses on bookkeeping (such as type-checking, managing scope, etc.) and more traditional automated theorem proving.
+This general idea dates back at least to the 1970s, most famously to work by Lenat \cite{lenat1977automated}, though the theory exploration tools used in functional programming trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
 
-Subsequent systems have increased the level of automation, and have shown promise in tasks such as lemma discovery.
 As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012} \cite{johansson2009isacosy} \cite{Hipster}.
-Since \qspec{} is used for conjecture generation in many of those systems, they can also benefit from our work with machine learning.
+Since \qspec{} is used for conjecture generation in many of these systems, they may also benefit from our machine learning approach.
 
@@
 
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}.
-The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the input size of a task by removing redundant or irrelevant information from the input.
-Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is the essentially the same idea being applied in a novel setting.
+The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the search space of a theorem proving task by removing redundant or irrelevant information (axioms, definitions, lemmas, etc.) from the input.
+Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is essentially the same idea applied in a novel setting.
 
-Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
 The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
@@
 
-Various approaches have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}, though the original approach is based on the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics).
+The original approach uses the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics), although various alternatives have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.
 
@@
 
-Our recurrent clustering approach takes inspiration from the ML4PG \cite{journals/corr/abs-1212-3618} and ACL2(ml) \cite{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
-Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
-Whilst it may be interesting to contrast our current algorithm with an alternative based on that of ACL2(ml), it is unclear how such summaries could be extended to include types, which seems the next logical step for our approach.
-The ML4PG algorithm extends trivially, by using (term, type) pairs instead of just terms.
+Our recurrent clustering algorithm is similar to that of ML4PG, as our transformation maps the elements of a syntax tree to distinct cells in a matrix.
+In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.
+
+Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it also ignores types.
+This is because obtaining the type of each component of a Core definition is more difficult than in ML4PG (which queries the current \textsc{Proof General} state), and is left for future work.
 
@@
 In ACL2(ml), the clusters are used to restrict the search space of a proof search, much like we restrict the scope of theory exploration.
-
-ACL2(ml) reasons by analogy: finding theorem statements which are similar to the current goal, and attempting to prove the goal in a similar way.
-In particular, the lemmas used to prove a theorem are mutated by substituting symbols for those which appear in the same cluster.
-For example, if \texttt{plus} and \texttt{multiply} are clustered together, and we are trying to prove a goal involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem involving \texttt{plus}.
-The lemmas used to prove that theorem will be mutated, for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an attempt to prove the goal.
-
-Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.
 

### 2016-03-02 CICM2016/check.sh
More checks

@@ -7,10 +7,14 @@ ERR=0 # Indicates failure, will be our exit code
 function checkLatex {
     FAIL=0 # A checkLatex-specific version of ERR
     for FUNC in checkLatexUndatedTodos checkLatexDatedTodos checkLatexRefs \
-                checkLatexCites checkLatexMacros checkLatexLabels          \
-                checkLatexFigs checkLatexCitep
+                checkLatexCites checkLatexCiteTodos checkLatexMacros       \
+                checkLatexLabels checkLatexFigs checkLatexCitep            \
+                checkLatexDupeCites
     do
-        mapFiles "$FUNC" || FAIL=1
+        for FILE in *.tex
+        do
+            "$FUNC" "$FILE" || FAIL=1
+        done
     done
     return "$FAIL"
 }
@@ -48,11 +52,10 @@ function checkWarnings {
     CITES=$(echo "$WARNINGS" | grep 'Citation `'  | sed -e 's/.*`\(.*\)'"'.*/\1/g")
     while read -r CITE
     do
-        grep "^@[^{]*{$CITE,$" < ~/Writing/Bibtex.bib > /dev/null ||
-            {
-                ERR=1
-                echo "Citation '$CITE' missing" >> /dev/stderr
-            }
+        grep "^@[^{]*{$CITE,$" < ~/Writing/Bibtex.bib > /dev/null || {
+            ERR=1
+            echo "Citation '$CITE' missing" >> /dev/stderr
+        }
     done < <(echo "$CITES")
 
     # Check for non-citation warnings
@@ -120,16 +123,52 @@ function checkLatexDatedTodos {
 }
 
 function checkLatexRefs {
-    # No forward-references
+    # Check references
+    LABELS=$(cat *.tex | grep -o '\\label{[^}]*}')
     [[ "x$1" = "xintro.tex" ]] || {
         while read -r REF
         do
-            echo "File '$1' contains reference '$REF'" >> /dev/stderr
-        done < <(grep -o '\ref{sec:[^}]*}' < "$1")
+            if echo "$LABELS" | grep -o "label{$REF}" > /dev/null
+            then
+                # Report references, to allow checking for forward references
+                echo "File '$1' contains reference '$REF'" >> /dev/stderr
+            else
[…]

### 2016-04-19 PhDSymposium2016/abstract.tex
Added PhDSymp2016

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}
\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{"#1"}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\title{Exploring Software with Symbolic and Statistical Algorithms}
  \author{
          Chris Warburton \\
      University of Dundee \\
      \texttt{cmwarburton@dundee.ac.uk}
    }
\date{}

\begin{document}
\maketitle
\begin{abstract}
We aim to aid programmers in understanding and maintaining software
by increasing the capability of \emph{theory exploration} systems for
discovering novel, unexpected properties of programs and libraries.
Existing theory exploration systems find equivalent program fragments via brute-force
search, which doesn't scale well to realistic codebases. We narrow down
the scope of each exploration run using statistics derived from the
semantic similarity of the source code being explored, using a process
called \emph{recurrent clustering}.
\end{abstract}

\section{Introduction}\label{introduction}

As computers become ever more important in research, industry and
society, so does the ability to program them effectively, economically
and \emph{correctly}. Whilst tools exist to help programmers understand
and manipulate software, ranging from linters \cite{Johnson78lint} through to formal
verification systems \cite{boyer1983proof}, these are often limited in scope (e.g. to syntax rather than semantics) or require precise guidance from the user. One
reason is that automating the search for patches \cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings,
optimisations \cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other routine programming tasks is
prohibitively expensive due to the size of the search space.

We investigate the use of machine learning techniques to restrict such
search spaces more intelligently than uninformed local search,
specifically the use of \emph{clustering} to aid \emph{theory
exploration} (TE) in discovering program equivalences.

\section{Background}\label{background}

\subsection{Theory Exploration}\label{theory-exploration}

Our work builds on the existing TE system
\textsc{QuickSpec}\cite{QuickSpec}, which finds equivalent expressions
in the Haskell programming language \cite{marlow2010haskell}, built from a user-provided
\emph{signature} of constants and variables. For example, given a
signature of the list-manipulating functions \texttt{reverse},
\texttt{length} and \texttt{append}, plus two list variables \texttt{x}
and \texttt{y}, \textsc{QuickSpec} will discover that
\texttt{reverse\ (reverse\ x)} is equivalent to \texttt{x}; that
\texttt{length\ (reverse\ x)} is equivalent to \texttt{length\ x}; that
\texttt{length\ (append\ x\ y)} is equivalent to
\texttt{length\ (append\ y\ x)}; and so on. These are discovered by
enumerating all well-typed combinations of the signature's constants and
variables, then randomly instantiating the variables \cite{claessen2011quickcheck} and comparing the
resulting closed terms. Any expressions which remain indistiguishable
after hundreds of instantiations are conjectured to be equivalent and,
after removing redundancies, are produced as output. These conjectures
can be sent through a theorem prover \cite{claessen2013automating} and used to inform tests,
optimisations, refactoring, verification, or simply to help the
programmer learn more about the code.

Due to its exponential complexity, this enumerating procedure is limited
to producing small expressions from signatures with only a few elements.
Such small signatures give little chance for serendipitous discoveries,
which undermines the algorithm's potential to present programmers with
new, unexpected information.

\subsection{Recurrent Clustering}\label{recurrent-clustering}

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g. a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.

Recurrent clustering attempts to cluster expressions based on their
Abstract Syntax Trees (ASTs), an example of which is shown in Figure \ref{fig:astexample}. First the recursively-structured ASTs are
transformed into vectors of fixed length, by truncating the tree
structure and listing the node labels in breadth-first order.

These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices} \cite{de1972lambda}). To replace global variables, $\phi$ first performs another round of clustering (hence the name
``recurrent''), without including the current expression; each global
variable is replaced by its cluster index found by this ``inner'' round
of clustering.

\begin{figure}
    \begin{small}
      \begin{tikzpicture}[sibling distance=0pt]
        \tikzset{sibling distance=0pt}
        \tikzset{level distance=20pt}
        \Tree[ .$\CLam$
                  $\id{a}$
                  [ .$\CCase$
                       [ .$\CVar$
                            [ .$\CLocal$
                                 $\id{a}$ ]]
                       $\id{b}$
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CVar$
                                 $\CConstructor$ ]]
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CApp$
                                 [ .$\CVar$
                                      [ .$\CGlobal$
                                           $\id{even}$ ]]
                                 [ .$\CVar$
                                      [ .$\CLocal$
                                           $\id{n}$ ]]]
                            $\id{n}$ ]]]
      \end{tikzpicture}
    \end{small}
    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc. are chosen arbitrarily.}
    \label{fig:astexample}
\end{figure}

These recursive calls to the clustering algorithm (we use a standard
\emph{k-means} implementation) allow expressions to be compared in a
principled way: the similarity of expressions depends on the similarity of the expressions they reference, and so on recursively. In practice, this recursive
algorithm can be implemented in an iterative way, by accumulating the set of expressions to cluster in topological order of their dependency graph.

Figure \ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:

\begin{lstlisting}[language=Haskell]
odd  Z    = False
odd (S n) = even n

even  Z    = True
even (S n) = odd n
\end{lstlisting}

The \texttt{odd} function references four global variables: \texttt{Z}, \texttt{S}, \texttt{False} and \texttt{even}. Our algorithm doesn't yet distinguish between data constructors, so the only global reference we replace is \texttt{even}. Likewise, the definition of \texttt{even} references \texttt{Z}, \texttt{S}, \texttt{True} and \texttt{odd}. For mutual recursion like this, there is no valid topological order, so we use a sentinel value as the feature.

These feature vectors are sent through a k-means clustering algorithm, using the \textsc{Weka} machine learning library. For example, the feature vector for \texttt{odd} (padded to 6 labels for each level of the tree) will begin:

\begin{equation*}
  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots
\end{equation*}

\section{Implementation}\label{implementation}

We obtain ASTs from Haskell projects using a bespoke plugin for the GHC
compiler. From these ASTs, we extract type information which is used to append variables to the signature; and dependency information, which is used to topologically sort the clustering rounds. We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered. Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which a) have an argument type which we can randomly generate and b) have an output type which we can compare. \textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.

\section{Results}\label{results}

We have developed a machine learning approach for analysing Haskell code, including a
bespoke feature extraction method and a full implementation pipeline for turning Haskell packages into sets of equations.

\section{Difficulties and Future Work}\label{future-work}

The most difficult aspect of performing this exploration was obtaining real code in a usable format, due to the myriad edge-cases encountered. We make extensive use of Haskell's existing infrastructure, including the \textsc{GHC} compiler, the \textsc{cabal} build system, the \textsc{Hackage} code repository and the \textsc{Nix} package manager. Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own. Whilst we have worked around these issues, e.g. using meta-programming, this introduces unnecessary latency, fragility and complexity.

Our investigation of recurrent clustering and theory exploration only scratches the surface of combining symbolic and statistical AI algorithms. As well as benchmarking our approach against other techniques, there are many similar opportunities to be explored, where the reasoning power of symbolic algorithms can be guided and supervised by statistical, data-driven processes.

\bibliographystyle{plain}
\bibliography{../Bibtex}

\end{document}

### 2016-04-19 PhDSymposium2016/pandoc.sh

@@ -1,0 +1,24 @@
+#!/usr/bin/env bash
+
+# Build slides.md with Beamer
+function slides {
+    pandoc -w dzslides --standalone --self-contained --filter pandoc-citeproc -o slides.html slides.md
+}
+
+# Build abstract.md
+function abstract {
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+    bibtex abstract
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+}
+
+pids=()
+trap 'kill "${pids[@]}"' EXIT
+slides &
+pids+=("$!")
+abstract &
+pids+=("$!")
+
+echo "Waiting for ${pids[*]}"
+wait

### 2016-04-19 PhDSymposium2016/render.sh

@@ -1,0 +1,2 @@
+#!/bin/sh
+nix-shell --command './pandoc.sh'

### 2017-02-15 benchmark2016/article.tex
Add benchmark2016

\title{An Approach to Benchmark Conjecture Generators}
\author{
        Chris Warburton \\
                University of Dundee\\
}
\date{\today}

\documentclass[12pt]{article}

\begin{document}
\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

Many algorithms and approaches have been developed to automate the discovery of new mathematical theorems, but it is difficult to compare their performance in a quantitative way. Existing approaches include precision/recall measurements on existing theories, but these rely on some form of ground truth, i.e. which theorems are ``worth'' producing.

We propose that the extensive test suites used to benchmark automated theorem provers can be repurposed to create benchmarks for theory exploration systems. We provide a tool to automatically perform this conversion, given an existing benchmark suite in SMT-LIB format \cite{BarFT-SMTLIB}.

\section{Previous approaches}\label{previous approaches}

Theory exploration tools such as IsaCoSy \cite{Johansson.Dixon.Bundy:conjecture-generation}, QuickSpec and IsaScheme have been evaluated using the standard library of the Isabelle theorem prover as their ground truth.

ALGORITHM:

Concatenate all definitions from all files together.
For each definition, prefix its name with the filename it came from and update all references.
Normalise all local variable names.
Remove any identical definitions, leaving the first one in place and updating any references to the removed definitions to use this one.
Repeat, until no further removals are possible.

\bibliographystyle{plain}
\bibliography{./Bibtex}

\end{document}

### 2017-02-15 PAAR16/render.sh
Various modifications

@@ -1,4 +1,4 @@
 #! /usr/bin/env nix-shell
-#! nix-shell -i bash -p bash gnumake
+#! nix-shell -i bash
 make clean
 make all

### 2017-02-15 TheoryExploration/render.sh

@@ -1,5 +1,6 @@
 #!/bin/sh
-nix-shell --show-trace --command "bash" <<'EOF'
+nix-shell -E 'with import <nixpkgs> {}; callPackage ./. {}' \
+          --show-trace --command "bash" <<'EOF'
   for src in *.md
   do
     doc=$(basename "$src" .md)

### 2017-02-15 PAAR16/paar/EasyChair3.4/easychair.tex
Added data for PAAR

% easychair.tex,v 3.4 2015/12/10

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{The {\easychair} Class File\\
       Documentation and Guide for Authors%
\thanks{Other people who contributed to this document include Maria Voronkov
  (Imperial College and EasyChair) and Graham Gough (The University of
  Manchester).}}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Serguei A. Mokhov\inst{1}\thanks{Designed and implemented the class style}
\and
    Geoff Sutcliffe\inst{2}\thanks{Did numerous tests and provided a lot of suggestions}
\and
   Andrei Voronkov\inst{3}\inst{4}\inst{5}\thanks{Masterminded EasyChair and created versions
     3.0--3.4 of the class style}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Concordia University,
  Montreal, Quebec, Canada\\
  \email{mokhov@cse.concordia.ca}
\and
   University of Miami,
   Miami, Florida, U.S.A.\\
   \email{geoff@cs.miami.edu}\\
\and
   University of Manchester,
   Manchester, U.K.\\
   \email{andrei@voronkov.com}\\
\and
   Chalmers University of Technology,
   Gothenburg, Sweden
\and
   EasyChair
 }

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Mokhov, Sutcliffe and Voronkov}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{The {\easychair} Class File}

\begin{document}

\maketitle

\begin{abstract}
  In order to ease the lives of authors, editors, and trees, we present an
  easy-to-read guide to the easy-to-use {\easychair} {\LaTeX2e} document style
  class for EasyChair-based electronic and on-paper publishing of workshop and conference
  proceedings.
\end{abstract}

% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC series

\setcounter{tocdepth}{2}
{\small
\tableofcontents}

%\section{To mention}
%
%Processing in EasyChair - number of pages.
%
%Examples of how EasyChair processes papers. Caveats (replacement of EC
%class, errors).

\pagestyle{empty}

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}

The {\easychair} class was designed to be easy to use, and specifically favors
electronic and on-paper publishing by the EasyChair conference system
\cite{easychair}, including the
\href{http://www.easychair.org/publications/EPiC}{EasyChair EPiC
  Series}\footnote{If you are writing a paper for the EPiC series, please
check Subsection~\ref{subsec:EPiC}.}. It is
also applicable to the Procedia publications series by Elsevier Science.
EasyChair is a free conference management system that is flexible, easy to use,
and has many features to make it suitable for various conference models.  It is
currently probably the most commonly used conference management system
\cite{easychair}.  The use of EasyChair and this style for creating Procedia
volumes is a pilot project between Elsevier Science and EasyChair.

The {\easychair} class was designed according to some requirements, which
are described in Appendix~\ref{sect:easychair-requirements}. 
An article that occupies approximately 15 LNCS-formatted pages
takes up approximately 14 {\easychair} pages.

\begin{figure}[tb]
	\begin{centering}
	\includegraphics[width=0.5\textwidth]{logoEC}
	\caption{EasyChair logo}
	\label{fig:easychair-logo}
	\end{centering}
\end{figure}

%------------------------------------------------------------------------------
\section{Typesetting}
\label{sect:typesetting}

Typesetting with {\easychair} is, well, easy.  Just by using the
document class entry in the document's preamble as follows:
\verb+\documentclass{easychair}+ the typesetting work is nearly done.
The {\easychair} class is a relatively conservative extension of the
standard \texttt{article} class, so most of the environments, section
headers, etc. defined by \texttt{article} are available.

%------------------------------------------------------------------------------
\subsection{Generalities}
\label{sect:generalities}

The following are the general default parameters {\easychair}
introduces into the typesetting aspect of articles. If you use
\easychair\ for proceedings or other kinds of publishing through
EasyChair, do not alter these -- papers deviating from the formatting
standards will be rejected by EasyChair.

\begin{enumerate}
\item
The default paper size is US letter. It can be explicitly set to A4 
(\texttt{a4paper}) or letter (\texttt{letterpaper}) paper in the
document class entry, e.g.:\\\verb+\documentclass[a4paper]{easychair}+

\item
The print area for both letter and A4 paper sizes is 145x224 mm. This size
has been selected to allow for inexpensive printing using our current
print-on-demand publisher.

\item
The base font is Computer Modern. The base font size is 10pt. If you
use any other font size, there is no guarantee that the produced
document will look nice or fit into our standard page size.

\item
The references list is condensed. The default bibliography styles, such as
\texttt{plain}, \texttt{abbrv}, and \texttt{alpha}, are suggested.

\item
PNG, JPG, and PDF images are supported, i.e., those that are supported
by the standard \texttt{graphicx} package \cite{graphicx-package}, and
render nicely in online versions of PDF documents.  This document
shows some examples of JPG and PDF images, for example in
Figure~\ref{fig:easychair-logo}. If the papers are designed for
publishing in print, the images should be at least 300dpi in
resolution.

\end{enumerate}

%------------------------------------------------------------------------------
\subsection{Front Matter}
\label{sect:front-matter}

The front matter of an {\easychair} article follows the \texttt{article}
style, augmented with the \verb+\titlerunning+ and \verb+\authorrunning+
commands for use by authors. 
For the \verb+\author+ command with multiple authors, use \verb+\and+ to
separate authors from different institutions, as done in this document.
Institutions are defined using \verb+\institute+ in a similar way and affiliations of authors assigned using \verb+\inst+.
Authors must set the \verb+\titlerunning+ and \verb+\authorrunning+.
For example, the front matter of this document defined the authors
and title as follows:

%---------------------------------------------------------------

\noindent\small
\begin{verbatim}
\title{The {\easychair} Class File\\
       Documentation and Guide for Authors%
\thanks{Other people who contributed to this document include Maria Voronkov
  (Imperial College and EasyChair) and Graham Gough (The University of
  Manchester).}}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Serguei A. Mokhov\inst{1}\thanks{Designed and implemented the class style}
\and
   Geoff Sutcliffe\inst{2}\thanks{Did numerous tests and provided a lot of suggestions}
\and
   Andrei Voronkov\inst{3}\inst{4}\inst{5}\thanks{Masterminded EasyChair and created 
     versions 3.0--3.4 of the class style}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Concordia University,
  Montreal, Quebec, Canada\\
  \email{mokhov@cse.concordia.ca}
\and
   University of Miami,
   Miami, Florida, U.S.A.\\
   \email{geoff@cs.miami.edu}\\
\and
   University of Manchester,
   Manchester, U.K.\\
   \email{andrei@voronkov.com}\\
\and
   Chalmers University of Technology,
   Gothenburg, Sweden
\and
   EasyChair
 }

\authorrunning{Mokhov, Sutcliffe and Voronkov}
\titlerunning{The {\easychair} Class File}
\end{verbatim}
\normalsize

%------------------------------------------------------------------------------
\subsection{Page Numbering}
\label{sect:page-numbering}

Page numbers are at the bottom of every page. Authors must leave the
page numbers in as-is. When EasyChair proceedings and Procedia volumes are processed by
EasyChair, the correct volume page numbers will be inserted
automatically.

%------------------------------------------------------------------------------
\subsection{Section Headings and Capitalization}
\label{sect:section-headings}

Section and paragraph headings in {\easychair} are invoked via the standard 
commands, such as
\verb+\section+,
\verb+\subsection+,
\verb+\subsubsection+, and
\verb+\paragraph+.
Generally, every non-trivial word in a heading must be capitalized according to
general capitalization guidelines. A reasonable rule to use is that
all prepositions, coordinating conjunctions and articles having four
or fewer letters should not be capitalized. If you do not know what it
means, simply do not capitalize the following words:
\textit{amid, anti, as, at, atop, but, by, down, for, from, in, into, like,
near, next, of, off, on, onto, out, over, pace, past, per, plus, qua,
save, than, till, to, up, upon, via, with, for, and, nor, but, or,
yet, so, a, an, the.} For example, if you want to call your paper
``oldest but goldest'', then the proper title for it is ``Oldest but
Goldest''. ``Oldest but goldest'' is wrong (since ``goldest'' is not
capitalized) and ``Oldest But Goldest'' is wrong (``but'' should stay
lower-cased since it belongs to the list of words above.) Needless to
say, ``OLDEST BUT GOLDEST'' is very wrong.

Paragraph headings should not be capitalized and should have a
trailing period. That is, you should write

\begin{quote}
\verb|\paragraph{EasyChair is cool.}|
\end{quote}
rather than

\begin{quote}
\verb|\paragraph{EasyChair is cool}|
\end{quote}
unless your aim was to write something like

\begin{quote}
\verb|\paragraph{EasyChair is cool} when you use it for publishing.| 
\end{quote}
Of course, you are welcome to replace the trailing period by any other
punctuation mark, for example

\begin{quote}
\verb|\paragraph{EasyChair is cool!}|
\end{quote}

See the examples in this document, e.g.,
Section~\ref{sect:typesetting} is a section, this 
(Section~\ref{sect:section-headings}) is a subsection, and
Section~\ref{sect:subsubsection-headings} is a subsubsection.

%------------------------------------------------------------------------------
\subsubsection{Subsubsection Header}
\label{sect:subsubsection-headings}

This is a subsubsection.

\paragraph{Paragraph header.}

This is a paragraph. 
One way of saving space when hyper-references are not essential is to 
use paragraphs instead of subsubsections.

%------------------------------------------------------------------------------
\subsection{Mathematics}
\label{sect:mathematics}

Mathematics can be done inline for simple things, e.g., an equation
$x = 0$, possibly with super and subscripts, e.g., $x^2_k \approx 27$,
Greek letters, e.g., $\alpha \cup \Theta \ne \gamma$, etc.
Larger formulae must be done using {\tt \verb|\|[}~~{\tt \verb|\|]}
bracketing, e.g.,
\[
\int_{0}^{1} x dx = \left[ \frac{1}{2}x^2 \right]_{0}^{1} = \frac{1}{2}
\]
or using {\tt \verb|\|begin\{equation\}} and {\tt \verb|\|end\{equation\}} for
numbered equations, e.g.,
\begin{equation}
e^x = \sum_{n=0}^\infty \frac{x^n}{n!} = \lim_{n\rightarrow\infty} (1+x/n)^n
\end{equation}

Use {\tt \verb|\|begin\{align*\}} and {\tt \verb|\|end\{align*\}} (or without
the {\tt *} include number) to align equations, e.g.,
\begin{align*}
x^2 + y^2 &= 1 \\
y &= \sqrt{1 - x^2}
\end{align*}

Fonts, using \verb|\matcal| and others can also be used in the math mode: $\mathcal{ALC}$.

%------------------------------------------------------------------------------
\subsection{Tables}
\label{sect:tables}

Table~\ref{tab:ltbexample} shows an example of a table of data that was
conveniently available (i.e., the data has nothing to do with
{\easychair}) apart from being related to the research of two of its
authors.

\begin{table}[htp]
	\begin{centering}
		\begin{tabular}{lrrrrrrrr}
		\hline
		ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
		                      & /100& time & out & Con. & Eff. & /35 & /40 & /25 \\
		\hline
		Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
		iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
		SInE 0.4              &  64 & 75.3 &  64 & 0.32 &  8.5 &  26 &  13 &  25 \\
		leanCoP-SInE 2.1      &  35 &110.8 &  35 & 0.23 &  3.2 &  23 &   1 &  11 \\
		E-LTB 1.1pre          &  18 & 63.4 &   0 & 0.21 &  2.8 &   7 &   9 &   2 \\
		EP-LTB 1.1pre         &  18 & 77.8 &  18 & 0.21 &  2.3 &   7 &   9 &   2 \\
		E-KRH'-LTB 1.1.3      &   0 &   -- &  -- &   -- &   -- &   0 &   0 &   0 \\
		\hline
		\end{tabular}
		\caption{LTB division results}
		\label{tab:ltbexample}
	\end{centering}
\end{table}

%------------------------------------------------------------------------------
\subsection{References}
\label{sect:references}

References must be provided in a {\tt .bib} file, so that \BibTeX\ can
be used to generate the references in a consistent style in a volume.
The preferred styles are {\tt plain} and {\tt alpha}.
For example, the references for this paper are generated from the
lines
\begin{verbatim}
    \bibliographystyle{plain}
    \bibliography{easychair}
\end{verbatim}
and a way to compose the entires, e.g. citing this class style~\cite{easychair-latex-class}
is below:
\small
\begin{verbatim}
    @misc
    {
      easychair-latex-class,
      author       = {Serguei A. Mokhov and Geoff Sutcliffe and Andrei Voronkov},
      title        = {The {\sf easychair} Class File Documentation and Guide
                      for Authors},
      year         = {2008--2011},
      howpublished = {[online]},
      note         = {Available at \url{http://easychair.org/easychair.zip}}
    }
\end{verbatim}
\normalsize

%------------------------------------------------------------------------------
\section{Installation and Usage Instructions}
\label{sect:installation-usage}

%------------------------------------------------------------------------------
\subsection{Installation}

The ``installation'' of the {\easychair} document class is easy.
Download the latest version of the \texttt{easychair.zip} package 
from \url{http://www.easychair.org/easychair.zip}
and unzip it in the directory where you will prepare your paper.
You will get the following files, out of which you may need to keep only 
the \texttt{easychair.cls} style class if you are familiar with the rest 
of the files and do not require them to get started.
We are also working to make {\easychair} available from CTAN \cite{ctan},
such that it can be installed with the popular \TeX Live \cite{texlive} and
{\miktex} \cite{miktex} \LaTeX\ package management systems.

\begin{itemize}
\item
\texttt{easychair.cls} -- the class file that this is all about.

\item
\texttt{easychair-a4.pdf} -- the PDF version of this guide rendered using 
\texttt{a4paper} option.

\item
\texttt{easychair.tex} -- the {\LaTeX} source of this guide, 
and
\texttt{easychair.bib} -- the supporting bibliography entries found starting 
on page~\pageref{sect:bib}.

\item
\texttt{logoEC.pdf} -- the PDF version of the EasyChair logo rendered in
Figure~\ref{fig:easychair-logo} and
\texttt{throneEC.jpg} -- the JPG version of the easy throne rendered in 
Figure~\ref{fig:easythrone}.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Required Packages}

The {\easychair} class relies only on packages deemed standard and shipped by
most {\LaTeX} distributions in the worlds of Linux (current \texttt{texlive} \cite{texlive}
or older \texttt{tetex}), MacOS X,
and Windows (via Cygwin or {\miktex}).
If for some reason your distribution is old or doesn't have the packages
listed below, you can always obtain a copy from CTAN \cite{ctan}. Note
that EasyChair loads this packages automatically so you don't have to
use \verb|\usepackage| to use any of these packages.

\begin{itemize}
\item
\texttt{url} \cite{url-package} (included also by \texttt{hyperref} automatically) -- to provide URL rendering support for the
monospaced font, which takes care of special characters as well as line 
wrapping.

\item
\texttt{hyperref} \cite{hyperref-package} -- to allow hyperlinking of URLs and
cross references within an article.
Its options are set to either \verb+letterpaper+ or \verb+a4paper+, depending
on the \verb+\documentclass+ options.

\item
\texttt{graphicx} \cite{graphicx-package} -- the standard package for rendering
PNG, JPG, and PDF graphic images, primarily in \texttt{figure} environments.

\item
optional \texttt{mathptmx} \cite{mathptmx-package} -- Times base font for compactness
(use with the \texttt{withtimes} {\easychair} option).

\item
\texttt{listings} \cite{listings-package} -- to allow highlighted source code 
listing styles.

\item
\texttt{latexsym} \cite{latexsym-package} -- to provide common math and other 
symbols.

\item
\texttt{amsthm} \cite{amsthm-package} -- to provide {\AmS} theorem-like 
environments.

\item
\texttt{empheq} \cite{empheq-package} -- to provide equation environments, etc.

\item
\texttt{geometry} \cite{geometry-package} -- to set {\easychair} margins, 
outlined in Section~\ref{sect:generalities}.

\item
\texttt{lastpage} \cite{lastpage-package} -- to allow computationally
referencing the last page.

\item
\texttt{fancyhdr} \cite{fancyhdr-package} -- for running heads.

\item
\texttt{footmisc} \cite{footmisc-package} -- to ensure that footnotes are
always at the bottom.

\item
optional \texttt{makeidx} \cite{makeidx-package} -- for index generation
(use with the \texttt{thesis} {\easychair} option).

\item
\texttt{eso-pic} \cite{eso-pic-package} -- for draft versions and checking page
overlflows vs. a border drawn around the headers, footers, and the main body of
the article.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Recommended Packages}
\label{sect:recommended-packages}

Here is a list of some packages that this guide's authors have experimented 
with, and which are suitable for inclusion if needed by article authors.
These packages must be loaded using \verb+\usepackage+.
In general, authors may use any standard packages provided they do not change
the basic layout and font settings established by the {\easychair} class.
Such packages must be provided with the submission of articles.

\begin{itemize}
\item
\texttt{rotating} \cite{rotating-package} -- to rotate floats (figures and
tables) on the page, when wide tables or figures do not fit in portrait layout.

\item
\texttt{pdflscape} \cite{pdflscape-package} -- similar to \texttt{rotating}, 
but also allows rotating text to make it conveniently viewable in a PDF 
viewer that supports individual rotated pages.
A possible disadvantage is that a page break is forced, which may create
gaps before or after the landscape page.

\item
\texttt{algorithm2e} \cite{algorithm2e-package} -- provides a figure-like
algorithm environment for formal algorithm presentation with highlighting.

\end{itemize}

%------------------------------------------------------------------------------
\subsection{Compiling}
\label{sect:compiling}

\texttt{pdflatex} \cite{pdflatex-instructions} is the preferred tool for
producing PDF files with {\easychair} class documents.
The author kit (\texttt{easychair.zip}) includes some minimal automation 
that authors can use at their discretion.

\begin{itemize}
\item
Linux and UNIX-like platforms (also works under Cygwin and MacOS X):
A {\makefile} is provided for the GNU \texttt{make} \cite{gmake} utility,
so this document can be compiled by typing \texttt{make} at the terminal 
prompt (on the systems where both GNU and non-GNU versions of \texttt{make} 
are installed, one may need to use \texttt{gmake}).

\item
Microsoft Windows:
{\texniccenter} \cite{texniccenter} or {\latexeditor} \cite{led} and
{\miktex} \cite{miktex} as their backend are common tools
for {\LaTeX} processing under Microsoft Windows. 
The former provide a GUI front-end to {\LaTeX}, and the latter is the 
Windows native-compiled binaries and standard packages with 
a comprehensive package update tool. 
The \texttt{easychair.tcp} project file is provided for {\texniccenter} users,
as well as \texttt{easychair.lpr} for {\latexeditor} users.

\item
MacOS X:
TeXShop \cite{texshop} is a tool for {\LaTeX} processing under Mac OS X.
It provides a GUI front-end to {\LaTeX}. The backend can be installed
through the \texttt{fink} \cite{fink} repository or the Darwin Ports.
\end{itemize}

Some authors use packages that require the use of \texttt{latex}
instead of \texttt{pdflatex}, most notably \texttt{pstricks} and its
derivatives. If switching to modern packages, such as
\texttt{tikz} is not an option, these authors can use the {\easychair}
class with \texttt{latex}.

%------------------------------------------------------------------------------
\subsection{Bug Reports}
\label{sec:bug-reports}

Please report bugs, errors, and omissions you find with the {\easychair} 
class to its current maintainer, Andrei Voronkov,
at \url{andrei@voronkov.com}. Any \emph{constructive} feedback is
always welcome. If anybody has a very hard question, we might be
unable to answer it without Sergei. If anybody knows his whereabouts,
Geoff and Andrei have the ransom money ready.

%------------------------------------------------------------------------------

\section{Avoiding Text Overflows}

When used in EasyChair, the {\easychair} class is intended for
publishing, including printing. This implies that the text of your
document should not overflow the page dimensions. This section gives
some practical advice on how you can avoid text overflows. When you
run \texttt{pdflatex} or \texttt{latex} on your files, you sometimes
get warnings like

\small
\begin{verbatim}
  LaTeX Warning: Overfull \hbox (7.43138pt too wide) in paragraph at lines 994--994
\end{verbatim}
\normalsize
or

\small
\begin{verbatim}
  LaTeX Warning: Float too large for page by 55.21666pt on input line 638.
\end{verbatim}
\normalsize
They mean what they say: your text overflows the allowed space. When
you prepare a PDF document that will only be used online or as a
draft, in many cases you can ignore these warnings, since they only
make your document uglier than it could be but it remains still
readable. However, for documents that are supposed to be printed,
overflows make a lot of difference. There is no universal recipe on
how to make your text fit into the page. However, there are some
common cases that can be easily fixed using the tricks described
here.

\subsection{Using the \texttt{debug} Option to Check for Overflows}

A general guideline is that you should first check how the alleged
overflows looks in reality. For example, some LaTeX overflows are
invisible or hardly noticeable for people and overlows by
less than 4pt look OK anyhow. The {\easychair} class has an option
\texttt{debug} that adds to each page of your documents a red frame
corresponding to the allowed page
dimensions. Figure~\ref{fig:redframe} on page~\pageref{fig:redframe}
shows an example of a page created using this style. With such a red
frame, you can clearly see any overflows apart from very minor ones.

\begin{figure}
  \begin{centering}
    \includegraphics[width=0.89\textwidth]{debug.pdf}
    \caption{A page of a document created using the \texttt{debug}
      option} 
    \label{fig:redframe}
  \end{centering}
\end{figure}

\subsection{Tables}

Many page overflows happen because of large tables. In many case these
overflows can be easily removed by slightly reducing padding added by
\LaTeX\ to every column. It is controlled by the \LaTeX\ command
\verb|\tabcolsep| whose value by default is 6pt. Even small changes in
the value of this command may give drastic reductions in the width of
tables. This is illustrated in Figure~\ref{fig:tabcolsep} on
page~\pageref{fig:tabcolsep}. Note though that there is no free lunch:
smaller values for this command may result in lower redability.

%------------------------------------------------------------
\begin{figure}[tb]\small
\begin{center}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{5pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{3pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}

\begin{center}\renewcommand{\tabcolsep}{1pt}
  \begin{tabular}{lrrrrrrrr}
    \hline
    ATP System            & LTB & Avg  &Prfs & SOTA & \multicolumn{1}{c}{$\mu$} & CYC & MZR & SMO \\
   \hline
    Vampire-LTB 11.0      &  69 & 24.5 &  69 & 0.37 & 28.1 &  23 &  22 &  24 \\
    iProver-SInE 0.7      &  67 & 76.5 &   0 & 0.36 &  8.8 &  28 &  14 &  25 \\
   \hline
  \end{tabular}
\end{center}
\normalsize

\caption{Original table and tables with \texttt{tabcolsep} set to 5pt,
  3pt, and 1pt
  \label{fig:tabcolsep}}

\end{figure}
%------------------------------------------------------------

\subsection{Images}

Images included using \verb|\includegraphics| are easy to resize since
one can specify the size of the result explicitly. For example,
Figure~\ref{fig:easythrone} shows three copies of the same image
having different sizes obtained using the following commands:

\begin{verbatim}
  \includegraphics[width=0.5\textwidth]{throneEC.jpg}
  \includegraphics[width=0.3\textwidth]{throneEC.jpg}
  \includegraphics[width=0.15\textwidth]{throneEC.jpg}
\end{verbatim}

\subsection{A Universal Recipe}

\LaTeX\ has a very powerful weapon for reducing the size of almost
anythings. More precisely, it can reduce anything producing what
\LaTeX\ considers a box. This weapon is called
\verb|\scalebox|. Consider an example (check the source of this file
to see how it was produced).

\begin{center}
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2010} \\
\hline
\end{tabular}
\end{center}
This is what happens when we put (almost) the same \LaTeX\ code in 
\verb|\scalebox{0.55923}{...}| to scale it down to the number of users
in 2009:

\begin{center}
\scalebox{0.55923}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2009} \\
\hline
\end{tabular}}
\end{center}
We can scale it down even further to the 2008 figure using
\verb|\scalebox{0.30856}{...}|:

\begin{center}
\scalebox{0.30856}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2008} \\
\hline
\end{tabular}}
\end{center}
or further down to 2007:

\begin{center}
\scalebox{0.12870}{%
\begin{tabular}{|c|c|}
\hline \Huge
\begin{tabular}[b]{cr}
year & users \\ \hline
2007 &    47,753 \\
 2008 &    114,494 \\
 2009 &    207,506 \\
 2010 &   371,054 \\
\end{tabular}
&
\includegraphics[width=0.28\textwidth]{chairEC}\\
\hline
\multicolumn{2}{|c|}{The number of users of EasyChair and one of its
  logos,}\\
\multicolumn{2}{|c|}{scaled to the number of users in 2008} \\
\hline
\end{tabular}}
\end{center}

This size reduction technique is very efficient: using the right scale
you may post your whole article on Twitter in a single tweet. However,
it may also may parts of your text virtually unreadable with an
unfortunate side effect of annoying reviewers.

\section{Submitting Your Article Through EasyChair}

This section is intended only for the authors and editors of EasyChair
proceedings and Procedia volumes.

When you prepare an article for either of these, it should be
submitted through EasyChair. EasyChair automates the submission
process as much as possible and goes to a great length to ensure that
your article can be published and printed. Publication for EasyChair
means much more than just putting a PDF of your article online. It
collects a some meta-information about the article to classify it,
find similar articles, make it easily searchable, and index it in
various Web services, such as DBLP. This section explains how
EasyChair processes your article.

\subsection{Submitting the Article}

Submitting the article is easy. All you should do is to put the source
of your article in a single zip file. The source must contain all
auxiliary files required to create a PDF file of your article: this
includes images, bibliography, and all non-standard \LaTeX\ packages
you used\footnote{ 
  A non-standard \LaTeX\ package is a package that is not included in
  CTAN. 
} For example, suppose that your main \LaTeX\ file is
\texttt{main.tex}, it inputs another file \texttt{macros.tex} and uses
the file \texttt{biblio.bib} to produce the bibliography. Suppose it
also uses two images \texttt{images/easy.jpg} and
\texttt{images/easy.jpg}. Then you should create a zip archive
containing all these files. Suppose all these files are put in a
directory \texttt{mypaper} on your computer, where \texttt{images} is
a subdirectory of \texttt{mypaper}

On almost any operating system (Linux,
Windows, or Mac) you can achieve this by using the following sequence
of commands:

\begin{verbatim}
cd mypaper
zip -r mypaper.zip *
\end{verbatim}
This will create a zip archive \texttt{mypaper.zip} including all
files in the directory \texttt{mypaper} and its subdirectories.

\subsection{EPiC Series}
\label{subsec:EPiC}

If you publish your article in the EasyChair EPiC series, our aim is to produce
a nicely looking EPiC Series header, see Figure~\ref{fig:epicheader} on page
\pageref{fig:epicheader} for an idea. The images and information in the header
will be added by EasyChair automatically and depend on the subseries and the
volume information.

\begin{figure}
  \begin{centering}
    \includegraphics[width=0.89\textwidth]{epic.pdf}
    \caption{This document with an EPiC Series header} 
    \label{fig:epicheader}
  \end{centering}
\end{figure}

To ensure that the layout of your document will not change when the header is
added, you should use the option \texttt{EPiC} as follows:

\small
\begin{verbatim}
\documentclass[EPiC]{easychair}
\end{verbatim}
\normalsize
This will reserve space for the header on the top of the first page of your
article.

%------------------------------------------------------------------------------
\section{Future Work}
\label{sect:future-work}

We plan to further strengthen the {\easychair} class and promote it for 
electronic publishing for EasyChair-powered conferences and workshops,
and take over the world, as shown in Figure~\ref{fig:easythrone}. We
aim at creating a new model of \emph{affordable publishing}, where
anybody can become a low-cost publisher. The
\href{http://www.easychair.org/publications/EPiC}{EasyChair EPiC Series} is our
first contribution to affordable open access publishing.

\begin{figure}[tb]
  \begin{centering}
    \includegraphics[width=0.5\textwidth]{throneEC.jpg}
    \includegraphics[width=0.3\textwidth]{throneEC.jpg}
    \includegraphics[width=0.15\textwidth]{throneEC.jpg}
  \end{centering}
  \caption{Why one should use EasyChair}
  \label{fig:easythrone}
\end{figure}

%------------------------------------------------------------------------------
\section{Acknowledgments}
\label{sect:acks}

\begin{itemize}
\item
Aleksander Kosenkov for the graphics that are used here and on the EasyChair 
website~\cite{easychair}.

\item
Graham Gough (The University of Manchester) for implementing title-related
commands.

\item
Thomas F. Sturm (Universit\"at der Bundeswehr M\"unchen) for implementing the
\LaTeX\ part of the headers used for the EasyChair EPiC series.

\item
The CTAN \cite{ctan} and {\LaTeX} communities \cite{texniccenter,miktex}.

\item
Guilin Qi, Jasmin Christian Blanchette, Leslie Lamport, Uwe Pfeiffer,
and others for constructive feedback on the style, most of which had been
incorporated in version 2 of the class style.
\end{itemize}

\label{sect:bib}
\bibliographystyle{plain}
%\bibliographystyle{alpha}
%\bibliographystyle{unsrt}
%\bibliographystyle{abbrv}
\bibliography{easychair}

%------------------------------------------------------------------------------
\appendix
\section{{\easychair} Requirements Specification}
\label{sect:easychair-requirements}

The following high-level requirements were set for the development of 
the {\easychair} class, and were refined as development went along.

\begin{enumerate}
\item
The style should be easy to use. 
The average {\LaTeX} user should not need to read a long manual.

\item
It should be economical in space but the text should be nice-to-read.

\item
It should use fonts producing a reasonable-quality PDF.

\item
The bibliography should produce hyperlinks.

\item
Sections should produce menu sections in PDF.

\item
The text should look good on both A4 and letter paper.

\item
The style should be single-column for convenience of scrolling.

\item
The print area should be convenient for printing using print-on-demand publishers.

\item
Running heads.
\end{enumerate}

%------------------------------------------------------------------------------
% Index
%\printindex

%------------------------------------------------------------------------------
\end{document}

### 2017-04-19 PhDSymposium2017/pandoc.sh
Added PHD symposium 2017 (copied from 2016)

@@ -1,0 +1,25 @@
+#!/usr/bin/env bash
+
+# Build slides.md with Beamer
+function slides {
+    pandoc -t slidy --standalone --self-contained --filter pandoc-citeproc \
+           --filter panpipe --filter panhandle -o slides.html slides.md
+}
+
+# Build abstract.md
+function abstract {
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+    bibtex abstract
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+    pdflatex -interaction=nonstopmode -halt-on-error --shell-escape abstract
+}
+
+pids=()
+trap 'kill "${pids[@]}"' EXIT
+slides &
+pids+=("$!")
+#abstract &
+#pids+=("$!")
+
+echo "Waiting for ${pids[*]}"
+wait

### 2017-04-19 PhDSymposium2017/render.sh

@@ -1,0 +1,6 @@
+#!/bin/sh
+nix-shell --command './pandoc.sh'
+          #-p pandoc \
+          #-p haskellPackages.pandoc-citeproc \
+          #-p panpipe \
+          #-p panhandle \

### 2017-04-19 benchmark2016/article.tex
Benchmark updates

@@
-\title{An Approach to Benchmark Conjecture Generators}
+\title{An Approach to Benchmarking Conjecture Generators}
 \author{
@@
 
+\usepackage{attrib}
+\usepackage{csquotes}
+
+\usepackage{graphicx}
+\graphicspath{ {support/} }
+
 \begin{document}
@@
 
+\begin{displayquote}
+  There is frequently more to be learn'd from the unexpected questions of a
+  child than the discourses of men.
+  \attrib{--- John Locke}
+\end{displayquote}
+
 \section{Introduction}
 
-Many algorithms and approaches have been developed to automate the discovery of new mathematical theorems, but it is difficult to compare their performance in a quantitative way.
-Existing approaches include precision/recall measurements on existing theories, but these rely on some form of ground truth, i.e.
-which theorems are ``worth'' producing.
+The field of mathematics has been transformed by the introduction of computers:
+calculations are now routinely automated, allowing humans to focus their efforts
+on higher-level, more creative activities like proving theorems; \emph{automated
+theorem proving} (ATP) seeks to automate this process too, and was one of the
+first tasks studied in the field of artificial intelligence
+\cite{newell1956logic, sutcliffe2001evaluating}. Despite this effort, ATP
+methods are not widely used by traditional mathematicians, although they have
+seen use in applied areas such as hardware and software verification
+\cite{Moore:2003}.
 
-We propose that the extensive test suites used to benchmark automated theorem provers can be repurposed to create benchmarks for theory exploration systems.
-We provide a tool to automatically perform this conversion, given an existing benchmark suite in SMT-LIB format \cite{BarFT-SMTLIB}.
+Even if ATP became widespread, the automation of theorem proving would still
+leave an important aspect of mathematics inaccessible to computers: conjecturing
+those theorems (and the definitions they involve) in the first place. Automated
+methods for posing good conjectures have obvious applications for those who must
+invent a constant stream of novel questions, e.g. for examinations or research.
+A larger impact may be found in areas such as computer programming, security and
+verification, where the benefits of formal methods are hindered by the
+difficulty of their application; tools to aid this process, e.g. by proposing
+test cases, can reduce this cost.
 
-\section{Previous approaches}\label{previous approaches}
+Many automated systems and algorithms have been developed to generate
+conjectures, but it is difficult to judge or compare their performance in a
+quantitative way. There are many aspects to consider when judging the quality of
+a conjecture: how likely it is to be true, how difficult it is to prove, its
+implications, and so on.
+
+One way to make these subjective, ambiguous properties more concrete is to
+choose a well-studied theory, with a set of existing theorems that are deemed to
+be useful or important in some way, and treat these as a \em{ground truth}
+against which we can compare conjectures generated from the same theory.
+
+A common choice of ground truth is the standard library of a theorem prover,
+such as Isabelle/HOL; the library authors have gone to the effort of stating,
+proving and including these theorems in every copy of their software, which is a
+good indication that they are useful or important.
+
+One problem with this approach is the small size of these libraries. For
+example, a benchmark based on Isabelle/HOL's theory of natural numbers contains
+only FIXME definitions and FIXME theorems. Whilst such benchmarks allow
+comparison between different approaches, their narrow scope doesn't provide much
+indication of performance in different, especially \emph{novel}, domains.
+
+We propose that the extensive test suites which are already used to benchmark
+automated theorem provers can be repurposed to create benchmarks for conjecture
+generation systems. We provide a benchmark suite, automatically derived from the
+Tons of Inductive Problems (TIP) benchmark suite, as well as software to convert
+other benchmarks written in the same TIP format (an extension of the SMT-LIB
+format \cite{BarFT-SMTLIB}).
+
+\section{Existing Evaluations}\label{previous approaches}
 
@@
 
+\section{Evaluation}
+
+\includegraphics{bucketing-graph}
+
 \bibliographystyle{plain}

### 2017-04-19 PhDSymposium2017/abstract/abstract.tex
Split up abstract from slides

--- 2016-04-19 PhDSymposium2016/abstract.tex
@@
             pdfauthor={},
-            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
+            pdftitle={Quantitative Benchmarks for Theory Exploration},
             colorlinks=true,
@@
 
-\title{Exploring Software with Symbolic and Statistical Algorithms}
-  \author{
-          Chris Warburton \\
+\title{Quantitative Benchmarks for Theory Exploration}
+  \author{Chris Warburton \\
           University of Dundee \\
-      \texttt{cmwarburton@dundee.ac.uk}
-    }
+          \texttt{cmwarburton@dundee.ac.uk}}
 \date{}
@@
 \begin{abstract}
-We aim to aid programmers in understanding and maintaining software
-by increasing the capability of \emph{theory exploration} systems for
-discovering novel, unexpected properties of programs and libraries.
-Existing theory exploration systems find equivalent program fragments via brute-force
-search, which doesn't scale well to realistic codebases. We narrow down
-the scope of each exploration run using statistics derived from the
-semantic similarity of the source code being explored, using a process
-called \emph{recurrent clustering}.
+  \emph{Automated theory exploration} has the potential to help scientists,
+  mathematicians and programmers discover unexpected properties of their models,
+  theories and programs; but has so far been limited to small-scale experiments,
+  with hand-picked data and qualitative results. We address these problems by
+  introducing a novel benchmark suite, orders of magnitude larger than previous
+  data sets, and use it to evaluate a variety of theory exploration approaches
+  (both existing and novel) and measure how they scale. It is our hope that the
+  availability of a robust method for comparing systems will encourage
+  competition and innovation in this field, as it has done for e.g.\ automated
+  theorem proving.
 \end{abstract}
@@
 
-As computers become ever more important in research, industry and
+Many fields represent knowledge using \emph{formal systems}, for example
+scientific models, mathematical theories and computer programs. As requirements
+grow, these systems can be modified in two ways: becoming more complicated, and
+becoming more numerous. Complication makes a system's behaviour harder to
+predict, whilst proliferation dilutes the attention that can be given to each
+particular system. This impacts aspects like safety and security, for example
+proving that a program will never leak confidential information, or that a model
+of some mechanism will never enter an unsafe configuration.
+
+One way to mitigate these problems is through automation: \emph{automated theory
+exploration} is the task of discovering interesting properties and behaviours of
+a given formal system, such as a software library or set of equations. Such
+tools can be used by domain experts, to discover structure, redundancy and
+patterns in a system; and, more importantly, to be \emph{surprised} by
+unexpected insights.
+
+\section{Background}\label{background}
+
+\begin{figure}
+  \begin{equation*}
+    plus(zero,    y) = y \\
+    plus(succ(x), y) = succ(plus(x, y)) \\
+    length(nil)        = zero \\
+    length(cons(x, y)) = succ(length(y)) \\
+    append(nil,       z) = z \\
+    append(cons(x,y), z) = cons(x, append(y, z)) \\
+    map(f, nil)        = nil \\
+    map(f, cons(x, y)) = cons(f(x), map(f, y))
+  \end{equation*}
+  \caption{Functions.}
+  \label{fig:defs}
+\end{figure}
+
+Our running example will be the software library\footnote{Our treatment of
+  software is directly applicable to mathematics \cite{wadler2015propositions},
+  and is also applicable to scientific models if we encode them as programs
+  which output predicted observations \cite{davies1990physical}
+} in figure~\ref{fig:defs}
+
+functions for manipulating data. This
+
+We can think of this as an axiomatic equational theory; it's also We will concentrate on the example of a software library, but this is
+
+The problem of \emph{automated theorem proving} (ATP) was one of the first tasks
+studied in the field of artificial intelligence \cite{newell1956logic,
+  sutcliffe2001evaluating}: given some statement about a formal system, prove
+that it is true. Many ATP approaches have been devised, and systems are compared
+by evaluating their performance on large benchmark problems, such as TPTP
+(Thousands of Problems for Theorem Provers).
+
+In contrast, theory \emph{exploration}progress in ATP has been through progTheory theorem proving a the  experts with a system for  discovering the behaviour of such systems has impact may  can exhibit Both present challenges to understandingare As these systems become more complexcomputers become ever more important in research, industry and
 society, so does the ability to program them effectively, economically
 and \emph{correctly}. Whilst tools exist to help programmers understand
-and manipulate software, ranging from linters \cite{Johnson78lint} through to formal
-verification systems \cite{boyer1983proof}, these are often limited in scope (e.g. to syntax rather than semantics) or require precise guidance from the user. One
-reason is that automating the search for patches \cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings,
-optimisations \cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other routine programming tasks is
-prohibitively expensive due to the size of the search space.
+and manipulate software, ranging from linters \cite{Johnson78lint} through to
+formal verification systems \cite{boyer1983proof}, these are often limited in
+scope (e.g.\ to syntax rather than semantics) or require precise guidance from
+the user. One reason is that automating the search for patches
+\cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings, optimisations
+\cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other
+routine programming tasks is prohibitively expensive due to the size of the
+search space.
 
@@
 To reduce the user's need to cherry-pick signatures, our approach
-accepts a large signature (e.g. a complete Haskell package) and uses
+accepts a large signature (e.g.\ a complete Haskell package) and uses
 statistical machine learning methods to select smaller signatures
@@
 Recurrent clustering attempts to cluster expressions based on their
-Abstract Syntax Trees (ASTs), an example of which is shown in Figure \ref{fig:astexample}. First the recursively-structured ASTs are
+Abstract Syntax Trees (ASTs), an example of which is shown in Figure~\ref{fig:astexample}. First the recursively-structured ASTs are
 transformed into vectors of fixed length, by truncating the tree
@@
     \end{small}
-    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc. are chosen arbitrarily.}
+    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc.\ are chosen arbitrarily.}
     \label{fig:astexample}
@@
 
-Figure \ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:
+Figure~\ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:
 
@@
 Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own.
-Whilst we have worked around these issues, e.g.
-using meta-programming, this introduces unnecessary latency, fragility and complexity.
+Whilst we have worked around these issues, e.g.\ using meta-programming, this introduces unnecessary latency, fragility and complexity.
 
@@
 \bibliographystyle{plain}
-\bibliography{../Bibtex}
+\bibliography{./Bibtex}
 

### 2017-04-24 PhDSymposium2017/abstract/abstract.tex
Symposium updates

@@
 \usepackage{tikz-qtree}
+\usepackage{enumitem}
 \usepackage{ifxetex,ifluatex}
 \usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\ifnum0\ifxetex{1}\fi\ifluatex{1}\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
@@
 \else % if luatex or xelatex
-  \ifxetex
+  \ifxetex{
     \usepackage{mathspec}
@@
     \usepackage{fontspec}
-  \fi
+  }\fi
   \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
@@
 }{}
-\ifxetex
+\ifxetex%
   \usepackage[setpagesize=false, % page size defined by xetex
@@
 \setcounter{secnumdepth}{5}
+
+\newcommand{\name}[1]{\mathrm{#1}}
+
 \newcommand{\feature}[1]{\phi(#1)}
-\newcommand{\id}[1]{\texttt{"#1"}}
+\newcommand{\id}[1]{\texttt{``#1''}}
 \newcommand{\CVar}{\texttt{Var}}
@@
 
+\newlist{inlinelist}{enumerate*}{1}
+\setlist*[inlinelist,1]{%
+  label= (\roman*),
+}
+
 \title{Quantitative Benchmarks for Theory Exploration}
@@
   availability of a robust method for comparing systems will encourage
-  competition and innovation in this field, as it has done for e.g.\ automated
-  theorem proving.
+  competition and innovation in this field.
 \end{abstract}
@@
 scientific models, mathematical theories and computer programs. As requirements
-grow, these systems can be modified in two ways: becoming more complicated, and
-becoming more numerous. Complication makes a system's behaviour harder to
+increase, these systems can grow in two ways: becoming more complex, and
+becoming more numerous. Complexification makes a system's behaviour harder to
 predict, whilst proliferation dilutes the attention that can be given to each
 particular system. This impacts aspects like safety and security, for example
-proving that a program will never leak confidential information, or that a model
-of some mechanism will never enter an unsafe configuration.
+proving that software will not leak confidential information, or that mechanisms
+will not enter unsafe configurations.
 
-One way to mitigate these problems is through automation: \emph{automated theory
-exploration} is the task of discovering interesting properties and behaviours of
-a given formal system, such as a software library or set of equations. Such
-tools can be used by domain experts, to discover structure, redundancy and
-patterns in a system; and, more importantly, to be \emph{surprised} by
-unexpected insights.
+One way to handle this growth is through automation: formal systems are easily
+represented in software, and techniques like \emph{automated theory exploration}
+(ATE) and \emph{automated theorem proving} (ATP) can help experts to discovery
+and verify the properties they do or do not want the system to exhibit. ATP is a
+mature field, with many competing implementations and competitive benchmarks. In
+contrast, ATE has seen little attention, and existing systems have mostly been
+evaluated qualitatively, on only a few hand-picked examples.
+
+We propose a benchmarking methodology for ATE systems, to evaluate and compare
+their performance on many large problems. We apply our benchmark to existing ATE
+systems, as well as two novel approaches.
 
@@
   \begin{equation*}
-    plus(zero,    y) = y \\
-    plus(succ(x), y) = succ(plus(x, y)) \\
-    length(nil)        = zero \\
-    length(cons(x, y)) = succ(length(y)) \\
-    append(nil,       z) = z \\
-    append(cons(x,y), z) = cons(x, append(y, z)) \\
-    map(f, nil)        = nil \\
-    map(f, cons(x, y)) = cons(f(x), map(f, y))
+    \begin{aligned}
+      \name{plus}(\name{zero},         y) &= y                                   \\
+      \name{plus}(\name{succ}(x),      y) &= \name{succ}(\name{plus}(x, y))      \\
+      \name{length}(\name{nil})           &= \name{zero}                         \\
+      \name{length}(\name{cons}(x, y))    &= \name{succ}(\name{length}(y))       \\
+      \name{append}(\name{nil},        z) &= z                                   \\
+      \name{append}(\name{cons}(x,y),  z) &= \name{cons}(x, \name{append}(y, z)) \\
+      \name{map}(f,           \name{nil}) &= \name{nil}                          \\
+      \name{map}(f,    \name{cons}(x, y)) &= \name{cons}(f(x), \name{map}(f, y))
+    \end{aligned}
   \end{equation*}
-  \caption{Functions.}
-  \label{fig:defs}
+  \caption{Definitions for $\name{plus}$, $\name{length}$, $\name{append}$ and
+    $\name{map}$. $\name{zero}$ and $\name{succ}$, $\name{nil}$ and
+    $\name{cons}$ are symbols encoding natural numbers and lists, respectively.
+    $\name{constants}$ and $variables$ are distinguished by
+    typeface.}\label{fig:defs}
 \end{figure}
 
-Our running example will be the software library\footnote{Our treatment of
-  software is directly applicable to mathematics \cite{wadler2015propositions},
-  and is also applicable to scientific models if we encode them as programs
-  which output predicted observations \cite{davies1990physical}
-} in figure~\ref{fig:defs}
+Our running example will be the small software library defined in
+figure~\ref{fig:defs}\footnote{Our approach can also be applied directly to
+  mathematical theories~\cite{wadler2015propositions}, and to scientific models
+  encoded as programs~\cite{davies1990physical}}. One property of this library
+is that the following programs are equivalent (for all $f$, $x$ and $y$):
 
-functions for manipulating data. This
+\begin{equation}\label{eq:mapappend}
+  \name{map}(f, \name{append}(x, y)) =
+    \name{append}(\name{map}(f, x), \name{map}(f, y))
+\end{equation}
 
-We can think of this as an axiomatic equational theory; it's also We will concentrate on the example of a software library, but this is
+The $\name{map}$ function applies an arbitrary function to all of the elements
+in a list; $\name{append}$ joins one list to the end of another. Property
+\ref{eq:mapappend} states that $\name{map}$ distributes over $\name{append}$,
+i.e.\ we can perform $\name{map}$ and $\name{append}$ in either order and get
+the same result. This is useful, since it may be much faster to run
+$\name{map}(f, x)$ and $\name{map}(f, y)$ in parallel on different machines,
+then $\name{append}$ the results; compared to $\name{append}$ing the inputs and
+running one big $\name{map}$.\footnote{This is one half of the ``MapReduce''
+  model for parallel programming.~\cite{dean2008mapreduce}}
 
-The problem of \emph{automated theorem proving} (ATP) was one of the first tasks
-studied in the field of artificial intelligence \cite{newell1956logic,
-  sutcliffe2001evaluating}: given some statement about a formal system, prove
-that it is true. Many ATP approaches have been devised, and systems are compared
-by evaluating their performance on large benchmark problems, such as TPTP
-(Thousands of Problems for Theorem Provers).
+There are two aspects to finding these properties:
 
-In contrast, theory \emph{exploration}progress in ATP has been through progTheory theorem proving a the  experts with a system for  discovering the behaviour of such systems has impact may  can exhibit Both present challenges to understandingare As these systems become more complexcomputers become ever more important in research, industry and
-society, so does the ability to program them effectively, economically
-and \emph{correctly}. Whilst tools exist to help programmers understand
-and manipulate software, ranging from linters \cite{Johnson78lint} through to
-formal verification systems \cite{boyer1983proof}, these are often limited in
-scope (e.g.\ to syntax rather than semantics) or require precise guidance from
-the user. One reason is that automating the search for patches
-\cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings, optimisations
-\cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other
-routine programming tasks is prohibitively expensive due to the size of the
-search space.
+\begin{enumerate}
+\item We must \emph{state} the property, singling it out for attention, because
+  it is somehow \emph{interesting}.
 
-We investigate the use of machine learning techniques to restrict such
-search spaces more intelligently than uninformed local search,
-specifically the use of \emph{clustering} to aid \emph{theory
-exploration} (TE) in discovering program equivalences.
+\item We must \emph{demonstrate} the property, providing evidence such as a
+  formal proof or empirical tests.
+\end{enumerate}
 
+\subsection{Automated Theorem Proving}
+
+Much work has been done on \emph{demonstrating} such properties, through the use
+of automated theorem proving. ATP has been an active area of study in artificial
+intelligence since the field was founded~\cite{newell1956logic,
+  sutcliffe2001evaluating}, and modern ATP systems such as CVC4, SPASS and
+Vampire take part in regular competitions~\cite{CASC}: attempting to prove as
+much as possible, as quickly as possible, from large benchmark suites such as
+TPTP (Thousands of Problems for Theorem Provers)~\cite{TPTP}.
+
+\subsection{Theory Exploration}
+
+We focus on the less-developed problem of \emph{conjecturing} properties:
+choosing some particular property, from the infinite number of possibilities,
+because it is somehow ``interesting''. Many measures of interestingness have
+been proposed\cite{INTERESTING}; John Conway proposed that a good conjecture
+should be ``outrageous''\cite{CONWAY}, but we also need some degree of
+\emph{plausibility} to avoid obvious falsehoods, even if we defer full proofs
+to a later stage.
+
+A particularly simple way to check if a property is interesting is to see if a
+human has already bothered to state it. If the theory given to an ATE system has
+already been studied by mathematicians, we can judge the quality of the system's
+output by comparing it against a corpus of known properties.
+
+Existing ATE systems have taken this approach, using theories and properties
+from the Isabelle proof assistant. Unfortunately, only two theories (natural
+numbers and lists) have been used to compare systems, and these only contain
+FIXME properties. This is not enough data to draw meaningful comparisons or
+extrapolations from.
+
+\section{Benchmarking}
+
+We propose a larger benchmark suite, containing many more definitions and
+properties, similar to those which drive competition in ATP. In fact, we can
+take the definitions and theorems from existing ATP benchmarks and repurpose
+them as the corpus of an ATE benchmark! We have chosen the TIP (Tons of
+Inductive Problems) benchmark~\cite{TIP} as the corpus for our approach, since
+it \begin{inlinelist}
+\item separates definitions from theorems
+\item allows higher-order functions (like $\name{map}$)
+\item incorporates many tricky problems from the theorem-proving literature
+\end{inlinelist}.
+
+Each of TIP's FIXME problems contains a set of definitions and a single theorem.
+We collect up all of the distinct definitions from all of the problems, to form
+one large theory. We collect up the theorems as our corpus, to compare ATE
+results against. Since FIXME is far larger than previously attempted theories,
+we also provide tools to uniformly sample a subset of definitions and their
+theorems.
+
+\section{Results}
+
+Figure FIXME compares the median run times of the existing
+\textsc{QuickSpec}\cite{QuickSpec} system across a variety of theory sizes,
+sampled from our benchmark. Figure FIXME shows the average proportion of corpus
+properties that were found for each sample. sizesFigure FIXME shows the run times of various
 \section{Background}\label{background}
@@
 Our work builds on the existing TE system
-\textsc{QuickSpec}\cite{QuickSpec}, which finds equivalent expressions
-in the Haskell programming language \cite{marlow2010haskell}, built from a user-provided
+, which finds equivalent expressions
+in the Haskell programming language~\cite{marlow2010haskell}, built from a user-provided
 \emph{signature} of constants and variables. For example, given a
@@
 enumerating all well-typed combinations of the signature's constants and
-variables, then randomly instantiating the variables \cite{claessen2011quickcheck} and comparing the
+variables, then randomly instantiating the variables~\cite{claessen2011quickcheck} and comparing the
 resulting closed terms. Any expressions which remain indistiguishable
@@
 after removing redundancies, are produced as output. These conjectures
-can be sent through a theorem prover \cite{claessen2013automating} and used to inform tests,
+can be sent through a theorem prover~\cite{claessen2013automating} and used to inform tests,
 optimisations, refactoring, verification, or simply to help the
@@
 To reduce the user's need to cherry-pick signatures, our approach
-accepts a large signature (e.g.\ a complete Haskell package) and uses
+accepts a large signature (e.g. a complete Haskell package) and uses
 statistical machine learning methods to select smaller signatures
 automatically, using a \emph{recurrent clustering} method inspired by
-those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.
+those of \textsc{ML4PG}~\cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)}~\cite{Heras.Komendantskaya.Johansson.ea:2013}.
 
@@
 
-These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices} \cite{de1972lambda}).
+These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices}~\cite{de1972lambda}).
 To replace global variables, $\phi$ first performs another round of clustering (hence the name
@@
 \begin{equation*}
-  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots
+  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots )
 \end{equation*}
@@
 We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered.
-Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which a) have an argument type which we can randomly generate and b) have an output type which we can compare.
-\textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.
+Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which
+\begin{inlinelist}
+\item have an argument type which we can randomly generate
+\item have an output type which we can compare
+\end{inlinelist}. \textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.
 

### 2017-04-25 PhDSymposium2017/abstract/abstract.tex
Submitted version for PhDSymp17

--- 2016-04-19 PhDSymposium2016/abstract.tex
@@
 
-\title{Exploring Software with Symbolic and Statistical Algorithms}
-  \author{
-          Chris Warburton \\
+\newlist{inlinelist}{enumerate*}{1}
+\setlist*[inlinelist,1]{%
+  label= (\roman*),
+}
+
+\title{Quantitative Benchmarks for Theory Exploration}
+  \author{Chris Warburton \\
           University of Dundee \\
-      \texttt{cmwarburton@dundee.ac.uk}
-    }
+          \texttt{cmwarburton@dundee.ac.uk}}
 \date{}
@@
 \begin{abstract}
-We aim to aid programmers in understanding and maintaining software
-by increasing the capability of \emph{theory exploration} systems for
-discovering novel, unexpected properties of programs and libraries.
-Existing theory exploration systems find equivalent program fragments via brute-force
-search, which doesn't scale well to realistic codebases. We narrow down
-the scope of each exploration run using statistics derived from the
-semantic similarity of the source code being explored, using a process
-called \emph{recurrent clustering}.
+  \emph{Automated theory exploration} has the potential to help scientists,
+  mathematicians and programmers discover unexpected properties of their models,
+  theories and programs; but has so far been limited to small-scale experiments,
+  with only a few, hand-picked, data points. We address this problem by
+  re-purposing existing benchmarks from the theorem proving domain, producing an
+  order of magnitude more data for evaluation, and use it to evaluate the
+  performance of a state of the art exploration system. To combat the long run
+  times of these systems, we also propose a pre-processing method to identify
+  promising subsets of a theory, sacrificing completeness to gain speed. It is
+  our hope that the availability of robust evaluation methods, and scalable
+  algorithms, will encourage competition and innovation, making theory
+  exploration more useful and relevant as a formal methods tool.
 \end{abstract}
@@
 
-As computers become ever more important in research, industry and
-society, so does the ability to program them effectively, economically
-and \emph{correctly}. Whilst tools exist to help programmers understand
-and manipulate software, ranging from linters \cite{Johnson78lint} through to formal
-verification systems \cite{boyer1983proof}, these are often limited in scope (e.g. to syntax rather than semantics) or require precise guidance from the user. One
-reason is that automating the search for patches \cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings,
-optimisations \cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other routine programming tasks is
-prohibitively expensive due to the size of the search space.
+Many fields represent knowledge using \emph{formal systems}, for example
+scientific models, mathematical theories and computer programs. These systems
+can be extended to meet new demands in two ways: the systems can be made more
+complex, or many new systems can be defined. Complexification makes a system's
+behaviour harder to predict, whilst proliferation dilutes the attention that can
+be given to each particular system. This presents a challenge for experts
+needing to characterise the behaviour of such systems, for example to prove
+properties related to protocol security or mechanical safety.
 
-We investigate the use of machine learning techniques to restrict such
-search spaces more intelligently than uninformed local search,
-specifically the use of \emph{clustering} to aid \emph{theory
-exploration} (TE) in discovering program equivalences.
+One way to help these experts is through automation: formal systems are easily
+represented as software, and techniques like \emph{automated theory exploration}
+can help experts to discover properties of a system, whether expected or
+surprising, desirable or not.
 
@@
 
-\subsection{Theory Exploration}\label{theory-exploration}
+An example of a simple formal system is the following software
+library:
 
-Our work builds on the existing TE system
-\textsc{QuickSpec}\cite{QuickSpec}, which finds equivalent expressions
-in the Haskell programming language \cite{marlow2010haskell}, built from a user-provided
-\emph{signature} of constants and variables. For example, given a
-signature of the list-manipulating functions \texttt{reverse},
-\texttt{length} and \texttt{append}, plus two list variables \texttt{x}
-and \texttt{y}, \textsc{QuickSpec} will discover that
-\texttt{reverse\ (reverse\ x)} is equivalent to \texttt{x}; that
-\texttt{length\ (reverse\ x)} is equivalent to \texttt{length\ x}; that
-\texttt{length\ (append\ x\ y)} is equivalent to
-\texttt{length\ (append\ y\ x)}; and so on. These are discovered by
-enumerating all well-typed combinations of the signature's constants and
-variables, then randomly instantiating the variables \cite{claessen2011quickcheck} and comparing the
-resulting closed terms. Any expressions which remain indistiguishable
-after hundreds of instantiations are conjectured to be equivalent and,
-after removing redundancies, are produced as output. These conjectures
-can be sent through a theorem prover \cite{claessen2013automating} and used to inform tests,
-optimisations, refactoring, verification, or simply to help the
-programmer learn more about the code.
+\begin{equation*}
+  \begin{aligned}
+    \name{append}(\name{nil},        z) &= z                                   \\
+    \name{append}(\name{cons}(x,y),  z) &= \name{cons}(x, \name{append}(y, z)) \\
+    \name{map}(f,           \name{nil}) &= \name{nil}                          \\
+    \name{map}(f,    \name{cons}(x, y)) &= \name{cons}(f(x), \name{map}(f, y))
+  \end{aligned}
+\end{equation*}
 
-Due to its exponential complexity, this enumerating procedure is limited
-to producing small expressions from signatures with only a few elements.
-Such small signatures give little chance for serendipitous discoveries,
-which undermines the algorithm's potential to present programmers with
-new, unexpected information.
+These equations define two functions, $\name{append}$ (for joining one list to
+another) and $\name{map}$ (for applying a function to the elements of a list),
+based on the way they rearrange particular constants ($\name{nil}$ and
+$\name{cons}$, which encode lists à la Lisp) and variables ($f$, $x$, $y$, $z$).
+One \emph{property} of this library is that the following expressions are
+equivalent:
 
-\subsection{Recurrent Clustering}\label{recurrent-clustering}
+\begin{equation}\label{eq:mapappend}
+  \name{map}(f, \name{append}(x, y)) =
+    \name{append}(\name{map}(f, x), \name{map}(f, y))
+\end{equation}
 
-To reduce the user's need to cherry-pick signatures, our approach
-accepts a large signature (e.g. a complete Haskell package) and uses
-statistical machine learning methods to select smaller signatures
-automatically, using a \emph{recurrent clustering} method inspired by
-those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.
+This states that $\name{map}$ distributes over $\name{append}$, i.e.\ we can
+perform $\name{map}$ and $\name{append}$ in either order and get the same
+result. This is useful, since it may be much faster to run $\name{map}(f, x)$
+and $\name{map}(f, y)$ in parallel on different machines, then $\name{append}$
+the results; compared to $\name{append}$ing the inputs and running one big
+$\name{map}$.\footnote{This is one half of the ``MapReduce'' model for parallel
+  programming.~\cite{dean2008mapreduce}.}
 
-Recurrent clustering attempts to cluster expressions based on their
-Abstract Syntax Trees (ASTs), an example of which is shown in Figure \ref{fig:astexample}. First the recursively-structured ASTs are
-transformed into vectors of fixed length, by truncating the tree
-structure and listing the node labels in breadth-first order.
+Most automation of formal systems focuses on \emph{automated theorem proving}
+(ATP), which dates back to the early days of artificial
+intelligence~\cite{newell1956logic}. In ATP, the user gives a particular
+property statement (like \ref{eq:mapappend}) to the program up front, and the
+program searches for a proof. Modern ATP systems, such as
+CVC4~\cite{barrett2011cvc4}, compete regularly to prove as many benchmark
+problems as they can, as quickly as possible~\cite{sutcliffe2006state}.
 
-These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices} \cite{de1972lambda}).
-To replace global variables, $\phi$ first performs another round of clustering (hence the name
-``recurrent''), without including the current expression; each global
-variable is replaced by its cluster index found by this ``inner'' round
-of clustering.
+\emph{Automated Theory Exploration} (ATE) is the task of taking in a formal
+system (such as the definitions of $\name{map}$ and $\name{append}$ above) and
+giving out interesting properties, like \ref{eq:mapappend}.\footnote{ATE doesn't
+  require a proof of these properties; we can use ATP tools for that.} ATE has
+seen little attention, partly because of its resource usage, and partly because
+it is difficult to evaluate: what makes a property ``interesting''? Many general
+measures have been proposed\cite{colton2000notion}, but for comparison purposes
+we can stick to known theories and simply check if a human has bothered to state
+these properties before!
+
+Existing ATE programs have been evaluated this way on theories of lists and
+natural numbers taken from the Isabelle proof
+assistant~\cite{claessen2013automating}. The discovered properties are compared
+against those in Isabelle, to see how many match up. Unfortunately these are the
+only data sets used, and each is rather small (containing 9 and 12 properties,
+respectively). This is not enough data from which to draw meaningful comparisons
+or extrapolations.
+
+\section{Contributions}
+
+\subsection{Benchmarking}
+
+We address the problem of ATE evaluation by implementing a much larger benchmark
+suite, containing 343 properties of 269 definitions. We take these from an
+existing ATP benchmark called TIP\cite{claessen2015tip}, since it
+\begin{inlinelist}
+\item separates definitions from theorems
+\item allows higher-order functions (like $\name{map}$)
+\item incorporates many tricky problems from the theorem-proving literature
+\item contains problems of immediate relevance to software verification
+\end{inlinelist}.
+
+Each TIP property is given alongside all of its required definitions. We
+re-purpose these for the ATE setting, by pooling together all of
+the definitions\footnote{Including custom functions to construct and destruct
+  constant values (like $\name{nil}$ and $\name{cons}$), we get 3595
+  definitions, most of which are redundant copies used in many problems.
+  Discarding these leaves 269 distinct definitions.} and all of the properties.
+
+We provide a tool to sample subsets of these definitions, along with the subset
+of properties which apply to that sample, so that theories of various sizes can
+be generated for evaluating ATE systems.
+
+\subsection{Scaling}\label{recurrent-clustering}
+
+In the worst case, ATE scales exponentially with the number of definitions in a
+theory. We have also investigated ways to mitigate this using \emph{relevance
+  filtering}~\cite{meng2009lightweight}: given a large theory, can we explore
+only those definitions which we expect to yield interesting properties? We have
+implemented such a filter based on
+\emph{recurrent clustering}~\cite{journals/corr/abs-1212-3618}, which compares
+the syntax trees of a theory's definitions to find those which are similar. We
+have also made a purely random filter, to control for the effect of
+\emph{filtering itself}. These can be used as a pre-processor for any ATE system
+and work is ongoing to evaluate their effect on property recall and performance.
+
+\section{Results}
 
 \begin{figure}
-    \begin{small}
-      \begin{tikzpicture}[sibling distance=0pt]
-        \tikzset{sibling distance=0pt}
-        \tikzset{level distance=20pt}
-        \Tree[ .$\CLam$
-                  $\id{a}$
-                  [ .$\CCase$
-                       [ .$\CVar$
-                            [ .$\CLocal$
-                                 $\id{a}$ ]]
-                       $\id{b}$
-                       [ .$\CAlt$
-                            $\CDataAlt$
-                            [ .$\CVar$
-                                 $\CConstructor$ ]]
-                       [ .$\CAlt$
-                            $\CDataAlt$
-                            [ .$\CApp$
-                                 [ .$\CVar$
-                                      [ .$\CGlobal$
-                                           $\id{even}$ ]]
-                                 [ .$\CVar$
-                                      [ .$\CLocal$
-                                           $\id{n}$ ]]]
-                            $\id{n}$ ]]]
-      \end{tikzpicture}
-    \end{small}
-    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc. are chosen arbitrarily.}
-    \label{fig:astexample}
+  \centering
+  \includegraphics[width=0.45\textwidth]{runtimes}
+  \caption{Median runtime of the QuickSpec ATE system for 30 iterations on
+    theories of various size, sampled from TIP. Error bars show the
+    interquartile range. We time out at 1000 seconds.}\label{fig:runtimes}
 \end{figure}
 
-These recursive calls to the clustering algorithm (we use a standard
-\emph{k-means} implementation) allow expressions to be compared in a
-principled way: the similarity of expressions depends on the similarity of the expressions they reference, and so on recursively. In practice, this recursive
-algorithm can be implemented in an iterative way, by accumulating the set of expressions to cluster in topological order of their dependency graph.
+Figure~\ref{fig:runtimes} shows the median run time of the existing
+\textsc{QuickSpec}\cite{QuickSpec} system as we explore theories of various size
+sampled from our benchmark. We have found that the time taken can vary greatly,
+even among theories of the same size, with most finishing after a few seconds
+but some taking hours; this can be seen by the increase in range for larger
+theories, despite the median remaining low.
 
-Figure \ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:
+This variance makes current ATE systems inconvenient for widespread use. Finding
+a pattern in these long-running inputs is current work; our hope is to make such
+tools more predictable and tractable using approaches like our relevance filter.
 
-\begin{lstlisting}[language=Haskell]
-odd  Z    = False
-odd (S n) = even n
+\section{Future Work}\label{future-work}
 
-even  Z    = True
-even (S n) = odd n
-\end{lstlisting}
+There are many promising directions that this work enables. Our benchmark is
+currently limited to ATE systems using Haskell; integrating other languages
+would allow wider comparisons, e.g. to IsaCoSy and IsaScheme which use Isabelle.
 
-The \texttt{odd} function references four global variables: \texttt{Z}, \texttt{S}, \texttt{False} and \texttt{even}.
-Our algorithm doesn't yet distinguish between data constructors, so the only global reference we replace is \texttt{even}.
-Likewise, the definition of \texttt{even} references \texttt{Z}, \texttt{S}, \texttt{True} and \texttt{odd}.
-For mutual recursion like this, there is no valid topological order, so we use a sentinel value as the feature.
+Machine learning is a promising direction for scaling ATE, since its approximate
+methods may be more efficient than existing searches, and its data-driven
+approach is well suited to handle fuzzy concepts such as ``interestingness''.
+Our benchmark can act as both a fitness function and a training corpus for such
+efforts.
 
-These feature vectors are sent through a k-means clustering algorithm, using the \textsc{Weka} machine learning library.
-For example, the feature vector for \texttt{odd} (padded to 6 labels for each level of the tree) will begin:
-
-\begin{equation*}
-  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots
-\end{equation*}
-
-\section{Implementation}\label{implementation}
-
-We obtain ASTs from Haskell projects using a bespoke plugin for the GHC
-compiler.
-From these ASTs, we extract type information which is used to append variables to the signature; and dependency information, which is used to topologically sort the clustering rounds.
-We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered.
-Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which a) have an argument type which we can randomly generate and b) have an output type which we can compare.
-\textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.
-
-\section{Results}\label{results}
-
-We have developed a machine learning approach for analysing Haskell code, including a
-bespoke feature extraction method and a full implementation pipeline for turning Haskell packages into sets of equations.
-
-\section{Difficulties and Future Work}\label{future-work}
-
-The most difficult aspect of performing this exploration was obtaining real code in a usable format, due to the myriad edge-cases encountered.
-We make extensive use of Haskell's existing infrastructure, including the \textsc{GHC} compiler, the \textsc{cabal} build system, the \textsc{Hackage} code repository and the \textsc{Nix} package manager.
-Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own.
-Whilst we have worked around these issues, e.g.
-using meta-programming, this introduces unnecessary latency, fragility and complexity.
-
-Our investigation of recurrent clustering and theory exploration only scratches the surface of combining symbolic and statistical AI algorithms.
-As well as benchmarking our approach against other techniques, there are many similar opportunities to be explored, where the reasoning power of symbolic algorithms can be guided and supervised by statistical, data-driven processes.
+Our system can explore \emph{arbitrary} code, not just our benchmarks. Besides
+listing properties, potential applications include tools for optimisation (if
+coupled to a profiler) and refactoring (e.g. using properties to propose new
+abstractions).
 
 \bibliographystyle{plain}
-\bibliography{../Bibtex}
+\bibliography{./Bibtex}
 

### 2017-05-05 PhDSymposium2017/abstract/abstract.tex
Final fixes

@@
 
+\setlength{\belowcaptionskip}{-10pt}
+
 \newlist{inlinelist}{enumerate*}{1}
@@
   algorithms, will encourage competition and innovation, making theory
-  exploration more useful and relevant as a formal methods tool.
+  exploration more useful and relevant in formal methods.
 \end{abstract}
+
+%\vspace{-2mm}
+
+\textbf{Keywords:} Theory exploration, automated reasoning, automated theorem
+proving
+
+%\vspace{-2mm}
 
@@
 
+\vspace{-2mm}
+
 \section{Background}\label{background}
@@
 library:
+
+\vspace{-5mm}
 
@@
 equivalent:
+
+\vspace{-3mm}
 
@@
 
-Existing ATE programs have been evaluated this way on theories of lists and
-natural numbers taken from the Isabelle proof
+Existing ATE programs such as \textsc{QuickSpec}~\cite{QuickSpec} and
+\textsc{IsaCoSy}~\cite{johansson2009isacosy} have been evaluated this way on
+theories of lists and natural numbers taken from the Isabelle proof
 assistant~\cite{claessen2013automating}. The discovered properties are compared
@@
   \includegraphics[width=0.45\textwidth]{runtimes}
-  \caption{Median runtime of the QuickSpec ATE system for 30 iterations on
-    theories of various size, sampled from TIP. Error bars show the
+  \vspace{-3mm}
+  \caption{Median runtime of the \textsc{QuickSpec} ATE system for 30 iterations
+    on theories of various size, sampled from TIP. Error bars show the
     interquartile range. We time out at 1000 seconds.}\label{fig:runtimes}
+  \vspace{-5mm}
 \end{figure}
 
-Figure~\ref{fig:runtimes} shows the median run time of the existing
-\textsc{QuickSpec}\cite{QuickSpec} system as we explore theories of various size
-sampled from our benchmark. We have found that the time taken can vary greatly,
-even among theories of the same size, with most finishing after a few seconds
-but some taking hours; this can be seen by the increase in range for larger
-theories, despite the median remaining low.
+Figure~\ref{fig:runtimes} shows the median run time of \textsc{QuickSpec} as we
+explore theories of various size sampled from our benchmark. We have found that
+the time taken can vary greatly, even among theories of the same size, with most
+finishing after a few seconds but some taking hours; this can be seen by the
+increase in range for larger theories, despite the median remaining low.
 
 This variance makes current ATE systems inconvenient for widespread use. Finding
-a pattern in these long-running inputs is current work; our hope is to make such
-tools more predictable and tractable using approaches like our relevance filter.
+a pattern in these long-running inputs is current work; our hope is that
+approaches like our relevance filter make ATE more predictable and tractable.
 

### 2017-08-07 TEBenchmarkPaper/article.tex
Rename benchmark paper

--- 2017-04-19 benchmark2016/article.tex
@@
-\title{An Approach to Benchmarking Conjecture Generators}
-\author{
-        Chris Warburton \\
-                University of Dundee\\
-}
-\date{\today}
-
-\documentclass[12pt]{article}
+% Derived from the template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+\begin{filecontents*}{example.eps}
+%!PS-Adobe-3.0 EPSF-3.0
+%%BoundingBox: 19 19 221 221
+%%CreationDate: Mon Sep 29 1997
+%%Creator: programmed by hand (JK)
+%%EndComments
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
+%
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
 
@@
 
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 \begin{document}
+
+\title{Quantitative Benchmarking for Theory Exploration%\thanks{Grants or other notes
+%about the article that should go on the front page should be
+%placed here. General acknowledgments should be placed at the end of the article.}
+}
+% \subtitle{Do you have a subtitle?\\ If so, write it here}
+
+% \titlerunning{Short form of title}        % if too long for running head
+
+\author{Chris Warburton% \and
+                %Second Author % etc
+}
+
+% \authorrunning{Short form of author list} % if too long for running head
+
+\institute{C. Warburton \at
+              University of Dundee \\
+              \email{cmwarburton@dundee.ac.uk}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           %\and
+           %S. Author \at
+           %   second address
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
@@
 \begin{abstract}
-
+Insert your abstract here. Include keywords, PACS and mathematical
+subject classification numbers as needed.
+\keywords{First keyword \and Second keyword \and More}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
 
-\begin{displayquote}
-  There is frequently more to be learn'd from the unexpected questions of a
-  child than the discourses of men.
-  \attrib{--- John Locke}
-\end{displayquote}
-
 \section{Introduction}
-
+\label{intro}
 The field of mathematics has been transformed by the introduction of computers:
@@
 first tasks studied in the field of artificial intelligence
-\cite{newell1956logic, sutcliffe2001evaluating}. Despite this effort, ATP
+\cite{newell1956logic,sutcliffe2001evaluating}. Despite this effort, ATP
 methods are not widely used by traditional mathematicians, although they have
@@
 choose a well-studied theory, with a set of existing theorems that are deemed to
-be useful or important in some way, and treat these as a \em{ground truth}
+be useful or important in some way, and treat these as a \emph{ground truth}
 against which we can compare conjectures generated from the same theory.
@@
 
-\section{Existing Evaluations}\label{previous approaches}
+\section{Existing Evaluations}
+\label{sec:previous}
 
-Theory exploration tools such as IsaCoSy \cite{Johansson.Dixon.Bundy:conjecture-generation}, QuickSpec and IsaScheme have been evaluated using the standard library of the Isabelle theorem prover as their ground truth.
+Theory exploration tools such as
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}, QuickSpec and
+IsaScheme have been evaluated using the standard library of the Isabelle theorem
+prover as their ground truth.
 
@@
 \section{Evaluation}
+\label{sec:evaluation}
 
-\includegraphics{bucketing-graph}
+%\includegraphics{bucketing-graph}
+
+\begin{acknowledgements}
+We are grateful for Jianguo Zhang
+\end{acknowledgements}
 

### 2017-08-07 TEBenchmarkPaper/article.tex
Added meeting notes

@@
 \author{Chris Warburton% \and
-                %Second Author % etc
+        Alison Pease
+        % TODO: Invite Katya? Even if she only offers some help, can acknowledge
 }
@@
 \maketitle
+IGNORE ISABELLE THEORIES FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+IGNORE CLUSTERING FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+
+intro (motivation, high-level background, eval is a problem),
+background,
+our proposal,
+application (evaluation),
+discussion (e.g. philosophical bits),
+future work, conclusions, etc.
+
+COMPARE COLTON ET AL.
+
+Evaluation of conjectures, sets of conjectures, systems
+what constitutes a good/desired output (in each author's terms)? This combines
+their evaluation with their (system-specific) assumptions (e.g. IsaCoSy doesn't
+generate redundant terms: non-redundancy is clearly a goal, but it's not measured
+in eval since it's assumed based on system's construction)
+
+Look carefully at each system's (paper's) eval section
+Make a table like tabl 1 of Colton et al
+
+Make it clear that methodology is the point (it's not just a detail, like in a
+chemistry paper for example)
 
 \begin{abstract}
+  % TODO: Write abstract sooner rather than layer
 Insert your abstract here. Include keywords, PACS and mathematical
@@
 \label{intro}
+
+% TODO: Preaching to the choir;
+% Jumping around a bit, rephrase
+% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
+% We can just jump straight in to automated conjecture generation
+
+% Maybe lay out argument in bullet points; flesh out sentences later
+
+% Lay out the field of TE: motivation, systems, evaluation, problems...
+% Compare to
+
 The field of mathematics has been transformed by the introduction of computers:
@@
 \cite{Moore:2003}.
+
+% FIXME: Be more specific about our contributions
 
@@
 example, a benchmark based on Isabelle/HOL's theory of natural numbers contains
-only FIXME definitions and FIXME theorems. Whilst such benchmarks allow
+only 4 definitions and 12 theorems. Whilst such benchmarks allow
 comparison between different approaches, their narrow scope doesn't provide much
@@
 Concatenate all definitions from all files together.
-For each definition, prefix its name with the filename it came from and update all references.
+For each definition, prefix its name with the filename it came from and update
+all references.
 Normalise all local variable names.
-Remove any identical definitions, leaving the first one in place and updating any references to the removed definitions to use this one.
+Remove any identical definitions, leaving the first one in place and updating
+any references to the removed definitions to use this one.
 Repeat, until no further removals are possible.
@@
 
-%\includegraphics{bucketing-graph}
+We have applied our benchmarking methodology to the QuickSpec and IsaCoSy theory
+exploration systems. To assess the scalability of their algorithms, we sampled
+theories of various sizes: from theories containing only a single definition, up
+to theories containing 20 definitions. These are shown in figure FIXME.
+
+To provide a more robust comparison of these two systems, we also applied a
+paired difference test: measuring, for each sampled theory, the difference in
+time, precision and recall between the two systems.
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+Our benchmark suite provides a unifying goal for the diverse approaches of
+theory exploration, whilst avoiding some of the philosophical complications of
+the field. Whilst the current implementation is quite limited, we welcome
+additions from other researchers, to more closely align the benchmark with their
+goals, and the strenghts and weaknesses of their systems.
+
+``Solving'' this benchmark suite would not solve the problem of theory exploration in general,
+so more ambitious goals must be set in the future; but we have found that there is still a
+long way to go until that problem arises.
 
 \begin{acknowledgements}
-We are grateful for Jianguo Zhang
+  We are grateful to Jianguo Zhang for help with our statistical analysis.
+
+  EPSRC
 \end{acknowledgements}

### 2017-09-07 TEBenchmarkPaper/article.tex
Change titlematter
Add abstract

@@
 
-\title{Quantitative Benchmarking for Theory Exploration%\thanks{Grants or other notes
+\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks{Grants or other notes
 %about the article that should go on the front page should be
@@
 
-\author{Chris Warburton% \and
+\author{Chris Warburton \and
         Alison Pease
@@
            University of Dundee \\
-              \email{cmwarburton@dundee.ac.uk}           %  \\
+           \email{c.m.warburton@dundee.ac.uk}           %  \\
 %             \emph{Present address:} of F. Author  %  if needed
-           %\and
-           %S. Author \at
-           %   second address
+           \and
+           A. Pease \at
+           University of Dundee \\
+           \email{a.pease@dundee.ac.uk}
 }
@@
 \begin{abstract}
-  % TODO: Write abstract sooner rather than layer
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
-\keywords{First keyword \and Second keyword \and More}
+  We propose a benchmark suite for evaluating the efficiency and effectiveness
+  of automated tools for \emph{theory exploration} or
+  \emph{conjecture formation} in higher-order, inductive theories; a domain
+  especially suitable for programming language libraries. By providing standard
+  tools and metrics, we hope to encourage innovation and comparison between the
+  disparate approaches currently being pursued, and spur improvements similar to
+  those seen in the competitive field of automated theorem proving.
+%\keywords{First keyword \and Second keyword \and More}
 % \PACS{PACS code1 \and PACS code2 \and more}

### 2017-09-07 TEBenchmarkPaper/article.tex
Add basic intro

@@
 % Maybe lay out argument in bullet points; flesh out sentences later
+Automated \emph{theory exploration} (ATE), also known as
+\emph{conjecture generation/formation}, is the open-ended problem of producing
+conjectures about a given logical theory which are somehow ``interesting''.
+
+This has applications wherever we find use for formal statements: for example in
+proof assistants and their libraries, in mathematics education and research, and
+(especially for our purposes) in the specification, verification, optimisation
+and testing of software.
+
+Existing attempts at tackling this problem are difficult to compare and study,
+due partially to the variety of approaches taken, but also because of the
+inherent ambiguity of the task and the different goals emphasised by the authors
+and their evaluation methods.
+
+Our contributions include:
+
+\begin{itemize}
+\item A general methodology for benchmarking ATE systems.
+\item Resolving the issue of ``interestingness'' through the use theorem-proving
+  benchmarks as a ground-truth.
+\item A specific instantiation of this methodology, using the Tons of Inductive
+  Problems benchmark as a corpus.
+\item Automated tooling to produce and analyse this benchmark, translating to
+  Haskell and Isabelle code for the QuickSpec and IsaCoSy systems, respectively.
+\end{itemize}
+
+Section \ref{sec:background} describes the ATE problem in more detail, along
+with existing approaches and their evaluation methodologies. We explain our
+proposal for a more general benchmark in section \ref{section:proposal} and
+section \ref{application} shows the results when applied to existing ATE tools.
+Analysis of these results is given in section \ref{section:discussion} and
+concluding remarks in \ref{section:conclusion}.
 

### 2017-09-07 TEBenchmarkPaper/article.tex
More fixing up

@@
 \maketitle
-IGNORE ISABELLE THEORIES FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-IGNORE CLUSTERING FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+%IGNORE ISABELLE THEORIES FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+%IGNORE CLUSTERING FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
 
-intro (motivation, high-level background, eval is a problem),
-background,
-our proposal,
-application (evaluation),
-discussion (e.g. philosophical bits),
-future work, conclusions, etc.
+%COMPARE COLTON ET AL.
 
-COMPARE COLTON ET AL.
+% Evaluation of conjectures, sets of conjectures, systems
+% what constitutes a good/desired output (in each author's terms)? This combines
+% their evaluation with their (system-specific) assumptions (e.g. IsaCoSy doesn't
+% generate redundant terms: non-redundancy is clearly a goal, but it's not measured
+% in eval since it's assumed based on system's construction)
 
-Evaluation of conjectures, sets of conjectures, systems
-what constitutes a good/desired output (in each author's terms)? This combines
-their evaluation with their (system-specific) assumptions (e.g. IsaCoSy doesn't
-generate redundant terms: non-redundancy is clearly a goal, but it's not measured
-in eval since it's assumed based on system's construction)
+% Look carefully at each system's (paper's) eval section
+% Make a table like tabl 1 of Colton et al
 
-Look carefully at each system's (paper's) eval section
-Make a table like tabl 1 of Colton et al
-
-Make it clear that methodology is the point (it's not just a detail, like in a
-chemistry paper for example)
+% Make it clear that methodology is the point (it's not just a detail, like in a
+% chemistry paper for example)
 
@@
 
-% TODO: Preaching to the choir;
-% Jumping around a bit, rephrase
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
+% Lay out the field of TE: motivation, systems, evaluation, problems...
 
-% Maybe lay out argument in bullet points; flesh out sentences later
 Automated \emph{theory exploration} (ATE), also known as
@@
 
-% Lay out the field of TE: motivation, systems, evaluation, problems...
-% Compare to
+%- Existing approaches are difficult to compare
+% - They want different things
+% - They're evaluated differently
+% - They're evaluated ``locally'' (as white boxen)
 
-The field of mathematics has been transformed by the introduction of computers:
-calculations are now routinely automated, allowing humans to focus their efforts
-on higher-level, more creative activities like proving theorems; \emph{automated
-theorem proving} (ATP) seeks to automate this process too, and was one of the
-first tasks studied in the field of artificial intelligence
-\cite{newell1956logic,sutcliffe2001evaluating}. Despite this effort, ATP
-methods are not widely used by traditional mathematicians, although they have
-seen use in applied areas such as hardware and software verification
-\cite{Moore:2003}.
+% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
+% We can just jump straight in to automated conjecture generation
 
-% FIXME: Be more specific about our contributions
+\section{Background}
+\label{section:background}
 
-Even if ATP became widespread, the automation of theorem proving would still
-leave an important aspect of mathematics inaccessible to computers: conjecturing
-those theorems (and the definitions they involve) in the first place. Automated
-methods for posing good conjectures have obvious applications for those who must
-invent a constant stream of novel questions, e.g. for examinations or research.
-A larger impact may be found in areas such as computer programming, security and
-verification, where the benefits of formal methods are hindered by the
-difficulty of their application; tools to aid this process, e.g. by proposing
-test cases, can reduce this cost.
+\begin{figure}
+  \label{figure:list_theory}
+  \begin{equation*}
+    Nil : \forall (a : Type). List a
+    Cons : \forall (a : Type). \forall (x : a). \forall (xs : List a). List a
 
-Many automated systems and algorithms have been developed to generate
+    Z : Nat
+    S : \forall (p : Nat). Nat
+
+    head(Cons(x, xs)) = x
+    tail(Cons(x, xs)) = xs
+    append(xs, Nil) = xs
+    append(Cons(x, xs), ys) = Cons x (append xs ys)
+    reverse(Nil) = Nil
+    reverse(Cons(x, xs)) = append(reverse(xs), Cons(x, Nil))
+    length(Nil) = Z
+    length(Cons(x, xs)) = S (length(xs))
+    map(f, Nil) = Nil
+    map(f, Cons(x, xs)) = Cons(f(x), map(f, xs))
+    foldl(f, x, Nil) = x
+    foldl(f, x, Cons(y, ys)) = foldl(f, f(x, y), xs)
+    foldr(f, Nil, y) = y
+    foldr(f, Cons(x, xs), y) = f(x, foldr(f, xs, y))
+  \end{equation*}
+  \caption{A simple theory defining $Nat$ and $List$ types and some associated
+    list operations, taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
+\end{figure}
+
+If we have a logical theory, for example the theory of lists shown in figure
+\ref{figure:list_theory} Many automated systems and algorithms have been developed to generate
 conjectures, but it is difficult to judge or compare their performance in a
-quantitative way. There are many aspects to consider when judging the quality of
+quantitative way. We consider three distinct aspects to evaluate:
+
+\begin{itemize}
+\item Each conjecture, to judge how ``interesting'' it is.
+\item The \emph{set} of conjectures produced, to assess things like consistency
+  and redundancy.
+\item Performance of the system, to find the balance struck between output
+  quality and time taken.
+\end{itemize}
+
+There are many aspects to consider when judging the quality of
 a conjecture: how likely it is to be true, how difficult it is to prove, its
@@
 One problem with this approach is the small size of these libraries. For
-example, a benchmark based on Isabelle/HOL's theory of natural numbers contains
+example, the benchmark based on Isabelle/HOL's theory of natural numbers given
+in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
 only 4 definitions and 12 theorems. Whilst such benchmarks allow
@@
 indication of performance in different, especially \emph{novel}, domains.
+
+\section{Our Proposal}
+\label{section:proposal}
 
@@
 
-\section{Existing Evaluations}
-\label{sec:previous}
+%ALGORITHM:
 
-Theory exploration tools such as
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}, QuickSpec and
-IsaScheme have been evaluated using the standard library of the Isabelle theorem
-prover as their ground truth.
+%Concatenate all definitions from all files together.
+%For each definition, prefix its name with the filename it came from and update
+%all references.
+%Normalise all local variable names.
+%Remove any identical definitions, leaving the first one in place and updating
+%any references to the removed definitions to use this one.
+%Repeat, until no further removals are possible.
 
-ALGORITHM:
+\section{Application}
+\label{section:application}
 
-Concatenate all definitions from all files together.
-For each definition, prefix its name with the filename it came from and update
-all references.
-Normalise all local variable names.
-Remove any identical definitions, leaving the first one in place and updating
-any references to the removed definitions to use this one.
-Repeat, until no further removals are possible.
-
-\section{Evaluation}
-\label{sec:evaluation}
+% evaluation
 
@@
 
+\section{Discussion}
+\label{section:application}
+
+% (e.g. philosophical bits),
+
 \section{Conclusion}
-\label{sec:conclusion}
+\label{section:conclusion}
+
+% future work, etc.
 

### 2017-09-07 TEBenchmarkPaper/article.tex
Fleshed out a bit

@@
 % insert here the call for the packages your document requires
-%\usepackage{latexsym}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{mathtools}
 % etc.
@@
 % Insert the name of "your journal" with
-% \journalname{myjournal}
-%
+\journalname{Journal of Automated Reasoning}
+
 \begin{document}
@@
   \emph{conjecture formation} in higher-order, inductive theories; a domain
-  especially suitable for programming language libraries. By providing standard
-  tools and metrics, we hope to encourage innovation and comparison between the
-  disparate approaches currently being pursued, and spur improvements similar to
-  those seen in the competitive field of automated theorem proving.
+  especially suited for analysing software. By providing standard tools and
+  metrics, we hope to encourage innovation and comparison between the disparate
+  approaches currently being pursued, and spur improvements similar to those
+  seen in the competitive field of automated theorem proving.
 %\keywords{First keyword \and Second keyword \and More}
@@
 
-Automated \emph{theory exploration} (ATE), also known as
+\emph{Automated theory exploration} (ATE), also known as
 \emph{conjecture generation/formation}, is the open-ended problem of producing
@@
 
-This has applications wherever we find use for formal statements: for example in
-proof assistants and their libraries, in mathematics education and research, and
-(especially for our purposes) in the specification, verification, optimisation
-and testing of software.
+This has applications wherever we find use for formal statements: in proof
+assistants and their libraries, in mathematics education and research, and
+(of special concern for the authors) in the specification, verification,
+optimisation and testing of software.
 
-Existing attempts at tackling this problem are difficult to compare and study,
-due partially to the variety of approaches taken, but also because of the
-inherent ambiguity of the task and the different goals emphasised by the authors
-and their evaluation methods.
+Existing attempts at tackling this problem have been found difficult to compare
+and study, due partially to the variety of approaches taken, but also because of
+the inherent ambiguity of the task and the different goals emphasised by their
+authors and evaluation methods.
 
-Our contributions include:
+We attempt to solve this discrepancy, at least for the foreseeable future, by
+defining a standard, unambiguous benchmarking approach with which to compare
+ATE systems. Our contributions include:
 
@@
   Problems benchmark as a corpus.
-\item Automated tooling to produce and analyse this benchmark, translating to
-  Haskell and Isabelle code for the QuickSpec and IsaCoSy systems, respectively.
+\item Automated tooling to produce this benchmark and analyse its output.
+\item Application of this methodology to the QuickSpec and IsaCoSy ATE systems,
+  and a discussion of the results.
 \end{itemize}
@@
 with existing approaches and their evaluation methodologies. We explain our
-proposal for a more general benchmark in section \ref{section:proposal} and
-section \ref{application} shows the results when applied to existing ATE tools.
-Analysis of these results is given in section \ref{section:discussion} and
-concluding remarks in \ref{section:conclusion}.
+proposal for a more general benchmark in section \ref{sec:proposal} and
+section \ref{sec:application} shows the results when applied to existing ATE tools.
+Analysis of these results is given in section \ref{sec:discussion} and
+concluding remarks in \ref{sec:conclusion}.
 
@@
 \section{Background}
-\label{section:background}
+\label{sec:background}
 
@@
   \begin{equation*}
-    Nil : \forall (a : Type). List a
-    Cons : \forall (a : Type). \forall (x : a). \forall (xs : List a). List a
-
-    Z : Nat
-    S : \forall (p : Nat). Nat
-
-    head(Cons(x, xs)) = x
-    tail(Cons(x, xs)) = xs
-    append(xs, Nil) = xs
-    append(Cons(x, xs), ys) = Cons x (append xs ys)
-    reverse(Nil) = Nil
-    reverse(Cons(x, xs)) = append(reverse(xs), Cons(x, Nil))
-    length(Nil) = Z
-    length(Cons(x, xs)) = S (length(xs))
-    map(f, Nil) = Nil
-    map(f, Cons(x, xs)) = Cons(f(x), map(f, xs))
-    foldl(f, x, Nil) = x
-    foldl(f, x, Cons(y, ys)) = foldl(f, f(x, y), xs)
-    foldr(f, Nil, y) = y
-    foldr(f, Cons(x, xs), y) = f(x, foldr(f, xs, y))
+    \begin{split}
+      \text{Nil} : \forall (a : Type). \text{List} a \\
+      \text{Cons} : \forall (a : Type). \forall (x : a). \forall (xs : \text{List} a). \text{List} a \\
+      \text{head}(\text{Cons}(x, xs)) &= x \\
+      \text{tail}(\text{Cons}(x, xs)) &= xs \\
+      \text{append}(xs, \text{Nil}) &= xs \\
+      \text{append}(\text{Cons}(x, xs), ys) &= \text{Cons}(x, \text{append}(xs, ys)) \\
+      \text{reverse}(\text{Nil}) &= \text{Nil} \\
+      \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
+      \text{length}(\text{Nil}) &= Z \\
+      \text{length}(\text{Cons}(x, xs)) &= S (\text{length}(xs)) \\
+      \text{map}(f, \text{Nil}) &= \text{Nil} \\
+      \text{map}(f, \text{Cons}(x, xs)) &= \text{Cons}(f(x), \text{map}(f, xs)) \\
+      \text{foldl}(f, x, \text{Nil}) &= x \\
+      \text{foldl}(f, x, \text{Cons}(y, ys)) &= \text{foldl}(f, f(x, y), xs) \\
+      \text{foldr}(f, \text{Nil}, y) &= y \\
+      \text{foldr}(f, \text{Cons}(x, xs), y) &= f(x, \text{foldr}(f, xs, y))
+    \end{split}
   \end{equation*}
-  \caption{A simple theory defining $Nat$ and $List$ types and some associated
-    list operations, taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
+  \caption{A simple theory defining a $\text{List}$ type and some associated
+    operations. $Z$ and $S$ are from a Peano encoding of the natural numbers.
+    Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
 \end{figure}
@@
 If we have a logical theory, for example the theory of lists shown in figure
-\ref{figure:list_theory} Many automated systems and algorithms have been developed to generate
-conjectures, but it is difficult to judge or compare their performance in a
-quantitative way. We consider three distinct aspects to evaluate:
+\ref{figure:list_theory}, we may want to find theorems which describe its
+behaviour. For example, if this is a software library then we may want to verify
+that certain (un)desirable properties do (not) hold; we may also want to
+\emph{optimise} programs using this library, rewriting expressions into a form
+which requires less time, memory, network usage, etc. To avoid altering the
+program's result, it is important to prove particular rewrite rules correct. For
+example, consider the following theorem:
+
+\begin{equation} \label{eq:mapreduce}
+  \forall f. \forall xs. \forall ys. \text{map}(f, \text{append}(xs, ys)) = \text{append}(\text{map}(f, xs), \text{map}(f, ys))
+\end{equation}
+
+This justifies a rewrite rule for splitting a single $\text{map}$ call into
+multiple independent calls dealing with different sections of the list. This is
+useful for the ``map/reduce'' programming paradigm, as each call can be
+evaluated by a separate machine in parallel.
+
+Discovering such theorems from an arbitrary theory requires two steps:
 
 \begin{itemize}
-\item Each conjecture, to judge how ``interesting'' it is.
+\item \emph{Exploring} the theory, to find patterns which we pose as
+  conjectures.
+\item \emph{Proving} that those conjectures are theorems.
+\end{itemize}
+
+Lots of research has been devoted to the second step, resulting in the field of
+Automated Theorem Proving (ATP) and spinoff applications (e.g. Prolog). Less
+attention has been paid to the first step, that of Automated Theory Exploration.
+
+Those ATE systems which do exist are difficult to evaluate and compare in a
+quantitative way. There are three aspects to consider:
+
+\begin{itemize}
+\item The quality of each conjecture, to judge how ``interesting'' it is.
 \item The \emph{set} of conjectures produced, to assess things like consistency
@@
 
-There are many aspects to consider when judging the quality of
-a conjecture: how likely it is to be true, how difficult it is to prove, its
-implications, and so on.
+In particular, we must choose how to measure what is ``interesting'', and since
+there is no objective answer we may find that the evaluation methods used by
+each system provide results which cannot be directly compared. It is also
+important to consider the goals of the system's designers, since that will
+affect for which measures it performs well.
 
-One way to make these subjective, ambiguous properties more concrete is to
-choose a well-studied theory, with a set of existing theorems that are deemed to
-be useful or important in some way, and treat these as a \emph{ground truth}
-against which we can compare conjectures generated from the same theory.
+There are many aspects to consider when judging the quality of a conjecture: how
+likely it is to be true, how difficult it is to prove, its implications, and so
+on.
+
+One method which is used to make these subjective, ambiguous properties more
+concrete is to choose a well-studied theory, with a set of existing theorems that
+are deemed to be useful or important in some way, and treat these as a
+\emph{ground truth} against which we can compare conjectures generated from the
+same theory.
 
@@
 in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
-only 4 definitions and 12 theorems. Whilst such benchmarks allow
-comparison between different approaches, their narrow scope doesn't provide much
-indication of performance in different, especially \emph{novel}, domains.
+only 4 definitions and 12 theorems. Whilst such benchmarks allow objective
+comparisons between different approaches, their narrow scope doesn't provide
+much indication of performance in different, especially \emph{novel}, domains.
 
 \section{Our Proposal}
-\label{section:proposal}
+\label{sec:proposal}
 
 We propose that the extensive test suites which are already used to benchmark
-automated theorem provers can be repurposed to create benchmarks for conjecture
-generation systems. We provide a benchmark suite, automatically derived from the
-Tons of Inductive Problems (TIP) benchmark suite, as well as software to convert
-other benchmarks written in the same TIP format (an extension of the SMT-LIB
-format \cite{BarFT-SMTLIB}).
+automated \emph{theorem provers} can be repurposed as ground truths in the
+theory \emph{exploration} setting. To do this we can take an ATP problem (for
+example proving equation \ref{eq:mapreduce} from the definitions in figure
+\ref{figure:list_theory}) and split apart the definitions (types and functions)
+from the theorem statement itself.
 
-%ALGORITHM:
+The former constitute a theory, suitable for exploration by an ATE system; the
+latter act as a ground truth, telling us which theorems about that theory were
+``interesting'' enough to write down in a benchmark.
 
-%Concatenate all definitions from all files together.
-%For each definition, prefix its name with the filename it came from and update
-%all references.
-%Normalise all local variable names.
-%Remove any identical definitions, leaving the first one in place and updating
-%any references to the removed definitions to use this one.
-%Repeat, until no further removals are possible.
+We focus on the Tons of Inductive Problems (TIP) theorem proving
+benchmark~\cite{claessen2015tip}, since it has several desirable properties:
+
+\begin{itemize}
+\item Each problem provides separate definitions and a theorem statement; hence
+  it's not difficult to tease them apart for our purposes.
+\item Higher-order functions and inductively defined types are provided, which
+  corresponds to our desired domain.
+\item All together, TIP provides 219 distinct function definitions and 343
+  theorem statements, which is enough to fully exercise current ATE systems.
+\item Benchmark problems include examples from the theorem proving literature
+  as well as common program verification tasks, ensuring the resulting corpus is
+  relevant to researchers and practitioners.
+\item TIP's maintainers provide a set of tools, including translators from TIP's
+  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
+  (which are used by ATE systems).
+\end{itemize}
+
+From TIP, we automatically derive our Theory Exploration Benchmark (TEB), and
+provide tools to benchmark ATE systems and compare their results. Our
+methodology is as follows:
+
+\begin{itemize}
+\item Collect together all definitions found in TIP into one large theory,
+  removing $\alpha$-equivalent duplicates.
+\item For each type, provide extra functions which act as constructors and
+  destructors. For example, the $\text{List}$ type would have extra constructor
+  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
+  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
+  This allows us to ignore constructors and destructors, and concentrate solely
+  on functions.
+\item Collect together all theorem statements found in TIP. Update any
+  references to removed definitions, and replace references to constructors with
+  the associated constructor function from above.
+\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
+  that we choose all of the names referenced by at least one of the theorems
+  (for example, theorem \ref{eq:mapreduce} references the names
+  $\text{map}$ and $\text{append}$).
+\item Provide this sub-set as an input to the ATE system, timing how long it
+  takes to complete.
+\item Perform precision/recall analysis on the resulting conjectures, against
+  the ground truth of TIP theorems which only reference names in this sample.
+\end{itemize}
 
 \section{Application}
-\label{section:application}
+\label{sec:application}
 
@@
 theories of various sizes: from theories containing only a single definition, up
-to theories containing 20 definitions. These are shown in figure FIXME.
+to theories containing 20 definitions.
 
@@
 \section{Discussion}
-\label{section:application}
+\label{sec:application}
 
@@
 \section{Conclusion}
-\label{section:conclusion}
+\label{sec:conclusion}
 

### 2017-09-11 TEBenchmarkPaper/article.tex
Notes, and fix copypasta reference
Add table from Colton et al

@@
 \usepackage{amssymb}
+\usepackage{array}
+\usepackage{hhline}
 \usepackage{mathtools}
+\usepackage{multirow}
 % etc.
@@
 
+\subsection{Motivation}
+\label{sec:motivation}
+
+% more detail about evaluation, interestingness, etc. Reference Simon/Alan
+% interestingness, e.g. examples from textbooks, etc.
+
+% split into sub-sections, e.g. motivation, existing approaches, etc.
+
 \begin{figure}
-  \label{figure:list_theory}
   \begin{equation*}
     \begin{split}
-      \text{Nil} : \forall (a : Type). \text{List} a \\
-      \text{Cons} : \forall (a : Type). \forall (x : a). \forall (xs : \text{List} a). \text{List} a \\
+      \forall a. \text{Nil} &: \text{List} \  a \\
+      \forall a. \text{Cons} &: a \rightarrow \text{List} \  a \rightarrow \text{List} \ a \\
       \text{head}(\text{Cons}(x, xs)) &= x \\
@@
       \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
-      \text{length}(\text{Nil}) &= Z \\
-      \text{length}(\text{Cons}(x, xs)) &= S (\text{length}(xs)) \\
+      \text{length}(\text{Nil}) &= \text{Z} \\
+      \text{length}(\text{Cons}(x, xs)) &= \text{S} (\text{length}(xs)) \\
       \text{map}(f, \text{Nil}) &= \text{Nil} \\
@@
   \caption{A simple theory defining a $\text{List}$ type and some associated
-    operations. $Z$ and $S$ are from a Peano encoding of the natural numbers.
+    operations. $\text{Z}$ and $\text{S}$ are from a Peano encoding of the natural numbers.
     Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
+  \label{figure:list_theory}
 \end{figure}
@@
 \ref{figure:list_theory}, we may want to find theorems which describe its
-behaviour. For example, if this is a software library then we may want to verify
-that certain (un)desirable properties do (not) hold; we may also want to
-\emph{optimise} programs using this library, rewriting expressions into a form
-which requires less time, memory, network usage, etc. To avoid altering the
-program's result, it is important to prove particular rewrite rules correct. For
-example, consider the following theorem:
+behaviour. For a software library we may want to verify that certain
+(un)desirable properties do (not) hold; we may also want to \emph{optimise}
+programs using this library, rewriting expressions into a form which requires
+less time, memory, network usage, etc. To avoid altering the program's result,
+it is important to prove particular rewrite rules correct. Consider the
+following theorem:
 
@@
 
-Discovering such theorems from an arbitrary theory requires two steps:
+\subsection{Theory Exploration}
+\label{sec:te}
+
+Discovering theorems from an arbitrary theory requires two steps:
 
 \begin{itemize}
-\item \emph{Exploring} the theory, to find patterns which we pose as
+\item \emph{Exploring} the theory, to find patterns suitable for posing as
   conjectures.
@@
 Lots of research has been devoted to the second step, resulting in the field of
-Automated Theorem Proving (ATP) and spinoff applications (e.g. Prolog). Less
-attention has been paid to the first step, that of Automated Theory Exploration.
+Automated Theorem Proving (ATP) and its spinoff applications. Less attention has
+been paid to the first step, that of Automated Theory Exploration (ATE).
 
-Those ATE systems which do exist are difficult to evaluate and compare in a
-quantitative way. There are three aspects to consider:
+A major challenge in ATE is the requirement for conjectures to be
+``interesting'', since this is an imprecise term with many different
+interpretations, which is reflected in the variety of disparate approaches taken
+so far.
+
+A survey of these systems and their notions of ``interestingness'' is given by
+Colton et al~\cite{colton2000notion}, for both conjectures and \emph{concepts}
+(the contents of a theory, which for ATE we consider to be provided as input,
+rather than part of the generating process). We summarise those aspects relevant
+to conjecture formation in table \ref{table:colton}, along with later systems
+we have studied.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{ |l|l|c|c|c|c|c|c| }
+      \hline
+      \multirow{2}{*}{\textbf{Program}}                 &
+      \multirow{2}{*}{\textbf{Conjecture Types}}        &
+      \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\
+      \hhline{~~------}
+         & & \textbf{E} & \textbf{N} & \textbf{S} & \textbf{A} & \textbf{C} & \textbf{U} \\
+
+      \hline
+      AM          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
+      GT          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
+      Graffiti    & inequalities                        & X & X & X &   & X & X \\ \hline
+      Bagai et al & non-exists                          &   & X &   & X & X &   \\ \hline
+      HR          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
+      QuickSpec   & equalities                          & X & X &   &   & X &   \\ \hline
+      IsaCoSy     & equalities                          &   & X &   &   & X &   \\ \hline
+      IsaScheme   & equalities                          &   &   &   &   & X &   \\ \hline
+    \end{tabular}
+  \end{center}
+  \caption{Classification of ATE systems from \cite{colton2000notion}, extended
+    to those compared in \cite{claessen2013automating} (QuickSpec is the
+    conjecture generation component of HipSpec). The interestingness measures
+    are \textbf{E}mpirical plausibility, \textbf{N}ovelty,
+    \textbf{S}urprisingness, \textbf{A}pplicability, \textbf{C}omprehensibility
+    (low complexity) and \textbf{U}tility.}
+  \label{table:colton}
+\end{table}
+
+One aspect mentioned in the survey is that ``interestingness'' is not just a
+measure of output quality, but can also impact the core design of a system. For
+example, the use of constraint solving in IsaCoSy is to prevent forming
+conjectures which are simple substitutions of previous ones; this is because the
+designers consider such conjectures to be uninteresting. This further
+complicates comparison of systems based on standalone evaluations, since
+assumptions about such behaviour may be implicit in the interpretation (e.g. an
+analysis of IsaCoSy conjectures does not need to consider substitutions), and
+each methodology will be using different assumptions.
+
+\subsection{Existing Evaluations}
+\label{sec:existing}
+
+The variety of existing ATE systems, their different domains, and their varying
+interpretations of what is interesting makes it difficult to directly compare
+their performance.
+
+Whilst all offer insights into the ATE problem, we will limit ourselves to those
+operating in higher-order logic and inductive theories, since this domain is the
+most relevant for our interest in analysing software. The relevant systems in
+this domain are QuickSpec, operating on functions in the Haskell programming
+language; and IsaCoSy and IsaScheme, which operate in the Isabelle proof
+assistant.
+
+There are three aspects to consider when we evaluate an ATE system:
 
 \begin{itemize}
-\item The quality of each conjecture, to judge how ``interesting'' it is.
+\item The quality of each conjecture, which is the ``interestingness'' discussed
+  above.
 \item The \emph{set} of conjectures produced, to assess things like consistency
@@
 
-In particular, we must choose how to measure what is ``interesting'', and since
-there is no objective answer we may find that the evaluation methods used by
-each system provide results which cannot be directly compared. It is also
-important to consider the goals of the system's designers, since that will
-affect for which measures it performs well.
-
-There are many aspects to consider when judging the quality of a conjecture: how
-likely it is to be true, how difficult it is to prove, its implications, and so
-on.
-
-One method which is used to make these subjective, ambiguous properties more
-concrete is to choose a well-studied theory, with a set of existing theorems that
-are deemed to be useful or important in some way, and treat these as a
-\emph{ground truth} against which we can compare conjectures generated from the
-same theory.
+Existing evaluation of these three systems have assessed the interestingness of
+their output by operating in well-studied theories, and taking a set of existing
+theorems as a \emph{ground truth} to compare against.
 
@@
 
-One problem with this approach is the small size of these libraries. For
-example, the benchmark based on Isabelle/HOL's theory of natural numbers given
-in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
-only 4 definitions and 12 theorems. Whilst such benchmarks allow objective
-comparisons between different approaches, their narrow scope doesn't provide
-much indication of performance in different, especially \emph{novel}, domains.
+One problem with this choice is the small size of these libraries. For example,
+the benchmark based on Isabelle/HOL's theory of natural numbers given
+in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains only 4
+definitions and a ground truth of 12 theorems. Whilst such benchmarks allow
+objective comparisons between different approaches, their narrow scope doesn't
+provide much indication of performance in different, especially \emph{novel}, domains.
 
-\section{Our Proposal}
+\section{Theory Exploration Benchmark}
 \label{sec:proposal}
 
-We propose that the extensive test suites which are already used to benchmark
-automated \emph{theorem provers} can be repurposed as ground truths in the
-theory \emph{exploration} setting. To do this we can take an ATP problem (for
-example proving equation \ref{eq:mapreduce} from the definitions in figure
-\ref{figure:list_theory}) and split apart the definitions (types and functions)
-from the theorem statement itself.
+We propose a larger, more ambitious method for evaluating ATE systems, which we
+call the Theory Exploration Benchmark (TEB). We take a similar approach to prior
+evaluations, but our ground truth comes from the extensive test suites which are
+already used to benchmark automated \emph{theorem provers}; which we repurpose
+for use in the theory \emph{exploration} setting.
 
-The former constitute a theory, suitable for exploration by an ATE system; the
-latter act as a ground truth, telling us which theorems about that theory were
-``interesting'' enough to write down in a benchmark.
+To do this we can take a set of ATP tasks, for example proving the equation
+\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
+split apart the definitions (to form a theory) from the theorem statements (to
+form our ground truth).
 
@@
 
+% Better presentation, flowchart?
 \begin{itemize}
@@
 \section{Discussion}
-\label{sec:application}
+\label{sec:discussion}
 
-% (e.g. philosophical bits),
+% philosophical bits
 

### 2017-09-13 TEBenchmarkPaper/article.tex
More TEBenchmarks fleshing out

@@
 
-Existing attempts at tackling this problem have been found difficult to compare
-and study, due partially to the variety of approaches taken, but also because of
-the inherent ambiguity of the task and the different goals emphasised by their
-authors and evaluation methods.
+Existing attempts at tackling this problem are difficult to compare, due
+partially to the variety of approaches taken, but also because of the inherent
+ambiguity of the task and the different goals emphasised by their designers and
+their choice of evaluation method.
 
@@
 \item A general methodology for benchmarking ATE systems.
-\item Resolving the issue of ``interestingness'' through the use theorem-proving
-  benchmarks as a ground-truth.
+\item Resolving the issue of ``interestingness'' through the use of
+  theorem-proving benchmarks as a ground-truth.
 \item A specific instantiation of this methodology, using the Tons of Inductive
   Problems benchmark as a corpus.
-\item Automated tooling to produce this benchmark and analyse its output.
-\item Application of this methodology to the QuickSpec and IsaCoSy ATE systems,
+\item Automated tooling to perform this benchmarking.
+\item Application of our methodology to the QuickSpec and IsaCoSy ATE systems,
   and a discussion of the results.
@@
 
-If we have a logical theory, for example the theory of lists shown in figure
+Given a logical theory, like the theory of lists shown in figure
 \ref{figure:list_theory}, we may want to find theorems which describe its
-behaviour. For a software library we may want to verify that certain
-(un)desirable properties do (not) hold; we may also want to \emph{optimise}
+behaviour. This could be for mathematical curiosity, or due to the theory's
+importance in some domain. In particular, for theories which capture the
+semantics of some software library, we may want to verify that certain
+(un)desirable properties do (not) hold; we might also want to \emph{optimise}
 programs using this library, rewriting expressions into a form which requires
-less time, memory, network usage, etc. To avoid altering the program's result,
-it is important to prove particular rewrite rules correct. Consider the
-following theorem:
+less time, memory, network usage, etc. To avoid altering a program's result,
+such rewrites should come with theorems proving their correctness, such as the
+following theorem for our theory of lists:
 
@@
 This justifies a rewrite rule for splitting a single $\text{map}$ call into
-multiple independent calls dealing with different sections of the list. This is
-useful for the ``map/reduce'' programming paradigm, as each call can be
-evaluated by a separate machine in parallel.
+multiple independent calls dealing with different sections of a list. Such rules
+are useful optimisations since each call can be evaluated in parallel, leading
+to the ``map/reduce'' programming paradigm.
+
+All of these example use cases require the ability to discover theorems about
+some particular theory. This is a hard problem in general, but presents
+opportunities for automation due to the precise, symbolic nature of the domain.
 
@@
 
-Discovering theorems from an arbitrary theory requires two steps:
+The task of discovering theorems in an arbitrary theory can be broken down into
+two steps:
 
@@
 
-Lots of research has been devoted to the second step, resulting in the field of
-Automated Theorem Proving (ATP) and its spinoff applications. Less attention has
-been paid to the first step, that of Automated Theory Exploration (ATE).
+This second step is studied extensively in the field of Automated Theorem
+Proving (ATP). Less attention has been paid to the first step, that of
+Automated Theory Exploration (ATE).
 
-A major challenge in ATE is the requirement for conjectures to be
-``interesting'', since this is an imprecise term with many different
-interpretations, which is reflected in the variety of disparate approaches taken
-so far.
+A major challenge in ATE is choosing how to narrow down the set of generated
+conjectures to those deemed ``interesting'', since this is an imprecise term
+with many different interpretations. For example, all existing approaches agree
+that simple tautologies are ``uninteresting'', but differ when it comes to more
+complex statements.
 
-A survey of these systems and their notions of ``interestingness'' is given by
-Colton et al~\cite{colton2000notion}, for both conjectures and \emph{concepts}
-(the contents of a theory, which for ATE we consider to be provided as input,
-rather than part of the generating process). We summarise those aspects relevant
-to conjecture formation in table \ref{table:colton}, along with later systems
-we have studied.
+A survey of ATE systems and their notions of ``interestingness'' is given by
+Colton et al~\cite{colton2000notion}. Their survey also includes \emph{concept
+  formation}, which corresponds to the automatic creation of theories, although
+We ignore this as a separate concern: the ATE task assumes that a theory is
+provided. We summarise those aspects relevant to conjecture formation in table
+\ref{table:colton}, and extend the comparison with three more recent systems.
 
@@
 
-One aspect mentioned in the survey is that ``interestingness'' is not just a
-measure of output quality, but can also impact the core design of a system. For
-example, the use of constraint solving in IsaCoSy is to prevent forming
-conjectures which are simple substitutions of previous ones; this is because the
-designers consider such conjectures to be uninteresting. This further
-complicates comparison of systems based on standalone evaluations, since
-assumptions about such behaviour may be implicit in the interpretation (e.g. an
-analysis of IsaCoSy conjectures does not need to consider substitutions), and
-each methodology will be using different assumptions.
-
 \subsection{Existing Evaluations}
@@
 
-The variety of existing ATE systems, their different domains, and their varying
-interpretations of what is interesting makes it difficult to directly compare
-their performance.
+The criteria for ``interestingness'' in table \ref{table:colton} not only allow
+the results of ATE systems to be evaluated, but they also affect the choice of
+algorithm used by those systems in the first place. This coupling between
+implementation detail and assessment method makes it difficult to compare
+existing approaches, since the assumptions made by each evaluation methodology
+may lead to incompatible results. For example, the design of IsaCoSy treats
+simple substitutions of existing conjectures as uninteresting; yet this
+criterion is absent from its evaluation, since the algorithm uses a constraint
+solver to entirely avoid such results.
 
-Whilst all offer insights into the ATE problem, we will limit ourselves to those
-operating in higher-order logic and inductive theories, since this domain is the
-most relevant for our interest in analysing software. The relevant systems in
-this domain are QuickSpec, operating on functions in the Haskell programming
-language; and IsaCoSy and IsaScheme, which operate in the Isabelle proof
-assistant.
+If we limit ourselves to exploring higher-order theories with inductive types
+(a domain closely matching our interest in analysing software), we do find a
+direct comparison of three systems~\cite{claessen2013automating}: QuickSpec (the
+conjecture-generating component of HipSpec), IsaCoSy and IsaScheme. This is a
+\emph{precision/recall} analysis, providing the same theories to each system and
+comparing the set of generated conjectures against a \emph{ground truth}.
 

### 2017-09-13 TransferReport/report.tex
Make transfer report render

@@
 
+\usepackage{rotating}
 \usepackage{hyperref}
@@
 \newcommand{\argmin}{\operatornamewithlimits{argmin}}
-\newcommand{\feature}[1]{\left\ulcorner #1 \right\urcorner}
+\newcommand{\feature}[1]{\phi(#1)}
 \newcommand{\id}[1]{\texttt{"#1"}}
@@
 It extends previously existing methods available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming.
-Current theory exploration techniques are limited by their use of exponential time algorithms, which we mitigate by intelligently reducing the size of large inputs, using machine learning to preserve the relationships needed for successful exploration.
+Current theory exploration techniques are limited by their use of exponential time algorithms, which although thorough are ultimately limited to finding simple properties of small systems.
+We propose a more powerful approach, which uses machine learning algorithms to intelligently choose which parts of a system to explore based on their similarity, hence focusing its efforts on areas which are most likely to lead to discoveries.
 
-\iffalse TODO: I would emphasise more selective (intelligent) approach, and as a result - better precision of TE. Rather than saying that it is merely a size/complexity problem. \fi
 \end{abstract}
@@
 \bibliographystyle{plain}
-\bibliography{../Bibtex}
+\bibliography{bibtex}
 

### 2017-09-22 TEBenchmarkPaper/article.tex
TEBenchmark additions and cleanup

@@
 % \newcommand{}{}
-%
+
+% For example theory of lists
+\newcommand{\function}{\rightarrow}
+\newcommand{\Zero}{\text{Z}}
+\newcommand{\Succ}{\text{S}}
+\newcommand{\List}{\text{List}}
+\newcommand{\ListA}{\text{List} \  a}
+\newcommand{\Nil}{\text{Nil}}
+\newcommand{\Cons}{\text{Cons}}
+\newcommand{\Head}{\text{head}}
+\newcommand{\Tail}{\text{tail}}
+\newcommand{\Append}{\text{append}}
+\newcommand{\Reverse}{\text{reverse}}
+\newcommand{\Length}{\text{length}}
+\newcommand{\Map}{\text{map}}
+\newcommand{\Foldl}{\text{foldl}}
+\newcommand{\Foldr}{\text{foldr}}
+
+% For interestingness table
+\newcommand{\iE}{\textbf{E}}
+\newcommand{\iN}{\textbf{N}}
+\newcommand{\iS}{\textbf{S}}
+\newcommand{\iA}{\textbf{A}}
+\newcommand{\iC}{\textbf{C}}
+\newcommand{\iU}{\textbf{U}}
+\newcommand{\tIFF}{if-and-only-if}
+\newcommand{\tNE}{non-exists}
+\newcommand{\tIMP}{implies}
+\newcommand{\tRow}[1]{#1 \\ \hline}
+
 % Insert the name of "your journal" with
@@
 
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks{Grants or other notes
-%about the article that should go on the front page should be
-%placed here. General acknowledgments should be placed at the end of the article.}
+\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
+% {Grants or other notes about the article that should go on the front page
+% should be placed here. General acknowledgments should be placed at the end of
+% the article.}
 }
@@
 \maketitle
-%IGNORE ISABELLE THEORIES FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-%IGNORE CLUSTERING FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-%COMPARE COLTON ET AL.
-
-% Evaluation of conjectures, sets of conjectures, systems
-% what constitutes a good/desired output (in each author's terms)? This combines
-% their evaluation with their (system-specific) assumptions (e.g. IsaCoSy doesn't
-% generate redundant terms: non-redundancy is clearly a goal, but it's not measured
-% in eval since it's assumed based on system's construction)
 
 % Look carefully at each system's (paper's) eval section
-% Make a table like tabl 1 of Colton et al
 
@@
 proposal for a more general benchmark in section \ref{sec:proposal} and
-section \ref{sec:application} shows the results when applied to existing ATE tools.
-Analysis of these results is given in section \ref{sec:discussion} and
+section \ref{sec:application} shows the results when applied to existing ATE
+tools. Analysis of these results is given in section \ref{sec:discussion} and
 concluding remarks in \ref{sec:conclusion}.
@@
 
-% split into sub-sections, e.g. motivation, existing approaches, etc.
-
 \begin{figure}
@@
     \begin{split}
-      \forall a. \text{Nil} &: \text{List} \  a \\
-      \forall a. \text{Cons} &: a \rightarrow \text{List} \  a \rightarrow \text{List} \ a \\
-      \text{head}(\text{Cons}(x, xs)) &= x \\
-      \text{tail}(\text{Cons}(x, xs)) &= xs \\
-      \text{append}(xs, \text{Nil}) &= xs \\
-      \text{append}(\text{Cons}(x, xs), ys) &= \text{Cons}(x, \text{append}(xs, ys)) \\
-      \text{reverse}(\text{Nil}) &= \text{Nil} \\
-      \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
-      \text{length}(\text{Nil}) &= \text{Z} \\
-      \text{length}(\text{Cons}(x, xs)) &= \text{S} (\text{length}(xs)) \\
-      \text{map}(f, \text{Nil}) &= \text{Nil} \\
-      \text{map}(f, \text{Cons}(x, xs)) &= \text{Cons}(f(x), \text{map}(f, xs)) \\
-      \text{foldl}(f, x, \text{Nil}) &= x \\
-      \text{foldl}(f, x, \text{Cons}(y, ys)) &= \text{foldl}(f, f(x, y), xs) \\
-      \text{foldr}(f, \text{Nil}, y) &= y \\
-      \text{foldr}(f, \text{Cons}(x, xs), y) &= f(x, \text{foldr}(f, xs, y))
+      \forall a. \Nil            &: \ListA                                  \\
+      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
+      \Head(\Cons(x, xs))        &= x                                       \\
+      \Tail(\Cons(x, xs))        &= xs                                      \\
+      \Append(\Nil,         ys)  &= ys                                      \\
+      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
+      \Reverse(\Nil)             &= \Nil                                    \\
+      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
+      \Length(\Nil)              &= \Zero                                   \\
+      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
+      \Map(f, \Nil)              &= \Nil                                    \\
+      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
+      \Foldl(f, x, \Nil)         &= x                                       \\
+      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), xs)                  \\
+      \Foldr(f, \Nil,         y) &= y                                       \\
+      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
     \end{split}
   \end{equation*}
-  \caption{A simple theory defining a $\text{List}$ type and some associated
-    operations. $\text{Z}$ and $\text{S}$ are from a Peano encoding of the natural numbers.
-    Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
+  \caption{A simple theory defining a $\List$ type and some associated
+    operations. $\Zero$ and $\Succ$ are from a Peano encoding of the
+    natural numbers. Taken from
+    \cite{Johansson.Dixon.Bundy:conjecture-generation}}
   \label{figure:list_theory}
@@
 \begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys. \text{map}(f, \text{append}(xs, ys)) = \text{append}(\text{map}(f, xs), \text{map}(f, ys))
+  \forall f. \forall xs. \forall ys.
+    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
 \end{equation}
 
-This justifies a rewrite rule for splitting a single $\text{map}$ call into
+This justifies a rewrite rule for splitting a single $\Map$ call into
 multiple independent calls dealing with different sections of a list. Such rules
@@
 
-The task of discovering theorems in an arbitrary theory can be broken down into
-two steps:
+The task of discovering theorems in an arbitrary theory can be described as the
+interplay of several processes:
+
+% Is it clear?
 
 \begin{itemize}
+\item \emph{Forming} or refining the theory, to add new definitions and
+  concepts.
 \item \emph{Exploring} the theory, to find patterns suitable for posing as
@@
 
-This second step is studied extensively in the field of Automated Theorem
-Proving (ATP). Less attention has been paid to the first step, that of
-Automated Theory Exploration (ATE).
+The latter is studied extensively in the field of Automated Theorem Proving
+(ATP). Less attention has been paid to automating the first two, those of
+\emph{automated theory formation} and \emph{automated theory exploration} (ATE).
 
-A major challenge in ATE is choosing how to narrow down the set of generated
-conjectures to those deemed ``interesting'', since this is an imprecise term
-with many different interpretations. For example, all existing approaches agree
-that simple tautologies are ``uninteresting'', but differ when it comes to more
-complex statements.
+We focus on ATE, where a major challenge is choosing how to narrow down the set
+of generated conjectures to those deemed ``interesting'', since this is an
+imprecise term with many different interpretations. For example, all existing
+approaches agree that simple tautologies are ``uninteresting'', but differ when
+it comes to more complex statements.
 
-A survey of ATE systems and their notions of ``interestingness'' is given by
-Colton et al~\cite{colton2000notion}. Their survey also includes \emph{concept
-  formation}, which corresponds to the automatic creation of theories, although
-We ignore this as a separate concern: the ATE task assumes that a theory is
-provided. We summarise those aspects relevant to conjecture formation in table
-\ref{table:colton}, and extend the comparison with three more recent systems.
+A survey of systems for theory formation and exploration, and their notions of
+``interestingness'' (for concepts and conjectures), is given by Colton et
+al~\cite{colton2000notion}. Six notions of ``interestingness'' are identified in
+these systems, their use by each is summarised in table \ref{table:colton},
+along with (our interpretation of) their use in three more recent ATE systems.
+These notions, as applied to ATE, are:
+
+\textbf{Empirical plausibility}, which checks whether a property holds across
+some specific examples. This is especially useful for avoiding  false
+conjectures, without resorting to a full proof search.
+
+\textbf{Novelty} is whether a conjecture, or one isomorphic or more general, has
+already been seen.
+
+\textbf{Surprisingness} of a conjecture is whether or not it is ``obvious'', for
+example if it is an instance of a tautology.
+
+\textbf{Applicability} depends on the number of models in which a conjecture
+holds. The system of Bagai et al conjectures the \emph{non-existence} of
+objects, and hence favours statements with \emph{zero} applicability. Other
+systems treat applicability as a positive aspect: the more applicable the
+statement, the more interesting it is.
+
+\textbf{Comprehensibility} is the \emph{complexity} of a statement. Simpler
+statements are considered more interesting, and many of the search algorithms
+explore simpler/smaller statements before complex/larger ones, to more
+efficiently find those which are interesting.
+
+\textbf{Utility} is the relevance or usefulness of a conjecture to the user's
+particular task. For example, if we want to find optimising rewrite rules such
+as equation \ref{eq:mapreduce}, then utility would include whether or not a
+conjecture justifies a rewrite rule, the difference in resource usage of the
+expressions involved, and how common those expressions are in real usage.
 
 \begin{table}
-  \begin{center}
+  \centering
   \begin{tabular}{ |l|l|c|c|c|c|c|c| }
@@
     \multirow{2}{*}{\textbf{Conjecture Types}}            &
-      \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\
-      \hhline{~~------}
-         & & \textbf{E} & \textbf{N} & \textbf{S} & \textbf{A} & \textbf{C} & \textbf{U} \\
-
-      \hline
-      AM          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
-      GT          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
-      Graffiti    & inequalities                        & X & X & X &   & X & X \\ \hline
-      Bagai et al & non-exists                          &   & X &   & X & X &   \\ \hline
-      HR          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
-      QuickSpec   & equalities                          & X & X &   &   & X &   \\ \hline
-      IsaCoSy     & equalities                          &   & X &   &   & X &   \\ \hline
-      IsaScheme   & equalities                          &   &   &   &   & X &   \\ \hline
+    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
+    \tRow{            &                    & \iE & \iN & \iS & \iA & \iC & \iU}
+    \tRow{AM          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
+    \tRow{GT          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
+    \tRow{Graffiti    & inequalities       &   X &   X &   X &     &   X &   X}
+    \tRow{Bagai et al & \tNE               &     &   X &     &   X &   X &    }
+    \tRow{HR          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
+    \tRow{QuickSpec   & equalities         &   X &   X &     &     &   X &    }
+    \tRow{IsaCoSy     & equalities         &     &   X &     &     &   X &    }
+    \tRow{IsaScheme   & equalities         &     &     &     &     &   X &    }
   \end{tabular}
-  \end{center}
   \caption{Classification of ATE systems from \cite{colton2000notion}, extended
@@
     conjecture generation component of HipSpec). The interestingness measures
-    are \textbf{E}mpirical plausibility, \textbf{N}ovelty,
-    \textbf{S}urprisingness, \textbf{A}pplicability, \textbf{C}omprehensibility
-    (low complexity) and \textbf{U}tility.}
+    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
+    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
   \label{table:colton}
 \end{table}
+
+These criteria, especially empirical plausibility and comprehensibility, are not
+only used for analysing results after the fact, but instead may form a core part
+of an algorithm's design decisions. For example, the designers of IsaCoSy
+consider simple substitutions of existing conjectures (i.e. those with low
+novelty) to be uninteresting, and hence their system includes a constraint
+solving component which avoids generating such statements entirely.
+
+With such ambiguous and varied goals, approaches and assessment criteria it is
+difficult to compare ATE systems in a quantitative way, and hence to form some
+measure of ``progress'' for the field.
 
@@
 
-The criteria for ``interestingness'' in table \ref{table:colton} not only allow
-the results of ATE systems to be evaluated, but they also affect the choice of
-algorithm used by those systems in the first place. This coupling between
-implementation detail and assessment method makes it difficult to compare
-existing approaches, since the assumptions made by each evaluation methodology
-may lead to incompatible results. For example, the design of IsaCoSy treats
-simple substitutions of existing conjectures as uninteresting; yet this
-criterion is absent from its evaluation, since the algorithm uses a constraint
-solver to entirely avoid such results.
+There are three aspects to consider when we evaluate an ATE system:
+
+\begin{itemize}
+\item The quality of each conjecture, which is the ``interestingness'' discussed
+  above.
+\item The \emph{set} of conjectures produced, to assess things like novelty, as
+  well as consistency of the results.
+\item Performance of the system, to find the balance struck between output
+  quality and time taken.
+\end{itemize}
+
+% TODO: Should "generality" be one? Or is that just an implicit "don't cheat"
+% assumption?
 
@@
 
-There are three aspects to consider when we evaluate an ATE system:
-
-\begin{itemize}
-\item The quality of each conjecture, which is the ``interestingness'' discussed
-  above.
-\item The \emph{set} of conjectures produced, to assess things like consistency
-  and redundancy.
-\item Performance of the system, to find the balance struck between output
-  quality and time taken.
-\end{itemize}
-
-Existing evaluation of these three systems have assessed the interestingness of
-their output by operating in well-studied theories, and taking a set of existing
-theorems as a \emph{ground truth} to compare against.
+% TODO: precision/recall is evaluating the set of conjectures. It assumes that
+% interestingness == membership in the ground truth, and ignores efficiency
+% completely
 
@@
 objective comparisons between different approaches, their narrow scope doesn't
-provide much indication of performance in different, especially \emph{novel}, domains.
+provide much indication of performance in different, especially \emph{novel},
+domains.
 
@@
 \item For each type, provide extra functions which act as constructors and
-  destructors. For example, the $\text{List}$ type would have extra constructor
+  destructors. For example, the $\List$ type would have extra constructor
   functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
@@
   (for example, theorem \ref{eq:mapreduce} references the names
-  $\text{map}$ and $\text{append}$).
+  $\Map$ and $\Append$).
 \item Provide this sub-set as an input to the ATE system, timing how long it
@@
 
-``Solving'' this benchmark suite would not solve the problem of theory exploration in general,
-so more ambitious goals must be set in the future; but we have found that there is still a
-long way to go until that problem arises.
+``Solving'' this benchmark suite would not solve the problem of theory
+exploration in general, so more ambitious goals must be set in the future; but
+we have found that there is still a long way to go until that problem arises.
 

### 2017-10-02 TEBenchmarkPaper/article.tex
Lots of fleshing out

@@
 \usepackage{multirow}
+\usepackage{tikz}
+\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
+  shapes.geometric}
 % etc.
@@
 
-\begin{itemize}
+\begin{enumerate}
 \item The quality of each conjecture, which is the ``interestingness'' discussed
   above.
-\item The \emph{set} of conjectures produced, to assess things like novelty, as
-  well as consistency of the results.
+\item The quality of the \emph{set} of conjectures produced, used for relative
+  interestingness criteria like novelty, as well as to assess consistency.
 \item Performance of the system, to find the balance struck between output
   quality and time taken.
-\end{itemize}
+\end{enumerate}
 
@@
 (a domain closely matching our interest in analysing software), we do find a
-direct comparison of three systems~\cite{claessen2013automating}: QuickSpec (the
-conjecture-generating component of HipSpec), IsaCoSy and IsaScheme. This is a
-\emph{precision/recall} analysis, providing the same theories to each system and
-comparing the set of generated conjectures against a \emph{ground truth}.
+direct comparison of three systems~\cite{claessen2013automating}: HipSpec,
+IsaCoSy and IsaScheme. QuickSpec, shown in table \ref{table:colton}, is the
+conjecture-generating component of HipSpec (which sends those conjectures into
+automated theorem provers).
 
-% TODO: precision/recall is evaluating the set of conjectures. It assumes that
-% interestingness == membership in the ground truth, and ignores efficiency
-% completely
+This comparison uses \emph{precision/recall analysis}: a particular set of
+definitions is chosen, such as those in figure \ref{figure:list_theory}, and a
+\emph{ground truth} is chosen as the 'ideal' set of conjectures which we would
+like an ATE system to produce from these definitions.
 
-A common choice of ground truth is the standard library of a theorem prover,
-such as Isabelle/HOL; the library authors have gone to the effort of stating,
+Each system is run, and their output compared to the ground truth. To score
+100\% on precision and recall, an ATE system must output all of the conjectures
+which appear in the ground truth, and nothing else:
+
+\begin{itemize}
+\item \emph{Precision} is the proportion of a system's output which appears in
+  the ground truth. This penalises systems which output large numbers of
+  conjectures in the hope that some turn out to be ``good''.
+\item \emph{Recall} is the proportion of the ground truth which appears in the
+  system's output. This penalises systems which overly restrict their output to
+  avoid generating ``bad'' conjectures.
+\end{itemize}
+
+Precision and recall fulfil our second requirement, by directly measuring the
+quality of the \emph{set} of conjectures. Their ``interestingness'' criterion is
+indirect: asking only whether or not a conjecture appears in the ground truth.
+A satisfactory solution to our first requirement hence requires careful choice
+of the ground truth set.
+
+This particular analysis takes its ground truths from the standard library of
+the Isabelle theorem prover, with one involving the natural numbers, and the
+other involving the list functions from figure \ref{figure:list_theory}. The library authors have gone to the effort of stating,
 proving and including these theorems in every copy of their software, which is a
@@
 
-We propose a larger, more ambitious method for evaluating ATE systems, which we
-call the Theory Exploration Benchmark (TEB). We take a similar approach to prior
-evaluations, but our ground truth comes from the extensive test suites which are
-already used to benchmark automated \emph{theorem provers}; which we repurpose
-for use in the theory \emph{exploration} setting.
+Our main contribution is a benchmarking methodology, shown in figure
+\ref{figure:flow_chart}, which generates both a large definition/ground-truth
+corpus (an order of magnitude larger than previous work) and a scalable,
+statistical approach to evaluating ATE systems using this corpus. We follow a
+precision/recall approach similar to the work described above, with the main
+difference being the source of definitions and ground truths: we take existing
+problem sets designed for automated \emph{theorem proving}, and adapt their
+content for use in the \emph{theory exploration} setting.
 
-To do this we can take a set of ATP tasks, for example proving the equation
-\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
-split apart the definitions (to form a theory) from the theorem statements (to
-form our ground truth).
+\subsection{Preparation}
 
-We focus on the Tons of Inductive Problems (TIP) theorem proving
-benchmark~\cite{claessen2015tip}, since it has several desirable properties:
+Automated theorem proving is an active area of research, with large problem sets
+and regular competitions to prove as much as possible, as fast as possible.
+These problem sets are an opportunity for theory exploration, as their
+definitions and theorems can be used as a corpus, in the same way that Isabelle
+libraries have been used in the past.
+
+Some problem sets are more amenable for this purpose than others. The most
+suitable are those which meet the following criteria:
 
 \begin{itemize}
-\item Each problem provides separate definitions and a theorem statement; hence
-  it's not difficult to tease them apart for our purposes.
-\item Higher-order functions and inductively defined types are provided, which
-  corresponds to our desired domain.
-\item All together, TIP provides 219 distinct function definitions and 343
-  theorem statements, which is enough to fully exercise current ATE systems.
-\item Benchmark problems include examples from the theorem proving literature
-  as well as common program verification tasks, ensuring the resulting corpus is
-  relevant to researchers and practitioners.
-\item TIP's maintainers provide a set of tools, including translators from TIP's
-  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
-  (which are used by ATE systems).
+\item For each problem, there should be a clear separation between the theorem
+  to be proved and the definitions involved, such that the two can be easily and
+  meaningfully separated. This rules out problem sets like SMTLIB, where many
+  problems \emph{declare} a function, but leave its behaviour \emph{implicit} in
+  the logical structure of the theorem statement.
+\item Definitions should be \emph{strongly typed}, such that they can be
+  translated into the native language of each ATE system (in this work, those
+  languages are Haskell and Isabelle).
+\item The problem set should be relevant to the desired domain. In our case, we
+  desire higher-order functions and inductively defined types, which rules out
+  first-order languages/logics.
+\item Since it will act as our ground truth, the problem set should ideally
+  contain \emph{every} ``interesting'' conjecture involving its included
+  definitions. Realistically, we should aim for each definition to appear in
+  \emph{many} theorem statements; rather than each problem having unique
+  definitions.
+\item The problem set should be as large as possible, for robustness of the
+  resulting statistics.
 \end{itemize}
 
-From TIP, we automatically derive our Theory Exploration Benchmark (TEB), and
-provide tools to benchmark ATE systems and compare their results. Our
-methodology is as follows:
+Once a theorem proving problem set has been chosen, we must separate the
+definitions referenced by the theorems from the theorem statements themselves.
+The former will be used as input to the ATE systems under evaluation, whilst the
+latter form a ground truth corpus against which to compare the ATE output.
+
+It is important to ensure that there are no duplicate definitions: theory
+exploration does not depend on the names of functions, so neither should our
+analysis. For example, consider a problem set which includes a commutativity
+theorem for a \texttt{plus} function, and an associativity theorem for an
+\texttt{add} function, where the definitions of \texttt{plus} and \texttt{add}
+are $\alpha$-equivalent. We would expect an ATE system to conjecture
+commutativity and associativity for \emph{both} functions, or for \emph{neither}
+function, since they're equivalent. Yet a na\"ive precision/recall analysis
+would treat commutativity of \texttt{add} and associativity of \texttt{plus} as
+\emph{uninteresting}, since they don't appear in the ground truth.
+
+For this reason, duplicates should be removed (e.g. leaving those which appear
+first alphabetically), and any references in the theorem statements updated to
+use the remaining definition. In the above example, the \texttt{plus} function
+would be removed, and the commutativity theorem updated to reference
+\texttt{add} instead.
+
+\subsection{Sampling}
+
+We could, in theory, send these de-duplicated definitions straight into an ATE
+system, and use the updated theorem statements as the ground truth for analysis.
+However, this would cause two problems:
+
+\begin{itemize}
+\item The result would be a single data point, which makes it difficult to
+  infer performance \emph{in general}.
+\item It is impractical to run any existing ATE system on ``large'' inputs,
+  containing more than around 10 definitions.
+\end{itemize}
+
+To solve both of these problems we instead \emph{sample} a subset of
+definitions. Given a sample size, we choose a subset of that many definitions,
+and use that as our ATE input. We generate a corresponding ground truth by
+selecting those theorems from the corpus which only contain references to
+definitions that appear in our sample.
+
+Unfortunately, uniform sampling of definitions gives rise to a lottery, where
+for a large corpus and a fixed sample size, it becomes highly unlikely that a
+sample will contain the particular definitions required by a theorem. This would
+cause most samples to have an empty ground truth, and hence $0$ precision and
+undefined recall \emph{independent} of what the ATE system produces. This is
+clearly undesirable as an evaluation method.
+
+Instead, we only choose samples which admit at least one of the corpus theorems.
+We could do this using rejection sampling, but it is more efficient to sample
+the \emph{theorem statements}, weighted in proportion to the number of
+definitions they reference. Once a theorem is chosen, we use the referenced
+definitions as our sample, padding uniformly up to the required size (we ignore
+theorems which reference more definitions than our sample size).
+
+We form the ground truth as above: selecting all applicable theorems from the
+corpus, which is now guaranteed to be non-empty.
+
+\subsection{Evaluation}
+
+Given a sample of definitions and a corresponding ground truth, the actual
+execution of the ATE system proceeds as in prior work. We must translate the
+definitions into the required input format, which depends on the particular
+tool being studied, then we time the execution, with a cutoff after e.g. 1 hour.
+
+In our experiments we have found that memory usage also plays an important role,
+but rather than introducing an extra dimension we can account for this by simply
+allowing the system
+
+\begin{figure}
+  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
+                           minimum height=1cm,text centered, draw=black]
+
+  \tikzstyle{io} = [trapezium, trapezium left angle=70,
+                    trapezium right angle=110, minimum width=1cm,
+                    minimum height=1cm, text centered, draw=black]
+
+  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
+                         text centered, draw=black]
+  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
+                          text centered, draw=black]
+
+  \tikzstyle{arrow} = [thick,->,>=stealth]
+
+  % Avoids too much padding in io shapes
+  \tikzset{trapezium stretches=true}
+
+  \centering
+  \begin{tikzpicture}[node distance=1cm]
+
+    % Preparation section
+
+    \node (in)  [io]{Theorem Proving Benchmark (TIP)};
+    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
+    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
+    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
+
+    \node (thy)  [startstop, below=1cm of def]{Full Theory};
+    \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};
+
+    \draw [arrow] (in)  -- (sep);
+    \draw [arrow] (def) -- (ref);
+    \draw [arrow] (def) -- (thy);
+    \draw [arrow] (ref) -- (thm);
+
+    % Arrows with labels
+    \path[->]
+        (sep) edge [arrow, sloped, above] node {definitions} (def)
+        (sep) edge [arrow, sloped, above] node {theorems}    (ref);
+
+    % Sampling section
+
+    \node (choose) [process, below=of thm   ]{Sample a theorem};
+    \node (deps)   [process, below=of choose]{List referenced definitions};
+
+    % Create dummy coordinate below deps, then use its y coordinate for pad
+    \coordinate [below=of deps] (padDummy);
+    \path let \p{dummy} = (padDummy),
+              \p{in}    = (in)
+              in coordinate (padPos) at (\x{in}, \y{dummy});
+    \node (pad)  [process, at=(padPos)]{Sample definitions to pad list};
+
+    \node (find) [process,   below right=1.5cm and -1cm of pad]{List all applicable theorems};
+    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
+
+    \node (sthm)  [startstop, below=of find]{Ground Truth};
+
+    % Calculate position of (size) using let, then define as normal
+    \path let \p{pad}    = (pad),
+              \p{choose} = (choose)
+           in coordinate (sizePos) at (\x{pad},\y{choose});
+    \node (size) [io, at=(sizePos)] {Sample size};
+
+    % Evaluation section
+    \path let \p{pad}  = (pad),
+              \p{sthm} = (sthm)
+           in coordinate (runPos) at (\x{pad}, \y{sthm});
+    \node (run) [process, below=of runPos]{Run ATE system};
+    \node (pr)  [process, below=of run]{Analysis};
+
+    \node (prec) [startstop, below=of pr  ]{Precision};
+    \node (time) [startstop,  left=of prec]{Time taken};
+    \node (rec)  [startstop, right=of prec]{Recall};
+
+    \draw [arrow] (thy)    |- (pad);
+    \draw [arrow] (thm)    -- (choose);
+    \draw [arrow] (choose) -- (deps);
+    \draw [arrow] (deps)   |- (pad);
+    \draw [arrow] (size)   -- (choose);
+    \draw [arrow] (size)   -- (pad);
+    \draw [arrow] (pad)    -- (find);
+    \draw [arrow] (pad)    -- (sthy);
+    \draw [arrow] (find)   -- (sthm);
+    \draw [arrow] (sthy)   |- (run);
+    \draw [arrow] (run)    -- (pr);
+    \draw [arrow] (sthm)   |- (pr);
+    \draw [arrow] (pr)     -- (prec);
+    \draw [arrow] (pr)     -- (rec);
+    \draw [arrow] (pr)     -- (time);
+
+    % Awkward arrow
+    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (find);
+
+    % Braces
+
+    % Preparation
+    \draw
+      let \p{thm} = (thm.east),
+          \p{in}  = (in.north),
+          \p{rec} = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
+         node [black, midway, right, xshift=0.3cm] {Preparation};
+
+    % Sampling
+    \draw
+      let \p{thm} = (thm.east),
+          \p{gt}  = (sthm.east),
+          \p{rec} = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
+         node [black, midway, right, xshift=0.3cm] {Sampling};
+
+    % Evaluation
+    \draw
+      let \p{sthm} = (sthm.east),
+          \p{rec}  = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
+         node [black, midway, right, xshift=0.3cm] {Evaluation};
+  \end{tikzpicture}
+  \caption[]{High-level view of our benchmarking methodology, showing
+    \begin{tikzpicture}
+      \node [rectangle, text centered, draw=black]{processes};
+    \end{tikzpicture},
+    \begin{tikzpicture}
+      \node [rectangle, rounded corners, text centered, draw=black]{data};
+    \end{tikzpicture} and
+    \begin{tikzpicture}
+      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
+             text centered, draw=black]{inputs};
+  \end{tikzpicture}}
+  \label{figure:flow_chart}
+\end{figure}
 
@@
 
+We call the Theory Exploration Benchmark (TEB).
+
+To do this we can take a set of ATP tasks, for example proving the equation
+\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
+split apart the definitions (to form a theory) from the theorem statements (to
+form our ground truth).
+
+We focus on the Tons of Inductive Problems (TIP) theorem proving
+benchmark~\cite{claessen2015tip}, since it has several desirable properties:
+
+All together, TIP provides 219 distinct function definitions and 343
+  theorem statements, which is enough to fully exercise current ATE systems.
+Benchmark problems include examples from the theorem proving literature
+  as well as common program verification tasks, ensuring the resulting corpus is
+  relevant to researchers and practitioners.
+TIP's maintainers provide a set of tools, including translators from TIP's
+  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
+  (which are used by ATE systems).
+
 % evaluation
@@
 
+We are assuming that the ATE system acts in ``batch mode'' (accepting
+definitions, producing conjectures, then halting), rather than running
+continuously or interactively. This holds for all of the systems we have
+analysed
+
 % philosophical bits
+
+% We're assuming there is general agreement on what is "good maths", when
+% actually this isn't at all clear-cut
 

### 2017-10-04 TEBenchmarkPaper/article.tex
Flesh out section 3

@@
 
-Unfortunately, uniform sampling of definitions gives rise to a lottery, where
-for a large corpus and a fixed sample size, it becomes highly unlikely that a
-sample will contain the particular definitions required by a theorem. This would
-cause most samples to have an empty ground truth, and hence $0$ precision and
+Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
+large corpus and a fixed sample size, it becomes highly unlikely that a sample
+will contain the particular definitions required by a theorem. This would cause
+most samples to have an empty ground truth, and hence $0$ precision and
 undefined recall \emph{independent} of what the ATE system produces. This is
@@
 definitions they reference. Once a theorem is chosen, we use the referenced
-definitions as our sample, padding uniformly up to the required size (we ignore
-theorems which reference more definitions than our sample size).
+definitions as our sample, padded up to the required size by uniformly sampling
+from the remaining definitions (we ignore theorems which reference more
+definitions than our sample size).
 
-We form the ground truth as above: selecting all applicable theorems from the
-corpus, which is now guaranteed to be non-empty.
+We form the ground truth as above: selecting all applicable theorems
+from the corpus, which is now guaranteed to be non-empty.
 
@@
 execution of the ATE system proceeds as in prior work. We must translate the
-definitions into the required input format, which depends on the particular
-tool being studied, then we time the execution, with a cutoff after e.g. 1 hour.
+chosen definitions into the particular input format of the tool under study,
+then we time the execution, with a timeout after e.g. 1 hour.
 
-In our experiments we have found that memory usage also plays an important role,
-but rather than introducing an extra dimension we can account for this by simply
-allowing the system
+In our experiments we have found that memory usage is also important, but rather
+than complicating our analysis with an extra dimension, we instead set a limit
+(as large as practical) and terminate program runs which cross it.~\footnote{We
+  could treat the hardware's RAM capacity as our limit, allow the operating
+  system to swap any overflow to disk, and wait for the inevitable slowdown to
+  trigger the timeout. This requires no work to implement, but wastes a lot of
+  time waiting for runs which could otherwise be terminated early.}
+
+This approach process produces a single runtime, precision and recall value for
+each sample. We propose two methods for analysing this data: \emph{summarising}
+the performance of an ATE system, or \emph{comparing} the performance of
+\emph{two} ATE systems.
+
+\subsection{Summarising}
+
+ - Giving each run a different sample is important for getting independent
+   values which test the generality of the system. This is preferable to e.g.
+   timing each sample multiple times.
+
+ - What sorts of aggregate data to report: median vs mean, variance, quartiles,
+   etc.
+
+\subsection{Comparison}
+
+To encourage competition in the field, it is important to compare the relative
+performance of different ATE systems on the same task. Since the aggregate
+statistics in the above summaries do not include details of specific runs, any
+comparison based on them would have very low statistical power.
+
+To produce more salient results, we should compare the data points directly, for
+example with a \emph{paired difference test}: for each individual sample, we
+find the \emph{difference} between the two systems' data \emph{for that sample}.
+We then aggregate these differences, to determine whether one system performs
+significantly better or worse than the other.
+
+ - Good because: handles the variance of results well
+ - Good because: approximates a normal distribution, making meaningful statistics easier
+ - Good because: conservative in the face of timeouts
+
+ - Question: do we use relative difference, e.g. [ 1.0:1.2, 1.0:1.5, ...], or do
+   we use absolute differences, e.g. [100s:120s, 1s:1.5s, ...]?
 
@@
 
+To aid reproducibility, we use a cryptographic hash as our source of randomness:
+the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
+corpus; the latter prevents tailoring the corpus to influence the choices,
+although ATE systems could .
+
 % evaluation

### 2017-10-27 TEBenchmarkPaper/article.tex
More work

@@
 \subsection{Preparation}
+\label{section:prep}
 
@@
 Some problem sets are more amenable for this purpose than others. The most
-suitable are those which meet the following criteria:
+suitable are those which have the following properties:
 
 \begin{itemize}
-\item For each problem, there should be a clear separation between the theorem
+\item For each problem, there should be a clear distinction between the theorem
   to be proved and the definitions involved, such that the two can be easily and
@@
 \item Since it will act as our ground truth, the problem set should ideally
-  contain \emph{every} ``interesting'' conjecture involving its included
+  contain every ``interesting'' conjecture involving its included
   definitions. Realistically, we should aim for each definition to appear in
@@
 
-Once a theorem proving problem set has been chosen, we must separate the
-definitions referenced by the theorems from the theorem statements themselves.
-The former will be used as input to the ATE systems under evaluation, whilst the
-latter form a ground truth corpus against which to compare the ATE output.
+Once such a problem set has been chosen, we must separate the definitions
+referenced by the theorems from the theorem statements themselves. The former
+will be used as input to the ATE systems under evaluation, whilst the latter
+form a ground truth corpus against which to compare the ATE output.
 
@@
 
-We form the ground truth as above: selecting all applicable theorems
-from the corpus, which is now guaranteed to be non-empty.
+When we form the ground truth for such samples, we're guaranteed to find at
+least one applicable theorem from the corpus (the one we chose as our starting
+point).
 
@@
 
-In our experiments we have found that memory usage is also important, but rather
-than complicating our analysis with an extra dimension, we instead set a limit
-(as large as practical) and terminate program runs which cross it.~\footnote{We
-  could treat the hardware's RAM capacity as our limit, allow the operating
-  system to swap any overflow to disk, and wait for the inevitable slowdown to
-  trigger the timeout. This requires no work to implement, but wastes a lot of
-  time waiting for runs which could otherwise be terminated early.}
+In our experiments we have found that memory usage is also an important part of
+a system's performance, but rather than complicating our analysis with an extra
+dimension, we instead set a limit (as large as practical) and terminate program
+runs which cross it.~\footnote{We could treat the hardware's RAM capacity as our
+  limit, allow the operating system to swap any overflow to disk, and wait for
+  the inevitable slowdown to trigger the timeout. This requires no work to
+  implement, but wastes a lot of time waiting for runs which could otherwise be
+  terminated early.} This is in line with the expected usage of these systems:
+either there is enough memory, or there isn't; implementations shouldn't be
+penalised for making use of available resources.
 
-This approach process produces a single runtime, precision and recall value for
-each sample. We propose two methods for analysing this data: \emph{summarising}
+This approach produces a single runtime, precision and recall value for each
+sample. We propose two methods for analysing this data: \emph{summarising}
 the performance of an ATE system, or \emph{comparing} the performance of
@@
 
- - Giving each run a different sample is important for getting independent
-   values which test the generality of the system. This is preferable to e.g.
-   timing each sample multiple times.
+Each sample is only explored once, so that we cover as many independent samples
+as possible to better estimate how a system's performance generalises to unseen
+inputs. To combine these data into an aggregate summary depends on what we are
+interested in measuring.
 
- - What sorts of aggregate data to report: median vs mean, variance, quartiles,
-   etc.
+One general question we might ask is how a system's performance scales with
+respect to the input size (the number of definitions in the theory). This is
+straightforward to measure by varying the sample size, but we need some way to
+combine the results from samples of the same size.
+
+The precision and recall for multiple samples $S$ can be combined in two ways:
+finding their mean value (the \emph{average of the ratios}) or by summing the
+output and ground-truth counts \emph{before} taking their ratios (the
+\emph{ratio of the averages}). The former estimates the precision and recall we
+can expect from any \emph{particular} run, which is of direct relevance to users
+of these tools. The latter estimates the proportion of discoveries that a tool
+can expect to find out of 'all possible', i.e. assigning higher weight to
+performance in theories which have more conjectures available to be discovered;
+whilst this is of theoretical interest, it is a more difficult claim to support
+empirically, so we leave its investigation to future work.
+
+We summarise the runtimes by choosing their median, as it is more robust against
+long-running outliers, and hence represents performance for a ``typical''
+theory.
+
+Another advantage to our statistical approach is that we can compute the
+\emph{variance} of our data, which is unknown in the one-off evaluations which
+have been performed previously. In particular, we might be concerned that such
+previous evaluations used ``popular'' theories, like those involving natural
+numbers, which may be unrepresentative of typical usage: any number theorist
+knows how amenable natural numbers are to conjecture and exploration, and we
+should not expect a typical user's theory (e.g. a software library) to be
+similarly rich.
 
@@
 statistics in the above summaries do not include details of specific runs, any
-comparison based on them would have very low statistical power.
+comparison based on them would have very low statistical power, and therefore it
+is appropriate to use an alternative approach when comparing.
 
-To produce more salient results, we should compare the data points directly, for
-example with a \emph{paired difference test}: for each individual sample, we
-find the \emph{difference} between the two systems' data \emph{for that sample}.
-We then aggregate these differences, to determine whether one system performs
-significantly better or worse than the other.
+We propose comparing the data points directly, using a \emph{paired difference
+  test}: for each individual sample, we find the \emph{difference} between the
+two systems' data \emph{for that sample}. We then aggregate these
+\emph{differences}, to determine whether one system performs significantly
+better or worse than the other.
 
- - Good because: handles the variance of results well
- - Good because: approximates a normal distribution, making meaningful statistics easier
- - Good because: conservative in the face of timeouts
-
- - Question: do we use relative difference, e.g. [ 1.0:1.2, 1.0:1.5, ...], or do
-   we use absolute differences, e.g. [100s:120s, 1s:1.5s, ...]?
+This neatly avoids the variance inherent in the summary statistics, and (with
+enough samples) even approaches a normal distribution. Paired difference tests
+are also robust to the censoring caused by timeouts: upper-bounding the run time
+gives an upper bound to the (observable) differences, biasing our inference
+towards the null hypothesis (indistinguishable performance) rather than
+unfounded conclusions.
 
@@
 
+\section{Application}
+\label{sec:application}
+
 % Better presentation, flowchart?
@@
 
-\section{Application}
-\label{sec:application}
+We have applied our benchmarking methodology to the Tons of Inductive Problems
+(TIP) problem set for automated theorem provers. TIP satisfies the desiderata of
+section\ref{section:prep} since each problem has standalone type and function
+definitions, making their separation trivial. The problem set includes known
+examples from the software verification and inductive theorem proving
+literature, including higher-order functions, common datatypes, etc., ensuring
+that benchmark results are relevant to those fields. It also includes tools to
+convert the problems into a variety of languages, including Haskell and
+Isabelle.
 
+TIP version 0.2 contains FIXME theorem statements, each of which is accompanied
+by definitions of the types and functions it uses (other than those for integers
+and booleans, which are ``built in''). There are FIXME function definitions, of
+which most are duplicates. We remove $\alpha$-equivalent definitions in the
+following way:
+
+\begin{enumerate}
+\item We prefix all type and function names with the filename in which we found
+  the definition, so that e.g. the theorem statement from
+  \texttt{tip2015/int\_add\_assoc.smt2} references the \texttt{plus} function from
+  that file, rather than e.g. the \texttt{plus} function from
+  \texttt{isaplanner/prop\_07.smt2}.
+\item For each definition, all local variables (including function arguments,
+  constructor parameters, pattern-matching variables and \texttt{let} bindings)
+  are replaced by standard names \texttt{local-1}, \texttt{local-2}, etc. in a
+  well-defined order.
+\item An $O(n^2)$ comparison of each definition to every other definition is
+  performed, to find any which are syntactically-identical modulo the names
+  being defined (but \emph{not} the names being referenced, i.e. ). For those which are,
+  we record which names are equivalent, replace all references to these names
+  with whichever is smallest lexicographically, and remove
+\end{enumerate}
+
+Altogether TIP problem
+by QuickSpec and IsaCoSy theory
+exploration systems. To assess the scalability of their algorithms, we sampled
+theories of various sizes: from theories containing only a single definition, up
+to theories containing 20 definitions.
  We call the Theory Exploration Benchmark (TEB).
@@
 although ATE systems could .
-
-% evaluation
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy theory
-exploration systems. To assess the scalability of their algorithms, we sampled
-theories of various sizes: from theories containing only a single definition, up
-to theories containing 20 definitions.
 

### 2017-10-27 TEBenchmarkPaper/article.tex
More citations, etc.

@@
 Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as possible.
-These problem sets are an opportunity for theory exploration, as their
-definitions and theorems can be used as a corpus, in the same way that Isabelle
-libraries have been used in the past.
+and regular competitions to prove as much as possible, as fast as
+possible\cite{pelletier2002development}. These problem sets are an opportunity
+for theory exploration, as their definitions and theorems can be used as a
+corpus, in the same way that Isabelle libraries have been used in the past.
 
@@
   to be proved and the definitions involved, such that the two can be easily and
-  meaningfully separated. This rules out problem sets like SMTLIB, where many
-  problems \emph{declare} a function, but leave its behaviour \emph{implicit} in
-  the logical structure of the theorem statement.
+  meaningfully separated. This rules out problem sets like those of
+  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
+  functions, whose behaviour is \emph{implicit} in the logical structure of the
+  theorem statement but not separately \emph{defined}.
 \item Definitions should be \emph{strongly typed}, such that they can be
@@
   desire higher-order functions and inductively defined types, which rules out
-  first-order languages/logics.
+  first-order languages/logics (such as TPTP~\cite{sutcliffe2009tptp}).
 \item Since it will act as our ground truth, the problem set should ideally
@@
 This neatly avoids the variance inherent in the summary statistics, and (with
-enough samples) even approaches a normal distribution. Paired difference tests
-are also robust to the censoring caused by timeouts: upper-bounding the run time
+enough samples) approaches a normal distribution. Paired difference tests are
+also robust to the censoring caused by timeouts: upper-bounding the run time
 gives an upper bound to the (observable) differences, biasing our inference

### 2017-10-30 TEBenchmarkPaper/atp-ate.tex
Add Alison sections (separate files ATM)

\documentclass[11pt]{article}
%\setlength{\textwidth}{6.5in}
%\addtolength{\voffset}{-5pt}
%\frenchspacing
\setlength\parindent{0pt}
\usepackage{parskip}
\usepackage{times}
\usepackage{amsfonts}
\usepackage[square]{natbib}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{amssymb}

% \usepackage[square]{natbib}

%\newcommand{\icarus}{\textsc{ICARuS}\xspace}
%\newcommand{\oants}{\mbox{$\Omega$}\textsc{ants}\xspace}
%\newcommand{\gc}{\textsc{GC}\xspace}
%\newcommand{\gcatf}{\textsc{GC-ATF}\xspace}
%\newcommand{\homer}{\textsc{Homer}\xspace}
%\newcommand{\gcicarus}{\textsc{GC-ICARuS}\xspace}

%\title{Section 2}

\begin{document}
\section{Background}

\subsection{Motivation}
Given a logical theory, such as the theory of lists shown in Figure 1,
we may want to find theorems which describe its behaviour. This could
be for mathematical curiosity, or due to the theory's importance in
some domain. In particular, for theories which capture the semantics
of some software library, we may want to verify that certain
(un)desirable properties do (not) hold. We might also want to optimise
programs using this library, rewriting expressions into a form which
requires less time, memory, network usage, etc. To avoid altering a
program's result, such rewrites should come with theorems proving
their correctness, such as the following theorem for our theory of
lists:

8f:8xs:8ys:map(f; append(xs; ys)) = append(map(f; xs); map(f; ys)) (1)
[**]

This justifies a rewrite rule for splitting a single map call into
multiple independent calls dealing with different sections of a
list. Such rules are useful optimisations since each call can be
evaluated in parallel, leading to the \map/reduce" programming
paradigm.  All of these example use cases require the ability to
discover theorems about some particular theory. This is a hard problem
in general, but presents opportunities for automation due to the
precise, symbolic nature of the domain.

\subsection{Automating Construction and Exploration of Theories}
The task of discovering theorems in an arbitrary theory can be
described as the interplay of several processes: (i) formulating new
definitions and concepts; (ii) finding patterns suitable for posing as
conjectures; and (iii) proving that those conjectures are theorems.
The latter is studied extensively in the field of Automated Theorem
Proving (ATP). Far less attention has been paid to automating the
first two tasks, which both Automated Theory Formation (ATF) and
Automated Theory Exploration (ATE) address.  We discuss the
relationship between ATF and ATE in historical context in
$\S$\ref{exploration-versus-formation}.

A major challenge for both ATF and ATE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'',
since this is an imprecise term with many different
interpretations. For example, all existing approaches agree that
simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton {\em et al.}  give a survey of
systems for theory formation and exploration, and their notions of
``interestingness'' for concepts and conjectures, identifying six
notions in total.

%methodology for getting these measures.

%interestingness paper

 are identified in these systems, their use by each
is summarised, along with (our interpretation of) their use
in three more recent ATE systems. These notions, as applied to ATE,
are:

{\bf Empirical plausibility}, which checks whether a property holds
across some specific examples. This is especially useful for avoiding
false conjectures, without resorting to a full proof search.

{\bf Novelty} is whether a conjecture, or one isomorphic or more
general, has already been seen.

{\bf Surprisingness} of a conjecture is whether or not it is
``obvious'', for example if it is an instance of a tautology.

{\bf Applicability} depends on the number of models in which a
conjecture holds. The system of Bagai et al conjectures the
non-existence of objects, and hence favours statements with zero
applicability. Other systems treat ap- plicability as a positive
aspect: the more applicable the statement, the more interesting it is.

{\bf Comprehensibility} is the complexity of a statement. Simpler
statements are considered more interesting, and many of the search
algorithms explore simpler/smaller statements before complex/larger
ones, to more efficiently find those which are interesting.

{\bf Utility} is the relevance or usefulness of a conjecture to the
user's particular task. For example, if we want to find optimising
rewrite rules such as equation 1, then utility would include whether
or not a conjecture justifies a rewrite rule, the difference in
resource usage of the expressions involved, and how common those
expressions are in real usage.

We summarise the criteria in Table 1 and state which were used in key
ATF/ATE systems, where Colton identified usage in AM, GT, Graffiti and
Bagai et al, and his own ATF system HR, and we identify usage in three
more recent ATE systems QuickSpec, IsaCoSy and IsaScheme. These
criteria, especially empirical plausibility and comprehensibility, are
not only used for analysing results after the fact, but instead may
form a core part of an algorithm's design decisions. For example, the
designers of IsaCoSy consider simple substitutions of existing
conjectures (i.e. those with low novelty) to be uninteresting, and
hence their system includes a constraint solving component which
avoids generating such statements entirely.

With such ambiguous and varied goals, approaches and assessment
criteria it is difficult to compare ATE systems in a quantitative way,
and hence to form some measure of ``progress'' for the field.

\section{Theory Exploration versus Theory
  Formation}\label{exploration-versus-formation}

In this section we put the Mathematical Theory Exploration approach
into context both historically, and with respect to related approaches
such as in Automated Theory Formation systems.

\subsection{A Short History of Automated Mathematics}

\begin{quote}
...``in his subsequent design for an Analytical Engine Mr. Babbage has
shown that material machinery is capable, in theory at least, of
rivalling the labours of the most practised mathematicians in all
branches of their science.'' \cite[p. 498]{jevons}
\end{quote}

The first automated mathematical system, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues \cite{d'ocagne}. Subsequent early systems include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees \cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable \cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
systems were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

Differences in background and motivation continue to be relevant
today. While the majority of work in automating mathematics has been
in symbolic manipulation and theorem proving, we are concerned here
with other aspects of mathematics, including the construction of
concepts, axioms, conjectures, examples and theorems. This is
varyingly known as ``Automated Theory Formation''
\cite{lenat:77,colton:book}, ``Automated Mathematical Theory
Exploration'', ``Mathematical Theory Exploration''
\cite{buchberger:06} (also sometimes prefaced with ``Computer-Aided''
or ``Algorithm-Supported''), ``Automated Mathematical Discovery''
\cite{epstein:91,colton:interestingness,esarm2008}, ``Concept
Formation in Discovery Systems'' \cite{haase}, and ``Automated Theorem
Discovery'' \cite{roy}. Such a plethora of terminology can be
unhelpful and can mask similarities between the different fields. In
particular, the twin strands of Automated Theory Formation and
Automated Mathematical Theory Exploration seem to be developing
somewhat independently without a clear differentiating
methodology. Below we discuss commonalities and differences between
the two schools of thought.

%we limit our discussion to systems which work in maths: there are
%others which work in scientific discovery which are also relevant.

\subsection{Automated Theory Formation}

% 1. history of terminology
Automated theory formation, the focus of this paper, derives its
terminology from psychology in which the term ``concept formation'' is
used (see, for example, \cite{bruner:67}) to describe the search for
features which differentiate exemplars from non-exemplars of various
categories. Lenat used this term in his 1977 paper: {\em Automated
  Theory Formation in Mathematics} \cite{lenat:77} and we have
extended it to describe a family of techniques used to construct and
evaluate concepts, conjectures, examples and proofs (embodied in the
HR, HRL and GC systems and derivatives, as described above).

% 4. example systems - AM, HR, ...
When Lenat built the AM system \cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in \cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral
\cite{buchanan:75}. No system could perform both of these tasks: Lenat
saw this as the next step. AM was designed to both construct new
concepts and conjecture relationships between them; fully automating
the cycle of discovery in mathematics. Lenat describes this as
follows:

\begin{quote} 
``What we are describing is a computer program which
defines new concepts, investigates them, notices
regularities in the data about them, and conjectures
relationships between them. This new information is used
by the program to evaluate the newly-defined concepts,
concentrate upon the most interesting ones, and iterate the
entire process.'' \cite[p. 834]{lenat:77}
\end{quote}

AM was a rule-based system which used a frame-like scheme to represent
its knowledge, enlarged its knowledge base via a collection of
heuristic rules, and controlled the firing of these rules via an
agenda mechanism. Lenat chose elementary arithmetic as the development
domain because he could use personal introspection for the heuristics
for constructing and evaluating concepts. Given the age of this
discipline, Lenat thought it unlikely that AM would make significant
discoveries, although he did cite its ``ultimate achievements'' as the
concepts and conjectures it discovered (or could have discovered). He
suggested various criteria by which his system could be evaluated,
many of which focused on an exploration of the techniques. For
instance, he considered generality (running AM in new domains) and how
finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). %most of which were qualitative.
Lenat saw his system and future developments in this field as having
implications for mathematics itself (finding results of significance),
for automating mathematics research (developing AI techniques), and
for designing ``scientist assistant'' programs (aids for
mathematicians). This shows a broad spread of motivation. Despite the
seeming success of the AM system, it is one of the most criticised
pieces of AI research. In their case study in methodology, Ritchie and
Hanna analysed Lenat's written work on AM and found that there was a
large discrepancy between his theoretical claims and the implemented
program \cite{partridge}. For instance, Lenat made claims about how AM
invented natural numbers from sets, whereas it used one heuristic
which was specifically written in order to make this connection (and
not used in any other context). Another problem was that the processes
were sometimes under-explained. For an argument of why many of the
claims made by Lenat about AM were false, see chapter 13 of
\cite{colton:book}.

\subsection{Mathematical Theory Exploration}

The phrase Mathematical Theory Exploration (MTE) is a recent term
which seems to have originated with Buchberger and colleagues (see,
for example, \cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm \cite{buchberger:04}.

Recent systems developed in this area include MATHsAiD \cite{roy},
IsaScheme \cite{MontanoRivas2011} and IsaCosy \cite{johansson}. The
goal of the MATHsAiD (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions) project is to build a tool which
takes in a set of axioms, concept definitions and a logic and applies
its inference rules to reason from the axioms to theorems. The
motivation is to produce a tool which will help mathematicians to
explore the consequences of a set of axioms or a particular concept
definition. One of the main challenges of the project has been to
automatically evaluate interestingness: to distinguish important
theorems from results which, although they follow from a set of
axioms, are of little mathematical interest.

%IsaScheme
Monta{\~n}o-Rivas {\em et al.} have implemented a scheme-based
approach to MTE in their IsaScheme system
\cite{MontanoRivas2011}. Schemes are higher-order formulae which can
be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention
process.  For instance, in the theory of natural numbers, given the
concepts of successor ($suc$), addition ($+$) and zero ($0$),
IsaScheme can use the following scheme to invent the concept of
multiplication:

\[
\text{def-scheme}(g,h,i,j) \equiv\\ 
  \exists f. \forall x y. \left\{
  \begin{array}{l l}
    f(g,y) = h (y)  \: \text{and}  & \\
    f(i(x),y) = j(y,f(x,y)) & \\
  \end{array} \right.
\]
where the existentially quantified variable $f$ stands for the new
function to be defined in terms of the variables g, h, i and j. Within
the theory of natural numbers, IsaScheme instantiates this scheme with
$\sigma_1 = \{g \mapsto 0, h \mapsto (\lambda x.0), i \mapsto suc, j
\mapsto +\}$. In this example, $f \mapsto *$ (multiplication), since:

\begin{align*} 
0*y  &= 0 \\
suc(x) * y &=y + (x*y).
\end{align*}
The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from
\cite{MontanoRivas2011}).

%IsaCosy
The IsaCosy system (Isabelle Conjecture Synthesis) is a program for
inductive theory formation, which synthesises conjectures from the
available constants and free variables. Only terms which do not
contain subterms that can be rewritten by a set of rewrite rules can
be synthesised.  Conjectures are tested by sending them to a
counterexample checker and, if no counterexamples are found, then sent
to IsaPlanner which attempts to prove them.

% Discussion.

\subsection{Commonalities and Differences between MTE and ATF}

We believe that ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches
seem to be that in MTE:
\begin{itemize}
\item the main proponents define themselves as mathematicians (they
  hold a PhD and have experience in research mathematics);
\item the primary motivation is to support mathematicians;
\item systems tend to be user-interactive;
\item systems tend to be specific to mathematics,
\end{itemize}

while in ATF:

\begin{itemize}
\item the main proponents define themselves as AI researchers (they
  often have a mathematics background up to Masters level, but their
  PhD and research experience is in AI);
\item the primary motivation is to extend AI techniques; a secondary
  motivation is to produce interesting new mathematics;
\item systems tend to be fully automated;
\item systems can often be applied to non-mathematical domains.
\end{itemize}

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to \cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and current
systems such as MATHsAiD \cite{roy}, IsaScheme \cite{MontanoRivas2011}
and IsaCosy \cite{johansson} are bridging the methodological gap
between ATF and MTE. This is particularly true in terms of
fully-automatic/user-interactive and the domain-specific/general
aspects.

\section{Discussion}

methodology behind Simon's paper

agreement in maths? matt inglis

other - zeki/two papers with ursula...

\bibliographystyle{plain} 
%\bibliography{biblio}
\bibliography{biblio}
\end{document}

### 2017-11-13 TEBenchmarkPaper/article.tex
Add timing graph
This is from example runs at the moment (few repetitions, short cutoffs).

The scripts should apply generally though.
Much addition

@@
 finding their mean value (the \emph{average of the ratios}) or by summing the
-output and ground-truth counts \emph{before} taking their ratios (the
-\emph{ratio of the averages}). The former estimates the precision and recall we
-can expect from any \emph{particular} run, which is of direct relevance to users
-of these tools. The latter estimates the proportion of discoveries that a tool
-can expect to find out of 'all possible', i.e. assigning higher weight to
-performance in theories which have more conjectures available to be discovered;
-whilst this is of theoretical interest, it is a more difficult claim to support
-empirically, so we leave its investigation to future work.
+numerators and denominators (the \emph{ratio of the averages}). The former gives
+the expected precision and recall for a run, which is of direct relevance to
+users of these tools, and is hence our preferred measure. The latter assigns more weight to theories which generate more
+conjectures, but that is not justification there is no reason to consider these theories ``better'',
 
@@
 
-% Better presentation, flowchart?
+We have applied our methodology to evaluating QuickSpec, %TODO:cite
+a library for conjecturing equations about functions in the Haskell programming
+language. We use the Tons of Inductive Problems (TIP) theorem proving benchmark
+to generate our corpus.
+
+\subsection{QuickSpec}
+
+We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
+is a library for the Haskell programming language, which is given a
+\emph{signature} containing Haskell functions and (universally quantified) typed
+variables, and outputs conjectured equations involving these terms.
+
+The QuickSpec algorithm first enumerates all well-typed expressions, up to a
+certain depth (3 by default), involving the given terms. Expressions of the same
+type are assumed to be equivalent, and grouped into equivalence classes. A
+testing process then tries to disprove this assumption by instantiating each
+variable with a randomly chosen value (using the QuickCheck testing library),
+evaluating the expressions and comparing the results. The equivalence classes
+are split to separate any expressions which are observed to be different, and
+the process repeats with new random values.
+
+After 500 rounds of testing, any expressions still sharing the same equivalence
+class are conjectured to be equal. Finally, a congruence closure algorithm is
+applied to these equations, to find a minimal set.
+
+In order to thoroughly benchmark QuickSpec, we need to automate some of the
+decisions which are normally left up to the user:
+
 \begin{itemize}
-\item Collect together all definitions found in TIP into one large theory,
-  removing $\alpha$-equivalent duplicates.
-\item For each type, provide extra functions which act as constructors and
-  destructors. For example, the $\List$ type would have extra constructor
-  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
-  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
-  This allows us to ignore constructors and destructors, and concentrate solely
-  on functions.
-\item Collect together all theorem statements found in TIP. Update any
-  references to removed definitions, and replace references to constructors with
-  the associated constructor function from above.
-\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
-  that we choose all of the names referenced by at least one of the theorems
-  (for example, theorem \ref{eq:mapreduce} references the names
-  $\Map$ and $\Append$).
-\item Provide this sub-set as an input to the ATE system, timing how long it
-  takes to complete.
-\item Perform precision/recall analysis on the resulting conjectures, against
-  the ground truth of TIP theorems which only reference names in this sample.
+\item We must decide what variables to include. We add three for each type that
+  appears as a function argument, except for types which have no QuickCheck data
+  generator.
+\item We must \emph{monomorphise} all types. For example, functions like
+  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
+  We need to pick a specific type in order to generate random values, so we
+  resolve this, arbitrarily, by picking
+  \emph{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
+    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
+    kind \texttt{* -> *}. If these wouldn't satisfy some constraint, we pick a
+    suitable type non-deterministically from the module scope, or skip these
+    variables completely if none is found.}
+\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
+  so functions of multiple arguments must be simulated by currying.
+  Unfortunately QuickSpec cannot compare functions for equality, so we must give
+  each function in our signature an explicit arity. Arities from zero to five
+  are supported, so we avoid partial application by picking the highest that is
+  type-correct.
 \end{itemize}
+
+To find out which functions to include in a signature, we use a plugin for the
+GHC compiler which tells us all of the functions defined in a package. We
+include all of those which type-check when passed to QuickSpec.
+
+\subsection{TIP}
+
+We use TIP version 0.2 which contains 343 problems, each stating a single
+theorem and together defining a total of 618 datatypes and 1498 functions. Most
+of these are duplicates, since each problem (re-)defines all of the datatypes
+and functions it involves. We prefix the names of all definitions with the
+filename it came from, and collect them into one big set.
+
+Datatypes can have several ``constructors'' and ``destructors''. For example the
+natural numbers can be defined as follows:
+
+\begin{verbatim}
+(declare-datatypes () ((Nat (Z) (S (p Nat)))))
+\end{verbatim}
+
+This defines the type \texttt{Nat}, along with the constructors \texttt{Z} (for
+zero) and \texttt{S} (the successor function), and destructor \texttt{p} (the
+predecessor function).
+
+Since our GHC plugin only lists function definitions, we add separate function
+definitions for each constructor and destructor (these are just
+$\eta$-expansions).
+
+TIP has ``native'' support for booleans and integers, allowing functions like
+\texttt{+} to be used without an accompanying definition. Without a definition
+these functions will not be included in the QuickSpec exploration, so we replace
+all occurrences with custom definitions.
+
+This gives us 3598 definitions in total, with 269 remaining after
+$\alpha$-equivalent duplicates are removed.
+
+TIP includes a tool to translate its definitions into Haskell code, including
+suitable comparison functions and random data generators. We translated all 269
+definitions into a single Haskell package, compiled it with GHC and extracted
+the function definitions with our plugin. Each benchmark run imported this
+entire package, but the signature given to QuickSpec only included functions
+chosen for the current sample.
+
+\subsection{Results}
+
+\begin{figure}
+  \includegraphics[width=8cm]{graphs/time}
+  \includegraphics[width=8cm]{graphs/precRec}
+  \caption{}
+  \label{figure:quickspec_results}
+\end{figure}
+
+We applied our sampling methodology to the 269 TIP definitions for sample sizes
+between 1 and 20, above which point QuickSpec becomes very slow. We chose 30
+samples for each size, but the smaller sizes included repetitions due to the
+requirement for samples to admit at least one theorem. The results are shown in
+figure~\ref{figure:quickspec_results}.
+
+Precision and recall were calculated by normalising each equation generated by
+QuickSpec: all variables are identified numerically, in ascending order from
+left to right; equations were also sorted such that the expression on the
+left-hand-side is lexicographically smaller than that on the right.
+
+Theorem statements from TIP were first updated to reference only the 269
+distinct names we explored, and references to constructors and destructors were
+updated to use the $\eta$-expanded versions instead. Equations were normalised
+in the same way as above (non-equations were kept unchanged), and elements of
+these two sets were deemed to match if they were syntactically identical.
 
@@
 
-Altogether TIP problem
-by QuickSpec and IsaCoSy theory
+Altogether TIP problem by QuickSpec and IsaCoSy theory
 exploration systems. To assess the scalability of their algorithms, we sampled

### 2017-11-16 TEBenchmarkPaper/article.tex
More changes

@@
 the expected precision and recall for a run, which is of direct relevance to
-users of these tools, and is hence our preferred measure. The latter assigns more weight to theories which generate more
-conjectures, but that is not justification there is no reason to consider these theories ``better'',
+users of these tools, and is hence our preferred measure. The latter assigns
+more weight to theories which generate more conjectures, but we avoid this since
+there is no \emph{a priori} justification for considering such theories as
+``better''.
 
@@
 Another advantage to our statistical approach is that we can compute the
-\emph{variance} of our data, which is unknown in the one-off evaluations which
-have been performed previously. In particular, we might be concerned that such
-previous evaluations used ``popular'' theories, like those involving natural
-numbers, which may be unrepresentative of typical usage: any number theorist
-knows how amenable natural numbers are to conjecture and exploration, and we
-should not expect a typical user's theory (e.g. a software library) to be
-similarly rich.
+\emph{variance} of our data, which is unknown in the previous one-off
+evaluations. In particular, we might be concerned that such previous evaluations
+used ``popular'' theories, like those involving natural numbers, which may be
+unrepresentative of typical usage: we should not expect a typical user's theory
+(e.g. a software library) to be as mathematically rich as number theory!
 
@@
 We have applied our methodology to evaluating QuickSpec, %TODO:cite
-a library for conjecturing equations about functions in the Haskell programming
-language. We use the Tons of Inductive Problems (TIP) theorem proving benchmark
-to generate our corpus.
+a library for conjecturing equations about functions written in the Haskell
+programming language. We use the Tons of Inductive Problems (TIP) theorem
+proving benchmark to generate our corpus.
 
@@
 We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
-is a library for the Haskell programming language, which is given a
-\emph{signature} containing Haskell functions and (universally quantified) typed
-variables, and outputs conjectured equations involving these terms.
+is a Haskell library which takes a \emph{signature} containing Haskell functions
+and (universally quantified) typed variables, and outputs conjectured equations
+involving these terms.
 
-The QuickSpec algorithm first enumerates all well-typed expressions, up to a
+The QuickSpec algorithm first enumerates all well-typed expressions up to a
 certain depth (3 by default), involving the given terms. Expressions of the same
@@
   We need to pick a specific type in order to generate random values, so we
-  resolve this, arbitrarily, by picking
-  \emph{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
+  resolve this (arbitrarily) by picking
+  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
     \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
-    kind \texttt{* -> *}. If these wouldn't satisfy some constraint, we pick a
-    suitable type non-deterministically from the module scope, or skip these
-    variables completely if none is found.}
+    kind \texttt{* -> *}. If these violate some constraint, we pick a suitable
+    type non-deterministically from the module scope, or skip these variables
+    completely if none is found.}
 \item Haskell functions only take one argument, as as in the $\lambda$-calculus,
-  so functions of multiple arguments must be simulated by currying.
+  so functions of multiple arguments are simulated by currying.
   Unfortunately QuickSpec cannot compare functions for equality, so we must give
@@
 
+We chose TIP for our ground truth since it satisfies the desiderata of
+section~\ref{section:prep}: each problem has standalone type and function
+definitions, making their separation trivial; known examples from the software
+verification and inductive theorem proving literature are included, ensuring
+relevance to those fields; the format includes higher-order functions and
+inductive datatypes, which we are interested in; plus it is accompanied by
+tooling to convert this format into a variety of languages, including Haskell
+and Isabelle.
+
 We use TIP version 0.2 which contains 343 problems, each stating a single
@@
 of these are duplicates, since each problem (re-)defines all of the datatypes
-and functions it involves. We prefix the names of all definitions with the
-filename it came from, and collect them into one big set.
+and functions it involves. To avoid ambiguity we prefix the names of all
+definitions with the filename it came from, and collect them into one big set.
 
-Datatypes can have several ``constructors'' and ``destructors''. For example the
-natural numbers can be defined as follows:
+TIP datatypes can have several ``constructors'' and ``destructors''. For example
+the natural numbers can be defined as follows:
 
@@
 
-Since our GHC plugin only lists function definitions, we add separate function
-definitions for each constructor and destructor (these are just
-$\eta$-expansions).
+Since our GHC plugin doesn't list constructors or destructors, we add separate
+function definitions for each (these are just $\eta$-expansions).
 
@@
 samples for each size, but the smaller sizes included repetitions due to the
-requirement for samples to admit at least one theorem. The results are shown in
+requirement for samples to admit at least one theorem. A timeout of 3 minutes
+was imposed, since preliminary experiments showed that most runs either finish
+before this time, or still do not finish after an hour. The results are shown in
 figure~\ref{figure:quickspec_results}.
@@
 updated to use the $\eta$-expanded versions instead. Equations were normalised
-in the same way as above (non-equations were kept unchanged), and elements of
-these two sets were deemed to match if they were syntactically identical.
+in the same way as above, whilst non-equations and those containing
+\textt{lambda} functions (which QuickSpec cannot synthesise) were replaced with
+null values. Elements of these two sets were deemed to match if they were
+syntactically identical.
 
-We have applied our benchmarking methodology to the Tons of Inductive Problems
-(TIP) problem set for automated theorem provers. TIP satisfies the desiderata of
-section\ref{section:prep} since each problem has standalone type and function
-definitions, making their separation trivial. The problem set includes known
-examples from the software verification and inductive theorem proving
-literature, including higher-order functions, common datatypes, etc., ensuring
-that benchmark results are relevant to those fields. It also includes tools to
-convert the problems into a variety of languages, including Haskell and
-Isabelle.
+% Altogether TIP problem by QuickSpec and IsaCoSy theory
+% exploration systems. To assess the scalability of their algorithms, we sampled
+% theories of various sizes: from theories containing only a single definition, up
+% to theories containing 20 definitions.
+%  We call the Theory Exploration Benchmark (TEB).
 
-TIP version 0.2 contains FIXME theorem statements, each of which is accompanied
-by definitions of the types and functions it uses (other than those for integers
-and booleans, which are ``built in''). There are FIXME function definitions, of
-which most are duplicates. We remove $\alpha$-equivalent definitions in the
-following way:
+% To do this we can take a set of ATP tasks, for example proving the equation
+% \ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
+% split apart the definitions (to form a theory) from the theorem statements (to
+% form our ground truth).
 
-\begin{enumerate}
-\item We prefix all type and function names with the filename in which we found
-  the definition, so that e.g. the theorem statement from
-  \texttt{tip2015/int\_add\_assoc.smt2} references the \texttt{plus} function from
-  that file, rather than e.g. the \texttt{plus} function from
-  \texttt{isaplanner/prop\_07.smt2}.
-\item For each definition, all local variables (including function arguments,
-  constructor parameters, pattern-matching variables and \texttt{let} bindings)
-  are replaced by standard names \texttt{local-1}, \texttt{local-2}, etc. in a
-  well-defined order.
-\item An $O(n^2)$ comparison of each definition to every other definition is
-  performed, to find any which are syntactically-identical modulo the names
-  being defined (but \emph{not} the names being referenced, i.e. ). For those which are,
-  we record which names are equivalent, replace all references to these names
-  with whichever is smallest lexicographically, and remove
-\end{enumerate}
+% We focus on the Tons of Inductive Problems (TIP) theorem proving
+% benchmark~\cite{claessen2015tip}, since it has several desirable properties:
 
-Altogether TIP problem by QuickSpec and IsaCoSy theory
-exploration systems. To assess the scalability of their algorithms, we sampled
-theories of various sizes: from theories containing only a single definition, up
-to theories containing 20 definitions.
- We call the Theory Exploration Benchmark (TEB).
+% All together, TIP provides 219 distinct function definitions and 343
+%   theorem statements, which is enough to fully exercise current ATE systems.
+% Benchmark problems include examples from the theorem proving literature
+%   as well as common program verification tasks, ensuring the resulting corpus is
+%   relevant to researchers and practitioners.
+% TIP's maintainers provide a set of tools, including translators from TIP's
+%   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
+%   (which are used by ATE systems).
 
-To do this we can take a set of ATP tasks, for example proving the equation
-\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
-split apart the definitions (to form a theory) from the theorem statements (to
-form our ground truth).
+% To aid reproducibility, we use a cryptographic hash as our source of randomness:
+% the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
+% corpus; the latter prevents tailoring the corpus to influence the choices,
+% although ATE systems could .
 
-We focus on the Tons of Inductive Problems (TIP) theorem proving
-benchmark~\cite{claessen2015tip}, since it has several desirable properties:
-
-All together, TIP provides 219 distinct function definitions and 343
-  theorem statements, which is enough to fully exercise current ATE systems.
-Benchmark problems include examples from the theorem proving literature
-  as well as common program verification tasks, ensuring the resulting corpus is
-  relevant to researchers and practitioners.
-TIP's maintainers provide a set of tools, including translators from TIP's
-  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
-  (which are used by ATE systems).
-
-To aid reproducibility, we use a cryptographic hash as our source of randomness:
-the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
-corpus; the latter prevents tailoring the corpus to influence the choices,
-although ATE systems could .
-
-To provide a more robust comparison of these two systems, we also applied a
-paired difference test: measuring, for each sampled theory, the difference in
-time, precision and recall between the two systems.
+% To provide a more robust comparison of these two systems, we also applied a
+% paired difference test: measuring, for each sampled theory, the difference in
+% time, precision and recall between the two systems.
 
@@
 
-\begin{acknowledgements}
-  We are grateful to Jianguo Zhang for help with our statistical analysis.
-
-  EPSRC
-\end{acknowledgements}
+%\begin{acknowledgements}
+%  We are grateful to Jianguo Zhang for help with our statistical analysis.
+%
+%  EPSRC
+%\end{acknowledgements}
 

### 2017-11-17 TEBenchmarkPaper/article.tex
Merge atp-ate.tex content into article.tex

@@
 
-Given a logical theory, like the theory of lists shown in figure
-\ref{figure:list_theory}, we may want to find theorems which describe its
+Given a logical theory, such as the theory of lists shown in
+Figure~\ref{figure:list_theory}, we may want to find theorems which describe its
 behaviour. This could be for mathematical curiosity, or due to the theory's
@@
 semantics of some software library, we may want to verify that certain
-(un)desirable properties do (not) hold; we might also want to \emph{optimise}
+(un)desirable properties do (not) hold. We might also want to \emph{optimise}
 programs using this library, rewriting expressions into a form which requires
@@
 
-This justifies a rewrite rule for splitting a single $\Map$ call into
-multiple independent calls dealing with different sections of a list. Such rules
-are useful optimisations since each call can be evaluated in parallel, leading
-to the ``map/reduce'' programming paradigm.
+This justifies a rewrite rule for splitting a single $\Map$ call into multiple
+independent calls dealing with different sections of a list. Such rules are
+useful optimisations since each call can be evaluated in parallel, leading to
+the ``map/reduce'' programming paradigm.
 
@@
 
-\subsection{Theory Exploration}
+\subsection{Automating Construction and Exploration of Theories}
 \label{sec:te}
 
-The task of discovering theorems in an arbitrary theory can be described as the
-interplay of several processes:
-
-% Is it clear?
-
-\begin{itemize}
-\item \emph{Forming} or refining the theory, to add new definitions and
-  concepts.
-\item \emph{Exploring} the theory, to find patterns suitable for posing as
-  conjectures.
-\item \emph{Proving} that those conjectures are theorems.
-\end{itemize}
-
-The latter is studied extensively in the field of Automated Theorem Proving
-(ATP). Less attention has been paid to automating the first two, those of
-\emph{automated theory formation} and \emph{automated theory exploration} (ATE).
-
-We focus on ATE, where a major challenge is choosing how to narrow down the set
-of generated conjectures to those deemed ``interesting'', since this is an
-imprecise term with many different interpretations. For example, all existing
-approaches agree that simple tautologies are ``uninteresting'', but differ when
-it comes to more complex statements.
-
-A survey of systems for theory formation and exploration, and their notions of
-``interestingness'' (for concepts and conjectures), is given by Colton et
-al~\cite{colton2000notion}. Six notions of ``interestingness'' are identified in
-these systems, their use by each is summarised in table \ref{table:colton},
-along with (our interpretation of) their use in three more recent ATE systems.
-These notions, as applied to ATE, are:
-
-\textbf{Empirical plausibility}, which checks whether a property holds across
-some specific examples. This is especially useful for avoiding  false
-conjectures, without resorting to a full proof search.
-
-\textbf{Novelty} is whether a conjecture, or one isomorphic or more general, has
-already been seen.
-
-\textbf{Surprisingness} of a conjecture is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-\textbf{Applicability} depends on the number of models in which a conjecture
-holds. The system of Bagai et al conjectures the \emph{non-existence} of
-objects, and hence favours statements with \emph{zero} applicability. Other
-systems treat applicability as a positive aspect: the more applicable the
-statement, the more interesting it is.
-
-\textbf{Comprehensibility} is the \emph{complexity} of a statement. Simpler
-statements are considered more interesting, and many of the search algorithms
-explore simpler/smaller statements before complex/larger ones, to more
-efficiently find those which are interesting.
-
-\textbf{Utility} is the relevance or usefulness of a conjecture to the user's
-particular task. For example, if we want to find optimising rewrite rules such
-as equation \ref{eq:mapreduce}, then utility would include whether or not a
-conjecture justifies a rewrite rule, the difference in resource usage of the
-expressions involved, and how common those expressions are in real usage.
+The task of discovering theorems in an arbitrary theory can be
+described as the interplay of several processes: (i) formulating new
+definitions and concepts; (ii) finding patterns suitable for posing as
+conjectures; and (iii) proving that those conjectures are theorems.
+The latter is studied extensively in the field of Automated Theorem
+Proving (ATP). Far less attention has been paid to automating the
+first two tasks, which both Automated Theory Formation (ATF) and
+Automated Theory Exploration (ATE) address.  We discuss the
+relationship between ATF and ATE in historical context in
+$\S$\ref{exploration-versus-formation}.
+
+A major challenge for both ATF and ATE is choosing how to narrow down
+the set of generated conjectures to those deemed ``interesting'',
+since this is an imprecise term with many different
+interpretations. For example, all existing approaches agree that
+simple tautologies are ``uninteresting'', but differ when it comes to
+more complex statements.  Colton {\em et al.}  give a survey of
+systems for theory formation and exploration, and their notions of
+``interestingness'' for concepts and conjectures, identifying six
+notions in total~\cite{colton2000notion}. These notions, as applied to ATE, are:
+
+{\bf Empirical plausibility}, which checks whether a property holds
+across some specific examples. This is especially useful for avoiding
+false conjectures, without resorting to a full proof search.
+
+{\bf Novelty} is whether a conjecture, or one isomorphic or more
+general, has already been seen.
+
+{\bf Surprisingness} of a conjecture is whether or not it is
+``obvious'', for example if it is an instance of a tautology.
+
+{\bf Applicability} depends on the number of models in which a
+conjecture holds. The system of Bagai et al conjectures the
+non-existence of objects, and hence favours statements with zero
+applicability. Other systems treat ap- plicability as a positive
+aspect: the more applicable the statement, the more interesting it is.
+
+{\bf Comprehensibility} is the complexity of a statement. Simpler
+statements are considered more interesting, and many of the search
+algorithms explore simpler/smaller statements before complex/larger
+ones, to more efficiently find those which are interesting.
+
+{\bf Utility} is the relevance or usefulness of a conjecture to the
+user's particular task. For example, if we want to find optimising
+rewrite rules such as equation 1, then utility would include whether
+or not a conjecture justifies a rewrite rule, the difference in
+resource usage of the expressions involved, and how common those
+expressions are in real usage.
 
@@
 
-These criteria, especially empirical plausibility and comprehensibility, are not
-only used for analysing results after the fact, but instead may form a core part
-of an algorithm's design decisions. For example, the designers of IsaCoSy
-consider simple substitutions of existing conjectures (i.e. those with low
-novelty) to be uninteresting, and hence their system includes a constraint
-solving component which avoids generating such statements entirely.
-
-With such ambiguous and varied goals, approaches and assessment criteria it is
-difficult to compare ATE systems in a quantitative way, and hence to form some
-measure of ``progress'' for the field.
+We summarise the criteria in Table 1 and state which were used in key
+ATF/ATE systems, where Colton identified usage in AM, GT, Graffiti and
+Bagai et al, and his own ATF system HR, and we identify usage in three
+more recent ATE systems QuickSpec, IsaCoSy and IsaScheme. These
+criteria, especially empirical plausibility and comprehensibility, are
+not only used for analysing results after the fact, but instead may
+form a core part of an algorithm's design decisions. For example, the
+designers of IsaCoSy consider simple substitutions of existing
+conjectures (i.e. those with low novelty) to be uninteresting, and
+hence their system includes a constraint solving component which
+avoids generating such statements entirely.
+
+With such ambiguous and varied goals, approaches and assessment
+criteria it is difficult to compare ATE systems in a quantitative way,
+and hence to form some measure of ``progress'' for the field.
+
+\section{Theory Exploration versus Theory
+  Formation}\label{exploration-versus-formation}
+
+In this section we put the Mathematical Theory Exploration approach
+into context both historically, and with respect to related approaches
+such as in Automated Theory Formation systems.
+
+\subsection{A Short History of Automated Mathematics}
+
+\begin{quote}
+...``in his subsequent design for an Analytical Engine Mr. Babbage has
+shown that material machinery is capable, in theory at least, of
+rivalling the labours of the most practised mathematicians in all
+branches of their science.'' \cite[p. 498]{jevons}
+\end{quote}
+
+The first automated mathematical system, the mechanical calculator
+(known as the Pascaline), was an adding machine that could perform
+additions and subtractions directly and multiplication and divisions
+by repetitions, and was conceived by Pascal in 1642 while reorganising
+tax revenues \cite{d'ocagne}. Subsequent early systems include
+M\"uller's universal calculating machine in 1784, which he built for
+the purpose of calculating and printing numerical tables: he invented
+this when he had to check and recalculate some tables relating to the
+volumes of trees \cite[p. 65]{lindgren}. Thirty seven years later,
+Babbage invented his famous difference engine: an automatic,
+mechanical calculator designed to tabulate polynomial functions. This
+was inspired by a flawed table of logarithms and the idea that
+machines would be quicker and more reliable \cite{bowden}. It is
+interesting to note background and motivation: while Pascal and
+Babbage were mathematicians, with an interest in engineering, M\"uller
+was an engineer with an interest in mathematical knowledge. All three
+systems were conceived as an aid to mathematicians, as well as
+scientists, accountants and surveyors.
+
+Differences in background and motivation continue to be relevant
+today. While the majority of work in automating mathematics has been
+in symbolic manipulation and theorem proving, we are concerned here
+with other aspects of mathematics, including the construction of
+concepts, axioms, conjectures, examples and theorems. This is
+varyingly known as ``Automated Theory Formation''
+\cite{lenat:77,colton:book}, ``Automated Mathematical Theory
+Exploration'', ``Mathematical Theory Exploration''
+\cite{buchberger:06} (also sometimes prefaced with ``Computer-Aided''
+or ``Algorithm-Supported''), ``Automated Mathematical Discovery''
+\cite{epstein:91,colton:interestingness,esarm2008}, ``Concept
+Formation in Discovery Systems'' \cite{haase}, and ``Automated Theorem
+Discovery'' \cite{roy}. Such a plethora of terminology can be
+unhelpful and can mask similarities between the different fields. In
+particular, the twin strands of Automated Theory Formation and
+Automated Mathematical Theory Exploration seem to be developing
+somewhat independently without a clear differentiating
+methodology. Below we discuss commonalities and differences between
+the two schools of thought.
+
+%we limit our discussion to systems which work in maths: there are
+%others which work in scientific discovery which are also relevant.
+
+\subsection{Automated Theory Formation}
+
+% 1. history of terminology
+Automated theory formation, the focus of this paper, derives its
+terminology from psychology in which the term ``concept formation'' is
+used (see, for example, \cite{bruner:67}) to describe the search for
+features which differentiate exemplars from non-exemplars of various
+categories. Lenat used this term in his 1977 paper: {\em Automated
+  Theory Formation in Mathematics} \cite{lenat:77} and we have
+extended it to describe a family of techniques used to construct and
+evaluate concepts, conjectures, examples and proofs (embodied in the
+HR, HRL and GC systems and derivatives, as described above).
+
+% 4. example systems - AM, HR, ...
+When Lenat built the AM system \cite{lenat:77}, there were systems
+which could define new concepts for investigation, such as those
+described in \cite{winston}, and systems which could discover
+relationships among known concepts, such as Meta-Dendral
+\cite{buchanan:75}. No system could perform both of these tasks: Lenat
+saw this as the next step. AM was designed to both construct new
+concepts and conjecture relationships between them; fully automating
+the cycle of discovery in mathematics. Lenat describes this as
+follows:
+
+\begin{quote}
+``What we are describing is a computer program which
+defines new concepts, investigates them, notices
+regularities in the data about them, and conjectures
+relationships between them. This new information is used
+by the program to evaluate the newly-defined concepts,
+concentrate upon the most interesting ones, and iterate the
+entire process.'' \cite[p. 834]{lenat:77}
+\end{quote}
+
+AM was a rule-based system which used a frame-like scheme to represent
+its knowledge, enlarged its knowledge base via a collection of
+heuristic rules, and controlled the firing of these rules via an
+agenda mechanism. Lenat chose elementary arithmetic as the development
+domain because he could use personal introspection for the heuristics
+for constructing and evaluating concepts. Given the age of this
+discipline, Lenat thought it unlikely that AM would make significant
+discoveries, although he did cite its ``ultimate achievements'' as the
+concepts and conjectures it discovered (or could have discovered). He
+suggested various criteria by which his system could be evaluated,
+many of which focused on an exploration of the techniques. For
+instance, he considered generality (running AM in new domains) and how
+finely-tuned various aspects of the program are (the agenda, the
+interaction of the heuristics, etc). %most of which were qualitative.
+Lenat saw his system and future developments in this field as having
+implications for mathematics itself (finding results of significance),
+for automating mathematics research (developing AI techniques), and
+for designing ``scientist assistant'' programs (aids for
+mathematicians). This shows a broad spread of motivation. Despite the
+seeming success of the AM system, it is one of the most criticised
+pieces of AI research. In their case study in methodology, Ritchie and
+Hanna analysed Lenat's written work on AM and found that there was a
+large discrepancy between his theoretical claims and the implemented
+program \cite{partridge}. For instance, Lenat made claims about how AM
+invented natural numbers from sets, whereas it used one heuristic
+which was specifically written in order to make this connection (and
+not used in any other context). Another problem was that the processes
+were sometimes under-explained. For an argument of why many of the
+claims made by Lenat about AM were false, see chapter 13 of
+\cite{colton:book}.
+
+\subsection{Mathematical Theory Exploration}
+
+The phrase Mathematical Theory Exploration (MTE) is a recent term
+which seems to have originated with Buchberger and colleagues (see,
+for example, \cite{buchberger}). His motivation is to support
+mathematicians during their exploration of mathematical theories. This
+support is intended to be for the straightforward reasoning, which he
+argues, covers most mathematical thought, rather than the ingenious
+points, which he leaves for human mathematicians. Buchberger's long
+term goal is to provide routine tools for the exploration activity of
+working mathematicians, to support the invention and structured
+build-up of mathematical knowledge. The Theorema project aims at
+prototyping features of a system for such a purpose. These features
+include Integration of the Functionality of Current Mathematical
+Systems (retention of the full power of current numerics and computer
+algebra systems, as well as enabling the user to add their own
+algorithms to the system); Attractive Syntax (input and output is
+readable and presented attractively, and can be personalised by the
+user); and Structured Mathematical Knowledge Bases (tools are provided
+for building and using large mathematical knowledge
+libraries). Buchberger has evaluated the potential of this strategy by
+illustrating the automated synthesis of his own Gr\"obner bases
+algorithm \cite{buchberger:04}.
+
+Recent systems developed in this area include MATHsAiD \cite{roy},
+IsaScheme \cite{MontanoRivas2011} and IsaCosy \cite{johansson}. The
+goal of the MATHsAiD (Mechanically Ascertaining Theorems from
+Hypotheses, Axioms and Definitions) project is to build a tool which
+takes in a set of axioms, concept definitions and a logic and applies
+its inference rules to reason from the axioms to theorems. The
+motivation is to produce a tool which will help mathematicians to
+explore the consequences of a set of axioms or a particular concept
+definition. One of the main challenges of the project has been to
+automatically evaluate interestingness: to distinguish important
+theorems from results which, although they follow from a set of
+axioms, are of little mathematical interest.
+
+%IsaScheme
+Monta{\~n}o-Rivas {\em et al.} have implemented a scheme-based
+approach to MTE in their IsaScheme system
+\cite{MontanoRivas2011}. Schemes are higher-order formulae which can
+be used to generate new concepts and conjectures; variables within the
+scheme are instantiated automatically and this drives the invention
+process.  For instance, in the theory of natural numbers, given the
+concepts of successor ($suc$), addition ($+$) and zero ($0$),
+IsaScheme can use the following scheme to invent the concept of
+multiplication:
+
+\[
+\text{def-scheme}(g,h,i,j) \equiv\\
+  \exists f. \forall x y. \left\{
+  \begin{array}{l l}
+    f(g,y) = h (y)  \: \text{and}  & \\
+    f(i(x),y) = j(y,f(x,y)) & \\
+  \end{array} \right.
+\]
+where the existentially quantified variable $f$ stands for the new
+function to be defined in terms of the variables g, h, i and j. Within
+the theory of natural numbers, IsaScheme instantiates this scheme with
+$\sigma_1 = \{g \mapsto 0, h \mapsto (\lambda x.0), i \mapsto suc, j
+\mapsto +\}$. In this example, $f \mapsto *$ (multiplication), since:
+
+\begin{align*}
+0*y  &= 0 \\
+suc(x) * y &=y + (x*y).
+\end{align*}
+The new multiplication function $f$ can itself be used to instantiate
+variables in the same scheme, resulting in the invention of the
+exponentiation concept (these examples are taken from
+\cite{MontanoRivas2011}).
+
+%IsaCosy
+The IsaCosy system (Isabelle Conjecture Synthesis) is a program for
+inductive theory formation, which synthesises conjectures from the
+available constants and free variables. Only terms which do not
+contain subterms that can be rewritten by a set of rewrite rules can
+be synthesised.  Conjectures are tested by sending them to a
+counterexample checker and, if no counterexamples are found, then sent
+to IsaPlanner which attempts to prove them.
+
+% Discussion.
+
+\subsection{Commonalities and Differences between MTE and ATF}
+
+We believe that ATF and MTE share many goals and would benefit from a
+closer alignment: both fields are highly specialised and by joining
+forces (while being explicit about differences) the techniques being
+developed would have greater impact. This closer alignment is already
+beginning to take place. The main commonalities between MTE and ATF
+are that both are interested in the same aspects of mathematical
+thinking, aiming to automatically construct and evaluate elements of a
+mathematical theory, including concepts, conjectures, theorems, axioms
+and examples. Both strands contrast themselves with Automated Theorem
+Proving. The main historical differences between these two approaches
+seem to be that in MTE:
+\begin{itemize}
+\item the main proponents define themselves as mathematicians (they
+  hold a PhD and have experience in research mathematics);
+\item the primary motivation is to support mathematicians;
+\item systems tend to be user-interactive;
+\item systems tend to be specific to mathematics,
+\end{itemize}
+
+while in ATF:
+
+\begin{itemize}
+\item the main proponents define themselves as AI researchers (they
+  often have a mathematics background up to Masters level, but their
+  PhD and research experience is in AI);
+\item the primary motivation is to extend AI techniques; a secondary
+  motivation is to produce interesting new mathematics;
+\item systems tend to be fully automated;
+\item systems can often be applied to non-mathematical domains.
+\end{itemize}
+
+The different backgrounds of the people who named the fields
+``automated theory formation'' (Lenat) and ``mathematical theory
+exploration'' (Buchberger) perhaps reflects different metaphysical
+perspectives on invention (in which case new mathematical ideas are
+being {\em formed}, created or produced) and discovery (in which case
+abstract mathematical objects exist independently of us and are {\em
+  explored} or investigated) in mathematics.\footnote{There is a vast
+  literature on the Platonic and the constructivist view of
+  mathematics: we shall not venture down this path here (interested
+  readers are referred to \cite{hersh:97,shapiro}). Instead, for now,
+  we make the pragmatic assumption that while there may well be
+  different cognitive processes involved in invention than those
+  involved in discovery, the fields of ATF and MTE are not currently
+  concerned with this level of detail and so the philosophical
+  distinction is not relevant for our purposes.} As can be seen above,
+some of these historical differences are disappearing, and current
+systems such as MATHsAiD \cite{roy}, IsaScheme \cite{MontanoRivas2011}
+and IsaCosy \cite{johansson} are bridging the methodological gap
+between ATF and MTE. This is particularly true in terms of
+fully-automatic/user-interactive and the domain-specific/general
+aspects.
 
@@
 in the same way as above, whilst non-equations and those containing
-\textt{lambda} functions (which QuickSpec cannot synthesise) were replaced with
+\texttt{lambda} functions (which QuickSpec cannot synthesise) were replaced with
 null values. Elements of these two sets were deemed to match if they were
@@
 
+% methodology behind Simon's paper
+
+% agreement in maths? matt inglis
+
+% other - zeki/two papers with ursula...
+
 \section{Conclusion}

### 2017-11-23 TEBenchmarkPaper/article.tex
Use matplotlib and seaborn to draw graphs
Add caption to time plot

@@
 \begin{figure}
-  \includegraphics[width=8cm]{graphs/time}
-  \includegraphics[width=8cm]{graphs/precRec}
+  \centering
+  \input{time.pgf}
+  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
+    is an individual run, with 30 per size except for 1 (14) and 2 (29). Colour
+    indicates the total number of conjectures produced by each run, with failing
+    processes shown in red. Some points are shifted horizontally to prevent
+    overlaps. Standard box plots are included, although many are obscured due to
+    quartiles near zero or timeout.}
+  \label{figure:quickspec_results}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \input{prec.pgf}
   \caption{}

### 2017-11-23 TEBenchmarkPaper/article.tex
Text additions

@@
 We applied our sampling methodology to the 269 TIP definitions for sample sizes
-between 1 and 20, above which point QuickSpec becomes very slow. We chose 30
-samples for each size, but the smaller sizes included repetitions due to the
-requirement for samples to admit at least one theorem. A timeout of 3 minutes
-was imposed, since preliminary experiments showed that most runs either finish
-before this time, or still do not finish after an hour. The results are shown in
-figure~\ref{figure:quickspec_results}.
+between 1 and 20. We chose 30 samples for each size, but due to the requirement
+for samples to admit at least one theorem, only 14 unique samples were obtained
+for size 1 and 29 for size 2. A timeout of 3 minutes was imposed, since
+preliminary experiments showed that most runs (around $\frac{4}{5}$) either
+finish before this time, or still do not finish after an hour. The results are
+shown in figure~\ref{figure:quickspec_results}.
+
+The only errors we encountered were timeouts, and these occurred for all sample
+sizes. Whilst the median runtimes are mostly in the range of a few seconds, the
+variance increased dramatically towards size 10, and $\frac{1}{4}$ of sample
+size 13 timed out.
+
+Most runs finished quickly and generated few conjectures, with the most being
+131 conjectures from a sample of size 20. Sizes 15 to 20 show a slight but
+consistent reduction in runtimes, which is likely an artefact of our testing
+setup.
 
@@
 
-We are assuming that the ATE system acts in ``batch mode'' (accepting
-definitions, producing conjectures, then halting), rather than running
-continuously or interactively. This holds for all of the systems we have
-analysed
+The key to our benchmarking methodology is the simplifying assumption that
+theorem proving problem sets are a good proxy for desirable ATE output. This
+allows us to side-step the philosophical quagmire of ``interestingness'' to
+produce concrete, measurable values; it is also a core weakness, as there are
+compelling reasons to refute it.
+
+We do not claim that the use of such corpora exactly captures all interesting
+conjectures of their definitions, or that those definitions exactly represent
+all theories we may wish to explore. Rather, we consider our approach to offer a
+pareto-optimal balance between theoretical rigour and experimental practicality,
+at least in the short term.
+
+If ATE matures to the point of providing \emph{novel} insights, even in
+well-studied domains, our assumption breaks down because such results cannot (by
+definition) appear in any existing corpus.
 
@@
 
+% agreement in maths? matt inglis
+
+% other - zeki/two papers with ursula...
+
+\subsection{Technical Difficulties}
+
+Our benchmarking approach only applies to systems which act in ``batch mode'',
+i.e. those which choose to halt after emitting some output. Whilst all of the
+systems we have encountered are of this form, an analogous benchmark may be
+desirable for systems which instead run continuously or interactively (perhaps
+akin to McCarthy's ``advice taker''%FIXME~\cite{}.
+
 % methodology behind Simon's paper
-
-% agreement in maths? matt inglis
-
-% other - zeki/two papers with ursula...
 

### 2017-11-27 TEBenchmarkPaper/article.tex
More text

@@
     overlaps. Standard box plots are included, although many are obscured due to
-    quartiles near zero or timeout.}
-  \label{figure:quickspec_results}
+    quartiles near zero or timeout. Whiskers show points at 1.5 $\times$ the
+    inter-quartile range.}
+  \label{figure:quickspec_runtimes}
 \end{figure}
@@
   \input{prec.pgf}
-  \caption{}
-  \label{figure:quickspec_results}
+  \caption{Precision and recall of all runs (spread horizontally to avoid
+    overlaps). Point colour shows how many conjectures were in the ground truth
+    set for each run. Lines show the mean value (average of ratios) and the
+    shaded region is standard deviation, assuming a binomial distribution.}
+  \label{figure:quickspec_precRec}
 \end{figure}
@@
 for size 1 and 29 for size 2. A timeout of 3 minutes was imposed, since
-preliminary experiments showed that most runs (around $\frac{4}{5}$) either
-finish before this time, or still do not finish after an hour. The results are
-shown in figure~\ref{figure:quickspec_results}.
-
-The only errors we encountered were timeouts, and these occurred for all sample
-sizes. Whilst the median runtimes are mostly in the range of a few seconds, the
-variance increased dramatically towards size 10, and $\frac{1}{4}$ of sample
-size 13 timed out.
+preliminary experiments showed that most runs either finish before this time, or
+still do not finish after an hour. The running times are shown in
+figure~\ref{figure:quickspec_runtimes}.
+
+The only errors we encountered were timeouts (shown in red), and these occurred
+for all sample sizes. Whilst the median runtimes are mostly in the range of a
+few seconds, the variance shows two distinct regions: sizes below 10 are mostly
+finished within 30 seconds, whilst sizes 10 and above have large variance; the
+whole upper quartile of size 13 timed out. There is also an unexpected downwards
+trend towards size 20, and a slight but consistent reduction in minimum runtimes
+from sizes 15 to 20 which is likely an artefact of our testing setup.
 
 Most runs finished quickly and generated few conjectures, with the most being
-131 conjectures from a sample of size 20. Sizes 15 to 20 show a slight but
-consistent reduction in runtimes, which is likely an artefact of our testing
-setup.
-
-Precision and recall were calculated by normalising each equation generated by
-QuickSpec: all variables are identified numerically, in ascending order from
-left to right; equations were also sorted such that the expression on the
-left-hand-side is lexicographically smaller than that on the right.
-
-Theorem statements from TIP were first updated to reference only the 269
-distinct names we explored, and references to constructors and destructors were
-updated to use the $\eta$-expanded versions instead. Equations were normalised
-in the same way as above, whilst non-equations and those containing
-\texttt{lambda} functions (which QuickSpec cannot synthesise) were replaced with
-null values. Elements of these two sets were deemed to match if they were
-syntactically identical.
+131 conjectures from a sample of size 20.
+
+Equations generated by QuickSpec and those from the ground truth were normalised
+by numbering their variables and putting the left- and right-hand-sides in
+lexicographical order. These two sets were then compared syntactically for
+precision and recall, with the results shown in
+figure~\ref{figure:quickspec_precRec}.
+
+We use a binomial distribution to model precision and recall, as if each
+conjecture is a ball drawn from an urn of wanted/unwanted results (in the case
+of precision) or found/missed ground truths (in the case of recall). The mean of
+these distributions estimates the overall proportion of wanted or found
+conjectures, and we use the standard deviation to determine how well this mean
+fits the data.
+
+There is a clear negative trend in precision as the sample size increases,
+indicating that a more aggressive post-processing filter may be useful as we
+explore larger theories. Some small samples reached full precision, but nothing
+above size 10 exceeded 50\%.
+
+Recall shows a slight downward trend but is mostly flat, averaging between 10\%
+and 40\%. This is expected, since adding more definitions to a QuickSpec
+signature will never reduce the conjectures it will find. Those conjectures
+which QuickSpec failed to find include terms which QuickSpec cannot synthesise
+(such as conditional equations, inequalities, terms larger than the search depth
+and those with anonymous functions) as well as equations which are removed as
+``uninteresting'' by the final congruence closure step.
 
@@
 
-If ATE matures to the point of providing \emph{novel} insights, even in
+If ATE matures to the point of providing \emph{truly novel} insights, even in
 well-studied domains, our assumption breaks down because such results cannot (by
@@
 
+A fundamental limitation of our benchmarking approach is that it only applies to
+systems which act in ``batch mode'', i.e. those which choose to halt after
+emitting some output. Whilst all of the systems we have encountered are of this
+form, an analogous benchmark may be desirable for systems which instead run
+continuously or interactively (perhaps akin to McCarthy's ``advice taker'').%FIXME~\cite{}.
+
 % philosophical bits
@@
 
-\subsection{Technical Difficulties}
-
-Our benchmarking approach only applies to systems which act in ``batch mode'',
-i.e. those which choose to halt after emitting some output. Whilst all of the
-systems we have encountered are of this form, an analogous benchmark may be
-desirable for systems which instead run continuously or interactively (perhaps
-akin to McCarthy's ``advice taker''%FIXME~\cite{}.
-
 % methodology behind Simon's paper
@@
 
-Our benchmark suite provides a unifying goal for the diverse approaches of
+Our benchmarking methodology, and accompanying suite based on TIP, provides a
+general way to evaluate and compare many diverse approaches to the problem of
 theory exploration, whilst avoiding some of the philosophical complications of
-the field. Whilst the current implementation is quite limited, we welcome
-additions from other researchers, to more closely align the benchmark with their
-goals, and the strenghts and weaknesses of their systems.
+the field. The methodology can be adapted to suit particular domains, for
+example our experiments have focused on inductive theories with higher-order
+functions, and the corpora used to generate a benchmark suite can always be
+improved and refined with the addition of new definitions and theorems.
+
+We believe that a standard approach to benchmarking and comparison will ease the
+burden on researchers wanting to evaluate systems, and provide a common goal to
+pursue in the short term.
 
@@
 exploration in general, so more ambitious goals must be set in the future; but
-we have found that there is still a long way to go until that problem arises.
+we believe that our approach will provide a compelling challenge, at least for
+the foreseeable future.
 

### 2017-11-30 TEBenchmarkPaper/article.tex
Add Jianguo as co-author
Fix typo in code example

@@
 \author{Chris Warburton \and
-        Alison Pease
+        Alison Pease    \and
+        Jianguo Zhang
         % TODO: Invite Katya? Even if she only offers some help, can acknowledge
@@
            \email{a.pease@dundee.ac.uk}
+           \and
+           J. Zhang \at
+           University of Dundee \\
+           \email{j.n.zhang@dundee.ac.uk}
 }
@@
       \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), xs)                  \\
+      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
       \Foldr(f, \Nil,         y) &= y                                       \\
@@
   \caption{A simple theory defining a $\List$ type and some associated
-    operations. $\Zero$ and $\Succ$ are from a Peano encoding of the
-    natural numbers. Taken from
-    \cite{Johansson.Dixon.Bundy:conjecture-generation}}
+    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
+    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
   \label{figure:list_theory}

### 2017-12-01 TEBenchmarkPaper/article.tex
Remove IsaCoSy contribution
Made a bunch of notes on TEBenchmark

@@
 \item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy ATE systems,
-  and a discussion of the results.
+\item Application of our methodology to the QuickSpec ATE system, and a
+  discussion of the results.
 \end{itemize}
@@
 % 1. history of terminology
-Automated theory formation, the focus of this paper, derives its
+FIXME Automated theory formation, the focus of this paper, derives its
 terminology from psychology in which the term ``concept formation'' is
@@
 Another advantage to our statistical approach is that we can compute the
-\emph{variance} of our data, which is unknown in the previous one-off
+\emph{spread} of our data, which is unknown in the previous one-off
 evaluations. In particular, we might be concerned that such previous evaluations

### 2017-12-01 Bucketing/contributions.tex
Rename bucketing paper

--- 2016-03-02 CICM2016/contributions.tex
@@
 
-We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, and suggest a new recurrent clustering and feature extraction algorithm for Haskell, which we then evaluate as a relevance filter technique for theory exploration.
-As a clustering algorithm, the aim of recurrent clustering is to identify similarities in a set of data points (in our case, Core expressions).
-Rather than clustering expressions directly, we first transform them into numeric vectors in a process called \emph{feature extraction}.
-The distinguishing characteristic of recurrent clustering is to \emph{combine} feature extraction and clustering into a single recursive algorithm (shown as Algorithm \ref{alg:recurrent}), which allows the feature vector of an expression to depend on those of the expressions it references.
-Here we describe our approach to recurrent clustering and compare its similarity and differences to those of ML4PG and ACL2(ml).
+Rather than clustering Core expressions directly, we first perform \emph{feature extraction} to transform them into numeric \emph{feature vectors}.
+We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature extraction and clustering into a single recursive algorithm, such that the feature vector of an expression depends on the feature vectors of those expressions it references.
+We suggest a new recurrent clustering and feature extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity and differences to those of ML4PG and ACL2(ml).
 
-We consider our algorithm in two stages: the first transforms the nested structure of expressions into a flat vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the function $\phi$.
+We describe our algorithm in two stages: the first transforms the nested structure of expressions into a flat vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the function $\phi$.
 
@@
 This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
-Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the $toTree$ function would ignore the constructor identifier $\id{C}$ and never produce the feature $\feature{\id{C}}$.
-However, this example is still accurate in terms of laying out the features as given.}
+Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the constructor identifier $\id{C}$ would not appear in the vector, but this example is still accurate in terms of laying out the features as given.}
 
@@
 
-To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the function $toTree$ given in Figure \ref{fig:totree}.
-Intuitively, these rose trees are simply s-expression representations of the ASTs, illustrated for the \hs{odd} function in Figure \ref{fig:rosetreeexample}.
+To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features.
+Intuitively, these are simply an s-expression representation of the Core syntax, with the feature extraction function $\phi$ mapped over the labels.
+The full transformation is given in Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the result for the \hs{odd} function.
 
@@
 \begin{sloppypar}
-Finally, matrices are turned into vectors by simply concatenating the rows from top to bottom, hence the matrix from Figure \ref{fig:matrixexample} will begin $(\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, \dots$.
+Finally, matrices are turned into vectors by concatenating the rows from top to bottom, hence Figure \ref{fig:matrixexample} will produce a vector beginning $(\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, \dots$.
 \end{sloppypar}
@@
 
-Since numerals, strings and constructor identifiers are discarded during conversion to rose trees, the only remaining case is global identifiers $\mathcal{G}$.
-Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as we did for local identifiers.
+We discard the contents of literals and constructor identifiers when converting to rose trees, so the only remaining case is global identifiers $\mathcal{G}$.
+Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as for local identifiers.
 We also cannot directly encode the form of the identifiers, e.g.
@@
 \fi
-
-\subsubsection{Comparison}
-
-Our algorithm is most similar to that of ML4PG, as our transformation maps the elements of a syntax tree to distinct cells in a matrix.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.
-
-There are two major differences between our algorithm and that of ML4PG: mutual-recursion and types.
-
-The special handling required for mutual recursion is discussed above (namely, topological sorting of expressions and the $f_{recursion}$ feature).
-Such handling is not present in ML4PG, since the Coq code it analyses must, by virtue of the language, be written in dependency order to begin with.
-Coq \emph{does} have limited support for mutually-recursive functions, of the following form:
-
-\begin{verbatim}
-Fixpoint even n := match n with
-                       | O   => true
-                       | S m => odd m
-                   end
-    with odd  n := match n with
-                       | O   => false
-                       | S m => even m
-                   end.
-\end{verbatim}
-
-However, this is relatively uncommon and unsupported by ML4PG.
-
-The more interesting differences come from our handling (or lack thereof) for types.
-Coq and ACL2 are at opposite ends of the typing spectrum, with the former treating types as first class entities of the language whilst the latter is untyped (or \emph{unityped}).
-In both cases, we have a \emph{single} language to analyse, by ML4PG and ACL2(ml) respectively.\footnote{ML4PG can also analyse Coq's \textsc{Ltac} meta-language.
-Haskell has its own meta-language, Template Haskell, but here we only consider the regular Haskell which it generates.}
-
-The situation is different for Haskell, where the type level is distinct from the value level, and there are strict rules for how they can influence each other.
-In particular, Haskell values can depend on types (via the type class mechanism) but types cannot depend on values.
-
-In our initial approach, we restrict ourselves to the value level. This has several consequences:
-
-\begin{itemize}
-  \item Although they are values, we cannot distinguish between data constructors, other than using exact equality. Hence they are discarded by $toTree$.
-  \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot always distinguish between them.
-This can cause a function's Core arity to be greater than its Haskell arity.
-\end{itemize}
-
-There is certainly promise in including types in our analysis, by pairing every term with its type as in ML4PG.
-This will allow fine-grained distinction of expressions which are otherwise identical, especially data constructors.
-Integrating types into our algorithm, and extracting them from Core expressions, is hence left as future work.

### 2017-12-13 PowerPlay/article.tex
Basic start of PowerPlay paper

Automation, Learning and Interestingness in Pure Mathematics

Humans have been practicing and researching mathematics, in one form or another,
for thousands of years. The invention of electronic computers has made it
routine to automate \emph{calculations}, but the autoamtion of more ``creative''
aspects of mathematics, such as theorem proving, conjecture generation and
concept formation, is still mostly limited to academic research rather than
being a routine task for computers.

We propose to address this by automating the \emph{discovery} and
\emph{learning} of mathematical concepts. We adapt the approach of Schmidhuber
et al., which has previously been used with artificial neural networks to learn
classifiers, and adapt it to a theorem proving setting.

%% System description

The PowerPlay architecture is based around inventing and solving problems. The
encoding and interpretation of these problems depends on the implementation, so
for now we leave it abstract by declaring it as a type:

%% FIXME: Problem is a type

We also keep solutions abstract, but we parameterise them by the particular
problem they're meant to solve:

%% FIXME: Solution is a function

Given the above, a problem solver is a function from problems to their
solutions, with an option type to represent failure:

%% FIXME: Solver

### 2018-01-15 TEBenchmarkPaper/article.tex
Fleshing out table more
More table justification

@@
 \newcommand{\tIMP}{implies}
+\newcommand{\tEQ}{equations}
 \newcommand{\tRow}[1]{#1 \\ \hline}
@@
     \tRow{HR          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
-    \tRow{QuickSpec   & equalities         &   X &   X &     &     &   X &    }
-    \tRow{IsaCoSy     & equalities         &     &   X &     &     &   X &    }
-    \tRow{IsaScheme   & equalities         &     &     &     &     &   X &    }
+    \tRow{QuickSpec   & \tEQ               &   X &   X &     &     &   X &    }
+    \tRow{Speculate   & conditional \tEQ   &   X &   X &     &   A &   X &    }
+    \tRow{IsaCoSy     & \tEQ               &   X &   X &     &     &   X &   X}
+    \tRow{IsaScheme   & \tEQ               &   X &   X &     &     &   X &   X}
   \end{tabular}
@@
 
+% JUSTIFICATIONS
+%
+% IsaCoSy uses counter-example checking; this ensures empirical plausibility
+% IsaCoSy uses constraints to avoid special-cases; this ensures novelty
+% IsaCoSy uses utility, since there are some hard-coded patterns which are
+% looked for
+%
+% Speculate uses a form of unification to ensure novelty, making sure one
+% equation is not a special-case of another
+% Speculate uses LeanCheck to enumerate values, looking for counterexamples
+% Speculate uses testing, to ensure Empirical Plausibility
+%
+% QuickSpec uses QuickCheck to ensure empirical plausibility
+% QuickSpec uses a congruence closure algorithm to ensure novelty
+%
+% IsaScheme uses utility, since it determines the "quality" of a definition
+% based on how many theorems it appears in?
+% IsaScheme uses novelty, using an equational rewrite system (Knuth-Bendix
+% completion) to remove redundancies
+% IsaScheme uses utility, since it focuses on the terms and definitions of a
+% user's theory. However, don't all of them?
+% IsaScheme uses counterexample checking for empirical plausibility.
+
+% ONLY APPLIES TO PRECONDITIONS
+% APPLICABILITY: ensure a conjecture holds in many models (or exactly zero, in
+% the case of Bagai et al). Does the testing-based approach of QuickSpec and
+% Speculate ensure that there's a model? After all, these are concrete values
+% rather than e.g. implications of some abstract specification.
+
 We summarise the criteria in Table 1 and state which were used in key
-ATF/ATE systems, where Colton identified usage in AM, GT, Graffiti and
-Bagai et al, and his own ATF system HR, and we identify usage in three
-more recent ATE systems QuickSpec, IsaCoSy and IsaScheme. These
-criteria, especially empirical plausibility and comprehensibility, are
-not only used for analysing results after the fact, but instead may
-form a core part of an algorithm's design decisions. For example, the
-designers of IsaCoSy consider simple substitutions of existing
-conjectures (i.e. those with low novelty) to be uninteresting, and
-hence their system includes a constraint solving component which
-avoids generating such statements entirely.
-
-With such ambiguous and varied goals, approaches and assessment
-criteria it is difficult to compare ATE systems in a quantitative way,
-and hence to form some measure of ``progress'' for the field.
+ATF/ATE systems. The entries for AM, GT, Graffiti, Bagai et al. and HR are based
+on this survey (the latter is Colton's own system). We extend this analysis to
+some more recent systems in the remaining rows (based on our understanding of
+their function): QuickSpec, Speculate, IsaCoSy and IsaScheme.
+
+All four of these systems use counterexample checking and a small-to-large
+search order, favouring empirical plausibile and comprehensibile output. IsaCoSy
+ensures novelty using a constraint system, such that no generated expression can
+be simplified using a previously-found equation. The other systems use post-hoc
+filters based on unification and rewriting (based on the Knuth-Bendix completion
+algorithm) to discard expressions or equations which are special-cases of some
+other, more general result.
+
+IsaCoSy is able to look for commonly desirable patterns, such as commutativity
+and associativity, which increases the utility of its output, whilst IsaScheme
+is completely based around such pattern-instantiation.
+
+This diversity of approaches makes it difficult to compare the existing
+evaluations of these systems directly. In particular, evaluation methods created
+for one system might not make sense for another, due to assumptions made about
+the algorithms' operation. For example, the novelty filters used by QuickSpec
+and Speculate are applied to the whole output set, guaranteeing that no
+special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
+since its constraint solver allows special cases if they're found \emph{before}
+the general case (it does not go back and discard previous results).
+
+With such ambiguous and varied goals, approaches and assessment criteria it is
+difficult to compare ATE systems in a quantitative way, and hence to form some
+measure of ``progress'' for the field. Our proposed benchmarking methodology
+attempts to remedy this by setting a clear target, hopefully in line with the
+myriad goals of existing researchers, whilst being generally applicable to
+systems of this type.
 

### 2018-01-15 TEBenchmarkPaper/article.tex
Remove out-of-context remarks

@@
 % 1. history of terminology
-FIXME Automated theory formation, the focus of this paper, derives its
-terminology from psychology in which the term ``concept formation'' is
-used (see, for example, \cite{bruner:67}) to describe the search for
-features which differentiate exemplars from non-exemplars of various
-categories. Lenat used this term in his 1977 paper: {\em Automated
-  Theory Formation in Mathematics} \cite{lenat:77} and we have
-extended it to describe a family of techniques used to construct and
-evaluate concepts, conjectures, examples and proofs (embodied in the
-HR, HRL and GC systems and derivatives, as described above).
+Automated theory formation derives its terminology from psychology in which the
+term ``concept formation'' is used (see, for example, \cite{bruner:67}) to
+describe the search for features which differentiate exemplars from
+non-exemplars of various categories. Lenat used this term in his 1977 paper:
+{\em Automated Theory Formation in Mathematics} \cite{lenat:77}.
 

### 2018-01-15 TEBenchmarkPaper/article.tex
Don't talk about Isabelle if/until we include IsaCoSy

@@
 the Isabelle theorem prover, with one involving the natural numbers, and the
-other involving the list functions from figure \ref{figure:list_theory}. The library authors have gone to the effort of stating,
-proving and including these theorems in every copy of their software, which is a
-good indication that they are useful or important.
+other involving the list functions from figure \ref{figure:list_theory}. The
+library authors have gone to the effort of stating, proving and including these
+theorems in every copy of their software, which is a good indication that they
+are useful or important.
 
@@
 \item Definitions should be \emph{strongly typed}, such that they can be
-  translated into the native language of each ATE system (in this work, those
-  languages are Haskell and Isabelle).
+  translated into the native language of each ATE system (in this work, that
+  includes Haskell\iffalse FIXME and Isabelle\fi).
 \item The problem set should be relevant to the desired domain. In our case, we

### 2018-01-15 TEBenchmarkPaper/article.tex
More table justification

@@
 \newcommand{\tEQ}{equations}
+\newcommand{\tINE}{inequalities}
+\newcommand{\tCON}{conditional}
 \newcommand{\tRow}[1]{#1 \\ \hline}
@@
 non-existence of objects, and hence favours statements with zero
-applicability. Other systems treat ap- plicability as a positive
+applicability. Other systems treat applicability as a positive
 aspect: the more applicable the statement, the more interesting it is.
@@
     \tRow{GT          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
-    \tRow{Graffiti    & inequalities       &   X &   X &   X &     &   X &   X}
+    \tRow{Graffiti    & \tINE              &   X &   X &   X &     &   X &   X}
     \tRow{Bagai et al & \tNE               &     &   X &     &   X &   X &    }
     \tRow{HR          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
-    \tRow{QuickSpec   & \tEQ               &   X &   X &     &     &   X &    }
-    \tRow{Speculate   & conditional \tEQ   &   X &   X &     &   A &   X &    }
-    \tRow{IsaCoSy     & \tEQ               &   X &   X &     &     &   X &   X}
-    \tRow{IsaScheme   & \tEQ               &   X &   X &     &     &   X &   X}
+    \tRow{QuickSpec   & \tEQ               &   X &   X &     &   X &   X &    }
+    \tRow{Speculate   & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
+    \tRow{IsaCoSy     & \tEQ               &   X &   X &     &   X &   X &   X}
+    \tRow{IsaScheme   & \tEQ               &   X &   X &     &   X &   X &   X}
   \end{tabular}
@@
 some more recent systems in the remaining rows (based on our understanding of
-their function): QuickSpec, Speculate, IsaCoSy and IsaScheme.
-
-All four of these systems use counterexample checking and a small-to-large
-search order, favouring empirical plausibile and comprehensibile output. IsaCoSy
-ensures novelty using a constraint system, such that no generated expression can
-be simplified using a previously-found equation. The other systems use post-hoc
-filters based on unification and rewriting (based on the Knuth-Bendix completion
-algorithm) to discard expressions or equations which are special-cases of some
-other, more general result.
+their function): QuickSpec, Speculate~\cite{braquehais2017speculate}, IsaCoSy
+and IsaScheme.
+
+All four of these systems use testing to check for counterexamples, which
+ensures results are empirically plausible and applicable (conditions are
+satisfiable, types are inhabited, etc.). They also use a small-to-large search
+order, ensuring more easily-comprehensibile conjectures are explored first.
+
+IsaCoSy ensures novelty using a constraint system, such that no generated
+expression can be simplified using a previously-found equation. The other
+systems use post-hoc filters based on unification and rewriting rules
+(derived via the Knuth-Bendix completion algorithm) to discard expressions or
+equations which are special-cases of some other, more general result.
 

### 2018-01-15 TEBenchmarkPaper/article.tex
Cite Advice Taker

@@
 form, an analogous benchmark may be desirable for systems which instead run
-continuously or interactively (perhaps akin to McCarthy's ``advice taker'').%FIXME~\cite{}.
+continuously or interactively (perhaps akin to McCarthy's
+``advice taker''~\cite{McCarthy_Programs59}).
 

### 2018-01-15 TEBenchmarkPaper/article.tex
Include Alison's remarks in discussion

@@
 
-We do not claim that the use of such corpora exactly captures all interesting
-conjectures of their definitions, or that those definitions exactly represent
-all theories we may wish to explore. Rather, we consider our approach to offer a
-pareto-optimal balance between theoretical rigour and experimental practicality,
-at least in the short term.
-
-If ATE matures to the point of providing \emph{truly novel} insights, even in
-well-studied domains, our assumption breaks down because such results cannot (by
-definition) appear in any existing corpus.
-
-A fundamental limitation of our benchmarking approach is that it only applies to
-systems which act in ``batch mode'', i.e. those which choose to halt after
-emitting some output. Whilst all of the systems we have encountered are of this
-form, an analogous benchmark may be desirable for systems which instead run
-continuously or interactively (perhaps akin to McCarthy's
+Fundamentally, any mathematical reasoning system must decide on, and formalise,
+what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
+``provable'' include trivial tautologies, while at the same time failing to
+capture the ``almost true'', which can be a valuable trigger for theory change,
+as demonstrated by Lakatos in his case studies of mathematical development
+\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly proposed
+metric. Methods for finding out what that means have come from neuro-scientists,
+such as the study by Zeki et al in which they test whether mathematicians'
+experiences of abstract beauty correlates with the same brain activity as
+experiences of sensory beauty \cite{zeki}; computer scientists, such as Colton's
+metrics above which are proposed based largely on a computer scientist's
+``intuition'' and argumentation about why a metric would be important (and
+previous use of such metrics, and - in one isolated case (surprisingingness) -
+on a single quote from a mathematician\footnote{Similarly, classification of a
+MATHsAiD result as either interesting or uninteresting \cite{roy} was made by
+the main system developer: in both cases the main developer had a background in
+mathematics.}; and mathematicians themselves, such as Gowers' suggestion that we
+can identify features which are commonly associated with good proofs
+\cite{gowers}. All of these approaches rest upon the assumption that it makes
+sense to speak of ``the good'' in mathematics. This assumption is supported by
+concepts such as Erdos's ``The Book'' -- a theoretical construct of a book which
+contains the most elegant proof of every theorem (brought to life in
+\cite{Martin Aigner and Günter M. Ziegler Proofs from THE BOOK}). However,
+empirical studies by psychologists such as Inglis call into question such
+assumptions: work by Inglis and colleagues has shown that there is not a single
+standard of validity among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
+nor beauty. %FIXME needs citation
+
+We do not claim that our use of corpora as a ground truth exactly captures all
+interesting conjectures of their definitions, or that those definitions exactly
+represent all theories we may wish to explore. Rather, we consider our approach
+to offer a pareto-optimal balance between theoretical rigour and experimental
+practicality, at least in the short term. Futhermore, since research is already
+on-going in these areas, we hope to at least improve on existing evaluation
+practices and offer a common ground for future research.
+
+Notably, if ATE matures to the point of providing \emph{truly novel} insights,
+even in well-studied domains, then the assumptions we rely on break down because
+such results cannot (by definition) appear in any existing corpus.
+
+Another practical limitation limitation of our benchmarking approach is that it
+only applies to systems which act in ``batch mode'', i.e. those which choose to
+halt after emitting some output. Whilst all of the systems we have encountered
+are of this form, an analogous benchmark may be desirable for systems which
+instead run continuously or interactively (perhaps akin to McCarthy's
 ``advice taker''~\cite{McCarthy_Programs59}).

### 2018-02-06 TEBenchmarkPaper/article.tex
Fix citation
Use Mathematical Theory Exploration
Reword, since we're not investigating e.g. concept formation

@@
 
-\emph{Automated theory exploration} (ATE), also known as
+\emph{Mathematical theory exploration} (MTE), also known as
 \emph{conjecture generation/formation}, is the open-ended problem of producing
@@
 in symbolic manipulation and theorem proving, we are concerned here
-with other aspects of mathematics, including the construction of
-concepts, axioms, conjectures, examples and theorems. This is
-varyingly known as ``Automated Theory Formation''
-\cite{lenat:77,colton:book}, ``Automated Mathematical Theory
-Exploration'', ``Mathematical Theory Exploration''
-\cite{buchberger:06} (also sometimes prefaced with ``Computer-Aided''
-or ``Algorithm-Supported''), ``Automated Mathematical Discovery''
-\cite{epstein:91,colton:interestingness,esarm2008}, ``Concept
-Formation in Discovery Systems'' \cite{haase}, and ``Automated Theorem
-Discovery'' \cite{roy}. Such a plethora of terminology can be
-unhelpful and can mask similarities between the different fields. In
+with other aspects of mathematics, including the construction of conjectures,
+(counter-)examples and theorems. These tasks, along with the formation of new
+definitions and concepts, are varyingly known as
+``Automated Theory Formation''~\cite{lenat:77,colton:book},
+``Automated Mathematical Theory Exploration'',
+``Mathematical Theory Exploration''~\cite{buchberger:06} (also sometimes
+prefaced with ``Computer-Aided'' or ``Algorithm-Supported''),
+``Automated Mathematical Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
+``Concept Formation in Discovery Systems''~\cite{haase}, and
+``Automated Theorem Discovery''~\cite{roy}. Such a plethora of terminology can
+be unhelpful and can mask similarities between the different fields. In
 particular, the twin strands of Automated Theory Formation and
@@
 contains the most elegant proof of every theorem (brought to life in
-\cite{Martin Aigner and Günter M. Ziegler Proofs from THE BOOK}). However,
+\cite{aigner2010proofs}). However,
 empirical studies by psychologists such as Inglis call into question such

### 2018-02-06 TEBenchmarkPaper/article.tex
Flesh out a comment

@@
 
-%we limit our discussion to systems which work in maths: there are
-%others which work in scientific discovery which are also relevant.
+We limit our discussion to applications in mathematics, although there are
+similar approaches being applied to the task of \emph{scientific} discovery.
+Scientific knowledge relies on inductive reasoning and experimental testing,
+which are not so important to a mathematical theory. However, these techniques
+do play an important role in many of the tools we investigate: in particular,
+inductive reasoning allows a system to conjecture statements which may be too
+difficult for it to deductively prove; whilst experimental testing, e.g. by
+checking universally-quantified statements against some specific values, allows
+many falsehoods to be quickly discarded without requiring more elaborate
+symbolic reasoning.
 

### 2018-02-26 TEBenchmarkPaper/article.tex
ATE -> MTE

@@
   We propose a benchmark suite for evaluating the efficiency and effectiveness
-  of automated tools for \emph{theory exploration} or
-  \emph{conjecture formation} in higher-order, inductive theories; a domain
-  especially suited for analysing software. By providing standard tools and
-  metrics, we hope to encourage innovation and comparison between the disparate
-  approaches currently being pursued, and spur improvements similar to those
-  seen in the competitive field of automated theorem proving.
+  of \emph{conjecture formation} by automated tools for \emph{mathematical
+    theory exploration} in higher-order, inductive theories; a domain especially
+  suited for analysing software. By providing standard tools and metrics, we
+  hope to encourage innovation and comparison between the disparate approaches
+  currently being pursued, and spur improvements similar to those seen in the
+  competitive field of automated theorem proving.
 %\keywords{First keyword \and Second keyword \and More}
@@
 
-\emph{Mathematical theory exploration} (MTE), also known as
-\emph{conjecture generation/formation}, is the open-ended problem of producing
-conjectures about a given logical theory which are somehow ``interesting''.
+\emph{Conjecture generation/formation}, a sub-field of \emph{mathematical theory
+  exploration} (MTE), is the open-ended problem of producing conjectures about a
+given logical theory which are somehow ``interesting''.
 
 This has applications wherever we find use for formal statements: in proof
-assistants and their libraries, in mathematics education and research, and
-(of special concern for the authors) in the specification, verification,
-optimisation and testing of software.
+assistants and their libraries, in mathematics education and research, and in
+the specification, verification, optimisation and testing of software.
 
@@
 defining a standard, unambiguous benchmarking approach with which to compare
-ATE systems. Our contributions include:
+the conjecture generation of MTE systems. Our contributions include:
 
 \begin{itemize}
-\item A general methodology for benchmarking ATE systems.
+\item A general methodology for benchmarking conjecture generation.
 \item Resolving the issue of ``interestingness'' through the use of
@@
 \item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec ATE system, and a
+\item Application of our methodology to the QuickSpec MTE system, and a
   discussion of the results.
@@
 
-Section \ref{sec:background} describes the ATE problem in more detail, along
-with existing approaches and their evaluation methodologies. We explain our
-proposal for a more general benchmark in section \ref{sec:proposal} and
-section \ref{sec:application} shows the results when applied to existing ATE
-tools. Analysis of these results is given in section \ref{sec:discussion} and
-concluding remarks in \ref{sec:conclusion}.
+Section \ref{sec:background} describes MTE and the conjecture generation problem
+in more detail, along with existing approaches and their evaluation
+methodologies. We explain our proposal for a more general benchmark in section
+\ref{sec:proposal} and section \ref{sec:application} shows the results when
+applied to existing MTE tools. Analysis of these results is given in section
+\ref{sec:discussion} and concluding remarks in \ref{sec:conclusion}.
 
@@
 first two tasks, which both Automated Theory Formation (ATF) and
-Automated Theory Exploration (ATE) address.  We discuss the
-relationship between ATF and ATE in historical context in
+Mathematical Theory Exploration (MTE) address.  We discuss the relationship
+between ATF and ATE in historical context in
 $\S$\ref{exploration-versus-formation}.
 
-A major challenge for both ATF and ATE is choosing how to narrow down
-the set of generated conjectures to those deemed ``interesting'',
-since this is an imprecise term with many different
-interpretations. For example, all existing approaches agree that
-simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton {\em et al.}  give a survey of
+A major challenge for both ATF and MTE is choosing how to narrow down
+the set of generated conjectures to those deemed ``interesting'', since this is
+an imprecise term with many different interpretations. For example, all existing
+approaches agree that simple tautologies are ``uninteresting'', but differ when
+it comes to more complex statements.  Colton {\em et al.} give a survey of
 systems for theory formation and exploration, and their notions of
-``interestingness'' for concepts and conjectures, identifying six
-notions in total~\cite{colton2000notion}. These notions, as applied to ATE, are:
+``interestingness'' for concepts and conjectures, identifying six notions in
+total~\cite{colton2000notion}. These notions, as applied to conjectures, are:
 
@@
   \end{tabular}
-  \caption{Classification of ATE systems from \cite{colton2000notion}, extended
+  \caption{Classification of MTE systems from \cite{colton2000notion}, extended
     to those compared in \cite{claessen2013automating} (QuickSpec is the
@@
 We summarise the criteria in Table 1 and state which were used in key
-ATF/ATE systems. The entries for AM, GT, Graffiti, Bagai et al. and HR are based
+ATF/MTE systems. The entries for AM, GT, Graffiti, Bagai et al. and HR are based
 on this survey (the latter is Colton's own system). We extend this analysis to
@@
 With such ambiguous and varied goals, approaches and assessment criteria it is
-difficult to compare ATE systems in a quantitative way, and hence to form some
+difficult to compare MTE systems in a quantitative way, and hence to form some
 measure of ``progress'' for the field. Our proposed benchmarking methodology
@@
 ``Automated Theory Formation''~\cite{lenat:77,colton:book},
-``Automated Mathematical Theory Exploration'',
 ``Mathematical Theory Exploration''~\cite{buchberger:06} (also sometimes
-prefaced with ``Computer-Aided'' or ``Algorithm-Supported''),
+prefaced with ``Computer-Aided'', ``Automated'' or ``Algorithm-Supported''),
 ``Automated Mathematical Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
@@
 
-There are three aspects to consider when we evaluate an ATE system:
+There are three aspects to consider when we evaluate an MTE system:
 
@@
 \emph{ground truth} is chosen as the 'ideal' set of conjectures which we would
-like an ATE system to produce from these definitions.
+like an MTE system to produce from these definitions.
 
 Each system is run, and their output compared to the ground truth. To score
-100\% on precision and recall, an ATE system must output all of the conjectures
+100\% on precision and recall, an MTE system must output all of the conjectures
 which appear in the ground truth, and nothing else:
@@
 corpus (an order of magnitude larger than previous work) and a scalable,
-statistical approach to evaluating ATE systems using this corpus. We follow a
+statistical approach to evaluating MTE systems using this corpus. We follow a
 precision/recall approach similar to the work described above, with the main
@@
 \item Definitions should be \emph{strongly typed}, such that they can be
-  translated into the native language of each ATE system (in this work, that
+  translated into the native language of each MTE system (in this work, that
   includes Haskell\iffalse FIXME and Isabelle\fi).
@@
 referenced by the theorems from the theorem statements themselves. The former
-will be used as input to the ATE systems under evaluation, whilst the latter
-form a ground truth corpus against which to compare the ATE output.
+will be used as input to the MTE systems under evaluation, whilst the latter
+form a ground truth corpus against which to compare the MTE output.
 
@@
 \texttt{add} function, where the definitions of \texttt{plus} and \texttt{add}
-are $\alpha$-equivalent. We would expect an ATE system to conjecture
+are $\alpha$-equivalent. We would expect an MTE system to conjecture
 commutativity and associativity for \emph{both} functions, or for \emph{neither}
@@
 
-We could, in theory, send these de-duplicated definitions straight into an ATE
+We could, in theory, send these de-duplicated definitions straight into an MTE
 system, and use the updated theorem statements as the ground truth for analysis.
@@
   infer performance \emph{in general}.
-\item It is impractical to run any existing ATE system on ``large'' inputs,
+\item It is impractical to run any existing MTE system on ``large'' inputs,
   containing more than around 10 definitions.
@@
 definitions. Given a sample size, we choose a subset of that many definitions,
-and use that as our ATE input. We generate a corresponding ground truth by
+and use that as our MTE input. We generate a corresponding ground truth by
 selecting those theorems from the corpus which only contain references to
@@
 most samples to have an empty ground truth, and hence $0$ precision and
-undefined recall \emph{independent} of what the ATE system produces. This is
+undefined recall \emph{independent} of what the MTE system produces. This is
 clearly undesirable as an evaluation method.
@@
 Given a sample of definitions and a corresponding ground truth, the actual
-execution of the ATE system proceeds as in prior work. We must translate the
+execution of the MTE system proceeds as in prior work. We must translate the
 chosen definitions into the particular input format of the tool under study,
@@
 sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of an ATE system, or \emph{comparing} the performance of
-\emph{two} ATE systems.
+the performance of an MTE system, or \emph{comparing} the performance of
+\emph{two} MTE systems.
 
@@
 To encourage competition in the field, it is important to compare the relative
-performance of different ATE systems on the same task. Since the aggregate
+performance of different MTE systems on the same task. Since the aggregate
 statistics in the above summaries do not include details of specific runs, any
@@
            in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run ATE system};
+    \node (run) [process, below=of runPos]{Run MTE system};
     \node (pr)  [process, below=of run]{Analysis};
@@
 % All together, TIP provides 219 distinct function definitions and 343
-%   theorem statements, which is enough to fully exercise current ATE systems.
+%   theorem statements, which is enough to fully exercise current MTE systems.
 % Benchmark problems include examples from the theorem proving literature
@@
 %   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
-%   (which are used by ATE systems).
+%   (which are used by MTE systems).
 
@@
 % corpus; the latter prevents tailoring the corpus to influence the choices,
-% although ATE systems could .
+% although MTE systems could .
 
@@
 The key to our benchmarking methodology is the simplifying assumption that
-theorem proving problem sets are a good proxy for desirable ATE output. This
+theorem proving problem sets are a good proxy for desirable MTE output. This
 allows us to side-step the philosophical quagmire of ``interestingness'' to
@@
 
-Notably, if ATE matures to the point of providing \emph{truly novel} insights,
+Notably, if MTE matures to the point of providing \emph{truly novel} insights,
 even in well-studied domains, then the assumptions we rely on break down because

### 2018-02-26 TEBenchmarkPaper/article.tex
Minor fixes

@@
 
-This justifies a rewrite rule for splitting a single $\Map$ call into multiple
-independent calls dealing with different sections of a list. Such rules are
-useful optimisations since each call can be evaluated in parallel, leading to
-the ``map/reduce'' programming paradigm.
+Equation \ref{eq:mapreduce} justifies a rewrite rule for splitting a single
+$\Map$ call into multiple independent calls dealing with different sections of a
+list. Such rules are useful optimisations since each call can be evaluated in
+parallel, leading to the ``map/reduce'' programming paradigm.
 
@@
 
-{\bf Comprehensibility} is the complexity of a statement. Simpler
-statements are considered more interesting, and many of the search
-algorithms explore simpler/smaller statements before complex/larger
-ones, to more efficiently find those which are interesting.
+{\bf Comprehensibility} depends on the complexity of a statement. Simpler
+statements are considered more interesting, hence many of the systems explore
+simpler/smaller statements before complex/larger ones, to more efficiently find
+those which are interesting.
 
@@
 
-All four of these systems use testing to check for counterexamples, which
-ensures results are empirically plausible and applicable (conditions are
-satisfiable, types are inhabited, etc.). They also use a small-to-large search
-order, ensuring more easily-comprehensibile conjectures are explored first.
+The latter systems all use testing to check for counterexamples, which ensures
+results are empirically plausible and applicable (conditions are satisfiable,
+types are inhabited, etc.). They also use a small-to-large search order,
+ensuring more easily-comprehensibile conjectures are explored first.
 
@@
 expression can be simplified using a previously-found equation. The other
-systems use post-hoc filters based on unification and rewriting rules
-(derived via the Knuth-Bendix completion algorithm) to discard expressions or
-equations which are special-cases of some other, more general result.
+systems use post-hoc filters based on unification and term rewriting to discard
+expressions or equations which are special-cases of some other, more general
+result.
 
@@
 with other aspects of mathematics, including the construction of conjectures,
-(counter-)examples and theorems. These tasks, along with the formation of new
-definitions and concepts, are varyingly known as
+(counter\nobreakdash-)examples and theorems. These tasks, along with the
+formation of new definitions and concepts, are varyingly known as
 ``Automated Theory Formation''~\cite{lenat:77,colton:book},
@@
 theorem and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re-)defines all of the datatypes
-and functions it involves. To avoid ambiguity we prefix the names of all
-definitions with the filename it came from, and collect them into one big set.
+of these are duplicates, since each problem (re\nobreakdash-)defines all of the
+datatypes and functions it involves. To avoid ambiguity we prefix the names of
+all definitions with the filename it came from, and collect them into one big set.
 

### 2018-02-26 TEBenchmarkPaper/article.tex
Use lists to demonstrate TIP

@@
 TIP datatypes can have several ``constructors'' and ``destructors''. For example
-the natural numbers can be defined as follows:
+the type of lists from figure~\ref{figure:list_theory} can be defined in TIP as
+follows:
 
 \begin{verbatim}
-(declare-datatypes () ((Nat (Z) (S (p Nat)))))
+(declare-datatypes (a) ((List (Nil) (Cons (head a) (tail (List a))))))
 \end{verbatim}
 
-This defines the type \texttt{Nat}, along with the constructors \texttt{Z} (for
-zero) and \texttt{S} (the successor function), and destructor \texttt{p} (the
-predecessor function).
-
-Since our GHC plugin doesn't list constructors or destructors, we add separate
-function definitions for each (these are just $\eta$-expansions).
+This defines a type \texttt{List} parameterised by a type variable \texttt{a}
+(the type of the list's elements), with constructors \texttt{Nil} and
+\texttt{Cons}, and destructors \texttt{head} and \texttt{tail}. Target languages
+(e.g. Haskell and Isabelle) handle constructors and destructors differently, so
+we avoid exploring them directly. Rather, we $\eta$-expand each constructor and
+destructor to create new function definitions, and rewrite the TIP theorems to
+reference these expanded forms instead.
 
 TIP has ``native'' support for booleans and integers, allowing functions like
-\texttt{+} to be used without an accompanying definition. Without a definition
-these functions will not be included in the QuickSpec exploration, so we replace
-all occurrences with custom definitions.
-
-This gives us 3598 definitions in total, with 269 remaining after
-$\alpha$-equivalent duplicates are removed.
+\texttt{+} to be used without an accompanying definition. To ensure consistency
+in the translations, we replace all occurrences of these ``native'' expressions
+with standard definitions using the same \texttt{declare-datatypes} and
+\texttt{define-fun} mechanisms as the problem set.
+
+This gives us a total of 3598 definitions, including those from the problem set
+and our automatically-generated additions. Removing $\alpha$-equivalent
+duplicates leaves 269 definitions, and we chose to only sample from those 182
+which occur in at least one theorem statement (this removes ambiguity about
+which \emph{definitions}, including our generated ones, are ``interesting'' and
+which are just ``implementation details'').
 

### 2018-02-26 TEBenchmarkPaper/article.tex
Use et al consistently

@@
 
+\newcommand{\etal}{{\em et al.}}
+\newcommand{\Bagai}{Bagai \etal{}}
+
 % For example theory of lists
@@
 approaches agree that simple tautologies are ``uninteresting'', but differ when
-it comes to more complex statements.  Colton {\em et al.} give a survey of
-systems for theory formation and exploration, and their notions of
-``interestingness'' for concepts and conjectures, identifying six notions in
+it comes to more complex statements.  Colton \etal{} give a survey of systems for
+theory formation and exploration, and their notions of ``interestingness'' for
+concepts and conjectures, identifying six notions in
 total~\cite{colton2000notion}. These notions, as applied to conjectures, are:
@@
 {\bf Applicability} depends on the number of models in which a
-conjecture holds. The system of Bagai et al conjectures the
-non-existence of objects, and hence favours statements with zero
-applicability. Other systems treat applicability as a positive
-aspect: the more applicable the statement, the more interesting it is.
+conjecture holds. The system of \Bagai{} conjectures the non-existence of
+objects, and hence favours statements with zero applicability. Other systems
+treat applicability as a positive aspect: the more applicable the statement, the
+more interesting it is.
 
@@
     \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
-    \tRow{Bagai et al & \tNE               &     &   X &     &   X &   X &    }
+    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
     \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
@@
 We summarise the criteria in Table 1 and state which were used in key
-ATF/MTE systems. The entries for AM, GT, Graffiti, Bagai et al. and HR are based
+ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
 on this survey (the latter is Colton's own system). We extend this analysis to
@@
 %IsaScheme
-Monta{\~n}o-Rivas {\em et al.} have implemented a scheme-based
-approach to MTE in their IsaScheme system
-\cite{MontanoRivas2011}. Schemes are higher-order formulae which can
-be used to generate new concepts and conjectures; variables within the
-scheme are instantiated automatically and this drives the invention
-process.  For instance, in the theory of natural numbers, given the
-concepts of successor ($suc$), addition ($+$) and zero ($0$),
-IsaScheme can use the following scheme to invent the concept of
-multiplication:
+Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
+IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
+which can be used to generate new concepts and conjectures; variables within the
+scheme are instantiated automatically and this drives the invention process.
+For instance, in the theory of natural numbers, given the concepts of successor
+($suc$), addition ($+$) and zero ($0$), IsaScheme can use the following scheme
+to invent the concept of multiplication:
 
@@
 capture the ``almost true'', which can be a valuable trigger for theory change,
-as demonstrated by Lakatos in his case studies of mathematical development
-\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly proposed
-metric. Methods for finding out what that means have come from neuro-scientists,
-such as the study by Zeki et al in which they test whether mathematicians'
-experiences of abstract beauty correlates with the same brain activity as
-experiences of sensory beauty \cite{zeki}; computer scientists, such as Colton's
-metrics above which are proposed based largely on a computer scientist's
-``intuition'' and argumentation about why a metric would be important (and
-previous use of such metrics, and - in one isolated case (surprisingingness) -
-on a single quote from a mathematician\footnote{Similarly, classification of a
-MATHsAiD result as either interesting or uninteresting \cite{roy} was made by
-the main system developer: in both cases the main developer had a background in
-mathematics.}; and mathematicians themselves, such as Gowers' suggestion that we
-can identify features which are commonly associated with good proofs
-\cite{gowers}. All of these approaches rest upon the assumption that it makes
-sense to speak of ``the good'' in mathematics. This assumption is supported by
-concepts such as Erdos's ``The Book'' -- a theoretical construct of a book which
-contains the most elegant proof of every theorem (brought to life in
-\cite{aigner2010proofs}). However,
-empirical studies by psychologists such as Inglis call into question such
-assumptions: work by Inglis and colleagues has shown that there is not a single
-standard of validity among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
+as demonstrated by Lakatos in his case studies of mathematical
+development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
+proposed metric. Methods for finding out what that means have come from
+neuro-scientists, such as the study by Zeki \etal{} in which they test whether
+mathematicians' experiences of abstract beauty correlates with the same brain
+activity as experiences of sensory beauty~\cite{zeki}; computer scientists, such
+as Colton's metrics above which are proposed based largely on a computer
+scientist's ``intuition'' and argumentation about why a metric would be
+important (and previous use of such metrics, and - in one isolated case
+(surprisingingness) - on a single quote from a mathematician\footnote{Similarly,
+  classification of a MATHsAiD result as either interesting or
+  uninteresting~\cite{roy} was made by the main system developer: in both cases
+  the main developer had a background in mathematics.}; and mathematicians
+themselves, such as Gowers' suggestion that we can identify features which are
+commonly associated with good proofs~\cite{gowers}. All of these approaches rest
+upon the assumption that it makes sense to speak of ``the good'' in mathematics.
+This assumption is supported by concepts such as Erdos's ``The Book'' -- a
+theoretical construct of a book which contains the most elegant proof of every
+theorem (brought to life in~\cite{aigner2010proofs}). However, empirical studies
+by psychologists such as Inglis call into question such assumptions: work by
+Inglis and colleagues has shown that there is not a single standard of validity
+among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
 nor beauty. %FIXME needs citation

### 2018-02-26 TEBenchmarkPaper/article.tex
Consistent citations

@@
   \end{tabular}
-  \caption{Classification of MTE systems from \cite{colton2000notion}, extended
-    to those compared in \cite{claessen2013automating} (QuickSpec is the
-    conjecture generation component of HipSpec). The interestingness measures
-    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
+  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
+    to include those from~\cite{claessen2013automating} (QuickSpec is the
+    conjecture generation component of HipSpec) and the similar Speculate
+    system~\cite{braquehais2017speculate}. The interestingness measures are
+    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
     \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
@@
 some more recent systems in the remaining rows (based on our understanding of
-their function): QuickSpec, Speculate~\cite{braquehais2017speculate}, IsaCoSy
-and IsaScheme.
+their function): QuickSpec~\cite{QuickSpec},
+Speculate~\cite{braquehais2017speculate},
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
+and IsaScheme~\cite{MontanoRivas2011}.
 
@@
 rivalling the labours of the most practised mathematicians in all
-branches of their science.'' \cite[p. 498]{jevons}
+branches of their science.''~\cite[p. 498]{jevons}
 \end{quote}
@@
 by repetitions, and was conceived by Pascal in 1642 while reorganising
-tax revenues \cite{d'ocagne}. Subsequent early systems include
+tax revenues~\cite{d'ocagne}. Subsequent early systems include
 M\"uller's universal calculating machine in 1784, which he built for
@@
 this when he had to check and recalculate some tables relating to the
-volumes of trees \cite[p. 65]{lindgren}. Thirty seven years later,
+volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
 Babbage invented his famous difference engine: an automatic,
@@
 was inspired by a flawed table of logarithms and the idea that
-machines would be quicker and more reliable \cite{bowden}. It is
+machines would be quicker and more reliable~\cite{bowden}. It is
 interesting to note background and motivation: while Pascal and
@@
 Automated theory formation derives its terminology from psychology in which the
-term ``concept formation'' is used (see, for example, \cite{bruner:67}) to
+term ``concept formation'' is used (see, for example,~\cite{bruner:67}) to
 describe the search for features which differentiate exemplars from
 non-exemplars of various categories. Lenat used this term in his 1977 paper:
-{\em Automated Theory Formation in Mathematics} \cite{lenat:77}.
+{\em Automated Theory Formation in Mathematics}~\cite{lenat:77}.
 
 % 4. example systems - AM, HR, ...
-When Lenat built the AM system \cite{lenat:77}, there were systems
+When Lenat built the AM system~\cite{lenat:77}, there were systems
 which could define new concepts for investigation, such as those
-described in \cite{winston}, and systems which could discover
-relationships among known concepts, such as Meta-Dendral
-\cite{buchanan:75}. No system could perform both of these tasks: Lenat
-saw this as the next step. AM was designed to both construct new
-concepts and conjecture relationships between them; fully automating
-the cycle of discovery in mathematics. Lenat describes this as
-follows:
+described in~\cite{winston}, and systems which could discover
+relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
+system could perform both of these tasks: Lenat saw this as the next step. AM
+was designed to both construct new concepts and conjecture relationships between
+them; fully automating the cycle of discovery in mathematics. Lenat describes
+this as follows:
 
@@
 concentrate upon the most interesting ones, and iterate the
-entire process.'' \cite[p. 834]{lenat:77}
+entire process.''~\cite[p. 834]{lenat:77}
 \end{quote}
@@
 large discrepancy between his theoretical claims and the implemented
-program \cite{partridge}. For instance, Lenat made claims about how AM
+program~\cite{partridge}. For instance, Lenat made claims about how AM
 invented natural numbers from sets, whereas it used one heuristic
@@
 were sometimes under-explained. For an argument of why many of the
-claims made by Lenat about AM were false, see chapter 13 of
-\cite{colton:book}.
+claims made by Lenat about AM were false, see chapter 13 of~\cite{colton:book}.
 
@@
 which seems to have originated with Buchberger and colleagues (see,
-for example, \cite{buchberger}). His motivation is to support
+for example,~\cite{buchberger}). His motivation is to support
 mathematicians during their exploration of mathematical theories. This
@@
 illustrating the automated synthesis of his own Gr\"obner bases
-algorithm \cite{buchberger:04}.
-
-Recent systems developed in this area include MATHsAiD \cite{roy},
-IsaScheme \cite{MontanoRivas2011} and IsaCosy \cite{johansson}. The
-goal of the MATHsAiD (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions) project is to build a tool which
-takes in a set of axioms, concept definitions and a logic and applies
-its inference rules to reason from the axioms to theorems. The
-motivation is to produce a tool which will help mathematicians to
-explore the consequences of a set of axioms or a particular concept
-definition. One of the main challenges of the project has been to
-automatically evaluate interestingness: to distinguish important
-theorems from results which, although they follow from a set of
-axioms, are of little mathematical interest.
+algorithm~\cite{buchberger:04}.
+
+Recent systems developed in this area include MATHsAiD~\cite{roy},
+IsaScheme~\cite{MontanoRivas2011} and
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}. The goal of the
+MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses, Axioms and
+Definitions) project is to build a tool which takes in a set of axioms, concept
+definitions and a logic and applies its inference rules to reason from the
+axioms to theorems. The motivation is to produce a tool which will help
+mathematicians to explore the consequences of a set of axioms or a particular
+concept definition. One of the main challenges of the project has been to
+automatically evaluate interestingness: to distinguish important theorems from
+results which, although they follow from a set of axioms, are of little
+mathematical interest.
 
@@
 variables in the same scheme, resulting in the invention of the
-exponentiation concept (these examples are taken from
-\cite{MontanoRivas2011}).
-
-%IsaCosy
-The IsaCosy system (Isabelle Conjecture Synthesis) is a program for
+exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).
+
+%IsaCoSy
+The IsaCoSy system (Isabelle Conjecture Synthesis) is a program for
 inductive theory formation, which synthesises conjectures from the
@@
   mathematics: we shall not venture down this path here (interested
-  readers are referred to \cite{hersh:97,shapiro}). Instead, for now,
+  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
   we make the pragmatic assumption that while there may well be
@@
 some of these historical differences are disappearing, and current
-systems such as MATHsAiD \cite{roy}, IsaScheme \cite{MontanoRivas2011}
-and IsaCosy \cite{johansson} are bridging the methodological gap
-between ATF and MTE. This is particularly true in terms of
+systems such as MATHsAiD~\cite{roy}, IsaScheme~\cite{MontanoRivas2011}
+and IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation} are bridging the
+methodological gap between ATF and MTE. This is particularly true in terms of
 fully-automatic/user-interactive and the domain-specific/general
@@
 If we limit ourselves to exploring higher-order theories with inductive types
-(a domain closely matching our interest in analysing software), we do find a
-direct comparison of three systems~\cite{claessen2013automating}: HipSpec,
-IsaCoSy and IsaScheme. QuickSpec, shown in table \ref{table:colton}, is the
-conjecture-generating component of HipSpec (which sends those conjectures into
-automated theorem provers).
+(a domain useful for analysing software), we do find a direct comparison of
+three systems~\cite{claessen2013automating}: HipSpec, IsaCoSy and IsaScheme.
+QuickSpec, shown in table \ref{table:colton}, is the conjecture-generating
+component of HipSpec.
 
@@
 and regular competitions to prove as much as possible, as fast as
-possible\cite{pelletier2002development}. These problem sets are an opportunity
+possible~\cite{pelletier2002development}. These problem sets are an opportunity
 for theory exploration, as their definitions and theorems can be used as a

### 2018-02-26 TEBenchmarkPaper/article.tex
Minor changes

@@
 Scientific knowledge relies on inductive reasoning and experimental testing,
-which are not so important to a mathematical theory. However, these techniques
-do play an important role in many of the tools we investigate: in particular,
-inductive reasoning allows a system to conjecture statements which may be too
-difficult for it to deductively prove; whilst experimental testing, e.g. by
-checking universally-quantified statements against some specific values, allows
-many falsehoods to be quickly discarded without requiring more elaborate
-symbolic reasoning.
+which (in principle) are not so important to a mathematical theory. However,
+these techniques \emph{do} play an important role in many existing formation and
+exploration tools: in particular, inductive reasoning allows a system to
+conjecture statements which may be too difficult for it to deductively prove;
+whilst experimental testing, e.g. by checking universally-quantified statements
+against some specific values, is a simple way to quickly discard many obvious
+falsehoods generated by powerful but imperfect algorithms (as opposed to, say,
+limiting ourselves to verified theorem generators).
 
@@
 
-There are three aspects to consider when we evaluate an MTE system:
+There are three aspects to consider when we evaluate a conjecture generation
+system:
 
@@
 
-% TODO: Should "generality" be one? Or is that just an implicit "don't cheat"
-% assumption?
+We also make the assumption that the system is intended to be \emph{general},
+and hence isn't hard-coded or fine-tuned to only perform well in a narrow
+domain. Examples where this does \emph{not} hold include the natural number
+heuristics of AM (discussed above), as well as machine learning models trained
+on some specific dataset.
 
@@
 
-Each system is run, and their output compared to the ground truth. To score
-100\% on precision and recall, an MTE system must output all of the conjectures
-which appear in the ground truth, and nothing else:
+Each system is run on these definitions, and their outputs are compared to the
+ground truth. To score 100\% on precision and recall, an MTE system must output
+all of the conjectures which appear in the ground truth, and nothing else:
 
@@
 the Isabelle theorem prover, with one involving the natural numbers, and the
-other involving the list functions from figure \ref{figure:list_theory}. The
-library authors have gone to the effort of stating, proving and including these
-theorems in every copy of their software, which is a good indication that they
-are useful or important.
+other involving the list functions from figure \ref{figure:list_theory}. Since
+the library authors have gone to the effort of stating, proving and including
+these theorems in every copy of their software, this is a good indication that
+they are useful or important.
 

### 2018-02-27 TEBenchmarkPaper/supporting-materials/mkQuickSpecGraphs.py
Use corrected data

@@ -7,10 +7,8 @@ msg = lambda x: sys.stderr.write(x + '\n')
 msg('Reading input')
 
 def readInput():
-    from gzip import open
     from json import load
-    with open(os.getenv('zipped'), 'r') as zipped:
-        return load(zipped)['results']['quickspectip.track_data']['result'][0]
+    return load(sys.stdin)['results']['quickspectip.track_data']['result'][0]
 
 data = readInput()
 del(readInput)
@@ -116,24 +114,9 @@ def aggregateData(data):
 
             knownEqs[sample] = got
 
-            if rdata['success']:
-                found   = len(got)
-                correct = len([x for x in rdata['wanted'] if x['found']])
-                wanted  = len(rdata['wanted'])
-            else:
-                found   = 0
-                correct = 0
-
-                # FIXME: HaskellTE should perform analysis even for failures
-                proc = subprocess.Popen(
-                    [os.getenv('conjectures_for_sample')],
-                    stdin  = subprocess.PIPE,
-                    stdout = subprocess.PIPE,
-                    env    = {'SAMPLED_NAMES' : '\n'.join(list(sample))})
-                (out, err) = proc.communicate('[]')
-                outData    = loads(out)
-                wanted = len(outData['wanted'])
-                del(proc, out, err)
+            found   = 0 if got is None else len(got)
+            correct = len([x for x in rdata['wanted'] if x['found']])
+            wanted  = len(rdata['wanted'])
 
             assert wanted > 0, repr({
                 'error'  : 'Sampling should guarantee a wanted conjecture',
@@ -309,7 +292,7 @@ def savePlot(name):
 def plotTime():
     newFigure('time')
 
-    plt.ylim(0, 180)
+    plt.ylim(0, 300)
 
     sns.boxplot(data      = agg,
                 x         = 'size',

### 2018-02-27 TEBenchmarkPaper/article.tex
Alter article text to reflect new data

@@
 
-It is important to ensure that there are no duplicate definitions: theory
-exploration does not depend on the names of functions, so neither should our
-analysis. For example, consider a problem set which includes a commutativity
-theorem for a \texttt{plus} function, and an associativity theorem for an
-\texttt{add} function, where the definitions of \texttt{plus} and \texttt{add}
-are $\alpha$-equivalent. We would expect an MTE system to conjecture
-commutativity and associativity for \emph{both} functions, or for \emph{neither}
-function, since they're equivalent. Yet a na\"ive precision/recall analysis
-would treat commutativity of \texttt{add} and associativity of \texttt{plus} as
-\emph{uninteresting}, since they don't appear in the ground truth.
-
-For this reason, duplicates should be removed (e.g. leaving those which appear
-first alphabetically), and any references in the theorem statements updated to
-use the remaining definition. In the above example, the \texttt{plus} function
-would be removed, and the commutativity theorem updated to reference
-\texttt{add} instead.
+It is important to ensure that there are no duplicate definitions: we are only
+concerned with the \emph{logical} content of the input, not the more arbitrary
+aspects like the names of functions. For example, consider a problem set which
+includes a commutativity theorem for a \texttt{plus} function, and an
+associativity theorem for an \texttt{add} function, where the definitions of
+\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
+system to conjecture commutativity and associativity for \emph{both} functions,
+or for \emph{neither} function, since they're logically equivalent. Yet a
+na\"ive precision/recall analysis would treat commutativity of \texttt{add} and
+associativity of \texttt{plus} as \emph{uninteresting}, since they don't appear
+in the ground truth.
+
+For this reason, duplicates should be removed, and any references in the theorem
+statements updated to use the remaining definition (e.g. chosen based on
+lexicographic order). In the above example, the \texttt{plus} function would be
+removed, and the commutativity theorem updated to reference \texttt{add}
+instead.
 
@@
 \item It is impractical to run any existing MTE system on ``large'' inputs,
-  containing more than around 10 definitions.
+  containing more than a few dozen definitions.
 \end{itemize}
@@
 definitions. Given a sample size, we choose a subset of that many definitions,
-and use that as our MTE input. We generate a corresponding ground truth by
-selecting those theorems from the corpus which only contain references to
-definitions that appear in our sample.
+and provide those definitions as our MTE input. We generate a corresponding
+ground truth by selecting those theorems from the corpus which only contain
+references to the sampled definitions.
 
 Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-large corpus and a fixed sample size, it becomes highly unlikely that a sample
-will contain the particular definitions required by a theorem. This would cause
-most samples to have an empty ground truth, and hence $0$ precision and
-undefined recall \emph{independent} of what the MTE system produces. This is
-clearly undesirable as an evaluation method.
+given sample size, increasing the size of the corpus (which, as detailed above,
+is advantageous) makes it less likely that a chosen sample will contain all of
+the definitions required by a theorem. The majority of such samples would hence
+have an empty ground truth, and thus $0$ precision and undefined recall
+\emph{independent} of what the MTE system produces! This is clearly undesirable
+as an evaluation method.
 
@@
 least one applicable theorem from the corpus (the one we chose as our starting
-point).
+point), and hence have a meaningful precision and recall.
 
@@
 chosen definitions into the particular input format of the tool under study,
-then we time the execution, with a timeout after e.g. 1 hour.
+then we time the execution, with a timeout after e.g. 5 minutes.
 
@@
 
-We summarise the runtimes by choosing their median, as it is more robust against
-long-running outliers, and hence represents performance for a ``typical''
-theory.
+We summarise the runtimes by choosing their median, as this is more robust
+against long-running outliers, and hence represents performance for a
+``typical'' theory.
 
@@
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 30 per size except for 1 (14) and 2 (29). Colour
-    indicates the total number of conjectures produced by each run, with failing
-    processes shown in red. Some points are shifted horizontally to prevent
-    overlaps. Standard box plots are included, although many are obscured due to
-    quartiles near zero or timeout. Whiskers show points at 1.5 $\times$ the
-    inter-quartile range.}
+    is an individual run, with 30 per size except for 2 (29). Colour indicates
+    the total number of conjectures produced by each run, with failing processes
+    shown in red. Some points are shifted horizontally to prevent overlaps.
+    Standard box plots are included, with whiskers showing points at
+    1.5~$\times$~the inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
 
-We applied our sampling methodology to the 269 TIP definitions for sample sizes
-between 1 and 20. We chose 30 samples for each size, but due to the requirement
-for samples to admit at least one theorem, only 14 unique samples were obtained
-for size 1 and 29 for size 2. A timeout of 3 minutes was imposed, since
-preliminary experiments showed that most runs either finish before this time, or
-still do not finish after an hour. The running times are shown in
+We applied our sampling methodology to the 182 definitions which appear in TIP
+theorems, for even sample sizes up to 20. We chose 30 samples for each size, but
+due to the requirement for samples to admit at least one theorem, only 29 unique
+samples were obtained for size 2. A timeout of 5 minutes was imposed, since
+preliminary experiments showed that most QuickSpec runs either finish before
+this time, or still do not finish after an hour. The running times are shown in
 figure~\ref{figure:quickspec_runtimes}.
@@
 The only errors we encountered were timeouts (shown in red), and these occurred
-for all sample sizes. Whilst the median runtimes are mostly in the range of a
-few seconds, the variance shows two distinct regions: sizes below 10 are mostly
-finished within 30 seconds, whilst sizes 10 and above have large variance; the
-whole upper quartile of size 13 timed out. There is also an unexpected downwards
-trend towards size 20, and a slight but consistent reduction in minimum runtimes
-from sizes 15 to 20 which is likely an artefact of our testing setup.
+for all sample sizes. The running times cluster, with many close to zero and
+many exceeding the timeout, but with few in between. There is a clear trend
+towards longer times and more timeouts as the sample size increases, such that
+most samples of size 20 timed out.
 
 Most runs finished quickly and generated few conjectures, with the most being
-131 conjectures from a sample of size 20.
+116 conjectures from a sample of size 16.
 
@@
 explore larger theories. Some small samples reached full precision, but nothing
-above size 10 exceeded 50\%.
-
-Recall shows a slight downward trend but is mostly flat, averaging between 10\%
-and 40\%. This is expected, since adding more definitions to a QuickSpec
+above size 6 exceeded 50\%.
+
+Recall shows a slight downward trend but is mostly flat, averaging between 20\%
+and 50\%. This is expected, since adding more definitions to a QuickSpec
 signature will never reduce the conjectures it will find. Those conjectures

### 2018-03-05 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Remove average of ratios

@@ -511,34 +511,9 @@ def aggProp(sizes=None, agg=None, key=None, total=None):
             'sample stddev'     : sStddev
         }
 
-    def averageOfRatios(size):
-        '''Assume that each run is a binomial experiment with the same p
-        parameter but differing n parameters. We calculate the mean and variance
-        for each run, then combine them together. This relies on the fact that
-        the sum of binomials with the same p parameter is also a binomial, hence
-        finding the mean by summing the individual means and dividing by the
-        number of runs will maintain our binomial assumption. The total variance
-        is just the sum of the individual variances, scaled due to the
-        division.
-
-        NOTE: This is mostly for comparison purposes.'''
-
-        # If a run has no total, then its mean and variance are undefined. We
-        # skip these runs.
-        indices = filter(lambda i: agg[total][i] not in [None, 0],
-                         sizeIndices(size))
-
-        means  = map(lambda i: noneToZero(agg[key][i]), indices)
-        mean   = float(sum(means)) / float(len(means))
-
-        counts    = map(lambda i: noneToZero(agg[total][i]), indices)
-        variances = [(p * (1.0 - p)) / n for p, n in zip(means, counts)]
-        variance  = sum(variances) / (float(len(means))**2)
-        return {'mean': mean, 'variance': variance}
-
     # Calculate values for graph
 
-    analysis = ratioOfAverages  # Switch this to use a different analysis
+    analysis = ratioOfAverages
 
     unzip = lambda xs: zip(*xs)
 

### 2018-03-05 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Only use successful runs in precision/recall

@@ -234,6 +234,7 @@ def aggregateData(data):
         'recall'    : [],
         'recHue'    : [],
         'size'      : [],
+        'success'   : [],
         'time'      : [],
         'timeHue'   : [],
         'wanted'    : []
@@ -289,6 +290,7 @@ def aggregateData(data):
             agg['recall'   ].append(r)
             agg['recHue'   ].append(wanted)
             agg['size'     ].append(size)
+            agg['success'  ].append(rdata['success'])
             agg['time'     ].append(t)
             agg['timeHue'  ].append(found  + 1 if rdata['success'] else 0)
             agg['wanted'   ].append(wanted)
@@ -478,11 +480,22 @@ def aggProp(sizes=None, agg=None, key=None, total=None):
         Under this assumption, the fact that conjectures come from different
         runs is irrelevant, so we pool them all together into one binomial
         experiment.'''
-        totals   = map(noneToZero, entriesOfSize(size, total))
-        corrects = map(noneToZero, entriesOfSize(size, 'correct'))
+        indices  = [i for i in sizeIndices(size) if agg['success'][i]]
+        totals   = [agg[total    ][i] for i in indices]
+        corrects = [agg['correct'][i] for i in indices]
 
-        count   = sum(totals)
-        correct = sum(corrects)
+        debug = lambda err: repr({
+            'error'    : err,
+            'key'      : key,
+            'totals'   : totals,
+            'corrects' : corrects,
+            'indices'  : indices
+        })
+        assert 0    not in totals, debug('Got 0 total, even for success')
+        assert None not in totals, debug('Got None total, even for success')
+
+        count   = float(sum(totals))
+        correct = float(sum(corrects))
 
         p       = correct / count
 
@@ -496,10 +509,9 @@ def aggProp(sizes=None, agg=None, key=None, total=None):
         # the correct proportion we got from the model (p).
         # Note that we filter out any runs with empty totals (i.e. we don't try
         # to work out the precision of a run which produced no conjectures)
-        n       = float(len(filter(lambda x: x > 0, totals)))
+        n       = float(len(totals))
         sVar    = (sum([((float(correct) / tot) - p)**2 \
-                        for tot, correct in zip(totals, corrects) \
-                        if  tot > 0])) / (n - 1)
+                        for tot, correct in zip(totals, corrects)])) / (n - 1)
         sStddev = math.sqrt(sVar)
 
         return {

### 2018-03-05 TEBenchmarkPaper/article.tex
Clarification

@@
 tooling to convert this format into a variety of languages, including Haskell
-and Isabelle.
+and Isabelle which are used by existing MTE tools.
 
@@
 of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves. To avoid ambiguity we prefix the names of
-all definitions with the filename it came from, and collect them into one big set.
-
-TIP datatypes can have several ``constructors'' and ``destructors''. For example
-the type of lists from figure~\ref{figure:list_theory} can be defined in TIP as
+datatypes and functions it involves.
+
+TIP datatypes can have several ``constructors'' (introduction forms) and
+``destructors'' (elimination forms; field accessors). For example the type of
+lists from figure~\ref{figure:list_theory} can be defined in the TIP format as
 follows:
@@
 \begin{verbatim}
-(declare-datatypes (a) ((List (Nil) (Cons (head a) (tail (List a))))))
+(declare-datatypes
+  (a)                       ;; Type variable (element type)
+  ((List                    ;; Type name
+     (Nil)                  ;; Constructor (nullary)
+     (Cons                  ;; Constructor (binary)
+       (head a)             ;; Field name and type
+       (tail (List a))))))  ;; Field name and type
 \end{verbatim}
 
-This defines a type \texttt{List} parameterised by a type variable \texttt{a}
-(the type of the list's elements), with constructors \texttt{Nil} and
-\texttt{Cons}, and destructors \texttt{head} and \texttt{tail}. Target languages
-(e.g. Haskell and Isabelle) handle constructors and destructors differently, so
-we avoid exploring them directly. Rather, we $\eta$-expand each constructor and
-destructor to create new function definitions, and rewrite the TIP theorems to
-reference these expanded forms instead.
-
-TIP has ``native'' support for booleans and integers, allowing functions like
-\texttt{+} to be used without an accompanying definition. To ensure consistency
-in the translations, we replace all occurrences of these ``native'' expressions
-with standard definitions using the same \texttt{declare-datatypes} and
-\texttt{define-fun} mechanisms as the problem set.
-
-This gives us a total of 3598 definitions, including those from the problem set
-and our automatically-generated additions. Removing $\alpha$-equivalent
-duplicates leaves 269 definitions, and we chose to only sample from those 182
-which occur in at least one theorem statement (this removes ambiguity about
-which \emph{definitions}, including our generated ones, are ``interesting'' and
-which are just ``implementation details'').
-
-TIP includes a tool to translate its definitions into Haskell code, including
-suitable comparison functions and random data generators. We translated all 269
-definitions into a single Haskell package, compiled it with GHC and extracted
-the function definitions with our plugin. Each benchmark run imported this
-entire package, but the signature given to QuickSpec only included functions
-chosen for the current sample.
+Our target languages (Haskell and Isabelle) differ in the way they handle
+constructors and destructors, which complicates comparisons. To avoid this, we
+generate a new function for each constructor (via $\eta$-expansion) and
+destructor (via pattern-matching) of the following form:
+
+\begin{verbatim}
+(define-fun
+  (par (a)                   ;; Type variable
+    (constructor-Cons        ;; Function name
+      ((x a) (xs (List a)))  ;; Argument names and types
+      (List a)               ;; Return type
+      (as                    ;; Type annotation
+        (Cons x xs)          ;; Return value
+        (List a)))))         ;; Return type
+
+(define-fun
+  (par (a)                        ;; Type variable
+    (destructor-head              ;; Function name
+      ((xs (List a)))             ;; Argument name and type
+      a                           ;; Return type
+      (match xs                   ;; Pattern-match
+        (case (Cons h t) h)))))   ;; Return relevant field
+\end{verbatim}
+
+\begin{sloppypar}
+  We rewrite the TIP theorems (our ground truth) to reference these expanded
+  forms instead of the raw constructors and destructors, and include these
+  functions in our samples in lieu of the raw expressions. Note that these
+  destructor wrappers are \emph{partial} functions (e.g.
+  \texttt{destructor-head} and \texttt{destructor-tail} are undefined for the
+  input \texttt{Nil}), which complicates their translation to proof assistants
+  like Isabelle.
+\end{sloppypar}
+
+Another complication is TIP's ``native'' support for booleans and integers,
+which allows numerals and symbols like \texttt{+} to appear without any
+accompanying definition. To ensure consistency in the translations, we replace
+all occurrences of such expressions with standard definitions written with the
+``user-level'' \texttt{declare-datatypes} and \texttt{define-fun} mechanisms.
+
+When we add all of these generated types and functions to those in TIP, we get a
+total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
+and we choose to only sample from those 182 which occur in at least one theorem
+statement (this removes ambiguity about which \emph{definitions} count as
+``interesting'' and which are just ``implementation details'' for other
+definitions).
+
+TIP includes a tool to translate its definitions into Haskell and Isabelle,
+including suitable comparison functions and random data generators. We translate
+all 269 unique definitions into a single module/theory which is imported on each
+run of the MTE tools (although only those functions which appear in the current
+sample are actually explored).
 
@@
 We applied our sampling methodology to the 182 definitions which appear in TIP
-theorems, for even sample sizes up to 20. We chose 30 samples for each size, but
-due to the requirement for samples to admit at least one theorem, only 29 unique
-samples were obtained for size 2. A timeout of 5 minutes was imposed, since
-preliminary experiments showed that most QuickSpec runs either finish before
-this time, or still do not finish after an hour. The running times are shown in
-figure~\ref{figure:quickspec_runtimes}.
-
-The only errors we encountered were timeouts (shown in red), and these occurred
-for all sample sizes. The running times cluster, with many close to zero and
-many exceeding the timeout, but with few in between. There is a clear trend
-towards longer times and more timeouts as the sample size increases, such that
-most samples of size 20 timed out.
-
-Most runs finished quickly and generated few conjectures, with the most being
-116 conjectures from a sample of size 16.
+theorems, for even sample sizes up to 20. We chose 30 samples for each size,
+although one sample of size 2 was discarded as it was a duplicate. A timeout of
+5 minutes was imposed, since preliminary experiments showed most QuickSpec runs
+to either finish before this time, or still be running after an hour. The
+running times are shown in figure~\ref{figure:quickspec_runtimes}.
+
+The only errors we encountered were timeouts (shown in red), which occurred for
+all sample sizes but are more common for larger samples (over half the runs for
+sample size 20 timed out). Runs which did not time out mostly finished within 30
+seconds, generating few or no conjectures; those taking longer generated more
+conjectures, with the most being 116 conjectures from a sample of size 16.
 
@@
 We use a binomial distribution to model precision and recall, as if each
-conjecture is a ball drawn from an urn of wanted/unwanted results (in the case
-of precision) or found/missed ground truths (in the case of recall). The mean of
-these distributions estimates the overall proportion of wanted or found
-conjectures, and we use the standard deviation to determine how well this mean
-fits the data.
+conjecture is drawn from a pool of wanted/unwanted results (in the case of
+precision) or found/missed ground truths (in the case of recall), with the mean
+of these distributions estimating the proportions of each pool's split.
 

### 2018-03-08 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Mark interrupted (OOM) IsaCoSy runs as failures

@@ -69,6 +69,19 @@ def normalise(data):
             if 'killed' in data[size][rep]:
                 del(data[size][rep]['killed'])
 
+            # If IsaCoSy gets interrupted, that indicates it died from OOM
+            stderr = data[size][rep]['stderr']
+            if stderr is not None:
+                interr = 'unable to increase stack'.replace(' ', '').lower()
+                stderr = stderr.replace('\n',  ' ').replace(' ', '').lower()
+                if interr in stderr:
+                    msg(repr({
+                        'warning' : 'Exploration was interrupted',
+                        'size'    : size,
+                        'rep'     : rep
+                    }))
+                    data[size][rep]['success'] = False
+
             # We should always analyse, so this is useless
             if 'analysed' in data[size][rep]:
                 assert data[size][rep]['analysed'], 'Run was not analysed'
@@ -165,6 +178,14 @@ def sanityCheck(data):
                     'Got different timeouts {} and {}'.format(
                         str(timeout),
                         str(rdata['timeout']))
+
+                if not rdata['success']:
+                    assert rdata['found'] == frozenset([]), repr({
+                        'error'  : 'Found conjectures for failing run',
+                        'system' : system,
+                        'size'   : size,
+                        'rep'    : rep
+                    })
 
                 # Make sure every system is using the same samples/ground truth
                 for sys in data.keys():

### 2018-03-08 TEBenchmarkPaper/article.tex
Fleshed out application section

@@
 definitions. Given a sample size, we choose a subset of that many definitions,
-and provide those definitions as our MTE input. We generate a corresponding
-ground truth by selecting those theorems from the corpus which only contain
-references to the sampled definitions.
+and provide those definitions as input to our MTE tool. We generate a
+corresponding ground truth by selecting those theorems from the corpus which
+``depend on'' (contain references to) \emph{only} the definitions in that
+sample. Transitive dependencies aren't required (e.g. a theorem which only
+references \texttt{times} does not depend on \texttt{plus}, even if
+\texttt{plus} occurs in the definition of \texttt{times}).
 
@@
 is advantageous) makes it less likely that a chosen sample will contain all of
-the definitions required by a theorem. The majority of such samples would hence
-have an empty ground truth, and thus $0$ precision and undefined recall
-\emph{independent} of what the MTE system produces! This is clearly undesirable
-as an evaluation method.
-
-Instead, we only choose samples which admit at least one of the corpus theorems.
-We could do this using rejection sampling, but it is more efficient to sample
-the \emph{theorem statements}, weighted in proportion to the number of
-definitions they reference. Once a theorem is chosen, we use the referenced
-definitions as our sample, padded up to the required size by uniformly sampling
-from the remaining definitions (we ignore theorems which reference more
-definitions than our sample size).
-
-When we form the ground truth for such samples, we're guaranteed to find at
-least one applicable theorem from the corpus (the one we chose as our starting
-point), and hence have a meaningful precision and recall.
+a theorem's dependencies. The majority of such samples would hence have an empty
+ground truth, and thus $0$ precision and undefined recall \emph{independent} of
+what the MTE tool produces! This is clearly undesirable as an evaluation
+method.
+
+Instead, we only allow a sample if it contains all dependencies of at least one
+theorem. We could do this using rejection sampling, but it is more efficient to
+pick a theorem, weighted in proportion to their number of dependencies (ignoring
+those with too many dependencies to fit in our sample). We use those
+dependencies as our sample, padded up to the required size with definitions
+chosen uniformly from the remainder.
+
+The ground truth for such samples is guaranteed to contain at least one theorem
+(the one we picked), and hence the precision and recall will depend meaningfully
+on the tool's output.
 
@@
 
-    \node (in)  [io]{Theorem Proving Benchmark (TIP)};
+    \node (in)  [io]{Theorem Proving Benchmark};
     \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
@@
 
-We have applied our methodology to evaluating QuickSpec, %TODO:cite
-a library for conjecturing equations about functions written in the Haskell
-programming language. We use the Tons of Inductive Problems (TIP) theorem
-proving benchmark to generate our corpus.
-
-\subsection{QuickSpec}
-
-We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
-is a Haskell library which takes a \emph{signature} containing Haskell functions
-and (universally quantified) typed variables, and outputs conjectured equations
-involving these terms.
-
-The QuickSpec algorithm first enumerates all well-typed expressions up to a
-certain depth (3 by default), involving the given terms. Expressions of the same
-type are assumed to be equivalent, and grouped into equivalence classes. A
-testing process then tries to disprove this assumption by instantiating each
-variable with a randomly chosen value (using the QuickCheck testing library),
-evaluating the expressions and comparing the results. The equivalence classes
-are split to separate any expressions which are observed to be different, and
-the process repeats with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal. Finally, a congruence closure algorithm is
-applied to these equations, to find a minimal set.
-
-In order to thoroughly benchmark QuickSpec, we need to automate some of the
-decisions which are normally left up to the user:
-
-\begin{itemize}
-\item We must decide what variables to include. We add three for each type that
-  appears as a function argument, except for types which have no QuickCheck data
-  generator.
-\item We must \emph{monomorphise} all types. For example, functions like
-  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
-  We need to pick a specific type in order to generate random values, so we
-  resolve this (arbitrarily) by picking
-  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
-    kind \texttt{* -> *}. If these violate some constraint, we pick a suitable
-    type non-deterministically from the module scope, or skip these variables
-    completely if none is found.}
-\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
-  so functions of multiple arguments are simulated by currying.
-  Unfortunately QuickSpec cannot compare functions for equality, so we must give
-  each function in our signature an explicit arity. Arities from zero to five
-  are supported, so we avoid partial application by picking the highest that is
-  type-correct.
-\end{itemize}
-
-To find out which functions to include in a signature, we use a plugin for the
-GHC compiler which tells us all of the functions defined in a package. We
-include all of those which type-check when passed to QuickSpec.
+We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
+tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
+benchmark as our ground truth corpus. TIP 0.2 contains 343 theorem statements,
+which depend on 182 distinct function definitions.
+
+To determine our benchmarking parameters we ran some initial tests on both tools
+for samples sized between 1 and 100, for an hour each. Most QuickSpec runs
+either finished within 200 seconds or not at all, and sizes above 20 mostly
+timed out. IsaCoSy mostly finished within 300 seconds on sample sizes up to 4,
+but by size 8 was mostly timing out; its few successes above this took thousands
+of seconds each, which we deemed infeasibly long.
+
+We decided to benchmark sample sizes up to 20, since neither tool seemed to
+perform well beyond this, and we followed the statistical ``rule of thumb'' of
+picking 30 samples for each size\footnote{Only 29 samples of size 2 were
+  analysed since one was a duplicate, due to the constraint that samples contain
+  at least one theorem's dependencies.}. This gave $20$ sizes $\times 30$
+samples $\times 2$ tools $= 1200$ runs. To keep the total time down to a few
+days we chose a timeout of 300 seconds, since that covered most of the
+successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
+diminishing returns.
+
+We normalised the conjectures generated by each tool, as well as the theorems
+from the ground truth, by numbering their free variables from left to right and
+using de Bruijn indices for bound variables. Since both tools can only output
+equations, we also sorted their left- and right-hand-sides into lexicographical
+order. Precision and recall were calculated by syntactically comparing the sets
+of generated conjectures to those of their ground truth. Other than equation
+symmetry, we specifically \emph{ignored} logical relationships between
+syntactically distinct statements, such as one equation being implied by
+another. The main reason is to avoid second-guessing the ground truth: the
+theorems' authors chose to state them that way, and it is not our place to
+decide whether an alternative form is ``just as good'' (for example, a more
+general statement is more widely applicable, but it might also be harder to
+understand).
 
@@
   We rewrite the TIP theorems (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and include these
-  functions in our samples in lieu of the raw expressions. Note that these
-  destructor wrappers are \emph{partial} functions (e.g.
-  \texttt{destructor-head} and \texttt{destructor-tail} are undefined for the
-  input \texttt{Nil}), which complicates their translation to proof assistants
-  like Isabelle.
+  forms instead of the raw constructors and destructors, and use these functions
+  in our samples in lieu of the raw expressions. Note that these destructor
+  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
+  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
+  complicates their translation to proof assistants like Isabelle.
 \end{sloppypar}
@@
 all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun} mechanisms.
+``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
+mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
+  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
+  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
+  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
+  symmetery.}
 
@@
 total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 which occur in at least one theorem
-statement (this removes ambiguity about which \emph{definitions} count as
-``interesting'' and which are just ``implementation details'' for other
+and we choose to only sample from those 182 functions which occur in at least
+one theorem statement (this removes ambiguity about which \emph{definitions}
+count as ``interesting'' and which are just ``implementation details'' for other
 definitions).
 
-TIP includes a tool to translate its definitions into Haskell and Isabelle,
-including suitable comparison functions and random data generators. We translate
-all 269 unique definitions into a single module/theory which is imported on each
-run of the MTE tools (although only those functions which appear in the current
-sample are actually explored).
-
-\subsection{Results}
+TIP comes with software to translate its definitions into Haskell and Isabelle
+code, including comparison functions and random data generators suitable for
+QuickSpec. We translate all 269 unique definitions into a single module/theory
+which is imported on each run of the tools, although only those functions which
+appear in the current sample are actually explored.
+
+\subsection{QuickSpec}
+
+We benchmarked QuickSpec version 0.9.6, which is tool written in Haskell for
+conjecturing equations involving a given set of Haskell functions and
+universally-quantified variables (a \emph{signature}).
+
+The QuickSpec algorithm first enumerates all well-typed expressions containing
+terms from the signature, up to a chosen depth (3 by default). Expressions of
+the same type are assumed to be equal, and grouped into equivalence classes. A
+testing process then tries to refute this assumption by instantiating each
+variable with a randomly chosen value (using the QuickCheck testing library),
+evaluating the expressions with those values and comparing the results. The
+equivalence classes are split up to separate any expressions which are observed
+to be different, and the process is repeated with new random values.
+
+After 500 rounds of testing, any expressions still sharing the same equivalence
+class are conjectured to be equal. Finally, a congruence closure algorithm is
+applied to these equations, to remove redundancies and special-cases.
+
+In order to thoroughly benchmark QuickSpec, we need to automate some of the
+decisions which are normally left up to the user:
+
+\begin{itemize}
+\item We must decide what variables to include. We choose to add three variables
+  for each type that appears as a function argument, except for types which have
+  no QuickCheck data generators.
+\item We must \emph{monomorphise} all types. For example, functions like
+  \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
+  element type, but we need to pick a specific type in order to know which
+  random value generator to use. We resolve this (arbitrarily) by picking
+  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
+    \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
+    \texttt{[]} (Haskell's list type constructor). If these violate some
+    type class constraint, we pick a suitable type non-deterministically from
+    those in scope during compilation; if no suitable type is found, we give up
+    and don't include the  variable.}
+\item Haskell functions are ``black boxes'', which QuickSpec can't compare
+  during its exploration process. They are also curried, always taking one
+  argument but potentially returning another function. QuickSpec lets us assign
+  an arity to each function in the signature, from 0 to 5, so we pick the
+  highest that is type-correct, since this avoids a proliferation of
+  incomparable, partially-applied functions.
+\end{itemize}
 
@@
   \centering
-  \input{time.pgf}
+  \input{quickspectime.pgf}
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
@@
     the total number of conjectures produced by each run, with failing processes
-    shown in red. Some points are shifted horizontally to prevent overlaps.
-    Standard box plots are included, with whiskers showing points at
-    1.5~$\times$~the inter-quartile range.}
+    shown in red. Points are spread out horizontally to prevent overlaps.
+    Box plots show median and quartiles, with whiskers showing points at
+    1.5$\times$~inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
   \centering
-  \input{prec.pgf}
-  \caption{Precision and recall of all runs (spread horizontally to avoid
-    overlaps). Point colour shows how many conjectures were in the ground truth
-    set for each run. Lines show the mean value (average of ratios) and the
-    shaded region is standard deviation, assuming a binomial distribution.}
+  \input{quickspecprec.pgf}
+  \caption{Precision and recall of successful QuickSpec runs (spread
+    horizontally to avoid overlaps). Point colour shows how many conjectures
+    were in the ground truth set for each run. Lines show the combined
+    proportion for each sample size (ratio of averages) and the shaded region is
+    the sample standard deviation.}
   \label{figure:quickspec_precRec}
@@
 
-We applied our sampling methodology to the 182 definitions which appear in TIP
-theorems, for even sample sizes up to 20. We chose 30 samples for each size,
-although one sample of size 2 was discarded as it was a duplicate. A timeout of
-5 minutes was imposed, since preliminary experiments showed most QuickSpec runs
-to either finish before this time, or still be running after an hour. The
-running times are shown in figure~\ref{figure:quickspec_runtimes}.
-
-The only errors we encountered were timeouts (shown in red), which occurred for
-all sample sizes but are more common for larger samples (over half the runs for
-sample size 20 timed out). Runs which did not time out mostly finished within 30
-seconds, generating few or no conjectures; those taking longer generated more
-conjectures, with the most being 116 conjectures from a sample of size 16.
-
-Equations generated by QuickSpec and those from the ground truth were normalised
-by numbering their variables and putting the left- and right-hand-sides in
-lexicographical order. These two sets were then compared syntactically for
-precision and recall, with the results shown in
-figure~\ref{figure:quickspec_precRec}.
-
-We use a binomial distribution to model precision and recall, as if each
-conjecture is drawn from a pool of wanted/unwanted results (in the case of
-precision) or found/missed ground truths (in the case of recall), with the mean
-of these distributions estimating the proportions of each pool's split.
-
-There is a clear negative trend in precision as the sample size increases,
-indicating that a more aggressive post-processing filter may be useful as we
-explore larger theories. Some small samples reached full precision, but nothing
-above size 6 exceeded 50\%.
-
-Recall shows a slight downward trend but is mostly flat, averaging between 20\%
-and 50\%. This is expected, since adding more definitions to a QuickSpec
-signature will never reduce the conjectures it will find. Those conjectures
-which QuickSpec failed to find include terms which QuickSpec cannot synthesise
-(such as conditional equations, inequalities, terms larger than the search depth
-and those with anonymous functions) as well as equations which are removed as
-``uninteresting'' by the final congruence closure step.
+The time taken to explore samples of different sizes is shown in
+\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red, and
+these occurred for all sample sizes but are more common for larger samples (over
+half the runs for sample size 20 timed out). Runs which succeeded mostly
+finished within 30 seconds, generating few or no conjectures; those taking
+longer generated more conjectures, with the most being 116 conjectures from a
+sample of size 16.
+
+We model precision and recall as simple Bernoulli processes: the generated
+conjectures (wanted or not) are like coin tosses whose bias is the precision;
+likewise, the ground truth theorems (found or not) are like coin tosses biased
+by the recall.
+
+QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
+generates monotonically more conjectures as definitions are added to a signature
+(assuming sufficient testing), it can't be finding fewer correct conjectures at
+larger sizes. This is supported by the relative flatness of the recall results.
+Rather, the number of conjectures generated is increasing at a higher rate than
+the size of the ground truth. These extra conjectures may involve the
+``padding'' definitions in a sample which don't contribute to its ground truth,
+or may be ``uninteresting'' relationships between the dependencies of different
+ground truth theorems.
+
+This indicates two potential improvements to the QuickSpec algorithm (as far as
+this benchmark is concerned). The deluge of generated conjectures could be
+filtered by another post-processing step down to a more desirable sub-set.
+Alternatively, large signatures could be filtered to remove ``padding''
+definitions (those whose conjectures are unlikely to be ``interesting'') and
+divided up into multiple smaller signatures to separate those definitions whose
+relationships are ``uninteresting''. Both of these methods could improve the
+precision, although care would be needed to avoid a large reduction in recall.
+The latter option could also improve the running time, since multiple smaller
+signatures may be faster to explore than a single large one.
+
+QuickSpec's recall is limited by two factors: the algorithm is unable to
+synthesise some theorem statements, such as conditional equations, inequalities,
+terms larger than the search depth and those containing anonymous functions.
+The congruence closure algorithm used as a post-processor may also be removing
+``interesting'' results, for example if we found an ``uninteresting'' result
+which is more general.
+
+\subsection{IsaCoSy}
+
+IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
+ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
+with the 2015 version of Isabelle.
+
+The following issues had to be overcome for the benchmarking to be successful:
+
+\begin{itemize}
+  polyrec can't be represented
+  constructors were being included
+  proofs had to be skipped
+  partial functions needed adding to the constraints
+\end{itemize}
+
+\begin{figure}
+  \centering
+  \input{isacosytime.pgf}
+  \caption{Running times of IsaCoSY on theories sampled from TIP. Each point
+    is an individual run, with 30 per size except for 2 (29). Colour indicates
+    the total number of conjectures produced by each run, with failing processes
+    shown in red (timeouts and out of memory). Points are spread out
+    horizontally to prevent overlaps. Box plots show median and quartiles, with
+    whiskers showing points at 1.5$\times$~inter-quartile range.}
+  \label{figure:isacosy_runtimes}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \input{isacosyprec.pgf}
+  \caption{Precision and recall of all successful IsaCoSy runs (spread
+    horizontally to avoid overlaps). Lines show the combined proportion for each
+    sample size (ratio of averages) and the shaded region is the sample standard
+    deviation.}
+  \label{figure:isacosy_precRec}
+\end{figure}
+
+The most striking result is how rapidly IsaCoSy's running time increases with
+sample size: from size 8 upwards, all runs failed (most timed out, a few aborted
+early due to running out of memory).
+
+\subsection{Comparison}
+
+We compare the performance of QuickSpec and IsaCoSy using a similar protocol to
+the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We differ
+by timing a wider variety of inputs: running many samples once, rather than a
+few samples many times and taking the medians. Since each sample is run through
+both tools, we can also gain statistical power by using the \emph{paired} form
+of the Wilcoxon signed-rank test.
 

### 2018-03-09 TEBenchmarkPaper/article.tex
Prose

@@
 
-The following issues had to be overcome for the benchmarking to be successful:
+The following issues had to be overcome to make our benchmark applicable to
+IsaCoSy:
 
 \begin{itemize}
-  polyrec can't be represented
-  constructors were being included
-  proofs had to be skipped
-  partial functions needed adding to the constraints
+\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
+  encoded in Isabelle. We strip out this type and the functions which depend on
+  it before translating.
+\item When using a type in an IsaCoSy signature, that type's constructors will
+  automatically be included in the exploration. Since those constructors will
+  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
+  even those may not be present in the current sample) this will unfairly reduce
+  the tool's precision. To avoid this, we add a post-processing step which
+  replaces all occurrences of a constructor with the corresponding wrapper, then
+  discards any conjectures which involve functions other than those in the
+  current sample.
+\item Since Isabelle is designed for theorem proving rather than programming, it
+  requires every definition to be accompanied by proofs of exhaustiveness and
+  termination. These are difficult to generate automatically, and don't exist in
+  the case of partial functions like destructor wrappers. Hence we use the
+  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
+  the \texttt{sorry} keyword.
+\item Partial functions cause problems during exploration, since they can throw
+  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
+  pre-populating IsaCoSy's constraint set with these undefined expressions
+  (for example \texttt{head Nil}), hence preventing IsaCoSy from ever generating
+  them.
 \end{itemize}
@@
 sample size: from size 8 upwards, all runs failed (most timed out, a few aborted
-early due to running out of memory).
+early due to running out of memory). As we found with our preliminary testing,
+this increase appears exponential, so even large increases to the timeout would
+not produce many more successes.
 
@@
 
-We compare the performance of QuickSpec and IsaCoSy using a similar protocol to
-the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We differ
-by timing a wider variety of inputs: running many samples once, rather than a
-few samples many times and taking the medians. Since each sample is run through
-both tools, we can also gain statistical power by using the \emph{paired} form
-of the Wilcoxon signed-rank test.
+We compare the running times of QuickSpec and IsaCoSy using a similar protocol
+to the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We
+differ by timing a wider variety of inputs: running many samples once, rather
+than a few samples many times and taking the medians. Since each sample is run
+through both tools, we can also gain statistical power by using the
+\emph{paired} form of the Wilcoxon signed-rank test. We compare each sample size
+is compared separately, to find the probability that QuickSpec is significantly
+faster.
 

### 2018-03-09 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Handle even and odd data

@@ -4,39 +4,34 @@ import os
 import sys
 msg = lambda x: sys.stderr.write(x + '\n')
 
-def readVar(var):
-    val = os.getenv(var)
-    assert val is not None, 'No {} env var given'.format(var)
-    return val
-
-sources = {
-    'isacosy'   : readVar(  'ISACOSY_DATA'),
-    'quickspec' : readVar('QUICKSPEC_DATA')
-}
-del(readVar)
-
 msg('Reading input')
 
 from json import load
 
-def readInput(source, path):
-    with open(sources[source], 'r') as f:
-        return load(f)['results'][path]['result'][0]
+def readInput(sys):
+    var = os.getenv(sys + 'Data')
+
+    assert var is not None, 'No {} env var given'.format(sys)
+
+    with open(var, 'r') as f:
+        return load(f)
 
 data = {
-    'isacosy'   : readInput('isacosy'  ,   'benchmarks.track_data'),
-    'quickspec' : readInput('quickspec', 'quickspectip.track_data')
+    'isacosy'   : readInput('isacosy'),
+    'quickspec' : readInput('quickspec')
 }
 del(readInput)
 
 msg('Normalising data')
 
-# Extract timeout from one of the QuickSpec runs, since IsaCoSy didn't store it
-timeout = data['quickspec'].values()[0]['reps'].values()[0]['timeout']
+# Extract timeout from one of the QuickSpec runs, since IsaCoSy may not have
+# stored it
+timeout = data['quickspec'].values()[0].values()[0]['timeout']
 
-def makeHashable(x):
-        '''Convert lists and dictionaries into tuples, recursively, so we can
-        turn them into sets for comparison.'''
+def normalise(data):
+    def makeHashable(x):
+        """Convert lists and dictionaries into tuples, recursively, so we can
+        turn them into sets for comparison."""
         xType    = type(x)
         reducers = {
             type([]) : lambda rest, elem: rest +  (makeHashable(elem),),
@@ -45,83 +40,14 @@ def makeHashable(x):
         }
         return reduce(reducers[xType], x, ()) if xType in reducers else x
 
-def normalise(data):
     for size in data:
-        # Unwrap useless 'reps' wrapper
-        if 'reps' in data[size]:
-            data[size] = data[size]['reps']
-
         for rep in data[size]:
-            # Remove more unnecessary wrappers
-            for key in ['result', 'analysis']:
-                if key in data[size][rep]:
-                    data[size][rep] = dict(  data[size][rep],
[…]

### 2018-03-12 TEBenchmarkPaper/article.tex
Update total runtime estimate

@@
 To determine our benchmarking parameters we ran some initial tests on both tools
-for samples sized between 1 and 100, for an hour each. Most QuickSpec runs
+for a few samples sized between 1 and 100, for an hour each. Most QuickSpec runs
 either finished within 200 seconds or not at all, and sizes above 20 mostly
@@
 We decided to benchmark sample sizes up to 20, since neither tool seemed to
-perform well beyond this, and we followed the statistical ``rule of thumb'' of
-picking 30 samples for each size\footnote{Only 29 samples of size 2 were
-  analysed since one was a duplicate, due to the constraint that samples contain
-  at least one theorem's dependencies.}. This gave $20$ sizes $\times 30$
-samples $\times 2$ tools $= 1200$ runs. To keep the total time down to a few
-days we chose a timeout of 300 seconds, since that covered most of the
-successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
-diminishing returns.
+perform well beyond this. Our results will be analysed by Speedup-test protocol,
+which follows the statistical ``rule of thumb'' of treating sample sizes
+$\leq$ 30 as ``small'', so we pick 31 samples of each size in order to cross
+this (arbitrary) threshold\footnote{Only 14 samples of size 1 were analysed, and
+  30 of size 2, since the rest were duplicates. This is due to the constraint
+  that samples contain at least one theorem's dependencies.}. This gave $20$
+sizes $\times 31$ samples $\times 2$ tools $= 1240$ runs. To keep the total time
+down to a few days we chose a timeout of 300 seconds, since that covered most of
+the successful QuickSpec and IsaCoSy results we saw, and longer times gave
+rapidly diminishing returns.
 

### 2018-03-12 TEBenchmarkPaper/article.tex
Better "second guessing" phrasing

@@
 syntactically distinct statements, such as one equation being implied by
-another. The main reason is to avoid second-guessing the ground truth: the
-theorems' authors chose to state them that way, and it is not our place to
-decide whether an alternative form is ``just as good'' (for example, a more
-general statement is more widely applicable, but it might also be harder to
-understand).
+another. Whilst logically sound, second-guessing the ground truth in this way
+would harm the other aspects which we care about which influence interestingness
+(for example, a more general statement is more widely applicable, but it might
+also be harder to understand).
 

### 2018-03-12 TEBenchmarkPaper/article.tex
IsaCoSy dies even earlier

@@
 The most striking result is how rapidly IsaCoSy's running time increases with
-sample size: from size 8 upwards, all runs failed (most timed out, a few aborted
+sample size: from size 7 upwards, all runs failed (most timed out, a few aborted
 early due to running out of memory). As we found with our preliminary testing,

### 2018-03-12 TEBenchmarkPaper/article.tex
IsaCoSy interpretation

@@
 this increase appears exponential, so even large increases to the timeout would
-not produce many more successes.
+not produce many more successful observations. A significant amount of IsaCoSy's
+time (around 50 seconds) was spent loading the generated theory (containing all
+distinct datatypes and functions from TIP); this overhead is unfortunate, but
+since it is constant across all sample sizes it does not affect our conclusions.
+
+With so few successful datapoints it is difficult to make strong claims about
+the precision/recall quality of IsaCoSy's output. It is clear from the quantity
+of non-zero proportions that IsaCoSy is capable of discovering interesting
+conjectures, and the graphs appear to follow the same shapes as those of
+QuickSpec (decreasing precision, flat recall), although the error margins are
+too wide to be definitive.
 

### 2018-03-12 TEBenchmarkPaper/article.tex
More Wilxocon

@@
 to the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We
-differ by timing a wider variety of inputs: running many samples once, rather
-than a few samples many times and taking the medians. Since each sample is run
-through both tools, we can also gain statistical power by using the
-\emph{paired} form of the Wilcoxon signed-rank test. We compare each sample size
-is compared separately, to find the probability that QuickSpec is significantly
-faster.
+treat each of our sample sizes as a separate ``benchmark'', but deviate from
+their approach by using a different input (sample) for each run. This reduces
+the sampling bias, but also reduces the accuracy of our confidence intervals.
+We used version 2 of the Speedup-Test \texttt{R} implementation, which we
+patched to use the \emph{paired} form of the (one-sided) Wilcoxon signed-rank
+test~\cite{wilcoxon1945individual}, since this increases our statistical power
+given that each sample is run through both tools. We use Pratt's
+method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
+out.
+
+For each sample size (``benchmark'') the Speedup-Test protocol compares the
+distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
+significantly different (with $\alpha < 0.05$), then the signed-rank test is
+only performed if more than 30 samples are available. This was the case for all
+sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
+two speedups were not deemed statistically significant. For all other sample
+sizes QuickSpec was found to be faster, with 95\% confidence. The speedup for
+each size is given in table~\ref{FIXME}. The 95\% confidence interval for the
+proportion of sample sizes where QuickSpec's median running time is faster than
+IsaCoSy's is 66.9\% to 98.2\%.
 

### 2018-03-12 TEBenchmarkPaper/article.tex
Add speedup table

@@
 \usepackage{array}
+\usepackage{csvsimple}
 \usepackage{hhline}
@@
 two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be faster, with 95\% confidence. The speedup for
-each size is given in table~\ref{FIXME}. The 95\% confidence interval for the
-proportion of sample sizes where QuickSpec's median running time is faster than
-IsaCoSy's is 66.9\% to 98.2\%.
+sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
+giving a 95\% confidence interval for the proportion of (median) speedups to be
+$[0.669;0.982]$. The magnitude of each speedup is is given in
+table~\ref{table:speedups}.
+
+\begin{table}
+  \begin{tabular}{ |r|l| }
+    \hline
+    \bfseries Sample Size & \bfseries Speedup
+    \csvreader[]{speedups.csv}{}
+    {\\\hline\csvcoli&\csvcolii} \\
+    \hline
+  \end{tabular}
+  \caption{Speedup when using QuickSpec compared to IsaCoSy for each sample
+    size. Those which aren't statistically significant with 95\% confidence are
+    marked as ``-''. These are conservative estimates, since the difference is
+    bounded by the timeout.}
+  \label{table:speedups}
+\end{table}
 

### 2018-03-14 TEBenchmarkPaper/article.tex
Background cleanup

@@
 \newcommand{\function}{\rightarrow}
+
 \newcommand{\Zero}{\text{Z}}
 \newcommand{\Succ}{\text{S}}
+\newcommand{\plus}{\text{plus}}
+\newcommand{\mult}{\text{times}}
+
 \newcommand{\List}{\text{List}}
@@
 \item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec MTE system, and a
-  discussion of the results.
+\item Application of our methodology to the QuickSpec and IsaCoSy MTE systems,
+  and a comparison and discussion of the results.
 \end{itemize}
@@
 methodologies. We explain our proposal for a more general benchmark in section
-\ref{sec:proposal} and section \ref{sec:application} shows the results when
+\ref{sec:proposal} whilst section \ref{sec:application} shows the results when
 applied to existing MTE tools. Analysis of these results is given in section
-\ref{sec:discussion} and concluding remarks in \ref{sec:conclusion}.
+\ref{sec:discussion} and concluding remarks in section \ref{sec:conclusion}.
 
@@
 $\Map$ call into multiple independent calls dealing with different sections of a
-list. Such rules are useful optimisations since each call can be evaluated in
-parallel, leading to the ``map/reduce'' programming paradigm.
+list. Such rules can be useful optimisations, for example in a distributed
+context where since each call can be evaluated in parallel (this is the basis of
+the ``map/reduce'' programming paradigm).
 
@@
 it comes to more complex statements.  Colton \etal{} give a survey of systems for
-theory formation and exploration, and their notions of ``interestingness'' for
-concepts and conjectures, identifying six notions in
-total~\cite{colton2000notion}. These notions, as applied to conjectures, are:
-
-{\bf Empirical plausibility}, which checks whether a property holds
-across some specific examples. This is especially useful for avoiding
-false conjectures, without resorting to a full proof search.
-
-{\bf Novelty} is whether a conjecture, or one isomorphic or more
+theory formation and exploration, and their associated notions of
+``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
+notions are identified, which are applied to conjectures as follows:
+
+{\bf Empirical plausibility} checks whether a property holds across some
+specific examples. This is especially useful for avoiding false conjectures,
+without resorting to a full proof search.
+
+{\bf Novelty} depends on whether a conjecture, or one isomorphic or more
 general, has already been seen.
@@
     system~\cite{braquehais2017speculate}. The interestingness measures are
+    to include four more recent systems. The interestingness measures are
     \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
@@
 
-The latter systems all use testing to check for counterexamples, which ensures
-results are empirically plausible and applicable (conditions are satisfiable,
-types are inhabited, etc.). They also use a small-to-large search order,
-ensuring more easily-comprehensibile conjectures are explored first.
-
-IsaCoSy ensures novelty using a constraint system, such that no generated
-expression can be simplified using a previously-found equation. The other
-systems use post-hoc filters based on unification and term rewriting to discard
-expressions or equations which are special-cases of some other, more general
-result.
-
-IsaCoSy is able to look for commonly desirable patterns, such as commutativity
-and associativity, which increases the utility of its output, whilst IsaScheme
-is completely based around such pattern-instantiation.
+The latter systems are described in more detail in section~\ref{section:mte},
+but all use testing to check for counterexamples, which ensures results are
+empirically plausible and applicable (conditions are satisfiable, types are
+inhabited, etc.). They also use a small-to-large search order, ensuring more
+easily-comprehensibile conjectures are explored first.
+
+IsaCoSy ensures novelty using a constraint system to prevent special-cases being
+generated, whilst the others use post-hoc filters based on unification and term
+rewriting to achieve the same result. IsaCoSy is able to look for commonly
+desirable patterns, such as commutativity and associativity, which increases the
+utility of its output, whilst IsaScheme is completely based around such
+pattern-instantiation.
 
@@
 \subsection{Mathematical Theory Exploration}
+\label{section:mte}
 
@@
 
-Recent systems developed in this area include MATHsAiD~\cite{roy},
-IsaScheme~\cite{MontanoRivas2011} and
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}. The goal of the
-MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses, Axioms and
-Definitions) project is to build a tool which takes in a set of axioms, concept
-definitions and a logic and applies its inference rules to reason from the
-axioms to theorems. The motivation is to produce a tool which will help
-mathematicians to explore the consequences of a set of axioms or a particular
-concept definition. One of the main challenges of the project has been to
-automatically evaluate interestingness: to distinguish important theorems from
-results which, although they follow from a set of axioms, are of little
-mathematical interest.
+Recent systems developed in this area include IsaScheme~\cite{MontanoRivas2011},
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
+QuickSpec~\cite{QuickSpec} and Speculate~\cite{braquehais2017speculate}
+(included in table~\ref{table:colton}) as well as MATHsAiD~\cite{roy}.
+
+The goal of the MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses,
+Axioms and Definitions) project is to build a tool which takes in a set of
+axioms, concept definitions and a logic and applies its inference rules to
+reason from the axioms to theorems. The motivation is to produce a tool which
+will help mathematicians to explore the consequences of a set of axioms or a
+particular concept definition. One of the main challenges of the project has
+been to automatically evaluate interestingness: to distinguish important
+theorems from results which, although they follow from a set of axioms, are of
+little mathematical interest.
 
@@
 scheme are instantiated automatically and this drives the invention process.
-For instance, in the theory of natural numbers, given the concepts of successor
-($suc$), addition ($+$) and zero ($0$), IsaScheme can use the following scheme
-to invent the concept of multiplication:
+For instance, in the theory of natural numbers, given the concepts of zero
+($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
+following scheme to invent the concept of multiplication:
 
@@
 \]
+
 where the existentially quantified variable $f$ stands for the new
-function to be defined in terms of the variables g, h, i and j. Within
+function to be defined in terms of the variables $g$, $h$, $i$ and $j$. Within
 the theory of natural numbers, IsaScheme instantiates this scheme with
-$\sigma_1 = \{g \mapsto 0, h \mapsto (\lambda x.0), i \mapsto suc, j
-\mapsto +\}$. In this example, $f \mapsto *$ (multiplication), since:
+$\sigma_1 = \{g \mapsto \Zero, h \mapsto (\lambda x.\Zero), i \mapsto \Succ, j
+\mapsto \plus\}$. In this example, $f \mapsto \mult$ (multiplication), since:
 
 \begin{align*}
-0*y  &= 0 \\
-suc(x) * y &=y + (x*y).
+\mult(\Zero, y)  &= \Zero \\
+\mult(\Succ(x), y) &= \plus(y, \mult(x, y))
 \end{align*}
+
 The new multiplication function $f$ can itself be used to instantiate
@@
 %IsaCoSy
-The IsaCoSy system (Isabelle Conjecture Synthesis) is a program for
-inductive theory formation, which synthesises conjectures from the
-available constants and free variables. Only terms which do not
-contain subterms that can be rewritten by a set of rewrite rules can
-be synthesised.  Conjectures are tested by sending them to a
-counterexample checker and, if no counterexamples are found, then sent
-to IsaPlanner which attempts to prove them.
-
-% Discussion.
+The IsaCoSy system (Isabelle Conjecture Synthesis) conjectures equations by
+enumerating expressions involving a given set of (typed) constants and free
+variables (a signature). A constraint solver forbids certain (sub)expressions
+from being synthesised, and these constraints are extended whenever a new
+conjecture is generated, to avoid generating any special-cases of this
+conjecture in the future. Conjectures are tested by sending them to the
+QuickCheck counterexample finder and, if no counterexamples are found, then
+sent to IsaPlanner which attempts to prove them.
+
+%QuickSpec
+QuickSpec emerged from work on the QuickCheck software testing framework for the
+Haskell programming language, and has been used to generate conjectures for the
+subsequent HipSpec and Hipster MTE systems. Like IsaCoSy, a signature is
+provided and terms are enumerated. Rather than a constraint system, QuickSpec
+version 1 collects expressions of the same type into equivalence classes,
+assuming them to be equal. QuickCheck is used to test this assumption by
+instantiating each variable with a randomly chosen value, evaluating the
+resulting expressions and comparing them. Any equivalence class whose elements
+don't compare equal are split up to separate those observably-different
+expressions, and the process is repeated with new random values.
+
+After 500 rounds of testing, any expressions still sharing the same equivalence
+class are conjectured to be equal for all values of their variables. Finally, a
+congruence closure algorithm is applied to these equations, to remove
+redundancies and special-cases.
+
+QuickSpec version 2 replaces this two-step process with a single, iterative
+algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
+Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
+As expressions are enumerated, they are simplified using these rules and
+discarded if equal to a known expression. If not, QuickCheck tests whether the
+new expression can be distinguished from the known expressions through random
+testing: those which can are added to the set of known expressions. Those which
+cannot be distinguished are conjectured to be equal, and the rewrite rules are
+updated.
+
+%Speculate
+Speculate is a Haskell tool which operates in a similar way to QuickSpec
+version 2. In addition to equations, Speculate uses the laws of total orders and
+Boolean algebra, along with testing (via LeanCheck), to conjecture inequalities
+and conditional relations between expressions, such as (for
+$x, y \in \mathbb{Z}$):
+
+\begin{equation*}
+  \begin{aligned}
+    x + y \leq \lvert x \rvert + \lvert y \rvert \\
+    x \leq 0 \rightarrow x + \lvert x \rvert = 0
+  \end{aligned}
+\end{equation*}
 
@@
   distinction is not relevant for our purposes.} As can be seen above,
-some of these historical differences are disappearing, and current
-systems such as MATHsAiD~\cite{roy}, IsaScheme~\cite{MontanoRivas2011}
-and IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation} are bridging the
-methodological gap between ATF and MTE. This is particularly true in terms of
-fully-automatic/user-interactive and the domain-specific/general
-aspects.
+some of these historical differences are disappearing, and recent
+implementations are bridging the methodological gap between ATF and MTE. This is
+particularly true in terms of fully-automatic/user-interactive and the
+domain-specific/general aspects.
 

### 2018-03-15 TEBenchmarkPaper/article.tex
Fix up table spacing
Fix table border
Lots of additions and changes

@@
 
-% Look carefully at each system's (paper's) eval section
+% Look carefully at each tool's (paper's) eval section
 
@@
 
-% Lay out the field of TE: motivation, systems, evaluation, problems...
+% Lay out the field of TE: motivation, tools, evaluation, problems...
 
@@
 defining a standard, unambiguous benchmarking approach with which to compare
-the conjecture generation of MTE systems. Our contributions include:
+the conjecture generation of MTE tools. Our contributions include:
 
@@
 \item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE systems,
+\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
   and a comparison and discussion of the results.
@@
 approaches agree that simple tautologies are ``uninteresting'', but differ when
-it comes to more complex statements.  Colton \etal{} give a survey of systems for
+it comes to more complex statements.  Colton \etal{} give a survey of tools for
 theory formation and exploration, and their associated notions of
@@
 {\bf Applicability} depends on the number of models in which a
-conjecture holds. The system of \Bagai{} conjectures the non-existence of
-objects, and hence favours statements with zero applicability. Other systems
-treat applicability as a positive aspect: the more applicable the statement, the
-more interesting it is.
+conjecture holds. The tool of \Bagai{} conjectures the non-existence of
+objects, and hence favours statements with zero applicability. Other tools treat
+applicability as a positive aspect: the more applicable the statement, the more
+interesting it is.
 
 {\bf Comprehensibility} depends on the complexity of a statement. Simpler
-statements are considered more interesting, hence many of the systems explore
+statements are considered more interesting, hence many of the tools explore
 simpler/smaller statements before complex/larger ones, to more efficiently find
@@
     \multirow{2}{*}{\textbf{Conjecture Types}}             &
-    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
+    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
     \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
@@
     \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
-    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
+    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
     \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
@@
   \end{tabular}
-  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
-    to include those from~\cite{claessen2013automating} (QuickSpec is the
-    conjecture generation component of HipSpec) and the similar Speculate
-    system~\cite{braquehais2017speculate}. The interestingness measures are
-    to include four more recent systems. The interestingness measures are
+  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
+    to include four more recent tools. The interestingness measures are
     \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
@@
 We summarise the criteria in Table 1 and state which were used in key
-ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
-on this survey (the latter is Colton's own system). We extend this analysis to
-some more recent systems in the remaining rows (based on our understanding of
+ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
+on this survey (the latter is Colton's own tool). We extend this analysis to
+some more recent tools in the remaining rows (based on our understanding of
 their function): QuickSpec~\cite{QuickSpec},
@@
 
-The latter systems are described in more detail in section~\ref{section:mte},
+The latter tools are described in more detail in section~\ref{section:mte},
 but all use testing to check for counterexamples, which ensures results are
@@
 This diversity of approaches makes it difficult to compare the existing
-evaluations of these systems directly. In particular, evaluation methods created
-for one system might not make sense for another, due to assumptions made about
+evaluations of these tools directly. In particular, evaluation methods created
+for one tool might not make sense for another, due to assumptions made about
 the algorithms' operation. For example, the novelty filters used by QuickSpec
@@
 With such ambiguous and varied goals, approaches and assessment criteria it is
-difficult to compare MTE systems in a quantitative way, and hence to form some
+difficult to compare MTE tools in a quantitative way, and hence to form some
 measure of ``progress'' for the field. Our proposed benchmarking methodology
@@
 myriad goals of existing researchers, whilst being generally applicable to
-systems of this type.
+tools of this type.
 
@@
 into context both historically, and with respect to related approaches
-such as in Automated Theory Formation systems.
+such as in Automated Theory Formation tools.
 
@@
 
-The first automated mathematical system, the mechanical calculator
+The first automated mathematical tool, the mechanical calculator
 (known as the Pascaline), was an adding machine that could perform
@@
 by repetitions, and was conceived by Pascal in 1642 while reorganising
-tax revenues~\cite{d'ocagne}. Subsequent early systems include
+tax revenues~\cite{d'ocagne}. Subsequent early tools include
 M\"uller's universal calculating machine in 1784, which he built for
@@
 was an engineer with an interest in mathematical knowledge. All three
-systems were conceived as an aid to mathematicians, as well as
+tools were conceived as an aid to mathematicians, as well as
 scientists, accountants and surveyors.
@@
 We limit our discussion to applications in mathematics, although there are
-similar approaches being applied to the task of \emph{scientific} discovery.
+similar approaches being applied to the task of Computational \emph{Scientific}
+Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}.
 Scientific knowledge relies on inductive reasoning and experimental testing,
 which (in principle) are not so important to a mathematical theory. However,
-these techniques \emph{do} play an important role in many existing formation and
-exploration tools: in particular, inductive reasoning allows a system to
-conjecture statements which may be too difficult for it to deductively prove;
-whilst experimental testing, e.g. by checking universally-quantified statements
-against some specific values, is a simple way to quickly discard many obvious
-falsehoods generated by powerful but imperfect algorithms (as opposed to, say,
-limiting ourselves to verified theorem generators).
+these techniques \emph{do} play an important role in many mathematical theory
+formation and exploration tools: in particular, inductive reasoning allows a
+tool to conjecture statements which may be too difficult for it to deductively
+prove; whilst experimental testing, e.g. by checking universally-quantified
+statements against some specific values, is a simple way to quickly discard many
+obvious falsehoods generated by powerful but imperfect algorithms (as opposed
+to, say, limiting ourselves to verified theorem generators).
 
 \subsection{Automated Theory Formation}
+\label{section:atf}
 
@@
 
-% 4. example systems - AM, HR, ...
+% 4. example tools - AM, HR, ...
 When Lenat built the AM system~\cite{lenat:77}, there were systems
@@
 seem to be that in MTE:
+
 \begin{itemize}
@@
 
-There are three aspects to consider when we evaluate a conjecture generation
-system:
+There are three aspects to consider when evaluating a tool for conjecture
+generation:
 
 \begin{enumerate}
-\item The quality of each conjecture, which is the ``interestingness'' discussed
-  above.
-\item The quality of the \emph{set} of conjectures produced, used for relative
-  interestingness criteria like novelty, as well as to assess consistency.
-\item Performance of the system, to find the balance struck between output
+\item The quality of each individual conjecture, which we will refer to as their
+  ``interestingness''.
+\item The quality of the \emph{set} of generated conjectures, which is important
+  for criteria like novelty, as well as to assess consistency.
+\item Resources used by the tool, to find the balance struck between output
   quality and time taken.
@@
 
-We also make the assumption that the system is intended to be \emph{general},
-and hence isn't hard-coded or fine-tuned to only perform well in a narrow
-domain. Examples where this does \emph{not} hold include the natural number
-heuristics of AM (discussed above), as well as machine learning models trained
-on some specific dataset.
-
-If we limit ourselves to exploring higher-order theories with inductive types
-(a domain useful for analysing software), we do find a direct comparison of
-three systems~\cite{claessen2013automating}: HipSpec, IsaCoSy and IsaScheme.
-QuickSpec, shown in table \ref{table:colton}, is the conjecture-generating
-component of HipSpec.
-
-This comparison uses \emph{precision/recall analysis}: a particular set of
-definitions is chosen, such as those in figure \ref{figure:list_theory}, and a
-\emph{ground truth} is chosen as the 'ideal' set of conjectures which we would
-like an MTE system to produce from these definitions.
-
-Each system is run on these definitions, and their outputs are compared to the
-ground truth. To score 100\% on precision and recall, an MTE system must output
-all of the conjectures which appear in the ground truth, and nothing else:
+We make the assumption that each tool is intended to be \emph{general}, and
+hence isn't hard-coded or fine-tuned to only perform well in a narrow domain.
+Examples where this does \emph{not} hold include the natural number heuristics
+of AM discussed in section~\ref{section:atf}, as well as machine learning models
+trained on some specific dataset (such as this benchmark).
+
+Evaluation of new tools and techniques for conjecture generation tends to be
+\emph{ad hoc} and hence difficult to compare, whilst comparative surveys (such
+as Colton's) are descriptive and qualitative rather than providing direct
+numerical measurements.
+
+One quantitative approach which has gained traction is \emph{precision/recall
+  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
+providing a direct comparison of these
+tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
+  reports prior results from the other tools, rather than reproducing them.
+  This makes the numbers less comparable, since each tool was tested with
+  slightly different definitions (e.g. whether or not the exponential and
+  maximum functions were included).} To measure precision and recall we first
+choose a set of definitions, such as those in figure \ref{figure:list_theory},
+and a \emph{ground truth} to treat as the 'ideal' set of conjectures we would
+like a tool to produce for those definitions. Each tool is run and their outputs
+are compared to the ground truth. To score 100\% on precision and recall, all of
+the conjectures which appear in the ground truth must be generated, and nothing
+else:
 
 \begin{itemize}
-\item \emph{Precision} is the proportion of a system's output which appears in
-  the ground truth. This penalises systems which output large numbers of
-  conjectures in the hope that some turn out to be ``good''.
+\item \emph{Precision} is the proportion of a tool's output which appears in
+  the ground truth. This penalises overly-liberal tools which output large
+  numbers of conjectures in the hope that some turn out to be ``good''.
 \item \emph{Recall} is the proportion of the ground truth which appears in the
-  system's output. This penalises systems which overly restrict their output to
-  avoid generating ``bad'' conjectures.
+  tool's output. This penalises overly-conservative tools which generate very
+  few conjectures as a way to avoid ``bad'' ones.
 \end{itemize}
 
-Precision and recall fulfil our second requirement, by directly measuring the
-quality of the \emph{set} of conjectures. Their ``interestingness'' criterion is
-indirect: asking only whether or not a conjecture appears in the ground truth.
-A satisfactory solution to our first requirement hence requires careful choice
-of the ground truth set.
-
-This particular analysis takes its ground truths from the standard library of
-the Isabelle theorem prover, with one involving the natural numbers, and the
-other involving the list functions from figure \ref{figure:list_theory}. Since
-the library authors have gone to the effort of stating, proving and including
-these theorems in every copy of their software, this is a good indication that
-they are useful or important.
-
-One problem with this choice is the small size of these libraries. For example,
-the benchmark based on Isabelle/HOL's theory of natural numbers given
-in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains only 4
-definitions and a ground truth of 12 theorems. Whilst such benchmarks allow
-objective comparisons between different approaches, their narrow scope doesn't
-provide much indication of performance in different, especially \emph{novel},
-domains.
+Precision and recall act as quality tests for sets of conjectures, fulfilling
+one of our requirements. They also provide a coarse quality assessment for
+individual conjectures: either a conjecture appears in the ground truth set or
+it doesn't.
+
+The analyses in~\cite{claessen2013automating} took their definitions and ground
+truths from the standard library of the Isabelle theorem prover: since the
+library authors have gone to the effort of stating, proving and including these
+theorems in every copy of their software, this is a good indication that they
+are interesting.
+
+Two tests were performed: one on a theory of natural numbers and one on a theory
+of lists (whose definitions are shown in figure~\ref{figure:list_theory}). These
+tests are limited both in number (it is difficult to draw general conclusions
+from two datapoints) and by the small size of these theories (e.g. the natural
+number theory from~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
+only 4 definitions and 12 theorems).
+
+We believe that this is a viable approach to drawing meaningful, objective
+conclusions about such tools, but larger data sets are needed to judge
+performance in different, especially \emph{novel}, domains (which, after all, is
+the purpose of such tools). In addition, we might be concerned that such
+``popular'' theories may be unrepresentative of typical usage. For example, we
+would not expect the library functions in a typical verification problem to be
+as mathematically rich as number theory.
 
@@
 Our main contribution is a benchmarking methodology, shown in figure
-\ref{figure:flow_chart}, which generates both a large definition/ground-truth
-corpus (an order of magnitude larger than previous work) and a scalable,
-statistical approach to evaluating MTE systems using this corpus. We follow a
-precision/recall approach similar to the work described above, with the main
-difference being the source of definitions and ground truths: we take existing
-problem sets designed for automated \emph{theorem proving}, and adapt their
-content for use in the \emph{theory exploration} setting.
+\ref{figure:flow_chart}, which both generates a large definition/ground-truth
+corpus, and provides a scalable, statistical approach to evaluating MTE tools
+using this corpus. We follow a precision/recall approach similar to prior work,
+with the main difference being the source of definitions and ground truths: we
+take existing problem sets designed for automated theorem proving, and adapt
+their content for use in the conjecture generation setting.
 
@@
 possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for theory exploration, as their definitions and theorems can be used as a
-corpus, in the same way that Isabelle libraries have been used in the past.
+for MTE, as their definitions and theorems can be used as a corpus in the same
+way that Isabelle libraries have been used in the past.
 
@@
   theorem statement but not separately \emph{defined}.
-\item Definitions should be \emph{strongly typed}, such that they can be
-  translated into the native language of each MTE system (in this work, that
-  includes Haskell\iffalse FIXME and Isabelle\fi).
-\item The problem set should be relevant to the desired domain. In our case, we
-  desire higher-order functions and inductively defined types, which rules out
-  first-order languages/logics (such as TPTP~\cite{sutcliffe2009tptp}).
+\item Definitions should be translatable into a form suitable for the MTE tools
+  under study. Our application translates to Haskell and Isabelle, which also
+  benefits from having definitions be strongly typed.
+\item The problem set should be relevant to the desired domain. Our focus on
+  functional programming requires higher-order functions and inductively defined
+  datatypes, which rules out first-order languages/logics (such as
+  TPTP~\cite{sutcliffe2009tptp}).
 \item Since it will act as our ground truth, the problem set should ideally
@@
 referenced by the theorems from the theorem statements themselves. The former
-will be used as input to the MTE systems under evaluation, whilst the latter
-form a ground truth corpus against which to compare the MTE output.
+will be used as input to the tools, whilst the latter form a ground truth corpus
+against which to compare the output.
 
@@
 \texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-system to conjecture commutativity and associativity for \emph{both} functions,
-or for \emph{neither} function, since they're logically equivalent. Yet a
-na\"ive precision/recall analysis would treat commutativity of \texttt{add} and
-associativity of \texttt{plus} as \emph{uninteresting}, since they don't appear
-in the ground truth.
+tool to either conjecture commutativity and associativity for \emph{both}
+functions, or for \emph{neither} function, since they're logically equivalent.
+Yet a na\"ive precision/recall analysis would treat commutativity of
+\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
+they don't appear in the ground truth.
 
@@
 We could, in theory, send these de-duplicated definitions straight into an MTE
-system, and use the updated theorem statements as the ground truth for analysis.
+tool and use the updated theorem statements as the ground truth for analysis.
 However, this would cause two problems:
@@
   infer performance \emph{in general}.
-\item It is impractical to run any existing MTE system on ``large'' inputs,
+\item It is impractical to run existing MTE tools on ``large'' inputs,
   containing more than a few dozen definitions.
@@
 To solve both of these problems we instead \emph{sample} a subset of
-definitions. Given a sample size, we choose a subset of that many definitions,
-and provide those definitions as input to our MTE tool. We generate a
-corresponding ground truth by selecting those theorems from the corpus which
-``depend on'' (contain references to) \emph{only} the definitions in that
-sample. Transitive dependencies aren't required (e.g. a theorem which only
-references \texttt{times} does not depend on \texttt{plus}, even if
-\texttt{plus} occurs in the definition of \texttt{times}).
+definitions.\footnote{This could be done randomly, but for reproducibility we
+  use a deterministic order (based on cryptographic hashes, to discourage
+  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
+and provide only those as input to the tool. We generate a corresponding
+ground truth by selecting those theorems from the corpus which ``depend on''
+(contain references to) \emph{only} the definitions in that sample. Transitive
+dependencies aren't required (e.g. a theorem which only references
+\texttt{times} does not depend on \texttt{plus}, even if \texttt{plus} occurs in
+the definition of \texttt{times}).
 
 Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which, as detailed above,
-is advantageous) makes it less likely that a chosen sample will contain all of
+given sample size, increasing the size of the corpus (which provides better
+statistics) makes it less likely that a chosen sample will contain all of
 a theorem's dependencies. The majority of such samples would hence have an empty
 ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-what the MTE tool produces! This is clearly undesirable as an evaluation
-method.
+the tool's output! This is clearly undesirable as an evaluation method.
 
@@
 theorem. We could do this using rejection sampling, but it is more efficient to
-pick a theorem, weighted in proportion to their number of dependencies (ignoring
-those with too many dependencies to fit in our sample). We use those
-dependencies as our sample, padded up to the required size with definitions
-chosen uniformly from the remainder.
+pick a theorem, weighted in proportion to their number of dependencies and
+ignoring those with more dependencies than our sample size. That theorem's
+dependencies become our sample, padded up to the required size with uniform
+choices from the remaining definitions.
 
@@
 Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE system proceeds as in prior work. We must translate the
-chosen definitions into the particular input format of the tool under study,
-then we time the execution, with a timeout after e.g. 5 minutes.
+execution of the MTE tool proceeds as in prior work. We must translate the
+chosen definitions into the required input format, then we time the execution
+with a timeout (e.g. 5 minutes).
 
 In our experiments we have found that memory usage is also an important part of
-a system's performance, but rather than complicating our analysis with an extra
-dimension, we instead set a limit (as large as practical) and terminate program
-runs which cross it.~\footnote{We could treat the hardware's RAM capacity as our
-  limit, allow the operating system to swap any overflow to disk, and wait for
-  the inevitable slowdown to trigger the timeout. This requires no work to
-  implement, but wastes a lot of time waiting for runs which could otherwise be
-  terminated early.} This is in line with the expected usage of these systems:
+a tool's performance, but rather than complicating our analysis with an extra
+dimension, we instead allow programs to use as much memory as they like, and
+either get killed by the operating system or slow down so much from swapping
+that they time out. This is in line with the expected usage of these tools:
 either there is enough memory, or there isn't; implementations shouldn't be
@@
 
-This approach produces a single runtime, precision and recall value for each
+To calculate precision and recall, the conjectures generated by each tool, as
+well as the ground-truth theorems, need to be parsed into a common format and
+compared syntactically after normalising away ``irrelevant'' details. We
+consider variable naming to be irrelevant, which can be normalised by numbering
+free variables from left to right and using de Bruijn indices for bound
+variables. We also consider the left/right order of equations to be irrelevant,
+which we normalise by choosing whichever order results in the
+lexicographically-smallest expression.
+
+Other than $\alpha$-equivalence and equation symmetry, we specifically
+\emph{ignore} logical relationships between syntactically distinct statements,
+such as one equation being implied by another. Whilst logically sound,
+second-guessing the ground truth in this way would harm other aspects which
+influence interestingness (for example, a more general statement is more widely
+applicable, but it might also be harder to comprehend).
+
+This procedure gives us a single runtime, precision and recall value for each
 sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of an MTE system, or \emph{comparing} the performance of
-\emph{two} MTE systems.
+the performance of a single MTE tool and \emph{comparing} the performance of two
+tools.
 
@@
 
-Each sample is only explored once, so that we cover as many independent samples
-as possible to better estimate how a system's performance generalises to unseen
-inputs. To combine these data into an aggregate summary depends on what we are
-interested in measuring.
-
-One general question we might ask is how a system's performance scales with
-respect to the input size (the number of definitions in the theory). This is
+Each sample is only explored once by each tool, so that we cover as many
+independent samples as possible to better estimate how a tool's performance
+generalises to unseen inputs. How we combine these data into an aggregate
+summary depends on what we are interested in measuring.
+
+One general question we might ask is how a tool's performance scales with
+respect to the input size (the number of given definitions). This is
 straightforward to measure by varying the sample size, but we need some way to
@@
 
-The precision and recall for multiple samples $S$ can be combined in two ways:
-finding their mean value (the \emph{average of the ratios}) or by summing the
-numerators and denominators (the \emph{ratio of the averages}). The former gives
-the expected precision and recall for a run, which is of direct relevance to
-users of these tools, and is hence our preferred measure. The latter assigns
-more weight to theories which generate more conjectures, but we avoid this since
-there is no \emph{a priori} justification for considering such theories as
-``better''.
-
-We summarise the runtimes by choosing their median, as this is more robust
+We can summarise runtimes by choosing their median, as this is more robust
 against long-running outliers, and hence represents performance for a
-``typical'' theory.
-
-Another advantage to our statistical approach is that we can compute the
-\emph{spread} of our data, which is unknown in the previous one-off
-evaluations. In particular, we might be concerned that such previous evaluations
-used ``popular'' theories, like those involving natural numbers, which may be
-unrepresentative of typical usage: we should not expect a typical user's theory
-(e.g. a software library) to be as mathematically rich as number theory!
+``typical'' input of that size. Unlike previous one-off evaluations, we can also
+compute the \emph{spread} of our data, for example with the inter-quartile
+range.
+
+The precision and recall for multiple samples can be combined in two ways. We
+can calculate their mean value, known as the \emph{average of the ratios}, to
+find the expected precision and recall for runs of this size. Alternatively, we
+can sum their numerators and denominators to get the \emph{ratio of the
+  averages}: this normalises the size of the generated and ground truth sets,
+letting us model precision and recall as simple Bernoulli processes.
 
@@
 
-To encourage competition in the field, it is important to compare the relative
-performance of different MTE systems on the same task. Since the aggregate
-statistics in the above summaries do not include details of specific runs, any
-comparison based on them would have very low statistical power, and therefore it
-is appropriate to use an alternative approach when comparing.
-
-We propose comparing the data points directly, using a \emph{paired difference
-  test}: for each individual sample, we find the \emph{difference} between the
-two systems' data \emph{for that sample}. We then aggregate these
-\emph{differences}, to determine whether one system performs significantly
-better or worse than the other.
-
-This neatly avoids the variance inherent in the summary statistics, and (with
-enough samples) approaches a normal distribution. Paired difference tests are
-also robust to the censoring caused by timeouts: upper-bounding the run time
-gives an upper bound to the (observable) differences, biasing our inference
-towards the null hypothesis (indistinguishable performance) rather than
-unfounded conclusions.
+To measure progress and encourage competition in the field, it is important to
+compare the relative performance of different tools on the same task. Since the
+aggregate statistics in the above summaries have obscured the details of
+specific runs, any comparison based on them (such as comparing average recall)
+would have very low statistical power. More direct comparisons can be made using
+\emph{paired} tests, since for each individual sample we have measurements for
+\emph{both} tools.
+
+We propose that running times are compared using a Wilcoxon signed-rank
+test~\cite{wilcoxon1945individual}, since it does not require that the data be
+normally distributed (unlike, say, the z-test or t-test). In our application we
+follow a variant of the Speedup-Test protocol of Touati, Worms and
+Briais~\cite{touati2013speedup}; the differences are our use of the paired form
+of the Wilcoxon test, and the fact that each run was performed on a different
+sample (to reduce sampling bias, at the expense of accuracy).
+
+Another benefit to such a paired difference test is its robustness to the
+censoring caused by timeouts: upper-bounding the run time gives an upper-bound
+to the observable difference, which biases our results \emph{towards} the null
+hypothesis of indistinguishable performance.
+
+To compare the recall we use the standard McNemar test for paired samples: each
+ground-truth theorem is categorised as either being found only by the first
+tool, only by the second, by both or by neither. Those found by both or neither
+are ignored, and the tool-specific numbers are tested for marginal homogeneity
+(i.e. whether their differences give rise to similar totals).
+
+We cannot use McNemar's test to compare precision, since each tool may generate
+different conjectures. Instead, we use Boschloo's form of Barnard's test to
+compare the number of wanted and unwanted conjectures generated by each tool.
 
@@
 
-    \node (thy)  [startstop, below=1cm of def]{Full Theory};
+    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
     \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};
@@
 
-    \node (choose) [process, below=of thm   ]{Sample a theorem};
-    \node (deps)   [process, below=of choose]{List referenced definitions};
+    \node (choose) [process, below=of thm   ]{Choose a theorem};
+    \node (deps)   [process, below=of choose]{List theorem dependencies};
 
@@
               in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Sample definitions to pad list};
-
-    \node (find) [process,   below right=1.5cm and -1cm of pad]{List all applicable theorems};
+    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
+
+    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
     \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    \node (sthm)  [startstop, below=of find]{Ground Truth};
 
@@
            in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE system};
+    \node (run) [process, below=of runPos]{Run MTE tool};
     \node (pr)  [process, below=of run]{Analysis};
@@
     \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (find);
+    \draw [arrow] (pad)    -- (sthm);
     \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (find)   -- (sthm);
     \draw [arrow] (sthy)   |- (run);
@@
     % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (find);
+    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
 
@@
 tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus. TIP 0.2 contains 343 theorem statements,
-which depend on 182 distinct function definitions.
+benchmark as our ground truth corpus.
 
@@
 
-We decided to benchmark sample sizes up to 20, since neither tool seemed to
-perform well beyond this. Our results will be analysed by Speedup-test protocol,
-which follows the statistical ``rule of thumb'' of treating sample sizes
-$\leq$ 30 as ``small'', so we pick 31 samples of each size in order to cross
-this (arbitrary) threshold\footnote{Only 14 samples of size 1 were analysed, and
-  30 of size 2, since the rest were duplicates. This is due to the constraint
-  that samples contain at least one theorem's dependencies.}. This gave $20$
-sizes $\times 31$ samples $\times 2$ tools $= 1240$ runs. To keep the total time
-down to a few days we chose a timeout of 300 seconds, since that covered most of
-the successful QuickSpec and IsaCoSy results we saw, and longer times gave
-rapidly diminishing returns.
-
-We normalised the conjectures generated by each tool, as well as the theorems
-from the ground truth, by numbering their free variables from left to right and
-using de Bruijn indices for bound variables. Since both tools can only output
-equations, we also sorted their left- and right-hand-sides into lexicographical
-order. Precision and recall were calculated by syntactically comparing the sets
-of generated conjectures to those of their ground truth. Other than equation
-symmetry, we specifically \emph{ignored} logical relationships between
-syntactically distinct statements, such as one equation being implied by
-another. Whilst logically sound, second-guessing the ground truth in this way
-would harm the other aspects which we care about which influence interestingness
-(for example, a more general statement is more widely applicable, but it might
-also be harder to understand).
+Based on this we decided to benchmark sample sizes up to 20, since neither tool
+seemed to perform well beyond that. The Speedup-test protocol follows the
+statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
+so we pick 31 samples of each size in order to cross this (arbitrary)
+threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
+a few days we chose a timeout of 300 seconds, since that covered most of the
+successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
+diminishing returns. During analysis, duplicate samples (caused by our sampling
+restrictions) were found and discarded, so only 14 samples of size 1 were used
+and 30 of size 2.
 
@@
 one theorem statement (this removes ambiguity about which \emph{definitions}
-count as ``interesting'' and which are just ``implementation details'' for other
+count as interesting and which are just ``implementation details'' for other
 definitions).
@@
 code, including comparison functions and random data generators suitable for
-QuickSpec. We translate all 269 unique definitions into a single module/theory
+QuickCheck. We translate all 269 unique definitions into a single module/theory
 which is imported on each run of the tools, although only those functions which
@@
 
-We benchmarked QuickSpec version 0.9.6, which is tool written in Haskell for
-conjecturing equations involving a given set of Haskell functions and
-universally-quantified variables (a \emph{signature}).
-
-The QuickSpec algorithm first enumerates all well-typed expressions containing
-terms from the signature, up to a chosen depth (3 by default). Expressions of
-the same type are assumed to be equal, and grouped into equivalence classes. A
-testing process then tries to refute this assumption by instantiating each
-variable with a randomly chosen value (using the QuickCheck testing library),
-evaluating the expressions with those values and comparing the results. The
-equivalence classes are split up to separate any expressions which are observed
-to be different, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal. Finally, a congruence closure algorithm is
-applied to these equations, to remove redundancies and special-cases.
-
-In order to thoroughly benchmark QuickSpec, we need to automate some of the
-decisions which are normally left up to the user:
+We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
+conjecturing equations involving Haskell functions, described in more detail in
+section~\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
+to automate some of the decisions which are normally left up to the user:
 
@@
   encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating.
+  it before translating. It still appears in samples and contributes to the
+  ground truth, which penalises IsaCoSy for being unable to explore such
+  definitions.
 \item When using a type in an IsaCoSy signature, that type's constructors will
@@
   even those may not be present in the current sample) this will unfairly reduce
-  the tool's precision. To avoid this, we add a post-processing step which
+  the calculated precision. To avoid this, we add a post-processing step which
   replaces all occurrences of a constructor with the corresponding wrapper, then
   discards any conjectures which involve functions other than those in the
-  current sample.
+  current sample. This presumably results in more work for IsaCoSy, exploring
+  constructors unnecessarily, but it at least does not bring down the quality
+  measures.
 \item Since Isabelle is designed for theorem proving rather than programming, it
@@
   pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{head Nil}), hence preventing IsaCoSy from ever generating
+  (for example $\Head(\Nil)$), hence preventing IsaCoSy from ever generating
   them.
@@
   \caption{Running times of IsaCoSY on theories sampled from TIP. Each point
-    is an individual run, with 30 per size except for 2 (29). Colour indicates
-    the total number of conjectures produced by each run, with failing processes
-    shown in red (timeouts and out of memory). Points are spread out
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
+    indicates the total number of conjectures produced by each run, with failing
+    processes (timeouts and out of memory) shown in red. Points are spread out
     horizontally to prevent overlaps. Box plots show median and quartiles, with
@@
 
-We compare the running times of QuickSpec and IsaCoSy using a similar protocol
-to the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We
-treat each of our sample sizes as a separate ``benchmark'', but deviate from
-their approach by using a different input (sample) for each run. This reduces
-the sampling bias, but also reduces the accuracy of our confidence intervals.
-We used version 2 of the Speedup-Test \texttt{R} implementation, which we
-patched to use the \emph{paired} form of the (one-sided) Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since this increases our statistical power
-given that each sample is run through both tools. We use Pratt's
+We compare the running times of QuickSpec and IsaCoSy using our paired variant
+of the Speedup-Test protocol, where each of our sample sizes is a separate
+``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
+which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
+signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
 method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
@@
 sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-giving a 95\% confidence interval for the proportion of (median) speedups to be
-$[0.669;0.982]$. The magnitude of each speedup is is given in
-table~\ref{table:speedups}.
+putting the proportion of sizes with faster median times between 66.9\% and
+98.2\% with 95\% confidence. The magnitude of each speedup is given in
+table~\ref{table:speedups}. We would predict the speedup to grow for larger
+sample sizes, but the increasing proportion of timeouts causes our estimates to
+become more conservative: by size 20 most runs are timing out, and hence are
+tied.
 
 \begin{table}
+  \centering
   \begin{tabular}{ |r|l| }
@@
   \end{tabular}
-  \caption{Speedup when using QuickSpec compared to IsaCoSy for each sample
-    size. Those which aren't statistically significant with 95\% confidence are
-    marked as ``-''. These are conservative estimates, since the difference is
-    bounded by the timeout.}
+  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
+    Speedup-test for a confidence level of 95\%. Sizes 1 and 2 were not
+    considered statistically significant due to having less than 30 samples.
+    These are conservative estimates, since the 300 second time limit acts to
+    reduce the measured time difference.}
   \label{table:speedups}
@@
 
-% Altogether TIP problem by QuickSpec and IsaCoSy theory
-% exploration systems. To assess the scalability of their algorithms, we sampled
-% theories of various sizes: from theories containing only a single definition, up
-% to theories containing 20 definitions.
-%  We call the Theory Exploration Benchmark (TEB).
+We compare the recall proportions by applying McNemar's test to only those
+samples where both tools succeeded. We pooled these together from all sample
+sizes to produce the following table, and found that QuickSpec is significantly
+different from IsaCoSy with a p value of 0.0026.
+
+{
+  \centering
+  \begin{tabular}{ |l|l|l| }
+    \hline
+                           & Found by IsaCoSy & Not found by IsaCoSy \\
+    \hline
+    Found by QuickSpec     & 28               & 19                   \\
+    Not found by QuickSpec & 4                & 28                   \\
+    \hline
+  \end{tabular}
+}
+
+Precision was also compared by pooling results from samples where both tools
+finished successfully. We cannot perform McNemar's test in this case, since each
+tool generated different sets of conjectures. Instead, we add up the number of
+results which are ``interesting'' (appear in the ground truth) and those which
+are ``uninteresting'' (do not appear in the ground truth):
+
+{
+  \centering
+  \begin{tabular}{ |l|l|l| }
+    \hline
+              & Interesting & Uninteresting \\
+    \hline
+    IsaCoSy   & 32          & 137      \\
+    QuickSpec & 47          & 301      \\
+    \hline
+  \end{tabular}
+}
+
+These totals were tested for independence using Boschloo's test, assuming a
+binomial distribution, and gave an p-value of 0.111. With our (arbitrary) choice
+of $\alpha = 0.05$, this is not significant enough to reject the null hypothesis
+that their precision is equal.
+
+Note that this is the only comparison where IsaCoSy performed better (although
+not significantly so) with 19\% precision compared to QuickSpec's 14\%. However,
+precision on its own is not the whole story: it can be increased at the expense
+of recall by simply generating fewer conjectures. IsaCoSy's significantly lower
+recall indicates that it may be overly-conservative in this way, compared to
+QuickSpec.
+
+More importantly, the poor time and memory usage of IsaCoSy meant that very few
+samples finished successfully. If we include failed runs in our tables, treating
+them as if they succeeded with no output, the resulting statistics lean heavily
+in favour of QuickSpec simply because it generated so much more than IsaCoSy in
+the available time.
 
@@
 % All together, TIP provides 219 distinct function definitions and 343
-%   theorem statements, which is enough to fully exercise current MTE systems.
+%   theorem statements, which is enough to fully exercise current MTE tools.
 % Benchmark problems include examples from the theorem proving literature
@@
 %   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
-%   (which are used by MTE systems).
+%   (which are used by MTE tools).
 
@@
 % corpus; the latter prevents tailoring the corpus to influence the choices,
-% although MTE systems could .
-
-% To provide a more robust comparison of these two systems, we also applied a
+% although MTE tools could .
+
+% To provide a more robust comparison of these two tools, we also applied a
 % paired difference test: measuring, for each sampled theory, the difference in
-% time, precision and recall between the two systems.
+% time, precision and recall between the two tools.
 
@@
 \label{sec:discussion}
-
-The key to our benchmarking methodology is the simplifying assumption that
-theorem proving problem sets are a good proxy for desirable MTE output. This
-allows us to side-step the philosophical quagmire of ``interestingness'' to
-produce concrete, measurable values; it is also a core weakness, as there are
-compelling reasons to refute it.
 
@@
 activity as experiences of sensory beauty~\cite{zeki}; computer scientists, such
-as Colton's metrics above which are proposed based largely on a computer
-scientist's ``intuition'' and argumentation about why a metric would be
-important (and previous use of such metrics, and - in one isolated case
-(surprisingingness) - on a single quote from a mathematician\footnote{Similarly,
+as Colton's metrics in table~\ref{table:colton} which are proposed based largely
+on a computer scientist's ``intuition'' and argumentation about why a metric
+would be important (and previous use of such metrics, and - in the case of
+``surprisingingness'' - on a single quote from a mathematician\footnote{Similarly,
   classification of a MATHsAiD result as either interesting or
@@
 
+The key to our benchmarking methodology is the simplifying assumption that
+theorem proving problem sets are a good proxy for desirable MTE output. This
+allows us to side-step this philosophical quagmire to produce concrete,
+measurable values; it is also a core weakness, as there are compelling reasons
+to refute it.
+
 We do not claim that our use of corpora as a ground truth exactly captures all
@@
 
-Another practical limitation limitation of our benchmarking approach is that it
-only applies to systems which act in ``batch mode'', i.e. those which choose to
-halt after emitting some output. Whilst all of the systems we have encountered
-are of this form, an analogous benchmark may be desirable for systems which
-instead run continuously or interactively (perhaps akin to McCarthy's
-``advice taker''~\cite{McCarthy_Programs59}).
-
-% philosophical bits
-
-% We're assuming there is general agreement on what is "good maths", when
-% actually this isn't at all clear-cut
-
-% agreement in maths? matt inglis
-
-% other - zeki/two papers with ursula...
-
-% methodology behind Simon's paper
+Another practical limitation of our benchmarking approach is that it only
+applies to tools which act in ``batch mode'', i.e. those which choose to halt
+after emitting some output. Whilst all of the systems we have encountered are of
+this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
+be run without a depth limit, and form part of the ``scientist assistant'' role
+envisaged by Lenat (or McCarthy's earlier
+``advice taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be
+designed for such long-running, potentially interactive programs, but that
+beyond the scope of this project.
 
@@
 %\begin{acknowledgements}
-%  We are grateful to Jianguo Zhang for help with our statistical analysis.
-%
 %  EPSRC

### 2018-03-15 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Limit colour bars to successes

@@ -287,7 +287,7 @@ import seaborn           as sns
 def makeColours(agg):
     def go(key):
         from matplotlib.colors import ListedColormap
-        maxVal    = max(agg[key])
+        maxVal    = max([agg[key][i] for i,s in enumerate(agg['success']) if s])
         colourMap = sns.color_palette('viridis', maxVal + 1).as_hex()
         msg('Highest {0} count is {1}'.format(key, maxVal))
         return {

### 2018-03-15 check.sh
Ignore warnings which can be triggered by 'https://' URLs

@@ -65,6 +65,7 @@ for PAT in 'Invalid checksum for ISBN' \
            'Suspicious year'           \
            'year ='                    \
            'pages ='                   \
+           'Unexpected protocol://'    \
            '?? "stdin", line 1: Expected comma after last field'
 do
     grep -v "$PAT" < "$DIR/bibclean.stderr" > "$DIR/bibclean.stderr2" || true

### 2018-03-15 TEBenchmarkPaper/article.tex
Keep tables in place
Final fixes

@@
 \usepackage{csvsimple}
+\usepackage{float}
+\restylefloat{table}
 \usepackage{hhline}
@@
   analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
-providing a direct comparison of these
+providing a direct comparison of these three
 tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
@@
   maximum functions were included).} To measure precision and recall we first
-choose a set of definitions, such as those in figure \ref{figure:list_theory},
-and a \emph{ground truth} to treat as the 'ideal' set of conjectures we would
-like a tool to produce for those definitions. Each tool is run and their outputs
-are compared to the ground truth. To score 100\% on precision and recall, all of
-the conjectures which appear in the ground truth must be generated, and nothing
-else:
+choose a set of definitions (such as those in figure \ref{figure:list_theory})
+and a \emph{ground truth} to treat as the ``ideal'' set of conjectures we would
+like a tool to produce for those definitions (e.g. this might include
+equation~\ref{eq:mapreduce}). Each tool is run and their outputs are compared to
+the ground truth. To score 100\% on precision and recall, all of the conjectures
+which appear in the ground truth must be generated, and nothing else:
 
@@
 
-We can summarise runtimes by choosing their median, as this is more robust
-against long-running outliers, and hence represents performance for a
+We can summarise runtimes by choosing their median, as this is more robust than
+the mean against long-running outliers, and hence represents performance for a
 ``typical'' input of that size. Unlike previous one-off evaluations, we can also
@@
 can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the size of the generated and ground truth sets,
-letting us model precision and recall as simple Bernoulli processes.
+  averages}: this normalises the output and ground truth cardinalities, allowing
+precision and recall to be modelled as simple Bernoulli processes.
 
@@
 test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed (unlike, say, the z-test or t-test). In our application we
-follow a variant of the Speedup-Test protocol of Touati, Worms and
-Briais~\cite{touati2013speedup}; the differences are our use of the paired form
-of the Wilcoxon test, and the fact that each run was performed on a different
-sample (to reduce sampling bias, at the expense of accuracy).
+normally distributed. In our application we follow a variant of the Speedup-Test
+protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the differences
+are our use of the paired form of the Wilcoxon test, and the fact that each run
+was performed on a different sample (to reduce sampling bias, at the expense of
+accuracy).
 
@@
 different conjectures. Instead, we use Boschloo's form of Barnard's test to
-compare the number of wanted and unwanted conjectures generated by each tool.
+compare the proportions of wanted and unwanted conjectures generated by each
+tool.
 
@@
 tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus.
+benchmark as our ground truth corpus~\cite{claessen2015tip}.
 
@@
 Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-test protocol follows the
+seemed to perform well beyond that. The Speedup-Test protocol follows the
 statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
@@
 verification and inductive theorem proving literature are included, ensuring
-relevance to those fields; the format includes higher-order functions and
-inductive datatypes, which we are interested in; plus it is accompanied by
-tooling to convert this format into a variety of languages, including Haskell
-and Isabelle which are used by existing MTE tools.
+relevance to those fields; the format includes the higher-order functions and
+inductive datatypes we are interested in; it is large enough to pose a challenge
+to current MTE tools; plus it is accompanied by tooling to convert its custom
+format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
+languages, including Haskell and Isabelle.
 
@@
 
+\begin{samepage}
 \begin{verbatim}
 (declare-datatypes
-  (a)                       ;; Type variable (element type)
+  (a)                       ;; Type parameter (element type)
   ((List                    ;; Type name
@@
 \end{verbatim}
+\end{samepage}
 
@@
 
+\begin{samepage}
 \begin{verbatim}
 (define-fun
-  (par (a)                   ;; Type variable
+  (par (a)                   ;; Type parameter
     (constructor-Cons        ;; Function name
@@
 (define-fun
-  (par (a)                        ;; Type variable
+  (par (a)                        ;; Type parameter
     (destructor-head              ;; Function name
@@
   constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetery.}
+  symmetry.}
 
@@
 which is imported on each run of the tools, although only those functions which
-appear in the current sample are actually explored.
+appear in the current sample are included in the signature and explored.
 
@@
 
+\begin{sloppypar}
   \begin{itemize}
-\item We must decide what variables to include. We choose to add three variables
-  for each type that appears as a function argument, except for types which have
-  no QuickCheck data generators.
+  \item We must decide what variables to include. We choose to add three
+    variables for each type that appears as a function argument, except for
+    types which have no QuickCheck data generators.
   \item We must \emph{monomorphise} all types. For example, functions like
@@
       type class constraint, we pick a suitable type non-deterministically from
-    those in scope during compilation; if no suitable type is found, we give up
-    and don't include the  variable.}
+      those in scope during compilation; if no suitable type is found, we give
+      up and don't include that function.}
   \item Haskell functions are ``black boxes'', which QuickSpec can't compare
     during its exploration process. They are also curried, always taking one
-  argument but potentially returning another function. QuickSpec lets us assign
-  an arity to each function in the signature, from 0 to 5, so we pick the
-  highest that is type-correct, since this avoids a proliferation of
+    argument but potentially returning another function. QuickSpec lets us
+    assign an arity to each function in the signature, from 0 to 5, so we pick
+    the highest that is type-correct, since this avoids a proliferation of
     incomparable, partially-applied functions.
   \end{itemize}
+\end{sloppypar}
 
@@
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 30 per size except for 2 (29). Colour indicates
-    the total number of conjectures produced by each run, with failing processes
-    shown in red. Points are spread out horizontally to prevent overlaps.
-    Box plots show median and quartiles, with whiskers showing points at
-    1.5$\times$~inter-quartile range.}
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
+    indicates the total number of conjectures produced by each run, with failing
+    processes shown in red. Points are spread out horizontally to prevent
+    overlaps. Box plots show median and quartiles, with whiskers showing points
+    at 1.5$\times$~inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
 The time taken to explore samples of different sizes is shown in
-\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red, and
-these occurred for all sample sizes but are more common for larger samples (over
-half the runs for sample size 20 timed out). Runs which succeeded mostly
+figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
+and these occurred for all sample sizes but are more common for larger samples
+(over half the runs for sample size 20 timed out). Runs which succeeded mostly
 finished within 30 seconds, generating few or no conjectures; those taking
@@
 
-We model precision and recall as simple Bernoulli processes: the generated
-conjectures (wanted or not) are like coin tosses whose bias is the precision;
-likewise, the ground truth theorems (found or not) are like coin tosses biased
-by the recall.
+Precision and recall are shown in figure~\ref{figure:quickspec_precRec}, which
+we model as simple Bernoulli processes: the generated conjectures (wanted or
+not) are like coin tosses whose bias is the precision; likewise, the ground
+truth theorems (found or not) are like coin tosses biased by the recall. Note
+that this simplification does not explain how many conjectures the tool decides
+to output for each run.
 
@@
 Rather, the number of conjectures generated is increasing at a higher rate than
-the size of the ground truth. These extra conjectures may involve the
+the size of the ground truth. These extra conjectures may involve those
 ``padding'' definitions in a sample which don't contribute to its ground truth,
@@
 this benchmark is concerned). The deluge of generated conjectures could be
-filtered by another post-processing step down to a more desirable sub-set.
-Alternatively, large signatures could be filtered to remove ``padding''
-definitions (those whose conjectures are unlikely to be ``interesting'') and
-divided up into multiple smaller signatures to separate those definitions whose
-relationships are ``uninteresting''. Both of these methods could improve the
-precision, although care would be needed to avoid a large reduction in recall.
-The latter option could also improve the running time, since multiple smaller
-signatures may be faster to explore than a single large one.
+filtered down to a more desirable sub-set by another post-processing step.
+Alternatively, rather than exploring all of the given definitions together,
+multiple smaller signatures could be selected from the input by predicting which
+combinations are likely to lead to interesting conjectures; this would avoid
+both the ``padding'' and the cross-dependency relationships. Both of these
+methods could improve the precision, although care would be needed to avoid a
+large reduction in recall. The latter option could also improve the running
+time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
+signatures may be faster to explore than a single large one. Such improvements
+would also need to be ``generic'', to avoid over-fitting to this particular
+benchmark.
 
@@
 ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle.
-
-The following issues had to be overcome to make our benchmark applicable to
-IsaCoSy:
+with the 2015 version of Isabelle. The following issues had to be overcome to
+make our benchmark applicable to IsaCoSy:
 
@@
   pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example $\Head(\Nil)$), hence preventing IsaCoSy from ever generating
-  them.
+  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
+  IsaCoSy from ever generating them.
 \end{itemize}
@@
   \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSY on theories sampled from TIP. Each point
+  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
     is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
@@
 
-The most striking result is how rapidly IsaCoSy's running time increases with
-sample size: from size 7 upwards, all runs failed (most timed out, a few aborted
-early due to running out of memory). As we found with our preliminary testing,
-this increase appears exponential, so even large increases to the timeout would
-not produce many more successful observations. A significant amount of IsaCoSy's
-time (around 50 seconds) was spent loading the generated theory (containing all
-distinct datatypes and functions from TIP); this overhead is unfortunate, but
-since it is constant across all sample sizes it does not affect our conclusions.
+The most striking result is how rapidly IsaCoSy's running time, shown in
+figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
+size 6 failed, with most timing out and a few aborting early due to running out
+of memory. Like in our preliminary testing, this increase appears exponential,
+so even large increases to the timeout would not produce many more successful
+observations. A significant amount of IsaCoSy's time (around 50 seconds) was
+spent loading the generated theory (containing all distinct datatypes and
+functions from TIP); this overhead is unfortunate, but since it is constant
+across all sample sizes it does not affect our conclusions.
 
 With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output. It is clear from the quantity
-of non-zero proportions that IsaCoSy is capable of discovering interesting
-conjectures, and the graphs appear to follow the same shapes as those of
-QuickSpec (decreasing precision, flat recall), although the error margins are
-too wide to be definitive.
+the precision/recall quality of IsaCoSy's output, shown in
+figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
+proportions that IsaCoSy is capable of discovering interesting conjectures, and
+the graphs appear to follow the same shapes as those of QuickSpec (decreasing
+precision, flat recall), although the error margins are too wide to be
+definitive.
 
@@
   \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-test for a confidence level of 95\%. Sizes 1 and 2 were not
+    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
     considered statistically significant due to having less than 30 samples.
@@
 
-{
+\begin{table}[H]
   \centering
-  \begin{tabular}{ |l|l|l| }
+  \begin{tabular}{ |r|l|l| }
     \hline
-                           & Found by IsaCoSy & Not found by IsaCoSy \\
+    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
     \hline
-    Found by QuickSpec     & 28               & 19                   \\
-    Not found by QuickSpec & 4                & 28                   \\
+    \bfseries Found by QuickSpec     & 28 & 19 \\
+    \bfseries Not found by QuickSpec & 4  & 28 \\
     \hline
   \end{tabular}
-}
+\end{table}
 
@@
 
-{
+\begin{table}[H]
   \centering
-  \begin{tabular}{ |l|l|l| }
+  \begin{tabular}{ |r|l|l| }
     \hline
-              & Interesting & Uninteresting \\
+              & \bfseries Interesting & \bfseries Uninteresting \\
     \hline
-    IsaCoSy   & 32          & 137      \\
-    QuickSpec & 47          & 301      \\
+    \bfseries IsaCoSy   & 32          & 137      \\
+    \bfseries QuickSpec & 47          & 301      \\
     \hline
   \end{tabular}
-}
-
-These totals were tested for independence using Boschloo's test, assuming a
-binomial distribution, and gave an p-value of 0.111. With our (arbitrary) choice
-of $\alpha = 0.05$, this is not significant enough to reject the null hypothesis
-that their precision is equal.
+\end{table}
+
+These totals were tested for independence using Boschloo's
+test~\cite{lydersen2009recommended}, assuming a binomial distribution and
+conditioning on the number of generated conjectures (as if this were fixed by
+design). This produces a p-value of 0.111, which given our (arbitrary) choice of
+$\alpha = 0.05$ is not significant enough to reject the null hypothesis that
+their precision is equal.
 
@@
 precision on its own is not the whole story: it can be increased at the expense
-of recall by simply generating fewer conjectures. IsaCoSy's significantly lower
-recall indicates that it may be overly-conservative in this way, compared to
-QuickSpec.
+of recall by making the tool more conservative. IsaCoSy's significantly lower
+recall indicates that it may be overly-conservative compared to QuickSpec, which
+would also explain its precision.
 
@@
 % form our ground truth).
-
-% We focus on the Tons of Inductive Problems (TIP) theorem proving
-% benchmark~\cite{claessen2015tip}, since it has several desirable properties:
-
-% All together, TIP provides 219 distinct function definitions and 343
-%   theorem statements, which is enough to fully exercise current MTE tools.
-% Benchmark problems include examples from the theorem proving literature
-%   as well as common program verification tasks, ensuring the resulting corpus is
-%   relevant to researchers and practitioners.
-% TIP's maintainers provide a set of tools, including translators from TIP's
-%   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
-%   (which are used by MTE tools).
-
-% To aid reproducibility, we use a cryptographic hash as our source of randomness:
-% the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
-% corpus; the latter prevents tailoring the corpus to influence the choices,
-% although MTE tools could .
-
-% To provide a more robust comparison of these two tools, we also applied a
-% paired difference test: measuring, for each sampled theory, the difference in
-% time, precision and recall between the two tools.
 
@@
 development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Methods for finding out what that means have come from
-neuro-scientists, such as the study by Zeki \etal{} in which they test whether
-mathematicians' experiences of abstract beauty correlates with the same brain
-activity as experiences of sensory beauty~\cite{zeki}; computer scientists, such
-as Colton's metrics in table~\ref{table:colton} which are proposed based largely
-on a computer scientist's ``intuition'' and argumentation about why a metric
-would be important (and previous use of such metrics, and - in the case of
-``surprisingingness'' - on a single quote from a mathematician\footnote{Similarly,
-  classification of a MATHsAiD result as either interesting or
-  uninteresting~\cite{roy} was made by the main system developer: in both cases
-  the main developer had a background in mathematics.}; and mathematicians
-themselves, such as Gowers' suggestion that we can identify features which are
-commonly associated with good proofs~\cite{gowers}. All of these approaches rest
-upon the assumption that it makes sense to speak of ``the good'' in mathematics.
-This assumption is supported by concepts such as Erdos's ``The Book'' -- a
-theoretical construct of a book which contains the most elegant proof of every
-theorem (brought to life in~\cite{aigner2010proofs}). However, empirical studies
-by psychologists such as Inglis call into question such assumptions: work by
-Inglis and colleagues has shown that there is not a single standard of validity
-among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
-nor beauty. %FIXME needs citation
-
-The key to our benchmarking methodology is the simplifying assumption that
-theorem proving problem sets are a good proxy for desirable MTE output. This
-allows us to side-step this philosophical quagmire to produce concrete,
-measurable values; it is also a core weakness, as there are compelling reasons
-to refute it.
+proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
+light on this by testing whether mathematicians' experiences of abstract beauty
+correlates with the same brain activity as experiences of sensory
+beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
+table~\ref{table:colton}, from computer scientists like Colton, are based
+largely on ``intuition'', plausibility arguments about why a metric would be
+important, and previous use of such metrics (in the case of
+``surprisingingness'', from a single quote from a mathematician).
+Opinions from mathematicians themselves include Gowers' suggestion that we can
+identify features which are commonly associated with good
+proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
+book containing the most elegant proof of every theorem, brought to life
+in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
+interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
+developer).
+
+All of these approaches rest upon the assumption that it even makes sense to
+speak of ``the good'' in mathematics. However, empirical psychological studies
+call into question such assumptions: for example, work by Inglis and colleagues
+has shown that there is not even a single standard of \emph{validity} among
+contemporary mathematicians~\cite{inglis2013mathematicians}.
+
+Whilst these difficulties are real and important, we cannot ignore the fact that
+mathematics is nevertheless being practised around the world; and similarly that
+researchers have forged ahead to develop a variety of tools for automated
+exploratory mathematics. If we wish to see these tools head in a useful,
+fruitful direction then \emph{some} method is needed to compare their
+approximate ``quality'' in a concrete, measurable way.
+
+The key to our benchmarking methodology is to side-step much of this
+philosophical quagmire using the simplifying assumption that theorem proving
+problem sets are a good proxy for desirable input/output behaviour of MTE tools.
+As an extension of existing precision/recall analysis, this should hopefully not
+prove too controversial, although we acknowledge that there are compelling
+reasons to refute it.
 
@@
 
-Notably, if MTE matures to the point of providing \emph{truly novel} insights,
-even in well-studied domains, then the assumptions we rely on break down because
-such results cannot (by definition) appear in any existing corpus.
+One notable weakness is that our methodology only makes sense in domains where
+existing human knowledge surpasses that discovered by the machine. Any
+\emph{truly novel} insights discovered during testing will not, by definition,
+appear in any existing corpus, and we would in fact \emph{penalise} tools for
+such output. However, we do not believe this to be a realistic problem for the
+time being (we would be delighted to be proved wrong!).
 
@@
 be run without a depth limit, and form part of the ``scientist assistant'' role
-envisaged by Lenat (or McCarthy's earlier
-``advice taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be
-designed for such long-running, potentially interactive programs, but that
-beyond the scope of this project.
+envisaged by Lenat (or McCarthy's earlier ``advice
+taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be designed for
+such long-running, potentially interactive programs, but that is beyond the
+scope of this project.
 
@@
 
-% future work, etc.
-
-Our benchmarking methodology, and accompanying suite based on TIP, provides a
-general way to evaluate and compare many diverse approaches to the problem of
-theory exploration, whilst avoiding some of the philosophical complications of
-the field. The methodology can be adapted to suit particular domains, for
-example our experiments have focused on inductive theories with higher-order
-functions, and the corpora used to generate a benchmark suite can always be
-improved and refined with the addition of new definitions and theorems.
-
-We believe that a standard approach to benchmarking and comparison will ease the
-burden on researchers wanting to evaluate systems, and provide a common goal to
-pursue in the short term.
+We propose a general benchmarking methodology for measuring, summarising and
+comparing performance of diverse approaches to the conjecture generation
+problem, whilst avoiding some of the philosophical complications and ambiguities
+of the field. This methodology can be tailored for specific interests, such as
+the choice of problem set and the focus of the resulting analysis.
+
+We have also presented an example application for the domain of higher-order
+inductive theories (which is immediately applicable to problems in functional
+programming). Using the TIP problem set as a corpus, we have evaluated the
+performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
+both significantly faster and to also output more desirable conjectures than
+IsaCoSy; although more \emph{undesirable} output may be generated as well. Based
+on these results we proposed two possible directions to improve the QuickSpec
+algorithm: a post-processing filter to remove more ``uninteresting'' conjectures
+and a pre-processing filter to ``focus'' on promising subsets of the input.
+
+Other promising directions for future work include the evaluation of other MTE
+tools, the use of other corpora more suited to different problem domains, and
+the extension of existing corpora with new definitions and theorems (which would
+also benefit the ATP benchmarks they originally came from).
+
+We believe that a standard approach to benchmarking and comparison such as ours
+will ease the burden on researchers wanting to evaluate different potential
+approaches to this task, and provide a common goal to pursue in the short term.
 
 ``Solving'' this benchmark suite would not solve the problem of theory
-exploration in general, so more ambitious goals must be set in the future; but
-we believe that our approach will provide a compelling challenge, at least for
-the foreseeable future.
+exploration in general, so more ambitious goals must be set in the future. For
+now, we believe that our approach provides a compelling challenge and will
+encourage healthy competition to improve the field.
 

### 2018-03-16 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Write out stats to file

@@ -290,6 +290,8 @@ def makeColours(agg):
         maxVal    = max([agg[key][i] for i,s in enumerate(agg['success']) if s])
         colourMap = sns.color_palette('viridis', maxVal + 1).as_hex()
         msg('Highest {0} count is {1}'.format(key, maxVal))
+        with open('highest_{0}_count'.format(key), 'w') as f:
+            f.write(str(maxVal))
         return {
             'cmap'    : ListedColormap(colourMap),
             'norm'    : mpl.colors.Normalize(vmin=0, vmax=maxVal),

### 2018-03-16 TEBenchmarkPaper/article.tex
Abstract

@@
 \begin{abstract}
-  We propose a benchmark suite for evaluating the efficiency and effectiveness
-  of \emph{conjecture formation} by automated tools for \emph{mathematical
-    theory exploration} in higher-order, inductive theories; a domain especially
-  suited for analysing software. By providing standard tools and metrics, we
-  hope to encourage innovation and comparison between the disparate approaches
-  currently being pursued, and spur improvements similar to those seen in the
-  competitive field of automated theorem proving.
+  We propose a benchmarking methodology to evaluate the efficiency and quality
+  of \emph{conjecture generation} by automated tools for \emph{mathematical
+    theory exploration}. Our approach uses widely available theorem proving
+  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
+  QuickSpec and IsaCoSy tools, finding that the former takes significantly less
+  time to produce significantly more ``interesting'' output. By providing a
+  standard, cross-tool evaluation technique we hope to encourage innovation and
+  comparison between the disparate approaches currently being pursued, and spur
+  improvements similar to those seen in the competitive field of automated
+  theorem proving.
 %\keywords{First keyword \and Second keyword \and More}

### 2018-03-16 TEBenchmarkPaper/article.tex
Sections

@@
 
-Section \ref{sec:background} describes MTE and the conjecture generation problem
-in more detail, along with existing approaches and their evaluation
-methodologies. We explain our proposal for a more general benchmark in section
-\ref{sec:proposal} whilst section \ref{sec:application} shows the results when
-applied to existing MTE tools. Analysis of these results is given in section
-\ref{sec:discussion} and concluding remarks in section \ref{sec:conclusion}.
+We introduce the conjecture generation problem in more detail, along with the
+difficulty of comparing existing solutions, in $\S$\ref{sec:background}. In
+$\S$\ref{exploration-versus-formation} we give a history of the MTE field, and
+its close cousin Automated Theory Formation (ATF), describing a variety of
+existing approaches and clarifying the diverse terminology found in the
+literature. We explain our proposal for a more general benchmark in
+$\S$\ref{sec:proposal} and demonstrate its application to existing MTE tools
+in $\S$\ref{sec:application}. Issues facing our approach, and the field in
+general, are discussed in $\S$\ref{sec:discussion} and concluding remarks
+are given in $\S$\ref{sec:conclusion}.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Background

@@
 
-Equation \ref{eq:mapreduce} justifies a rewrite rule for splitting a single
+Equation~\ref{eq:mapreduce} justifies a rewrite rule for splitting a single
 $\Map$ call into multiple independent calls dealing with different sections of a
 list. Such rules can be useful optimisations, for example in a distributed
-context where since each call can be evaluated in parallel (this is the basis of
-the ``map/reduce'' programming paradigm).
+context where each call can be evaluated in parallel (this is the basis of the
+``map/reduce'' programming paradigm).
 
 All of these example use cases require the ability to discover theorems about
-some particular theory. This is a hard problem in general, but presents
+some particular definitions. This is a hard problem in general, but presents
 opportunities for automation due to the precise, symbolic nature of the domain.
@@
 user's particular task. For example, if we want to find optimising
-rewrite rules such as equation 1, then utility would include whether
-or not a conjecture justifies a rewrite rule, the difference in
+rewrite rules such as equation~\ref{eq:mapreduce}, then utility would include
+whether or not a conjecture justifies a rewrite rule, the difference in
 resource usage of the expressions involved, and how common those
@@
 
-The latter tools are described in more detail in section~\ref{section:mte},
+The latter tools are described in more detail in $\S$\ref{section:mte},
 but all use testing to check for counterexamples, which ensures results are
@@
 and examples. Both strands contrast themselves with Automated Theorem
-Proving. The main historical differences between these two approaches
-seem to be that in MTE:
-
-\begin{itemize}
-\item the main proponents define themselves as mathematicians (they
-  hold a PhD and have experience in research mathematics);
-\item the primary motivation is to support mathematicians;
-\item systems tend to be user-interactive;
-\item systems tend to be specific to mathematics,
-\end{itemize}
-
-while in ATF:
-
-\begin{itemize}
-\item the main proponents define themselves as AI researchers (they
-  often have a mathematics background up to Masters level, but their
-  PhD and research experience is in AI);
-\item the primary motivation is to extend AI techniques; a secondary
-  motivation is to produce interesting new mathematics;
-\item systems tend to be fully automated;
-\item systems can often be applied to non-mathematical domains.
-\end{itemize}
+Proving. The main historical differences between these two approaches seem to be
+that MTE was initiated by mathematicians as an attempt to support existing
+mathematical research, perhaps with user interaction; while ATF is a product of
+AI researchers trying to extend AI techniques, with full automation and often
+applications to non-mathematical domains.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Existing evaluations

@@
 Examples where this does \emph{not} hold include the natural number heuristics
-of AM discussed in section~\ref{section:atf}, as well as machine learning models
-trained on some specific dataset (such as this benchmark).
+of AM discussed in $\S$\ref{section:atf}, as well as machine learning models
+trained on some specific dataset (such as the benchmark we describe).
 
@@
 The analyses in~\cite{claessen2013automating} took their definitions and ground
-truths from the standard library of the Isabelle theorem prover: since the
+truths from the standard library of the Isabelle proof assistant: since the
 library authors have gone to the effort of stating, proving and including these
@@
 
-Two tests were performed: one on a theory of natural numbers and one on a theory
-of lists (whose definitions are shown in figure~\ref{figure:list_theory}). These
-tests are limited both in number (it is difficult to draw general conclusions
-from two datapoints) and by the small size of these theories (e.g. the natural
-number theory from~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
-only 4 definitions and 12 theorems).
-
-We believe that this is a viable approach to drawing meaningful, objective
-conclusions about such tools, but larger data sets are needed to judge
-performance in different, especially \emph{novel}, domains (which, after all, is
-the purpose of such tools). In addition, we might be concerned that such
-``popular'' theories may be unrepresentative of typical usage. For example, we
-would not expect the library functions in a typical verification problem to be
-as mathematically rich as number theory.
+Tests were performed with a theory of natural numbers and a theory of lists (the
+definitions shown in figure~\ref{figure:list_theory}), but it is difficult to
+draw general conclusions from only two datapoints. These theories were also
+quite small, the natural number theory
+from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
+definitions and 12 theorems.
+
+Despite these limitations we believe that this is a viable approach to drawing
+meaningful, objective conclusions about such tools; but only with larger data
+sets can we judge performance in different, especially \emph{novel}, domains
+(which, after all, is the purpose of such tools). In addition, we might be
+concerned that such ``popular'' theories may be unrepresentative of typical
+usage. For example, we would not expect the library functions in a typical
+verification problem to be as mathematically rich as number theory.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Application

@@
 \item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application translates to Haskell and Isabelle, which also
-  benefits from having definitions be strongly typed.
+  under study. Our application in $\S$\ref{sec:application} requires Haskell and
+  Isabelle translations, and also benefits from having definitions be strongly
+  typed.
 \item The problem set should be relevant to the desired domain. Our focus on
@@
   TPTP~\cite{sutcliffe2009tptp}).
-\item Since it will act as our ground truth, the problem set should ideally
-  contain every ``interesting'' conjecture involving its included
-  definitions. Realistically, we should aim for each definition to appear in
-  \emph{many} theorem statements; rather than each problem having unique
-  definitions.
+\item Since it will act as our ground truth, the problem set should
+  \emph{ideally} contain every ``interesting'' conjecture involving its included
+  definitions. More \emph{realistically}, we should aim for each definition to
+  appear in \emph{many} theorem statements; rather than each problem having
+  unique definitions.
 \item The problem set should be as large as possible, for robustness of the
@@
 
+We prefer the latter, where each generated conjecture (wanted or not) is like a
+coin toss whose bias is the precision; likewise, the ground truth theorems
+(found or not) are like coin tosses biased by the recall. Note that this simple
+model does not explain how many conjectures the tool decides to output for each
+run.
+
 \subsection{Comparison}
@@
 test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In our application we follow a variant of the Speedup-Test
-protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the differences
-are our use of the paired form of the Wilcoxon test, and the fact that each run
-was performed on a different sample (to reduce sampling bias, at the expense of
-accuracy).
-
-Another benefit to such a paired difference test is its robustness to the
+normally distributed. In $\S$\ref{sec:application} we follow a variant of the
+Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
+differences are our use of the paired form of the Wilcoxon test, and the fact
+that each run was performed on a different sample (to reduce sampling bias, at
+the expense of accuracy). This paired difference test is also robust to the
 censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results \emph{towards} the null
-hypothesis of indistinguishable performance.
-
-To compare the recall we use the standard McNemar test for paired samples: each
+to the observable difference, which biases our results in a conservative way
+(\emph{towards} the null hypothesis of indistinguishable performance).
+
+For simplicity, we suggest comparing quality by pooling together results from
+all samples where both tools finished successfully. Recall can be compared using
+the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
 ground-truth theorem is categorised as either being found only by the first
@@
 We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we use Boschloo's form of Barnard's test to
-compare the proportions of wanted and unwanted conjectures generated by each
-tool.
+different conjectures. Instead, we can use Boschloo's form of Barnard's
+test~\cite{lydersen2009recommended} to compare the proportions of wanted and
+unwanted conjectures generated by each tool, assuming a binomial distribution
+and conditioning on the number of generated conjectures (as if this were fixed
+by design).
 
@@
 To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each. Most QuickSpec runs
-either finished within 200 seconds or not at all, and sizes above 20 mostly
-timed out. IsaCoSy mostly finished within 300 seconds on sample sizes up to 4,
-but by size 8 was mostly timing out; its few successes above this took thousands
-of seconds each, which we deemed infeasibly long.
+for a few samples sized between 1 and 100, for an hour each on our benchmarking
+machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
+of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
+and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
+on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
+above this took thousands of seconds each, which we deemed infeasibly long.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Analysis

@@
 We chose TIP for our ground truth since it satisfies the desiderata of
-section~\ref{section:prep}: each problem has standalone type and function
+$\S$\ref{section:prep}: each problem has standalone type and function
 definitions, making their separation trivial; known examples from the software
@@
         (List a)))))         ;; Return type
-
+\end{verbatim}
+\end{samepage}
+
+\begin{samepage}
+\begin{verbatim}
 (define-fun
@@
 conjecturing equations involving Haskell functions, described in more detail in
-section~\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
+$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
 to automate some of the decisions which are normally left up to the user:
@@
 finished within 30 seconds, generating few or no conjectures; those taking
-longer generated more conjectures, with the most being 116 conjectures from a
-sample of size 16.
-
-Precision and recall are shown in figure~\ref{figure:quickspec_precRec}, which
-we model as simple Bernoulli processes: the generated conjectures (wanted or
-not) are like coin tosses whose bias is the precision; likewise, the ground
-truth theorems (found or not) are like coin tosses biased by the recall. Note
-that this simplification does not explain how many conjectures the tool decides
-to output for each run.
-
-QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
-generates monotonically more conjectures as definitions are added to a signature
-(assuming sufficient testing), it can't be finding fewer correct conjectures at
-larger sizes. This is supported by the relative flatness of the recall results.
-Rather, the number of conjectures generated is increasing at a higher rate than
-the size of the ground truth. These extra conjectures may involve those
-``padding'' definitions in a sample which don't contribute to its ground truth,
-or may be ``uninteresting'' relationships between the dependencies of different
-ground truth theorems.
+longer generated more conjectures, with the most being 133 conjectures from a
+sample of size 19.
+
+Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
+Since QuickSpec generates monotonically more conjectures as definitions are
+added to a signature (assuming sufficient testing), its decreasing precision
+can't be due to finding fewer wanted conjectures at larger sizes. This is
+supported by the relative flatness of the recall results. Rather, the number of
+conjectures generated is increasing at a higher rate than the size of the ground
+truth. These extra conjectures may involve those ``padding'' definitions in a
+sample which don't contribute to its ground truth, or may be ``uninteresting''
+relationships between the dependencies of different ground truth theorems.
 
@@
     \hline
+    \hline
   \end{tabular}
@@
 
-Precision was also compared by pooling results from samples where both tools
-finished successfully. We cannot perform McNemar's test in this case, since each
-tool generated different sets of conjectures. Instead, we add up the number of
+McNemar's test is not applicable for comparing precision, since each tool
+generated different sets of conjectures. Instead, we add up the number of
 results which are ``interesting'' (appear in the ground truth) and those which
@@
     \bfseries IsaCoSy   & 32          & 137      \\
+    \hline
     \bfseries QuickSpec & 47          & 301      \\
@@
 
-These totals were tested for independence using Boschloo's
-test~\cite{lydersen2009recommended}, assuming a binomial distribution and
-conditioning on the number of generated conjectures (as if this were fixed by
-design). This produces a p-value of 0.111, which given our (arbitrary) choice of
-$\alpha = 0.05$ is not significant enough to reject the null hypothesis that
-their precision is equal.
-
-Note that this is the only comparison where IsaCoSy performed better (although
-not significantly so) with 19\% precision compared to QuickSpec's 14\%. However,
-precision on its own is not the whole story: it can be increased at the expense
-of recall by making the tool more conservative. IsaCoSy's significantly lower
-recall indicates that it may be overly-conservative compared to QuickSpec, which
-would also explain its precision.
+We test these totals for independence using Boschloo's test and find a p-value
+of 0.111, which exceeds our (arbitrary) choice of $\alpha = 0.05$; hence we do
+not consider the difference between these proportions (14\% for QuickSpec, 19\%
+for IsaCoSy) to be significant.
+
+Note that precision is the only comparison where IsaCoSy scored higher, and even
+that was not found to be significant. Also, precision on its own is not the
+whole story: it can be increased at the expense of recall by making a tool more
+conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
+given that its recall is significantly lower.
 
@@
 samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean heavily
-in favour of QuickSpec simply because it generated so much more than IsaCoSy in
-the available time.
-
-% To do this we can take a set of ATP tasks, for example proving the equation
-% \ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
-% split apart the definitions (to form a theory) from the theorem statements (to
-% form our ground truth).
+them as if they succeeded with no output, the resulting statistics lean
+overwhelmingly in favour of QuickSpec simply because it generated so much more
+than IsaCoSy in the available time.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Fix table values
Discussion and conclusion

@@
     \bfseries Found by QuickSpec     & 28 & 19 \\
-    \bfseries Not found by QuickSpec & 4  & 28 \\
     \hline
+    \bfseries Not found by QuickSpec & 4  & 75 \\
     \hline
@@
 on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future research.
+practices and offer a common ground for future endeavours.
 
@@
 appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. However, we do not believe this to be a realistic problem for the
-time being (we would be delighted to be proved wrong!).
+such output. We do not believe this to be a realistic problem for the time
+being, although we would be delighted if proved wrong.
 
@@
 the extension of existing corpora with new definitions and theorems (which would
-also benefit the ATP benchmarks they originally came from).
+also be of benefit to the original ATP benchmarks).
 
@@
 
-``Solving'' this benchmark suite would not solve the problem of theory
-exploration in general, so more ambitious goals must be set in the future. For
-now, we believe that our approach provides a compelling challenge and will
-encourage healthy competition to improve the field.
+``Solving'' this benchmark would not solve the problem of conjecture generation
+in general, so more ambitious goals must be set in the future. For now, we
+believe that our approach provides a compelling challenge and will encourage
+healthy competition to improve the field.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Reference QuickSpec, HipSpec and Hipster

@@
 enumerating expressions involving a given set of (typed) constants and free
-variables (a signature). A constraint solver forbids certain (sub)expressions
-from being synthesised, and these constraints are extended whenever a new
-conjecture is generated, to avoid generating any special-cases of this
-conjecture in the future. Conjectures are tested by sending them to the
-QuickCheck counterexample finder and, if no counterexamples are found, then
-sent to IsaPlanner which attempts to prove them.
+variables (a \emph{signature}). A constraint solver forbids certain
+(sub)expressions from being synthesised, and these constraints are extended
+whenever a new conjecture is generated, to avoid generating any special-cases of
+this conjecture in the future. Conjectures are tested by looking for
+counterexamples and, if none are found, sent to IsaPlanner which attempts to
+prove them.
 
@@
 QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language, and has been used to generate conjectures for the
-subsequent HipSpec and Hipster MTE systems. Like IsaCoSy, a signature is
-provided and terms are enumerated. Rather than a constraint system, QuickSpec
-version 1 collects expressions of the same type into equivalence classes,
-assuming them to be equal. QuickCheck is used to test this assumption by
-instantiating each variable with a randomly chosen value, evaluating the
-resulting expressions and comparing them. Any equivalence class whose elements
-don't compare equal are split up to separate those observably-different
-expressions, and the process is repeated with new random values.
+Haskell programming language~\cite{claessen2011quickcheck}, and has been used to
+generate conjectures for the subsequent
+HipSpec~\cite{Claessen_hipspec:automating} and Hipster~\cite{Hipster} MTE
+systems. Like IsaCoSy, a signature is provided and terms are enumerated. Rather
+than a constraint system, QuickSpec version 1 collects expressions of the same
+type into equivalence classes, assuming them to be equal. QuickCheck is used to
+test this assumption by instantiating each variable with a randomly chosen
+value, evaluating the resulting expressions and comparing them. Any equivalence
+class whose elements don't compare equal are split up to separate those
+observably-different expressions, and the process is repeated with new random
+values.
 

### 2018-03-16 TEBenchmarkPaper/article.tex
Acknowledgements

@@
 
-%\begin{acknowledgements}
+\begin{acknowledgements}
+  We are grateful to the authors of the systems we have used in our experiments,
+  especially for help in obtaining and configuring their tools. We would
+  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
+  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
+  their software to our purposes.
 %  EPSRC
-%\end{acknowledgements}
+\end{acknowledgements}
 

### 2018-03-21 PhDThesis/outline.tex
Make outline into LaTeX

Title: Use of "Interestingness" in Scaling Mathematical Theory Exploration

Introduction:

 - Description of problem
 - Motivation

Background (/Literature review)

 - History of automated mathematical discovery
 - History of what counts as "interesting"
 - Relevance filtering

Contributions:

 - Empirical proposal for measuring what is "interesting" (TEBenchmark)
 - Evaluation and comparison of existing systems using this benchmark (QuickSpec
   v1 and IsaCoSy)

Application of machine learning to focus existing systems on what is "interesting"

 - Overview of machine learning techniques:
  - Supervised vs unsupervised
  - Clustering
  - Representing structured inputs
 - Bucketing algorithm
  - Use with pseudorandom similarity metric
  - Use with "flattened tree" similarity metric
  - Potentially some other comparison?

Practical uses of automated mathematical discovery systems

 - Optimisation
  - Self-optimising programs
  - Use with Hutter search?
 - Finding abstractions (anti-unification)
  - Concept formation?

Application of discovery system to existing software packages

 - Packages from Hackage (may need extra Arbitrary instances)
 - GHC libraries (base, maybe more)
 - Maybe use test suites as ground truth?

Conclusions

---

We have a few things sat around gathering dust. Can we fit them in somewhere?

 - Purely-functional self-modifying code. This could be an implementation
   vehicle for open-ended self-optimising code.

 - Powerplay. This uses self-invented problems. Could we use "interesting"
   conjectures from a theory exploration system to guide the self-improvement of
   an automated theorem prover?

Things which have influenced my thinking, which would be nice to include:

 - Interestingness based on compression progress. The "compression" could
   perhaps come from writing proof tactics? For example, try to prove any given
   conjecture using a single 'prove' tactic. Given a (conjecture, tactic) pair,
   and an existing tactic, we deem the conjecture to be "interesting" if it
   can't be proved by the existing tactic, it can be proved by the new tactic,
   and the new tactic is smaller than the old tactic. Hmm... This wouldn't allow
   tactics to increase in size. The compression progress idea relies on a
   lossless compression of all 'past experience', which in this situation would
   just be a list of all conjectures we've tried. This just passes the buck on
   what's "interesting", since those conjectures need to be chosen somehow...

 - Artificial curiosity systems, like problem generators rewarded by the
   variance of how difficult a population of solvers found the problem to be.

 - Recursive self-improvement: take an algorithm which searches for improvements
   to itself, and run it for a few self-improvement iterations. How do the
   improvements to the search algorithm compare against the difficulty of
   improving an optimised system ('recalcitrance')?

 - Can we create a compiler by providing a property discovery algorithm with a
   specification and an model interpreter (e.g. for x86 machine code)?

 - A result I'd really like to have is a system which collapses down abstraction
   layers. This would be especially nice if it worked across languages, and we
   could either use staged execution/compilation and/or a database of
   compiled/transformed code.

### 2018-04-06 TEBenchmarkPaper/article.tex
Reword

@@
   time to produce significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we hope to encourage innovation and
-  comparison between the disparate approaches currently being pursued, and spur
-  improvements similar to those seen in the competitive field of automated
-  theorem proving.
+  standard, cross-tool evaluation technique we allow the disparate approaches
+  currently being pursued to be more directly compared. Our hope is to encourage
+  innovation and improvements similar to those seen in fields like automated
+  theorem proving, where the availability of benchmarks encourages healthy
+  competition.
 %\keywords{First keyword \and Second keyword \and More}
@@
 
-\emph{Conjecture generation/formation}, a sub-field of \emph{mathematical theory
-  exploration} (MTE), is the open-ended problem of producing conjectures about a
-given logical theory which are somehow ``interesting''.
+\emph{Conjecture generation} is the open-ended task of producing conjectures
+about a given logical theory which are somehow ``interesting'', and is studied
+as a sub-field of \emph{mathematical theory exploration} (MTE).
 
@@
 
-The task of discovering theorems in an arbitrary theory can be
-described as the interplay of several processes: (i) formulating new
-definitions and concepts; (ii) finding patterns suitable for posing as
-conjectures; and (iii) proving that those conjectures are theorems.
-The latter is studied extensively in the field of Automated Theorem
-Proving (ATP). Far less attention has been paid to automating the
-first two tasks, which both Automated Theory Formation (ATF) and
-Mathematical Theory Exploration (MTE) address.  We discuss the relationship
-between ATF and ATE in historical context in
+The task of automatically proving or disproving a \emph{given} conjecture has
+been studied extensively in the field of Automated Theorem Proving (ATP). Less
+attention has been paid to \emph{generating} such conjectures automatically,
+either from a given set of definitions (the task we will focus on) or where the
+definitions are also generated. The fields of Automated Theory Formation (ATF)
+and Mathematical Theory Exploration (MTE) both tackle these generation tasks,
+and we discuss their relationship in historical context in
 $\S$\ref{exploration-versus-formation}.

### 2018-04-06 PhDThesis/background.tex
Add files to thesis dir

\section{Background}

\subsection{Conjecture Generation}

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
  \label{figure:list_theory}
\end{figure}

We focus on the task of generating conjectures from some given mathematical
theory. This theory may be, for example, a scientific model, an executable
computer program or simply an object of mathematical curiosity. A theory
defines (perhaps implicitly) a particular system of logic in which to work (for
example higher-order logic), along with a finite set of \emph{definitions} (like
those in figure~\ref{figure:list_theory}) which assign structure or meaning to
particular terms. Theoretically, this is all of the information we need to
derive the truth or falsity of all decidable statements involving these terms,
for example by enumerating all valid proofs. However, due to the exponential
search spaces involved, such brute-force enumeration is infeasible in practice;
plus the resulting proofs would be difficult to comprehend and the processed
statements would be overwhelming in number and mostly uninteresting in content.

The impracticality of enumeration demonstrates that a statement being ``true''
or ``provable'' is not \emph{sufficient} to warrant the (limited) attention of a
human operator. We would also argue that it is not \emph{necessary}, since there
are many statements considered to be interesting whose truth is either unknown,
or which later turned out to be false. We thus focus on the automated production
of \emph{interesting conjectures}, with the primary focus being on efficient
generation and interest to the user.

As an example, we use the theory of lists shown in
figure~\ref{figure:list_theory}. These definitions are so widely used in
software that they appear in the standard libraries of many programming
languages. An example of a conjecture involving these definitions is the
following universally quantified equation:

\begin{equation} \label{eq:mapreduce}
  \forall f. \forall xs. \forall ys.
    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
\end{equation}

Equation~\ref{eq:mapreduce} states that combining small lists into a larger one
using $\Append$ then transforming the elements with $\Map$ is the same as
transforming the elements of the small lists then combining. This is interesting
due to its applicability as an optimisation: if a program calls $\Map$ with an
expensive transformation $f$, we can divide up the work across multiple machines
in parallel (this is the basis of the ``map/reduce'' programming paradigm).
Whether or not this equation holds depends on the choice of logical framework,
and hence (in the case of software) on the semantics of the programming
language: in particular this equation and optimisation are invalid for systems
with strict evaluation order and side-effects (common in imperative programming)
since any error/exception in the calculation of $ys$ will cause the surrounding
calls to abort; on the left-hand-side this will abort $\Append$ then $\Map$, so
$f$ will never be executed; on the right-hand-side there is an additional
$\Map(f, xs)$ call which will \emph{not} be aborted, which may execute $f$ many
times and trigger arbitrary observable effects.

\section{Machine Learning}

Machine learning (ML) is a sub-field of artificial intelligence (AI) which
emphasises the use of statistical methods to find patterns and optimise
functions, in contrast with symbolic reasoning approaches like automated theorem
proving (commonly referred to as ``Good Old-Fashioned AI'').

The machine learning umbrella spans many techniques and applications, but one
common distinction is between \emph{supervised} and \emph{unsupervised} tasks.
In supervised learning we allow the system to perform some behaviour, then an
external process assesses whether it behaved well or poorly (akin to Justice
Potter Stewart's test for obscenity: ``I know it when I see it''). The system's
parameters are optimised to try and score highly on the test. Common supervised
learning tasks include function approximation using input/output examples (e.g.
classifying scanned images of numerals) where the test is the proportion of
example inputs which got the correct output, and reinforcement learning where
the system chooses from a set of ``actions'', receives a pair of ``observation''
and ``reward'' (test score) in response, and must accumulate the most reward.

Unsupervised learning does not have the same testable ``success'' criteria as
supervised learning. Instead, the goal is to find patterns and form models of
the target domain, which is more ambitious and open-ended than supervised
learning but also much harder to define, measure and optimise. Examples of this
are prediction and compression tasks such as auto-encoders
(learned compressor/decompressor pairs, rewarded by how well they preserve data
through a round-trip) and adversarial or co-evolution tasks (coupled systems
where each is rewarded by exploiting mis-prediction in the others).

### 2018-04-06 PhDThesis/outline.tex
Thesis Nix changes

@@
-Title: Use of "Interestingness" in Scaling Mathematical Theory Exploration
+\title{``Interestingness'' in Mathematical Theory Exploration}
+\documentclass{article}
+\begin{document}
+\maketitle
 
-Introduction:
+\begin{itemize}
+\item Introduction
+  \begin{itemize}
+  \item Description of problem (simple examples)
+  \item Motivation (education, research, formalising, verification, ...)
+  \item Difficulties (ambiguity, open-endedness, large search space, ...)
+  \end{itemize}
 
- - Description of problem
- - Motivation
+\item Contributions:
+  \begin{itemize}
+  \item TEBenchmark (large, unambiguous empirical measurement and comparison)
+  \item MLSpec (bucketing methods, recurrent clustering, etc.)
+  \end{itemize}
 
-Background (/Literature review)
+\item Background/Literature review
+  \begin{itemize}
+  \item History of automated discovery/creativity
+    \begin{itemize}
+    \item Theorem proving (Logic Theorist)
+    \item Theory formation/exploration (AM, HR, Graffiti, IsaCoSy, IsaScheme,
+      QuickSpec, Speculate, ...)
+    \item Scientific discovery (Adam/Eve, invariant discovery)
+    \item Artificial curiosity (model building, motor babbling, play, language invention, ...)
+    \item Artistic creativity?
+    \end{itemize}
+  \item History of "interestingness"
+    \begin{itemize}
+    \item From philosophical perspective (surprising, novel, useful, Wundt
+      curves, ...)
+    \item From real implementations (non-redundancy, precision/recall,
+      compression progress, learnability, ...)
+    \end{itemize}
+  \item Overview of machine learning techniques:
+    \begin{itemize}
+    \item Models (neural networks, generative models, probabilistic programming,
+      gradient descent, kernel methods, population models, ...)
+    \item Supervised (classification, reinforcement learning, ...)
+    \item Unsupervised (clustering, co-evolution, adversarial, ...)
+    \item Feature engineering
+      \begin{itemize}
+      \item Hand-crafted features (comparison with probabalistic programming?)
+      \item Distributed representations
+      \item Learned representations (auto-encoding, deep learning, convolutional
+        neural networks, ...)
+      \end{itemize}
+    \item Structured representations (recurrent NNs, recursive NNs, tree
+      kernels, Tree Based CNN, generative models, ...)
+    \end{itemize}
+  \item (Statistical) machine learning in formal systems
+    \begin{itemize}
+    \item Relevance filtering (Sledgehammer, Josef Urban's alternatives)
+    \item Recurrent clustering
+    \item Program manipulation (Learning to Execute, Tree Based CNN tasks
+      (program classification, bubble-sort identification), auto-generated bug
+      fixes, auto-generated compiler test cases, ...
+    \end{itemize}
+  \end{itemize}
 
- - History of automated mathematical discovery
- - History of what counts as "interesting"
- - Relevance filtering
+\item TEBenchmark (mostly from paper)
+  \begin{itemize}
+    \item Methodology
+    \item TIP corpus
+    \item Haskell translation (specific issues, etc.)
+    \item Isabelle translation (specific issues, etc.)
+    \item QuickSpec analysis
+    \item IsaCoSy analysis
+    \item QuickSpec/IsaCoSy comparison
+    \item Discussion of pros and cons
+  \end{itemize}
 
-Contributions:
+\item Bucketing (mostly paper and transfer report)
+  \begin{itemize}
+  \item Justification (exponential running times, large numbers of discarded
+    equations, low precision, ...)
+  \item Bucketing as a solution (big-O analysis, oracles, ...)
+  \item Bucketing algorithms
+    \begin{itemize}
+    \item Pseudorandom features (baseline)
+    \item Circular convolution (language agnostic, distributed representation,
+      not learned)
+    \item Recurrent clustering (Haskell-specific, not distributed, hand-crafted,
+      solves global name problem)
+    \item Non-recurrent clustering?
+    \end{itemize}
+  \item Application to TEBenchmark (impact on total recall)
+  \item Discussion
+    \begin{itemize}
+    \item Pros and cons
+    \end{itemize}
+  \end{itemize}
 
- - Empirical proposal for measuring what is "interesting" (TEBenchmark)
- - Evaluation and comparison of existing systems using this benchmark (QuickSpec
-   v1 and IsaCoSy)
+\item Other contributions (may splice into the other sections)
+  \begin{itemize}
+    \item Sandboxed evaluation of Haskell code (nix-eval)
+    \item Haskell AST extraction and annotation (GHC plugin, standalone command)
+    \item Exploration of popular Hackage packages
+  \end{itemize}
 
-Application of machine learning to focus existing systems on what is "interesting"
+\item Conclusions
+  \begin{itemize}
+  \item Exploration is an important problem with many applications
+  \item Hampered by ambiguous goals, small sample sizes, etc.
+  \item Hampered by (lack of) efficiency
+  \item Unambiguous benchmark is provided for comparison
+  \item ``Smart'' bucketing method to improve efficiency
+  \item Lots of room for future work (other corpora, other tools, other
+    languages, applications of results, optimisation through profiling,
+    abstraction invention through anti-unification, self-modelling, ...)
+  \end{itemize}
+\end{itemize}
 
- - Overview of machine learning techniques:
-  - Supervised vs unsupervised
-  - Clustering
-  - Representing structured inputs
- - Bucketing algorithm
-  - Use with pseudorandom similarity metric
-  - Use with "flattened tree" similarity metric
-  - Potentially some other comparison?
-
-Practical uses of automated mathematical discovery systems
-
- - Optimisation
-  - Self-optimising programs
-  - Use with Hutter search?
- - Finding abstractions (anti-unification)
-  - Concept formation?
-
-Application of discovery system to existing software packages
-
- - Packages from Hackage (may need extra Arbitrary instances)
- - GHC libraries (base, maybe more)
- - Maybe use test suites as ground truth?
-
-Conclusions
-
----
-
-We have a few things sat around gathering dust. Can we fit them in somewhere?
-
- - Purely-functional self-modifying code. This could be an implementation
-   vehicle for open-ended self-optimising code.
-
- - Powerplay. This uses self-invented problems. Could we use "interesting"
-   conjectures from a theory exploration system to guide the self-improvement of
-   an automated theorem prover?
-
-Things which have influenced my thinking, which would be nice to include:
-
- - Interestingness based on compression progress. The "compression" could
-   perhaps come from writing proof tactics? For example, try to prove any given
-   conjecture using a single 'prove' tactic. Given a (conjecture, tactic) pair,
-   and an existing tactic, we deem the conjecture to be "interesting" if it
-   can't be proved by the existing tactic, it can be proved by the new tactic,
-   and the new tactic is smaller than the old tactic. Hmm... This wouldn't allow
-   tactics to increase in size. The compression progress idea relies on a
-   lossless compression of all 'past experience', which in this situation would
-   just be a list of all conjectures we've tried. This just passes the buck on
-   what's "interesting", since those conjectures need to be chosen somehow...
-
- - Artificial curiosity systems, like problem generators rewarded by the
-   variance of how difficult a population of solvers found the problem to be.
-
- - Recursive self-improvement: take an algorithm which searches for improvements
-   to itself, and run it for a few self-improvement iterations. How do the
-   improvements to the search algorithm compare against the difficulty of
-   improving an optimised system ('recalcitrance')?
-
- - Can we create a compiler by providing a property discovery algorithm with a
-   specification and an model interpreter (e.g. for x86 machine code)?
-
- - A result I'd really like to have is a system which collapses down abstraction
-   layers. This would be especially nice if it worked across languages, and we
-   could either use staged execution/compilation and/or a database of
-   compiled/transformed code.
+\end{document}

### 2018-04-06 Bucketing/background.tex
Bucketing changes (mostly long lines)

--- 2016-03-02 CICM2016/background.tex
@@
 
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and \hspec{} \cite{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
+We decided to focus on the Haskell programming language since it has direct
+applications in industry, a large selection of library code for testing and
+mature, state-of-the-art MTE tools including \qspec{}~\cite{QuickSpec},
+\hspec{}~\cite{claessen2013automating} and
+\textsc{Speculate}~\cite{braquehais2017speculate}. Our overall approach is
+agnostic to the underlying system, so could also be applied to other targets
+such as the Isabelle/HOL, which has the
+\textsc{IsaCoSy}~\cite{johansson2009isacosy}, \textsc{IsaScheme}~\cite{omar} and
+more recent \textsc{Hipster}~\cite{Hipster} MTE tools. However, since
+\textsc{Hipster} works by translating to Haskell and invoking \hspec{}, it seems
+that improving the Haskell tools will give the most benefit.
 
-Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-To avoid the relatively large and complex syntax of Haskell proper, we will focus on an intermediate representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described in Appendix \ref{sec:core}.
+Like most functional programming languages, Haskell builds upon
+$\lambda$-calculus, with extra features such as a strong type system and
+``syntactic sugar'' to improve readability. To avoid the relatively large and
+complex syntax of Haskell proper, we will focus on an intermediate
+representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described
+in Appendix \ref{sec:core}.
 

### 2018-04-06 Bucketing/intro.tex
Bucketing changes (mostly long lines)

\section{Introduction}

Functional programming claims to be more ``mathematically tractable'' than other
``conventional'' approaches, due to properties like \emph{referential
transparency} and \emph{immutability}, which allow equational reasoning without
many of the complications caused by side effects and strict evaluation
order~\cite{hughes1989functional}. Languages such as Haskell combine these
advantages with static type systems to restrict the possible values of every
expression, making programs amenable to powerful analysis tools such as property
checking~\cite{claessen2011quickcheck} and theorem
proving~\cite{rosen2012proving}.

Whilst such tools can support or refute user-provided properties, we are
interested in the related problem of \emph{discovering} novel properties
automatically; a task addressed by the field of \emph{Mathematical Theory
  Exploration}. Whilst clever algorithms are used to guide their search,
existing MTE tools, including those applied to Haskell, ultimately rely on
enumerating all (well-typed) expressions. This limits their usefulness, as it is
only feasible to explore a small collection of functions at a time, relying on
the user to select promising combinations. We investigate the use of machine
learning methods to perform this selection automatically, removing the need for
user intervention whilst maintaining the ability to discover desirable
properties.

Our contributions are:

\begin{enumerate}
  \item The application of machine learning algorithms to theory exploration,
    for intelligently discovering interesting sub-sets of Haskell libraries,
    which are more tractable to explore.
  \item A novel feature extraction method for transforming Haskell expressions
    into a form amenable to off-the-shelf learning algorithms.
  \item An implementation of these feature extraction and theory exploration
    approaches.
  \item A comparison of our methods with existing approaches, both for theory
    exploration in Haskell, and for machine learning in other languages.
\end{enumerate}

We describe our treatment of Haskell expressions in \S~\ref{sec:haskell} and
describe the \qspec{} theory exploration system we build upon in
\S~\ref{sec:theoryexploration}. We discuss our contributions in more depth in
\S~\ref{sec:contributions}, and provide implementation details in
\S~\ref{sec:implementation}. A variety of related work is surveyed in
\S~\ref{sec:related}, we briefly evaluate our implementations in
\S~\ref{sec:evaluation} and give several potential directions for future
research in \S~\ref{sec:future}.

### 2018-04-10 TEBenchmarkPaper/article.tex
Move historical section into discussion
Added FIXMEs for things brought up in meeting

@@
 % Lay out the field of TE: motivation, tools, evaluation, problems...
+% FIXME: Strengthen motivation:
+%  - Automated Scientific Discovery (e.g. Dundee drug discovery "centaur" story)
+%  - "Mixed initiative" (social machine?); maybe Kasparov
+%  - Gowers calling for better AI in maths
 
@@
 assistants and their libraries, in mathematics education and research, and in
-the specification, verification, optimisation and testing of software.
-
-Existing attempts at tackling this problem are difficult to compare, due
-partially to the variety of approaches taken, but also because of the inherent
-ambiguity of the task and the different goals emphasised by their designers and
-their choice of evaluation method.
-
-We attempt to solve this discrepancy, at least for the foreseeable future, by
-defining a standard, unambiguous benchmarking approach with which to compare
-the conjecture generation of MTE tools. Our contributions include:
+the specification, verification, optimisation and testing of software. Similar
+tasks are also studied in Computational \emph{Scientific}
+Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling},
+although that relies on inductive reasoning and experimental testing which (in
+principle) are not so important for pure mathematics. Whilst we \emph{have}
+found these methods to also be important for MTE tools (in particular for the
+\emph{plausibility} of generated conjectures), we leave a more thorough
+application of our results to scientific reasoning for future work.
+
+Even limiting ourselves to mathematics applications, it is difficult to measure
+and compare the success of approaches and tools. This is partially due to their
+diversity, but also because of the inherent ambiguity of the task, the different
+goals emphasised by their designers and the variety of evaluation methods
+employed. We attempt to solve this discrepancy, at least for the foreseeable
+future, by defining a standard, unambiguous benchmarking approach with which to
+compare the conjecture generation of MTE tools. Our contributions include:
 
@@
 We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. In
-$\S$\ref{exploration-versus-formation} we give a history of the MTE field, and
-its close cousin Automated Theory Formation (ATF), describing a variety of
-existing approaches and clarifying the diverse terminology found in the
-literature. We explain our proposal for a more general benchmark in
+difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We explain our proposal for a more general benchmark in
 $\S$\ref{sec:proposal} and demonstrate its application to existing MTE tools
 in $\S$\ref{sec:application}. Issues facing our approach, and the field in
-general, are discussed in $\S$\ref{sec:discussion} and concluding remarks
-are given in $\S$\ref{sec:conclusion}.
+general, are discussed in $\S$\ref{sec:discussion}; related work is surveyed in
+$\S$\ref{sec:related-work}, including the diverse terminology found in the
+literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
 
@@
 % interestingness, e.g. examples from textbooks, etc.
+
+% FIXME: Perhaps mention examples like "plus(x, 0) = 0" as motivating examples
+% of things which are "simple" but which proof assistants still don't do
+% automatically (reference Moa's use of TE for background lemma generation)
 
@@
 
+% FIXME: We jump from conjectures to theorems. Maybe to prevent confusion we
+% should talk more generally about "interesting properties" of a formal system,
+% and use that as motivation for conjecturing and proving?
+
 Given a logical theory, such as the theory of lists shown in
@@
 and we discuss their relationship in historical context in
-$\S$\ref{exploration-versus-formation}.
+$\S$\ref{related-work}.
 
@@
 
+% FIXME: This is original work, so should be in a different section
 \begin{table}
@@
 
+% FIXME: Give brief description and reference for each system in the text.
+% We could include the more lengthy descriptions of IsaCoSy and QuickSpec here?
+
 We summarise the criteria in Table 1 and state which were used in key
@@
 
-\section{Theory Exploration versus Theory
-  Formation}\label{exploration-versus-formation}
-
-In this section we put the Mathematical Theory Exploration approach
-into context both historically, and with respect to related approaches
-such as in Automated Theory Formation tools.
-
-\subsection{A Short History of Automated Mathematics}
-
-\begin{quote}
-...``in his subsequent design for an Analytical Engine Mr. Babbage has
-shown that material machinery is capable, in theory at least, of
-rivalling the labours of the most practised mathematicians in all
-branches of their science.''~\cite[p. 498]{jevons}
-\end{quote}
-
+\subsection{Existing Evaluations}
+\label{sec:existing}
+
+There are three aspects to consider when evaluating a tool for conjecture
+generation:
+
+% FIXME: How are these referred to/considered in the paper?
+\begin{enumerate}
+\item The quality of each individual conjecture, which we will refer to as their
+  ``interestingness''.
+\item The quality of the \emph{set} of generated conjectures, which is important
+  for criteria like novelty, as well as to assess consistency.
+\item Resources used by the tool, to find the balance struck between output
+  quality and time taken.
+\end{enumerate}
+
+We make the assumption that each tool is intended to be \emph{general}, and
+hence isn't hard-coded or fine-tuned to only perform well in a narrow domain.
+Examples where this does \emph{not} hold include the natural number heuristics
+of AM discussed in $\S$\ref{section:atf}, as well as machine learning models
+trained on some specific dataset (such as the benchmark we describe).
+
+Evaluation of new tools and techniques for conjecture generation tends to be
+\emph{ad hoc} and hence difficult to compare, whilst comparative surveys (such
+as Colton's) are descriptive and qualitative rather than providing direct
+numerical measurements.
+
+One quantitative approach which has gained traction is \emph{precision/recall
+  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
+providing a direct comparison of these three
+tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
+  reports prior results from the other tools, rather than reproducing them.
+  This makes the numbers less comparable, since each tool was tested with
+  slightly different definitions (e.g. whether or not the exponential and
+  maximum functions were included).} To measure precision and recall we first
+choose a set of definitions (such as those in figure \ref{figure:list_theory})
+and a \emph{ground truth} to treat as the ``ideal'' set of conjectures we would
+like a tool to produce for those definitions (e.g. this might include
+equation~\ref{eq:mapreduce}). Each tool is run and their outputs are compared to
+the ground truth. To score 100\% on precision and recall, all of the conjectures
+which appear in the ground truth must be generated, and nothing else:
+
+\begin{itemize}
+\item \emph{Precision} is the proportion of a tool's output which appears in
+  the ground truth. This penalises overly-liberal tools which output large
+  numbers of conjectures in the hope that some turn out to be ``good''.
+\item \emph{Recall} is the proportion of the ground truth which appears in the
+  tool's output. This penalises overly-conservative tools which generate very
+  few conjectures as a way to avoid ``bad'' ones.
+\end{itemize}
+
+Precision and recall act as quality tests for sets of conjectures, fulfilling
+one of our requirements. They also provide a coarse quality assessment for
+individual conjectures: either a conjecture appears in the ground truth set or
+it doesn't.
+
+The analyses in~\cite{claessen2013automating} took their definitions and ground
+truths from the standard library of the Isabelle proof assistant: since the
+library authors have gone to the effort of stating, proving and including these
+theorems in every copy of their software, this is a good indication that they
+are interesting.
+
+Tests were performed with a theory of natural numbers and a theory of lists (the
+definitions shown in figure~\ref{figure:list_theory}), %FIXME: add page ref
+but it is difficult to
+draw general conclusions from only two datapoints. These theories were also
+quite small, the natural number theory
+from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
+definitions and 12 theorems.
+
+Despite these limitations we believe that this is a viable approach to drawing
+meaningful, objective conclusions about such tools; but only with larger data
+sets can we judge performance in different, especially \emph{novel}, domains
+(which, after all, is the purpose of such tools). In addition, we might be
+concerned that such ``popular'' theories may be unrepresentative of typical
+usage. For example, we would not expect the library functions in a typical
+verification problem to be as mathematically rich as number theory.
+
+\section{Theory Exploration Benchmark}
+\label{sec:proposal}
+
+Our main contribution is a benchmarking methodology, shown in figure
+\ref{figure:flow_chart}, which both generates a large definition/ground-truth
+corpus, and provides a scalable, statistical approach to evaluating MTE tools
+using this corpus. We follow a precision/recall approach similar to prior work,
+with the main difference being the source of definitions and ground truths: we
+take existing problem sets designed for automated theorem proving, and adapt
+their content for use in the conjecture generation setting.
+
+\subsection{Preparation}
+\label{section:prep}
+
+Automated theorem proving is an active area of research, with large problem sets
+and regular competitions to prove as much as possible, as fast as
+possible~\cite{pelletier2002development}. These problem sets are an opportunity
+for MTE, as their definitions and theorems can be used as a corpus in the same
+way that Isabelle libraries have been used in the past.
+
+Some problem sets are more amenable for this purpose than others. The most
+suitable are those which have the following properties:
+
+\begin{itemize}
+\item For each problem, there should be a clear distinction between the theorem
+  to be proved and the definitions involved, such that the two can be easily and
+  meaningfully separated. This rules out problem sets like those of
+  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
+  functions, whose behaviour is \emph{implicit} in the logical structure of the
+  theorem statement but not separately \emph{defined}.
+\item Definitions should be translatable into a form suitable for the MTE tools
+  under study. Our application in $\S$\ref{sec:application} requires Haskell and
+  Isabelle translations, and also benefits from having definitions be strongly
+  typed.
+\item The problem set should be relevant to the desired domain. Our focus on
+  functional programming requires higher-order functions and inductively defined
+  datatypes, which rules out first-order languages/logics (such as
+  TPTP~\cite{sutcliffe2009tptp}).
+\item Since it will act as our ground truth, the problem set should
+  \emph{ideally} contain every ``interesting'' conjecture involving its included
+  definitions. More \emph{realistically}, we should aim for each definition to
+  appear in \emph{many} theorem statements; rather than each problem having
+  unique definitions. % FIXME: Reword: make clear that non-membership means
+  % uninteresting, and hence the things we *don't* include are just as important
+  % as the things we do
+\item The problem set should be as large as possible, for robustness of the
+  resulting statistics. % FIXME: Prioritising quantity over quality?
+\end{itemize}
+
+Once such a problem set has been chosen, we must separate the definitions
+referenced by the theorems from the theorem statements themselves. The former
+will be used as input to the tools, whilst the latter form a ground truth corpus
+against which to compare the output.
+
+It is important to ensure that there are no duplicate definitions: we are only
+concerned with the \emph{logical} content of the input, not the more arbitrary
+aspects like the names of functions. For example, consider a problem set which
+includes a commutativity theorem for a \texttt{plus} function, and an
+associativity theorem for an \texttt{add} function, where the definitions of
+\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
+tool to either conjecture commutativity and associativity for \emph{both}
+functions, or for \emph{neither} function, since they are logically equivalent.
+Yet a na\"ive precision/recall analysis would treat commutativity of
+\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
+they don't appear in the ground truth.
+
+For this reason, duplicates should be removed, and any references in the theorem
+statements updated to use the remaining definition (e.g. chosen based on
+lexicographic order). In the above example, the \texttt{plus} function would be
+removed, and the commutativity theorem updated to reference \texttt{add}
+instead.
+
+\subsection{Sampling}
+
+We could, in theory, send these de-duplicated definitions straight into an MTE
+tool and use the updated theorem statements as the ground truth for analysis.
+However, this would cause two problems:
+
+\begin{itemize}
+\item The result would be a single data point, which makes it difficult to
+  infer performance \emph{in general}.
+\item It is impractical to run existing MTE tools on ``large'' inputs,
+  containing more than a few dozen definitions.
+\end{itemize}
+
+To solve both of these problems we instead \emph{sample} a subset of
+definitions.\footnote{This could be done randomly, but for reproducibility we
+  use a deterministic order (based on cryptographic hashes, to discourage
+  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
+and provide only those as input to the tool. We generate a corresponding
+ground truth by selecting those theorems from the corpus which ``depend on''
+(contain references to) \emph{only} the definitions in that sample. Transitive
+dependencies aren't required (e.g. a theorem which only references
+\texttt{times} does not depend on \texttt{plus}, even if \texttt{plus} occurs in
+the definition of \texttt{times}).
+
+Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
+given sample size, increasing the size of the corpus (which provides better
+statistics) makes it less likely that a chosen sample will contain all of
+a theorem's dependencies. The majority of such samples would hence have an empty
+ground truth, and thus $0$ precision and undefined recall \emph{independent} of
+the tool's output! This is clearly undesirable as an evaluation method.
+
+Instead, we only allow a sample if it contains all dependencies of at least one
+theorem. We could do this using rejection sampling, but it is more efficient to
+pick a theorem, weighted in proportion to their number of dependencies and
+ignoring those with more dependencies than our sample size. That theorem's
+dependencies become our sample, padded up to the required size with uniform
+choices from the remaining definitions.
+
+The ground truth for such samples is guaranteed to contain at least one theorem
+(the one we picked), and hence the precision and recall will depend meaningfully
+on the tool's output.
+
+\subsection{Evaluation}
+
+Given a sample of definitions and a corresponding ground truth, the actual
+execution of the MTE tool proceeds as in prior work. We must translate the
+chosen definitions into the required input format, then we time the execution
+with a timeout (e.g. 5 minutes).
+
+In our experiments we have found that memory usage is also an important part of
+a tool's performance, but rather than complicating our analysis with an extra
+dimension, we instead allow programs to use as much memory as they like, and
+either get killed by the operating system or slow down so much from swapping
+that they time out. This is in line with the expected usage of these tools:
+either there is enough memory, or there isn't; implementations shouldn't be
+penalised for making use of available resources.
+
+To calculate precision and recall, the conjectures generated by each tool, as
+well as the ground-truth theorems, need to be parsed into a common format and
+compared syntactically after normalising away ``irrelevant'' details. We
+consider variable naming to be irrelevant, which can be normalised by numbering
+free variables from left to right and using de Bruijn indices for bound
+variables. We also consider the left/right order of equations to be irrelevant,
+which we normalise by choosing whichever order results in the
+lexicographically-smallest expression.
+
+Other than $\alpha$-equivalence and equation symmetry, we specifically
+\emph{ignore} logical relationships between syntactically distinct statements,
+such as one equation being implied by another. Whilst logically sound,
+second-guessing the ground truth in this way would harm other aspects which
+influence interestingness (for example, a more general statement is more widely
+applicable, but it might also be harder to comprehend).
+
+This procedure gives us a single runtime, precision and recall value for each
+sample. We propose two methods for analysing this data: \emph{summarising}
+the performance of a single MTE tool and \emph{comparing} the performance of two
+tools.
+
+\subsection{Summarising}
+
+Each sample is only explored once by each tool, so that we cover as many
+independent samples as possible to better estimate how a tool's performance
+generalises to unseen inputs. How we combine these data into an aggregate
+summary depends on what we are interested in measuring.
+
+One general question we might ask is how a tool's performance scales with
+respect to the input size (the number of given definitions). This is
+straightforward to measure by varying the sample size, but we need some way to
+combine the results from samples of the same size.
+
+We can summarise runtimes by choosing their median, as this is more robust than
+the mean against long-running outliers, and hence represents performance for a
+``typical'' input of that size. Unlike previous one-off evaluations, we can also
+compute the \emph{spread} of our data, for example with the inter-quartile
+range.
+
+The precision and recall for multiple samples can be combined in two ways. We
+can calculate their mean value, known as the \emph{average of the ratios}, to
+find the expected precision and recall for runs of this size. Alternatively, we
+can sum their numerators and denominators to get the \emph{ratio of the
+  averages}: this normalises the output and ground truth cardinalities, allowing
+precision and recall to be modelled as simple Bernoulli processes.
+
+We prefer the latter, where each generated conjecture (wanted or not) is like a
+coin toss whose bias is the precision; likewise, the ground truth theorems
+(found or not) are like coin tosses biased by the recall. Note that this simple
+model does not explain how many conjectures the tool decides to output for each
+run.
+
+\subsection{Comparison}
+
+To measure progress and encourage competition in the field, it is important to
+compare the relative performance of different tools on the same task. Since the
+aggregate statistics in the above summaries have obscured the details of
+specific runs, any comparison based on them (such as comparing average recall)
+would have very low statistical power. More direct comparisons can be made using
+\emph{paired} tests, since for each individual sample we have measurements for
+\emph{both} tools.
+
+We propose that running times are compared using a Wilcoxon signed-rank
+test~\cite{wilcoxon1945individual}, since it does not require that the data be
+normally distributed. In $\S$\ref{sec:application} we follow a variant of the
+Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
+differences are our use of the paired form of the Wilcoxon test, and the fact
+that each run was performed on a different sample (to reduce sampling bias, at
+the expense of accuracy). This paired difference test is also robust to the
+censoring caused by timeouts: upper-bounding the run time gives an upper-bound
+to the observable difference, which biases our results in a conservative way
+(\emph{towards} the null hypothesis of indistinguishable performance).
+
+For simplicity, we suggest comparing quality by pooling together results from
+all samples where both tools finished successfully. Recall can be compared using
+the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
+ground-truth theorem is categorised as either being found only by the first
+tool, only by the second, by both or by neither. Those found by both or neither
+are ignored, and the tool-specific numbers are tested for marginal homogeneity
+(i.e. whether their differences give rise to similar totals).
+
+We cannot use McNemar's test to compare precision, since each tool may generate
+different conjectures. Instead, we can use Boschloo's form of Barnard's
+test~\cite{lydersen2009recommended} to compare the proportions of wanted and
+unwanted conjectures generated by each tool, assuming a binomial distribution
+and conditioning on the number of generated conjectures (as if this were fixed
+by design).
+
+% FIXME: Add section references to the braces
+% FIXME: Maybe rethink "analysis" bit; maybe add summary and comparison
+% references (although maybe not as full blown steps)
+\begin{figure}
+  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
+                           minimum height=1cm,text centered, draw=black]
+
+  \tikzstyle{io} = [trapezium, trapezium left angle=70,
+                    trapezium right angle=110, minimum width=1cm,
+                    minimum height=1cm, text centered, draw=black]
+
+  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
+                         text centered, draw=black]
+  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
+                          text centered, draw=black]
+
+  \tikzstyle{arrow} = [thick,->,>=stealth]
+
+  % Avoids too much padding in io shapes
+  \tikzset{trapezium stretches=true}
+
+  \centering
+  \begin{tikzpicture}[node distance=1cm]
+
+    % Preparation section
+
+    \node (in)  [io]{Theorem Proving Benchmark};
+    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
+    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
+    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
+
+    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
+    \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};
+
+    \draw [arrow] (in)  -- (sep);
+    \draw [arrow] (def) -- (ref);
+    \draw [arrow] (def) -- (thy);
+    \draw [arrow] (ref) -- (thm);
+
+    % Arrows with labels
+    \path[->]
+        (sep) edge [arrow, sloped, above] node {definitions} (def)
+        (sep) edge [arrow, sloped, above] node {theorems}    (ref);
+
+    % Sampling section
+
+    \node (choose) [process, below=of thm   ]{Choose a theorem};
+    \node (deps)   [process, below=of choose]{List theorem dependencies};
+
+    % Create dummy coordinate below deps, then use its y coordinate for pad
+    \coordinate [below=of deps] (padDummy);
+    \path let \p{dummy} = (padDummy),
+              \p{in}    = (in)
+              in coordinate (padPos) at (\x{in}, \y{dummy});
+    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
+
+    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
+    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
+
+    % Calculate position of (size) using let, then define as normal
+    \path let \p{pad}    = (pad),
+              \p{choose} = (choose)
+           in coordinate (sizePos) at (\x{pad},\y{choose});
+    \node (size) [io, at=(sizePos)] {Sample size};
+
+    % Evaluation section
+    \path let \p{pad}  = (pad),
+              \p{sthm} = (sthm)
+           in coordinate (runPos) at (\x{pad}, \y{sthm});
+    \node (run) [process, below=of runPos]{Run MTE tool};
+    \node (pr)  [process, below=of run]{Analysis};
+
+    \node (prec) [startstop, below=of pr  ]{Precision};
+    \node (time) [startstop,  left=of prec]{Time taken};
+    \node (rec)  [startstop, right=of prec]{Recall};
+
+    \draw [arrow] (thy)    |- (pad);
+    \draw [arrow] (thm)    -- (choose);
+    \draw [arrow] (choose) -- (deps);
+    \draw [arrow] (deps)   |- (pad);
+    \draw [arrow] (size)   -- (choose);
+    \draw [arrow] (size)   -- (pad);
+    \draw [arrow] (pad)    -- (sthm);
+    \draw [arrow] (pad)    -- (sthy);
+    \draw [arrow] (sthy)   |- (run);
+    \draw [arrow] (run)    -- (pr);
+    \draw [arrow] (sthm)   |- (pr);
+    \draw [arrow] (pr)     -- (prec);
+    \draw [arrow] (pr)     -- (rec);
+    \draw [arrow] (pr)     -- (time);
+
+    % Awkward arrow
+    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
+
+    % Braces
+
+    % Preparation
+    \draw
+      let \p{thm} = (thm.east),
+          \p{in}  = (in.north),
+          \p{rec} = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
+         node [black, midway, right, xshift=0.3cm] {Preparation};
+
+    % Sampling
+    \draw
+      let \p{thm} = (thm.east),
+          \p{gt}  = (sthm.east),
+          \p{rec} = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
+         node [black, midway, right, xshift=0.3cm] {Sampling};
+
+    % Evaluation
+    \draw
+      let \p{sthm} = (sthm.east),
+          \p{rec}  = (rec.south east)
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
+         node [black, midway, right, xshift=0.3cm] {Evaluation};
+  \end{tikzpicture}
+  \caption[]{High-level view of our benchmarking methodology, showing
+    \begin{tikzpicture}
+      \node [rectangle, text centered, draw=black]{processes};
+    \end{tikzpicture},
+    \begin{tikzpicture}
+      \node [rectangle, rounded corners, text centered, draw=black]{data};
+    \end{tikzpicture} and
+    \begin{tikzpicture}
+      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
+             text centered, draw=black]{inputs};
+  \end{tikzpicture}}
+  \label{figure:flow_chart}
+\end{figure}
+
+\section{Application}
+\label{sec:application}
+% FIXME: Maybe use (red?) crosses rather than red dots? More useful for black and white
+We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
+tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
+benchmark as our ground truth corpus~\cite{claessen2015tip}.
+
+To determine our benchmarking parameters we ran some initial tests on both tools
+for a few samples sized between 1 and 100, for an hour each on our benchmarking
+machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
+of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
+and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
+on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
+above this took thousands of seconds each, which we deemed infeasibly long.
+
+Based on this we decided to benchmark sample sizes up to 20, since neither tool
+seemed to perform well beyond that. The Speedup-Test protocol follows the
+statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
+so we pick 31 samples of each size in order to cross this (arbitrary)
+threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
+a few days we chose a timeout of 300 seconds, since that covered most of the
+successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
+diminishing returns. During analysis, duplicate samples (caused by our sampling
+restrictions) were found and discarded, so only 14 samples of size 1 were used
+and 30 of size 2.
+
+\subsection{TIP} % FIXME: Expand acronym
+
+We chose TIP for our ground truth since it satisfies the desiderata of
+$\S$\ref{section:prep}: each problem has standalone type and function
+definitions, making their separation trivial; known examples from the software
+verification and inductive theorem proving literature are included, ensuring
+relevance to those fields; the format includes the higher-order functions and
+inductive datatypes we are interested in; it is large enough to pose a challenge
+to current MTE tools; plus it is accompanied by tooling to convert its custom
+format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
+languages, including Haskell and Isabelle.
+
+We use TIP version 0.2 which contains 343 problems, each stating a single
+theorem and together defining a total of 618 datatypes and 1498 functions. Most
+of these are duplicates, since each problem (re\nobreakdash-)defines all of the
+datatypes and functions it involves.
+
+TIP datatypes can have several ``constructors'' (introduction forms) and
+``destructors'' (elimination forms; field accessors). For example the type of
+lists from figure~\ref{figure:list_theory} can be defined in the TIP format as
+follows:
+
+\begin{samepage}
+\begin{verbatim}
+(declare-datatypes
+  (a)                       ;; Type parameter (element type)
+  ((List                    ;; Type name
+     (Nil)                  ;; Constructor (nullary)
+     (Cons                  ;; Constructor (binary)
+       (head a)             ;; Field name and type
+       (tail (List a))))))  ;; Field name and type
+\end{verbatim}
+\end{samepage}
+
+Our target languages (Haskell and Isabelle) differ in the way they handle
+constructors and destructors, which complicates comparisons. To avoid this, we
+generate a new function for each constructor (via $\eta$-expansion) and
+destructor (via pattern-matching) of the following form:
+
+\begin{samepage}
+\begin{verbatim}
+(define-fun
+  (par (a)                   ;; Type parameter
+    (constructor-Cons        ;; Function name
+      ((x a) (xs (List a)))  ;; Argument names and types
+      (List a)               ;; Return type
+      (as                    ;; Type annotation
+        (Cons x xs)          ;; Return value
+        (List a)))))         ;; Return type
+\end{verbatim}
+\end{samepage}
+
+\begin{samepage}
+\begin{verbatim}
+(define-fun
+  (par (a)                        ;; Type parameter
+    (destructor-head              ;; Function name
+      ((xs (List a)))             ;; Argument name and type
+      a                           ;; Return type
+      (match xs                   ;; Pattern-match
+        (case (Cons h t) h)))))   ;; Return relevant field
+\end{verbatim}
+\end{samepage}
+
+\begin{sloppypar}
+  We rewrite the TIP theorems (our ground truth) to reference these expanded
+  forms instead of the raw constructors and destructors, and use these functions
+  in our samples in lieu of the raw expressions. Note that these destructor
+  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
+  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
+  complicates their translation to proof assistants like Isabelle.
+\end{sloppypar}
+
+Another complication is TIP's ``native'' support for booleans and integers,
+which allows numerals and symbols like \texttt{+} to appear without any
+accompanying definition. To ensure consistency in the translations, we replace
+all occurrences of such expressions with standard definitions written with the
+``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
+mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
+  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
+  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
+  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
+  symmetry.}
+
+When we add all of these generated types and functions to those in TIP, we get a
+total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
+and we choose to only sample from those 182 functions which occur in at least
+one theorem statement (this removes ambiguity about which \emph{definitions}
+count as interesting and which are just ``implementation details'' for other
+definitions).
+
+TIP comes with software to translate its definitions into Haskell and Isabelle
+code, including comparison functions and random data generators suitable for
+QuickCheck. We translate all 269 unique definitions into a single module/theory
+which is imported on each run of the tools, although only those functions which
+appear in the current sample are included in the signature and explored.
+
+\subsection{QuickSpec}
+
+We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
+conjecturing equations involving Haskell functions, described in more detail in
+$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
+to automate some of the decisions which are normally left up to the user:
+
+\begin{sloppypar}
+  \begin{itemize}
+  \item We must decide what variables to include. We choose to add three
+    variables for each type that appears as a function argument, except for
+    types which have no QuickCheck data generators.
+  \item We must \emph{monomorphise} all types. For example, functions like
+    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
+    element type, but we need to pick a specific type in order to know which
+    random value generator to use. We resolve this (arbitrarily) by picking
+    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
+      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
+      \texttt{[]} (Haskell's list type constructor). If these violate some
+      type class constraint, we pick a suitable type non-deterministically from
+      those in scope during compilation; if no suitable type is found, we give
+      up and don't include that function.}
+  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
+    during its exploration process. They are also curried, always taking one
+    argument but potentially returning another function. QuickSpec lets us
+    assign an arity to each function in the signature, from 0 to 5, so we pick
+    the highest that is type-correct, since this avoids a proliferation of
+    incomparable, partially-applied functions.
+  \end{itemize}
+\end{sloppypar}
+
+\begin{figure}
+  \centering
+  \input{quickspectime.pgf}
+  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
+    indicates the total number of conjectures produced by each run, with failing
+    processes shown in red. Points are spread out horizontally to prevent
+    overlaps. Box plots show median and quartiles, with whiskers showing points
+    at 1.5$\times$~inter-quartile range.}
+  \label{figure:quickspec_runtimes}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \input{quickspecprec.pgf}
+  \caption{Precision and recall of successful QuickSpec runs (spread
+    horizontally to avoid overlaps). Point colour shows how many conjectures
+    were in the ground truth set for each run. Lines show the combined
+    proportion for each sample size (ratio of averages) and the shaded region is
+    the sample standard deviation.}
+  \label{figure:quickspec_precRec}
+\end{figure}
+
+The time taken to explore samples of different sizes is shown in
+figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
+and these occurred for all sample sizes but are more common for larger samples
+(over half the runs for sample size 20 timed out). Runs which succeeded mostly
+finished within 30 seconds, generating few or no conjectures; those taking
+longer generated more conjectures, with the most being 133 conjectures from a
+sample of size 19.
+
+Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
+Since QuickSpec generates monotonically more conjectures as definitions are
+added to a signature (assuming sufficient testing), its decreasing precision
+can't be due to finding fewer wanted conjectures at larger sizes. This is
+supported by the relative flatness of the recall results. Rather, the number of
+conjectures generated is increasing at a higher rate than the size of the ground
+truth. These extra conjectures may involve those ``padding'' definitions in a
+sample which don't contribute to its ground truth, or may be ``uninteresting''
+relationships between the dependencies of different ground truth theorems.
+
+This indicates two potential improvements to the QuickSpec algorithm (as far as
+this benchmark is concerned). The deluge of generated conjectures could be
+filtered down to a more desirable sub-set by another post-processing step.
+Alternatively, rather than exploring all of the given definitions together,
+multiple smaller signatures could be selected from the input by predicting which
+combinations are likely to lead to interesting conjectures; this would avoid
+both the ``padding'' and the cross-dependency relationships. Both of these
+methods could improve the precision, although care would be needed to avoid a
+large reduction in recall. The latter option could also improve the running
+time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
+signatures may be faster to explore than a single large one. Such improvements
+would also need to be ``generic'', to avoid over-fitting to this particular
+benchmark.
+
+QuickSpec's recall is limited by two factors: the algorithm is unable to
+synthesise some theorem statements, such as conditional equations, inequalities,
+terms larger than the search depth and those containing anonymous functions.
+The congruence closure algorithm used as a post-processor may also be removing
+``interesting'' results, for example if we found an ``uninteresting'' result
+which is more general.
+
+\subsection{IsaCoSy}
+
+IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
+ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
+with the 2015 version of Isabelle. The following issues had to be overcome to
+make our benchmark applicable to IsaCoSy:
+
+\begin{itemize}
+\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
+  encoded in Isabelle. We strip out this type and the functions which depend on
+  it before translating. It still appears in samples and contributes to the
+  ground truth, which penalises IsaCoSy for being unable to explore such
+  definitions.
+\item When using a type in an IsaCoSy signature, that type's constructors will
+  automatically be included in the exploration. Since those constructors will
+  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
+  even those may not be present in the current sample) this will unfairly reduce
+  the calculated precision. To avoid this, we add a post-processing step which
+  replaces all occurrences of a constructor with the corresponding wrapper, then
+  discards any conjectures which involve functions other than those in the
+  current sample. This presumably results in more work for IsaCoSy, exploring
+  constructors unnecessarily, but it at least does not bring down the quality
+  measures.
+\item Since Isabelle is designed for theorem proving rather than programming, it
+  requires every definition to be accompanied by proofs of exhaustiveness and
+  termination. These are difficult to generate automatically, and don't exist in
+  the case of partial functions like destructor wrappers. Hence we use the
+  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
+  the \texttt{sorry} keyword.
+\item Partial functions cause problems during exploration, since they can throw
+  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
+  pre-populating IsaCoSy's constraint set with these undefined expressions
+  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
+  IsaCoSy from ever generating them.
+\end{itemize}
+
+\begin{figure}
+  \centering
+  \input{isacosytime.pgf}
+  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
+    indicates the total number of conjectures produced by each run, with failing
+    processes (timeouts and out of memory) shown in red. Points are spread out
+    horizontally to prevent overlaps. Box plots show median and quartiles, with
+    whiskers showing points at 1.5$\times$~inter-quartile range.}
+  \label{figure:isacosy_runtimes}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \input{isacosyprec.pgf}
+  \caption{Precision and recall of all successful IsaCoSy runs (spread
+    horizontally to avoid overlaps). Lines show the combined proportion for each
+    sample size (ratio of averages) and the shaded region is the sample standard
+    deviation.}
+  \label{figure:isacosy_precRec}
+\end{figure}
+
+The most striking result is how rapidly IsaCoSy's running time, shown in
+figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
+size 6 failed, with most timing out and a few aborting early due to running out
+of memory. Like in our preliminary testing, this increase appears exponential,
+so even large increases to the timeout would not produce many more successful
+observations. A significant amount of IsaCoSy's time (around 50 seconds) was
+spent loading the generated theory (containing all distinct datatypes and
+functions from TIP); this overhead is unfortunate, but since it is constant
+across all sample sizes it does not affect our conclusions.
+
+With so few successful datapoints it is difficult to make strong claims about
+the precision/recall quality of IsaCoSy's output, shown in
+figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
+proportions that IsaCoSy is capable of discovering interesting conjectures, and
+the graphs appear to follow the same shapes as those of QuickSpec (decreasing
+precision, flat recall), although the error margins are too wide to be
+definitive.
+
+\subsection{Comparison}
+
+We compare the running times of QuickSpec and IsaCoSy using our paired variant
+of the Speedup-Test protocol, where each of our sample sizes is a separate
+``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
+which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
+signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
+method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
+out.
+
+For each sample size (``benchmark'') the Speedup-Test protocol compares the
+distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
+significantly different (with $\alpha < 0.05$), then the signed-rank test is
+only performed if more than 30 samples are available. This was the case for all
+sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
+two speedups were not deemed statistically significant. For all other sample
+sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
+putting the proportion of sizes with faster median times between 66.9\% and
+98.2\% with 95\% confidence. The magnitude of each speedup is given in
+table~\ref{table:speedups}. We would predict the speedup to grow for larger
+sample sizes, but the increasing proportion of timeouts causes our estimates to
+become more conservative: by size 20 most runs are timing out, and hence are
+tied.
+
+\begin{table}
+  \centering
+  \begin{tabular}{ |r|l| }
+    \hline
+    \bfseries Sample Size & \bfseries Speedup
+    \csvreader[]{speedups.csv}{}
+    {\\\hline\csvcoli&\csvcolii} \\
+    \hline
+  \end{tabular}
+  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
+    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
+    considered statistically significant due to having less than 30 samples.
+    These are conservative estimates, since the 300 second time limit acts to
+    reduce the measured time difference.}
+  \label{table:speedups}
+\end{table}
+
+We compare the recall proportions by applying McNemar's test to only those
+samples where both tools succeeded. We pooled these together from all sample
+sizes to produce the following table, and found that QuickSpec is significantly
+different from IsaCoSy with a p value of 0.0026.
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{ |r|l|l| }
+    \hline
+    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
+    \hline
+    \bfseries Found by QuickSpec     & 28 & 19 \\
+    \hline
+    \bfseries Not found by QuickSpec & 4  & 75 \\
+    \hline
+  \end{tabular}
+\end{table}
+
+McNemar's test is not applicable for comparing precision, since each tool
+generated different sets of conjectures. Instead, we add up the number of
+results which are ``interesting'' (appear in the ground truth) and those which
+are ``uninteresting'' (do not appear in the ground truth):
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{ |r|l|l| }
+    \hline
+              & \bfseries Interesting & \bfseries Uninteresting \\
+    \hline
+    \bfseries IsaCoSy   & 32          & 137      \\
+    \hline
+    \bfseries QuickSpec & 47          & 301      \\
+    \hline
+  \end{tabular}
+\end{table}
+
+We test these totals for independence using Boschloo's test and find a p-value
+of 0.111, which exceeds our (arbitrary) choice of $\alpha = 0.05$; hence we do
+not consider the difference between these proportions (14\% for QuickSpec, 19\%
+for IsaCoSy) to be significant.
+
+Note that precision is the only comparison where IsaCoSy scored higher, and even
+that was not found to be significant. Also, precision on its own is not the
+whole story: it can be increased at the expense of recall by making a tool more
+conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
+given that its recall is significantly lower.
+
+More importantly, the poor time and memory usage of IsaCoSy meant that very few
+samples finished successfully. If we include failed runs in our tables, treating
+them as if they succeeded with no output, the resulting statistics lean
+overwhelmingly in favour of QuickSpec simply because it generated so much more
+than IsaCoSy in the available time.
+
+\section{Discussion}
+\label{sec:discussion}
+
+% FIXME: The designers must decide on, not the system
+Fundamentally, any mathematical reasoning system must decide on, and formalise,
+what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
+``provable'' include trivial tautologies, while at the same time failing to
+capture the ``almost true'', which can be a valuable trigger for theory change,
+as demonstrated by Lakatos in his case studies of mathematical
+development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
+proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
+light on this by testing whether mathematicians' experiences of abstract beauty
+correlates with the same brain activity as experiences of sensory
+beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
+table~\ref{table:colton}, from computer scientists like Colton, are based
+largely on ``intuition'', plausibility arguments about why a metric would be
+important, and previous use of such metrics (in the case of
+``surprisingingness'', from a single quote from a mathematician).
+Opinions from mathematicians themselves include Gowers' suggestion that we can
+identify features which are commonly associated with good
+proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
+book containing the most elegant proof of every theorem, brought to life
+in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
+interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
+developer).
+
+All of these approaches rest upon the assumption that it makes sense to speak of
+``the good'' in mathematics. However, empirical psychological studies
+call into question such assumptions: for example, work by Inglis and colleagues
+has shown that there is not even a single standard of \emph{validity} among
+contemporary mathematicians~\cite{inglis2013mathematicians}.
+
+Whilst these difficulties are real and important, we cannot ignore the fact that
+mathematics is nevertheless being practised around the world; and similarly that
+researchers have forged ahead to develop a variety of tools for automated
+exploratory mathematics. If we wish to see these tools head in a useful,
+fruitful direction then \emph{some} method is needed to compare their
+approximate ``quality'' in a concrete, measurable way.
+
+The key to our benchmarking methodology is to side-step much of this
+philosophical quagmire using the simplifying assumption that theorem proving
+problem sets are a good proxy for desirable input/output behaviour of MTE tools.
+As an extension of existing precision/recall analysis, this should hopefully not
+prove too controversial, although we acknowledge that there are compelling
+reasons to refute it.
+
+We do not claim that our use of corpora as a ground truth exactly captures all
+interesting conjectures of their definitions, or that those definitions exactly
+represent all theories we may wish to explore. Rather, we consider our approach
+to offer a pareto-optimal balance between theoretical rigour and experimental
+practicality, at least in the short term. Futhermore, since research is already
+on-going in these areas, we hope to at least improve on existing evaluation
+practices and offer a common ground for future endeavours.
+
+One notable weakness is that our methodology only makes sense in domains where
+existing human knowledge surpasses that discovered by the machine. Any
+\emph{truly novel} insights discovered during testing will not, by definition,
+appear in any existing corpus, and we would in fact \emph{penalise} tools for
+such output. We do not believe this to be a realistic problem for the time
+being, although we would be delighted if proved wrong. % FIXME: Is there a solution? Mention that this is common to most tests
+
+Another practical limitation of our benchmarking approach is that it only
+applies to tools which act in ``batch mode'', i.e. those which choose to halt
+after emitting some output. Whilst all of the systems we have encountered are of
+this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
+be run without a depth limit, and form part of the ``scientist assistant'' role
+envisaged by Lenat (or McCarthy's earlier ``advice
+taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be designed for
+such long-running, potentially interactive programs, but that is beyond the
+scope of this project.
+
+\section{Related Work}
+\label{sec:related-work}
+
+% FIXME: Reword into a short intro, e.g. "There is a long history of automating
+% mathematics, going back to... A recurring theme is separation between the
+% efforts of pure mathematicians and those wanting practical engineering
+% applications....
 The first automated mathematical tool, the mechanical calculator
@@
 
-We limit our discussion to applications in mathematics, although there are
-similar approaches being applied to the task of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}.
-Scientific knowledge relies on inductive reasoning and experimental testing,
-which (in principle) are not so important to a mathematical theory. However,
-these techniques \emph{do} play an important role in many mathematical theory
-formation and exploration tools: in particular, inductive reasoning allows a
-tool to conjecture statements which may be too difficult for it to deductively
-prove; whilst experimental testing, e.g. by checking universally-quantified
-statements against some specific values, is a simple way to quickly discard many
-obvious falsehoods generated by powerful but imperfect algorithms (as opposed
-to, say, limiting ourselves to verified theorem generators).
-
 \subsection{Automated Theory Formation}
 \label{section:atf}
+
+% FIXME: Cut this down a lot. Currently it's not even an ATF section, it's just
+% an AM section. We should maybe remove its status as a section, and instead
+% just remark on the influence of AM on the field
 
@@
 
-The phrase Mathematical Theory Exploration (MTE) is a recent term
-which seems to have originated with Buchberger and colleagues (see,
+The phrase Mathematical Theory Exploration (MTE) is a recent term % FIXME: When?
+which seems to have originated with Buchberger and colleagues (see, % FIXME: informal
 for example,~\cite{buchberger}). His motivation is to support
@@
 
-%IsaScheme
+% IsaScheme FIXME: Maybe strip down this description? We could use prose instead,
+% like "schemes capture common patterns such as iteration, which can be
+% instantiated with an addition function to produce multiplication, or with
+% multiplication to produce exponentiation.
 Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
@@
 
+% FIXME: CITE
 QuickSpec version 2 replaces this two-step process with a single, iterative
@@
 
-We believe that ATF and MTE share many goals and would benefit from a
+% FIXME: Condense
+ATF and MTE share many goals and would benefit from a
 closer alignment: both fields are highly specialised and by joining
@@
 
-\subsection{Existing Evaluations}
-\label{sec:existing}
-
-There are three aspects to consider when evaluating a tool for conjecture
-generation:
-
-\begin{enumerate}
-\item The quality of each individual conjecture, which we will refer to as their
-  ``interestingness''.
-\item The quality of the \emph{set} of generated conjectures, which is important
-  for criteria like novelty, as well as to assess consistency.
-\item Resources used by the tool, to find the balance struck between output
-  quality and time taken.
-\end{enumerate}
-
-We make the assumption that each tool is intended to be \emph{general}, and
-hence isn't hard-coded or fine-tuned to only perform well in a narrow domain.
-Examples where this does \emph{not} hold include the natural number heuristics
-of AM discussed in $\S$\ref{section:atf}, as well as machine learning models
-trained on some specific dataset (such as the benchmark we describe).
-
-Evaluation of new tools and techniques for conjecture generation tends to be
-\emph{ad hoc} and hence difficult to compare, whilst comparative surveys (such
-as Colton's) are descriptive and qualitative rather than providing direct
-numerical measurements.
-
-One quantitative approach which has gained traction is \emph{precision/recall
-  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
-providing a direct comparison of these three
-tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
-  reports prior results from the other tools, rather than reproducing them.
-  This makes the numbers less comparable, since each tool was tested with
-  slightly different definitions (e.g. whether or not the exponential and
-  maximum functions were included).} To measure precision and recall we first
-choose a set of definitions (such as those in figure \ref{figure:list_theory})
-and a \emph{ground truth} to treat as the ``ideal'' set of conjectures we would
-like a tool to produce for those definitions (e.g. this might include
-equation~\ref{eq:mapreduce}). Each tool is run and their outputs are compared to
-the ground truth. To score 100\% on precision and recall, all of the conjectures
-which appear in the ground truth must be generated, and nothing else:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output large
-  numbers of conjectures in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few conjectures as a way to avoid ``bad'' ones.
-\end{itemize}
-
-Precision and recall act as quality tests for sets of conjectures, fulfilling
-one of our requirements. They also provide a coarse quality assessment for
-individual conjectures: either a conjecture appears in the ground truth set or
-it doesn't.
-
-The analyses in~\cite{claessen2013automating} took their definitions and ground
-truths from the standard library of the Isabelle proof assistant: since the
-library authors have gone to the effort of stating, proving and including these
-theorems in every copy of their software, this is a good indication that they
-are interesting.
-
-Tests were performed with a theory of natural numbers and a theory of lists (the
-definitions shown in figure~\ref{figure:list_theory}), but it is difficult to
-draw general conclusions from only two datapoints. These theories were also
-quite small, the natural number theory
-from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
-definitions and 12 theorems.
-
-Despite these limitations we believe that this is a viable approach to drawing
-meaningful, objective conclusions about such tools; but only with larger data
-sets can we judge performance in different, especially \emph{novel}, domains
-(which, after all, is the purpose of such tools). In addition, we might be
-concerned that such ``popular'' theories may be unrepresentative of typical
-usage. For example, we would not expect the library functions in a typical
-verification problem to be as mathematically rich as number theory.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in figure
-\ref{figure:flow_chart}, which both generates a large definition/ground-truth
-corpus, and provides a scalable, statistical approach to evaluating MTE tools
-using this corpus. We follow a precision/recall approach similar to prior work,
-with the main difference being the source of definitions and ground truths: we
-take existing problem sets designed for automated theorem proving, and adapt
-their content for use in the conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those which have the following properties:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the theorem
-  to be proved and the definitions involved, such that the two can be easily and
-  meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item Since it will act as our ground truth, the problem set should
-  \emph{ideally} contain every ``interesting'' conjecture involving its included
-  definitions. More \emph{realistically}, we should aim for each definition to
-  appear in \emph{many} theorem statements; rather than each problem having
-  unique definitions.
-\item The problem set should be as large as possible, for robustness of the
-  resulting statistics.
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the definitions
-referenced by the theorems from the theorem statements themselves. The former
-will be used as input to the tools, whilst the latter form a ground truth corpus
-against which to compare the output.
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a commutativity theorem for a \texttt{plus} function, and an
-associativity theorem for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they're logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references in the theorem
-statements updated to use the remaining definition (e.g. chosen based on
-lexicographic order). In the above example, the \texttt{plus} function would be
-removed, and the commutativity theorem updated to reference \texttt{add}
-instead.
-
-\subsection{Sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the updated theorem statements as the ground truth for analysis.
-However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those theorems from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a theorem which only references
-\texttt{times} does not depend on \texttt{plus}, even if \texttt{plus} occurs in
-the definition of \texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a theorem's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
-
-Instead, we only allow a sample if it contains all dependencies of at least one
-theorem. We could do this using rejection sampling, but it is more efficient to
-pick a theorem, weighted in proportion to their number of dependencies and
-ignoring those with more dependencies than our sample size. That theorem's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one theorem
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
-
-\subsection{Evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth theorems, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-Other than $\alpha$-equivalence and equation symmetry, we specifically
-\emph{ignore} logical relationships between syntactically distinct statements,
-such as one equation being implied by another. Whilst logically sound,
-second-guessing the ground truth in this way would harm other aspects which
-influence interestingness (for example, a more general statement is more widely
-applicable, but it might also be harder to comprehend).
-
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring.
-
-One general question we might ask is how a tool's performance scales with
-respect to the input size (the number of given definitions). This is
-straightforward to measure by varying the sample size, but we need some way to
-combine the results from samples of the same size.
-
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth theorems
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth theorem is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
-
-\begin{figure}
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \centering
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {theorems}    (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a theorem};
-    \node (deps)   [process, below=of choose]{List theorem dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm] {Preparation};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm] {Sampling};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm] {Evaluation};
-  \end{tikzpicture}
-  \caption[]{High-level view of our benchmarking methodology, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this (arbitrary)
-threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
-a few days we chose a timeout of 300 seconds, since that covered most of the
-successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
-diminishing returns. During analysis, duplicate samples (caused by our sampling
-restrictions) were found and discarded, so only 14 samples of size 1 were used
-and 30 of size 2.
-
-\subsection{TIP}
-
-We chose TIP for our ground truth since it satisfies the desiderata of
-$\S$\ref{section:prep}: each problem has standalone type and function
-definitions, making their separation trivial; known examples from the software
-verification and inductive theorem proving literature are included, ensuring
-relevance to those fields; the format includes the higher-order functions and
-inductive datatypes we are interested in; it is large enough to pose a challenge
-to current MTE tools; plus it is accompanied by tooling to convert its custom
-format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
-languages, including Haskell and Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-theorem and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP theorems (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which occur in at least
-one theorem statement (this removes ambiguity about which \emph{definitions}
-count as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
-to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
-    indicates the total number of conjectures produced by each run, with failing
-    processes shown in red. Points are spread out horizontally to prevent
-    overlaps. Box plots show median and quartiles, with whiskers showing points
-    at 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to avoid overlaps). Point colour shows how many conjectures
-    were in the ground truth set for each run. Lines show the combined
-    proportion for each sample size (ratio of averages) and the shaded region is
-    the sample standard deviation.}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
-and these occurred for all sample sizes but are more common for larger samples
-(over half the runs for sample size 20 timed out). Runs which succeeded mostly
-finished within 30 seconds, generating few or no conjectures; those taking
-longer generated more conjectures, with the most being 133 conjectures from a
-sample of size 19.
-
-Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth theorems.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some theorem statements, such as conditional equations, inequalities,
-terms larger than the search depth and those containing anonymous functions.
-The congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
-ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
-    indicates the total number of conjectures produced by each run, with failing
-    processes (timeouts and out of memory) shown in red. Points are spread out
-    horizontally to prevent overlaps. Box plots show median and quartiles, with
-    whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to avoid overlaps). Lines show the combined proportion for each
-    sample size (ratio of averages) and the shaded region is the sample standard
-    deviation.}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce the following table, and found that QuickSpec is significantly
-different from IsaCoSy with a p value of 0.0026.
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-\end{table}
-
-We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (arbitrary) choice of $\alpha = 0.05$; hence we do
-not consider the difference between these proportions (14\% for QuickSpec, 19\%
-for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, any mathematical reasoning system must decide on, and formalise,
-what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
-``provable'' include trivial tautologies, while at the same time failing to
-capture the ``almost true'', which can be a valuable trigger for theory change,
-as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
-table~\ref{table:colton}, from computer scientists like Colton, are based
-largely on ``intuition'', plausibility arguments about why a metric would be
-important, and previous use of such metrics (in the case of
-``surprisingingness'', from a single quote from a mathematician).
-Opinions from mathematicians themselves include Gowers' suggestion that we can
-identify features which are commonly associated with good
-proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
-book containing the most elegant proof of every theorem, brought to life
-in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it even makes sense to
-speak of ``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although we would be delighted if proved wrong.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-envisaged by Lenat (or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
 \section{Conclusion}
@@
   their software to our purposes.
-%  EPSRC
+%  FIXME: EPSRC, Katya, Jianguo fundung, Alison funding
 \end{acknowledgements}

### 2018-04-12 TEBenchmarkPaper/article.tex
Improve justification

@@
 
-This has applications wherever we find use for formal statements: in proof
+This has applications wherever we find uses for formal methods: in proof
 assistants and their libraries, in mathematics education and research, and in
-the specification, verification, optimisation and testing of software. Similar
-tasks are also studied in Computational \emph{Scientific}
+the specification and verification of software. Although formal methods, and
+automated reasoning more generally, have been advocated in mathematics (notably
+by Voevodsky~\cite{voevodsky2010univalent} and
+Gowers~\cite{ganesalingam2013fully}) and software engineering (for example via
+dependently-typed
+programming~\cite{McKinna:2006, chlipala2011certified, Xi2003}), they have a
+reputation for being difficult and expensive, which has historically limited
+their use to high-assurance domains such as aerospace and microprocessor design.
+Increased automation is one way to address these problems, and conjecture
+generation has already been shown to aid in specification and verification of
+software properties~\cite{Claessen.Johansson.Rosen.ea:2013}).
+
+Similar tasks are also studied in the domain of Computational \emph{Scientific}
 Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling},

### 2018-04-13 TEBenchmarkPaper/article.tex
TEB: Better motivation

@@
 %  - "Mixed initiative" (social machine?); maybe Kasparov
-%  - Gowers calling for better AI in maths
 
@@
 their use to high-assurance domains such as aerospace and microprocessor design.
-Increased automation is one way to address these problems, and conjecture
-generation has already been shown to aid in specification and verification of
-software properties~\cite{Claessen.Johansson.Rosen.ea:2013}).
+One way to reduce these hurdles is to increase automation, which in the case of
+software verification can benefit from the introduction of conjecture generation
+tools~\cite{Claessen.Johansson.Rosen.ea:2013}. Another approach is to tailor the
+division of labour between humans and machines such that each is playing to
+their strengths, which has been referred to as ``centaur
+teams''~\cite{harari2017reboot, davenport2015beyond}. In this respect a
+conjecture generating tool can search a much larger space of relationships than
+a human, whilst leaving the ultimate judgement of which to investigate in more
+detail to the user.
 
 Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling},
-although that relies on inductive reasoning and experimental testing which (in
-principle) are not so important for pure mathematics. Whilst we \emph{have}
-found these methods to also be important for MTE tools (in particular for the
-\emph{plausibility} of generated conjectures), we leave a more thorough
-application of our results to scientific reasoning for future work.
-
-Even limiting ourselves to mathematics applications, it is difficult to measure
-and compare the success of approaches and tools. This is partially due to their
-diversity, but also because of the inherent ambiguity of the task, the different
-goals emphasised by their designers and the variety of evaluation methods
-employed. We attempt to solve this discrepancy, at least for the foreseeable
-future, by defining a standard, unambiguous benchmarking approach with which to
-compare the conjecture generation of MTE tools. Our contributions include:
+Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
+example searching for new drugs. The major difference between scientific and
+mathematical discovery is that inductive reasoning and experimental testing are,
+in principle, more important for the former, although we note that they are also
+core principles in the mathematical tools we have investigated (in particular
+for ensuring the \emph{plausibility} of conjectures).
+
+We limit ourselves to applications in mathematics, but even here it is difficult
+to measure and compare the success of approaches and tools. This is partially
+due to their diversity, but also because of the inherent ambiguity of the task
+(what counts as ``interesting''?), the different goals emphasised by their
+designers and the variety of evaluation methods employed. We attempt to solve
+this discrepancy, at least for the foreseeable future, by defining a standard,
+unambiguous benchmarking approach with which to compare the conjecture
+generation of MTE tools. Our contributions include:
 

### 2018-04-16 PhDSymposium2018/abstract.tex
Added PhDSymposium2018

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

\title{Exploring Software with Symbolic and Statistical Algorithms}
  \author{
          Chris Warburton \\
      University of Dundee \\
      \texttt{cmwarburton@dundee.ac.uk}
    }
\date{}

\begin{document}
\maketitle
\begin{abstract}
  We propose a benchmarking methodology to evaluate the efficiency and quality
  of \emph{conjecture generation} by automated tools for \emph{mathematical
    theory exploration}. Our approach uses widely available theorem proving
  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
  QuickSpec and IsaCoSy tools, finding that the former takes significantly less
  time to produce significantly more ``interesting'' output. By providing a
  standard, cross-tool evaluation technique we allow the disparate approaches
  currently being pursued to be more directly compared. Our hope is to encourage
  innovation and improvements similar to those seen in fields like automated
  theorem proving, where the availability of benchmarks encourages healthy
  competition.
\end{abstract}

\section{Introduction}\label{introduction}

\emph{Conjecture generation} (CG) is the open-ended task of producing
conjectures about a given logical theory (such as a software library) which are
somehow ``interesting'', and is studied as a sub-field of \emph{mathematical
  theory exploration}. CG tools can discover formal specifications for software,
and complement theorem provers by spotting crucial lemmas needed in verification
tasks~\cite{Claessen.Johansson.Rosen.ea:2013}. Increasing the automation of
formal methods in this way can reduce their cost and difficulty, which have
historically limited their use to mission- or life-critical domains like
aerospace and microprocessor design. Promising applications which would benefit
from more practical automated reasoning include correctness proofs for software
engineering~\cite{McKinna:2006,chlipala2011certified,Xi2003}, automated
``assistants'' for STEM
research~\cite{McCarthy_Programs59,lenat:77,benzmuller1997omegamega} and more
powerful education software~\cite{conf/ijcai/TrybulecB85,hendriks2010teaching}.

Progress in this field is hard to ascertain, since there is no standard measure
of ``success'' which we can compare across different approaches. This is
partially due to the inherent ambiguity of the task, i.e. what counts as
``interesting''? Different researchers may have different goals, and a variety
of evaluation methods are employed. We attempt to solve this discrepancy, at
least for the foreseeable future, by defining a standard, unambiguous
benchmarking approach with which to compare conjecture generation tools. Our
contributions include:

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
  and a comparison and discussion of the results.
\end{itemize}

\bibliographystyle{plain}
\bibliography{../Bibtex}

\end{document}

### 2018-04-19 PhDSymposium2018/abstract.tex
Finish symposium text

@@
 \usepackage{paralist}
+\usepackage{tikz}
 \usepackage{ifxetex,ifluatex}
@@
             pdfauthor={},
-            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
+            pdftitle={Quantitative Benchmarking for Conjecture Generation},
             colorlinks=true,
@@
 \setcounter{secnumdepth}{5}
+\usepackage{subcaption}
+\newcommand{\etal}{{\em et al.}}
 
-\title{Exploring Software with Symbolic and Statistical Algorithms}
+\title{Quantitative Benchmarking for Conjecture Generation}
   \author{
@@
   We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools, finding that the former takes significantly less
-  time to produce significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
+  of \emph{automated conjecture generation}. Our approach uses widely available
+  theorem proving tasks as a \emph{ground-truth} corpus, and we demonstrate its
+  use on the \textsc{QuickSpec} and \textsc{IsaCoSy} tools, finding that the
+  former takes significantly less time to produce significantly more
+  ``interesting'' output. By providing a standard, cross-tool evaluation
+  technique we allow the disparate approaches currently being pursued to be more
+  directly compared. Our hope is to encourage innovation and improvements
+  similar to those seen in fields like automated theorem proving, where the
+  availability of benchmarks encourages healthy competition.
 \end{abstract}
@@
 
-\emph{Conjecture generation} (CG) is the open-ended task of producing
-conjectures about a given logical theory (such as a software library) which are
-somehow ``interesting'', and is studied as a sub-field of \emph{mathematical
-  theory exploration}. CG tools can discover formal specifications for software,
-and complement theorem provers by spotting crucial lemmas needed in verification
-tasks~\cite{Claessen.Johansson.Rosen.ea:2013}. Increasing the automation of
-formal methods in this way can reduce their cost and difficulty, which have
-historically limited their use to mission- or life-critical domains like
-aerospace and microprocessor design. Promising applications which would benefit
-from more practical automated reasoning include correctness proofs for software
-engineering~\cite{McKinna:2006,chlipala2011certified,Xi2003}, automated
-``assistants'' for STEM
-research~\cite{McCarthy_Programs59,lenat:77,benzmuller1997omegamega} and more
-powerful education software~\cite{conf/ijcai/TrybulecB85,hendriks2010teaching}.
+\emph{Conjecture generation} (CG) is the open-ended task of conjecturing
+``interesting'' properties of a given logical theory (such as a software
+library), and is studied as a sub-field of \emph{mathematical theory
+  exploration}. Such conjectures can be sent to automated theorem provers (ATP),
+have helped prove software correctness by finding key
+lemmas~\cite{Claessen.Johansson.Rosen.ea:2013}. Automating formal methods this
+way can reduce the cost and difficulty which have historically limited their use
+to mission- or life-critical domains. More practical automated reasoning shows
+promise in software
+engineering~\cite{McKinna:2006,chlipala2011certified,Xi2003}, automated research
+``assistants''~\cite{McCarthy_Programs59,lenat:77,benzmuller1997omegamega} and
+educational software~\cite{conf/ijcai/TrybulecB85,hendriks2010teaching}.
 
-Progress in this field is hard to ascertain, since there is no standard measure
-of ``success'' which we can compare across different approaches. This is
-partially due to the inherent ambiguity of the task, i.e. what counts as
-``interesting''? Different researchers may have different goals, and a variety
-of evaluation methods are employed. We attempt to solve this discrepancy, at
-least for the foreseeable future, by defining a standard, unambiguous
-benchmarking approach with which to compare conjecture generation tools. Our
-contributions include:
+CG progress is hard to ascertain since there is no standard measure of
+``success'', partially due to the ambiguity of the task, i.e. what counts as
+``interesting''? Researchers may have different goals, and a variety of
+evaluation methods appear in the literature. We attempt to solve this
+discrepancy (in the short-term) by defining a standard, unambiguous, extensible
+benchmarking approach for CG tools. Our contributions include:
 
 \begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
+\item A benchmarking methodology for CG tools.
+\item Resolving the ``interestingness'' question by adapting ATP benchmarks into
+  ground-truth input/output examples.
+\item Benchmarking and comparing the \textsc{QuickSpec} and \textsc{IsaCoSy}
+  tools.
 \end{itemize}
+
+\begin{figure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \input{quickspectime.pgf}
+    \caption{\textsc{QuickSpec}}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \input{isacosytime.pgf}
+    \caption{\textsc{IsaCoSy}}
+  \end{subfigure}
+
+  \caption{Time taken for CG on samples from TIP, with 300 second timeout. Lines
+    show medians and error bars show inter-quartile range, with 31 samples of
+    each size except 1 (14) and 2 (30). }
+  \label{figure:quickspec_runtimes}
+\end{figure}
+
+\begin{figure*}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \input{quickspecprec.pgf}
+    \caption{\textsc{QuickSpec}}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \input{isacosyprec.pgf}
+    \caption{\textsc{IsaCoSy}}
+  \end{subfigure}
+
+  \caption{Precision and recall of successful runs. Lines show average
+    proportion, with sample standard deviation shaded.}
+  \label{figure:quickspec_precRec}
+\end{figure*}
+
+\section{Background}
+
+Consider a software library for inductive (i.e. singly-linked) lists, containing
+simple functions like \texttt{append}, more complex ones like \texttt{mergesort}
+and even higher-order functions like \texttt{map} and \texttt{reduce}. The large
+input space of these functions limits \emph{testing} to a tiny fraction of the
+possible executions, whilst higher-order functions and induction makes
+\emph{proving} correctness properties undecidable (requiring significant user
+guidance). We may also \emph{miss} key properties (especially if the same person
+writes the spec and the code), e.g. that calling \texttt{mergesort} on something
+already sorted can be optimised away. Similar tasks arise in mathematics, and
+scientific models more generally, when exploring the consequences of axioms and
+definitions.
+
+The aim of CG is to read such definitions and produce a set of ``interesting''
+properties which they appear to satisfy. This can help formalise the
+``background knowledge'' required for proving, and present the user with ideas
+they hadn't considered.
+
+Crucial to CG is how we define ``interesting''. The literature gives many
+alternatives, for example Colton \etal{} survey existing
+tools~\cite{colton2000notion} and find six common qualities: empirical
+plausibility, novelty, surprisingness, applicability, comprehensibility and
+utility.
+
+\emph{Quantitative} evaluation has compared the
+\textsc{IsaCoSy}~\cite{Johansson.Dixon.Bundy:conjecture-generation},
+\textsc{IsaScheme}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
+\textsc{HipSpec}~\cite{claessen2013automating} tools, by selecting sets of
+definitions and properties from the standard library of the \textsc{Isabelle}
+proof assistant, and treating those as a \emph{ground-truth} corpus, i.e. using
+``membership in the corpus'' as a proxy for ``interesting''. The proportion of
+conjectures appearing in the corpus is the \emph{precision}, the proportion of
+the corpus appearing in the conjectures is the \emph{recall}.
+
+This is a pragmatic solution to the ambiguity, but only two datapoints were
+taken: one with 12 properties, the other with 9. This makes it hard to draw
+robust conclusions.
+
+\section{Methodology}
+
+Building on existing precision/recall work, we define a benchmarking methodology
+with a much larger ground-truth corpus, and provide a scalable, statistical
+approach to evaluation using this corpus. Inspired by the availability of large
+benchmarks for ATP tools~\cite{pelletier2002development}, we take those problem
+sets and automatically convert them into input/output examples for the CG task.
+
+We choose an ATP benchmark relevant to our domain. For our demonstration we
+chose Tons of Inductive Problems (TIP)~\cite{claessen2015tip}, which is relevant
+to software verification due to using higher-order logic, inductive definitions
+and relevant problems (arithmetic, sorting, etc.). TIP 0.2 contains 343 theorem
+proving problems, each defining the types and functions involved and stating a
+single property of those definitions. A total of 618 datatypes and 1498
+functions are defined, but most of these are duplicates (e.g. every problem
+involving \texttt{multiply} will include its own definitions of
+\texttt{multiply}, \texttt{add}, \texttt{successor}, \texttt{Natural}, etc.).
+Removing these redundancies, normalising the definitions and only selecting
+those which appear in at least one property leaves 182 functions for use in our
+corpus.
+
+We could, in theory, send these de-duplicated definitions straight into a CG
+tool and use the 343 properties as the ground truth for analysis. However, this
+number of definitions is too large to be feasible for current tools, and we
+would only get a single datapoint this way. Instead we \emph{sample} a subset of
+definitions, and form the ground truth by selecting those properties which only
+reference the sampled definitions. To avoid undefined precisions, we only use
+samples with at least one ground-truth property.
+
+We translate the function definitions into each tool's input language (Isabelle
+for \textsc{IsaCoSy} and Haskell for \textsc{QuickSpec}), run the tools with a
+timeout (5 minutes, based on preliminary experiments) and translate the outputs
+into a standard form for comparison with the ground-truths.
+
+\section{Demonstration}
+
+We applied our methodology to two existing CG tools:
+\textsc{QuickSpec}~\cite{QuickSpec} (a component of
+\textsc{HipSpec}~\cite{Claessen_hipspec:automating} and
+\textsc{Hipster}~\cite{Hipster}) enumerates expressions in
+Haskell~\cite{marlow2010haskell}, conjecturing that they're all equal unless
+proven otherwise by a counter-example finder;
+\textsc{IsaCoSy}~\cite{johansson2009isacosy} enumerates equations in
+\textsc{Isabelle}~\cite{nipkow2002isabelle}, using a constraint solver to avoid
+repeating itself.
+
+Their running times are shown in Figure~\ref{figure:quickspec_runtimes}, for
+sample of up to 20 functions. The measurements are paired, and we followed the
+Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup} to
+find that \textsc{QuickSpec} was faster for $67\%-98\%$ of sizes (with 95\%
+confidence). Only sizes 1 and 2 weren't significantly faster, due to too few
+samples. Timeouts bounding the difference make this a conservative estimate.
+
+Precision and recall are shown in Figure~\ref{figure:quickspec_precRec},
+although few \textsc{IsaCoSy} runs succeeded. Recalls remain roughly flat, as
+the algorithms are monotonic (adding definitions can't prevent conjectures),
+whilst precisions decrease due to the buildup of ``uninteresting''
+relationships. We pooled all successful samples together and compared: McNemar's
+test~\cite{mcnemar1947note} found \textsc{QuickSpec}'s recall significantly
+higher ($p = 0.0026$), and Boschloo's test~\cite{lydersen2009recommended} found
+no significant difference between the precisions ($p = 0.111$).
+
+\section{Conclusion}
+
+``Interestingness'' remains a difficult philosophical question, but we work
+around this to provide a practical benchmark which researchers can use for the
+foreseeable future to characterise and, importantly, \emph{compare} their
+solutions to the CG task. Although it does not capture every facet of the CG
+goal, our methodology builds on existing work and is easily extended to other
+corpora, tools and analyses.
+
+We have demonstrated that \textsc{QuickSpec} significantly out-performs
+\textsc{IsaCoSy}, and hope that such healthy competition will accelerate future
+developments, as has been seen in the field of ATP.
 

### 2018-04-19 PhDSymposium2018/supporting-materials/mkBenchmarkGraphs.py
Symposium graphs and build rules

@@ -1,0 +1,636 @@
+#!/usr/bin/env python
+from json import loads, dumps
+import os
+import sys
+msg = lambda x: sys.stderr.write(x + '\n')
+
+msg('Reading input')
+
+from json import load
+
+def readInput(sys):
+    var = os.getenv(sys + 'Data')
+
+    assert var is not None, 'No {} env var given'.format(sys)
+
+    with open(var, 'r') as f:
+        return load(f)
+
+data = {
+    'isacosy'   : readInput('isacosy'),
+    'quickspec' : readInput('quickspec')
+}
+del(readInput)
+
+msg('Normalising data')
+
+# Extract timeout from one of the QuickSpec runs, since IsaCoSy may not have
+# stored it
+timeout = data['quickspec'].values()[0].values()[0]['timeout']
+
+def normalise(data):
+    def makeHashable(x):
+        """Convert lists and dictionaries into tuples, recursively, so we can
+        turn them into sets for comparison."""
+        xType    = type(x)
+        reducers = {
+            type([]) : lambda rest, elem: rest +  (makeHashable(elem),),
+            type({}) : lambda rest, key:  rest + ((makeHashable(key),
+                                                   makeHashable(x[key])),)
+        }
+        return reduce(reducers[xType], x, ()) if xType in reducers else x
+
+    for size in data:
+        for rep in data[size]:
+            # Samples should be sets
+            data[size][rep]['sample'] = frozenset(data[size][rep]['sample'])
+
+            # So should found equations
+            data[size][rep]['found'] = frozenset(
+                makeHashable(data[size][rep]['found']))
+    return data
+
+data = {
+    'isacosy'   : normalise(data['isacosy'  ]),
+    'quickspec' : normalise(data['quickspec'])
+}
+
+del(normalise)
+
+def quickspecCheck(data):
+    foundResults = {}
+
+    for size in data:
+        for rep in data[size]:
+            rdata = data[size][rep]
+            found = rdata['found']
+
+            if len(found) > 0:
+                for known in foundResults:
+                    if len(foundResults[known]) > 0:
+                        if rdata['sample'].issubset(known):
+                            assert found.issubset(foundResults[known]), repr({
+                                'error'  : 'More names should give more output',
+                                'got'    : list(found),
+                                'known'  : list(known),
+                                'prev'   : list(foundResults[known]),
+                                'sample' : list(sample),
+                                'size'   : size
+                            })
[…]

### 2018-04-23 PhDSymposium2018/abstract.tex
Symposium 2018 abstract tweaks

@@
   exploration}. Such conjectures can be sent to automated theorem provers (ATP),
-have helped prove software correctness by finding key
+and have helped software correctness proofs by finding key
 lemmas~\cite{Claessen.Johansson.Rosen.ea:2013}. Automating formal methods this
@@
 promise in software
-engineering~\cite{McKinna:2006,chlipala2011certified,Xi2003}, automated research
+engineering~\cite{McKinna:2006,chlipala2011certified}, automated research
 ``assistants''~\cite{McCarthy_Programs59,lenat:77,benzmuller1997omegamega} and
@@
 
-  \caption{Time taken for CG on samples from TIP, with 300 second timeout. Lines
-    show medians and error bars show inter-quartile range, with 31 samples of
-    each size except 1 (14) and 2 (30). }
+  \caption{CG running time on definitions sampled from TIP (31 of each size,
+    except 1 (14) and 2 (30)). Failures (out-of-memory or 300 second timeout)
+    shown in red. Lines show medians, error bars are inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
 properties which they appear to satisfy. This can help formalise the
-``background knowledge'' required for proving, and present the user with ideas
-they hadn't considered.
+``background knowledge'' required for proving, and present users with novel
+ideas.
 
@@
 Their running times are shown in Figure~\ref{figure:quickspec_runtimes}, for
-sample of up to 20 functions. The measurements are paired, and we followed the
+samples of up to 20 functions. The measurements are paired, and we followed the
 Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup} to
@@
 confidence). Only sizes 1 and 2 weren't significantly faster, due to too few
-samples. Timeouts bounding the difference make this a conservative estimate.
+samples. This is also conservative, since the difference is bounded by the
+timeout.
 

### 2018-04-23 PhDSymposium2018/supporting-materials/mkBenchmarkGraphs.py

@@ -336,17 +336,11 @@ def plotTime(system, agg):
     plt.grid(color='grey', which='major', axis='y', linestyle='solid')
     plt.grid(axis='x', linestyle=' ', linewidth=0)
 
-    plt.xlim(0, 20 )
+    plt.xlim(1, 20 )
     plt.ylim(0, 300)
 
-    plt.xticks(range(0, 21))
+    plt.xticks(range(1, 21))
     plt.yticks([0, 100, 200, 300])
-
-    mean   = lambda l: sum(l) / len(l)
-
-    odd    = lambda n: (n // 2) == 1
-
-    median = lambda l: np.percentile(sorted(l), 50.0)
 
     sizes = sorted(list(set(agg['size'])))
 
@@ -360,40 +354,23 @@ def plotTime(system, agg):
 
     diff = lambda xs, ys: [xs[i] - ys[i] for i, _ in enumerate(xs)]
 
-    plt.errorbar(x=sizes, y=p50, yerr=(diff(p50, p25), diff(p75, p50)),
-                 #marker='+',
-                 elinewidth=0.5,# width of error bar line
-                 ecolor='k',    # color of error bar
-                 capsize=1.5,     # cap length for error bar
-                 capthick=0.5)   # cap thickness for error bar
-
-    # Filter data, if asked
-    newAgg = agg
-
-    keepIndices = [i for i,_ in enumerate(agg['size']) if succeeded(agg)(i)]
-    keepers     = lambda key: [agg[key][i] for i in keepIndices]
-    newAgg = {
-        'size'    : keepers('size'),
-        'time'    : keepers('time'),
-        'timeHue' : keepers('timeHue')
+    newAgg      = {
+        'size' : agg['size'],
+        'time' : agg['time'],
+        'hue'  : [0 if s else 1 for s in agg['success']]
     }
-
-    #plt.xlim((0, 21))
-    #plt.ylim((0, 300))
-    #newAx.set_xlim((0, 21))
-    #plt.xticks(range(1, 21))
 
     # Alternatively, we could use stripplot with jitter
     newAx = sns.swarmplot(data      = newAgg,
                           x         = 'size',
                           y         = 'time',
-                          size      = 2,  # Marker size
-                          marker    = '+',
-                          #edgecolor = 'k',
-                          color     = 'k',
+                          size      = 3,  # Marker size
+                          marker    = 'x',
+                          palette   = ['k', 'r'],
                           linewidth = 0.3,
-                          #hue       = 'timeHue',
+                          hue       = 'hue',
[…]

### 2018-04-24 TEBenchmarkPaper/article.tex
Strengthen motivation

@@
 % Lay out the field of TE: motivation, tools, evaluation, problems...
-% FIXME: Strengthen motivation:
-%  - Automated Scientific Discovery (e.g. Dundee drug discovery "centaur" story)
-%  - "Mixed initiative" (social machine?); maybe Kasparov
 
 \emph{Conjecture generation} is the open-ended task of producing conjectures
-about a given logical theory which are somehow ``interesting'', and is studied
-as a sub-field of \emph{mathematical theory exploration} (MTE).
-
-This has applications wherever we find uses for formal methods: in proof
-assistants and their libraries, in mathematics education and research, and in
-the specification and verification of software. Although formal methods, and
-automated reasoning more generally, have been advocated in mathematics (notably
-by Voevodsky~\cite{voevodsky2010univalent} and
-Gowers~\cite{ganesalingam2013fully}) and software engineering (for example via
-dependently-typed
-programming~\cite{McKinna:2006, chlipala2011certified, Xi2003}), they have a
-reputation for being difficult and expensive, which has historically limited
-their use to high-assurance domains such as aerospace and microprocessor design.
-One way to reduce these hurdles is to increase automation, which in the case of
-software verification can benefit from the introduction of conjecture generation
-tools~\cite{Claessen.Johansson.Rosen.ea:2013}. Another approach is to tailor the
-division of labour between humans and machines such that each is playing to
-their strengths, which has been referred to as ``centaur
-teams''~\cite{harari2017reboot, davenport2015beyond}. In this respect a
-conjecture generating tool can search a much larger space of relationships than
-a human, whilst leaving the ultimate judgement of which to investigate in more
-detail to the user.
+concerning a given logical theory which are somehow ``interesting'', and is
+studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
+applications wherever we find uses for formal methods: in proof assistants and
+their libraries, in mathematics education and research, and in the specification
+and verification of software. Although formal methods, and automated reasoning
+more generally, have been advocated in mathematics (notably by
+Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
+and software engineering (for example via dependently-typed
+programming~\cite{McKinna:2006}), they have a reputation for being difficult and
+expensive, which has historically limited their use to high-assurance domains
+such as aerospace and microprocessor design.
+
+One path to reducing these barriers is to increase the level of automation.
+Tools for automated theorem proving (ATP) have been shown to benefit from the
+addition of a conjecture generator, improving their ability to prove properties
+relevant to software verification~\cite{Claessen.Johansson.Rosen.ea:2013}.
+Another path is to tailor the division of labour between humans and machines
+such that each is playing to their strengths, which has been referred to as
+``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
+systematically search a much larger space of possible relationships than humans,
+potentially bringing to light novel and surprising connections, whilst allowing
+the user to make the ultimate judgement over which of the most promising results
+are worth investigating further.
 
@@
 Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example searching for new drugs. The major difference between scientific and
+example in the search for new drugs. The major difference between scientific and
 mathematical discovery is that inductive reasoning and experimental testing are,
@@
 
-We limit ourselves to applications in mathematics, but even here it is difficult
+We limit ourselves to mathematical applications, but even here it is difficult
 to measure and compare the success of approaches and tools. This is partially

### 2018-04-24 TEBenchmarkPaper/article.tex
TEB: Say "properties" rather than "theorems"

@@
 One path to reducing these barriers is to increase the level of automation.
-Tools for automated theorem proving (ATP) have been shown to benefit from the
-addition of a conjecture generator, improving their ability to prove properties
-relevant to software verification~\cite{Claessen.Johansson.Rosen.ea:2013}.
+Software verification tools have been shown to benefit from the addition of a
+conjecture generation step for finding ``background lemmas'', which are useful
+when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
 Another path is to tailor the division of labour between humans and machines
@@
 Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find theorems which describe its
-behaviour. This could be for mathematical curiosity, or due to the theory's
-importance in some domain. In particular, for theories which capture the
-semantics of some software library, we may want to verify that certain
-(un)desirable properties do (not) hold. We might also want to \emph{optimise}
-programs using this library, rewriting expressions into a form which requires
-less time, memory, network usage, etc. To avoid altering a program's result,
-such rewrites should come with theorems proving their correctness, such as the
-following theorem for our theory of lists:
+Figure~\ref{figure:list_theory}, we may want to find properties (such as
+invariants) describing its behaviour. This could be for mathematical curiosity,
+or due to the theory's importance in some domain. In particular, when dealing
+with definitions in a software library, we may regard a set of such properties
+as the code's \emph{specification}. Even those properties which only hold by
+coincidence, rather than by design, may still be useful for \emph{optimising}
+uses of this library, for example by rewriting expressions into equivalent forms
+which require less time, memory, network usage, etc. For example, our theory of
+lists turns out to satisfy the following invariant:
 
@@
 
-Equation~\ref{eq:mapreduce} justifies a rewrite rule for splitting a single
-$\Map$ call into multiple independent calls dealing with different sections of a
-list. Such rules can be useful optimisations, for example in a distributed
-context where each call can be evaluated in parallel (this is the basis of the
-``map/reduce'' programming paradigm).
-
-All of these example use cases require the ability to discover theorems about
-some particular definitions. This is a hard problem in general, but presents
-opportunities for automation due to the precise, symbolic nature of the domain.
+Whilst these two expressions produce equivalent results, the form on the right
+can execute both calls to $\Map$ in parallel (in a pure functional language, at
+least), giving rise to the ``map/reduce'' paradigm for parallel programming.
+Such use cases require the ability to \emph{discover} properties about some
+particular definitions; this requires not only \emph{proving} their correctness,
+but also \emph{posing} them as conjectures in the first place. This is a hard
+problem in general, but presents opportunities for automation due to the
+precise, symbolic nature of the domain.
 
@@
 theorems in every copy of their software, this is a good indication that they
-are interesting.
+are interesting properties of those definitions.
 
@@
 will be used as input to the tools, whilst the latter form a ground truth corpus
-against which to compare the output.
+of properties against which to compare the output.
 
@@
 aspects like the names of functions. For example, consider a problem set which
-includes a commutativity theorem for a \texttt{plus} function, and an
-associativity theorem for an \texttt{add} function, where the definitions of
+includes a statement of commutativity for a \texttt{plus} function, and of
+associativity for an \texttt{add} function, where the definitions of
 \texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
@@
 lexicographic order). In the above example, the \texttt{plus} function would be
-removed, and the commutativity theorem updated to reference \texttt{add}
+removed, and the commutativity statement updated to reference \texttt{add}
 instead.
@@
 We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the updated theorem statements as the ground truth for analysis.
-However, this would cause two problems:
+tool and use the entire set of properties (taken from the theorem statements) as
+the ground truth for analysis. However, this would cause two problems:
 
@@
 and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those theorems from the corpus which ``depend on''
+ground truth by selecting those properties from the corpus which ``depend on''
 (contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a theorem which only references
-\texttt{times} does not depend on \texttt{plus}, even if \texttt{plus} occurs in
-the definition of \texttt{times}).
+dependencies aren't required (e.g. a property involving only a \texttt{times}
+function would not depend on a \texttt{plus} function, even if \texttt{plus}
+occurs in the definition of \texttt{times}).
 
@@
 statistics) makes it less likely that a chosen sample will contain all of
-a theorem's dependencies. The majority of such samples would hence have an empty
+a property's dependencies. The majority of such samples would hence have an empty
 ground truth, and thus $0$ precision and undefined recall \emph{independent} of
@@
 Instead, we only allow a sample if it contains all dependencies of at least one
-theorem. We could do this using rejection sampling, but it is more efficient to
-pick a theorem, weighted in proportion to their number of dependencies and
-ignoring those with more dependencies than our sample size. That theorem's
+property. We could do this using rejection sampling, but it is more efficient to
+pick a property, weighted in proportion to their number of dependencies
+(ignoring those with more dependencies than our sample size). That property's
 dependencies become our sample, padded up to the required size with uniform
@@
 
-The ground truth for such samples is guaranteed to contain at least one theorem
+The ground truth for such samples is guaranteed to contain at least one property
 (the one we picked), and hence the precision and recall will depend meaningfully
@@
 To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth theorems, need to be parsed into a common format and
+well as the ground-truth properties, need to be parsed into a common format and
 compared syntactically after normalising away ``irrelevant'' details. We
@@
 We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth theorems
+coin toss whose bias is the precision; likewise, the ground truth properties
 (found or not) are like coin tosses biased by the recall. Note that this simple
@@
 the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth theorem is categorised as either being found only by the first
+ground-truth property is categorised as either being found only by the first
 tool, only by the second, by both or by neither. Those found by both or neither
@@
     \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};
+    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
 
@@
         (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {theorems}    (ref);
+        (sep) edge [arrow, sloped, above] node {properties}  (ref);
 
@@
 
-    \node (choose) [process, below=of thm   ]{Choose a theorem};
-    \node (deps)   [process, below=of choose]{List theorem dependencies};
+    \node (choose) [process, below=of thm   ]{Choose a property};
+    \node (deps)   [process, below=of choose]{List property's dependencies};
 
@@
 We chose TIP for our ground truth since it satisfies the desiderata of
-$\S$\ref{section:prep}: each problem has standalone type and function
+$\S$\ref{section:prep}: each benchmark problem has standalone type and function
 definitions, making their separation trivial; known examples from the software
@@
 We use TIP version 0.2 which contains 343 problems, each stating a single
-theorem and together defining a total of 618 datatypes and 1498 functions. Most
+property and together defining a total of 618 datatypes and 1498 functions. Most
 of these are duplicates, since each problem (re\nobreakdash-)defines all of the
@@
 ``destructors'' (elimination forms; field accessors). For example the type of
-lists from figure~\ref{figure:list_theory} can be defined in the TIP format as
+lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
 follows:
@@
 \begin{sloppypar}
-  We rewrite the TIP theorems (our ground truth) to reference these expanded
+  We rewrite the TIP properties (our ground truth) to reference these expanded
   forms instead of the raw constructors and destructors, and use these functions
@@
 total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which occur in at least
-one theorem statement (this removes ambiguity about which \emph{definitions}
-count as interesting and which are just ``implementation details'' for other
+and we choose to only sample from those 182 functions which are referenced by at
+least one property (this removes ambiguity about which \emph{definitions} count
+as interesting and which are just ``implementation details'' for other
 definitions).
@@
 sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth theorems.
+relationships between the dependencies of different ground truth properties.
 
@@
 QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some theorem statements, such as conditional equations, inequalities,
-terms larger than the search depth and those containing anonymous functions.
-The congruence closure algorithm used as a post-processor may also be removing
+synthesise some properties, such as conditional equations, inequalities, terms
+larger than the search depth and those containing anonymous functions. The
+congruence closure algorithm used as a post-processor may also be removing
 ``interesting'' results, for example if we found an ``uninteresting'' result
@@
 tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and theorems (which would
-also be of benefit to the original ATP benchmarks).
+the extension of existing corpora with new definitions and properties (which
+would also be of benefit to the original ATP benchmarks).
 

### 2018-04-26 TEBenchmarkPaper/article.tex
Fix a bunch of FIXMEs

@@
 We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We explain our proposal for a more general benchmark in
-$\S$\ref{sec:proposal} and demonstrate its application to existing MTE tools
-in $\S$\ref{sec:application}. Issues facing our approach, and the field in
-general, are discussed in $\S$\ref{sec:discussion}; related work is surveyed in
+difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
+explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
+demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
+Issues facing our approach, and the field in general, are discussed in
+$\S$\ref{sec:discussion}; related work is surveyed in
 $\S$\ref{sec:related-work}, including the diverse terminology found in the
@@
 % interestingness, e.g. examples from textbooks, etc.
-
-% FIXME: Perhaps mention examples like "plus(x, 0) = 0" as motivating examples
-% of things which are "simple" but which proof assistants still don't do
-% automatically (reference Moa's use of TE for background lemma generation)
 
@@
 
-% FIXME: We jump from conjectures to theorems. Maybe to prevent confusion we
-% should talk more generally about "interesting properties" of a formal system,
-% and use that as motivation for conjecturing and proving?
-
 Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (such as
+Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
 invariants) describing its behaviour. This could be for mathematical curiosity,
@@
 coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, for example by rewriting expressions into equivalent forms
-which require less time, memory, network usage, etc. For example, our theory of
-lists turns out to satisfy the following invariant:
+uses of this library, by rewriting expressions into equivalent forms which
+require less time, memory, network usage, etc. For example, our theory of lists
+turns out to satisfy the following invariant:
 
@@
 can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least), giving rise to the ``map/reduce'' paradigm for parallel programming.
-Such use cases require the ability to \emph{discover} properties about some
+least); a fact which is exploited by the ``map/reduce'' parallel programming
+paradigm.
+
+Such use cases require the ability to \emph{discover} true properties of some
 particular definitions; this requires not only \emph{proving} their correctness,
@@
 problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain.
+precise, symbolic nature of the domain. These processes can also feed into each
+other, since proven properties can be used as lemmas for subsequent proofs. As
+an example, proving Equation~\ref{eq:mapreduce} may require ``background
+knowledge'' which is tedious to produce by hand, such as
+$\forall x. \Append(x, \Nil) = x$.
 
@@
 either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. The fields of Automated Theory Formation (ATF)
-and Mathematical Theory Exploration (MTE) both tackle these generation tasks,
-and we discuss their relationship in historical context in
-$\S$\ref{related-work}.
-
-A major challenge for both ATF and MTE is choosing how to narrow down
-the set of generated conjectures to those deemed ``interesting'', since this is
-an imprecise term with many different interpretations. For example, all existing
-approaches agree that simple tautologies are ``uninteresting'', but differ when
-it comes to more complex statements.  Colton \etal{} give a survey of tools for
-theory formation and exploration, and their associated notions of
+definitions are also generated. This research has been pursued under the names
+of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
+we will use the MTE terminology, and discuss their relationship further in
+$\S$\ref{sec:related-work}.
+
+A major challenge when generating conjectures is choosing how to narrow down the
+resulting set to only those deemed ``interesting'', since this is an imprecise
+term with many different interpretations. For example, all existing approaches
+agree that simple tautologies are ``uninteresting'', but differ when it comes to
+more complex statements.  Colton \etal{} give a survey of tools for theory
+formation and exploration, and their associated notions of
 ``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
-notions are identified, which are applied to conjectures as follows:
+notions are identified, which are applied to conjectured properties as follows:
 
 {\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding false conjectures,
-without resorting to a full proof search.
-
-{\bf Novelty} depends on whether a conjecture, or one isomorphic or more
+specific examples. This is especially useful for avoiding falsehoods, without
+resorting to a deductive proof search.
+
+{\bf Novelty} depends on whether the property, or one isomorphic or more
 general, has already been seen.
 
-{\bf Surprisingness} of a conjecture is whether or not it is
-``obvious'', for example if it is an instance of a tautology.
+{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
+example if it is an instance of a tautology.
 
 {\bf Applicability} depends on the number of models in which a
-conjecture holds. The tool of \Bagai{} conjectures the non-existence of
-objects, and hence favours statements with zero applicability. Other tools treat
-applicability as a positive aspect: the more applicable the statement, the more
-interesting it is.
-
-{\bf Comprehensibility} depends on the complexity of a statement. Simpler
-statements are considered more interesting, hence many of the tools explore
-simpler/smaller statements before complex/larger ones, to more efficiently find
-those which are interesting.
-
-{\bf Utility} is the relevance or usefulness of a conjecture to the
-user's particular task. For example, if we want to find optimising
-rewrite rules such as equation~\ref{eq:mapreduce}, then utility would include
-whether or not a conjecture justifies a rewrite rule, the difference in
-resource usage of the expressions involved, and how common those
-expressions are in real usage.
-
-% FIXME: This is original work, so should be in a different section
-\begin{table}
-  \centering
-  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
-    \hline
-    \multirow{2}{*}{\textbf{Program}}                      &
-    \multirow{2}{*}{\textbf{Conjecture Types}}             &
-    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
-    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
-    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
-    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
-    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
-    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
-    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
-    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
-    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
-    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
-    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
-  \end{tabular}
-  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
-    to include four more recent tools. The interestingness measures are
-    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
-    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
-  \label{table:colton}
-\end{table}
-
-% JUSTIFICATIONS
-%
-% IsaCoSy uses counter-example checking; this ensures empirical plausibility
-% IsaCoSy uses constraints to avoid special-cases; this ensures novelty
-% IsaCoSy uses utility, since there are some hard-coded patterns which are
-% looked for
-%
-% Speculate uses a form of unification to ensure novelty, making sure one
-% equation is not a special-case of another
-% Speculate uses LeanCheck to enumerate values, looking for counterexamples
-% Speculate uses testing, to ensure Empirical Plausibility
-%
-% QuickSpec uses QuickCheck to ensure empirical plausibility
-% QuickSpec uses a congruence closure algorithm to ensure novelty
-%
-% IsaScheme uses utility, since it determines the "quality" of a definition
-% based on how many theorems it appears in?
-% IsaScheme uses novelty, using an equational rewrite system (Knuth-Bendix
-% completion) to remove redundancies
-% IsaScheme uses utility, since it focuses on the terms and definitions of a
-% user's theory. However, don't all of them?
-% IsaScheme uses counterexample checking for empirical plausibility.
-
-% ONLY APPLIES TO PRECONDITIONS
-% APPLICABILITY: ensure a conjecture holds in many models (or exactly zero, in
-% the case of Bagai et al). Does the testing-based approach of QuickSpec and
-% Speculate ensure that there's a model? After all, these are concrete values
-% rather than e.g. implications of some abstract specification.
-
-% FIXME: Give brief description and reference for each system in the text.
-% We could include the more lengthy descriptions of IsaCoSy and QuickSpec here?
-
-We summarise the criteria in Table 1 and state which were used in key
-ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
-on this survey (the latter is Colton's own tool). We extend this analysis to
-some more recent tools in the remaining rows (based on our understanding of
-their function): QuickSpec~\cite{QuickSpec},
-Speculate~\cite{braquehais2017speculate},
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
-and IsaScheme~\cite{MontanoRivas2011}.
-
-The latter tools are described in more detail in $\S$\ref{section:mte},
-but all use testing to check for counterexamples, which ensures results are
-empirically plausible and applicable (conditions are satisfiable, types are
-inhabited, etc.). They also use a small-to-large search order, ensuring more
-easily-comprehensibile conjectures are explored first.
-
-IsaCoSy ensures novelty using a constraint system to prevent special-cases being
-generated, whilst the others use post-hoc filters based on unification and term
-rewriting to achieve the same result. IsaCoSy is able to look for commonly
-desirable patterns, such as commutativity and associativity, which increases the
-utility of its output, whilst IsaScheme is completely based around such
-pattern-instantiation.
-
-This diversity of approaches makes it difficult to compare the existing
-evaluations of these tools directly. In particular, evaluation methods created
-for one tool might not make sense for another, due to assumptions made about
-the algorithms' operation. For example, the novelty filters used by QuickSpec
-and Speculate are applied to the whole output set, guaranteeing that no
-special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
-since its constraint solver allows special cases if they're found \emph{before}
-the general case (it does not go back and discard previous results).
-
-With such ambiguous and varied goals, approaches and assessment criteria it is
-difficult to compare MTE tools in a quantitative way, and hence to form some
-measure of ``progress'' for the field. Our proposed benchmarking methodology
-attempts to remedy this by setting a clear target, hopefully in line with the
-myriad goals of existing researchers, whilst being generally applicable to
-tools of this type.
-
-\subsection{Existing Evaluations}
-\label{sec:existing}
-
-There are three aspects to consider when evaluating a tool for conjecture
-generation:
-
-% FIXME: How are these referred to/considered in the paper?
-\begin{enumerate}
-\item The quality of each individual conjecture, which we will refer to as their
-  ``interestingness''.
-\item The quality of the \emph{set} of generated conjectures, which is important
-  for criteria like novelty, as well as to assess consistency.
-\item Resources used by the tool, to find the balance struck between output
-  quality and time taken.
-\end{enumerate}
-
-We make the assumption that each tool is intended to be \emph{general}, and
-hence isn't hard-coded or fine-tuned to only perform well in a narrow domain.
-Examples where this does \emph{not} hold include the natural number heuristics
-of AM discussed in $\S$\ref{section:atf}, as well as machine learning models
-trained on some specific dataset (such as the benchmark we describe).
-
-Evaluation of new tools and techniques for conjecture generation tends to be
-\emph{ad hoc} and hence difficult to compare, whilst comparative surveys (such
-as Colton's) are descriptive and qualitative rather than providing direct
-numerical measurements.
-
-One quantitative approach which has gained traction is \emph{precision/recall
-  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
-providing a direct comparison of these three
-tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
-  reports prior results from the other tools, rather than reproducing them.
-  This makes the numbers less comparable, since each tool was tested with
-  slightly different definitions (e.g. whether or not the exponential and
-  maximum functions were included).} To measure precision and recall we first
-choose a set of definitions (such as those in figure \ref{figure:list_theory})
-and a \emph{ground truth} to treat as the ``ideal'' set of conjectures we would
-like a tool to produce for those definitions (e.g. this might include
-equation~\ref{eq:mapreduce}). Each tool is run and their outputs are compared to
-the ground truth. To score 100\% on precision and recall, all of the conjectures
-which appear in the ground truth must be generated, and nothing else:
+property holds. High applicability makes a property more interesting, but so
+does zero applicability (i.e. non-existence).
+
+{\bf Comprehensibility} depends on the syntactic complexity of the property's
+statement. Simpler statements are considered more interesting (which favours
+tools adopting a small-to-large search order).
+
+{\bf Utility} is the relevance or usefulness of a property to the user's
+particular task. For example, if we want to find optimisations like
+Equation~\ref{eq:mapreduce}, then utility would include whether the property
+allows rewriting, the difference in resource usage of the expressions involved,
+and how common those expressions are in real usage.
+
+Attempting to \emph{measure} such qualities is difficult, and having too many
+measures complicates comparisons. System developers have employed a more
+practical alternative in their evaluations, which is to perform
+\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
+choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
+and a set of properties (the ground truth) which represent the ``ideal'' result
+of conjecture generation for those definitions (e.g. this might include
+Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
+run it on these chosen definitions and compare its output to the ground truth:
 
@@
 \item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output large
-  numbers of conjectures in the hope that some turn out to be ``good''.
+  the ground truth. This penalises overly-liberal tools which output a large
+  number of properties in the hope that some turn out to be ``good''.
 \item \emph{Recall} is the proportion of the ground truth which appears in the
   tool's output. This penalises overly-conservative tools which generate very
-  few conjectures as a way to avoid ``bad'' ones.
+  few properties as a way to avoid ``bad'' ones.
 \end{itemize}
 
-Precision and recall act as quality tests for sets of conjectures, fulfilling
-one of our requirements. They also provide a coarse quality assessment for
-individual conjectures: either a conjecture appears in the ground truth set or
-it doesn't.
-
-The analyses in~\cite{claessen2013automating} took their definitions and ground
-truths from the standard library of the Isabelle proof assistant: since the
-library authors have gone to the effort of stating, proving and including these
-theorems in every copy of their software, this is a good indication that they
-are interesting properties of those definitions.
-
-Tests were performed with a theory of natural numbers and a theory of lists (the
-definitions shown in figure~\ref{figure:list_theory}), %FIXME: add page ref
-but it is difficult to
-draw general conclusions from only two datapoints. These theories were also
-quite small, the natural number theory
-from~\cite{Johansson.Dixon.Bundy:conjecture-generation} only containing 4
-definitions and 12 theorems.
-
-Despite these limitations we believe that this is a viable approach to drawing
-meaningful, objective conclusions about such tools; but only with larger data
-sets can we judge performance in different, especially \emph{novel}, domains
-(which, after all, is the purpose of such tools). In addition, we might be
-concerned that such ``popular'' theories may be unrepresentative of typical
-usage. For example, we would not expect the library functions in a typical
-verification problem to be as mathematically rich as number theory.
+To score 100\% on precision and recall, all of the properties which appear in
+the ground truth must have been conjectured, and nothing else. This gives us a
+simple method to evaluate and compare tools without requiring a general solution
+to the question of what is ``interesting''; although we must still decide what
+to put in the ground truth set, and what to leave out, for each measured set of
+definitions.
+
+The precision and recall of three MTE tools are compared
+in~\cite{claessen2013automating}:
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
+IsaScheme~\cite{MontanoRivas2011} and
+HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
+  comparison only reports prior results for each tool, rather than reproducing
+  them. This makes the numbers less comparable, since each tool was tested with
+  slightly different definitions (e.g. whether or not the exponential and
+  maximum functions were included).} The definitions and ground truths for all
+three were taken from the standard library of the Isabelle proof assistant:
+since the library authors have gone to the effort of stating and proving these
+properties, and including them with every copy of their software, this is a good
+indication that they are interesting. Two sets of definitions were used: a
+theory of natural numbers and of lists (which we show in
+Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
+two datapoints containing only a few definitions and properties (e.g. the
+natural number theory contained 4 functions with 12 properties), it is difficult
+to draw general conclusions; yet despite these limitations we believe that this
+is a viable method to performing meaningful, objective analyses and comparisons
+for such tools. It is only with larger data sets that we can judge performance
+in different, especially \emph{novel}, domains (which, after all, is the purpose
+of MTE tools). In addition, we might be concerned that such ``popular'' theories
+as the natural numbers may be unrepresentative of typical usage. For example, we
+would not expect the library functions in a typical verification problem to be
+as mathematically rich as number theory.
+
+\subsection{Existing Tools}
+\label{sec:existing-tools}
+
+Since our benchmarking methodology builds on the precision/recall analysis
+described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
+and QuickSpec (the conjecture generating component of HipSpec), as our
+demonstration. Other notable tools, including IsaScheme, are described under
+related work in $\S$\ref{sec:related-work}.
+
+IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
+assistant, mostly in Standard
+ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
+by enumerating expressions involving a given set of (typed) constants and free
+variables (a \emph{signature}). A constraint solver forbids certain
+(sub)expressions from being synthesised, and these constraints are extended
+whenever a new property is conjectured, to avoid generating any special-cases of
+this property in the future. Conjectures are tested by looking for
+counterexamples and, if none are found, sent to IsaPlanner which attempts to
+prove them.
+
+QuickSpec emerged from work on the QuickCheck software testing framework for the
+Haskell programming language~\cite{claessen2011quickcheck}. As well as
+generating conjectures for HipSpec (also a Haskell tool) it has also been
+applied to Isabelle via the Hipster tool~\cite{Hipster}. Like IsaCoSy, QuickSpec
+takes a signature and enumerates well-typed terms, collecting together those of
+the same type into equivalence classes, assuming them to be equal. QuickCheck is
+then used to find counterexamples to this assumption, which it does by randomly
+instantiating the variables and comparing the resulting closed expressions. Any
+equivalence class whose elements don't compare equal are split up, and the
+process is repeated with new random values.
+
+After 500 rounds of testing, any expressions still sharing the same equivalence
+class are conjectured to be equal for all values of their variables. Rather than
+using a constraint system to prevent redundancies (such as special-cases of
+other properties), QuickSpec instead passes its output through a congruence
+closure algorithm to achieve the same effect.
 
@@
 
-Our main contribution is a benchmarking methodology, shown in figure
-\ref{figure:flow_chart}, which both generates a large definition/ground-truth
-corpus, and provides a scalable, statistical approach to evaluating MTE tools
-using this corpus. We follow a precision/recall approach similar to prior work,
-with the main difference being the source of definitions and ground truths: we
-take existing problem sets designed for automated theorem proving, and adapt
-their content for use in the conjecture generation setting.
+Our main contribution is a benchmarking methodology, shown in
+Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
+generates a large definition/ground-truth corpus, and provides a scalable,
+statistical approach to evaluating MTE tools using this corpus. We follow a
+precision/recall approach similar to prior work, with the main difference being
+the source of definitions and ground truths: we take existing problem sets
+designed for automated theorem proving, and adapt their content for use in the
+conjecture generation setting.
 
@@
   TPTP~\cite{sutcliffe2009tptp}).
-\item Since it will act as our ground truth, the problem set should
-  \emph{ideally} contain every ``interesting'' conjecture involving its included
-  definitions. More \emph{realistically}, we should aim for each definition to
-  appear in \emph{many} theorem statements; rather than each problem having
-  unique definitions. % FIXME: Reword: make clear that non-membership means
-  % uninteresting, and hence the things we *don't* include are just as important
-  % as the things we do
-\item The problem set should be as large as possible, for robustness of the
-  resulting statistics. % FIXME: Prioritising quantity over quality?
+\item The problem set should ideally contain \emph{every} ``interesting''
+  property involving its included definitions, since non-membership in the
+  ground truth will be treated as being ``uninteresting''. More realistically,
+  we should at least aim for each definition to have \emph{multiple} properties;
+  rather than a mixture of mostly unrelated properties.
+\item Larger problem sets are preferable, all else being equal, as they give
+  more robust statistics.
 \end{itemize}
@@
 \subsection{Sampling}
+\label{section:sampling}
 
@@
 \subsection{Evaluation}
+\label{section:evaluation}
 
@@
        (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm] {Preparation};
+         node [black, midway, right, xshift=0.3cm]
+              {Preparation $\S$\ref{section:prep}};
 
@@
        (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm] {Sampling};
+         node [black, midway, right, xshift=0.3cm]
+              {Sampling $\S$\ref{section:sampling}};
 
@@
        (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm] {Evaluation};
+         node [black, midway, right, xshift=0.3cm]
+              {Evaluation $\S$\ref{section:evaluation}};
+
   \end{tikzpicture}
@@
 
-IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
-ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
+We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
 with the 2015 version of Isabelle. The following issues had to be overcome to
@@
 
-%IsaCoSy
-The IsaCoSy system (Isabelle Conjecture Synthesis) conjectures equations by
-enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new conjecture is generated, to avoid generating any special-cases of
-this conjecture in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-%QuickSpec
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}, and has been used to
-generate conjectures for the subsequent
-HipSpec~\cite{Claessen_hipspec:automating} and Hipster~\cite{Hipster} MTE
-systems. Like IsaCoSy, a signature is provided and terms are enumerated. Rather
-than a constraint system, QuickSpec version 1 collects expressions of the same
-type into equivalence classes, assuming them to be equal. QuickCheck is used to
-test this assumption by instantiating each variable with a randomly chosen
-value, evaluating the resulting expressions and comparing them. Any equivalence
-class whose elements don't compare equal are split up to separate those
-observably-different expressions, and the process is repeated with new random
-values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Finally, a
-congruence closure algorithm is applied to these equations, to remove
-redundancies and special-cases.
-
 % FIXME: CITE

### 2018-04-30 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Inline swarmplots into prec/rec and time functions.

@@ -212,7 +212,7 @@ def aggregateData(data):
             r = r or 0
             assert num(r), 'Non-numeric rec %r' % type(r)
 
-            # Store in "wide format". We +1 the hues and use 0 for failure
+            # Store in "wide format"
             agg['correct'  ].append(correct)
             agg['found'    ].append(found)
             agg['precision'].append(p)
@@ -222,7 +222,7 @@ def aggregateData(data):
             agg['size'     ].append(size)
             agg['success'  ].append(rdata['success'])
             agg['time'     ].append(t)
-            agg['timeHue'  ].append(found + 1 if rdata['success'] else 0)
+            agg['timeHue'  ].append(found)
             agg['wanted'   ].append(wanted)
 
     return agg
@@ -295,8 +295,7 @@ def makeColours(agg):
         return {
             'cmap'    : ListedColormap(colourMap),
             'norm'    : mpl.colors.Normalize(vmin=0, vmax=maxVal),
-            'palette' : dict(enumerate(['#ff0000'] + colourMap \
-                                       if key == 'found' else colourMap))
+            'palette' : dict(enumerate(colourMap))
         }
     return go
 
@@ -320,45 +319,6 @@ def newFigure(name):
     sns.set_context('paper')
 
     return fig
-
-def drawPoints(agg, y=None, colours=None, hue=None, yLabel=None, ax=None,
-               condition=None):
-    '''Draw a scatter plot of agg[y][n] against agg['size'][n], for each point
-    n. Points will be spread out horizontally to avoid overlaps, and the colour
-    of each will be colours['palette'][agg[hue][n]].
-
-    If condition is given, we filter the data to only keep those where filter(n)
-    returns True, where n is the index as above.
-
-    The the x-axis will be labelled with "Sample size", the y-axis with yLabel.
-
-    The resulting plot (axes) are returned. To draw on existing axes, pass them
-    in as ax.'''
-
-    # Filter data, if asked
-    if condition is not None:
-        keepIndices = filter(condition, [i for i, s in enumerate(agg['size'])])
-        keepers     = lambda key: [agg[key][i] for i in keepIndices]
-        agg = {
-            'size' : keepers('size'),
-            y      : keepers(y),
-            hue    : keepers(hue)
-        }
-
-    # Alternatively, we could use stripplot with jitter
-    newAx = sns.swarmplot(data      = agg,
-                          x         = 'size',
-                          y         = y,
-                          size      = 2,  # Marker size
[…]

### 2018-04-30 TEBenchmarkPaper/article.tex
More fixes

@@
 
-% FIXME: Add section references to the braces
 % FIXME: Maybe rethink "analysis" bit; maybe add summary and comparison
@@
 \label{sec:application}
-% FIXME: Maybe use (red?) crosses rather than red dots? More useful for black and white
+
 We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
@@
 
-\subsection{TIP} % FIXME: Expand acronym
-
-We chose TIP for our ground truth since it satisfies the desiderata of
-$\S$\ref{section:prep}: each benchmark problem has standalone type and function
-definitions, making their separation trivial; known examples from the software
-verification and inductive theorem proving literature are included, ensuring
-relevance to those fields; the format includes the higher-order functions and
-inductive datatypes we are interested in; it is large enough to pose a challenge
-to current MTE tools; plus it is accompanied by tooling to convert its custom
-format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
-languages, including Haskell and Isabelle.
+\subsection{Tons of Inductive Problems}
+
+We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
+since it satisfies the desiderata of $\S$\ref{section:prep}: each benchmark
+problem has standalone type and function definitions, making their separation
+trivial; known examples from the software verification and inductive theorem
+proving literature are included, ensuring relevance to those fields; the format
+includes the higher-order functions and inductive datatypes we are interested
+in; it is large enough to pose a challenge to current MTE tools; plus it is
+accompanied by tooling to convert its custom format (an extension of
+SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
+Isabelle.
 
@@
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
-    indicates the total number of conjectures produced by each run, with failing
-    processes shown in red. Points are spread out horizontally to prevent
-    overlaps. Box plots show median and quartiles, with whiskers showing points
-    at 1.5$\times$~inter-quartile range.}
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
+    processes (timeouts or out-of-memory) are shown in red. Points are spread
+    out horizontally to reduce overlapping. Box plots show median and quartiles,
+    with whiskers showing points at 1.5$\times$~inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
   \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to avoid overlaps). Point colour shows how many conjectures
-    were in the ground truth set for each run. Lines show the combined
-    proportion for each sample size (ratio of averages) and the shaded region is
-    the sample standard deviation.}
+    horizontally to reduce overlapping). Lines show the combined proportion for
+    each sample size (ratio of averages) and errorbars show the sample standard
+    deviation.}
   \label{figure:quickspec_precRec}
@@
   \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
-    indicates the total number of conjectures produced by each run, with failing
-    processes (timeouts and out of memory) shown in red. Points are spread out
-    horizontally to prevent overlaps. Box plots show median and quartiles, with
-    whiskers showing points at 1.5$\times$~inter-quartile range.}
+    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
+    processes (timeouts and out of memory) are shown in red. Points are spread
+    out horizontally to reduce overlapping. Box plots show median and quartiles,
+    with whiskers showing points at 1.5$\times$~inter-quartile range.}
   \label{figure:isacosy_runtimes}
@@
   \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to avoid overlaps). Lines show the combined proportion for each
-    sample size (ratio of averages) and the shaded region is the sample standard
+    horizontally to reduce overlapping). Lines show the combined proportion for
+    each sample size (ratio of averages) and errorbars show the sample standard
     deviation.}

### 2018-04-30 TEBenchmarkPaper/article.tex
TEB:More fixes

@@
 
-% FIXME: Maybe rethink "analysis" bit; maybe add summary and comparison
-% references (although maybe not as full blown steps)
 \begin{figure}
@@
 \subsection{Tons of Inductive Problems}
+\label{sec:tip}
 
@@
 
-% FIXME: The designers must decide on, not the system
-Fundamentally, any mathematical reasoning system must decide on, and formalise,
-what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
-``provable'' include trivial tautologies, while at the same time failing to
-capture the ``almost true'', which can be a valuable trigger for theory change,
-as demonstrated by Lakatos in his case studies of mathematical
+Fundamentally, when designing any mathematical reasoning system we must decide
+on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
+such as ``true'' or ``provable'' include trivial tautologies, while at the same
+time failing to capture the ``almost true'', which can be a valuable trigger for
+theory change, as demonstrated by Lakatos in his case studies of mathematical
 development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
@@
 correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
-table~\ref{table:colton}, from computer scientists like Colton, are based
-largely on ``intuition'', plausibility arguments about why a metric would be
-important, and previous use of such metrics (in the case of
-``surprisingingness'', from a single quote from a mathematician).
-Opinions from mathematicians themselves include Gowers' suggestion that we can
-identify features which are commonly associated with good
-proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
-book containing the most elegant proof of every theorem, brought to life
-in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
+beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
+Colton (such as those in~\cite{colton2000notion}) are based largely on
+``intuition'', plausibility arguments about why a metric would be important, and
+previous use of such metrics (in the case of ``surprisingingness'', from a
+single quote from a mathematician). Opinions from mathematicians themselves
+include Gowers' suggestion that we can identify features which are commonly
+associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
+Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
 interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
@@
 such output. We do not believe this to be a realistic problem for the time
-being, although we would be delighted if proved wrong. % FIXME: Is there a solution? Mention that this is common to most tests
+being, although it does emphasise the continued importance of testing these
+tools with real human users, rather than relying solely on artificial benchmarks
+such as ours.
 
@@
 be run without a depth limit, and form part of the ``scientist assistant'' role
-envisaged by Lenat (or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be designed for
+which Lenat envisaged for AM, or McCarthy's earlier ``advice
+taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
 such long-running, potentially interactive programs, but that is beyond the
@@
 
-% FIXME: Reword into a short intro, e.g. "There is a long history of automating
-% mathematics, going back to... A recurring theme is separation between the
-% efforts of pure mathematicians and those wanting practical engineering
-% applications....
-The first automated mathematical tool, the mechanical calculator
-(known as the Pascaline), was an adding machine that could perform
-additions and subtractions directly and multiplication and divisions
-by repetitions, and was conceived by Pascal in 1642 while reorganising
-tax revenues~\cite{d'ocagne}. Subsequent early tools include
-M\"uller's universal calculating machine in 1784, which he built for
-the purpose of calculating and printing numerical tables: he invented
-this when he had to check and recalculate some tables relating to the
-volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
-Babbage invented his famous difference engine: an automatic,
-mechanical calculator designed to tabulate polynomial functions. This
-was inspired by a flawed table of logarithms and the idea that
-machines would be quicker and more reliable~\cite{bowden}. It is
-interesting to note background and motivation: while Pascal and
-Babbage were mathematicians, with an interest in engineering, M\"uller
-was an engineer with an interest in mathematical knowledge. All three
-tools were conceived as an aid to mathematicians, as well as
-scientists, accountants and surveyors.
-
-Differences in background and motivation continue to be relevant
-today. While the majority of work in automating mathematics has been
-in symbolic manipulation and theorem proving, we are concerned here
-with other aspects of mathematics, including the construction of conjectures,
-(counter\nobreakdash-)examples and theorems. These tasks, along with the
-formation of new definitions and concepts, are varyingly known as
-``Automated Theory Formation''~\cite{lenat:77,colton:book},
-``Mathematical Theory Exploration''~\cite{buchberger:06} (also sometimes
-prefaced with ``Computer-Aided'', ``Automated'' or ``Algorithm-Supported''),
-``Automated Mathematical Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
+The automation of mathematical tasks has been pursued since at least the time of
+mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
+these efforts is the separation between those undertaken by mathematicians like
+Pascal and Babbage~\cite{bowden}, and those of engineers such as
+M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
+are concerned with (automatically constructing and evaluating concepts,
+conjectures, theorems, axioms, examples, etc.) being divided into two main
+fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
+sometimes prefaced with ``Computer-Aided'', ``Automated'' or
+``Algorithm-Supported''), which is championed by mathematicians such as
+Buchberger~\cite{buchberger}; and Automated Theory Formation
+(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
+Other related terms include ``Automated Mathematical
+Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
 ``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}. Such a plethora of terminology can
-be unhelpful and can mask similarities between the different fields. In
-particular, the twin strands of Automated Theory Formation and
-Automated Mathematical Theory Exploration seem to be developing
-somewhat independently without a clear differentiating
-methodology. Below we discuss commonalities and differences between
-the two schools of thought.
-
-\subsection{Automated Theory Formation}
-\label{section:atf}
-
-% FIXME: Cut this down a lot. Currently it's not even an ATF section, it's just
-% an AM section. We should maybe remove its status as a section, and instead
-% just remark on the influence of AM on the field
-
-% 1. history of terminology
-Automated theory formation derives its terminology from psychology in which the
-term ``concept formation'' is used (see, for example,~\cite{bruner:67}) to
-describe the search for features which differentiate exemplars from
-non-exemplars of various categories. Lenat used this term in his 1977 paper:
-{\em Automated Theory Formation in Mathematics}~\cite{lenat:77}.
-
-% 4. example tools - AM, HR, ...
-When Lenat built the AM system~\cite{lenat:77}, there were systems
-which could define new concepts for investigation, such as those
-described in~\cite{winston}, and systems which could discover
-relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
-system could perform both of these tasks: Lenat saw this as the next step. AM
-was designed to both construct new concepts and conjecture relationships between
-them; fully automating the cycle of discovery in mathematics. Lenat describes
-this as follows:
-
-\begin{quote}
-``What we are describing is a computer program which
-defines new concepts, investigates them, notices
-regularities in the data about them, and conjectures
-relationships between them. This new information is used
-by the program to evaluate the newly-defined concepts,
-concentrate upon the most interesting ones, and iterate the
-entire process.''~\cite[p. 834]{lenat:77}
-\end{quote}
-
-AM was a rule-based system which used a frame-like scheme to represent
-its knowledge, enlarged its knowledge base via a collection of
-heuristic rules, and controlled the firing of these rules via an
-agenda mechanism. Lenat chose elementary arithmetic as the development
-domain because he could use personal introspection for the heuristics
-for constructing and evaluating concepts. Given the age of this
-discipline, Lenat thought it unlikely that AM would make significant
-discoveries, although he did cite its ``ultimate achievements'' as the
-concepts and conjectures it discovered (or could have discovered). He
-suggested various criteria by which his system could be evaluated,
-many of which focused on an exploration of the techniques. For
-instance, he considered generality (running AM in new domains) and how
-finely-tuned various aspects of the program are (the agenda, the
-interaction of the heuristics, etc). %most of which were qualitative.
-Lenat saw his system and future developments in this field as having
-implications for mathematics itself (finding results of significance),
-for automating mathematics research (developing AI techniques), and
-for designing ``scientist assistant'' programs (aids for
-mathematicians). This shows a broad spread of motivation. Despite the
-seeming success of the AM system, it is one of the most criticised
-pieces of AI research. In their case study in methodology, Ritchie and
-Hanna analysed Lenat's written work on AM and found that there was a
-large discrepancy between his theoretical claims and the implemented
-program~\cite{partridge}. For instance, Lenat made claims about how AM
-invented natural numbers from sets, whereas it used one heuristic
-which was specifically written in order to make this connection (and
-not used in any other context). Another problem was that the processes
-were sometimes under-explained. For an argument of why many of the
-claims made by Lenat about AM were false, see chapter 13 of~\cite{colton:book}.
-
-\subsection{Mathematical Theory Exploration}
-\label{section:mte}
-
-The phrase Mathematical Theory Exploration (MTE) is a recent term % FIXME: When?
-which seems to have originated with Buchberger and colleagues (see, % FIXME: informal
-for example,~\cite{buchberger}). His motivation is to support
-mathematicians during their exploration of mathematical theories. This
-support is intended to be for the straightforward reasoning, which he
-argues, covers most mathematical thought, rather than the ingenious
-points, which he leaves for human mathematicians. Buchberger's long
-term goal is to provide routine tools for the exploration activity of
-working mathematicians, to support the invention and structured
-build-up of mathematical knowledge. The Theorema project aims at
-prototyping features of a system for such a purpose. These features
-include Integration of the Functionality of Current Mathematical
-Systems (retention of the full power of current numerics and computer
-algebra systems, as well as enabling the user to add their own
-algorithms to the system); Attractive Syntax (input and output is
-readable and presented attractively, and can be personalised by the
-user); and Structured Mathematical Knowledge Bases (tools are provided
-for building and using large mathematical knowledge
-libraries). Buchberger has evaluated the potential of this strategy by
-illustrating the automated synthesis of his own Gr\"obner bases
-algorithm~\cite{buchberger:04}.
-
-Recent systems developed in this area include IsaScheme~\cite{MontanoRivas2011},
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-QuickSpec~\cite{QuickSpec} and Speculate~\cite{braquehais2017speculate}
-(included in table~\ref{table:colton}) as well as MATHsAiD~\cite{roy}.
-
-The goal of the MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses,
-Axioms and Definitions) project is to build a tool which takes in a set of
-axioms, concept definitions and a logic and applies its inference rules to
-reason from the axioms to theorems. The motivation is to produce a tool which
-will help mathematicians to explore the consequences of a set of axioms or a
-particular concept definition. One of the main challenges of the project has
-been to automatically evaluate interestingness: to distinguish important
-theorems from results which, although they follow from a set of axioms, are of
-little mathematical interest.
-
-% IsaScheme FIXME: Maybe strip down this description? We could use prose instead,
-% like "schemes capture common patterns such as iteration, which can be
-% instantiated with an addition function to produce multiplication, or with
-% multiplication to produce exponentiation.
-Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
-IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
-which can be used to generate new concepts and conjectures; variables within the
-scheme are instantiated automatically and this drives the invention process.
-For instance, in the theory of natural numbers, given the concepts of zero
-($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
-following scheme to invent the concept of multiplication:
-
-\[
-\text{def-scheme}(g,h,i,j) \equiv\\
-  \exists f. \forall x y. \left\{
-  \begin{array}{l l}
-    f(g,y) = h (y)  \: \text{and}  & \\
-    f(i(x),y) = j(y,f(x,y)) & \\
-  \end{array} \right.
-\]
-
-where the existentially quantified variable $f$ stands for the new
-function to be defined in terms of the variables $g$, $h$, $i$ and $j$. Within
-the theory of natural numbers, IsaScheme instantiates this scheme with
-$\sigma_1 = \{g \mapsto \Zero, h \mapsto (\lambda x.\Zero), i \mapsto \Succ, j
-\mapsto \plus\}$. In this example, $f \mapsto \mult$ (multiplication), since:
-
-\begin{align*}
-\mult(\Zero, y)  &= \Zero \\
-\mult(\Succ(x), y) &= \plus(y, \mult(x, y))
-\end{align*}
-
-The new multiplication function $f$ can itself be used to instantiate
-variables in the same scheme, resulting in the invention of the
-exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).
-
-% FIXME: CITE
-QuickSpec version 2 replaces this two-step process with a single, iterative
-algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
-Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
-As expressions are enumerated, they are simplified using these rules and
+``Automated Theorem Discovery''~\cite{roy}.
+
+Such a plethora of terminology can mask similarities and shared goals between
+these fields. Even notable historical differences, such as the emphasis of MTE
+on user-interaction and mathematical domains, in contrast to the full automation
+and more general applications targeted by ATF, are disappearing in recent
+implementations.
+
+An important historical implementation of ATF is Lenat's AM system. Unlike prior
+work, such as Meta-Dendral~\cite{buchanan:75} and those described
+in~\cite{winston}, AM aims to be a general purpose mathematical discovery
+system, designed to both construct new concepts and conjecture relationships
+between them. AM is a rule-based system which represents knowledge using a
+frame-like scheme, enlarges its knowledge base via a collection of heuristic
+rules, and controls the firing of these rules via an agenda
+mechanism. Evaluation of AM considered generality (performance in new domains)
+and how finely-tuned various aspects of the program are (the agenda, the
+interaction of the heuristics, etc). Most of this evaluation was qualitative,
+and has subsequently been criticised~\cite[chap.~13]{colton:book}. In their
+case study in methodology, Ritchie and Hanna found a large discrepancy between
+the theoretical claims made of AM and the implemented program~\cite{partridge};
+for example, AM ``invented'' natural numbers from sets, but did so using a
+heuristic specifically designed to make this connection.
+
+The prototypical implementation of MTE is the Theorema system of Buchberger and
+colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
+emphasis on user interface and output presentation. Theory exploration in the
+Theorema system involves the user formalising their definitions in a consistent,
+layered approach; such that reasoning algorithms can exploit this structure in
+subsequent proofs, calculations, etc. The potential of this strategy was
+evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
+bases algorithm~\cite{buchberger:04}.
+
+A similar ``layering'' approach is found in the IsaScheme system of
+Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which is compared against
+IsaCoSy and HipSpec is compared using precision/recall
+in~\cite{claessen2013automating}. The name comes from its embedding in the
+Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
+can be used to generate new concepts and conjectures. Variables within a scheme
+are instantiated automatically and this drives the invention process. For
+example, the concept of ``repetition'' can be encoded as a scheme, and
+instantiated with existing encodings of zero, successor and addition to produce
+a definition of multiplication. The same scheme can be instantiated with this
+new multiplication function to produce exponentiation.
+
+IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
+described in more detail in $\S$\ref{sec:existing-tools}, since these are the
+tools we chose to evaluate and compare to demonstrate our benchmarking
+methodology. QuickSpec has since evolved to version 2~\cite{smallbone2017quick},
+which replaces the distinct enumeration and testing steps with a single,
+iterative algorithm similar to that of IsaCoSy. Generated conjectures are fed
+into a Knuth-Bendix completion algorithm to form a corresponding set of rewrite
+rules. As expressions are enumerated, they are simplified using these rules and
 discarded if equal to a known expression. If not, QuickCheck tests whether the
@@
 
-%Speculate
-Speculate is a Haskell tool which operates in a similar way to QuickSpec
-version 2. In addition to equations, Speculate uses the laws of total orders and
-Boolean algebra, along with testing (via LeanCheck), to conjecture inequalities
-and conditional relations between expressions, such as (for
-$x, y \in \mathbb{Z}$):
-
-\begin{equation*}
-  \begin{aligned}
-    x + y \leq \lvert x \rvert + \lvert y \rvert \\
-    x \leq 0 \rightarrow x + \lvert x \rvert = 0
-  \end{aligned}
-\end{equation*}
-
-\subsection{Commonalities and Differences between MTE and ATF}
-
-% FIXME: Condense
-ATF and MTE share many goals and would benefit from a
-closer alignment: both fields are highly specialised and by joining
-forces (while being explicit about differences) the techniques being
-developed would have greater impact. This closer alignment is already
-beginning to take place. The main commonalities between MTE and ATF
-are that both are interested in the same aspects of mathematical
-thinking, aiming to automatically construct and evaluate elements of a
-mathematical theory, including concepts, conjectures, theorems, axioms
-and examples. Both strands contrast themselves with Automated Theorem
-Proving. The main historical differences between these two approaches seem to be
-that MTE was initiated by mathematicians as an attempt to support existing
-mathematical research, perhaps with user interaction; while ATF is a product of
-AI researchers trying to extend AI techniques, with full automation and often
-applications to non-mathematical domains.
-
-The different backgrounds of the people who named the fields
-``automated theory formation'' (Lenat) and ``mathematical theory
-exploration'' (Buchberger) perhaps reflects different metaphysical
-perspectives on invention (in which case new mathematical ideas are
-being {\em formed}, created or produced) and discovery (in which case
-abstract mathematical objects exist independently of us and are {\em
-  explored} or investigated) in mathematics.\footnote{There is a vast
-  literature on the Platonic and the constructivist view of
-  mathematics: we shall not venture down this path here (interested
-  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
-  we make the pragmatic assumption that while there may well be
-  different cognitive processes involved in invention than those
-  involved in discovery, the fields of ATF and MTE are not currently
-  concerned with this level of detail and so the philosophical
-  distinction is not relevant for our purposes.} As can be seen above,
-some of these historical differences are disappearing, and recent
-implementations are bridging the methodological gap between ATF and MTE. This is
-particularly true in terms of fully-automatic/user-interactive and the
-domain-specific/general aspects.
+QuickSpec has also inspired another MTE tool for Haskell called
+Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
+also makes use of the laws of total orders and Boolean algebra to conjecture
+\emph{in}equalities and conditional relations between expressions.
+
+Another notable MTE implementation, distinct from those based in Isabelle and
+Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
+Hypotheses, Axioms and Definitions)~\cite{roy}. The project's goal is to help
+mathematicians explore the consequences of a set of axioms or a particular
+concept definition by deducing, via the chosen logic, a set of theorems which
+they entail. Like other MTE approaches, one of its main challenges is to
+automatically evaluate interestingness: to distinguish important theorems from
+results which, although they follow from a set of axioms, are of little
+mathematical interest. Evaluation of these theorems was performed by the
+system's developer, which highlights the need for a general method of evaluation
+and comparison for these tools.
+
+Whilst there are many reasonably objective benchmarks for mathematical tasks
+such as automated theorem proving, the precision/recall analysis shown
+in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
+the only quantitative comparison of these recent tools we have found in the
+literature. Our work is essentially an extension of this approach, to a larger
+and more diverse set of examples. The suitability of the TIP theorem proving
+benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
+since its developers are also those of the IsaCoSy and QuickSpec tools we have
+tested. This goes some way to ensuring that our demonstration is a faithful
+representation of the task these tools were designed for, whilst our independent
+repurposing of the problem set reduces the risk that the benchmark is
+tailor-made for these tools; or, conversely, that the tools over-fit to these
+particular problems.
 

### 2018-04-30 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py

@@ -362,10 +362,10 @@ def plotTime(system, agg):
                                                       for s in agg['success']]),
                        x         = 'size',
                        y         = 'time',
-                       size      = 4,  # Marker size
+                       size      = 2.5,  # Marker size
                        edgecolor = 'k',
                        linewidth = 0.35,
-                       marker    = 'x',
+                       marker    = 'o',
                        palette   = ['k', 'r'],
                        hue       = 'timeHue')
     if ax.legend_: ax.legend_.remove()
@@ -566,10 +566,10 @@ def plotPrecRec(system, agg):
         newAx = sns.swarmplot(data      = newAgg,
                               x         = 'size',
                               y         = args['y'],
-                              size      = 4,  # Marker size
+                              size      = 2.5,  # Marker size
                               edgecolor = 'k',
                               linewidth = 0.35,
-                              marker    = 'x',
+                              marker    = 'o',
                               color     = 'k',
                               ax        = args['ax'])
         if newAx.legend_: newAx.legend_.remove()

### 2018-04-30 TEBenchmarkPaper/article.tex
Once-over to make sure we're consistent, etc.

@@
         Alison Pease    \and
-        Jianguo Zhang
-        % TODO: Invite Katya? Even if she only offers some help, can acknowledge
-}
+        Jianguo Zhang}
 
@@
 
-\emph{Conjecture generation} is the open-ended task of producing conjectures
-concerning a given logical theory which are somehow ``interesting'', and is
+\emph{Conjecture generation} is the open-ended task of conjecturing properties
+of a given logical theory which are somehow ``interesting'', and is
 studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
@@
 an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'' which is tedious to produce by hand, such as
-$\forall x. \Append(x, \Nil) = x$.
+knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
+be tedious to produce by hand but possible to discover automatically.
 
@@
 agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} give a survey of tools for theory
-formation and exploration, and their associated notions of
-``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
-notions are identified, which are applied to conjectured properties as follows:
+more complex statements.  Colton \etal{} surveyed tools for theory formation
+and exploration, and their associated notions of ``interestingness'' for
+concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
+which are applied to conjectured properties as follows:
 
@@
 {\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find optimisations like
-Equation~\ref{eq:mapreduce}, then utility would include whether the property
-allows rewriting, the difference in resource usage of the expressions involved,
-and how common those expressions are in real usage.
+particular task. For example, if we want to find properties which are useful for
+optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
+whether the property justifies some rewrite rule, the difference in resource
+usage of the expressions involved, and how common those expressions are in real
+usage.
 
@@
 indication that they are interesting. Two sets of definitions were used: a
-theory of natural numbers and of lists (which we show in
+theory of natural numbers; and a theory of lists (which we show in
 Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
-two datapoints containing only a few definitions and properties (e.g. the
-natural number theory contained 4 functions with 12 properties), it is difficult
+two datapoints, each containing only a few definitions and properties (4
+functions and 12 properties, in the natural number example), it is difficult
 to draw general conclusions; yet despite these limitations we believe that this
 is a viable method to performing meaningful, objective analyses and comparisons
-for such tools. It is only with larger data sets that we can judge performance
-in different, especially \emph{novel}, domains (which, after all, is the purpose
-of MTE tools). In addition, we might be concerned that such ``popular'' theories
-as the natural numbers may be unrepresentative of typical usage. For example, we
-would not expect the library functions in a typical verification problem to be
-as mathematically rich as number theory.
+for such tools. It is only with larger data sets that we can generalise the
+measured performance to different, especially \emph{novel}, domains (which,
+after all, is the purpose of MTE tools). In addition, we might be concerned that
+such ``popular'' theories as the natural numbers may be unrepresentative of
+typical usage. For example, we would not expect the library functions in a
+typical verification problem to be as mathematically rich as number theory.
 
@@
 and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration. Other notable tools, including IsaScheme, are described under
-related work in $\S$\ref{sec:related-work}.
+demonstration.
 
@@
 Haskell programming language~\cite{claessen2011quickcheck}. As well as
-generating conjectures for HipSpec (also a Haskell tool) it has also been
-applied to Isabelle via the Hipster tool~\cite{Hipster}. Like IsaCoSy, QuickSpec
-takes a signature and enumerates well-typed terms, collecting together those of
-the same type into equivalence classes, assuming them to be equal. QuickCheck is
-then used to find counterexamples to this assumption, which it does by randomly
-instantiating the variables and comparing the resulting closed expressions. Any
-equivalence class whose elements don't compare equal are split up, and the
-process is repeated with new random values.
+conjecturing properties of Haskell definitions for HipSpec, it has been applied
+to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
+QuickSpec takes a signature and enumerates well-typed terms, collecting together
+those of the same type into equivalence classes, assuming them to be
+equal. QuickCheck is then used to find counterexamples to this assumption, which
+it does by randomly instantiating the variables and comparing the resulting
+closed expressions. Any equivalence class whose elements don't compare equal are
+split up, and the process is repeated with new random values.
 
@@
 Some problem sets are more amenable for this purpose than others. The most
-suitable are those which have the following properties:
+suitable are those meeting the following criteria:
 
 \begin{itemize}
-\item For each problem, there should be a clear distinction between the theorem
-  to be proved and the definitions involved, such that the two can be easily and
-  meaningfully separated. This rules out problem sets like those of
+\item For each problem, there should be a clear distinction between the
+  statement to be proved and the definitions involved, such that the two can be
+  easily and meaningfully separated. This rules out problem sets like those of
   SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
@@
   ground truth will be treated as being ``uninteresting''. More realistically,
-  we should at least aim for each definition to have \emph{multiple} properties;
-  rather than a mixture of mostly unrelated properties.
-\item Larger problem sets are preferable, all else being equal, as they give
-  more robust statistics.
+  we should aim for each definition to have \emph{multiple} properties; rather
+  than a mixture of unrelated problems, cherry-picked from different domains.
+\item Larger problem set are preferable as they give more robust statistics, all
+  else being equal (i.e. when this does not sacrifice quality).
 \end{itemize}
 
-Once such a problem set has been chosen, we must separate the definitions
-referenced by the theorems from the theorem statements themselves. The former
-will be used as input to the tools, whilst the latter form a ground truth corpus
-of properties against which to compare the output.
+Once such a problem set has been chosen, we must separate the theorem statements
+from the definitions they reference. The latter will be used as input to the
+MTE tools, whilst the former constitute the ground truth corpus of properties
+against which to compare the output.
 
@@
 
-For this reason, duplicates should be removed, and any references in the theorem
-statements updated to use the remaining definition (e.g. chosen based on
-lexicographic order). In the above example, the \texttt{plus} function would be
-removed, and the commutativity statement updated to reference \texttt{add}
-instead.
+For this reason, duplicates should be removed, and any references to them
+updated to use the remaining definition (e.g. chosen based on lexicographic
+order). In the above example, the \texttt{plus} function would be removed, and
+the commutativity statement updated to reference \texttt{add} instead.
 
@@
   \end{tikzpicture}
-  \caption[]{High-level view of our benchmarking methodology, showing
+  \caption[]{High-level view of the benchmarking methodology described in
+    $\S$\ref{sec:proposal}, showing
     \begin{tikzpicture}
@@
 
-An important historical implementation of ATF is Lenat's AM system. Unlike prior
-work, such as Meta-Dendral~\cite{buchanan:75} and those described
-in~\cite{winston}, AM aims to be a general purpose mathematical discovery
-system, designed to both construct new concepts and conjecture relationships
-between them. AM is a rule-based system which represents knowledge using a
-frame-like scheme, enlarges its knowledge base via a collection of heuristic
-rules, and controls the firing of these rules via an agenda
-mechanism. Evaluation of AM considered generality (performance in new domains)
-and how finely-tuned various aspects of the program are (the agenda, the
-interaction of the heuristics, etc). Most of this evaluation was qualitative,
-and has subsequently been criticised~\cite[chap.~13]{colton:book}. In their
-case study in methodology, Ritchie and Hanna found a large discrepancy between
-the theoretical claims made of AM and the implemented program~\cite{partridge};
-for example, AM ``invented'' natural numbers from sets, but did so using a
-heuristic specifically designed to make this connection.
+An important historical implementation of ATF is Lenat's AM (Automated
+Mathematician) system. Unlike prior work, such as
+Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
+to be a general purpose mathematical discovery system, designed to both
+construct new concepts and conjecture relationships between them. AM is a
+rule-based system which represents knowledge using a frame-like scheme, enlarges
+its knowledge base via a collection of heuristic rules, and controls the firing
+of these rules via an agenda mechanism. Evaluation of AM considered generality
+(performance in new domains) and how finely-tuned various aspects of the program
+are (the agenda, the interaction of the heuristics, etc). Most of this
+evaluation was qualitative, and has subsequently been
+criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
+Ritchie and Hanna found a large discrepancy between the theoretical claims made
+of AM and the implemented program~\cite{partridge}; for example, AM ``invented''
+natural numbers from sets, but did so using a heuristic specifically designed to
+make this connection.
 
@@
   their software to our purposes.
-%  FIXME: EPSRC, Katya, Jianguo fundung, Alison funding
+  (FIXME: Funding details go here)
 \end{acknowledgements}

### 2018-05-01 TEBenchmarkPaper/article.tex
(Semi-)final fiddling

@@
 
-Other than $\alpha$-equivalence and equation symmetry, we specifically
-\emph{ignore} logical relationships between syntactically distinct statements,
-such as one equation being implied by another. Whilst logically sound,
-second-guessing the ground truth in this way would harm other aspects which
-influence interestingness (for example, a more general statement is more widely
-applicable, but it might also be harder to comprehend).
+We specifically \emph{ignore} other logical relationships between syntactically
+distinct statements, such as one equation being implied by another. Whilst
+logically sound, second-guessing the ground truth in this way would harm other
+aspects which influence interestingness (for example, a more general statement
+is more widely applicable, but it might also be harder to comprehend).
 
@@
 generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring.
-
-One general question we might ask is how a tool's performance scales with
-respect to the input size (the number of given definitions). This is
-straightforward to measure by varying the sample size, but we need some way to
-combine the results from samples of the same size.
+summary depends on what we are interested in measuring. One general question we
+might ask is how a tool's performance scales with respect to the input size (the
+number of given definitions). This is straightforward to measure by varying the
+sample size, but we need some way to combine the results from samples of the
+same size.
 
@@
     \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
+    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
     \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
@@
 successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
-diminishing returns. During analysis, duplicate samples (caused by our sampling
-restrictions) were found and discarded, so only 14 samples of size 1 were used
-and 30 of size 2.
+diminishing returns. During analysis, duplicate samples (caused by our
+requirement that samples have a non-empty ground truth) were found and
+discarded, so only 14 samples of size 1 were used and 30 of size 2.
 
@@
 We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the desiderata of $\S$\ref{section:prep}: each benchmark
+since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
 problem has standalone type and function definitions, making their separation
@@
 conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
-to automate some of the decisions which are normally left up to the user:
+$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
+need to automate some of the decisions which are normally left up to the user:
 
@@
 The time taken to explore samples of different sizes is shown in
-figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
-and these occurred for all sample sizes but are more common for larger samples
-(over half the runs for sample size 20 timed out). Runs which succeeded mostly
-finished within 30 seconds, generating few or no conjectures; those taking
-longer generated more conjectures, with the most being 133 conjectures from a
-sample of size 19.
-
-Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
+Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
+failures were due to timing out, and these occurred for all sample sizes but are
+more common for larger samples (over half the runs for sample sizes 13 and 20
+timed out). Runs which succeeded mostly finished within 30 seconds, generating
+few or no conjectures; those taking longer generated more conjectures, with the
+most being 133 conjectures from a sample of size 19.
+
+Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
 Since QuickSpec generates monotonically more conjectures as definitions are
@@
 large reduction in recall. The latter option could also improve the running
-time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
+time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
 signatures may be faster to explore than a single large one. Such improvements
@@
 The most striking result is how rapidly IsaCoSy's running time, shown in
-figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
+Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
 size 6 failed, with most timing out and a few aborting early due to running out
@@
 the precision/recall quality of IsaCoSy's output, shown in
-figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
+Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
 proportions that IsaCoSy is capable of discovering interesting conjectures, and
@@
 A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which is compared against
-IsaCoSy and HipSpec is compared using precision/recall
-in~\cite{claessen2013automating}. The name comes from its embedding in the
+Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which has also been
+quantitatively compared against IsaCoSy and HipSpec using precision/recall
+analysis~\cite{claessen2013automating}. The name comes from its embedding in the
 Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
@@
 described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare to demonstrate our benchmarking
-methodology. QuickSpec has since evolved to version 2~\cite{smallbone2017quick},
-which replaces the distinct enumeration and testing steps with a single,
-iterative algorithm similar to that of IsaCoSy. Generated conjectures are fed
-into a Knuth-Bendix completion algorithm to form a corresponding set of rewrite
-rules. As expressions are enumerated, they are simplified using these rules and
-discarded if equal to a known expression. If not, QuickCheck tests whether the
-new expression can be distinguished from the known expressions through random
+tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
+has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
+distinct enumeration and testing steps with a single, iterative algorithm
+similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
+completion algorithm to form a corresponding set of rewrite rules. As
+expressions are enumerated, they are simplified using these rules and discarded
+if equal to a known expression. If not, QuickCheck tests whether the new
+expression can be distinguished from the known expressions through random
 testing: those which can are added to the set of known expressions. Those which
@@
 Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. The project's goal is to help
-mathematicians explore the consequences of a set of axioms or a particular
-concept definition by deducing, via the chosen logic, a set of theorems which
-they entail. Like other MTE approaches, one of its main challenges is to
-automatically evaluate interestingness: to distinguish important theorems from
-results which, although they follow from a set of axioms, are of little
-mathematical interest. Evaluation of these theorems was performed by the
-system's developer, which highlights the need for a general method of evaluation
-and comparison for these tools.
+Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
+generate \emph{conjectures} that may later be sent to automated provers,
+MATHsAiD directly generates \emph{theorems}, by making logically valid
+inferences from a given set of axioms and definitions. Evaluation of the
+interestingness of these theorems was performed qualitatively by the system's
+developer, which highlights how these tools could benefit from the availability
+of an objective, repeatable, quantitative method of evaluation and comparison
+such as ours.
 
@@
 in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent tools we have found in the
+the only quantitative comparison of these recent MTE tools we have found in the
 literature. Our work is essentially an extension of this approach, to a larger
@@
 tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were designed for, whilst our independent
-repurposing of the problem set reduces the risk that the benchmark is
-tailor-made for these tools; or, conversely, that the tools over-fit to these
-particular problems.
+representation of the task these tools were intended to solve; our independent
+repurposing of this problem set, in a way it was not designed for, reduces the
+risk that the benchmark is tailor-made for these tools (or that the tools
+over-fit to these particular problems).
 

### 2018-05-14 TEBenchmarkPaper/article.tex
TEB fixes

@@
   tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools, finding that the former takes significantly less
-  time to produce significantly more ``interesting'' output. By providing a
+  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
+  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
+  and produces significantly more ``interesting'' output. By providing a
   standard, cross-tool evaluation technique we allow the disparate approaches
@@
 statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this (arbitrary)
-threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
-a few days we chose a timeout of 300 seconds, since that covered most of the
-successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
-diminishing returns. During analysis, duplicate samples (caused by our
-requirement that samples have a non-empty ground truth) were found and
-discarded, so only 14 samples of size 1 were used and 30 of size 2.
+so we pick 31 samples of each size in order to cross this threshold. This gave a
+total of 1240 runs. To keep the benchmarking time down to a few days we chose a
+timeout of 300 seconds, since that covered most of the successful QuickSpec and
+IsaCoSy results we saw, and longer times gave rapidly diminishing
+returns. During analysis, duplicate samples (caused by our requirement that
+samples have a non-empty ground truth) were found and discarded, so only 14
+samples of size 1 were used and 30 of size 2.
 
@@
 samples where both tools succeeded. We pooled these together from all sample
-sizes to produce the following table, and found that QuickSpec is significantly
-different from IsaCoSy with a p value of 0.0026.
-
-\begin{table}[H]
+sizes to produce table~\ref{table:recall}, and found that the recall of
+QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
+of 0.0026.
+
+\begin{table}
   \centering
@@
   \end{tabular}
+  \caption{Contingency table for ground truth properties, combined from those
+    samples where both QuickSpec and IsaCoSy finished successfully. The combined
+    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
+  \label{table:recall}
 \end{table}
@@
 
-\begin{table}[H]
+\begin{table}
   \centering
@@
   \end{tabular}
+  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
+    IsaCoSy, combined from those samples where both both finished successfully.
+    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
+  \label{table:precision}
 \end{table}
@@
 We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (arbitrary) choice of $\alpha = 0.05$; hence we do
-not consider the difference between these proportions (14\% for QuickSpec, 19\%
-for IsaCoSy) to be significant.
+of 0.111, which exceeds our (conventional) significance threshold of
+$\alpha = 0.05$; hence we do not consider the difference between these
+proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
 
@@
 both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. Based
-on these results we proposed two possible directions to improve the QuickSpec
-algorithm: a post-processing filter to remove more ``uninteresting'' conjectures
-and a pre-processing filter to ``focus'' on promising subsets of the input.
+IsaCoSy; although more \emph{undesirable} output may be generated as well. We
+found that they both fail, due to time and memory constraints, for a large
+proportion of inputs; that this gets worse as input size increases; and that
+IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
+possible directions to improve the QuickSpec algorithm: a post-processing filter
+to remove more ``uninteresting'' conjectures and a pre-processing filter to
+``focus'' on promising subsets of the input.
 
@@
   their software to our purposes.
-  (FIXME: Funding details go here)
+
+  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
 \end{acknowledgements}

### 2018-05-16 TEBenchmarkPaper/article.tex
TEB: Close many issues

@@
 \usepackage{hhline}
+\usepackage{hyperref}
 \usepackage{mathtools}
@@
 \item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems benchmark as a corpus.
+  Problems (TIP) benchmark as a corpus.
 \item Automated tooling to perform this benchmarking.
@@
 literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
+
+\begin{sloppypar}
+  Our benchmark, including the scripts used to generate it from the TIP problem
+  set, is available online at
+  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
+  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
+  \url{https://github.com/warbo/isaplanner-tip}.
+\end{sloppypar}
 
@@
 
+\begin{sloppypar}
   The precision and recall of three MTE tools are compared
@@
   IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-IsaScheme~\cite{MontanoRivas2011} and
+  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
   HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
     comparison only reports prior results for each tool, rather than reproducing
-  them. This makes the numbers less comparable, since each tool was tested with
-  slightly different definitions (e.g. whether or not the exponential and
+    them. This makes the numbers less comparable, since each tool was tested
+    with slightly different definitions (e.g. whether or not the exponential and
     maximum functions were included).} The definitions and ground truths for all
@@
   theory of natural numbers; and a theory of lists (which we show in
-Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
-two datapoints, each containing only a few definitions and properties (4
+  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
+  only two datapoints, each containing only a few definitions and properties (4
   functions and 12 properties, in the natural number example), it is difficult
-to draw general conclusions; yet despite these limitations we believe that this
-is a viable method to performing meaningful, objective analyses and comparisons
-for such tools. It is only with larger data sets that we can generalise the
-measured performance to different, especially \emph{novel}, domains (which,
-after all, is the purpose of MTE tools). In addition, we might be concerned that
-such ``popular'' theories as the natural numbers may be unrepresentative of
-typical usage. For example, we would not expect the library functions in a
-typical verification problem to be as mathematically rich as number theory.
+  to draw general conclusions; yet despite these limitations we believe that
+  this is a viable method to performing meaningful, objective analyses and
+  comparisons for such tools. It is only with larger data sets that we can
+  generalise the measured performance to different, especially \emph{novel},
+  domains (which, after all, is the purpose of MTE tools). In addition, we might
+  be concerned that such ``popular'' theories as the natural numbers may be
+  unrepresentative of typical usage. For example, we would not expect the
+  library functions in a typical verification problem to be as mathematically
+  rich as number theory.
+\end{sloppypar}
 
@@
 Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
+Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
 ``Concept Formation in Discovery Systems''~\cite{haase}, and
@@
 Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{partridge}; for example, AM ``invented''
+of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
 natural numbers from sets, but did so using a heuristic specifically designed to
@@
 A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which has also been
-quantitatively compared against IsaCoSy and HipSpec using precision/recall
-analysis~\cite{claessen2013automating}. The name comes from its embedding in the
-Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
-can be used to generate new concepts and conjectures. Variables within a scheme
-are instantiated automatically and this drives the invention process. For
-example, the concept of ``repetition'' can be encoded as a scheme, and
-instantiated with existing encodings of zero, successor and addition to produce
-a definition of multiplication. The same scheme can be instantiated with this
-new multiplication function to produce exponentiation.
+Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
+has also been quantitatively compared against IsaCoSy and HipSpec using
+precision/recall analysis~\cite{claessen2013automating}. The name comes from its
+embedding in the Isabelle proof assistant and its use of ``schemes'':
+higher-order formulae which can be used to generate new concepts and
+conjectures. Variables within a scheme are instantiated automatically and this
+drives the invention process. For example, the concept of ``repetition'' can be
+encoded as a scheme, and instantiated with existing encodings of zero, successor
+and addition to produce a definition of multiplication. The same scheme can be
+instantiated with this new multiplication function to produce exponentiation.
 

### 2018-05-16 TEBenchmarkPaper/article.tex
TEB: "Final" final, last, finished, done (honest this time!) changes (yeah right...)

@@
 chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
+with a timeout (e.g. 5 minutes). We use wall-clock time since this is most
+relevant to a user's experience, it has a straightforward interpretation and
+tracking it does not require intrusive alterations to a tool's implementation.
+The downside is that comparisons must take hardware performance into account,
+which would not be the case if time were measured by some proxy like number of
+expressions evaluated.
 
@@
   \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
+  \caption{Speedup from using QuickSpec compared to IsaCoSy, i.e. the median
+    time taken by IsaCoSy divided by that of QuickSpec. All were deemed
+    statistically significant QuickSpec was significantly faster
+    ($\alpha = 0.05$) for all sizes, except 1 and 2 due to too few samples.
     These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
+    reduce the measured time difference (e.g. sizes 13 and 20, where the median
+    for both tools timed out).}
   \label{table:speedups}

### 2018-05-17 PhDThesis/background.tex
Drag a bunch of stuff into thesis folder

@@
 
-\subsection{Conjecture Generation}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-We focus on the task of generating conjectures from some given mathematical
-theory. This theory may be, for example, a scientific model, an executable
-computer program or simply an object of mathematical curiosity. A theory
-defines (perhaps implicitly) a particular system of logic in which to work (for
-example higher-order logic), along with a finite set of \emph{definitions} (like
-those in figure~\ref{figure:list_theory}) which assign structure or meaning to
-particular terms. Theoretically, this is all of the information we need to
-derive the truth or falsity of all decidable statements involving these terms,
-for example by enumerating all valid proofs. However, due to the exponential
-search spaces involved, such brute-force enumeration is infeasible in practice;
-plus the resulting proofs would be difficult to comprehend and the processed
-statements would be overwhelming in number and mostly uninteresting in content.
-
-The impracticality of enumeration demonstrates that a statement being ``true''
-or ``provable'' is not \emph{sufficient} to warrant the (limited) attention of a
-human operator. We would also argue that it is not \emph{necessary}, since there
-are many statements considered to be interesting whose truth is either unknown,
-or which later turned out to be false. We thus focus on the automated production
-of \emph{interesting conjectures}, with the primary focus being on efficient
-generation and interest to the user.
-
-As an example, we use the theory of lists shown in
-figure~\ref{figure:list_theory}. These definitions are so widely used in
-software that they appear in the standard libraries of many programming
-languages. An example of a conjecture involving these definitions is the
-following universally quantified equation:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Equation~\ref{eq:mapreduce} states that combining small lists into a larger one
-using $\Append$ then transforming the elements with $\Map$ is the same as
-transforming the elements of the small lists then combining. This is interesting
-due to its applicability as an optimisation: if a program calls $\Map$ with an
-expensive transformation $f$, we can divide up the work across multiple machines
-in parallel (this is the basis of the ``map/reduce'' programming paradigm).
-Whether or not this equation holds depends on the choice of logical framework,
-and hence (in the case of software) on the semantics of the programming
-language: in particular this equation and optimisation are invalid for systems
-with strict evaluation order and side-effects (common in imperative programming)
-since any error/exception in the calculation of $ys$ will cause the surrounding
-calls to abort; on the left-hand-side this will abort $\Append$ then $\Map$, so
-$f$ will never be executed; on the right-hand-side there is an additional
-$\Map(f, xs)$ call which will \emph{not} be aborted, which may execute $f$ many
-times and trigger arbitrary observable effects.
-
-\section{Machine Learning}
-
-Machine learning (ML) is a sub-field of artificial intelligence (AI) which
-emphasises the use of statistical methods to find patterns and optimise
-functions, in contrast with symbolic reasoning approaches like automated theorem
-proving (commonly referred to as ``Good Old-Fashioned AI'').
-
-The machine learning umbrella spans many techniques and applications, but one
-common distinction is between \emph{supervised} and \emph{unsupervised} tasks.
-In supervised learning we allow the system to perform some behaviour, then an
-external process assesses whether it behaved well or poorly (akin to Justice
-Potter Stewart's test for obscenity: ``I know it when I see it''). The system's
-parameters are optimised to try and score highly on the test. Common supervised
-learning tasks include function approximation using input/output examples (e.g.
-classifying scanned images of numerals) where the test is the proportion of
-example inputs which got the correct output, and reinforcement learning where
-the system chooses from a set of ``actions'', receives a pair of ``observation''
-and ``reward'' (test score) in response, and must accumulate the most reward.
-
-Unsupervised learning does not have the same testable ``success'' criteria as
-supervised learning. Instead, the goal is to find patterns and form models of
-the target domain, which is more ambitious and open-ended than supervised
-learning but also much harder to define, measure and optimise. Examples of this
-are prediction and compression tasks such as auto-encoders
-(learned compressor/decompressor pairs, rewarded by how well they preserve data
-through a round-trip) and adversarial or co-evolution tasks (coupled systems
-where each is rewarded by exploiting mis-prediction in the others).
+\input{background_haskell.tex}
+\input{background_quickcheck.tex}
+\input{background_theoryexploration.tex}
+\input{background_conjecturegeneration.tex}
+\input{background_machinelearning.tex}
+\input{background_recurrentclustering.tex}

### 2018-05-17 PhDThesis/background_clustering.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-15 TransferReport/background.tex
@@
-\section{Background}
-\label{sec:background}
+Our approach to scaling up \qspec{} takes inspiration from two sources. The
+first is relevance filtering, which makes expensive algorithms used in theorem
+proving more practical by limiting the size of their inputs. We describe this
+approach in more details in \S \ref{sec:relevance}. Relevance filtering is a
+practical tool which has existing applications in software, such as the
+\emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
 
-\subsection{Haskell}
-\label{sec:haskell}
+Despite the idea's promise, we cannot simply invoke existing relevance filter
+algorithms in our theory exploration setting. The reason is that relevance
+filtering is a supervised learning method, i.e. it would require a distinguished
+expression to compare everything against. Theory exploration does not have such
+a distinguished expression; instead, we are interested in relationships between
+\emph{any} terms generated from a signature, and hence we must consider the
+relevance of \emph{all terms} to \emph{all other terms}.
 
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
+A natural fit for this task is \emph{clustering}, which attempts to group
+similar inputs together in an unsupervised way. Based on their success in
+discovering relationships and patterns between expressions in Coq and ACL2 (in
+the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
+methods can fulfil the role of relevance filters for theory exploration:
+intelligently breaking up large signatures into smaller ones more amenable to
+brute force enumeration, such that related expressions are explored together.
 
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \CVar\ id                          \\
-                         |\ & \CLit\ literal                     \\
-                         |\ & \CApp\ expr\ expr                  \\
-                         |\ & \CLam\ \mathcal{L}\ expr           \\
-                         |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                         |\ & \CType                             \\
-      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                         |\ & \CGlobal\      \mathcal{G}         \\
-                         |\ & \CConstructor\ \mathcal{D}         \\
-      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                         |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                         |\ & \CLitAlt\ literal                  \\
-                         |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                         |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-    \end{split}
-  \end{equation*}
-  Where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers \\
-    $\mathcal{D}$ & constructor identifiers
-  \end{tabular}
+\subsection{K-Means}
 
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
+We use the Weka tool to perform k-means clustering
+\cite{Holmes.Donkin.Witten:1994}, since we are more concerned with the
+application of feature extraction to Haskell and its use in theory exploration,
+rather than precise tuning of learning algorithms. Since k-means is a standard
+method, there are many other implementations available. More interestingly,
+there are many other clustering algorithms we could use, such as
+\emph{expectation maximisation} \footnote{In fact, k-means is very similar to
+  expectation-maximisation, as it alternates between an \emph{expectation step}
+  (finding the mean value of each cluster) and a \emph{maximisation step}
+  (assigning points to the cluster they're most similar to; or alternatively,
+  \emph{minimising} the distance of each point to the centre of its cluster, as
+  per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have
+shown little difference in their results; in effect, the quality of our features
+is the bottleneck to learning, so there is no reason to avoid a fast algorithm
+like k-means.
 
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
-
-Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most \emph{functional} programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
-Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-
-The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major restriction is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
-For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
-We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
-
-\begin{figure}
-  \begin{haskell}
-    data Nat = Z
-             | S Nat
-
-    plus :: Nat -> Nat -> Nat
-    plus    Z  y = y
-    plus (S x) y = S (plus x y)
-
-    mult :: Nat -> Nat -> Nat
-    mult    Z  y = Z
-    mult (S x) y = plus y (mult x y)
-
-    odd :: Nat -> Bool
-    odd    Z  = False
-    odd (S n) = even n
-
-    even :: Nat -> Bool
-    even    Z  = True
-    even (S n) = odd n
-  \end{haskell}
-  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
-\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
-  \label{fig:haskellexample}
-\end{figure}
-
-\begin{figure}
-  \begin{subfigure}[plus]{\textwidth}
-    \begin{small}
-      \underline{\texttt{plus}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
-                                               (App (App (Var (Global "plus"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[mult]{\textwidth}
-    \begin{small}
-      \underline{\texttt{mult}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
-                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                    (Var (Local  "y")))
-                                               (App (App (Var (Global "mult"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local  "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[odd]{\textwidth}
-    \begin{small}
-      \underline{\texttt{odd}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "False")))
-              (Alt (DataAlt "S") (App (Var (Global "even"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[even]{\textwidth}
-    \begin{small}
-      \underline{\texttt{even}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "True")))
-              (Alt (DataAlt "S") (App (Var (Global "odd"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}.
-Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
-  \label{fig:coreexample}
-\end{figure}
-
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
-The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
-It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
-
-Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}.
-GHC can represent these, but in this work we only consider reducible expressions (i.e.
-value-level bindings of the form \mbox{\hs{f a b ...
-= ...}}).
-
-\subsection{QuickCheck}
-\label{sec:quickcheck}
-
-Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g.
-life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = n * factorial (n-1)
-
-fact_base      = factorial 0 == factorial 1
-fact_increases = factorial 3 <= factorial 4
-fact_div       = factorial 4 == factorial 5 `div` 5
-\end{lstlisting}
-
-The intent of the function is to map an input $n$ to an output $n!$.
-The tests check a few properties of the implementation, including the base case, that the function is monotonically increasing, and a relationship between adjacent outputs.
-However, these tests will \emph{not} expose a serious problem with the implementation: it diverges on half of its possible inputs!
-
-All of Haskell's built-in numeric types allow negative numbers, which this implementation doesn't take into account.
-Whilst this is a rather trivial example, it highlights a common problem: unit tests are insufficient to expose incorrect assumptions.
-In this case, our assumption that numbers are positive has caused a bug in the implementation \emph{and} limited the tests we've written.
-
-If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = let nPos = abs n
-               in nPos * factorial (nPos - 1)
-
-fact_neg = factorial 1 == factorial (-1)
-\end{lstlisting}
-
-However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t.
-Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
-In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge.
-Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
-
-Haskell offers a partial solution to this problem in the form of \emph{property checking}.
-Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
-Here is how we might restate our unit tests as properties:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-fact_base        = factorial 0 == factorial 1
-fact_increases n = factorial n <= factorial (n+1)
-fact_div       n = factorial n == factorial (n+1) `div` (n+1)
-fact_neg       n = factorial n == factorial (-n)
-\end{lstlisting}
-
-The free variables (all called \hs{n} in this case) are abstracted as function parameters; these parameters are implicitly \emph{universally quantified}, i.e.
-we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
-Notice that unit tests like \hs{fact\_base} are valid properties; they just assert rather weak statements.
-
-To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
-For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
-The default stopping criterion for \qcheck{} (for each test) is when a single generated test fails, or when 100 generated tests pass.
-
-The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:
-
-\begin{itemize}
-  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type).
-This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
-
-  \item The generators used by \qcheck{} depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
-This will expose the problem with negative numbers, which we weren't expecting.
-\end{itemize}
-
-Property checking is certainly an improvement over unit testing, but the problem of tests being biased towards expected cases remains, since we are manually specifying the properties to be checked.
-
-We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}.
-These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.
-
-\subsection{Theory Exploration}
-\label{sec:theoryexploration}
-
-In this work we consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
-
-We focus on \qspec{} \citep{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
-These conjectures are arrived at through the following stages:
-
-\iffalse TODO: Make this more formal?
- V \in Var
- F \in Fun
- T \in Term
- T ::= V | F | T1 T2
-
- Term ::= VAR | Const | Fun (Term)
-or
- Term t ::= x | f | t t'
-\fi
-
-\begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
-To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
-  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item Each variable is instantiated to a particular value, generated randomly by \qcheck{}.
-  \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
-If so, the class is split up to separate such distinguishable members.
-  \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
-  \item A set of equations are then conjectured, relating each class's members.
-\end{enumerate}
-
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
-
-As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
-
-\begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
-\end{align*}
-
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
-
-\begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
-\end{align*}
-
-Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments.
-In addition, Haskell curries functions, so these binary functions will be treated as unary functions which return unary functions.
-This is required as the construction of $terms$ applies functions to one argument at a time.
-
-\begin{figure}
-  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
-  \begin{haskell}
-                  plus a b = plus b a
-                  plus a Z = a
-         plus a (plus b c) = plus b (plus a c)
-                  mult a b = mult b a
-                  mult a Z = Z
-         mult a (mult b c) = mult b (mult a c)
-              plus a (S b) = S (plus a b)
-              mult a (S b) = plus a (mult a b)
-         mult a (plus b b) = mult b (plus a a)
-                 odd (S a) = even a
-            odd (plus a a) = odd Z
-           odd (times a a) = odd a
-                even (S a) = odd a
-           even (plus a a) = even Z
-          even (times a a) = even a
-plus (mult a b) (mult a c) = mult a (plus b c)
-  \end{haskell}
-  \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
-  \label{fig:qspecresult}
-\end{figure}
-
-These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
-
-Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-Here we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
-
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-
-\subsection{Clustering}
-\label{sec:clustering}
-
-Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{sec:relevance}.
-Relevance filtering is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
-
-Despite the idea's promise, we cannot simply invoke existing relevance filter algorithms in our theory exploration setting.
-The reason is that relevance filtering is a supervised learning method, i.e.
-it would require a distinguished expression to compare everything against.
-Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
-
-A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
-
-\subsubsection{Feature Extraction}
-\label{sec:featureextraction}
-
-Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
-This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms.
-By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
-
-We use a standard representation of features as a \emph{feature vector} of numbers $\vect{x} = (x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$ (we use \textbf{bold face} to represent vectors, including feature vectors).
-\footnote{In fact, practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
-
-\begin{itemize}
-  \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $d$.
-Many machine learning algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
-  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
-This enables many powerful machine learning algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
-  \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
-This reduces the input size of the machine learning problem, improving efficiency.
-\end{itemize}
+In any case there are many conservative improvements to the standard k-means
+algorithm, which could be applied to our setup. For example, a more efficient
+approach like \emph{yinyang k-means} \cite{conf/icml/DingZSMM15} could make
+larger input sizes more practical to cluster, especially since recurrent
+clustering invokes k-means many times. The \emph{k-means++} approach
+\cite{arthur2007k, bahmani2012scalable} can be used to more carefully select
+the ``seed'' values for the first timestep, and the \emph{x-means} algorithm
+\cite{pelleg2000x} is able to estimate how many clusters to use (our
+\emph{final} clusters should be tuned to maximise the performance of the
+subsequent theory exploration step, but x-means could still be useful in the
+recurrent clustering steps).
 
@@
 
-Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:
+Clustering is an unsupervised machine learning task for grouping $n$ data points
+using a similarity metric. There are many variations on this theme, but in our
+case we make the following choices:
 
 \begin{itemize}
-  \item For simplicity, we use the ``rule of thumb'' given in \citep[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
+\item For simplicity, we use the ``rule of thumb'' given in
+  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
+  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
 \item Data points will be $d$-dimensional feature vectors, as defined above.
 \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \citep{lloyd1982least}.
+\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
+  \cite{lloyd1982least}.
 \end{itemize}
 
-This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \citep{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
+This is a standard setup, supported by off-the-shelf tools; in particular we use
+the implementation provided by Weka \cite{Holmes.Donkin.Witten:1994}, due to
+its use by ML4PG, which makes our results more easily comparable.
 
-Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$.
-As the name suggests, k-means uses the mean value of each cluster, which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
+Since k-means is iterative, we will use function notation to denote time steps,
+so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
+to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
+which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
 
@@
 
-Before k-means starts, we must choose \emph{seed} values for $\vect{m}^i(0)$.
-Many methods have been proposed for choosing these values \citep{arthur2007k}.
-For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
+Before k-means starts, we must choose \emph{seed} values for
+$\vect{m}^i(0)$. Many methods have been proposed for choosing these values
+\cite{arthur2007k}. For simplicity, we will choose values randomly from our
+data set $S$; this is known as the Forgy method.
 
-The elements of each cluster $C^i(t)$ are those data points closest to the mean value at the previous time step:
+The elements of each cluster $C^i(t)$ are those data points closest to the mean
+value at the previous time step:
 
@@
 
-As $t$ increases, the clusters $C^i$ move from their initial location around the ``seeds'', to converge on a local minimum of the ``within-cluster sum of squared error'' objective:
+As $t$ increases, the clusters $C^i$ move from their initial location around the
+``seeds'', to converge on a local minimum of the ``within-cluster sum of squared
+error'' objective:
 

### 2018-05-17 PhDThesis/background_conjecturegeneration.tex
Drag a bunch of stuff into thesis folder

% For example theory of lists
\newcommand{\function}{\rightarrow}

\newcommand{\Zero}{\text{Z}}
\newcommand{\Succ}{\text{S}}
\newcommand{\plus}{\text{plus}}
\newcommand{\mult}{\text{times}}

\newcommand{\List}{\text{List}}
\newcommand{\ListA}{\text{List} \  a}
\newcommand{\Nil}{\text{Nil}}
\newcommand{\Cons}{\text{Cons}}
\newcommand{\Head}{\text{head}}
\newcommand{\Tail}{\text{tail}}
\newcommand{\Append}{\text{append}}
\newcommand{\Reverse}{\text{reverse}}
\newcommand{\Length}{\text{length}}
\newcommand{\Map}{\text{map}}
\newcommand{\Foldl}{\text{foldl}}
\newcommand{\Foldr}{\text{foldr}}

\subsection{Conjecture Generation}

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
  \label{figure:list_theory}
\end{figure}

We focus on the task of generating conjectures from some given mathematical
theory. This theory may be, for example, a scientific model, an executable
computer program or simply an object of mathematical curiosity. A theory
defines (perhaps implicitly) a particular system of logic in which to work (for
example higher-order logic), along with a finite set of \emph{definitions} (like
those in figure~\ref{figure:list_theory}) which assign structure or meaning to
particular terms. Theoretically, this is all of the information we need to
derive the truth or falsity of all decidable statements involving these terms,
for example by enumerating all valid proofs. However, due to the exponential
search spaces involved, such brute-force enumeration is infeasible in practice;
plus the resulting proofs would be difficult to comprehend and the processed
statements would be overwhelming in number and mostly uninteresting in content.

The impracticality of enumeration demonstrates that a statement being ``true''
or ``provable'' is not \emph{sufficient} to warrant the (limited) attention of a
human operator. We would also argue that it is not \emph{necessary}, since there
are many statements considered to be interesting whose truth is either unknown,
or which later turned out to be false. We thus focus on the automated production
of \emph{interesting conjectures}, with the primary focus being on efficient
generation and interest to the user.

As an example, we use the theory of lists shown in
figure~\ref{figure:list_theory}. These definitions are so widely used in
software that they appear in the standard libraries of many programming
languages. An example of a conjecture involving these definitions is the
following universally quantified equation:

\begin{equation} \label{eq:mapreduce}
  \forall f. \forall xs. \forall ys.
    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
\end{equation}

Equation~\ref{eq:mapreduce} states that combining small lists into a larger one
using $\Append$ then transforming the elements with $\Map$ is the same as
transforming the elements of the small lists then combining. This is interesting
due to its applicability as an optimisation: if a program calls $\Map$ with an
expensive transformation $f$, we can divide up the work across multiple machines
in parallel (this is the basis of the ``map/reduce'' programming paradigm).
Whether or not this equation holds depends on the choice of logical framework,
and hence (in the case of software) on the semantics of the programming
language: in particular this equation and optimisation are invalid for systems
with strict evaluation order and side-effects (common in imperative programming)
since any error/exception in the calculation of $ys$ will cause the surrounding
calls to abort; on the left-hand-side this will abort $\Append$ then $\Map$, so
$f$ will never be executed; on the right-hand-side there is an additional
$\Map(f, xs)$ call which will \emph{not} be aborted, which may execute $f$ many
times and trigger arbitrary observable effects.

### 2018-05-17 PhDThesis/background_featureextraction.tex
Drag a bunch of stuff into thesis folder

One major difficulty when applying statistical machine learning algorithms to
\emph{languages}, such as Haskell, is the appearance of recursive
structures. This can lead to nested expressions of arbitrary depth, which are
difficult to compare in numerical ways. One solution, as described in \S
\ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our
method is not the only possible way to encode recursive structures as fixed-size
features.

The simplest way to encode such inputs is to simply choose a desired size, then
pad anything smaller and truncate anything larger. We use this to make our
matrices a uniform size, borrowing the idea from ML4PG. Care must be taken to
ensure that we are not discarding too much information, that we are not
producing features with too many dimensions to be practical, and that there is a
uniform ``meaning'' to each feature across different feature vectors. In our
case, we avoid many of these problems by transforming the recursive structure of
expressions into matrices first; this gives each feature a stable meaning such
as ``the $i$th token from the left at the $j$th level of nesting''.

Truncation works best when the input data is arranged with the most significant
data first (in a sense, it is ``big-endian''). This is the case for Haskell
expressions, since the higher levels of the syntax tree are the most
semantically significant; for example, the lower levels may never even be
evaluated due to laziness. This allows us to truncate more aggressively than if
the leaves were most significant.

By modelling our inputs as points in high-dimensional spaces, we can consider
feature extraction as projection into a lower-dimensional space (known as
\emph{dimension reduction}). Truncation is a trivial dimension reduction
technique; more sophisticated projection functions consider the
\emph{distribution} of the input points, and project with the hyperplane which
preserves as much of the variance as possible (or, equivalently, reduces the
\emph{mutual information} between the points).

Techniques such as \emph{principle component analysis} (PCA) can be used to find
these hyperplanes, but unfortunately require their inputs to already have a
fixed, integer number of dimensions. In the case of our recursive expressions
(which we may consider to have \emph{fractal} dimension), we would need another
pre-processing stage to satisfy this requirement.

There are machine learning algorithms which can handle variable-size input, but
these are often \emph{supervised} algorithms which require an
externally-provided error function to minimise. Error functions can be given for
clustering, for example k-means implicitly minimises the function given in
equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred
for efficiency as they are more direct.

One example of learning from variable-size input is to use \emph{recurrent
  neural networks} (RNNs). These contain cyclic connections between neurons,
unlike the traditional acyclic ``feed-forward'' NNs, allowing state to persist
between observations. In this way, each data point can be divided into a
sequence of arbitrary length, for example an s-expression, and fed into an RNN
one token at a time for processing.

Unfortunately RNNs are difficult to train. The standard way to train NNs is the
back-propagation algorithm; when this is extended to handle cycles we get the
\emph{backpropagation through time} algorithm
\cite{werbos1990backpropagation}. However, this suffers a problem known as the
\emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change
exponentially as they propagate back through the cycles, which prevents
effective learning of correlations across a sequence, undermining the main
advantage of RNNs. The vanishing gradient problem is the subject of current
research, with countermeasures including \emph{neuroevolution} (using
evolutionary computation techniques rather than back-propagation) and \emph{long
  short-term memory} (LSTM; introducing special nodes to ``store'' state, rather
than having them loop around a cycle \cite{hochreiter1997long}).

Using sequences to represent recursive structures is also problematic: if we
want our learning algorithm to exploit structure (such as the depth of a token),
it will have to discover how to parse the sequences for itself, which seems
wasteful. The \emph{back-propagation through structure} approach
\cite{goller1996learning} is a more direct solution to this problem, using a
feed-forward NN to learn recursive distributed representations
\cite{pollack1990recursive} which correspond to the recursive structure of the
inputs. Such distributed representations can also be used for sequences, which
we can use to encode sub-trees when the branching factor of nodes is not uniform
\cite{kwasny1995tail}. More recent work has investigated storing recursive
structures inside LSTM cells \cite{zhu2015long}.

A simpler alternative for generating recursive distributed representations is to
use circular convolution \cite{conf/ijcai/Plate91}. Although promising results
are shown for its use in \emph{distributed tree kernels}
\cite{zanzotto2012distributed}, our preliminary experiments in applying
circular convolution to functional programming expressions found most of the
information to be lost in the process; presumably as the expressions are too
small.

\emph{Kernel methods} have also been applied to structured information, for
example in \cite{Gartner2003} the input data (including sequences, trees and
graphs) are represented using \emph{generative models}, such as hidden Markov
models, of a fixed size suitable for learning. Many more applications of kernel
methods to structured domains are given in \cite{bakir2007predicting}, which
could be used to learn more subtle relations between expressions than recurrent
clustering alone.

%% FIXME: MERGE THE FOLLOWING INTO THE ABOVE

\subsubsection{Feature Extraction}
\label{sec:featureextraction}

Before describing clustering in detail, we must introduce the idea of
\emph{feature extraction}. This is the conversion of ``raw'' input data, such as
audio, images or (in our case) Core expressions, into a form more suited to
machine learning algorithms. By pre-processing our data in this way, we can
re-use the same ``off-the-shelf'' machine learning algorithms in a variety of
domains.

We use a standard representation of features as a \emph{feature vector} of
numbers $\vect{x} = (x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$ (we use
\textbf{bold face} to represent vectors, including feature
vectors). \footnote{In fact, practical implementations will use an approximate
  format such as IEEE 754 floating point numbers.} For learning purposes this
has some important advantages over raw expressions:

\begin{itemize}
\item All of our feature vectors will be the same size, i.e. they will all have
  length (or \emph{dimension}) $d$. Many machine learning algorithms only work
  with inputs of a uniform size; feature extraction allows us to use these
  algorithms in domains where the size of each input is not known, may vary or
  may even be unbounded. For example, element-wise comparison of feature vectors
  is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions
  this is not so straightforward, as their nesting may give rise to very
  different shapes.
\item Unlike our expressions, which are discrete, we can continuously transform
  one feature vector into another. This enables many powerful machine learning
  algorithms to be used, such as those based on \emph{gradient descent} or, in
  our case, arithmetic means.
\item Feature vectors can be chosen to represent the relevant information in a
  more compressed form than the raw data; for example, we might replace verbose,
  descriptive identifiers with sequential numbers. This reduces the input size
  of the machine learning problem, improving efficiency.
\end{itemize}

### 2018-05-17 PhDThesis/background_haskell.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-09 TransferReport/related.tex
@@
-\section{Related Work}
-\label{sec:related}
+\subsection{Haskell}
+\label{sec:haskell}
 
-\subsection{Haskell}
-\label{sec:haskelldesc}
+\newcommand{\CVar}{\texttt{Var}}
+\newcommand{\CLit}{\texttt{Lit}}
+\newcommand{\CApp}{\texttt{App}}
+\newcommand{\CLam}{\texttt{Lam}}
+\newcommand{\CLet}{\texttt{Let}}
+\newcommand{\CCase}{\texttt{Case}}
+\newcommand{\CType}{\texttt{Type}}
+\newcommand{\CLocal}{\texttt{Local}}
+\newcommand{\CGlobal}{\texttt{Global}}
+\newcommand{\CConstructor}{\texttt{Constructor}}
+\newcommand{\CLitNum}{\texttt{LitNum}}
+\newcommand{\CLitStr}{\texttt{LitStr}}
+\newcommand{\CAlt}{\texttt{Alt}}
+\newcommand{\CDataAlt}{\texttt{DataAlt}}
+\newcommand{\CLitAlt}{\texttt{LitAlt}}
+\newcommand{\CDefault}{\texttt{Default}}
+\newcommand{\CNonRec}{\texttt{NonRec}}
+\newcommand{\CRec}{\texttt{Rec}}
+\newcommand{\CBind}{\texttt{Bind}}
 
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
-Here we discuss the relevant language features from a high-level, which motivated our choice:
+\begin{figure}
+  \begin{equation*}
+    \begin{split}
+      expr\    \rightarrow\ & \CVar\ id                          \\
+                         |\ & \CLit\ literal                     \\
+                         |\ & \CApp\ expr\ expr                  \\
+                         |\ & \CLam\ \mathcal{L}\ expr           \\
+                         |\ & \CLet\ bind\ expr                  \\
+                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
+                         |\ & \CType                             \\
+      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
+                         |\ & \CGlobal\      \mathcal{G}         \\
+                         |\ & \CConstructor\ \mathcal{D}         \\
+      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
+                         |\ & \CLitStr\ \mathcal{S}              \\
+      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
+      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
+                         |\ & \CLitAlt\ literal                  \\
+                         |\ & \CDefault                          \\
+      bind\    \rightarrow\ & \CNonRec\ binder                   \\
+                         |\ & \CRec\ [binder]                    \\
+      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
+    \end{split}
+  \end{equation*}
+  Where:
+  \begin{tabular}[t]{l @{ $=$ } l}
+    $\mathcal{S}$ & string literals    \\
+    $\mathcal{N}$ & numeric literals   \\
+    $\mathcal{L}$ & local identifiers  \\
+    $\mathcal{G}$ & global identifiers \\
+    $\mathcal{D}$ & constructor identifiers
+  \end{tabular}
+
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
+  \label{fig:coresyntax}
+\end{figure}
+
+We decided to focus on theory exploration in the Haskell programming language as
+it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and
+\hspec{} \cite{claessen2013automating}). This is evident from the fact that the
+state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster}
+\cite{Hipster} system, is actually implemented by translating to Haskell and
+invoking \hspec{}.
+
+Haskell is well-suited to programming language research; indeed, this was a goal
+of the language's creators \cite{marlow2010haskell}. Like most
+\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
+with extra features such as a strong type system and ``syntactic sugar'' to
+improve readability. For simplicity, we will focus on an intermediate
+representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather
+than the relatively large and complex syntax of Haskell proper. Core is based on
+\fc{}, but for our machine learning purposes we are mostly interested in its
+syntax; for a more thorough treatment of \fc{} and its use in GHC, see
+\cite[Appendix C]{sulzmann2007system}.
+
+The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax};
+compared to the full language \footnote{As of GHC version 7.10.2, the latest at
+  the time of writing.} our major restriction is to ignore type hints (such as
+explicit casts, and differences between types/kinds/coercions). For brevity, we
+also omit several other forms of literal (machine words of various sizes,
+individual characters, etc.), as their treatment is similar to those of strings
+and numerals. We will use quoted strings to denote names and literals,
+e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"},
+\texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can
+be compared for equality.
+
+\begin{figure}
+  \begin{haskell}
+    data Nat = Z
+             | S Nat
+
+    plus :: Nat -> Nat -> Nat
+    plus    Z  y = y
+    plus (S x) y = S (plus x y)
+
+    mult :: Nat -> Nat -> Nat
+    mult    Z  y = Z
+    mult (S x) y = plus y (mult x y)
+
+    odd :: Nat -> Bool
+    odd    Z  = False
+    odd (S n) = even n
+
+    even :: Nat -> Bool
+    even    Z  = True
+    even (S n) = odd n
+  \end{haskell}
+  \caption{A Haskell datatype for Peano numerals with some simple arithmetic
+    functions, including mutually-recursive definitions for \hs{odd} and
+    \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
+    regarded as \hs{data Bool \equal{} True | False}.}
+  \label{fig:haskellexample}
+\end{figure}
+
+\begin{figure}
+  \begin{subfigure}[plus]{\textwidth}
+    \begin{small}
+      \underline{\texttt{plus}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       (Alt (DataAlt "Z") (Var (Local "y")))
+                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
+                                               (App (App (Var (Global "plus"))
+                                                         (Var (Local  "x")))
+                                                    (Var (Local "y"))))
+                                          "x")))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[mult]{\textwidth}
+    \begin{small}
+      \underline{\texttt{mult}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
+                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
+                                                    (Var (Local  "y")))
+                                               (App (App (Var (Global "mult"))
+                                                         (Var (Local  "x")))
+                                                    (Var (Local  "y"))))
+                                          "x")))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[odd]{\textwidth}
+    \begin{small}
+      \underline{\texttt{odd}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              (Alt (DataAlt "Z") (Var (Constructor "False")))
+              (Alt (DataAlt "S") (App (Var (Global "even"))
+                                      (Var (Local  "n")))
+                                 "n"))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[even]{\textwidth}
+    \begin{small}
+      \underline{\texttt{even}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              (Alt (DataAlt "Z") (Var (Constructor "True")))
+              (Alt (DataAlt "S") (App (Var (Global "odd"))
+                                      (Var (Local  "n")))
+                                 "n"))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the
+    Core syntax of Figure \ref{fig:coresyntax}. Notice the introduction of
+    explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case}
+    to represent piecewise definitions. Fresh variables are chosen arbitrarily
+    as \texttt{"a"}, \texttt{"b"}, etc.}
+  \label{fig:coreexample}
+\end{figure}
+
+Figure \ref{fig:haskellexample} shows some simple Haskell function definitions,
+along with a custom datatype for Peano numerals. The translation to our Core
+syntax is routine, and shown in Figure \ref{fig:coreexample}. Although the Core
+is more verbose, we can see that similar structure in the Haskell definitions
+gives rise to similar structure in the Core; for example, the definitions of
+\hs{odd} and \hs{even} are identical in both languages, except for the
+particular identifiers used. It is this close correspondence which allows us to
+analyse Core expressions in place of their more complicated Haskell source.
+
+Note that we exclude representations for type-level entities, including datatype
+definitions like that of \hs{Nat}. GHC can represent these, but in this work we
+only consider reducible expressions (i.e. value-level bindings of the form
+\mbox{\hs{f a b ... = ...}}).
+
+%% FIXME: MERGE THE FOLLOWING WITH THE ABOVE
+
+Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little
+explanation was given for why we chose this language rather than, for example
+Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more
+widely used language like Java. Here we discuss the relevant language features
+from a high-level, which motivated our choice:
 
@@
 
-\item{Functional}: All control flow in Haskell is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
+\item{Functional}: All control flow in Haskell is performed by function
+  abstraction and application, which we can reason about using standard rules of
+  inference such as \emph{modus ponens}.
 
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
+\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation
+  of expressions; hence our reasoning can safely ignore complicated external and
+  non-local interactions.
 
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
+\item{Statically Typed}: Expression are constrained by \emph{types}, which can
+  be used to eliminate unwanted combinations of values, and hence reduce search
+  spaces; \emph{static} types can be deduced syntactically, without having to
+  execute the code.
 
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
+\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an
+  expression (i.e. performing function calls) without diverging, then a
+  non-strict evaluation strategy will not diverge when evaluating that
+  expression. This is rather technical, but in simple terms it allows us to
+  reason effectively about a Turing-complete language, where evaluation may not
+  terminate. For example, when reasoning about \emph{pairs} of values \hs{(x,
+    y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an
+  ``obvious'' rule such as
+  $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$. Haskell's
+  non-strict semantics makes this equation valid; whilst it would \emph{not} be
+  valid in the strict setting common to most other languages, where the
+  expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter
+  the semantics, if \hs{x} doesn't diverge).
 
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
+\item{Algebraic Data Types}: These provide a rich grammar for building up
+  user-defined data representations, and an inverse mechanism to inspect these
+  data by \emph{pattern-matching}. For our purposes, the useful consequences of
+  ADTs and pattern-matching include their amenability for inductive proofs and
+  the fact they are \emph{closed}; i.e. an ADT's declaration specifies all of
+  the normal forms for that type. This makes exhaustive case analysis trivial,
+  which would be impossible for \emph{open} types (for example, consider classes
+  in an object oriented language, where new subclasses can be introduced at any
+  time).
 
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
+\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over
+  \emph{type-level} objects; provided those objects are never inspected. This
+  has the \emph{practical} benefit of enabling \emph{polymorphism}: for example,
+  we can write a polymorphic identity function \hs{id :: forall t. t ->
+    t}. \footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a
+      -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually,
+  this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of
+  type \hs{t}; yet only the latter is available in the function body,
+  e.g. \hs{id x = x}. This inability to inspect type-level arguments gives us
+  the \emph{theoretical} benefit of being able to characterise the behaviour of
+  polymorphic functions from their type alone, a technique known as
+  \emph{theorems for free} \cite{wadler1989theorems}.
 
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
+\item{Type classes}: Along with their various extensions, type classes are
+  interfaces which specify a set of operations over a type (or other type-level
+  object, such as a \emph{type constructor}). Many type classes also specify a
+  set of \emph{laws} which their operations should obey but, lacking a simple
+  mechanism to enforce this, laws are usually considered as documentation. As a
+  simple example, we can define a type class \hs{Semigroup} with the following
+  operation and associativity law:
 
@@
 
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
+The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which
+restricts the possible types \hs{t} to only those which implement
+\hs{Semigroup}. \footnote{Alternatively, we can consider \hs{Semigroup t} as the
+  type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case
+  \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take
+  \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t}
+  and two values of type \hs{t}. As with parameteric polymorphism, this extra
+  \hs{Semigroup t} parameter is not available at the value level. Even if it
+  were, we could not alter our behaviour by inspecting it, since Haskell only
+  allows types to implement each type class in at most one way, so there would
+  be no information to branch on.} There are many \emph{instances} of
+\hs{Semigroup} (types which may be substituted for \hs{t}), e.g. \hs{Integer}
+with \hs{op} performing addition. Many more examples can be found in the
+\emph{typeclassopedia} \cite{yorgey2009typeclassopedia}. This ability to
+constrain types, and the existence of laws, helps us reason about code
+generically, rather than repeating the same arguments for each particular pair
+of \hs{t} and \hs{op}.
 
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
+\item{Equational}: Haskell uses equations at the value level, for definitions;
+  at the type level, for coercions; at the documentation level, for typeclass
+  laws; and at the compiler level, for ad-hoc rewrite rules. This provides us
+  with many \emph{sources} of equations, as well as many possible \emph{uses}
+  for any equations we might discover. Along with their support in existing
+  tools such as SMT solvers, this makes equational conjectures a natural target
+  for theory exploration.
 
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
+\item{Modularity}: Haskell has a module system, where each module may specify an
+  \emph{export list} containing the names which should be made available for
+  other modules to import. When such a list is given, any expressions \emph{not}
+  on the list are considered \emph{private} to that module, and are hence
+  inaccessible from elsewhere. This mechanism allows modules to provide more
+  guarantees than are available just in their types. For example, a module may
+  represent email addresses in the following way:
 
@@
 
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
+The \hs{Email} type guarantees that its elements have both a user part and a
+host part (modulo divergence), but it does not provide any guarantees about
+those parts. We also define the \hs{at} function, a so-called ``smart
+constructor'', which has the additional guarantee that the \hs{Email}s it
+returns contain non-empty \hs{String}s. By ommitting the \hs{E} constructor from
+the export list on the first line \footnote{The syntax \hs{Email()} means we're
+  exporting the \hs{Email} type, but not any of its constructors.}, the only way
+\emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces
+the non-empty guarantee to hold globally.
 
@@
 
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-This puts Haskell in a rather special position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
+Together, these features make Haskell code highly structured, amenable to
+logical analysis and subject to many algebraic laws. However, as mentioned with
+regards to type classes, Haskell itself is incapable of expressing or enforcing
+these laws (at least, without difficulty \cite{lindley2014hasochism}). This
+reduces the incentive to manually discover, state and prove theorems about
+Haskell code, e.g. in the style of interactive theorem proving, as these results
+may be invalidated by seemingly innocuous code changes. This puts Haskell in a
+rather special position with regards to the discovery of interesting theorems;
+namely that many discoveries may be available with very little work, simply
+because the code's authors are focused on \emph{software} development rather
+than \emph{proof} development. The same cannot be said, for example, of ITP
+systems; although our reasoning capabilities may be stronger in an ITP setting,
+much of the ``low hanging fruit'' will have already been found through the
+user's dedicated efforts, and hence theory exploration would be unlikely to
+discover unexpected properties.
 
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
+Other empirical advantages to studying Haskell, compared to other programming
+languages or theorem proving systems, include:
 
 \begin{itemize}
+\item The large amount of Haskell code which is freely available online, e.g. in
+  repositories like \href{http://hackage.haskell.org}{Hackage}, with which we
+  can experiment.
 
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
+\item The existence of theory exploration systems such as \hspec{}, and related
+  tools which we may be able to re-use, including conjecture generators like
+  \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and
+  \textsc{SmartCheck}; theorem provers like \textsc{Hip}
+  \cite{rosen2012proving}; and other related testing and term-generating
+  systems like \textsc{MuCheck} \cite{le2014mucheck}, \textsc{MagicHaskeller}
+  \cite{katayama2011magichaskeller} and \textsc{Djinn}
+  \cite{augustsson2005djinn}.
 
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
+\item The remarkable amount of infrastructure which exists for working with
+  Haskell code, including package managers, compilers, interpreters, parsers,
+  static analysers, etc.
 \end{itemize}
-
-\subsection{Relevance Filtering}
-\label{sec:relevance}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering} (or, \emph{premise selection}, when viewed as the \emph{addition} of relevant information to an initially-empty problem).
-This is the core idea behind our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time.
-
-Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach taken by Sledgehammer, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, for example, the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
-Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}.
-
-\subsection{Recurrent Clustering}
-\label{sec:clusteringexpressions}
-
-Our recurrent clustering approach takes inspiration from the ML4PG \citep{journals/corr/abs-1212-3618} and ACL2(ml) \citep{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
-Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
-
-However, the way we \emph{use} our clusters to inform theory exploration is similar to that of ACL2(ml).
-In ACL2(ml), the clusters are used to reason by analogy: finding theorem statements which are similar to the current goal being proved.
-Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
-For example, if we are trying to prove a theorem \texttt{multiply\_commutes}, we might find it appears in the same cluster as \texttt{plus\_commutes} which was proved using the lemma \texttt{plus\_identity} relating \texttt{plus} and \texttt{0}.
-In this case ACL2(ml) might try mutating \texttt{plus\_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \texttt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply\_identity}) would be given to ACL2 to try and prove \texttt{multiply\_commutes}.
-
-Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.
-
-\iffalse
-We could expand this a bit, e.g. talking about how we both use Weka, etc.
-\fi
-
-\subsection{Feature Extraction}
-
-One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
-This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways.
-The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
-
-The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger.
-We use this to make our matrices a uniform size, borrowing the idea from ML4PG.
-Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
-In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.
-
-Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian'').
-This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness.
-This allows us to truncate more aggressively than if the leaves were most significant.
-
-By modelling our inputs as points in high-dimensional spaces, we can consider feature extraction as projection into a lower-dimensional space (known as \emph{dimension reduction}).
-Truncation is a trivial dimension reduction technique; more sophisticated projection functions consider the \emph{distribution} of the input points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-Techniques such as \emph{principle component analysis} (PCA) can be used to find these hyperplanes, but unfortunately require their inputs to already have a fixed, integer number of dimensions.
-In the case of our recursive expressions (which we may consider to have \emph{fractal} dimension), we would need another pre-processing stage to satisfy this requirement.
-
-There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise.
-Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.
-
-One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs).
-These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations.
-In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.
-
-Unfortunately RNNs are difficult to train.
-The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
-
-Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}.
-More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
-
-A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
-Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
-
-\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.

### 2018-05-17 PhDThesis/background_machinelearning.tex
Drag a bunch of stuff into thesis folder

\section{Machine Learning}

Machine learning (ML) is a sub-field of artificial intelligence (AI) which
emphasises the use of statistical methods to find patterns and optimise
functions, in contrast with symbolic reasoning approaches like automated theorem
proving (commonly referred to as ``Good Old-Fashioned AI'').

The machine learning umbrella spans many techniques and applications, but one
common distinction is between \emph{supervised} and \emph{unsupervised} tasks.
In supervised learning we allow the system to perform some behaviour, then an
external process assesses whether it behaved well or poorly (akin to Justice
Potter Stewart's test for obscenity: ``I know it when I see it''). The system's
parameters are optimised to try and score highly on the test. Common supervised
learning tasks include function approximation using input/output examples (e.g.
classifying scanned images of numerals) where the test is the proportion of
example inputs which got the correct output, and reinforcement learning where
the system chooses from a set of ``actions'', receives a pair of ``observation''
and ``reward'' (test score) in response, and must accumulate the most reward.

Unsupervised learning does not have the same testable ``success'' criteria as
supervised learning. Instead, the goal is to find patterns and form models of
the target domain, which is more ambitious and open-ended than supervised
learning but also much harder to define, measure and optimise. Examples of this
are prediction and compression tasks such as auto-encoders
(learned compressor/decompressor pairs, rewarded by how well they preserve data
through a round-trip) and adversarial or co-evolution tasks (coupled systems
where each is rewarded by exploiting mis-prediction in the others).

\input{background_clustering.tex}
\input{background_featureextraction.tex}

### 2018-05-17 PhDThesis/background_quickcheck.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-08 TransferReport/background.tex
@@
-\section{Background}
-\label{sec:background}
-
-\subsection{Haskell}
-\label{sec:haskell}
-
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \CVar\ id                          \\
-                         |\ & \CLit\ literal                     \\
-                         |\ & \CApp\ expr\ expr                  \\
-                         |\ & \CLam\ \mathcal{L}\ expr           \\
-                         |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                         |\ & \CType                             \\
-      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                         |\ & \CGlobal\      \mathcal{G}         \\
-                         |\ & \CConstructor\ \mathcal{D}         \\
-      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                         |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                         |\ & \CLitAlt\ literal                  \\
-                         |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                         |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-    \end{split}
-  \end{equation*}
-  Where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers \\
-    $\mathcal{D}$ & constructor identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
-
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
-
-Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most \emph{functional} programming languages, Haskell is essentially a variant of $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
-Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-
-The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
-For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
-We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
-
-\begin{figure}
-  \begin{haskell}
-    data Nat = Z
-             | S Nat
-
-    plus :: Nat -> Nat -> Nat
-    plus    Z  y = y
-    plus (S x) y = S (plus x y)
-
-    mult :: Nat -> Nat -> Nat
-    mult    Z  y = Z
-    mult (S x) y = plus y (mult x y)
-
-    odd :: Nat -> Bool
-    odd    Z  = False
-    odd (S n) = even n
-
-    even :: Nat -> Bool
-    even    Z  = True
-    even (S n) = odd n
-  \end{haskell}
-  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
-\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
-  \label{fig:haskellexample}
-\end{figure}
-
-\begin{figure}
-  \begin{subfigure}[plus]{\textwidth}
-    \begin{small}
-      \underline{\texttt{plus}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
-                                               (App (App (Var (Global "plus"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[mult]{\textwidth}
-    \begin{small}
-      \underline{\texttt{mult}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
-                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                    (Var (Local  "y")))
-                                               (App (App (Var (Global "mult"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local  "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[odd]{\textwidth}
-    \begin{small}
-      \underline{\texttt{odd}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "False")))
-              (Alt (DataAlt "S") (App (Var (Global "even"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[even]{\textwidth}
-    \begin{small}
-      \underline{\texttt{even}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "True")))
-              (Alt (DataAlt "S") (App (Var (Global "odd"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}.
-Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
-  \label{fig:coreexample}
-\end{figure}
-
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
-The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the global variables.
-This correspondence allows us to analyse
-
-Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}.
-GHC can represent these, but in this work we only consider reducible expressions (i.e.
-value-level bindings of the form \hs{f a b ...
-= ...}).
-
 \subsection{QuickCheck}
@@
 
-Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g.
-life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests:
+Although unit testing is the de facto industry standard for quality assurance in
+non-critical systems, the level of confidence it provides is rather low, and
+totally inadequate for many (e.g. life-) critical systems. To see why, consider
+the following Haskell function, along with some unit tests:
 
@@
 
-The intent of the function is to map an input $n$ to an output $n!$.
-The tests check a few properties of the implementation, including the base case, that the function is monotonically increasing, and a relationship between adjacent outputs.
-However, these tests will \emph{not} expose a serious problem with the implementation: it diverges on half of its possible inputs!
+The intent of the function is to map an input $n$ to an output $n!$. The tests
+check a few properties of the implementation, including the base case, that the
+function is monotonically increasing, and a relationship between adjacent
+outputs. However, these tests will \emph{not} expose a serious problem with the
+implementation: it diverges on half of its possible inputs!
 
-All of Haskell's built-in numeric types allow negative numbers, which this implementation doesn't take into account.
-Whilst this is a rather trivial example, it highlights a common problem: unit tests are insufficient to expose incorrect assumptions.
-In this case, our assumption that numbers are positive has caused a bug in the implementation \emph{and} limited the tests we've written.
+All of Haskell's built-in numeric types allow negative numbers, which this
+implementation doesn't take into account. Whilst this is a rather trivial
+example, it highlights a common problem: unit tests are insufficient to expose
+incorrect assumptions. In this case, our assumption that numbers are positive
+has caused a bug in the implementation \emph{and} limited the tests we've
+written.
 
-If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
+If we do manage to spot this error, we might capture it in a \emph{regression
+  test} and update the definition of \hs{factorial} to handle negative numbers,
+e.g. by taking their absolute value:
 
@@
 
-However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t.
-Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
-In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge.
-Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
+However, this is \emph{still} not enough, since this function will also accept
+fractional values\footnote{Since we only use generic numeric operations, the
+  function will be polymorphic with a type of the form \hs{forall t. Num t => t
+    -> t}, where \hs{Num t} constrains the type variable \hs{t} to be
+  numeric. In fact, Haskell will infer extra constraints such as \hs{Eq t} since
+  we have used \hs{==} in the unit tests.}, which will also cause it to
+diverge. Clearly, by choosing what to test we are biasing the test suite towards
+those cases we've already taken into account, whilst neglecting the problems we
+did not expect.
 
-Haskell offers a partial solution to this problem in the form of \emph{property checking}.
-Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
-Here is how we might restate our unit tests as properties:
+Haskell offers a partial solution to this problem in the form of \emph{property
+  checking}. Tools such as \qcheck{} separate tests into three components: a
+\emph{property} to check, which unlike a unit test may contain \emph{free
+  variables}; a source of values to instantiate these free variables; and a
+stopping criterion. Here is how we might restate our unit tests as properties:
 
@@
 
-The free variables (all called \hs{n} in this case) are abstracted as function parameters; these parameters are implicitly \emph{universally quantified}, i.e.
-we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
-Notice that unit tests like \hs{fact\_base} are valid properties; they just assert rather weak statements.
+The free variables (all called \hs{n} in this case) are abstracted as function
+parameters; these parameters are implicitly \emph{universally quantified},
+i.e. we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a
+property asserting $\forall n, factorial(n) \leq factorial(n+1)$. Notice that
+unit tests like \hs{fact\_base} are valid properties; they just assert rather
+weak statements.
 
-To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
-For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
-The default stopping criterion for \qcheck{} (for each test) is when a single generated test fails, or when 100 generated tests pass.
+To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base})
+just like unit tests: pass if they evaluate to \hs{True}, fail otherwise. For
+open terms, a random selection of values are generated and passed in via the
+function parameter; the results are then treated in the same way as closed
+terms. The default stopping criterion for \qcheck{} (for each test) is when a
+single generated test fails, or when 100 generated tests pass.
 
-The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:
+The ability to state \emph{universal} properties in this way avoids some of the
+bias we encountered with unit tests. In the \hs{factorial} example, this
+manifests in two ways:
 
 \begin{itemize}
-  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type).
-This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
+\item \qcheck{} cannot test polymorphic functions; they must be
+  \emph{monomorphised} first (instantiated to a particular concrete type). This
+  is a technical limitation, since \qcheck{} must know which type of values to
+  generate, but in our example it would bring the issue with fractional values
+  to our attention.
 
-  \item The generators used by \qcheck{} depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
-This will expose the problem with negative numbers, which we weren't expecting.
+\item The generators used by \qcheck{} depend only on the \emph{type} of value
+  they are generating: since \hs{Int} includes positive and negative values, the
+  \hs{Int} generator will output both. This will expose the problem with
+  negative numbers, which we weren't expecting.
 \end{itemize}
 
-Property checking is certainly an improvement over unit testing, but the problem of tests being biased towards expected cases remains, since we are manually specifying the properties to be checked.
+Property checking is certainly an improvement over unit testing, but the problem
+of tests being biased towards expected cases remains, since we are manually
+specifying the properties to be checked.
 
-We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}.
-These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.
-
-\subsection{Theory Exploration}
-\label{sec:theoryexploration}
-
-In this work we consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
-
-We focus on \qspec{} \citep{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
-These conjectures are arrived at through the following stages:
-
-\iffalse TODO: Make this more formal?
- V \in Var
- F \in Fun
- T \in Term
- T ::= V | F | T1 T2
-
- Term ::= VAR | Const | Fun (Term)
-or
- Term t ::= x | f | t t'
-\fi
-
-\begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the functions and constants from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
-To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
-  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item Each variable is instantiated to a particular value, generated randomly by \qcheck{}.
-  \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
-If so, the class is split up to separate such distinguishable members.
-  \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
-  \item A set of equations are then conjectured, relating each class's members.
-\end{enumerate}
-
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
-
-As an example, we can consider a simple signature containing the expressions from figure \ref{fig:haskellexample}:
-
-\begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
-\end{align*}
-
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
-
-\begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
-\end{align*}
-
-Notice that Haskell curries functions, so the binary functions \hs{plus} and \hs{mult} can be treated as unary functions which return unary functions.
-This is required as the construction of $terms$ applies functions to one argument at a time.
-\iffalse TODO: Maybe mention that not all definitions of a term would regard plus as a well-formed term on its own; maybe make it more explicit that yes, we do regard functions as first-class values, etc \fi
-
-\begin{figure}
-  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
-  \begin{haskell}
-                  plus a b = plus b a
-                  plus a Z = a
-         plus a (plus b c) = plus b (plus a c)
-                  mult a b = mult b a
-                  mult a Z = Z
-         mult a (mult b c) = mult b (mult a c)
-              plus a (S b) = S (plus a b)
-              mult a (S b) = plus a (mult a b)
-         mult a (plus b b) = mult b (plus a a)
-                 odd (S a) = even a
-            odd (plus a a) = odd Z
-           odd (times a a) = odd a
-                even (S a) = odd a
-           even (plus a a) = even Z
-          even (times a a) = even a
-plus (mult a b) (mult a c) = mult a (plus b c)
-  \end{haskell}
-  \caption{Equations conjectured by \qspec{} for the functions in figure \ref{fig:haskellexample}; after simplification.}
-  \label{fig:qspecresult}
-\end{figure}
-
-These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.
-
-Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-Here we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
-
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
-
-In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-
-\subsection{Clustering}
-\label{sec:clustering}
-
-Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{sec:relevance}.
-Premise selection is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
-
-Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting.
-The reason is that premise selection requires a distinguished expression to compare everything against; in practice this is the current goal of a theorem prover.
-Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
-
-A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
-
-\subsubsection{Feature Extraction}
-\label{sec:featureextraction}
-
-Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
-This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms.
-By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
-
-We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$.
-\footnote{In fact, ractical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
-
-\begin{itemize}
-  \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $d$.
-Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
-  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
-This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
-  \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
-This reduces the input size of the machine learning problem, improving efficiency.
-\end{itemize}
-
-\subsubsection{k-Means}
-\label{sec:kmeans}
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:
-
-\begin{itemize}
-  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$.
-  \item Data points will be $d$-dimensional feature vectors, as defined above.
-  \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm.
-\end{itemize}
-
-This is a standard setup, supported by off-the-shelf tools. In particular, it is similar to that used by the ML4PG and ACL2(ml), which makes our results more easily comparable.
-
-k-means works, as the name suggests, by calculating the \emph{mean value} of each cluster, which we define as: \iffalse TODO: Define environment? \fi
-
-\begin{equation*}
-  \mean{X}_i = \frac{\sum_{x \in X} x_i}{|X|}
-\end{equation*}
-
-Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$ and their mean values as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation*}
-  \vect{m}^i(t) = \mean{C}^i(t) \quad \text{for $t > 0$}
-\end{equation*}
-
-Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^i(0)$.
-Many methods have been proposed for choosing these values \citep{arthur2007k}.
-For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
-
-The elements of each cluster $C^i(t)$ are those data points closest to the mean value at the previous time step:
-
-\begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
-\end{equation*}
-
-As $t$ increases, the clusters $C^i$ move from their initial location around the ``seeds'', to converge on a local minimum of the ``within-cluster sum of squared error'' objective:
-
-\begin{equation*}
-  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation*}
-
-\iffalse TODO: subsection Available Software \fi
+We can reduce this bias further through the use of \emph{theory exploration}
+tools, such as \qspec{} and \hspec{}. These programs \emph{discover} properties
+of a ``theory'' (e.g. a library), through a combination of brute-force
+enumeration, random testing and (in the case of \hspec{}) automated theorem
+proving.

### 2018-05-17 PhDThesis/background_recurrentclustering.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-09 TransferReport/related.tex
@@
-\section{Related Work}
-\label{sec:related}
+Our recurrent clustering approach takes inspiration from the ML4PG
+\cite{journals/corr/abs-1212-3618} and ACL2(ml) \cite{heras2013proof} tools,
+used for analysing proofs in Coq and ACL2, respectively. Whilst both transform
+syntax trees into matrices, the algorithm of ML4PG most closely resembles ours
+as it assigns tokens directly to matrix elements. In contrast, the matrices
+produced by ACL2(ml) \emph{summarise} information about the tree; for example,
+one column counts the number of variables appearing at each tree level, others
+count the number of function symbols which are nullary, unary, binary,
+etc. Whilst it may be interesting to contrast our current algorithm with an
+alternative based on that of ACL2(ml), it is unclear how such summaries could be
+extended to include types, which seems the next logical step for our
+approach. The ML4PG algorithm extends trivially, by using (term, type) pairs
+instead of just terms.
 
-\subsection{Haskell}
-\label{sec:haskelldesc}
+The way we \emph{use} our clusters to inform theory exploration is actually more
+similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
+user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
+clusters are used to restrict the search space of a proof search, much like we
+restrict the scope of theory exploration.
 
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
-Here we discuss the relevant language features from a high-level, which motivated our choice:
+ACL2(ml) reasons by analogy: finding theorem statements which are similar to the
+current goal, and attempting to prove the goal in a similar way. In particular,
+the lemmas used to prove a theorem are mutated by substituting symbols for those
+which appear in the same cluster. For example, if \texttt{plus} and
+\texttt{multiply} are clustered together, and we are trying to prove a goal
+involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem
+involving \texttt{plus}. The lemmas used to prove that theorem will be mutated,
+for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an
+attempt to prove the goal.
 
-\begin{description}
-
-\item{Functional}: All control flow in Haskell is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-This puts Haskell in a rather special position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
-\subsection{Relevance Filtering}
-\label{sec:relevance}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering} (or, \emph{premise selection}, when viewed as the \emph{addition} of relevant information to an initially-empty problem).
-This is the core idea behind our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time.
-
-Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach taken by Sledgehammer, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, for example, the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
-Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}.
-
-\subsection{Recurrent Clustering}
-\label{sec:clusteringexpressions}
-
-Our recurrent clustering approach takes inspiration from the ML4PG \citep{journals/corr/abs-1212-3618} and ACL2(ml) \citep{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
-Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
-
-However, the way we \emph{use} our clusters to inform theory exploration is similar to that of ACL2(ml).
-In ACL2(ml), the clusters are used to reason by analogy: finding theorem statements which are similar to the current goal being proved.
-Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
-For example, if we are trying to prove a theorem \texttt{multiply\_commutes}, we might find it appears in the same cluster as \texttt{plus\_commutes} which was proved using the lemma \texttt{plus\_identity} relating \texttt{plus} and \texttt{0}.
-In this case ACL2(ml) might try mutating \texttt{plus\_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \texttt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply\_identity}) would be given to ACL2 to try and prove \texttt{multiply\_commutes}.
-
-Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.
+Whilst we do not currently reason by analogy, this is an interesting area for
+future work in theory exploration: given a set of theorems relating particular
+terms, we might form conjectures regarding similar terms found through
+clustering.
 
@@
 \fi
-
-\subsection{Feature Extraction}
-
-One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
-This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways.
-The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
-
-The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger.
-We use this to make our matrices a uniform size, borrowing the idea from ML4PG.
-Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
-In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.
-
-Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian'').
-This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness.
-This allows us to truncate more aggressively than if the leaves were most significant.
-
-By modelling our inputs as points in high-dimensional spaces, we can consider feature extraction as projection into a lower-dimensional space (known as \emph{dimension reduction}).
-Truncation is a trivial dimension reduction technique; more sophisticated projection functions consider the \emph{distribution} of the input points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-Techniques such as \emph{principle component analysis} (PCA) can be used to find these hyperplanes, but unfortunately require their inputs to already have a fixed, integer number of dimensions.
-In the case of our recursive expressions (which we may consider to have \emph{fractal} dimension), we would need another pre-processing stage to satisfy this requirement.
-
-There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise.
-Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.
-
-One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs).
-These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations.
-In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.
-
-Unfortunately RNNs are difficult to train.
-The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
-
-Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}.
-More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
-
-A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
-Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
-
-\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.

### 2018-05-17 PhDThesis/background_theoryexploration.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-15 TransferReport/background.tex
@@
-\section{Background}
-\label{sec:background}
-
-\subsection{Haskell}
-\label{sec:haskell}
-
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \CVar\ id                          \\
-                         |\ & \CLit\ literal                     \\
-                         |\ & \CApp\ expr\ expr                  \\
-                         |\ & \CLam\ \mathcal{L}\ expr           \\
-                         |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                         |\ & \CType                             \\
-      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                         |\ & \CGlobal\      \mathcal{G}         \\
-                         |\ & \CConstructor\ \mathcal{D}         \\
-      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                         |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                         |\ & \CLitAlt\ literal                  \\
-                         |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                         |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-    \end{split}
-  \end{equation*}
-  Where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers \\
-    $\mathcal{D}$ & constructor identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
-
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
-
-Haskell is well-suited to programming language research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}.
-Like most \emph{functional} programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-For simplicity, we will focus on an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively large and complex syntax of Haskell proper.
-Core is based on \fc{}, but for our machine learning purposes we are mostly interested in its syntax; for a more thorough treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.
-
-The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major restriction is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions).
-For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals.
-We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
-
-\begin{figure}
-  \begin{haskell}
-    data Nat = Z
-             | S Nat
-
-    plus :: Nat -> Nat -> Nat
-    plus    Z  y = y
-    plus (S x) y = S (plus x y)
-
-    mult :: Nat -> Nat -> Nat
-    mult    Z  y = Z
-    mult (S x) y = plus y (mult x y)
-
-    odd :: Nat -> Bool
-    odd    Z  = False
-    odd (S n) = even n
-
-    even :: Nat -> Bool
-    even    Z  = True
-    even (S n) = odd n
-  \end{haskell}
-  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}.
-\hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool \equal{} True | False}.}
-  \label{fig:haskellexample}
-\end{figure}
-
-\begin{figure}
-  \begin{subfigure}[plus]{\textwidth}
-    \begin{small}
-      \underline{\texttt{plus}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
-                                               (App (App (Var (Global "plus"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[mult]{\textwidth}
-    \begin{small}
-      \underline{\texttt{mult}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
-                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                    (Var (Local  "y")))
-                                               (App (App (Var (Global "mult"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local  "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[odd]{\textwidth}
-    \begin{small}
-      \underline{\texttt{odd}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "False")))
-              (Alt (DataAlt "S") (App (Var (Global "even"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[even]{\textwidth}
-    \begin{small}
-      \underline{\texttt{even}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "True")))
-              (Alt (DataAlt "S") (App (Var (Global "odd"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the Core syntax of Figure \ref{fig:coresyntax}.
-Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
-  \label{fig:coreexample}
-\end{figure}
-
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals.
-The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
-Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
-It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
-
-Note that we exclude representations for type-level entities, including datatype definitions like that of \hs{Nat}.
-GHC can represent these, but in this work we only consider reducible expressions (i.e.
-value-level bindings of the form \mbox{\hs{f a b ...
-= ...}}).
-
-\subsection{QuickCheck}
-\label{sec:quickcheck}
-
-Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g.
-life-) critical systems.
-To see why, consider the following Haskell function, along with some unit tests:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = n * factorial (n-1)
-
-fact_base      = factorial 0 == factorial 1
-fact_increases = factorial 3 <= factorial 4
-fact_div       = factorial 4 == factorial 5 `div` 5
-\end{lstlisting}
-
-The intent of the function is to map an input $n$ to an output $n!$.
-The tests check a few properties of the implementation, including the base case, that the function is monotonically increasing, and a relationship between adjacent outputs.
-However, these tests will \emph{not} expose a serious problem with the implementation: it diverges on half of its possible inputs!
-
-All of Haskell's built-in numeric types allow negative numbers, which this implementation doesn't take into account.
-Whilst this is a rather trivial example, it highlights a common problem: unit tests are insufficient to expose incorrect assumptions.
-In this case, our assumption that numbers are positive has caused a bug in the implementation \emph{and} limited the tests we've written.
-
-If we do manage to spot this error, we might capture it in a \emph{regression test} and update the definition of \hs{factorial} to handle negative numbers, e.g. by taking their absolute value:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-factorial 0 = 1
-factorial n = let nPos = abs n
-               in nPos * factorial (nPos - 1)
-
-fact_neg = factorial 1 == factorial (-1)
-\end{lstlisting}
-
-However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t.
-Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric.
-In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge.
-Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.
-
-Haskell offers a partial solution to this problem in the form of \emph{property checking}.
-Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion.
-Here is how we might restate our unit tests as properties:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-fact_base        = factorial 0 == factorial 1
-fact_increases n = factorial n <= factorial (n+1)
-fact_div       n = factorial n == factorial (n+1) `div` (n+1)
-fact_neg       n = factorial n == factorial (-n)
-\end{lstlisting}
-
-The free variables (all called \hs{n} in this case) are abstracted as function parameters; these parameters are implicitly \emph{universally quantified}, i.e.
-we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a property asserting $\forall n, factorial(n) \leq factorial(n+1)$.
-Notice that unit tests like \hs{fact\_base} are valid properties; they just assert rather weak statements.
-
-To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base}) just like unit tests: pass if they evaluate to \hs{True}, fail otherwise.
-For open terms, a random selection of values are generated and passed in via the function parameter; the results are then treated in the same way as closed terms.
-The default stopping criterion for \qcheck{} (for each test) is when a single generated test fails, or when 100 generated tests pass.
-
-The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:
-
-\begin{itemize}
-  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type).
-This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.
-
-  \item The generators used by \qcheck{} depend only on the \emph{type} of value they are generating: since \hs{Int} includes positive and negative values, the \hs{Int} generator will output both.
-This will expose the problem with negative numbers, which we weren't expecting.
-\end{itemize}
-
-Property checking is certainly an improvement over unit testing, but the problem of tests being biased towards expected cases remains, since we are manually specifying the properties to be checked.
-
-We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}.
-These programs \emph{discover} properties of a ``theory'' (e.g.
-a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.
-
 \subsection{Theory Exploration}
@@
 
-In this work we consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
+In this work we consider the problem of \emph{(automated) theory exploration},
+which includes the ability to \emph{generate} conjectures about code, to
+\emph{prove} those conjectures, and hence output \emph{novel} theorems without
+guidance from the user. The method of conjecture generation is a key
+characteristic of any theory exploration system, although all existing
+implementations rely on brute force enumeration to some degree.
 
-We focus on \qspec{} \citep{QuickSpec}, which conjectures equations about Haskell code (these may be fed into another tool, such as \hspec{}, for proving).
-These conjectures are arrived at through the following stages:
+We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about
+Haskell code (these may be fed into another tool, such as \hspec{}, for
+proving). These conjectures are arrived at through the following stages:
 
@@
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
-To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
-  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item Each variable is instantiated to a particular value, generated randomly by \qcheck{}.
-  \item For each class, the members are compared (using a pre-specified function, such as equality \hs{==}) to see if these instantiations have caused an observable difference between members.
-If so, the class is split up to separate such distinguishable members.
-  \item The previous steps of variable instantiation and comparison are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
+\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
+  generates a list $terms$ containing the constants (including functions) from
+  $\Sigma$, the variables from $V$ and type-correct function applications
+  $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little
+  awkward; maybe use the above notation? \fi. To ensure the list is finite,
+  function applications are only nested up to a specified depth (by default, 3).
+\item The elements of $terms$ are grouped into equivalence classes, based on
+  their type.
+\item Each variable is instantiated to a particular value, generated randomly by
+  \qcheck{}.
+\item For each class, the members are compared (using a pre-specified function,
+  such as equality \hs{==}) to see if these instantiations have caused an
+  observable difference between members. If so, the class is split up to
+  separate such distinguishable members.
+\item The previous steps of variable instantiation and comparison are repeated
+  until the classes stabilise (i.e. no differences have been observed for some
+  specified number of repetitions).
 \item A set of equations are then conjectured, relating each class's members.
@@
 
-Such $conjectures$ can be used in several ways: they can be simplified for direct presentation to the user (by removing any equation which can be derived from the others by rewriting), sent to a more rigorous system like \hspec{} or \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.
+Such $conjectures$ can be used in several ways: they can be simplified for
+direct presentation to the user (by removing any equation which can be derived
+from the others by rewriting), sent to a more rigorous system like \hspec{} or
+\textsc{Hipster} for proving, or even serve as a background theory for an
+automated theorem prover \cite{claessen2013automating}.
 
-As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
+As an example, we can consider a simple signature containing the expressions
+from Figure \ref{fig:haskellexample}:
 
@@
 
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
+Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$,
+\qspec{}'s enumeration will resemble the following:
 
@@
 
-Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite not being applied to any arguments.
-In addition, Haskell curries functions, so these binary functions will be treated as unary functions which return unary functions.
-This is required as the construction of $terms$ applies functions to one argument at a time.
+Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite
+not being applied to any arguments. In addition, Haskell curries functions, so
+these binary functions will be treated as unary functions which return unary
+functions. This is required as the construction of $terms$ applies functions to
+one argument at a time.
 
@@
 
-These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in Figure \ref{fig:qspecresult} are conjectured.
+These terms will be grouped into five classes, one each for \hs{Nat},
+\hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}. As the
+variables $a$, $b$ and $c$ are instantiated to various randomly-generated
+numbers, these equivalence classes will be divided, until eventually the
+equations in Figure \ref{fig:qspecresult} are conjectured.
 
-Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-Here we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which nevertheless have the potential to give rise to many equations.
+Although complete, this enumeration approach is wasteful: many terms are
+unlikely to appear in theorems, which requires careful choice by the user of
+what to include in the signature. Here we know that addition and multiplication
+are closely related, and hence obey many algebraic laws. Our machine learning
+technique aims to predict these kinds of relations between functions, so we can
+create small signatures which nevertheless have the potential to give rise to
+many equations.
 
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.
+\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing
+infrastructure, such that an invocation of \texttt{cabal test} can run these
+tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and
+\textsc{Criterion}.
 
-In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-
-\subsection{Clustering}
-\label{sec:clustering}
-
-Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs.
-We describe this approach in more details in \S \ref{sec:relevance}.
-Relevance filtering is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
-
-Despite the idea's promise, we cannot simply invoke existing relevance filter algorithms in our theory exploration setting.
-The reason is that relevance filtering is a supervised learning method, i.e.
-it would require a distinguished expression to compare everything against.
-Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.
-
-A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
-
-\subsubsection{Feature Extraction}
-\label{sec:featureextraction}
-
-Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}.
-This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms.
-By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.
-
-We use a standard representation of features as a \emph{feature vector} of numbers $\vect{x} = (x_1, \dots, x_d)$ where $x_i \in \mathbb{R}$ (we use \textbf{bold face} to represent vectors, including feature vectors).
-\footnote{In fact, practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has some important advantages over raw expressions:
-
-\begin{itemize}
-  \item All of our feature vectors will be the same size, i.e.
-they will all have length (or \emph{dimension}) $d$.
-Many machine learning algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.
-For example, element-wise comparison of feature vectors is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions this is not so straightforward, as their nesting may give rise to very different shapes.
-  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another.
-This enables many powerful machine learning algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
-  \item Feature vectors can be chosen to represent the relevant information in a more compressed form than the raw data; for example, we might replace verbose, descriptive identifiers with sequential numbers.
-This reduces the input size of the machine learning problem, improving efficiency.
-\end{itemize}
-
-\subsubsection{K-Means}
-\label{sec:kmeans}
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:
-
-\begin{itemize}
-  \item For simplicity, we use the ``rule of thumb'' given in \citep[pp. 365]{mardia1979multivariate} to fix the number of clusters at $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
-  \item Data points will be $d$-dimensional feature vectors, as defined above.
-  \item We will use euclidean distance (denoted $e$) as our similarity metric.
-  \item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm \citep{lloyd1982least}.
-\end{itemize}
-
-This is a standard setup, supported by off-the-shelf tools; in particular we use the implementation provided by Weka \citep{Holmes.Donkin.Witten:1994}, due to its use by ML4PG, which makes our results more easily comparable.
-
-Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$.
-We denote the clusters as $C^1$ to $C^k$.
-As the name suggests, k-means uses the mean value of each cluster, which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation*}
-  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
-\end{equation*}
-
-Before k-means starts, we must choose \emph{seed} values for $\vect{m}^i(0)$.
-Many methods have been proposed for choosing these values \citep{arthur2007k}.
-For simplicity, we will choose values randomly from our data set $S$; this is known as the Forgy method.
-
-The elements of each cluster $C^i(t)$ are those data points closest to the mean value at the previous time step:
-
-\begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
-\end{equation*}
-
-As $t$ increases, the clusters $C^i$ move from their initial location around the ``seeds'', to converge on a local minimum of the ``within-cluster sum of squared error'' objective:
-
-\begin{equation} \label{eq:kmeansobjective}
-  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation}
+In fact, there are similarities between the way a TE system like \qspec{} can
+generalise from checking \emph{particular} properties to \emph{inventing} new
+ones, and the way counterexample finders like \qcheck{} can generalise from
+testing \emph{particular} expressions to \emph{inventing} expressions to
+test. One of our aims is to understand the implications of this generalisation,
+the lessons that each can learn from the other's approach to term generation,
+and the consequences for testing and QA in general.

### 2018-05-17 PhDThesis/evaluation_existingcode.tex
Drag a bunch of stuff into thesis folder

--- 2016-02-29 CICM2016/evaluation.tex
@@
 
-We have applied our recurrent clustering algorithm to several scenarios, with mixed results.
-A major difficulty in evaluating these clusters is that we have no ``ground truth'', i.e.
-there is no objectively correct way to compare expressions.
-Instead, we provide a qualitative overview of the more interesting characteristics.
+We have applied our recurrent clustering algorithm to several scenarios, with
+mixed results. A major difficulty in evaluating these clusters is that we have
+no ``ground truth'', i.e. there is no objectively correct way to compare
+expressions. Instead, we provide a qualitative overview of the more interesting
+characteristics.
 
-As a simple example, we clustered (Haskell equivalents of) the running examples used to present ACL2(ml) \cite{heras2013proof}, shown in Figure \ref{fig:lisp}.
-These include tail-recursive and non-tail-recursive implementations of several functions.
-We expect those with similar \emph{structure} to be clustered together, rather than those which implement the same function.
-The results are shown in Figure \ref{fig:haskellcluster}, where we can see the ``\hs{Tail}'' functions clearly distinguished, with little distinction between the tail recursive and na\"{\i}ve implementations.
+As a simple example, we clustered (Haskell equivalents of) the running examples
+used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
+\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
+implementations of several functions. We expect those with similar
+\emph{structure} to be clustered together, rather than those which implement the
+same function. The results are shown in Figure \ref{fig:haskellcluster}, where
+we can see the ``\hs{Tail}'' functions clearly distinguished, with little
+distinction between the tail recursive and na\"{\i}ve implementations.
 
-Next we tested whether these same functions would be clustered together when mixed with seemingly-unrelated functions, in this case 207 functions from Haskell's \hs{text} library.
-In fact, the \hs{helperFib} and \hs{fibTail} functions appeared together in a separate cluster from the rest.
-This was unexpected, with no obvious semantic connection between these two functions and the others in their cluster (although most are recursive, due to the nature of the \hs{text} library).
+Next we tested whether these same functions would be clustered together when
+mixed with seemingly-unrelated functions, in this case 207 functions from
+Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail}
+functions appeared together in a separate cluster from the rest. This was
+unexpected, with no obvious semantic connection between these two functions and
+the others in their cluster (although most are recursive, due to the nature of
+the \hs{text} library).
 
@@
 
-We have also applied our recurrent clustering algorithm to a variety of the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including \texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-Whilst we expected functions with a similar purpose to appear together, such as the various reader and writer functions of \hs{pandoc}, there were always a few exceptions which became separate for reasons which are still unclear.
+We have also applied our recurrent clustering algorithm to a variety of the
+most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
+\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
+\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
+expected functions with a similar purpose to appear together, such as the
+various reader and writer functions of \hs{pandoc}, there were always a few
+exceptions which became separate for reasons which are still unclear.
 
-When clustering the \hs{yesod} Web framework, the clustering did seem to match our intuitions, in particular since all 15 of Yesod's MIME type identifiers appeared in the same cluster.
+When clustering the \hs{yesod} Web framework, the clustering did seem to match
+our intuitions, in particular since all 15 of Yesod's MIME type identifiers
+appeared in the same cluster.
 
-Whilst recurrent clustering has produced results which merit further investigation, the application to theory exploration has yet to be tested empirically.
-This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type class to generate random values for instantiating variables.
-Whilst we can automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not all types have \hs{Arbitrary} instances; those without cannot be given any variables in our signature, which severely limits the possible combinations which can be explored.
-In many cases, no variables can be included at all, leaving just equations involving constants.
-This has so far prevented us from measuring the direct impact on \qspec{} performance, either directly by exploring the sub-sets identified through recurrent clustering, or indirectly by comparing the equations generated by a full brute-force search to our recurrent clusters: those equations relating terms from different clusters would not be discovered by our method.
-It is this ratio of equations found through brute force to those found after narrowing-down by clusters which is one of our key objectives to maximise at this stage; until we begin to pursue the \emph{interestingness} of the properties.
+Whilst recurrent clustering has produced results which merit further
+investigation, the application to theory exploration has yet to be tested
+empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type
+class to generate random values for instantiating variables. Whilst we can
+automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
+all types have \hs{Arbitrary} instances; those without cannot be given any
+variables in our signature, which severely limits the possible combinations
+which can be explored. In many cases, no variables can be included at all,
+leaving just equations involving constants. This has so far prevented us from
+measuring the direct impact on \qspec{} performance, either directly by
+exploring the sub-sets identified through recurrent clustering, or indirectly by
+comparing the equations generated by a full brute-force search to our recurrent
+clusters: those equations relating terms from different clusters would not be
+discovered by our method. It is this ratio of equations found through brute
+force to those found after narrowing-down by clusters which is one of our key
+objectives to maximise at this stage; until we begin to pursue the
+\emph{interestingness} of the properties.
 
-The following less-serious problems were also encountered while applying \textsc{ML4HS} to \textsc{Hackage} packages:
+The following less-serious problems were also encountered while applying
+\textsc{ML4HS} to \textsc{Hackage} packages:
 
 \begin{itemize}
-  \item Some packages, such as \texttt{warp} and \texttt{conduits}, get no declarations to cluster.
-This is because they make all of their declarations privately, e.g.
-in ``internal'' modules, then use separate modules to export the public declarations.
-GHC's renaming phase makes all references to such exports canonical, by pointing them to the private declarations.
-This forces us to ignore such declarations, as \qspec{} will not be able to access them.
+\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
+  declarations to cluster. This is because they make all of their declarations
+  privately, e.g. in ``internal'' modules, then use separate modules to export
+  the public declarations. GHC's renaming phase makes all references to such
+  exports canonical, by pointing them to the private declarations. This forces
+  us to ignore such declarations, as \qspec{} will not be able to access them.
 
-  \item Since we do not support type-level entities, we ignore type classes.
-Unfortunately, this also means ignoring any value-level bindings (AKA ``methods'') which occur in a type class instance.
-Instead of being clustered, these result in references getting $f_{recursion}$ features.
-This is especially noticable in libraries like \hs{scientific}, where only the functions for constructing and destructing numbers in scientific notation are clustered; all of the arithmetic is defined in type classes.
-One difficulty with supporting methods is that their namespace in Core is disjoint from that of regular Haskell identifiers: a transformation layer would be required, along with explicit type annotations to avoid ambiguity.
+\item Since we do not support type-level entities, we ignore type
+  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
+  ``methods'') which occur in a type class instance. Instead of being clustered,
+  these result in references getting $f_{recursion}$ features. This is
+  especially noticable in libraries like \hs{scientific}, where only the
+  functions for constructing and destructing numbers in scientific notation are
+  clustered; all of the arithmetic is defined in type classes. One difficulty
+  with supporting methods is that their namespace in Core is disjoint from that
+  of regular Haskell identifiers: a transformation layer would be required,
+  along with explicit type annotations to avoid ambiguity.
 \end{itemize}
 
-It seems like this recurrent clustering method has promise, although it will require a more thorough exploration of the parameters to obtain more intuitively reliable results.
-These clusters can then be used in several ways to perform theory exploration; the most na\"{\i}ve way being to explore each cluster as \qspec{} signature in its own right.
-\textsc{ML4HS} already provides this functionality, although the lack of test generators severely limits what can be discovered.
+It seems like this recurrent clustering method has promise, although it will
+require a more thorough exploration of the parameters to obtain more intuitively
+reliable results. These clusters can then be used in several ways to perform
+theory exploration; the most na\"{\i}ve way being to explore each cluster as
+\qspec{} signature in its own right. \textsc{ML4HS} already provides this
+functionality, although the lack of test generators severely limits what can be
+discovered.
 
-As alluded to previously, we also have the opportunity to reason by \emph{analogy}.
-Similar to the work on ACL2(ml) \cite{heras2013proof}, we could produce a general ``scheme'' from each equation we find (either through \qspec{} or by data mining test suites); like Isabelle approaches have shown \cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be instantiated to a variety of similar values, in an attempt to find new theorems which are analogues of existing results from a different context.
-``Mutating'' existing theorem statements in such a way would also increase the chance of any result being considered interesting; since it's likely that the unmodified statement was deemed interesting, and the new result would not in general follow as a simple logical consequence.
+As alluded to previously, we also have the opportunity to reason by
+\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
+produce a general ``scheme'' from each equation we find (either through \qspec{}
+or by data mining test suites); like Isabelle approaches have shown
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
+instantiated to a variety of similar values, in an attempt to find new theorems
+which are analogues of existing results from a different context. ``Mutating''
+existing theorem statements in such a way would also increase the chance of any
+result being considered interesting; since it's likely that the unmodified
+statement was deemed interesting, and the new result would not in general follow
+as a simple logical consequence.

### 2018-05-17 PhDThesis/futurework.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-15 TransferReport/future.tex
@@
 
-\begin{sidewaysfigure}
-  \centering
-  \includegraphics[trim=3cm 2cm 3cm 2cm, width=1.00\textwidth]{Gantt.pdf}
-  \caption{Gantt chart showing research direction, sub-tasks and approximate timeframes. CICM is the \emph{Conference on Intelligent Computer Mathematics}.}
-  \label{fig:gantt}
-\end{sidewaysfigure}
-
-Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made.
-Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further.
-In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique.
-The Gantt chart in Figure \ref{fig:gantt} shows how these relate to the short- and long-term direction being taken by this research.
-Below, we elaborate on the details, background and motivation for these choices.
+Our use of clustering to pre-process \qspec{} signatures has required many
+decisions and tradeoffs to be made. Hence our approach is just one possibility
+out of many alternatives which could be investigated to push this work
+further. In addition, there are other ways in which machine learning could aid
+theory exploration besides our relevance filter technique. Below, we elaborate
+on the details, background and motivation for these choices.
 
@@
 
-The most glaring omission in our algorithm is its disregard for types.
-By ignoring types, not only are we losing valuable information about expressions, but we also lose the ability to distinguish between constructors.
-This is because a constructor, like \hs{True} or \hs{Just}, has no internal structure; it is just a token.
-The distinguishing features of constructors are their types, which not only tell us which data type they construct, but also their arity, the types of their arguments, etc.
+The most glaring omission in our algorithm is its disregard for types. By
+ignoring types, not only are we losing valuable information about expressions,
+but we also lose the ability to distinguish between constructors. This is
+because a constructor, like \hs{True} or \hs{Just}, has no internal structure;
+it is just a token. The distinguishing features of constructors are their types,
+which not only tell us which data type they construct, but also their arity, the
+types of their arguments, etc.
 
-Our algorithm closely follows that of ML4PG, which \emph{does} support types.
-This is handled by populating matrix cells with tokens \emph{and} their types.
-Unfortunately this is more complicated in Haskell than it is in Coq, since types form a separate part of the language from terms, and we do not have an interactive Core environment to query for types (unlike ML4PG, which runs inside the Proof General environment).
+Our algorithm closely follows that of ML4PG, which \emph{does} support
+types. This is handled by populating matrix cells with tokens \emph{and} their
+types. Unfortunately this is more complicated in Haskell than it is in Coq,
+since types form a separate part of the language from terms, and we do not have
+an interactive Core environment to query for types (unlike ML4PG, which runs
+inside the Proof General environment).
 
-One partial solution would be leave most Core expressions without types, but to include them for non-local identifiers (i.e.
-globals and constructors), which we can look up in a database.
-In fact, our ML4HS framework already includes such type information in its database, alongside the Core syntax trees.
-Integrating this information into our algorithm is the next logical step.
+One partial solution would be leave most Core expressions without types, but to
+include them for non-local identifiers (i.e. globals and constructors), which we
+can look up in a database. In fact, our ML4HS framework already includes such
+type information in its database, alongside the Core syntax trees. Integrating
+this information into our algorithm is the next logical step.
 
-We can also compare the performance of our hand-selected features with \emph{learned} representations, like those reviewed in \citep{bengio2013representation}.
-This may provide an indication of how important it is to understand the language when identifying salient aspects of expressions, and how difficult various aspects of it might be to learn.
+We can also compare the performance of our hand-selected features with
+\emph{learned} representations, like those reviewed in
+\cite{bengio2013representation}. This may provide an indication of how
+important it is to understand the language when identifying salient aspects of
+expressions, and how difficult various aspects of it might be to learn.
 
-With more expressive features, it may also be useful to experiment with more powerful learning algorithms.
-An interesting possibility is to add a feedback loop between the theory exploration phase and the clustering phase, to more directly base the similarity of expressions on whether they (are predicted to) occur together in equations.
+With more expressive features, it may also be useful to experiment with more
+powerful learning algorithms. An interesting possibility is to add a feedback
+loop between the theory exploration phase and the clustering phase, to more
+directly base the similarity of expressions on whether they (are predicted to)
+occur together in equations.
 
@@
 
-Our current approach is a rather conservative change to the existing theory exploration approaches, as it is essentially a wrapper around \qspec{}.
-There is potential for more radical changes to be made, which alter the search process itself.
+Our current approach is a rather conservative change to the existing theory
+exploration approaches, as it is essentially a wrapper around \qspec{}. There is
+potential for more radical changes to be made, which alter the search process
+itself.
 
@@
 
-\qcheck{} is certainly the most popular property checker for Haskell, which motivates its use in \qspec{} to instantiate variables to random values.
-However, this task of finding type inhabitants has also been solved in many other ways, which may be worth investigating in place of \qcheck{} (or perhaps even as part of an ensemble).
+\qcheck{} is certainly the most popular property checker for Haskell, which
+motivates its use in \qspec{} to instantiate variables to random
+values. However, this task of finding type inhabitants has also been solved in
+many other ways, which may be worth investigating in place of \qcheck{} (or
+perhaps even as part of an ensemble).
 
-The \textsc{SmallCheck} system \citep{runciman2008smallcheck} \emph{enumerates} values rather than sampling them randomly.
-Whilst this does not make \textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage is that it may use much less memory, as the generated values are built up incrementally.
-In contrast, \qcheck{} may generate very large values; in particular, generating tree structures na\"{\i}vely can cause them to grow exponentially.
-For example, here is a potential generator for \hs{RoseTree}s:
+The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates}
+values rather than sampling them randomly. Whilst this does not make
+\textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage
+is that it may use much less memory, as the generated values are built up
+incrementally. In contrast, \qcheck{} may generate very large values; in
+particular, generating tree structures na\"{\i}vely can cause them to grow
+exponentially. For example, here is a potential generator for \hs{RoseTree}s:
 
@@
 
-The \hs{listOf genRoseTree} call will return a list of arbitrary length, where each element is generated by \hs{genRoseTree}.
-This allows an arbitrary number of recursive calls to be made for each invocation of \hs{genRoseTree}, which will quickly exhaust the resources of any machine.
-Whilst such problems may be anticipated, or quickly spotted, in a property checking setting, this can be more difficult for our automated approach.
-For example, if a type does not have a generator available, we cannot use a library like \hs{derive} to create one automatically, as it suffers from this na\"{\i}vity problem.
+The \hs{listOf genRoseTree} call will return a list of arbitrary length, where
+each element is generated by \hs{genRoseTree}. This allows an arbitrary number
+of recursive calls to be made for each invocation of \hs{genRoseTree}, which
+will quickly exhaust the resources of any machine. Whilst such problems may be
+anticipated, or quickly spotted, in a property checking setting, this can be
+more difficult for our automated approach. For example, if a type does not have
+a generator available, we cannot use a library like \hs{derive} to create one
+automatically, as it suffers from this na\"{\i}vity problem.
 
-A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck} \citep{reich2013advances}, which uses laziness to only produce parts of a datastructure as they are demanded.
-This may narrow down our search procedures greatly, especially when predicates are involved.
-\qcheck{} allows predicates to restrict the values it tests with, and hence allows \emph{conditional} equations to be discovered.
-However, its implementation uses a simple rejection sampling technique: values are generated just as if the predicate were not there, and afterwards are filtered to reject any which do not satisfy the predicate.
-This makes it difficult to use very specific predicates, as it is unlikely that many of our random samples will exactly match our criteria.
-On the other hand, \textsc{Lazy SmallCheck} will focus its search on exactly those parts of the datastructure which are checked by the predicate, as those are the parts being forced to evaluate.
-This makes it much more likely that we will find values which satisfy the predicate, allowing us to effectively explore more specific conditional properties.
+A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck}
+\cite{reich2013advances}, which uses laziness to only produce parts of a
+datastructure as they are demanded. This may narrow down our search procedures
+greatly, especially when predicates are involved. \qcheck{} allows predicates to
+restrict the values it tests with, and hence allows \emph{conditional} equations
+to be discovered. However, its implementation uses a simple rejection sampling
+technique: values are generated just as if the predicate were not there, and
+afterwards are filtered to reject any which do not satisfy the predicate. This
+makes it difficult to use very specific predicates, as it is unlikely that many
+of our random samples will exactly match our criteria. On the other hand,
+\textsc{Lazy SmallCheck} will focus its search on exactly those parts of the
+datastructure which are checked by the predicate, as those are the parts being
+forced to evaluate. This makes it much more likely that we will find values
+which satisfy the predicate, allowing us to effectively explore more specific
+conditional properties.
 
-Other approaches to generating inhabitants include \textsc{Djinn} \citep{augustsson2005djinn}, which uses a decision procedure for a sub-set of Haskell types which in particular can generate and apply functions (unlike the above tools, which generate values ``bottom-up'' from constructors, and only use functions when they have been explicitly written in a generator).
-\textsc{MuCheck} \citep{le2014mucheck} is designed for \emph{mutation testing}, and contains combinators for altering functions in common ways (e.g.
-changing the order of pattern match clauses); whilst not as exhaustive as the other approaches, mutating existing values in this way is claimed to yield values which correspond more closely to what a programmer might write.
-This is an interesting possibility for focusing theory exploration on to more ``realistic'' areas of the search space, and hence avoiding some of the more useless or bizarre expressions that random search and enumeration may produce.
+Other approaches to generating inhabitants include \textsc{Djinn}
+\cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of
+Haskell types which in particular can generate and apply functions (unlike the
+above tools, which generate values ``bottom-up'' from constructors, and only use
+functions when they have been explicitly written in a
+generator). \textsc{MuCheck} \cite{le2014mucheck} is designed for
+\emph{mutation testing}, and contains combinators for altering functions in
+common ways (e.g. changing the order of pattern match clauses); whilst not as
+exhaustive as the other approaches, mutating existing values in this way is
+claimed to yield values which correspond more closely to what a programmer might
+write. This is an interesting possibility for focusing theory exploration on to
+more ``realistic'' areas of the search space, and hence avoiding some of the
+more useless or bizarre expressions that random search and enumeration may
+produce.
 
-In fact, the database generated by our \textsc{AstPlugin} may prove helpful in generating values, since its type information can be fed to a tool like \textsc{Djinn} to discover chains of function applications for building values, which would be particularly useful in cases where constructors are private, like in our email example.
-This is similar to the \textsc{Hoogle} tool, but also offers the ability to use dependency information to avoid potentially infinite recursion.
+In fact, the database generated by our \textsc{AstPlugin} may prove helpful in
+generating values, since its type information can be fed to a tool like
+\textsc{Djinn} to discover chains of function applications for building values,
+which would be particularly useful in cases where constructors are private, like
+in our email example. This is similar to the \textsc{Hoogle} tool, but also
+offers the ability to use dependency information to avoid potentially infinite
+recursion.
 
-The Core syntax trees in our database could also be used to generate theories for automated theorem provers.
-\hspec{} currently uses the GHC API to transform Core within its own process, however that approach suffers from the problems described in \S \ref{sec:astplugin}.
+The Core syntax trees in our database could also be used to generate theories
+for automated theorem provers. \hspec{} currently uses the GHC API to transform
+Core within its own process, however that approach suffers from the problems
+described in \S \ref{sec:astplugin}.
 
@@
 
-If we do succeed in producing a fast theory exploration system, which chooses productive combinations of terms and finds a large number of properties, we encounter the problem of managing the output: finding the needles we are interested in among the haystack of trivialities and coincidences.
+If we do succeed in producing a fast theory exploration system, which chooses
+productive combinations of terms and finds a large number of properties, we
+encounter the problem of managing the output: finding the needles we are
+interested in among the haystack of trivialities and coincidences.
 
-This is governed by the ``interestingness'' criteria of the theory exploration system: what to keep and what to discard, and even what areas of the search space to prioritise.
-\qspec{}'s approach, briefly mentioned in \S \ref{sec:theoryexploration}, is very simple: we discard equations which are direct consequences of others, and keep all the rest.
-Different, and more sophisticated notions of interestingness have been widely studied in other fields, which may be applied in the context of theory exploration.
+This is governed by the ``interestingness'' criteria of the theory exploration
+system: what to keep and what to discard, and even what areas of the search
+space to prioritise. \qspec{}'s approach, briefly mentioned in \S
+\ref{sec:theoryexploration}, is very simple: we discard equations which are
+direct consequences of others, and keep all the rest. Different, and more
+sophisticated notions of interestingness have been widely studied in other
+fields, which may be applied in the context of theory exploration.
 
@@
 
-One directly applicable area to consider is \emph{concept formation}, which considers the (automatic) generation of new definitions and axioms.
-Unlike theory exploration, such systems are not constrained by the requirement that their output be provable, and hence interestingness is an important way to judge the quality of the results.
+One directly applicable area to consider is \emph{concept formation}, which
+considers the (automatic) generation of new definitions and axioms. Unlike
+theory exploration, such systems are not constrained by the requirement that
+their output be provable, and hence interestingness is an important way to judge
+the quality of the results.
 
-Approaches vary, from those which are directly related to theory exploration (such as the scheme-instantiating approach of \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory exploration system), to others which are more closely related to theories of human learning and discovery \citep{Piantadosi.Tenenbaum.Goodman:2012, mullerunderstanding}.
-Those based on finding patterns in data, such as \citep{Wille:2005}, may be useful in tandem with our expression database, and the results of value generators like \qcheck{}.
+Approaches vary, from those which are directly related to theory exploration
+(such as the scheme-instantiating approach of
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory
+exploration system), to others which are more closely related to theories of
+human learning and discovery \cite{Piantadosi.Tenenbaum.Goodman:2012,
+  mullerunderstanding}. Those based on finding patterns in data, such as
+\cite{Wille:2005}, may be useful in tandem with our expression database, and
+the results of value generators like \qcheck{}.
 
-Since tools like \hspec{} already call out to third-party automated theorem provers, and indeed our own ML4HS system uses the external Weka program, there may also be merit in using external concept formation or conjecture generation tools (or reimplementations of their ideas), in order to build up more structure on top of that provided by the code we are exploring.
-For example, the approaches taken by AM \citep{lenat1977automated, lenat1979automated}, Graffiti \citep{delavina2005some, delavina2005graffiti} and HR \citep{colton1999automatic, colton2000agent} could be used alongside those of \qspec{}.
+Since tools like \hspec{} already call out to third-party automated theorem
+provers, and indeed our own ML4HS system uses the external Weka program, there
+may also be merit in using external concept formation or conjecture generation
+tools (or reimplementations of their ideas), in order to build up more structure
+on top of that provided by the code we are exploring. For example, the
+approaches taken by AM \cite{lenat1977automated, lenat1979automated}, Graffiti
+\cite{delavina2005some, delavina2005graffiti} and HR
+\cite{colton1999automatic, colton2000agent} could be used alongside those of
+\qspec{}.
 
@@
 
-  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}.
+  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
 The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
@@
 
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this reliance on interest is clearly relevant to our theory exploration setting.
+\emph{Artificial curiosity} (AC) describes active learning systems which are
+rewarded based on how interesting the input or data they discover is
+\cite{schmidhuber2006developmental}. Although framed in the context of
+\emph{reinforcement learning}, this reliance on interest is clearly relevant to
+our theory exploration setting.
 
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, in a similar way to our recurrent clustering algorithm.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.
-This same behaviour may be applicable to the theorems produced by a theory exploration system.
+As an unsupervised learning task, AC has no access to labels or meanings
+associated with its input; the only features it can learn are the structure and
+relationships inherent in the data, in a similar way to our recurrent clustering
+algorithm. The unifying principle of AC methods is to force systems away from
+inputs which are not amenable to learning; either because they are so familiar
+that there is nothing left to learn, or so unfamiliar that they are
+unintelligible. The resulting behaviour is characterised by the \emph{Wundt
+  curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures
+  avoid negative values for simplicity, in which cases we replace all negative
+  points on the curve with zero.}, which has been used in psychology to explain
+human aesthetics and preferences \cite{berlyne1970novelty}. This same behaviour
+may be applicable to the theorems produced by a theory exploration system.
 
-We can divide AC approaches into two groups: those which make \emph{explicit} use of interestingness, learning from signals which follow a Wundt curve; whilst \emph{implicit} approaches modify the \emph{output} of their learning algorithm(s), to engineer the overall system behaviour to follow a Wundt curve as an emergent property.
+We can divide AC approaches into two groups: those which make \emph{explicit}
+use of interestingness, learning from signals which follow a Wundt curve; whilst
+\emph{implicit} approaches modify the \emph{output} of their learning
+algorithm(s), to engineer the overall system behaviour to follow a Wundt curve
+as an emergent property.
 
-A framework encompassing many examples of the explicit approach is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning; for comparison, many similar measures are surveyed in a data mining context in \citep{geng2006interestingness}.
-Many more reinforcement learning examples can be found in \citep{Kaplan2006, Lipson2007, Luciw2011, Macedo2000, Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991, oudeyer2004intelligent}; whilst more general descriptions are given in \citep{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be more amenable to our theory exploration setting.
+A framework encompassing many examples of the explicit approach is given in
+\cite{oudeyer2007intrinsic} in the context of reinforcement learning; for
+comparison, many similar measures are surveyed in a data mining context in
+\cite{geng2006interestingness}. Many more reinforcement learning examples can
+be found in \cite{Kaplan2006, Lipson2007, Luciw2011, Macedo2000,
+  Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991,
+  oudeyer2004intelligent}; whilst more general descriptions are given in
+\cite{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be
+more amenable to our theory exploration setting.
 
-Many of these reward signals are based on information theory, with a prominent example being \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new insights into the structure of past experience can provide a space saving when compressed together.
-This seems particularly relevant for our problem of identifying interesting theorems: those new theorems (``observations'') which shorten the proofs of previously discovered theorems may be more general, more powerful and therefore more \emph{interesting} and hence worth keeping.
-In fact this is very similar to \qspec{}'s interestingness criterion.
+Many of these reward signals are based on information theory, with a prominent
+example being \emph{compression progress}: given a compressed representation of
+our previous observations, the ``progress'' is the space saved if we include the
+current observation. Observations which are incompressible or trivially
+compressible don't save any space, whilst observations which provide new
+insights into the structure of past experience can provide a space saving when
+compressed together. This seems particularly relevant for our problem of
+identifying interesting theorems: those new theorems (``observations'') which
+shorten the proofs of previously discovered theorems may be more general, more
+powerful and therefore more \emph{interesting} and hence worth keeping. In fact
+this is very similar to \qspec{}'s interestingness criterion.
 
-Another example of explicit artificial curiosity is given in \citep{Hester.Stone:2012}, where world states which cause \emph{disagreement} among a population of decision trees (a \emph{random forest} \citep{randomforests}) are considered interesting.
-Since the models make stochastic predictions, the disagreement follows a Wundt curve as the complexity of state transitions increases: for parts of the state space which have been fully learned, the models will agree on accurate predictions; for parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the same average value.
-Whilst the latter predictions may not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the interestingness of states which are too complex.
+Another example of explicit artificial curiosity is given in
+\cite{Hester.Stone:2012}, where world states which cause \emph{disagreement}
+among a population of decision trees (a \emph{random forest}
+\cite{randomforests}) are considered interesting. Since the models make
+stochastic predictions, the disagreement follows a Wundt curve as the complexity
+of state transitions increases: for parts of the state space which have been
+fully learned, the models will agree on accurate predictions; for parts which
+are unlearnable, the models cannot infer any structure, and will converge to
+reporting the same average value. Whilst the latter predictions may not be
+\emph{accurate}, they will be \emph{in agreement}, hence pushing down the
+interestingness of states which are too complex.
 
-Many examples of the implicit case are based on \emph{coevolution}: rewarding one part of the system for exploiting another part, and vice versa.
-In \citep{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on the outcome of actions, and the winner is rewarded at the expense of the loser.
-Due to the risk involved, each algorithm will only bet when it is confident in its prediction, and bets will only be actioned when each algorithm is confident in a \emph{different} outcome.
-The overall behaviour of this system is therefore similar to the explicit measure of disagreement used in the random forest example.
-In terms of theory exploration, such a scheme could be used to find theorems which are \emph{non-obvious}, and hence informative in some way to the user.
+Many examples of the implicit case are based on \emph{coevolution}: rewarding
+one part of the system for exploiting another part, and vice versa. In
+\cite{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on
+the outcome of actions, and the winner is rewarded at the expense of the
+loser. Due to the risk involved, each algorithm will only bet when it is
+confident in its prediction, and bets will only be actioned when each algorithm
+is confident in a \emph{different} outcome. The overall behaviour of this system
+is therefore similar to the explicit measure of disagreement used in the random
+forest example. In terms of theory exploration, such a scheme could be used to
+find theorems which are \emph{non-obvious}, and hence informative in some way to
+the user.
 
-Another case is the ``darwinian brain'' of Fernando et al.
-\citep{fernando2013design1, fernando2013design2}.
-This coevolves a population of problem generators and problem solvers, rewarding the solvers based on their speed, and rewarding the generators based on the \emph{variance} of the solvers' speed.
-This avoids trivial problems (which all solvers can quickly overcome) and complex problems (which no solver can manage), and focuses on those with the most possibility for learning.
-It is easy to imagine such a general architecture being populated by conjecture generators and theorem provers to form a theory exploration system.
+Another case is the ``darwinian brain'' of Fernando et
+al. \cite{fernando2013design1, fernando2013design2}. This coevolves a
+population of problem generators and problem solvers, rewarding the solvers
+based on their speed, and rewarding the generators based on the \emph{variance}
+of the solvers' speed. This avoids trivial problems (which all solvers can
+quickly overcome) and complex problems (which no solver can manage), and focuses
+on those with the most possibility for learning. It is easy to imagine such a
+general architecture being populated by conjecture generators and theorem
+provers to form a theory exploration system.
 
@@
 
-Coevolution is a form of \emph{evolutionary computation}; an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting (there are also precedents for using evolutionary computation in a theorem proving domain \citep{spector2008genetic}).
+Coevolution is a form of \emph{evolutionary computation}; an umbrella term for
+heuristic search algorithms which mimic the process of evolution by natural
+selection among a population of candidate solutions
+\cite{back1997evolutionary}. Whilst \emph{genetic algorithms} are perhaps the
+most well-known instance of evolutionary computation, their use of
+\emph{strings} to represent solutions causes complications when comparing to a
+domain like theory exploration, where recursive structures of unbounded depth
+arise. Thankfully these problems are not insurmountable, for example
+\emph{genetic programming} can operate on tree-structures natively
+\cite{banzhaf1998genetic}, which makes evolutionary computation a useful source
+of ideas for reuse in our theory exploration setting (there are also precedents
+for using evolutionary computation in a theorem proving domain
+\cite{spector2008genetic}).
 
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
+Traditionally, evolutionary approaches assign solutions a \emph{fitness} value,
+using a user-supplied \emph{fitness function}. Fitness should correlate with how
+well a solution solves the user's problem; for example, the fitness of a
+solution to some engineering problem may depend on the estimated materials
+cost. If we frame the task of theory exploration in evolutionary computation
+terms, the fitness function would be our interestingness measure.
 
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\citep{baum2000evolution}).
+Pure exploration (i.e. for its own sake) has been studied in evolutionary
+computation for two main reasons: \emph{artificial life} and \emph{deceptive
+  problems}. The former attempts to gain insight into the nature of life and
+biology through competition over limited resources. Whilst this may have utility
+in resource allocation, e.g. efficient scheduling of a portfolio of ATP
+programs, there is no direct connection to interestingness in theory
+exploration, so we will not consider it further (note that similar
+resource-usage ideas can also be found in the literature on \emph{artificial
+  economies}, e.g. \cite{baum2000evolution}).
 
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.
+On the other hand, work on deceptive problems is highly relevant, as it has lead
+to studying various notions of intrinsic fitness, which are analogous to the
+interestingness measures we want. Deceptive problems are those where
+\textquote{pursuing the objective may prevent the objective from being reached}
+\cite{lehman2011abandoning}, which is caused by the fitness (objective)
+function having many local optima which are easy to find (e.g. by hill
+climbing), but few global optima which are hard to find. Many approaches try to
+avoid deception by augmenting the given fitness function to promote
+\emph{diversity} and \emph{novelty}, such as \emph{niching methods}
+\cite{sareni1998fitness}.
 
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
+One example is \emph{fitness sharing}, which divides up fitness values between
+identical or similar solutions. Say we have a user-provided fitness function
+$f$, and a population containing two identical solutions $s_1$ and $s_2$; hence
+$f(s_1) = f(s_2)$. In a fitness sharing scheme, we interpret fitness as a fixed
+resource, distributed according to $f$; when multiple individuals occupy the
+same point in the solution space, they must \emph{share} the fitness available
+there. We can describe the fitness \emph{allocated} to a solution by augmenting
+$f$, e.g. if we allocate fitness uniformly between identical solutions we get:
 
@@
 
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
+Where $n$ is the population size, $s_i$ is the $i$th solution in the population
+and $\delta$ is the Kronecker delta function. In the example above, assuming
+there are no other copies in the population, then
+$f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$. By sharing in this
+way, the fitness of each solution is balanced against redundancy in the
+population: there may still be many copies of a solution, but only when the
+fitness is high enough to justify all of them.
 
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
+There are many variations on this theme, such as sharing between ``close''
+solutions rather than just identical ones and judging distance based on fitness
+(AKA phenotypically) rather than based on the location in solution space (AKA
+genetically). Yet the underlying principle is always the same: penalise
+duplication in order to promote diversity. This lesson can be carried over to
+our theory exploration context, where a theorem should be considered less
+interesting if it is ``close'' to others which have been found.
 
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
+In a similar way, we can bias our search procedure, rather than our fitness
+function, towards diversity. The search procedure in population-based
+evolutionary algorithms consists of \emph{selecting} one or more individuals
+from the population, e.g. via truncation (select the best $n$ individuals,
+discarding the rest); then \emph{transforming} the selected individuals,
+e.g. via mutation and crossover, to obtain new solutions.
 
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$
+Traditional selection methods are biased towards high fitness individuals (this
+is especially clear for truncation). Alternative schemes have been proposed
+which favour diversity \emph{at the expense of} fitness. For example, the
+fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a
+target fitness $f_t$ uniformly from the interval
+$\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the
+population. An individual $s$ is then selected with fitness closest to $f_t$,
+i.e. $s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$
 
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
+In this way, the fitness function $f$ is used to assign comparable quantities to
+solutions, but it is not treated as the objective; instead, the implicit
+objective is to maintain a diverse population, with individuals spread out
+uniformly in fitness space. This approach seems useful for informing our work in
+theory exploration, as it supports search criteria which \emph{describe}
+solutions, but which we may not want to \emph{optimise}. As a simple example, we
+might distinguish different forms of theorem by measuring how balanced their
+syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced);
+but it would be senseless to \emph{maximise} how far they lean.
 
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
+Once we begin this process of augmenting fitness functions, or abandoning their
+use as objectives, an obvious question arises: what happens if our new function
+contains nothing of the original? This kind of pure exploration scenario leads
+to a variety of ideas for \emph{instrinsic} fitness, such as novelty
+\cite{lehman2011abandoning}, which can lead to learning useful ``stepping
+stones'' even in objective-driven domains. Such intrinsic notions of fitness are
+direct analogues of the interestingness measures we seek for theory exploration.

### 2018-05-17 PhDThesis/litreview.tex
Drag a bunch of stuff into thesis folder

\section{Literature Review}

\input{litreview_mathhistory.tex}
\input{litreview_interestingness.tex}
\input{litreview_theoryformation.tex}
\input{litreview_theoryexploration.tex}
\input{litreview_premiseselection.tex}

### 2018-05-17 PhDThesis/litreview_interestingness.tex
Drag a bunch of stuff into thesis folder

--- 2018-05-14 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
-\usepackage{csvsimple}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
-
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
-
-% For example theory of lists
-\newcommand{\function}{\rightarrow}
-
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
-%\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{intro}
-
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-The precision and recall of three MTE tools are compared
-in~\cite{claessen2013automating}:
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-IsaScheme~\cite{MontanoRivas2011} and
-HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-  comparison only reports prior results for each tool, rather than reproducing
-  them. This makes the numbers less comparable, since each tool was tested with
-  slightly different definitions (e.g. whether or not the exponential and
-  maximum functions were included).} The definitions and ground truths for all
-three were taken from the standard library of the Isabelle proof assistant:
-since the library authors have gone to the effort of stating and proving these
-properties, and including them with every copy of their software, this is a good
-indication that they are interesting. Two sets of definitions were used: a
-theory of natural numbers; and a theory of lists (which we show in
-Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
-two datapoints, each containing only a few definitions and properties (4
-functions and 12 properties, in the natural number example), it is difficult
-to draw general conclusions; yet despite these limitations we believe that this
-is a viable method to performing meaningful, objective analyses and comparisons
-for such tools. It is only with larger data sets that we can generalise the
-measured performance to different, especially \emph{novel}, domains (which,
-after all, is the purpose of MTE tools). In addition, we might be concerned that
-such ``popular'' theories as the natural numbers may be unrepresentative of
-typical usage. For example, we would not expect the library functions in a
-typical verification problem to be as mathematically rich as number theory.
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those meeting the following criteria:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the
-  statement to be proved and the definitions involved, such that the two can be
-  easily and meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item The problem set should ideally contain \emph{every} ``interesting''
-  property involving its included definitions, since non-membership in the
-  ground truth will be treated as being ``uninteresting''. More realistically,
-  we should aim for each definition to have \emph{multiple} properties; rather
-  than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem set are preferable as they give more robust statistics, all
-  else being equal (i.e. when this does not sacrifice quality).
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the theorem statements
-from the definitions they reference. The latter will be used as input to the
-MTE tools, whilst the former constitute the ground truth corpus of properties
-against which to compare the output.
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a statement of commutativity for a \texttt{plus} function, and of
-associativity for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they are logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references to them
-updated to use the remaining definition (e.g. chosen based on lexicographic
-order). In the above example, the \texttt{plus} function would be removed, and
-the commutativity statement updated to reference \texttt{add} instead.
-
-\subsection{Sampling}
-\label{section:sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the entire set of properties (taken from the theorem statements) as
-the ground truth for analysis. However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those properties from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a property involving only a \texttt{times}
-function would not depend on a \texttt{plus} function, even if \texttt{plus}
-occurs in the definition of \texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
-
-Instead, we only allow a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
-
-\subsection{Evaluation}
-\label{section:evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth properties, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-We specifically \emph{ignore} other logical relationships between syntactically
-distinct statements, such as one equation being implied by another. Whilst
-logically sound, second-guessing the ground truth in this way would harm other
-aspects which influence interestingness (for example, a more general statement
-is more widely applicable, but it might also be harder to comprehend).
-
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring. One general question we
-might ask is how a tool's performance scales with respect to the input size (the
-number of given definitions). This is straightforward to measure by varying the
-sample size, but we need some way to combine the results from samples of the
-same size.
-
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
-
-\begin{figure}
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \centering
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {properties}  (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a property};
-    \node (deps)   [process, below=of choose]{List property's dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
-              {Preparation $\S$\ref{section:prep}};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
-              {Sampling $\S$\ref{section:sampling}};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
-              {Evaluation $\S$\ref{section:evaluation}};
-
-  \end{tikzpicture}
-  \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this threshold. This gave a
-total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
-returns. During analysis, duplicate samples (caused by our requirement that
-samples have a non-empty ground truth) were found and discarded, so only 14
-samples of size 1 were used and 30 of size 2.
-
-\subsection{Tons of Inductive Problems}
-\label{sec:tip}
-
-We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
-problem has standalone type and function definitions, making their separation
-trivial; known examples from the software verification and inductive theorem
-proving literature are included, ensuring relevance to those fields; the format
-includes the higher-order functions and inductive datatypes we are interested
-in; it is large enough to pose a challenge to current MTE tools; plus it is
-accompanied by tooling to convert its custom format (an extension of
-SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
-Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-property and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP properties (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which are referenced by at
-least one property (this removes ambiguity about which \emph{definitions} count
-as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
-need to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
-failures were due to timing out, and these occurred for all sample sizes but are
-more common for larger samples (over half the runs for sample sizes 13 and 20
-timed out). Runs which succeeded mostly finished within 30 seconds, generating
-few or no conjectures; those taking longer generated more conjectures, with the
-most being 133 conjectures from a sample of size 19.
-
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth properties.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some properties, such as conditional equations, inequalities, terms
-larger than the search depth and those containing anonymous functions. The
-congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
-of 0.0026.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-  \caption{Contingency table for ground truth properties, combined from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
-  \label{table:recall}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, combined from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
-  \label{table:precision}
-\end{table}
-
-We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (conventional) significance threshold of
-$\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, when designing any mathematical reasoning system we must decide
-on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
-such as ``true'' or ``provable'' include trivial tautologies, while at the same
-time failing to capture the ``almost true'', which can be a valuable trigger for
-theory change, as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
-Colton (such as those in~\cite{colton2000notion}) are based largely on
-``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingingness'', from a
-single quote from a mathematician). Opinions from mathematicians themselves
-include Gowers' suggestion that we can identify features which are commonly
-associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
-Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it makes sense to speak of
-``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although it does emphasise the continued importance of testing these
-tools with real human users, rather than relying solely on artificial benchmarks
-such as ours.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
-
-The automation of mathematical tasks has been pursued since at least the time of
-mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
-these efforts is the separation between those undertaken by mathematicians like
-Pascal and Babbage~\cite{bowden}, and those of engineers such as
-M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
-are concerned with (automatically constructing and evaluating concepts,
-conjectures, theorems, axioms, examples, etc.) being divided into two main
-fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
-sometimes prefaced with ``Computer-Aided'', ``Automated'' or
-``Algorithm-Supported''), which is championed by mathematicians such as
-Buchberger~\cite{buchberger}; and Automated Theory Formation
-(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
-Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
-``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}.
-
-Such a plethora of terminology can mask similarities and shared goals between
-these fields. Even notable historical differences, such as the emphasis of MTE
-on user-interaction and mathematical domains, in contrast to the full automation
-and more general applications targeted by ATF, are disappearing in recent
-implementations.
-
-An important historical implementation of ATF is Lenat's AM (Automated
-Mathematician) system. Unlike prior work, such as
-Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
-to be a general purpose mathematical discovery system, designed to both
-construct new concepts and conjecture relationships between them. AM is a
-rule-based system which represents knowledge using a frame-like scheme, enlarges
-its knowledge base via a collection of heuristic rules, and controls the firing
-of these rules via an agenda mechanism. Evaluation of AM considered generality
-(performance in new domains) and how finely-tuned various aspects of the program
-are (the agenda, the interaction of the heuristics, etc). Most of this
-evaluation was qualitative, and has subsequently been
-criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
-Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{partridge}; for example, AM ``invented''
-natural numbers from sets, but did so using a heuristic specifically designed to
-make this connection.
-
-The prototypical implementation of MTE is the Theorema system of Buchberger and
-colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
-emphasis on user interface and output presentation. Theory exploration in the
-Theorema system involves the user formalising their definitions in a consistent,
-layered approach; such that reasoning algorithms can exploit this structure in
-subsequent proofs, calculations, etc. The potential of this strategy was
-evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
-bases algorithm~\cite{buchberger:04}.
-
-A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which has also been
-quantitatively compared against IsaCoSy and HipSpec using precision/recall
-analysis~\cite{claessen2013automating}. The name comes from its embedding in the
-Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
-can be used to generate new concepts and conjectures. Variables within a scheme
-are instantiated automatically and this drives the invention process. For
-example, the concept of ``repetition'' can be encoded as a scheme, and
-instantiated with existing encodings of zero, successor and addition to produce
-a definition of multiplication. The same scheme can be instantiated with this
-new multiplication function to produce exponentiation.
-
-IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
-described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
-has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
-distinct enumeration and testing steps with a single, iterative algorithm
-similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
-completion algorithm to form a corresponding set of rewrite rules. As
-expressions are enumerated, they are simplified using these rules and discarded
-if equal to a known expression. If not, QuickCheck tests whether the new
-expression can be distinguished from the known expressions through random
-testing: those which can are added to the set of known expressions. Those which
-cannot be distinguished are conjectured to be equal, and the rewrite rules are
-updated.
-
-QuickSpec has also inspired another MTE tool for Haskell called
-Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
-also makes use of the laws of total orders and Boolean algebra to conjecture
-\emph{in}equalities and conditional relations between expressions.
-
-Another notable MTE implementation, distinct from those based in Isabelle and
-Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
-generate \emph{conjectures} that may later be sent to automated provers,
-MATHsAiD directly generates \emph{theorems}, by making logically valid
-inferences from a given set of axioms and definitions. Evaluation of the
-interestingness of these theorems was performed qualitatively by the system's
-developer, which highlights how these tools could benefit from the availability
-of an objective, repeatable, quantitative method of evaluation and comparison
-such as ours.
-
 Whilst there are many reasonably objective benchmarks for mathematical tasks
@@
 over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
-\end{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/litreview_mathhistory.tex
Drag a bunch of stuff into thesis folder

--- 2018-05-16 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
-\usepackage{csvsimple}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
-\usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
-
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
-
-% For example theory of lists
-\newcommand{\function}{\rightarrow}
-
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
-%\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{intro}
-
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems (TIP) benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-\begin{sloppypar}
-  Our benchmark, including the scripts used to generate it from the TIP problem
-  set, is available online at
-  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
-  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
-  \url{https://github.com/warbo/isaplanner-tip}.
-\end{sloppypar}
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-\begin{sloppypar}
-  The precision and recall of three MTE tools are compared
-  in~\cite{claessen2013automating}:
-  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
-  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-    comparison only reports prior results for each tool, rather than reproducing
-    them. This makes the numbers less comparable, since each tool was tested
-    with slightly different definitions (e.g. whether or not the exponential and
-    maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant:
-  since the library authors have gone to the effort of stating and proving these
-  properties, and including them with every copy of their software, this is a good
-  indication that they are interesting. Two sets of definitions were used: a
-  theory of natural numbers; and a theory of lists (which we show in
-  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
-  only two datapoints, each containing only a few definitions and properties (4
-  functions and 12 properties, in the natural number example), it is difficult
-  to draw general conclusions; yet despite these limitations we believe that
-  this is a viable method to performing meaningful, objective analyses and
-  comparisons for such tools. It is only with larger data sets that we can
-  generalise the measured performance to different, especially \emph{novel},
-  domains (which, after all, is the purpose of MTE tools). In addition, we might
-  be concerned that such ``popular'' theories as the natural numbers may be
-  unrepresentative of typical usage. For example, we would not expect the
-  library functions in a typical verification problem to be as mathematically
-  rich as number theory.
-\end{sloppypar}
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those meeting the following criteria:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the
-  statement to be proved and the definitions involved, such that the two can be
-  easily and meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item The problem set should ideally contain \emph{every} ``interesting''
-  property involving its included definitions, since non-membership in the
-  ground truth will be treated as being ``uninteresting''. More realistically,
-  we should aim for each definition to have \emph{multiple} properties; rather
-  than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem set are preferable as they give more robust statistics, all
-  else being equal (i.e. when this does not sacrifice quality).
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the theorem statements
-from the definitions they reference. The latter will be used as input to the
-MTE tools, whilst the former constitute the ground truth corpus of properties
-against which to compare the output.
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a statement of commutativity for a \texttt{plus} function, and of
-associativity for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they are logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references to them
-updated to use the remaining definition (e.g. chosen based on lexicographic
-order). In the above example, the \texttt{plus} function would be removed, and
-the commutativity statement updated to reference \texttt{add} instead.
-
-\subsection{Sampling}
-\label{section:sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the entire set of properties (taken from the theorem statements) as
-the ground truth for analysis. However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those properties from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a property involving only a \texttt{times}
-function would not depend on a \texttt{plus} function, even if \texttt{plus}
-occurs in the definition of \texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
-
-Instead, we only allow a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
-
-\subsection{Evaluation}
-\label{section:evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth properties, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-We specifically \emph{ignore} other logical relationships between syntactically
-distinct statements, such as one equation being implied by another. Whilst
-logically sound, second-guessing the ground truth in this way would harm other
-aspects which influence interestingness (for example, a more general statement
-is more widely applicable, but it might also be harder to comprehend).
-
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring. One general question we
-might ask is how a tool's performance scales with respect to the input size (the
-number of given definitions). This is straightforward to measure by varying the
-sample size, but we need some way to combine the results from samples of the
-same size.
-
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
-
-\begin{figure}
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \centering
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {properties}  (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a property};
-    \node (deps)   [process, below=of choose]{List property's dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
-              {Preparation $\S$\ref{section:prep}};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
-              {Sampling $\S$\ref{section:sampling}};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
-              {Evaluation $\S$\ref{section:evaluation}};
-
-  \end{tikzpicture}
-  \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this threshold. This gave a
-total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
-returns. During analysis, duplicate samples (caused by our requirement that
-samples have a non-empty ground truth) were found and discarded, so only 14
-samples of size 1 were used and 30 of size 2.
-
-\subsection{Tons of Inductive Problems}
-\label{sec:tip}
-
-We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
-problem has standalone type and function definitions, making their separation
-trivial; known examples from the software verification and inductive theorem
-proving literature are included, ensuring relevance to those fields; the format
-includes the higher-order functions and inductive datatypes we are interested
-in; it is large enough to pose a challenge to current MTE tools; plus it is
-accompanied by tooling to convert its custom format (an extension of
-SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
-Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-property and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP properties (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which are referenced by at
-least one property (this removes ambiguity about which \emph{definitions} count
-as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
-need to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
-failures were due to timing out, and these occurred for all sample sizes but are
-more common for larger samples (over half the runs for sample sizes 13 and 20
-timed out). Runs which succeeded mostly finished within 30 seconds, generating
-few or no conjectures; those taking longer generated more conjectures, with the
-most being 133 conjectures from a sample of size 19.
-
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth properties.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some properties, such as conditional equations, inequalities, terms
-larger than the search depth and those containing anonymous functions. The
-congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
-of 0.0026.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-  \caption{Contingency table for ground truth properties, combined from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
-  \label{table:recall}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, combined from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
-  \label{table:precision}
-\end{table}
-
-We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (conventional) significance threshold of
-$\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, when designing any mathematical reasoning system we must decide
-on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
-such as ``true'' or ``provable'' include trivial tautologies, while at the same
-time failing to capture the ``almost true'', which can be a valuable trigger for
-theory change, as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
-Colton (such as those in~\cite{colton2000notion}) are based largely on
-``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingingness'', from a
-single quote from a mathematician). Opinions from mathematicians themselves
-include Gowers' suggestion that we can identify features which are commonly
-associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
-Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it makes sense to speak of
-``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although it does emphasise the continued importance of testing these
-tools with real human users, rather than relying solely on artificial benchmarks
-such as ours.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
+%% FIXME: This is copypasta from benchmark paper
 
@@
 implementations.
-
-An important historical implementation of ATF is Lenat's AM (Automated
-Mathematician) system. Unlike prior work, such as
-Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
-to be a general purpose mathematical discovery system, designed to both
-construct new concepts and conjecture relationships between them. AM is a
-rule-based system which represents knowledge using a frame-like scheme, enlarges
-its knowledge base via a collection of heuristic rules, and controls the firing
-of these rules via an agenda mechanism. Evaluation of AM considered generality
-(performance in new domains) and how finely-tuned various aspects of the program
-are (the agenda, the interaction of the heuristics, etc). Most of this
-evaluation was qualitative, and has subsequently been
-criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
-Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
-natural numbers from sets, but did so using a heuristic specifically designed to
-make this connection.
-
-The prototypical implementation of MTE is the Theorema system of Buchberger and
-colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
-emphasis on user interface and output presentation. Theory exploration in the
-Theorema system involves the user formalising their definitions in a consistent,
-layered approach; such that reasoning algorithms can exploit this structure in
-subsequent proofs, calculations, etc. The potential of this strategy was
-evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
-bases algorithm~\cite{buchberger:04}.
-
-A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
-has also been quantitatively compared against IsaCoSy and HipSpec using
-precision/recall analysis~\cite{claessen2013automating}. The name comes from its
-embedding in the Isabelle proof assistant and its use of ``schemes'':
-higher-order formulae which can be used to generate new concepts and
-conjectures. Variables within a scheme are instantiated automatically and this
-drives the invention process. For example, the concept of ``repetition'' can be
-encoded as a scheme, and instantiated with existing encodings of zero, successor
-and addition to produce a definition of multiplication. The same scheme can be
-instantiated with this new multiplication function to produce exponentiation.
-
-IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
-described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
-has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
-distinct enumeration and testing steps with a single, iterative algorithm
-similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
-completion algorithm to form a corresponding set of rewrite rules. As
-expressions are enumerated, they are simplified using these rules and discarded
-if equal to a known expression. If not, QuickCheck tests whether the new
-expression can be distinguished from the known expressions through random
-testing: those which can are added to the set of known expressions. Those which
-cannot be distinguished are conjectured to be equal, and the rewrite rules are
-updated.
-
-QuickSpec has also inspired another MTE tool for Haskell called
-Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
-also makes use of the laws of total orders and Boolean algebra to conjecture
-\emph{in}equalities and conditional relations between expressions.
-
-Another notable MTE implementation, distinct from those based in Isabelle and
-Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
-generate \emph{conjectures} that may later be sent to automated provers,
-MATHsAiD directly generates \emph{theorems}, by making logically valid
-inferences from a given set of axioms and definitions. Evaluation of the
-interestingness of these theorems was performed qualitatively by the system's
-developer, which highlights how these tools could benefit from the availability
-of an objective, repeatable, quantitative method of evaluation and comparison
-such as ours.
-
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
-\end{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/litreview_premiseselection.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-09 TransferReport/related.tex
@@
-\section{Related Work}
-\label{sec:related}
-
-\subsection{Haskell}
-\label{sec:haskelldesc}
-
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little consideration was given for why we chose this language rather than, for example Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more widespread language like Java.
-Here we discuss the relevant language features from a high-level, which motivated our choice:
-
-\begin{description}
-
-\item{Functional}: All control flow in Haskell is performed by function abstraction and application, which we can reason about using standard rules of inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation of expressions; hence our reasoning can safely ignore complicated external and non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can be used to eliminate unwanted combinations of values, and hence reduce search spaces; \emph{static} types can be deduced syntactically, without having to execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an expression (i.e.
-performing function calls) without diverging, then a non-strict evaluation strategy will not diverge when evaluating that expression.
-This is rather technical, but in simple terms it allows us to reason effectively about a Turing-complete language, where evaluation may not terminate.
-For example, when reasoning about \emph{pairs} of values \hs{(x, y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an ``obvious'' rule such as $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$.
-Haskell's non-strict semantics makes this equation valid; whilst it would \emph{not} be valid in the strict setting common to most other languages, where the expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up user-defined data representations, and an inverse mechanism to inspect these data by \emph{pattern-matching}.
-For our purposes, the useful consequences of ADTs and pattern-matching include their amenability for inductive proofs and the fact they are \emph{closed}; i.e.
-an ADT's declaration specifies all of the normal forms for that type.
-This makes exhaustive case analysis trivial, which would be impossible for \emph{open} types (for example, consider classes in an object oriented language, where new subclasses can be introduced at any time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over \emph{type-level} objects; provided those objects are never inspected.
-This has the \emph{practical} benefit of enabling \emph{polymorphism}: for example, we can write a polymorphic identity function \hs{id :: forall t.
-t -> t}.
-\footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually, this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of type \hs{t}; yet only the latter is available in the function body, e.g.
-\hs{id x = x}.
-This inability to inspect type-level arguments gives us the \emph{theoretical} benefit of being able to characterise the behaviour of polymorphic functions from their type alone, a technique known as \emph{theorems for free} \citep{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are interfaces which specify a set of operations over a type (or other type-level object, such as a \emph{type constructor}).
-Many type classes also specify a set of \emph{laws} which their operations should obey but, lacking a simple mechanism to enforce this, laws are usually considered as documentation.
-As a simple example, we can define a type class \hs{Semigroup} with the following operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which restricts the possible types \hs{t} to only those which implement \hs{Semigroup}.
-\footnote{Alternatively, we can consider \hs{Semigroup t} as the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t} and two values of type \hs{t}.
-As with parameteric polymorphism, this extra \hs{Semigroup t} parameter is not available at the value level.
-Even if it were, we could not alter our behaviour by inspecting it, since Haskell only allows types to implement each type class in at most one way, so there would be no information to branch on.} There are many \emph{instances} of \hs{Semigroup} (types which may be substituted for \hs{t}), e.g.
-\hs{Integer} with \hs{op} performing addition.
-Many more examples can be found in the \emph{typeclassopedia} \citep{yorgey2009typeclassopedia}.
-This ability to constrain types, and the existence of laws, helps us reason about code generically, rather than repeating the same arguments for each particular pair of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions; at the type level, for coercions; at the documentation level, for typeclass laws; and at the compiler level, for ad-hoc rewrite rules.
-This provides us with many \emph{sources} of equations, as well as many possible \emph{uses} for any equations we might discover.
-Along with their support in existing tools such as SMT solvers, this makes equational conjectures a natural target for our investigation.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an \emph{export list} containing the names which should be made available for other modules to import.
-When such a list is given, any expressions \emph{not} on the list are considered \emph{private} to that module, and are hence inaccessible from elsewhere.
-This mechanism allows modules to provide more guarantees than are available just in their types.
-For example, a module may represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a host part (modulo divergence), but it does not provide any guarantees about those parts.
-We also define the \hs{at} function, a so-called ``smart constructor'', which has the additional guarantee that the \hs{Email}s it returns contain non-empty \hs{String}s.
-By ommitting the \hs{E} constructor from the export list on the first line \footnote{The syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any of its constructors.}, the only way \emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws.
-However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}).
-This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g.
-in the style of interactive theorem proving, as these results may be invalidated by seemingly innocuous code changes.
-This puts Haskell in a rather special position with regards to the discovery of interesting theorems.
-Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development.
-The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.
-
-Other empirical advantages to studying Haskell, compared to either other programming languages or theorem proving systems, include:
-
-\begin{itemize}
-
-\item The large amount of Haskell code which is freely available online, e.g. in repositories like \href{http://hackage.haskell.org}{Hackage}, with which we can experiment.
-
-\item The existence of conjecture generation and theory exploration systems for Haskell, such as \qspec{} and \hspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and \textsc{SmartCheck}; theorem provers like \textsc{Hip} \citep{rosen2012proving}; and other related testing and term-generating systems like \textsc{MuCheck} \citep{le2014mucheck}, \textsc{MagicHaskeller} \citep{katayama2011magichaskeller} and \textsc{Djinn} \citep{augustsson2005djinn}.
-
-\item The remarkable amount of infrastructure which exists for working with Haskell code, including package managers, compilers, interpreters, parsers, static analysers, etc.
-
-\end{itemize}
-
 \subsection{Relevance Filtering}
@@
 
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \citep{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering} (or, \emph{premise selection}, when viewed as the \emph{addition} of relevant information to an initially-empty problem).
-This is the core idea behind our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time.
+The combinatorial nature of formal systems causes many proof search methods,
+such as resolution, to have exponential complexity
+\cite{haken1985intractability}; hence even a modest size increase can turn a
+trivial problem into an intractable one. Finding efficient alternatives for such
+algorithms, especially those which are NP-complete (e.g. determining
+satisfiability) or co-NP-complete (e.g. determining tautologies), seems
+unlikely, as it would imply progress on the famously intractable open problems
+of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$. On the other hand, we
+can turn this difficulty around: a modest \emph{decrease} in size may turn an
+intractable problem into a solvable one. We can ensure that the solutions to
+these reduced problems coincide with the original if we only remove
+\emph{redundant} information. This leads to the idea of \emph{relevance
+  filtering} (or, \emph{premise selection}, when viewed as the \emph{addition}
+of relevant information to an initially-empty problem). This is the core idea
+behind our restriction of theory exploration to intelligently-selected clusters
+of symbols, rather than whole libraries at a time.
 
-Relevance filtering has mostly been used in automated proof search, where it simplifies problems by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
+Relevance filtering has mostly been used in automated proof search, where it
+simplifies problems by removing from consideration those clauses (axioms,
+definitions, lemmas, etc.) which are deemed \emph{irrelevant}. The technique is
+used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL
+theories to statements in first order logic: rather than translating the entire
+theory, only a sub-set of relevant clauses are included. This reduces the size
+of the problem and speeds up the proof search, but it creates the new problem of
+determining when a clause is relevant: how do we know what will be required,
+before we have the proof?
 
-The initial approach taken by Sledgehammer, known as \textsc{MePO} \iffalse TODO: Check name \fi (from \emph{Meng-Paulson} \citep{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
+The initial approach taken by Sledgehammer, known as \textsc{MePo} (from
+\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score
+based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant''
+(where $n$ is the number of symbols in the clause and $m$ is the number which
+are relevant). Initially, the relevant symbols are those which occur in the goal
+to be proved, but whenever a clause is found which scores more than a particular
+threshold, all of its symbols are then also considered relevant. There are other
+heuristics applied too, such as increasing the score of user-provided facts
+(e.g. given by keywords like \texttt{using}), locally-scoped facts, first-order
+facts and rarely-occuring facts. To choose $r$ relevant clauses for an ATP
+invocation, we simply order the clauses by decreasing score and take the first
+$r$ of them.
 
-Recently, a variety of alternative algorithms have also been investigated, for example, the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses.
-Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features).
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
+Recently, a variety of alternative algorithms have also been investigated, for
+example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer)
+\cite{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to
+determine the relevance of clauses. Visibility is essentially a dependency graph
+of which theorems were used in the proofs of which other theorems (although the
+theorems are actually represented as abstract sets of features). To select
+relevant clauses for a goal, the set of clauses which are visible from the
+goal's components is generated; this is further reduced by (an efficient
+approximation of) a na\"{\i}ve Bayes algorithm.
 
-Another example is \emph{multi-output ranking} (MOR), which uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \citep{alama2014premise}.
-Many more approaches are described and evaluated in \citep{kuhlwein2012overview}.
-
-\subsection{Recurrent Clustering}
-\label{sec:clusteringexpressions}
-
-Our recurrent clustering approach takes inspiration from the ML4PG \citep{journals/corr/abs-1212-3618} and ACL2(ml) \citep{heras2013proof} tools, used for analysing proofs in Coq and ACL2, respectively.
-Whilst both transform syntax trees into matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} information about the tree; for example, one column counts the number of variables appearing at each tree level, others count the number of function symbols which are nullary, unary, binary, etc.
-
-However, the way we \emph{use} our clusters to inform theory exploration is similar to that of ACL2(ml).
-In ACL2(ml), the clusters are used to reason by analogy: finding theorem statements which are similar to the current goal being proved.
-Once a similar theorem is found, the lemma statements used in its proof are mutated, which includes substituting symbols for those which appear in the same cluster.
-For example, if we are trying to prove a theorem \texttt{multiply\_commutes}, we might find it appears in the same cluster as \texttt{plus\_commutes} which was proved using the lemma \texttt{plus\_identity} relating \texttt{plus} and \texttt{0}.
-In this case ACL2(ml) might try mutating \texttt{plus\_identity}, for example replacing \texttt{plus} by \texttt{multiply} and \texttt{0} by \texttt{1} (assuming those appear in the same clusters); the resulting lemma (which we might call \texttt{multiply\_identity}) would be given to ACL2 to try and prove \texttt{multiply\_commutes}.
-
-Whilst we do not currently reason by analogy, this is an interesting area for future work in theory exploration: given a set of theorems relating particular terms, we might form conjectures regarding similar terms found through clustering.
-
-\iffalse
-We could expand this a bit, e.g. talking about how we both use Weka, etc.
-\fi
-
-\subsection{Feature Extraction}
-
-One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
-This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways.
-The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.
-
-The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger.
-We use this to make our matrices a uniform size, borrowing the idea from ML4PG.
-Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors.
-In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.
-
-Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian'').
-This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness.
-This allows us to truncate more aggressively than if the leaves were most significant.
-
-By modelling our inputs as points in high-dimensional spaces, we can consider feature extraction as projection into a lower-dimensional space (known as \emph{dimension reduction}).
-Truncation is a trivial dimension reduction technique; more sophisticated projection functions consider the \emph{distribution} of the input points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-Techniques such as \emph{principle component analysis} (PCA) can be used to find these hyperplanes, but unfortunately require their inputs to already have a fixed, integer number of dimensions.
-In the case of our recursive expressions (which we may consider to have \emph{fractal} dimension), we would need another pre-processing stage to satisfy this requirement.
-
-There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise.
-Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.
-
-One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs).
-These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations.
-In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.
-
-Unfortunately RNNs are difficult to train.
-The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).
-
-Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful.
-The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs.
-Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}.
-More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.
-
-A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}.
-Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.
-
-\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning.
-Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.
+Another example is \emph{multi-output ranking} (MOR), which uses a support
+vector machine (SVM) approach for selecting relevant axioms from the Mizar
+Mathematical Library for use by the Vampire ATP system
+\cite{alama2014premise}. Many more approaches are described and evaluated in
+\cite{kuhlwein2012overview}, some of which may be directly applicable in the
+context of theory exploration.

### 2018-05-17 PhDThesis/litreview_theoryexploration.tex
Drag a bunch of stuff into thesis folder

--- 2018-05-16 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
-\usepackage{csvsimple}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
-\usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
-
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
-
-% For example theory of lists
-\newcommand{\function}{\rightarrow}
-
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
-%\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{intro}
-
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems (TIP) benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-\begin{sloppypar}
-  Our benchmark, including the scripts used to generate it from the TIP problem
-  set, is available online at
-  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
-  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
-  \url{https://github.com/warbo/isaplanner-tip}.
-\end{sloppypar}
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-\begin{sloppypar}
-  The precision and recall of three MTE tools are compared
-  in~\cite{claessen2013automating}:
-  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
-  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-    comparison only reports prior results for each tool, rather than reproducing
-    them. This makes the numbers less comparable, since each tool was tested
-    with slightly different definitions (e.g. whether or not the exponential and
-    maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant:
-  since the library authors have gone to the effort of stating and proving these
-  properties, and including them with every copy of their software, this is a good
-  indication that they are interesting. Two sets of definitions were used: a
-  theory of natural numbers; and a theory of lists (which we show in
-  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
-  only two datapoints, each containing only a few definitions and properties (4
-  functions and 12 properties, in the natural number example), it is difficult
-  to draw general conclusions; yet despite these limitations we believe that
-  this is a viable method to performing meaningful, objective analyses and
-  comparisons for such tools. It is only with larger data sets that we can
-  generalise the measured performance to different, especially \emph{novel},
-  domains (which, after all, is the purpose of MTE tools). In addition, we might
-  be concerned that such ``popular'' theories as the natural numbers may be
-  unrepresentative of typical usage. For example, we would not expect the
-  library functions in a typical verification problem to be as mathematically
-  rich as number theory.
-\end{sloppypar}
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those meeting the following criteria:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the
-  statement to be proved and the definitions involved, such that the two can be
-  easily and meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item The problem set should ideally contain \emph{every} ``interesting''
-  property involving its included definitions, since non-membership in the
-  ground truth will be treated as being ``uninteresting''. More realistically,
-  we should aim for each definition to have \emph{multiple} properties; rather
-  than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem set are preferable as they give more robust statistics, all
-  else being equal (i.e. when this does not sacrifice quality).
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the theorem statements
-from the definitions they reference. The latter will be used as input to the
-MTE tools, whilst the former constitute the ground truth corpus of properties
-against which to compare the output.
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a statement of commutativity for a \texttt{plus} function, and of
-associativity for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they are logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references to them
-updated to use the remaining definition (e.g. chosen based on lexicographic
-order). In the above example, the \texttt{plus} function would be removed, and
-the commutativity statement updated to reference \texttt{add} instead.
-
-\subsection{Sampling}
-\label{section:sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the entire set of properties (taken from the theorem statements) as
-the ground truth for analysis. However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those properties from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a property involving only a \texttt{times}
-function would not depend on a \texttt{plus} function, even if \texttt{plus}
-occurs in the definition of \texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
-
-Instead, we only allow a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
-
-\subsection{Evaluation}
-\label{section:evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth properties, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-We specifically \emph{ignore} other logical relationships between syntactically
-distinct statements, such as one equation being implied by another. Whilst
-logically sound, second-guessing the ground truth in this way would harm other
-aspects which influence interestingness (for example, a more general statement
-is more widely applicable, but it might also be harder to comprehend).
-
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring. One general question we
-might ask is how a tool's performance scales with respect to the input size (the
-number of given definitions). This is straightforward to measure by varying the
-sample size, but we need some way to combine the results from samples of the
-same size.
-
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
-
-\begin{figure}
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \centering
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {properties}  (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a property};
-    \node (deps)   [process, below=of choose]{List property's dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
-              {Preparation $\S$\ref{section:prep}};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
-              {Sampling $\S$\ref{section:sampling}};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
-              {Evaluation $\S$\ref{section:evaluation}};
-
-  \end{tikzpicture}
-  \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this threshold. This gave a
-total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
-returns. During analysis, duplicate samples (caused by our requirement that
-samples have a non-empty ground truth) were found and discarded, so only 14
-samples of size 1 were used and 30 of size 2.
-
-\subsection{Tons of Inductive Problems}
-\label{sec:tip}
-
-We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
-problem has standalone type and function definitions, making their separation
-trivial; known examples from the software verification and inductive theorem
-proving literature are included, ensuring relevance to those fields; the format
-includes the higher-order functions and inductive datatypes we are interested
-in; it is large enough to pose a challenge to current MTE tools; plus it is
-accompanied by tooling to convert its custom format (an extension of
-SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
-Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-property and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP properties (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which are referenced by at
-least one property (this removes ambiguity about which \emph{definitions} count
-as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
-need to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
-failures were due to timing out, and these occurred for all sample sizes but are
-more common for larger samples (over half the runs for sample sizes 13 and 20
-timed out). Runs which succeeded mostly finished within 30 seconds, generating
-few or no conjectures; those taking longer generated more conjectures, with the
-most being 133 conjectures from a sample of size 19.
-
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth properties.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some properties, such as conditional equations, inequalities, terms
-larger than the search depth and those containing anonymous functions. The
-congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
-of 0.0026.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-  \caption{Contingency table for ground truth properties, combined from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
-  \label{table:recall}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, combined from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
-  \label{table:precision}
-\end{table}
-
-We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (conventional) significance threshold of
-$\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, when designing any mathematical reasoning system we must decide
-on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
-such as ``true'' or ``provable'' include trivial tautologies, while at the same
-time failing to capture the ``almost true'', which can be a valuable trigger for
-theory change, as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
-Colton (such as those in~\cite{colton2000notion}) are based largely on
-``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingingness'', from a
-single quote from a mathematician). Opinions from mathematicians themselves
-include Gowers' suggestion that we can identify features which are commonly
-associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
-Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it makes sense to speak of
-``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although it does emphasise the continued importance of testing these
-tools with real human users, rather than relying solely on artificial benchmarks
-such as ours.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
-
-The automation of mathematical tasks has been pursued since at least the time of
-mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
-these efforts is the separation between those undertaken by mathematicians like
-Pascal and Babbage~\cite{bowden}, and those of engineers such as
-M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
-are concerned with (automatically constructing and evaluating concepts,
-conjectures, theorems, axioms, examples, etc.) being divided into two main
-fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
-sometimes prefaced with ``Computer-Aided'', ``Automated'' or
-``Algorithm-Supported''), which is championed by mathematicians such as
-Buchberger~\cite{buchberger}; and Automated Theory Formation
-(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
-Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
-``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}.
-
-Such a plethora of terminology can mask similarities and shared goals between
-these fields. Even notable historical differences, such as the emphasis of MTE
-on user-interaction and mathematical domains, in contrast to the full automation
-and more general applications targeted by ATF, are disappearing in recent
-implementations.
-
-An important historical implementation of ATF is Lenat's AM (Automated
-Mathematician) system. Unlike prior work, such as
-Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
-to be a general purpose mathematical discovery system, designed to both
-construct new concepts and conjecture relationships between them. AM is a
-rule-based system which represents knowledge using a frame-like scheme, enlarges
-its knowledge base via a collection of heuristic rules, and controls the firing
-of these rules via an agenda mechanism. Evaluation of AM considered generality
-(performance in new domains) and how finely-tuned various aspects of the program
-are (the agenda, the interaction of the heuristics, etc). Most of this
-evaluation was qualitative, and has subsequently been
-criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
-Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
-natural numbers from sets, but did so using a heuristic specifically designed to
-make this connection.
-
 The prototypical implementation of MTE is the Theorema system of Buchberger and
@@
 such as ours.
-
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
-\end{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/litreview_theoryformation.tex
Drag a bunch of stuff into thesis folder

--- 2018-05-16 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
-\usepackage{csvsimple}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
-\usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
-
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
-
-% For example theory of lists
-\newcommand{\function}{\rightarrow}
-
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
-%\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{intro}
-
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems (TIP) benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-\begin{sloppypar}
-  Our benchmark, including the scripts used to generate it from the TIP problem
-  set, is available online at
-  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
-  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
-  \url{https://github.com/warbo/isaplanner-tip}.
-\end{sloppypar}
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-\begin{sloppypar}
-  The precision and recall of three MTE tools are compared
-  in~\cite{claessen2013automating}:
-  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
-  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-    comparison only reports prior results for each tool, rather than reproducing
-    them. This makes the numbers less comparable, since each tool was tested
-    with slightly different definitions (e.g. whether or not the exponential and
-    maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant:
-  since the library authors have gone to the effort of stating and proving these
-  properties, and including them with every copy of their software, this is a good
-  indication that they are interesting. Two sets of definitions were used: a
-  theory of natural numbers; and a theory of lists (which we show in
-  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
-  only two datapoints, each containing only a few definitions and properties (4
-  functions and 12 properties, in the natural number example), it is difficult
-  to draw general conclusions; yet despite these limitations we believe that
-  this is a viable method to performing meaningful, objective analyses and
-  comparisons for such tools. It is only with larger data sets that we can
-  generalise the measured performance to different, especially \emph{novel},
-  domains (which, after all, is the purpose of MTE tools). In addition, we might
-  be concerned that such ``popular'' theories as the natural numbers may be
-  unrepresentative of typical usage. For example, we would not expect the
-  library functions in a typical verification problem to be as mathematically
-  rich as number theory.
-\end{sloppypar}
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those meeting the following criteria:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the
-  statement to be proved and the definitions involved, such that the two can be
-  easily and meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item The problem set should ideally contain \emph{every} ``interesting''
-  property involving its included definitions, since non-membership in the
-  ground truth will be treated as being ``uninteresting''. More realistically,
-  we should aim for each definition to have \emph{multiple} properties; rather
-  than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem set are preferable as they give more robust statistics, all
-  else being equal (i.e. when this does not sacrifice quality).
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the theorem statements
-from the definitions they reference. The latter will be used as input to the
-MTE tools, whilst the former constitute the ground truth corpus of properties
-against which to compare the output.
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a statement of commutativity for a \texttt{plus} function, and of
-associativity for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they are logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references to them
-updated to use the remaining definition (e.g. chosen based on lexicographic
-order). In the above example, the \texttt{plus} function would be removed, and
-the commutativity statement updated to reference \texttt{add} instead.
-
-\subsection{Sampling}
-\label{section:sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the entire set of properties (taken from the theorem statements) as
-the ground truth for analysis. However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those properties from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a property involving only a \texttt{times}
-function would not depend on a \texttt{plus} function, even if \texttt{plus}
-occurs in the definition of \texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
-
-Instead, we only allow a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
-
-\subsection{Evaluation}
-\label{section:evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes).
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth properties, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-We specifically \emph{ignore} other logical relationships between syntactically
-distinct statements, such as one equation being implied by another. Whilst
-logically sound, second-guessing the ground truth in this way would harm other
-aspects which influence interestingness (for example, a more general statement
-is more widely applicable, but it might also be harder to comprehend).
-
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring. One general question we
-might ask is how a tool's performance scales with respect to the input size (the
-number of given definitions). This is straightforward to measure by varying the
-sample size, but we need some way to combine the results from samples of the
-same size.
-
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
-
-\begin{figure}
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \centering
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {properties}  (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a property};
-    \node (deps)   [process, below=of choose]{List property's dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
-              {Preparation $\S$\ref{section:prep}};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
-              {Sampling $\S$\ref{section:sampling}};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
-              {Evaluation $\S$\ref{section:evaluation}};
-
-  \end{tikzpicture}
-  \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this threshold. This gave a
-total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
-returns. During analysis, duplicate samples (caused by our requirement that
-samples have a non-empty ground truth) were found and discarded, so only 14
-samples of size 1 were used and 30 of size 2.
-
-\subsection{Tons of Inductive Problems}
-\label{sec:tip}
-
-We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
-problem has standalone type and function definitions, making their separation
-trivial; known examples from the software verification and inductive theorem
-proving literature are included, ensuring relevance to those fields; the format
-includes the higher-order functions and inductive datatypes we are interested
-in; it is large enough to pose a challenge to current MTE tools; plus it is
-accompanied by tooling to convert its custom format (an extension of
-SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
-Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-property and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP properties (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which are referenced by at
-least one property (this removes ambiguity about which \emph{definitions} count
-as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
-need to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
-failures were due to timing out, and these occurred for all sample sizes but are
-more common for larger samples (over half the runs for sample sizes 13 and 20
-timed out). Runs which succeeded mostly finished within 30 seconds, generating
-few or no conjectures; those taking longer generated more conjectures, with the
-most being 133 conjectures from a sample of size 19.
-
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth properties.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some properties, such as conditional equations, inequalities, terms
-larger than the search depth and those containing anonymous functions. The
-congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
-    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
-    considered statistically significant due to having less than 30 samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference.}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
-of 0.0026.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-  \caption{Contingency table for ground truth properties, combined from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
-  \label{table:recall}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, combined from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
-  \label{table:precision}
-\end{table}
-
-We test these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (conventional) significance threshold of
-$\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, when designing any mathematical reasoning system we must decide
-on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
-such as ``true'' or ``provable'' include trivial tautologies, while at the same
-time failing to capture the ``almost true'', which can be a valuable trigger for
-theory change, as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
-Colton (such as those in~\cite{colton2000notion}) are based largely on
-``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingingness'', from a
-single quote from a mathematician). Opinions from mathematicians themselves
-include Gowers' suggestion that we can identify features which are commonly
-associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
-Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it makes sense to speak of
-``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although it does emphasise the continued importance of testing these
-tools with real human users, rather than relying solely on artificial benchmarks
-such as ours.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
-
-The automation of mathematical tasks has been pursued since at least the time of
-mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
-these efforts is the separation between those undertaken by mathematicians like
-Pascal and Babbage~\cite{bowden}, and those of engineers such as
-M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
-are concerned with (automatically constructing and evaluating concepts,
-conjectures, theorems, axioms, examples, etc.) being divided into two main
-fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
-sometimes prefaced with ``Computer-Aided'', ``Automated'' or
-``Algorithm-Supported''), which is championed by mathematicians such as
-Buchberger~\cite{buchberger}; and Automated Theory Formation
-(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
-Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
-``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}.
-
-Such a plethora of terminology can mask similarities and shared goals between
-these fields. Even notable historical differences, such as the emphasis of MTE
-on user-interaction and mathematical domains, in contrast to the full automation
-and more general applications targeted by ATF, are disappearing in recent
-implementations.
-
 An important historical implementation of ATF is Lenat's AM (Automated
@@
 make this connection.
-
-The prototypical implementation of MTE is the Theorema system of Buchberger and
-colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
-emphasis on user interface and output presentation. Theory exploration in the
-Theorema system involves the user formalising their definitions in a consistent,
-layered approach; such that reasoning algorithms can exploit this structure in
-subsequent proofs, calculations, etc. The potential of this strategy was
-evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
-bases algorithm~\cite{buchberger:04}.
-
-A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
-has also been quantitatively compared against IsaCoSy and HipSpec using
-precision/recall analysis~\cite{claessen2013automating}. The name comes from its
-embedding in the Isabelle proof assistant and its use of ``schemes'':
-higher-order formulae which can be used to generate new concepts and
-conjectures. Variables within a scheme are instantiated automatically and this
-drives the invention process. For example, the concept of ``repetition'' can be
-encoded as a scheme, and instantiated with existing encodings of zero, successor
-and addition to produce a definition of multiplication. The same scheme can be
-instantiated with this new multiplication function to produce exponentiation.
-
-IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
-described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
-has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
-distinct enumeration and testing steps with a single, iterative algorithm
-similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
-completion algorithm to form a corresponding set of rewrite rules. As
-expressions are enumerated, they are simplified using these rules and discarded
-if equal to a known expression. If not, QuickCheck tests whether the new
-expression can be distinguished from the known expressions through random
-testing: those which can are added to the set of known expressions. Those which
-cannot be distinguished are conjectured to be equal, and the rewrite rules are
-updated.
-
-QuickSpec has also inspired another MTE tool for Haskell called
-Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
-also makes use of the laws of total orders and Boolean algebra to conjecture
-\emph{in}equalities and conditional relations between expressions.
-
-Another notable MTE implementation, distinct from those based in Isabelle and
-Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
-generate \emph{conjectures} that may later be sent to automated provers,
-MATHsAiD directly generates \emph{theorems}, by making logically valid
-inferences from a given set of axioms and definitions. Evaluation of the
-interestingness of these theorems was performed qualitatively by the system's
-developer, which highlights how these tools could benefit from the availability
-of an objective, repeatable, quantitative method of evaluation and comparison
-such as ours.
-
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
-\end{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/ml4hs.tex
Drag a bunch of stuff into thesis folder

--- 2015-12-15 TransferReport/implementation.tex
@@
-\section{Implementation}
+\section{ML4HS Implementation}
 \label{sec:implementation}
 
-We provide an implementation of our recurrent clustering algorithm in a tool called \textsc{ML4HS}, which consists of a loose collection of components shown in Figure \ref{fig:ml4hs}.
-This arrangement makes it easy to swap out parts for experimentation.
-In the following, we describe the custom components in the order they appear in the diagram.
+We provide an implementation of our recurrent clustering algorithm in a tool
+called \textsc{ML4HS}, which consists of a loose collection of components shown
+in Figure \ref{fig:ml4hs}. This arrangement makes it easy to swap out parts for
+experimentation. In the following, we describe the custom components in the
+order they appear in the diagram.
 
@@
 
-The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core.
-It also includes a \emph{renaming} transformation, which resolves global identifiers into a canonical form.
-This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.
+The GHC compiler provides mechanisms for parsing Haskell source code and
+converting it to Core. It also includes a \emph{renaming} transformation, which
+resolves global identifiers into a canonical form. This allows us to spot
+repeated use of a term, across multiple modules and packages, with a simple
+syntactic equality check.
 
-Since we are interested in comparing definitions based on the terms they reference, building our framework on top of GHC seems like a promising approach.
-Indeed, \hspec{} already invokes GHC's API to obtain the definitions of Haskell functions, in order to transform them into a form suitable for ATP systems.
-However, our initial experiments showed that this technique is too fragile for use on many real Haskell projects.
+Since we are interested in comparing definitions based on the terms they
+reference, building our framework on top of GHC seems like a promising
+approach. Indeed, \hspec{} already invokes GHC's API to obtain the definitions
+of Haskell functions, in order to transform them into a form suitable for ATP
+systems. However, our initial experiments showed that this technique is too
+fragile for use on many real Haskell projects.
 
-This is due to many projects having a complex module structure, requiring particular GHC flags to be given, or using pre-processors such as \texttt{cpp} and Template Haskell to generate parts of their code.
-All of this complexity means that invoking GHC ``manually'' via its API is unlikely to obtain the definitions we require.
+This is due to many projects having a complex module structure, requiring
+particular GHC flags to be given, or using pre-processors such as \texttt{cpp}
+and Template Haskell to generate parts of their code. All of this complexity
+means that invoking GHC ``manually'' via its API is unlikely to obtain the
+definitions we require.
 
-Thankfully there is one implementation detail which most Haskell packages agree on: the Cabal build system.
-All of the above complexities will be specified in a package's ``Cabal file'', such that the \texttt{cabal configure} and \texttt{cabal build} commands are very likely to work for most packages, without any extra effort.
-This shifted our focus to augmenting GHC and Cabal, such that definitions can be collected during the normal Haskell build process.
+Thankfully there is one implementation detail which most Haskell packages agree
+on: the Cabal build system. All of the above complexities will be specified in a
+package's ``Cabal file'', such that the \texttt{cabal configure} and
+\texttt{cabal build} commands are very likely to work for most packages, without
+any extra effort. This shifted our focus to augmenting GHC and Cabal, such that
+definitions can be collected during the normal Haskell build process.
 
-GHC provides a plugin mechanism for manipulating Core during a build, intended for optimisation passes, which we use to inspect definitions as they are being compiled.
-We provide a plugin called \textsc{AstPlugin} which emits a serialised version of each Core definition to the console (to satisfy the type system, it also implements a dummy ``optimisation'' which returns the Core unchanged).
+GHC provides a plugin mechanism for manipulating Core during a build, intended
+for optimisation passes, which we use to inspect definitions as they are being
+compiled. We provide a plugin called \textsc{AstPlugin} which emits a serialised
+version of each Core definition to the console (to satisfy the type system, it
+also implements a dummy ``optimisation'' which returns the Core unchanged).
 
-Compared to Haskell, Core is a much simpler language and its representation is relatively stable compared to many existing representations of Haskell (which often change to support various language extensions).
-Three areas which make Core difficult to handle are:
+Compared to Haskell, Core is a much simpler language and its representation is
+relatively stable compared to many existing representations of Haskell (which
+often change to support various language extensions). Three areas which make
+Core difficult to handle are:
 
 \begin{description}
-  \item{Type variables}: Parametric polymorphism (described in more detail in \S \ref{sec:haskelldesc}) can be thought of as values being parameterised by type-level objects.
-In System F, this is represented explicitly by a special abstraction form $\Lambda$, distinct from the $\lambda$ used for values.
-Core only has one abstraction form, \CLam, for both types and values.
-This alters function properties like arity.
+\item{Type variables}: Parametric polymorphism (described in more detail in \S
+  \ref{sec:haskelldesc}) can be thought of as values being parameterised by
+  type-level objects. In System F, this is represented explicitly by a special
+  abstraction form $\Lambda$, distinct from the $\lambda$ used for values. Core
+  only has one abstraction form, \CLam, for both types and values. This alters
+  function properties like arity.
 
-  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc.
-Since Core does not make these distinctions, names may become ambiguous.
-For example, a type parameter \hs{t} may be confused with a function argument \hs{t}.
-To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core, which is guaranteed not to conflict since this name would be invalid in Haskell.
-This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.
+\item{Unified namespace}: Haskell has distinct namespaces for values, types,
+  data constructors, etc. Since Core does not make these distinctions, names may
+  become ambiguous. For example, a type parameter \hs{t} may be confused with a
+  function argument \hs{t}. To prevent this, overlapping namespaces are
+  distinguished by prefices which are distinct from the available names; for
+  example a type class constraint \hs{Ord t} may give rise to a binder
+  $\CLam\ \hs{"\$dOrd"}$ in Core, which is guaranteed not to conflict since this
+  name would be invalid in Haskell. This causes difficulties when looking up
+  names, as these prefixed forms do not easily map back to the Haskell source.
 
-  \item{Violating encapsulation}: Although Haskell allows names to be \emph{private} to a module, when compiling Core we have full access to private definitions, as well as references to private names from within other definitions.
-Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
+\item{Violating encapsulation}: Although Haskell allows names to be
+  \emph{private} to a module, when compiling Core we have full access to private
+  definitions, as well as references to private names from within other
+  definitions. Hence the definitions we receive from \textsc{AstPlugin} will
+  include private values which we cannot import into a theory exploration tool.
 \end{description}
 
-In practice, we work around these issues with a post-processing stage: for each named definition appearing in the output of \textsc{AstPlugin}, we attempt to reference that name within the GHCi interpreter.
-Names with the above problems will cause an error, and are discarded.
+In practice, we work around these issues with a post-processing stage: for each
+named definition appearing in the output of \textsc{AstPlugin}, we attempt to
+reference that name within the GHCi interpreter. Names with the above problems
+will cause an error, and are discarded.
 
-The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \citep{mitchell2008hoogle}.
-Definitions are indexed by a combination of their package name, module name and binding name.
-The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which makes it trivial to look up references in the database.
-Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.
+The result of building a Haskell package with \textsc{AstPlugin} enabled is a
+database of Haskell definitions, similar in some respects to \textsc{Hoogle}
+\cite{mitchell2008hoogle}. Definitions are indexed by a combination of their
+package name, module name and binding name. The definitions themselves are
+s-expressions representing the Core AST, with non-local references replaced by a
+combination of package name, module name and binding name, which makes it
+trivial to look up references in the database. Each definition also has an
+associated arity and type, obtained during the post-processing step mentioned
+above.
 
@@
 
-As described in \S \ref{sec:symbolstofeatures}, we must topologically sort the output of \textsc{AstPlugin} in order for our recurrent clustering to be well-founded.
-Since our database keys (containing the package, module and binding names, as described above) match our representation of non-local references, it is simple to walk each syntax tree to obtain the set of references it makes.
-In addition, the resulting set of (identifier, list-of-referenced-identifiers) pairs exactly matches the (vertex, list-of-successor-vertices) format used to represent directed graphs by the popular \hs{containers} library which ships with GHC.
-This provides an implementation of topological sort for strongly connected components, which we use as-is.
-A simple shell script loops through these SCCs, invoking the recurrent clustering component for each and appending the resulting features and clusters to the database.
+As described in \S \ref{sec:symbolstofeatures}, we must topologically sort the
+output of \textsc{AstPlugin} in order for our recurrent clustering to be
+well-founded. Since our database keys (containing the package, module and
+binding names, as described above) match our representation of non-local
+references, it is simple to walk each syntax tree to obtain the set of
+references it makes. In addition, the resulting set of (identifier,
+list-of-referenced-identifiers) pairs exactly matches the (vertex,
+list-of-successor-vertices) format used to represent directed graphs by the
+popular \hs{containers} library which ships with GHC. This provides an
+implementation of topological sort for strongly connected components, which we
+use as-is. A simple shell script loops through these SCCs, invoking the
+recurrent clustering component for each and appending the resulting features and
+clusters to the database.
 
@@
 
-The implementation of our feature extraction algorithm is a rather direct translation of the description given in \S \ref{sec:contributions} into Haskell.
-We parse the s-expressions generated by \textsc{AstPlugin} into algebraic data types which correspond directly to the definitions in Figure \ref{fig:coresyntax}; this is routine, so we omit the details for brevity.
-Similarly, we can represent rose trees with a datatype corresponding to the definition given in \S \ref{sec:expressionstovectors}:
+The implementation of our feature extraction algorithm is a rather direct
+translation of the description given in \S \ref{sec:contributions} into
+Haskell. We parse the s-expressions generated by \textsc{AstPlugin} into
+algebraic data types which correspond directly to the definitions in Figure
+\ref{fig:coresyntax}; this is routine, so we omit the details for
+brevity. Similarly, we can represent rose trees with a datatype corresponding to
+the definition given in \S \ref{sec:expressionstovectors}:
 
@@
 
-For simplicity we use the representation \hs{Feature = Int}, as we do not have fractional values.
-Since we represent the symbols $expr$, $id$, etc.
-from Figure \ref{fig:coresyntax} using different datatypes, we cannot write one big definition of $toTree$ or $\phi$ which works on all tokens.
-Each case shown in Figure \ref{fig:totree} and equations \ref{eq:feature}, \ref{eq:localfeature} and \ref{eq:globalfeature} appears in the implementation, although they are spread across several functions.
+For simplicity we use the representation \hs{Feature = Int}, as we do not have
+fractional values. Since we represent the symbols $expr$, $id$, etc. from Figure
+\ref{fig:coresyntax} using different datatypes, we cannot write one big
+definition of $toTree$ or $\phi$ which works on all tokens. Each case shown in
+Figure \ref{fig:totree} and equations \ref{eq:feature}, \ref{eq:localfeature}
+and \ref{eq:globalfeature} appears in the implementation, although they are
+spread across several functions.
 
-To support looking up local identifiers, our implementation of $toTree$ takes a context as argument, extending it as required.
-As an example of the complexity this adds, here is the \CCase\ branch of $toTree$:
+To support looking up local identifiers, our implementation of $toTree$ takes a
+context as argument, extending it as required. As an example of the complexity
+this adds, here is the \CCase\ branch of $toTree$:
 
@@
 
-Breaking this down we can see \hs{fCase} representing the value of $\feature{\CCase}$, and a list of sub-trees defined in parentheses.
-The first subtree is a straightforward recursive call in an unmodified context: \mbox{\hs{toTree ctx e}}.
-The rest of the list is formed by applying the function \mbox{\hs{toTreeAlt (l:ctx)}} to each element of the list \hs{as} of \CAlt\ clauses.
+Breaking this down we can see \hs{fCase} representing the value of
+$\feature{\CCase}$, and a list of sub-trees defined in parentheses. The first
+subtree is a straightforward recursive call in an unmodified context:
+\mbox{\hs{toTree ctx e}}. The rest of the list is formed by applying the
+function \mbox{\hs{toTreeAlt (l:ctx)}} to each element of the list \hs{as} of
+\CAlt\ clauses.
 
-The \hs{toTreeAlt} function contains those cases of $toTree$ which handle symbols in $alt$.
-We prepend the identifier \hs{l} to the context, to get the extended context \hs{l:ctx}.
-This is because \hs{l} will be bound the value of \hs{e}, in order to avoid re-computing its value several times.
+The \hs{toTreeAlt} function contains those cases of $toTree$ which handle
+symbols in $alt$. We prepend the identifier \hs{l} to the context, to get the
+extended context \hs{l:ctx}. This is because \hs{l} will be bound the value of
+\hs{e}, in order to avoid re-computing its value several times.
 
-The other clauses are handled in a similar way.
-The trickiest is the \CLet\ clause, since the local identifiers aren't directly available; we must extract them from their \CRec, \CNonRec\ and \CBind\ wrappers first, which we do using helper functions.
+The other clauses are handled in a similar way. The trickiest is the \CLet\
+clause, since the local identifiers aren't directly available; we must extract
+them from their \CRec, \CNonRec\ and \CBind\ wrappers first, which we do using
+helper functions.
 
-As shown above, the values from equations \ref{eq:feature} are encoded directly in $toTree$.
-For $phi(l \in \mathcal{L})$ we use standard Haskell functions to look up the required indices in the context:
+As shown above, the values from equations \ref{eq:feature} are encoded directly
+in $toTree$. For $phi(l \in \mathcal{L})$ we use standard Haskell functions to
+look up the required indices in the context:
 
@@
 
-As explained in \S \ref{sec:symbolstofeatures}, local identifiers should always exist in the context. If this precondition doesn't hold, we abort the program with an error rather than continuing.
+As explained in \S \ref{sec:symbolstofeatures}, local identifiers should always
+exist in the context. If this precondition doesn't hold, we abort the program
+with an error rather than continuing.
 
-Global identifiers are kept as-is until we have access to the clusters from the last iteration. This takes place outside Haskell, using the \hs{jq} data processing tool.
+Global identifiers are kept as-is until we have access to the clusters from the
+last iteration. This takes place outside Haskell, using the \hs{jq} data
+processing tool.
 
@@
 
-To produce feature vectors, we do not directly construct the matrix; instead we generate the rows and concatenate them together in one step, using the \hs{concatMap} function:
+To produce feature vectors, we do not directly construct the matrix; instead we
+generate the rows and concatenate them together in one step, using the
+\hs{concatMap} function:
 
@@
 
-By providing \hs{featureVec} with the latest set of clusters, read from the \textsc{AstPlugin} database, we turn Core expressions into feature vectors, which are appended to the database.
+By providing \hs{featureVec} with the latest set of clusters, read from the
+\textsc{AstPlugin} database, we turn Core expressions into feature vectors,
+which are appended to the database.
 
-We use the Weka system to perform our k-means clustering, as it is widely used, including by ML4PG.
-We select all feature vectors from our database, and write them in CSV format for Weka to process.
-The Weka CLI command is invoked, which appends a cluster number to each of these feature vectors; we read these off and append them to the database.
-As long as more SCCs remain unprocessed, we keep looping this process, using the database to communicate between the feature extraction and clustering phases.
+We use the Weka system to perform our k-means clustering, as it is widely used,
+including by ML4PG. We select all feature vectors from our database, and write
+them in CSV format for Weka to process. The Weka CLI command is invoked, which
+appends a cluster number to each of these feature vectors; we read these off and
+append them to the database. As long as more SCCs remain unprocessed, we keep
+looping this process, using the database to communicate between the feature
+extraction and clustering phases.
 
@@
 
-We cannot supply these clusters as-is to \qspec{}, since it must be provided with a \emph{signature}.
-These are constructed by our \textsc{MLSpec} tool, using information from the \textsc{AstPlugin} database.
-Tasks performed by \textsc{MLSpec} include:
+We cannot supply these clusters as-is to \qspec{}, since it must be provided
+with a \emph{signature}. These are constructed by our \textsc{MLSpec} tool,
+using information from the \textsc{AstPlugin} database. Tasks performed by
+\textsc{MLSpec} include:
 
 \begin{itemize}
-  \item{Monomorphising}: Given values of polymorphic type, e.g.
-\hs{safeHead :: forall t.
-[t] -> Maybe t} and \hs{[] :: forall t.
-[t]}, a testing-based system like \qspec{} is unable to evaluate these expressions without instantiating the variable \hs{t} to a specific type.
-Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
-We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g.
-\hs{forall c.
-c Bool -> c Bool}) or incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}).
+\item{Monomorphising}: Given values of polymorphic type, e.g. \hs{safeHead ::
+    forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based
+  system like \qspec{} is unable to evaluate these expressions without
+  instantiating the variable \hs{t} to a specific type. Such an instantiation is
+  called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on
+  previous work in \qcheck{} by attempting to instantiate all type variables to
+  \hs{Integer}. We discard those cases where this is invalid, such as variable
+  \emph{type constructors} (e.g. \hs{forall c. c Bool -> c Bool}) or
+  incompatible class constraints (e.g. \hs{forall t. IsString t => t}).
 
-  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules.
-In such cases the exploration process simply aborts.
+\item{Qualification}: All names are \emph{qualified} (prefixed by their module's
+  name), to avoid most ambiguity. There is still the possibility that multiple
+  packages will declare modules of the same name, although this is rare as it
+  causes problems for any Haskell programmer trying to use those modules. In
+  such cases the exploration process simply aborts.
 
-  \item{Variable definition}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).
+\item{Variable definition}: Once a \qspec{} theory has been defined containing
+  all of the given terms, we inspect the types it references and append three
+  variables for each to the theory (enough to discover laws such as
+  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
+  search).
 
-  \item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is that all required packages and modules must be provided up-front, usually by specification in a Cabal file.
-Since \textsc{MLSpec} builds signatures \emph{dynamically}, depending on the cluster information it is given, we do not know what packages it may need.
-To work around this problem, \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have built called \texttt{nix-eval}.
-This provides an \texttt{eval} function, like those commonly found in dynamic languages such as Python and Javascript, for evaluating Haskell expressions.
-The key feature of \texttt{nix-eval} is that these Haskell expressions may reference packages that are not installed on the system.
-When such expressions are evaluated, these packages will be automatically downloaded and installed into a sandbox using the Nix package manager, and GHC will be invoked in this sandbox to perform the evaluation.
-
+\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
+  that all required packages and modules must be provided up-front, usually by
+  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
+  \emph{dynamically}, depending on the cluster information it is given, we do
+  not know what packages it may need. To work around this problem,
+  \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have
+  built called \texttt{nix-eval}. This provides an \texttt{eval} function, like
+  those commonly found in dynamic languages such as Python and Javascript, for
+  evaluating Haskell expressions. The key feature of \texttt{nix-eval} is that
+  these Haskell expressions may reference packages that are not installed on the
+  system. When such expressions are evaluated, these packages will be
+  automatically downloaded and installed into a sandbox using the Nix package
+  manager, and GHC will be invoked in this sandbox to perform the evaluation.
 \end{itemize}

### 2018-05-17 PhDThesis/background_theoryexploration.tex
More imports

@@
 and the consequences for testing and QA in general.
+
+%%%%%
+
+\begin{description}
+\item{Interestingness} Various alternative interestingness criteria have been
+  proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing
+  the criteria may be useful, for example to distinguish useful relationships
+  from incidental coincidences; or to prevent surprising, insightful equations
+  from being discarded because they can be simplified.
+\end{description}
+
+% Theory exploration is similar to \emph{experimental mathematics}
+
+% - Relation to Science
+%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
+% - Relation to AI tasks: exploring surroundings, etc.
+
+% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+
+% Theory formation: Alison? Others.
+% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
+% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
+% Communication: Latex, Wikis, APIs, communicating with aliens
+
+\section{Exploration in Theorem Proving}
+\label{sec:examples}
+
+Before exploring abstract definitions of interestingness, we can first consider
+some scenarios which arise during formal proof where we are forced to generate
+conjectures. An analysis of these situations, and the subsequent theorems they
+produce, will contribute towards an empirical justification for what is
+interesting (at least from a utilitarian point of view) and inform our later
+exploration of the literature.
+
+\subsection{Generalisation}
+
+\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+
+When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
+which $S$ is a special case. Although it seems counterintuitive, a generalised
+statement can sometimes be \emph{easier} to prove than the original. This arises
+often in inductive proofs, since the specific obligations which arise in the
+proof may be incompatible with the available inductive hypotheses.
+
+However, we cannot blindly generalise \emph{all} obligations we encounter, since
+\emph{over-generalising} results in obligations which are so strong that they
+are unprovable, or even false. We must therefore rely on heuristics to guide the
+generation of generalised conjectures, and hence perform a kind of exploration.
+
+An informative example is given by Boyer and Moore of the associativity of
+multiplication in ACL2 \cite{boyer1983proof}:
+
+$$(x * y) * z = x * (y * z)$$
+
+During the course of the proof, the following obligation arises:
+
+\begin{equation}
+  \tag{conc3}
+  (y + (x * y)) * z = (y * z) + ((x * y) * z)
+  \label{eq:conc3}
+\end{equation}
+
+ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
+sub-term $x * y$ with a fresh variable $w$:
+
+\begin{equation}
+  \tag{conc4}
+  (y + w) * z = (y * z) + (w * z)
+  \label{eq:conc4}
+\end{equation}
+
+This generalised form is clearly the distributivity law for multiplication and
+addition, which can be proved separately to the original goal of
+associativity. It would not be controversial to claim that this distributivity
+law is interesting in its own right (relative to associativity, at least), in
+addition to its usefulness in making this proof go through.
+
+% TODO: Describe the ACL2 heuristics
+
+Generalisation also occurs frequently when reasoning about \emph{tail-recursive}
+definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can
+  be executed in constant space using a loop, whereas recursion in non-tail
+  positions may require a growing number of stack frames or nested closures. See
+  \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
+
+\subsection{Analogy}
+
+One way to characterise the interestingness of a statement is by \emph{analogy}
+to existing interesting statements. By finding lemmas analogous to those of a
+different theory, we may be able to re-use tactics and other forms of
+meta-programming across both.
+
+Existing theory exploration systems have been successfully applied to this
+problem, however the use of pure exploration misses opportunities to
+\emph{focus} the search, since we know which lemmas are used in those theories
+where a technique succeeded. If we can find an analogy to map from such solved
+problems to our unsolved goal, we can infer the approximate form of the lemmas
+we require, and target these specifically.
+
+The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
+to solving a goal $G$ by making analogies via unsupervised clustering
+\cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are used in two
+ways:
+
+\begin{itemize}
+\item First, we use the cluster $C_G$ containing $G$ to identify analogous
+  theorems.
+
+\item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
+  $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
+  to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
+  the cluster $C_s$ containing $s$.
+\end{itemize}
+
+The running examples for demonstrating \textsc{ACL2(ml)} are equivalence
+theorems for tail-recursive and non-tail-recursive calculations, as well as the
+effect of repeating certain list operations:
+
+\begin{itemize}
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
+
+  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
+
+\end{itemize}
+
+\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
+
+One consideration when generating conjectures is the difference between
+theorems, lemmas, corollaries, etc. From a logical point of view, these are all
+equivalent, and hence most proof assistants do not distinguish between
+them. However, their \emph{intention} may be different: in a sense, theorems are
+the interesting results; whilst lemmas are useful results, required for proving
+the theorems.
+
+Some systems, like Coq, allow users to \emph{label} each statement as being a
+\coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
+the same. This shows us immediately that lemmas outnumber theorems; in the Coq
+standard library there are over five times as many lemmas as theorems
+\footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when
+  excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of
+  \coq{Lemma} in its \texttt{theories/} directory.}.
+
+% TODO: Analyse them
+
+% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
+
+We can find a need for auxiliary lemmas, once again, in the context of
+tail-recursive functions. Consider proving the (pointwise) equality of the
+following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Inductive Nat : Set := Z : Nat
+                     | S : Nat -> Nat.
+
+Fixpoint plus      (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => S (plus n' m)
+                                  end.
+
+Fixpoint plus_tail (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => plus_tail n' (S m)
+                                  end.
+\end{lstlisting}
+
+\begin{haskell}
+Haskell equivalent:
+
+plus :: Nat -> Nat -> Nat
+plus      n  Z    = n
+plus      n (S m) = S (plus n m)
+
+plus_tail :: Nat -> Nat -> Nat
+plus_tail n  Z    = n
+plus_tail n (S m) = plus_tail (S n) m
+\end{haskell}
+
+Both of these functions implement addition, but the \coq{plus_tail} variant is
+tail-recursive. However, if we want to \emph{prove} that the definitions are
+(pointwise) equal, we run into difficulties. In particular, when the inductive
+step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
+reasonable), we cannot make this go through using another inductive argument.
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+(* Solve equalities by beta-normalising both sides *)
+Ltac triv := try (simpl; reflexivity).
+
+(* Prove equivalence of plus and plus_tail *)
+Theorem equiv : forall n m, plus n m = plus_tail n m.
+  induction n; triv. (* Base case is trivial *)
+
+  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
+  intro m.
+
+  (* Beta-reduce the right-hand-side (justification is trivial) *)
+  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
+
+  (* Use induction hypothesis to replace plus_tail with plus *)
+  rewrite <- (IHn (S m)).
+\end{lstlisting}
+
+Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
+the equation we need:
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
+\end{lstlisting}
+
+Yet we cannot provide it with the argument it needs, as our original induction
+hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+IHn : forall x, plus (S n') x = plus_tail (S n') x
+\end{lstlisting}
+
+We are forced to abandon the proof, despite such a reasonable-looking
+intermediate goal.
+
+In fact, if we attempt to prove that goal \emph{separately}, we can use a
+straightforward argument by induction; even though it is actually
+\emph{stronger} due to the absence of the \coq{IHn} assumption. Using this
+separate result as a lemma, the pointwise equality is proven easily.
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+Lemma gen n m : plus (S n) m = plus n (S m).
+  induction n; triv. (* Base case is trivial *)
+
+  (* Move all S constructors outside *)
+  simpl. rewrite <- IHn. simpl.
+
+  (* Trivial *)
+  reflexivity.
+Defined.
+\end{lstlisting}
+
+\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  rewrite (gen n m).
+  reflexivity.
+Defined.
+\end{lstlisting}

### 2018-05-17 PhDThesis/intro.tex
More imports

--- 2015-11-23 TheoryExploration/interesting.tex
@@
-% Since writing this up, it was decided to leave as future work. Still, worth
-% having around for editing into such future work.
+As computers and software become more capable, and as our reliance on them
+increases, the importance of \emph{understanding}, \emph{predicting} and
+\emph{verifying} these systems grows; which is undermined by their
+ever-increasing complexity. Mathematics provides us with powerful, systematic
+methods of reasoning, which we can bring to bear on this challenge; in
+particular those of \emph{formal logic} and \emph{statistics}. By (partially)
+mechanising these approaches, in the fields of \emph{theorem proving} and
+\emph{machine learning}, respectively, we can leverage these increasing machine
+capabilities and direct them for the purpose of analysis. However, the question
+still remains: on what should we focus that analysis?
 
-\documentclass[]{article}
-\begin{document}
+In this work, we investigate the notion of \emph{interestingness} in the
+exploration of formal systems (an area known as \emph{theory exploration}) as a
+way to make productive use of resources in an often intractable domain. To keep
+things concrete, we focus our formal analysis on equational formulae describing
+programs in the Haskell language, for reasons elaborated in \S
+\ref{sec:haskell}. Conceptually, we maintain a broader view, and survey many
+related areas which may offer insights on the problem.
 
-\title{Interestingness Measures for Theory Exploration}
+Appeals to interestingness arise when more direct measures, such as utility, are
+not available. For example, the inclusion of particular statements in a program
+or proof development can be easily justified based on their contribution to the
+overall solution; however, in a \emph{library} there is no particular problem
+being solved, in which case we must judge statements on less direct criteria,
+such as how ``interesting'' they may be to our users. Appeals to interestingness
+abound in the history of computer-assisted reasoning; for example, in 1971
+Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering
+  theorems $T$ from a system of axioms $Ax$}, and in particular the questions
+\textquote{Under what conditions is $T$ an interesting, possible theorem in the
+  system $Ax$?} and \textquote{Is there a way to generate (most) interesting
+  possible theorems?}. Despite such widespread use of the term, there is no
+standard definition of what makes a formal object, whether it is an axiom, a
+conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have
+been proposed.
 
-\author{Chris Warburton}
+We begin our undertaking in \S \ref{sec:background} by introducing the Haskell
+language, as well as the relevant fields of verification for context. We define
+a formal framework for our investigation, and show how it relates to the
+existing theorem proving landscape. A selection of theorem proving scenarios
+which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst
+related work, including existing defintions of interestingness, is surveyed in
+\S \ref{sec:related}. We also review the use of exploration in other fields of
+Artificial Intelligence and Machine Learning, where researchers are
+experimenting with replacing \emph{explicit} goals and rewards with
+\emph{implicit} alternatives such as interestingness. Recent efforts in this
+area have lead to the emergence of principled theories, mostly based around
+(algorithmic) information theory, which may be adapted to our theory exploration
+context.
 
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{sec:haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{sec:background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{sec:related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{sec:current} and future research directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.
-
-\begin{description}
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{sec:relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{sec:examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \cite{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{sec:tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \cite{claessen2011quickcheck} and \textsc{SmartCheck} \cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \cite{reich2013advances}) and logic programming (\textsc{SparseCheck} \cite{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Related Work}
-\label{sec:related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\cite{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \cite{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \cite{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{sec:curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{fig:wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \cite{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\cite{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \cite{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\cite{Kaplan2006}
-\cite{Lipson2007}
-\cite{Luciw2011}
-\cite{Macedo2000}
-\cite{Ramik.Sabourin.Madani:2013}
-\cite{Roa.Kruijff.Jacobsson:2009}
-\cite{Schaul.Sun.Wierstra.ea:2011}
-\cite{Schmidhuber1999}
-\cite{Schmidhuber:1991}
-\cite{Scott1989}
-\cite{Steunebrink.Koutnik.Thorisson.ea:2013}
-\cite{maher2008achieving}
-\cite{meyer1991possibility}
-\cite{oudeyer2004intelligent}
-\cite{oudeyer2014evolution}
-\cite{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{sec:relevance}
-
-% TODO
-\cite{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \cite{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{sec:clustering}
-
-% TODO: ML4PG
-\cite{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\cite{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\cite{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{sec:conceptformation}
-
-% TODO:
-\cite{Montano-Rivas.McCasland.Dixon.ea:2012}
-\cite{Piantadosi.Tenenbaum.Goodman:2012}
-\cite{Wille:2005}
-\cite{colton1999automatic}
-\cite{colton2000agent}
-\cite{colton2012automated}
-\cite{lenat1977automated}
-\cite{mullerunderstanding}
-\cite{Bundy.Cavallo.Dixon.ea:2015}
-\cite{johansson2009isacosy}
-\cite{spector2008genetic}
-\cite{colton2012automated}
- \cite{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{sec:featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \cite{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\cite{Gartner2003}
-\cite{Oveisi.Oveisi.Erfanian.ea:2012}
-\cite{bakir2007predicting}
-\cite{conf/ijcai/Plate91}
-\cite{goller1996learning}
-\cite{kwasny1995tail}
-\cite{pollack1990recursive}
-\cite{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}
+We discuss our present contributions in \S \ref{sec:current} and future research
+directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.

### 2018-05-17 PhDThesis/related_formalstats.tex
More imports

--- 2015-11-23 TheoryExploration/interesting.tex
@@
-% Since writing this up, it was decided to leave as future work. Still, worth
-% having around for editing into such future work.
-
-\documentclass[]{article}
-\begin{document}
-
-\title{Interestingness Measures for Theory Exploration}
-
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{sec:haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{sec:background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{sec:related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{sec:current} and future research directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.
-
-\begin{description}
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{sec:relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{sec:examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \cite{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{sec:tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \cite{claessen2011quickcheck} and \textsc{SmartCheck} \cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \cite{reich2013advances}) and logic programming (\textsc{SparseCheck} \cite{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-
-% TODO: Analysis
-
-\section{Related Work}
-\label{sec:related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\cite{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \cite{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \cite{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{sec:curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{fig:wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \cite{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\cite{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \cite{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\cite{Kaplan2006}
-\cite{Lipson2007}
-\cite{Luciw2011}
-\cite{Macedo2000}
-\cite{Ramik.Sabourin.Madani:2013}
-\cite{Roa.Kruijff.Jacobsson:2009}
-\cite{Schaul.Sun.Wierstra.ea:2011}
-\cite{Schmidhuber1999}
-\cite{Schmidhuber:1991}
-\cite{Scott1989}
-\cite{Steunebrink.Koutnik.Thorisson.ea:2013}
-\cite{maher2008achieving}
-\cite{meyer1991possibility}
-\cite{oudeyer2004intelligent}
-\cite{oudeyer2014evolution}
-\cite{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
 \subsection{Statistics of Formal Systems}
 
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
+The core problem of assigning ``interestingness'' to logical formulae is the
+application of statistical reasoning to the discrete, semantically-rich domain
+of formal systems. This problem has been tackled from various directions for a
+variety of reasons; here we summarise those contributions which seem of
+particular importance for theory exploration.
 
@@
 
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
+The combinatorial nature of formal systems causes many proof search methods,
+such as resolution, to have exponential complexity
+\cite{haken1985intractability}; hence even a modest size increase can turn a
+trivial problem into an intractable one. Finding efficient alternatives for such
+algorithms, especially those which are NP-complete (e.g. determining
+satisfiability) or co-NP-complete (e.g. determining tautologies), seems
+unlikely, as it would imply progress on the famously intractable open problems
+of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$. On the other hand, we
+can turn this difficulty around: a modest \emph{decrease} in size may turn an
+intractable problem into a solvable one. We can ensure that the solutions to
+these reduced problems coincide with the original if we only remove
+\emph{redundant} information. This leads to the idea of \emph{relevance
+  filtering}.
 
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
+Relevance filtering simplifies a proof search problem by removing from
+consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed
+\emph{irrelevant}. The technique is used in Sledgehammer during its translation
+of Isabelle/HOL theories to statements in first order logic: rather than
+translating the entire theory, only a sub-set of relevant clauses are
+included. This reduces the size of the problem and speeds up the proof search,
+but it creates the new problem of determining when a clause is relevant: how do
+we know what will be required, before we have the proof?
 
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
+The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson}
+\cite{meng2009lightweight}), gives each clause a score based on the proportion
+$m / n$ of its symbols which are ``relevant'' (where $n$ is the number of
+symbols in the clause and $m$ is the number which are relevant). Initially, the
+relevant symbols are those which occur in the goal, but whenever a clause is
+found which scores more than a particular threshold, all of its symbols are then
+also considered relevant. There are other heuristics applied too, such as
+increasing the score of user-provided facts (e.g. given by keywords like
+\texttt{using}), locally-scoped facts, first-order facts and rarely-occuring
+facts. To choose $r$ relevant clauses for an ATP invocation, we simply order the
+clauses by decreasing score and take the first $r$ of them.
 
-Recently, a variety of alternative algorithms have also been investigated, including:
+Recently, a variety of alternative algorithms have also been investigated,
+including:
 
 \begin{description}
+\item{\textsc{MaSH}}: Machine Learning for SledgeHammer
+  \cite{kuhlwein2013mash}. The distinguishing feature of \textsc{MaSH} is its
+  use of ``visibility'', which is essentially a dependency graph of which
+  theorems were used in the proofs of which other theorems; although theorems
+  are represented as abstract sets of features. To select relevant clauses for a
+  goal, the set of clauses which are visible from the goal's components is
+  generated; this is further reduced by (an efficient approximation of) a naive
+  Bayes algorithm.
 
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \cite{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
+\item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine
+  (SVM) approach for selecting relevant axioms from the Mizar Mathematical
+  Library for use by the Vampire ATP system \cite{alama2014premise}. \iffalse
+  TODO: describe the kernel, as that's the interesting bit \fi It compares
+  favourably to \textsc{SNoW} and \textsc{SInE}.
 
@@
 
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
+The most important property of a logical formula is its truth value. Although we
+may be able to determine some truth values exactly, e.g. using decision or
+semi-decision procedures, it may be more efficient to \emph{approximate} truth
+values. One straightforward extension of truth values is \emph{probabilities},
+where we can assign probability $1$ to formulae which are known to be true, $0$
+to formulae known to be false, and intermediate values to those which we do not
+yet know.
 
@@
 
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
+% However, this search space grows exponentially in the length of the proofs,
+% which is unfortunate since proof length has been proposed as an approximate
+% measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
 
@@
 
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
+One major difficulty with formal mathematics as a domain in which to apply
+statistical machine learning is the use of \emph{structure} to encode
+information in objects. In particular, \emph{trees} appear in many places: from
+inductive datatypes, to recursive function definitions; from theorem statements,
+to proof objects. Such nested structures may extend to arbitrary depth, which
+makes them difficult to represent with a fixed number of features, as is
+expected by most machine learning algorithms. Here we review a selection of
+solutions to this problem, and compare their distinguishing properties.
 
@@
 
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
+\emph{Feature extraction} is a common pre-processing step for machine learning
+(ML). Rather than feeding ``raw'' data straight into our ML algorithm, we only
+learn a sample of \emph{relevant} details, known as \emph{features}. This has
+two benefits:
 
 \begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
+\item \emph{Feature vectors} (ordered sets of features) are chosen to be more
+  compact than the data they're extracted from: feature extraction is
+  \emph{lossy compression}. This reduces the size of the ML problem, improving
+  efficiency (e.g. running time).
+\item We avoid learning irrelevant details such as the encoding system used,
+  improving \emph{data} efficiency (the number of samples required to spot a
+  pattern).
 \end{itemize}
 
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
+Another benefit of feature extraction is to \emph{normalise} the input data to a
+fixed-size representation. Many ML algorithms only work with fixed-size inputs;
+for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294}
+algorithm works on models with \emph{fixed} topology (e.g. \emph{artificial
+  neural networks} with fixed connections between nodes). This requires some
+form of pre-processing in domains where the size of each input is not known, may
+vary or may even be unbounded.
 
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
+For example, in the case of \emph{online} learning we must make
+predictions/decisions before seeing all of the inputs. Unbounded input appears
+in domains such as programming and theorem proving, where individual term may be
+trees of unbounded depth. In these situations we need a mapping from arbitrary
+inputs to a fixed representation which is amenable to learning.
 
-As an example, say we want to learn relationships between the following program fragments:
+As an example, say we want to learn relationships between the following program
+fragments:
 
@@
   \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
+  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a
+      = Either () a} and \hs{Nothing = Left ()}).
+  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to
+    \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
   \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
@@
 
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
+However, this is too optimistic. Without our domain-knowledge of Haskell, an ML
+algorithm cannot impose any structure on these fragments, and will treat them as
+strings of bits. Our high-level hopes are obscured by low-level details: the
+desirable patterns of Haskell types are mixed with undesirable patterns of ASCII
+bytes, of letter frequency in English words, and so on.
 
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
+In theory we could throw more computing resources and data at a problem, but
+available hardware and corpora are always limited. Instead, feature extraction
+lets us narrow the ML problem to what we, with our domain knowlege, consider
+important.
 
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
+There is no \emph{fundamental} difference between raw representations and
+features: the identity function is a valid feature extractor. Likewise, there is
+no crisp distinction between feature extraction and machine learning: a
+sufficiently-powerful learner doesn't require feature extraction, and a
+sufficiently-powerful feature extractor doesn't require any learning!
+\footnote{Consider a classification problem, to assign a label $l \in L$ to each
+  input. If we only extract a single feature $f \in L$, we have solved the
+  classification problem without using a separate learning step.}
 
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
+Rather, the terms are distinguished for purely \emph{practical} reasons: by
+separating feature extraction from learning, we can distinguish straightforward,
+fast data transformation (feature extraction) from complex, slow statistical
+analysis (learning). This allows for modularity, separation of concerns, and in
+particular allows ``off-the-shelf'' ML to be re-used across a variety of
+different domains.
 
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
+Even if we have no domain knowledge, we can still use a feature extraction phase
+to improve efficiency: first we learn a compact representation for our data, for
+example using \emph{autoencoding}; then we use that encoder as a feature
+extractor for our main learning task. This stacking of one learning algorithm on
+top of another, especially with greedy learning of each layer, has lead to the
+recent trend of \emph{deep learning}.
 
@@
 
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
+The simplest way to limit the size of our inputs is to truncate anything larger
+than a particular size (and pad anything smaller). This is the approach taken by
+ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at
+most 10 levels and 10 elements per level; each tree is converted to a
+$30 \times 10$ matrix (3 values per tree node) and learning takes place on these
+normalised representations.
 
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
+Truncation is unsatisfactory in the way it balances \emph{data} efficiency with
+\emph{time} efficiency. Specifically, truncation works best when the input data
+contains no redundancy and is arranged with the most significant data first (in
+a sense, it is ``big-endian''). The less these assumptions hold, the less we can
+truncate. Since many ML algorithms scale poorly with input size, we would prefer
+to eliminate the redundancy using a more aggressive algorithm, to keep the
+resulting feature size as low as possible.
 
@@
 
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
+A more sophisticated approach to the problem of reducing input size is to view
+it as a \emph{dimension reduction} technique: our inputs can be modelled as
+points in high-dimensional spaces, which we want to project into a
+lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit
+vectors).
 
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
+Truncation is a trivial dimension reduction technique: take the first $N$
+coordinates (bits). More sophisticated projection functions consider the
+\emph{distribution} of the points, and project with the hyperplane which
+preserves as much of the variance as possible (or, equivalently, reduces the
+\emph{mutual information} between the points).
 
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
+There are many techniques to find these hyperplanes, such as \emph{principle
+  component analysis} (PCA) and \emph{autoencoding}; however, since these
+techniques are effectively ML algorithms in their own right, they suffer some of
+the same constraints we're trying to avoid:
 
@@
 
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
+In particular, the second constraint is precisely what we're trying to
+avoid. Sophisticated dimension reduction is still useful for \emph{compressing}
+large, redundant features into smaller, information-dense representations, and
+as such provides a good complement to truncation.
 
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
+The requirement for offline ``batch'' processing is more difficult to overcome,
+since any learning we perform for feature extraction will interfere with the
+core learning algorithm that's consuming these features (this is why deep
+learning is often done greedily).
 
@@
 
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
+The task of dimension reduction changes when we consider \emph{structured}
+data. Recursive structures, like trees and lists, have \emph{fractal} dimension:
+adding layers to a recursive structure gives us more \emph{fine-grained}
+features, rather than \emph{orthogonal} features. For data mining context-free
+languages (e.g. those of programming and theorem-proving systems), we will
+mainly be concerned with tree structures of variable size.
 
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
+Any investigation of variable-size input would be incomplete without mentioning
+\emph{sequencing}. This is a lossless approach, which splits the input into
+fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at
+a time. The sequence is terminated by a sentinel; an ``end-of-sequence'' marker
+which, by construction, is distinguishable from the data chunks. This technique
+allows us to trade \emph{space} (the size of our input) for \emph{time} (the
+number of chunks in a sequence).
 
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
+Not all ML algorithms can be adapted to accept sequences. One notable approach
+is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections
+between nodes, including cycles. Compared to \emph{feed-forward} ANNs (FFANNs),
+which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on
+its \emph{past inputs} (in fact, RANNs are universal computers).
 
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \cite{hochreiter1997long}).
+The main problem with RANNs, compared to the more widely-used FFANNs, is the
+difficulty of training them. If we extend the standard backpropagation algorithm
+to handle cycles, we get the \emph{backpropagation through time} algorithm
+\cite{werbos1990backpropagation}. However, this suffers a problem known as the
+\emph{vanishing gradient}: error values decay exponentially as they propagate
+back through the cycles, which prevents effective learning of delayed
+dependencies, undermining the main advantage of RANNs. The vanishing gradient
+problem is the subject of current research, with countermeasures including
+\emph{neuroevolution} (using evolutionary computation techniques to train an
+ANN) and \emph{long short-term memory} (LSTM; introducing a few special,
+untrainable nodes to persist values for long time periods
+\cite{hochreiter1997long}).
 
-The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
+The application of \emph{kernel methods} to structured information is discussed
+in \cite{Gartner2003}, where the input data (including sequences, trees and
+graphs) are represented using \emph{generative models}, such as hidden Markov
+models, of a fixed size.
 
@@
 
-\iffalse
-
 Machine learning over structured data:
@@
 Backpropagation through structure
-
-\fi
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/related_tests.tex
More imports

--- 2015-11-06 TransferReport/interesting.tex
@@
-% Since writing this up, it was decided to leave as future work. Still, worth
-% having around for editing into such future work.
+\subsection{Tests} \label{sec:tests}
 
-\documentclass[]{article}
-\begin{document}
+Since we are working in the domain of Haskell programs, an abundant source of
+statements are available in the form of \emph{tests}. For manually-written
+tests, the effort required to write them implies that they must be of some
+interest to their author. Many forms of test are \emph{not} suitable for our
+purposes, such as \emph{unit tests} which are always trivially provable by
+$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
+the external environment. One reason to study Haskell is its widespread use of
+\emph{property checking}, which does give us useful data. Many Haskell property
+checkers exist, based on random testing (\textsc{QuickCheck}
+\cite{claessen2011quickcheck} and \textsc{SmartCheck}
+\cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck}
+\cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck}
+\cite{reich2013advances}) and logic programming (\textsc{SparseCheck}
+\cite{sparsecheck}).
 
-\title{Interestingness Measures for Theory Exploration}
+Thankfully the major differences between these systems are in the way they
+instantiate test arguments; their representations of properties are largely the
+same (modulo renaming of functions and types).
 
-\author{Chris Warburton}
-
-\affil{University of Dundee,\\
-\texttt{http://tocai.computing.dundee.ac.uk}}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
-Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on.
-Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
-\end{abstract}
-
-\section{Introduction}
-
-As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity.
-Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}.
-By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis.
-However, the question still remains: on what should we focus that analysis?
-
-In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain.
-To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}.
-Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.
-
-Appeals to interestingness arise when more direct measures, such as utility, are not available.
-For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users.
-Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \cite{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}.
-Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.
-
-We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context.
-We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape.
-A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}.
-We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness.
-Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.
-
-We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.
-
-\begin{description}
-\item{Interestingness}
-  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}.
-Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
-\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
-
-\section{Exploration in Theorem Proving}
-\label{examples}
-
-Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures.
-An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.
-
-\subsection{Generalisation}
-
-\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case.
-Although it seems counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original.
-This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.
-
-However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false.
-We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.
-
-An informative example is given by Boyer and Moore of the associativity of multiplication in ACL2 \cite{boyer1983proof}:
-
-$$(x * y) * z = x * (y * z)$$
-
-During the course of the proof, the following obligation arises:
-
-\begin{equation}
-  \tag{conc3}
-  (y + (x * y)) * z = (y * z) + ((x * y) * z)
-  \label{eq:conc3}
-\end{equation}
-
-ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:
-
-\begin{equation}
-  \tag{conc4}
-  (y + w) * z = (y * z) + (w * z)
-  \label{eq:conc4}
-\end{equation}
-
-This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.
-It would not be controversial to claim that this distributivity law is interesting in its own right (relative to associativity, at least), in addition to its usefulness in making this proof go through.
-
-% TODO: Describe the ACL2 heuristics
-
-Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}.
-\footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures.
-See \S \ref{auxiliarylemmas} for example definitions of each type.}
-
-\subsection{Analogy}
-
-One way to characterise the interestingness of a statement is by \emph{analogy} to existing interesting statements.
-By finding lemmas analogous to those of a different theory, we may be able to re-use tactics and other forms of meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this problem, however the use of pure exploration misses opportunities to \emph{focus} the search, since we know which lemmas are used in those theories where a technique succeeded.
-If we can find an analogy to map from such solved problems to our unsolved goal, we can infer the approximate form of the lemmas we require, and target these specifically.
-
-The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant to solving a goal $G$ by making analogies via unsupervised clustering \cite{Heras.Komendantskaya.Johansson.ea:2013}.
-These clusters are used in two ways:
-
-\begin{itemize}
-
-  \item First, we use the cluster $C_G$ containing $G$ to identify analogous theorems.
-
-  \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols $S_T$ which occur in $T$ but not in $G$.
-Our analogous lemmas are those used to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of the cluster $C_s$ containing $s$.
-
-\end{itemize}
-
-The running examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for tail-recursive and non-tail-recursive calculations, as well as the effect of repeating certain list operations:
-
-\begin{itemize}
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-\end{itemize}
-
-\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}
-
-One consideration when generating conjectures is the difference between theorems, lemmas, corollaries, etc.
-From a logical point of view, these are all equivalent, and hence most proof assistants do not distinguish between them.
-However, their \emph{intention} may be different: in a sense, theorems are the interesting results; whilst lemmas are useful results, required for proving the theorems.
-
-Some systems, like Coq, allow users to \emph{label} each statement as being a \coq{Theorem}, a \coq{Lemma}, etc.
-despite their internal representations being the same.
-This shows us immediately that lemmas outnumber theorems; in the Coq standard library there are over five times as many lemmas as theorems \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% TODO: Analyse them
-
-% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-We can find a need for auxiliary lemmas, once again, in the context of tail-recursive functions.
-Consider proving the (pointwise) equality of the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Inductive Nat : Set := Z : Nat
-                     | S : Nat -> Nat.
-
-Fixpoint plus      (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => S (plus n' m)
-                                  end.
-
-Fixpoint plus_tail (n m : Nat) := match n with
-                                      | Z    => m
-                                      | S n' => plus_tail n' (S m)
-                                  end.
-\end{lstlisting}
-
-\iffalse
-Haskell equivalent:
-
-plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
-
-plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
-\fi
-
-Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive.
-However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties.
-In particular, when the inductive step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems reasonable), we cannot make this go through using another inductive argument.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-(* Solve equalities by beta-normalising both sides *)
-Ltac triv := try (simpl; reflexivity).
-
-(* Prove equivalence of plus and plus_tail *)
-Theorem equiv : forall n m, plus n m = plus_tail n m.
-  induction n; triv. (* Base case is trivial *)
-
-  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-  intro m.
-
-  (* Beta-reduce the right-hand-side (justification is trivial) *)
-  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-  (* Use induction hypothesis to replace plus_tail with plus *)
-  rewrite <- (IHn (S m)).
-\end{lstlisting}
-
-\fi
-
-Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly the equation we need:
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-\end{lstlisting}
-
-Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-IHn : forall x, plus (S n') x = plus_tail (S n') x
-\end{lstlisting}
-
-We are forced to abandon the proof, despite such a reasonable-looking intermediate goal.
-
-In fact, if we attempt to prove that goal \emph{separately}, we can use a straightforward argument by induction; even though it is actually \emph{stronger} due to the absence of the \coq{IHn} assumption.
-Using this separate result as a lemma, the pointwise equality is proven easily.
-
-\iffalse
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-Lemma gen n m : plus (S n) m = plus n (S m).
-  induction n; triv. (* Base case is trivial *)
-
-  (* Move all S constructors outside *)
-  simpl. rewrite <- IHn. simpl.
-
-  (* Trivial *)
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-  rewrite (gen n m).
-  reflexivity.
-Defined.
-\end{lstlisting}
-
-\fi
-
-\subsection{Tests} \label{tests}
-
-Since we are working in the domain of Haskell programs, an abundant source of statements are available in the form of \emph{tests}.
-For manually-written tests, the effort required to write them implies that they must be of some interest to their author.
-Many forms of test are \emph{not} suitable for our purposes, such as \emph{unit tests} which are always trivially provable by $\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of the external environment.
-One reason to study Haskell is its widespread use of \emph{property checking}, which does give us useful data.
-Many Haskell property checkers exist, based on random testing (\textsc{QuickCheck} \cite{claessen2011quickcheck} and \textsc{SmartCheck} \cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck} \cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck} \cite{reich2013advances}) and logic programming (\textsc{SparseCheck} \cite{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they instantiate test arguments; their representations of properties are largely the same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson}, \texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
+Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of
+2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
+\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
+\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
+\texttt{blaze-html}.
 
 % TODO: Analysis
-
-\section{Related Work}
-\label{related}
-
-\iffalse
-
-\subsection{Conjecture Generation}
-
-The task of \emph{conjecture generation} lies at the heart of theory exploration, and
-
-\subsubsection{Lemma generation}
-
-\fi
-
-\subsection{Exploration in Artificial Intelligence}
-
-Theorem proving is not the only area of AI to have automation extended from problem \emph{solving} to problem \emph{invention}.
-Here we consider similar approaches in a variety of domains, and identify techniques which could be re-purposed in the context of theory exploration.
-
-\subsubsection{Evolutionary Computation}
-
-\emph{Evolutionary computation} is an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \cite{back1997evolutionary}.
-Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise.
-Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \cite{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting.
-
-Traditionally, evolutionary approaches assign solutions a \emph{fitness} value, using a user-supplied \emph{fitness function}.
-Fitness should correlate with how well a solution solves the user's problem; for example, the fitness of a solution to some engineering problem may depend on the estimated materials cost.
-If we frame the task of theory exploration in evolutionary computation terms, the fitness function would be our interestingness measure.
-
-Pure exploration (i.e.
-for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}.
-The former attempts to gain insight into the nature of life and biology through competition over limited resources.
-Whilst this may have utility in resource allocation, e.g.
-efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g.
-\cite{baum2000evolution}).
-
-On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want.
-Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \cite{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g.
-by hill climbing), but few global optima which are hard to find.
-Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \cite{sareni1998fitness}.
-
-One example is \emph{fitness sharing}, which divides up fitness values between identical or similar solutions.
-Say we have a user-provided fitness function $f$, and a population containing two identical solutions $s_1$ and $s_2$; hence $f(s_1) = f(s_2)$.
-In a fitness sharing scheme, we interpret fitness as a fixed resource, distributed according to $f$; when multiple individuals occupy the same point in the solution space, they must \emph{share} the fitness available there.
-We can describe the fitness \emph{allocated} to a solution by augmenting $f$, e.g.
-if we allocate fitness uniformly between identical solutions we get:
-
-$$f'(x) = \frac{f(x)}{\sum_{i=1}^n \delta_{s_i x}}$$
-
-Where $n$ is the population size, $s_i$ is the $i$th solution in the population and $\delta$ is the Kronecker delta function.
-In the example above, assuming there are no other copies in the population, then $f'(s_1) = \frac{f(s_1)}{2} = \frac{f(s_2)}{2} = f'(s_2)$.
-By sharing in this way, the fitness of each solution is balanced against redundancy in the population: there may still be many copies of a solution, but only when the fitness is high enough to justify all of them.
-
-There are many variations on this theme, such as sharing between ``close'' solutions rather than just identical ones and judging distance based on fitness (AKA phenotypically) rather than based on the location in solution space (AKA genetically).
-Yet the underlying principle is always the same: penalise duplication in order to promote diversity.
-This lesson can be carried over to our theory exploration context, where a theorem should be considered less interesting if it is ``close'' to others which have been found.
-
-In a similar way, we can bias our search procedure, rather than our fitness function, towards diversity.
-The search procedure in population-based evolutionary algorithms consists of \emph{selecting} one or more individuals from the population, e.g.
-via truncation (select the best $n$ individuals, discarding the rest); then \emph{transforming} the selected individuals, e.g.
-via mutation and crossover, to obtain new solutions.
-
-\newcommand{\argmin}{\operatornamewithlimits{argmin}}
-
-Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation).
-Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness.
-For example, the fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population.
-An individual $s$ is then selected with fitness closest to $f_t$, i.e.
-$s = \operatornamewithlimits{argmin}\limits_{x} \lvert f(x) - f_t \rvert$
-
-In this way, the fitness function $f$ is used to assign comparable quantities to solutions, but it is not treated as the objective; instead, the implicit objective is to maintain a diverse population, with individuals spread out uniformly in fitness space.
-This approach seems useful for informing our work in theory exploration, as it supports search criteria which \emph{describe} solutions, but which we may not want to \emph{optimise}.
-As a simple example, we might distinguish different forms of theorem by measuring how balanced their syntax trees are (-1 for left-leaning, +1 for right leaning, 0 for balanced); but it would be senseless to \emph{maximise} how far they lean.
-
-Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original?
-This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \cite{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains.
-Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.
-
-\subsubsection{Artificial Curiosity}
-\label{curiosity}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \cite{schmidhuber2006developmental}.
-Although framed in the context of \emph{reinforcement learning}, this is clearly relevant to our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do.
-The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible.
-The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.
-
-We can divide AC approaches into two groups: the first, which we call \emph{explicit}, send inputs which follow a Wundt curve to their learning algorithm; the second, the \emph{implicit} approaches, instead modify the \emph{output} of their learning algorithm(s), such that the overall system follows a Wundt curve as an emergent property.
-
-In the explicit case, the \emph{implicit reward} signals being learned are analogous to our notion of interestingness.
-A framework encompassing many examples is given in \cite{oudeyer2007intrinsic} in the context of reinforcement learning.
-
-One particularly general measure is \emph{compression progress}: given a compressed representation of our previous observations, the ``progress'' is the space saved if we include the current observation.
-Observations which are incompressible or trivially compressible don't save any space, whilst observations which provide new information relevant to past experience can provide a saving.
-This can be translated to a theorem proving context very naturally: our observations are theorems and their proofs, whilst new theorems which generalise known results will allow us to compress their proofs.
-
-% TODO: Examples
-\cite{Schmidhuber1999}
-
-Two sources of intrinsic reward are proposed in \cite{Hester.Stone:2012} for \emph{random forests}.
-A random forest is a population of decision trees, where each tree is trained on a sub-set of the available examples, each decision is made using a sub-set of the available features, and the predictions of every tree are averaged to obtain that of the forest \cite{randomforests}.
-The first intrinsic reward is the \emph{disagreement} between predictions; for a forest with $m$ models (trees), predicting features $x_1$ \dots $x_n$ of the state resulting from taking action $a$ in state $s$, we simply sum the Kullback-Leibler divergence $D_{\rm KL}$ of each prediction $P_1$ \dots $P_m$ from every other prediction:
-
-\begin{equation}
-  D(s,a) = \sum_{i = 1}^n \sum_{j = 1}^m \sum_{k = 1}^m D_{KL}(P_j(x_i|s,a) || P_k(x_i|s,a))
-\end{equation}
-
-$D(s,a)$ is an explicit AC reward, as it follows a Wundt curve as the complexity of transitions increases.
-For parts of the state space which have been fully learned, the models will agree on accurate predictions.
-For parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the average of past observations; these predictions may not be accurate, but they will be in agreement.
-Hence it is the states which are amenable to learning which produce the largest disagreement.
-
-The second intrinsic reward is simply a measure of distance from previous observations, which pushes the system towards unseen states regardless of how learnable they are (similar to the $R_{max}$ technique).
-This is too simple to meet our definition of AC, but it does force the models to generalise their predictions to unexplored states, acting to increase disagreement in the forest.
-
-A key advantage of random forests is that their models are \emph{inspectable}: they not only give predictions, but also \emph{reasons} for those predictions (i.e.
-we can see which paths are taken through each decision tree).
-% TODO: The accuracy of these random forest models are compared
-
-% TODO
-\cite{Kaplan2006}
-\cite{Lipson2007}
-\cite{Luciw2011}
-\cite{Macedo2000}
-\cite{Ramik.Sabourin.Madani:2013}
-\cite{Roa.Kruijff.Jacobsson:2009}
-\cite{Schaul.Sun.Wierstra.ea:2011}
-\cite{Schmidhuber1999}
-\cite{Schmidhuber:1991}
-\cite{Scott1989}
-\cite{Steunebrink.Koutnik.Thorisson.ea:2013}
-\cite{maher2008achieving}
-\cite{meyer1991possibility}
-\cite{oudeyer2004intelligent}
-\cite{oudeyer2014evolution}
-\cite{schmidhuber2006developmental}
-
-% TODO: Coevolution
-
-Whilst clearly of relevance to theory exploration, artificial curiosity is usually framed in the context of a \emph{reinforcement learning} and \emph{intrinsic reward}, especially in the field of developmental robotics.
-This requires non-trivial choices to be made in deciding which of its concepts are of relevance to our domain, and how they may be translated across.
-For example, much of developmental robotics studies continuous, real-valued sensorimotor signals which may not have any direct analogue in the manipulation of logical formulae.
-However, if we take a higher-level view, the study of such signals may provide insight for predicting and tuning the behaviour of off-the-shelf ATP algorithms.
-
-The most obvious contrast between developmental robotics and theory exploration is that the latter is not physically embodied (e.g.
-in a robot).
-Embodiment has been proposed as a necessary property of intelligent systems, as it provides \emph{grounding} \cite{anderson2003embodied}.
-Embodiment emerged as a response to the symbolic techniques of GOFAI, and in this sense the fields of theory exploration and developmental robotics seem incompatible.
-Nevertheless, TE can be seen to avoid the problems of GOFAI in two ways:
-
-\begin{itemize}
-
-  \item Firstly, the abstract, mathematical domain being explored is not a \emph{model} of some external, physical environment; the domain \emph{is} our environment; hence there is no issue of grounding terms with some external meaning.
-
-  \item Secondly, there is a physical aspect of TE in that \emph{resource usage} is a critical factor.
-If it weren't, then brute force enumeration of proofs would be a viable solution.
-In this sense, we can provide physical inputs to our algorithms, such as measures of time and space used.
-
-\end{itemize}
-
-\iffalse
-
-\subsubsection{Universal Drives}
-
-PhysRevLett.110.168702.pdf
-Omohundro? Too physical.
-\emph{Universal drives} are those
-
-\fi
-
-\subsection{Statistics of Formal Systems}
-
-The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems.
-This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{relevance}
-
-% TODO
-\cite{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods, such as resolution, to have exponential complexity \cite{haken1985intractability}; hence even a modest size increase can turn a trivial problem into an intractable one.
-Finding efficient alternatives for such algorithms, especially those which are NP-complete (e.g.
-determining satisfiability) or co-NP-complete (e.g.
-determining tautologies), seems unlikely, as it would imply progress on the famously intractable open problems of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$.
-On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one.
-We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information.
-This leads to the idea of \emph{relevance filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed \emph{irrelevant}.
-The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is the number of symbols in the clause and $m$ is the number which are relevant).
-Initially, the relevant symbols are those which occur in the goal, but whenever a clause is found which scores more than a particular threshold, all of its symbols are then also considered relevant.
-There are other heuristics applied too, such as increasing the score of user-provided facts (e.g.
-given by keywords like \texttt{using}), locally-scoped facts, first-order facts and rarely-occuring facts.
-To choose $r$ relevant clauses for an ATP invocation, we simply order the clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated, including:
-
-\begin{description}
-
-  \item{\textsc{MaSH}}: Machine Learning for SledgeHammer \cite{kuhlwein2013mash}.
-The distinguishing feature of \textsc{MaSH} is its use of ``visibility'', which is essentially a dependency graph of which theorems were used in the proofs of which other theorems; although theorems are represented as abstract sets of features.
-To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.
-
-  \item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine (SVM) approach for selecting relevant axioms from the Mizar Mathematical Library for use by the Vampire ATP system \cite{alama2014premise}.
-\iffalse TODO: describe the kernel, as that's the interesting bit \fi It compares favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-  \item{\textsc{SInE}}
-  \item{\textsc{BliStr}}
-  \item{\textsc{HOLyHammer}}
-  \item{\textsc{MoMM}}
-  \item{\textsc{SNoW}}
-  \item{\textsc{MPTP 0.2}}
-  \item{\textsc{MaLARea}}
-  \item{\textsc{MaLARea SG1}}
-
-\end{description}
-
-\subsubsection{Clustering}
-\label{clustering}
-
-% TODO: ML4PG
-\cite{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\cite{heras2013proof}
-
-\subsubsection{Probability of Sentences}
-
-The most important property of a logical formula is its truth value.
-Although we may be able to determine some truth values exactly, e.g.
-using decision or semi-decision procedures, it may be more efficient to \emph{approximate} truth values.
-One straightforward extension of truth values is \emph{probabilities}, where we can assign probability $1$ to formulae which are known to be true, $0$ to formulae known to be false, and intermediate values to those which we do not yet know.
-
-% TODO
-\cite{Hutter.Lloyd.Ng.ea:2013}
-
-\subsubsection{Interestingness in Concept Formation}
-\label{conceptformation}
-
-% TODO:
-\cite{Montano-Rivas.McCasland.Dixon.ea:2012}
-\cite{Piantadosi.Tenenbaum.Goodman:2012}
-\cite{Wille:2005}
-\cite{colton1999automatic}
-\cite{colton2000agent}
-\cite{colton2012automated}
-\cite{lenat1977automated}
-\cite{mullerunderstanding}
-\cite{Bundy.Cavallo.Dixon.ea:2015}
-\cite{johansson2009isacosy}
-\cite{spector2008genetic}
-\cite{colton2012automated}
- \cite{geng2006interestingness}
-% TODO: How does https en.wikipedia.org/wiki/Discovery system relate?
-
-% However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.
-
-% Alan Bundy et al
-
-% Eurisko, AM, etc.?
-
-\subsubsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects.
-In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects.
-Such nested structures may extend to arbitrary depth, which makes them difficult to represent with a fixed number of features, as is expected by most machine learning algorithms.
-Here we review a selection of solutions to this problem, and compare their distinguishing properties.
-
-\paragraph{Feature Extraction}\label{featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning (ML).
-Rather than feeding ``raw'' data straight into our ML algorithm, we only learn a sample of \emph{relevant} details, known as \emph{features}.
-This has two benefits:
-
-\begin{itemize}
-  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}.
-This reduces the size of the ML problem, improving efficiency (e.g.
-running time).
-  \item We avoid learning irrelevant details such as the encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation.
-Many ML algorithms only work with fixed-size inputs; for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294} algorithm works on models with \emph{fixed} topology (e.g.
-\emph{artificial neural networks} with fixed connections between nodes).
-This requires some form of pre-processing in domains where the size of each input is not known, may vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make predictions/decisions before seeing all of the inputs.
-Unbounded input appears in domains such as programming and theorem proving, where individual term may be trees of unbounded depth.
-In these situations we need a mapping from arbitrary inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program fragments:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{lstlisting}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
-\end{itemize}
-
-However, this is too optimistic.
-Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits.
-Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited.
-Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.
-
-There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor.
-Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each input.
-If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning).
-This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase to improve efficiency: first we learn a compact representation for our data, for example using \emph{autoencoding}; then we use that encoder as a feature extractor for our main learning task.
-This stacking of one learning algorithm on top of another, especially with greedy learning of each layer, has lead to the recent trend of \emph{deep learning}.
-
-\paragraph{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger than a particular size (and pad anything smaller).
-This is the approach taken by ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at most 10 levels and 10 elements per level; each tree is converted to a $30 \times 10$ matrix (3 values per tree node) and learning takes place on these normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with \emph{time} efficiency.
-Specifically, truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian'').
-The less these assumptions hold, the less we can truncate.
-Since many ML algorithms scale poorly with input size, we would prefer to eliminate the redundancy using a more aggressive algorithm, to keep the resulting feature size as low as possible.
-
-\paragraph{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view it as a \emph{dimension reduction} technique: our inputs can be modelled as points in high-dimensional spaces, which we want to project into a lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit vectors).
-
-Truncation is a trivial dimension reduction technique: take the first $N$ coordinates (bits).
-More sophisticated projection functions consider the \emph{distribution} of the points, and project with the hyperplane which preserves as much of the variance as possible (or, equivalently, reduces the \emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle component analysis} (PCA) and \emph{autoencoding}; however, since these techniques are effectively ML algorithms in their own right, they suffer some of the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to avoid.
-Sophisticated dimension reduction is still useful for \emph{compressing} large, redundant features into smaller, information-dense representations, and as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome, since any learning we perform for feature extraction will interfere with the core learning algorithm that's consuming these features (this is why deep learning is often done greedily).
-
-\paragraph{Sequencing}
-
-The task of dimension reduction changes when we consider \emph{structured} data.
-Recursive structures, like trees and lists, have \emph{fractal} dimension: adding layers to a recursive structure gives us more \emph{fine-grained} features, rather than \emph{orthogonal} features.
-For data mining context-free languages (e.g.
-those of programming and theorem-proving systems), we will mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning \emph{sequencing}.
-This is a lossless approach, which splits the input into fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at a time.
-The sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by construction, is distinguishable from the data chunks.
-This technique allows us to trade \emph{space} (the size of our input) for \emph{time} (the number of chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences.
-One notable approach is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections between nodes, including cycles.
-Compared to \emph{feed-forward} ANNs (FFANNs), which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the difficulty of training them.
-If we extend the standard backpropagation algorithm to handle cycles, we get the \emph{backpropagation through time} algorithm \cite{werbos1990backpropagation}.
-However, this suffers a problem known as the \emph{vanishing gradient}: error values decay exponentially as they propagate back through the cycles, which prevents effective learning of delayed dependencies, undermining the main advantage of RANNs.
-The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train an ANN) and \emph{long short-term memory} (LSTM; introducing a few special, untrainable nodes to persist values for long time periods \cite{hochreiter1997long}).
-
-The application of \emph{kernel methods} to structured information is discussed in \cite{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.
-
-% TODO
-\cite{Gartner2003}
-\cite{Oveisi.Oveisi.Erfanian.ea:2012}
-\cite{bakir2007predicting}
-\cite{conf/ijcai/Plate91}
-\cite{goller1996learning}
-\cite{kwasny1995tail}
-\cite{pollack1990recursive}
-\cite{zanzotto2012distributed}
-
-\iffalse
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure
-
-\fi
-
-\bibliographystyle{plain}
-\bibliography{../Bibtex}
-
-\end{document}

### 2018-05-17 PhDThesis/litreview_premiseselection.tex
Reduce some duplication

@@
-\subsection{Relevance Filtering}
+\subsubsection{Relevance Filtering}
 \label{sec:relevance}
+
+% TODO
+\cite{kuhlwein2012overview}
 
@@
 \emph{redundant} information. This leads to the idea of \emph{relevance
-  filtering} (or, \emph{premise selection}, when viewed as the \emph{addition}
-of relevant information to an initially-empty problem). This is the core idea
-behind our restriction of theory exploration to intelligently-selected clusters
-of symbols, rather than whole libraries at a time.
+  filtering}.
 
-Relevance filtering has mostly been used in automated proof search, where it
-simplifies problems by removing from consideration those clauses (axioms,
-definitions, lemmas, etc.) which are deemed \emph{irrelevant}. The technique is
-used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL
-theories to statements in first order logic: rather than translating the entire
-theory, only a sub-set of relevant clauses are included. This reduces the size
-of the problem and speeds up the proof search, but it creates the new problem of
-determining when a clause is relevant: how do we know what will be required,
-before we have the proof?
+Relevance filtering simplifies a proof search problem by removing from
+consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed
+\emph{irrelevant}. The technique is used in Sledgehammer during its translation
+of Isabelle/HOL theories to statements in first order logic: rather than
+translating the entire theory, only a sub-set of relevant clauses are
+included. This reduces the size of the problem and speeds up the proof search,
+but it creates the new problem of determining when a clause is relevant: how do
+we know what will be required, before we have the proof?
 
-The initial approach taken by Sledgehammer, known as \textsc{MePo} (from
-\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score
-based on the proportion $\frac{m}{n}$ of its symbols which are ``relevant''
-(where $n$ is the number of symbols in the clause and $m$ is the number which
-are relevant). Initially, the relevant symbols are those which occur in the goal
-to be proved, but whenever a clause is found which scores more than a particular
-threshold, all of its symbols are then also considered relevant. There are other
-heuristics applied too, such as increasing the score of user-provided facts
-(e.g. given by keywords like \texttt{using}), locally-scoped facts, first-order
-facts and rarely-occuring facts. To choose $r$ relevant clauses for an ATP
-invocation, we simply order the clauses by decreasing score and take the first
-$r$ of them.
+The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson}
+\cite{meng2009lightweight}), gives each clause a score based on the proportion
+$m / n$ of its symbols which are ``relevant'' (where $n$ is the number of
+symbols in the clause and $m$ is the number which are relevant). Initially, the
+relevant symbols are those which occur in the goal, but whenever a clause is
+found which scores more than a particular threshold, all of its symbols are then
+also considered relevant. There are other heuristics applied too, such as
+increasing the score of user-provided facts (e.g. given by keywords like
+\texttt{using}), locally-scoped facts, first-order facts and rarely-occuring
+facts. To choose $r$ relevant clauses for an ATP invocation, we simply order the
+clauses by decreasing score and take the first $r$ of them.
 
-Recently, a variety of alternative algorithms have also been investigated, for
-example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer)
-\cite{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to
-determine the relevance of clauses. Visibility is essentially a dependency graph
-of which theorems were used in the proofs of which other theorems (although the
-theorems are actually represented as abstract sets of features). To select
-relevant clauses for a goal, the set of clauses which are visible from the
-goal's components is generated; this is further reduced by (an efficient
-approximation of) a na\"{\i}ve Bayes algorithm.
+Recently, a variety of alternative algorithms have also been investigated,
+including:
 
-Another example is \emph{multi-output ranking} (MOR), which uses a support
-vector machine (SVM) approach for selecting relevant axioms from the Mizar
-Mathematical Library for use by the Vampire ATP system
-\cite{alama2014premise}. Many more approaches are described and evaluated in
-\cite{kuhlwein2012overview}, some of which may be directly applicable in the
-context of theory exploration.
+\begin{description}
+\item{\textsc{MaSH}}: Machine Learning for SledgeHammer
+  \cite{kuhlwein2013mash}. The distinguishing feature of \textsc{MaSH} is its
+  use of ``visibility'', which is essentially a dependency graph of which
+  theorems were used in the proofs of which other theorems; although theorems
+  are represented as abstract sets of features. To select relevant clauses for a
+  goal, the set of clauses which are visible from the goal's components is
+  generated; this is further reduced by (an efficient approximation of) a naive
+  Bayes algorithm.
+
+\item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine
+  (SVM) approach for selecting relevant axioms from the Mizar Mathematical
+  Library for use by the Vampire ATP system \cite{alama2014premise}. \iffalse
+  TODO: describe the kernel, as that's the interesting bit \fi It compares
+  favourably to \textsc{SNoW} and \textsc{SInE}.
+
+  % TODO:
+\item{\textsc{SInE}}
+\item{\textsc{BliStr}}
+\item{\textsc{HOLyHammer}}
+\item{\textsc{MoMM}}
+\item{\textsc{SNoW}}
+\item{\textsc{MPTP 0.2}}
+\item{\textsc{MaLARea}}
+\item{\textsc{MaLARea SG1}}
+
+\end{description}

### 2018-05-17 PhDThesis/related_formalstats.tex
Reduce some duplication

@@
 particular importance for theory exploration.
-
-\subsubsection{Relevance Filtering}
-\label{sec:relevance}
-
-% TODO
-\cite{kuhlwein2012overview}
-
-The combinatorial nature of formal systems causes many proof search methods,
-such as resolution, to have exponential complexity
-\cite{haken1985intractability}; hence even a modest size increase can turn a
-trivial problem into an intractable one. Finding efficient alternatives for such
-algorithms, especially those which are NP-complete (e.g. determining
-satisfiability) or co-NP-complete (e.g. determining tautologies), seems
-unlikely, as it would imply progress on the famously intractable open problems
-of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$. On the other hand, we
-can turn this difficulty around: a modest \emph{decrease} in size may turn an
-intractable problem into a solvable one. We can ensure that the solutions to
-these reduced problems coincide with the original if we only remove
-\emph{redundant} information. This leads to the idea of \emph{relevance
-  filtering}.
-
-Relevance filtering simplifies a proof search problem by removing from
-consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed
-\emph{irrelevant}. The technique is used in Sledgehammer during its translation
-of Isabelle/HOL theories to statements in first order logic: rather than
-translating the entire theory, only a sub-set of relevant clauses are
-included. This reduces the size of the problem and speeds up the proof search,
-but it creates the new problem of determining when a clause is relevant: how do
-we know what will be required, before we have the proof?
-
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson}
-\cite{meng2009lightweight}), gives each clause a score based on the proportion
-$m / n$ of its symbols which are ``relevant'' (where $n$ is the number of
-symbols in the clause and $m$ is the number which are relevant). Initially, the
-relevant symbols are those which occur in the goal, but whenever a clause is
-found which scores more than a particular threshold, all of its symbols are then
-also considered relevant. There are other heuristics applied too, such as
-increasing the score of user-provided facts (e.g. given by keywords like
-\texttt{using}), locally-scoped facts, first-order facts and rarely-occuring
-facts. To choose $r$ relevant clauses for an ATP invocation, we simply order the
-clauses by decreasing score and take the first $r$ of them.
-
-Recently, a variety of alternative algorithms have also been investigated,
-including:
-
-\begin{description}
-\item{\textsc{MaSH}}: Machine Learning for SledgeHammer
-  \cite{kuhlwein2013mash}. The distinguishing feature of \textsc{MaSH} is its
-  use of ``visibility'', which is essentially a dependency graph of which
-  theorems were used in the proofs of which other theorems; although theorems
-  are represented as abstract sets of features. To select relevant clauses for a
-  goal, the set of clauses which are visible from the goal's components is
-  generated; this is further reduced by (an efficient approximation of) a naive
-  Bayes algorithm.
-
-\item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine
-  (SVM) approach for selecting relevant axioms from the Mizar Mathematical
-  Library for use by the Vampire ATP system \cite{alama2014premise}. \iffalse
-  TODO: describe the kernel, as that's the interesting bit \fi It compares
-  favourably to \textsc{SNoW} and \textsc{SInE}.
-
-  % TODO:
-\item{\textsc{SInE}}
-\item{\textsc{BliStr}}
-\item{\textsc{HOLyHammer}}
-\item{\textsc{MoMM}}
-\item{\textsc{SNoW}}
-\item{\textsc{MPTP 0.2}}
-\item{\textsc{MaLARea}}
-\item{\textsc{MaLARea SG1}}
-
-\end{description}
 

### 2018-05-17 PhDThesis/related_formalstats.tex
More deduplication

@@
   pattern).
+\item All of our feature vectors will be the same size, i.e. they will all have
+  length (or \emph{dimension}) $d$. Many machine learning algorithms only work
+  with inputs of a uniform size; feature extraction allows us to use these
+  algorithms in domains where the size of each input is not known, may vary or
+  may even be unbounded. For example, element-wise comparison of feature vectors
+  is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions
+  this is not so straightforward, as their nesting may give rise to very
+  different shapes.
+\item Unlike our expressions, which are discrete, we can continuously transform
+  one feature vector into another. This enables many powerful machine learning
+  algorithms to be used, such as those based on \emph{gradient descent} or, in
+  our case, arithmetic means.
+\item Feature vectors can be chosen to represent the relevant information in a
+  more compressed form than the raw data; for example, we might replace verbose,
+  descriptive identifiers with sequential numbers. This reduces the input size
+  of the machine learning problem, improving efficiency.
 \end{itemize}
@@
 
-\paragraph{Sequencing}
+\paragraph{Structured Data}
 
@@
 
+Using sequences to represent recursive structures is also problematic: if we
+want our learning algorithm to exploit structure (such as the depth of a token),
+it will have to discover how to parse the sequences for itself, which seems
+wasteful. The \emph{back-propagation through structure} approach
+\cite{goller1996learning} is a more direct solution to this problem, using a
+feed-forward NN to learn recursive distributed representations
+\cite{pollack1990recursive} which correspond to the recursive structure of the
+inputs. Such distributed representations can also be used for sequences, which
+we can use to encode sub-trees when the branching factor of nodes is not uniform
+\cite{kwasny1995tail}. More recent work has investigated storing recursive
+structures inside LSTM cells \cite{zhu2015long}.
+
+A simpler alternative for generating recursive distributed representations is to
+use circular convolution \cite{conf/ijcai/Plate91}. Although promising results
+are shown for its use in \emph{distributed tree kernels}
+\cite{zanzotto2012distributed}, our preliminary experiments in applying
+circular convolution to functional programming expressions found most of the
+information to be lost in the process; presumably as the expressions are too
+small.
+
 The application of \emph{kernel methods} to structured information is discussed
@@
 models, of a fixed size.
+
+\emph{Kernel methods} have also been applied to structured information, for
+example in \cite{Gartner2003} the input data (including sequences, trees and
+graphs) are represented using \emph{generative models}, such as hidden Markov
+models, of a fixed size suitable for learning. Many more applications of kernel
+methods to structured domains are given in \cite{bakir2007predicting}, which
+could be used to learn more subtle relations between expressions than recurrent
+clustering alone.
 
@@
 \cite{zanzotto2012distributed}
-
-Machine learning over structured data:
-1D is common: parsing natural language
-2D is common; images
-Trees are fractal
-Backpropagation through structure
-LSTM with recursive structure
-Most work tries to identify structure; we already have it
-
-Recurrent neural networks
-Backpropagation through structure

### 2018-05-17 PhDThesis/related_formalstats.tex
More dedupe

@@
 
-\paragraph{Structured Data}
+\paragraph{Sequences}
 
-The task of dimension reduction changes when we consider \emph{structured}
-data. Recursive structures, like trees and lists, have \emph{fractal} dimension:
-adding layers to a recursive structure gives us more \emph{fine-grained}
-features, rather than \emph{orthogonal} features. For data mining context-free
-languages (e.g. those of programming and theorem-proving systems), we will
-mainly be concerned with tree structures of variable size.
-
-Any investigation of variable-size input would be incomplete without mentioning
-\emph{sequencing}. This is a lossless approach, which splits the input into
-fixed-size \emph{chunks}, which are fed into an appropriate ML algorithm one at
-a time. The sequence is terminated by a sentinel; an ``end-of-sequence'' marker
-which, by construction, is distinguishable from the data chunks. This technique
-allows us to trade \emph{space} (the size of our input) for \emph{time} (the
-number of chunks in a sequence).
+To handle input of \emph{variable} size, most attention has been given to the
+handling of \emph{sequences}. This is a lossless approach, which splits the
+input into fixed-size \emph{chunks}, which are fed into an appropriate ML
+algorithm one at a time. The sequence is terminated by a sentinel; an
+``end-of-sequence'' marker which, by construction, is distinguishable from the
+data chunks. This technique allows us to trade \emph{space} (the size of our
+input) for \emph{time} (the number of chunks in a sequence).
 
@@
 \cite{hochreiter1997long}).
+
+\paragraph{Recursive Structure}
+
+Recursive structures, like trees and lists, have \emph{fractal} dimension:
+adding layers to a recursive structure gives us more \emph{fine-grained}
+features, rather than \emph{orthogonal} features. For data mining context-free
+languages (e.g. those of programming and theorem-proving systems), we will
+mainly be concerned with tree and graph structures of variable size.
 

### 2018-05-17 TEBenchmarkPaper/article.tex
Fiddling

@@
 putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup is given in
-table~\ref{table:speedups}. We would predict the speedup to grow for larger
-sample sizes, but the increasing proportion of timeouts causes our estimates to
-become more conservative: by size 20 most runs are timing out, and hence are
-tied.
+98.2\% with 95\% confidence. The magnitude of each speedup (median IsaCoSy time
+divided by median QuickSpec time) is given in table~\ref{table:speedups}. We
+would predict the speedup to grow for larger sample sizes, but the increasing
+proportion of timeouts causes our estimates to become more conservative: by size
+20 most runs are timing out, and hence are tied.
 
@@
   \caption{Speedup from using QuickSpec compared to IsaCoSy, i.e. the median
-    time taken by IsaCoSy divided by that of QuickSpec. All were deemed
-    statistically significant QuickSpec was significantly faster
-    ($\alpha = 0.05$) for all sizes, except 1 and 2 due to too few samples.
-    These are conservative estimates, since the 300 second time limit acts to
-    reduce the measured time difference (e.g. sizes 13 and 20, where the median
-    for both tools timed out).}
+    time taken by IsaCoSy divided by that of QuickSpec. QuickSpec was found to
+    be significantly faster ($\alpha = 0.05$) for all sizes, except 1 and 2 due
+    to too few samples. These are conservative estimates, since the 300 second
+    time limit acts to reduce the measured time difference (e.g. sizes 13 and
+    20, where the median for both tools timed out).}
   \label{table:speedups}
@@
 results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth):
+are ``uninteresting'' (do not appear in the ground truth), with the results
+shown in table~\ref{table:precision}
 
@@
 
-We test these totals for independence using Boschloo's test and find a p-value
+We tested these totals for independence using Boschloo's test and find a p-value
 of 0.111, which exceeds our (conventional) significance threshold of

### 2018-05-21 TEBenchmarkPaper/article.tex
Minor changes

@@
 \usepackage{csvsimple}
+\usepackage[inline, shortlabels]{enumitem}
 \usepackage{float}
@@
     maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant:
-  since the library authors have gone to the effort of stating and proving these
-  properties, and including them with every copy of their software, this is a good
+  three were taken from the standard library of the Isabelle proof assistant.
+  The fact that the library authors expended the effort of stating, proving and
+  bundling these properties with every copy of their software is a good
   indication that they are interesting. Two sets of definitions were used: a
@@
 
-Once such a problem set has been chosen, we must separate the theorem statements
-from the definitions they reference. The latter will be used as input to the
-MTE tools, whilst the former constitute the ground truth corpus of properties
-against which to compare the output.
+Once such a problem set has been chosen, we must separate the definitions from
+the theorem statements which reference those definitions. The definitions will
+be used as input to the MTE tools, whilst the theorem statements form the ground
+truth corpus of properties (which tool output will be compared against).
 
@@
   infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on ``large'' inputs,
-  containing more than a few dozen definitions.
+\item It is impractical to run existing MTE tools on inputs containing more
+  than a few dozen definitions.
 \end{itemize}
@@
 chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes). We use wall-clock time since this is most
-relevant to a user's experience, it has a straightforward interpretation and
-tracking it does not require intrusive alterations to a tool's implementation.
+with a timeout (e.g. 5 minutes). We use wall-clock time since
+\begin {enumerate*} [i) ]%
+\item this is most relevant to a user's experience,
+\item it has a straightforward interpretation and
+\item measuring it does not require intrusive alterations to a tool's
+  implementation.
+\end {enumerate*}
 The downside is that comparisons must take hardware performance into account,

### 2018-05-22 TEBenchmarkPaper/article.tex
TEB tweaks (from Jianguo)

@@
   inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By providing a
-  standard, cross-tool evaluation technique we allow the disparate approaches
-  currently being pursued to be more directly compared. Our hope is to encourage
-  innovation and improvements similar to those seen in fields like automated
-  theorem proving, where the availability of benchmarks encourages healthy
-  competition.
+  and produces significantly more ``interesting'' output. By defining a standard
+  benchmark we provide a measure of progress for the field, encourage
+  innovation through healthy competition and allow direct comparisons to be made
+  between the disparate approaches currently being pursued for this task.
 %\keywords{First keyword \and Second keyword \and More}
@@
   use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
-and provide only those as input to the tool. We generate a corresponding
-ground truth by selecting those properties from the corpus which ``depend on''
-(contain references to) \emph{only} the definitions in that sample. Transitive
-dependencies aren't required (e.g. a property involving only a \texttt{times}
-function would not depend on a \texttt{plus} function, even if \texttt{plus}
-occurs in the definition of \texttt{times}).
+  ``cheating'').} Given a sample size, we choose a subset of that many
+definitions, and provide only those as input to the tool. We generate a
+corresponding ground truth by selecting those properties from the corpus which
+``depend on'' (contain references to) \emph{only} the definitions in that
+sample. Transitive dependencies aren't required (e.g. a property involving only
+a \texttt{times} function would not depend on a \texttt{plus} function, even if
+\texttt{plus} occurs in the definition of \texttt{times}).
 
@@
 statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an empty
-ground truth, and thus $0$ precision and undefined recall \emph{independent} of
-the tool's output! This is clearly undesirable as an evaluation method.
+a property's dependencies. The majority of such samples would hence have an
+empty ground truth, and thus $0$ precision and undefined recall
+\emph{independent} of the tool's output! This is clearly undesirable as an
+evaluation method.
 

### 2018-05-28 PhDThesis/benchmark.tex
Add supporting materials, taken from TEB

--- 2018-05-22 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
-\usepackage{csvsimple}
-\usepackage[inline, shortlabels]{enumitem}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
-\usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
-
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
-
-% For example theory of lists
-\newcommand{\function}{\rightarrow}
-
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By defining a standard
-  benchmark we provide a measure of progress for the field, encourage
-  innovation through healthy competition and allow direct comparisons to be made
-  between the disparate approaches currently being pursued for this task.
-%\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{intro}
-
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems (TIP) benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We introduce the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-\begin{sloppypar}
-  Our benchmark, including the scripts used to generate it from the TIP problem
-  set, is available online at
-  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
-  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
-  \url{https://github.com/warbo/isaplanner-tip}.
-\end{sloppypar}
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-\begin{sloppypar}
-  The precision and recall of three MTE tools are compared
-  in~\cite{claessen2013automating}:
-  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
-  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-    comparison only reports prior results for each tool, rather than reproducing
-    them. This makes the numbers less comparable, since each tool was tested
-    with slightly different definitions (e.g. whether or not the exponential and
-    maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant.
-  The fact that the library authors expended the effort of stating, proving and
-  bundling these properties with every copy of their software is a good
-  indication that they are interesting. Two sets of definitions were used: a
-  theory of natural numbers; and a theory of lists (which we show in
-  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
-  only two datapoints, each containing only a few definitions and properties (4
-  functions and 12 properties, in the natural number example), it is difficult
-  to draw general conclusions; yet despite these limitations we believe that
-  this is a viable method to performing meaningful, objective analyses and
-  comparisons for such tools. It is only with larger data sets that we can
-  generalise the measured performance to different, especially \emph{novel},
-  domains (which, after all, is the purpose of MTE tools). In addition, we might
-  be concerned that such ``popular'' theories as the natural numbers may be
-  unrepresentative of typical usage. For example, we would not expect the
-  library functions in a typical verification problem to be as mathematically
-  rich as number theory.
-\end{sloppypar}
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
 \section{Theory Exploration Benchmark}
@@
 scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
-
-The automation of mathematical tasks has been pursued since at least the time of
-mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
-these efforts is the separation between those undertaken by mathematicians like
-Pascal and Babbage~\cite{bowden}, and those of engineers such as
-M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
-are concerned with (automatically constructing and evaluating concepts,
-conjectures, theorems, axioms, examples, etc.) being divided into two main
-fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
-sometimes prefaced with ``Computer-Aided'', ``Automated'' or
-``Algorithm-Supported''), which is championed by mathematicians such as
-Buchberger~\cite{buchberger}; and Automated Theory Formation
-(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
-Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
-``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}.
-
-Such a plethora of terminology can mask similarities and shared goals between
-these fields. Even notable historical differences, such as the emphasis of MTE
-on user-interaction and mathematical domains, in contrast to the full automation
-and more general applications targeted by ATF, are disappearing in recent
-implementations.
-
-An important historical implementation of ATF is Lenat's AM (Automated
-Mathematician) system. Unlike prior work, such as
-Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
-to be a general purpose mathematical discovery system, designed to both
-construct new concepts and conjecture relationships between them. AM is a
-rule-based system which represents knowledge using a frame-like scheme, enlarges
-its knowledge base via a collection of heuristic rules, and controls the firing
-of these rules via an agenda mechanism. Evaluation of AM considered generality
-(performance in new domains) and how finely-tuned various aspects of the program
-are (the agenda, the interaction of the heuristics, etc). Most of this
-evaluation was qualitative, and has subsequently been
-criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
-Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
-natural numbers from sets, but did so using a heuristic specifically designed to
-make this connection.
-
-The prototypical implementation of MTE is the Theorema system of Buchberger and
-colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
-emphasis on user interface and output presentation. Theory exploration in the
-Theorema system involves the user formalising their definitions in a consistent,
-layered approach; such that reasoning algorithms can exploit this structure in
-subsequent proofs, calculations, etc. The potential of this strategy was
-evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
-bases algorithm~\cite{buchberger:04}.
-
-A similar ``layering'' approach is found in the IsaScheme system of
-Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
-has also been quantitatively compared against IsaCoSy and HipSpec using
-precision/recall analysis~\cite{claessen2013automating}. The name comes from its
-embedding in the Isabelle proof assistant and its use of ``schemes'':
-higher-order formulae which can be used to generate new concepts and
-conjectures. Variables within a scheme are instantiated automatically and this
-drives the invention process. For example, the concept of ``repetition'' can be
-encoded as a scheme, and instantiated with existing encodings of zero, successor
-and addition to produce a definition of multiplication. The same scheme can be
-instantiated with this new multiplication function to produce exponentiation.
-
-IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
-described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
-has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
-distinct enumeration and testing steps with a single, iterative algorithm
-similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
-completion algorithm to form a corresponding set of rewrite rules. As
-expressions are enumerated, they are simplified using these rules and discarded
-if equal to a known expression. If not, QuickCheck tests whether the new
-expression can be distinguished from the known expressions through random
-testing: those which can are added to the set of known expressions. Those which
-cannot be distinguished are conjectured to be equal, and the rewrite rules are
-updated.
-
-QuickSpec has also inspired another MTE tool for Haskell called
-Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
-also makes use of the laws of total orders and Boolean algebra to conjecture
-\emph{in}equalities and conditional relations between expressions.
-
-Another notable MTE implementation, distinct from those based in Isabelle and
-Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
-generate \emph{conjectures} that may later be sent to automated provers,
-MATHsAiD directly generates \emph{theorems}, by making logically valid
-inferences from a given set of axioms and definitions. Evaluation of the
-interestingness of these theorems was performed qualitatively by the system's
-developer, which highlights how these tools could benefit from the availability
-of an objective, repeatable, quantitative method of evaluation and comparison
-such as ours.
-
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
-\end{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-05-28 PhDThesis/support-for-benchmarks/mkBenchmarkGraphs.py

@@ -1,0 +1,598 @@
+#!/usr/bin/env python
+from json import loads, dumps
+import os
+import sys
+msg = lambda x: sys.stderr.write(x + '\n')
+
+msg('Reading input')
+
+from json import load
+
+def readInput(sys):
+    var = os.getenv(sys + 'Data')
+
+    assert var is not None, 'No {} env var given'.format(sys)
+
+    with open(var, 'r') as f:
+        return load(f)
+
+data = {
+    'isacosy'   : readInput('isacosy'),
+    'quickspec' : readInput('quickspec')
+}
+del(readInput)
+
+msg('Normalising data')
+
+# Extract timeout from one of the QuickSpec runs, since IsaCoSy may not have
+# stored it
+timeout = data['quickspec'].values()[0].values()[0]['timeout']
+
+def normalise(data):
+    def makeHashable(x):
+        """Convert lists and dictionaries into tuples, recursively, so we can
+        turn them into sets for comparison."""
+        xType    = type(x)
+        reducers = {
+            type([]) : lambda rest, elem: rest +  (makeHashable(elem),),
+            type({}) : lambda rest, key:  rest + ((makeHashable(key),
+                                                   makeHashable(x[key])),)
+        }
+        return reduce(reducers[xType], x, ()) if xType in reducers else x
+
+    for size in data:
+        for rep in data[size]:
+            # Samples should be sets
+            data[size][rep]['sample'] = frozenset(data[size][rep]['sample'])
+
+            # So should found equations
+            data[size][rep]['found'] = frozenset(
+                makeHashable(data[size][rep]['found']))
+    return data
+
+data = {
+    'isacosy'   : normalise(data['isacosy'  ]),
+    'quickspec' : normalise(data['quickspec'])
+}
+
+del(normalise)
+
+def quickspecCheck(data):
+    foundResults = {}
+
+    for size in data:
+        for rep in data[size]:
+            rdata = data[size][rep]
+            found = rdata['found']
+
+            if len(found) > 0:
+                for known in foundResults:
+                    if len(foundResults[known]) > 0:
+                        if rdata['sample'].issubset(known):
+                            assert found.issubset(foundResults[known]), repr({
+                                'error'  : 'More names should give more output',
+                                'got'    : list(found),
+                                'known'  : list(known),
+                                'prev'   : list(foundResults[known]),
+                                'sample' : list(sample),
+                                'size'   : size
+                            })
[…]

### 2018-05-30 TEBenchmarkPaper/article.tex
Fix some TEB issues, e.g. explaining the different averages

@@
 \item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth. This penalises overly-liberal tools which output a large
-  number of properties in the hope that some turn out to be ``good''.
+  the ground truth (the ratio of true positives to all positives). This
+  penalises overly-liberal tools which output a large number of properties in
+  the hope that some turn out to be ``good''.
 \item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output. This penalises overly-conservative tools which generate very
-  few properties as a way to avoid ``bad'' ones.
+  tool's output (the ratio of true positives to actual positives). This
+  penalises overly-conservative tools which generate very few properties as a
+  way to avoid ``bad'' ones.
 \end{itemize}
@@
 to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms, collecting together
-those of the same type into equivalence classes, assuming them to be
-equal. QuickCheck is then used to find counterexamples to this assumption, which
-it does by randomly instantiating the variables and comparing the resulting
-closed expressions. Any equivalence class whose elements don't compare equal are
-split up, and the process is repeated with new random values.
+QuickSpec takes a signature and enumerates well-typed terms. It collects
+together those of the same type into equivalence classes, assuming them to be
+equal, then uses QuickCheck to find counterexamples to this assumption by
+randomly instantiating the variables and comparing the resulting closed
+expressions. Any equivalence class whose elements don't compare equal are split
+up, and the process is repeated with new random values.
 
@@
   typed.
-\item The problem set should be relevant to the desired domain. Our focus on
-  functional programming requires higher-order functions and inductively defined
-  datatypes, which rules out first-order languages/logics (such as
+\item The problem set should be relevant to the desired domain. We focus on pure
+  functional programming, which requires higher-order functions and inductively
+  defined datatypes, which rules out first-order languages/logics (such as
   TPTP~\cite{sutcliffe2009tptp}).
@@
   than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem set are preferable as they give more robust statistics, all
+\item Larger problem sets are preferable as they give more robust statistics, all
   else being equal (i.e. when this does not sacrifice quality).
@@
 concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects like the names of functions. For example, consider a problem set which
-includes a statement of commutativity for a \texttt{plus} function, and of
-associativity for an \texttt{add} function, where the definitions of
-\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
-tool to either conjecture commutativity and associativity for \emph{both}
-functions, or for \emph{neither} function, since they are logically equivalent.
-Yet a na\"ive precision/recall analysis would treat commutativity of
+aspects of their presentation like the names of functions. For example, consider
+a problem set which includes a statement of commutativity for a \texttt{plus}
+function, and of associativity for an \texttt{add} function, where the
+definitions of \texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would
+expect an MTE tool to either conjecture commutativity and associativity for
+\emph{both} functions, or for \emph{neither} function, since they are logically
+equivalent. Yet a na\"ive precision/recall analysis would treat commutativity of
 \texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
@@
 definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order (based on cryptographic hashes, to discourage
-  ``cheating'').} Given a sample size, we choose a subset of that many
-definitions, and provide only those as input to the tool. We generate a
-corresponding ground truth by selecting those properties from the corpus which
-``depend on'' (contain references to) \emph{only} the definitions in that
-sample. Transitive dependencies aren't required (e.g. a property involving only
-a \texttt{times} function would not depend on a \texttt{plus} function, even if
-\texttt{plus} occurs in the definition of \texttt{times}).
+  use a deterministic order based on cryptographic hashes.} Given a sample size,
+we choose a subset of that many definitions, and provide only those as input to
+the tool. We generate a corresponding ground truth by selecting those properties
+from the corpus which ``depend on'' (contain references to) \emph{only} the
+definitions in that sample. Transitive dependencies aren't required (e.g. a
+property involving only a \texttt{times} function would not depend on a
+\texttt{plus} function, even if \texttt{plus} occurs in the definition of
+\texttt{times}).
 
@@
 
-Instead, we only allow a sample if it contains all dependencies of at least one
+Other measurements could be used for such cases, but our approach is to avoid
+them by only allowing a sample if it contains all dependencies of at least one
 property. We could do this using rejection sampling, but it is more efficient to
@@
 (the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output.
+on the tool's output. One downside of this restriction is that we limit the
+number of possible samples to measure; this is significant for sample sizes less
+than 3.
 
@@
 the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
-
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
-
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
+``typical'' input of that size. Since our methodology measures performance over
+many samples, we can also compute the \emph{spread} of our data, for example the
+inter-quartile range. This is not possible using the one-off measurements common
+in prior evaluation methods.
+
+There are two ways to average the measured precision and (separately) recall for
+samples of a particular size. Since these quantities are ratios, their mean
+values are known as the \emph{average of the ratios} (AOR), which we can define
+for $S$ samples indexed by $1 \leq i \leq S$ as follows:
+
+\begin{align*}
+       \overline{\text{precision}}_{AOR}
+    &= \frac{1}{S} \sum_i{\text{precision}_i} \\
+    &= \frac{1}{S} \sum_i{
+         \frac{\#\text{interesting}_i}
+              {\#\text{generated}_i}}         \\[10pt]
+       \overline{\text{recall}}_{AOR}
+    &= \frac{1}{S} \sum_i{\text{recall}_i} \\
+    &= \frac{1}{S} \sum_i{
+         \frac{\#\text{interesting}_i}
+              {\#\text{groundtruth}_i}}
+\end{align*}
+
+We interpret these as the \emph{expected} values of precision and recall for
+samples of this size, i.e. the expected quality of the \emph{set} of conjectures
+that will be generated from such a sample.
+
+The alternative is to calculate \emph{pooled} averages, by combining the
+(multi)sets from each sample before taking the ratios. If we multiply the
+numerator and denominator by a factor $\frac{1}{S}$ we find that this is
+equivalent to using the mean set sizes, hence this is known as the
+\emph{ratio of the averages} (ROA):
+
+\begin{align*}
+       \overline{\text{precision}}_{ROA}
+    &= \frac{\sum_i{\#\text{interesting}_i}}
+            {\sum_i{\#\text{generated}_i}}                              \\
+    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
+            {\frac{1}{S} \sum_i{\#\text{generated}_i}}                  \\
+    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
+            {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
+       \overline{\text{recall}}_{ROA}
+    &= \frac{\sum_i{\#\text{interesting}_i}}
+            {\sum_i{\#\text{groundtruth}_i}}                            \\
+    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
+            {\frac{1}{S} \sum_i{\#\text{groundtruth}_i}}                \\
+    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
+            {\hspace{5pt} \overline{\#\text{groundtruth}} \hspace{5pt}}
+\end{align*}
+
+Larger sets contribute more to a ratio of averages, unlike the mean which treats
+each set equally. This weighting gives us expected values of \emph{conjectures}
+rather than sets: the ROA for precision is the expected chance of a generated
+conjecture being interesting; the ROA for recall is the expected chance that a
+ground truth property will be generated. Note that both of these averaging
+methods deal with \emph{fractions}: we may also want to know the \emph{absolute}
+number of conjectures to expect, which is the mean output size
+$\overline{\#\text{generated}}$.
+
+If we instead want to combine precision values \emph{with} recall values (either
+individually or pooled) we can calculate their \emph{F-score}, defined as:
+
+\begin{equation*}
+  F = 2 \times \frac{\text{precision} \times \text{recall}}
+                    {\text{precision} +      \text{recall}}
+\end{equation*}
+
+This is the harmonic mean of precision and recall, ranging between 0 and 1. The
+F-score summarises performance into a single number, which is useful for
+purposes like ranking, but it obscures insights we might gain from looking at
+precision and recall separately (e.g. whether a tool is generating too much
+output or too little).
 
@@
 aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing average recall)
+specific runs, any comparison based on them (such as comparing mean recall)
 would have very low statistical power. More direct comparisons can be made using
@@
 
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
+For simplicity, we suggest comparing quality by pooling results from all samples
+where both tools finished successfully. Recall can be compared using the
+standard McNemar test for paired samples~\cite{mcnemar1947note}: each
 ground-truth property is categorised as either being found only by the first
@@
 We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
-problem has standalone type and function definitions, making their separation
-trivial; known examples from the software verification and inductive theorem
-proving literature are included, ensuring relevance to those fields; the format
-includes the higher-order functions and inductive datatypes we are interested
-in; it is large enough to pose a challenge to current MTE tools; plus it is
-accompanied by tooling to convert its custom format (an extension of
+since it satisfies the criteria specified in $\S$\ref{section:prep}: each
+benchmark problem has standalone type and function definitions, making their
+separation trivial; known examples from the software verification and inductive
+theorem proving literature are included, ensuring relevance to those fields; the
+format includes the higher-order functions and inductive datatypes we are
+interested in; it is large enough to pose a challenge to current MTE tools; plus
+it is accompanied by tooling to convert its custom format (an extension of
 SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
@@
 which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
+appear in the current sample are included in the signature and explored. We also
+encode all names in hexadecimal to avoid problems with language-specific naming
+rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
+distinguishes these from local variables and prevents names from beginning with
+a digit). This ensures that the generated conjectures will be using the same
+names as the ground truth, rather than some lossy variant.
 
@@
 
-We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
+We benchmarked QuickSpec version 0.9.6, a tool written in Haskell for
 conjecturing equations involving Haskell functions, described in more detail in
@@
 sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
+QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p-value
 of 0.0026.
@@
   \end{tabular}
-  \caption{Contingency table for ground truth properties, combined from those
+  \caption{Contingency table for ground truth properties, pooled from those
     samples where both QuickSpec and IsaCoSy finished successfully. The combined
@@
   \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, combined from those samples where both both finished successfully.
+    IsaCoSy, pooled from those samples where both both finished successfully.
     The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
@@
 
-One notable weakness is that our methodology only makes sense in domains where
-existing human knowledge surpasses that discovered by the machine. Any
-\emph{truly novel} insights discovered during testing will not, by definition,
-appear in any existing corpus, and we would in fact \emph{penalise} tools for
-such output. We do not believe this to be a realistic problem for the time
-being, although it does emphasise the continued importance of testing these
-tools with real human users, rather than relying solely on artificial benchmarks
-such as ours.
+One notable weakness is that our methodology does not allow negative examples,
+such as particularly dull properties that tools should never produce. Everything
+outside the ground truth set is treated equally, whether it's genuinely
+uninteresting or was only left out due to oversight. In particular this limits
+the use of our approach to domains where existing human knowledge surpasses that
+discovered by the machine. Any \emph{truly novel} insights discovered during
+testing will not, by definition, appear in any existing corpus, and we would in
+fact \emph{penalise} tools for such output. We do not believe this to be a
+realistic problem for the time being, as long as evaluation is limited to
+well-studied domains and results can be generalised to real areas of
+application. This does emphasise the continued importance of testing these tools
+with real human users, rather than relying solely on artificial benchmarks.
 

### 2018-05-30 TEBenchmarkPaper/article.tex
More detail on aggregation

@@
 
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Since our methodology measures performance over
-many samples, we can also compute the \emph{spread} of our data, for example the
-inter-quartile range. This is not possible using the one-off measurements common
-in prior evaluation methods.
-
-There are two ways to average the measured precision and (separately) recall for
-samples of a particular size. Since these quantities are ratios, their mean
-values are known as the \emph{average of the ratios} (AOR), which we can define
-for $S$ samples indexed by $1 \leq i \leq S$ as follows:
+We can summarise a set of runtimes by choosing the median, as this is more
+robust than the mean against long-running outliers, and hence represents
+performance for a ``typical'' input of that size. Since our methodology measures
+performance over many samples, we can also compute the \emph{spread} of our
+data, for example the inter-quartile range. This is not possible using the
+one-off measurements common in prior evaluation methods.
+
+Considering precision and recall separately, there are two ways to average these
+for a set of samples. Since both of these measurements are ratios, their means
+are known as the \emph{average of the ratios} (AOR). Using the definitions of
+precision and recall given in $\S$\ref{sec:te}, we can define the average of the
+ratios for $S$ samples indexed by $1 \leq i \leq S$ as follows:
 
@@
 
-We interpret these as the \emph{expected} values of precision and recall for
-samples of this size, i.e. the expected quality of the \emph{set} of conjectures
-that will be generated from such a sample.
+We interpret these mean values as the \emph{expected} precision and recall for
+samples of this size. In particular, these tell us the quality of the \emph{set}
+of conjectures that will be generated if we were to run the tool on such a
+sample, but they don't give us direct information about the individual
+conjectures themselves. Since these are mean values, they are the reference
+point for measures of spread such as the variance, standard deviation and mean
+absolute deviation.
 
 The alternative is to calculate \emph{pooled} averages, by combining the
-(multi)sets from each sample before taking the ratios. If we multiply the
-numerator and denominator by a factor $\frac{1}{S}$ we find that this is
-equivalent to using the mean set sizes, hence this is known as the
+(multi)sets from each sample before taking the ratios. The size of these pooled
+sets is simply the sum of the individual set sizes, and ratios of these sums are
+equal to ratios of the means (since the normalising factor $\frac{1}{S}$ appears
+in both numerator and denominator). Hence these values are known as the
 \emph{ratio of the averages} (ROA):
@@
 Larger sets contribute more to a ratio of averages, unlike the mean which treats
-each set equally. This weighting gives us expected values of \emph{conjectures}
-rather than sets: the ROA for precision is the expected chance of a generated
-conjecture being interesting; the ROA for recall is the expected chance that a
-ground truth property will be generated. Note that both of these averaging
-methods deal with \emph{fractions}: we may also want to know the \emph{absolute}
-number of conjectures to expect, which is the mean output size
-$\overline{\#\text{generated}}$.
-
-If we instead want to combine precision values \emph{with} recall values (either
-individually or pooled) we can calculate their \emph{F-score}, defined as:
+each set equally. This weighting gives us expected qualities of
+\emph{conjectures} rather than sets: the ROA for precision is the expected
+chance that some particular generated conjecture will appear in the ground
+truth; the ROA for recall is the expected chance that some particular ground
+truth property will appear in the generated output.
+
+These averages can differ when there is a large variation in set size. For
+example, if we generate a single, interesting conjecture for one sample and 99
+uninteresting conjectures for another sample, the AOR (mean) precision will be
+$\frac{1}{2}$ but the ROA precision will be $\frac{1}{100}$. This highlights the
+importance of knowing \emph{absolute} numbers of conjectures, such as the mean
+output size $\overline{\#\text{generated}}$ and its spread, alongside the
+proportions given by precision and recall.
+
+It is also possible to combine precision values \emph{with} recall values
+(either individually or aggregated) by calculating their \emph{F-score}, defined
+as:
 

### 2018-05-30 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Plot mean values of precision and recall, since that's what std. dev. uses

@@ -393,13 +393,8 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
 
     noneToZero = lambda x: 0.0 if x is None else float(x)
 
-    def ratioOfAverages(size):
-        '''Assume that the measure of interest (precision or recall) is fixed
-        for this size, and each particular conjecture is a bernoulli trial with
-        this measure as its p parameter.
-        Under this assumption, the fact that conjectures come from different
-        runs is irrelevant, so we pool them all together into one binomial
-        experiment.'''
+    def averageOfRatios(size):
+        '''Mean value across samples of the given size.'''
 
         # Skip runs which timed out, since we'd rather keep the measurements of
         # running time separate to "quality"
@@ -415,11 +410,15 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
         totals   = [agg[total    ][i] for i in indices]
         corrects = [agg['correct'][i] for i in indices]
 
+        ratios = [float(agg['correct'][i]) / float(agg[total][i]) \
+                  for i in indices]
+
         debug = lambda err: repr({
             'error'    : err,
             'key'      : key,
             'totals'   : totals,
             'corrects' : corrects,
+            'ratios'   : ratios,
             'indices'  : indices,
             'system'   : system
         })
@@ -433,15 +432,7 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
                 'size'       : size
             }
 
-        count   = float(sum(totals))
-        correct = float(sum(corrects))
-
-        p       = correct / count
-
-        # Analytic values, calculated from our model
-        anVar     = (p * (1.0 - p)) / count
-        anStddev  = math.sqrt(anVar)
-        anStderr  = anStddev / math.sqrt(count)
+        p = sum(ratios) / len(ratios)
 
         # Sample variability, calculated from our data.
         if len(totals) == 1:
@@ -459,20 +450,14 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
             sStddev = 1.0
         else:
             # Variance is the average squared difference between each run's
-            # actual correct proportion and the correct proportion we got from
-            # the model (p).
-            denom   = float(len(totals)) - 1.0
-            sVar    = (sum([((float(correct) / tot) - p)**2 \
-                            for tot, correct in zip(totals, corrects)])) / denom
+            # actual ratio and the mean ratio
+            sVar    = sum([(r - p)**2 for r in ratios]) / len(ratios)
[…]

### 2018-06-19 TEBenchmarkPaper/article.tex
Plot mean values of precision and recall, since that's what std. dev. uses
Fix TEB graphs

@@
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
+    is a successful run (spread out horizontally to reduce overlaps). Red
+    crosses are failed runs (timeouts or out-of-memory). Each size has 31 runs
+    total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
+    restrictions. Box plots show inter-quartile range, with lines at the median
+    and whiskers at 1.5$\times$~inter-quartile range. Medians below size 10 are
+    near zero, whilst medians of size 13 and 20 timed out.}
   \label{figure:quickspec_runtimes}
@@
   \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
+    horizontally to reduce overlapping). Lines show the mean proportion for each
+    sample size (average of the ratios) and errorbars show the sample standard
     deviation.}
@@
   \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
+    is a successful run (spread out horizontally to reduce overlaps). Red
+    crosses are failed runs (timeouts or out-of-memory). Each size has 31 runs
+    total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
+    restrictions. Box plots show inter-quartile range, with lines at the median
+    and whiskers at 1.5$\times$~inter-quartile range. All medians above size 4
+    timed out, and all runs above size 7 failed.}
   \label{figure:isacosy_runtimes}
@@
   \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
+    horizontally to reduce overlapping). Lines show the mean proportion for each
+    sample size (average of the ratios) and errorbars show the sample standard
     deviation.}

### 2018-06-19 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py

@@ -305,59 +305,12 @@ def savePlot(name):
     plt.tight_layout()
     plt.savefig(name + '.pgf', bbox_inches='tight', pad_inches=0.0)
 
-def plotTime(system, agg):
-    newFigure('time')
-
-    plt.ylim(0, 300)
-
-    sns.boxplot(data      = agg,
-                x         = 'size',
-                y         = 'time',
-                color     = '0.95',
-                fliersize = 0)
-
-    # Alternatively, we could use stripplot with jitter
-    successes = [i for i,s in enumerate(agg['success']) if     s]
-    failures  = [i for i,s in enumerate(agg['success']) if not s]
-    listAtIndices = lambda idc, lst: [x for i,x in enumerate(lst) if i in idc]
-    dictAtIndices = lambda idc: dict(agg,
-                                     size=listAtIndices(idc, agg['size']),
-                                     time=listAtIndices(idc, agg['time']))
-    ax = sns.swarmplot(data      = dictAtIndices(successes),
-                       x         = 'size',
-                       y         = 'time',
-                       size      = point['size'],
-                       edgecolor = 'k',
-                       linewidth = point['linewidth'],
-                       marker    = 'o',
-                       color     = 'k')
-    ax = sns.swarmplot(data      = dictAtIndices(failures),
-                       x         = 'size',
-                       y         = 'time',
-                       size      = cross['size'],
-                       edgecolor = 'k',
-                       linewidth = cross['linewidth'],
-                       marker    = 'x',
-                       color     = 'r')
-    if ax.legend_: ax.legend_.remove()
-    ax.set_xlabel('Sample size')
-    ax.set_ylabel('Runtime (seconds)')
-
-    savePlot(system + 'time')
-
 def aggProp(system, sizes=None, agg=None, key=None, total=None):
-    '''We assume the output conjectures from each run are bernoulli trials, i.e.
-    whether it's good or bad is modelled as a (biased) coin toss, where the bias
-    depends on the "quality" of the algorithm.'''
-
     import math
 
     sizeIndices   = lambda s: [i for i, x in enumerate(agg['size']) if x == s]
     entriesOfSize = lambda s, k: [agg[k][i] for i in sizeIndices(s)]
     runsOfSize    = lambda s: entriesOfSize(s, key)
-
-    # There are many analyses we could perform. We implement a few, so we can
-    # check that the results behave as we expect.
 
[…]

### 2018-07-04 TEBenchmarkPaper/article.tex
Fix acknowledgements
Use condensed style, as author guidelines ask
More detail about comparison methods

@@
 %\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
 %\documentclass[twocolumn]{svjour3}          % twocolumn
@@
 
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
-
-For simplicity, we suggest comparing quality by pooling results from all samples
-where both tools finished successfully. Recall can be compared using the
-standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
-
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
+Running times do not follow a normal distribution, in particular due to the
+lower bound at 0, which complicates any application of standard comparisons like
+the paired z-test. An alternative which does not assume normality is the
+Wilcoxon signed-rank test~\cite{wilcoxon1945individual}, which has been applied
+to software benchmarking in the Speedup-Test protocol of Touati, Worms and
+Briais~\cite{touati2013speedup}. Our methodology differs by measuring with a
+different sample each time, rather than repeatedly measuring one sample of each
+size; measuring both tools on the \emph{same} samples allows the paired form of
+the Wilcoxon test to be used.
+
+Another complication with our time measurements is the censoring effect caused
+by timeouts. Paired difference tests are robust to this, since the upper-bound
+on total time imposes an upper-bound to the observable difference; this biases
+our conclusions in a conservative way, \emph{towards} the null hypothesis of
+indistinguishable performance.
+
+Quality can be compared using precision and recall, but we must choose how to
+handle samples where one tool succeeded and the other failed (e.g. by timing
+out). Failed runs can be treated as if they had succeeded with an empty set of
+conjectures; this penalises failure-prone tools and simplifies analysis by
+ensuring every sample produces a data point. An alternative is to discard
+samples which caused either tool to fail, and compare only those where both
+tools finished successfully. We follow this second approach, since it is more
+charitable to the tools being evaluated (which, after all, were not designed
+with our benchmark in mind). With fewer data points to analyse, we pool together
+samples of all sizes (allowing duplicates) and compare proportions from these
+aggregations.
+
+Recall relies on a fixed set of properties, the ground truth, so we can compare
+tools using McNemar's test for paired samples~\cite{mcnemar1947note}. For each
+tool, we count how many of the ground-truth properties it found which were
+\emph{not} found by the other tool; i.e. those occasions where one tool was
+objectively better than the other. McNemar's test determines whether we can
+reject the null hypothesis that both counts are the same.
+
+Comparing precision cannot be done in the same way as recall, since we do not
+have a fixed set of properties to compare both tools on (there are an unlimited
+number of properties that may or may not appear in each tool's output). Instead,
+we can count how many of the generated conjectures were interesting and
+uninteresting for each tool, and use Boschloo's form of Barnard's
+test~\cite{lydersen2009recommended} to determine if these proportions are
+independent. This assumes that precision follows a binomial distribution, and
+the test is conditioned on the number of generated conjectures (as if this were
+fixed by design, rather than determined by the tool).
 
@@
 
-  We acknowledge the support of EPSRC grant EP/P017320/1 and award 1526504.
+  This work was supported by the Engineering and Physical Sciences Research
+  Council grant EP/P017320/1 and studentship 1526504.
 \end{acknowledgements}

### 2018-07-05 TEBenchmarkPaper/article.tex
TEB: Minor changes

@@
 number of possible samples to measure; this is significant for sample sizes less
-than 3.
+than~3.
 
@@
 that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations shouldn't be
+either there is enough memory, or there isn't; implementations should not be
 penalised for making use of available resources.
@@
 a digit). This ensures that the generated conjectures will be using the same
-names as the ground truth, rather than some lossy variant.
+names as the ground truth, rather than some language-specific variant.
 

### 2018-07-05 PhDThesis/benchmark.tex
Thesis: Bring in some changes from TEB

@@
 
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
+If we wish to make measurable progress on the task of conjecture generation, we
+must quantify what it is we are trying to achieve. Various approaches can then
+be measured and compared to optimise toward our goal. To this end we define a
+benchmarking methodology, shown in Figure~\ref{figure:flow_chart} on
+page~\pageref{figure:flow_chart}, which both generates a large
+definition/ground-truth corpus, and provides a scalable, statistical approach to
+evaluating MTE tools using this corpus. We follow a precision/recall approach
+similar to prior work, with the main difference being the source of definitions
+and ground truths: we take existing problem sets designed for automated theorem
+proving, and adapt their content for use in the conjecture generation setting.
 
@@
 
-We can summarise runtimes by choosing their median, as this is more robust than
-the mean against long-running outliers, and hence represents performance for a
-``typical'' input of that size. Unlike previous one-off evaluations, we can also
-compute the \emph{spread} of our data, for example with the inter-quartile
-range.
+We can summarise a set of runtimes by choosing the median, as this is more
+robust than the mean against long-running outliers, and hence represents
+performance for a ``typical'' input of that size. Since our methodology measures
+performance over many samples, we can also compute the \emph{spread} of our
+data, for example the inter-quartile range. This is not possible using the
+one-off measurements common in prior evaluation methods.
 
-The precision and recall for multiple samples can be combined in two ways. We
-can calculate their mean value, known as the \emph{average of the ratios}, to
-find the expected precision and recall for runs of this size. Alternatively, we
-can sum their numerators and denominators to get the \emph{ratio of the
-  averages}: this normalises the output and ground truth cardinalities, allowing
-precision and recall to be modelled as simple Bernoulli processes.
+Considering precision and recall separately, there are two ways to average these
+for a set of samples. Since both of these measurements are ratios, their means
+are known as the \emph{average of the ratios} (AOR). Using the definitions of
+precision and recall given in $\S$\ref{sec:te}, we can define the average of the
+ratios for $S$ samples indexed by $1 \leq i \leq S$ as follows:
 
-We prefer the latter, where each generated conjecture (wanted or not) is like a
-coin toss whose bias is the precision; likewise, the ground truth properties
-(found or not) are like coin tosses biased by the recall. Note that this simple
-model does not explain how many conjectures the tool decides to output for each
-run.
+\begin{align*}
+       \overline{\text{precision}}_{AOR}
+    &= \frac{1}{S} \sum_i{\text{precision}_i} \\
+    &= \frac{1}{S} \sum_i{
+         \frac{\#\text{interesting}_i}
+              {\#\text{generated}_i}}         \\[10pt]
+       \overline{\text{recall}}_{AOR}
+    &= \frac{1}{S} \sum_i{\text{recall}_i} \\
+    &= \frac{1}{S} \sum_i{
+         \frac{\#\text{interesting}_i}
+              {\#\text{groundtruth}_i}}
+\end{align*}
+
+We interpret these mean values as the \emph{expected} precision and recall for
+samples of this size. In particular, these tell us the quality of the \emph{set}
+of conjectures that will be generated if we were to run the tool on such a
+sample, but they don't give us direct information about the individual
+conjectures themselves. Since these are mean values, they are the reference
+point for measures of spread such as the variance, standard deviation and mean
+absolute deviation.
+
+The alternative is to calculate \emph{pooled} averages, by combining the
+(multi)sets from each sample before taking the ratios. The size of these pooled
+sets is simply the sum of the individual set sizes, and ratios of these sums are
+equal to ratios of the means (since the normalising factor $\frac{1}{S}$ appears
+in both numerator and denominator). Hence these values are known as the
+\emph{ratio of the averages} (ROA):
+
+\begin{align*}
+       \overline{\text{precision}}_{ROA}
+    &= \frac{\sum_i{\#\text{interesting}_i}}
+            {\sum_i{\#\text{generated}_i}}                              \\
+    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
+            {\frac{1}{S} \sum_i{\#\text{generated}_i}}                  \\
+    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
+            {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
+       \overline{\text{recall}}_{ROA}
+    &= \frac{\sum_i{\#\text{interesting}_i}}
+            {\sum_i{\#\text{groundtruth}_i}}                            \\
+    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
+            {\frac{1}{S} \sum_i{\#\text{groundtruth}_i}}                \\
+    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
+            {\hspace{5pt} \overline{\#\text{groundtruth}} \hspace{5pt}}
+\end{align*}
+
+Larger sets contribute more to a ratio of averages, unlike the mean which treats
+each set equally. This weighting gives us expected qualities of
+\emph{conjectures} rather than sets: the ROA for precision is the expected
+chance that some particular generated conjecture will appear in the ground
+truth; the ROA for recall is the expected chance that some particular ground
+truth property will appear in the generated output.
+
+These averages can differ when there is a large variation in set size. For
+example, if we generate a single, interesting conjecture for one sample and 99
+uninteresting conjectures for another sample, the AOR (mean) precision will be
+$\frac{1}{2}$ but the ROA precision will be $\frac{1}{100}$. This highlights the
+importance of knowing \emph{absolute} numbers of conjectures, such as the mean
+output size $\overline{\#\text{generated}}$ and its spread, alongside the
+proportions given by precision and recall.
+
+It is also possible to combine precision values \emph{with} recall values
+(either individually or aggregated) by calculating their \emph{F-score}, defined
+as:
+
+\begin{equation*}
+  F = 2 \times \frac{\text{precision} \times \text{recall}}
+                    {\text{precision} +      \text{recall}}
+\end{equation*}
+
+This is the harmonic mean of precision and recall, ranging between 0 and 1. The
+F-score summarises performance into a single number, which is useful for
+purposes like ranking, but it obscures insights we might gain from looking at
+precision and recall separately (e.g. whether a tool is generating too much
+output or too little).
 
@@
 
-We propose that running times are compared using a Wilcoxon signed-rank
-test~\cite{wilcoxon1945individual}, since it does not require that the data be
-normally distributed. In $\S$\ref{sec:application} we follow a variant of the
-Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
-differences are our use of the paired form of the Wilcoxon test, and the fact
-that each run was performed on a different sample (to reduce sampling bias, at
-the expense of accuracy). This paired difference test is also robust to the
-censoring caused by timeouts: upper-bounding the run time gives an upper-bound
-to the observable difference, which biases our results in a conservative way
-(\emph{towards} the null hypothesis of indistinguishable performance).
+Running times do not follow a normal distribution, in particular due to the
+lower bound at 0, which complicates any application of standard comparisons like
+the paired z-test. An alternative which does not assume normality is the
+Wilcoxon signed-rank test~\cite{wilcoxon1945individual}, which has been applied
+to software benchmarking in the Speedup-Test protocol of Touati, Worms and
+Briais~\cite{touati2013speedup}. Our methodology differs by measuring with a
+different sample each time, rather than repeatedly measuring one sample of each
+size; this also allows us to use the paired form of the Wilcoxon test.
 
-For simplicity, we suggest comparing quality by pooling together results from
-all samples where both tools finished successfully. Recall can be compared using
-the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
-ground-truth property is categorised as either being found only by the first
-tool, only by the second, by both or by neither. Those found by both or neither
-are ignored, and the tool-specific numbers are tested for marginal homogeneity
-(i.e. whether their differences give rise to similar totals).
+Another complication with our time measurements is the censoring effect caused
+by timeouts. Paired difference tests are robust to this, since the upper-bound
+on total time imposes an upper-bound to the observable difference; this biases
+our results in a conservative way, \emph{towards} the null hypothesis of
+indistinguishable performance.
 
-We cannot use McNemar's test to compare precision, since each tool may generate
-different conjectures. Instead, we can use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to compare the proportions of wanted and
-unwanted conjectures generated by each tool, assuming a binomial distribution
-and conditioning on the number of generated conjectures (as if this were fixed
-by design).
+Quality can be compared using precision and recall, but we must choose how to
+handle samples where one tool succeeded and the other failed (e.g. timing out).
+One approach would be to consider failed runs as if they generated an empty set
+of conjectures; this conflates quality with failure, which may be desirable, but
+we prefer the more charitable approach of comparing only those samples where
+both tools finished successfully. We pool the results from such samples together
+(allowing duplicates) and compare the resulting proportions.
+
+Recall relies on a fixed set of properties, the ground truth, so we can compare
+tools using McNemar's test for paired samples~\cite{mcnemar1947note}. For each
+tool, we count how many of the ground-truth properties it found which were
+\emph{not} found by the other tool. McNemar's test determines whether we can
+reject the null hypothesis that both tools have the same count (i.e. both tools
+have a similar contribution to the symmetric difference of their interesting
+outputs).
+
+Comparing precision cannot be done in the same way as recall, since we do not
+have a fixed set of properties to divide up (there are an unlimited number of
+properties that may or may not appear in a tool's output). Instead, we can count
+how many conjectures were interesting and uninteresting for each tool and use
+Boschloo's form of Barnard's test~\cite{lydersen2009recommended} to determine if
+these proportions are independent. This assumes a binomial distribution and is
+conditioned on the number of generated conjectures (as if this were fixed by
+design).
 
@@
 which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored.
+appear in the current sample are included in the signature and explored. We also
+encode all names in hexadecimal to avoid problems with language-specific naming
+rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
+distinguishes these from local variables and prevents names from beginning with
+a digit). This ensures that the generated conjectures will be using the same
+names as the ground truth, rather than some language-specific variant.
 
@@
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts or out-of-memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
+    is a successful run (spread out horizontally to reduce overlaps). Red
+    crosses show runs which timed out, which occurred for every size. Each size
+    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
+    sampling restrictions. Box plots show inter-quartile range, which reaches
+    the timeout for sizes above 8. Medians are marked with a line and remain
+    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
+    show 1.5$\times$~inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
   \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
-    deviation.}
+    horizontally to reduce overlapping). Lines show the mean proportion for each
+    sample size (average of the ratios) and errorbars show the sample standard
+    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
+    around $0.1$ for size 20, whilst mean recall remains roughly flat between
+    $0.2$ and $0.5$"}
   \label{figure:quickspec_precRec}
@@
   \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
-    processes (timeouts and out of memory) are shown in red. Points are spread
-    out horizontally to reduce overlapping. Box plots show median and quartiles,
-    with whiskers showing points at 1.5$\times$~inter-quartile range.}
+    is a successful run (spread out horizontally to reduce overlaps). Red
+    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
+    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
+    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
+    restrictions. Box plots show inter-quartile range, which grows with size,
+    and lines mark the median time, which increases rapidly from around 100
+    seconds at size 1 to timing out at size 4 and above. Whiskers show
+    1.5$\times$~inter-quartile range.}
   \label{figure:isacosy_runtimes}
@@
   \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the combined proportion for
-    each sample size (ratio of averages) and errorbars show the sample standard
+    horizontally to reduce overlapping; all runs above size 6 failed). Lines
+    show the mean proportion for each sample size (average of the ratios), which
+    trends downwards for precision and appears flat for recall, although there
+    are too few datapoints to be definitive. Errorbars show the sample standard
     deviation.}

### 2018-07-05 trigger-build.sh
Add trigger-build.sh

@@ -1,0 +1,11 @@
+#!/usr/bin/env bash
+# Trigger a build on Laminar continuous integration server. Useful as a git hook
+# for both post-commit (to check that local changes work) and post-receive (to
+# perform the "real" build).
+if command -v laminarc > /dev/null
+then
+  echo "Queueing laminar build" 1>&2
+  LAMINAR_REASON="Git commit" laminarc queue writing
+else
+  echo "Skipping laminar build, laminarc not found" 1>&2
+fi

### 2018-07-06 TEBenchmarkPaper/article.tex
TEB: Remove unused cruft
TEB: Mention Nix, give link to TE overview page and github

@@
 
-% For interestingness table
-\newcommand{\iE}{\textbf{E}}
-\newcommand{\iN}{\textbf{N}}
-\newcommand{\iS}{\textbf{S}}
-\newcommand{\iA}{\textbf{A}}
-\newcommand{\iC}{\textbf{C}}
-\newcommand{\iU}{\textbf{U}}
-\newcommand{\tIFF}{if-and-only-if}
-\newcommand{\tNE}{non-exists}
-\newcommand{\tIMP}{implies}
-\newcommand{\tEQ}{equations}
-\newcommand{\tINE}{inequalities}
-\newcommand{\tCON}{conditional}
-\newcommand{\tRow}[1]{#1 \\ \hline}
-
 % Insert the name of "your journal" with
@@
 \maketitle
-
-% Look carefully at each tool's (paper's) eval section
-
-% Make it clear that methodology is the point (it's not just a detail, like in a
-% chemistry paper for example)
 
@@
 
-% Lay out the field of TE: motivation, tools, evaluation, problems...
-
 \emph{Conjecture generation} is the open-ended task of conjecturing properties
@@
 \begin{sloppypar}
-  Our benchmark, including the scripts used to generate it from the TIP problem
-  set, is available online at
-  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
-  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
-  \url{https://github.com/warbo/isaplanner-tip}.
+  Our benchmark implementation is available at
+  \url{http://chriswarbo.net/projects/theory-exploration} and
+  \url{https://github.com/warbo}. All processes, including benchmark generation,
+  experimental runs and generation of this paper, are specified using the
+  Nix~\cite{dolstra2004nix} system to aid reproducibility.
 \end{sloppypar}
-
-%- Existing approaches are difficult to compare
-% - They want different things
-% - They're evaluated differently
-% - They're evaluated ``locally'' (as white boxen)
-
-% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
-% We can just jump straight in to automated conjecture generation
 
@@
 \label{sec:motivation}
-
-% more detail about evaluation, interestingness, etc. Reference Simon/Alan
-% interestingness, e.g. examples from textbooks, etc.
 

### 2018-07-06 TEBenchmarkPaper/article.tex
TEB: Phrasing

@@
 
-We introduce the conjecture generation problem in more detail, along with the
+We describe the conjecture generation problem in more detail, along with the
 difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
@@
 
-This procedure gives us a single runtime, precision and recall value for each
-sample. We propose two methods for analysing this data: \emph{summarising}
-the performance of a single MTE tool and \emph{comparing} the performance of two
-tools.
+This procedure gives us, for each sample, a set of generated conjectures and a
+set of ground truth properties (from which precision and recall can be
+calculated) and a single runtime. We propose two methods for analysing this
+data: \emph{summarising} the performance of a single MTE tool and
+\emph{comparing} the performance of two tools.
 
@@
     is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses are failed runs (timeouts or out-of-memory). Each size has 31 runs
-    total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
-    restrictions. Box plots show inter-quartile range, with lines at the median
-    and whiskers at 1.5$\times$~inter-quartile range. Medians below size 10 are
-    near zero, whilst medians of size 13 and 20 timed out.}
+    crosses show runs which timed out, which occurred for every size. Each size
+    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
+    sampling restrictions. Box plots show inter-quartile range, which reaches
+    the timeout for sizes above 8. Medians are marked with a line and remain
+    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
+    show 1.5$\times$~inter-quartile range.}
   \label{figure:quickspec_runtimes}
@@
   \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlapping). Lines show the mean proportion for each
-    sample size (average of the ratios) and errorbars show the sample standard
-    deviation.}
+    horizontally to reduce overlaps). Lines show the mean proportion for each
+    sample size (Average of Ratios) and errorbars show the sample standard
+    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
+    around $0.1$ for size 20, whilst mean recall remains roughly flat between
+    $0.2$ and $0.5$}
   \label{figure:quickspec_precRec}
@@
     is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses are failed runs (timeouts or out-of-memory). Each size has 31 runs
-    total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
-    restrictions. Box plots show inter-quartile range, with lines at the median
-    and whiskers at 1.5$\times$~inter-quartile range. All medians above size 4
-    timed out, and all runs above size 7 failed.}
+    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
+    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
+    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
+    restrictions. Box plots show inter-quartile range, which grows with size,
+    and lines mark the median time, which increases rapidly from around 100
+    seconds at size 1 to timing out at size 4 and above. Whiskers show
+    1.5$\times$~inter-quartile range.}
   \label{figure:isacosy_runtimes}
@@
   \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlapping). Lines show the mean proportion for each
-    sample size (average of the ratios) and errorbars show the sample standard
-    deviation.}
+    horizontally to reduce overlaps); there were no successes above size 6.
+    Lines show the mean proportion for each sample size (Average of Ratios),
+    which trends downwards for precision and appears flat for recall, although
+    there are too few datapoints to be definitive. Errorbars show the sample
+    standard deviation.}
   \label{figure:isacosy_precRec}

### 2018-07-06 TEBenchmarkPaper/article.tex
TEB: Typo

@@
 ``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingingness'', from a
-single quote from a mathematician). Opinions from mathematicians themselves
-include Gowers' suggestion that we can identify features which are commonly
-associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
+previous use of such metrics (in the case of ``surprisingness'', from a single
+quote from a mathematician). Opinions from mathematicians themselves include
+Gowers' suggestion that we can identify features which are commonly associated
+with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
 Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the

### 2018-07-06 TEBenchmarkPaper/article.tex
TEB: Better description of averages, including reference

@@
 
-Considering precision and recall separately, there are two ways to average these
-for a set of samples. Since both of these measurements are ratios, their means
-are known as the \emph{average of the ratios} (AOR). Using the definitions of
-precision and recall given in $\S$\ref{sec:te}, we can define the average of the
-ratios for $S$ samples indexed by $1 \leq i \leq S$ as follows:
+Precision and recall are ratios, which can be averaged (separately) in two ways,
+known as the \emph{Average of Ratios} (AoR) and the \emph{Ratio of Averages}
+(RoA)~\cite{egghe2012averages}. The Average of Ratios is the mean value. Using
+the definitions of precision and recall given in $\S$\ref{sec:te}, we can define
+their means for $S$ samples indexed by $1 \leq i \leq S$ as follows:
 
 \begin{align*}
-       \overline{\text{precision}}_{AOR}
+       \overline{\text{precision}}_{AoR}
     &= \frac{1}{S} \sum_i{\text{precision}_i} \\
@@
               {\#\text{generated}_i}}         \\[10pt]
-       \overline{\text{recall}}_{AOR}
+       \overline{\text{recall}}_{AoR}
     &= \frac{1}{S} \sum_i{\text{recall}_i} \\
@@
 
-The alternative is to calculate \emph{pooled} averages, by combining the
-(multi)sets from each sample before taking the ratios. The size of these pooled
-sets is simply the sum of the individual set sizes, and ratios of these sums are
-equal to ratios of the means (since the normalising factor $\frac{1}{S}$ appears
-in both numerator and denominator). Hence these values are known as the
-\emph{ratio of the averages} (ROA):
+The Ratio of Averages is \emph{pooled}, combining the (multi)sets from each
+sample before taking the ratios. The size of these pooled sets is simply the sum
+of the individual set sizes, and ratios of these sums are equal to ratios of the
+means (since the normalising factor $\frac{1}{S}$ appears in both numerator and
+denominator):
 
 \begin{align*}
-       \overline{\text{precision}}_{ROA}
+       \overline{\text{precision}}_{RoA}
     &= \frac{\sum_i{\#\text{interesting}_i}}
@@
             {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
-       \overline{\text{recall}}_{ROA}
+       \overline{\text{recall}}_{RoA}
     &= \frac{\sum_i{\#\text{interesting}_i}}
@@
 
-Larger sets contribute more to a ratio of averages, unlike the mean which treats
+Larger sets contribute more to a Ratio of Averages, unlike the mean which treats
 each set equally. This weighting gives us expected qualities of
-\emph{conjectures} rather than sets: the ROA for precision is the expected
+\emph{conjectures} rather than sets: the RoA for precision is the expected
 chance that some particular generated conjecture will appear in the ground
-truth; the ROA for recall is the expected chance that some particular ground
+truth; the RoA for recall is the expected chance that some particular ground
 truth property will appear in the generated output.
@@
 example, if we generate a single, interesting conjecture for one sample and 99
-uninteresting conjectures for another sample, the AOR (mean) precision will be
-$\frac{1}{2}$ but the ROA precision will be $\frac{1}{100}$. This highlights the
+uninteresting conjectures for another sample, the AoR (mean) precision will be
+$\frac{1}{2}$ but the RoA precision will be $\frac{1}{100}$. This highlights the
 importance of knowing \emph{absolute} numbers of conjectures, such as the mean

### 2018-07-11 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Render matplotlib to EPS rather than PGF

@@ -303,7 +303,7 @@ def newFigure(name):
 def savePlot(name):
     '''Write our the current figure as LaTeX.'''
     plt.tight_layout()
-    plt.savefig(name + '.pgf', bbox_inches='tight', pad_inches=0.0)
+    plt.savefig(name + '.eps', bbox_inches='tight', pad_inches=0.0)
 
 def aggProp(system, sizes=None, agg=None, key=None, total=None):
     import math

### 2018-07-11 TEBenchmarkPaper/supporting-materials/mkBenchmarkGraphs.py
Don't render chart text with TeX
It causes labels to be chopped off e.g. see
https://github.com/matplotlib/matplotlib/issues/7075

@@ -255,13 +255,13 @@ def figSize(widthFraction, height=None):
 # Taken from http://bkanuka.com/articles/native-latex-plots
 mpl.rcParams.update({
     'pgf.texsystem' : 'pdflatex',
-    'text.usetex'   : True,  # Use LaTeX to write all text
+    #'text.usetex'   : True,  # Use LaTeX to write all text
 
     # '[]' causes fonts to be inherited from document
     'font.family'     : 'serif',
-    'font.serif'      : [],
-    'font.sans-serif' : [],
-    'font.monospace'  : [],
+    #'font.serif'      : [],
+    #'font.sans-serif' : [],
+    #'font.monospace'  : [],
 
     # LaTeX default is 10pt font.
     'axes.labelsize'  : 10,

### 2018-07-11 TEBenchmarkPaper/article.tex
TEB: Close off some issues
Looks like we didn't need a bunch of stuff after all (maybe Colton table related?)
TEB: Minor changes (e.g. no full stop in captions)

@@
 % insert here the call for the packages your document requires
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{array}
 \usepackage{csvsimple}
 \usepackage[inline, shortlabels]{enumitem}
-\usepackage{float}
-\restylefloat{table}
-\usepackage{hhline}
 \usepackage{hyperref}
 \usepackage{mathtools}
-\usepackage{multirow}
 \usepackage{tikz}
@@
 
-\usepackage{attrib}
-\usepackage{csquotes}
-
-\usepackage{graphicx}
-\graphicspath{ {support/} }
-
 % please place your own definitions here and don't use \def but
@@
 \newcommand{\etal}{{\em et al.}}
-\newcommand{\Bagai}{Bagai \etal{}}
 
 % For example theory of lists
-\newcommand{\function}{\rightarrow}
-
 \newcommand{\Zero}{\text{Z}}
 \newcommand{\Succ}{\text{S}}
-\newcommand{\plus}{\text{plus}}
-\newcommand{\mult}{\text{times}}
 
@@
            University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk}           %  \\
+           \email{c.m.warburton@dundee.ac.uk} \\
+           ORCID: 0000-0002-4878-6319 \\
+           Phone: +44 (0)1382 386968
 %             \emph{Present address:} of F. Author  %  if needed
@@
            University of Dundee \\
-           \email{a.pease@dundee.ac.uk}
+           \email{a.pease@dundee.ac.uk} \\
+           ORCID: 0000-0003-1856-9599
            \and
@@
            University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk}
+           \email{j.n.zhang@dundee.ac.uk} \\
+           ORCID: 0000-0001-9317-0268
 }
@@
   between the disparate approaches currently being pursued for this task.
-%\keywords{First keyword \and Second keyword \and More}
+  \keywords{Theory Formation \and Theorem Proving \and Lemma Discovery \and Conjecture Generation \and Benchmarking}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 \end{abstract}
+
+\begin{acknowledgements}
+  We are grateful to the authors of the systems we have used in our experiments,
+  especially for help in obtaining and configuring their tools. We would
+  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
+  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
+  their software to our purposes.
+
+  This work was supported by the Engineering and Physical Sciences Research
+  Council grant EP/P017320/1 and studentship 1526504.
+\end{acknowledgements}
+
+\pagebreak
 
@@
     operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
+    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers}
   \label{figure:list_theory}
@@
 \begin{figure}
+  \centering
   \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
@@
 
-  \centering
   \begin{tikzpicture}[node distance=1cm]
@@
     near zero until size 10, with those of size 13 and 20 timing out. Whiskers
-    show 1.5$\times$~inter-quartile range.}
+    show 1.5$\times$~inter-quartile range}
   \label{figure:quickspec_runtimes}
@@
     seconds at size 1 to timing out at size 4 and above. Whiskers show
-    1.5$\times$~inter-quartile range.}
+    1.5$\times$~inter-quartile range}
   \label{figure:isacosy_runtimes}
@@
     there are too few datapoints to be definitive. Errorbars show the sample
-    standard deviation.}
+    standard deviation}
   \label{figure:isacosy_precRec}
@@
 
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  This work was supported by the Engineering and Physical Sciences Research
-  Council grant EP/P017320/1 and studentship 1526504.
-\end{acknowledgements}
-
 \bibliographystyle{plain}

### 2018-07-11 TEBenchmarkPaper/article.tex
Binaries, etc. to convert EPS and the like
Remove end of caption punctuation

@@
 \usepackage[inline, shortlabels]{enumitem}
+\usepackage{epstopdf}
 \usepackage{hyperref}
@@
   \centering
-  \input{quickspectime.pgf}
+  \includegraphics[scale=1]{Fig3}
   \caption{Running times of QuickSpec on theories sampled from TIP. Each point
@@
   \centering
-  \input{quickspecprec.pgf}
+  \includegraphics[scale=1]{Fig4}
   \caption{Precision and recall of successful QuickSpec runs (spread
@@
   \centering
-  \input{isacosytime.pgf}
+  \includegraphics[scale=1]{Fig5}
   \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
@@
   \centering
-  \input{isacosyprec.pgf}
+  \includegraphics[scale=1]{Fig6}
   \caption{Precision and recall of all successful IsaCoSy runs (spread
@@
     time limit acts to reduce the measured time difference (e.g. sizes 13 and
-    20, where the median for both tools timed out).}
+    20, where the median for both tools timed out)}
   \label{table:speedups}
@@
     samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
+    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$}
   \label{table:recall}
@@
     IsaCoSy, pooled from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
+    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$}
   \label{table:precision}

### 2018-08-20 PhDThesis/background_haskell.tex
Lit review and background

@@
 \label{sec:haskell}
+
+We mostly focus our attention on the Haskell programming language, both as an
+implementation vehicle and as our representation of functions, properties, etc.
+to explore. We choose Haskell since it combines formal, logical underpinnings
+which aid reasoning (compared to more popular languages like Java or C), yet it
+is still popular enough to sustain a rich ecosystem of tooling and a large
+corpus of existing code (compared to more formal languages like Coq or
+Isabelle).
+
+The following features make Haskell especially useful for our purposes, although
+many are also present in other languages such as StandardML and Coq (which we
+also use, but only when needed for compatibility with prior work):
+
+\begin{description}
+
+\item{Functional}: All control flow in Haskell is performed by function
+  abstraction and application, which we can reason about using standard rules of
+  inference such as \emph{modus ponens}.
+
+\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation
+  of expressions; hence our reasoning can safely ignore complicated external and
+  non-local interactions.
+
+\item{Statically Typed}: Expressions are constrained by \emph{types}, which can
+  be used to eliminate unwanted combinations of values, and hence reduce search
+  spaces; \emph{static} types can be deduced syntactically, without having to
+  execute the code.
+
+\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an
+  expression (i.e. performing function calls) without diverging, then a
+  non-strict evaluation strategy will not diverge when evaluating that
+  expression. This is rather technical, but in simple terms it allows us to
+  reason effectively about a Turing-complete language, where evaluation may not
+  terminate. For example, when reasoning about \emph{pairs} of values \hs{(x,
+    y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an
+  ``obvious'' rule such as
+  $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$. Haskell's
+  non-strict semantics makes this equation valid; whilst it would \emph{not} be
+  valid in the strict setting common to most other languages, where the
+  expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter
+  the semantics, if \hs{x} doesn't diverge).
+
+\item{Algebraic Data Types}: These provide a rich grammar for building up
+  user-defined data representations, and an inverse mechanism to inspect these
+  data by \emph{pattern-matching}. For our purposes, the useful consequences of
+  ADTs and pattern-matching include their amenability for inductive proofs and
+  the fact they are \emph{closed}; i.e. an ADT's declaration specifies all of
+  the normal forms for that type. This makes exhaustive case analysis trivial,
+  which would be impossible for \emph{open} types (for example, consider classes
+  in an object oriented language, where new subclasses can be introduced at any
+  time).
+
+\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over
+  \emph{type-level} objects; provided those objects are never inspected. This
+  has the \emph{practical} benefit of enabling \emph{polymorphism}: for example,
+  we can write a polymorphic identity function \hs{id :: forall t. t ->
+    t}. \footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a
+      -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually,
+  this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of
+  type \hs{t}; yet only the latter is available in the function body,
+  e.g. \hs{id x = x}. This inability to inspect type-level arguments gives us
+  the \emph{theoretical} benefit of being able to characterise the behaviour of
+  polymorphic functions from their type alone, a technique known as
+  \emph{theorems for free} \cite{wadler1989theorems}.
+
+\item{Type classes}: Along with their various extensions, type classes are
+  interfaces which specify a set of operations over a type (or other type-level
+  object, such as a \emph{type constructor}). Many type classes also specify a
+  set of \emph{laws} which their operations should obey but, lacking a simple
+  mechanism to enforce this, laws are usually considered as documentation. As a
+  simple example, we can define a type class \hs{Semigroup} with the following
+  operation and associativity law:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+op :: forall t. Semigroup t => t -> t -> t
+\end{lstlisting}
+
+$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
+
+The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which
+restricts the possible types \hs{t} to only those which implement
+\hs{Semigroup}. \footnote{Alternatively, we can consider \hs{Semigroup t} as the
+  type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case
+  \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take
+  \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t}
+  and two values of type \hs{t}. As with parameteric polymorphism, this extra
+  \hs{Semigroup t} parameter is not available at the value level. Even if it
+  were, we could not alter our behaviour by inspecting it, since Haskell only
+  allows types to implement each type class in at most one way, so there would
+  be no information to branch on.} There are many \emph{instances} of
+\hs{Semigroup} (types which may be substituted for \hs{t}), e.g. \hs{Integer}
+with \hs{op} performing addition. Many more examples can be found in the
+\emph{typeclassopedia} \cite{yorgey2009typeclassopedia}. This ability to
+constrain types, and the existence of laws, helps us reason about code
+generically, rather than repeating the same arguments for each particular pair
+of \hs{t} and \hs{op}.
+
+\item{Equational}: Haskell uses equations at the value level, for definitions;
+  at the type level, for coercions; at the documentation level, for typeclass
+  laws; and at the compiler level, for ad-hoc rewrite rules. This provides us
+  with many \emph{sources} of equations, as well as many possible \emph{uses}
+  for any equations we might discover. Along with their support in existing
+  tools such as SMT solvers, this makes equational conjectures a natural target
+  for theory exploration.
+
+\item{Modularity}: Haskell has a module system, where each module may specify an
+  \emph{export list} containing the names which should be made available for
+  other modules to import. When such a list is given, any expressions \emph{not}
+  on the list are considered \emph{private} to that module, and are hence
+  inaccessible from elsewhere. This mechanism allows modules to provide more
+  guarantees than are available just in their types. For example, a module may
+  represent email addresses in the following way:
+
+\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
+module Email (Email(), at, render) where
+
+data Email = E String String
+
+render :: Email -> String
+render (E u h) = u ++ "@" ++ h
+
+at :: String -> String -> Maybe Email
+at "" h  = Nothing
+at u  "" = Nothing
+at u  h  = Just (E u h)
+\end{lstlisting}
+
+The \hs{Email} type guarantees that its elements have both a user part and a
+host part (modulo divergence), but it does not provide any guarantees about
+those parts. We also define the \hs{at} function, a so-called ``smart
+constructor'', which has the additional guarantee that the \hs{Email}s it
+returns contain non-empty \hs{String}s. By ommitting the \hs{E} constructor from
+the export list on the first line \footnote{The syntax \hs{Email()} means we're
+  exporting the \hs{Email} type, but not any of its constructors.}, the only way
+\emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces
+the non-empty guarantee to hold globally.
+
+\end{description}
+
+Together, these features make Haskell code highly structured, amenable to
+logical analysis and subject to many algebraic laws. However, as mentioned with
+regards to type classes, Haskell itself is incapable of expressing or enforcing
+these laws (at least, without difficulty \cite{lindley2014hasochism}). This
+reduces the incentive to manually discover, state and prove theorems about
+Haskell code, e.g. in the style of interactive theorem proving, as these results
+may be invalidated by seemingly innocuous code changes. This puts Haskell in a
+rather special position with regards to the discovery of interesting theorems;
+namely that many discoveries may be available with very little work, simply
+because the code's authors are focused on \emph{software} development rather
+than \emph{proof} development. The same cannot be said, for example, of ITP
+systems; although our reasoning capabilities may be stronger in an ITP setting,
+much of the ``low hanging fruit'' will have already been found through the
+user's dedicated efforts, and hence theory exploration would be less likely to
+discover unexpected properties.
+
+Other empirical advantages to studying Haskell, compared to other programming
+languages or theorem proving systems, include:
+
+\begin{itemize}
+\item The large amount of Haskell code which is freely available online, e.g. in
+  repositories like \href{http://hackage.haskell.org}{Hackage}, with which we
+  can experiment.
+
+\item The existence of theory exploration systems such as \hspec{}, and related
+  tools which we can to re-use, including conjecture generators like \qspec{}
+  and \speculate{}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
+  \textsc{SmartCheck}, \textsc{LeanCheck}, \textsc{Hedgehog}, etc.; theorem
+  provers like \textsc{Hip} \cite{rosen2012proving}; and other related testing
+  and term-generating systems like \textsc{MuCheck} \cite{le2014mucheck},
+  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
+  \textsc{Djinn}~\cite{augustsson2005djinn}.
+
+\item The remarkable amount of infrastructure which exists for working with
+  Haskell code, including package managers, compilers, interpreters, parsers,
+  static analysers, etc.
+\end{itemize}
+
+%% FIXME: MERGE THE FOLLOWING WITH THE ABOVE
 
@@
 \mbox{\hs{f a b ... = ...}}).
-
-%% FIXME: MERGE THE FOLLOWING WITH THE ABOVE
-
-Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little
-explanation was given for why we chose this language rather than, for example
-Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more
-widely used language like Java. Here we discuss the relevant language features
-from a high-level, which motivated our choice:
-
-\begin{description}
-
-\item{Functional}: All control flow in Haskell is performed by function
-  abstraction and application, which we can reason about using standard rules of
-  inference such as \emph{modus ponens}.
-
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation
-  of expressions; hence our reasoning can safely ignore complicated external and
-  non-local interactions.
-
-\item{Statically Typed}: Expression are constrained by \emph{types}, which can
-  be used to eliminate unwanted combinations of values, and hence reduce search
-  spaces; \emph{static} types can be deduced syntactically, without having to
-  execute the code.
-
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an
-  expression (i.e. performing function calls) without diverging, then a
-  non-strict evaluation strategy will not diverge when evaluating that
-  expression. This is rather technical, but in simple terms it allows us to
-  reason effectively about a Turing-complete language, where evaluation may not
-  terminate. For example, when reasoning about \emph{pairs} of values \hs{(x,
-    y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an
-  ``obvious'' rule such as
-  $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$. Haskell's
-  non-strict semantics makes this equation valid; whilst it would \emph{not} be
-  valid in the strict setting common to most other languages, where the
-  expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter
-  the semantics, if \hs{x} doesn't diverge).
-
-\item{Algebraic Data Types}: These provide a rich grammar for building up
-  user-defined data representations, and an inverse mechanism to inspect these
-  data by \emph{pattern-matching}. For our purposes, the useful consequences of
-  ADTs and pattern-matching include their amenability for inductive proofs and
-  the fact they are \emph{closed}; i.e. an ADT's declaration specifies all of
-  the normal forms for that type. This makes exhaustive case analysis trivial,
-  which would be impossible for \emph{open} types (for example, consider classes
-  in an object oriented language, where new subclasses can be introduced at any
-  time).
-
-\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over
-  \emph{type-level} objects; provided those objects are never inspected. This
-  has the \emph{practical} benefit of enabling \emph{polymorphism}: for example,
-  we can write a polymorphic identity function \hs{id :: forall t. t ->
-    t}. \footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a
-      -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually,
-  this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of
-  type \hs{t}; yet only the latter is available in the function body,
-  e.g. \hs{id x = x}. This inability to inspect type-level arguments gives us
-  the \emph{theoretical} benefit of being able to characterise the behaviour of
-  polymorphic functions from their type alone, a technique known as
-  \emph{theorems for free} \cite{wadler1989theorems}.
-
-\item{Type classes}: Along with their various extensions, type classes are
-  interfaces which specify a set of operations over a type (or other type-level
-  object, such as a \emph{type constructor}). Many type classes also specify a
-  set of \emph{laws} which their operations should obey but, lacking a simple
-  mechanism to enforce this, laws are usually considered as documentation. As a
-  simple example, we can define a type class \hs{Semigroup} with the following
-  operation and associativity law:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
-op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
-
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
-
-The notation \hs{Semigroup t =>} is a \emph{type class constraint}, which
-restricts the possible types \hs{t} to only those which implement
-\hs{Semigroup}. \footnote{Alternatively, we can consider \hs{Semigroup t} as the
-  type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case
-  \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take
-  \emph{four} parameters: a type \hs{t}, an implementation of \hs{Semigroup t}
-  and two values of type \hs{t}. As with parameteric polymorphism, this extra
-  \hs{Semigroup t} parameter is not available at the value level. Even if it
-  were, we could not alter our behaviour by inspecting it, since Haskell only
-  allows types to implement each type class in at most one way, so there would
-  be no information to branch on.} There are many \emph{instances} of
-\hs{Semigroup} (types which may be substituted for \hs{t}), e.g. \hs{Integer}
-with \hs{op} performing addition. Many more examples can be found in the
-\emph{typeclassopedia} \cite{yorgey2009typeclassopedia}. This ability to
-constrain types, and the existence of laws, helps us reason about code
-generically, rather than repeating the same arguments for each particular pair
-of \hs{t} and \hs{op}.
-
-\item{Equational}: Haskell uses equations at the value level, for definitions;
-  at the type level, for coercions; at the documentation level, for typeclass
-  laws; and at the compiler level, for ad-hoc rewrite rules. This provides us
-  with many \emph{sources} of equations, as well as many possible \emph{uses}
-  for any equations we might discover. Along with their support in existing
-  tools such as SMT solvers, this makes equational conjectures a natural target
-  for theory exploration.
-
-\item{Modularity}: Haskell has a module system, where each module may specify an
-  \emph{export list} containing the names which should be made available for
-  other modules to import. When such a list is given, any expressions \emph{not}
-  on the list are considered \emph{private} to that module, and are hence
-  inaccessible from elsewhere. This mechanism allows modules to provide more
-  guarantees than are available just in their types. For example, a module may
-  represent email addresses in the following way:
-
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
-module Email (Email(), at, render) where
-
-data Email = E String String
-
-render :: Email -> String
-render (E u h) = u ++ "@" ++ h
-
-at :: String -> String -> Maybe Email
-at "" h  = Nothing
-at u  "" = Nothing
-at u  h  = Just (E u h)
-\end{lstlisting}
-
-The \hs{Email} type guarantees that its elements have both a user part and a
-host part (modulo divergence), but it does not provide any guarantees about
-those parts. We also define the \hs{at} function, a so-called ``smart
-constructor'', which has the additional guarantee that the \hs{Email}s it
-returns contain non-empty \hs{String}s. By ommitting the \hs{E} constructor from
-the export list on the first line \footnote{The syntax \hs{Email()} means we're
-  exporting the \hs{Email} type, but not any of its constructors.}, the only way
-\emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces
-the non-empty guarantee to hold globally.
-
-\end{description}
-
-Together, these features make Haskell code highly structured, amenable to
-logical analysis and subject to many algebraic laws. However, as mentioned with
-regards to type classes, Haskell itself is incapable of expressing or enforcing
-these laws (at least, without difficulty \cite{lindley2014hasochism}). This
-reduces the incentive to manually discover, state and prove theorems about
-Haskell code, e.g. in the style of interactive theorem proving, as these results
-may be invalidated by seemingly innocuous code changes. This puts Haskell in a
-rather special position with regards to the discovery of interesting theorems;
-namely that many discoveries may be available with very little work, simply
-because the code's authors are focused on \emph{software} development rather
-than \emph{proof} development. The same cannot be said, for example, of ITP
-systems; although our reasoning capabilities may be stronger in an ITP setting,
-much of the ``low hanging fruit'' will have already been found through the
-user's dedicated efforts, and hence theory exploration would be unlikely to
-discover unexpected properties.
-
-Other empirical advantages to studying Haskell, compared to other programming
-languages or theorem proving systems, include:
-
-\begin{itemize}
-\item The large amount of Haskell code which is freely available online, e.g. in
-  repositories like \href{http://hackage.haskell.org}{Hackage}, with which we
-  can experiment.
-
-\item The existence of theory exploration systems such as \hspec{}, and related
-  tools which we may be able to re-use, including conjecture generators like
-  \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and
-  \textsc{SmartCheck}; theorem provers like \textsc{Hip}
-  \cite{rosen2012proving}; and other related testing and term-generating
-  systems like \textsc{MuCheck} \cite{le2014mucheck}, \textsc{MagicHaskeller}
-  \cite{katayama2011magichaskeller} and \textsc{Djinn}
-  \cite{augustsson2005djinn}.
-
-\item The remarkable amount of infrastructure which exists for working with
-  Haskell code, including package managers, compilers, interpreters, parsers,
-  static analysers, etc.
-\end{itemize}

### 2018-08-20 PhDThesis/litreview.tex
Lit review and background

@@
 
-\input{litreview_mathhistory.tex}
+The automation of mathematical tasks and reasoning has been pursued since at
+least the time of mechanical calculators like the Pascaline~\cite{d'ocagne}. A
+recurring theme in these efforts is the separation between those undertaken by
+mathematicians like Pascal and Babbage~\cite{bowden}, and those of engineers
+such as M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the
+tasks we are concerned with (automatically constructing and evaluating concepts,
+conjectures, theorems, axioms, examples, etc.) being divided into two main
+fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
+sometimes prefaced with ``Computer-Aided'', ``Automated'' or
+``Algorithm-Supported''), which is championed by mathematicians such as
+Buchberger~\cite{buchberger}; and Automated Theory Formation
+(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
+Other related terms include ``Automated Mathematical
+Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
+``Concept Formation in Discovery Systems''~\cite{haase}, and
+``Automated Theorem Discovery''~\cite{roy}.
+
+Such a plethora of terminology can mask similarities and shared goals between
+these fields. Even notable historical differences, such as the emphasis of MTE
+on user-interaction and mathematical domains, in contrast to the full automation
+and more general applications targeted by ATF, are disappearing in recent
+implementations.
+
+\subsection{Theory Formation}
+
+An important historical implementation of ATF is Lenat's AM (Automated
+Mathematician) system. Unlike prior work, such as
+Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
+to be a general purpose mathematical discovery system, designed to both
+construct new concepts and conjecture relationships between them. AM is a
+rule-based system which represents knowledge using a frame-like scheme, enlarges
+its knowledge base via a collection of heuristic rules, and controls the firing
+of these rules via an agenda mechanism.
+
+The performance of AM was evaluated based on its generality (performance in new
+domains) and how finely-tuned various aspects of the program are (the agenda,
+the interaction of the heuristics, etc). Most of this evaluation was
+qualitative, and has subsequently been criticised~\cite[chap.~13]{colton:book}.
+In their case study in methodology, Ritchie and Hanna found a large discrepancy
+between the theoretical claims made of AM and the implemented
+program~\cite{ritchie1984case}; for example, AM ``invented'' natural numbers
+from sets, but did so using a heuristic specifically designed to make this
+connection.
+
+The successor of AM is Eurisko, a discovery system intended to be useful in more
+general domains than just mathematical theory formation. Despite claiming some
+early successes % FIXME Traveller TCS
+work on Eurisko was put on hold due to a lack of machine-readable data about
+real world systems for it to work with % FIXME Find something quotable
+. This lead to the Cyc project, in an attempt to encode such knowledge in a
+semantically rich and consistent form. Cyc is an ongoing project, with parts of
+its database made available via OpenCyc.
+
+\subsection{Theory Exploration}
+
+The prototypical implementation of MTE is the Theorema system of Buchberger and
+colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
+emphasis on user interface and output presentation. Theory exploration in the
+Theorema system involves the user formalising their definitions in a consistent,
+layered approach; such that reasoning algorithms can exploit this structure in
+subsequent proofs, calculations, etc. The potential of this strategy was
+evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
+bases algorithm~\cite{buchberger:04}.
+
+A similar ``layering'' approach is found in the IsaScheme system of
+Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
+has also been quantitatively compared against IsaCoSy and HipSpec using
+precision/recall analysis~\cite{claessen2013automating}. The name comes from its
+embedding in the Isabelle proof assistant and its use of ``schemes'':
+higher-order formulae which can be used to generate new concepts and
+conjectures. Variables within a scheme are instantiated automatically and this
+drives the invention process. For example, the concept of ``repetition'' can be
+encoded as a scheme, and instantiated with existing encodings of zero, successor
+and addition to produce a definition of multiplication. The same scheme can be
+instantiated with this new multiplication function to produce exponentiation.
+
+IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
+described in more detail in $\S$\ref{sec:existing-tools}, since these are the
+tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
+has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
+distinct enumeration and testing steps with a single, iterative algorithm
+similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
+completion algorithm to form a corresponding set of rewrite rules. As
+expressions are enumerated, they are simplified using these rules and discarded
+if equal to a known expression. If not, QuickCheck tests whether the new
+expression can be distinguished from the known expressions through random
+testing: those which can are added to the set of known expressions. Those which
+cannot be distinguished are conjectured to be equal, and the rewrite rules are
+updated.
+
+QuickSpec has also inspired another MTE tool for Haskell called
+Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
+also makes use of the laws of total orders and Boolean algebra to conjecture
+\emph{in}equalities and conditional relations between expressions.
+
+Another notable MTE implementation, distinct from those based in Isabelle and
+Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
+Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
+generate \emph{conjectures} that may later be sent to automated provers,
+MATHsAiD directly generates \emph{theorems}, by making logically valid
+inferences from a given set of axioms and definitions. Evaluation of the
+interestingness of these theorems was performed qualitatively by the system's
+developer, which highlights how these tools could benefit from the availability
+of an objective, repeatable, quantitative method of evaluation and comparison
+such as ours.
+
+\input{litreview_staticanalysis.tex}
+\input{litreview_premiseselection.tex}
 \input{litreview_interestingness.tex}
-\input{litreview_theoryformation.tex}
-\input{litreview_theoryexploration.tex}
-\input{litreview_premiseselection.tex}

### 2018-08-20 PhDThesis/litreview_premiseselection.tex
Lit review and background

@@
-\subsubsection{Relevance Filtering}
+\subsection{Relevance Filtering}
 \label{sec:relevance}
@@
 
-The combinatorial nature of formal systems causes many proof search methods,
-such as resolution, to have exponential complexity
-\cite{haken1985intractability}; hence even a modest size increase can turn a
-trivial problem into an intractable one. Finding efficient alternatives for such
-algorithms, especially those which are NP-complete (e.g. determining
-satisfiability) or co-NP-complete (e.g. determining tautologies), seems
-unlikely, as it would imply progress on the famously intractable open problems
-of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$. On the other hand, we
-can turn this difficulty around: a modest \emph{decrease} in size may turn an
-intractable problem into a solvable one. We can ensure that the solutions to
-these reduced problems coincide with the original if we only remove
-\emph{redundant} information. This leads to the idea of \emph{relevance
-  filtering}.
+One fruitful application of machine learning methods to formal mathematics is
+\emph{relevance filtering}, which is used for \emph{premise selection} in
+automated theorem provers.
 
@@
 
-The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson}
-\cite{meng2009lightweight}), gives each clause a score based on the proportion
-$m / n$ of its symbols which are ``relevant'' (where $n$ is the number of
-symbols in the clause and $m$ is the number which are relevant). Initially, the
-relevant symbols are those which occur in the goal, but whenever a clause is
-found which scores more than a particular threshold, all of its symbols are then
-also considered relevant. There are other heuristics applied too, such as
-increasing the score of user-provided facts (e.g. given by keywords like
-\texttt{using}), locally-scoped facts, first-order facts and rarely-occuring
-facts. To choose $r$ relevant clauses for an ATP invocation, we simply order the
-clauses by decreasing score and take the first $r$ of them.
+Sledgehammer's initial approach, known as \textsc{MePO} (from
+\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based
+on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is
+the number of symbols in the clause and $m$ is the number which are relevant).
+Initially, the relevant symbols are those which occur in the goal, but whenever
+a clause is found which scores more than a particular threshold, all of its
+symbols are then also considered relevant. There are other heuristics applied
+too, such as increasing the score of user-provided facts (e.g. given by keywords
+like \texttt{using}), locally-scoped facts, first-order facts and
+rarely-occuring facts. To choose $r$ relevant clauses for an ATP invocation, we
+simply order the clauses by decreasing score and take the first $r$ of them.
 
 Recently, a variety of alternative algorithms have also been investigated,
-including:
+dubbed \textsc{MaSH}, \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
+\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
+\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
+algorithm with a custom feature extraction approach, is trained using a corpus
+of existing theorems and their proofs, and evaluated by trying to re-prove a
+selection of unseen theorems from the corpus (i.e. given only the theorem
+statement and possible lemmas, not the existing proof).
 
-\begin{description}
-\item{\textsc{MaSH}}: Machine Learning for SledgeHammer
-  \cite{kuhlwein2013mash}. The distinguishing feature of \textsc{MaSH} is its
-  use of ``visibility'', which is essentially a dependency graph of which
-  theorems were used in the proofs of which other theorems; although theorems
-  are represented as abstract sets of features. To select relevant clauses for a
-  goal, the set of clauses which are visible from the goal's components is
-  generated; this is further reduced by (an efficient approximation of) a naive
-  Bayes algorithm.
+For example the Machine Learning for SledgeHammer (\textsc{MaSH}) system
+defines a notion of ``visibility''~\cite{kuhlwein2013mash}. This is essentially
+a dependency graph of which theorems were used in the proofs of which other
+theorems (theorems are represented as abstract sets of features, rather than
+e.g. sentences or syntax trees). To select relevant clauses for a goal, the
+set of clauses which are ``visible'' from the goal's components is generated;
+this is further reduced by (an efficient approximation of) a naive Bayes
+algorithm.
 
-\item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine
-  (SVM) approach for selecting relevant axioms from the Mizar Mathematical
-  Library for use by the Vampire ATP system \cite{alama2014premise}. \iffalse
-  TODO: describe the kernel, as that's the interesting bit \fi It compares
-  favourably to \textsc{SNoW} and \textsc{SInE}.
+The Multi-output ranking (\textsc{MOR}) implementation uses a support vector
+machine (SVM) to select relevant axioms from the Mizar Mathematical Library for
+use by the Vampire ATP system~\cite{alama2014premise}.
+%TODO: describe the kernel, as that's the interesting bit.
 
-  % TODO:
-\item{\textsc{SInE}}
-\item{\textsc{BliStr}}
-\item{\textsc{HOLyHammer}}
-\item{\textsc{MoMM}}
-\item{\textsc{SNoW}}
-\item{\textsc{MPTP 0.2}}
-\item{\textsc{MaLARea}}
-\item{\textsc{MaLARea SG1}}
-
-\end{description}
+The premise selection problem for theorem proving is related to the signature
+selection problem for theory exploration, in that both are trying to pick a
+small number of terms (depending on the speed of the subsequent algorithm and
+how long the user is willing to wait) from a large set, to optimise the
+outcome of that algorithm (e.g. finding a proof quickly; or finding many
+interesting relationships quickly). The major difference is that premise
+selection has a particular ``target'' theorem statement, against which to
+compare each term (e.g. for the ``relevance'' of their sub-terms). This is not
+the case for theory exploration, so we must consider the interactions of each
+term with all of the others when deciding what to select.

### 2018-08-20 PhDThesis/litreview_staticanalysis.tex
Lit review and background

\subsection{Automated Software Improvement}

Software for analysing and improving code can be roughly divided into two
categories. Those which contribute to the running of a program, such as
(optimising) compilers, linkers and interpreters must produce correct,
predictable, desirable results in a timely manner; for example, compiler
optimisations should not alter program semantics, should provide improvements in
the majority of cases (i.e. they should not be \emph{pessimisations}) and under
which they apply should be predictable enough to prevent spurious performance
regressions when making seemingly innocuous refactorings. Optimisations in
ahead-of-time (AOT) compilers must be fast enough to run during every build, and
should have a large enough impact on the generated code's performance that they
``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
ideals become hard requirements, since any runtime transformation which doesn't
pay for itself is, by definition, a pessimisation.

The other category is that of ``standalone'' tools, which are run separately to
building or executing the target code, and whose results do not contribute
directly to the code's output. These include tools for testing, profiling,
verification, standards compliance, etc. Our application of theory exploration
to software libraries fits into this latter category, whose constraints are not
as restricting as those applying to tools like compilers. The only real
restrictions on our tools are that their resource requirements are acceptable to
users, and that output is sufficiently high quality to be worth waiting for and
sifting through.

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used or
``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

A popular linter for Haskell code is \texttt{HLint}~\cite{mitchell2014hlint},
which focuses mostly on simplifying code; for example by spotting use-cases for
common library functions, e.g. replacing a pattern match \texttt{case A of \{
  Nothing -> B; Just C -> D C; \}} with the function call \texttt{maybe B D A}.
Linting can be seen as the inverse of theory exploration: rather than inferring
general relationships from repeated patterns in the code, linting begins with
known relationships and looks through the code for specific instances. Whilst
linting is a useful programming aid, and fast enough to apply regularly (e.g. as
a background process in a code editor), it is not an open-ended process: once
the predetermined patterns are exhausted, no more improvements will be found.

If fast but limited linters live at one end of a code improvement spectrum, at
the other end we find \emph{superoptimisers}~\cite{massalin1987superoptimizer}.
These treat the target program as a specification, and perform an open-ended
search through the space of all programs (in some specific language, usually a
subset of machine code which is free of branches and side-effects) to find all
possible implementations. Superoptimisers need some way to check program
equivalence, such as a test suite (to quickly eliminate incorrect programs) and
a theorem prover. Proving equivalence of machine code programs is helped by the
uniform, finite nature of their domain (e.g. a fixed set of 32bit registers).
Programs in higher level languages, like Haskell, may have unlimited input
domains (say, the natural numbers), which makes effective testing more difficult
and proofs potentially undecidable. A decidable method of comparison is also
required, to choose which of two equivalent programs is ``better''. This is
typically a cost model for the language, with a trivial example being program
length (i.e. assuming all instructions have the same cost).

The biggest hindrance to superoptimisation is how long it takes to search such a
large space. Whilst pruning methods can help~\cite{phothilimthana2016scaling},
the power of superoptimisation comes from its ``bottom up'' nature, which makes
no assumptions about the structure of the generated program. Maintaining this
generality is expensive, and current superoptimisers typically scale to programs
only a few dozen machine instructions long. Nevertheless, this still finds use
in compiler construction, as it is an effective method for finding ``peephole
optimisations'' (faster replacements for small, common patterns of generated
machine code)~\cite{Bansal.Aiken:2006}.

### 2018-08-20 PhDThesis/evaluation_existingcode.tex
Some existing code changes

@@
-\section{Evaluation}
-\label{sec:evaluation}
-
-We have applied our recurrent clustering algorithm to several scenarios, with
-mixed results. A major difficulty in evaluating these clusters is that we have
-no ``ground truth'', i.e. there is no objectively correct way to compare
-expressions. Instead, we provide a qualitative overview of the more interesting
-characteristics.
+To see how our recurrent clustering implementation performs outside our own
+benchmark we have also applied it to existing code, with mixed results. Unlike
+our benchmark, these scenarios have no ground truth or measure of success to
+compare against (there is no objectively correct way to compare expressions).
+Instead, we provide a qualitative overview of the more interesting
+characteristics noticed in these explorations.
 
@@
 
-We have also applied our recurrent clustering algorithm to a variety of the
-most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
+For a more realistic test we also applied recurrent clustering to a variety of
+the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
 \texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
@@
 
-Whilst recurrent clustering has produced results which merit further
-investigation, the application to theory exploration has yet to be tested
-empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type
-class to generate random values for instantiating variables. Whilst we can
+It is currently infeasible to extend these clustering applications into full
+theory exploration runs, due to the use of constraints like \qcheck{}'s
+\hs{Arbitrary} type class (or LeanCheck's , in the case of Speculate) to
+generate suitable values for instantiating variables. Whilst we can
 automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
@@
 variables in our signature, which severely limits the possible combinations
-which can be explored. In many cases, no variables can be included at all,
+which can be explored. In many cases no variables can be included at all,
 leaving just equations involving constants. This has so far prevented us from
-measuring the direct impact on \qspec{} performance, either directly by
-exploring the sub-sets identified through recurrent clustering, or indirectly by
-comparing the equations generated by a full brute-force search to our recurrent
-clusters: those equations relating terms from different clusters would not be
-discovered by our method. It is this ratio of equations found through brute
-force to those found after narrowing-down by clusters which is one of our key
-objectives to maximise at this stage; until we begin to pursue the
-\emph{interestingness} of the properties.
+measuring the direct impact on \qspec{} performance on real code examples,
+either directly by exploring the sub-sets identified through recurrent
+clustering or indirectly by comparing the equations generated by a full
+brute-force search to our recurrent clusters.
 

### 2018-09-12 PhDThesis/background_haskell.tex
Thesis changes (e.g. lit review)

@@
   tools which we can to re-use, including conjecture generators like \qspec{}
-  and \speculate{}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
+  and \textsc{Speculate}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
   \textsc{SmartCheck}, \textsc{LeanCheck}, \textsc{Hedgehog}, etc.; theorem

### 2018-09-12 PhDThesis/litreview_automatedreasoning.tex
Thesis changes (e.g. lit review)

--- 2018-07-11 TEBenchmarkPaper/article.tex
@@
-% Derived from the template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{csvsimple}
-\usepackage[inline, shortlabels]{enumitem}
-\usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{tikz}
-\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
-  shapes.geometric}
-% etc.
+\subsection{Automated Mathematical Discovery and Invention}
 
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-\newcommand{\etal}{{\em et al.}}
-
-% For example theory of lists
-\newcommand{\Zero}{\text{Z}}
-\newcommand{\Succ}{\text{S}}
-
-\newcommand{\List}{\text{List}}
-\newcommand{\ListA}{\text{List} \  a}
-\newcommand{\Nil}{\text{Nil}}
-\newcommand{\Cons}{\text{Cons}}
-\newcommand{\Head}{\text{head}}
-\newcommand{\Tail}{\text{tail}}
-\newcommand{\Append}{\text{append}}
-\newcommand{\Reverse}{\text{reverse}}
-\newcommand{\Length}{\text{length}}
-\newcommand{\Map}{\text{map}}
-\newcommand{\Foldl}{\text{foldl}}
-\newcommand{\Foldr}{\text{foldr}}
-
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
-
-\begin{document}
-
-\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
-% {Grants or other notes about the article that should go on the front page
-% should be placed here. General acknowledgments should be placed at the end of
-% the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-% \titlerunning{Short form of title}        % if too long for running head
-
-\author{Chris Warburton \and
-        Alison Pease    \and
-        Jianguo Zhang}
-
-% \authorrunning{Short form of author list} % if too long for running head
-
-\institute{C. Warburton \at
-           University of Dundee \\
-           \email{c.m.warburton@dundee.ac.uk} \\
-           ORCID: 0000-0002-4878-6319 \\
-           Phone: +44 (0)1382 386968
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           A. Pease \at
-           University of Dundee \\
-           \email{a.pease@dundee.ac.uk} \\
-           ORCID: 0000-0003-1856-9599
-           \and
-           J. Zhang \at
-           University of Dundee \\
-           \email{j.n.zhang@dundee.ac.uk} \\
-           ORCID: 0000-0001-9317-0268
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-  We propose a benchmarking methodology to evaluate the efficiency and quality
-  of \emph{conjecture generation} by automated tools for \emph{mathematical
-    theory exploration}. Our approach uses widely available theorem proving
-  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
-  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
-  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
-  and produces significantly more ``interesting'' output. By defining a standard
-  benchmark we provide a measure of progress for the field, encourage
-  innovation through healthy competition and allow direct comparisons to be made
-  between the disparate approaches currently being pursued for this task.
-  \keywords{Theory Formation \and Theorem Proving \and Lemma Discovery \and Conjecture Generation \and Benchmarking}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\begin{acknowledgements}
-  We are grateful to the authors of the systems we have used in our experiments,
-  especially for help in obtaining and configuring their tools. We would
-  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
-  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
-  their software to our purposes.
-
-  This work was supported by the Engineering and Physical Sciences Research
-  Council grant EP/P017320/1 and studentship 1526504.
-\end{acknowledgements}
-
-\pagebreak
-
-\section{Introduction}
-\label{intro}
-
-\emph{Conjecture generation} is the open-ended task of conjecturing properties
-of a given logical theory which are somehow ``interesting'', and is
-studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
-applications wherever we find uses for formal methods: in proof assistants and
-their libraries, in mathematics education and research, and in the specification
-and verification of software. Although formal methods, and automated reasoning
-more generally, have been advocated in mathematics (notably by
-Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
-and software engineering (for example via dependently-typed
-programming~\cite{McKinna:2006}), they have a reputation for being difficult and
-expensive, which has historically limited their use to high-assurance domains
-such as aerospace and microprocessor design.
-
-One path to reducing these barriers is to increase the level of automation.
-Software verification tools have been shown to benefit from the addition of a
-conjecture generation step for finding ``background lemmas'', which are useful
-when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
-Another path is to tailor the division of labour between humans and machines
-such that each is playing to their strengths, which has been referred to as
-``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
-systematically search a much larger space of possible relationships than humans,
-potentially bringing to light novel and surprising connections, whilst allowing
-the user to make the ultimate judgement over which of the most promising results
-are worth investigating further.
-
-Similar tasks are also studied in the domain of Computational \emph{Scientific}
-Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
-example in the search for new drugs. The major difference between scientific and
-mathematical discovery is that inductive reasoning and experimental testing are,
-in principle, more important for the former, although we note that they are also
-core principles in the mathematical tools we have investigated (in particular
-for ensuring the \emph{plausibility} of conjectures).
-
-We limit ourselves to mathematical applications, but even here it is difficult
-to measure and compare the success of approaches and tools. This is partially
-due to their diversity, but also because of the inherent ambiguity of the task
-(what counts as ``interesting''?), the different goals emphasised by their
-designers and the variety of evaluation methods employed. We attempt to solve
-this discrepancy, at least for the foreseeable future, by defining a standard,
-unambiguous benchmarking approach with which to compare the conjecture
-generation of MTE tools. Our contributions include:
-
-\begin{itemize}
-\item A general methodology for benchmarking conjecture generation.
-\item Resolving the issue of ``interestingness'' through the use of
-  theorem-proving benchmarks as a ground-truth.
-\item A specific instantiation of this methodology, using the Tons of Inductive
-  Problems (TIP) benchmark as a corpus.
-\item Automated tooling to perform this benchmarking.
-\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
-  and a comparison and discussion of the results.
-\end{itemize}
-
-We describe the conjecture generation problem in more detail, along with the
-difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
-explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
-demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
-Issues facing our approach, and the field in general, are discussed in
-$\S$\ref{sec:discussion}; related work is surveyed in
-$\S$\ref{sec:related-work}, including the diverse terminology found in the
-literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.
-
-\begin{sloppypar}
-  Our benchmark implementation is available at
-  \url{http://chriswarbo.net/projects/theory-exploration} and
-  \url{https://github.com/warbo}. All processes, including benchmark generation,
-  experimental runs and generation of this paper, are specified using the
-  Nix~\cite{dolstra2004nix} system to aid reproducibility.
-\end{sloppypar}
-
-\section{Background}
-\label{sec:background}
-
-\subsection{Motivation}
-\label{sec:motivation}
-
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      \forall a. \Nil            &: \ListA                                  \\
-      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
-      \Head(\Cons(x, xs))        &= x                                       \\
-      \Tail(\Cons(x, xs))        &= xs                                      \\
-      \Append(\Nil,         ys)  &= ys                                      \\
-      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
-      \Reverse(\Nil)             &= \Nil                                    \\
-      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
-      \Length(\Nil)              &= \Zero                                   \\
-      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
-      \Map(f, \Nil)              &= \Nil                                    \\
-      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
-      \Foldl(f, x, \Nil)         &= x                                       \\
-      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
-      \Foldr(f, \Nil,         y) &= y                                       \\
-      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
-    \end{split}
-  \end{equation*}
-  \caption{A simple theory defining a $\List$ type and some associated
-    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
-    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers}
-  \label{figure:list_theory}
-\end{figure}
-
-Given a logical theory, such as the theory of lists shown in
-Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
-invariants) describing its behaviour. This could be for mathematical curiosity,
-or due to the theory's importance in some domain. In particular, when dealing
-with definitions in a software library, we may regard a set of such properties
-as the code's \emph{specification}. Even those properties which only hold by
-coincidence, rather than by design, may still be useful for \emph{optimising}
-uses of this library, by rewriting expressions into equivalent forms which
-require less time, memory, network usage, etc. For example, our theory of lists
-turns out to satisfy the following invariant:
-
-\begin{equation} \label{eq:mapreduce}
-  \forall f. \forall xs. \forall ys.
-    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
-\end{equation}
-
-Whilst these two expressions produce equivalent results, the form on the right
-can execute both calls to $\Map$ in parallel (in a pure functional language, at
-least); a fact which is exploited by the ``map/reduce'' parallel programming
-paradigm.
-
-Such use cases require the ability to \emph{discover} true properties of some
-particular definitions; this requires not only \emph{proving} their correctness,
-but also \emph{posing} them as conjectures in the first place. This is a hard
-problem in general, but presents opportunities for automation due to the
-precise, symbolic nature of the domain. These processes can also feed into each
-other, since proven properties can be used as lemmas for subsequent proofs. As
-an example, proving Equation~\ref{eq:mapreduce} may require ``background
-knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
-be tedious to produce by hand but possible to discover automatically.
-
-\subsection{Automating Construction and Exploration of Theories}
-\label{sec:te}
-
-The task of automatically proving or disproving a \emph{given} conjecture has
-been studied extensively in the field of Automated Theorem Proving (ATP). Less
-attention has been paid to \emph{generating} such conjectures automatically,
-either from a given set of definitions (the task we will focus on) or where the
-definitions are also generated. This research has been pursued under the names
-of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
-we will use the MTE terminology, and discuss their relationship further in
-$\S$\ref{sec:related-work}.
-
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
-term with many different interpretations. For example, all existing approaches
-agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
-concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
-which are applied to conjectured properties as follows:
-
-{\bf Empirical plausibility} checks whether a property holds across some
-specific examples. This is especially useful for avoiding falsehoods, without
-resorting to a deductive proof search.
-
-{\bf Novelty} depends on whether the property, or one isomorphic or more
-general, has already been seen.
-
-{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
-example if it is an instance of a tautology.
-
-{\bf Applicability} depends on the number of models in which a
-property holds. High applicability makes a property more interesting, but so
-does zero applicability (i.e. non-existence).
-
-{\bf Comprehensibility} depends on the syntactic complexity of the property's
-statement. Simpler statements are considered more interesting (which favours
-tools adopting a small-to-large search order).
-
-{\bf Utility} is the relevance or usefulness of a property to the user's
-particular task. For example, if we want to find properties which are useful for
-optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
-whether the property justifies some rewrite rule, the difference in resource
-usage of the expressions involved, and how common those expressions are in real
-usage.
-
-Attempting to \emph{measure} such qualities is difficult, and having too many
-measures complicates comparisons. System developers have employed a more
-practical alternative in their evaluations, which is to perform
-\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
-choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
-and a set of properties (the ground truth) which represent the ``ideal'' result
-of conjecture generation for those definitions (e.g. this might include
-Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
-run it on these chosen definitions and compare its output to the ground truth:
-
-\begin{itemize}
-\item \emph{Precision} is the proportion of a tool's output which appears in
-  the ground truth (the ratio of true positives to all positives). This
-  penalises overly-liberal tools which output a large number of properties in
-  the hope that some turn out to be ``good''.
-\item \emph{Recall} is the proportion of the ground truth which appears in the
-  tool's output (the ratio of true positives to actual positives). This
-  penalises overly-conservative tools which generate very few properties as a
-  way to avoid ``bad'' ones.
-\end{itemize}
-
-To score 100\% on precision and recall, all of the properties which appear in
-the ground truth must have been conjectured, and nothing else. This gives us a
-simple method to evaluate and compare tools without requiring a general solution
-to the question of what is ``interesting''; although we must still decide what
-to put in the ground truth set, and what to leave out, for each measured set of
-definitions.
-
-\begin{sloppypar}
-  The precision and recall of three MTE tools are compared
-  in~\cite{claessen2013automating}:
-  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
-  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
-  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
-    comparison only reports prior results for each tool, rather than reproducing
-    them. This makes the numbers less comparable, since each tool was tested
-    with slightly different definitions (e.g. whether or not the exponential and
-    maximum functions were included).} The definitions and ground truths for all
-  three were taken from the standard library of the Isabelle proof assistant.
-  The fact that the library authors expended the effort of stating, proving and
-  bundling these properties with every copy of their software is a good
-  indication that they are interesting. Two sets of definitions were used: a
-  theory of natural numbers; and a theory of lists (which we show in
-  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
-  only two datapoints, each containing only a few definitions and properties (4
-  functions and 12 properties, in the natural number example), it is difficult
-  to draw general conclusions; yet despite these limitations we believe that
-  this is a viable method to performing meaningful, objective analyses and
-  comparisons for such tools. It is only with larger data sets that we can
-  generalise the measured performance to different, especially \emph{novel},
-  domains (which, after all, is the purpose of MTE tools). In addition, we might
-  be concerned that such ``popular'' theories as the natural numbers may be
-  unrepresentative of typical usage. For example, we would not expect the
-  library functions in a typical verification problem to be as mathematically
-  rich as number theory.
-\end{sloppypar}
-
-\subsection{Existing Tools}
-\label{sec:existing-tools}
-
-Since our benchmarking methodology builds on the precision/recall analysis
-described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
-and QuickSpec (the conjecture generating component of HipSpec), as our
-demonstration.
-
-IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
-assistant, mostly in Standard
-ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
-by enumerating expressions involving a given set of (typed) constants and free
-variables (a \emph{signature}). A constraint solver forbids certain
-(sub)expressions from being synthesised, and these constraints are extended
-whenever a new property is conjectured, to avoid generating any special-cases of
-this property in the future. Conjectures are tested by looking for
-counterexamples and, if none are found, sent to IsaPlanner which attempts to
-prove them.
-
-QuickSpec emerged from work on the QuickCheck software testing framework for the
-Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for HipSpec, it has been applied
-to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
-QuickSpec takes a signature and enumerates well-typed terms. It collects
-together those of the same type into equivalence classes, assuming them to be
-equal, then uses QuickCheck to find counterexamples to this assumption by
-randomly instantiating the variables and comparing the resulting closed
-expressions. Any equivalence class whose elements don't compare equal are split
-up, and the process is repeated with new random values.
-
-After 500 rounds of testing, any expressions still sharing the same equivalence
-class are conjectured to be equal for all values of their variables. Rather than
-using a constraint system to prevent redundancies (such as special-cases of
-other properties), QuickSpec instead passes its output through a congruence
-closure algorithm to achieve the same effect.
-
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
-
-Our main contribution is a benchmarking methodology, shown in
-Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
-generates a large definition/ground-truth corpus, and provides a scalable,
-statistical approach to evaluating MTE tools using this corpus. We follow a
-precision/recall approach similar to prior work, with the main difference being
-the source of definitions and ground truths: we take existing problem sets
-designed for automated theorem proving, and adapt their content for use in the
-conjecture generation setting.
-
-\subsection{Preparation}
-\label{section:prep}
-
-Automated theorem proving is an active area of research, with large problem sets
-and regular competitions to prove as much as possible, as fast as
-possible~\cite{pelletier2002development}. These problem sets are an opportunity
-for MTE, as their definitions and theorems can be used as a corpus in the same
-way that Isabelle libraries have been used in the past.
-
-Some problem sets are more amenable for this purpose than others. The most
-suitable are those meeting the following criteria:
-
-\begin{itemize}
-\item For each problem, there should be a clear distinction between the
-  statement to be proved and the definitions involved, such that the two can be
-  easily and meaningfully separated. This rules out problem sets like those of
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions, whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item Definitions should be translatable into a form suitable for the MTE tools
-  under study. Our application in $\S$\ref{sec:application} requires Haskell and
-  Isabelle translations, and also benefits from having definitions be strongly
-  typed.
-\item The problem set should be relevant to the desired domain. We focus on pure
-  functional programming, which requires higher-order functions and inductively
-  defined datatypes, which rules out first-order languages/logics (such as
-  TPTP~\cite{sutcliffe2009tptp}).
-\item The problem set should ideally contain \emph{every} ``interesting''
-  property involving its included definitions, since non-membership in the
-  ground truth will be treated as being ``uninteresting''. More realistically,
-  we should aim for each definition to have \emph{multiple} properties; rather
-  than a mixture of unrelated problems, cherry-picked from different domains.
-\item Larger problem sets are preferable as they give more robust statistics, all
-  else being equal (i.e. when this does not sacrifice quality).
-\end{itemize}
-
-Once such a problem set has been chosen, we must separate the definitions from
-the theorem statements which reference those definitions. The definitions will
-be used as input to the MTE tools, whilst the theorem statements form the ground
-truth corpus of properties (which tool output will be compared against).
-
-It is important to ensure that there are no duplicate definitions: we are only
-concerned with the \emph{logical} content of the input, not the more arbitrary
-aspects of their presentation like the names of functions. For example, consider
-a problem set which includes a statement of commutativity for a \texttt{plus}
-function, and of associativity for an \texttt{add} function, where the
-definitions of \texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would
-expect an MTE tool to either conjecture commutativity and associativity for
-\emph{both} functions, or for \emph{neither} function, since they are logically
-equivalent. Yet a na\"ive precision/recall analysis would treat commutativity of
-\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
-they don't appear in the ground truth.
-
-For this reason, duplicates should be removed, and any references to them
-updated to use the remaining definition (e.g. chosen based on lexicographic
-order). In the above example, the \texttt{plus} function would be removed, and
-the commutativity statement updated to reference \texttt{add} instead.
-
-\subsection{Sampling}
-\label{section:sampling}
-
-We could, in theory, send these de-duplicated definitions straight into an MTE
-tool and use the entire set of properties (taken from the theorem statements) as
-the ground truth for analysis. However, this would cause two problems:
-
-\begin{itemize}
-\item The result would be a single data point, which makes it difficult to
-  infer performance \emph{in general}.
-\item It is impractical to run existing MTE tools on inputs containing more
-  than a few dozen definitions.
-\end{itemize}
-
-To solve both of these problems we instead \emph{sample} a subset of
-definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order based on cryptographic hashes.} Given a sample size,
-we choose a subset of that many definitions, and provide only those as input to
-the tool. We generate a corresponding ground truth by selecting those properties
-from the corpus which ``depend on'' (contain references to) \emph{only} the
-definitions in that sample. Transitive dependencies aren't required (e.g. a
-property involving only a \texttt{times} function would not depend on a
-\texttt{plus} function, even if \texttt{plus} occurs in the definition of
-\texttt{times}).
-
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an
-empty ground truth, and thus $0$ precision and undefined recall
-\emph{independent} of the tool's output! This is clearly undesirable as an
-evaluation method.
-
-Other measurements could be used for such cases, but our approach is to avoid
-them by only allowing a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output. One downside of this restriction is that we limit the
-number of possible samples to measure; this is significant for sample sizes less
-than~3.
-
-\subsection{Evaluation}
-\label{section:evaluation}
-
-Given a sample of definitions and a corresponding ground truth, the actual
-execution of the MTE tool proceeds as in prior work. We must translate the
-chosen definitions into the required input format, then we time the execution
-with a timeout (e.g. 5 minutes). We use wall-clock time since
-\begin {enumerate*} [i) ]%
-\item this is most relevant to a user's experience,
-\item it has a straightforward interpretation and
-\item measuring it does not require intrusive alterations to a tool's
-  implementation.
-\end {enumerate*}
-The downside is that comparisons must take hardware performance into account,
-which would not be the case if time were measured by some proxy like number of
-expressions evaluated.
-
-In our experiments we have found that memory usage is also an important part of
-a tool's performance, but rather than complicating our analysis with an extra
-dimension, we instead allow programs to use as much memory as they like, and
-either get killed by the operating system or slow down so much from swapping
-that they time out. This is in line with the expected usage of these tools:
-either there is enough memory, or there isn't; implementations should not be
-penalised for making use of available resources.
-
-To calculate precision and recall, the conjectures generated by each tool, as
-well as the ground-truth properties, need to be parsed into a common format and
-compared syntactically after normalising away ``irrelevant'' details. We
-consider variable naming to be irrelevant, which can be normalised by numbering
-free variables from left to right and using de Bruijn indices for bound
-variables. We also consider the left/right order of equations to be irrelevant,
-which we normalise by choosing whichever order results in the
-lexicographically-smallest expression.
-
-We specifically \emph{ignore} other logical relationships between syntactically
-distinct statements, such as one equation being implied by another. Whilst
-logically sound, second-guessing the ground truth in this way would harm other
-aspects which influence interestingness (for example, a more general statement
-is more widely applicable, but it might also be harder to comprehend).
-
-This procedure gives us, for each sample, a set of generated conjectures and a
-set of ground truth properties (from which precision and recall can be
-calculated) and a single runtime. We propose two methods for analysing this
-data: \emph{summarising} the performance of a single MTE tool and
-\emph{comparing} the performance of two tools.
-
-\subsection{Summarising}
-
-Each sample is only explored once by each tool, so that we cover as many
-independent samples as possible to better estimate how a tool's performance
-generalises to unseen inputs. How we combine these data into an aggregate
-summary depends on what we are interested in measuring. One general question we
-might ask is how a tool's performance scales with respect to the input size (the
-number of given definitions). This is straightforward to measure by varying the
-sample size, but we need some way to combine the results from samples of the
-same size.
-
-We can summarise a set of runtimes by choosing the median, as this is more
-robust than the mean against long-running outliers, and hence represents
-performance for a ``typical'' input of that size. Since our methodology measures
-performance over many samples, we can also compute the \emph{spread} of our
-data, for example the inter-quartile range. This is not possible using the
-one-off measurements common in prior evaluation methods.
-
-Precision and recall are ratios, which can be averaged (separately) in two ways,
-known as the \emph{Average of Ratios} (AoR) and the \emph{Ratio of Averages}
-(RoA)~\cite{egghe2012averages}. The Average of Ratios is the mean value. Using
-the definitions of precision and recall given in $\S$\ref{sec:te}, we can define
-their means for $S$ samples indexed by $1 \leq i \leq S$ as follows:
-
-\begin{align*}
-       \overline{\text{precision}}_{AoR}
-    &= \frac{1}{S} \sum_i{\text{precision}_i} \\
-    &= \frac{1}{S} \sum_i{
-         \frac{\#\text{interesting}_i}
-              {\#\text{generated}_i}}         \\[10pt]
-       \overline{\text{recall}}_{AoR}
-    &= \frac{1}{S} \sum_i{\text{recall}_i} \\
-    &= \frac{1}{S} \sum_i{
-         \frac{\#\text{interesting}_i}
-              {\#\text{groundtruth}_i}}
-\end{align*}
-
-We interpret these mean values as the \emph{expected} precision and recall for
-samples of this size. In particular, these tell us the quality of the \emph{set}
-of conjectures that will be generated if we were to run the tool on such a
-sample, but they don't give us direct information about the individual
-conjectures themselves. Since these are mean values, they are the reference
-point for measures of spread such as the variance, standard deviation and mean
-absolute deviation.
-
-The Ratio of Averages is \emph{pooled}, combining the (multi)sets from each
-sample before taking the ratios. The size of these pooled sets is simply the sum
-of the individual set sizes, and ratios of these sums are equal to ratios of the
-means (since the normalising factor $\frac{1}{S}$ appears in both numerator and
-denominator):
-
-\begin{align*}
-       \overline{\text{precision}}_{RoA}
-    &= \frac{\sum_i{\#\text{interesting}_i}}
-            {\sum_i{\#\text{generated}_i}}                              \\
-    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
-            {\frac{1}{S} \sum_i{\#\text{generated}_i}}                  \\
-    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
-            {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
-       \overline{\text{recall}}_{RoA}
-    &= \frac{\sum_i{\#\text{interesting}_i}}
-            {\sum_i{\#\text{groundtruth}_i}}                            \\
-    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
-            {\frac{1}{S} \sum_i{\#\text{groundtruth}_i}}                \\
-    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
-            {\hspace{5pt} \overline{\#\text{groundtruth}} \hspace{5pt}}
-\end{align*}
-
-Larger sets contribute more to a Ratio of Averages, unlike the mean which treats
-each set equally. This weighting gives us expected qualities of
-\emph{conjectures} rather than sets: the RoA for precision is the expected
-chance that some particular generated conjecture will appear in the ground
-truth; the RoA for recall is the expected chance that some particular ground
-truth property will appear in the generated output.
-
-These averages can differ when there is a large variation in set size. For
-example, if we generate a single, interesting conjecture for one sample and 99
-uninteresting conjectures for another sample, the AoR (mean) precision will be
-$\frac{1}{2}$ but the RoA precision will be $\frac{1}{100}$. This highlights the
-importance of knowing \emph{absolute} numbers of conjectures, such as the mean
-output size $\overline{\#\text{generated}}$ and its spread, alongside the
-proportions given by precision and recall.
-
-It is also possible to combine precision values \emph{with} recall values
-(either individually or aggregated) by calculating their \emph{F-score}, defined
-as:
-
-\begin{equation*}
-  F = 2 \times \frac{\text{precision} \times \text{recall}}
-                    {\text{precision} +      \text{recall}}
-\end{equation*}
-
-This is the harmonic mean of precision and recall, ranging between 0 and 1. The
-F-score summarises performance into a single number, which is useful for
-purposes like ranking, but it obscures insights we might gain from looking at
-precision and recall separately (e.g. whether a tool is generating too much
-output or too little).
-
-\subsection{Comparison}
-
-To measure progress and encourage competition in the field, it is important to
-compare the relative performance of different tools on the same task. Since the
-aggregate statistics in the above summaries have obscured the details of
-specific runs, any comparison based on them (such as comparing mean recall)
-would have very low statistical power. More direct comparisons can be made using
-\emph{paired} tests, since for each individual sample we have measurements for
-\emph{both} tools.
-
-Running times do not follow a normal distribution, in particular due to the
-lower bound at 0, which complicates any application of standard comparisons like
-the paired z-test. An alternative which does not assume normality is the
-Wilcoxon signed-rank test~\cite{wilcoxon1945individual}, which has been applied
-to software benchmarking in the Speedup-Test protocol of Touati, Worms and
-Briais~\cite{touati2013speedup}. Our methodology differs by measuring with a
-different sample each time, rather than repeatedly measuring one sample of each
-size; measuring both tools on the \emph{same} samples allows the paired form of
-the Wilcoxon test to be used.
-
-Another complication with our time measurements is the censoring effect caused
-by timeouts. Paired difference tests are robust to this, since the upper-bound
-on total time imposes an upper-bound to the observable difference; this biases
-our conclusions in a conservative way, \emph{towards} the null hypothesis of
-indistinguishable performance.
-
-Quality can be compared using precision and recall, but we must choose how to
-handle samples where one tool succeeded and the other failed (e.g. by timing
-out). Failed runs can be treated as if they had succeeded with an empty set of
-conjectures; this penalises failure-prone tools and simplifies analysis by
-ensuring every sample produces a data point. An alternative is to discard
-samples which caused either tool to fail, and compare only those where both
-tools finished successfully. We follow this second approach, since it is more
-charitable to the tools being evaluated (which, after all, were not designed
-with our benchmark in mind). With fewer data points to analyse, we pool together
-samples of all sizes (allowing duplicates) and compare proportions from these
-aggregations.
-
-Recall relies on a fixed set of properties, the ground truth, so we can compare
-tools using McNemar's test for paired samples~\cite{mcnemar1947note}. For each
-tool, we count how many of the ground-truth properties it found which were
-\emph{not} found by the other tool; i.e. those occasions where one tool was
-objectively better than the other. McNemar's test determines whether we can
-reject the null hypothesis that both counts are the same.
-
-Comparing precision cannot be done in the same way as recall, since we do not
-have a fixed set of properties to compare both tools on (there are an unlimited
-number of properties that may or may not appear in each tool's output). Instead,
-we can count how many of the generated conjectures were interesting and
-uninteresting for each tool, and use Boschloo's form of Barnard's
-test~\cite{lydersen2009recommended} to determine if these proportions are
-independent. This assumes that precision follows a binomial distribution, and
-the test is conditioned on the number of generated conjectures (as if this were
-fixed by design, rather than determined by the tool).
-
-\begin{figure}
-  \centering
-  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
-                           minimum height=1cm,text centered, draw=black]
-
-  \tikzstyle{io} = [trapezium, trapezium left angle=70,
-                    trapezium right angle=110, minimum width=1cm,
-                    minimum height=1cm, text centered, draw=black]
-
-  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
-                         text centered, draw=black]
-  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
-                          text centered, draw=black]
-
-  \tikzstyle{arrow} = [thick,->,>=stealth]
-
-  % Avoids too much padding in io shapes
-  \tikzset{trapezium stretches=true}
-
-  \begin{tikzpicture}[node distance=1cm]
-
-    % Preparation section
-
-    \node (in)  [io]{Theorem Proving Benchmark};
-    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
-    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
-    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};
-
-    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
-    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};
-
-    \draw [arrow] (in)  -- (sep);
-    \draw [arrow] (def) -- (ref);
-    \draw [arrow] (def) -- (thy);
-    \draw [arrow] (ref) -- (thm);
-
-    % Arrows with labels
-    \path[->]
-        (sep) edge [arrow, sloped, above] node {definitions} (def)
-        (sep) edge [arrow, sloped, above] node {properties}  (ref);
-
-    % Sampling section
-
-    \node (choose) [process, below=of thm   ]{Choose a property};
-    \node (deps)   [process, below=of choose]{List property's dependencies};
-
-    % Create dummy coordinate below deps, then use its y coordinate for pad
-    \coordinate [below=of deps] (padDummy);
-    \path let \p{dummy} = (padDummy),
-              \p{in}    = (in)
-              in coordinate (padPos) at (\x{in}, \y{dummy});
-    \node (pad)  [process, at=(padPos)]{Pad list to form sample};
-
-    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
-    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};
-
-    % Calculate position of (size) using let, then define as normal
-    \path let \p{pad}    = (pad),
-              \p{choose} = (choose)
-           in coordinate (sizePos) at (\x{pad},\y{choose});
-    \node (size) [io, at=(sizePos)] {Sample size};
-
-    % Evaluation section
-    \path let \p{pad}  = (pad),
-              \p{sthm} = (sthm)
-           in coordinate (runPos) at (\x{pad}, \y{sthm});
-    \node (run) [process, below=of runPos]{Run MTE tool};
-    \node (pr)  [process, below=of run]{Analysis};
-
-    \node (prec) [startstop, below=of pr  ]{Precision};
-    \node (time) [startstop,  left=of prec]{Time taken};
-    \node (rec)  [startstop, right=of prec]{Recall};
-
-    \draw [arrow] (thy)    |- (pad);
-    \draw [arrow] (thm)    -- (choose);
-    \draw [arrow] (choose) -- (deps);
-    \draw [arrow] (deps)   |- (pad);
-    \draw [arrow] (size)   -- (choose);
-    \draw [arrow] (size)   -- (pad);
-    \draw [arrow] (pad)    -- (sthm);
-    \draw [arrow] (pad)    -- (sthy);
-    \draw [arrow] (sthy)   |- (run);
-    \draw [arrow] (run)    -- (pr);
-    \draw [arrow] (sthm)   |- (pr);
-    \draw [arrow] (pr)     -- (prec);
-    \draw [arrow] (pr)     -- (rec);
-    \draw [arrow] (pr)     -- (time);
-
-    % Awkward arrow
-    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);
-
-    % Braces
-
-    % Preparation
-    \draw
-      let \p{thm} = (thm.east),
-          \p{in}  = (in.north),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
-              {Preparation $\S$\ref{section:prep}};
-
-    % Sampling
-    \draw
-      let \p{thm} = (thm.east),
-          \p{gt}  = (sthm.east),
-          \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
-              {Sampling $\S$\ref{section:sampling}};
-
-    % Evaluation
-    \draw
-      let \p{sthm} = (sthm.east),
-          \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
-       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
-              {Evaluation $\S$\ref{section:evaluation}};
-
-  \end{tikzpicture}
-  \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
-    \begin{tikzpicture}
-      \node [rectangle, text centered, draw=black]{processes};
-    \end{tikzpicture},
-    \begin{tikzpicture}
-      \node [rectangle, rounded corners, text centered, draw=black]{data};
-    \end{tikzpicture} and
-    \begin{tikzpicture}
-      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
-             text centered, draw=black]{inputs};
-  \end{tikzpicture}}
-  \label{figure:flow_chart}
-\end{figure}
-
-\section{Application}
-\label{sec:application}
-
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
-
-To determine our benchmarking parameters we ran some initial tests on both tools
-for a few samples sized between 1 and 100, for an hour each on our benchmarking
-machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
-on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
-above this took thousands of seconds each, which we deemed infeasibly long.
-
-Based on this we decided to benchmark sample sizes up to 20, since neither tool
-seemed to perform well beyond that. The Speedup-Test protocol follows the
-statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
-so we pick 31 samples of each size in order to cross this threshold. This gave a
-total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
-returns. During analysis, duplicate samples (caused by our requirement that
-samples have a non-empty ground truth) were found and discarded, so only 14
-samples of size 1 were used and 30 of size 2.
-
-\subsection{Tons of Inductive Problems}
-\label{sec:tip}
-
-We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
-since it satisfies the criteria specified in $\S$\ref{section:prep}: each
-benchmark problem has standalone type and function definitions, making their
-separation trivial; known examples from the software verification and inductive
-theorem proving literature are included, ensuring relevance to those fields; the
-format includes the higher-order functions and inductive datatypes we are
-interested in; it is large enough to pose a challenge to current MTE tools; plus
-it is accompanied by tooling to convert its custom format (an extension of
-SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
-Isabelle.
-
-We use TIP version 0.2 which contains 343 problems, each stating a single
-property and together defining a total of 618 datatypes and 1498 functions. Most
-of these are duplicates, since each problem (re\nobreakdash-)defines all of the
-datatypes and functions it involves.
-
-TIP datatypes can have several ``constructors'' (introduction forms) and
-``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
-follows:
-
-\begin{samepage}
-\begin{verbatim}
-(declare-datatypes
-  (a)                       ;; Type parameter (element type)
-  ((List                    ;; Type name
-     (Nil)                  ;; Constructor (nullary)
-     (Cons                  ;; Constructor (binary)
-       (head a)             ;; Field name and type
-       (tail (List a))))))  ;; Field name and type
-\end{verbatim}
-\end{samepage}
-
-Our target languages (Haskell and Isabelle) differ in the way they handle
-constructors and destructors, which complicates comparisons. To avoid this, we
-generate a new function for each constructor (via $\eta$-expansion) and
-destructor (via pattern-matching) of the following form:
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                   ;; Type parameter
-    (constructor-Cons        ;; Function name
-      ((x a) (xs (List a)))  ;; Argument names and types
-      (List a)               ;; Return type
-      (as                    ;; Type annotation
-        (Cons x xs)          ;; Return value
-        (List a)))))         ;; Return type
-\end{verbatim}
-\end{samepage}
-
-\begin{samepage}
-\begin{verbatim}
-(define-fun
-  (par (a)                        ;; Type parameter
-    (destructor-head              ;; Function name
-      ((xs (List a)))             ;; Argument name and type
-      a                           ;; Return type
-      (match xs                   ;; Pattern-match
-        (case (Cons h t) h)))))   ;; Return relevant field
-\end{verbatim}
-\end{samepage}
-
-\begin{sloppypar}
-  We rewrite the TIP properties (our ground truth) to reference these expanded
-  forms instead of the raw constructors and destructors, and use these functions
-  in our samples in lieu of the raw expressions. Note that these destructor
-  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
-  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
-  complicates their translation to proof assistants like Isabelle.
-\end{sloppypar}
-
-Another complication is TIP's ``native'' support for booleans and integers,
-which allows numerals and symbols like \texttt{+} to appear without any
-accompanying definition. To ensure consistency in the translations, we replace
-all occurrences of such expressions with standard definitions written with the
-``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
-mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
-  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
-  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
-  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
-  symmetry.}
-
-When we add all of these generated types and functions to those in TIP, we get a
-total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
-and we choose to only sample from those 182 functions which are referenced by at
-least one property (this removes ambiguity about which \emph{definitions} count
-as interesting and which are just ``implementation details'' for other
-definitions).
-
-TIP comes with software to translate its definitions into Haskell and Isabelle
-code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
-which is imported on each run of the tools, although only those functions which
-appear in the current sample are included in the signature and explored. We also
-encode all names in hexadecimal to avoid problems with language-specific naming
-rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
-distinguishes these from local variables and prevents names from beginning with
-a digit). This ensures that the generated conjectures will be using the same
-names as the ground truth, rather than some language-specific variant.
-
-\subsection{QuickSpec}
-
-We benchmarked QuickSpec version 0.9.6, a tool written in Haskell for
-conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
-need to automate some of the decisions which are normally left up to the user:
-
-\begin{sloppypar}
-  \begin{itemize}
-  \item We must decide what variables to include. We choose to add three
-    variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
-  \item We must \emph{monomorphise} all types. For example, functions like
-    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
-    element type, but we need to pick a specific type in order to know which
-    random value generator to use. We resolve this (arbitrarily) by picking
-    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
-      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
-      \texttt{[]} (Haskell's list type constructor). If these violate some
-      type class constraint, we pick a suitable type non-deterministically from
-      those in scope during compilation; if no suitable type is found, we give
-      up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
-    during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
-    assign an arity to each function in the signature, from 0 to 5, so we pick
-    the highest that is type-correct, since this avoids a proliferation of
-    incomparable, partially-applied functions.
-  \end{itemize}
-\end{sloppypar}
-
-\begin{figure}
-  \centering
-  \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
-    is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses show runs which timed out, which occurred for every size. Each size
-    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
-    sampling restrictions. Box plots show inter-quartile range, which reaches
-    the timeout for sizes above 8. Medians are marked with a line and remain
-    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
-    show 1.5$\times$~inter-quartile range}
-  \label{figure:quickspec_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
-    horizontally to reduce overlaps). Lines show the mean proportion for each
-    sample size (Average of Ratios) and errorbars show the sample standard
-    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
-    around $0.1$ for size 20, whilst mean recall remains roughly flat between
-    $0.2$ and $0.5$}
-  \label{figure:quickspec_precRec}
-\end{figure}
-
-The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
-failures were due to timing out, and these occurred for all sample sizes but are
-more common for larger samples (over half the runs for sample sizes 13 and 20
-timed out). Runs which succeeded mostly finished within 30 seconds, generating
-few or no conjectures; those taking longer generated more conjectures, with the
-most being 133 conjectures from a sample of size 19.
-
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
-added to a signature (assuming sufficient testing), its decreasing precision
-can't be due to finding fewer wanted conjectures at larger sizes. This is
-supported by the relative flatness of the recall results. Rather, the number of
-conjectures generated is increasing at a higher rate than the size of the ground
-truth. These extra conjectures may involve those ``padding'' definitions in a
-sample which don't contribute to its ground truth, or may be ``uninteresting''
-relationships between the dependencies of different ground truth properties.
-
-This indicates two potential improvements to the QuickSpec algorithm (as far as
-this benchmark is concerned). The deluge of generated conjectures could be
-filtered down to a more desirable sub-set by another post-processing step.
-Alternatively, rather than exploring all of the given definitions together,
-multiple smaller signatures could be selected from the input by predicting which
-combinations are likely to lead to interesting conjectures; this would avoid
-both the ``padding'' and the cross-dependency relationships. Both of these
-methods could improve the precision, although care would be needed to avoid a
-large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
-signatures may be faster to explore than a single large one. Such improvements
-would also need to be ``generic'', to avoid over-fitting to this particular
-benchmark.
-
-QuickSpec's recall is limited by two factors: the algorithm is unable to
-synthesise some properties, such as conditional equations, inequalities, terms
-larger than the search depth and those containing anonymous functions. The
-congruence closure algorithm used as a post-processor may also be removing
-``interesting'' results, for example if we found an ``uninteresting'' result
-which is more general.
-
-\subsection{IsaCoSy}
-
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
-with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
-
-\begin{itemize}
-\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
-  encoded in Isabelle. We strip out this type and the functions which depend on
-  it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
-  definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
-  automatically be included in the exploration. Since those constructors will
-  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
-  even those may not be present in the current sample) this will unfairly reduce
-  the calculated precision. To avoid this, we add a post-processing step which
-  replaces all occurrences of a constructor with the corresponding wrapper, then
-  discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
-  constructors unnecessarily, but it at least does not bring down the quality
-  measures.
-\item Since Isabelle is designed for theorem proving rather than programming, it
-  requires every definition to be accompanied by proofs of exhaustiveness and
-  termination. These are difficult to generate automatically, and don't exist in
-  the case of partial functions like destructor wrappers. Hence we use the
-  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
-  the \texttt{sorry} keyword.
-\item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
-  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
-\end{itemize}
-
-\begin{figure}
-  \centering
-  \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
-    is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
-    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
-    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
-    restrictions. Box plots show inter-quartile range, which grows with size,
-    and lines mark the median time, which increases rapidly from around 100
-    seconds at size 1 to timing out at size 4 and above. Whiskers show
-    1.5$\times$~inter-quartile range}
-  \label{figure:isacosy_runtimes}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
-    horizontally to reduce overlaps); there were no successes above size 6.
-    Lines show the mean proportion for each sample size (Average of Ratios),
-    which trends downwards for precision and appears flat for recall, although
-    there are too few datapoints to be definitive. Errorbars show the sample
-    standard deviation}
-  \label{figure:isacosy_precRec}
-\end{figure}
-
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
-size 6 failed, with most timing out and a few aborting early due to running out
-of memory. Like in our preliminary testing, this increase appears exponential,
-so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
-spent loading the generated theory (containing all distinct datatypes and
-functions from TIP); this overhead is unfortunate, but since it is constant
-across all sample sizes it does not affect our conclusions.
-
-With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
-precision, flat recall), although the error margins are too wide to be
-definitive.
-
-\subsection{Comparison}
-
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
-of the Speedup-Test protocol, where each of our sample sizes is a separate
-``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
-which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
-signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
-method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
-out.
-
-For each sample size (``benchmark'') the Speedup-Test protocol compares the
-distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
-significantly different (with $\alpha < 0.05$), then the signed-rank test is
-only performed if more than 30 samples are available. This was the case for all
-sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
-two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
-putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup (median IsaCoSy time
-divided by median QuickSpec time) is given in table~\ref{table:speedups}. We
-would predict the speedup to grow for larger sample sizes, but the increasing
-proportion of timeouts causes our estimates to become more conservative: by size
-20 most runs are timing out, and hence are tied.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l| }
-    \hline
-    \bfseries Sample Size & \bfseries Speedup
-    \csvreader[]{speedups.csv}{}
-    {\\\hline\csvcoli&\csvcolii} \\
-    \hline
-  \end{tabular}
-  \caption{Speedup from using QuickSpec compared to IsaCoSy, i.e. the median
-    time taken by IsaCoSy divided by that of QuickSpec. QuickSpec was found to
-    be significantly faster ($\alpha = 0.05$) for all sizes, except 1 and 2 due
-    to too few samples. These are conservative estimates, since the 300 second
-    time limit acts to reduce the measured time difference (e.g. sizes 13 and
-    20, where the median for both tools timed out).}
-  \label{table:speedups}
-\end{table}
-
-We compare the recall proportions by applying McNemar's test to only those
-samples where both tools succeeded. We pooled these together from all sample
-sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p-value
-of 0.0026.
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
-    \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
-    \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
-    \hline
-  \end{tabular}
-  \caption{Contingency table for ground truth properties, pooled from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
-  \label{table:recall}
-\end{table}
-
-McNemar's test is not applicable for comparing precision, since each tool
-generated different sets of conjectures. Instead, we add up the number of
-results which are ``interesting'' (appear in the ground truth) and those which
-are ``uninteresting'' (do not appear in the ground truth), with the results
-shown in table~\ref{table:precision}
-
-\begin{table}
-  \centering
-  \begin{tabular}{ |r|l|l| }
-    \hline
-              & \bfseries Interesting & \bfseries Uninteresting \\
-    \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
-    \hline
-    \bfseries QuickSpec & 47          & 301      \\
-    \hline
-  \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, pooled from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
-  \label{table:precision}
-\end{table}
-
-We tested these totals for independence using Boschloo's test and find a p-value
-of 0.111, which exceeds our (conventional) significance threshold of
-$\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
-
-Note that precision is the only comparison where IsaCoSy scored higher, and even
-that was not found to be significant. Also, precision on its own is not the
-whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
-given that its recall is significantly lower.
-
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
-samples finished successfully. If we include failed runs in our tables, treating
-them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
-
-\section{Discussion}
-\label{sec:discussion}
-
-Fundamentally, when designing any mathematical reasoning system we must decide
-on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
-such as ``true'' or ``provable'' include trivial tautologies, while at the same
-time failing to capture the ``almost true'', which can be a valuable trigger for
-theory change, as demonstrated by Lakatos in his case studies of mathematical
-development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
-proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
-light on this by testing whether mathematicians' experiences of abstract beauty
-correlates with the same brain activity as experiences of sensory
-beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
-Colton (such as those in~\cite{colton2000notion}) are based largely on
-``intuition'', plausibility arguments about why a metric would be important, and
-previous use of such metrics (in the case of ``surprisingness'', from a single
-quote from a mathematician). Opinions from mathematicians themselves include
-Gowers' suggestion that we can identify features which are commonly associated
-with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
-Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
-interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
-developer).
-
-All of these approaches rest upon the assumption that it makes sense to speak of
-``the good'' in mathematics. However, empirical psychological studies
-call into question such assumptions: for example, work by Inglis and colleagues
-has shown that there is not even a single standard of \emph{validity} among
-contemporary mathematicians~\cite{inglis2013mathematicians}.
-
-Whilst these difficulties are real and important, we cannot ignore the fact that
-mathematics is nevertheless being practised around the world; and similarly that
-researchers have forged ahead to develop a variety of tools for automated
-exploratory mathematics. If we wish to see these tools head in a useful,
-fruitful direction then \emph{some} method is needed to compare their
-approximate ``quality'' in a concrete, measurable way.
-
-The key to our benchmarking methodology is to side-step much of this
-philosophical quagmire using the simplifying assumption that theorem proving
-problem sets are a good proxy for desirable input/output behaviour of MTE tools.
-As an extension of existing precision/recall analysis, this should hopefully not
-prove too controversial, although we acknowledge that there are compelling
-reasons to refute it.
-
-We do not claim that our use of corpora as a ground truth exactly captures all
-interesting conjectures of their definitions, or that those definitions exactly
-represent all theories we may wish to explore. Rather, we consider our approach
-to offer a pareto-optimal balance between theoretical rigour and experimental
-practicality, at least in the short term. Futhermore, since research is already
-on-going in these areas, we hope to at least improve on existing evaluation
-practices and offer a common ground for future endeavours.
-
-One notable weakness is that our methodology does not allow negative examples,
-such as particularly dull properties that tools should never produce. Everything
-outside the ground truth set is treated equally, whether it's genuinely
-uninteresting or was only left out due to oversight. In particular this limits
-the use of our approach to domains where existing human knowledge surpasses that
-discovered by the machine. Any \emph{truly novel} insights discovered during
-testing will not, by definition, appear in any existing corpus, and we would in
-fact \emph{penalise} tools for such output. We do not believe this to be a
-realistic problem for the time being, as long as evaluation is limited to
-well-studied domains and results can be generalised to real areas of
-application. This does emphasise the continued importance of testing these tools
-with real human users, rather than relying solely on artificial benchmarks.
-
-Another practical limitation of our benchmarking approach is that it only
-applies to tools which act in ``batch mode'', i.e. those which choose to halt
-after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
-taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
-such long-running, potentially interactive programs, but that is beyond the
-scope of this project.
-
-\section{Related Work}
-\label{sec:related-work}
-
-The automation of mathematical tasks has been pursued since at least the time of
-mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
-these efforts is the separation between those undertaken by mathematicians like
-Pascal and Babbage~\cite{bowden}, and those of engineers such as
-M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
-are concerned with (automatically constructing and evaluating concepts,
+Automated logical and mathematical reasoning has been pursued since at least the
+time of mechanical calculators like the Pascaline~\cite{d'ocagne}. A
+recurring theme in these efforts is the separation between those undertaken by
+mathematicians like Pascal and Babbage~\cite{bowden}, and those of engineers
+such as M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the
+tasks we are concerned with (automatically constructing and evaluating concepts,
 conjectures, theorems, axioms, examples, etc.) being divided into two main
@@
 
+\subsection{Theory Formation}
+
 An important historical implementation of ATF is Lenat's AM (Automated
@@
 its knowledge base via a collection of heuristic rules, and controls the firing
-of these rules via an agenda mechanism. Evaluation of AM considered generality
-(performance in new domains) and how finely-tuned various aspects of the program
-are (the agenda, the interaction of the heuristics, etc). Most of this
-evaluation was qualitative, and has subsequently been
-criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
-Ritchie and Hanna found a large discrepancy between the theoretical claims made
-of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
-natural numbers from sets, but did so using a heuristic specifically designed to
-make this connection.
+of these rules via an agenda mechanism.
+
+The performance of AM was evaluated based on its generality (performance in new
+domains) and how finely-tuned various aspects of the program are (the agenda,
+the interaction of the heuristics, etc). Most of this evaluation was
+qualitative, and has subsequently been criticised~\cite[chap.~13]{colton:book}.
+In their case study in methodology, Ritchie and Hanna found a large discrepancy
+between the theoretical claims made of AM and the implemented
+program~\cite{ritchie1984case}; for example, AM ``invented'' natural numbers
+from sets, but did so using a heuristic specifically designed to make this
+connection.
+
+The successor of AM is Eurisko, a discovery system intended to be useful in more
+general domains than just mathematical theory formation. Despite claiming some
+early successes % FIXME Traveller TCS
+work on Eurisko was put on hold due to a lack of machine-readable data about
+real world systems for it to work with % FIXME Find something quotable
+. This lead to the Cyc project, in an attempt to encode such knowledge in a
+semantically rich and consistent form. Cyc is an ongoing project, with parts of
+its database made available via OpenCyc.
+
+\subsection{Theory Exploration}
 
@@
 such as ours.
-
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We propose a general benchmarking methodology for measuring, summarising and
-comparing performance of diverse approaches to the conjecture generation
-problem, whilst avoiding some of the philosophical complications and ambiguities
-of the field. This methodology can be tailored for specific interests, such as
-the choice of problem set and the focus of the resulting analysis.
-
-We have also presented an example application for the domain of higher-order
-inductive theories (which is immediately applicable to problems in functional
-programming). Using the TIP problem set as a corpus, we have evaluated the
-performance of the QuickSpec and IsaCoSy tools and demonstrated QuickSpec to be
-both significantly faster and to also output more desirable conjectures than
-IsaCoSy; although more \emph{undesirable} output may be generated as well. We
-found that they both fail, due to time and memory constraints, for a large
-proportion of inputs; that this gets worse as input size increases; and that
-IsaCoSy fails more often than QuickSpec. Based on these results we proposed two
-possible directions to improve the QuickSpec algorithm: a post-processing filter
-to remove more ``uninteresting'' conjectures and a pre-processing filter to
-``focus'' on promising subsets of the input.
-
-Other promising directions for future work include the evaluation of other MTE
-tools, the use of other corpora more suited to different problem domains, and
-the extension of existing corpora with new definitions and properties (which
-would also be of benefit to the original ATP benchmarks).
-
-We believe that a standard approach to benchmarking and comparison such as ours
-will ease the burden on researchers wanting to evaluate different potential
-approaches to this task, and provide a common goal to pursue in the short term.
-
-``Solving'' this benchmark would not solve the problem of conjecture generation
-in general, so more ambitious goals must be set in the future. For now, we
-believe that our approach provides a compelling challenge and will encourage
-healthy competition to improve the field.
-
-\bibliographystyle{plain}
-\bibliography{./Bibtex}
-
-\end{document}

### 2018-09-12 PhDThesis/litreview_machinelearning.tex
Thesis changes (e.g. lit review)

\subsection{Machine Learning Applications in Mathematics, Logic and Programming Language Theory}

Machine learning techniques, used for finding statistical patterns and numerical
correlations in arbitrary data, are characterised by their statistical pattern-finding

AutoTuning

DeepTune analyses raw program text to predict the best options for an optimising
compiler. FIXME

Fuzzing AFL DeepSmith

### 2018-09-12 PhDThesis/litreview_staticanalysis.tex
Thesis changes (e.g. lit review)

@@
 Software for analysing and improving code can be roughly divided into two
-categories. Those which contribute to the running of a program, such as
-(optimising) compilers, linkers and interpreters must produce correct,
-predictable, desirable results in a timely manner; for example, compiler
-optimisations should not alter program semantics, should provide improvements in
-the majority of cases (i.e. they should not be \emph{pessimisations}) and under
-which they apply should be predictable enough to prevent spurious performance
-regressions when making seemingly innocuous refactorings. Optimisations in
-ahead-of-time (AOT) compilers must be fast enough to run during every build, and
-should have a large enough impact on the generated code's performance that they
+categories. Some contribute to the running of a program, such as (optimising)
+compilers, linkers and interpreters. These must produce correct, predictable,
+desirable results in a timely manner. For example, compiler optimisations should
+not alter program semantics, should provide monotonic improvements in the
+majority of cases (i.e. they should not be \emph{pessimisations}) and conditions
+under which they apply should be predictable enough to prevent spurious
+performance regressions when making seemingly innocuous
+refactorings~\cite{robison2001impact}. Optimisations in ahead-of-time (AOT)
+compilers must be fast enough to run during every build, and should have a large
+enough impact on the generated code's performance that they
 ``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
@@
 to software libraries fits into this latter category, whose constraints are not
-as restricting as those applying to tools like compilers. The only real
+as stringent as those applying to tools like compilers. The only real
 restrictions on our tools are that their resource requirements are acceptable to
@@
 Linters are characterised by parsing the given code and looking for occurrences
-of certain syntax patterns, for example definitions which are never used or
+of certain syntax patterns, for example definitions which are never used, or
 ``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
@@
 
-A popular linter for Haskell code is \texttt{HLint}~\cite{mitchell2014hlint},
+A popular linter for Haskell code is \textsc{HLint}~\cite{mitchell2014hlint},
 which focuses mostly on simplifying code; for example by spotting use-cases for
-common library functions, e.g. replacing a pattern match \texttt{case A of \{
-  Nothing -> B; Just C -> D C; \}} with the function call \texttt{maybe B D A}.
+common library functions like \texttt{null x} instead of \texttt{length x == 0}.
 Linting can be seen as the inverse of theory exploration: rather than inferring
-general relationships from repeated patterns in the code, linting begins with
-known relationships and looks through the code for specific instances. Whilst
-linting is a useful programming aid, and fast enough to apply regularly (e.g. as
-a background process in a code editor), it is not an open-ended process: once
-the predetermined patterns are exhausted, no more improvements will be found.
+novel, general relationships from patterns found in the code, linting begins
+with known relationships and looks through the code for specific instances.
+Whilst linting is a useful programming aid, and fast enough to apply regularly
+(e.g. as a background process in a code editor), it is not an open-ended
+process: once the predetermined patterns are exhausted, no more improvements
+will be found.
 
@@
 the other end we find \emph{superoptimisers}~\cite{massalin1987superoptimizer}.
-These treat the target program as a specification, and perform an open-ended
-search through the space of all programs (in some specific language, usually a
-subset of machine code which is free of branches and side-effects) to find all
-possible implementations. Superoptimisers need some way to check program
-equivalence, such as a test suite (to quickly eliminate incorrect programs) and
-a theorem prover. Proving equivalence of machine code programs is helped by the
-uniform, finite nature of their domain (e.g. a fixed set of 32bit registers).
-Programs in higher level languages, like Haskell, may have unlimited input
-domains (say, the natural numbers), which makes effective testing more difficult
-and proofs potentially undecidable. A decidable method of comparison is also
-required, to choose which of two equivalent programs is ``better''. This is
-typically a cost model for the language, with a trivial example being program
-length (i.e. assuming all instructions have the same cost).
+These treat the given code as a specification, and perform an open-ended search
+through the space of all programs (in some specific language, usually a subset
+of machine code which is free of branches and side-effects) to find the optimal
+implementation according to some quality measure or cost model, such as
+instruction count. Superoptimisers need some way to check program equivalence,
+such as a test suite (to quickly eliminate incorrect programs) and/or a theorem
+prover. Proving equivalence of machine code programs is helped by the uniform,
+finite nature of their domain (e.g. a fixed set of 32bit registers). Programs in
+higher level languages, like Haskell, may have unlimited input domains (say, the
+natural numbers), which makes effective testing more difficult and proofs
+potentially undecidable. Measuring and comparing efficiency of implementations
+is also more complicated in such languages, potentially requiring some form of
+cost semantics to derive an algebraic cost
+formulation~\cite{danner2015denotational} and empirical testing to determine the
+constants.
 

### 2018-09-12 PhDThesis/litreview_testing.tex
Thesis changes (e.g. lit review)

\subsection{Property Checking}

Automated testing is a common method of software quality assurance. In Haskell,
some of the most popular testing frameworks go beyond the industry standard of
\emph{unit testing} (asserting that components of a system behave as expected on
a fixed set of inputs) to instead perform \emph{property checking}. A
``property'', in this sense, is a statement about a program's components, which
may include free variables. Such properties are typically encoded in Haskell as
functions, accepting values for the free variables as arguments, and returning a
boolean value (perhaps adorned with useful metadata). A property checker
proceeds by
\begin{enumerate*}
\item generating suitable values for the function's arguments,
\item calling the function with these values,
\item inspecting the return value and
\item repeating until some stop condition is satisfied
\end{enumerate*}.

Most frameworks will test each property a fixed number of times (e.g. 100), and
stop early if the function returns a false value. The major difference between
implementations is the way they instantiate free variables (or, in encoded form,
how they generate arguments for the property functions).

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \texttt{forAll}), or by
choosing a default generator for types implementing its \texttt{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way).

Random testing is effective at thoroughly exercising the branches of a program,
since inputs may be quite large and cover many cases, especially for functions
which recurse on each \emph{sub-}expressions of their input as well. The output
of a random generator will also vary across runs, \emph{increasing} our
confidence each time we check a property, rather than simply reiterating past
results. However, there are two common problems with random property checking.
Firstly we must be able to generate random values, which can be tricky for
recursive datatypes. Consider a type of binary trees, with constructors
\texttt{Leaf~::~Tree} and \texttt{Node~::~Tree~->~Tree~->~Tree}. We might
generate a \texttt{Tree} by picking uniformly between these two constructors
and, if a \texttt{Node} was picked, recursing to generate the two
sub\texttt{Tree}s. Yet this simple scheme will generate \texttt{Tree}s whose
expected size is \emph{unbounded}! This is due to the 50\% chance of generating
a \texttt{Node}, which requires two recursive calls; hence we can expect each
call to recurse $0.5~\times~2~=~1$ more time on average, and that call will also
recurse once more on average, and so on; this can easily exhaust all available
memory. Many automated data generators, such as those from Haskell's
\texttt{derive} package, suffer from this problem. It can be avoided by biasing
the generators (in the \texttt{Tree} example we would produce a \texttt{Leaf}
more than half of the time), but this bias compounds at each level of recursion,
limiting testing to very small values. More sophisticated solutions exist, such
as the \texttt{boltzmann-sampling} package (see \cite{duchon2004boltzmann}) and
even domain-specific languages~\cite{lampropoulos2017beginner}, but all of these
add complications to the property checking process.

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \texttt{shrink} function along with their
generator, which is given any counterexample found during testing and returns a
list of smaller values. In the above example we might shrink a \texttt{Node} by
returning each sub-tree, and return an empty list when asked to shrink a
\texttt{Leaf}. After a counterexample has been found, the same property is
applied to each of the shrunk values in an attempt to narrow down the source of
the problem. When a value cannot be shrunk any more, or if none of its shrunken
versions are counterexamples, that value is presented to the user in the hope
that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumerating function or using its
\texttt{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package). In the \texttt{Tree} example this might
check \texttt{Leaf}, then \texttt{Node Leaf Leaf}, then
\texttt{Node (Node Leaf Leaf) Leaf}, and so on. This ensures that
counterexamples are always small, that tests are not repeated with the same
values and that memory is not exhausted (since the stopping condition will
trigger long before the enumeration reaches such large values). The disadvantage
is that large inputs (which exercise properties, especially recursive ones, the
most) may occur so late in the enumeration that testing never reaches them.

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \texttt{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which checks the
same property on refined versions of the previous value. For example, a lazily
generated \texttt{Tree} may initially be equivalent to \texttt{undefined}. If
the property aborts, it will be run again with inputs equivalent to
\texttt{Leaf}
and \texttt{Node undefined undefined}; if the latter aborts due to inpuecting
the left sub-tree, then \texttt{Node Leaf undefined} and
\texttt{Node (Node undefined undefined) undefined} will be tried (and likewise
for the right sub-tree). This refinement continues until the property no longer
aborts or until a predefined depth limit is reached. If the property returns a
true value when given such partially-defined input then it must succeed for
\emph{all} refinements of that input (since the output was determined without
evaluating that sub-term). If the property returns a false value, the input is a
\emph{minimal} counterexample, since it only specifies those sub-terms which are
required to violate the property; any undefined subterms are irrelevant to the
failure.

\textsc{LazySmallCheck}'s automatic elaboration of free variables is similar in
operation to the use of search procedures in logic programming languages like
Kanren, Prolog (which has its own \textsc{QuickCheck} port) and Curry (a
functional logic language, whose \textsc{EasyCheck} framework explicitly
unifies the concepts of generators and logic
variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
is also reminiscent of \emph{symbolic execution}~\cite{}, a form of
\emph{abstract interpretation} where unknown runtime values are represented by
opaque symbols, computation proceeds to arrive at a result, which may be a
complicated expression involving many symbols. Importantly, we can  the in terms of the until the result is
by accumulating
combinations of these symbols. For example, we may not know the output of a function

### 2018-09-22 PhDThesis/background_machinelearning.tex
Remove reference to deleted file
Add a bit about distributed representations

@@
 
+\subsection{Distributed Representations}
+
+A \emph{distributed representation} does not isolate each semantic feature (for
+example, presence of absence of some lexeme) to separate, independent
+``computing elements'' (e.g. memory locations, such as the elements of a feature
+vector). Instead, features are represented as \emph{patterns} across multiple
+elements: each feature uses many elements and each element contributes to many
+features~\cite{hinton1984distributed}. Such encodings provide simple mechanisms
+to represent \emph{combinations} of features (combining their patterns
+element-wise), \emph{partial} features (presence of the pattern across a subset
+of its elements) and \emph{generalisation} (patterns which are present as
+sub-patterns of multiple features).
+
+% Distributed representations may be learned from data, for example by
+% \emph{auto-encoding}; or they may be calculated ``from whole cloth'' using
+% techniques like hashing. We are particularly interested in distributed
+% representations for storing structured data (like syntax trees) in fixed-length
+% feature vectors.
+
 \input{background_clustering.tex}
-\input{background_featureextraction.tex}

### 2018-10-15 PhDThesis/litreview_staticanalysis.tex
Content

@@
 as stringent as those applying to tools like compilers. The only real
-restrictions on our tools are that their resource requirements are acceptable to
-users, and that output is sufficiently high quality to be worth waiting for and
-sifting through.
+restrictions on these tools are that their resource requirements are acceptable
+to users, and that output is sufficiently high quality to be worth waiting for
+and sifting through.
 
@@
 
-If fast but limited linters live at one end of a code improvement spectrum, at
-the other end we find \emph{superoptimisers}~\cite{massalin1987superoptimizer}.
-These treat the given code as a specification, and perform an open-ended search
-through the space of all programs (in some specific language, usually a subset
-of machine code which is free of branches and side-effects) to find the optimal
-implementation according to some quality measure or cost model, such as
-instruction count. Superoptimisers need some way to check program equivalence,
-such as a test suite (to quickly eliminate incorrect programs) and/or a theorem
-prover. Proving equivalence of machine code programs is helped by the uniform,
-finite nature of their domain (e.g. a fixed set of 32bit registers). Programs in
-higher level languages, like Haskell, may have unlimited input domains (say, the
-natural numbers), which makes effective testing more difficult and proofs
-potentially undecidable. Measuring and comparing efficiency of implementations
-is also more complicated in such languages, potentially requiring some form of
-cost semantics to derive an algebraic cost
-formulation~\cite{danner2015denotational} and empirical testing to determine the
-constants.
+At the other end of the code improvement spectrum we find
+\emph{superoptimisers}~\cite{}, which can take an
+arbitrarily long time to find improvements, but whose deductive power is only
+limited by the undecidability of the source language. Superoptimisers, like ,
+treat the given code as a specification, and perform an open-ended search
+through the space of all programs to find the optimal implementation according
+to some quality measure or cost model (often a branch- and effect-free subset of
+machine code is used, and measured by counting instructions or
+cycles~\cite{massalin1987superoptimizer}). Superoptimisers need some method to
+check whether a synthesised program implements the required specification, such
+as a test suite and/or a theorem prover.
 
-The biggest hindrance to superoptimisation is how long it takes to search such a
+Proving equivalence of machine code programs is helped by the uniform, finite
+nature of their domain (e.g. a fixed set of instructions, operating on a fixed
+set of 32bit registers). Higher level languages, like Haskell, are more
+difficult to reason about and measure, since their behaviour and efficiency may
+depend on unknown runtime parameters. For example, a higher-order function may
+choose to memoise its argument, which avoids repeated function calls but uses
+more memory and comparison checks; whether this is an optimisation or
+pessimisation depends on the runtime data at each call site, including the
+function argument (whether it's cheaper than performing a comparison) and the
+data it will be applied to (whether it actually contains any repeated values).
+In addition to this uncertainty, each function's input domain may be infinite
+(such as the natural numbers), and programs may use abstract interfaces and
+datatypes with a variety of possible implementations (such as lists or key/value
+mappings). All these factors makes effective testing more difficult, and proofs
+potentially undecidable.
+
+We can reduce the amount of unknown parameters in a concrete program by
+propagating statically-known values, for example using
+\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
+dramatically increase code size, as each function may be duplicated into
+multiple specialised instances. Efficiency can be calculated
+using techniques like cost semantics~\cite{danner2015denotational}, with
+constant factors determined through empirical testing. This cannot solve the
+problem of optimality depending on the characteristics of runtime data, but it
+can expose these dependencies as parameters allowing the user to select
+between pareto-optimal strategies.
+
+Even if these problems are mitigated, the biggest difficulty with
+superoptimisation remains: the length of time it takes to search such a
 large space. Whilst pruning methods can help~\cite{phothilimthana2016scaling},
 the power of superoptimisation comes from its ``bottom up'' nature, which makes
-no assumptions about the structure of the generated program. Maintaining this
-generality is expensive, and current superoptimisers typically scale to programs
-only a few dozen machine instructions long. Nevertheless, this still finds use
-in compiler construction, as it is an effective method for finding ``peephole
-optimisations'' (faster replacements for small, common patterns of generated
-machine code)~\cite{Bansal.Aiken:2006}.
+no assumptions about the structure of the generated program. This generality is
+expensive, with current superoptimisers typically scaling to code fragments only
+a few dozen machine instructions long. Contemporary research has focused on the
+use of stochastic search algorithms, such as Markov Chain Monte-Carlo (MCMC), to
+synthesise promising improvements more quickly, whilst still finding optimal
+implementations in the limit~\cite{schkufza2013stochastic}; guiding such
+algorithms with \emph{learned} policies is also a promising area of
+study~\cite{mudigonda2017learning}.
+
+The open-ended search process of such superoptimisers is similar to application
+of mathematical theory exploration to software property discovery, especially
+when exploration is limited to equality properties (which is the case for
+\textsc{IsaCoSy} and \textsc{QuickSpec}). Quality metrics like efficiency (e.g.
+relative instruction count) can be considered a subset of the interestingness
+criteria of theory exploration. The \emph{aims} of these analyses are in
+opposition, since theory exploration \emph{builds up} structure and abstraction
+latent in a set of definitions, whilst superoptimisation \emph{tears down} such
+layers to find minimal implementations; yet the practical necessity of
+superoptimising only small snippets at a time (e.g. to find ``peephole
+optimisations''~\cite{Bansal.Aiken:2006}) results in remarkably similar
+discovery processes.

### 2018-10-15 PhDThesis/background_haskell.tex
More content

@@
 corpus of existing code (compared to more formal languages like Coq or
-Isabelle).
+Isabelle). Haskell is well-suited to programming language research; indeed, this
+was a goal of the language's creators \cite{marlow2010haskell}. Like most
+\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
+with extra features such as a strong type system and ``syntactic sugar'' to
+improve readability.
 
@@
 
-%% FIXME: MERGE THE FOLLOWING WITH THE ABOVE
-
 \newcommand{\CVar}{\texttt{Var}}
@@
 
-We decided to focus on theory exploration in the Haskell programming language as
-it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and
-\hspec{} \cite{claessen2013automating}). This is evident from the fact that the
-state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster}
-\cite{Hipster} system, is actually implemented by translating to Haskell and
-invoking \hspec{}.
+Further evidence of Haskell's suitability for theory exploration is given by the
+fact that the state-of-the-art implementation for Isabelle/HOL, the
+\textsc{Hipster}~\cite{Hipster} system, is actually implemented by translating
+to Haskell and invoking \hspec{}~\cite{claessen2013automating}.
 
-Haskell is well-suited to programming language research; indeed, this was a goal
-of the language's creators \cite{marlow2010haskell}. Like most
-\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
-with extra features such as a strong type system and ``syntactic sugar'' to
-improve readability. For simplicity, we will focus on an intermediate
-representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather
-than the relatively large and complex syntax of Haskell proper. Core is based on
-\fc{}, but for our machine learning purposes we are mostly interested in its
-syntax; for a more thorough treatment of \fc{} and its use in GHC, see
-\cite[Appendix C]{sulzmann2007system}.
+Whilst we will use normal Haskell code in this text and many of our examples,
+for simplicity we perform many of our analyses on an intermediate representation
+of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the
+relatively large and complex syntax of Haskell proper. Core is based on \fc{},
+which is described in detail in~\cite[Appendix C]{sulzmann2007system}.
 
-The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax};
-compared to the full language \footnote{As of GHC version 7.10.2, the latest at
-  the time of writing.} our major restriction is to ignore type hints (such as
-explicit casts, and differences between types/kinds/coercions). For brevity, we
-also omit several other forms of literal (machine words of various sizes,
-individual characters, etc.), as their treatment is similar to those of strings
-and numerals. We will use quoted strings to denote names and literals,
-e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"},
-\texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can
-be compared for equality.
+Core contains explicit type annotations and coercions, which we can omit as they
+have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
+  of GHC version 7.10.2, the latest at the time of writing.} is shown in
+Figure~\ref{fig:coresyntax}; for brevity, we also omit several other forms of
+literal (machine words of various sizes, individual characters, etc.), since
+their treatment is similar to those of strings and numerals. We use quoted
+strings to denote names and literals, e.g. \texttt{Local "foo"},
+\texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and
+\texttt{LitNum "42"}, and require only that they can be compared for equality.
 
@@
 
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions,
-along with a custom datatype for Peano numerals. The translation to our Core
-syntax is routine, and shown in Figure \ref{fig:coreexample}. Although the Core
-is more verbose, we can see that similar structure in the Haskell definitions
-gives rise to similar structure in the Core; for example, the definitions of
-\hs{odd} and \hs{even} are identical in both languages, except for the
-particular identifiers used. It is this close correspondence which allows us to
-analyse Core expressions in place of their more complicated Haskell source.
+Figure~\ref{fig:haskellexample} shows some simple Haskell function definitions,
+along with a custom datatype for Peano numerals. The translation to Core syntax
+is routine, and shown in Figure~\ref{fig:coreexample}. Although the Core is more
+verbose, we can see that similar structure in the Haskell definitions gives rise
+to similar structure in the Core; for example, the definitions of \hs{odd} and
+\hs{even} are identical in both languages, except for the particular identifiers
+used. It is this close correspondence which allows us to analyse Core
+expressions in place of their more complicated Haskell source.
 

### 2018-10-15 PhDThesis/background_theoryexploration.tex
More content

@@
 
-We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about
+We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
 Haskell code (these may be fed into another tool, such as \hspec{}, for

### 2018-10-15 PhDThesis/litreview_premiseselection.tex
More content

@@
 
-% TODO
-\cite{kuhlwein2012overview}
-
 One fruitful application of machine learning methods to formal mathematics is
 \emph{relevance filtering}, which is used for \emph{premise selection} in
-automated theorem provers.
+automated theorem provers~\cite{kuhlwein2012overview}.
 
@@
 
-The Multi-output ranking (\textsc{MOR}) implementation uses a support vector
-machine (SVM) to select relevant axioms from the Mizar Mathematical Library for
-use by the Vampire ATP system~\cite{alama2014premise}.
+Premise selection has also been applied to the re-proving problem in other
+systems, such as the Multi-Output Ranking (\textsc{MOR}) tool, which uses a
+support vector machine (SVM) to select relevant axioms from the Mizar
+Mathematical Library for use by the Vampire ATP system~\cite{alama2014premise}.
 %TODO: describe the kernel, as that's the interesting bit.

### 2018-10-15 PhDThesis/litreview_staticanalysis.tex
More content

@@
 directly to the code's output. These include tools for testing, profiling,
-verification, standards compliance, etc. Our application of theory exploration
+verification, standards compliance, etc. The application of theory exploration
 to software libraries fits into this latter category, whose constraints are not
@@
 of certain syntax patterns, for example definitions which are never used, or
-``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
+``no-op'' constructs (e.g. \texttt{(x~<~0)} when \texttt{x} is of
 unsigned/non-negative type, making this comparison always false).
 
-A popular linter for Haskell code is \textsc{HLint}~\cite{mitchell2014hlint},
-which focuses mostly on simplifying code; for example by spotting use-cases for
-common library functions like \texttt{null x} instead of \texttt{length x == 0}.
+A popular linter for Haskell is \textsc{HLint}~\cite{mitchell2014hlint}, which
+focuses mostly on simplifying code; for example by spotting use-cases for common
+library functions like \texttt{null x} instead of \texttt{length x == 0}.
 Linting can be seen as the inverse of theory exploration: rather than inferring
@@
 At the other end of the code improvement spectrum we find
-\emph{superoptimisers}~\cite{}, which can take an
+\emph{superoptimisers}~\cite{massalin1987superoptimizer}, which can take an
 arbitrarily long time to find improvements, but whose deductive power is only
-limited by the undecidability of the source language. Superoptimisers, like ,
-treat the given code as a specification, and perform an open-ended search
-through the space of all programs to find the optimal implementation according
-to some quality measure or cost model (often a branch- and effect-free subset of
-machine code is used, and measured by counting instructions or
-cycles~\cite{massalin1987superoptimizer}). Superoptimisers need some method to
-check whether a synthesised program implements the required specification, such
-as a test suite and/or a theorem prover.
+limited by the undecidability of the source language. Superoptimisers, such as
+\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
+Assistant'', developed to accompany \emph{Hacker's
+  Delight}~\cite{warren2013hacker}) treat the given code as a specification, and
+perform an open-ended search through the space of all programs to find the
+optimal implementation according to some quality measure or cost model (often a
+branch- and effect-free subset of machine code is used, and measured by counting
+instructions or cycles).  Superoptimisers need some method to check whether a
+synthesised program implements the required specification, such as a test suite
+and/or a theorem prover.
 
-Proving equivalence of machine code programs is helped by the uniform, finite
+Proving correctness of machine code programs is helped by the uniform, finite
 nature of their domain (e.g. a fixed set of instructions, operating on a fixed
-set of 32bit registers). Higher level languages, like Haskell, are more
+set of fixed-size registers). Higher level languages like Haskell are more
 difficult to reason about and measure, since their behaviour and efficiency may
-depend on unknown runtime parameters. For example, a higher-order function may
-choose to memoise its argument, which avoids repeated function calls but uses
-more memory and comparison checks; whether this is an optimisation or
+depend on unknown runtime parameters. For example, a higher-order function $f$
+accepting another function $g$ as an argument may choose to memoise calls to $g$
+in order to avoid evaluating identical calls multiple times; however, storing
+previous results requires more memory, and each input must be looked up in the
+memo table to see if it's been used before. Whether this is an optimisation or
 pessimisation depends on the runtime data at each call site, including the
@@
 dramatically increase code size, as each function may be duplicated into
-multiple specialised instances. Efficiency can be calculated
-using techniques like cost semantics~\cite{danner2015denotational}, with
-constant factors determined through empirical testing. This cannot solve the
-problem of optimality depending on the characteristics of runtime data, but it
-can expose these dependencies as parameters allowing the user to select
-between pareto-optimal strategies.
+multiple specialised instances. Related techniques, like \emph{symbolic
+  execution} (a form of \emph{abstract interpretation}), allow approximate
+execution of code, where unknown runtime data is represented as opaque
+metavariables. Property- and model-checkers are similar, except they work
+empirically by performing multiple runs, instantiating metavariables with
+synthetic data (\emph{QuickCheck} does this randomly, \emph{SmallCheck}
+enumerates values, etc.).
+
+Efficiency can be calculated using techniques like cost
+semantics~\cite{danner2015denotational}, with constant factors determined
+through empirical testing. This cannot solve the problem of optimality depending
+on the characteristics of runtime data, but it can expose these dependencies as
+parameters allowing the user to select between pareto-optimal strategies.
 

### 2018-10-15 PhDThesis/litreview_testing.tex
More content

@@
 same property on refined versions of the previous value. For example, a lazily
-generated \texttt{Tree} may initially be equivalent to \texttt{undefined}. If
-the property aborts, it will be run again with inputs equivalent to
-\texttt{Leaf}
-and \texttt{Node undefined undefined}; if the latter aborts due to inpuecting
-the left sub-tree, then \texttt{Node Leaf undefined} and
-\texttt{Node (Node undefined undefined) undefined} will be tried (and likewise
-for the right sub-tree). This refinement continues until the property no longer
+generated \texttt{Tree} may initially be equivalent to $\bot$. If the property
+aborts, it will be run again with each possible refinement of $\bot$: in this
+case inputs equivalent to \texttt{Leaf} and \texttt{Node~$\bot_1$~$\bot_2$}. If
+the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
+give inputs \texttt{Node~Leaf~$\bot_2$} and
+\texttt{Node~(Node~$\bot_3$~$\bot_4$)~$\bot_2$} (and likewise if $\bot_2$ had
+caused the exception). This refinement continues until the property no longer
 aborts or until a predefined depth limit is reached. If the property returns a
@@
 \emph{all} refinements of that input (since the output was determined without
-evaluating that sub-term). If the property returns a false value, the input is a
-\emph{minimal} counterexample, since it only specifies those sub-terms which are
-required to violate the property; any undefined subterms are irrelevant to the
-failure.
+evaluating the undefined sub-terms). If the property returns a false value, the
+input is a \emph{minimal} counterexample, since it only specifies those
+sub-terms which are required to violate the property; any undefined sub-terms
+are irrelevant to the failure.
 
@@
 variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
-is also reminiscent of \emph{symbolic execution}~\cite{}, a form of
-\emph{abstract interpretation} where unknown runtime values are represented by
-opaque symbols, computation proceeds to arrive at a result, which may be a
-complicated expression involving many symbols. Importantly, we can  the in terms of the until the result is
-by accumulating
-combinations of these symbols. For example, we may not know the output of a function
+is also reminiscent of \emph{symbolic execution}~\cite{king1976symbolic}, a form
+of \emph{abstract interpretation} where unknown runtime values are represented
+by opaque symbols and computation proceeds to arrive at a result, which may be a
+complicated expression involving many symbols.

### 2019-02-05 Bucketing/check.sh
Turn bucketing check.sh into Nix derivations

@@ -1,343 +1,0 @@
-#!/usr/bin/env bash
-
-ERR=0 # Indicates failure, will be our exit code
-
-# "Top-level" tests, invoked directly
-
-function checkLatex {
-    FAIL=0 # A checkLatex-specific version of ERR
-    for FUNC in checkLatexUndatedTodos checkLatexDatedTodos checkLatexRefs \
-                checkLatexCites checkLatexCiteTodos checkLatexMacros       \
-                checkLatexLabels checkLatexFigs checkLatexCitep            \
-                checkLatexDupeCites
-    do
-        for FILE in *.tex
-        do
-            "$FUNC" "$FILE" || FAIL=1
-        done
-    done
-    return "$FAIL"
-}
-
-function checkRender {
-    render > /dev/null || {
-        ERR=1
-        echo "Couldn't render PDF" >> /dev/stderr
-        return 1
-    }
-    [[ -e report.pdf ]] || {
-        ERR=1
-        echo "report.pdf wasn't created" >> /dev/stderr
-        return 1
-    }
-}
-
-function checkBibtex {
-    if [[ -z "$(render | getBibStart)" ]]
-    then
-        ERR=1
-        echo "Bibtex didn't run?" >> /dev/stderr
-        return 1
-    fi
-}
-
-function checkWarnings {
-    BIBSTART=$(render | getBibStart)
-
-    [[ -z "$BIBSTART" ]] && return 1
-
-    WARNINGS=$(render | tail "-n$BIBSTART" | grep -ai "warning")
-
-    # Citations take two passes, so double-check
-    CITES=$(echo "$WARNINGS" | grep 'Citation `' |
-                               sed -e 's/.*`\(.*\)'"'.*/\\1/g")
-    while read -r CITE
-    do
-        grep "^@[^{]*{$CITE,$" < ~/Writing/Bibtex.bib > /dev/null || {
-            ERR=1
-            echo "Citation '$CITE' missing" >> /dev/stderr
-        }
-    done < <(echo "$CITES")
-
-    # Check for non-citation warnings
-    NONCIT=$(echo "$WARNINGS" | grep -v 'Citation `'                      |
-                                grep -v "There were undefined citations." |
-                                grep -v "Citation(s) may have changed."   |
[…]

### 2019-02-06 Bucketing/background.tex
Add dates to TODOs

@@
 
-\iffalse TODO: Make this more formal?
+\iffalse % TODO{2019-02-07}: Make this more formal?
  V \in Var
@@
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
+  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$
+% TODO{2019-02-07}: A little awkward; maybe use the above notation?
 To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
@@
 \iffalse
-TODO
+TODO{2019-02-07}
 In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.

### 2019-02-08 TEBenchmarkPaper/article.tex
Embed CSV data in .tex file
Embed all files except figures in latex
Try to address sampling points

@@
 \end{filecontents*}
-%
+
+%INCLUDE STYLES HERE
+
+%INCLUDE SPEEDUP CSV HERE
+
+%INCLUDE BIBTEX HERE
+
 \RequirePackage{fix-cm}
@@
 definitions.\footnote{This could be done randomly, but for reproducibility we
-  use a deterministic order based on cryptographic hashes.} Given a sample size,
-we choose a subset of that many definitions, and provide only those as input to
-the tool. We generate a corresponding ground truth by selecting those properties
+  use a deterministic order based on cryptographic hashes.} The sample size is a
+free parameter so that users have some control over how long the runs will take
+($\S$\ref{sec:application} shows that larger samples take longer, but the
+correlation is very noisy). Given a sample size, we need to choose a subset of
+that many definitions from the corpus, and provide only those as input to the
+tool.
+
+We cannot expect an MTE tool to produce all of the properties found in the
+corpus, if it is only given a subset of definitions as input. Hence for each
+sample we generate a corresponding ground truth by selecting those properties
 from the corpus which ``depend on'' (contain references to) \emph{only} the
@@
 
-Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
-given sample size, increasing the size of the corpus (which provides better
-statistics) makes it less likely that a chosen sample will contain all of
-a property's dependencies. The majority of such samples would hence have an
-empty ground truth, and thus $0$ precision and undefined recall
-\emph{independent} of the tool's output! This is clearly undesirable as an
-evaluation method.
-
-Other measurements could be used for such cases, but our approach is to avoid
-them by only allowing a sample if it contains all dependencies of at least one
-property. We could do this using rejection sampling, but it is more efficient to
-pick a property, weighted in proportion to their number of dependencies
-(ignoring those with more dependencies than our sample size). That property's
-dependencies become our sample, padded up to the required size with uniform
-choices from the remaining definitions.
-
-The ground truth for such samples is guaranteed to contain at least one property
-(the one we picked), and hence the precision and recall will depend meaningfully
-on the tool's output. One downside of this restriction is that we limit the
-number of possible samples to measure; this is significant for sample sizes less
-than~3.
+The question remains of how to pick definitions for each sample. One simple way
+is to choose uniformly without replacement, but this will give many samples
+whose corresponding ground truths are empty. The effect is similar to a lottery,
+where our sample is a ticket and must contain all of some property's
+dependencies in order to ``win'' (i.e. contain that property in its ground
+truth). In particular, if we choose a sample of size $S$ uniformly at random
+from a corpus containing $N$ definitions and $P$ properties which depend on an
+average of $D$ definitions each (assuming no properties have the exact same
+dependencies), then the probability that our sample contains the dependencies of
+some particular property $p$ follows a hypergeometric distribution (similar to
+the binomial distribution, but without replacement):
+
+\begin{equation*}
+  Pr(p) = \frac{\binom{S}{D} \binom{N - S}{S - D}}
+               {\binom{N}{S}}
+\end{equation*}
+
+Choosing at least one property's dependencies is a Bernoulli process with $P$
+repetitions, with probability $1 - (1 - Pr(p))^P$. This gets larger if we
+increase the sample size $S$ or the number of properties $P$, but it gets
+smaller if we increase the number of definitions $N$ or the average number of
+dependencies $D$. This is problematic because we would like large corpora for
+better statistics, but this would make empty ground truths more likely.
+
+Samples whose corresponding ground truth is empty are problematic because they
+have $0$ precision and undefined recall (due to division by $0$). Such samples
+are hence unsuitable for our benchmark, especially since these results would be
+\emph{independent} of the tool's output (all tools would fare the same, so no
+comparisons could be made). Other measurements could be used for such cases, but
+since our choice of analysis comes from existing work we would rather avoid such
+samples by only allowing those with a non-empty ground truth (i.e. containing
+all dependencies of at least one corpus property).
+
+We can achieve this by instead performing uniform sampling \emph{with
+  rejection}, where samples with empty ground truth are rejected and a different
+sample is drawn instead.
+
+One problem with performing rejection sampling is that it becomes very
+inefficient when the probability of rejection is high. In our case there is a
+more direct method of sampling, whose output is exactly equivalent to rejection
+sampling but is far more efficient.
+
+Since non-rejected samples must contain the dependencies of some property, we
+can consider each such sample to be the union of those dependencies and some
+other ``padding'' definitions. By choosing these two components separately and
+returning their union, we can obtain an identical distribution without having to
+perform any rejection.
+
+We first choose the set of dependencies by picking from those of all corpus
+properties, weighted according to the probability that they would appear under
+uniform sampling. To do this we first discard any dependency sets which are
+larger than the given sample size (since they cannot appear as subsets); we then
+discard any dependency set which is a superset of another (since only the subset
+is required to avoid rejection). Their probabilities under uniform sampling
+depend only on their size, so we assign weights by calculating the least
+common multiple (LCM) of their sizes, and dividing that by each set's size. For
+example, given dependency sets $X = \{a, b, c\}$, $Y = \{b, d\}$ and
+$Z = \{a, b, e, f\}$ the LCM of their lengths ($3$, $2$ and $4$) is 12, so the
+weighting will be:
+
+\begin{equation*}
+  w(X) = \frac{12}{3} = 4
+  w(Y) = \frac{12}{2} = 6
+  w(Z) = \frac{12}{4} = 3
+\end{equation*}
+
+Once we've chosen a dependency set, we know how much padding to choose (sample
+size minus dependency set size). The padding will be uniformly distributed,
+since it does not affect the rejection criterion.
+
+Using rejection sampling or, equivalently, our direct method, ensures that the
+analysis is well defined and depends meaningfully on the tool's output. One
+downside of applying this non-empty ground truth restriction is that we limit
+the number of possible samples to measure; this is significant for sample sizes
+less than~3.
 

### 2019-02-28 Bucketing/render.sh
Metadata

@@ -1,4 +1,0 @@
-#! /usr/bin/env nix-shell
-#! nix-shell -i bash
-make clean
-make all

### 2019-02-28 UnderstandabilityPaper/machinelearning.tex
Add bare bones of understandability paper

I meant to send this earlier, in case it's useful. These are just some
things that came to mind regarding "explainable" AI and machine
learning. Lots of rambling, but hopefully some insight into current
machine-learning thinking (as I understand it, as somewhat of an
outsider!).

I think a key idea is "differentiable programming", which captures the
essence of how machine learning models like neural networks work, and
precisely what makes them different from traditional computer programs
(i.e. they're differentiable!), without ignoring the details ("black
box") or using flimsy analogies ("brain-like"):

https://fluxml.ai/2019/02/07/what-is-differentiable-programming.html

When run on an input, a differentiable program doesn't just give an
output, it also gives us the partial derivatives of that output
w.r.t. each of the program's parameters. This tells us exactly how to
adjust those parameters (via backpropagation) to reduce the output error
for that particular input. Doing this enough times for enough inputs
will hopefully find a general solution.

A neural network is an example of a differentiable program, one which is
just a big collection of identical expressions ("neurons") composed
together. This is incredibly general (they are universal function
approximators), but it doesn't have much "knowledge" or "structure"
baked in: the behaviour of this program depends almost completely on the
parameters, so everything it "learns" must come from input/output
examples.

The convolutional neural networks that have come to dominate image
processing are also differentiable programs, but they contain more
knowledge/structure. Instead of general-purpose 'neurons', these use
'convolutional layers' which are also expressions that get composed
together. The difference is that these expressions hard-code the 2D grid
structure that we expect to find in raster images. This makes them less
general than neural networks, since they assume the data fits this 2D
grid structure, but they also don't have to learn that structure from
examples. Hence they need fewer examples, but we can also be more
confident that the input/output relationships they learn involve 2D
patches of image content (like shapes or textures), rather than some
arbitrary numerical relationship between pixels.

Taking things much further, there are differentiable programs which do
ray tracing ( https://people.csail.mit.edu/tzumao/diffrt/ ). These can
be used to learn the position of 3D objects in an image, the camera
position, illumination, etc. Again, this is less general since it
assumes the input is an image of a 3D scene, but again it doesn't have
to learn this knowledge from examples. The learned parameters are easily
"explained", as they are the usual rendering parameters of camera
position, rotation, etc.

Another example is learning to recognise and generate hand-written
characters by using a differentiable program for drawing lines:
https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf‎

In this way, there are two tradeoffs being made:

 - By hard-coding our knowledge of the domain into a program, like using
   a ray-tracer for images of the 3D world, or pen-based drawing for
   images of hand-written characters, we don't need as much training
   data for the system to 'learn' what's going on. Unfortunately, we end
   up with less general systems, e.g. we might apply the same
   neural-network architecture to all sorts of problems (with enough
   data), but we can't apply the ray-tracer to e.g. audio data.

 - This same tradeoff also has implications for understandability and
   explainability: the parameters of a ray-tracer are directly
   understandable, based on how they're used in the program, e.g. "how
   bright is this light source" or "how big is this object". In
   contrast, the parameters of something like a neural network make
   little sense on their own, they just mean "how important is this
   output for that neuron?".

STUFF BELOW IS MORE TANGENTIAL

There are probably parallels to be drawn w.r.t. feature engineering
too. "Feature extraction" turns raw data into numbers for the learning
algorithm. This used to be done by hand, so we knew what the numbers
meant (e.g. average brightness, number of edges, etc. this is what
Katya's 'recurrent clustering' algorithm does for symbolic expressions).
With "deep nets", such features are figured out automatically, and we
don't know what they mean. Hence we gain generality (automatically
discovered features tend to out-perform hand-engineered ones), but we
lose understanding/explainability, and we become more dependent on
training data.

Things like "deep dream" can try to demonstrate what these features
seem to represent, but not much more. As far as I understand it, deep
dream works by picking a particular neuron/feature, providing an
arbitrary input image to the system, and iteratively tweaking that image
so that feature increases. The "explanation" of that feature is the
changes that were made to the image, e.g. if the changes look
eye-shaped, then we say that feature is for detecting eyes. This is, of
course, open to interpretation.

Interestingly, exactly the same technique is used for generating
adversarial inputs, where an input (usually to a classifier) is tweaked
until it gets classified incorrectly:

https://en.wikipedia.org/wiki/Adversarial_machine_learning

Adversarial inputs show how limited this sort of "explanation" is, since
these changes might look like random noise:
https://blog.openai.com/adversarial-example-research/

Or might even just be a single pixel:
https://arxiv.org/abs/1710.08864

Another nice example of "explaining" a machine learning model by
demonstrating what its features represent is the variational
autoencoder.

Normal autoencoders 'compress' inputs down to a sequence of numbers
representing their important features. Unfortunately these are hard to
understand in isolation (sometimes called "holographic reduced
representations", all of the components depend on all of the others).

Variational autoencoders compress down to a multi-dimensional gaussian
distribution instead. By randomly sampling from this gaussian, the
representation is forced to be robust against small changes, and this
imposes a meaningful structure on the space; such that adjusting the
centre of the gaussian will alter the resulting data, e.g. by making a
person's hair darker; or changing their gender; or giving them glasses;
etc.

https://towardsdatascience.com/intuitively-understanding-variational-autoencoders-1bfe67eb5daf
(Especially the "Vector arithmetic" section)

Although this property can't actually be relied upon!
http://paulrubenstein.co.uk/variational-autoencoders-are-not-autoencoders/

A similar effect to the structured "latent space" of the VAE has been
found with word2vec:
https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships

word2vec turns words into vectors of numbers, based on co-occurrences of
words in some training data. These vectors turn out to have some
"meaning" in the sense that vector arithmetic can make semantic changes.
For example if we do the sum:

  vector("king") - vector("man") + vector("woman")

The resulting vector is very close to vector("queen").

These sorts of machine learning system are unsupervised, so they don't
solve any particular problem. However, the representations they learn
can be useful for other task-specific systems.

### 2019-02-28 UnderstandabilityPaper/machinelearning.tex
Add default.nix to understandability

@@
+% STRUCTURE:
+
+%  - Understandability in AI
+%  - Theorem proving as example:
+%   - Least informative
+
+\documentclass{article}
+\begin{document}
 I meant to send this earlier, in case it's useful. These are just some
@@
 characters by using a differentiable program for drawing lines:
-https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf‎
+https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf
 
@@
 
-https://en.wikipedia.org/wiki/Adversarial_machine_learning
+% https://en.wikipedia.org/wiki/Adversarial_machine_learning
 
@@
 found with word2vec:
-https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships
+
+% https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships
 
@@
 can be useful for other task-specific systems.
+
+\bibliographystyle{plain}
+\bibliography{./Bibtex}
+
+\end{document}

### 2019-02-28 UnderstandabilityPaper/machinelearning.tex
Fleshed out automated theorem proving section

@@
 %  - Theorem proving as example:
-%   - Least informative
+%   - Least informative is a yes/no/unknown answer
+%   -
 
@@
 \begin{document}
+The problem of making machine-generated artefacts more understandable to human
+users is quite clear in the field of automated theorem proving~CITE: the task of
+deciding whether a given formal statement follows from a given set of premeses.
+The least informative approaches, such as those based on the popular \emph{SLD
+  resolution} algorithm~CITE, produce nothing more than a ``yes'', ``no'' or
+``unknown'' response. Not only is this devoid of \emph{explanation}, but it also
+hides the effects of any bugs; requiring the user to either trust the results,
+or verify the implementation.
+
+Both of these issues can be mitigated by having the system instead generate a
+\emph{proof object}: a formal argument for \emph{why} a given statement follows
+or does not follow. Once generated, a proof object's validity can be checked
+without requiring any knowledge of how it was created, thus avoiding the need to
+trust or verify the (potentially complicated) search and generation
+procedures. The \emph{de Bruijn criterion}~CITE describes theorem provers which
+produce proof objects that are trivial to check by independent \emph{proof
+  checker} programs (which are themselves easily verified, due to their
+simplicity); examples are Coq~CITE and Isabelle~CITE.
+
+Proof objects are not a complete solution to understandability, since they can
+still be quite inscrutable to human users. This often depends on how closely
+the chosen formal system is able to encode the user's ideas: for example, the
+formal proof of Kepler's Conjecture was performed using a system of Higher Order
+Logic (HOL)~CITE whose proof objects (natural-deduction style derivations),
+whilst tedious, are in principle understandable to a user experienced with both
+the software and problem domain. The same cannot be said of the Boolean
+Pythagorean Triples problem, a statement of Ramsey theory involving the
+structure of the natural numbers. Rather than taking a high-level approach like
+HOL, the authors of~CITE analysed sets $\{0..n\}$ for larger and larger $n$,
+encoding these restrictions of the problem into the language of boolean
+satisfiability (SAT), and finding that the problem is unsatisfiable for
+$n= 7825$, and hence for the natural numbers as a whole. The proof object for
+$n = 7825$ contains 200 terabytes (compressable to 68 gigabytes) of
+propositional logic clauses, demonstrating the unsatisfiability of the encoded
+problem instance. Not only is this far too much for any human to comprehend, but
+the concepts used in the proof (boolean formulae) are several layers removed
+from the actual problem statement (natural numbers, subsets and pythagorean
+triples).
+
+Whilst ``low level'' formalisms like SAT are less understandable or explanatory
+for users, they are far more amenable to automation than more expressive logics.
+Despite the proof for the Boolean Pythagorean Triples problem being many orders
+of magnitude larger than that of the Kepler Conjecture, the latter is well
+beyond the ability of today's automated theorem provers due to its encoding in
+HOL. Instead, it took 22 collaborators 9 years just to formalise the proof
+(Hales had previously produced a less formal proof, hundreds of pages long and
+accompanied by unverified software).
+
 I meant to send this earlier, in case it's useful. These are just some

### 2019-02-28 UnderstandabilityPaper/machinelearning.tex
Finished draft of understandability

--- 2019-02-28 UnderstandabilityPaper/machinelearning.tex
@@
 \begin{document}
-I meant to send this earlier, in case it's useful. These are just some
-things that came to mind regarding "explainable" AI and machine
-learning. Lots of rambling, but hopefully some insight into current
-machine-learning thinking (as I understand it, as somewhat of an
-outsider!).
+The problem of making machine-generated artefacts more understandable to human
+users is quite clear in the field of \emph{automated theorem
+  proving}~\cite{sutcliffe2001evaluating}: the task of deciding whether a given
+formal statement follows from a given set of premeses.  The least informative
+approaches, such algorithms based on the popular \emph{resolution}
+rule~\cite{robinson1965machine}, produce nothing more than a ``yes'', ``no'' or
+``unknown'' response. Not only is this devoid of \emph{explanation}, but it also
+hides the effects of any bugs; requiring the user to either trust the results,
+or verify the implementation.
 
-I think a key idea is "differentiable programming", which captures the
-essence of how machine learning models like neural networks work, and
-precisely what makes them different from traditional computer programs
-(i.e. they're differentiable!), without ignoring the details ("black
-box") or using flimsy analogies ("brain-like"):
+Both of these issues can be mitigated by having the system instead generate a
+\emph{proof object}: a formal argument for \emph{why} a given statement follows
+or does not follow. Once generated, a proof object's validity can be checked
+without requiring any knowledge of how it was created, thus avoiding the need to
+trust or verify the (potentially complicated) search and generation
+procedures. The \emph{de Bruijn criterion}~\cite{barendregt2005challenge}
+describes theorem provers which produce proof objects that are trivial to check
+by independent \emph{proof checker} programs (which are themselves easily
+verified, due to their simplicity); examples are Coq~\cite{barras1997coq} and
+Isabelle~\cite{nipkow2002isabelle}.
 
-https://fluxml.ai/2019/02/07/what-is-differentiable-programming.html
+Proof objects are not a complete solution to understandability, since they can
+still be quite inscrutable to human users. This often depends on how closely the
+chosen formal system is able to encode the user's ideas: for example, the formal
+proof of the Kepler Conjecture was performed using a system of Higher Order
+Logic (HOL)~\cite{hales2015formal} whose proof objects (natural-deduction style
+derivations), whilst tedious, are in principle understandable to a user
+experienced with both the software and problem domain. The same cannot be said
+of the Boolean Pythagorean Triples problem, a statement of Ramsey theory
+involving the structure of the natural numbers. Rather than taking a high-level
+approach like HOL, the authors of~\cite{heule2016solving} analysed sets
+$\{0..n\}$ for larger and larger $n$, encoding these restricted versions of the
+problem into the language of boolean satisfiability (SAT), and found that the
+problem is unsatisfiable for $n= 7825$, and hence for the natural numbers as a
+whole. In this case, the proof object demonstrates this unsatisfiability using
+200 terabytes of propositional logic clauses (compressable to 68 gigabytes).
+Not only is this far too much for any human to comprehend, but the concepts used
+in the proof (boolean formulae) are several layers removed from the actual
+problem statement (natural numbers, subsets and pythagorean triples).
 
-When run on an input, a differentiable program doesn't just give an
-output, it also gives us the partial derivatives of that output
-w.r.t. each of the program's parameters. This tells us exactly how to
-adjust those parameters (via backpropagation) to reduce the output error
-for that particular input. Doing this enough times for enough inputs
-will hopefully find a general solution.
+Whilst ``low level'' formalisms like SAT are less understandable or explanatory
+for users, they are far more amenable to automation than more expressive logics.
+Despite the proof for the Boolean Pythagorean Triples problem being many orders
+of magnitude larger than that of the Kepler Conjecture, the latter is well
+beyond the ability of today's automated theorem provers due to its encoding in
+HOL. Instead, it took 22 collaborators 9 years just to formalise the proof
+(Hales had previously produced a less formal proof, hundreds of pages long and
+accompanied by unverified software; a reminder that human-generated artefacts
+are not necessarily understandable either!).
 
-A neural network is an example of a differentiable program, one which is
-just a big collection of identical expressions ("neurons") composed
-together. This is incredibly general (they are universal function
-approximators), but it doesn't have much "knowledge" or "structure"
-baked in: the behaviour of this program depends almost completely on the
-parameters, so everything it "learns" must come from input/output
-examples.
+% A similar pattern has emerged when it comes to understanding
 
-The convolutional neural networks that have come to dominate image
-processing are also differentiable programs, but they contain more
-knowledge/structure. Instead of general-purpose 'neurons', these use
-'convolutional layers' which are also expressions that get composed
-together. The difference is that these expressions hard-code the 2D grid
-structure that we expect to find in raster images. This makes them less
-general than neural networks, since they assume the data fits this 2D
-grid structure, but they also don't have to learn that structure from
-examples. Hence they need fewer examples, but we can also be more
-confident that the input/output relationships they learn involve 2D
-patches of image content (like shapes or textures), rather than some
-arbitrary numerical relationship between pixels.
+% I meant to send this earlier, in case it's useful. These are just some
+% things that came to mind regarding "explainable" AI and machine
+% learning. Lots of rambling, but hopefully some insight into current
+% machine-learning thinking (as I understand it, as somewhat of an
+% outsider!).
 
-Taking things much further, there are differentiable programs which do
-ray tracing ( https://people.csail.mit.edu/tzumao/diffrt/ ). These can
-be used to learn the position of 3D objects in an image, the camera
-position, illumination, etc. Again, this is less general since it
-assumes the input is an image of a 3D scene, but again it doesn't have
-to learn this knowledge from examples. The learned parameters are easily
-"explained", as they are the usual rendering parameters of camera
-position, rotation, etc.
+% I think a key idea is "differentiable programming", which captures the
+% essence of how machine learning models like neural networks work, and
+% precisely what makes them different from traditional computer programs
+% (i.e. they're differentiable!), without ignoring the details ("black
+% box") or using flimsy analogies ("brain-like"):
 
-Another example is learning to recognise and generate hand-written
-characters by using a differentiable program for drawing lines:
-https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf
+% https://fluxml.ai/2019/02/07/what-is-differentiable-programming.html
 
-In this way, there are two tradeoffs being made:
+% When run on an input, a differentiable program doesn't just give an
+% output, it also gives us the partial derivatives of that output
+% w.r.t. each of the program's parameters. This tells us exactly how to
+% adjust those parameters (via backpropagation) to reduce the output error
+% for that particular input. Doing this enough times for enough inputs
+% will hopefully find a general solution.
 
- - By hard-coding our knowledge of the domain into a program, like using
-   a ray-tracer for images of the 3D world, or pen-based drawing for
-   images of hand-written characters, we don't need as much training
-   data for the system to 'learn' what's going on. Unfortunately, we end
-   up with less general systems, e.g. we might apply the same
-   neural-network architecture to all sorts of problems (with enough
-   data), but we can't apply the ray-tracer to e.g. audio data.
+% A neural network is an example of a differentiable program, one which is
+% just a big collection of identical expressions ("neurons") composed
+% together. This is incredibly general (they are universal function
+% approximators), but it doesn't have much "knowledge" or "structure"
+% baked in: the behaviour of this program depends almost completely on the
+% parameters, so everything it "learns" must come from input/output
+% examples.
 
- - This same tradeoff also has implications for understandability and
-   explainability: the parameters of a ray-tracer are directly
-   understandable, based on how they're used in the program, e.g. "how
-   bright is this light source" or "how big is this object". In
-   contrast, the parameters of something like a neural network make
-   little sense on their own, they just mean "how important is this
-   output for that neuron?".
+% The convolutional neural networks that have come to dominate image
+% processing are also differentiable programs, but they contain more
+% knowledge/structure. Instead of general-purpose 'neurons', these use
+% 'convolutional layers' which are also expressions that get composed
+% together. The difference is that these expressions hard-code the 2D grid
+% structure that we expect to find in raster images. This makes them less
+% general than neural networks, since they assume the data fits this 2D
+% grid structure, but they also don't have to learn that structure from
+% examples. Hence they need fewer examples, but we can also be more
+% confident that the input/output relationships they learn involve 2D
+% patches of image content (like shapes or textures), rather than some
+% arbitrary numerical relationship between pixels.
 
-STUFF BELOW IS MORE TANGENTIAL
+% Taking things much further, there are differentiable programs which do
+% ray tracing ( https://people.csail.mit.edu/tzumao/diffrt/ ). These can
+% be used to learn the position of 3D objects in an image, the camera
+% position, illumination, etc. Again, this is less general since it
+% assumes the input is an image of a 3D scene, but again it doesn't have
+% to learn this knowledge from examples. The learned parameters are easily
+% "explained", as they are the usual rendering parameters of camera
+% position, rotation, etc.
 
-There are probably parallels to be drawn w.r.t. feature engineering
-too. "Feature extraction" turns raw data into numbers for the learning
-algorithm. This used to be done by hand, so we knew what the numbers
-meant (e.g. average brightness, number of edges, etc. this is what
-Katya's 'recurrent clustering' algorithm does for symbolic expressions).
-With "deep nets", such features are figured out automatically, and we
-don't know what they mean. Hence we gain generality (automatically
-discovered features tend to out-perform hand-engineered ones), but we
-lose understanding/explainability, and we become more dependent on
-training data.
+% Another example is learning to recognise and generate hand-written
+% characters by using a differentiable program for drawing lines:
+% https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf
 
-Things like "deep dream" can try to demonstrate what these features
-seem to represent, but not much more. As far as I understand it, deep
-dream works by picking a particular neuron/feature, providing an
-arbitrary input image to the system, and iteratively tweaking that image
-so that feature increases. The "explanation" of that feature is the
-changes that were made to the image, e.g. if the changes look
-eye-shaped, then we say that feature is for detecting eyes. This is, of
-course, open to interpretation.
+% In this way, there are two tradeoffs being made:
 
-Interestingly, exactly the same technique is used for generating
-adversarial inputs, where an input (usually to a classifier) is tweaked
-until it gets classified incorrectly:
+%  - By hard-coding our knowledge of the domain into a program, like using
+%    a ray-tracer for images of the 3D world, or pen-based drawing for
+%    images of hand-written characters, we don't need as much training
+%    data for the system to 'learn' what's going on. Unfortunately, we end
+%    up with less general systems, e.g. we might apply the same
+%    neural-network architecture to all sorts of problems (with enough
+%    data), but we can't apply the ray-tracer to e.g. audio data.
 
-% https://en.wikipedia.org/wiki/Adversarial_machine_learning
+%  - This same tradeoff also has implications for understandability and
+%    explainability: the parameters of a ray-tracer are directly
+%    understandable, based on how they're used in the program, e.g. "how
+%    bright is this light source" or "how big is this object". In
+%    contrast, the parameters of something like a neural network make
+%    little sense on their own, they just mean "how important is this
+%    output for that neuron?".
 
-Adversarial inputs show how limited this sort of "explanation" is, since
-these changes might look like random noise:
-https://blog.openai.com/adversarial-example-research/
+% STUFF BELOW IS MORE TANGENTIAL
 
-Or might even just be a single pixel:
-https://arxiv.org/abs/1710.08864
+% There are probably parallels to be drawn w.r.t. feature engineering
+% too. "Feature extraction" turns raw data into numbers for the learning
+% algorithm. This used to be done by hand, so we knew what the numbers
+% meant (e.g. average brightness, number of edges, etc. this is what
+% Katya's 'recurrent clustering' algorithm does for symbolic expressions).
+% With "deep nets", such features are figured out automatically, and we
+% don't know what they mean. Hence we gain generality (automatically
+% discovered features tend to out-perform hand-engineered ones), but we
+% lose understanding/explainability, and we become more dependent on
+% training data.
 
-Another nice example of "explaining" a machine learning model by
-demonstrating what its features represent is the variational
-autoencoder.
+% Things like "deep dream" can try to demonstrate what these features
+% seem to represent, but not much more. As far as I understand it, deep
+% dream works by picking a particular neuron/feature, providing an
+% arbitrary input image to the system, and iteratively tweaking that image
+% so that feature increases. The "explanation" of that feature is the
+% changes that were made to the image, e.g. if the changes look
+% eye-shaped, then we say that feature is for detecting eyes. This is, of
+% course, open to interpretation.
 
-Normal autoencoders 'compress' inputs down to a sequence of numbers
-representing their important features. Unfortunately these are hard to
-understand in isolation (sometimes called "holographic reduced
-representations", all of the components depend on all of the others).
+% Interestingly, exactly the same technique is used for generating
+% adversarial inputs, where an input (usually to a classifier) is tweaked
+% until it gets classified incorrectly:
 
-Variational autoencoders compress down to a multi-dimensional gaussian
-distribution instead. By randomly sampling from this gaussian, the
-representation is forced to be robust against small changes, and this
-imposes a meaningful structure on the space; such that adjusting the
-centre of the gaussian will alter the resulting data, e.g. by making a
-person's hair darker; or changing their gender; or giving them glasses;
-etc.
+% % https://en.wikipedia.org/wiki/Adversarial_machine_learning
 
-https://towardsdatascience.com/intuitively-understanding-variational-autoencoders-1bfe67eb5daf
-(Especially the "Vector arithmetic" section)
+% Adversarial inputs show how limited this sort of "explanation" is, since
+% these changes might look like random noise:
+% https://blog.openai.com/adversarial-example-research/
 
-Although this property can't actually be relied upon!
-http://paulrubenstein.co.uk/variational-autoencoders-are-not-autoencoders/
+% Or might even just be a single pixel:
+% https://arxiv.org/abs/1710.08864
 
-A similar effect to the structured "latent space" of the VAE has been
-found with word2vec:
+% Another nice example of "explaining" a machine learning model by
+% demonstrating what its features represent is the variational
+% autoencoder.
 
-% https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships
+% Normal autoencoders 'compress' inputs down to a sequence of numbers
+% representing their important features. Unfortunately these are hard to
+% understand in isolation (sometimes called "holographic reduced
+% representations", all of the components depend on all of the others).
 
-word2vec turns words into vectors of numbers, based on co-occurrences of
-words in some training data. These vectors turn out to have some
-"meaning" in the sense that vector arithmetic can make semantic changes.
-For example if we do the sum:
+% Variational autoencoders compress down to a multi-dimensional gaussian
+% distribution instead. By randomly sampling from this gaussian, the
+% representation is forced to be robust against small changes, and this
+% imposes a meaningful structure on the space; such that adjusting the
+% centre of the gaussian will alter the resulting data, e.g. by making a
+% person's hair darker; or changing their gender; or giving them glasses;
+% etc.
 
-  vector("king") - vector("man") + vector("woman")
+% https://towardsdatascience.com/intuitively-understanding-variational-autoencoders-1bfe67eb5daf
+% (Especially the "Vector arithmetic" section)
 
-The resulting vector is very close to vector("queen").
+% Although this property can't actually be relied upon!
+% http://paulrubenstein.co.uk/variational-autoencoders-are-not-autoencoders/
 
-These sorts of machine learning system are unsupervised, so they don't
-solve any particular problem. However, the representations they learn
-can be useful for other task-specific systems.
+% A similar effect to the structured "latent space" of the VAE has been
+% found with word2vec:
+
+% % https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships
+
+% word2vec turns words into vectors of numbers, based on co-occurrences of
+% words in some training data. These vectors turn out to have some
+% "meaning" in the sense that vector arithmetic can make semantic changes.
+% For example if we do the sum:
+
+%   vector("king") - vector("man") + vector("woman")
+
+% The resulting vector is very close to vector("queen").
+
+% These sorts of machine learning system are unsupervised, so they don't
+% solve any particular problem. However, the representations they learn
+% can be useful for other task-specific systems.
 
 \bibliographystyle{plain}
-\bibliography{./Bibtex}
+\bibliography{./bibtex}
 

### 2019-03-01 UnderstandabilityPaper/machinelearning.tex
Fix up understandability section

@@
-% STRUCTURE:
-
-%  - Understandability in AI
-%  - Theorem proving as example:
-%   - Least informative is a yes/no/unknown answer
-%   -
-
 \documentclass{article}
+\usepackage{cite}
 \begin{document}
@@
 
-% A similar pattern has emerged when it comes to understanding
+A similar problem of understandability has emerged in the field of \emph{machine
+  learning} (ML), where statistical methods are used to make predictions or
+decisions about data, often without human
+involvement~\cite{alpaydin2014introduction}. In particular there is a tradeoff
+between the generality of an approach and how easily its resulting behaviour can
+be understood.
 
-% I meant to send this earlier, in case it's useful. These are just some
-% things that came to mind regarding "explainable" AI and machine
-% learning. Lots of rambling, but hopefully some insight into current
-% machine-learning thinking (as I understand it, as somewhat of an
-% outsider!).
+Many ML approaches can be characterised as constructing a computer program (or
+``model'') consisting of two parts: an overall structure or \emph{architecture},
+which remains fixed; and a set of adjustable \emph{parameters}, which are
+inferred or ``learned'' from data (e.g. \emph{training data} of desired
+input/output examples). One particularly simple architecture is the
+\emph{decision tree}: nested boolean queries of the input, often used for
+\emph{classification}~\cite{safavian1991survey}. These queries are parameters,
+and are chosen based on how efficiently they separate the classes given in the
+training data.  Decision trees are nominally understandable, since their
+behaviour on a given input traces a single path through these queries; yet they
+perform poorly compared to other ML algorithms. The \emph{random forest}
+approach yields better results by combining many decision trees and having them
+vote on the overall outcome~\cite{randomforests}, although such ensemble
+behaviour is more difficult to reason about than that of a single tree.
 
-% I think a key idea is "differentiable programming", which captures the
-% essence of how machine learning models like neural networks work, and
-% precisely what makes them different from traditional computer programs
-% (i.e. they're differentiable!), without ignoring the details ("black
-% box") or using flimsy analogies ("brain-like"):
+Recent research has focused on highly expressive classes of models such as
+\emph{differentiable programming}~\cite{wang2018demystifying}, whose
+architectures output not only a (numerical) answer, but also partial derivatives
+with respect to the parameters; and \emph{probabilistic programming}, which
+samples from a \emph{distribution} conditioned on the training data. Both
+frameworks allow arbitrary architectures, specified via Turing-complete
+languages, and provide efficient, composable methods for optimising the
+parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
+respectively) to minimise arbitrary loss functions (e.g. the squared output
+error on a training data set).
 
-% https://fluxml.ai/2019/02/07/what-is-differentiable-programming.html
+With such expressive formalisms, the conflict between the generality of a model
+and its understandability becomes clear. Task-specific architectures require
+fewer parameters than general-purpose approaches and perform well with little
+training data, for example hand-written characters can be classified based on a
+single example if we allow our model to assume the given images are generated by
+pen strokes~\cite{lake2015human}. Likewise we can infer parameters of a 3D scene
+(such as object position and lighting) from images if we embed a ray-tracer into
+our model~\cite{li2018differentiable}; embedding a physics engine can provide
+predictions about these scenes, which are useful e.g. for robot
+controllers~\cite{degrave2016differentiable}.
 
-% When run on an input, a differentiable program doesn't just give an
-% output, it also gives us the partial derivatives of that output
-% w.r.t. each of the program's parameters. This tells us exactly how to
-% adjust those parameters (via backpropagation) to reduce the output error
-% for that particular input. Doing this enough times for enough inputs
-% will hopefully find a general solution.
+The parameters of such programs are easily understood: they describe the pen
+strokes, geometric and physical properties of the simulated system, inferred
+from the given images. However, such specificity makes these implementations
+unsuitable for anything else. The choice of such high-level, task-specific
+components is performed by the user, and encodes some of their domain knowledge
+into the structure of the solution, such that it doesn't have to be learned.
+This is similar to how logics like HOL can directly represent high-level
+concepts (e.g. natural numbers and sets) but whose proof methods have limited
+reusability due to the difficulty of automating such high-level reasoning.
 
-% A neural network is an example of a differentiable program, one which is
-% just a big collection of identical expressions ("neurons") composed
-% together. This is incredibly general (they are universal function
-% approximators), but it doesn't have much "knowledge" or "structure"
-% baked in: the behaviour of this program depends almost completely on the
-% parameters, so everything it "learns" must come from input/output
-% examples.
+At the other end of the spectrum we have differentiable programs with general
+purpose architectures, like \emph{neural networks} (capable of universal
+function approximation~\cite{funahashi1989approximate}). These are compositions
+of a large number of identical sub-expressions (``neurons''), whose parameters
+(``weights'') scale the input values, and hence the contribution of each
+sub-expression to the whole. Such architectures encode essentially no domain
+knowledge, requiring much larger training sets than task-specific algorithms in
+order to ``learn'' these details.  The behaviour of general purpose models
+depends so heavily on their (many) parameters, that understanding their
+behaviour becomes difficult and they are often treated as inscrutable ``black
+boxes''; akin to the large (un)SAT proof described above.
 
-% The convolutional neural networks that have come to dominate image
-% processing are also differentiable programs, but they contain more
-% knowledge/structure. Instead of general-purpose 'neurons', these use
-% 'convolutional layers' which are also expressions that get composed
-% together. The difference is that these expressions hard-code the 2D grid
-% structure that we expect to find in raster images. This makes them less
-% general than neural networks, since they assume the data fits this 2D
-% grid structure, but they also don't have to learn that structure from
-% examples. Hence they need fewer examples, but we can also be more
-% confident that the input/output relationships they learn involve 2D
-% patches of image content (like shapes or textures), rather than some
-% arbitrary numerical relationship between pixels.
+Full generality is seldom required: even if we don't want to constrain a model's
+internal operation, the input and output data formats will always be known. Such
+knowledge can be encoded in the architecture through the use of
+\emph{convolutional layers}, for example forcing a 2D grid structure when
+dealing with images~\cite{krizhevsky2012imagenet}. Introducing task-specific
+components like this can aid in understanding the resulting models, since they
+are constrained to operate on 2D patches of image content (like shapes or
+textures), rather than potentially arbitrary numerical relationships between
+pixels.
 
-% Taking things much further, there are differentiable programs which do
-% ray tracing ( https://people.csail.mit.edu/tzumao/diffrt/ ). These can
-% be used to learn the position of 3D objects in an image, the camera
-% position, illumination, etc. Again, this is less general since it
-% assumes the input is an image of a 3D scene, but again it doesn't have
-% to learn this knowledge from examples. The learned parameters are easily
-% "explained", as they are the usual rendering parameters of camera
-% position, rotation, etc.
-
-% Another example is learning to recognise and generate hand-written
-% characters by using a differentiable program for drawing lines:
-% https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf
-
-% In this way, there are two tradeoffs being made:
-
-%  - By hard-coding our knowledge of the domain into a program, like using
-%    a ray-tracer for images of the 3D world, or pen-based drawing for
-%    images of hand-written characters, we don't need as much training
-%    data for the system to 'learn' what's going on. Unfortunately, we end
-%    up with less general systems, e.g. we might apply the same
-%    neural-network architecture to all sorts of problems (with enough
-%    data), but we can't apply the ray-tracer to e.g. audio data.
-
-%  - This same tradeoff also has implications for understandability and
-%    explainability: the parameters of a ray-tracer are directly
-%    understandable, based on how they're used in the program, e.g. "how
-%    bright is this light source" or "how big is this object". In
-%    contrast, the parameters of something like a neural network make
-%    little sense on their own, they just mean "how important is this
-%    output for that neuron?".
-
-% STUFF BELOW IS MORE TANGENTIAL
-
-% There are probably parallels to be drawn w.r.t. feature engineering
-% too. "Feature extraction" turns raw data into numbers for the learning
-% algorithm. This used to be done by hand, so we knew what the numbers
-% meant (e.g. average brightness, number of edges, etc. this is what
-% Katya's 'recurrent clustering' algorithm does for symbolic expressions).
-% With "deep nets", such features are figured out automatically, and we
-% don't know what they mean. Hence we gain generality (automatically
-% discovered features tend to out-perform hand-engineered ones), but we
-% lose understanding/explainability, and we become more dependent on
-% training data.
-
-% Things like "deep dream" can try to demonstrate what these features
-% seem to represent, but not much more. As far as I understand it, deep
-% dream works by picking a particular neuron/feature, providing an
-% arbitrary input image to the system, and iteratively tweaking that image
-% so that feature increases. The "explanation" of that feature is the
-% changes that were made to the image, e.g. if the changes look
-% eye-shaped, then we say that feature is for detecting eyes. This is, of
-% course, open to interpretation.
-
-% Interestingly, exactly the same technique is used for generating
-% adversarial inputs, where an input (usually to a classifier) is tweaked
-% until it gets classified incorrectly:
-
-% % https://en.wikipedia.org/wiki/Adversarial_machine_learning
-
-% Adversarial inputs show how limited this sort of "explanation" is, since
-% these changes might look like random noise:
-% https://blog.openai.com/adversarial-example-research/
-
-% Or might even just be a single pixel:
-% https://arxiv.org/abs/1710.08864
-
-% Another nice example of "explaining" a machine learning model by
-% demonstrating what its features represent is the variational
-% autoencoder.
-
-% Normal autoencoders 'compress' inputs down to a sequence of numbers
-% representing their important features. Unfortunately these are hard to
-% understand in isolation (sometimes called "holographic reduced
-% representations", all of the components depend on all of the others).
-
-% Variational autoencoders compress down to a multi-dimensional gaussian
-% distribution instead. By randomly sampling from this gaussian, the
-% representation is forced to be robust against small changes, and this
-% imposes a meaningful structure on the space; such that adjusting the
-% centre of the gaussian will alter the resulting data, e.g. by making a
-% person's hair darker; or changing their gender; or giving them glasses;
-% etc.
-
-% https://towardsdatascience.com/intuitively-understanding-variational-autoencoders-1bfe67eb5daf
-% (Especially the "Vector arithmetic" section)
-
-% Although this property can't actually be relied upon!
-% http://paulrubenstein.co.uk/variational-autoencoders-are-not-autoencoders/
-
-% A similar effect to the structured "latent space" of the VAE has been
-% found with word2vec:
-
-% % https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships
-
-% word2vec turns words into vectors of numbers, based on co-occurrences of
-% words in some training data. These vectors turn out to have some
-% "meaning" in the sense that vector arithmetic can make semantic changes.
-% For example if we do the sum:
-
-%   vector("king") - vector("man") + vector("woman")
-
-% The resulting vector is very close to vector("queen").
-
-% These sorts of machine learning system are unsupervised, so they don't
-% solve any particular problem. However, the representations they learn
-% can be useful for other task-specific systems.
+Understanding exactly how these programs make their decisions is an active area
+of research, known as \emph{explainable artificial intelligence}
+(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable},
+with popular approaches such as saliency maps~\cite{simonyan2013deep} tending to
+reverse-engineer the factors which lead to a model's decisions (judging the
+saliency of each pixel based on how strongly it would effect the output
+prediction if adjusted). These methods appear intuitive, e.g. producing
+visualisations highlighting a particular object in a scene as the reason for its
+classification; akin to the proof objects emitted by theorem provers. Yet this
+can obscure the difficulty of interpreting such high-dimensional decision
+boundaries. In particular, these explanations or justifications can be
+fundamentally altered by imperceptible adjustments to the
+input~\cite{ghorbani2017interpretation}. Similar adjustments can also change a
+model's output, leading to the field of \emph{adversarial machine
+  learning}~\cite{goodfellow2014explaining}; adjustments to even a single pixel
+can not only cause a system to mislabel an input, but to give high confidence to
+its erroneous result~\cite{su2019one}. Such fragility and opaqueness has
+important implications as these techniques begin to have decision making
+applications in life-critical domains such as automotive
+control~\cite{bojarski2016end} and medical
+diagnosis~\cite{esteva2017dermatologist}.
 

### 2019-03-05 UnderstandabilityPaper/iccc-science.tex
Switch out our section for the combined tex

% This file is iccc.tex.  It contains the formatting instructions for and acts as a template for submissions to ICCC.  It borrows liberally from the AAAI and IJCAI formats and instructions.  It uses the files iccc.sty, iccc.bst and iccc.bib, the first two of which also borrow liberally from the same sources.

\documentclass[letterpaper]{article}
\usepackage{iccc}

\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\pdfinfo{
/Title (Applying CC to scientific domains)
/Subject (Proceedings of ICCC)
/Author (ICCC)}
% The file iccc.sty is the style file for ICCC proceedings.
%
\title{The importance of applying computational creativity\\ to
  scientific and mathematical domains\\
Position Paper}
%\author{Alison Pease, Simon Colton, Chris Warburton, Athanasios Nathanail, Irina Preda, \\
%  Daniel Arnold, Daniel Winterstein, Mike Cook}

\setcounter{secnumdepth}{0}

\begin{document} 
\maketitle
\begin{abstract}
\begin{quote}
  Science and mathematics are currently vastly under-represented in
  the computational creativity (CC) community. This is at best a
  wasted opportunity, and at worst a significant problem for
  computational creativity. We discuss why the CC community should
  apply their work to scientific and mathematical domains, and argue
  that this would be mutually beneficial for the domains in question,
  demonstrating our position throughout the paper by reference to
  mathematics and geology. We propose a research programme for
  building closer relationships between CC and the scientific
  community, focusing on understandability of science and the role
  that AI is currently playing in scientific research.
\end{quote}
\end{abstract}

\section{Introduction}

Despite the best efforts of successive ICCC organising committees and
the computational creativity (CC) community, CC has always attracted
substantially more interest from researchers in artistic than
scientific and mathematical domains. In their 2017 study of
application domains in CC, \cite{loughran17} found that, of 16
categories, papers on Maths, Science and Logic accounted for only 3\%
of the 353 papers on CC across 12 years. Of course, some work is
domain independent, or at least not easily assigned to an academic
discipline, such as the body of work on CC and curiosity (for instance
\cite{grace}). Even taking this into account, it is clear that science
and mathematics are vastly under-represented in our community.

There are potentially many reasons why this may be the case. Firstly,
AI researchers in scientific domains may well be doing
creativity-related work in other contexts but not engaging with the CC
community. Automated reasoning and automated scientific discovery are
both thriving subfields of AI, with internationally recognised
journals as outlets for publication and engagement; certainly these
will contain work relevant to our field but couched in different
terminology with different methodologies. Secondly, it may be that
other, practical, priorities in scientific domains have led to a focus
on techniques such as search, data-mining and automated
deduction. Since these generate results of interest to domain experts,
the more difficult, fluid and tenuous concept of creativity may be
seen as unnecessary, risky or simply not a priority. This may
particularly be the case given the various ``AI winters'' in the
twentieth century (the second of which ended in 1993, just six years
before the first workshop on CC), and the need for AI to ``prove
itself''. Thirdly, CC researchers may consider that even if generation
is possible within scientific domains, evaluation is too
difficult. How we should evaluate our work and our systems has always
been a contentious -- albeit important -- issue in CC, with few
proposed evaluation metrics and the majority of researchers still
arguing for value along the lines of ``we/people liked the system's
output'' or ``we/people couldn't distinguish the system's output from
human produced work''. It might be the case that in science, the main
evaluation metric -- ``is it true?'', or ``does it work?'' -- is
considered simply too expensive or difficult to demonstrate.

We argue here that neglecting scientific and mathematics domains in CC
is at best a wasted opportunity, and at worst a significant problem
for CC. While computationally-produced work in the arts is currently
novel, this {\em may} have a limited lifespan. We call this the {\em
  Saturation Problem}. In ten or twenty years, people may ask
themselves whether we still want more computer-generated paintings or
poems. There may or may not be a good answer to this. The question as
to whether we still want more computer-generated science or
mathematics, however, seems less likely to be asked: we {\em always}
want more science and mathematics. We believe that it is essential to
the health of our field that we reach out as a community at this
stage, both to domain experts in science and maths and to those in
related AI areas. The benefits of doing so will go both ways. As AI is
used more and more in science, there is greater dependence on blackbox
machine learning systems. While providing greater predictive power,
this often comes at the cost of understanding. We call this the {\em
  Understandability Problem} and argue that it will become a big issue
in science, which we will have to address.  Twenty years of thinking
about computational creativity has provided us with valuable tools for
thinking about these problems. This paper is a call to arms to CC
researchers to apply their work to science, thus solving both the
saturation problem and the understandability problem.

Loughran and O'Neill argue that ``tackling scientific, logical or
realistic issues could help bring the reputation of CC away from a
purely aesthetic domain towards developing solutions for real world
problems.'' \cite[p.7]{loughran17} and that ``It is imperative that
the field remains balanced as it grows and that we remember to reflect
on all areas of growth.'' [{\em Ibid.}]. In this paper we support and
present further arguments for this position, alongside practical
recommendations for doing so.

% We start by looking at work in related fields, arguing that in fact CC
% research {\em is} being carried out in science and maths, but under
% different umbrellas, with different terminologies and priorities.  We
% then argue both that it is imperative for the CC community to apply
% their work to scientific and mathematical domains, and that this would
% be beneficial for the domains in question. We conclude by proposing a
% research programme for doing so, using examples from mathematics and
% geology throughout. We conclude by proposing a research programme for
% doing so, using examples from mathematics and geology throughout.

\section{What is science?}\label{science}
%\subsection{what is science?}
The concept of science is not a straightforward one. The division of
the origins of learning and systematic production of new knowledge
into disciplines as we know them tends to takes into account at least
some of the following: methodologies, objects of study (which can be
shared with other disciplines), a body of accumulated knowledge (which
is generally not shared with other disciplines), theories and
concepts, terminology and an institutional manifesto (so that it can
reproduce itself) \cite[p9]{krishnan2009academic}. Sciences include
{\em Natural sciences}, which are subdivided into {\em physical
  sciences} (chemistry, physics, astronomy), {\em life sciences}, or
biology (zoology, botany, ecology, genetics) and {\em earth science}
(geology, oceanography, meteorology, palaeontology); {\em Social
  sciences} (psychology, sociology, economics, law, political
science); {\em Formal sciences} (mathematics, logic, theoretical
computer science, statistics); and {\em Applied sciences}, which are
subdivided into {\em engineering} (computer science, civil
engineering, electrical engineering, mechanical engineering); {\em
  health sciences} (medicine, dentistry, pharmacy) and {\em
  agriculture}. The number and variety of sciences makes
generalisations difficult, and core values vary accordingly.  However,
values commonly associated in particular with the (rather unhelpfully
named) ``hard sciences'' include repeatability, reproducibility,
predictability, generality and understandability. This last value is
particularly cherished: for instance, Roger G. Newton sums it up as
``The primary aim of most physical scientists is to understand and
explain the workings of Nature.''  \cite[p. 4]{newton:2000}.

%\subsection{What are the arts?}

The arts are just as difficult to define.  These are sometimes divided
into the ``seven arts'': Architecture, Sculpture, Painting,
Literature, Music, Performing, and Film.\footnote{A visualisation of
  the fields of knowledge based on Wikipedia's list of academic
  disciplines and sub-disciplines can be found here:
  http://www.thingsmadethinkable.com/item/fields\_of\_knowledge.php}
Julie Van Camp, writing in the context of United States Congressional 
policy on arts education, provides the following extensional
definition:

\begin{quote}
  The term `the arts' includes, but is not limited to, music
  (instrumental and vocal), dance, drama, folk art, creative writing,
  architecture and allied fields, painting, sculpture, photography,
  graphic and craft arts, industrial design, costume and fashion
  design, motion pictures, television, radio, film, video, tape and
  sound recording, the arts related to the presentation, performance,
  execution, and exhibition of such major art forms, all those
  traditional arts practiced by the diverse peoples of this
  country. ({\em sic}) and the study and application of the arts to
  the human
  environment.\footnote{http://web.csulb.edu/~jvancamp/361\_r8.html}
\end{quote}

%\subsection{Connection between the sciences and the arts}

%\subsection{Traditional boundaries might not serve us}

\begin{table}
\begin{center}
 \begin{tabular}{||l c c ||} 
 \hline
 & {\bf Science} & {\bf Arts} \\[0.5ex] 
 \hline\hline
{\em Aesthetic:} & truth & beauty \\ \hline

{\em Approach:} & problem-driven & artefact-driven \\ \hline

{\em Task:} & analytic & generative \\ \hline

{\em Terminology:} & discover & create \\ \hline

{\em Status:} & objective & subjective \\ \hline

{\em Goal:} & knowledge & self-expression \\ \hline

 \end{tabular} \caption{Possible differences between the sciences and the arts}
\end{center}
\label{Table:differences}
\end{table}

As a starting point -- generalising (and controversially) -- we could
suggest some of the differences between the sciences and the arts as
shown in Table 1. Of course, the real-world everyday lived experience
of {\em doing} science or {\em doing} art is far more complex than
Table 1 would suggest. Studies of interpretations of seismic data in
geology, for instance, show the large number of different expert
interpretations of the same seismic section, highlighting the
subjectivity involved \cite{bond}. These interpretations are used to
analyse subsurface geology, and form the basis for many exploration
and extraction decisions. Even in cases where interpreters report that
an interpretation is relatively straightforward, there are significant
differences in interpretation, leading to significantly different
predictions, for instance about gross pore volume in a pool or gross
rock volume \cite{rankey}. While clearly objectivity is the goal here,
such studies may suggest that this aspect of geological practice is
closer to visual art interpretation than it is to some other
scientific domains. Similarly, studies of the back stage production of
mathematics show that beauty is often a guiding value
\cite{inglis:15}; there is a high level of disagreement amongst
experts about the validity of certain proofs \cite{inglis}; and proofs
and theories are often considered to be constructed rather than
discovered \cite{lakatos}. Less structured knowledge such as our
ability to reason logically has been shown to be highly context
dependent (for instance, participants in the Wason Selection Task were
unable to solve a problem at an abstract level but could solve it
correctly when it was framed in a familiar context \cite{wason});
constructing grounding metaphors to the physical world and abstract
linking metaphors found to be fundamental to our understanding and
construction of mathematical knowledge \cite{Lak00}; and even the
language in which reasoning occurs affecting our preconceptions,
perceptions and assumptions \cite{dehaene,barton}.

Dibbets expresses the relationship between arts and sciences well:
\begin{quote}
  But in the end, we all do very much of the same. All scientists,
  artists, composers and writers are intensively occupied imagining
  something that does not yet exist. They find themselves at the
  borders of areas where up to then hardly anyone found himself,
  trying to solve problems that are incomprehensible to others, trying
  to answer questions no one has ever asked. Here, they share a vision
  on things that are not yet real. \cite[p. 1]{dibbets}
\end{quote}
Some of these interdisciplinary features are recognised in curriculum
design and teaching featuring transferable skills, in which one skill
may be learned within a scientific context and developed or employed
in an arts context, or vice versa (see for instance \cite{gaff}).
Indeed, the STEM to STEAM movement (expanding the acronym Science,
Technology, Engineering, and Mathematical disciplines to include Art)
explores the role of arts integration, collaboration, and experience
centered learning in knowledge creation \cite{ghanbari}. Of course,
the need for so many interdisciplinary initiatives (and related
concepts such as transdisciplinarity, pluridisciplinarity, and
multidisciplinarity) may suggest that some traditional discipline
boundaries are no longer drawn in a helpful way. The evolving role and
functionality of AI systems further complicates things. The focus of
AI researchers, particularly in machine learning, is often on the
skills they hope to simulate rather than a particular domain in which
they are usually employed. This may be more a more productive approach
than the typical CC focus on domain over skill.

%1-3, programme chairs should design the call for papers with these in
%mind, collaborators in scientific domains should be invited as keynote
%speakers, workshops on science should be co-located with
%ICCC.... programmes should be themed around skill rather than domain,
%[ Common themes include linguistic, music, visual arts, etc,
%creativity, with science/mathematical creativity lumped into a single
%theme if that.  ]

\section{The Saturation Problem in Computational Creativity}

The saturation problem in CC is potentially an identity problem for
us. Recent developments in other areas of AI -- principally machine
learning (ML) -- have led to astonishingly rapid progress in
generative processes.  Research in Constructive Machine Learning has
led to impressive generative results in both the arts and sciences,
including painting, music, poetry, gaming, drug design, and gene
design -- usually in collaboration with domain experts. Our concern is
that the sheer size and combined resources of the ML community may
render generative work in CC untenable.

CC has long been seen as more than ``mere generation''\footnote{The
  slogan at the 2012 ICCC conference was ``scoffing at mere generation
  for more than a decade.'' - although this has been challenged, for
  instance by \cite{ventura})}, with many other aspects of the
creative act modelled, in particular aesthetic judgements, but also
(more controversially) the importance of framing information and
meta-level processes. As generative results in other areas of AI
become more sophisticated, we may need to focus on these other aspects
of the creative act. However, even with this in mind, we are concerned
that the arts domains {\em may} reach a saturation point for CC: as
the novelty and backstory of computer-generated art grows old, society
may question whether and why we want more computer-produced artistic
artefacts. In order to keep the field alive, we propose that as well
as focusing on other aspects of creativity, we find other application
domains.

\section{The Understandability Problem in Science}
Roger Newton's quote above about the primary aim of physical
scientists being to understand and explain nature is uncontroversial,
but difficult to unpack. Ever since the entirety of our collective
scientific knowledge became too big for a single polymath to
comprehend, we have had to outsource our understanding to others. The
institutionalised ways in which trust of others' understanding and
progress is handled started with the early universities, and developed
with the invention of the printing press, academic journals, the peer
review process and so on. Knowledge and understanding is a social
process, as argued in \cite{martin:2013}, but even in the human-only
case this gets complicated. The longest proof in history, of the
Classification of finite simple groups, is over 10,000 pages, spread
across 500 or so journal articles, by over 100 different authors from
around the world, and took 110 years to complete. What does
understanding mean here? Perhaps a handful of people understand the
proof in its entirety, and when they die it is not obvious that any
single person will ever again understand the entire proof (in part
because it may be replaced with a simpler proof).

Shinichi Mochizuki's 500-page proof in 2012 of the abc conjecture
provides another perspective on the importance of understanding in
mathematics. The abc conjecture, proposed in 1985, on relationships
between prime numbers, is considered to be one of the most important
conjectures in number theory (more significant than Fermat’s Last
Theorem -- in fact Fermat’s Last Theorem would be a corollary of the
proof). A proof would be ``one of the most astounding achievements of
mathematics of the twenty-first century.''  [Goldfeld, in
\cite{ball:12}]. Mochizuki has a good track record as a
mathematician, having proved ``extremely deep'' theorems in the past
[Conrad in \cite{ball:12}]. The problem is that the techniques and
mathematical objects which Mochizuki has developed to use in his proof
are so strange and new that it would take a reviewer or mathematical
colleague most of their career to understand them, before they were
able to understand and verify the proof. Despite some efforts from
Mochizuki and a handful of his followers to make his work accessible,
currently his proof is neither published nor accepted by mainstream
mathematicians, for the simple reason that they don't understand
it.

Crowd-sourced mathematics, in which open conjectures are solved
collaboratively via online fora (amongst other things), has been used
for around ten years now by a subset of the mathematical community as
a new way of producing mathematics through collaboration and sharing
\cite{massively-collaborative}. Neilsen argues that this has resulted
in ``amplifying collective intelligence'' in his book {\em Reinventing
  Discovery} \cite{nielsen}. It has certainly resulted in some
original and significant new proofs. Here it is perfectly possible (in
fact it would be surprising if it were not the case) for a person to
be a co-author but not fully understand the proof in their
paper.

Adding computers to the social process, to form a combination of
people, computers, and mathematical archives to create and apply
mathematics -- a ``mathematics social machine'' \cite{martin:2013} --
further complicates matters. \emph{Automated theorem proving} is the
task of deciding whether a given formal statement follows from a given
set of premises ~\cite{sutcliffe2001evaluating}.  The least
informative approaches, algorithms based on the \emph{resolution}
rule~\cite{robinson1965machine}, produce merely a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation},
but it also hides the effects of any bugs; requiring the user to
either trust the results, or verify the implementation.

Both of these issues can be mitigated by having the system instead
generate a \emph{proof object}: a formal argument for \emph{why} a
given statement follows or does not follow. Once generated, a proof
object's validity can be checked without requiring any knowledge of
how it was created, thus avoiding the need to trust or verify the
(potentially complicated) search and generation procedures. The
\emph{de Bruijn criterion}~\cite{barendregt2005challenge} describes
theorem provers which produce proof objects that are trivial to check
by independent \emph{proof checker} programs (which are themselves
easily verified, due to their simplicity); examples are
Coq~\cite{barras1997coq} and Isabelle~\cite{nipkow2002isabelle}.

Proof objects are not a complete solution to understandability, since
they can still be quite inscrutable to human users. This often depends
on how closely the chosen formal system is able to encode the user's
ideas: for example, the formal proof of the Kepler Conjecture was
performed using a system of Higher Order Logic
(HOL)~\cite{hales2015formal} whose proof objects (natural-deduction
style derivations), whilst tedious, are in principle understandable to
a user experienced with both the software and problem domain. The same
cannot be said of the Boolean Pythagorean Triples problem, a statement
of Ramsey theory involving the structure of the natural
numbers. Rather than taking a high-level approach like HOL, the
authors of~\cite{heule2016solving} analysed sets $\{0..n\}$ for larger
and larger $n$, encoding these restricted versions of the problem into
the language of boolean satisfiability (SAT), and found that the
problem is unsatisfiable for $n= 7825$, and hence for the natural
numbers as a whole. In this case, the proof object demonstrates this
unsatisfiability using 200 terabytes of propositional logic clauses
(compressable to 68 gigabytes).  Not only is this far too much for any
human to comprehend, but the concepts used in the proof (boolean
formulae) are several layers removed from the actual problem statement
(natural numbers, subsets and pythagorean triples).

Whilst ``low level'' formalisms like SAT are less understandable or
explanatory for users, they are far more amenable to automation than
more expressive logics.  Despite the proof for the Boolean Pythagorean
Triples problem being many orders of magnitude larger than that of the
Kepler Conjecture, the latter is well beyond the ability of today's
automated theorem provers due to its encoding in HOL. Instead, it took
22 collaborators 9 years just to formalise the proof (Hales had
previously produced a less formal proof, hundreds of pages long and
accompanied by unverified software; yet another reminder that
human-generated artefacts are not necessarily understandable either).

The situation is even worse in automated scientific discovery. The
widespread use of \emph{machine learning} (ML) to find patterns in
scientific data has been criticised by Genevera Allen in her recent
talk at the 2019 Annual Meeting of the American Association for the
Advancement of Science
(AAAS)\footnote{https://eurekalert.org/pub\_releases/2019-02/ru-cwt021119.php}. She
highlighted accuracy and reproducibility issues with scientific
discoveries made by machine-learning techniques. Understandability is
another huge problem with such discoveries; often with a tradeoff
between the generality of an approach and how easily its resulting
behaviour can be understood. The understandability problem concerns
this last issue.

\section{Proposed solutions}
Here we propose three different approaches that the CC community can
take to address the understandability problem.  Each approach, at the
same time, would solve the saturation problem, by focusing on
science. We end each subsection with a concrete recommendation towards
a new Research Programme in CC.

\subsection{The ``human-like computing'' approach}

{\em Human-like computing} research is a research programme developed
for a UK funding initiative by the Engineering and Physical Sciences
Research Council. This programme ``aims to endow machines with
human-like perceptual, reasoning and learning abilities, which support
collaboration and communication with human beings.'', with one of the
stated motivations to ``inspire new forms of computation based on
human cognition, especially on tasks where humans currently exhibit
superior
abilities.''\footnote{https://epsrc.ukri.org/newsevents/pubs/human-like-computing-strategy-roadmap/}

We believe that a focus on more human-like computing would result in
systems whose output more closely resembled human produced work. If
this were the case, then we hypothesise that there would be greater
understandability.  In the context of science and mathematics, there
is much theoretical work on how people work in these domains and how
knowledge is constructed. For instance, the philosophy of informal
mathematics and the study of mathematical practice and cultures are
thriving communities with annual research events and a good number of
published books and papers. We can build on this work by building
human-like computing systems which model theoretical findings.

The CC community is particularly well equipped to work in this
research programme. The FACE evaluation model \cite{colton11} is
based on the multiple aspects involved in the human creative act,
including aesthetic judgements, concept development, framing
information and meta-level processes which can generate, for instance,
the means by which an artefact is generated. This might be reflected
in automated theory formation approaches to Automated Reasoning, which
consider a far wider view of mathematical knowledge production than
the traditional narrow focus of Automated Theorem Proving, including
the automatic generation and evaluation of new conjectures, concepts,
examples explanations, and so on (see \cite{atf-the-next-generation}
for an example). In some ways, given the closeness of some (aspects of
some) sciences to artistic domains, this may be low hanging fruit for
system developers to apply their systems to scientific and
mathematical domains.

%in our own work, we follow this ....

\begin{quote}
%\noindent\fbox{%
 % \parbox{\textwidth}{%
    {\bf Recommendation 1:} Apply your system to scientific domains.

%}%
%}
\end{quote}

\subsection{The ``Framing '' approach} 
Enhancing software with explanatory functionality would also help to
mitigate the understandability problem. The ``F'' from the FACE model
stands for {\em Framing}, and we advocate a dual-approach of software
generating framing information alongside an artefact, problem solution
or new data pattern. We foresee this being an increasingly important
area of research in CC, with an increasing level of sophistication:
from explanation to justification to argument and dialogue with a user
about the value, method of production, motivation etc behind
output. How framing information should be developed is a research
programme in its own right. For now, we discuss greater and lesser
understandability in terms of describing the processes underlying the
generative act and consider these for ML approaches.

Many ML approaches can be characterised as constructing a computer
program (or ``model'') consisting of two parts: an overall structure
or \emph{architecture}, which remains fixed; and a set of adjustable
\emph{parameters}, which are inferred or ``learned'' from data
(e.g. \emph{training data} of desired input/output examples). One
particularly simple architecture is the \emph{decision tree}: nested
boolean queries of the input, often used for
\emph{classification}~\cite{safavian1991survey}. These queries are
parameters, and are chosen based on how efficiently they separate the
classes given in the training data.  Decision trees are nominally
understandable, since their behaviour on a given input traces a single
path through these queries; yet they perform poorly compared to other
ML algorithms. The \emph{random forest} approach yields better results
by combining many decision trees and having them vote on the overall
outcome~\cite{randomforests}, although such ensemble behaviour is more
difficult to reason about than that of a single tree.

Recent research has focused on highly expressive classes of models such as
\emph{differentiable programming}~\cite{wang2018demystifying}, whose
architectures output not only a (numerical) answer, but also partial derivatives
with respect to the parameters; and \emph{probabilistic programming}, which
samples from a \emph{distribution} conditioned on the training data. Both
frameworks allow arbitrary architectures, specified via Turing-complete
languages, and provide efficient, composable methods for optimising the
parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
respectively) to minimise arbitrary loss functions (e.g. the squared output
error on a training data set).

With such expressive formalisms, the conflict between the generality
of a model and its understandability becomes clear. Task-specific
architectures require fewer parameters than general-purpose approaches
and perform well with little training data, for example hand-written
characters can be classified based on a single example if we allow our
model to assume the given images are generated by pen
strokes~\cite{lake2015human}. Likewise we can infer parameters of a 3D
scene (such as object position and lighting) from images if we embed a
ray-tracer into our model~\cite{li2018differentiable}; embedding a
physics engine can provide predictions about these scenes, which are
useful e.g. for robot controllers~\cite{degrave2016differentiable}.

The parameters of such programs are easily understood: they describe
the pen strokes, geometric and physical properties of the simulated
system, inferred from the given images. Framing information which
describes the processes underlying the generative act could be
generated; for instance:

\begin{itemize}
\item Decision trees: ``Class {\em x} was chosen because {\em y} was
  greater than {\em z} and ...''
\item Random forests: ``Class {\em x} was chosen because most trees
  looking at {\em y} and {\em z} voted for it, ...''
\item Handwriting recognition: ``{\em B} was chosen because we saw one
  long stroke and two curved strokes''
\item The robot controller: ``The motor was engaged because the
  pendulum had begun to rotate...''
\end{itemize}

However, such specificity makes these implementations
unsuitable for anything else. The choice of such high-level, task-specific
components is performed by the user, and encodes some of their domain knowledge
into the structure of the solution, such that it doesn't have to be learned.
This is similar to how logics like HOL can directly represent high-level
concepts (e.g. natural numbers and sets) but whose proof methods have limited
reusability due to the difficulty of automating such high-level reasoning.

At the other end of the spectrum we have differentiable programs with general
purpose architectures, like \emph{neural networks} (capable of universal
function approximation~\cite{funahashi1989approximate}). These are compositions
of a large number of identical sub-expressions (``neurons''), whose parameters
(``weights'') scale the input values, and hence the contribution of each
sub-expression to the whole. Such architectures encode essentially no domain
knowledge, requiring much larger training sets than task-specific algorithms in
order to ``learn'' these details.  The behaviour of general purpose models
depends so heavily on their (many) parameters, that understanding their
behaviour becomes difficult and they are often treated as inscrutable ``black
boxes''; akin to the large (un)SAT proof described above.

Understanding exactly how these programs make their decisions is an
active area of research, known as \emph{explainable artificial
  intelligence}
(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable},
with popular approaches such as saliency maps~\cite{simonyan2013deep}
tending to reverse-engineer the factors which lead to a model's
decisions (judging the saliency of each pixel based on how strongly it
would effect the output prediction if adjusted). These methods appear
intuitive, e.g. producing visualisations highlighting a particular
object in a scene as the reason for its classification; akin to the
proof objects emitted by theorem provers. Yet this can obscure the
difficulty of interpreting such high-dimensional decision
boundaries. In particular, these explanations or justifications can be
fundamentally altered by imperceptible adjustments to the
input~\cite{ghorbani2017interpretation}. Similar adjustments can also
change a model's output, leading to the field of \emph{adversarial
  machine learning}~\cite{goodfellow2014explaining}; adjustments to
even a single pixel can not only cause a system to mislabel an input,
but to give high confidence to its erroneous
result~\cite{su2019one}. Such fragility and opaqueness has important
implications as these techniques begin to have decision making
applications in life-critical domains such as automotive
control~\cite{bojarski2016end} and medical
diagnosis~\cite{esteva2017dermatologist}.

\begin{quote}
%\noindent\fbox{%
%    \parbox{\textwidth}{%
{\bf Recommendation 2:} Enhance your system with framing capabilities.

%}%
%}
\end{quote}

\subsection{The ``forgoing understandability'' approach}
It may be the case that, given the increase in power, generality and
predictiveness that ML approaches give, we decide to forgo
understandability in science. As a community we would be in a unique
position to develop thinking on this, and to answer questions such as
whether we should try to replace understandability with something
else. Previous work such as \cite{colton:2015} identifies stakeholder
groups in CC research and practice: this can be extended to scientists
and mathematicians to ensure that we develop in directions which will
be fruitful and useful to society.

\begin{quote}
%\noindent\fbox{%
 %   \parbox{\textwidth}{%

  {\bf Recommendation 3:} Produce philosophical work on what
  computational creativity should mean, and what science done with
  computers should entail.

%}%
%}
\end{quote}

\section{Concluding Remarks}

%\subsection{shakeup - we have to do more than generation}
%[move]
Deep learning and ML are making inroads everywhere: generative arts,
processors, windfarms, Go, machine vision, and so on, and we need to
consider as a community where this leaves us. Our suggestion in this
paper is to focus on science and mathematics, where we have much to
contribute.

%\subsection{science is actually a more natural domain for cc than the
 % arts}
%AI is a good idea in this domain

Of course, both science and CC are fluid disciplines and will evolve
as technology and culture change. One possible solution to the
problems described here would be to adapt our view of what is
valuable, and what is an artefact. For instance, could a neural
network itself be considered to be a scientific discovery?  It may be
that AI systems become objects of study in the same way as the human
brain is currently an object of study, with methods and approaches
from neural science, psychology, cognitive science and so on employed
to understand an AI system and its behaviour and interactions.

People are not naturally good at science. The history of science and
scientific methodology, the length of time it takes to train a
scientist and the high number of published research findings in
science which are considered to be false or
sub-standard\footnote{Meta-scientific studies suggest that 85\% of
  biomedical research efforts are wasted \cite{macleod} and 90\% of
  respondents to a recent survey in Nature agreed that there is a
  `reproducibility crisis' \cite{baker} (see \cite{munafo,ioannidis}
  for further details).} all hint at the difficulty of the scientific
enterprise. This is partially due to political and institutional
factors such as pressure to publish, conflicts of interest and a
culture which is often more competitive than collaborative; but also
partially due to the constant battle to avoid the large number of
cognitive biases that adversely affect our reasoning and judgements
\cite{haselton,sutherland}. On the other hand, the arts -- while also
difficult to do well -- do not usually go against our natural way of
thinking, and can be seen as a celebration of our humanity.  In many
ways science should be an obvious application domain for computational
creativity.  This paper is a call to arms for the whole CC community,
to apply their systems to scientific and mathematical domains, to
enhance their systems with framing functionality, and to produce
philosophical work on new directions in our field.

%The psychology of AI.

%\subsection{The psychology of AI}

% Our hope is that we as a community design a research space which is
% friendly and welcoming and positively encouraging of applications to
% science and mathematics.

\bibliographystyle{iccc}
\bibliography{biblio}
\end{document}

***

where statistical methods are used to
make predictions or decisions about data, often without human
involvement~\cite{alpaydin2014introduction}.

In particular there is a

A similar problem of understandability has emerged in the field of

*******

Automated reasoning ....

****************
The problem of making machine-generated artefacts more understandable
to human users is quite clear in the field of

***********

**

bit from Chris.

***

 Crisis in science work.

 Are we in the last throes where humans understand science? Right now
 we have to chose between understandability and power,
 understandability and generality, understandability and predictive
 accuracy. ...

I need to do/know/predict X. I don't need to understand it.

phil of science here.

Part three focuses upon specific academic disciplines including; arts
and science, humanities, natural sciences and mathematics, social
sciences, the arts, professional and occupational education. Part four
describes directions for reform including; teaching across the
curriculum, interdisciplinary studies, internationalizing the
curriculum, transforming the curriculum through diversity, creating
learning communities and using technology.

Eisner, E., & Powell, K. (2002). Special Series on Arts-Based Educational Research: Art In
Science?. Curriculum Inquiry, 32(2), 131-159.

questioned the notion that art and science belong in different worlds, and noted synergies
across the different disciplines.

social scientists who were fellows at the Center for Advanced Study in the Behavioral Sciences to secure insight into the role that artistry might play in the course of their work.

Maeda, J. (2013). STEM + Art = STEAM. STEAM Journal, 1(1).

### 2019-03-05 UnderstandabilityPaper/iccc-science.tex
My own submitted edits

@@
 mathematics -- a ``mathematics social machine'' \cite{martin:2013} --
-further complicates matters. \emph{Automated theorem proving} is the
+further complicates matters. Take \emph{automated theorem proving}, the
 task of deciding whether a given formal statement follows from a given
@@
 
-Both of these issues can be mitigated by having the system instead
+This can be mitigated by having the system instead
 generate a \emph{proof object}: a formal argument for \emph{why} a
@@
 how it was created, thus avoiding the need to trust or verify the
-(potentially complicated) search and generation procedures. The
-\emph{de Bruijn criterion}~\cite{barendregt2005challenge} describes
-theorem provers which produce proof objects that are trivial to check
-by independent \emph{proof checker} programs (which are themselves
-easily verified, due to their simplicity); examples are
-Coq~\cite{barras1997coq} and Isabelle~\cite{nipkow2002isabelle}.
+(potentially complicated) search and generation procedures. Theorem provers
+which produce proof objects that are trivial to check by independent
+\emph{proof checker} programs (which are themselves
+easily verified, due to their simplicity) satisfy the
+\emph{de Bruijn criterion}~\cite{barendregt2005challenge}; examples are
+Coq~\cite{barras1997coq} and Isabelle/HOL~\cite{nipkow2002isabelle}.
 
@@
 generating framing information alongside an artefact, problem solution
-or new data pattern. We foresee this being an increasingly important
-area of research in CC, with an increasing level of sophistication:
+or new data pattern (proof objects can be seen as a limited form of framing
+information). We foresee this being an increasingly important area of
+research in CC, with an increasing level of sophistication:
 from explanation to justification to argument and dialogue with a user
 about the value, method of production, motivation etc behind
-output. How framing information should be developed is a research
-programme in its own right. For now, we discuss greater and lesser
-understandability in terms of describing the processes underlying the
-generative act and consider these for ML approaches.
+output. How framing information should be developed is a research programme in
+its own right. For now, we discuss greater and lesser understandability in
+terms of describing the processes underlying the generative act and consider
+these for ML approaches.
 
@@
 parameters, and are chosen based on how efficiently they separate the
-classes given in the training data.  Decision trees are nominally
-understandable, since their behaviour on a given input traces a single
-path through these queries; yet they perform poorly compared to other
-ML algorithms. The \emph{random forest} approach yields better results
-by combining many decision trees and having them vote on the overall
+classes given in the training data.  Decision trees perform poorly compared to
+other ML algorithms, but are nominally understandable since their behaviour on a
+given input traces a single path through these queries, which could be turned
+into framing information such as ``Class {\em x} was chosen because {\em y} was
+greater than {\em z}...''.  The \emph{random forest} approach gives better
+performance by combining many decision trees and having them vote on the overall
 outcome~\cite{randomforests}, although such ensemble behaviour is more
-difficult to reason about than that of a single tree.
+difficult to reason about than that of a single tree, and is hence harder to
+frame in an understandable way. One approach might be to find patterns in the
+votes, such as ``Class {\em x} was chosen because most trees looking at features
+{\em y} and {\em z} voted for it''.
 
@@
 parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
-respectively) to minimise arbitrary loss functions (e.g. the squared output
-error on a training data set).
+respectively) to minimise arbitrary loss functions (e.g. output error for the
+training data).
 
@@
 of a model and its understandability becomes clear. Task-specific
-architectures require fewer parameters than general-purpose approaches
-and perform well with little training data, for example hand-written
-characters can be classified based on a single example if we allow our
-model to assume the given images are generated by pen
-strokes~\cite{lake2015human}. Likewise we can infer parameters of a 3D
-scene (such as object position and lighting) from images if we embed a
-ray-tracer into our model~\cite{li2018differentiable}; embedding a
-physics engine can provide predictions about these scenes, which are
-useful e.g. for robot controllers~\cite{degrave2016differentiable}.
+architectures require fewer parameters than general-purpose approaches, perform
+well with little training data, and are amenable to descriptive framing
+information. For example, hand-written characters can be classified based on a
+single example if we allow our model to assume the given images are generated by
+pen strokes~\cite{lake2015human}, and these models may allow descriptions such
+as ``Character {\em c} was chosen because there appear to be {\em x} long
+strokes, {\em y} curved strokes, etc''. Likewise the parameters of a 3D scene
+(such as object position and lighting) can be inferred from images if a
+ray-tracer is embedded in the model~\cite{li2018differentiable}; embedding a
+physics engine enables predictions about these scenes, which are useful e.g. for
+robot controllers~\cite{degrave2016differentiable}. Whilst more complicated than
+the previous examples, such a controller could (in principle) justify its
+actions based on interpretations of the model, such as ``The motor was engaged
+because the pendulum appeared to be falling to the left''.
 
-The parameters of such programs are easily understood: they describe
-the pen strokes, geometric and physical properties of the simulated
-system, inferred from the given images. Framing information which
-describes the processes underlying the generative act could be
-generated; for instance:
+However, the specificity that makes these implementations understandable also
+makes them unsuitable for any other task. The choice of such high-level,
+task-specific components is performed by the user, and encodes some of their
+domain knowledge into the structure of the solution, such that it doesn't have
+to be learned. This is similar to how high-level logics can represent relevant
+domain concepts (like natural numbers and sets), yet proof methods making use of
+this have limited reusability due to the difficulty of automating such
+high-level reasoning.
 
-\begin{itemize}
-\item Decision trees: ``Class {\em x} was chosen because {\em y} was
-  greater than {\em z} and ...''
-\item Random forests: ``Class {\em x} was chosen because most trees
-  looking at {\em y} and {\em z} voted for it, ...''
-\item Handwriting recognition: ``{\em B} was chosen because we saw one
-  long stroke and two curved strokes''
-\item The robot controller: ``The motor was engaged because the
-  pendulum had begun to rotate...''
-\end{itemize}
-
-However, such specificity makes these implementations
-unsuitable for anything else. The choice of such high-level, task-specific
-components is performed by the user, and encodes some of their domain knowledge
-into the structure of the solution, such that it doesn't have to be learned.
-This is similar to how logics like HOL can directly represent high-level
-concepts (e.g. natural numbers and sets) but whose proof methods have limited
-reusability due to the difficulty of automating such high-level reasoning.
-
-At the other end of the spectrum we have differentiable programs with general
-purpose architectures, like \emph{neural networks} (capable of universal
-function approximation~\cite{funahashi1989approximate}). These are compositions
-of a large number of identical sub-expressions (``neurons''), whose parameters
-(``weights'') scale the input values, and hence the contribution of each
+At the other end of the spectrum are general purpose architectures, like
+\emph{neural networks} (differentiable programs capable of universal function
+approximation~\cite{funahashi1989approximate}). These are compositions of a
+large number of identical sub-expressions (``neurons''), whose parameters
+(``weights'') scale their input values, and hence the contribution of each
 sub-expression to the whole. Such architectures encode essentially no domain
 knowledge, requiring much larger training sets than task-specific algorithms in
-order to ``learn'' these details.  The behaviour of general purpose models
-depends so heavily on their (many) parameters, that understanding their
-behaviour becomes difficult and they are often treated as inscrutable ``black
-boxes''; akin to the large (un)SAT proof described above.
+order to ``learn'' these details.  So much of these general purpose models'
+behaviour comes from tuning their (many) parameters, that understanding or
+describing their high-level behaviour is difficult; indeed they are often
+treated as inscrutable ``black boxes'', akin to the large (un)SAT proof
+described above.
 
-Understanding exactly how these programs make their decisions is an
+Understanding exactly how such programs make their decisions is an
 active area of research, known as \emph{explainable artificial
   intelligence}
-(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable},
-with popular approaches such as saliency maps~\cite{simonyan2013deep}
-tending to reverse-engineer the factors which lead to a model's
-decisions (judging the saliency of each pixel based on how strongly it
-would effect the output prediction if adjusted). These methods appear
-intuitive, e.g. producing visualisations highlighting a particular
-object in a scene as the reason for its classification; akin to the
-proof objects emitted by theorem provers. Yet this can obscure the
-difficulty of interpreting such high-dimensional decision
-boundaries. In particular, these explanations or justifications can be
+(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable}.
+Saliency maps are a popular form of framing information~\cite{simonyan2013deep},
+which reverse-engineer the factors which lead to a model's decisions (for
+example judging the saliency of input pixels by how strongly they each effect
+the output prediction if adjusted). These methods appear intuitive, e.g.
+producing visualisations highlighting a particular object in a scene as the
+reason for its classification; yet this can obscure the difficulty of
+interpreting such high-dimensional decision boundaries. In particular,
+reasonable justifications (such as classifying an image as a butterfly, with
+high saliency given to those pixels which show the butterfly) can be
 fundamentally altered by imperceptible adjustments to the
-input~\cite{ghorbani2017interpretation}. Similar adjustments can also
-change a model's output, leading to the field of \emph{adversarial
-  machine learning}~\cite{goodfellow2014explaining}; adjustments to
-even a single pixel can not only cause a system to mislabel an input,
-but to give high confidence to its erroneous
-result~\cite{su2019one}. Such fragility and opaqueness has important
-implications as these techniques begin to have decision making
-applications in life-critical domains such as automotive
-control~\cite{bojarski2016end} and medical
-diagnosis~\cite{esteva2017dermatologist}.
+input~\cite{ghorbani2017interpretation} (in this case choosing butterfly based
+only on the background vegetation). Similar adjustments can also change a
+model's output, leading to the field of \emph{adversarial machine
+  learning}~\cite{goodfellow2014explaining}; adjustments to even a single pixel
+can not only cause a system to mislabel an input, but to give high confidence to
+its erroneous result~\cite{su2019one}. Whilst much attention has been focused on
+making the outputs of ML systems more accurate and robust, there is also a need
+for framing information which explains more, is more understandable to users and
+less prone to misinterpretation.
 

### 2019-03-14 Bucketing/intro.tex
A bunch of bucketing metadata, and new intro

@@
 
-Functional programming claims to be more ``mathematically tractable'' than other
-``conventional'' approaches, due to properties like \emph{referential
-transparency} and \emph{immutability}, which allow equational reasoning without
-many of the complications caused by side effects and strict evaluation
-order~\cite{hughes1989functional}. Languages such as Haskell combine these
-advantages with static type systems to restrict the possible values of every
-expression, making programs amenable to powerful analysis tools such as property
-checking~\cite{claessen2011quickcheck} and theorem
-proving~\cite{rosen2012proving}.
+\emph{Theory exploration} (TE) describes an open-ended search for
+interesting properties (conjectured or proven) of a given set of formal
+definitions, such as a mathematical theory, a scientific model or a software
+library. It has been applied to generating ``background lemmas'' in the
+interactive proof assistant \textsc{Isabelle/HOL}~TODO{CITE ISACOSY AND HIPSTER}
+and for automated theorem proving~TODO{CITE PAPER WHICH SHOWS IMPROVED PROVING},
+as well as for generating specifications for (pure functional) programming
+libraries~TODO{CITE QUICKSPEC 1, 2 AND SPECULATE}.
 
-Whilst such tools can support or refute user-provided properties, we are
-interested in the related problem of \emph{discovering} novel properties
-automatically; a task addressed by the field of \emph{Mathematical Theory
-  Exploration}. Whilst clever algorithms are used to guide their search,
-existing MTE tools, including those applied to Haskell, ultimately rely on
-enumerating all (well-typed) expressions. This limits their usefulness, as it is
-only feasible to explore a small collection of functions at a time, relying on
-the user to select promising combinations. We investigate the use of machine
-learning methods to perform this selection automatically, removing the need for
-user intervention whilst maintaining the ability to discover desirable
-properties.
+Despite the clever search strategies employed by these tools, they all rely on
+enumerating combinations of definitions in some way, which can cause infeasible
+running times on larger inputs (in our experience, a few dozen definitions). To
+avoid this, users of these systems must carefully select only small subsets of
+their definitions to explore at a time.
+
+We dub such selection the \emph{bucketing problem}, and analyse its effect on
+the running time of theory exploration tools on a particular problem set, and on
+the quality of the statements they are able to generate. An automated solution
+to the bucketing problem would accept arbitrarily many definitions, and
+efficiently select sub-sets which are small enough to reasonably explore whilst
+still enabling interesting properties to be discovered.
 
@@
 \begin{enumerate}
-  \item The application of machine learning algorithms to theory exploration,
-    for intelligently discovering interesting sub-sets of Haskell libraries,
-    which are more tractable to explore.
-  \item A novel feature extraction method for transforming Haskell expressions
-    into a form amenable to off-the-shelf learning algorithms.
-  \item An implementation of these feature extraction and theory exploration
-    approaches.
-  \item A comparison of our methods with existing approaches, both for theory
-    exploration in Haskell, and for machine learning in other languages.
+  \item Defining the \emph{bucketing problem} for theory exploration.
+  \item Statistical analysis of theory exploration performance on a large
+    problem set, derived from existing theorem proving problem sets.
+  \item Empirical analysis of the effect of bucketing on the quantity of
+    desirable properties attainable by theory exploration.
+  \item Experimental application of unsupervised machine learning to the
+    bucketing problem.
+  % \item A novel feature extraction method for transforming Haskell expressions
+  %   into a form amenable to off-the-shelf learning algorithms.
+  % \item An implementation of these feature extraction and theory exploration
+  %   approaches.
+  % \item A comparison of our methods with existing approaches, both for theory
+  %   exploration in Haskell, and for machine learning in other languages.
 \end{enumerate}
 
-We describe our treatment of Haskell expressions in \S~\ref{sec:haskell} and
-describe the \qspec{} theory exploration system we build upon in
-\S~\ref{sec:theoryexploration}. We discuss our contributions in more depth in
-\S~\ref{sec:contributions}, and provide implementation details in
-\S~\ref{sec:implementation}. A variety of related work is surveyed in
-\S~\ref{sec:related}, we briefly evaluate our implementations in
-\S~\ref{sec:evaluation} and give several potential directions for future
-research in \S~\ref{sec:future}.
+We describe theory exploration in more detail, as well as the \qspec{} tool we
+use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
+bucketing problem in \S~\ref{sec:bucketing} and measure its effect on theory
+exploration systems in \S~\ref{sec:implementation}. We attempt (unsuccessfully)
+to apply machine learning to the bucketing problem in \S~\ref{sec:clustering}.
+A variety of related work is surveyed in \S~\ref{sec:related} and we conclude in
+\S~\ref{sec:conclusion} with potential directions for future research.

### 2019-03-14 Bucketing/conclusion.tex
Change import, now that filename has changed

--- 2016-03-02 CICM2016/future.tex
@@
-\section{Future Work}
-\label{sec:future}
+\section{Conclusion and Future Work}
+\label{sec:conclusion}
 
-Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made.
-Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further.
+Our use of clustering to pre-process \qspec{} signatures has required many
+decisions and tradeoffs to be made. Hence our approach is just one possibility
+out of many alternatives which could be investigated to push this work further.
 
-The most obvious next step is to incorporate types.
-Types contain valuable information about an expression, and would allow us to distinguish between constructors.
-Since our algorithm closely follows that of ML4PG, which \emph{does} support types, the only barrier is the practical issue of propagating annotations through all of the Core definitions.
+The most obvious next step is to incorporate types. Types contain valuable
+information about an expression, and would allow us to distinguish between
+constructors. Since our algorithm closely follows that of ML4PG, which
+\emph{does} support types, the only barrier is the practical issue of
+propagating annotations through all of the Core definitions.
 
-More speculative directions include the use of \emph{learned} representations, rather than our hand-crafted features \cite{bengio2013representation}.
-This would provide an interesting comparison, as well as being more robust in the face of language evolution.
-Another intriguing possibility is to extend the recurrent nature of our algorithm to make use of the discovered properties during clustering; for example, by treating the discovered equations as rewrite rules to reduce the ASTs prior to feature extraction.
+More speculative directions include the use of \emph{learned} representations,
+rather than our hand-crafted features \cite{bengio2013representation}. This
+would provide an interesting comparison, as well as being more robust in the
+face of language evolution. Another intriguing possibility is to extend the
+recurrent nature of our algorithm to make use of the discovered properties
+during clustering; for example, by treating the discovered equations as rewrite
+rules to reduce the ASTs prior to feature extraction.

### 2019-03-14 Bucketing/intro.tex
Add citations
Also add the cite package so we can put multiple citations together

@@
 library. It has been applied to generating ``background lemmas'' in the
-interactive proof assistant \textsc{Isabelle/HOL}~TODO{CITE ISACOSY AND HIPSTER}
-and for automated theorem proving~TODO{CITE PAPER WHICH SHOWS IMPROVED PROVING},
-as well as for generating specifications for (pure functional) programming
-libraries~TODO{CITE QUICKSPEC 1, 2 AND SPECULATE}.
+interactive proof assistant
+\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and for automated
+theorem proving~\cite{claessen2013automating}, as well as for generating
+specifications for (pure functional) programming
+libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.
 

### 2019-04-08 Bucketing/background.tex
Add bucketing section to background
Bucketing changes

--- 2016-03-02 CICM2016/background.tex
@@
 \label{sec:background}
-
-\subsection{Haskell}
-\label{sec:haskell}
-
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
-
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and \hspec{} \cite{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
-
-Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-To avoid the relatively large and complex syntax of Haskell proper, we will focus on an intermediate representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described in Appendix \ref{sec:core}.
-
-We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
-
-\begin{figure}
-  \begin{haskell}\begin{verbatim}
-data Bool = True | False
-
-data Nat = Z | S Nat
-
-not :: Bool -> Bool
-not True  = False
-not False = True
-
-odd :: Nat -> Bool
-odd    Z  = False
-odd (S n) = even n
-
-even :: Nat -> Bool
-even    Z  = True
-even (S n) = odd n\end{verbatim}
-  \end{haskell}
-  \caption{Haskell datatypes for booleans and natural numbers, followed by some simple function definitions. Note that \hs{odd} and \hs{even} are mutually recursive.}
-  \label{fig:haskellexample}
-\end{figure}
-
-\begin{figure}
-  \begin{small}
-    \underline{\texttt{not}}
-    \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "True")  (Var (Constructor "False")))
-              (Alt (DataAlt "False") (Var (Constructor "True"))))
-    \end{verbatim}
-
-    \underline{\texttt{odd}}
-    \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "False")))
-              (Alt (DataAlt "S") (App (Var (Global "even"))
-                                      (Var (Local  "n")))
-                                 "n"))
-    \end{verbatim}
-
-    \underline{\texttt{even}}
-    \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "True")))
-              (Alt (DataAlt "S") (App (Var (Global "odd"))
-                                      (Var (Local  "n")))
-                                 "n"))
-    \end{verbatim}
-  \end{small}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into our Core syntax, with fresh variables chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.
-Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-The \texttt{"b"} variables are introduced to preserve sharing.}
-  \label{fig:coreexample}
-\end{figure}
-
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with common definitions for boolean and natural number datatypes (analogous to those in Haskell's standard library).
-The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
-Although Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
-It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
 
@@
 
-We consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
+Theory exploration is the task of taking a \emph{signature} of definitions in
+some formal system (for example a programming language) and automatically
+generating a set of formal statements (properties) involving those
+definitions. These may be conjectures or theorems (proven either by sending
+conjectures to an automated theorem prover, or by having the generating
+procedure proceed in logically sound steps); in either case, these statements
+must also be ``interesting'' in some way. It is beyond the scope of this paper
+to define what makes a mathematical statement ``interesting'', but this problem
+has been tackled extensively in the literature~\cite{colton2000notion}.
 
-We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about Haskell code.
-These may be fed into another tool, such as \hspec{}, for subsequent proving.
-These conjectures are arrived at through the following stages:
+We will follow the tradition of prior practitioners and use an existing
+corpus of theorems as a \emph{ground truth} against which to compare results.
+Prior work has used libraries of the \textsc{Isabelle/HOL} proof assistant as a
+ground
+truth~\cite{Montano-Rivas.McCasland.Dixon.ea:2012,johansson2009isacosy,claessen2013automating},
+but these only contained a few definitions and theorems each. To analyse
+behaviour on a larger scale, as required to study the effects of clustering, we
+have chosen the \emph{Tons of Inductive Problems} (TIP) problem set for
+automated theorem provers~\cite{claessen2015tip}.  Reasons for choosing TIP
+include:
 
-\iffalse TODO: Make this more formal?
+\begin{itemize}
+\item Each problem in TIP clearly distinguishes between the statement to be
+  proved and the definitions involved, such that the two can be easily and
+  meaningfully separated to act as inputs and desired outputs in a TE setting.
+  This would not be the case for problem sets like
+  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
+  functions whose behaviour is \emph{implicit} in the logical structure of the
+  theorem statement but not separately \emph{defined}.
+\item The TIP problem set includes a tool to translate from the language it
+  uses to define problems (a variant of SMTLIB) into various other languages
+  including Haskell and Isabelle, for which TE tools already exist.
+\item TIP problems are relevant to the domains we are interested in, including
+  pure functional programming, higher-order functions and inductively defined
+  datatypes. This makes TIP more attractive than problem sets for first-order
+  languages/logics such as TPTP~\cite{sutcliffe2009tptp}.
+\item Many of TIP's problem statements re-use the same definitions, which makes
+  it more useful as a ground-truth than a mixture of unrelated problems. This is
+  because non-membership in the ground truth will be treated as being
+  ``uninteresting''.
+\end{itemize}
+
+All existing theory exploration systems rely on brute force enumeration to some
+degree. We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
+Haskell code. These may be fed into another tool, such as \hspec{}, for
+subsequent proving. We have used version 1 of \qspec{} in our experiments, due
+to its availability, stability and tooling integration, and we will drop the
+version number from now on. At the time of writing there is a \qspec{} version 2
+available, which has a much improved generation procedure that is significantly
+faster than its predecessor. However, it still scales poorly in our experience,
+and has less integration with the tooling we have used. We hence leave analysis
+of \qspec{} 2 as future work, pending the necessary infrastructure changes that
+would require.
+
+\qspec{} (version 1) is written in the Haskell programming language, and works
+with definitions which are also written in Haskell. Some simple Haskell
+definitions are shown in Figure~\ref{fig:haskellexample}. The following
+procedure is used to generate conjectures, which are both plausible (due to
+testing on many examples) and potentially interesting (due to being mutually
+irreducible):
+
+\iffalse % TODO{2019-03-15}: Make this more formal?
  V \in Var
@@
 
+\begin{figure}
+  \centering
+  \begin{minted}{haskell}
+    -- A datatype with two constructors
+    data Bool = True | False
+
+    -- A recursive datatype (S requires a Nat as argument)
+    data Nat = Z | S Nat
+
+    -- A function turning a Bool into a Bool
+    not :: Bool -> Bool
+    not True  = False
+    not False = True
+
+    -- A pair of mutually recursive functions
+
+    odd :: Nat -> Bool
+    odd    Z  = False
+    odd (S n) = even n
+
+    even :: Nat -> Bool
+    even    Z  = True
+    even (S n) = odd n
+  \end{minted}
+  \caption{Haskell datatypes for booleans and natural numbers, followed by some
+    simple function definitions (with type annotations).}
+  \label{fig:haskellexample}
+\end{figure}
+
 \begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
-To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
-  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item The equivalence of terms in each class is tested using \qcheck{}: variables are instantiated to particular values, generated randomly, and the resulting closed expressions are evaluated and compared for equality.
-  \item If a class is found to have non-equal members, it is split up to separate those members.
-  \item The previous steps of testing and splitting are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
-  \item For each class, one member is selected and equations are conjectured that it is equal to each of the other members.
+\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
+  generates a list $terms$ containing the constants (including functions) from
+  $\Sigma$, the variables from $V$ and type-correct function applications
+  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
+  is finite, function applications are only nested up to a specified depth (by
+  default, 3).
+\item The elements of $terms$ are grouped into equivalence classes, based on
+  their type.
+\item The equivalence of terms in each class is tested using \qcheck{}:
+  variables are instantiated to particular values, generated randomly, and the
+  resulting closed expressions are evaluated and compared for equality.
+\item If a class is found to have non-equal members, it is split up to separate
+  those members.
+\item The previous steps of testing and splitting are repeated until the classes
+  stabilise (i.e. no differences have been observed for some specified number of
+  repetitions).
+\item For each class, one member is selected and equations are conjectured that
+  it is equal to each of the other members.
+\item A congruence closure algorithm is applied to these equations, to discard
+  any which are implied by the others.
 \end{enumerate}
 
-Such conjectures can be used in several ways: they can be presented to the user, sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
-As an example, giving the definitions from Figure \ref{fig:haskellexample} to \qspec{}.
-along with some suitable \hs{Nat} and \hs{Bool} variables, produces equations including \hs{not (not a) == a} and \hs{not (odd x) == even x}.
+As an example, giving the definitions from Figure~\ref{fig:haskellexample} to
+\qspec{}, along with suitable comparison functions, random data generators and
+variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
+class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
+\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
+(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
+\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
+them up, and repeat until no more splits are found. Equations are then generated
+between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
+\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
+to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
+instance of the more general \hs{b == not (not b)}). The remaining equations
+(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
+output, either for the user to peruse or for another system to process, like the
+\textsc{HipSpec} automated theorem prover.
 
-\iffalse
-As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
+Note that we must be able to generate values of a type in order to include
+variables of that type. Also, only equivalence classes whose element type can be
+compared are split and conjectured about in this way; in particular, functions
+(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
+equivalence class, but they will not be subject to testing and comparison since
+Haskell has no generic notion of function equality~\footnote{In
+  fact \qspec{} compares expressions using a total ordering rather than equality;
+  this is even more restrictive than requiring a notion of equality, but reduces
+  the number of required comparisons}. \qspec{} can hence conjecture that two
+functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
+that the functions themselves are equal (e.g. \hs{f == g}).
 
-\begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{odd}, \texttt{even}\}
-\end{align*}
+Although complete, this enumeration approach is wasteful: many terms are
+unlikely to appear in theorems, which requires careful choice by the user of
+what to include in the signature. For example, we know that addition and
+multiplication are closely related, and hence obey many algebraic laws. Our
+machine learning technique aims to predict these kinds of relations between
+functions, so we can create small signatures which can be explored quickly, yet
+have the potential to give rise to many equations.
 
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
-
-\begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{odd Z},\ \texttt{odd}\ a,\ \dots ]
-\end{align*}
-
-Notice that functions such as \hs{odd} and \hs{even} are valid terms, despite not being applied to any arguments.
-In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.
-
-These terms will be grouped into four classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations such as $odd (S a) = even a$ and $even (S a) = odd a$ are conjectured.
-\fi
-
-Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which can be explored quickly, yet have the potential to give rise to many equations.
-
-\iffalse
-TODO
-In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-\fi
+\iffalse TODO{2019-03-15} In fact, there are similarities between the way a TE
+system like \qspec{} can generalise from checking \emph{particular} properties
+to \emph{inventing} new ones, and the way counterexample finders like \qcheck{}
+can generalise from testing \emph{particular} expressions to \emph{inventing}
+expressions to test. One of our aims is to understand the implications of this
+generalisation, the lessons that each can learn from the other's approach to
+term generation, and the consequences for testing and QA in general.  \fi
 
@@
 365]{mardia1979multivariate} of clustering $n$ data points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
+
+\subsection{The Clustering Problem}
+\label{sec:clustering}
+
+% TODO{2019-03-15}

### 2019-04-08 Bucketing/haskell.tex
Bucketing changes

--- 2016-03-02 CICM2016/background.tex
@@
-\section{Background}
-\label{sec:background}
-
 \subsection{Haskell}
@@
 
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
+We decided to focus on the Haskell programming language since it has direct
+applications in industry, a large selection of library code for testing and
+mature, state-of-the-art MTE tools including \qspec{}~\cite{QuickSpec},
+\hspec{}~\cite{claessen2013automating} and
+\textsc{Speculate}~\cite{braquehais2017speculate}. Our overall approach is
+agnostic to the underlying system, so could also be applied to other targets
+such as the Isabelle/HOL, which has the
+\textsc{IsaCoSy}~\cite{johansson2009isacosy}, \textsc{IsaScheme}~\cite{omar} and
+more recent \textsc{Hipster}~\cite{Hipster} MTE tools. However, since
+\textsc{Hipster} works by translating to Haskell and invoking \hspec{}, it seems
+that improving the Haskell tools will give the most benefit.
 
-We decided to focus on theory exploration in the Haskell programming language as it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and \hspec{} \cite{claessen2013automating}).
-This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \cite{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.
+Like most functional programming languages, Haskell builds upon
+$\lambda$-calculus, with extra features such as a strong type system and
+``syntactic sugar'' to improve readability. To avoid the relatively large and
+complex syntax of Haskell proper, we will focus on an intermediate
+representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described
+in Appendix \ref{sec:core}.
 
-Like most functional programming languages, Haskell builds upon $\lambda$-calculus, with extra features such as a strong type system and ``syntactic sugar'' to improve readability.
-To avoid the relatively large and complex syntax of Haskell proper, we will focus on an intermediate representation of the \textsc{GHC} compiler known as \emph{GHC Core}, described in Appendix \ref{sec:core}.
-
-We will use quoted strings to denote names and literals, e.g.
-\texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.
-
-\begin{figure}
-  \begin{haskell}\begin{verbatim}
-data Bool = True | False
-
-data Nat = Z | S Nat
-
-not :: Bool -> Bool
-not True  = False
-not False = True
-
-odd :: Nat -> Bool
-odd    Z  = False
-odd (S n) = even n
-
-even :: Nat -> Bool
-even    Z  = True
-even (S n) = odd n\end{verbatim}
-  \end{haskell}
-  \caption{Haskell datatypes for booleans and natural numbers, followed by some simple function definitions. Note that \hs{odd} and \hs{even} are mutually recursive.}
-  \label{fig:haskellexample}
-\end{figure}
+We will use quoted strings to denote names and literals, e.g. \texttt{Local
+  "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr
+  "quux"} and \texttt{LitNum "42"}, and require only that they can be compared
+for equality.
 
@@
   \end{small}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into our Core syntax, with fresh variables chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.
-Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions.
-The \texttt{"b"} variables are introduced to preserve sharing.}
+  \caption{Translations of functions in Figure \ref{fig:haskellexample} into our
+    Core syntax, with fresh variables chosen arbitrarily as \texttt{"a"},
+    \texttt{"b"}, etc. Notice the introduction of explicit $\lambda$
+    abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent
+    piecewise definitions. The \texttt{"b"} variables are introduced to preserve
+    sharing.}
   \label{fig:coreexample}
@@
 
-Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with common definitions for boolean and natural number datatypes (analogous to those in Haskell's standard library).
-The translation to our Core syntax is routine, and shown in Figure \ref{fig:coreexample}.
-Although Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the particular identifiers used.
-It is this close correspondence which allows us to analyse Core expressions in place of their more complicated Haskell source.
-
-\subsection{Theory Exploration}
-\label{sec:theoryexploration}
-
-We consider the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user.
-The method of conjecture generation is a key characteristic of any theory exploration system, although all existing implementations rely on brute force enumeration to some degree.
-
-We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about Haskell code.
-These may be fed into another tool, such as \hspec{}, for subsequent proving.
-These conjectures are arrived at through the following stages:
-
-\iffalse TODO: Make this more formal?
- V \in Var
- F \in Fun
- T \in Term
- T ::= V | F | T1 T2
-
- Term ::= VAR | Const | Fun (Term)
-or
- Term t ::= x | f | t t'
-\fi
-
-\begin{enumerate}
-  \item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{} generates a list $terms$ containing the constants (including functions) from $\Sigma$, the variables from $V$ and type-correct function applications $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little awkward; maybe use the above notation?
-\fi.
-To ensure the list is finite, function applications are only nested up to a specified depth (by default, 3).
-  \item The elements of $terms$ are grouped into equivalence classes, based on their type.
-  \item The equivalence of terms in each class is tested using \qcheck{}: variables are instantiated to particular values, generated randomly, and the resulting closed expressions are evaluated and compared for equality.
-  \item If a class is found to have non-equal members, it is split up to separate those members.
-  \item The previous steps of testing and splitting are repeated until the classes stabilise (i.e. no differences have been observed for some specified number of repetitions).
-  \item For each class, one member is selected and equations are conjectured that it is equal to each of the other members.
-\end{enumerate}
-
-Such conjectures can be used in several ways: they can be presented to the user, sent to a more rigorous system like \hspec{} for proving, or even serve as a background theory for an automated theorem prover \cite{claessen2013automating}.
-As an example, giving the definitions from Figure \ref{fig:haskellexample} to \qspec{}.
-along with some suitable \hs{Nat} and \hs{Bool} variables, produces equations including \hs{not (not a) == a} and \hs{not (odd x) == even x}.
-
-\iffalse
-As an example, we can consider a simple signature containing the expressions from Figure \ref{fig:haskellexample}:
-
-\begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{odd}, \texttt{even}\}
-\end{align*}
-
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:
-
-\begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{odd Z},\ \texttt{odd}\ a,\ \dots ]
-\end{align*}
-
-Notice that functions such as \hs{odd} and \hs{even} are valid terms, despite not being applied to any arguments.
-In addition, Haskell curries multi-argument functions, allowing them to be applied to one argument at a time, as used in the construction of $terms$.
-
-These terms will be grouped into four classes, one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}.
-As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations such as $odd (S a) = even a$ and $even (S a) = odd a$ are conjectured.
-\fi
-
-Although complete, this enumeration approach is wasteful: many terms are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature.
-For example, we know that addition and multiplication are closely related, and hence obey many algebraic laws.
-Our machine learning technique aims to predict these kinds of relations between functions, so we can create small signatures which can be explored quickly, yet have the potential to give rise to many equations.
-
-\iffalse
-TODO
-In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test.
-One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.
-\fi
-
-\subsection{Clustering}
-\label{sec:clustering}
-
-Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by only considering clauses deemed ``relevant'' to the problem \cite{meng2009lightweight}.
-
-Relevance is determined by comparing clauses to the target theorem, but theory exploration does not have such a distinguished term.
-Instead, we are interested in relationships between \emph{all} terms in a signature, and hence we need a different algorithm for considering the relevance of \emph{all terms} to \emph{all other terms}.
-
-A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
-
-Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, with randomly-selected input elements as the initial clusters.
-Rather than relying on the user to provide the number of clusters $k$, we use the ``rule of thumb'' given in \cite[pp.
-365]{mardia1979multivariate} of clustering $n$ data points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
+Figure \ref{fig:haskellexample} shows some simple Haskell function definitions,
+along with common definitions for boolean and natural number datatypes
+(analogous to those in Haskell's standard library). The translation to our Core
+syntax is routine, and shown in Figure \ref{fig:coreexample}. Although Core is
+more verbose, we can see that similar structure in the Haskell definitions gives
+rise to similar structure in the Core; for example, the definitions of \hs{odd}
+and \hs{even} are identical in both languages, except for the particular
+identifiers used. It is this close correspondence which allows us to analyse
+Core expressions in place of their more complicated Haskell source.

### 2019-04-08 Bucketing/intro.tex
Bucketing changes

--- 2019-03-14 Bucketing/intro.tex
@@
 
-\emph{Theory exploration} (TE) describes an open-ended search for
-interesting properties (conjectured or proven) of a given set of formal
-definitions, such as a mathematical theory, a scientific model or a software
-library. It has been applied to generating ``background lemmas'' in the
-interactive proof assistant \textsc{Isabelle/HOL}~TODO{CITE ISACOSY AND HIPSTER}
-and for automated theorem proving~TODO{CITE PAPER WHICH SHOWS IMPROVED PROVING},
-as well as for generating specifications for (pure functional) programming
-libraries~TODO{CITE QUICKSPEC 1, 2 AND SPECULATE}.
+\emph{(Mathematical) Theory Exploration} (TE) describes an open-ended search for
+interesting properties (proven or merely conjectured, depending on the approach)
+of a given set of formal definitions, such as a mathematical theory, a
+scientific model or a software library. It has been applied to generating
+``background lemmas'' for the interactive proof assistant
+\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and automated
+theorem provers~\cite{claessen2013automating}, as well as for generating
+specifications for (pure functional) programming
+libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.
 
@@
 
-We dub such selection the \emph{bucketing problem}, and analyse its effect on
-the running time of theory exploration tools on a particular problem set, and on
-the quality of the statements they are able to generate. An automated solution
-to the bucketing problem would accept arbitrarily many definitions, and
-efficiently select sub-sets which are small enough to reasonably explore whilst
-still enabling interesting properties to be discovered.
+We dub such selection the \emph{clustering problem} for theory exploration and
+analyse its effect on the running time of theory exploration tools on a
+particular problem set, and on the quality of the statements they are able to
+generate. An automated solution to this clustering problem would accept a large
+number of definitions and efficiently select sub-sets which are both small
+enough to reasonably explore, whilst still enabling interesting properties to be
+discovered.
 
@@
 \begin{enumerate}
-  \item Defining the \emph{bucketing problem} for theory exploration.
+  \item Defining the \emph{clustering problem} for theory exploration.
   \item Statistical analysis of theory exploration performance on a large
     problem set, derived from existing theorem proving problem sets.
-  \item Empirical analysis of the effect of bucketing on the quantity of
+  \item Empirical analysis of the effect of clustering on the quantity of
     desirable properties attainable by theory exploration.
   \item Experimental application of unsupervised machine learning to the
-    bucketing problem.
+    clustering problem.
   % \item A novel feature extraction method for transforming Haskell expressions
@@
 use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
-bucketing problem in \S~\ref{sec:bucketing} and measure its effect on theory
-exploration systems in \S~\ref{sec:implementation}. We attempt (unsuccessfully)
-to apply machine learning to the bucketing problem in \S~\ref{sec:clustering}.
-A variety of related work is surveyed in \S~\ref{sec:related} and we conclude in
-\S~\ref{sec:conclusion} with potential directions for future research.
+clustering problem more formally in \S~\ref{sec:clustering} and measure its
+effect on theory exploration systems in \S~\ref{sec:implementation}. We attempt
+(unsuccessfully) to apply machine learning to the clustering problem in
+\S~\ref{sec:clustering}.  A variety of related work is surveyed in
+\S~\ref{sec:related} and we conclude in \S~\ref{sec:conclusion} with potential
+directions for future research.

### 2019-04-08 Bucketing/supporting-material/survivalGraph.py

@@ -1,0 +1,73 @@
+#!/usr/bin/env python3
+
+from os     import getenv
+from pandas import read_csv
+with open(getenv('csv'), 'r') as f:
+    times = read_csv(f, header=0, index_col=0)
+del(getenv)
+del(read_csv)
+
+from lifelines import KaplanMeierFitter
+import matplotlib.pyplot as plt
+
+axes = plt.subplot(111, label="stepped")
+kmf  = KaplanMeierFitter()
+for s, grouped_df in times.groupby('size'):
+    if s not in [1, 5, 10, 15, 20]:
+        continue
+    kmf.fit(grouped_df['time'], grouped_df['success'], label=str(s))
+    axes = kmf.plot(ax=axes)
+axes.get_figure().savefig('stepped.pdf')
+
+def crossValidate(name, fitter):
+    from lifelines.utils import k_fold_cross_validation
+    import numpy as np
+    print("Cross Validating " + name)
+    print(np.mean(k_fold_cross_validation(fitter, times, duration_col='time', event_col='success')))
+    print("End cross-validation of " + name)
+
+from lifelines import CoxPHFitter
+cph = CoxPHFitter()
+cph.fit(times, duration_col='time', event_col='success', show_progress=True)
+
+cph.print_summary()
+
+cph.plot().get_figure().savefig('coxph.pdf')
+#cph.check_assumptions(times, show_plots=True)
+
+crossValidate('cox', cph)
+
+fitters = {'cox':cph}
+
+from lifelines import WeibullAFTFitter, LogNormalAFTFitter, LogLogisticAFTFitter
+for (name, fitter) in [("weibull"    , WeibullAFTFitter    ),
+                       ("lognormal"  , LogNormalAFTFitter  ),
+                       ("loglogistic", LogLogisticAFTFitter)]:
+    print("BEGIN " + name)
+    aft = fitter()
+    aft.fit(times, duration_col='time', event_col='success')
+    aft.print_summary(3)
+
+    #aft = WeibullAFTFitter().fit(times, 'time', 'success', ancillary_df=True)
+    aft.plot().get_figure().savefig(name + 'aft.pdf')
+    fitters[name] = aft
+    crossValidate(name, aft)
+    print("END " + name)
+
+print('EXAMPLE DATA FOLLOWS')
+from lifelines import AalenAdditiveFitter, CoxPHFitter
+from lifelines.datasets import load_regression_dataset
[…]

### 2019-04-08 Bucketing/supporting-material/timeoutGraph.py

@@ -1,0 +1,57 @@
+#!/usr/bin/env python3
+from os  import getenv
+from csv import DictReader
+
+def msg(x):
+    from sys import stderr
+    stderr.write(repr(x))
+    stderr.flush()
+
+with open(getenv('csv'), 'r') as f:
+    times = DictReader(f)
+    rows  = [row for row in times]
+
+del(getenv  )
+del(DictReader)
+
+counts = {}
+fails  = {}
+for row in rows:
+    s = row['size']
+    if s not in counts:
+        counts[s] = 0
+    if s not in fails:
+        fails[s] = 0
+    counts[s] += 1
+    if row['success'] == '0':
+        fails[s] += 1
+
+keys  = sorted(counts.keys(), key=int)
+fracs = [fails[s] / counts[s] for s in keys]
+
+msg({
+    'len(fracs)': len(fracs),
+    'len(range(1,21))': len(range(1,21))
+})
+
+from scipy import stats
+slope, intercept, r_value, p_value, std_err = stats.linregress(range(1,21),
+                                                               fracs)
+
+import matplotlib.pyplot as plt
+
+plt.plot(fracs, '.')
+
+f  = lambda x: slope*x + intercept
+xs = [  x  for x in range(1,21)]
+ys = [f(x) for x in xs         ]
+
+ax = plt.plot(xs, ys)
+
+plt.xlim((1,20))
+plt.ylim((0,1 ))
+plt.xticks(xs)
+plt.xlabel('size')
+plt.ylabel('failure proportion')
+plt.title('Failures as size grows')
+plt.savefig('timeouts.pdf')

### 2019-04-08 PhDThesis/background_theoryexploration.tex
Thesis bits

--- 2018-05-17 PhDThesis/background_theoryexploration.tex
@@
 
-In this work we consider the problem of \emph{(automated) theory exploration},
-which includes the ability to \emph{generate} conjectures about code, to
+In this work we consider the application of \emph{(automated) theory
+  exploration} to generate conjectures about code. Existing tools are able to
 \emph{prove} those conjectures, and hence output \emph{novel} theorems without
@@
 what to include in the signature. Here we know that addition and multiplication
-are closely related, and hence obey many algebraic laws. Our machine learning
-technique aims to predict these kinds of relations between functions, so we can
-create small signatures which nevertheless have the potential to give rise to
-many equations.
+are closely related, and hence obey many algebraic laws.
 
@@
 and the consequences for testing and QA in general.
+
+%%%%% THE BELOW SEEMS TO BE INTERESTINGNESS?
+
+%\begin{description}
+%\item{Interestingness} Various alternative interestingness criteria have been
+%  proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing
+%  the criteria may be useful, for example to distinguish useful relationships
+%  from incidental coincidences; or to prevent surprising, insightful equations
+%  from being discarded because they can be simplified.
+%\end{description}
+
+% Theory exploration is similar to \emph{experimental mathematics}
+
+% - Relation to Science
+%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
+% - Relation to AI tasks: exploring surroundings, etc.
+
+% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
+
+% Theory formation: Alison? Others.
+% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
+% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
+% Communication: Latex, Wikis, APIs, communicating with aliens
+
+% \section{Exploration in Theorem Proving}
+% \label{sec:examples}
+
+% Before exploring abstract definitions of interestingness, we can first consider
+% some scenarios which arise during formal proof where we are forced to generate
+% conjectures. An analysis of these situations, and the subsequent theorems they
+% produce, will contribute towards an empirical justification for what is
+% interesting (at least from a utilitarian point of view) and inform our later
+% exploration of the literature.
+
+% \subsection{Generalisation}
+
+% \providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+
+% When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
+% which $S$ is a special case. Although it seems counterintuitive, a generalised
+% statement can sometimes be \emph{easier} to prove than the original. This arises
+% often in inductive proofs, since the specific obligations which arise in the
+% proof may be incompatible with the available inductive hypotheses.
+
+% However, we cannot blindly generalise \emph{all} obligations we encounter, since
+% \emph{over-generalising} results in obligations which are so strong that they
+% are unprovable, or even false. We must therefore rely on heuristics to guide the
+% generation of generalised conjectures, and hence perform a kind of exploration.
+
+% An informative example is given by Boyer and Moore of the associativity of
+% multiplication in ACL2 \cite{boyer1983proof}:
+
+% $$(x * y) * z = x * (y * z)$$
+
+% During the course of the proof, the following obligation arises:
+
+% \begin{equation}
+%   \tag{conc3}
+%   (y + (x * y)) * z = (y * z) + ((x * y) * z)
+%   \label{eq:conc3}
+% \end{equation}
+
+% ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
+% sub-term $x * y$ with a fresh variable $w$:
+
+% \begin{equation}
+%   \tag{conc4}
+%   (y + w) * z = (y * z) + (w * z)
+%   \label{eq:conc4}
+% \end{equation}
+
+% This generalised form is clearly the distributivity law for multiplication and
+% addition, which can be proved separately to the original goal of
+% associativity. It would not be controversial to claim that this distributivity
+% law is interesting in its own right (relative to associativity, at least), in
+% addition to its usefulness in making this proof go through.
+
+% % TODO: Describe the ACL2 heuristics
+
+% Generalisation also occurs frequently when reasoning about \emph{tail-recursive}
+% definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can
+%   be executed in constant space using a loop, whereas recursion in non-tail
+%   positions may require a growing number of stack frames or nested closures. See
+%   \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
+
+% \subsection{Analogy}
+
+% One way to characterise the interestingness of a statement is by \emph{analogy}
+% to existing interesting statements. By finding lemmas analogous to those of a
+% different theory, we may be able to re-use tactics and other forms of
+% meta-programming across both.
+
+% Existing theory exploration systems have been successfully applied to this
+% problem, however the use of pure exploration misses opportunities to
+% \emph{focus} the search, since we know which lemmas are used in those theories
+% where a technique succeeded. If we can find an analogy to map from such solved
+% problems to our unsolved goal, we can infer the approximate form of the lemmas
+% we require, and target these specifically.
+
+% The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
+% to solving a goal $G$ by making analogies via unsupervised clustering
+% \cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are used in two
+% ways:
+
+% \begin{itemize}
+% \item First, we use the cluster $C_G$ containing $G$ to identify analogous
+%   theorems.
+
+% \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
+%   $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
+%   to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
+%   the cluster $C_s$ containing $s$.
+% \end{itemize}
+
+% The running examples for demonstrating \textsc{ACL2(ml)} are equivalence
+% theorems for tail-recursive and non-tail-recursive calculations, as well as the
+% effect of repeating certain list operations:
+
+% \begin{itemize}
+
+%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
+
+%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
+
+%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
+
+%   \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
+
+%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
+
+%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
+
+% \end{itemize}
+
+% \subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
+
+% One consideration when generating conjectures is the difference between
+% theorems, lemmas, corollaries, etc. From a logical point of view, these are all
+% equivalent, and hence most proof assistants do not distinguish between
+% them. However, their \emph{intention} may be different: in a sense, theorems are
+% the interesting results; whilst lemmas are useful results, required for proving
+% the theorems.
+
+% Some systems, like Coq, allow users to \emph{label} each statement as being a
+% \coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
+% the same. This shows us immediately that lemmas outnumber theorems; in the Coq
+% standard library there are over five times as many lemmas as theorems
+% \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when
+%   excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of
+%   \coq{Lemma} in its \texttt{theories/} directory.}.
+
+% % TODO: Analyse them
+
+% % TODO: Theory exploration as lemma generation; give example from a HipSpec paper
+
+% We can find a need for auxiliary lemmas, once again, in the context of
+% tail-recursive functions. Consider proving the (pointwise) equality of the
+% following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% Inductive Nat : Set := Z : Nat
+%                      | S : Nat -> Nat.
+
+% Fixpoint plus      (n m : Nat) := match n with
+%                                       | Z    => m
+%                                       | S n' => S (plus n' m)
+%                                   end.
+
+% Fixpoint plus_tail (n m : Nat) := match n with
+%                                       | Z    => m
+%                                       | S n' => plus_tail n' (S m)
+%                                   end.
+% \end{lstlisting}
+
+% \begin{haskell}
+% Haskell equivalent:
+
+% plus :: Nat -> Nat -> Nat
+% plus      n  Z    = n
+% plus      n (S m) = S (plus n m)
+
+% plus_tail :: Nat -> Nat -> Nat
+% plus_tail n  Z    = n
+% plus_tail n (S m) = plus_tail (S n) m
+% \end{haskell}
+
+% Both of these functions implement addition, but the \coq{plus_tail} variant is
+% tail-recursive. However, if we want to \emph{prove} that the definitions are
+% (pointwise) equal, we run into difficulties. In particular, when the inductive
+% step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
+% reasonable), we cannot make this go through using another inductive argument.
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% (* Solve equalities by beta-normalising both sides *)
+% Ltac triv := try (simpl; reflexivity).
+
+% (* Prove equivalence of plus and plus_tail *)
+% Theorem equiv : forall n m, plus n m = plus_tail n m.
+%   induction n; triv. (* Base case is trivial *)
+
+%   (* Inductive case: plus (S n) m = plus_tail (S n) m *)
+%   intro m.
+
+%   (* Beta-reduce the right-hand-side (justification is trivial) *)
+%   replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
+
+%   (* Use induction hypothesis to replace plus_tail with plus *)
+%   rewrite <- (IHn (S m)).
+% \end{lstlisting}
+
+% Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
+% the equation we need:
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
+% \end{lstlisting}
+
+% Yet we cannot provide it with the argument it needs, as our original induction
+% hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% IHn : forall x, plus (S n') x = plus_tail (S n') x
+% \end{lstlisting}
+
+% We are forced to abandon the proof, despite such a reasonable-looking
+% intermediate goal.
+
+% In fact, if we attempt to prove that goal \emph{separately}, we can use a
+% straightforward argument by induction; even though it is actually
+% \emph{stronger} due to the absence of the \coq{IHn} assumption. Using this
+% separate result as a lemma, the pointwise equality is proven easily.
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% Lemma gen n m : plus (S n) m = plus n (S m).
+%   induction n; triv. (* Base case is trivial *)
+
+%   (* Move all S constructors outside *)
+%   simpl. rewrite <- IHn. simpl.
+
+%   (* Trivial *)
+%   reflexivity.
+% Defined.
+% \end{lstlisting}
+
+% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+%   rewrite (gen n m).
+%   reflexivity.
+% Defined.
+% \end{lstlisting}

### 2019-04-08 PhDThesis/litreview_automatedreasoning.tex
Thesis bits

@@
 
-A similar ``layering'' approach is found in the IsaScheme system of
+A similar ``layering'' approach is found in the \textsc{IsaScheme} system of
 Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
-has also been quantitatively compared against IsaCoSy and HipSpec using
-precision/recall analysis~\cite{claessen2013automating}. The name comes from its
-embedding in the Isabelle proof assistant and its use of ``schemes'':
-higher-order formulae which can be used to generate new concepts and
-conjectures. Variables within a scheme are instantiated automatically and this
-drives the invention process. For example, the concept of ``repetition'' can be
-encoded as a scheme, and instantiated with existing encodings of zero, successor
-and addition to produce a definition of multiplication. The same scheme can be
-instantiated with this new multiplication function to produce exponentiation.
+has also been quantitatively compared against \textsc{IsaCoSy} and
+\textsc{HipSpec} using precision/recall analysis~\cite{claessen2013automating}.
+The name comes from its embedding in the Isabelle proof assistant and its use of
+``schemes'': higher-order formulae which can be used to generate new concepts
+and conjectures. Variables within a scheme are instantiated automatically and
+this drives the invention process. For example, the concept of ``repetition''
+can be encoded as a scheme, and instantiated with existing encodings of zero,
+successor and addition to produce a definition of multiplication. The same
+scheme can be instantiated with this new multiplication function to produce
+exponentiation.
 
-IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
-described in more detail in $\S$\ref{sec:existing-tools}, since these are the
-tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
-has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
-distinct enumeration and testing steps with a single, iterative algorithm
-similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
-completion algorithm to form a corresponding set of rewrite rules. As
-expressions are enumerated, they are simplified using these rules and discarded
-if equal to a known expression. If not, QuickCheck tests whether the new
-expression can be distinguished from the known expressions through random
-testing: those which can are added to the set of known expressions. Those which
-cannot be distinguished are conjectured to be equal, and the rewrite rules are
-updated.
+\textsc{IsaCoSy} (Isabelle Conjecture Synthesis) is written for the Isabelle
+proof assistant, mostly in
+StandardML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures
+equations by enumerating expressions involving a given set of (typed) constants
+and free variables (a \emph{signature}). A constraint solver forbids certain
+(sub)expressions from being synthesised, and these constraints are extended
+whenever a new property is conjectured, to avoid generating any special-cases of
+this property in the future. Conjectures are tested by looking for
+counterexamples and, if none are found, sent to \textsc{IsaPlanner} which
+attempts to prove them.
 
-QuickSpec has also inspired another MTE tool for Haskell called
-Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
-also makes use of the laws of total orders and Boolean algebra to conjecture
-\emph{in}equalities and conditional relations between expressions.
+\textsc{QuickSpec} (the conjecture generation component of \textsc{HipSpec})
+emerged from work on the \textsc{QuickCheck} software testing framework for the
+Haskell programming language~\cite{claessen2011quickcheck}. As well as
+conjecturing properties of Haskell definitions for \textsc{HipSpec}, it has been
+applied to Isabelle definitions via the \textsc{Hipster} tool~\cite{Hipster}.
+Like \textsc{IsaCoSy}, \textsc{QuickSpec} takes a signature and enumerates
+well-typed terms. It collects together those of the same type into equivalence
+classes, assuming them to be equal, then uses \textsc{QuickCheck} to find
+counterexamples to this assumption by randomly instantiating the variables and
+comparing the resulting closed expressions. Any equivalence class whose elements
+don't compare equal are split up, and the process is repeated with new random
+values.
+
+After 500 rounds of testing, any expressions still sharing the same equivalence
+class are conjectured to be equal for all values of their variables. Rather than
+using a constraint system to prevent redundancies (such as special-cases of
+other properties), \textsc{QuickSpec} instead passes its output through a
+congruence closure algorithm to achieve the same effect.
+
+\textsc{QuickSpec} has since evolved to version 2~\cite{smallbone2017quick},
+which replaces the distinct enumeration and testing steps with a single,
+iterative algorithm similar to that of \textsc{IsaCoSy}. Generated conjectures
+are fed into a Knuth-Bendix completion algorithm to form a corresponding set of
+rewrite rules. As expressions are enumerated, they are simplified using these
+rules and discarded if equal to a known expression. If not, \textsc{QuickCheck}
+tests whether the new expression can be distinguished from the known expressions
+through random testing: those which can are added to the set of known
+expressions. Those which cannot be distinguished are conjectured to be equal,
+and the rewrite rules are updated.
+
+\textsc{QuickSpec} has also inspired another MTE tool for Haskell called
+\textsc{Speculate}~\cite{braquehais2017speculate}, which operates in a similar
+way but also makes use of the laws of total orders and Boolean algebra to
+conjecture \emph{in}equalities and conditional relations between expressions.
 
 Another notable MTE implementation, distinct from those based in Isabelle and
-Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
-Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
-generate \emph{conjectures} that may later be sent to automated provers,
-MATHsAiD directly generates \emph{theorems}, by making logically valid
+Haskell, is the \textsc{MATHsAiD} project (Mechanically Ascertaining Theorems
+from Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above,
+which generate \emph{conjectures} that may later be sent to automated provers,
+\textsc{MATHsAiD} directly generates \emph{theorems}, by making logically valid
 inferences from a given set of axioms and definitions. Evaluation of the
@@
 developer, which highlights how these tools could benefit from the availability
-of an objective, repeatable, quantitative method of evaluation and comparison
-such as ours.
+of an objective, repeatable, quantitative method of evaluation and comparison.

### 2019-04-09 Bucketing/sizes.tex
Use vector output

\begin{figure}
  \input{images/stepped.pgf}
\end{figure}

\begin{figure}
  \input{images/timeouts.pgf}
\end{figure}

### 2019-04-09 Bucketing/supporting-material/survivalGraph.py

@@ -10,6 +10,10 @@ del(read_csv)
 from lifelines import KaplanMeierFitter
 import matplotlib.pyplot as plt
 
+def save(name, axes):
+    axes.get_figure().savefig(name + '.pdf')
+    axes.get_figure().savefig(name + '.pgf')
+
 axes = plt.subplot(111, label="stepped")
 kmf  = KaplanMeierFitter()
 for s, grouped_df in times.groupby('size'):
@@ -17,7 +21,7 @@ for s, grouped_df in times.groupby('size'):
         continue
     kmf.fit(grouped_df['time'], grouped_df['success'], label=str(s))
     axes = kmf.plot(ax=axes)
-axes.get_figure().savefig('stepped.pdf')
+save('stepped', axes)
 
 def crossValidate(name, fitter):
     from lifelines.utils import k_fold_cross_validation
@@ -32,7 +36,7 @@ cph.fit(times, duration_col='time', event_col='success', show_progress=True)
 
 cph.print_summary()
 
-cph.plot().get_figure().savefig('coxph.pdf')
+save('coxph', cph.plot())
 #cph.check_assumptions(times, show_plots=True)
 
 crossValidate('cox', cph)
@@ -49,7 +53,7 @@ for (name, fitter) in [("weibull"    , WeibullAFTFitter    ),
     aft.print_summary(3)
 
     #aft = WeibullAFTFitter().fit(times, 'time', 'success', ancillary_df=True)
-    aft.plot().get_figure().savefig(name + 'aft.pdf')
+    save(name + 'aft', aft.plot())
     fitters[name] = aft
     crossValidate(name, aft)
     print("END " + name)

### 2019-04-09 Bucketing/supporting-material/timeoutGraph.py

@@ -7,6 +7,10 @@ def msg(x):
     stderr.write(repr(x))
     stderr.flush()
 
+def save(name, axes):
+    axes.get_figure().savefig(name + '.pdf')
+    axes.get_figure().savefig(name + '.pgf')
+
 with open(getenv('csv'), 'r') as f:
     times = DictReader(f)
     rows  = [row for row in times]
@@ -54,4 +58,5 @@ plt.xticks(xs)
 plt.xlabel('size')
 plt.ylabel('failure proportion')
 plt.title('Failures as size grows')
-plt.savefig('timeouts.pdf')
+save('timeouts', plt.gca())
+#plt.savefig('timeouts.pdf')

### 2019-04-09 Bucketing/background.tex
Include sizes in paper

@@
 
-\iffalse % TODO{2019-03-15}: Make this more formal?
+\iffalse % TODO{2019-04-09}: Make this more formal?
  V \in Var
@@
 
-\iffalse TODO{2019-03-15} In fact, there are similarities between the way a TE
+\iffalse TODO{2019-04-09} In fact, there are similarities between the way a TE
 system like \qspec{} can generalise from checking \emph{particular} properties
@@
 
-% TODO{2019-03-15}
+% TODO{2019-04-09}

### 2019-04-10 Bucketing/sizes.tex
Added captions

@@
   \input{images/stepped.pgf}
+  \caption{Survival plot % TODO: NAME OF PLOT
+    for running QuickSpec on inputs containing various numbers of definitions,
+    sampled from TIP. Most samples finished quickly, leaving a heavy tail which
+    didn't finish within the 300 second timeout.}
+  \label{fig:survival}
 \end{figure}
@@
   \input{images/timeouts.pgf}
+  \caption{Proportion of samples which timed out per size, with least-squares
+    linear regression.}
+  \label{fig:tailsize}
 \end{figure}

### 2019-04-10 Bucketing/supporting-material/contentsGraph.py
Graph contents

@@ -1,0 +1,67 @@
+#!/usr/bin/env python3
+from os  import getenv
+from csv import DictReader
+
+def msg(x):
+    from sys import stderr
+    stderr.write(repr(x))
+    stderr.flush()
+
+def save(name, axes):
+    axes.get_figure().savefig(name + '.pdf')
+    axes.get_figure().savefig(name + '.pgf')
+
+with open(getenv('csv'), 'r') as f:
+    counts = DictReader(f)
+    rows   = [{**row,
+               **{'successes': int(row['successes']),
+                  'failures' : int(row['failures' ])}} for row in counts]
+
+del(getenv    )
+del(DictReader)
+
+rows.sort(key=lambda x: x['successes'] / (x['successes'] + x['failures']),
+          reverse=True)
+
+names     = [row['name'     ] for row in rows]
+successes = [row['successes'] for row in rows]
+failures  = [row['failures' ] for row in rows]
+
+prop = lambda key: lambda row: row[key] / (row['successes'] + row['failures'])
+
+successProp = list(map(prop('successes'), rows))
+failureProp = list(map(prop('failures' ), rows))
+
+import matplotlib.pyplot as plt
+
+barWidth = 0.85
+r        = [x for x in range(0, len(names))]
+
+# Create green Bars
+plt.bar(r, successProp,                     color='#00ff00', edgecolor='white', width=barWidth)
+# Create orange Bars
+plt.bar(r, failureProp, bottom=successProp, color='#ff0000', edgecolor='white', width=barWidth)
+
+# Custom x axis
+plt.xticks(r, names)
+plt.xlabel("Name")
+
+#ax = plt.subplot(111)
+#ax.bar(names, successes, width=0.3, color='g', align='center')
+#ax.bar(names, failures , width=0.3, color='r', align='center')
+#ax.xaxis_date()
+
+#plt.bar([ row['name'     ]                   for row in rows],
+#        [[row['successes'], row['failures']] for row in rows])
+
+#plt.xlim((1,20))
+#plt.ylim((0,1 ))
+#plt.xticks(xs)
+plt.xlabel('Name')
+plt.ylabel('Occurrences')
+plt.title('Name distribution between successful and failed runs')
+
+#ax.set_xlabel('Name')
+#ax.set_ylabel('Occurrences')
+#ax.set_title('Name distribution between successful and failed runs')
+save('contents', plt.gca())

### 2019-04-11 Bucketing/sizes.tex
Add problematic definitions

@@
   \input{images/stepped.pgf}
-  \caption{Survival plot % TODO: NAME OF PLOT
+  \caption{Survival plot % TODO{2019-04-15}: NAME OF PLOT
     for running QuickSpec on inputs containing various numbers of definitions,
@@
 \end{figure}
+
+\begin{figure}
+  \input{images/proportions.pgf}
+  \caption{}
+  \label{fig:proportions}
+\end{figure}
+
+Five definitions appeared only in failing samples, these are named
+\texttt{mult2}, \texttt{qexp}, \texttt{op}, \texttt{mul3acc} and \texttt{mul3};
+their definitions appear in Figure~\ref{fig:faildefs}. All of these are
+functions of natural numbers (\texttt{Nat}). The \texttt{mul3} and
+\texttt{mul3acc} functions are rather pathological implementations of
+multiplication with an accumulator parameter, involving many (non-tail)
+recursive calls which can cause space leaks (or stack overflow, in languages
+with eager evaluation). This is likely the reason \textsc{QuickSpec} performs
+poorly with these definitions in the input.
+
+The \texttt{op}, \texttt{mult2} and \texttt{qexp} functions are less easily
+dismissed since they only recurse in tail position; the latter are in fact
+standard definitions of multiplication and exponentiation, respectively. One
+potential problem with these definitions, when translated into Haskell for
+\textsc{QuickSpec}, is that their accumulating parameter will be lazy by
+default, and hence cause space leaks.
+
+\begin{figure}
+  \texttt{
+    (define-fun-rec mult2 ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case  Z z)
+        (case (S x2)
+          (mult2 x2 y (plus y z)))))
+  }
+
+  \texttt{
+    (define-fun-rec qexp ((x Nat) (y Nat) (z Nat)) Nat
+      (match y
+        (case  Z z)
+        (case (S n)
+          (qexp x n (mult x z)))))
+  }
+
+  \texttt{
+    (define-fun-rec op ((x Nat) (y Nat) (z Nat) (x2 Nat)) Nat
+      (match x
+        (case Z
+          (match z
+            (case  Z x2)
+            (case (S x3)
+              (op Z y x3 (S x2)))))
+        (case (S x4)
+          (match z
+            (case Z
+              (op x4 y y x2))
+            (case (S c)
+              (op x y c (S x2)))))))
+  }
+
+  \texttt{
+    (define-fun-rec mul3acc ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case  Z Z)                         ;; Base case for 0 * y * z
+        (case (S x2)
+          (match y
+            (case  Z Z)                     ;; Base case for x * 0 * z
+            (case (S x3)
+              (match z
+                (case  Z Z)                 ;; Base case for x * y * 0
+                (case (S x4)
+                  (match x2
+                    (case Z
+                      (match x3
+                        (case Z
+                          (match x4
+                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
+                            (case (S x5)
+                              (S (add3acc (mul3acc Z Z x4)
+                                          (add3acc (mul3acc (S Z) Z x4)
+                                                   (mul3acc Z (S Z) x4)
+                                                   (mul3acc Z Z (S Z)))
+                                          (add3acc Z Z x4))))))
+                        (case (S x6)
+                          (S (add3acc (mul3acc Z x3 x4)
+                                      (add3acc (mul3acc (S Z) x3 x4)
+                                               (mul3acc Z (S Z) x4)
+                                               (mul3acc Z x3 (S Z)))
+                                      (add3acc Z x3 x4))))))
+                    (case (S x7)
+                      (S (add3acc (mul3acc x2 x3 x4)
+                                  (add3acc (mul3acc (S Z) x3 x4)
+                                           (mul3acc x2 (S Z) x4)
+                                           (mul3acc x2 x3 (S Z)))
+                                  (add3acc x2 x3 x4))))))))))))
+  }
+
+  \texttt{
+    (define-fun-rec mul3 ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case Z Z)                    ;; Base case for 0 * y * z
+        (case (S x2)
+          (match y
+            (case Z Z)                ;; Base case for x * 0 * z
+            (case (S x3)
+              (match z
+                (case Z Z)            ;; Base case for x * y * 0
+                (case (S x4)
+                  (match x2
+                    (case Z
+                      (match x3
+                        (case Z
+                          (match x4
+                            (case Z
+                              (S Z))  ;; Base case for 1 * 1 * 1
+                            (case (S x5)
+                              (S (add3 (mul3 Z Z x4)
+                                       (add3 (mul3 (S Z) Z x4)
+                                             (mul3 Z (S Z) x4)
+                                             (mul3 Z Z (S Z)))
+                                       (add3 Z Z x4))))))
+                        (case (S x6)
+                          (S (add3 (mul3 Z x3 x4)
+                                   (add3 (mul3 (S Z) x3 x4)
+                                         (mul3 Z (S Z) x4)
+                                         (mul3 Z x3 (S Z)))
+                                   (add3 Z x3 x4))))))
+                    (case (S x7)
+                      (S (add3 (mul3 x2 x3 x4)
+                               (add3 (mul3 (S Z) x3 x4)
+                                     (mul3 x2 (S Z) x4)
+                                     (mul3 x2 x3 (S Z)))
+                               (add3 x2 x3 x4))))))))))))
+  }
+  \label{fig:faildefs}
+\end{figure}

### 2019-04-11 Bucketing/supporting-material/StrictAccumulators/Main.hs
Add our tests of strict/lazy definitions

@@ -1,0 +1,46 @@
+{-# LANGUAGE BangPatterns #-}
+
+import Test.QuickCheck
+import Test.QuickSpec
+
+-- Peano numerals
+
+data Nat = Z | S Nat deriving (Eq)
+
+i2n :: Integer -> Nat
+i2n 0 = Z
+i2n n = S (i2n (abs n - 1))
+
+n2i :: Nat -> Integer
+n2i  Z    = 0
+n2i (S n) = 1 + n2i n
+
+instance Show Nat where
+  show = show . n2i
+
+instance Ord Nat where
+  Z   <= _   = True
+  S x <= Z   = False
+  S x <= S y = x <= y
+
+instance Arbitrary Nat where
+  arbitrary = i2n <$> arbitrary
+
+plus :: Nat -> Nat -> Nat
+plus  Z    y = y
+plus (S x) y = S (plus x y)
+
+mult2_lazy, mult2_strict :: Nat -> Nat -> Nat -> Nat
+mult2_lazy x y z = case x of
+                     Z    -> z
+                     S x2 -> mult2_lazy x2 y (plus y z)
+
+mult2_strict x y !z = case x of
+                        Z    -> z
+                        S x2 -> mult2_strict x2 y (plus y z)
+
+sig_mult2_lazy = fun3 "mult2_lazy" mult2_lazy :
+                map (gvars ["nat1", "nat2", "nat3"]) [g, g, g]
+  where g = arbitrary :: Gen Nat
+
+qs_mult2_lazy = quickSpec sig_mult2_lazy

### 2019-04-11 Bucketing/supporting-material/contentsGraph.py
Change name

@@ -64,4 +64,4 @@ plt.title('Name distribution between successful and failed runs')
 #ax.set_xlabel('Name')
 #ax.set_ylabel('Occurrences')
 #ax.set_title('Name distribution between successful and failed runs')
-save('contents', plt.gca())
+save('proportions', plt.gca())

### 2019-04-12 Bucketing/sizes.tex
More bucketing stuff

@@
 Five definitions appeared only in failing samples, these are named
-\texttt{mult2}, \texttt{qexp}, \texttt{op}, \texttt{mul3acc} and \texttt{mul3};
+\texttt{mul3}, \texttt{mul3acc}, \texttt{mult2}, \texttt{op} and \texttt{qexp};
 their definitions appear in Figure~\ref{fig:faildefs}. All of these are
-functions of natural numbers (\texttt{Nat}). The \texttt{mul3} and
-\texttt{mul3acc} functions are rather pathological implementations of
-multiplication with an accumulator parameter, involving many (non-tail)
-recursive calls which can cause space leaks (or stack overflow, in languages
-with eager evaluation). This is likely the reason \textsc{QuickSpec} performs
-poorly with these definitions in the input.
+functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
+exploration to time out by generating very large outputs.
 
-The \texttt{op}, \texttt{mult2} and \texttt{qexp} functions are less easily
-dismissed since they only recurse in tail position; the latter are in fact
-standard definitions of multiplication and exponentiation, respectively. One
-potential problem with these definitions, when translated into Haskell for
-\textsc{QuickSpec}, is that their accumulating parameter will be lazy by
-default, and hence cause space leaks.
+\texttt{mul3} and \texttt{mul3acc} are rather pathological implementations of
+multiplication with an accumulator parameter, with many (non-tail) recursive
+calls. \texttt{mult2} and \texttt{qexp} are standard tail-recursive definitions
+of multiplication and exponentiation, respectively. The \texttt{op} function
+seems to exist only for testing purposes, appearing in files named
+\texttt{weird_nat_op} which assert its commutativity and associativity.
+
+Exploring each of these functions on its own does not require much memory, since
+Haskell generates the output lazily. However, comparing such large numbers for
+equality takes a lot of CPU time as the \texttt{S} constructors are successively
+unwrapped from each side, and this is why the timeout is reached. We confirmed
+this hypothesis by exploring with a custom data generator which only generates
+the values \texttt{Z}, \texttt{S Z} and \texttt{S (S Z)} (0, 1 and 2); this
+caused the exploration to finish quickly. Other interventions, like making the
+accumulator arguments strict (to prevent space leaks), did not prevent timeouts.
 
@@
         (case Z      z)
-        (case (S x2)
-          (mult2 x2 y (plus y z)))))
+        (case (S x2) (mult2 x2 y (plus y z)))))
   }
@@
         (case Z     z)
-        (case (S n)
-          (qexp x n (mult x z)))))
+        (case (S n) (qexp x n (mult x z)))))
   }
@@
             (case Z      x2)
-            (case (S x3)
-              (op Z y x3 (S x2)))))
+            (case (S x3) (op Z  y x3 (S x2)))))
         (case (S x4)
           (match z
-            (case Z
-              (op x4 y y x2))
-            (case (S c)
-              (op x y c (S x2)))))))
+            (case Z      (op x4 y y  x2))
+            (case (S c ) (op x  y c  (S x2)))))))
   }
@@
                           (match x4
-                            (case Z
-                              (S Z))  ;; Base case for 1 * 1 * 1
+                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
                             (case (S x5)
@@
 \end{figure}
+
+To assess the impact of these problematic definitions, we removed any samples
+containing them and repeated our analysis.

### 2019-04-12 Bucketing/supporting-material/StrictAccumulators/Defs.hs

@@ -27,9 +27,20 @@ instance Ord Nat where
 instance Arbitrary Nat where
   arbitrary = i2n <$> arbitrary
 
+smallGen = i2n <$> choose (0, 2)
+
+-- Helper functions
+
 plus :: Nat -> Nat -> Nat
-plus  Z    y = y
-plus (S x) y = S (plus x y)
+plus x y = case x of
+             Z   -> y
+             S z -> S (plus z y)
+
+mult x y = case x of
+             Z   -> Z
+             S z -> plus y (mult z y)
+
+-- Functions to explore
 
 mult2_lazy, mult2_strict :: Nat -> Nat -> Nat -> Nat
 mult2_lazy x y z = case x of
@@ -39,3 +50,29 @@ mult2_lazy x y z = case x of
 mult2_strict x y !z = case x of
                         Z    -> z
                         S x2 -> mult2_strict x2 y (plus y z)
+
+qexp_lazy, qexp_strict :: Nat -> Nat -> Nat -> Nat
+qexp_lazy x y z = case y of
+                    Z   -> z
+                    S n -> qexp_lazy x n (mult x z)
+
+qexp_strict x y !z = case y of
+                       Z   -> z
+                       S n -> qexp_strict x n (mult x z)
+
+op_lazy, op_strict :: Nat -> Nat -> Nat -> Nat -> Nat
+op_lazy x y z x2 = case x of
+                     Z    -> case z of
+                               Z    -> x2
+                               S x3 -> op_lazy Z  y x3 (S x2)
+                     S x4 -> case z of
+                               Z    -> op_lazy x4 y y     x2
+                               S c  -> op_lazy x  y c  (S x2)
+
+op_strict x y z x2 = case x of
+                       Z    -> case z of
+                                 Z    -> x2
+                                 S x3 -> op_strict Z  y x3 (S x2)
+                       S x4 -> case z of
+                                 Z    -> op_strict x4 y y     x2
+                                 S c  -> op_strict x  y c  (S x2)

### 2019-04-12 Bucketing/supporting-material/StrictAccumulators/QS1Runners.hs

@@ -1,13 +1,39 @@
 module Runners where
 
 import Defs
+import Test.QuickCheck
 import Test.QuickSpec
 
 vs = map (gvars ["nat1", "nat2", "nat3"]) [g, g, g]
   where g = arbitrary :: Gen Nat
 
-[qs_mult2_lazy, qs_mult2_strict, qs_plus_lazy] = map (quickSpec . (:vs))
-  [ fun3 "mult2_lazy"   mult2_lazy
-  , fun3 "mult2_strict" mult2_strict
-  , fun2 "plus_lazy"    plus
-  ]
+[qs_mult2_lazy  ,
+ qs_mult2_strict,
+ qs_qexp_lazy   ,
+ qs_qexp_strict ,
+ qs_op_lazy     ,
+ qs_op_strict   ] =
+  map (quickSpec . (:vs)) [ fun3 "mult2_lazy"   mult2_lazy
+                          , fun3 "mult2_strict" mult2_strict
+                          , fun3 "qexp_lazy"    qexp_lazy
+                          , fun3 "qexp_strict"  qexp_strict
+                          , fun4 "op_lazy"      op_lazy
+                          , fun4 "op_strict"    op_strict
+                          ]
+
+vsmall = map (gvars ["small1", "small2", "small3"]) [g, g, g]
+  where g = smallGen
+
+[small_mult2_lazy  ,
+ small_mult2_strict,
+ small_qexp_lazy   ,
+ small_qexp_strict ,
+ small_op_lazy     ,
+ small_op_strict   ] =
+  map (quickSpec . (:vsmall)) [ fun3 "mult2_lazy"   mult2_lazy
+                              , fun3 "mult2_strict" mult2_strict
+                              , fun3 "qexp_lazy"    qexp_lazy
+                              , fun3 "qexp_strict"  qexp_strict
+                              , fun4 "op_lazy"      op_lazy
+                              , fun4 "op_strict"    op_strict
+                              ]

### 2019-04-17 Bucketing/background.tex
Remove some commented-out code

@@
 irreducible):
-
-\iffalse % TODO{2019-04-09}: Make this more formal?
- V \in Var
- F \in Fun
- T \in Term
- T ::= V | F | T1 T2
-
- Term ::= VAR | Const | Fun (Term)
-or
- Term t ::= x | f | t t'
-\fi
 
@@
 
-\iffalse TODO{2019-04-09} In fact, there are similarities between the way a TE
-system like \qspec{} can generalise from checking \emph{particular} properties
-to \emph{inventing} new ones, and the way counterexample finders like \qcheck{}
-can generalise from testing \emph{particular} expressions to \emph{inventing}
-expressions to test. One of our aims is to understand the implications of this
-generalisation, the lessons that each can learn from the other's approach to
-term generation, and the consequences for testing and QA in general.  \fi
-
 \subsection{Clustering}
@@
 
-% TODO{2019-04-09}
+% TODO{2019-04-18}

### 2019-04-17 Bucketing/sizes.tex
Content analysis

--- 2019-04-10 Bucketing/sizes.tex
@@
   \input{images/stepped.pgf}
-  \caption{Survival plot % TODO: NAME OF PLOT
-    for running QuickSpec on inputs containing various numbers of definitions,
-    sampled from TIP. Most samples finished quickly, leaving a heavy tail which
-    didn't finish within the 300 second timeout.}
+  \caption{Kaplan-Meier survival plot for running QuickSpec on inputs containing
+    various numbers of definitions, sampled from TIP. Most runs finished
+    quickly, leaving a heavy tail which didn't finish within the 300 second
+    timeout.}
   \label{fig:survival}
 \end{figure}
+
+Figure~\ref{fig:survival} shows a Kaplan-Meier plot of \qspec{} running
+times, when given inputs containing different numbers of definitions. There is a
+clear pattern, with many runs finishing quickly and the remainder occupying a
+``long tail'' which we cut off after 300 seconds (based on preliminary
+experiments, which showed little difference in survival between 300 seconds and
+1 hour).
+
+There is an overall trend for larger inputs to have more runs time-out, except
+for the very smallest inputs (which are more constrained by the sampling
+procedure). This relationship is linear, which can be explained if the presence
+of certain definitions is causing the exploration process to fail (since larger
+inputs have a higher chance of containing one of these definitions).
 
@@
 \end{figure}
+
+\begin{figure}
+  \input{images/proportions.pgf}
+  \caption{}
+  \label{fig:proportions}
+\end{figure}
+
+\begin{figure}
+  \begin{minted}{scheme}
+    (define-fun-rec mult2 ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case Z      z)
+        (case (S x2) (mult2 x2 y (plus y z)))))
+  \end{minted}
+
+  \begin{minted}{scheme}
+    (define-fun-rec qexp ((x Nat) (y Nat) (z Nat)) Nat
+      (match y
+        (case Z     z)
+        (case (S n) (qexp x n (mult x z)))))
+  \end{minted}
+
+  \begin{minted}{scheme}
+    (define-fun-rec op ((x Nat) (y Nat) (z Nat) (x2 Nat)) Nat
+      (match x
+        (case Z
+          (match z
+            (case Z      x2)
+            (case (S x3) (op Z  y x3 (S x2)))))
+        (case (S x4)
+          (match z
+            (case Z      (op x4 y y  x2))
+            (case (S c ) (op x  y c  (S x2)))))))
+  \end{minted}
+
+  \iffalse
+  \begin{minted}{scheme}
+    (define-fun-rec mul3acc ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case Z Z)                          ;; Base case for 0 * y * z
+        (case (S x2)
+          (match y
+            (case Z Z)                      ;; Base case for x * 0 * z
+            (case (S x3)
+              (match z
+                (case Z Z)                  ;; Base case for x * y * 0
+                (case (S x4)
+                  (match x2
+                    (case Z
+                      (match x3
+                        (case Z
+                          (match x4
+                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
+                            (case (S x5)
+                              (S (add3acc (mul3acc Z Z x4)
+                                          (add3acc (mul3acc (S Z) Z x4)
+                                                   (mul3acc Z (S Z) x4)
+                                                   (mul3acc Z Z (S Z)))
+                                          (add3acc Z Z x4))))))
+                        (case (S x6)
+                          (S (add3acc (mul3acc Z x3 x4)
+                                      (add3acc (mul3acc (S Z) x3 x4)
+                                               (mul3acc Z (S Z) x4)
+                                               (mul3acc Z x3 (S Z)))
+                                      (add3acc Z x3 x4))))))
+                    (case (S x7)
+                      (S (add3acc (mul3acc x2 x3 x4)
+                                  (add3acc (mul3acc (S Z) x3 x4)
+                                           (mul3acc x2 (S Z) x4)
+                                           (mul3acc x2 x3 (S Z)))
+                                  (add3acc x2 x3 x4))))))))))))
+  \end{minted}
+  \if
+
+  \iffalse
+  \begin{minted}{scheme}
+    (define-fun-rec mul3 ((x Nat) (y Nat) (z Nat)) Nat
+      (match x
+        (case Z Z)                          ;; Base case for 0 * y * z
+        (case (S x2)
+          (match y
+            (case Z Z)                      ;; Base case for x * 0 * z
+            (case (S x3)
+              (match z
+                (case Z Z)                  ;; Base case for x * y * 0
+                (case (S x4)
+                  (match x2
+                    (case Z
+                      (match x3
+                        (case Z
+                          (match x4
+                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
+                            (case (S x5)
+                              (S (add3 (mul3 Z Z x4)
+                                       (add3 (mul3 (S Z) Z x4)
+                                             (mul3 Z (S Z) x4)
+                                             (mul3 Z Z (S Z)))
+                                       (add3 Z Z x4))))))
+                        (case (S x6)
+                          (S (add3 (mul3 Z x3 x4)
+                                   (add3 (mul3 (S Z) x3 x4)
+                                         (mul3 Z (S Z) x4)
+                                         (mul3 Z x3 (S Z)))
+                                   (add3 Z x3 x4))))))
+                    (case (S x7)
+                      (S (add3 (mul3 x2 x3 x4)
+                               (add3 (mul3 (S Z) x3 x4)
+                                     (mul3 x2 (S Z) x4)
+                                     (mul3 x2 x3 (S Z)))
+                               (add3 x2 x3 x4))))))))))))
+  \end{minted}
+  \fi
+  \caption{Definitions which consistently cause \qspec{} to fail. Two other
+    definitions (\texttt{mul3} and \texttt{mul3acc}) are ommitted due to their
+    verbosity.}
+  \label{fig:faildefs}
+\end{figure}
+
+This does appear to be the case, as Figure~\ref{fig:proportions} shows that five
+definitions appeared \emph{only} in failing inputs; these are named
+\texttt{mul3}, \texttt{mul3acc}, \texttt{mult2}, \texttt{op} and \texttt{qexp};
+their definitions appear in Figure~\ref{fig:faildefs}. All of these are
+functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
+exploration to time out by generating very large outputs.
+
+\texttt{mul3} and \texttt{mul3acc} are rather pathological implementations of
+multiplication with an accumulator parameter, with many (non-tail) recursive
+calls. The \texttt{op} function appears in files named \texttt{weird\_nat\_op},
+which assert its commutativity and associativity. Finally, the \texttt{mult2}
+and \texttt{qexp} functions are standard tail-recursive definitions of
+multiplication and exponentiation, respectively.
+
+Exploring each of these functions on its own does not require much memory, since
+Haskell generates the output lazily. However, comparing such large numbers for
+equality takes a lot of CPU time as the \texttt{S} constructors are successively
+unwrapped from each side, and this is why the timeout is reached. We confirmed
+this hypothesis by exploring with a custom data generator which only generates
+the values \texttt{Z}, \texttt{S Z} and \texttt{S (S Z)} (0, 1 and 2); this
+caused the exploration to finish quickly. Other interventions, like making the
+accumulator arguments strict (to prevent space leaks), did not prevent timeouts.
+
+To assess the impact of these problematic definitions, we removed any samples
+containing them and repeated our analysis.

### 2019-04-26 Bucketing/sizes.tex
Add "labels" to distinguish between graphs for "all" and "nontoxic"

@@
 \begin{figure}
-  \input{images/stepped.pgf}
-  \caption{Kaplan-Meier survival plot for running QuickSpec on inputs containing
-    various numbers of definitions, sampled from TIP. Most runs finished
-    quickly, leaving a heavy tail which didn't finish within the 300 second
-    timeout.}
+  \scalebox{0.45}{\input{images/steppedall.pgf}}
+  \scalebox{0.45}{\input{images/steppednontoxic.pgf}}
+  \caption{Kaplan-Meier survival plot for running QuickSpec on inputs
+    containing various numbers of definitions, sampled from TIP. The x axis
+    denotes running time, which we cut short after 300 seconds. The height of
+    each line shows the proportion of \qspec{} runs which were still going at
+    that time (lower is better). First plot is for all TIP definitions, second
+    plot removes runs given ``toxic'' definitions.}
   \label{fig:survival}
@@
 \begin{figure}
-  \input{images/timeouts.pgf}
+  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
+  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
   \caption{Proportion of samples which timed out per size, with least-squares
-    linear regression.}
+    linear regression. First plot is for all TIP definitions, second removes
+    runs given ``toxic'' definitions.}
   \label{fig:tailsize}
@@
 \begin{figure}
-  \input{images/proportions.pgf}
-  \caption{}
+  \scalebox{0.45}{\input{images/proportionsall.pgf}}
+  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
   \label{fig:proportions}
+  \caption{Definitions, ordered by the ratio of successes to failures of the
+    runs they appeared in. First graph contains all TIP definitions, showing
+    ``toxic'' definitions which always failed. Second graph only contains runs
+    without any toxic definitions.}
 \end{figure}
@@
   \fi
-  \caption{Definitions which consistently cause \qspec{} to fail. Two other
-    definitions (\texttt{mul3} and \texttt{mul3acc}) are ommitted due to their
-    verbosity.}
+  \caption{``Toxic'' definitions, which consistently cause \qspec{} to fail. Two
+    other definitions (\texttt{mul3} and \texttt{mul3acc}) are ommitted due to
+    their verbosity.}
   \label{fig:faildefs}

### 2019-04-26 Bucketing/supporting-material/contentsGraph.py

@@ -2,14 +2,16 @@
 from os  import getenv
 from csv import DictReader
 
+label = getenv('name')
+
 def msg(x):
     from sys import stderr
     stderr.write(repr(x))
     stderr.flush()
 
 def save(name, axes):
-    axes.get_figure().savefig(name + '.pdf')
-    axes.get_figure().savefig(name + '.pgf')
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
 
 with open(getenv('csv'), 'r') as f:
     counts = DictReader(f)

### 2019-04-26 Bucketing/supporting-material/survivalGraph.py

@@ -4,6 +4,9 @@ from os     import getenv
 from pandas import read_csv
 with open(getenv('csv'), 'r') as f:
     times = read_csv(f, header=0, index_col=0)
+
+label = getenv('label')
+
 del(getenv)
 del(read_csv)
 
@@ -11,8 +14,8 @@ from lifelines import KaplanMeierFitter
 import matplotlib.pyplot as plt
 
 def save(name, axes):
-    axes.get_figure().savefig(name + '.pdf')
-    axes.get_figure().savefig(name + '.pgf')
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
 
 axes = plt.subplot(111, label="stepped")
 kmf  = KaplanMeierFitter()
@@ -27,7 +30,10 @@ def crossValidate(name, fitter):
     from lifelines.utils import k_fold_cross_validation
     import numpy as np
     print("Cross Validating " + name)
-    print(np.mean(k_fold_cross_validation(fitter, times, duration_col='time', event_col='success')))
+    print(np.mean(k_fold_cross_validation(fitter,
+                                          times,
+                                          duration_col='time',
+                                          event_col='success')))
     print("End cross-validation of " + name)
 
 from lifelines import CoxPHFitter

### 2019-04-26 Bucketing/supporting-material/timeoutGraph.py

@@ -2,14 +2,16 @@
 from os  import getenv
 from csv import DictReader
 
+label = getenv('label')
+
 def msg(x):
     from sys import stderr
     stderr.write(repr(x))
     stderr.flush()
 
 def save(name, axes):
-    axes.get_figure().savefig(name + '.pdf')
-    axes.get_figure().savefig(name + '.pgf')
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
 
 with open(getenv('csv'), 'r') as f:
     times = DictReader(f)

### 2019-04-26 Bucketing/background.tex
Use less-confusing "bucketing problem" rather than "clustering problem"

@@
 but these only contained a few definitions and theorems each. To analyse
-behaviour on a larger scale, as required to study the effects of clustering, we
+behaviour on a larger scale, as required to study the effects of bucketing, we
 have chosen the \emph{Tons of Inductive Problems} (TIP) problem set for
@@
 
-\subsection{The Clustering Problem}
-\label{sec:clustering}
+\subsection{The Bucketing Problem}
+\label{sec:bucketing}
 
-% TODO{2019-04-18}
+% TODO{2019-04-19}

### 2019-04-26 Bucketing/intro.tex
Use less-confusing "bucketing problem" rather than "clustering problem"

@@
 
-We dub such selection the \emph{clustering problem} for theory exploration and
+We dub such selection the \emph{bucketing problem} for theory exploration and
 analyse its effect on the running time of theory exploration tools on a
 particular problem set, and on the quality of the statements they are able to
-generate. An automated solution to this clustering problem would accept a large
+generate. An automated solution to this bucketing problem would accept a large
 number of definitions and efficiently select sub-sets which are both small
@@
 \begin{enumerate}
-  \item Defining the \emph{clustering problem} for theory exploration.
+  \item Defining the \emph{bucketing problem} for theory exploration.
   \item Statistical analysis of theory exploration performance on a large
     problem set, derived from existing theorem proving problem sets.
-  \item Empirical analysis of the effect of clustering on the quantity of
+  \item Empirical analysis of the effect of bucketing on the quantity of
     desirable properties attainable by theory exploration.
-  \item Experimental application of unsupervised machine learning to the
-    clustering problem.
+  \item Experimental application of machine learning to the bucketing problem,
+    via unsupervised \emph{clustering}.
   % \item A novel feature extraction method for transforming Haskell expressions
@@
 use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
-clustering problem more formally in \S~\ref{sec:clustering} and measure its
-effect on theory exploration systems in \S~\ref{sec:implementation}. We attempt
-(unsuccessfully) to apply machine learning to the clustering problem in
-\S~\ref{sec:clustering}.  A variety of related work is surveyed in
-\S~\ref{sec:related} and we conclude in \S~\ref{sec:conclusion} with potential
-directions for future research.
+bucketing problem more formally in \S~\ref{sec:bucketing} and measure its
+effect on theory exploration systems in \S~\ref{sec:implementation}. In
+\S~\ref{sec:clustering} we highlight the difference betwee bucketing and the
+well-known \emph{clustering problem} from machine learning, by demonstrating
+that clustering algorithms perform poorly at bucketing.  A variety of related
+work is surveyed in \S~\ref{sec:related} and we conclude in
+\S~\ref{sec:conclusion} with potential directions for future research.

### 2019-04-26 Bucketing/appendices.tex
Don't need sentinels in appendix anymore
Split long lines

--- 2016-03-02 CICM2016/appendices.tex
@@
-%\iffalse  % WITHOUT APPENDICES
-\iftrue    % WITH APPENDICES
-
 \appendix
@@
 
-The GHC Core language is based on \fc{}, and described in detail in \cite[Appendix C]{sulzmann2007system}.
-For our machine learning purposes we are mostly interested in the syntax of reducible expressions (representing Haskell of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
+The GHC Core language is based on \fc{}, and described in detail in
+\cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are
+mostly interested in the syntax of reducible expressions (representing Haskell
+of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below
+in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
 
@@
 
-The full Core language emitted by GHC (as of version 7.10.2, the latest at the time of writing) is translated automatically to this simplified form prior to recurrent clustering.
-Our major restriction is to ignore type-level entities (such as datatype definitions, explicit casts, and differences between types/kinds/coercions).
-Our implementation also handles several other forms of literal (machine words of various sizes, individual characters, etc.), but we omit them here for brevity as their treatment is similar to those of strings and numerals.
+The full Core language emitted by GHC (as of version 7.10.2, the latest at the
+time of writing) is translated automatically to this simplified form prior to
+recurrent clustering. Our major restriction is to ignore type-level entities
+(such as datatype definitions, explicit casts, and differences between
+types/kinds/coercions). Our implementation also handles several other forms of
+literal (machine words of various sizes, individual characters, etc.), but we
+omit them here for brevity as their treatment is similar to those of strings and
+numerals.
 
@@
 
-The $toTree$ function shown below transforms Core expressions, described in Appendix \ref{sec:core} to rose trees.
-We follow the presentation in \cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees.
-$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
-$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
+The $toTree$ function shown below transforms Core expressions, described in
+Appendix \ref{sec:core} to rose trees. We follow the presentation in
+\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is
+a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
+$T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the
+\emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where
+$n_T = 0$ are \emph{leaves}.
 
-The recursive definition is mostly routine; each repeated element (shown as $\dots$) has an example to indicate their handling, e.g.
-for $\CRec$ we apply $toTree$ to each $e_i$.
-We ignore values of $\mathcal{D}$, since constructors have no internal structure for us to compare; they can only be compared based on their types, which we do not currently support.
-We also ignore values from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$, and we conjecture that the effect of such ``magic values'' on clustering real code is low.
+The recursive definition is mostly routine; each repeated element (shown as
+$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
+$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
+have no internal structure for us to compare; they can only be compared based on
+their types, which we do not currently support. We also ignore values from
+$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
+and we conjecture that the effect of such ``magic values'' on clustering real
+code is low.
 
@@
 %\end{figure}
-
-\fi

### 2019-04-26 Bucketing/background.tex
More fleshing out

@@
 corpus of theorems as a \emph{ground truth} against which to compare results.
-Prior work has used libraries of the \textsc{Isabelle/HOL} proof assistant as a
-ground
+Existing work has used libraries of the \textsc{Isabelle/HOL} proof assistant as
+a ground
 truth~\cite{Montano-Rivas.McCasland.Dixon.ea:2012,johansson2009isacosy,claessen2013automating},
@@
 
-Our approach to scaling up \qspec{} takes inspiration from two sources.
-The first is relevance filtering, which makes expensive algorithms used in theorem proving more practical by only considering clauses deemed ``relevant'' to the problem \cite{meng2009lightweight}.
+Our approach to scaling up \qspec{} takes inspiration from two sources. The
+first is relevance filtering, which makes expensive algorithms used in theorem
+proving more practical by only considering clauses deemed ``relevant'' to the
+problem \cite{meng2009lightweight}.
 
-Relevance is determined by comparing clauses to the target theorem, but theory exploration does not have such a distinguished term.
-Instead, we are interested in relationships between \emph{all} terms in a signature, and hence we need a different algorithm for considering the relevance of \emph{all terms} to \emph{all other terms}.
+Relevance is determined by comparing clauses to the target theorem, but theory
+exploration does not have such a distinguished term. Instead, we are interested
+in relationships between \emph{all} terms in a signature, and hence we need a
+different algorithm for considering the relevance of \emph{all terms} to
+\emph{all other terms}.
 
-A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way.
-Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.
+A natural fit for this task is \emph{clustering}, which attempts to group
+similar inputs together in an unsupervised way. Based on their success in
+discovering relationships and patterns between expressions in Coq and ACL2 (in
+the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
+methods can fulfil the role of relevance filters for theory exploration:
+intelligently breaking up large signatures into smaller ones more amenable to
+brute force enumeration, such that related expressions are explored together.
 
-Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering, implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's algorithm \cite{lloyd1982least}, with randomly-selected input elements as the initial clusters.
-Rather than relying on the user to provide the number of clusters $k$, we use the ``rule of thumb'' given in \cite[pp.
-365]{mardia1979multivariate} of clustering $n$ data points into $k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
+Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering,
+implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's
+algorithm \cite{lloyd1982least}, with randomly-selected input elements as the
+initial clusters. Rather than relying on the user to provide the number of
+clusters $k$, we use the ``rule of thumb'' given in
+\cite[pp. 365]{mardia1979multivariate} of clustering $n$ data points into
+$k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
 

### 2019-04-26 Bucketing/contributions.tex
More fleshing out

@@
 
-Rather than clustering Core expressions directly, we first perform \emph{feature extraction} to transform them into numeric \emph{feature vectors}.
-We adapt the methodology of \emph{recurrent clustering} proposed in \cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature extraction and clustering into a single recursive algorithm, such that the feature vector of an expression depends on the feature vectors of those expressions it references.
-We suggest a new recurrent clustering and feature extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity and differences to those of ML4PG and ACL2(ml).
+Rather than clustering Core expressions directly, we first perform \emph{feature
+  extraction} to transform them into numeric \emph{feature vectors}. We adapt
+the methodology of \emph{recurrent clustering} proposed in
+\cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature
+extraction and clustering into a single recursive algorithm, such that the
+feature vector of an expression depends on the feature vectors of those
+expressions it references. We suggest a new recurrent clustering and feature
+extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and
+compare its similarity and differences to those of ML4PG and ACL2(ml).
 
-We describe our algorithm in two stages: the first transforms the nested structure of expressions into a flat vector representation; the second converts the discrete symbols of Core syntax into features (real numbers), which we will denote using the function $\phi$.
+We describe our algorithm in two stages: the first transforms the nested
+structure of expressions into a flat vector representation; the second converts
+the discrete symbols of Core syntax into features (real numbers), which we will
+denote using the function $\phi$.
 
@@
 
-Our recurrent clustering algorithm is based on k-means clustering, which considers the elements of a feature vector to be \emph{orthogonal}.
-Hence we must ensure that similar expressions not only give rise to similar numerical values, but crucially that these values appear \emph{at the same position} in the feature vectors.
-Since different patterns of nesting can alter the ``shape'' of expressions, simple traversals (breadth-first, depth-first, post-order, etc.) may cause features from equivalent sub-expressions to be mis-aligned.
-For example, consider the following expressions, which represent pattern-match clauses with different patterns but the same body (\hs{\vlocal{y}}):
+Our recurrent clustering algorithm is based on k-means clustering, which
+considers the elements of a feature vector to be \emph{orthogonal}. Hence we
+must ensure that similar expressions not only give rise to similar numerical
+values, but crucially that these values appear \emph{at the same position} in
+the feature vectors. Since different patterns of nesting can alter the ``shape''
+of expressions, simple traversals (breadth-first, depth-first, post-order, etc.)
+may cause features from equivalent sub-expressions to be mis-aligned. For
+example, consider the following expressions, which represent pattern-match
+clauses with different patterns but the same body (\hs{\vlocal{y}}):
 
@@
 
-If we traverse these expressions in breadth-first order, converting each token to a feature using $\phi$ and padding to the same length with $0$, we would get the following feature vectors:
+If we traverse these expressions in breadth-first order, converting each token
+to a feature using $\phi$ and padding to the same length with $0$, we would get
+the following feature vectors:
 
@@
 \begin{sloppypar}
-Finally, matrices are turned into vectors by concatenating the rows from top to bottom, hence Figure \ref{fig:matrixexample} will produce a vector beginning $(\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, \dots$.
+  Finally, matrices are turned into vectors by concatenating the rows from top
+  to bottom, hence Figure \ref{fig:matrixexample} will produce a vector
+  beginning
+  $(\feature{\CLam} ,
+    0               ,
+    0               ,
+    0               ,
+    0               ,
+    0               ,
+    \feature{\id{a}},
+    \feature{\CCase},
+    0               ,
+    0               ,
+    0               ,
+    0               ,
+    \feature{\CVar} ,
+    \feature{\id{b}},
+    \feature{\CAlt} ,
+    \feature{\CAlt} ,
+    \dots$.
 \end{sloppypar}
@@
 
-We now define the function $\phi$, which turns terminal symbols of Core syntax into features (real numbers).
-For known language features, such as $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14} in Coq.
-We use a constant $\alpha$ to separate these values from those of other tokens (e.g.
-identifiers), but the order is essentially arbitrary: \footnote{In \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix} and \coq{cofix} are grouped together to reduce redundancy; we do not group tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and \CGlobal.}
+We now define the function $\phi$, which turns terminal symbols of Core syntax
+into features (real numbers). For known language features, such as
+$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
+assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14}
+in Coq. We use a constant $\alpha$ to separate these values from those of other
+tokens (e.g. identifiers), but the order is essentially arbitrary: \footnote{In
+  \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix}
+  and \coq{cofix} are grouped together to reduce redundancy; we do not group
+  tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and
+  \CGlobal.}
 
@@
 
-To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which gives equal values for $\alpha$-equivalent expressions (i.e.
-renaming an identifier shouldn't affect the feature).
-To do this, we use the \emph{de Bruijn index} of the identifier \cite{de1972lambda}, denoted $i_l$:
+To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which
+gives equal values for $\alpha$-equivalent expressions (i.e. renaming an
+identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
+  index} of the identifier \cite{de1972lambda}, denoted $i_l$:
 
@@
 
-We discard the contents of literals and constructor identifiers when converting to rose trees, so the only remaining case is global identifiers $\mathcal{G}$.
-Since these are declared \emph{outside} the body of an expression, we cannot perform the same indexing trick as for local identifiers.
-We also cannot directly encode the form of the identifiers, e.g.
-using a scheme like G{\"o}del numbering, since this is essentially arbitrary and has no effect on their semantic meaning (referencing other expressions).
+We discard the contents of literals and constructor identifiers when converting
+to rose trees, so the only remaining case is global identifiers
+$\mathcal{G}$. Since these are declared \emph{outside} the body of an
+expression, we cannot perform the same indexing trick as for local
+identifiers. We also cannot directly encode the form of the identifiers,
+e.g. using a scheme like G{\"o}del numbering, since this is essentially
+arbitrary and has no effect on their semantic meaning (referencing other
+expressions).
 
-Instead, we use the approach taken in the latest versions of ML4PG and encode global identifiers \emph{indirectly}, by looking up the expressions which they \emph{reference}:
+Instead, we use the approach taken in the latest versions of ML4PG and encode
+global identifiers \emph{indirectly}, by looking up the expressions which they
+\emph{reference}:
 
@@
 
-Where $C_i$ are our clusters (in some arbitrary order).
-This is where the recurrent nature of the algorithm appears: to determine the contents of $C_i$, we must perform k-means clustering \emph{during} feature extraction; yet that clustering step, in turn, requires that we perform feature extraction.
+Where $C_i$ are our clusters (in some arbitrary order). This is where the
+recurrent nature of the algorithm appears: to determine the contents of $C_i$,
+we must perform k-means clustering \emph{during} feature extraction; yet that
+clustering step, in turn, requires that we perform feature extraction.
 
-For this recursive process to be well-founded, we perform a topological sort on declarations based on their dependencies (the expressions they reference).
-In this way, we can avoid looking up expressions which haven't been clustered yet.
-To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth} to produce a sorted list of \emph{strongly connected components} (SCCs), where each SCC is a mutually-recursive sub-set of the declarations.
-If an identifier cannot be found amongst those clustered so far, it must appear in the same SCC as the expression we are processing; hence we give that identifier the constant feature value $f_{recursion}$.
-
-By working through the sorted list of SCCs, storing the features of each top-level expression as they are calculated, our algorithm can be computed \emph{iteratively} rather than recursively, as shown in Algorithm \ref{alg:recurrent}.
+For this recursive process to be well-founded, we perform a topological sort on
+declarations based on their dependencies (the expressions they reference). In
+this way, we can avoid looking up expressions which haven't been clustered
+yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth}
+to produce a sorted list of \emph{strongly connected components} (SCCs), where
+each SCC is a mutually-recursive sub-set of the declarations. If an identifier
+cannot be found amongst those clustered so far, it must appear in the same SCC
+as the expression we are processing; hence we give that identifier the constant
+feature value $f_{recursion}$.
 
@@
   \begin{algorithmic}[1]
-    \Require List $d$ contains SCCs of (identifier, expression) pairs, in dependency order.
+    \Require List $d$ contains SCCs of (identifier, expression) pairs, in
+    dependency order.
     \Procedure{RecurrentCluster}{}
@@
 
+By working through the sorted list of SCCs, storing the features of each
+top-level expression as they are calculated, our algorithm can be computed
+\emph{iteratively} rather than recursively, as shown in Algorithm
+\ref{alg:recurrent}.
+
 \iffalse
@@
 \begin{equation}
-  d = [\{(\hs{plus}, \dots)\}, \{(\hs{odd}, \dots), (\hs{even}, \dots)\}, \{(\hs{mult}, \dots)\}]
+  d = [\{(\hs{plus}, \dots)\},
+       \{(\hs{odd}, \dots), (\hs{even}, \dots)\},
+       \{(\hs{mult}, \dots)\}]
 \end{equation}
@@
 \begin{itemize}
-  \item The first iteration through \textsc{RecurrentCluster}'s loop will set $scc \gets \{(\hs{plus}, \dots)\}$.
-  \item With $i = \hs{plus}$ and $e$ as its Core expression, calculating $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$ will become $f_{recursion}$ (since \hs{plus} doesn't appear in $\vect{C}$).
-  \item The call to $kMeans$ will produce $\vect{C} \gets [\{\hs{plus}\}]$, i.e. a single cluster containing \hs{plus}.
-  \item The next iteration will set $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
-  \item With $i = \hs{odd}$ and $e$ as its Core expression, the call to \hs{even} will result in $f_{recursion}$.
+\item The first iteration through \textsc{RecurrentCluster}'s loop will set
+  $scc \gets \{(\hs{plus}, \dots)\}$.
+\item With $i = \hs{plus}$ and $e$ as its Core expression, calculating
+  $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$
+  will become $f_{recursion}$ (since \hs{plus} doesn't appear in $\vect{C}$).
+\item The call to $kMeans$ will produce $\vect{C} \gets [\{\hs{plus}\}]$, i.e. a
+  single cluster containing \hs{plus}.
+\item The next iteration will set
+  $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
+\item With $i = \hs{odd}$ and $e$ as its Core expression, the call to \hs{even}
+  will result in $f_{recursion}$.
 \item Likewise for the call to $\hs{even}$ when $i = \hs{odd}$.
-  \item Since the feature vectors for \hs{odd} and \hs{even} will be identical, $kMeans$ will put them in the same cluster.
-To avoid the degenerate case of a single cluster, for this example we will assume that $k = 2$; in which case the other cluster must contain \hs{plus}.
-Their order is arbitrary, so one possibility is $\vect{C} = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
-  \item Finally \hs{mult} will be clustered. The recursive call will become $f_{recursion}$ whilst the call to \hs{plus} will become $2 + 3 \alpha$, since $\hs{plus} \in C_2$.
-  \item \begin{sloppypar}Again assuming that $k = 2$, the resulting clusters will be \mbox{$\vect{C} \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus}, \hs{mult}\}]$}.\end{sloppypar}
+\item Since the feature vectors for \hs{odd} and \hs{even} will be identical,
+  $kMeans$ will put them in the same cluster. To avoid the degenerate case of a
+  single cluster, for this example we will assume that $k = 2$; in which case
+  the other cluster must contain \hs{plus}. Their order is arbitrary, so one
+  possibility is $\vect{C} = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
+\item Finally \hs{mult} will be clustered. The recursive call will become
+  $f_{recursion}$ whilst the call to \hs{plus} will become $2 + 3 \alpha$, since
+  $\hs{plus} \in C_2$.
+\item \begin{sloppypar}Again assuming that $k = 2$, the resulting clusters will
+    be
+    \mbox{$\vect{C} \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus},
+      \hs{mult}\}]$}.\end{sloppypar}
 \end{itemize}
 
-Even in this very simple example we can see a few features of our algorithm emerge.
-For example, \hs{odd} and \hs{even} will always appear in the same cluster, since they only differ in their choice of constructor names (which are discarded by $toTree$) and recursive calls (which are replaced by $f_{recursion}$).
-A more extensive investigation of these features requires a concrete implementation, in particular to pin down values for the parameters such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
-\fi
+Even in this very simple example we can see a few features of our algorithm
+emerge. For example, \hs{odd} and \hs{even} will always appear in the same
+cluster, since they only differ in their choice of constructor names (which are
+discarded by $toTree$) and recursive calls (which are replaced by
+$f_{recursion}$). A more extensive investigation of these features requires a
+concrete implementation, in particular to pin down values for the parameters
+such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.  \fi

### 2019-04-26 Bucketing/evaluation.tex
Rename bucketing paper
More fleshing out

--- 2016-02-29 CICM2016/evaluation.tex
@@
 \label{sec:evaluation}
+\iffalse
 
-We have applied our recurrent clustering algorithm to several scenarios, with mixed results.
-A major difficulty in evaluating these clusters is that we have no ``ground truth'', i.e.
-there is no objectively correct way to compare expressions.
-Instead, we provide a qualitative overview of the more interesting characteristics.
+We have applied our recurrent clustering algorithm to several scenarios, with
+mixed results. A major difficulty in evaluating these clusters is that we have
+no ``ground truth'', i.e. there is no objectively correct way to compare
+expressions. Instead, we provide a qualitative overview of the more interesting
+characteristics.
 
-As a simple example, we clustered (Haskell equivalents of) the running examples used to present ACL2(ml) \cite{heras2013proof}, shown in Figure \ref{fig:lisp}.
-These include tail-recursive and non-tail-recursive implementations of several functions.
-We expect those with similar \emph{structure} to be clustered together, rather than those which implement the same function.
-The results are shown in Figure \ref{fig:haskellcluster}, where we can see the ``\hs{Tail}'' functions clearly distinguished, with little distinction between the tail recursive and na\"{\i}ve implementations.
+As a simple example, we clustered (Haskell equivalents of) the running examples
+used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
+\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
+implementations of several functions. We expect those with similar
+\emph{structure} to be clustered together, rather than those which implement the
+same function. The results are shown in Figure \ref{fig:haskellcluster}, where
+we can see the ``\hs{Tail}'' functions clearly distinguished, with little
+distinction between the tail recursive and na\"{\i}ve implementations.
 
-Next we tested whether these same functions would be clustered together when mixed with seemingly-unrelated functions, in this case 207 functions from Haskell's \hs{text} library.
-In fact, the \hs{helperFib} and \hs{fibTail} functions appeared together in a separate cluster from the rest.
-This was unexpected, with no obvious semantic connection between these two functions and the others in their cluster (although most are recursive, due to the nature of the \hs{text} library).
+Next we tested whether these same functions would be clustered together when
+mixed with seemingly-unrelated functions, in this case 207 functions from
+Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail}
+functions appeared together in a separate cluster from the rest. This was
+unexpected, with no obvious semantic connection between these two functions and
+the others in their cluster (although most are recursive, due to the nature of
+the \hs{text} library).
 
@@
 
-We have also applied our recurrent clustering algorithm to a variety of the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including \texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}.
-Whilst we expected functions with a similar purpose to appear together, such as the various reader and writer functions of \hs{pandoc}, there were always a few exceptions which became separate for reasons which are still unclear.
+We have also applied our recurrent clustering algorithm to a variety of the
+most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
+\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
+\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
+expected functions with a similar purpose to appear together, such as the
+various reader and writer functions of \hs{pandoc}, there were always a few
+exceptions which became separate for reasons which are still unclear.
 
-When clustering the \hs{yesod} Web framework, the clustering did seem to match our intuitions, in particular since all 15 of Yesod's MIME type identifiers appeared in the same cluster.
+When clustering the \hs{yesod} Web framework, the clustering did seem to match
+our intuitions, in particular since all 15 of Yesod's MIME type identifiers
+appeared in the same cluster.
 
-Whilst recurrent clustering has produced results which merit further investigation, the application to theory exploration has yet to be tested empirically.
-This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type class to generate random values for instantiating variables.
-Whilst we can automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not all types have \hs{Arbitrary} instances; those without cannot be given any variables in our signature, which severely limits the possible combinations which can be explored.
-In many cases, no variables can be included at all, leaving just equations involving constants.
-This has so far prevented us from measuring the direct impact on \qspec{} performance, either directly by exploring the sub-sets identified through recurrent clustering, or indirectly by comparing the equations generated by a full brute-force search to our recurrent clusters: those equations relating terms from different clusters would not be discovered by our method.
-It is this ratio of equations found through brute force to those found after narrowing-down by clusters which is one of our key objectives to maximise at this stage; until we begin to pursue the \emph{interestingness} of the properties.
+Whilst recurrent clustering has produced results which merit further
+investigation, the application to theory exploration has yet to be tested
+empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type
+class to generate random values for instantiating variables. Whilst we can
+automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
+all types have \hs{Arbitrary} instances; those without cannot be given any
+variables in our signature, which severely limits the possible combinations
+which can be explored. In many cases, no variables can be included at all,
+leaving just equations involving constants. This has so far prevented us from
+measuring the direct impact on \qspec{} performance, either directly by
+exploring the sub-sets identified through recurrent clustering, or indirectly by
+comparing the equations generated by a full brute-force search to our recurrent
+clusters: those equations relating terms from different clusters would not be
+discovered by our method. It is this ratio of equations found through brute
+force to those found after narrowing-down by clusters which is one of our key
+objectives to maximise at this stage; until we begin to pursue the
+\emph{interestingness} of the properties.
 
-The following less-serious problems were also encountered while applying \textsc{ML4HS} to \textsc{Hackage} packages:
+The following less-serious problems were also encountered while applying
+\textsc{ML4HS} to \textsc{Hackage} packages:
 
 \begin{itemize}
-  \item Some packages, such as \texttt{warp} and \texttt{conduits}, get no declarations to cluster.
-This is because they make all of their declarations privately, e.g.
-in ``internal'' modules, then use separate modules to export the public declarations.
-GHC's renaming phase makes all references to such exports canonical, by pointing them to the private declarations.
-This forces us to ignore such declarations, as \qspec{} will not be able to access them.
+\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
+  declarations to cluster. This is because they make all of their declarations
+  privately, e.g. in ``internal'' modules, then use separate modules to export
+  the public declarations. GHC's renaming phase makes all references to such
+  exports canonical, by pointing them to the private declarations. This forces
+  us to ignore such declarations, as \qspec{} will not be able to access them.
 
-  \item Since we do not support type-level entities, we ignore type classes.
-Unfortunately, this also means ignoring any value-level bindings (AKA ``methods'') which occur in a type class instance.
-Instead of being clustered, these result in references getting $f_{recursion}$ features.
-This is especially noticable in libraries like \hs{scientific}, where only the functions for constructing and destructing numbers in scientific notation are clustered; all of the arithmetic is defined in type classes.
-One difficulty with supporting methods is that their namespace in Core is disjoint from that of regular Haskell identifiers: a transformation layer would be required, along with explicit type annotations to avoid ambiguity.
+\item Since we do not support type-level entities, we ignore type
+  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
+  ``methods'') which occur in a type class instance. Instead of being clustered,
+  these result in references getting $f_{recursion}$ features. This is
+  especially noticable in libraries like \hs{scientific}, where only the
+  functions for constructing and destructing numbers in scientific notation are
+  clustered; all of the arithmetic is defined in type classes. One difficulty
+  with supporting methods is that their namespace in Core is disjoint from that
+  of regular Haskell identifiers: a transformation layer would be required,
+  along with explicit type annotations to avoid ambiguity.
 \end{itemize}
 
-It seems like this recurrent clustering method has promise, although it will require a more thorough exploration of the parameters to obtain more intuitively reliable results.
-These clusters can then be used in several ways to perform theory exploration; the most na\"{\i}ve way being to explore each cluster as \qspec{} signature in its own right.
-\textsc{ML4HS} already provides this functionality, although the lack of test generators severely limits what can be discovered.
+It seems like this recurrent clustering method has promise, although it will
+require a more thorough exploration of the parameters to obtain more intuitively
+reliable results. These clusters can then be used in several ways to perform
+theory exploration; the most na\"{\i}ve way being to explore each cluster as
+\qspec{} signature in its own right. \textsc{ML4HS} already provides this
+functionality, although the lack of test generators severely limits what can be
+discovered.
 
-As alluded to previously, we also have the opportunity to reason by \emph{analogy}.
-Similar to the work on ACL2(ml) \cite{heras2013proof}, we could produce a general ``scheme'' from each equation we find (either through \qspec{} or by data mining test suites); like Isabelle approaches have shown \cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be instantiated to a variety of similar values, in an attempt to find new theorems which are analogues of existing results from a different context.
-``Mutating'' existing theorem statements in such a way would also increase the chance of any result being considered interesting; since it's likely that the unmodified statement was deemed interesting, and the new result would not in general follow as a simple logical consequence.
+As alluded to previously, we also have the opportunity to reason by
+\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
+produce a general ``scheme'' from each equation we find (either through \qspec{}
+or by data mining test suites); like Isabelle approaches have shown
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
+instantiated to a variety of similar values, in an attempt to find new theorems
+which are analogues of existing results from a different context. ``Mutating''
+existing theorem statements in such a way would also increase the chance of any
+result being considered interesting; since it's likely that the unmodified
+statement was deemed interesting, and the new result would not in general follow
+as a simple logical consequence.
+
+\fi

### 2019-04-26 Bucketing/implementation.tex
More fleshing out

--- 2016-03-02 CICM2016/implementation.tex
@@
 
-We provide an implementation of our recurrent clustering algorithm in a tool called \textsc{ML4HS}.
-We obtain Core ASTs from Haskell packages using a plugin for the GHC compiler, which emits a serialised version of each Core definition as it is being compiled.
-This approach is more robust than, for example, parsing source files, since it avoids the complications of preprocessors and language extensions altering the syntax.
+We provide an implementation of our recurrent clustering algorithm in a tool
+called \textsc{ML4HS}. We obtain Core ASTs from Haskell packages using a plugin
+for the GHC compiler, which emits a serialised version of each Core definition
+as it is being compiled. This approach is more robust than, for example, parsing
+source files, since it avoids the complications of preprocessors and language
+extensions altering the syntax.
 
-A post-processing stage determines which Core definitions can be used by \qspec{}, based on their type, visibility (whether they are encapsulated inside their module or visible to importers), etc.
-For each definition, we simply use the GHCi interpreter to check if calling \qspec{} with that argument is well-typed.
-This information, along with type signatures, arity, etc.
-are stored alongside the Core definitions in JSON format.
+A post-processing stage determines which Core definitions can be used by
+\qspec{}, based on their type, visibility (whether they are encapsulated inside
+their module or visible to importers), etc. For each definition, we simply use
+the GHCi interpreter to check if calling \qspec{} with that argument is
+well-typed. This information, along with type signatures, arity, etc. are stored
+alongside the Core definitions in JSON format.
 
-The definitions are then sorted topologically, based on the non-local identifiers appearing in their ASTs, and feature vectors are constructed using a Haskell implementation of the approach described in \S \ref{sec:recurrentclustering}.
-Since the features associated with each identifier may vary between iterations (as the clusters change), we leave the raw identifiers in the vector so their features can be extracted in the correct context.
+The definitions are then sorted topologically, based on the non-local
+identifiers appearing in their ASTs, and feature vectors are constructed using a
+Haskell implementation of the approach described in \S
+\ref{sec:recurrentclustering}. Since the features associated with each
+identifier may vary between iterations (as the clusters change), we leave the
+raw identifiers in the vector so their features can be extracted in the correct
+context.
 
-We implement Algorithm \ref{alg:recurrent} as a simple shell script, which invokes Weka for clustering and associates each Core definition with a cluster number.
-These numbers are used to finish the deferred feature extraction of identifiers, the resulting feature vectors are clustered, and the process is repeated until all SCCs have been processed.
+We implement Algorithm \ref{alg:recurrent} as a simple shell script, which
+invokes Weka for clustering and associates each Core definition with a cluster
+number. These numbers are used to finish the deferred feature extraction of
+identifiers, the resulting feature vectors are clustered, and the process is
+repeated until all SCCs have been processed.
 
-Once the recurrent clustering is complete, we generate a string of Haskell code for each of the final clusters, which will invoke \qspec{} with a suitable signature. This involves:
+Once the recurrent clustering is complete, we generate a string of Haskell code
+for each of the final clusters, which will invoke \qspec{} with a suitable
+signature. This involves:
 
 \begin{itemize}
-  \item{Monomorphising}: If a value has polymorphic type, e.g.
-\hs{equal :: forall t.
-t -> t -> Bool}, we must choose a concrete representation to use (in this case for \hs{t}), in order to know which random generator to use.
-We take the approach used in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}.
-Any values where this is invalid, such as those with incompatible class constraints (e.g.
-\hs{forall t.
-IsString t => t}, where \hs{Integer} does not implement \hs{IsString}) will not be included in the signature (this is checked at the AST post-processing stage).
+\item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall
+    t. t -> t -> Bool}, we must choose a concrete representation to use (in this
+  case for \hs{t}), in order to know which random generator to use. We take the
+  approach used in \qcheck{} by attempting to instantiate all type variables to
+  \hs{Integer}. Any values where this is invalid, such as those with
+  incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where
+  \hs{Integer} does not implement \hs{IsString}) will not be included in the
+  signature (this is checked at the AST post-processing stage).
 
-  \item{Qualifying}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity.
-There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules.
-In such cases the exploration process simply aborts.
+\item{Qualifying}: All names are \emph{qualified} (prefixed by their module's
+  name), to avoid most ambiguity. There is still the possibility that multiple
+  packages will declare modules of the same name, although this is rare as it
+  causes problems for any Haskell programmer trying to use those modules. In
+  such cases the exploration process simply aborts.
 
-  \item{Appending variables}: Once a \qspec{} theory has been defined containing all of the given terms, we inspect the types it references and append three variables for each to the theory (enough to discover laws such as associativity, but not too many to overflow the limit of \qspec{}'s exhaustive search).
+\item{Appending variables}: Once a \qspec{} theory has been defined containing
+  all of the given terms, we inspect the types it references and append three
+  variables for each to the theory (enough to discover laws such as
+  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
+  search).
 
-  \item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is that all required packages and modules must be provided up-front, usually by specification in a Cabal file.
-Since \textsc{MLSpec} builds signatures \emph{dynamically}, depending on the cluster information it is given, we do not know what packages it may need.
-To work around this problem, we evaluate these strings of Haskell using a custom library called \texttt{nix-eval}.
-This uses the Nix package manager to obtain all of the required packages and make them available to GHC.
+\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
+  that all required packages and modules must be provided up-front, usually by
+  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
+  \emph{dynamically}, depending on the cluster information it is given, we do
+  not know what packages it may need. To work around this problem, we evaluate
+  these strings of Haskell using a custom library called \texttt{nix-eval}. This
+  uses the Nix package manager to obtain all of the required packages and make
+  them available to GHC.
 
@@
 
-The equations resulting from evaluating these strings are collected and outputted as JSON values, to ease further processing (e.g. displaying in some form, sending to a theorem prover, etc.).
+The equations resulting from evaluating these strings are collected and
+outputted as JSON values, to ease further processing (e.g. displaying in some
+form, sending to a theorem prover, etc.).

### 2019-04-26 Bucketing/related.tex
More fleshing out

--- 2016-03-02 CICM2016/related.tex
@@
 
-We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements.
-This general idea dates back at least to the 1970s, most famously to work by Lenat \cite{lenat1977automated}, though the theory exploration tools used in functional programming trace their roots back to the interactive \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
+We briefly described theory exploration in \S \ref{sec:theoryexploration}, as
+the task of discovering \emph{new} theorems in a software or proof library,
+rather than proving/disproving user-provided statements. This general idea dates
+back at least to the 1970s, most famously to work by Lenat
+\cite{lenat1977automated}, though the theory exploration tools used in
+functional programming trace their roots back to the interactive
+\textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
 
-As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012} \cite{johansson2009isacosy} \cite{Hipster}.
-Since \qspec{} is used for conjecture generation in many of these systems, they may also benefit from our machine learning approach.
+As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has
+been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
+\cite{johansson2009isacosy} \cite{Hipster}. Since \qspec{} is used for
+conjecture generation in many of these systems, they may also benefit from our
+machine learning approach.
 
@@
 
-The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to reduce the search space of a theorem proving task by removing redundant or irrelevant information (axioms, definitions, lemmas, etc.) from the input.
-Our restriction of theory exploration to intelligently-selected clusters of symbols, rather than whole libraries at a time, is essentially the same idea applied in a novel setting.
+The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to
+reduce the search space of a theorem proving task by removing redundant or
+irrelevant information (axioms, definitions, lemmas, etc.) from the input. Our
+restriction of theory exploration to intelligently-selected clusters of symbols,
+rather than whole libraries at a time, is essentially the same idea applied in a
+novel setting.
 
-The technique is used in Isabelle's Sledgehammer tool, during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only a sub-set of relevant clauses are included.
-This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a clause is relevant: how do we know what will be required, before we have the proof?
+The technique is used in Isabelle's Sledgehammer tool, during its translation of
+Isabelle/HOL theories to statements in first order logic: rather than
+translating the entire theory, only a sub-set of relevant clauses are
+included. This reduces the size of the problem and speeds up the proof search,
+but it creates the new problem of determining when a clause is relevant: how do
+we know what will be required, before we have the proof?
 
-The original approach uses the proportion of symbols which a clause has in common with the problem statement \cite{meng2009lightweight} (plus various heuristics), although various alternatives have been proposed in the literature \cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.
+The original approach uses the proportion of symbols which a clause has in
+common with the problem statement \cite{meng2009lightweight} (plus various
+heuristics), although various alternatives have been proposed in the literature
+\cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.
 
@@
 
-Our recurrent clustering algorithm is similar to that of ML4PG, as our transformation maps the elements of a syntax tree to distinct cells in a matrix.
-In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree elements: providing, for each level of the tree, the number of variables, nullary symbols, unary symbols, etc.
+Our recurrent clustering algorithm is similar to that of ML4PG, as our
+transformation maps the elements of a syntax tree to distinct cells in a
+matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree
+elements: providing, for each level of the tree, the number of variables,
+nullary symbols, unary symbols, etc.
 
-Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it also ignores types.
-This is because obtaining the type of each component of a Core definition is more difficult than in ML4PG (which queries the current \textsc{Proof General} state), and is left for future work.
+Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it
+also ignores types. This is because obtaining the type of each component of a
+Core definition is more difficult than in ML4PG (which queries the current
+\textsc{Proof General} state), and is left for future work.
 
-The way we \emph{use} our clusters to inform theory exploration is actually more similar to that of ACL2(ml) than ML4PG.
-ML4PG can either present clusters to the user for inspection, or produce automata for recreating proofs.
-In ACL2(ml), the clusters are used to restrict the search space of a proof search, much like we restrict the scope of theory exploration.
+The way we \emph{use} our clusters to inform theory exploration is actually more
+similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
+user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
+clusters are used to restrict the search space of a proof search, much like we
+restrict the scope of theory exploration.
 
@@
 
-One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures.
-This can lead to nested expressions of arbitrary depth, which are difficult to compare in numerical ways.
-One common approach to this problem is to represent such structures as \emph{sequences}.
-\emph{Recurrent neural networks} (RNNs) are a popular choice for processing sequences, especially when combined with mechanisms such as \emph{long short-term memory} (LSTM) for preserving information across long sequences \cite{hochreiter1997long}.
-Such systems have been used, for example, to parse and execute computer programs \cite{zaremba2014learning}.
-However, learning to parse sequences seems inefficient considering that we already have correctly-parsed ASTs.
+One major difficulty when applying statistical machine learning algorithms to
+\emph{languages}, such as Haskell, is the appearance of recursive
+structures. This can lead to nested expressions of arbitrary depth, which are
+difficult to compare in numerical ways. One common approach to this problem is
+to represent such structures as \emph{sequences}. \emph{Recurrent neural
+  networks} (RNNs) are a popular choice for processing sequences, especially
+when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
+preserving information across long sequences \cite{hochreiter1997long}. Such
+systems have been used, for example, to parse and execute computer programs
+\cite{zaremba2014learning}. However, learning to parse sequences seems
+inefficient considering that we already have correctly-parsed ASTs.
 
-Whilst neural networks have been applied directly to recursive structures \cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more popular approach is to use \emph{kernel methods} \cite{bakir2007predicting}.
-These are promising as a more principled alternative to our current hand-crafted translation of ASTs to vectors.
+Whilst neural networks have been applied directly to recursive structures
+\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
+popular approach is to use \emph{kernel methods}
+\cite{bakir2007predicting}. These are promising as a more principled alternative
+to our current hand-crafted translation of ASTs to vectors.

### 2019-04-26 Bucketing/sizes.tex
More details of toxic definitions

@@
 
-Figure~\ref{fig:survival} shows a Kaplan-Meier plot of \qspec{} running
-times, when given inputs containing different numbers of definitions. There is a
-clear pattern, with many runs finishing quickly and the remainder occupying a
-``long tail'' which we cut off after 300 seconds (based on preliminary
-experiments, which showed little difference in survival between 300 seconds and
-1 hour).
+Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \qspec{} running
+times, when given inputs containing different numbers of definitions. Many runs
+finish quickly, with the remainder occupying a ``long tail'' which we cut off
+after 300 seconds\footnote{Chosen based on preliminary experiments, which showed
+  little difference in survival between 300 seconds and 1 hour}.
 
-There is an overall trend for larger inputs to have more runs time-out, except
-for the very smallest inputs (which are more constrained by the sampling
-procedure). This relationship is linear, which can be explained if the presence
-of certain definitions is causing the exploration process to fail (since larger
-inputs have a higher chance of containing one of these definitions).
+The number of definitions in the input is linearly correlated with the amount of
+timeouts, except for the very smallest inputs (which are more constrained by the
+sampling procedure). One explanation for this is ``toxic'' definitions, whose
+presence in the input always leads to the exploration failing (with larger
+inputs being more likely to have sampled a toxic definition).
 
+\iffalse
 \begin{figure}
@@
 \end{figure}
+\fi
 
@@
   \end{minted}
-  \if
+  \fi
 
@@
 functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
-exploration to time out by generating very large outputs.
+exploration to time out by either exhausting the RAM with too many expressions
+to explore, % TODO{2019-04-27} CAN QUICKSPEC2 AVOID THIS?
+or by generating such deeply-nested outputs that comparing them takes too long.
+% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?
 
-\texttt{mul3} and \texttt{mul3acc} are rather pathological implementations of
-multiplication with an accumulator parameter, with many (non-tail) recursive
-calls. The \texttt{op} function appears in files named \texttt{weird\_nat\_op},
-which assert its commutativity and associativity. Finally, the \texttt{mult2}
-and \texttt{qexp} functions are standard tail-recursive definitions of
-multiplication and exponentiation, respectively.
+The \texttt{mul3} and \texttt{mul3acc} definitions are rather pathological
+implementations of multiplication with an accumulator parameter, with many
+(non-tail) recursive calls. The \texttt{op} function appears in files named
+\texttt{weird\_nat\_op}, which assert its commutativity and associativity.
+Finally, the \texttt{mult2} and \texttt{qexp} functions are standard
+tail-recursive definitions of multiplication and exponentiation, respectively.
+All of these functions have an extra ``accumulator'' argument, which increases
+the number of possible expressions to explore compared to those without.
 

### 2019-04-26 PhDThesis/bucketing.tex
Copypasta bucketing paper into thesis (needs reworking)

\section{Introduction}

\emph{(Mathematical) Theory Exploration} (TE) describes an open-ended search for
interesting properties (proven or merely conjectured, depending on the approach)
of a given set of formal definitions, such as a mathematical theory, a
scientific model or a software library. It has been applied to generating
``background lemmas'' for the interactive proof assistant
\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and automated
theorem provers~\cite{claessen2013automating}, as well as for generating
specifications for (pure functional) programming
libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.

Despite the clever search strategies employed by these tools, they all rely on
enumerating combinations of definitions in some way, which can cause infeasible
running times on larger inputs (in our experience, a few dozen definitions). To
avoid this, users of these systems must carefully select only small subsets of
their definitions to explore at a time.

We dub such selection the \emph{bucketing problem} for theory exploration and
analyse its effect on the running time of theory exploration tools on a
particular problem set, and on the quality of the statements they are able to
generate. An automated solution to this bucketing problem would accept a large
number of definitions and efficiently select sub-sets which are both small
enough to reasonably explore, whilst still enabling interesting properties to be
discovered.

Our contributions are:

\begin{enumerate}
  \item Defining the \emph{bucketing problem} for theory exploration.
  \item Statistical analysis of theory exploration performance on a large
    problem set, derived from existing theorem proving problem sets.
  \item Empirical analysis of the effect of bucketing on the quantity of
    desirable properties attainable by theory exploration.
  \item Experimental application of machine learning to the bucketing problem,
    via unsupervised \emph{clustering}.
  % \item A novel feature extraction method for transforming Haskell expressions
  %   into a form amenable to off-the-shelf learning algorithms.
  % \item An implementation of these feature extraction and theory exploration
  %   approaches.
  % \item A comparison of our methods with existing approaches, both for theory
  %   exploration in Haskell, and for machine learning in other languages.
\end{enumerate}

We describe theory exploration in more detail, as well as the \qspec{} tool we
use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
bucketing problem more formally in \S~\ref{sec:bucketing} and measure its
effect on theory exploration systems in \S~\ref{sec:implementation}. In
\S~\ref{sec:clustering} we highlight the difference betwee bucketing and the
well-known \emph{clustering problem} from machine learning, by demonstrating
that clustering algorithms perform poorly at bucketing.  A variety of related
work is surveyed in \S~\ref{sec:related} and we conclude in
\S~\ref{sec:conclusion} with potential directions for future research.

\section{Background}
\label{sec:background}

\subsection{Theory Exploration}
\label{sec:theoryexploration}

Theory exploration is the task of taking a \emph{signature} of definitions in
some formal system (for example a programming language) and automatically
generating a set of formal statements (properties) involving those
definitions. These may be conjectures or theorems (proven either by sending
conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps); in either case, these statements
must also be ``interesting'' in some way. It is beyond the scope of this paper
to define what makes a mathematical statement ``interesting'', but this problem
has been tackled extensively in the literature~\cite{colton2000notion}.

We will follow the tradition of prior practitioners and use an existing
corpus of theorems as a \emph{ground truth} against which to compare results.
Existing work has used libraries of the \textsc{Isabelle/HOL} proof assistant as
a ground
truth~\cite{Montano-Rivas.McCasland.Dixon.ea:2012,johansson2009isacosy,claessen2013automating},
but these only contained a few definitions and theorems each. To analyse
behaviour on a larger scale, as required to study the effects of bucketing, we
have chosen the \emph{Tons of Inductive Problems} (TIP) problem set for
automated theorem provers~\cite{claessen2015tip}.  Reasons for choosing TIP
include:

\begin{itemize}
\item Each problem in TIP clearly distinguishes between the statement to be
  proved and the definitions involved, such that the two can be easily and
  meaningfully separated to act as inputs and desired outputs in a TE setting.
  This would not be the case for problem sets like
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item The TIP problem set includes a tool to translate from the language it
  uses to define problems (a variant of SMTLIB) into various other languages
  including Haskell and Isabelle, for which TE tools already exist.
\item TIP problems are relevant to the domains we are interested in, including
  pure functional programming, higher-order functions and inductively defined
  datatypes. This makes TIP more attractive than problem sets for first-order
  languages/logics such as TPTP~\cite{sutcliffe2009tptp}.
\item Many of TIP's problem statements re-use the same definitions, which makes
  it more useful as a ground-truth than a mixture of unrelated problems. This is
  because non-membership in the ground truth will be treated as being
  ``uninteresting''.
\end{itemize}

All existing theory exploration systems rely on brute force enumeration to some
degree. We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
Haskell code. These may be fed into another tool, such as \hspec{}, for
subsequent proving. We have used version 1 of \qspec{} in our experiments, due
to its availability, stability and tooling integration, and we will drop the
version number from now on. At the time of writing there is a \qspec{} version 2
available, which has a much improved generation procedure that is significantly
faster than its predecessor. However, it still scales poorly in our experience,
and has less integration with the tooling we have used. We hence leave analysis
of \qspec{} 2 as future work, pending the necessary infrastructure changes that
would require.

\qspec{} (version 1) is written in the Haskell programming language, and works
with definitions which are also written in Haskell. Some simple Haskell
definitions are shown in Figure~\ref{fig:haskellexample}. The following
procedure is used to generate conjectures, which are both plausible (due to
testing on many examples) and potentially interesting (due to being mutually
irreducible):

\begin{figure}
  \centering
  \begin{minted}{haskell}
    -- A datatype with two constructors
    data Bool = True | False

    -- A recursive datatype (S requires a Nat as argument)
    data Nat = Z | S Nat

    -- A function turning a Bool into a Bool
    not :: Bool -> Bool
    not True  = False
    not False = True

    -- A pair of mutually recursive functions

    odd :: Nat -> Bool
    odd    Z  = False
    odd (S n) = even n

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{minted}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some
    simple function definitions (with type annotations).}
  \label{fig:haskellexample}
\end{figure}

\begin{enumerate}
\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
  generates a list $terms$ containing the constants (including functions) from
  $\Sigma$, the variables from $V$ and type-correct function applications
  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
  is finite, function applications are only nested up to a specified depth (by
  default, 3).
\item The elements of $terms$ are grouped into equivalence classes, based on
  their type.
\item The equivalence of terms in each class is tested using \qcheck{}:
  variables are instantiated to particular values, generated randomly, and the
  resulting closed expressions are evaluated and compared for equality.
\item If a class is found to have non-equal members, it is split up to separate
  those members.
\item The previous steps of testing and splitting are repeated until the classes
  stabilise (i.e. no differences have been observed for some specified number of
  repetitions).
\item For each class, one member is selected and equations are conjectured that
  it is equal to each of the other members.
\item A congruence closure algorithm is applied to these equations, to discard
  any which are implied by the others.
\end{enumerate}

As an example, giving the definitions from Figure~\ref{fig:haskellexample} to
\qspec{}, along with suitable comparison functions, random data generators and
variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
them up, and repeat until no more splits are found. Equations are then generated
between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
instance of the more general \hs{b == not (not b)}). The remaining equations
(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
output, either for the user to peruse or for another system to process, like the
\textsc{HipSpec} automated theorem prover.

Note that we must be able to generate values of a type in order to include
variables of that type. Also, only equivalence classes whose element type can be
compared are split and conjectured about in this way; in particular, functions
(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
equivalence class, but they will not be subject to testing and comparison since
Haskell has no generic notion of function equality~\footnote{In
  fact \qspec{} compares expressions using a total ordering rather than equality;
  this is even more restrictive than requiring a notion of equality, but reduces
  the number of required comparisons}. \qspec{} can hence conjecture that two
functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
that the functions themselves are equal (e.g. \hs{f == g}).

Although complete, this enumeration approach is wasteful: many terms are
unlikely to appear in theorems, which requires careful choice by the user of
what to include in the signature. For example, we know that addition and
multiplication are closely related, and hence obey many algebraic laws. Our
machine learning technique aims to predict these kinds of relations between
functions, so we can create small signatures which can be explored quickly, yet
have the potential to give rise to many equations.

\subsection{Clustering}
\label{sec:clustering}

Our approach to scaling up \qspec{} takes inspiration from two sources. The
first is relevance filtering, which makes expensive algorithms used in theorem
proving more practical by only considering clauses deemed ``relevant'' to the
problem \cite{meng2009lightweight}.

Relevance is determined by comparing clauses to the target theorem, but theory
exploration does not have such a distinguished term. Instead, we are interested
in relationships between \emph{all} terms in a signature, and hence we need a
different algorithm for considering the relevance of \emph{all terms} to
\emph{all other terms}.

A natural fit for this task is \emph{clustering}, which attempts to group
similar inputs together in an unsupervised way. Based on their success in
discovering relationships and patterns between expressions in Coq and ACL2 (in
the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
methods can fulfil the role of relevance filters for theory exploration:
intelligently breaking up large signatures into smaller ones more amenable to
brute force enumeration, such that related expressions are explored together.

Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering,
implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's
algorithm \cite{lloyd1982least}, with randomly-selected input elements as the
initial clusters. Rather than relying on the user to provide the number of
clusters $k$, we use the ``rule of thumb'' given in
\cite[pp. 365]{mardia1979multivariate} of clustering $n$ data points into
$k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.

\subsection{The Bucketing Problem}
\label{sec:bucketing}

% TODO{2019-04-19}

\section{Contributions}
\label{sec:contributions}

\subsection{Recurrent Clustering}
\label{sec:recurrentclustering}

Rather than clustering Core expressions directly, we first perform \emph{feature
  extraction} to transform them into numeric \emph{feature vectors}. We adapt
the methodology of \emph{recurrent clustering} proposed in
\cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature
extraction and clustering into a single recursive algorithm, such that the
feature vector of an expression depends on the feature vectors of those
expressions it references. We suggest a new recurrent clustering and feature
extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and
compare its similarity and differences to those of ML4PG and ACL2(ml).

We describe our algorithm in two stages: the first transforms the nested
structure of expressions into a flat vector representation; the second converts
the discrete symbols of Core syntax into features (real numbers), which we will
denote using the function $\phi$.

\subsubsection{Expressions to Vectors}
\label{sec:expressionstovectors}

Our recurrent clustering algorithm is based on k-means clustering, which
considers the elements of a feature vector to be \emph{orthogonal}. Hence we
must ensure that similar expressions not only give rise to similar numerical
values, but crucially that these values appear \emph{at the same position} in
the feature vectors. Since different patterns of nesting can alter the ``shape''
of expressions, simple traversals (breadth-first, depth-first, post-order, etc.)
may cause features from equivalent sub-expressions to be mis-aligned. For
example, consider the following expressions, which represent pattern-match
clauses with different patterns but the same body (\hs{\vlocal{y}}):

\begin{equation}\label{eq:xy}
  \begin{array}{r@{}l@{}l@{}}
    X\ &=\ \CAlt\ (\CDataAlt\ \id{C})\ & (\vlocal{y}) \\
    Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
  \end{array}
\end{equation}

If we traverse these expressions in breadth-first order, converting each token
to a feature using $\phi$ and padding to the same length with $0$, we would get
the following feature vectors:

\begin{small}
  \begin{equation}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
      breadthFirst(Y)\ &=\ (\feature{\CAlt},\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &)
    \end{array}
  \end{equation}
\end{small}

Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\phi$). These feature vectors might be deemed very dissimilar during clustering, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.

If we were to align these features optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.

The method we use to ``flatten'' expressions, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size $c$ (for \emph{columns}). This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting. Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the constructor identifier $\id{C}$ would not appear in the vector, but this example is still accurate in terms of laying out the features as given.}

\begin{small}
  \begin{equation}\label{eq:flattened}
    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
    \end{array}
  \end{equation}
\end{small}

Here $\frac{1}{2}$ of the original 6 features align, which is more than $breadthFirst$ but not optimal. Both vectors have also been padded by an extra 2 zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.

To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features. Intuitively, these are simply an s-expression representation of the Core syntax, with the feature extraction function $\phi$ mapped over the labels. The full transformation is given in Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the result for the \hs{odd} function.

\begin{figure}
  \centering
  \begin{scriptsize}
      \Tree[ .$\feature{\CLam}$
                $\feature{\id{a}}$
                [ .$\feature{\CCase}$
                     [ .$\feature{\CVar}$
                          [ .$\feature{\CLocal}$
                               $\feature{\id{a}}$ ]]
                     $\feature{\id{b}}$
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CVar}$
                               $\feature{\CConstructor}$ ]]
                     [ .$\feature{\CAlt}$
                          $\feature{\CDataAlt}$
                          [ .$\feature{\CApp}$
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CGlobal}$
                                         $\feature{\id{even}}$ ]]
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CLocal}$
                                         $\feature{\id{n}}$ ]]]
                          $\feature{\id{n}}$ ]]]
  \end{scriptsize}
  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
\end{figure}

\begin{figure}
    \begin{equation*}
      \begin{bmatrix}
        \feature{\CLam}      & 0                       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{a}}     & \feature{\CCase}        & 0                 & 0                   & 0               & 0                \\
        \feature{\CVar}      & \feature{\id{b}}        & \feature{\CAlt}   & \feature{\CAlt}     & 0               & 0                \\
        \feature{\CLocal}    & \feature{\CDataAlt}     & \feature{\CVar}   & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
        \feature{\id{a}}     & \feature{\CConstructor} & \feature{\CVar}   & \feature{\CVar}     & 0               & 0                \\
        \feature{\CGlobal}   & \feature{\CLocal}       & 0                 & 0                   & 0               & 0                \\
        \feature{\id{even}}  & \feature{\id{n}}        & 0                 & 0                   & 0               & 0
      \end{bmatrix}
    \end{equation*}
    \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
    \label{fig:matrixexample}
\end{figure}

These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, read left-to-right, followed by any required padding. These matrices are then either truncated, or padded with rows (on the bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.

\begin{sloppypar}
  Finally, matrices are turned into vectors by concatenating the rows from top
  to bottom, hence Figure \ref{fig:matrixexample} will produce a vector
  beginning
  $(\feature{\CLam} ,
    0               ,
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\id{a}},
    \feature{\CCase},
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\CVar} ,
    \feature{\id{b}},
    \feature{\CAlt} ,
    \feature{\CAlt} ,
    \dots$.
\end{sloppypar}

\subsubsection{Symbols to Features}
\label{sec:symbolstofeatures}

We now define the function $\phi$, which turns terminal symbols of Core syntax
into features (real numbers). For known language features, such as
$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14}
in Coq. We use a constant $\alpha$ to separate these values from those of other
tokens (e.g. identifiers), but the order is essentially arbitrary: \footnote{In
  \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix}
  and \coq{cofix} are grouped together to reduce redundancy; we do not group
  tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and
  \CGlobal.}

\begin{equation} \label{eq:feature}
  \begin{aligned}
    \feature{\CAlt}          &= \alpha      &
    \feature{\CDataAlt}      &= \alpha + 1  &
    \feature{\CLitAlt}       &= \alpha + 2  \\
    \feature{\CDefault}      &= \alpha + 3  &
    \feature{\CNonRec}       &= \alpha + 4  &
    \feature{\CRec}          &= \alpha + 5  \\
    \feature{\CBind}         &= \alpha + 6  &
    \feature{\CLet}          &= \alpha + 7  &
    \feature{\CCase}         &= \alpha + 8  \\
    \feature{\CLocal}        &= \alpha + 9  &
    \feature{\CGlobal}       &= \alpha + 10 &
    \feature{\CConstructor}  &= \alpha + 11 \\
    \feature{\CVar}          &= \alpha + 12 &
    \feature{\CLam}          &= \alpha + 13 &
    \feature{\CApp}          &= \alpha + 14 \\
    \feature{\CType}         &= \alpha + 15 &
    \feature{\CLit}          &= \alpha + 16 &
    \feature{\CLitNum}       &= \alpha + 17 \\
    \feature{\CLitStr}       &= \alpha + 18
  \end{aligned}
\end{equation}

To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which
gives equal values for $\alpha$-equivalent expressions (i.e. renaming an
identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
  index} of the identifier \cite{de1972lambda}, denoted $i_l$:

\begin{equation} \label{eq:localfeature}
  \feature{l} = i_l + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
\end{equation}

We again use $\alpha$ to separate these features from those of other constructs.

We discard the contents of literals and constructor identifiers when converting
to rose trees, so the only remaining case is global identifiers
$\mathcal{G}$. Since these are declared \emph{outside} the body of an
expression, we cannot perform the same indexing trick as for local
identifiers. We also cannot directly encode the form of the identifiers,
e.g. using a scheme like G{\"o}del numbering, since this is essentially
arbitrary and has no effect on their semantic meaning (referencing other
expressions).

Instead, we use the approach taken in the latest versions of ML4PG and encode
global identifiers \emph{indirectly}, by looking up the expressions which they
\emph{reference}:

\begin{equation} \label{eq:globalfeature}
  \feature{g \in \mathcal{G}} =
    \begin{cases}
      i + 3 \alpha \quad & \text{if $g \in C_i$} \\
      f_{recursion}         & \text{otherwise}
    \end{cases}
\end{equation}

Where $C_i$ are our clusters (in some arbitrary order). This is where the
recurrent nature of the algorithm appears: to determine the contents of $C_i$,
we must perform k-means clustering \emph{during} feature extraction; yet that
clustering step, in turn, requires that we perform feature extraction.

For this recursive process to be well-founded, we perform a topological sort on
declarations based on their dependencies (the expressions they reference). In
this way, we can avoid looking up expressions which haven't been clustered
yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth}
to produce a sorted list of \emph{strongly connected components} (SCCs), where
each SCC is a mutually-recursive sub-set of the declarations. If an identifier
cannot be found amongst those clustered so far, it must appear in the same SCC
as the expression we are processing; hence we give that identifier the constant
feature value $f_{recursion}$.

\begin{algorithm}
  \begin{algorithmic}[1]
    \Require List $d$ contains SCCs of (identifier, expression) pairs, in
    dependency order.
    \Procedure{RecurrentCluster}{}
      \State $\vect{C} \gets []$
      \State $DB \gets \varnothing$
      \ForAll{$scc$ \textbf{in} $d$}
        \State $DB \gets DB \cup \{(i, featureVec(e)) \mid (i, e) \in scc\}$
        \State $\vect{C} \gets kMeans(DB)$
      \EndFor
      \Return $\vect{C}$
    \EndProcedure
  \end{algorithmic}
  \caption{Recurrent clustering of Core expressions.}
  \label{alg:recurrent}
\end{algorithm}

By working through the sorted list of SCCs, storing the features of each
top-level expression as they are calculated, our algorithm can be computed
\emph{iteratively} rather than recursively, as shown in Algorithm
\ref{alg:recurrent}.

\iffalse
As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}. A valid topological ordering is given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding Core expressions to save space):

\begin{equation}
  d = [\{(\hs{plus}, \dots)\},
       \{(\hs{odd}, \dots), (\hs{even}, \dots)\},
       \{(\hs{mult}, \dots)\}]
\end{equation}

We can then trace the execution of Algorithm \ref{alg:recurrent} as follows:

\begin{itemize}
\item The first iteration through \textsc{RecurrentCluster}'s loop will set
  $scc \gets \{(\hs{plus}, \dots)\}$.
\item With $i = \hs{plus}$ and $e$ as its Core expression, calculating
  $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$
  will become $f_{recursion}$ (since \hs{plus} doesn't appear in $\vect{C}$).
\item The call to $kMeans$ will produce $\vect{C} \gets [\{\hs{plus}\}]$, i.e. a
  single cluster containing \hs{plus}.
\item The next iteration will set
  $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
\item With $i = \hs{odd}$ and $e$ as its Core expression, the call to \hs{even}
  will result in $f_{recursion}$.
\item Likewise for the call to $\hs{even}$ when $i = \hs{odd}$.
\item Since the feature vectors for \hs{odd} and \hs{even} will be identical,
  $kMeans$ will put them in the same cluster. To avoid the degenerate case of a
  single cluster, for this example we will assume that $k = 2$; in which case
  the other cluster must contain \hs{plus}. Their order is arbitrary, so one
  possibility is $\vect{C} = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
\item Finally \hs{mult} will be clustered. The recursive call will become
  $f_{recursion}$ whilst the call to \hs{plus} will become $2 + 3 \alpha$, since
  $\hs{plus} \in C_2$.
\item \begin{sloppypar}Again assuming that $k = 2$, the resulting clusters will
    be
    \mbox{$\vect{C} \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus},
      \hs{mult}\}]$}.\end{sloppypar}
\end{itemize}

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$). A more extensive investigation of these features requires a
concrete implementation, in particular to pin down values for the parameters
such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.  \fi

\begin{figure}
  \scalebox{0.45}{\input{images/steppedall.pgf}}
  \scalebox{0.45}{\input{images/steppednontoxic.pgf}}
  \caption{Kaplan-Meier survival plot for running QuickSpec on inputs
    containing various numbers of definitions, sampled from TIP. The x axis
    denotes running time, which we cut short after 300 seconds. The height of
    each line shows the proportion of \qspec{} runs which were still going at
    that time (lower is better). First plot is for all TIP definitions, second
    plot removes runs given ``toxic'' definitions.}
  \label{fig:survival}
\end{figure}

Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \qspec{} running
times, when given inputs containing different numbers of definitions. Many runs
finish quickly, with the remainder occupying a ``long tail'' which we cut off
after 300 seconds\footnote{Chosen based on preliminary experiments, which showed
  little difference in survival between 300 seconds and 1 hour}.

The number of definitions in the input is linearly correlated with the amount of
timeouts, except for the very smallest inputs (which are more constrained by the
sampling procedure). One explanation for this is ``toxic'' definitions, whose
presence in the input always leads to the exploration failing (with larger
inputs being more likely to have sampled a toxic definition).

\iffalse
\begin{figure}
  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
  \caption{Proportion of samples which timed out per size, with least-squares
    linear regression. First plot is for all TIP definitions, second removes
    runs given ``toxic'' definitions.}
  \label{fig:tailsize}
\end{figure}
\fi

\begin{figure}
  \scalebox{0.45}{\input{images/proportionsall.pgf}}
  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
  \label{fig:proportions}
  \caption{Definitions, ordered by the ratio of successes to failures of the
    runs they appeared in. First graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. Second graph only contains runs
    without any toxic definitions.}
\end{figure}

\begin{figure}
  \begin{minted}{scheme}
    (define-fun-rec mult2 ((x Nat) (y Nat) (z Nat)) Nat
      (match x
        (case Z      z)
        (case (S x2) (mult2 x2 y (plus y z)))))
  \end{minted}

  \begin{minted}{scheme}
    (define-fun-rec qexp ((x Nat) (y Nat) (z Nat)) Nat
      (match y
        (case Z     z)
        (case (S n) (qexp x n (mult x z)))))
  \end{minted}

  \begin{minted}{scheme}
    (define-fun-rec op ((x Nat) (y Nat) (z Nat) (x2 Nat)) Nat
      (match x
        (case Z
          (match z
            (case Z      x2)
            (case (S x3) (op Z  y x3 (S x2)))))
        (case (S x4)
          (match z
            (case Z      (op x4 y y  x2))
            (case (S c ) (op x  y c  (S x2)))))))
  \end{minted}

  \iffalse
  \begin{minted}{scheme}
    (define-fun-rec mul3acc ((x Nat) (y Nat) (z Nat)) Nat
      (match x
        (case Z Z)                          ;; Base case for 0 * y * z
        (case (S x2)
          (match y
            (case Z Z)                      ;; Base case for x * 0 * z
            (case (S x3)
              (match z
                (case Z Z)                  ;; Base case for x * y * 0
                (case (S x4)
                  (match x2
                    (case Z
                      (match x3
                        (case Z
                          (match x4
                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
                            (case (S x5)
                              (S (add3acc (mul3acc Z Z x4)
                                          (add3acc (mul3acc (S Z) Z x4)
                                                   (mul3acc Z (S Z) x4)
                                                   (mul3acc Z Z (S Z)))
                                          (add3acc Z Z x4))))))
                        (case (S x6)
                          (S (add3acc (mul3acc Z x3 x4)
                                      (add3acc (mul3acc (S Z) x3 x4)
                                               (mul3acc Z (S Z) x4)
                                               (mul3acc Z x3 (S Z)))
                                      (add3acc Z x3 x4))))))
                    (case (S x7)
                      (S (add3acc (mul3acc x2 x3 x4)
                                  (add3acc (mul3acc (S Z) x3 x4)
                                           (mul3acc x2 (S Z) x4)
                                           (mul3acc x2 x3 (S Z)))
                                  (add3acc x2 x3 x4))))))))))))
  \end{minted}
  \fi

  \iffalse
  \begin{minted}{scheme}
    (define-fun-rec mul3 ((x Nat) (y Nat) (z Nat)) Nat
      (match x
        (case Z Z)                          ;; Base case for 0 * y * z
        (case (S x2)
          (match y
            (case Z Z)                      ;; Base case for x * 0 * z
            (case (S x3)
              (match z
                (case Z Z)                  ;; Base case for x * y * 0
                (case (S x4)
                  (match x2
                    (case Z
                      (match x3
                        (case Z
                          (match x4
                            (case Z (S Z))  ;; Base case for 1 * 1 * 1
                            (case (S x5)
                              (S (add3 (mul3 Z Z x4)
                                       (add3 (mul3 (S Z) Z x4)
                                             (mul3 Z (S Z) x4)
                                             (mul3 Z Z (S Z)))
                                       (add3 Z Z x4))))))
                        (case (S x6)
                          (S (add3 (mul3 Z x3 x4)
                                   (add3 (mul3 (S Z) x3 x4)
                                         (mul3 Z (S Z) x4)
                                         (mul3 Z x3 (S Z)))
                                   (add3 Z x3 x4))))))
                    (case (S x7)
                      (S (add3 (mul3 x2 x3 x4)
                               (add3 (mul3 (S Z) x3 x4)
                                     (mul3 x2 (S Z) x4)
                                     (mul3 x2 x3 (S Z)))
                               (add3 x2 x3 x4))))))))))))
  \end{minted}
  \fi
  \caption{``Toxic'' definitions, which consistently cause \qspec{} to fail. Two
    other definitions (\texttt{mul3} and \texttt{mul3acc}) are ommitted due to
    their verbosity.}
  \label{fig:faildefs}
\end{figure}

This does appear to be the case, as Figure~\ref{fig:proportions} shows that five
definitions appeared \emph{only} in failing inputs; these are named
\texttt{mul3}, \texttt{mul3acc}, \texttt{mult2}, \texttt{op} and \texttt{qexp};
their definitions appear in Figure~\ref{fig:faildefs}. All of these are
functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
exploration to time out by either exhausting the RAM with too many expressions
to explore, % TODO{2019-04-27} CAN QUICKSPEC2 AVOID THIS?
or by generating such deeply-nested outputs that comparing them takes too long.
% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?

The \texttt{mul3} and \texttt{mul3acc} definitions are rather pathological
implementations of multiplication with an accumulator parameter, with many
(non-tail) recursive calls. The \texttt{op} function appears in files named
\texttt{weird\_nat\_op}, which assert its commutativity and associativity.
Finally, the \texttt{mult2} and \texttt{qexp} functions are standard
tail-recursive definitions of multiplication and exponentiation, respectively.
All of these functions have an extra ``accumulator'' argument, which increases
the number of possible expressions to explore compared to those without.

Exploring each of these functions on its own does not require much memory, since
Haskell generates the output lazily. However, comparing such large numbers for
equality takes a lot of CPU time as the \texttt{S} constructors are successively
unwrapped from each side, and this is why the timeout is reached. We confirmed
this hypothesis by exploring with a custom data generator which only generates
the values \texttt{Z}, \texttt{S Z} and \texttt{S (S Z)} (0, 1 and 2); this
caused the exploration to finish quickly. Other interventions, like making the
accumulator arguments strict (to prevent space leaks), did not prevent timeouts.

To assess the impact of these problematic definitions, we removed any samples
containing them and repeated our analysis.

\section{Implementation}
\label{sec:implementation}

We provide an implementation of our recurrent clustering algorithm in a tool
called \textsc{ML4HS}. We obtain Core ASTs from Haskell packages using a plugin
for the GHC compiler, which emits a serialised version of each Core definition
as it is being compiled. This approach is more robust than, for example, parsing
source files, since it avoids the complications of preprocessors and language
extensions altering the syntax.

A post-processing stage determines which Core definitions can be used by
\qspec{}, based on their type, visibility (whether they are encapsulated inside
their module or visible to importers), etc. For each definition, we simply use
the GHCi interpreter to check if calling \qspec{} with that argument is
well-typed. This information, along with type signatures, arity, etc. are stored
alongside the Core definitions in JSON format.

The definitions are then sorted topologically, based on the non-local
identifiers appearing in their ASTs, and feature vectors are constructed using a
Haskell implementation of the approach described in \S
\ref{sec:recurrentclustering}. Since the features associated with each
identifier may vary between iterations (as the clusters change), we leave the
raw identifiers in the vector so their features can be extracted in the correct
context.

We implement Algorithm \ref{alg:recurrent} as a simple shell script, which
invokes Weka for clustering and associates each Core definition with a cluster
number. These numbers are used to finish the deferred feature extraction of
identifiers, the resulting feature vectors are clustered, and the process is
repeated until all SCCs have been processed.

Once the recurrent clustering is complete, we generate a string of Haskell code
for each of the final clusters, which will invoke \qspec{} with a suitable
signature. This involves:

\begin{itemize}
\item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall
    t. t -> t -> Bool}, we must choose a concrete representation to use (in this
  case for \hs{t}), in order to know which random generator to use. We take the
  approach used in \qcheck{} by attempting to instantiate all type variables to
  \hs{Integer}. Any values where this is invalid, such as those with
  incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where
  \hs{Integer} does not implement \hs{IsString}) will not be included in the
  signature (this is checked at the AST post-processing stage).

\item{Qualifying}: All names are \emph{qualified} (prefixed by their module's
  name), to avoid most ambiguity. There is still the possibility that multiple
  packages will declare modules of the same name, although this is rare as it
  causes problems for any Haskell programmer trying to use those modules. In
  such cases the exploration process simply aborts.

\item{Appending variables}: Once a \qspec{} theory has been defined containing
  all of the given terms, we inspect the types it references and append three
  variables for each to the theory (enough to discover laws such as
  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
  search).

\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
  that all required packages and modules must be provided up-front, usually by
  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
  \emph{dynamically}, depending on the cluster information it is given, we do
  not know what packages it may need. To work around this problem, we evaluate
  these strings of Haskell using a custom library called \texttt{nix-eval}. This
  uses the Nix package manager to obtain all of the required packages and make
  them available to GHC.

\end{itemize}

The equations resulting from evaluating these strings are collected and
outputted as JSON values, to ease further processing (e.g. displaying in some
form, sending to a theorem prover, etc.).

\section{Evaluation}
\label{sec:evaluation}
\iffalse

We have applied our recurrent clustering algorithm to several scenarios, with
mixed results. A major difficulty in evaluating these clusters is that we have
no ``ground truth'', i.e. there is no objectively correct way to compare
expressions. Instead, we provide a qualitative overview of the more interesting
characteristics.

As a simple example, we clustered (Haskell equivalents of) the running examples
used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
implementations of several functions. We expect those with similar
\emph{structure} to be clustered together, rather than those which implement the
same function. The results are shown in Figure \ref{fig:haskellcluster}, where
we can see the ``\hs{Tail}'' functions clearly distinguished, with little
distinction between the tail recursive and na\"{\i}ve implementations.

Next we tested whether these same functions would be clustered together when
mixed with seemingly-unrelated functions, in this case 207 functions from
Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail}
functions appeared together in a separate cluster from the rest. This was
unexpected, with no obvious semantic connection between these two functions and
the others in their cluster (although most are recursive, due to the nature of
the \hs{text} library).

\begin{figure}
  \begin{haskell}
(defun fact (n)
  (if (zp n) 1 (* n (fact (- n 1)))))

(defun helper-fact (n a)
  (if (zp n) a (helper-fact (- n 1) (* a n))))

(defun fact-tail (n)
  (helper-fact n 1))

(defun power (n)
  (if (zp n) 1 (* 2 (power (- n 1)))))

(defun helper-power (n a)
  (if (zp n) a (helper-power (- n 1) (+ a a))))

(defun power-tail (n)
  (helper-power n 1))

(defun fib (n)
  (if (zp n)
      0
      (if (equal n 1)
          1
          (+ (fib (- n 1)) (fib (- n 2))))))

(defun helper-fib (n j k)
  (if (zp n)
      j
      (if (equal n 1)
          k
          (helper-fib (- n 1) k (+ j k)))))

(defun fib-tail (n)
  (helper-fib n 0 1))
  \end{haskell}
  \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
  \label{fig:lisp}
\end{figure}

\begin{figure}
  \begin{haskell}
fact n = if n == 0
            then 1
            else n * fact (n - 1)

helperFact n a = if n == 0
                    then a
                    else helperFact (n - 1) (a * n)

factTail n = helperFact n 1

power n = if n == 0
             then 1
             else 2 * power (n - 1)

helperPower n a = if n == 0
                     then a
                     else helperPower (n - 1) (a + a)

powerTail n = helperPower n 1

fib n = if n == 0
           then 0
           else if n == 1
                   then 1
                   else fib (n - 1) + fib (n - 2)

helperFib n j k = if n == 0
                     then j
                     else if n == 1
                          then k
                          else helperFib (n - 1) k (j + k)

fibTail n = helperFib n 0 1
  \end{haskell}
  \caption{Haskell equivalents of the Common Lisp functions in Figure \ref{fig:lisp}.}
  \label{fig:haskelllisp}
\end{figure}

\begin{figure}
  \centering
    \tikzstyle{block} = [rectangle, draw]
    \tikzstyle{line}  = [draw, -latex']
    \begin{tikzpicture}[node distance=3cm]
      \node [block] (c1) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                    (factTail)  {\hs{factTail}};
          \node [block, below of=factTail] (fibTail)   {\hs{fibTail}};
          \node [block, below of=fibTail]  (powerTail) {\hs{powerTail}};
        \end{tikzpicture}
      };
      \node [block, right of=c1] (c2) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                (fact)       {\hs{fact}};
          \node [block, below of=fact] (helperFact) {\hs{helperFact}};
        \end{tikzpicture}
      };
      \node [block, right of=c2] (c3) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]                       (fib)         {\hs{fib}};
          \node [block, below of=fib]         (helperFib)   {\hs{helperFib}};
          \node [block, below of=helperFib]   (helperPower) {\hs{helperPower}};
          \node [block, below of=helperPower] (power)       {\hs{power}};
        \end{tikzpicture}
      };
  \end{tikzpicture}
  \caption{Typical clusters for the functions in Figure \ref{fig:haskelllisp}.}
  \label{fig:haskellcluster}
\end{figure}

We have also applied our recurrent clustering algorithm to a variety of the
most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
expected functions with a similar purpose to appear together, such as the
various reader and writer functions of \hs{pandoc}, there were always a few
exceptions which became separate for reasons which are still unclear.

When clustering the \hs{yesod} Web framework, the clustering did seem to match
our intuitions, in particular since all 15 of Yesod's MIME type identifiers
appeared in the same cluster.

Whilst recurrent clustering has produced results which merit further
investigation, the application to theory exploration has yet to be tested
empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type
class to generate random values for instantiating variables. Whilst we can
automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
all types have \hs{Arbitrary} instances; those without cannot be given any
variables in our signature, which severely limits the possible combinations
which can be explored. In many cases, no variables can be included at all,
leaving just equations involving constants. This has so far prevented us from
measuring the direct impact on \qspec{} performance, either directly by
exploring the sub-sets identified through recurrent clustering, or indirectly by
comparing the equations generated by a full brute-force search to our recurrent
clusters: those equations relating terms from different clusters would not be
discovered by our method. It is this ratio of equations found through brute
force to those found after narrowing-down by clusters which is one of our key
objectives to maximise at this stage; until we begin to pursue the
\emph{interestingness} of the properties.

The following less-serious problems were also encountered while applying
\textsc{ML4HS} to \textsc{Hackage} packages:

\begin{itemize}
\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
  declarations to cluster. This is because they make all of their declarations
  privately, e.g. in ``internal'' modules, then use separate modules to export
  the public declarations. GHC's renaming phase makes all references to such
  exports canonical, by pointing them to the private declarations. This forces
  us to ignore such declarations, as \qspec{} will not be able to access them.

\item Since we do not support type-level entities, we ignore type
  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
  ``methods'') which occur in a type class instance. Instead of being clustered,
  these result in references getting $f_{recursion}$ features. This is
  especially noticable in libraries like \hs{scientific}, where only the
  functions for constructing and destructing numbers in scientific notation are
  clustered; all of the arithmetic is defined in type classes. One difficulty
  with supporting methods is that their namespace in Core is disjoint from that
  of regular Haskell identifiers: a transformation layer would be required,
  along with explicit type annotations to avoid ambiguity.
\end{itemize}

It seems like this recurrent clustering method has promise, although it will
require a more thorough exploration of the parameters to obtain more intuitively
reliable results. These clusters can then be used in several ways to perform
theory exploration; the most na\"{\i}ve way being to explore each cluster as
\qspec{} signature in its own right. \textsc{ML4HS} already provides this
functionality, although the lack of test generators severely limits what can be
discovered.

As alluded to previously, we also have the opportunity to reason by
\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
produce a general ``scheme'' from each equation we find (either through \qspec{}
or by data mining test suites); like Isabelle approaches have shown
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
instantiated to a variety of similar values, in an attempt to find new theorems
which are analogues of existing results from a different context. ``Mutating''
existing theorem statements in such a way would also increase the chance of any
result being considered interesting; since it's likely that the unmodified
statement was deemed interesting, and the new result would not in general follow
as a simple logical consequence.

\fi

\section{Related Work}
\label{sec:related}

\subsection{Theory Exploration}

We briefly described theory exploration in \S \ref{sec:theoryexploration}, as
the task of discovering \emph{new} theorems in a software or proof library,
rather than proving/disproving user-provided statements. This general idea dates
back at least to the 1970s, most famously to work by Lenat
\cite{lenat1977automated}, though the theory exploration tools used in
functional programming trace their roots back to the interactive
\textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.

As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has
been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
\cite{johansson2009isacosy} \cite{Hipster}. Since \qspec{} is used for
conjecture generation in many of these systems, they may also benefit from our
machine learning approach.

\subsection{Relevance Filtering}
\label{sec:relevance}

The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to
reduce the search space of a theorem proving task by removing redundant or
irrelevant information (axioms, definitions, lemmas, etc.) from the input. Our
restriction of theory exploration to intelligently-selected clusters of symbols,
rather than whole libraries at a time, is essentially the same idea applied in a
novel setting.

The technique is used in Isabelle's Sledgehammer tool, during its translation of
Isabelle/HOL theories to statements in first order logic: rather than
translating the entire theory, only a sub-set of relevant clauses are
included. This reduces the size of the problem and speeds up the proof search,
but it creates the new problem of determining when a clause is relevant: how do
we know what will be required, before we have the proof?

The original approach uses the proportion of symbols which a clause has in
common with the problem statement \cite{meng2009lightweight} (plus various
heuristics), although various alternatives have been proposed in the literature
\cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.

\subsection{Recurrent Clustering}
\label{sec:clusteringexpressions}

Our recurrent clustering algorithm is similar to that of ML4PG, as our
transformation maps the elements of a syntax tree to distinct cells in a
matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree
elements: providing, for each level of the tree, the number of variables,
nullary symbols, unary symbols, etc.

Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it
also ignores types. This is because obtaining the type of each component of a
Core definition is more difficult than in ML4PG (which queries the current
\textsc{Proof General} state), and is left for future work.

The way we \emph{use} our clusters to inform theory exploration is actually more
similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
clusters are used to restrict the search space of a proof search, much like we
restrict the scope of theory exploration.

\subsection{Feature Extraction}

One major difficulty when applying statistical machine learning algorithms to
\emph{languages}, such as Haskell, is the appearance of recursive
structures. This can lead to nested expressions of arbitrary depth, which are
difficult to compare in numerical ways. One common approach to this problem is
to represent such structures as \emph{sequences}. \emph{Recurrent neural
  networks} (RNNs) are a popular choice for processing sequences, especially
when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
preserving information across long sequences \cite{hochreiter1997long}. Such
systems have been used, for example, to parse and execute computer programs
\cite{zaremba2014learning}. However, learning to parse sequences seems
inefficient considering that we already have correctly-parsed ASTs.

Whilst neural networks have been applied directly to recursive structures
\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
popular approach is to use \emph{kernel methods}
\cite{bakir2007predicting}. These are promising as a more principled alternative
to our current hand-crafted translation of ASTs to vectors.

\section{Conclusion and Future Work}
\label{sec:conclusion}

Our use of clustering to pre-process \qspec{} signatures has required many
decisions and tradeoffs to be made. Hence our approach is just one possibility
out of many alternatives which could be investigated to push this work further.

The most obvious next step is to incorporate types. Types contain valuable
information about an expression, and would allow us to distinguish between
constructors. Since our algorithm closely follows that of ML4PG, which
\emph{does} support types, the only barrier is the practical issue of
propagating annotations through all of the Core definitions.

More speculative directions include the use of \emph{learned} representations,
rather than our hand-crafted features \cite{bengio2013representation}. This
would provide an interesting comparison, as well as being more robust in the
face of language evolution. Another intriguing possibility is to extend the
recurrent nature of our algorithm to make use of the discovered properties
during clustering; for example, by treating the discovered equations as rewrite
rules to reduce the ASTs prior to feature extraction.

\appendix

\section{Core Syntax}\label{sec:core}

The GHC Core language is based on \fc{}, and described in detail in
\cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are
mostly interested in the syntax of reducible expressions (representing Haskell
of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below
in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):

\begin{equation}\label{fig:coresyntax}
  \begin{split}
    expr\    \rightarrow\ & \CVar\ id                          \\
                       |\ & \CLit\ literal                     \\
                       |\ & \CApp\ expr\ expr                  \\
                       |\ & \CLam\ \mathcal{L}\ expr           \\
                       |\ & \CLet\ bind\ expr                  \\
                       |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
                       |\ & \CType                             \\
    id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
                       |\ & \CGlobal\      \mathcal{G}         \\
                       |\ & \CConstructor\ \mathcal{D}         \\
    literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
                       |\ & \CLitStr\ \mathcal{S}              \\
    alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
    altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
                       |\ & \CLitAlt\ literal                  \\
                       |\ & \CDefault                          \\
    bind\    \rightarrow\ & \CNonRec\ binder                   \\
                       |\ & \CRec\ [binder]                    \\
    binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
  \end{split}
\end{equation}
Where:
\begin{tabular}[t]{l @{ $=$ } l}
  $\mathcal{S}$ & string literals    \\
  $\mathcal{N}$ & numeric literals   \\
  $\mathcal{L}$ & local identifiers  \\
  $\mathcal{G}$ & global identifiers \\
  $\mathcal{D}$ & constructor identifiers
\end{tabular}

The full Core language emitted by GHC (as of version 7.10.2, the latest at the
time of writing) is translated automatically to this simplified form prior to
recurrent clustering. Our major restriction is to ignore type-level entities
(such as datatype definitions, explicit casts, and differences between
types/kinds/coercions). Our implementation also handles several other forms of
literal (machine words of various sizes, individual characters, etc.), but we
omit them here for brevity as their treatment is similar to those of strings and
numerals.

\section{Rose Trees}\label{sec:rosetree}

The $toTree$ function shown below transforms Core expressions, described in
Appendix \ref{sec:core} to rose trees. We follow the presentation in
\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is
a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
$T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the
\emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where
$n_T = 0$ are \emph{leaves}.

The recursive definition is mostly routine; each repeated element (shown as
$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
have no internal structure for us to compare; they can only be compared based on
their types, which we do not currently support. We also ignore values from
$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
and we conjecture that the effect of such ``magic values'' on clustering real
code is low.

%\begin{figure}
  \begin{align*}\label{fig:totree}
    toTree(e) &=
    \begin{cases}
      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
      (\feature{\CType})                                                & \text{if $e = \CType$} \\
      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
    \end{cases}
  \end{align*}

%\end{figure}

### 2019-04-26 Bucketing/background.tex
Add more intro fluff to bucketing

@@
 has been tackled extensively in the literature~\cite{colton2000notion}.
+
+%TODO work into flow
+\emph{(Mathematical) Theory Exploration} (TE) describes an open-ended search for
+interesting properties (proven or merely conjectured, depending on the approach)
+of a given set of formal definitions, such as a mathematical theory, a
+scientific model or a software library. It has been applied to generating
+``background lemmas'' for the interactive proof assistant
+\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and automated
+theorem provers~\cite{claessen2013automating}, as well as for generating
+specifications for (pure functional) programming
+libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.
+% END TODO
 

### 2019-04-26 PhDThesis/outline.tex
Thesis bits
Change bucketing to signature selection

@@
   \item TEBenchmark (large, unambiguous empirical measurement and comparison)
-  \item MLSpec (bucketing methods, recurrent clustering, etc.)
+  \item MLSpec (signature selection methods, recurrent clustering, etc.)
   \end{itemize}
@@
 
-\item Bucketing (mostly paper and transfer report)
+\item Signature selection (mostly paper and transfer report)
   \begin{itemize}
@@
     equations, low precision, ...)
-  \item Bucketing as a solution (big-O analysis, oracles, ...)
-  \item Bucketing algorithms
+  \item Signature selection as a solution (big-O analysis, oracles, ...)
+  \item Signature selection algorithms
     \begin{itemize}
@@
 
+\item Future Work
+
 \item Conclusions
@@
   \item Unambiguous benchmark is provided for comparison
-  \item ``Smart'' bucketing method to improve efficiency
+  \item ``Smart'' signature selection method to improve efficiency
   \item Lots of room for future work (other corpora, other tools, other

### 2019-04-26 PhDThesis/background_theoryexploration.tex
Import a bunch of text and fiddle

@@
 \label{sec:theoryexploration}
+
+% TODO Work this in
+\iffalse
+Theory exploration is the task of taking a \emph{signature} of definitions in
+some formal system (for example a programming language) and automatically
+generating a set of formal statements (properties) involving those
+definitions. These may be conjectures or theorems (proven either by sending
+conjectures to an automated theorem prover, or by having the generating
+procedure proceed in logically sound steps); in either case, these statements
+must also be ``interesting'' in some way. It is beyond the scope of this paper
+to define what makes a mathematical statement ``interesting'', but this problem
+has been tackled extensively in the literature~\cite{colton2000notion}.
+\fi
+% END TOTO
 
@@
 
-We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
-Haskell code (these may be fed into another tool, such as \hspec{}, for
-proving). These conjectures are arrived at through the following stages:
+We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about Haskell
+code (these may be fed into another tool, such as \hspec{}, for proving). We
+have used version 1 of \qspec{} in our experiments, due to its availability,
+stability and tooling integration; hence references to \qspec{} are to version 1
+unless stated otherwise. At the time of writing there is a \qspec{} version 2
+available, which has a much improved generation procedure which can be
+significantly faster than its predecessor. Whilst \qspec{} 2 has advanced the
+state of the art in theory exploration, our preliminary experience has found
+that it still suffers the scaling issues we identify in this work; indeed, we
+even encountered memory exhaustion from some of the examples included with the
+\qspec{} 2 source code! Since \qspec{} 2 has less integration with the tooling
+we have used, we leave a more thorough analysis of it (and related systems such
+as \speculate{}) for future work, pending the necessary infrastructure changes
+that would require.
 
-\iffalse TODO: Make this more formal?
- V \in Var
- F \in Fun
- T \in Term
- T ::= V | F | T1 T2
+\begin{figure}
+  \centering
+  \begin{minted}{haskell}
+    -- A datatype with two constructors (Note that -- introduces a comment)
+    data Bool = True | False
 
- Term ::= VAR | Const | Fun (Term)
-or
- Term t ::= x | f | t t'
-\fi
+    -- A recursive datatype (S requires a Nat as argument)
+    data Nat = Z | S Nat
+
+    -- A function turning a Bool into a Bool
+    not :: Bool -> Bool
+    not True  = False
+    not False = True
+
+    -- A pair of mutually recursive functions
+
+    odd :: Nat -> Bool
+    odd    Z  = False
+    odd (S n) = even n
+
+    even :: Nat -> Bool
+    even    Z  = True
+    even (S n) = odd n
+  \end{minted}
+  \caption{Haskell datatypes for booleans and natural numbers, followed by some
+    simple function definitions (with type annotations).}
+  \label{fig:haskellexample}
+\end{figure}
+
+\qspec{} (version 1) is written in the Haskell programming language, and works
+with definitions which are also written in Haskell. A thorough description of
+Haskell, including its suitability for theory exploration, can be found
+in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
+are shown in Figure~\ref{fig:haskellexample}. The following procedure is used to
+generate conjectures, which are both plausible (due to testing on many examples)
+and potentially interesting (due to being mutually irreducible):
 
@@
 
-Such $conjectures$ can be used in several ways: they can be simplified for
-direct presentation to the user (by removing any equation which can be derived
-from the others by rewriting), sent to a more rigorous system like \hspec{} or
-\textsc{Hipster} for proving, or even serve as a background theory for an
+\iffalse
+% TODO: Compare the above to this; merge/swap anything that's better
+\begin{enumerate}
+\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
+  generates a list $terms$ containing the constants (including functions) from
+  $\Sigma$, the variables from $V$ and type-correct function applications
+  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
+  is finite, function applications are only nested up to a specified depth (by
+  default, 3).
+\item The elements of $terms$ are grouped into equivalence classes, based on
+  their type.
+\item The equivalence of terms in each class is tested using \qcheck{}:
+  variables are instantiated to particular values, generated randomly, and the
+  resulting closed expressions are evaluated and compared for equality.
+\item If a class is found to have non-equal members, it is split up to separate
+  those members.
+\item The previous steps of testing and splitting are repeated until the classes
+  stabilise (i.e. no differences have been observed for some specified number of
+  repetitions).
+\item For each class, one member is selected and equations are conjectured that
+  it is equal to each of the other members.
+\item A congruence closure algorithm is applied to these equations, to discard
+  any which are implied by the others.
+\end{enumerate}
+\fi
+
+Such conjectures can be used in several ways: they can be simplified for direct
+presentation to the user (for example via the congruence closure algorithm
+\iffalse\cite{TODO}\fi), sent to a more rigorous system like \hspec{} or
+\hipster{} for proving, or even serve as a background theory for an
 automated theorem prover \cite{claessen2013automating}.
@@
   \end{haskell}
-  \caption{Equations conjectured by \qspec{} for the functions in Figure \ref{fig:haskellexample}; after simplification.}
+  \caption{Equations conjectured by \qspec{} for the functions in Figure
+    \ref{fig:haskellexample}; after simplification.}
   \label{fig:qspecresult}

### 2019-04-26 PhDThesis/bucketing.tex
Import a bunch of text and fiddle

@@
-\section{Introduction}
-
-\emph{(Mathematical) Theory Exploration} (TE) describes an open-ended search for
-interesting properties (proven or merely conjectured, depending on the approach)
-of a given set of formal definitions, such as a mathematical theory, a
-scientific model or a software library. It has been applied to generating
-``background lemmas'' for the interactive proof assistant
-\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and automated
-theorem provers~\cite{claessen2013automating}, as well as for generating
-specifications for (pure functional) programming
-libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.
-
-Despite the clever search strategies employed by these tools, they all rely on
-enumerating combinations of definitions in some way, which can cause infeasible
-running times on larger inputs (in our experience, a few dozen definitions). To
-avoid this, users of these systems must carefully select only small subsets of
-their definitions to explore at a time.
-
-We dub such selection the \emph{bucketing problem} for theory exploration and
-analyse its effect on the running time of theory exploration tools on a
-particular problem set, and on the quality of the statements they are able to
-generate. An automated solution to this bucketing problem would accept a large
-number of definitions and efficiently select sub-sets which are both small
-enough to reasonably explore, whilst still enabling interesting properties to be
-discovered.
-
-Our contributions are:
+\section{The Signature Selection Problem}
+
+Despite the clever search strategies employed by theory exploration tools, they
+all rely on enumerating combinations of definitions in some way, which can cause
+infeasible running times on larger inputs (in our experience, a few dozen
+definitions). To avoid this, users of these systems must carefully select only
+small subsets of their definitions to explore at a time.
+
+We dub such selection the \emph{signature selection problem} for theory
+exploration, and we analyse its effect on the running time of theory exploration
+tools on the TIP particular problem set, and on the quality of the statements
+they are able to generate. An automated solution to this signature selection problem would
+accept a large number of definitions and efficiently select sub-sets which are
+both small enough to reasonably explore, whilst still enabling interesting
+properties to be discovered.
+
+In this chapter we will:
 
 \begin{enumerate}
-  \item Defining the \emph{bucketing problem} for theory exploration.
-  \item Statistical analysis of theory exploration performance on a large
-    problem set, derived from existing theorem proving problem sets.
-  \item Empirical analysis of the effect of bucketing on the quantity of
-    desirable properties attainable by theory exploration.
-  \item Experimental application of machine learning to the bucketing problem,
-    via unsupervised \emph{clustering}.
-  % \item A novel feature extraction method for transforming Haskell expressions
-  %   into a form amenable to off-the-shelf learning algorithms.
-  % \item An implementation of these feature extraction and theory exploration
-  %   approaches.
-  % \item A comparison of our methods with existing approaches, both for theory
-  %   exploration in Haskell, and for machine learning in other languages.
+  \item Define the \emph{signature selection problem} for theory exploration.
+  \item Analyse the performance of theory exploration on the TEBenchmark problem
+    set.
+  \item Analyse the effect of signature selection on the quantity of desirable properties
+    attainable by theory exploration on this problem set.
+  \item Demonstrate how the signature selection problem is distinct from the well-know
+    \emph{clustering problem} in machine learning, and how clustering algorithms
+    cannot be directly applied to signature selection.
 \end{enumerate}
 
+Some bespoke components required for these analyses and comparisons may also be
+useful on their own, or re-purposed for other tasks. These include:
+
+\begin{enumerate}
+\item A novel feature extraction method for transforming Haskell expressions
+  into a form amenable to off-the-shelf machine-learning algorithms, based on
+  the existing \emph{recurrent clustering} algorithm from other languages.
+\item An executable implementation of this feature extraction
+  approach\footnote{https://github.com/warbo/ml4hsfe}.
+\item End-to-end automation for exploration arbitrary, user-provided Haskell
+  packages\footnote{https://github.com/warbo/haskell-te}.
+\item Dynamic evaluation of Haskell code, with access to dynamically installed
+  Haskell packages\footnote{http://hackage.haskell.org/package/nix-eval}.
+\item Translation of the signature selection problem to the constrain-solving domain, and
+  executable oracles for optimal (and pessimal) signature selection, when the desired
+  properties are already known\footnote{Part of
+    https://github.com/warbo/bucketing-algorithms}.
+\end{enumerate}
+
+\iffalse
+% TODO: Do we need this?
 We describe theory exploration in more detail, as well as the \qspec{} tool we
 use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
-bucketing problem more formally in \S~\ref{sec:bucketing} and measure its
+signature selection problem more formally in \S~\ref{sec:sigselect} and measure its
 effect on theory exploration systems in \S~\ref{sec:implementation}. In
-\S~\ref{sec:clustering} we highlight the difference betwee bucketing and the
+\S~\ref{sec:clustering} we highlight the difference between signature selection and the
 well-known \emph{clustering problem} from machine learning, by demonstrating
-that clustering algorithms perform poorly at bucketing.  A variety of related
+that clustering algorithms perform poorly at signature selection.  A variety of related
 work is surveyed in \S~\ref{sec:related} and we conclude in
 \S~\ref{sec:conclusion} with potential directions for future research.
-
-\section{Background}
+\fi
+
+\subsection{Background}
 \label{sec:background}
-
-\subsection{Theory Exploration}
-\label{sec:theoryexploration}
-
-Theory exploration is the task of taking a \emph{signature} of definitions in
-some formal system (for example a programming language) and automatically
-generating a set of formal statements (properties) involving those
-definitions. These may be conjectures or theorems (proven either by sending
-conjectures to an automated theorem prover, or by having the generating
-procedure proceed in logically sound steps); in either case, these statements
-must also be ``interesting'' in some way. It is beyond the scope of this paper
-to define what makes a mathematical statement ``interesting'', but this problem
-has been tackled extensively in the literature~\cite{colton2000notion}.
-
-We will follow the tradition of prior practitioners and use an existing
-corpus of theorems as a \emph{ground truth} against which to compare results.
-Existing work has used libraries of the \textsc{Isabelle/HOL} proof assistant as
-a ground
-truth~\cite{Montano-Rivas.McCasland.Dixon.ea:2012,johansson2009isacosy,claessen2013automating},
-but these only contained a few definitions and theorems each. To analyse
-behaviour on a larger scale, as required to study the effects of bucketing, we
-have chosen the \emph{Tons of Inductive Problems} (TIP) problem set for
-automated theorem provers~\cite{claessen2015tip}.  Reasons for choosing TIP
-include:
-
-\begin{itemize}
-\item Each problem in TIP clearly distinguishes between the statement to be
-  proved and the definitions involved, such that the two can be easily and
-  meaningfully separated to act as inputs and desired outputs in a TE setting.
-  This would not be the case for problem sets like
-  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
-  functions whose behaviour is \emph{implicit} in the logical structure of the
-  theorem statement but not separately \emph{defined}.
-\item The TIP problem set includes a tool to translate from the language it
-  uses to define problems (a variant of SMTLIB) into various other languages
-  including Haskell and Isabelle, for which TE tools already exist.
-\item TIP problems are relevant to the domains we are interested in, including
-  pure functional programming, higher-order functions and inductively defined
-  datatypes. This makes TIP more attractive than problem sets for first-order
-  languages/logics such as TPTP~\cite{sutcliffe2009tptp}.
-\item Many of TIP's problem statements re-use the same definitions, which makes
-  it more useful as a ground-truth than a mixture of unrelated problems. This is
-  because non-membership in the ground truth will be treated as being
-  ``uninteresting''.
-\end{itemize}
-
-All existing theory exploration systems rely on brute force enumeration to some
-degree. We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
-Haskell code. These may be fed into another tool, such as \hspec{}, for
-subsequent proving. We have used version 1 of \qspec{} in our experiments, due
-to its availability, stability and tooling integration, and we will drop the
-version number from now on. At the time of writing there is a \qspec{} version 2
-available, which has a much improved generation procedure that is significantly
-faster than its predecessor. However, it still scales poorly in our experience,
-and has less integration with the tooling we have used. We hence leave analysis
-of \qspec{} 2 as future work, pending the necessary infrastructure changes that
-would require.
-
-\qspec{} (version 1) is written in the Haskell programming language, and works
-with definitions which are also written in Haskell. Some simple Haskell
-definitions are shown in Figure~\ref{fig:haskellexample}. The following
-procedure is used to generate conjectures, which are both plausible (due to
-testing on many examples) and potentially interesting (due to being mutually
-irreducible):
-
-\begin{figure}
-  \centering
-  \begin{minted}{haskell}
-    -- A datatype with two constructors
-    data Bool = True | False
-
-    -- A recursive datatype (S requires a Nat as argument)
-    data Nat = Z | S Nat
-
-    -- A function turning a Bool into a Bool
-    not :: Bool -> Bool
-    not True  = False
-    not False = True
-
-    -- A pair of mutually recursive functions
-
-    odd :: Nat -> Bool
-    odd    Z  = False
-    odd (S n) = even n
-
-    even :: Nat -> Bool
-    even    Z  = True
-    even (S n) = odd n
-  \end{minted}
-  \caption{Haskell datatypes for booleans and natural numbers, followed by some
-    simple function definitions (with type annotations).}
-  \label{fig:haskellexample}
-\end{figure}
-
-\begin{enumerate}
-\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
-  generates a list $terms$ containing the constants (including functions) from
-  $\Sigma$, the variables from $V$ and type-correct function applications
-  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
-  is finite, function applications are only nested up to a specified depth (by
-  default, 3).
-\item The elements of $terms$ are grouped into equivalence classes, based on
-  their type.
-\item The equivalence of terms in each class is tested using \qcheck{}:
-  variables are instantiated to particular values, generated randomly, and the
-  resulting closed expressions are evaluated and compared for equality.
-\item If a class is found to have non-equal members, it is split up to separate
-  those members.
-\item The previous steps of testing and splitting are repeated until the classes
-  stabilise (i.e. no differences have been observed for some specified number of
-  repetitions).
-\item For each class, one member is selected and equations are conjectured that
-  it is equal to each of the other members.
-\item A congruence closure algorithm is applied to these equations, to discard
-  any which are implied by the others.
-\end{enumerate}
 
@@
 
-\subsection{The Bucketing Problem}
-\label{sec:bucketing}
+\subsection{The Signature Selection Problem}
+\label{sec:sigselect}
 
@@
 
-Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\phi$).
-These feature vectors might be deemed very dissimilar during clustering, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
-
-If we were to align these features optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
-
-The method we use to ``flatten'' expressions, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size $c$ (for \emph{columns}).
-This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
-Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the constructor identifier $\id{C}$ would not appear in the vector, but this example is still accurate in terms of laying out the features as given.}
+Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$
+are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match
+(others may match by coincidence, depending on $\phi$). These feature vectors
+might be deemed very dissimilar during clustering, despite the intuitive
+similarity of the expressions $X$ and $Y$ from which they derive.
+
+If we were to align these features optimally, by padding the fourth column
+rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to
+match, making the similarity of the vectors more closely match our intuition and
+depend less on coincidence.
+
+The method we use to ``flatten'' expressions, described below, is a variation of
+breadth-first traversal which pads each level of nesting to a fixed size $c$
+(for \emph{columns}). This doesn't guarantee alignment, but it does prevent
+mis-alignment from accumulating across different levels of nesting. Our method
+would align these features into the following vectors, if $c = 2$: \footnote{In
+  fact, the constructor identifier $\id{C}$ would not appear in the vector, but
+  this example is still accurate in terms of laying out the features as given.}
 
@@
 
-Here $\frac{1}{2}$ of the original 6 features align, which is more than $breadthFirst$ but not optimal.
-Both vectors have also been padded by an extra 2 zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
-
-To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features.
-Intuitively, these are simply an s-expression representation of the Core syntax, with the feature extraction function $\phi$ mapped over the labels.
-The full transformation is given in Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the result for the \hs{odd} function.
+Here $\frac{1}{2}$ of the original 6 features align, which is more than
+$breadthFirst$ but not optimal. Both vectors have also been padded by an extra 2
+zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
+
+To perform this flattening we first transform the nested tokens of an expression
+into a \emph{rose tree} of features. Intuitively, these are simply an
+s-expression representation of the Core syntax, with the feature extraction
+function $\phi$ mapped over the labels. The full transformation is given in
+Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the
+result for the \hs{odd} function.
 
@@
   \end{scriptsize}
-  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the expression \hs{odd} from Figure \ref{fig:coreexample}.
-Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
+  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the
+    expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree
+    is rendered with its feature at the node and sub-trees beneath.}
 \end{figure}
@@
 
-These rose trees are then turned into matrices, as shown in Figure \ref{fig:matrixexample}.
-Each row $i$ of the matrix contains the features at depth $i$ in the rose tree, read left-to-right, followed by any required padding.
-These matrices are then either truncated, or padded with rows (on the bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.
+These rose trees are then turned into matrices, as shown in Figure
+\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at
+depth $i$ in the rose tree, read left-to-right, followed by any required
+padding. These matrices are then either truncated, or padded with rows (on the
+bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.
 
@@
 \iffalse
-As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}.
-A valid topological ordering is given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding Core expressions to save space):
+
+As an example of this recurrent process, we can consider the Peano arithmetic
+functions from Figure \ref{fig:coreexample}. A valid topological ordering is
+given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
+Core expressions to save space):
 
@@
 concrete implementation, in particular to pin down values for the parameters
-such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.  \fi
+such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
+
+\fi
+
+\subsection{Experiments}
+
+% TODO: This was taken from the Background section, so probably doesn't fit here
+% without some fiddling.
+% TODO: Describe all of our experimental setups:
+%
+%  - The use of the TEBenchmark methodology for running QuickSpec, including the
+%    sampling process
+%
+%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
+%    including how we picked the sizes
+%
+%  - The definition of HashSpec and its use as control; how we ran all of the
+%    same samples through it
+%
+%  - The use of optimal and pessimal oracles, defined via constraint
+%    satisfaction. How we it was only feasible to run these up to size 10 (11?)
+
+We follow the tradition of prior practitioners and use an existing corpus of
+theorems as a \emph{ground truth} against which to compare results. To analyse
+behaviour on large inputs (i.e. those for which exploration is currently
+infeasible), as required to study the effects of signature selection, we have used the
+\tebenchmark{} of section~\ref{sec:tebenchmark}, which is based on the Tons of
+Inductive Problems (TIP) problem set for theorem provers~\cite{claessen2015tip}.
 

### 2019-04-26 PhDThesis/intro.tex
Import a bunch of text and fiddle

@@
 
+\iffalse
 We begin our undertaking in \S \ref{sec:background} by introducing the Haskell
@@
 
+% TODO: Do we need this?
 We discuss our present contributions in \S \ref{sec:current} and future research
 directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.
+\fi

### 2019-04-26 PhDThesis/litreview_staticanalysis.tex
Import a bunch of text and fiddle

@@
 empirically by performing multiple runs, instantiating metavariables with
-synthetic data (\emph{QuickCheck} does this randomly, \emph{SmallCheck}
-enumerates values, etc.).
+synthetic data (\qcheck{} does this randomly, \textsc{SmallCheck} enumerates
+values, etc.).
 

### 2019-04-26 PhDThesis/litreview_testing.tex
Import a bunch of text and fiddle

@@
 
-Automated testing is a common method of software quality assurance. In Haskell,
-some of the most popular testing frameworks go beyond the industry standard of
-\emph{unit testing} (asserting that components of a system behave as expected on
-a fixed set of inputs) to instead perform \emph{property checking}. A
-``property'', in this sense, is a statement about a program's components, which
-may include free variables. Such properties are typically encoded in Haskell as
-functions, accepting values for the free variables as arguments, and returning a
-boolean value (perhaps adorned with useful metadata). A property checker
-proceeds by
-\begin{enumerate*}
-\item generating suitable values for the function's arguments,
-\item calling the function with these values,
-\item inspecting the return value and
-\item repeating until some stop condition is satisfied
-\end{enumerate*}.
+Automated testing is a common method of software quality assurance. Typical
+industry standards (outside of specialised domains like those which are
+life-critical) include \emph{unit testing} \iffalse \cite{TODO} \fi: asserting
+that components of a system behave as expected on a fixed set of inputs. In
+Haskell, some of the most popular testing frameworks go beyond this to instead
+perform \emph{property checking}. A ``property'', in this sense, is a statement
+about a program's components, which may include free variables. We can think of
+a unit test as proving an \emph{existential} property: that the given statement
+holds for at least one assignment of the variables, and those values are stated
+as part of the unit test. However, the phrase ``property checking'' is typically
+reserved for \emph{universal} properties: those which should hold \emph{for all}
+values of the variables.
+
+Universal properties are typically encoded in Haskell as \emph{functions},
+accepting values for the free variables as arguments, and returning a boolean
+value (perhaps adorned with useful metadata). If the type of the variables is
+finite (for example booleans), it might be feasible to check all possible
+values. Typically this is not feasible, especially when multiple variables are
+involved, or the input types contain an infinite number of values (for example
+lists). Property checkers avoid this by only testing a sub-set of values, in the
+hope that a counterexample will be found (if one exists). They work by:
+
+\begin{enumerate}
+\item Generating suitable values for the function's arguments
+\item Calling the function with these values
+\item Inspecting the return value
+\item Repeating until some stop condition is satisfied
+\end{enumerate}.
 
@@
 stop early if the function returns a false value. The major difference between
-implementations is the way they instantiate free variables (or, in encoded form,
-how they generate arguments for the property functions).
+implementations is the way they generate values to test with.
 
@@
 choosing a default generator for types implementing its \texttt{Arbitrary}
-typeclass (the recent \texttt{hedgehog} package works in a similar way).
+typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
+\cite{TODO} \fi).
 
@@
 which recurse on each \emph{sub-}expressions of their input as well. The output
-of a random generator will also vary across runs, \emph{increasing} our
-confidence each time we check a property, rather than simply reiterating past
-results. However, there are two common problems with random property checking.
+of a random generator will also vary across runs, rather than checking the same
+values every time. Hence every invocation of a property checker, e.g. as part of
+a development or debugging workflow, allows more of the input space to be
+tested. However, there are two common problems with random property checking.
+
 Firstly we must be able to generate random values, which can be tricky for
@@
 tests pass, such that failures are likely to be triggered by small, more easily
-understood counterexamples. Additionally, a mechanism for ``shrinking'' is
+understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
 provided: users can supply a \texttt{shrink} function along with their
@@
 returning each sub-tree, and return an empty list when asked to shrink a
-\texttt{Leaf}. After a counterexample has been found, the same property is
-applied to each of the shrunk values in an attempt to narrow down the source of
-the problem. When a value cannot be shrunk any more, or if none of its shrunken
-versions are counterexamples, that value is presented to the user in the hope
-that it is simple enough to understand.
+\texttt{Leaf}. If one of these smaller values is also a counterexample, it too
+is shrunk and the process repeats, otherwise it is discarded and the next shrunk
+value is tried. When a value cannot be shrunk any more, or if none of its
+shrunken versions are counterexamples, that value is presented to the user in
+the hope that it is simple enough to understand.
 
 The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
-values (either using a user-provided enumerating function or using its
+values (either using a user-provided enumeration function or using its
 \texttt{Serial} typeclass), rather than generating values
 randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
-the \texttt{testing-feat} package). In the \texttt{Tree} example this might
-check \texttt{Leaf}, then \texttt{Node Leaf Leaf}, then
-\texttt{Node (Node Leaf Leaf) Leaf}, and so on. This ensures that
-counterexamples are always small, that tests are not repeated with the same
-values and that memory is not exhausted (since the stopping condition will
-trigger long before the enumeration reaches such large values). The disadvantage
-is that large inputs (which exercise properties, especially recursive ones, the
-most) may occur so late in the enumeration that testing never reaches them.
+the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
+\texttt{Tree} example this might check \texttt{Leaf}, then \texttt{Node Leaf
+  Leaf}, then \texttt{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
+order of ``size'' ensures that counterexamples are always small, that tests are
+not repeated with the same values and that memory is not exhausted (since the
+stopping condition will trigger long before the enumeration reaches such large
+values). The disadvantage is that large inputs (which exercise properties,
+especially recursive ones, the most) may occur so late in the enumeration that
+testing never reaches them.
 

### 2019-04-29 PhDThesis/bucketing.tex
Include bucketing tex file in thesis

@@
   \item Define the \emph{signature selection problem} for theory exploration.
-  \item Analyse the performance of theory exploration on the TEBenchmark problem
-    set.
+  \item Analyse the performance of theory exploration on the Theory Exploration
+    Benchmark.
   \item Analyse the effect of signature selection on the quantity of desirable properties
@@
 
-As an example, giving the definitions from Figure~\ref{fig:haskellexample} to
+As an example, giving the definitions from Figure~\ref{fig:haskellteexample} to
 \qspec{}, along with suitable comparison functions, random data generators and
@@
 behaviour on large inputs (i.e. those for which exploration is currently
-infeasible), as required to study the effects of signature selection, we have used the
-\tebenchmark{} of section~\ref{sec:tebenchmark}, which is based on the Tons of
-Inductive Problems (TIP) problem set for theorem provers~\cite{claessen2015tip}.
+infeasible), as required to study the effects of signature selection, we have
+used the Theory Exploration Benchmark of section~\ref{sec:tebenchmark}, which is
+based on the Tons of Inductive Problems (TIP) problem set for theorem
+provers~\cite{claessen2015tip}.
 

### 2019-04-29 PhDThesis/background_haskell.tex
Bunch more changes to thesis

@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
+  \begin{haskell}
     module Email (Email(), at, render) where
@@
     at u  h  = Just (E u h)
-\end{lstlisting}
+  \end{haskell}
 
-The \hs{Email} type guarantees that its elements have both a user part and a
-host part (modulo divergence), but it does not provide any guarantees about
-those parts. We also define the \hs{at} function, a so-called ``smart
-constructor'', which has the additional guarantee that the \hs{Email}s it
-returns contain non-empty \hs{String}s. By ommitting the \hs{E} constructor from
-the export list on the first line \footnote{The syntax \hs{Email()} means we're
-  exporting the \hs{Email} type, but not any of its constructors.}, the only way
-\emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces
-the non-empty guarantee to hold globally.
+  This \hs{Email} type can be constructed in two ways: directly, via the \hs{E}
+  constructor, which requires both a user part and a host part (both
+  \hs{String}s, for simplicity); or indirectly, via the \hs{at} function, which
+  will only call \hs{E} if both \hs{String}s are non-empty (a programming
+  pattern known as a ``smart constructor''). The export list in the first line
+  contains the \hs{at} function, but not the \hs{E} constructor~\footnote{The
+    syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any
+    of its constructors.}, so any code using this module is only able to create
+  \hs{Email} values with non-empty parts.
 
@@
 
-\item The existence of theory exploration systems such as \hspec{}, and related
-  tools which we can to re-use, including conjecture generators like \qspec{}
-  and \textsc{Speculate}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
-  \textsc{SmartCheck}, \textsc{LeanCheck}, \textsc{Hedgehog}, etc.; theorem
-  provers like \textsc{Hip} \cite{rosen2012proving}; and other related testing
-  and term-generating systems like \textsc{MuCheck} \cite{le2014mucheck},
+\item The existence of theory exploration systems such as \hspec{}, \qspec{} and
+  \textsc{Speculate}.
+
+\item Related tooling we can re-use such as counterexample finders (\qcheck{},
+  \textsc{SmallCheck}, \textsc{SmartCheck}, \textsc{LeanCheck},
+  \textsc{Hedgehog}, etc.), theorem provers
+  (e.g. \textsc{Hip}~\cite{rosen2012proving}), and other testing and
+  term-generating systems like \textsc{MuCheck}~\cite{le2014mucheck},
   \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
@@
 
-Whilst we will use normal Haskell code in this text and many of our examples,
-for simplicity we perform many of our analyses on an intermediate representation
-of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the
-relatively large and complex syntax of Haskell proper. Core is based on \fc{},
-which is described in detail in~\cite[Appendix C]{sulzmann2007system}.
+Whilst our systems and experiments use normal Haskell code, for simplicity we
+perform some of our analyses on an intermediate representation of the
+\textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively
+large and complex syntax of Haskell proper. Core is based on \fc{}, which is
+described in detail in~\cite[Appendix C]{sulzmann2007system}.
 
-Core contains explicit type annotations and coercions, which we can omit as they
+Core contains explicit type annotations and coercions, which we omit as they
 have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
-  of GHC version 7.10.2, the latest at the time of writing.} is shown in
-Figure~\ref{fig:coresyntax}; for brevity, we also omit several other forms of
-literal (machine words of various sizes, individual characters, etc.), since
-their treatment is similar to those of strings and numerals. We use quoted
-strings to denote names and literals, e.g. \texttt{Local "foo"},
-\texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and
-\texttt{LitNum "42"}, and require only that they can be compared for equality.
+  of GHC version 7.10.2.} is shown in Figure~\ref{fig:coresyntax}; for brevity,
+we also omit several other forms of literal (machine words of various sizes,
+individual characters, etc.), since their treatment is similar to those of
+strings and numerals. We use quoted strings to denote names and literals,
+e.g. \hs{Local "foo"}, \hs{Global "bar"}, \hs{Constructor "Baz"}, \hs{LitStr
+  "quux"} and \hs{LitNum "42"}, and require only that they can be compared for
+equality.
 
@@
     \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
-    regarded as \hs{data Bool \equal{} True | False}.}
+    regarded as \hs{data Bool = True | False}.}
   \label{fig:haskellexample}

### 2019-04-29 PhDThesis/background_quickcheck.tex
Bunch more changes to thesis

@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+\begin{haskell}
   factorial 0 = 1
@@
   fact_div       = factorial 4 == factorial 5 `div` 5
-\end{lstlisting}
+\end{haskell}
 
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+\begin{haskell}
 factorial 0 = 1
@@
 fact_neg = factorial 1 == factorial (-1)
-\end{lstlisting}
+\end{haskell}
 

### 2019-04-29 PhDThesis/background_theoryexploration.tex
Bunch more changes to thesis

@@
     simple function definitions (with type annotations).}
-  \label{fig:haskellexample}
+  \label{fig:haskellteexample}
 \end{figure}
@@
 in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
-are shown in Figure~\ref{fig:haskellexample}. The following procedure is used to
+are shown in Figure~\ref{fig:haskellteexample}. The following procedure is used to
 generate conjectures, which are both plausible (due to testing on many examples)
@@
 As an example, we can consider a simple signature containing the expressions
-from Figure \ref{fig:haskellexample}:
+from Figure \ref{fig:haskellteexample}:
 
@@
   \caption{Equations conjectured by \qspec{} for the functions in Figure
-    \ref{fig:haskellexample}; after simplification.}
+    \ref{fig:haskellteexample}; after simplification.}
   \label{fig:qspecresult}

### 2019-04-29 PhDThesis/litreview_premiseselection.tex
Bunch more changes to thesis

@@
 
-Recently, a variety of alternative algorithms have also been investigated,
-dubbed \textsc{MaSH}, \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
+Recently, a variety of alternative algorithms have also been
+investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr},
+%                  MaSH             SInE,         BliStr
+dubbed , \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
 \textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
@@
 small number of terms (depending on the speed of the subsequent algorithm and
-how long the user is willing to wait) from a large set, to optimise the
-outcome of that algorithm (e.g. finding a proof quickly; or finding many
-interesting relationships quickly). The major difference is that premise
-selection has a particular ``target'' theorem statement, against which to
-compare each term (e.g. for the ``relevance'' of their sub-terms). This is not
-the case for theory exploration, so we must consider the interactions of each
-term with all of the others when deciding what to select.
+how long the user is willing to wait) from a large set, to optimise the outcome
+of that algorithm (e.g. finding a proof quickly; or finding many interesting
+relationships quickly). The major difference is that premise selection has a
+particular ``target'' theorem statement, against which to compare each
+term. This is not the case for theory exploration, so we must consider the
+interactions of each term with all of the others when deciding what to select.

### 2019-04-29 PhDThesis/support-for-benchmarks/mkBenchmarkGraphs.py

@@ -265,7 +265,7 @@ mpl.rcParams.update({
 
     # LaTeX default is 10pt font.
     'axes.labelsize'  : 10,
-    'text.fontsize'   : 10,
+    'font.size'       : 10,
     'legend.fontsize' : 8,
     'xtick.labelsize' : 8,
     'ytick.labelsize' : 8,

### 2019-04-29 PhDThesis/support-for-bucketing/contentsGraph.py
Bucketing stuff for thesis

@@ -1,0 +1,69 @@
+#!/usr/bin/env python3
+from os  import getenv
+from csv import DictReader
+
+label = getenv('name')
+
+def msg(x):
+    from sys import stderr
+    stderr.write(repr(x))
+    stderr.flush()
+
+def save(name, axes):
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
+
+with open(getenv('csv'), 'r') as f:
+    counts = DictReader(f)
+    rows   = [{**row,
+               **{'successes': int(row['successes']),
+                  'failures' : int(row['failures' ])}} for row in counts]
+
+del(getenv    )
+del(DictReader)
+
+rows.sort(key=lambda x: x['successes'] / (x['successes'] + x['failures']),
+          reverse=True)
+
+names     = [row['name'     ] for row in rows]
+successes = [row['successes'] for row in rows]
+failures  = [row['failures' ] for row in rows]
+
+prop = lambda key: lambda row: row[key] / (row['successes'] + row['failures'])
+
+successProp = list(map(prop('successes'), rows))
+failureProp = list(map(prop('failures' ), rows))
+
+import matplotlib.pyplot as plt
+
+barWidth = 0.85
+r        = [x for x in range(0, len(names))]
+
+# Create green Bars
+plt.bar(r, successProp,                     color='#00ff00', edgecolor='white', width=barWidth)
+# Create orange Bars
+plt.bar(r, failureProp, bottom=successProp, color='#ff0000', edgecolor='white', width=barWidth)
+
+# Custom x axis
+plt.xticks(r, names)
+plt.xlabel("Name")
+
+#ax = plt.subplot(111)
+#ax.bar(names, successes, width=0.3, color='g', align='center')
+#ax.bar(names, failures , width=0.3, color='r', align='center')
+#ax.xaxis_date()
+
+#plt.bar([ row['name'     ]                   for row in rows],
+#        [[row['successes'], row['failures']] for row in rows])
+
+#plt.xlim((1,20))
+#plt.ylim((0,1 ))
+#plt.xticks(xs)
+plt.xlabel('Name')
+plt.ylabel('Occurrences')
+plt.title('Name distribution between successful and failed runs')
+
+#ax.set_xlabel('Name')
+#ax.set_ylabel('Occurrences')
+#ax.set_title('Name distribution between successful and failed runs')
+save('proportions', plt.gca())

### 2019-04-29 PhDThesis/support-for-bucketing/survivalGraph.py

@@ -1,0 +1,83 @@
+#!/usr/bin/env python3
+
+from os     import getenv
+from pandas import read_csv
+with open(getenv('csv'), 'r') as f:
+    times = read_csv(f, header=0, index_col=0)
+
+label = getenv('label')
+
+del(getenv)
+del(read_csv)
+
+from lifelines import KaplanMeierFitter
+import matplotlib.pyplot as plt
+
+def save(name, axes):
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
+
+axes = plt.subplot(111, label="stepped")
+kmf  = KaplanMeierFitter()
+for s, grouped_df in times.groupby('size'):
+    if s not in [1, 5, 10, 15, 20]:
+        continue
+    kmf.fit(grouped_df['time'], grouped_df['success'], label=str(s))
+    axes = kmf.plot(ax=axes)
+save('stepped', axes)
+
+def crossValidate(name, fitter):
+    from lifelines.utils import k_fold_cross_validation
+    import numpy as np
+    print("Cross Validating " + name)
+    print(np.mean(k_fold_cross_validation(fitter,
+                                          times,
+                                          duration_col='time',
+                                          event_col='success')))
+    print("End cross-validation of " + name)
+
+from lifelines import CoxPHFitter
+cph = CoxPHFitter()
+cph.fit(times, duration_col='time', event_col='success', show_progress=True)
+
+cph.print_summary()
+
+save('coxph', cph.plot())
+#cph.check_assumptions(times, show_plots=True)
+
+crossValidate('cox', cph)
+
+fitters = {'cox':cph}
+
+from lifelines import WeibullAFTFitter, LogNormalAFTFitter, LogLogisticAFTFitter
+for (name, fitter) in [("weibull"    , WeibullAFTFitter    ),
+                       ("lognormal"  , LogNormalAFTFitter  ),
+                       ("loglogistic", LogLogisticAFTFitter)]:
+    print("BEGIN " + name)
+    aft = fitter()
+    aft.fit(times, duration_col='time', event_col='success')
+    aft.print_summary(3)
+
+    #aft = WeibullAFTFitter().fit(times, 'time', 'success', ancillary_df=True)
+    save(name + 'aft', aft.plot())
+    fitters[name] = aft
+    crossValidate(name, aft)
+    print("END " + name)
+
[…]

### 2019-04-29 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -1,0 +1,64 @@
+#!/usr/bin/env python3
+from os  import getenv
+from csv import DictReader
+
+label = getenv('label')
+
+def msg(x):
+    from sys import stderr
+    stderr.write(repr(x))
+    stderr.flush()
+
+def save(name, axes):
+    axes.get_figure().savefig(name + label + '.pdf')
+    axes.get_figure().savefig(name + label + '.pgf')
+
+with open(getenv('csv'), 'r') as f:
+    times = DictReader(f)
+    rows  = [row for row in times]
+
+del(getenv  )
+del(DictReader)
+
+counts = {}
+fails  = {}
+for row in rows:
+    s = row['size']
+    if s not in counts:
+        counts[s] = 0
+    if s not in fails:
+        fails[s] = 0
+    counts[s] += 1
+    if row['success'] == '0':
+        fails[s] += 1
+
+keys  = sorted(counts.keys(), key=int)
+fracs = [fails[s] / counts[s] for s in keys]
+
+msg({
+    'len(fracs)': len(fracs),
+    'len(range(1,21))': len(range(1,21))
+})
+
+from scipy import stats
+slope, intercept, r_value, p_value, std_err = stats.linregress(range(1,21),
+                                                               fracs)
+
+import matplotlib.pyplot as plt
+
+plt.plot(fracs, '.')
+
+f  = lambda x: slope*x + intercept
+xs = [  x  for x in range(1,21)]
+ys = [f(x) for x in xs         ]
+
+ax = plt.plot(xs, ys)
+
+plt.xlim((1,20))
+plt.ylim((0,1 ))
+plt.xticks(xs)
+plt.xlabel('size')
+plt.ylabel('failure proportion')
+plt.title('Failures as size grows')
+save('timeouts', plt.gca())
+#plt.savefig('timeouts.pdf')

### 2019-04-29 PhDThesis/support-for-bucketing/toxic/Defs.hs

@@ -1,0 +1,78 @@
+{-# LANGUAGE BangPatterns #-}
+module Defs where
+
+import Test.QuickCheck
+import Test.QuickSpec
+
+-- Peano numerals
+
+data Nat = Z | S Nat deriving (Eq)
+
+i2n :: Integer -> Nat
+i2n 0 = Z
+i2n n = S (i2n (abs n - 1))
+
+n2i :: Nat -> Integer
+n2i  Z    = 0
+n2i (S n) = 1 + n2i n
+
+instance Show Nat where
+  show = show . n2i
+
+instance Ord Nat where
+  Z   <= _   = True
+  S x <= Z   = False
+  S x <= S y = x <= y
+
+instance Arbitrary Nat where
+  arbitrary = i2n <$> arbitrary
+
+smallGen = i2n <$> choose (0, 2)
+
+-- Helper functions
+
+plus :: Nat -> Nat -> Nat
+plus x y = case x of
+             Z   -> y
+             S z -> S (plus z y)
+
+mult x y = case x of
+             Z   -> Z
+             S z -> plus y (mult z y)
+
+-- Functions to explore
+
+mult2_lazy, mult2_strict :: Nat -> Nat -> Nat -> Nat
+mult2_lazy x y z = case x of
+                     Z    -> z
+                     S x2 -> mult2_lazy x2 y (plus y z)
+
+mult2_strict x y !z = case x of
+                        Z    -> z
+                        S x2 -> mult2_strict x2 y (plus y z)
+
+qexp_lazy, qexp_strict :: Nat -> Nat -> Nat -> Nat
+qexp_lazy x y z = case y of
+                    Z   -> z
+                    S n -> qexp_lazy x n (mult x z)
+
+qexp_strict x y !z = case y of
+                       Z   -> z
+                       S n -> qexp_strict x n (mult x z)
+
+op_lazy, op_strict :: Nat -> Nat -> Nat -> Nat -> Nat
+op_lazy x y z x2 = case x of
+                     Z    -> case z of
+                               Z    -> x2
+                               S x3 -> op_lazy Z  y x3 (S x2)
+                     S x4 -> case z of
+                               Z    -> op_lazy x4 y y     x2
+                               S c  -> op_lazy x  y c  (S x2)
+
+op_strict x y z x2 = case x of
+                       Z    -> case z of
+                                 Z    -> x2
+                                 S x3 -> op_strict Z  y x3 (S x2)
+                       S x4 -> case z of
[…]

### 2019-04-29 PhDThesis/support-for-bucketing/toxic/QS1Runners.hs

@@ -1,0 +1,39 @@
+module Runners where
+
+import Defs
+import Test.QuickCheck
+import Test.QuickSpec
+
+vs = map (gvars ["nat1", "nat2", "nat3"]) [g, g, g]
+  where g = arbitrary :: Gen Nat
+
+[qs_mult2_lazy  ,
+ qs_mult2_strict,
+ qs_qexp_lazy   ,
+ qs_qexp_strict ,
+ qs_op_lazy     ,
+ qs_op_strict   ] =
+  map (quickSpec . (:vs)) [ fun3 "mult2_lazy"   mult2_lazy
+                          , fun3 "mult2_strict" mult2_strict
+                          , fun3 "qexp_lazy"    qexp_lazy
+                          , fun3 "qexp_strict"  qexp_strict
+                          , fun4 "op_lazy"      op_lazy
+                          , fun4 "op_strict"    op_strict
+                          ]
+
+vsmall = map (gvars ["small1", "small2", "small3"]) [g, g, g]
+  where g = smallGen
+
+[small_mult2_lazy  ,
+ small_mult2_strict,
+ small_qexp_lazy   ,
+ small_qexp_strict ,
+ small_op_lazy     ,
+ small_op_strict   ] =
+  map (quickSpec . (:vsmall)) [ fun3 "mult2_lazy"   mult2_lazy
+                              , fun3 "mult2_strict" mult2_strict
+                              , fun3 "qexp_lazy"    qexp_lazy
+                              , fun3 "qexp_strict"  qexp_strict
+                              , fun4 "op_lazy"      op_lazy
+                              , fun4 "op_strict"    op_strict
+                              ]

### 2019-04-29 PhDThesis/litreview_premiseselection.tex
Clean up

@@
 
-One fruitful application of machine learning methods to formal mathematics is
-\emph{relevance filtering}, which is used for \emph{premise selection} in
-automated theorem provers~\cite{kuhlwein2012overview}.
+\emph{Relevance filtering} takes a specified ``target'' (such as the current
+goal of a theorem prover), a large list of inputs (e.g. existing lemmas from a
+theory) and produces a sub-set of those inputs which are predicted to be useful
+or relevant to the target. The example of choosing a sub-set of lemmas to help
+prove a theorem is known as \emph{premise
+  selection}~\cite{kuhlwein2012overview}.
 
-Relevance filtering simplifies a proof search problem by removing from
-consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed
-\emph{irrelevant}. The technique is used in Sledgehammer during its translation
-of Isabelle/HOL theories to statements in first order logic: rather than
-translating the entire theory, only a sub-set of relevant clauses are
-included. This reduces the size of the problem and speeds up the proof search,
-but it creates the new problem of determining when a clause is relevant: how do
-we know what will be required, before we have the proof?
+Relevance filtering simplifies such proof search by focusing on only those
+clauses (axioms, definitions, lemmas, etc.) which are deemed relevant, shrinking
+the search space. The technique is used in \textsc{Sledgehammer} during its
+translation of Isabelle/HOL theories to statements in first order logic: rather
+than translating the entire theory, only a sub-set of relevant clauses are
+included. This speeds up the proof search, but it creates the new problem of
+determining when a clause is relevant: how do we know what will be required,
+before we have the proof?
 
-Sledgehammer's initial approach, known as \textsc{MePO} (from
+The initial approach used in \textsc{Sledgehammer}, known as \textsc{MePO} (from
 \emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based
@@
 symbols are then also considered relevant. There are other heuristics applied
-too, such as increasing the score of user-provided facts (e.g. given by keywords
+too, such as increasing the score of user-provided facts (given by keywords
 like \texttt{using}), locally-scoped facts, first-order facts and
@@
 Recently, a variety of alternative algorithms have also been
-investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr},
-%                  MaSH             SInE,         BliStr
-dubbed , \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
-\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
-\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
-algorithm with a custom feature extraction approach, is trained using a corpus
-of existing theorems and their proofs, and evaluated by trying to re-prove a
-selection of unseen theorems from the corpus (i.e. given only the theorem
-statement and possible lemmas, not the existing proof).
+investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr,gauthier2015premise,alama2014premise,carlson1999snow,kaliszyk2014machinelearner,kaliszyk2015femalecop}.
+%                  MaSH             SInE,         BliStr          HOLyHammer          MOR,             SNoW,           MaLARea
+Each combines an ``off the shelf'' machine learning algorithm (e.g. a
+naive-Bayes classifier, support vector machines, etc.) with a custom feature
+extraction approach; is trained using a corpus of existing theorems and their
+proofs; and evaluated by trying to \emph{re-prove} a selection of unseen
+theorems from the corpus (i.e. given only the theorem statement and possible
+lemmas, not the existing proof).
 
-For example the Machine Learning for SledgeHammer (\textsc{MaSH}) system
-defines a notion of ``visibility''~\cite{kuhlwein2013mash}. This is essentially
-a dependency graph of which theorems were used in the proofs of which other
-theorems (theorems are represented as abstract sets of features, rather than
-e.g. sentences or syntax trees). To select relevant clauses for a goal, the
-set of clauses which are ``visible'' from the goal's components is generated;
-this is further reduced by (an efficient approximation of) a naive Bayes
-algorithm.
+For example the ``Machine Learning for SledgeHammer'' (\textsc{MaSH}) system
+defines a notion of ``visibility''~\cite{kuhlwein2013mash} for determining how
+relevant one statement is to another. Visibility is essentially a dependency
+graph of which theorems were used in the proofs of which other theorems
+(represented as abstract sets of features, rather than concrete sentences or
+syntax trees). To select relevant clauses for a goal, the set of clauses which
+are ``visible'' from the goal's components is generated; and this is filtered
+further by (an efficient approximation of) a naive-Bayes algorithm.
 
@@
 support vector machine (SVM) to select relevant axioms from the Mizar
-Mathematical Library for use by the Vampire ATP system~\cite{alama2014premise}.
-%TODO: describe the kernel, as that's the interesting bit.
+Mathematical Library (MML) for use by the Vampire ATP
+system~\cite{alama2014premise}. SVM requires a ``kernel function'', effectively
+a similarity measure between formulas. The \textsc{MOR} tool measures similarity
+based on sub-terms which two formulas have in common: first a list of all $m$
+terms appearing in the input space is extracted, and each formula is converted
+to an $m$-dimensional binary vector, with the $n$th element of the vector
+indicating whether the $n$th term of the input space occurs in this formula. The
+kernel used by \textsc{MOR} is a Gaussian function of the dot-products of
+these binary vectors (parameterised by a hyperparameter $\sigma$, chosen via
+grid search).
 
@@
 particular ``target'' theorem statement, against which to compare each
-term. This is not the case for theory exploration, so we must consider the
+term. Indeed, machine learning approaches like \textsc{MOR} learn a separate
+ranking function for each target formula.
+
+This is not the case for theory exploration, which must consider the
 interactions of each term with all of the others when deciding what to select.

### 2019-04-29 PhDThesis/litreview_testing.tex
Clean up

@@
 
-Automated testing is a common method of software quality assurance. Typical
-industry standards (outside of specialised domains like those which are
-life-critical) include \emph{unit testing} \iffalse \cite{TODO} \fi: asserting
-that components of a system behave as expected on a fixed set of inputs. In
-Haskell, some of the most popular testing frameworks go beyond this to instead
-perform \emph{property checking}. A ``property'', in this sense, is a statement
-about a program's components, which may include free variables. We can think of
-a unit test as proving an \emph{existential} property: that the given statement
-holds for at least one assignment of the variables, and those values are stated
-as part of the unit test. However, the phrase ``property checking'' is typically
-reserved for \emph{universal} properties: those which should hold \emph{for all}
-values of the variables.
+Some prominent tools for mathematical theory exploration, like \qspec{}, were
+originally developed in the context of \emph{property checking} (although for
+\emph{discovering} software properties, rather than checking them), and only
+later applied more directly to fields like formal mathematics (e.g. via the
+\hipster{} tool for Isabelle/HOL). In particular, the logical domain used by
+\qspec{} (inductive logic with higher-order functions) comes directly from its
+application to Haskell; and its search algorithm makes heavy use of the
+\qcheck{} property checker. The boundary between these two fields is hence
+rather fuzzy, and cross-pollination of techniques is a compelling target for
+research.
+
+Property checking attempts to establish the truth or falsity of a statement (a
+``property''), usually involving components of a software system and, crucially,
+may contain free variables. \emph{Unit testing} is a common method of software
+quality assurance (outside of specialised domains like those which are
+life-critical), which can be characterised as proving \emph{existential}
+properties: each test includes concrete values for any variables, and checking
+is simply a matter of evaluating the code to get a boolean value (pass or fail).
+The phrase ``property checking'' is typically reserved for \emph{universal}
+properties: those which should hold \emph{for all} values of the variables.
+This is a popular method of quality assurance for Haskell software, giving rise
+to a number of packages which approach this challenge in different ways.
 
@@
 accepting values for the free variables as arguments, and returning a boolean
-value (perhaps adorned with useful metadata). If the type of the variables is
-finite (for example booleans), it might be feasible to check all possible
-values. Typically this is not feasible, especially when multiple variables are
-involved, or the input types contain an infinite number of values (for example
+value (perhaps adorned with useful metadata). If the type of the input is finite
+(for example booleans), it might be feasible to check all possible
+values. Typically this is not the case, especially when multiple variables are
+involved, or the types have an infinite number of members (for example
 lists). Property checkers avoid this by only testing a sub-set of values, in the
@@
 \item Repeating until some stop condition is satisfied
-\end{enumerate}.
+\end{enumerate}
 
@@
 programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
-either using explicit generators (via a function called \texttt{forAll}), or by
-choosing a default generator for types implementing its \texttt{Arbitrary}
+either using explicit generators (via a function called \hs{forAll}), or by
+choosing a default generator for types implementing its \hs{Arbitrary}
 typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
@@
 since inputs may be quite large and cover many cases, especially for functions
-which recurse on each \emph{sub-}expressions of their input as well. The output
+which recurse on each \emph{sub-}expression of their input as well. The output
 of a random generator will also vary across runs, rather than checking the same
@@
 recursive datatypes. Consider a type of binary trees, with constructors
-\texttt{Leaf~::~Tree} and \texttt{Node~::~Tree~->~Tree~->~Tree}. We might
-generate a \texttt{Tree} by picking uniformly between these two constructors
-and, if a \texttt{Node} was picked, recursing to generate the two
-sub\texttt{Tree}s. Yet this simple scheme will generate \texttt{Tree}s whose
+\hs{Leaf :: Tree} and \hs{Node :: Tree -> Tree -> Tree}. We might
+generate a \hs{Tree} by picking uniformly between these two constructors
+and, if a \hs{Node} was picked, recursing to generate the two
+sub\hs{Tree}s. Yet this simple scheme will generate \hs{Tree}s whose
 expected size is \emph{unbounded}! This is due to the 50\% chance of generating
-a \texttt{Node}, which requires two recursive calls; hence we can expect each
-call to recurse $0.5~\times~2~=~1$ more time on average, and that call will also
+a \hs{Node}, which requires two recursive calls; hence we can expect each
+call to recurse $0.5~\times~0~+~0.5~\times~2~=~1$ more time on average, and that call will also
 recurse once more on average, and so on; this can easily exhaust all available
@@
 \texttt{derive} package, suffer from this problem. It can be avoided by biasing
-the generators (in the \texttt{Tree} example we would produce a \texttt{Leaf}
+the generators (in the \hs{Tree} example we would produce a \hs{Leaf}
 more than half of the time), but this bias compounds at each level of recursion,
-limiting testing to very small values. More sophisticated solutions exist, such
-as the \texttt{boltzmann-sampling} package (see \cite{duchon2004boltzmann}) and
-even domain-specific languages~\cite{lampropoulos2017beginner}, but all of these
-add complications to the property checking process.
+limiting testing to exponentially small values. More sophisticated solutions
+exist, such as the \texttt{boltzmann-sampling} package (see
+\cite{duchon2004boltzmann}) and even domain-specific
+languages~\cite{lampropoulos2017beginner}, but all of these add complications to
+the property checking process.
 
@@
 understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
-provided: users can supply a \texttt{shrink} function along with their
-generator, which is given any counterexample found during testing and returns a
-list of smaller values. In the above example we might shrink a \texttt{Node} by
-returning each sub-tree, and return an empty list when asked to shrink a
-\texttt{Leaf}. If one of these smaller values is also a counterexample, it too
-is shrunk and the process repeats, otherwise it is discarded and the next shrunk
-value is tried. When a value cannot be shrunk any more, or if none of its
-shrunken versions are counterexamples, that value is presented to the user in
-the hope that it is simple enough to understand.
+provided: users can supply a \hs{shrink} function along with their generator,
+which is given any counterexample found during testing and returns a list of
+smaller values. In the above example we might shrink a \hs{Node} by returning
+each sub-tree, and return an empty list when asked to shrink a \hs{Leaf}. If one
+of these smaller values is also a counterexample, it too is shrunk and the
+process repeats, otherwise it is discarded and the next element of the list is
+tried. When a value cannot be shrunk any more, or if none of its shrunken
+versions are counterexamples, that value is presented to the user in the hope
+that it is simple enough to understand.
 
@@
 values (either using a user-provided enumeration function or using its
-\texttt{Serial} typeclass), rather than generating values
+\hs{Serial} typeclass), rather than generating values
 randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
 the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
-\texttt{Tree} example this might check \texttt{Leaf}, then \texttt{Node Leaf
-  Leaf}, then \texttt{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
+\hs{Tree} example this might check \hs{Leaf}, then \hs{Node Leaf
+  Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
 order of ``size'' ensures that counterexamples are always small, that tests are
@@
 which will trigger an exception if evaluated. Since catching a Haskell exception
-is an \texttt{IO} action, pure properties which inspect such values will always
-be aborted, handing control back to \textsc{LazySmallCheck} which checks the
-same property on refined versions of the previous value. For example, a lazily
-generated \texttt{Tree} may initially be equivalent to $\bot$. If the property
+is an \hs{IO} action, pure properties which inspect such values will always
+be aborted, handing control back to \textsc{LazySmallCheck} which refines the
+previous value and tries again. For example, a lazily generated \hs{Tree} may
+initially be equivalent to $\bot$ (an undefined value). If the property
 aborts, it will be run again with each possible refinement of $\bot$: in this
-case inputs equivalent to \texttt{Leaf} and \texttt{Node~$\bot_1$~$\bot_2$}. If
+case inputs equivalent to \hs{Leaf} and \hs{Node}~$\bot_1$~$\bot_2$. If
 the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
-give inputs \texttt{Node~Leaf~$\bot_2$} and
-\texttt{Node~(Node~$\bot_3$~$\bot_4$)~$\bot_2$} (and likewise if $\bot_2$ had
-caused the exception). This refinement continues until the property no longer
-aborts or until a predefined depth limit is reached. If the property returns a
-true value when given such partially-defined input then it must succeed for
-\emph{all} refinements of that input (since the output was determined without
-evaluating the undefined sub-terms). If the property returns a false value, the
-input is a \emph{minimal} counterexample, since it only specifies those
-sub-terms which are required to violate the property; any undefined sub-terms
-are irrelevant to the failure.
+give inputs \hs{Node Leaf}~$\bot_2$ and
+\hs{Node (Node}~$\bot_3$~$\bot_4$\hs{)}~$\bot_2$ (and likewise if $\bot_2$
+had caused the exception). This refinement continues until the property no
+longer aborts or until a predefined depth limit is reached. If the property
+returns a true value when given such partially-defined input then it must
+succeed for \emph{all} refinements of that input (since the output was
+determined without evaluating the undefined sub-terms). If the property returns
+a false value, the input is a \emph{minimal} counterexample, since it only
+specifies those sub-terms which are required to violate the property; any
+undefined sub-terms are irrelevant to the failure.
 

### 2019-04-29 PhDThesis/litreview_automatedreasoning.tex
Fixes

@@
 Automated logical and mathematical reasoning has been pursued since at least the
-time of mechanical calculators like the Pascaline~\cite{d'ocagne}. A
+time of mechanical calculators like the Pascaline~\cite{ocagne93:_le_calcul_simpl}. A
 recurring theme in these efforts is the separation between those undertaken by

### 2019-05-14 PhDThesis/background_haskell.tex
Changed Haskell background

@@
 \label{sec:haskell}
+
+\begin{figure}
+  \centering
+  \begin{haskell}
+-- A datatype with two constructors
+data Bool = True | False
+
+-- A recursive datatype (S requires a Nat as argument)
+data Nat = Z | S Nat
+
+-- A polymorphic (AKA "generic") datatype
+data List t = Nil | Cons t (List t)
+
+-- Arithmetic functions
+
+plus :: Nat -> Nat -> Nat
+plus    Z  y = y
+plus (S x) y = S (plus x y)
+
+mult :: Nat -> Nat -> Nat
+mult    Z  y = Z
+mult (S x) y = plus y (mult x y)
+
+-- Mutually-recursive functions
+
+odd :: Nat -> Bool
+odd    Z  = False
+odd (S n) = even n
+
+even :: Nat -> Bool
+even    Z  = True
+even (S n) = odd n
+  \end{haskell}
+  \caption{Haskell code, defining datatypes and functions involving them (note
+    that \hs{--} introduces a comment). \hs{Bool} is the Booleans, \hs{Nat} is a
+    Peano encoding of the natural numbers and \hs{List t} are polymorphic lists
+    with elements of type \hs{t}. Functions are called using whitespace, e.g.
+    \hs{f x}, and may be defined by pattern-matching (case analysis) on their
+    arguments. \hs{A :: B} is an optional type annotation, stating that value
+    \hs{A} has type \hs{B}. \hs{A -> B} is the type of functions from \hs{A} to
+    \hs{B}.}
+  \label{fig:haskellexample}
+\end{figure}
 
@@
 Isabelle). Haskell is well-suited to programming language research; indeed, this
-was a goal of the language's creators \cite{marlow2010haskell}. Like most
-\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
-with extra features such as a strong type system and ``syntactic sugar'' to
-improve readability.
+was a goal of the language's creators \cite{marlow2010haskell}. An example of
+Haskell code is given in Figure~\ref{fig:haskellexample}.
 
-The following features make Haskell especially useful for our purposes, although
-many are also present in other languages such as StandardML and Coq (which we
-also use, but only when needed for compatibility with prior work):
+Like most \emph{functional} programming languages, Haskell builds upon
+$\lambda$-calculus, with extra features such as a strong type system and
+``syntactic sugar'' to improve readability. The following features make Haskell
+especially useful for our purposes, although many are also present in other
+languages such as StandardML and Coq (which we also use, but only when needed
+for compatibility with prior work):
 
@@
 
-\item{Functional}: All control flow in Haskell is performed by function
-  abstraction and application, which we can reason about using standard rules of
-  inference such as \emph{modus ponens}.
+\item{Functional}: Most control flow in Haskell (all except
+  \emph{pattern-matching}, described below) is performed by function abstraction
+  and application , which we can reason about using standard rules of inference
+  such as \emph{modus ponens}. For example, reading from external sources (like
+  environment variables) is \emph{impure}, so the only way to parameterise a
+  value of type \hs{B} with a value of type \hs{A} is using a function of type
+  \hs{A -> B}. Conversely, applying a function of type \hs{A -> B} to a value of
+  type \hs{A} can only ever produce a value of type \hs{B} (it may instead crash
+  the program or loop forever, but neither of those are \emph{observable}, i.e.
+  we can't branch on them).
 
-\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation
-  of expressions; hence our reasoning can safely ignore complicated external and
-  non-local interactions.
+\item{Pure}: Execution of actions (e.g. reading or writing files) is separate to
+  evaluation of expressions; hence our reasoning can safely ignore complicated
+  external and non-local interactions. Purity ensures \emph{referential
+    transparency}: references can be substituted for their referent with
+  no change in semantics (as in standard mathematical practice). This implies
+  that calling a function twice with the same argument must produce the same
+  result (modulo crashes or infinite loops). For example, consider applying the
+  function \hs{pair x = (x, x)} to some arbitrary function call \hs{foo y}. The
+  references \hs{x} in the resulting pair \hs{(x, x)} both refer to \hs{foo y}
+  so, by referential transparency, can be substituted to get
+  \hs{(foo y, foo y)}. Both members of \hs{(x, x)} are identical, by definition;
+  hence, to preserve the semantics, both (and indeed \emph{all}) calls to
+  \hs{foo y} must also be identical. This holds regardless of what we choose for
+  \hs{foo} and \hs{y}, and implies that function behaviour cannot depend on
+  external interactions or non-deterministic processes (which may change between
+  calls).
 
 \item{Statically Typed}: Expressions are constrained by \emph{types}, which can
-  be used to eliminate unwanted combinations of values, and hence reduce search
-  spaces; \emph{static} types can be deduced syntactically, without having to
-  execute the code.
+  be used to eliminate unwanted combinations of values (for example
+  \hs{plus True} is not a valid expression), and hence reduce search spaces;
+  \emph{static} types can be deduced syntactically, without having to execute
+  the code. The type of Haskell expressions can also be \emph{inferred}
+  automatically.\footnote{Except for certain cases, such as those caused by the
+    \emph{monomorphism restriction}~\cite{marlow2010haskell}[$\S$4.5.5]}
 
-\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an
-  expression (i.e. performing function calls) without diverging, then a
-  non-strict evaluation strategy will not diverge when evaluating that
-  expression. This is rather technical, but in simple terms it allows us to
-  reason effectively about a Turing-complete language, where evaluation may not
-  terminate. For example, when reasoning about \emph{pairs} of values \hs{(x,
-    y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an
-  ``obvious'' rule such as
-  $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$. Haskell's
-  non-strict semantics makes this equation valid; whilst it would \emph{not} be
-  valid in the strict setting common to most other languages, where the
-  expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter
-  the semantics, if \hs{x} doesn't diverge).
+\item{Curried}: All functions in Haskell take a single argument (as in
+  $\lambda$-calculus), which makes them easier to manipulate programatically.
+  Currying allows multi-argument functions to be simulated, by accepting one
+  argument and returning a function to accept the rest. The \hs{mult} function
+  in Figure~\ref{fig:haskellexample} has type \hs{Nat -> Nat -> Nat} meaning it
+  takes a \hs{Nat} as argument and returns a function of type
+  \hs{Nat -> Nat}. Function calls are written with whitespace, so \hs{mult x y}
+  calls the \hs{mult} function with the argument \hs{x}, then calls the result
+  of that with the argument \hs{y}. This allows \emph{partial application} such
+  as \hs{double = mult (S (S Z))}, but for our purposes it is important for
+  unifying calling conventions. For example, in Javascript \hs{mult x y} would
+  be either \js{mult(x, y)} or \js{mult(x)(y)}, and depends on the definition of
+  \js{mult} (the problem compounds with more arguments). In Haskell there is no
+  distinction between these forms.
+
+\item{Non-strict}: ``Strict'' evaluation strategies evaluate the arguments of a
+  function call before the function body; non-strict does the opposite. The
+  Haskell standards do not specify a particular evaluation strategy, but they
+  do require that it be non-strict (for efficiency, most implementations use
+  \emph{lazy} evaluation to avoid duplicating work). Strictness can result in
+  infinite loops and other errors which may be avoided by non-strictness,
+  making the latter more useful for reasoning in the face of such errors.
+  For example, given a pair of values \hs{(x, y)} and a projection function
+  \hs{fst}, we might make the ``obvious''  conjecture that
+  \hs{fst (x, y) = x}.\footnote{Incidentally, this is also a valid definition of
+    the \hs{fst} function} This statement is true for non-strict languages, but
+  \emph{not} for strict languages. Crucially, a strict language will attempt to
+  calculate the value of \hs{y}, which may cause an infinite loop or other
+  error; a non-strict language like Haskell will ignore \hs{y} since it isn't
+  used in the body of the \hs{fst} function.
 
@@
   user-defined data representations, and an inverse mechanism to inspect these
-  data by \emph{pattern-matching}. For our purposes, the useful consequences of
-  ADTs and pattern-matching include their amenability for inductive proofs and
-  the fact they are \emph{closed}; i.e. an ADT's declaration specifies all of
-  the normal forms for that type. This makes exhaustive case analysis trivial,
-  which would be impossible for \emph{open} types (for example, consider classes
-  in an object oriented language, where new subclasses can be introduced at any
-  time).
+  data by \emph{pattern-matching} (Haskell's other form of control flow). The
+  \hs{Bool}, \hs{Nat} and \hs{List t} definitions in
+  Figure~\ref{fig:haskellexample} are ADTs; whilst the functions use
+  pattern-matching to branch on their first argument. For our purposes, the
+  useful consequences of ADTs and pattern-matching include their amenability for
+  inductive proofs and the fact they are \emph{closed}; i.e. an ADT's
+  declaration specifies all of the normal forms for that type. This makes
+  exhaustive case analysis trivial, which would be impossible for \emph{open}
+  types (for example, consider classes in an object oriented language, where new
+  subclasses can be introduced at any time).
 
@@
   \emph{type-level} objects; provided those objects are never inspected. This
-  has the \emph{practical} benefit of enabling \emph{polymorphism}: for example,
-  we can write a polymorphic identity function \hs{id :: forall t. t ->
-    t}. \footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a
-      -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually,
-  this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of
-  type \hs{t}; yet only the latter is available in the function body,
-  e.g. \hs{id x = x}. This inability to inspect type-level arguments gives us
-  the \emph{theoretical} benefit of being able to characterise the behaviour of
-  polymorphic functions from their type alone, a technique known as
-  \emph{theorems for free} \cite{wadler1989theorems}.
+  enables \emph{polymorphism}. The \hs{List t} type in
+  Figure~\ref{fig:haskellexample} is an example: there are many useful functions
+  involving lists which work in the same way regardless of the element type
+  (e.g. getting the length, appending, reversing, etc.). An even simpler example
+  is the polymorphic identity function \hs{id x = x}. The type of \hs{id} is
+  \hs{forall t. t -> t}\footnote{The \hs{forall t.} is optional; type-level
+    identifiers beginning with a lowercase letter are assumed to be universally
+    quantified variables.}, which we can view as taking \emph{two} parameters:
+  a type \hs{t} and a value of type \hs{t}. Only the latter argument appears in
+  the definition (as \hs{x}), meaning that we can't use the type \hs{t} to
+  determine the function's behaviour. Indeed, in the case of \hs{id} we can't
+  branch on the value of \hs{x} either, since we don't know what type it might
+  have (our definition must work \emph{for all} types \hs{t}); the only
+  functions we can call on \hs{x} must also be polymorphic, and hence also
+  incapable of branching. The type of \hs{id} states that it returns a value of
+  type \hs{t}; without knowing what that type is, the only type-correct value it
+  can return is the argument \hs{x}. Hence the \emph{type} of \hs{id} tells us
+  everything about its behaviour, with this style of reasoning known as
+  \emph{theorems for free}~\cite{wadler1989theorems}. Haskell definitions are
+  commonly made polymorphic like this, to prevent incorrect implementations
+  passing the type checker, e.g. \hs{fst :: (Nat, Nat) -> Nat} might return the
+  wrong element, but \hs{fst :: (t1, t2) -> t1} can't.
 
 \item{Type classes}: Along with their various extensions, type classes are
-  interfaces which specify a set of operations over a type (or other type-level
-  object, such as a \emph{type constructor}). Many type classes also specify a
-  set of \emph{laws} which their operations should obey but, lacking a simple
-  mechanism to enforce this, laws are usually considered as documentation. As a
-  simple example, we can define a type class \hs{Semigroup} with the following
-  operation and associativity law:
+  interfaces which specify a set of operations over a type or other type-level
+  object (like a \emph{type constructor}, e.g. \hs{List}). Many type classes
+  also specify a set of \emph{laws} which their operations should obey but,
+  lacking a simple mechanism to enforce this, laws are usually considered as
+  documentation. As a simple example, we can define a type class \hs{Semigroup}
+  with the following operation and an associativity law:
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+  \begin{center}
+    \begin{haskell}
 op :: forall t. Semigroup t => t -> t -> t
-\end{lstlisting}
+    \end{haskell}
 
-$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$
+    $\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} =
+                                \text{\hs{op (op x y) z}}$
+  \end{center}
 
@@
   restricts the possible types \hs{t} to only those which implement
-\hs{Semigroup}. \footnote{Alternatively, we can consider \hs{Semigroup t} as the
-  type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case
+  \hs{Semigroup}. \footnote{Alternatively, we can consider \hs{Semigroup t} as
+    the type of ``implementations of \hs{Semigroup} for \hs{t}'', in which case
   \hs{=>} has a similar role to \hs{->} and we can consider \hs{op} to take
@@
 
-\item{Modularity}: Haskell has a module system, where each module may specify an
-  \emph{export list} containing the names which should be made available for
-  other modules to import. When such a list is given, any expressions \emph{not}
-  on the list are considered \emph{private} to that module, and are hence
-  inaccessible from elsewhere. This mechanism allows modules to provide more
-  guarantees than are available just in their types. For example, a module may
-  represent email addresses in the following way:
+\item{Modularity}: Haskell's module system allows definitions to be kept
+  private. This mechanism allows modules to provide more guarantees than are
+  available just in their types, by constraining the ways that values can be
+  constructed. For example, the following module represents email addresses as a
+  pair of \hs{String}s, one for the user part and one for the host:
 
   \begin{haskell}
+-- Exports appear between the parentheses
 module Email (Email(), at, render) where
@@
 render :: Email -> String
-    render (E u h) = u ++ "@" ++ h
+render (E user host) = user ++ "@" ++ host
 
+-- if/then/else is sugar for pattern-matching a Bool
 at :: String -> String -> Maybe Email
-    at "" h  = Nothing
-    at u  "" = Nothing
-    at u  h  = Just (E u h)
+at user host = if user == "" || host == ""
+                  then Nothing
+                  else Just (E user host)
   \end{haskell}
 
-  This \hs{Email} type can be constructed in two ways: directly, via the \hs{E}
-  constructor, which requires both a user part and a host part (both
-  \hs{String}s, for simplicity); or indirectly, via the \hs{at} function, which
-  will only call \hs{E} if both \hs{String}s are non-empty (a programming
-  pattern known as a ``smart constructor''). The export list in the first line
-  contains the \hs{at} function, but not the \hs{E} constructor~\footnote{The
-    syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any
-    of its constructors.}, so any code using this module is only able to create
-  \hs{Email} values with non-empty parts.
+  An \hs{Email} value can be constructed by passing any two \hs{String}s to
+  \hs{E}, but \hs{E} is private (not exported). The \hs{at} function is
+  exported, but only passes its arguments to \hs{E} iff they are not empty.\footnote{\hs{Maybe t} is a
+    safer alternative to the \textsc{Null} construct of other
+    languages~\cite{hoare2009null}. It is defined as
+    \hs{data Maybe t = Nothing | Just t} and can be understood as an optional
+    value, or a computation which may fail, or as a list with at most one
+    element, or as a degenerate search tree with no
+    backtracking~\cite{wadler1985replace}} Since this module never calls \hs{E}
+  with empty \hs{String}s, and other modules must use \hs{at}, we're guaranteed
+  that \emph{all} \hs{Email} values will have non-empty \hs{String}s. Such
+  ``smart constructors'' can guarantee \emph{any} decidable property, at the
+  cost of performing run-time checks on each invocation.\footnote{We can
+    guarantee non-emptiness ``by construction'', without run-time checks or
+    \hs{Maybe} wrappers, by changing the type to require at least one element.
+    \hs{String} is equivalent to \hs{List Char}, with \hs{""} represented as
+    \hs{Nil}. Changing \hs{Nil} to require a \hs{Char} would eliminate empty
+    \hs{String}s, e.g. \hs{data NonEmpty t = Nil t | Cons t (NonEmpty t)}. We
+    could also use a pair like \hs{data NonEmpty t = NE t (List t)} instead.
+    Such precise types are often more desirable than smart constructors, but are
+    less general since ad-hoc representations need to be invented to enforce
+    each guarantee.}
 
@@
 \end{itemize}
-
-\newcommand{\CVar}{\texttt{Var}}
-\newcommand{\CLit}{\texttt{Lit}}
-\newcommand{\CApp}{\texttt{App}}
-\newcommand{\CLam}{\texttt{Lam}}
-\newcommand{\CLet}{\texttt{Let}}
-\newcommand{\CCase}{\texttt{Case}}
-\newcommand{\CType}{\texttt{Type}}
-\newcommand{\CLocal}{\texttt{Local}}
-\newcommand{\CGlobal}{\texttt{Global}}
-\newcommand{\CConstructor}{\texttt{Constructor}}
-\newcommand{\CLitNum}{\texttt{LitNum}}
-\newcommand{\CLitStr}{\texttt{LitStr}}
-\newcommand{\CAlt}{\texttt{Alt}}
-\newcommand{\CDataAlt}{\texttt{DataAlt}}
-\newcommand{\CLitAlt}{\texttt{LitAlt}}
-\newcommand{\CDefault}{\texttt{Default}}
-\newcommand{\CNonRec}{\texttt{NonRec}}
-\newcommand{\CRec}{\texttt{Rec}}
-\newcommand{\CBind}{\texttt{Bind}}
 
@@
 equality.
-
-\begin{figure}
-  \begin{haskell}
-    data Nat = Z
-             | S Nat
-
-    plus :: Nat -> Nat -> Nat
-    plus    Z  y = y
-    plus (S x) y = S (plus x y)
-
-    mult :: Nat -> Nat -> Nat
-    mult    Z  y = Z
-    mult (S x) y = plus y (mult x y)
-
-    odd :: Nat -> Bool
-    odd    Z  = False
-    odd (S n) = even n
-
-    even :: Nat -> Bool
-    even    Z  = True
-    even (S n) = odd n
-  \end{haskell}
-  \caption{A Haskell datatype for Peano numerals with some simple arithmetic
-    functions, including mutually-recursive definitions for \hs{odd} and
-    \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
-    regarded as \hs{data Bool = True | False}.}
-  \label{fig:haskellexample}
-\end{figure}
 
@@
 
-Figure~\ref{fig:haskellexample} shows some simple Haskell function definitions,
-along with a custom datatype for Peano numerals. The translation to Core syntax
-is routine, and shown in Figure~\ref{fig:coreexample}. Although the Core is more
-verbose, we can see that similar structure in the Haskell definitions gives rise
-to similar structure in the Core; for example, the definitions of \hs{odd} and
-\hs{even} are identical in both languages, except for the particular identifiers
-used. It is this close correspondence which allows us to analyse Core
-expressions in place of their more complicated Haskell source.
+GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
+shows the Core representation of functions from Figure~\ref{fig:haskellexample}.
+Although the Core is more verbose, we can see that similar structure in the
+Haskell definitions gives rise to similar structure in the Core; for example,
+the definitions of \hs{odd} and \hs{even} are identical in both languages,
+except for the particular identifiers used. It is this close correspondence
+which allows us to analyse Core expressions in place of their more complicated
+Haskell source.
 
 Note that we exclude representations for type-level entities, including datatype
-definitions like that of \hs{Nat}. GHC can represent these, but in this work we
-only consider reducible expressions (i.e. value-level bindings of the form
-\mbox{\hs{f a b ... = ...}}).
+definitions like that of \hs{Nat}. GHC can represent these, but we are mostly
+concerned with dynamic behaviour, which excludes types (they are \emph{erased}
+during compilation). For simplicity we also avoid using constructors directly,
+choosing instead to wrap them with normal functions (e.g. \hs{s x = S x}), since
+this makes our domain more symmetric and reduces cross-language differences.

### 2019-05-14 PhDThesis/background_quickcheck.tex
Minor changes to QuickCheck background

@@
 totally inadequate for many (e.g. life-) critical systems. To see why, consider
-the following Haskell function, along with some unit tests:
+the following Haskell function, along with some unit
+tests\footnote{Backticks allow functions to be written as infix operators, so
+  \hs{x `div` y} is equivalent to \hs{div x y}.}:
 
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+\begin{haskell}
 fact_base        = factorial 0 == factorial 1
@@
 fact_neg       n = factorial n == factorial (-n)
-\end{lstlisting}
+\end{haskell}
 
@@
 property asserting $\forall n, factorial(n) \leq factorial(n+1)$. Notice that
-unit tests like \hs{fact\_base} are valid properties; they just assert rather
+unit tests like \hs{fact_base} are valid properties; they just assert rather
 weak statements.
 
-To check these properties, \qcheck{} treats closed terms (like \hs{fact\_base})
+To check these properties, \qcheck{} treats closed terms (like \hs{fact_base})
 just like unit tests: pass if they evaluate to \hs{True}, fail otherwise. For

### 2019-05-14 PhDThesis/ml4hs.tex
Fix listings

@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=0pt, xrightmargin=0pt]
+\begin{haskell}
 toTree :: Context -> Expr -> RoseTree
@@
   ...
-\end{lstlisting}
+\end{haskell}
 
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+\begin{haskell}
 phiL :: Context -> Local -> Feature
@@
   Just i  -> (2 * alpha) + i
-\end{lstlisting}
+\end{haskell}
 
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
+\begin{haskell}
 featureVec :: Expr -> [Feature]
@@
         pad xs = take c (xs ++ repeat 0)
-\end{lstlisting}
+\end{haskell}
 

### 2019-05-14 PhDThesis/appendix_nixeval.tex
Add nix-eval appendix

\subsection{\texttt{nix-eval}}

Haskell's support for runtime evaluation of code (a la Lisp's \eval) is
limited. The most popular mechanism is the \hint package, which is built
on top of GHC's API and provides functions like
\hs{eval :: MonadInterpreter m => String -> m String}. The \hs{MonadInterpreter}
constraint ensures that GHC can be invoked and for managing its ``session'',
which includes details like compiler flags, a list of imported modules and paths
to compiled packages.

This function doesn't return arbitrary types of value; only \hs{String}s. This
is due to the given expression being passed to the \hs{show} function (e.g.
\hs{eval "length []"} will produce \hs{"0" :: String}, not
\hs{0 :: Int}).\footnote{\hs{show :: Show a => a -> String} is a method of the
  \hs{Show} type class, and is the standard way to equip Haskell types values
  with a printable representation (akin to Java's \textt{toString} method).
  \hs{show} has no associated laws, but it is conventional to make it the
  inverse of the \hs{read :: Read a => String -> a} method of the \hs{Read} type
  class (where possible).} Serialising results in this way simplifies the
interface between evaluation contexts (in particular, type checking and
inference doesn't need to cross the interface); at the cost of having to
explicitly deserialise results into the desired type as needed.

One major problem with the \hint package is that the available packages
(and GHC version) are hard-coded into the ``outer'' Haskell context. For
example, a Haskell interpreter based on \hint would fail to evaluate
expressions that use packages or GHC features other than those the interpreter
was compiled against.\footnote{The package database could be overridden in
  principle, but compatibility issues and version conflicts would make this
  unreliable and error-prone in practice.} For our research we want a single
program for exploring \hs{arbitrary} Haskell code, which makes approaches like
\hint undesirable.

Our solution is the \nixeval package. This provides a function
\hs{eval :: Expr -> IO (Maybe String)}, which returns a \hs{String} for the same
reasons as \hint, but differs in its handling of expressions and context. Rather
than operating on a raw \hs{String}, \nixeval defines a type \hs{Expr}
containing several fields to keep track of the context in which the expression
is to be evaluated:

\begin{itemize}
\item \hs{eExpr :: String} is the raw code itself.
\item \hs{ePkgs :: [Pkg]} lists the packages which need to be installed.
\item \hs{eMods :: [Mod]} lists the modules which need to be imported.
\item \hs{eFlags :: [Flag]} lists flags which need passing to the compiler (e.g.
  to enable certain language features).
\item \hs{ePreamble :: [String]} lists code that needs to be defined before
  evaluating (for example, custom function definitions).
\end{itemize}

The types \hs{Pkg}, \hs{Mod} and \hs{Flag} are just wrappers containing a
\hs{String}; they are kept distinct to allow the type checker to catch
mistakes.\footnote{Avoiding such distinctions (i.e. using types to track
  representations but not semantics) is known as ``stringly-typed programming''
  and contributes to widespread errors like injection attacks. In the case of
  binary choices this principle is known as ``boolean blindness''.} These
wrapper types, as well as \hs{Expr} itself, implement the \hs{IsString} type
class, which allows GHC's \texttt{OverloadedStrings} language extension to parse
them from quoted literals, for example:

\begin{haskell}
textToUpper :: Expr
textToUpper = withPkgs       ["text"     ]  -- This is a  Pkg
                (withMods    ["Data.Text"]  -- This is a  Mod
                  (withFlags ["-O2"      ]  -- This is a  Flag
                    "toUpper"))             -- This is an Expr
\end{haskell}

The main advantage over \hint is the reification of packages, which can be
computed in arbitrary ways (including from user input). When an \hs{Expr} is
passed to our \hs{eval} function, a new \texttt{runhaskell} process is launched
with the given \hs{eFlags} (hence the presence of \hs{IO}). The elements of
\hs{eMods} are prefixed with \hs{"import "} and fed into its standard input
stream, followed by the contents of \hs{ePreamble}. The \hs{eExpr} code is
wrapped in parentheses and prefixed with \hs{"main = Prelude.putStr"} which acts
as the entry point of the generated code (\hs{Prelude.putStr} writes a
\hs{String} to standard output).\footnote{The code generation for \hs{eExpr} is
  actually user-configurable, in case the default is too simplistic. Unlike
  \hint we never force calls to the \hs{show} method, since normal function
  calls are easier to specify and override than type class instances.} If the
subprocess exits successfully, its standard output is returned (wrapped in
\hs{Just}); otherwise it is written to the standard error stream and a
\hs{Nothing} value is returned (the subprocess inherits the parent's standard
error stream).

The \texttt{runhaskell} command can be invoked in two ways: if
\texttt{runhaskell} is available via the \texttt{PATH} environment variable, and
the contents of \hs{ePkgs} are all installed in its package database, then we
invoke that command directly. If not, we use the \texttt{nix-shell} command of
the Nix package manager to invoke \texttt{runhaskell} in a sandbox~\cite{TODO}.
We use the Nix language to request a sandbox containing GHC and all of the
elements of \hs{ePkgs}. We also allow the default Nix packages to be overridden
via an environment variable, which allows the user to define their own packages,
select particular versions, or perform any other arbitrary setup (since the Nix
language is Turing-complete).

Nix packages are immutable and (mostly) reproducible, which allows sandboxes and
their components to be re-used, and even shared across machines. However, any
package which isn't cached in this way must be built from source, which can
cause \hs{eval} to take an arbitrarily long time. Even a cached sandbox can take
a few seconds to initialise (largely due to I/O latency as Nix computes and
inspects the sandbox components). To avoid too much overhead, it is advisable to
invoke \hs{eval} as few times as possible (preferably only once). If we need to
dynamically evaluate multiple parts of a program, it might be preferable to move
them all into one big \hs{Expr} value.

\begin{figure}
  \begin{haskell}
-- An Expr representing a list of the given Exprs
asList :: [Expr] -> Expr
asList = foldr (\x -> (("(:)" $$ x) $$)) "[]"

-- Composes the given list of functions using the (.) operator
compose :: [Expr] -> Expr
compose [x]    = x
compose (x:xs) = ("(.)" $$ x) $$ compose xs

-- Apply x to n undefined arguments; result is undefined with x's output type
addCalls :: Int -> Expr -> Expr
addCalls 0 x = x
addCalls n x = addCalls (n-1) (x $$ "undefined")

-- Coerces values to have the output type of a known function "f" (of arity a)
converter :: Expr
converter = ("flip" $$ "asTypeOf") $$ addCalls a "f"

-- Coerces a value to the output type of "f", serialises it then hashes it
hasher :: Expr
hasher = compose [withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "asWord32"),
                  withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "hash32"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "runPut"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "put"),
                  converter]
  \end{haskell}
  \caption{Example usage of \nixeval functions, taken from \textsc{ML4HS}, which
    invokes \qspec on arbitrary user-provided functions.}
  \label{fig:nix-eval}
\end{figure}

Using \hs{String} functions to manipulate such complicated values soon becomes
unwieldy, so a suite of combinators are provided for composing \hs{Expr} values
in useful ways; examples are shown in Figure~\ref{fig:nix-eval}. Besides the
\hs{withFoo} functions shown before, the \hs{qualified} function can reduce the
ambiguity of names while avoiding repetition of module names: a call like
\hs{qualified "Data.Text" "toUpper"} returns an \hs{Expr} of the name
\hs{"Data.Text.toUpper"} with the \hs{"Data.Text"} \hs{Mod}ule in its
\hs{eMods} list. \hs{foo $$ bar} denotes function application (analogous to
Haskell's standard \hs{$}%$ operator), by juxtaposing the \hs{eExpr} of \hs{foo}
with that of \hs{bar} (wrapping both in parentheses to ensure precedence); the
contexts are appended. Finally, the \hs{asString :: Show a => a -> Expr}
function calls \hs{show} on the given value and wraps the resulting \hs{String}
into an \hs{Expr}; this is especially useful for representing literal
\hs{String}s, since their \hs{show} instance will wrap them in quotes and
perform any required escaping.

Despite being more structured than raw \hs{String}s, the the above API still
suffers from a form of "stringly-typed programming"; notably, that every encoded
expression has the same type: \hs{Expr}. We can avoid this through the use of
\emph{phantom types}: overly-restrictive type annotations, which are unrelated
to the encoding but must still match up in order to pass the type checker. In
particular we can represent typed expressions like this:

\begin{haskell}
newtype TypedExpr t = TE Expr
\end{haskell}

Notice that the parameter \hs{t} doesn't appear on the right hand side: the
values of \hs{TypedExpr} are just those of \hs{Expr} (the phantom type is
\emph{erased} after type-checking). The \hs{TE} constructor is hence polymorphic
in \hs{t}, so \hs{TE x :: TypedExpr T} is valid for \emph{any} choice of \hs{x}
and \hs{T} (similar to how \hs{Nil :: List T} is valid for any choice of
\hs{T}). In particular, this includes the type of the expression
\emph{represented by} \hs{x}, for example \hs{TE "True" :: TypedExpr Bool} is a
type-correct expression.

Haskell will infer the most general type of an expression if none is provided
(i.e. \hs{forall t. TypedExpr t}), but will honour more restricted annotations
if given. This allows the types represented by literal expressions to be
asserted, for example:

\begin{haskell}
nothing' :: TypedExpr (Maybe a)
nothing' = TE (qualified "Data.Maybe" "Nothing")

just' :: TypedExpr (a -> Maybe a)
just' = TE (qualified "Data.Maybe" "Just")
\end{haskell}

Annotations on the combinator library can propagate and check these phantom
types, most notably:

\begin{haskell}
-- Typed function application
($$$) :: TypedExpr (a -> b) -> TypedExpr a -> TypedExpr b
TE f $$$ TE x = TE (f $$ x)

-- Only expressions of type String can be evaluated, since we use Prelude.putStr
tEval :: TypedExpr String -> IO (Maybe String)
tEval (TE x) = eval x
\end{haskell}

Haskell will accept any annotation, so care must be taken to ensure that they
correspond correctly to the encoded expressions. If user-provided literals
aren't required then the \hs{TE} constructor can be kept private to prevent
unsound assertions or coercions. However, there is a more subtle problem with
such \hs{TypedExpr}s: the whole point of \nixeval is to manipulate and evaluate
code whose dependencies are unknown or unavailable at compile time. For example,
we cannot assign a meaningful type at compile time to an expression which has
been generated from user input. Even if we know the value of an expression at
compile time, its type may come from a dependency that our program doesn't have
access to. In these cases the \hs{TypedExpr} approach can still be useful for
auxiliary code whose types are known to us; those unknown or unavailable parts
can either be unwrapped into raw \hs{Expr}s, or dummy types can be used instead.
For example, if we are manipulating code involving the \hs{Data.Text.Text} type,
but we don't have or want the \textt{text} package in our dependencies, we can
declare an empty type like \hs{data TEXT} (note that this declaration has no
right-hand side) and use this in our annotations in place of
\hs{Data.Text.Text}. The compiler will type-check our code as desired, ensuring
functions are called on appropriate arguments, \hs{tEval} is given on a
\hs{String}, etc.

In the course of writing MLSpec we experimented with both raw \hs{Expr} and
\hs{TypedExpr} values. Whilst we found the latter easier for constructing large
expressions, we ultimately found it even more productive to move as much
functionality as possible into a standalone Cabal package
(\texttt{mlspec-helper}) and have MLSpec call into that code from simple
\hs{Expr} values.

### 2019-05-14 PhDThesis/appendix_nixeval.tex
Fix markup

@@
   \hs{Show} type class, and is the standard way to equip Haskell types values
-  with a printable representation (akin to Java's \textt{toString} method).
+  with a printable representation (akin to Java's \texttt{toString} method).
   \hs{show} has no associated laws, but it is conventional to make it the
@@
 For example, if we are manipulating code involving the \hs{Data.Text.Text} type,
-but we don't have or want the \textt{text} package in our dependencies, we can
+but we don't have or want the \texttt{text} package in our dependencies, we can
 declare an empty type like \hs{data TEXT} (note that this declaration has no

### 2019-05-16 PhDThesis/appendix_nixeval.tex
Appendix changes

--- 2019-05-14 PhDThesis/appendix_nixeval.tex
@@
-\subsection{\texttt{nix-eval}}
+\section{\texttt{nix-eval}}
+\label{sec:nix-eval}
 
-Haskell's support for runtime evaluation of code (a la Lisp's \eval) is
-limited. The most popular mechanism is the \hint package, which is built
-on top of GHC's API and provides functions like
-\hs{eval :: MonadInterpreter m => String -> m String}. The \hs{MonadInterpreter}
-constraint ensures that GHC can be invoked and for managing its ``session'',
-which includes details like compiler flags, a list of imported modules and paths
-to compiled packages.
+\begin{sloppypar}
+  Haskell's support for runtime evaluation of code (\'a la Lisp's
+  \eval{}~\cite[\S~1.6]{edwards1962lisp}) is limited. The most popular mechanism
+  is the \hint{} package\footnote{https://hackage.haskell.org/package/hint},
+  which is built on top of GHC's API and provides functions like
+  \hs{eval :: MonadInterpreter m => String -> m String}. The
+  \hs{MonadInterpreter} constraint provides mechanisms for calling into GHC and
+  managing its ``session'', which includes details like compiler flags, a list
+  of imported modules and paths to compiled packages.
 
@@
   is due to the given expression being passed to the \hs{show} function (e.g.
-\hs{eval "length []"} will produce \hs{"0" :: String}, not
-\hs{0 :: Int}).\footnote{\hs{show :: Show a => a -> String} is a method of the
-  \hs{Show} type class, and is the standard way to equip Haskell types values
-  with a printable representation (akin to Java's \textt{toString} method).
-  \hs{show} has no associated laws, but it is conventional to make it the
-  inverse of the \hs{read :: Read a => String -> a} method of the \hs{Read} type
-  class (where possible).} Serialising results in this way simplifies the
+  \hs{eval "not True"} will produce \hs{"False" :: String}, not
+  \hs{False :: Bool}).\footnote{\hs{show :: Show a => a -> String} is a method
+    of the \hs{Show} type class, and is the standard way to equip Haskell types
+    with a printable representation (akin to Java's \texttt{toString} method).
+    \hs{show} has no associated laws, but convention is for \hs{show} to be the
+    inverse of the \hs{read :: Read a => String -> a} method of the \hs{Read}
+    type class (where possible).} Serialising results in this way simplifies the
   interface between evaluation contexts (in particular, type checking and
@@
 
-One major problem with the \hint package is that the available packages
+  One major problem with \hint{} is that the available packages
   (and GHC version) are hard-coded into the ``outer'' Haskell context. For
-example, a Haskell interpreter based on \hint would fail to evaluate
-expressions that use packages or GHC features other than those the interpreter
-was compiled against.\footnote{The package database could be overridden in
-  principle, but compatibility issues and version conflicts would make this
-  unreliable and error-prone in practice.} For our research we want a single
-program for exploring \hs{arbitrary} Haskell code, which makes approaches like
-\hint undesirable.
+  example, a Haskell interpreter based on \hint{} would be restricted to only
+  those packages and GHC features that are available during its
+  compilation.\footnote{The package database could be overridden in principle,
+    but compatibility issues and version conflicts would make this unreliable
+    and error-prone in practice.} For our research we want a single program for
+  exploring \emph{arbitrary} Haskell code, regardless of what package it comes
+  from, which makes approaches like those of \hint{} undesirable.
 
-Our solution is the \nixeval package. This provides a function
-\hs{eval :: Expr -> IO (Maybe String)}, which returns a \hs{String} for the same
-reasons as \hint, but differs in its handling of expressions and context. Rather
-than operating on a raw \hs{String}, \nixeval defines a type \hs{Expr}
-containing several fields to keep track of the context in which the expression
-is to be evaluated:
+  To solve this problem we created the \nixeval{}
+  package\footnote{https://hackage.haskell.org/package/nix-eval}. This provides
+  a function \hs{eval :: Expr -> IO (Maybe String)}, which produces \hs{String}s
+  for the same reasons as \hint{}, but differs in its handling of expressions
+  and context. Rather than operating on a raw \hs{String}, \nixeval{} defines
+  the \hs{Expr} type, containing several fields to keep track of the context in
+  which the expression is to be evaluated:
+\end{sloppypar}
 
@@
 mistakes.\footnote{Avoiding such distinctions (i.e. using types to track
-  representations but not semantics) is known as ``stringly-typed programming''
-  and contributes to widespread errors like injection attacks. In the case of
-  binary choices this principle is known as ``boolean blindness''.} These
+  representations but not semantics) is known as ``stringly-typed
+  programming''~\cite{c2stringlytyped}. In the case of binary choices this
+  practice is known as ``boolean blindness''~\cite[\S~26.2]{Harper:2009}.} These
 wrapper types, as well as \hs{Expr} itself, implement the \hs{IsString} type
@@
 
-The main advantage over \hint is the reification of packages, which can be
+The main advantage over \hint{} is the reification of packages, which can be
 computed in arbitrary ways (including from user input). When an \hs{Expr} is
@@
 wrapped in parentheses and prefixed with \hs{"main = Prelude.putStr"} which acts
-as the entry point of the generated code (\hs{Prelude.putStr} writes a
-\hs{String} to standard output).\footnote{The code generation for \hs{eExpr} is
-  actually user-configurable, in case the default is too simplistic. Unlike
-  \hint we never force calls to the \hs{show} method, since normal function
-  calls are easier to specify and override than type class instances.} If the
+as the entry point of the generated code (writing the expression, which must be
+a \hs{String}, to standard output).\footnote{The code generation for \hs{eExpr}
+  is actually user-configurable, in case the default is too simplistic. Unlike
+  \hint{} we never force calls to the \hs{show} method, since normal function
+  calls are easier to specify and override than type class methods.} If the
 subprocess exits successfully, its standard output is returned (wrapped in
@@
 invoke that command directly. If not, we use the \texttt{nix-shell} command of
-the Nix package manager to invoke \texttt{runhaskell} in a sandbox~\cite{TODO}.
-We use the Nix language to request a sandbox containing GHC and all of the
-elements of \hs{ePkgs}. We also allow the default Nix packages to be overridden
-via an environment variable, which allows the user to define their own packages,
-select particular versions, or perform any other arbitrary setup (since the Nix
-language is Turing-complete).
+the Nix package manager to invoke \texttt{runhaskell} in a
+sandbox~\cite{dolstra2004nix}. We use the Nix language to request a sandbox
+containing GHC and all of the elements of \hs{ePkgs}. We also allow the default
+Nix packages to be overridden via an environment variable, which allows the user
+to define their own packages, select particular versions, or perform any other
+arbitrary setup (since the Nix language is Turing-complete).
 
@@
 inspects the sandbox components). To avoid too much overhead, it is advisable to
-invoke \hs{eval} as few times as possible (preferably only once). If we need to
+invoke \hs{eval} as few times as possible (ideally only once). If we need to
 dynamically evaluate multiple parts of a program, it might be preferable to move
@@
   \begin{haskell}
--- An Expr representing a list of the given Exprs
-asList :: [Expr] -> Expr
-asList = foldr (\x -> (("(:)" $$ x) $$)) "[]"
+-- Apply x to n undefined arguments; result is undefined with x's output type
+addCalls :: Int -> Expr -> Expr
+addCalls 0 x = x
+addCalls n x = addCalls (n-1) (x $$ "undefined")
+
+-- Identity function, monomorphised to the output type of "f" (of arity a)
+converter :: Expr
+converter = ("flip" $$ "asTypeOf") $$ addCalls a "f"
 
@@
 
--- Apply x to n undefined arguments; result is undefined with x's output type
-addCalls :: Int -> Expr -> Expr
-addCalls 0 x = x
-addCalls n x = addCalls (n-1) (x $$ "undefined")
-
--- Coerces values to have the output type of a known function "f" (of arity a)
-converter :: Expr
-converter = ("flip" $$ "asTypeOf") $$ addCalls a "f"
-
--- Coerces a value to the output type of "f", serialises it then hashes it
+-- Forces a value's type, serialises it then hashes it
 hasher :: Expr
@@
   \end{haskell}
-  \caption{Example usage of \nixeval functions, taken from \textsc{ML4HS}, which
-    invokes \qspec on arbitrary user-provided functions.}
+  \caption{Example usage of \nixeval{} functions, taken from \textsc{ML4HS}.
+    These are used to dynamically generate \qspec{} signatures of user-provided
+    functions.}
   \label{fig:nix-eval}
@@
 
+\begin{sloppypar}
   Using \hs{String} functions to manipulate such complicated values soon becomes
-unwieldy, so a suite of combinators are provided for composing \hs{Expr} values
-in useful ways; examples are shown in Figure~\ref{fig:nix-eval}. Besides the
-\hs{withFoo} functions shown before, the \hs{qualified} function can reduce the
-ambiguity of names while avoiding repetition of module names: a call like
-\hs{qualified "Data.Text" "toUpper"} returns an \hs{Expr} of the name
-\hs{"Data.Text.toUpper"} with the \hs{"Data.Text"} \hs{Mod}ule in its
-\hs{eMods} list. \hs{foo $$ bar} denotes function application (analogous to
-Haskell's standard \hs{$}%$ operator), by juxtaposing the \hs{eExpr} of \hs{foo}
-with that of \hs{bar} (wrapping both in parentheses to ensure precedence); the
-contexts are appended. Finally, the \hs{asString :: Show a => a -> Expr}
-function calls \hs{show} on the given value and wraps the resulting \hs{String}
-into an \hs{Expr}; this is especially useful for representing literal
-\hs{String}s, since their \hs{show} instance will wrap them in quotes and
-perform any required escaping.
+  unwieldy, so a suite of combinators (such as the \hs{withFoo} functions above)
+  are provided for composing \hs{Expr} values in useful ways; examples are shown
+  in Figure~\ref{fig:nix-eval}. The \hs{qualified} function is particularly
+  useful, for qualifying names with their module (reducing ambiguity and making
+  composition more robust) and appending to \hs{eMods}, i.e.
+  \hs{qualified "Data.Text" "toUpper"} is equivalent to
+  \hs{withMods ["Data.Text"] "Data.Text.toUpper"}. Function calls are also very
+  common, which can be written as \hs{foo $$ bar} (analogous to Haskell's
+  standard \hs{$} %$
+  operator); this wraps both \hs{Expr}essions in parentheses
+  and juxtaposes them, appending both contexts.
+\end{sloppypar}
 
 Despite being more structured than raw \hs{String}s, the the above API still
-suffers from a form of "stringly-typed programming"; notably, that every encoded
-expression has the same type: \hs{Expr}. We can avoid this through the use of
-\emph{phantom types}: overly-restrictive type annotations, which are unrelated
-to the encoding but must still match up in order to pass the type checker. In
-particular we can represent typed expressions like this:
+suffers from a form of ``stringly-typed programming''; notably, that every
+encoded expression has the same type: \hs{Expr}. We can avoid this through the
+use of \emph{phantom types}: overly-restrictive type annotations, which are
+unrelated to the encoding but must still match up in order to pass the type
+checker~\cite[\S~5.1]{leijen1999domain}. In particular we can represent typed
+expressions like this:
 
@@
 Notice that the parameter \hs{t} doesn't appear on the right hand side: the
-values of \hs{TypedExpr} are just those of \hs{Expr} (the phantom type is
-\emph{erased} after type-checking). The \hs{TE} constructor is hence polymorphic
-in \hs{t}, so \hs{TE x :: TypedExpr T} is valid for \emph{any} choice of \hs{x}
-and \hs{T} (similar to how \hs{Nil :: List T} is valid for any choice of
-\hs{T}). In particular, this includes the type of the expression
-\emph{represented by} \hs{x}, for example \hs{TE "True" :: TypedExpr Bool} is a
-type-correct expression.
+values of \hs{TypedExpr} are just those of \hs{Expr} (the wrapper, along with
+the phantom type, is \emph{erased} after type-checking). The \hs{TE} constructor
+is hence polymorphic in \hs{t}, so \hs{TE x :: TypedExpr T} is valid for
+\emph{any} choice of \hs{x} and \hs{T} (similar to how \hs{Nil :: List T} is
+valid for any choice of \hs{T}). In particular, this includes the type of the
+expression \emph{represented by} \hs{x}, for example
+\hs{TE "True" :: TypedExpr Bool} is a type-correct expression.
 
@@
 
--- Only expressions of type String can be evaluated, since we use Prelude.putStr
+-- Only expressions of type String can be evaluated
 tEval :: TypedExpr String -> IO (Maybe String)
@@
 
-Haskell will accept any annotation, so care must be taken to ensure that they
-correspond correctly to the encoded expressions. If user-provided literals
-aren't required then the \hs{TE} constructor can be kept private to prevent
-unsound assertions or coercions. However, there is a more subtle problem with
-such \hs{TypedExpr}s: the whole point of \nixeval is to manipulate and evaluate
-code whose dependencies are unknown or unavailable at compile time. For example,
-we cannot assign a meaningful type at compile time to an expression which has
-been generated from user input. Even if we know the value of an expression at
-compile time, its type may come from a dependency that our program doesn't have
-access to. In these cases the \hs{TypedExpr} approach can still be useful for
-auxiliary code whose types are known to us; those unknown or unavailable parts
-can either be unwrapped into raw \hs{Expr}s, or dummy types can be used instead.
+Haskell will not check that these annotation match the encoded expressions, so
+care must be taken to ensure that they correspond correctly. Unannotated values
+will remain polymorphic and unify with anything, which reduces the verbosity and
+correctness burden (e.g. for intermediate values) whilst still guaranteeing
+consistency (unification will fail if multiple incompatible types are used). If
+user-provided literals aren't required then the \hs{TE} constructor can be kept
+private to prevent unsound assertions or coercions.
+
+However, there is a more subtle problem with such \hs{TypedExpr}s: the whole
+point of \nixeval{} is to manipulate and evaluate code whose dependencies are
+unknown or unavailable at compile time. For example, we cannot assign a
+meaningful type at compile time to an expression which has been generated from
+user input. Even if we know the value of an expression at compile time, its type
+may come from a dependency that our program doesn't have access to. In these
+cases the \hs{TypedExpr} approach can still be useful for auxiliary code whose
+types are known to us; those unknown or unavailable parts can either be left
+polymorphic, unwrapped into raw \hs{Expr}s, or dummy types can be used instead.
 For example, if we are manipulating code involving the \hs{Data.Text.Text} type,
-but we don't have or want the \textt{text} package in our dependencies, we can
-declare an empty type like \hs{data TEXT} (note that this declaration has no
-right-hand side) and use this in our annotations in place of
-\hs{Data.Text.Text}. The compiler will type-check our code as desired, ensuring
-functions are called on appropriate arguments, \hs{tEval} is given on a
-\hs{String}, etc.
+but we don't have or want the \texttt{text} package in our dependencies, we can
+declare an empty type like:
 
-In the course of writing MLSpec we experimented with both raw \hs{Expr} and
+\begin{haskell}
+data TEXT  -- No constructors given, making this type empty
+\end{haskell}
+
+We can use this in our annotations in place of \hs{Data.Text.Text}. The compiler
+will type-check our code as desired, ensuring functions are called on
+appropriate arguments, \hs{tEval} is called with a \hs{String}, etc.
+
+In the course of writing \mlspec{} we experimented with both raw \hs{Expr} and
 \hs{TypedExpr} values. Whilst we found the latter easier for constructing large
@@
 functionality as possible into a standalone Cabal package
-(\texttt{mlspec-helper}) and have MLSpec call into that code from simple
-\hs{Expr} values.
+(\texttt{mlspec-helper}), add that package to the Nix package set and have
+\mlspec{} call it using simple \hs{Expr} values.

### 2019-05-16 PhDThesis/background_haskell.tex
Headingy things

@@
-\subsection{Haskell}
+\section{Haskell}
 \label{sec:haskell}
@@
     Peano encoding of the natural numbers and \hs{List t} are polymorphic lists
-    with elements of type \hs{t}. Functions are called using whitespace, e.g.
-    \hs{f x}, and may be defined by pattern-matching (case analysis) on their
-    arguments. \hs{A :: B} is an optional type annotation, stating that value
-    \hs{A} has type \hs{B}. \hs{A -> B} is the type of functions from \hs{A} to
-    \hs{B}.}
+    with elements of type \hs{t}. Juxtaposition denotes a function call, e.g.
+    \hs{f x}, and functions may be defined by pattern-matching (case analysis)
+    on their arguments. \hs{A :: B} is an optional type annotation, stating that
+    value \hs{A} has type \hs{B}. \hs{A -> B} is the type of functions from
+    \hs{A} to \hs{B}.}
   \label{fig:haskellexample}

### 2019-05-16 PhDThesis/bucketing.tex
Fix metadata
Headingy things

@@
-\section{The Signature Selection Problem}
+\chapter{The Signature Selection Problem}
 
@@
 
-\subsection{Background}
+\section{Background}
 \label{sec:background}
@@
 
-\subsection{Clustering}
+\section{Clustering}
 \label{sec:clustering}
@@
 % TODO{2019-04-19}
-
-\section{Contributions}
-\label{sec:contributions}
 
@@
 
-\subsection{Experiments}
+\section{Experiments}
 
@@
   uses the Nix package manager to obtain all of the required packages and make
-  them available to GHC.
+  them available to GHC, and is described further in
+  Appendix~\ref{sec:nix-eval}.
 
@@
 \begin{figure}
-  \begin{haskell}
+  \begin{common-lisp}
 (defun fact (n)
@@
   (helper-fib n 0 1))
-  \end{haskell}
+  \end{common-lisp}
   \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
@@
 
-\begin{equation}\label{fig:coresyntax}
+\begin{equation}\label{eq:coresyntax}
   \begin{split}

### 2019-05-16 PhDThesis/background_machinelearning.tex
Rearrangement

@@
 
-\subsection{Distributed Representations}
+\subsection{Feature Extraction}
+
+One major difficulty when applying statistical machine learning algorithms to
+\emph{languages}, such as Haskell, is the appearance of recursive
+structures. This can lead to nested expressions of arbitrary depth, which are
+difficult to compare in numerical ways. One common approach to this problem is
+to represent such structures as \emph{sequences}. \emph{Recurrent neural
+  networks} (RNNs) are a popular choice for processing sequences, especially
+when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
+preserving information across long sequences \cite{hochreiter1997long}. Such
+systems have been used, for example, to parse and execute computer programs
+\cite{zaremba2014learning}. However, learning to parse sequences seems
+inefficient considering that we already have correctly-parsed ASTs.
+
+Whilst neural networks have been applied directly to recursive structures
+\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
+popular approach is to use \emph{kernel methods}
+\cite{bakir2007predicting}. These are promising as a more principled alternative
+to our current hand-crafted translation of ASTs to vectors.
+
+\subsubsection{Distributed Representations}
 

### 2019-05-16 PhDThesis/bucketing.tex
Rearrangement

@@
 \label{sec:evaluation}
-\iffalse
-
+
+% TODO: Move recurrent clustering to its own chapter, with this as its
+% evaluation?
 We have applied our recurrent clustering algorithm to several scenarios, with
@@
 
-\fi
-
-\section{Related Work}
-\label{sec:related}
-
-\subsection{Theory Exploration}
-
-We briefly described theory exploration in \S \ref{sec:theoryexploration}, as
-the task of discovering \emph{new} theorems in a software or proof library,
-rather than proving/disproving user-provided statements. This general idea dates
-back at least to the 1970s, most famously to work by Lenat
-\cite{lenat1977automated}, though the theory exploration tools used in
-functional programming trace their roots back to the interactive
-\textsc{Theorema} \cite{buchberger2000theory} system of Buchberger.
-
-As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has
-been applied to Isabelle \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
-\cite{johansson2009isacosy} \cite{Hipster}. Since \qspec{} is used for
-conjecture generation in many of these systems, they may also benefit from our
-machine learning approach.
-
-\subsection{Relevance Filtering}
-\label{sec:relevance}
-
-The idea of \emph{relevance filtering} (or, \emph{premise selection}) is to
-reduce the search space of a theorem proving task by removing redundant or
-irrelevant information (axioms, definitions, lemmas, etc.) from the input. Our
-restriction of theory exploration to intelligently-selected clusters of symbols,
-rather than whole libraries at a time, is essentially the same idea applied in a
-novel setting.
-
-The technique is used in Isabelle's Sledgehammer tool, during its translation of
-Isabelle/HOL theories to statements in first order logic: rather than
-translating the entire theory, only a sub-set of relevant clauses are
-included. This reduces the size of the problem and speeds up the proof search,
-but it creates the new problem of determining when a clause is relevant: how do
-we know what will be required, before we have the proof?
-
-The original approach uses the proportion of symbols which a clause has in
-common with the problem statement \cite{meng2009lightweight} (plus various
-heuristics), although various alternatives have been proposed in the literature
-\cite{kuhlwein2013mash} \cite{kuhlwein2012overview} \cite{alama2014premise}.
-
 \subsection{Recurrent Clustering}
@@
 restrict the scope of theory exploration.
-
-\subsection{Feature Extraction}
-
-One major difficulty when applying statistical machine learning algorithms to
-\emph{languages}, such as Haskell, is the appearance of recursive
-structures. This can lead to nested expressions of arbitrary depth, which are
-difficult to compare in numerical ways. One common approach to this problem is
-to represent such structures as \emph{sequences}. \emph{Recurrent neural
-  networks} (RNNs) are a popular choice for processing sequences, especially
-when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
-preserving information across long sequences \cite{hochreiter1997long}. Such
-systems have been used, for example, to parse and execute computer programs
-\cite{zaremba2014learning}. However, learning to parse sequences seems
-inefficient considering that we already have correctly-parsed ASTs.
-
-Whilst neural networks have been applied directly to recursive structures
-\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
-popular approach is to use \emph{kernel methods}
-\cite{bakir2007predicting}. These are promising as a more principled alternative
-to our current hand-crafted translation of ASTs to vectors.
 

### 2019-05-16 PhDThesis/litreview_interestingness.tex
More content
Bunch more changes to thesis
Rearrangement

@@
-Whilst there are many reasonably objective benchmarks for mathematical tasks
-such as automated theorem proving, the precision/recall analysis shown
-in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
-the only quantitative comparison of these recent MTE tools we have found in the
-literature. Our work is essentially an extension of this approach, to a larger
-and more diverse set of examples. The suitability of the TIP theorem proving
-benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
-since its developers are also those of the IsaCoSy and QuickSpec tools we have
-tested. This goes some way to ensuring that our demonstration is a faithful
-representation of the task these tools were intended to solve; our independent
-repurposing of this problem set, in a way it was not designed for, reduces the
-risk that the benchmark is tailor-made for these tools (or that the tools
-over-fit to these particular problems).
+\subsection{Interestingness}
+
+A major challenge when generating conjectures is choosing how to narrow down the
+resulting set to only those deemed ``interesting'', since this is an imprecise
+term with many different interpretations. For example, all existing approaches
+agree that simple tautologies are ``uninteresting'', but differ when it comes to
+more complex statements.  Colton \etal{} surveyed tools for theory formation
+and exploration, and their associated notions of ``interestingness'' for
+concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
+which are applied to conjectured properties as follows:
+
+{\bf Empirical plausibility} checks whether a property holds across some
+specific examples. This is especially useful for avoiding falsehoods, without
+resorting to a deductive proof search.
+
+{\bf Novelty} depends on whether the property, or one isomorphic or more
+general, has already been seen.
+
+{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
+example if it is an instance of a tautology.
+
+{\bf Applicability} depends on the number of models in which a
+property holds. High applicability makes a property more interesting, but so
+does zero applicability (i.e. non-existence).
+
+{\bf Comprehensibility} depends on the syntactic complexity of the property's
+statement. Simpler statements are considered more interesting (which favours
+tools adopting a small-to-large search order).
+
+{\bf Utility} is the relevance or usefulness of a property to the user's
+particular task. For example, if we want to find properties which are useful for
+optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
+whether the property justifies some rewrite rule, the difference in resource
+usage of the expressions involved, and how common those expressions are in real
+usage.
+
+% FIXME: This is original work, so should be in a different section
+\begin{table}
+  \centering
+  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
+    \hline
+    \multirow{2}{*}{\textbf{Program}}                      &
+    \multirow{2}{*}{\textbf{Conjecture Types}}             &
+    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
+    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
+    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
+    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
+    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
+    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
+    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
+    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
+    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
+    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
+    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
+  \end{tabular}
+  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
+    to include four more recent tools. The interestingness measures are
+    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
+    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
+  \label{table:colton}
+\end{table}
+
+% JUSTIFICATIONS
+%
+% IsaCoSy uses counter-example checking; this ensures empirical plausibility
+% IsaCoSy uses constraints to avoid special-cases; this ensures novelty
+% IsaCoSy uses utility, since there are some hard-coded patterns which are
+% looked for
+%
+% Speculate uses a form of unification to ensure novelty, making sure one
+% equation is not a special-case of another
+% Speculate uses LeanCheck to enumerate values, looking for counterexamples
+% Speculate uses testing, to ensure Empirical Plausibility
+%
+% QuickSpec uses QuickCheck to ensure empirical plausibility
+% QuickSpec uses a congruence closure algorithm to ensure novelty
+%
+% IsaScheme uses utility, since it determines the "quality" of a definition
+% based on how many theorems it appears in?
+% IsaScheme uses novelty, using an equational rewrite system (Knuth-Bendix
+% completion) to remove redundancies
+% IsaScheme uses utility, since it focuses on the terms and definitions of a
+% user's theory. However, don't all of them?
+% IsaScheme uses counterexample checking for empirical plausibility.
+
+% ONLY APPLIES TO PRECONDITIONS
+% APPLICABILITY: ensure a conjecture holds in many models (or exactly zero, in
+% the case of Bagai et al). Does the testing-based approach of QuickSpec and
+% Speculate ensure that there's a model? After all, these are concrete values
+% rather than e.g. implications of some abstract specification.
+
+% FIXME: Give brief description and reference for each system in the text.
+% We could include the more lengthy descriptions of IsaCoSy and QuickSpec here?
+
+We summarise the criteria in Table~\ref{table:colton} and state which were used
+in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
+based on this survey (the latter is Colton's own tool). We extend this analysis
+to some more recent tools in the remaining rows (based on our understanding of
+their function): \qspec{}~\cite{QuickSpec},
+Speculate~\cite{braquehais2017speculate},
+IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
+and IsaScheme~\cite{MontanoRivas2011}.
+
+These recent tools all use testing to check for counterexamples, which ensures
+results are empirically plausible and applicable (conditions are satisfiable,
+types are inhabited, etc.). They also use a small-to-large search order,
+ensuring more easily-comprehensibile conjectures are explored first.
+
+IsaCoSy ensures novelty using a constraint system to prevent special-cases being
+generated, whilst the others use post-hoc filters based on unification and term
+rewriting to achieve the same result. IsaCoSy is able to look for commonly
+desirable patterns, such as commutativity and associativity, which increases the
+utility of its output, whilst IsaScheme is completely based around such
+pattern-instantiation.
+
+This diversity of approaches makes it difficult to compare the existing
+evaluations of these tools directly. In particular, evaluation methods created
+for one tool might not make sense for another, due to assumptions made about
+the algorithms' operation. For example, the novelty filters used by \qspec{}
+and Speculate are applied to the whole output set, guaranteeing that no
+special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
+since its constraint solver allows special cases if they're found \emph{before}
+the general case (it does not go back and discard previous results).
+
+In addition, attempting to \emph{measure} these qualities directly is difficult,
+and having too many measures complicates comparisons. System developers have
+employed a more practical alternative in their evaluations, which is to perform
+\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
+choosing a set of definitions and a set of properties (the ground truth) which
+represent the ``ideal'' result of conjecture generation for those definitions.
+To analyse the quality of a tool's conjectures, we run it on these chosen
+definitions and compare its output to the ground truth:
+
+\begin{itemize}
+\item \emph{Precision} is the proportion of a tool's output which appears in
+  the ground truth (the ratio of true positives to all positives). This
+  penalises overly-liberal tools which output a large number of properties in
+  the hope that some turn out to be ``good''.
+\item \emph{Recall} is the proportion of the ground truth which appears in the
+  tool's output (the ratio of true positives to actual positives). This
+  penalises overly-conservative tools which generate very few properties as a
+  way to avoid ``bad'' ones.
+\end{itemize}
+
+To score 100\% on precision and recall, all of the properties which appear in
+the ground truth must have been conjectured, and nothing else. This gives us a
+simple method to evaluate and compare tools without requiring a general solution
+to the question of what is ``interesting''; although we must still decide what
+to put in the ground truth set, and what to leave out, for each measured set of
+definitions.
+
+The precision/recall analysis shown in~\cite{claessen2013automating} is the only
+quantitative comparison of recent MTE tools we have found in the literature.
+Our Theory Exploration Benchmark, described in section~\ref{sec:benchmark}, is
+essentially an extension of this approach, to a larger and more diverse set of
+examples.
+
+% TODO Include Colton/Bundy et al
+
+% TODO Include something about intrinsic motivation, compression progress,
+% variance among experts (decision tree stuff as well as more established
+% splitting functions stuff)
+
+% TODO Make sure we mention the recent reinforcement learning task that includes
+% a constantly-changing "TV" in the 3D world, shown in some TwoMinutePapers
+% videos

### 2019-05-22 PhDThesis/background_quickcheck.tex
Headingy things
Property checking rather than QuickCheck

@@
-\subsection{QuickCheck}
-\label{sec:quickcheck}
+\section{Property Checking}
+\label{sec:propertychecking}
+
+% TODO: Automated testing
+% TODO: Agile
+% TODO: Unit testing
+% TODO:
 
@@
 outputs. However, these tests will \emph{not} expose a serious problem with the
-implementation: it diverges on half of its possible inputs!
+implementation: it diverges on half of its possible inputs.
 
@@
   variables}; a source of values to instantiate these free variables; and a
-stopping criterion. Here is how we might restate our unit tests as properties:
+stopping criterion.
+
+\subsection{\qcheck{}}
+\label{sec:quickcheck}
+
+\qcheck{}~\cite{quickcheck} is the most widely used property checking library
+for
+Haskell~\footnote{According to downloads counted at
+  \href{http://hackage.haskell.org/packages/browse}, and packages depending on
+  it at \href{https://packdeps.haskellers.com/reverse/QuickCheck}, accessed on
+  2019-05-21.}. Here is how we might restate our unit tests as \qcheck{}
+properties:
 
@@
 
-Property checking is certainly an improvement over unit testing, but the problem
-of tests being biased towards expected cases remains, since we are manually
-specifying the properties to be checked.
-
-We can reduce this bias further through the use of \emph{theory exploration}
-tools, such as \qspec{} and \hspec{}. These programs \emph{discover} properties
-of a ``theory'' (e.g. a library), through a combination of brute-force
-enumeration, random testing and (in the case of \hspec{}) automated theorem
-proving.
+Whilst property checking generalises and improves upon unit testing, the problem
+of tests being biased towards expected cases remains, since the properties to be
+checked are still manually specified. Property checking can be complemented by
+\emph{theory exploration} (TE), which avoids this bias by \emph{discovering}
+such properties; through a combination of brute-force enumeration, random
+testing and (in the case of \hspec{} and \hipster{}) automated theorem proving.
+These TE systems make use of property checkers as part of their search process,
+to try and disprove generated conjectures before output (to the user, or a
+subsequent tool).

### 2019-05-22 PhDThesis/background_theoryexploration.tex
Change listing type in theoryexploration
Headingy things
Better TE intro

@@
-\subsection{Theory Exploration}
+\section{Theory Exploration}
 \label{sec:theoryexploration}
@@
 In this work we consider the application of \emph{(automated) theory
-  exploration} to generate conjectures about code. Existing tools are able to
-\emph{prove} those conjectures, and hence output \emph{novel} theorems without
-guidance from the user. The method of conjecture generation is a key
-characteristic of any theory exploration system, although all existing
-implementations rely on brute force enumeration to some degree.
+  exploration} (TE) to generate conjectures about code\footnote{In principle TE
+  can be used with any formal system, but appropriate methods can differ
+  according to the underlying logic, etc. that is in use. We focus on Haskell
+  code to anchor these variables, and to ensure our results have immediate
+  utility.}. Existing tools are able to \emph{prove} those conjectures, and
+hence output \emph{novel} theorems without guidance from the user. The method of
+conjecture generation is a key characteristic of any theory exploration system,
+although all existing implementations rely on brute force enumeration to some
+degree.
 
@@
 
+% TODO
 %%%%% THE BELOW SEEMS TO BE INTERESTINGNESS?
@@
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 % Inductive Nat : Set := Z : Nat
@@
 %                                   end.
-% \end{lstlisting}
+% \end{coqblock}
 
@@
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 % (* Solve equalities by beta-normalising both sides *)
@@
 %   rewrite <- (IHn (S m)).
-% \end{lstlisting}
+% \end{coqblock}
 
@@
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 % IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-% \end{lstlisting}
+% \end{coqblock}
 
@@
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 % IHn : forall x, plus (S n') x = plus_tail (S n') x
-% \end{lstlisting}
+% \end{coqblock}
 
@@
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 % Lemma gen n m : plus (S n) m = plus n (S m).
@@
 % Defined.
-% \end{lstlisting}
+% \end{coqblock}
 
-% \begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+% \begin{coqblock}
 %   rewrite (gen n m).
@@
 % Defined.
-% \end{lstlisting}
+% \end{coqblock}

### 2019-06-02 PhDThesis/appendix_nixeval.tex
Thesis updates

@@
   \eval{}~\cite[\S~1.6]{edwards1962lisp}) is limited. The most popular mechanism
-  is the \hint{} package\footnote{https://hackage.haskell.org/package/hint},
-  which is built on top of GHC's API and provides functions like
+  is the \hint{} package\footnote{\hPackage{hint}}, which is built on top of
+  GHC's API and provides functions like
   \hs{eval :: MonadInterpreter m => String -> m String}. The
@@
   To solve this problem we created the \nixeval{}
-  package\footnote{https://hackage.haskell.org/package/nix-eval}. This provides
-  a function \hs{eval :: Expr -> IO (Maybe String)}, which produces \hs{String}s
-  for the same reasons as \hint{}, but differs in its handling of expressions
-  and context. Rather than operating on a raw \hs{String}, \nixeval{} defines
-  the \hs{Expr} type, containing several fields to keep track of the context in
+  package\footnote{\hPackage{nix-eval}}. This provides a function
+  \hs{eval :: Expr -> IO (Maybe String)}, which produces \hs{String}s for the
+  same reasons as \hint{}, but differs in its handling of expressions and
+  context. Rather than operating on a raw \hs{String}, \nixeval{} defines the
+  \hs{Expr} type, containing several fields to keep track of the context in
   which the expression is to be evaluated:
@@
   \end{haskell}
-  \caption{Example usage of \nixeval{} functions, taken from \textsc{ML4HS}.
-    These are used to dynamically generate \qspec{} signatures of user-provided
-    functions.}
+  \caption{Example usage of \nixeval{} functions, taken from \mlforhs{}.
+    These are used to dynamically generate \quickspec{} signatures of
+    user-provided functions.}
   \label{fig:nix-eval}

### 2019-06-02 PhDThesis/background_conjecturegeneration.tex
Headingy things
Thesis updates

@@
 
-\subsection{Conjecture Generation}
+\section{Conjecture Generation}
 
@@
     $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
-  \label{figure:list_theory}
+  \label{fig:list_theory}
 \end{figure}
@@
 example higher-order logic), along with a finite set of \emph{definitions} (like
-those in figure~\ref{figure:list_theory}) which assign structure or meaning to
+those in figure~\ref{fig:list_theory}) which assign structure or meaning to
 particular terms. Theoretically, this is all of the information we need to
@@
 As an example, we use the theory of lists shown in
-figure~\ref{figure:list_theory}. These definitions are so widely used in
+figure~\ref{fig:list_theory}. These definitions are so widely used in
 software that they appear in the standard libraries of many programming

### 2019-06-02 PhDThesis/background_haskell.tex
Slight fiddling
Thesis updates

@@
   \hs{E}, but \hs{E} is private (not exported). The \hs{at} function is
-  exported, but only passes its arguments to \hs{E} iff they are not empty.\footnote{\hs{Maybe t} is a
-    safer alternative to the \textsc{Null} construct of other
-    languages~\cite{hoare2009null}. It is defined as
+  exported, but only passes its arguments to \hs{E} iff they are not
+  empty.\footnote{\hs{Maybe t} is a safer alternative to the \textsc{Null}
+    construct of other languages~\cite{hoare2009null}. It is defined as
     \hs{data Maybe t = Nothing | Just t} and can be understood as an optional
@@
 \item The large amount of Haskell code which is freely available online, e.g. in
-  repositories like \href{http://hackage.haskell.org}{Hackage}, with which we
+  repositories like \hackage{}, with which we
   can experiment.
 
-\item The existence of theory exploration systems such as \hspec{}, \qspec{} and
-  \textsc{Speculate}.
+\item The existence of theory exploration systems such as \hipspec{},
+  \quickspec{} and \speculate{}.
 
-\item Related tooling we can re-use such as counterexample finders (\qcheck{},
-  \textsc{SmallCheck}, \textsc{SmartCheck}, \textsc{LeanCheck},
-  \textsc{Hedgehog}, etc.), theorem provers
-  (e.g. \textsc{Hip}~\cite{rosen2012proving}), and other testing and
-  term-generating systems like \textsc{MuCheck}~\cite{le2014mucheck},
-  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
-  \textsc{Djinn}~\cite{augustsson2005djinn}.
+\item Related tooling we can re-use such as counterexample finders (\quickcheck{},
+  \smallcheck{}, \smartcheck{}, \leancheck{},
+  \hedgehog{}, etc.), theorem provers
+  (e.g. \hip{}~\cite{rosen2012proving}), and other testing and
+  term-generating systems like \mucheck{}~\cite{le2014mucheck},
+  \magichaskeller{}~\cite{katayama2011magichaskeller} and
+  \djinn{}~\cite{augustsson2005djinn}.
 
@@
 
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
+    repetition and grouping, respectively.}
   \label{fig:coresyntax}
@@
 fact that the state-of-the-art implementation for Isabelle/HOL, the
-\textsc{Hipster}~\cite{Hipster} system, is actually implemented by translating
-to Haskell and invoking \hspec{}~\cite{claessen2013automating}.
+\hipster{}~\cite{Hipster} system, is actually implemented by translating
+to Haskell and invoking \hipspec{}~\cite{claessen2013automating}.
+
+\subsection{GHC Core}
+\label{sec:ghccore}
 

### 2019-06-02 PhDThesis/background_quickcheck.tex
Thesis updates

@@
 the following Haskell function, along with some unit
-tests\footnote{Backticks allow functions to be written as infix operators, so
-  \hs{x `div` y} is equivalent to \hs{div x y}.}:
+tests\footnote{Haskell functions can be prefix or infix: prefix functions have
+  alphanumeric names, like \hs{div} and \hs{factorial}, and appear in function
+  calls before their arguments, e.g. \hs{div 10 5}; infix functions have
+  non-alphanumeric names, like \hs{+} and \hs{==}, and appear in function calls
+  between their first two arguments, e.g. \hs{4 + 3}. Prefix functions can
+  appear infix using backticks, like \hs{10 `div` 5}, and infix functions can
+  appear prefix using parentheses, like \hs{(+) 4 3}.}:
 
+% BEGIN unit
 \begin{haskell}
 factorial 0 = 1
-factorial n = n * factorial (n-1)
+factorial n = n * factorial (n - 1)
 
@@
 \end{haskell}
+% END unit
 
@@
 
-All of Haskell's built-in numeric types allow negative numbers, which this
+Haskell's built-in numeric types allow negative numbers, which this
 implementation doesn't take into account. Whilst this is a rather trivial
@@
 
-If we do manage to spot this error, we might capture it in a \emph{regression
-  test} and update the definition of \hs{factorial} to handle negative numbers,
-e.g. by taking their absolute value:
+One way to avoid this problem, if we do manage to spot it, would make results
+symmetric around zero, which can be checked by a \emph{regression test} such as:
 
+% BEGIN neg
 \begin{haskell}
-factorial 0 = 1
-factorial n = let nPos = abs n
-               in nPos * factorial (nPos - 1)
-
 fact_neg = factorial 1 == factorial (-1)
 \end{haskell}
+% END neg
+
+The implementation can be modified by composing if with the \hs{abs}olute value
+function:
+
+% BEGIN abs
+\begin{haskell}
+factorial = original . abs
+  where original 0 = 1
+        original n = n * naive (n - 1)
+\end{haskell}
+% END abs
 
@@
 fractional values\footnote{Since we only use generic numeric operations, the
-  function will be polymorphic with a type of the form \hs{forall t. Num t => t
-    -> t}, where \hs{Num t} constrains the type variable \hs{t} to be
-  numeric. In fact, Haskell will infer extra constraints such as \hs{Eq t} since
-  we have used \hs{==} in the unit tests.}, which will also cause it to
-diverge. Clearly, by choosing what to test we are biasing the test suite towards
-those cases we've already taken into account, whilst neglecting the problems we
+  function will be polymorphic with a type of the form
+  \hs{forall t. Num t => t -> t}, where \hs{Num t} constrains the type variable
+  \hs{t} to be numeric.}, which will also cause it to diverge. This problem can
+be fixed by restricting \hs{factorial}'s type with an explicit annotation, e.g.
+to only whole numbers: \hs{factorial :: forall t. Integral t => t -> t}
+
+Clearly, by choosing which scenarios to check, we are biasing the test suite
+towards those cases we've already thought of, whilst neglecting the problems we
 did not expect.
-
-Haskell offers a partial solution to this problem in the form of \emph{property
-  checking}. Tools such as \qcheck{} separate tests into three components: a
-\emph{property} to check, which unlike a unit test may contain \emph{free
-  variables}; a source of values to instantiate these free variables; and a
+Haskell offers a partial solution to this problem in the form of \emph{property checking}.
+Tools such as \quickcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain
+\emph{free variables}; a source of values to instantiate these free variables; and a
 stopping criterion.
 
-\subsection{\qcheck{}}
+\subsection{\quickcheck{}}
 \label{sec:quickcheck}
 
-\qcheck{}~\cite{quickcheck} is the most widely used property checking library
-for
+\quickcheck{}~\cite{claessen2011quickcheck} is the most widely used property
+checking library for
 Haskell~\footnote{According to downloads counted at
-  \href{http://hackage.haskell.org/packages/browse}, and packages depending on
-  it at \href{https://packdeps.haskellers.com/reverse/QuickCheck}, accessed on
-  2019-05-21.}. Here is how we might restate our unit tests as \qcheck{}
+  \url{http://hackage.haskell.org/packages/browse}, and packages depending on
+  it at \url{https://packdeps.haskellers.com/reverse/QuickCheck}, accessed on
+  2019-05-21.}. Here is how we might restate our unit tests as \quickcheck{}
 properties:
 
+% BEGIN quickchecked
 \begin{haskell}
 fact_base        = factorial 0 == factorial 1
-fact_increases n = factorial n <= factorial (n+1)
-fact_div       n = factorial n == factorial (n+1) `div` (n+1)
+fact_increases n = factorial n <= factorial (n + 1)
+fact_div       n = factorial n == factorial (n + 1) `div` (n + 1)
 fact_neg       n = factorial n == factorial (-n)
 \end{haskell}
+% END quickchecked
 
@@
 property asserting $\forall n, factorial(n) \leq factorial(n+1)$. Notice that
-unit tests like \hs{fact_base} are valid properties; they just assert rather
-weak statements.
+unit tests like \hs{fact_base} are valid properties; they just happen to assert
+rather weak statements, since they contain no free variables.
 
-To check these properties, \qcheck{} treats closed terms (like \hs{fact_base})
+To check these properties, \quickcheck{} treats closed terms (like \hs{fact_base})
 just like unit tests: pass if they evaluate to \hs{True}, fail otherwise. For
@@
 function parameter; the results are then treated in the same way as closed
-terms. The default stopping criterion for \qcheck{} (for each test) is when a
+terms. The default stopping criterion for \quickcheck{} (for each test) is when a
 single generated test fails, or when 100 generated tests pass.
@@
 \begin{itemize}
-\item \qcheck{} cannot test polymorphic functions; they must be
-  \emph{monomorphised} first (instantiated to a particular concrete type). This
-  is a technical limitation, since \qcheck{} must know which type of values to
-  generate, but in our example it would bring the issue with fractional values
-  to our attention.
+\item We are forced to confront the generality of \hs{factorial}'s type, since
+  \quickcheck{} cannot test polymorphic functions directly; they must be
+  \emph{monomorphised} first (instantiated to a particular concrete type), to
+  decide which data generator to use (i.e. which instance of \quickcheck{}'s
+  \hs{Arbitrary} type class). This can be done either by specifying a generator
+  directly (avoiding the type class mechanism), or by adding explicit type
+  annotations, or by calling \quickcheck{}'s \hs{monomorphic}
+  function.\footnote{\hs{monomorphic} picks \hs{Integer} for all polymorphic
+    type variables. This na\"ive choice instantiates many common type classes
+    like \hs{Num}, \hs{Show} and \hs{Eq}; variables whose constraints are not
+    satisfied will fail to type check (e.g. \hs{Bounded}, since \hs{Integer} is
+    a bignum).} This will bring the problem of fractional values. Instead of
+  monomorphising the properties, we might choose to monomorphise \hs{factorial}
 
-\item The generators used by \qcheck{} depend only on the \emph{type} of value
-  they are generating: since \hs{Int} includes positive and negative values, the
-  \hs{Int} generator will output both. This will expose the problem with
+\item By default, \quickcheck{} picks a generator based on the \emph{type} of
+  value to be generated: since \hs{Int} includes positive and negative values,
+  the \hs{Int} generator will output both. This will expose the problem with
   negative numbers, which we weren't expecting.
@@
 such properties; through a combination of brute-force enumeration, random
-testing and (in the case of \hspec{} and \hipster{}) automated theorem proving.
-These TE systems make use of property checkers as part of their search process,
-to try and disprove generated conjectures before output (to the user, or a
-subsequent tool).
+testing and (in the case of \hipspec{} and \hipster{}) automated theorem
+proving. Property checkers like \quickcheck{} are an important component of TE
+systems, since their data generators can the search process, and they can
+prevent ``obvious'' falsehoods from being output by checking them first.

### 2019-06-02 PhDThesis/background_theoryexploration.tex
Thesis updates

--- 2019-04-29 PhDThesis/background_theoryexploration.tex
@@
-\subsection{Theory Exploration}
+\section{Theory Exploration}
 \label{sec:theoryexploration}
 
-% TODO Work this in
+(Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
+of definitions in some formal system (for example a programming language) and
+automatically generating a set of formal statements (properties) involving
+those definitions. These may be conjectures or theorems (proven either by
+sending conjectures to an automated theorem prover, or by having the generating
+procedure proceed in logically sound steps). These statements must also be
+``interesting'' in some way, to rule out unhelpful trivialities such as
+iterating a pattern over and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x},
+\dots).
+
 \iffalse
-Theory exploration is the task of taking a \emph{signature} of definitions in
-some formal system (for example a programming language) and automatically
-generating a set of formal statements (properties) involving those
-definitions. These may be conjectures or theorems (proven either by sending
-conjectures to an automated theorem prover, or by having the generating
-procedure proceed in logically sound steps); in either case, these statements
-must also be ``interesting'' in some way. It is beyond the scope of this paper
+It is beyond the scope of this paper
 to define what makes a mathematical statement ``interesting'', but this problem
@@
 
-In this work we consider the application of \emph{(automated) theory
-  exploration} to generate conjectures about code. Existing tools are able to
-\emph{prove} those conjectures, and hence output \emph{novel} theorems without
-guidance from the user. The method of conjecture generation is a key
-characteristic of any theory exploration system, although all existing
-implementations rely on brute force enumeration to some degree.
+The choice of appropriate methods can differ according to the underlying logic,
+etc. that is in use, so in this work we ground ourselves by applying TE to
+discovering properties of pure functional software libraries, mostly in the
+Haskell programming language. This also gives our results immediate utility to
+software engineering. We ignore the issue of \emph{proving} such conjectures,
+deferring that task to existing tools (such as \hipspec{}).
 
-We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about Haskell
-code (these may be fed into another tool, such as \hspec{}, for proving). We
-have used version 1 of \qspec{} in our experiments, due to its availability,
-stability and tooling integration; hence references to \qspec{} are to version 1
-unless stated otherwise. At the time of writing there is a \qspec{} version 2
-available, which has a much improved generation procedure which can be
-significantly faster than its predecessor. Whilst \qspec{} 2 has advanced the
-state of the art in theory exploration, our preliminary experience has found
-that it still suffers the scaling issues we identify in this work; indeed, we
-even encountered memory exhaustion from some of the examples included with the
-\qspec{} 2 source code! Since \qspec{} 2 has less integration with the tooling
-we have used, we leave a more thorough analysis of it (and related systems such
-as \speculate{}) for future work, pending the necessary infrastructure changes
-that would require.
+The method of conjecture generation is a key characteristic of any theory
+exploration system, although all existing implementations rely on brute force
+enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
+conjectures equations about Haskell code. We have used version 1 of
+\quickspec{} in our experiments, due to its availability, stability and tooling
+integration; hence references to \quickspec{} are to version 1 unless stated
+otherwise. At the time of writing there is a \quickspec{} version 2 available,
+which has a much improved generation procedure which can be significantly faster
+than its predecessor. Whilst \quickspec{} 2 has advanced the state of the art in
+theory exploration, our preliminary experience has found that it still suffers
+the scaling issues we identify in this work\footnote{We encountered memory
+  exhaustion when trying to run some of the examples included with the
+  \quickspec{} 2 source code; let alone auto-generated signatures.} Since
+\quickspec{} 2 has less integration with the tooling we have used, we leave a
+more thorough analysis of it (and related systems such as \speculate{}) for
+future work, pending the necessary infrastructure changes that would require.
 
@@
 
-\qspec{} (version 1) is written in the Haskell programming language, and works
-with definitions which are also written in Haskell. A thorough description of
-Haskell, including its suitability for theory exploration, can be found
-in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
-are shown in Figure~\ref{fig:haskellteexample}. The following procedure is used to
-generate conjectures, which are both plausible (due to testing on many examples)
-and potentially interesting (due to being mutually irreducible):
+\quickspec{} (version 1) is written in the Haskell programming language, and
+works with definitions which are also written in Haskell. A thorough description
+of Haskell, including its suitability for theory exploration, can be found
+in \S~\ref{sec:haskell}. For illustrative purposes, some simple Haskell
+definitions are shown in Figure~\ref{fig:haskellteexample}. \quickspec{} uses
+the following procedure to generate conjectures, which are both plausible (due
+to testing on many examples) and potentially interesting (due to being mutually irreducible):
 
 \begin{enumerate}
-\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
-  generates a list $terms$ containing the constants (including functions) from
-  $\Sigma$, the variables from $V$ and type-correct function applications
-  $f(x)$, where $f$ and $x$ are elements of $terms$ \iffalse TODO: A little
-  awkward; maybe use the above notation? \fi. To ensure the list is finite,
-  function applications are only nested up to a specified depth (by default, 3).
+\item Given a signature $\Sigma$ of typed expressions and set of variables $V$,
+  \quickspec{} generates a list $terms$ containing the expressions from
+  $\Sigma$ (including functions), the variables from $V$ and type-correct
+  function applications \hs{f x}, where \hs{f} and \hs{x} are elements of
+  $terms$. To ensure the list is finite, function applications are only nested
+  up to a specified depth (by default, 3).
 \item The elements of $terms$ are grouped into equivalence classes, based on
   their type.
-\item Each variable is instantiated to a particular value, generated randomly by
-  \qcheck{}.
-\item For each class, the members are compared (using a pre-specified function,
-  such as equality \hs{==}) to see if these instantiations have caused an
-  observable difference between members. If so, the class is split up to
-  separate such distinguishable members.
-\item The previous steps of variable instantiation and comparison are repeated
-  until the classes stabilise (i.e. no differences have been observed for some
-  specified number of repetitions).
-\item A set of equations are then conjectured, relating each class's members.
-\end{enumerate}
-
-\iffalse
-% TODO: Compare the above to this; merge/swap anything that's better
-\begin{enumerate}
-\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
-  generates a list $terms$ containing the constants (including functions) from
-  $\Sigma$, the variables from $V$ and type-correct function applications
-  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
-  is finite, function applications are only nested up to a specified depth (by
-  default, 3).
-\item The elements of $terms$ are grouped into equivalence classes, based on
-  their type.
-\item The equivalence of terms in each class is tested using \qcheck{}:
+\item The equivalence of terms in each class is tested using \quickcheck{}:
   variables are instantiated to particular values, generated randomly, and the
@@
   those members.
-\item The previous steps of testing and splitting are repeated until the classes
-  stabilise (i.e. no differences have been observed for some specified number of
-  repetitions).
+\item The previous steps of variable instantiation and comparison are repeated
+  until the classes stabilise (i.e. no differences have been observed for some
+  specified number of repetitions).
 \item For each class, one member is selected and equations are conjectured that
@@
 \end{enumerate}
-\fi
 
-Such conjectures can be used in several ways: they can be simplified for direct
-presentation to the user (for example via the congruence closure algorithm
-\iffalse\cite{TODO}\fi), sent to a more rigorous system like \hspec{} or
-\hipster{} for proving, or even serve as a background theory for an
-automated theorem prover \cite{claessen2013automating}.
+Such conjectures can be used in several ways: they can be presented directly
+to the user, sent to a more rigorous system like \hipspec{} or \hipster{} for
+proving, or even serve as a background theory for an automated theorem
+prover~\cite{claessen2013automating}.
 
 As an example, we can consider a simple signature containing the expressions
-from Figure \ref{fig:haskellteexample}:
+from Figure~\ref{fig:haskellteexample}, and some suitable variables:
 
 \begin{align*}
-  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
+  \Sigma_{\hs{Nat}} &= \{\hs{Z},\ \hs{S},\ \hs{plus},\ \hs{mult},\ \hs{odd},
+                       \ \hs{even}\} \\
+  V_{\hs{Nat}}     &= \{\hs{a :: Nat},\ \hs{b :: Nat},\ \hs{c :: Nat}\}
 \end{align*}
 
-Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$,
-\qspec{}'s enumeration will resemble the following:
+\quickspec{}'s enumeration of these terms will resemble the following:
 
 \begin{align*}
-  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
-                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
+  terms_{\hs{Nat}} = [& \hs{Z},\ \hs{S},\ \hs{plus},\ \hs{mult},\ \hs{odd},
+                     \ \hs{even},\ \hs{a},\ \hs{b},\ \hs{c},\ \hs{S Z},
+                     \ \hs{S a},\ \hs{S b}, \\
+                     & \hs{S c},\ \hs{plus Z},\ \hs{plus a},\ \dots ]
 \end{align*}
@@
 Notice that functions such as \hs{plus} and \hs{mult} are valid terms, despite
-not being applied to any arguments. In addition, Haskell curries functions, so
-these binary functions will be treated as unary functions which return unary
-functions. This is required as the construction of $terms$ applies functions to
-one argument at a time.
+not being applied to any arguments. In addition, all Haskell functions are unary
+(due to currying), which makes it valid to apply them one argument at a time as
+we construct $terms_{\hs{Nat}}$.
 
@@
   \end{haskell}
-  \caption{Equations conjectured by \qspec{} for the functions in Figure
+  \caption{Equations conjectured by \quickspec{} for the functions in Figure
     \ref{fig:haskellteexample}; after simplification.}
@@
 
-These terms will be grouped into five classes, one each for \hs{Nat},
-\hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}. As the
-variables $a$, $b$ and $c$ are instantiated to various randomly-generated
-numbers, these equivalence classes will be divided, until eventually the
-equations in Figure \ref{fig:qspecresult} are conjectured.
+The elements of $terms_{\hs{Nat}}$ will be grouped into five classes, one each
+for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and
+\hs{Bool}. As the variables \hs{a}, \hs{b} and \hs{c} are instantiated to
+various randomly-generated numbers, the differences between these terms will
+be discovered and the equivalence classes will be divided, until eventually
+the equations in Figure \ref{fig:qspecresult} are conjectured.
 
@@
 what to include in the signature. Here we know that addition and multiplication
-are closely related, and hence obey many algebraic laws.
+are closely related, and hence obey many algebraic laws; arbitrary definitions
+from a typical software library or proof development are unlikely to have be
+related so strongly.
 
-\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing
+\iffalse
+\quickspec{} (and \hipspec{}) is also compatible with Haskell's existing testing
 infrastructure, such that an invocation of \texttt{cabal test} can run these
-tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and
+tools alongside more traditional QA tools like \quickcheck{}, \textsc{HUnit} and
 \textsc{Criterion}.
 
-In fact, there are similarities between the way a TE system like \qspec{} can
+In fact, there are similarities between the way a TE system like \quickspec{} can
 generalise from checking \emph{particular} properties to \emph{inventing} new
-ones, and the way counterexample finders like \qcheck{} can generalise from
+ones, and the way counterexample finders like \quickcheck{} can generalise from
 testing \emph{particular} expressions to \emph{inventing} expressions to
@@
 and the consequences for testing and QA in general.
+\fi
 
+\iffalse
+% TODO: Work this into the example
+
+As an example, giving the definitions from Figure~\ref{fig:haskellteexample} to
+\quickspec{}, along with suitable comparison functions, random data generators and
+variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
+class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
+\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
+(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
+\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
+them up, and repeat until no more splits are found. Equations are then generated
+between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
+\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
+to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
+instance of the more general \hs{b == not (not b)}). The remaining equations
+(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
+output, either for the user to peruse or for another system to process, like the
+\hipspec{} automated theorem prover.
+
+Note that we must be able to generate values of a type in order to include
+variables of that type. Also, only equivalence classes whose element type can be
+compared are split and conjectured about in this way; in particular, functions
+(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
+equivalence class, but they will not be subject to testing and comparison since
+Haskell has no generic notion of function equality~\footnote{In
+  fact \quickspec{} compares expressions using a total ordering rather than equality;
+  this is even more restrictive than requiring a notion of equality, but reduces
+  the number of required comparisons}. \quickspec{} can hence conjecture that two
+functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
+that the functions themselves are equal (e.g. \hs{f == g}).
+\fi
+
+% TODO
 %%%%% THE BELOW SEEMS TO BE INTERESTINGNESS?

### 2019-06-02 PhDThesis/benchmark.tex
Bunch more changes to thesis
Headingy things
Thesis updates

@@
-\section{Theory Exploration Benchmark}
-\label{sec:proposal}
+\chapter{Theory Exploration Benchmark}
+\label{sec:benchmark}
 
@@
 be measured and compared to optimise toward our goal. To this end we define a
-benchmarking methodology, shown in Figure~\ref{figure:flow_chart} on
-page~\pageref{figure:flow_chart}, which both generates a large
+benchmarking methodology, shown in Figure~\ref{fig:flow_chart} on
+page~\pageref{fig:flow_chart}, which both generates a large
 definition/ground-truth corpus, and provides a scalable, statistical approach to
@@
 
-\subsection{Preparation}
+\section{Preparation}
 \label{section:prep}
@@
 
-\subsection{Sampling}
+\section{Sampling}
 \label{section:sampling}
@@
 
-\subsection{Evaluation}
+\section{Evaluation}
 \label{section:evaluation}
@@
   \caption[]{High-level view of the benchmarking methodology described in
-    $\S$\ref{sec:proposal}, showing
+    $\S$\ref{sec:benchmark}, showing
     \begin{tikzpicture}
@@
   \end{tikzpicture}}
-  \label{figure:flow_chart}
+  \label{fig:flow_chart}
 \end{figure}
@@
 
-We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
-tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
-benchmark as our ground truth corpus~\cite{claessen2015tip}.
+We have applied our benchmarking methodology to the \quickspec{} and \isacosy{}
+MTE tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem
+proving benchmark as our ground truth corpus~\cite{claessen2015tip}.
 
@@
 machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
-of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
-and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
+of RAM. Most \quickspec{} runs either finished within 200 seconds or not at all,
+and sizes above 20 mostly timed out. \isacosy{} mostly finished within 300 seconds
 on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
@@
 total of 1240 runs. To keep the benchmarking time down to a few days we chose a
-timeout of 300 seconds, since that covered most of the successful QuickSpec and
-IsaCoSy results we saw, and longer times gave rapidly diminishing
+timeout of 300 seconds, since that covered most of the successful \quickspec{} and
+\isacosy{} results we saw, and longer times gave rapidly diminishing
 returns. During analysis, duplicate samples (caused by our requirement that
@@
 ``destructors'' (elimination forms; field accessors). For example the type of
-lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
+lists from Figure~\ref{fig:list_theory} can be defined in the TIP format as
 follows:
@@
 code, including comparison functions and random data generators suitable for
-QuickCheck. We translate all 269 unique definitions into a single module/theory
+\quickcheck{}. We translate all 269 unique definitions into a single module/theory
 which is imported on each run of the tools, although only those functions which
@@
 
-\subsection{QuickSpec}
+\subsection{\quickspec{}}
 
-We benchmarked QuickSpec version 0.9.6, a tool written in Haskell for
+We benchmarked \quickspec{} version 0.9.6, a tool written in Haskell for
 conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
+$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark \quickspec{}, we
 need to automate some of the decisions which are normally left up to the user:
@@
     variables for each type that appears as a function argument, except for
-    types which have no QuickCheck data generators.
+    types which have no \quickcheck{} data generators.
   \item We must \emph{monomorphise} all types. For example, functions like
@@
       up and don't include that function.}
-  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
+  \item Haskell functions are ``black boxes'', which \quickspec{} can't compare
     during its exploration process. They are also curried, always taking one
-    argument but potentially returning another function. QuickSpec lets us
+    argument but potentially returning another function. \quickspec{} lets us
     assign an arity to each function in the signature, from 0 to 5, so we pick
@@
   \input{quickspectime.pgf}
-  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
+  \caption{Running times of \quickspec{} on theories sampled from TIP. Each point
     is a successful run (spread out horizontally to reduce overlaps). Red
@@
     show 1.5$\times$~inter-quartile range.}
-  \label{figure:quickspec_runtimes}
+  \label{fig:quickspec_runtimes}
 \end{figure}
@@
   \input{quickspecprec.pgf}
-  \caption{Precision and recall of successful QuickSpec runs (spread
+  \caption{Precision and recall of successful \quickspec{} runs (spread
     horizontally to reduce overlapping). Lines show the mean proportion for each
@@
     $0.2$ and $0.5$"}
-  \label{figure:quickspec_precRec}
+  \label{fig:quickspec_precRec}
 \end{figure}
@@
 The time taken to explore samples of different sizes is shown in
-Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
+Figure~\ref{fig:quickspec_runtimes}. Failed runs are shown in red: all
 failures were due to timing out, and these occurred for all sample sizes but are
@@
 
-Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
-Since QuickSpec generates monotonically more conjectures as definitions are
+Precision and recall are shown in Figure~\ref{fig:quickspec_precRec}.
+Since \quickspec{} generates monotonically more conjectures as definitions are
 added to a signature (assuming sufficient testing), its decreasing precision
@@
 
-This indicates two potential improvements to the QuickSpec algorithm (as far as
+This indicates two potential improvements to the \quickspec{} algorithm (as far as
 this benchmark is concerned). The deluge of generated conjectures could be
@@
 large reduction in recall. The latter option could also improve the running
-time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
+time, since (based on Figure~\ref{fig:quickspec_runtimes}) multiple smaller
 signatures may be faster to explore than a single large one. Such improvements
@@
 
-QuickSpec's recall is limited by two factors: the algorithm is unable to
+\quickspec{}'s recall is limited by two factors: the algorithm is unable to
 synthesise some properties, such as conditional equations, inequalities, terms
@@
 
-\subsection{IsaCoSy}
+\subsection{\isacosy{}}
 
-We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
+We took \isacosy{} from version 2015.0.3 of the IsaPlanner project, and ran it
 with the 2015 version of Isabelle. The following issues had to be overcome to
-make our benchmark applicable to IsaCoSy:
+make our benchmark applicable to \isacosy{}:
 
@@
   it before translating. It still appears in samples and contributes to the
-  ground truth, which penalises IsaCoSy for being unable to explore such
+  ground truth, which penalises \isacosy{} for being unable to explore such
   definitions.
-\item When using a type in an IsaCoSy signature, that type's constructors will
+\item When using a type in an \isacosy{} signature, that type's constructors will
   automatically be included in the exploration. Since those constructors will
@@
   discards any conjectures which involve functions other than those in the
-  current sample. This presumably results in more work for IsaCoSy, exploring
+  current sample. This presumably results in more work for \isacosy{}, exploring
   constructors unnecessarily, but it at least does not bring down the quality
@@
 \item Partial functions cause problems during exploration, since they can throw
-  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
-  pre-populating IsaCoSy's constraint set with these undefined expressions
+  an ``undefined'' exception which causes \isacosy{} to abort. We avoid this by
+  pre-populating \isacosy{}'s constraint set with these undefined expressions
   (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
-  IsaCoSy from ever generating them.
+  \isacosy{} from ever generating them.
 \end{itemize}
@@
   \input{isacosytime.pgf}
-  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
+  \caption{Running times of \isacosy{} on theories sampled from TIP. Each point
     is a successful run (spread out horizontally to reduce overlaps). Red
@@
     1.5$\times$~inter-quartile range.}
-  \label{figure:isacosy_runtimes}
+  \label{fig:isacosy_runtimes}
 \end{figure}
@@
   \input{isacosyprec.pgf}
-  \caption{Precision and recall of all successful IsaCoSy runs (spread
+  \caption{Precision and recall of all successful \isacosy{} runs (spread
     horizontally to reduce overlapping; all runs above size 6 failed). Lines
@@
     deviation.}
-  \label{figure:isacosy_precRec}
+  \label{fig:isacosy_precRec}
 \end{figure}
 
-The most striking result is how rapidly IsaCoSy's running time, shown in
-Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
+The most striking result is how rapidly \isacosy{}'s running time, shown in
+Figure~\ref{fig:isacosy_runtimes}, increases with sample size: all runs above
 size 6 failed, with most timing out and a few aborting early due to running out
@@
 so even large increases to the timeout would not produce many more successful
-observations. A significant amount of IsaCoSy's time (around 50 seconds) was
+observations. A significant amount of \isacosy{}'s time (around 50 seconds) was
 spent loading the generated theory (containing all distinct datatypes and
@@
 With so few successful datapoints it is difficult to make strong claims about
-the precision/recall quality of IsaCoSy's output, shown in
-Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
-proportions that IsaCoSy is capable of discovering interesting conjectures, and
-the graphs appear to follow the same shapes as those of QuickSpec (decreasing
+the precision/recall quality of \isacosy{}'s output, shown in
+Figure~\ref{fig:isacosy_precRec}. It is clear from the quantity of non-zero
+proportions that \isacosy{} is capable of discovering interesting conjectures, and
+the graphs appear to follow the same shapes as those of \quickspec{} (decreasing
 precision, flat recall), although the error margins are too wide to be
@@
 
-We compare the running times of QuickSpec and IsaCoSy using our paired variant
+We compare the running times of \quickspec{} and \isacosy{} using our paired variant
 of the Speedup-Test protocol, where each of our sample sizes is a separate
@@
 two speedups were not deemed statistically significant. For all other sample
-sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
+sizes \quickspec{} was found to be significantly faster with $\alpha = 0.05$,
 putting the proportion of sizes with faster median times between 66.9\% and
-98.2\% with 95\% confidence. The magnitude of each speedup (median IsaCoSy time
-divided by median QuickSpec time) is given in table~\ref{table:speedups}. We
+98.2\% with 95\% confidence. The magnitude of each speedup (median \isacosy{} time
+divided by median \quickspec{} time) is given in table~\ref{table:speedups}. We
 would predict the speedup to grow for larger sample sizes, but the increasing
@@
   \end{tabular}
-  \caption{Speedup from using QuickSpec compared to IsaCoSy, i.e. the median
-    time taken by IsaCoSy divided by that of QuickSpec. QuickSpec was found to
+  \caption{Speedup from using \quickspec{} compared to \isacosy{}, i.e. the median
+    time taken by \isacosy{} divided by that of \quickspec{}. \quickspec{} was found to
     be significantly faster ($\alpha = 0.05$) for all sizes, except 1 and 2 due
@@
 sizes to produce table~\ref{table:recall}, and found that the recall of
-QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p-value
+\quickspec{} ($37\%$) is significantly higher than \isacosy{} ($25\%$) with a p-value
 of 0.0026.
@@
     \hline
-    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
+    & \bfseries Found by \isacosy{} & \bfseries Not found by \isacosy{} \\
     \hline
-    \bfseries Found by QuickSpec     & 28 & 19 \\
+    \bfseries Found by \quickspec{}     & 28 & 19 \\
     \hline
-    \bfseries Not found by QuickSpec & 4  & 75 \\
+    \bfseries Not found by \quickspec{} & 4  & 75 \\
     \hline
@@
   \caption{Contingency table for ground truth properties, pooled from those
-    samples where both QuickSpec and IsaCoSy finished successfully. The combined
-    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
+    samples where both \quickspec{} and \isacosy{} finished successfully. The combined
+    recall for \quickspec{} is $37\%$ and for \isacosy{} is $25\%$.}
   \label{table:recall}
@@
     \hline
-    \bfseries IsaCoSy   & 32          & 137      \\
+    \bfseries \isacosy{}   & 32          & 137      \\
     \hline
-    \bfseries QuickSpec & 47          & 301      \\
+    \bfseries \quickspec{} & 47          & 301      \\
     \hline
   \end{tabular}
-  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
-    IsaCoSy, pooled from those samples where both both finished successfully.
-    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
+  \caption{Interesting and uninteresting conjectures generated by \quickspec{} and
+    \isacosy{}, pooled from those samples where both both finished successfully.
+    The combined precision for \quickspec{} is $14\%$ and for \isacosy{} is $19\%$.}
   \label{table:precision}
@@
 $\alpha = 0.05$; hence we do not consider the difference between these
-proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.
+proportions (14\% for \quickspec{}, 19\% for \isacosy{}) to be significant.
 
-Note that precision is the only comparison where IsaCoSy scored higher, and even
+Note that precision is the only comparison where \isacosy{} scored higher, and even
 that was not found to be significant. Also, precision on its own is not the
 whole story: it can be increased at the expense of recall by making a tool more
-conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
+conservative. \isacosy{} may already be overly-conservative compared to \quickspec{},
 given that its recall is significantly lower.
 
-More importantly, the poor time and memory usage of IsaCoSy meant that very few
+More importantly, the poor time and memory usage of \isacosy{} meant that very few
 samples finished successfully. If we include failed runs in our tables, treating
 them as if they succeeded with no output, the resulting statistics lean
-overwhelmingly in favour of QuickSpec simply because it generated so much more
-than IsaCoSy in the available time.
+overwhelmingly in favour of \quickspec{} simply because it generated so much more
+than \isacosy{} in the available time.
 
-\section{Discussion}
+\subsection{Discussion}
 \label{sec:discussion}
@@
 after emitting some output. Whilst all of the systems we have encountered are of
-this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
-be run without a depth limit, and form part of the ``scientist assistant'' role
-which Lenat envisaged for AM, or McCarthy's earlier ``advice
+this form, some (such as \isacosy{}, \quickspec{} 2 and \speculate{}) could
+conceivably be run without a depth limit, and form part of the ``scientist
+assistant'' role which Lenat envisaged for AM, or McCarthy's earlier ``advice
 taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for

### 2019-06-02 PhDThesis/bucketing.tex
Thesis updates

@@
 \chapter{The Signature Selection Problem}
+\label{sec:signatureselection}
 
@@
 all rely on enumerating combinations of definitions in some way, which can cause
-infeasible running times on larger inputs (in our experience, a few dozen
+infeasible running times on larger inputs (based on results of the Theory
+Exploration Benchmark in \S~\ref{sec:benchmark}, this happens above a few dozen
 definitions). To avoid this, users of these systems must carefully select only
@@
 
-We dub such selection the \emph{signature selection problem} for theory
+We dub such cherry-picking the \emph{signature selection problem} for theory
 exploration, and we analyse its effect on the running time of theory exploration
-tools on the TIP particular problem set, and on the quality of the statements
-they are able to generate. An automated solution to this signature selection problem would
-accept a large number of definitions and efficiently select sub-sets which are
-both small enough to reasonably explore, whilst still enabling interesting
-properties to be discovered.
+tools on the Theory Exploration Benchmark, and on the quality of the statements
+they are able to generate. An automated solution to this signature selection
+problem would accept a large number of definitions and efficiently select
+sub-sets which are both small enough to reasonably explore, whilst still
+enabling interesting properties to be discovered.
 
@@
     Benchmark.
-  \item Analyse the effect of signature selection on the quantity of desirable properties
-    attainable by theory exploration on this problem set.
-  \item Demonstrate how the signature selection problem is distinct from the well-know
-    \emph{clustering problem} in machine learning, and how clustering algorithms
-    cannot be directly applied to signature selection.
+  \item Analyse the effect of signature selection on the quantity of desirable
+    properties attainable by theory exploration on this problem set.
+  \item Demonstrate how the signature selection problem is distinct from the
+    well-known \emph{clustering problem} in machine learning, and how clustering
+    algorithms cannot be directly applied to signature selection.
 \end{enumerate}
 
+\iffalse
+% TODO: Split out
 Some bespoke components required for these analyses and comparisons may also be
@@
 \item Dynamic evaluation of Haskell code, with access to dynamically installed
-  Haskell packages\footnote{http://hackage.haskell.org/package/nix-eval}.
+  Haskell packages\footnote{\hPackage{nix-eval}}.
 \item Translation of the signature selection problem to the constrain-solving domain, and
@@
 \end{enumerate}
-
-\iffalse
-% TODO: Do we need this?
-We describe theory exploration in more detail, as well as the \qspec{} tool we
-use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
-signature selection problem more formally in \S~\ref{sec:sigselect} and measure its
-effect on theory exploration systems in \S~\ref{sec:implementation}. In
-\S~\ref{sec:clustering} we highlight the difference between signature selection and the
-well-known \emph{clustering problem} from machine learning, by demonstrating
-that clustering algorithms perform poorly at signature selection.  A variety of related
-work is surveyed in \S~\ref{sec:related} and we conclude in
-\S~\ref{sec:conclusion} with potential directions for future research.
 \fi
 
-\section{Background}
-\label{sec:background}
+\section{The Signature Selection Problem}
+\label{sec:sigselect}
 
-As an example, giving the definitions from Figure~\ref{fig:haskellteexample} to
-\qspec{}, along with suitable comparison functions, random data generators and
-variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
-class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
-\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
-(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
-\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
-them up, and repeat until no more splits are found. Equations are then generated
-between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
-\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
-to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
-instance of the more general \hs{b == not (not b)}). The remaining equations
-(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
-output, either for the user to peruse or for another system to process, like the
-\textsc{HipSpec} automated theorem prover.
+Although complete, the enumeration approaches used by tools like \quickspec{}
+are wasteful: many terms are unlikely to appear in theorems, which requires
+careful choice by the user of what to include in the signature. For example, we
+know that addition and multiplication are closely related, and hence obey many
+algebraic laws; it would be prudent to explore these functions together. On the
+other hand, functions such as HTTP parsers and spell-checkers will not be
+related in many interesting ways; exploring their combinations is wasteful.
 
-Note that we must be able to generate values of a type in order to include
-variables of that type. Also, only equivalence classes whose element type can be
-compared are split and conjectured about in this way; in particular, functions
-(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
-equivalence class, but they will not be subject to testing and comparison since
-Haskell has no generic notion of function equality~\footnote{In
-  fact \qspec{} compares expressions using a total ordering rather than equality;
-  this is even more restrictive than requiring a notion of equality, but reduces
-  the number of required comparisons}. \qspec{} can hence conjecture that two
-functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
-that the functions themselves are equal (e.g. \hs{f == g}).
+The signature selection problem is that of selecting sub-sets of a large
+signature, such that:
 
-Although complete, this enumeration approach is wasteful: many terms are
-unlikely to appear in theorems, which requires careful choice by the user of
-what to include in the signature. For example, we know that addition and
-multiplication are closely related, and hence obey many algebraic laws. Our
-machine learning technique aims to predict these kinds of relations between
-functions, so we can create small signatures which can be explored quickly, yet
-have the potential to give rise to many equations.
+\begin{itemize}
+\item Few subsets are selected
+\item Each sub-set is small
+\item As many of the total interesting relationships are still findable
+\end{itemize}
 
@@
 
-Our approach to scaling up \qspec{} takes inspiration from two sources. The
+Our approach to scaling up \quickspec{} takes inspiration from two sources. The
 first is relevance filtering, which makes expensive algorithms used in theorem
@@
 
-\subsection{The Signature Selection Problem}
-\label{sec:sigselect}
-
-% TODO{2019-04-19}
-
-\subsection{Recurrent Clustering}
-\label{sec:recurrentclustering}
-
-Rather than clustering Core expressions directly, we first perform \emph{feature
-  extraction} to transform them into numeric \emph{feature vectors}. We adapt
-the methodology of \emph{recurrent clustering} proposed in
-\cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature
-extraction and clustering into a single recursive algorithm, such that the
-feature vector of an expression depends on the feature vectors of those
-expressions it references. We suggest a new recurrent clustering and feature
-extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and
-compare its similarity and differences to those of ML4PG and ACL2(ml).
-
-We describe our algorithm in two stages: the first transforms the nested
-structure of expressions into a flat vector representation; the second converts
-the discrete symbols of Core syntax into features (real numbers), which we will
-denote using the function $\phi$.
-
-\subsubsection{Expressions to Vectors}
-\label{sec:expressionstovectors}
-
-Our recurrent clustering algorithm is based on k-means clustering, which
-considers the elements of a feature vector to be \emph{orthogonal}. Hence we
-must ensure that similar expressions not only give rise to similar numerical
-values, but crucially that these values appear \emph{at the same position} in
-the feature vectors. Since different patterns of nesting can alter the ``shape''
-of expressions, simple traversals (breadth-first, depth-first, post-order, etc.)
-may cause features from equivalent sub-expressions to be mis-aligned. For
-example, consider the following expressions, which represent pattern-match
-clauses with different patterns but the same body (\hs{\vlocal{y}}):
-
-\begin{equation}\label{eq:xy}
-  \begin{array}{r@{}l@{}l@{}}
-    X\ &=\ \CAlt\ (\CDataAlt\ \id{C})\ & (\vlocal{y}) \\
-    Y\ &=\ \CAlt\ \CDefault\           & (\vlocal{y})
-  \end{array}
-\end{equation}
-
-If we traverse these expressions in breadth-first order, converting each token
-to a feature using $\phi$ and padding to the same length with $0$, we would get
-the following feature vectors:
-
-\begin{small}
-  \begin{equation}
-    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
-      breadthFirst(Y)\ &=\ (\feature{\CAlt},\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &)
-    \end{array}
-  \end{equation}
-\end{small}
-
-Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$
-are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match
-(others may match by coincidence, depending on $\phi$). These feature vectors
-might be deemed very dissimilar during clustering, despite the intuitive
-similarity of the expressions $X$ and $Y$ from which they derive.
-
-If we were to align these features optimally, by padding the fourth column
-rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to
-match, making the similarity of the vectors more closely match our intuition and
-depend less on coincidence.
-
-The method we use to ``flatten'' expressions, described below, is a variation of
-breadth-first traversal which pads each level of nesting to a fixed size $c$
-(for \emph{columns}). This doesn't guarantee alignment, but it does prevent
-mis-alignment from accumulating across different levels of nesting. Our method
-would align these features into the following vectors, if $c = 2$: \footnote{In
-  fact, the constructor identifier $\id{C}$ would not appear in the vector, but
-  this example is still accurate in terms of laying out the features as given.}
-
-\begin{small}
-  \begin{equation}\label{eq:flattened}
-    \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
-      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
-    \end{array}
-  \end{equation}
-\end{small}
-
-Here $\frac{1}{2}$ of the original 6 features align, which is more than
-$breadthFirst$ but not optimal. Both vectors have also been padded by an extra 2
-zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
-
-To perform this flattening we first transform the nested tokens of an expression
-into a \emph{rose tree} of features. Intuitively, these are simply an
-s-expression representation of the Core syntax, with the feature extraction
-function $\phi$ mapped over the labels. The full transformation is given in
-Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the
-result for the \hs{odd} function.
-
-\begin{figure}
-  \centering
-  \begin{scriptsize}
-      \Tree[ .$\feature{\CLam}$
-                $\feature{\id{a}}$
-                [ .$\feature{\CCase}$
-                     [ .$\feature{\CVar}$
-                          [ .$\feature{\CLocal}$
-                               $\feature{\id{a}}$ ]]
-                     $\feature{\id{b}}$
-                     [ .$\feature{\CAlt}$
-                          $\feature{\CDataAlt}$
-                          [ .$\feature{\CVar}$
-                               $\feature{\CConstructor}$ ]]
-                     [ .$\feature{\CAlt}$
-                          $\feature{\CDataAlt}$
-                          [ .$\feature{\CApp}$
-                               [ .$\feature{\CVar}$
-                                    [ .$\feature{\CGlobal}$
-                                         $\feature{\id{even}}$ ]]
-                               [ .$\feature{\CVar}$
-                                    [ .$\feature{\CLocal}$
-                                         $\feature{\id{n}}$ ]]]
-                          $\feature{\id{n}}$ ]]]
-  \end{scriptsize}
-  \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the
-    expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree
-    is rendered with its feature at the node and sub-trees beneath.}
-\end{figure}
-
-\begin{figure}
-    \begin{equation*}
-      \begin{bmatrix}
-        \feature{\CLam}      & 0                       & 0                 & 0                   & 0               & 0                \\
-        \feature{\id{a}}     & \feature{\CCase}        & 0                 & 0                   & 0               & 0                \\
-        \feature{\CVar}      & \feature{\id{b}}        & \feature{\CAlt}   & \feature{\CAlt}     & 0               & 0                \\
-        \feature{\CLocal}    & \feature{\CDataAlt}     & \feature{\CVar}   & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
-        \feature{\id{a}}     & \feature{\CConstructor} & \feature{\CVar}   & \feature{\CVar}     & 0               & 0                \\
-        \feature{\CGlobal}   & \feature{\CLocal}       & 0                 & 0                   & 0               & 0                \\
-        \feature{\id{even}}  & \feature{\id{n}}        & 0                 & 0                   & 0               & 0
-      \end{bmatrix}
-    \end{equation*}
-    \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
-    \label{fig:matrixexample}
-\end{figure}
-
-These rose trees are then turned into matrices, as shown in Figure
-\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at
-depth $i$ in the rose tree, read left-to-right, followed by any required
-padding. These matrices are then either truncated, or padded with rows (on the
-bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.
-
-\begin{sloppypar}
-  Finally, matrices are turned into vectors by concatenating the rows from top
-  to bottom, hence Figure \ref{fig:matrixexample} will produce a vector
-  beginning
-  $(\feature{\CLam} ,
-    0               ,
-    0               ,
-    0               ,
-    0               ,
-    0               ,
-    \feature{\id{a}},
-    \feature{\CCase},
-    0               ,
-    0               ,
-    0               ,
-    0               ,
-    \feature{\CVar} ,
-    \feature{\id{b}},
-    \feature{\CAlt} ,
-    \feature{\CAlt} ,
-    \dots$.
-\end{sloppypar}
-
-\subsubsection{Symbols to Features}
-\label{sec:symbolstofeatures}
-
-We now define the function $\phi$, which turns terminal symbols of Core syntax
-into features (real numbers). For known language features, such as
-$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
-assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14}
-in Coq. We use a constant $\alpha$ to separate these values from those of other
-tokens (e.g. identifiers), but the order is essentially arbitrary: \footnote{In
-  \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix}
-  and \coq{cofix} are grouped together to reduce redundancy; we do not group
-  tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and
-  \CGlobal.}
-
-\begin{equation} \label{eq:feature}
-  \begin{aligned}
-    \feature{\CAlt}          &= \alpha      &
-    \feature{\CDataAlt}      &= \alpha + 1  &
-    \feature{\CLitAlt}       &= \alpha + 2  \\
-    \feature{\CDefault}      &= \alpha + 3  &
-    \feature{\CNonRec}       &= \alpha + 4  &
-    \feature{\CRec}          &= \alpha + 5  \\
-    \feature{\CBind}         &= \alpha + 6  &
-    \feature{\CLet}          &= \alpha + 7  &
-    \feature{\CCase}         &= \alpha + 8  \\
-    \feature{\CLocal}        &= \alpha + 9  &
-    \feature{\CGlobal}       &= \alpha + 10 &
-    \feature{\CConstructor}  &= \alpha + 11 \\
-    \feature{\CVar}          &= \alpha + 12 &
-    \feature{\CLam}          &= \alpha + 13 &
-    \feature{\CApp}          &= \alpha + 14 \\
-    \feature{\CType}         &= \alpha + 15 &
-    \feature{\CLit}          &= \alpha + 16 &
-    \feature{\CLitNum}       &= \alpha + 17 \\
-    \feature{\CLitStr}       &= \alpha + 18
-  \end{aligned}
-\end{equation}
-
-To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which
-gives equal values for $\alpha$-equivalent expressions (i.e. renaming an
-identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
-  index} of the identifier \cite{de1972lambda}, denoted $i_l$:
-
-\begin{equation} \label{eq:localfeature}
-  \feature{l} = i_l + 2 \alpha \quad \text{if $l \in \mathcal{L}$}
-\end{equation}
-
-We again use $\alpha$ to separate these features from those of other constructs.
-
-We discard the contents of literals and constructor identifiers when converting
-to rose trees, so the only remaining case is global identifiers
-$\mathcal{G}$. Since these are declared \emph{outside} the body of an
-expression, we cannot perform the same indexing trick as for local
-identifiers. We also cannot directly encode the form of the identifiers,
-e.g. using a scheme like G{\"o}del numbering, since this is essentially
-arbitrary and has no effect on their semantic meaning (referencing other
-expressions).
-
-Instead, we use the approach taken in the latest versions of ML4PG and encode
-global identifiers \emph{indirectly}, by looking up the expressions which they
-\emph{reference}:
-
-\begin{equation} \label{eq:globalfeature}
-  \feature{g \in \mathcal{G}} =
-    \begin{cases}
-      i + 3 \alpha \quad & \text{if $g \in C_i$} \\
-      f_{recursion}         & \text{otherwise}
-    \end{cases}
-\end{equation}
-
-Where $C_i$ are our clusters (in some arbitrary order). This is where the
-recurrent nature of the algorithm appears: to determine the contents of $C_i$,
-we must perform k-means clustering \emph{during} feature extraction; yet that
-clustering step, in turn, requires that we perform feature extraction.
-
-For this recursive process to be well-founded, we perform a topological sort on
-declarations based on their dependencies (the expressions they reference). In
-this way, we can avoid looking up expressions which haven't been clustered
-yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth}
-to produce a sorted list of \emph{strongly connected components} (SCCs), where
-each SCC is a mutually-recursive sub-set of the declarations. If an identifier
-cannot be found amongst those clustered so far, it must appear in the same SCC
-as the expression we are processing; hence we give that identifier the constant
-feature value $f_{recursion}$.
-
-\begin{algorithm}
-  \begin{algorithmic}[1]
-    \Require List $d$ contains SCCs of (identifier, expression) pairs, in
-    dependency order.
-    \Procedure{RecurrentCluster}{}
-      \State $\vect{C} \gets []$
-      \State $DB \gets \varnothing$
-      \ForAll{$scc$ \textbf{in} $d$}
-        \State $DB \gets DB \cup \{(i, featureVec(e)) \mid (i, e) \in scc\}$
-        \State $\vect{C} \gets kMeans(DB)$
-      \EndFor
-      \Return $\vect{C}$
-    \EndProcedure
-  \end{algorithmic}
-  \caption{Recurrent clustering of Core expressions.}
-  \label{alg:recurrent}
-\end{algorithm}
-
-By working through the sorted list of SCCs, storing the features of each
-top-level expression as they are calculated, our algorithm can be computed
-\emph{iteratively} rather than recursively, as shown in Algorithm
-\ref{alg:recurrent}.
-
-\iffalse
-
-As an example of this recurrent process, we can consider the Peano arithmetic
-functions from Figure \ref{fig:coreexample}. A valid topological ordering is
-given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
-Core expressions to save space):
-
-\begin{equation}
-  d = [\{(\hs{plus}, \dots)\},
-       \{(\hs{odd}, \dots), (\hs{even}, \dots)\},
-       \{(\hs{mult}, \dots)\}]
-\end{equation}
-
-We can then trace the execution of Algorithm \ref{alg:recurrent} as follows:
-
-\begin{itemize}
-\item The first iteration through \textsc{RecurrentCluster}'s loop will set
-  $scc \gets \{(\hs{plus}, \dots)\}$.
-\item With $i = \hs{plus}$ and $e$ as its Core expression, calculating
-  $featureVec(e)$ is straightforward; the recursive call $\feature{\hs{plus}}$
-  will become $f_{recursion}$ (since \hs{plus} doesn't appear in $\vect{C}$).
-\item The call to $kMeans$ will produce $\vect{C} \gets [\{\hs{plus}\}]$, i.e. a
-  single cluster containing \hs{plus}.
-\item The next iteration will set
-  $scc \gets \{(\hs{odd}, \dots), (\hs{even}, \dots)\}$.
-\item With $i = \hs{odd}$ and $e$ as its Core expression, the call to \hs{even}
-  will result in $f_{recursion}$.
-\item Likewise for the call to $\hs{even}$ when $i = \hs{odd}$.
-\item Since the feature vectors for \hs{odd} and \hs{even} will be identical,
-  $kMeans$ will put them in the same cluster. To avoid the degenerate case of a
-  single cluster, for this example we will assume that $k = 2$; in which case
-  the other cluster must contain \hs{plus}. Their order is arbitrary, so one
-  possibility is $\vect{C} = [\{\hs{odd}, \hs{even}\}, \{\hs{plus}\}]$.
-\item Finally \hs{mult} will be clustered. The recursive call will become
-  $f_{recursion}$ whilst the call to \hs{plus} will become $2 + 3 \alpha$, since
-  $\hs{plus} \in C_2$.
-\item \begin{sloppypar}Again assuming that $k = 2$, the resulting clusters will
-    be
-    \mbox{$\vect{C} \gets [\{\hs{odd}, \hs{even}\}, \{\hs{plus},
-      \hs{mult}\}]$}.\end{sloppypar}
-\end{itemize}
-
-Even in this very simple example we can see a few features of our algorithm
-emerge. For example, \hs{odd} and \hs{even} will always appear in the same
-cluster, since they only differ in their choice of constructor names (which are
-discarded by $toTree$) and recursive calls (which are replaced by
-$f_{recursion}$). A more extensive investigation of these features requires a
-concrete implementation, in particular to pin down values for the parameters
-such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
-
-\fi
-
 \section{Experiments}
@@
 %
-%  - The use of the TEBenchmark methodology for running QuickSpec, including the
+%  - The use of the TEBenchmark methodology for running QSpec, including the
 %    sampling process
@@
   \scalebox{0.45}{\input{images/steppednontoxic.pgf}}
-  \caption{Kaplan-Meier survival plot for running QuickSpec on inputs
+  \caption{Kaplan-Meier survival plot for running \quickspec{} on inputs
     containing various numbers of definitions, sampled from TIP. The x axis
     denotes running time, which we cut short after 300 seconds. The height of
-    each line shows the proportion of \qspec{} runs which were still going at
+    each line shows the proportion of \quickspec{} runs which were still going at
     that time (lower is better). First plot is for all TIP definitions, second
@@
 
-Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \qspec{} running
+Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{} running
 times, when given inputs containing different numbers of definitions. Many runs
@@
   \fi
-  \caption{``Toxic'' definitions, which consistently cause \qspec{} to fail. Two
+  \caption{``Toxic'' definitions, which consistently cause \quickspec{} to fail. Two
     other definitions (\texttt{mul3} and \texttt{mul3acc}) are ommitted due to
@@
 exploration to time out by either exhausting the RAM with too many expressions
-to explore, % TODO{2019-04-27} CAN QUICKSPEC2 AVOID THIS?
+to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
 or by generating such deeply-nested outputs that comparing them takes too long.
@@
 containing them and repeated our analysis.
-
-\section{Implementation}
-\label{sec:implementation}
-
-We provide an implementation of our recurrent clustering algorithm in a tool
-called \textsc{ML4HS}. We obtain Core ASTs from Haskell packages using a plugin
-for the GHC compiler, which emits a serialised version of each Core definition
-as it is being compiled. This approach is more robust than, for example, parsing
-source files, since it avoids the complications of preprocessors and language
-extensions altering the syntax.
-
-A post-processing stage determines which Core definitions can be used by
-\qspec{}, based on their type, visibility (whether they are encapsulated inside
-their module or visible to importers), etc. For each definition, we simply use
-the GHCi interpreter to check if calling \qspec{} with that argument is
-well-typed. This information, along with type signatures, arity, etc. are stored
-alongside the Core definitions in JSON format.
-
-The definitions are then sorted topologically, based on the non-local
-identifiers appearing in their ASTs, and feature vectors are constructed using a
-Haskell implementation of the approach described in \S
-\ref{sec:recurrentclustering}. Since the features associated with each
-identifier may vary between iterations (as the clusters change), we leave the
-raw identifiers in the vector so their features can be extracted in the correct
-context.
-
-We implement Algorithm \ref{alg:recurrent} as a simple shell script, which
-invokes Weka for clustering and associates each Core definition with a cluster
-number. These numbers are used to finish the deferred feature extraction of
-identifiers, the resulting feature vectors are clustered, and the process is
-repeated until all SCCs have been processed.
-
-Once the recurrent clustering is complete, we generate a string of Haskell code
-for each of the final clusters, which will invoke \qspec{} with a suitable
-signature. This involves:
-
-\begin{itemize}
-\item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall
-    t. t -> t -> Bool}, we must choose a concrete representation to use (in this
-  case for \hs{t}), in order to know which random generator to use. We take the
-  approach used in \qcheck{} by attempting to instantiate all type variables to
-  \hs{Integer}. Any values where this is invalid, such as those with
-  incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where
-  \hs{Integer} does not implement \hs{IsString}) will not be included in the
-  signature (this is checked at the AST post-processing stage).
-
-\item{Qualifying}: All names are \emph{qualified} (prefixed by their module's
-  name), to avoid most ambiguity. There is still the possibility that multiple
-  packages will declare modules of the same name, although this is rare as it
-  causes problems for any Haskell programmer trying to use those modules. In
-  such cases the exploration process simply aborts.
-
-\item{Appending variables}: Once a \qspec{} theory has been defined containing
-  all of the given terms, we inspect the types it references and append three
-  variables for each to the theory (enough to discover laws such as
-  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
-  search).
-
-\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
-  that all required packages and modules must be provided up-front, usually by
-  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
-  \emph{dynamically}, depending on the cluster information it is given, we do
-  not know what packages it may need. To work around this problem, we evaluate
-  these strings of Haskell using a custom library called \texttt{nix-eval}. This
-  uses the Nix package manager to obtain all of the required packages and make
-  them available to GHC, and is described further in
-  Appendix~\ref{sec:nix-eval}.
-
-\end{itemize}
-
-The equations resulting from evaluating these strings are collected and
-outputted as JSON values, to ease further processing (e.g. displaying in some
-form, sending to a theorem prover, etc.).
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-% TODO: Move recurrent clustering to its own chapter, with this as its
-% evaluation?
-We have applied our recurrent clustering algorithm to several scenarios, with
-mixed results. A major difficulty in evaluating these clusters is that we have
-no ``ground truth'', i.e. there is no objectively correct way to compare
-expressions. Instead, we provide a qualitative overview of the more interesting
-characteristics.
-
-As a simple example, we clustered (Haskell equivalents of) the running examples
-used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
-\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
-implementations of several functions. We expect those with similar
-\emph{structure} to be clustered together, rather than those which implement the
-same function. The results are shown in Figure \ref{fig:haskellcluster}, where
-we can see the ``\hs{Tail}'' functions clearly distinguished, with little
-distinction between the tail recursive and na\"{\i}ve implementations.
-
-Next we tested whether these same functions would be clustered together when
-mixed with seemingly-unrelated functions, in this case 207 functions from
-Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail}
-functions appeared together in a separate cluster from the rest. This was
-unexpected, with no obvious semantic connection between these two functions and
-the others in their cluster (although most are recursive, due to the nature of
-the \hs{text} library).
-
-\begin{figure}
-  \begin{common-lisp}
-(defun fact (n)
-  (if (zp n) 1 (* n (fact (- n 1)))))
-
-(defun helper-fact (n a)
-  (if (zp n) a (helper-fact (- n 1) (* a n))))
-
-(defun fact-tail (n)
-  (helper-fact n 1))
-
-(defun power (n)
-  (if (zp n) 1 (* 2 (power (- n 1)))))
-
-(defun helper-power (n a)
-  (if (zp n) a (helper-power (- n 1) (+ a a))))
-
-(defun power-tail (n)
-  (helper-power n 1))
-
-(defun fib (n)
-  (if (zp n)
-      0
-      (if (equal n 1)
-          1
-          (+ (fib (- n 1)) (fib (- n 2))))))
-
-(defun helper-fib (n j k)
-  (if (zp n)
-      j
-      (if (equal n 1)
-          k
-          (helper-fib (- n 1) k (+ j k)))))
-
-(defun fib-tail (n)
-  (helper-fib n 0 1))
-  \end{common-lisp}
-  \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
-  \label{fig:lisp}
-\end{figure}
-
-\begin{figure}
-  \begin{haskell}
-fact n = if n == 0
-            then 1
-            else n * fact (n - 1)
-
-helperFact n a = if n == 0
-                    then a
-                    else helperFact (n - 1) (a * n)
-
-factTail n = helperFact n 1
-
-power n = if n == 0
-             then 1
-             else 2 * power (n - 1)
-
-helperPower n a = if n == 0
-                     then a
-                     else helperPower (n - 1) (a + a)
-
-powerTail n = helperPower n 1
-
-fib n = if n == 0
-           then 0
-           else if n == 1
-                   then 1
-                   else fib (n - 1) + fib (n - 2)
-
-helperFib n j k = if n == 0
-                     then j
-                     else if n == 1
-                          then k
-                          else helperFib (n - 1) k (j + k)
-
-fibTail n = helperFib n 0 1
-  \end{haskell}
-  \caption{Haskell equivalents of the Common Lisp functions in Figure \ref{fig:lisp}.}
-  \label{fig:haskelllisp}
-\end{figure}
-
-\begin{figure}
-  \centering
-    \tikzstyle{block} = [rectangle, draw]
-    \tikzstyle{line}  = [draw, -latex']
-    \begin{tikzpicture}[node distance=3cm]
-      \node [block] (c1) {
-        \begin{tikzpicture}[node distance = 1cm]
-          \node [block]                    (factTail)  {\hs{factTail}};
-          \node [block, below of=factTail] (fibTail)   {\hs{fibTail}};
-          \node [block, below of=fibTail]  (powerTail) {\hs{powerTail}};
-        \end{tikzpicture}
-      };
-      \node [block, right of=c1] (c2) {
-        \begin{tikzpicture}[node distance = 1cm]
-          \node [block]                (fact)       {\hs{fact}};
-          \node [block, below of=fact] (helperFact) {\hs{helperFact}};
-        \end{tikzpicture}
-      };
-      \node [block, right of=c2] (c3) {
-        \begin{tikzpicture}[node distance = 1cm]
-          \node [block]                       (fib)         {\hs{fib}};
-          \node [block, below of=fib]         (helperFib)   {\hs{helperFib}};
-          \node [block, below of=helperFib]   (helperPower) {\hs{helperPower}};
-          \node [block, below of=helperPower] (power)       {\hs{power}};
-        \end{tikzpicture}
-      };
-  \end{tikzpicture}
-  \caption{Typical clusters for the functions in Figure \ref{fig:haskelllisp}.}
-  \label{fig:haskellcluster}
-\end{figure}
-
-We have also applied our recurrent clustering algorithm to a variety of the
-most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
-\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
-\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
-expected functions with a similar purpose to appear together, such as the
-various reader and writer functions of \hs{pandoc}, there were always a few
-exceptions which became separate for reasons which are still unclear.
-
-When clustering the \hs{yesod} Web framework, the clustering did seem to match
-our intuitions, in particular since all 15 of Yesod's MIME type identifiers
-appeared in the same cluster.
-
-Whilst recurrent clustering has produced results which merit further
-investigation, the application to theory exploration has yet to be tested
-empirically. This is due to \qspec{}'s use of \qcheck{}'s \hs{Arbitrary} type
-class to generate random values for instantiating variables. Whilst we can
-automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
-all types have \hs{Arbitrary} instances; those without cannot be given any
-variables in our signature, which severely limits the possible combinations
-which can be explored. In many cases, no variables can be included at all,
-leaving just equations involving constants. This has so far prevented us from
-measuring the direct impact on \qspec{} performance, either directly by
-exploring the sub-sets identified through recurrent clustering, or indirectly by
-comparing the equations generated by a full brute-force search to our recurrent
-clusters: those equations relating terms from different clusters would not be
-discovered by our method. It is this ratio of equations found through brute
-force to those found after narrowing-down by clusters which is one of our key
-objectives to maximise at this stage; until we begin to pursue the
-\emph{interestingness} of the properties.
-
-The following less-serious problems were also encountered while applying
-\textsc{ML4HS} to \textsc{Hackage} packages:
-
-\begin{itemize}
-\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
-  declarations to cluster. This is because they make all of their declarations
-  privately, e.g. in ``internal'' modules, then use separate modules to export
-  the public declarations. GHC's renaming phase makes all references to such
-  exports canonical, by pointing them to the private declarations. This forces
-  us to ignore such declarations, as \qspec{} will not be able to access them.
-
-\item Since we do not support type-level entities, we ignore type
-  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
-  ``methods'') which occur in a type class instance. Instead of being clustered,
-  these result in references getting $f_{recursion}$ features. This is
-  especially noticable in libraries like \hs{scientific}, where only the
-  functions for constructing and destructing numbers in scientific notation are
-  clustered; all of the arithmetic is defined in type classes. One difficulty
-  with supporting methods is that their namespace in Core is disjoint from that
-  of regular Haskell identifiers: a transformation layer would be required,
-  along with explicit type annotations to avoid ambiguity.
-\end{itemize}
-
-It seems like this recurrent clustering method has promise, although it will
-require a more thorough exploration of the parameters to obtain more intuitively
-reliable results. These clusters can then be used in several ways to perform
-theory exploration; the most na\"{\i}ve way being to explore each cluster as
-\qspec{} signature in its own right. \textsc{ML4HS} already provides this
-functionality, although the lack of test generators severely limits what can be
-discovered.
-
-As alluded to previously, we also have the opportunity to reason by
-\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
-produce a general ``scheme'' from each equation we find (either through \qspec{}
-or by data mining test suites); like Isabelle approaches have shown
-\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
-instantiated to a variety of similar values, in an attempt to find new theorems
-which are analogues of existing results from a different context. ``Mutating''
-existing theorem statements in such a way would also increase the chance of any
-result being considered interesting; since it's likely that the unmodified
-statement was deemed interesting, and the new result would not in general follow
-as a simple logical consequence.
-
-\subsection{Recurrent Clustering}
-\label{sec:clusteringexpressions}
-
-Our recurrent clustering algorithm is similar to that of ML4PG, as our
-transformation maps the elements of a syntax tree to distinct cells in a
-matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree
-elements: providing, for each level of the tree, the number of variables,
-nullary symbols, unary symbols, etc.
-
-Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it
-also ignores types. This is because obtaining the type of each component of a
-Core definition is more difficult than in ML4PG (which queries the current
-\textsc{Proof General} state), and is left for future work.
-
-The way we \emph{use} our clusters to inform theory exploration is actually more
-similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
-user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
-clusters are used to restrict the search space of a proof search, much like we
-restrict the scope of theory exploration.
-
-\section{Conclusion and Future Work}
-\label{sec:conclusion}
-
-Our use of clustering to pre-process \qspec{} signatures has required many
-decisions and tradeoffs to be made. Hence our approach is just one possibility
-out of many alternatives which could be investigated to push this work further.
-
-The most obvious next step is to incorporate types. Types contain valuable
-information about an expression, and would allow us to distinguish between
-constructors. Since our algorithm closely follows that of ML4PG, which
-\emph{does} support types, the only barrier is the practical issue of
-propagating annotations through all of the Core definitions.
-
-More speculative directions include the use of \emph{learned} representations,
-rather than our hand-crafted features \cite{bengio2013representation}. This
-would provide an interesting comparison, as well as being more robust in the
-face of language evolution. Another intriguing possibility is to extend the
-recurrent nature of our algorithm to make use of the discovered properties
-during clustering; for example, by treating the discovered equations as rewrite
-rules to reduce the ASTs prior to feature extraction.
-
-\appendix
-
-\section{Core Syntax}\label{sec:core}
-
-The GHC Core language is based on \fc{}, and described in detail in
-\cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are
-mostly interested in the syntax of reducible expressions (representing Haskell
-of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below
-in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
-
-\begin{equation}\label{eq:coresyntax}
-  \begin{split}
-    expr\    \rightarrow\ & \CVar\ id                          \\
-                       |\ & \CLit\ literal                     \\
-                       |\ & \CApp\ expr\ expr                  \\
-                       |\ & \CLam\ \mathcal{L}\ expr           \\
-                       |\ & \CLet\ bind\ expr                  \\
-                       |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                       |\ & \CType                             \\
-    id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                       |\ & \CGlobal\      \mathcal{G}         \\
-                       |\ & \CConstructor\ \mathcal{D}         \\
-    literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                       |\ & \CLitStr\ \mathcal{S}              \\
-    alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-    altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                       |\ & \CLitAlt\ literal                  \\
-                       |\ & \CDefault                          \\
-    bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                       |\ & \CRec\ [binder]                    \\
-    binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-  \end{split}
-\end{equation}
-Where:
-\begin{tabular}[t]{l @{ $=$ } l}
-  $\mathcal{S}$ & string literals    \\
-  $\mathcal{N}$ & numeric literals   \\
-  $\mathcal{L}$ & local identifiers  \\
-  $\mathcal{G}$ & global identifiers \\
-  $\mathcal{D}$ & constructor identifiers
-\end{tabular}
-
-The full Core language emitted by GHC (as of version 7.10.2, the latest at the
-time of writing) is translated automatically to this simplified form prior to
-recurrent clustering. Our major restriction is to ignore type-level entities
-(such as datatype definitions, explicit casts, and differences between
-types/kinds/coercions). Our implementation also handles several other forms of
-literal (machine words of various sizes, individual characters, etc.), but we
-omit them here for brevity as their treatment is similar to those of strings and
-numerals.
-
-\section{Rose Trees}\label{sec:rosetree}
-
-The $toTree$ function shown below transforms Core expressions, described in
-Appendix \ref{sec:core} to rose trees. We follow the presentation in
-\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is
-a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
-$T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the
-\emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where
-$n_T = 0$ are \emph{leaves}.
-
-The recursive definition is mostly routine; each repeated element (shown as
-$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
-$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
-have no internal structure for us to compare; they can only be compared based on
-their types, which we do not currently support. We also ignore values from
-$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
-and we conjecture that the effect of such ``magic values'' on clustering real
-code is low.
-
-%\begin{figure}
-  \begin{align*}\label{fig:totree}
-    toTree(e) &=
-    \begin{cases}
-      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
-      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
-      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
-      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
-      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
-      (\feature{\CType})                                                & \text{if $e = \CType$} \\
-      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
-      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
-      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
-      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
-      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
-      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
-      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
-      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
-      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
-      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
-      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
-    \end{cases}
-  \end{align*}
-
-%\end{figure}

### 2019-06-02 PhDThesis/evaluation_existingcode.tex
Fix metadata
Thesis updates

@@
 For a more realistic test we also applied recurrent clustering to a variety of
-the most-downloaded packages from \textsc{Hackage} (as of 2015-10-30), including
+the most-downloaded packages from \hackage{} (as of 2015-10-30), including
 \texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
@@
 It is currently infeasible to extend these clustering applications into full
-theory exploration runs, due to the use of constraints like \qcheck{}'s
-\hs{Arbitrary} type class (or LeanCheck's , in the case of Speculate) to
+theory exploration runs, due to the use of constraints like \quickcheck{}'s
+\hs{Arbitrary} type class (or \leancheck{}'s , in the case of \speculate{}) to
 generate suitable values for instantiating variables. Whilst we can
-automatically define \qspec{} theories and invoke them with \hs{nix-eval}, not
-all types have \hs{Arbitrary} instances; those without cannot be given any
+automatically define \quickspec{} theories and invoke them with \hs{nix-eval},
+not all types have \hs{Arbitrary} instances; those without cannot be given any
 variables in our signature, which severely limits the possible combinations
@@
 leaving just equations involving constants. This has so far prevented us from
-measuring the direct impact on \qspec{} performance on real code examples,
+measuring the direct impact on \quickspec{} performance on real code examples,
 either directly by exploring the sub-sets identified through recurrent
@@
 The following less-serious problems were also encountered while applying
-\textsc{ML4HS} to \textsc{Hackage} packages:
+\mlforhs{} to \hackage{} packages:
 
@@
   exports canonical, by pointing them to the private declarations. This forces
-  us to ignore such declarations, as \qspec{} will not be able to access them.
+  us to ignore such declarations, as \quickspec{} will not be able to access them.
 
@@
 theory exploration; the most na\"{\i}ve way being to explore each cluster as
-\qspec{} signature in its own right. \textsc{ML4HS} already provides this
+\quickspec{} signature in its own right. \mlforhs{} already provides this
 functionality, although the lack of test generators severely limits what can be
@@
 \emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
-produce a general ``scheme'' from each equation we find (either through \qspec{}
+produce a general ``scheme'' from each equation we find (either through \quickspec{}
 or by data mining test suites); like Isabelle approaches have shown

### 2019-06-02 PhDThesis/futurework.tex
Fix metadata of listings
Thesis updates

@@
 
-Our use of clustering to pre-process \qspec{} signatures has required many
+Our use of clustering to pre-process \quickspec{} signatures has required many
 decisions and tradeoffs to be made. Hence our approach is just one possibility
@@
 include them for non-local identifiers (i.e. globals and constructors), which we
-can look up in a database. In fact, our ML4HS framework already includes such
+can look up in a database. In fact, our \mlforhs{} framework already includes such
 type information in its database, alongside the Core syntax trees. Integrating
@@
 Our current approach is a rather conservative change to the existing theory
-exploration approaches, as it is essentially a wrapper around \qspec{}. There is
-potential for more radical changes to be made, which alter the search process
-itself.
+exploration approaches, as it is essentially a wrapper around \quickspec{}.
+There is potential for more radical changes to be made, which alter the search
+process itself.
 
@@
 
-\qcheck{} is certainly the most popular property checker for Haskell, which
-motivates its use in \qspec{} to instantiate variables to random
+\quickcheck{} is certainly the most popular property checker for Haskell, which
+motivates its use in \quickspec{} to instantiate variables to random
 values. However, this task of finding type inhabitants has also been solved in
-many other ways, which may be worth investigating in place of \qcheck{} (or
+many other ways, which may be worth investigating in place of \quickcheck{} (or
 perhaps even as part of an ensemble).
 
-The \textsc{SmallCheck} system \cite{runciman2008smallcheck} \emph{enumerates}
+The \smallcheck{} system \cite{runciman2008smallcheck} \emph{enumerates}
 values rather than sampling them randomly. Whilst this does not make
-\textsc{SmallCheck} objectively ``better'' than \qcheck{}, one major advantage
+\smallcheck{} objectively ``better'' than \quickcheck{}, one major advantage
 is that it may use much less memory, as the generated values are built up
-incrementally. In contrast, \qcheck{} may generate very large values; in
+incrementally. In contrast, \quickcheck{} may generate very large values; in
 particular, generating tree structures na\"{\i}vely can cause them to grow
@@
 
-A relative of \textsc{SmallCheck} is \textsc{Lazy SmallCheck}
-\cite{reich2013advances}, which uses laziness to only produce parts of a
-datastructure as they are demanded. This may narrow down our search procedures
-greatly, especially when predicates are involved. \qcheck{} allows predicates to
-restrict the values it tests with, and hence allows \emph{conditional} equations
-to be discovered. However, its implementation uses a simple rejection sampling
-technique: values are generated just as if the predicate were not there, and
-afterwards are filtered to reject any which do not satisfy the predicate. This
-makes it difficult to use very specific predicates, as it is unlikely that many
-of our random samples will exactly match our criteria. On the other hand,
-\textsc{Lazy SmallCheck} will focus its search on exactly those parts of the
-datastructure which are checked by the predicate, as those are the parts being
-forced to evaluate. This makes it much more likely that we will find values
-which satisfy the predicate, allowing us to effectively explore more specific
-conditional properties.
+A relative of \smallcheck{} is \lazysmallcheck{}~\cite{reich2013advances}, which
+uses laziness to only produce parts of a datastructure as they are demanded.
+This may narrow down our search procedures greatly, especially when predicates
+are involved. \quickcheck{} allows predicates to restrict the values it tests
+with, and hence allows \emph{conditional} equations to be discovered. However,
+its implementation uses a simple rejection sampling technique: values are
+generated just as if the predicate were not there, and afterwards are filtered
+to reject any which do not satisfy the predicate. This makes it difficult to use
+very specific predicates, as it is unlikely that many of our random samples will
+exactly match our criteria. On the other hand, \lazysmallcheck{} will focus its
+search on exactly those parts of the datastructure which are checked by the
+predicate, as those are the parts being forced to evaluate. This makes it much
+more likely that we will find values which satisfy the predicate, allowing us to
+effectively explore more specific conditional properties.
 
-Other approaches to generating inhabitants include \textsc{Djinn}
-\cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of
+Other approaches to generating inhabitants include
+\djinn{}~\cite{augustsson2005djinn}, which uses a decision procedure for a sub-set of
 Haskell types which in particular can generate and apply functions (unlike the
@@
 functions when they have been explicitly written in a
-generator). \textsc{MuCheck} \cite{le2014mucheck} is designed for
+generator). \mucheck{}~\cite{le2014mucheck} is designed for
 \emph{mutation testing}, and contains combinators for altering functions in
@@
 
-In fact, the database generated by our \textsc{AstPlugin} may prove helpful in
+In fact, the database generated by our \astplugin{} may prove helpful in
 generating values, since its type information can be fed to a tool like
-\textsc{Djinn} to discover chains of function applications for building values,
+\djinn{} to discover chains of function applications for building values,
 which would be particularly useful in cases where constructors are private, like
-in our email example. This is similar to the \textsc{Hoogle} tool, but also
-offers the ability to use dependency information to avoid potentially infinite
-recursion.
+in our email example. This is similar to the \hoogle{} tool, but also offers the
+ability to use dependency information to avoid potentially infinite recursion.
 
 The Core syntax trees in our database could also be used to generate theories
-for automated theorem provers. \hspec{} currently uses the GHC API to transform
+for automated theorem provers. \hipspec{} currently uses the GHC API to transform
 Core within its own process, however that approach suffers from the problems
-described in \S \ref{sec:astplugin}.
+described in \S~\ref{sec:astplugin}.
 
@@
 system: what to keep and what to discard, and even what areas of the search
-space to prioritise. \qspec{}'s approach, briefly mentioned in \S
-\ref{sec:theoryexploration}, is very simple: we discard equations which are
+space to prioritise. \quickspec{}'s approach, briefly mentioned in
+\S~\ref{sec:theoryexploration}, is very simple: we discard equations which are
 direct consequences of others, and keep all the rest. Different, and more
@@
 \cite{Wille:2005}, may be useful in tandem with our expression database, and
-the results of value generators like \qcheck{}.
+the results of value generators like \quickcheck{}.
 
-Since tools like \hspec{} already call out to third-party automated theorem
-provers, and indeed our own ML4HS system uses the external Weka program, there
+Since tools like \hipspec{} already call out to third-party automated theorem
+provers, and indeed our own \mlforhs{} system uses the external Weka program, there
 may also be merit in using external concept formation or conjecture generation
@@
 \cite{colton1999automatic, colton2000agent} could be used alongside those of
-\qspec{}.
+\quickspec{}.
 
@@
 powerful and therefore more \emph{interesting} and hence worth keeping. In fact
-this is very similar to \qspec{}'s interestingness criterion.
+this is very similar to \quickspec{}'s interestingness criterion.
 

### 2019-06-02 PhDThesis/litreview_automatedreasoning.tex
Fix references
Headingy things
Thesis updates

@@
-\subsection{Automated Mathematical Discovery and Invention}
+\section{Automated Mathematical Discovery and Invention}
 
 Automated logical and mathematical reasoning has been pursued since at least the
-time of mechanical calculators like the Pascaline~\cite{ocagne93:_le_calcul_simpl}. A
-recurring theme in these efforts is the separation between those undertaken by
-mathematicians like Pascal and Babbage~\cite{bowden}, and those of engineers
-such as M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the
-tasks we are concerned with (automatically constructing and evaluating concepts,
+time of mechanical calculators like the
+Pascaline~\cite{ocagne93:_le_calcul_simpl}. A recurring theme in these efforts
+is the separation between those undertaken by mathematicians like Pascal and
+Babbage~\cite{bowden}, and those of engineers such as
+M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
+are concerned with (automatically constructing and evaluating concepts,
 conjectures, theorems, axioms, examples, etc.) being divided into two main
@@
 Other related terms include ``Automated Mathematical
-Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
-``Concept Formation in Discovery Systems''~\cite{haase}, and
-``Automated Theorem Discovery''~\cite{roy}.
+Discovery''~\cite{epstein:91,colton2000notion,esarm2008}, ``Concept Formation in
+Discovery Systems''~\cite{haase}, and ``Automated Theorem
+Discovery''~\cite{roy}.
 
@@
 
-A similar ``layering'' approach is found in the \textsc{IsaScheme} system of
+A similar ``layering'' approach is found in the \isascheme{} system of
 Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
-has also been quantitatively compared against \textsc{IsaCoSy} and
-\textsc{HipSpec} using precision/recall analysis~\cite{claessen2013automating}.
+has also been quantitatively compared against \isacosy{} and \hipspec{} using
+precision/recall analysis~\cite{claessen2013automating}.
 The name comes from its embedding in the Isabelle proof assistant and its use of
@@
 
-\textsc{IsaCoSy} (Isabelle Conjecture Synthesis) is written for the Isabelle
+\isacosy{} (Isabelle Conjecture Synthesis) is written for the Isabelle
 proof assistant, mostly in
@@
 
-\textsc{QuickSpec} (the conjecture generation component of \textsc{HipSpec})
-emerged from work on the \textsc{QuickCheck} software testing framework for the
+\quickspec{} (the conjecture generation component of \hipspec{})
+emerged from work on the \quickcheck{} software testing framework for the
 Haskell programming language~\cite{claessen2011quickcheck}. As well as
-conjecturing properties of Haskell definitions for \textsc{HipSpec}, it has been
-applied to Isabelle definitions via the \textsc{Hipster} tool~\cite{Hipster}.
-Like \textsc{IsaCoSy}, \textsc{QuickSpec} takes a signature and enumerates
+conjecturing properties of Haskell definitions for \hipspec{}, it has been
+applied to Isabelle definitions via the \hipster{} tool~\cite{Hipster}.
+Like \isacosy{}, \quickspec{} takes a signature and enumerates
 well-typed terms. It collects together those of the same type into equivalence
-classes, assuming them to be equal, then uses \textsc{QuickCheck} to find
+classes, assuming them to be equal, then uses \quickcheck{} to find
 counterexamples to this assumption by randomly instantiating the variables and
@@
 using a constraint system to prevent redundancies (such as special-cases of
-other properties), \textsc{QuickSpec} instead passes its output through a
+other properties), \quickspec{} instead passes its output through a
 congruence closure algorithm to achieve the same effect.
 
-\textsc{QuickSpec} has since evolved to version 2~\cite{smallbone2017quick},
+\quickspec{} has since evolved to version 2~\cite{smallbone2017quick},
 which replaces the distinct enumeration and testing steps with a single,
-iterative algorithm similar to that of \textsc{IsaCoSy}. Generated conjectures
+iterative algorithm similar to that of \isacosy{}. Generated conjectures
 are fed into a Knuth-Bendix completion algorithm to form a corresponding set of
 rewrite rules. As expressions are enumerated, they are simplified using these
-rules and discarded if equal to a known expression. If not, \textsc{QuickCheck}
+rules and discarded if equal to a known expression. If not, \quickcheck{}
 tests whether the new expression can be distinguished from the known expressions
@@
 
-\textsc{QuickSpec} has also inspired another MTE tool for Haskell called
-\textsc{Speculate}~\cite{braquehais2017speculate}, which operates in a similar
+\quickspec{} has also inspired another MTE tool for Haskell called
+\speculate{}~\cite{braquehais2017speculate}, which operates in a similar
 way but also makes use of the laws of total orders and Boolean algebra to

### 2019-06-02 PhDThesis/litreview_interestingness.tex
Thesis updates

@@
     \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
-    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
-    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
-    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
-    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
+    \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
+    \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
+    \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
+    \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
   \end{tabular}
@@
 %
-% IsaCoSy uses counter-example checking; this ensures empirical plausibility
-% IsaCoSy uses constraints to avoid special-cases; this ensures novelty
-% IsaCoSy uses utility, since there are some hard-coded patterns which are
+% ICoSy uses counter-example checking; this ensures empirical plausibility
+% ICoSy uses constraints to avoid special-cases; this ensures novelty
+% ICoSy uses utility, since there are some hard-coded patterns which are
 % looked for
 %
-% Speculate uses a form of unification to ensure novelty, making sure one
+% Speculat uses a form of unification to ensure novelty, making sure one
 % equation is not a special-case of another
-% Speculate uses LeanCheck to enumerate values, looking for counterexamples
-% Speculate uses testing, to ensure Empirical Plausibility
+% Speculat uses LeanCheck to enumerate values, looking for counterexamples
+% Speculat uses testing, to ensure Empirical Plausibility
 %
-% QuickSpec uses QuickCheck to ensure empirical plausibility
-% QuickSpec uses a congruence closure algorithm to ensure novelty
+% QSpec uses QCheck to ensure empirical plausibility
+% QSpec uses a congruence closure algorithm to ensure novelty
 %
-% IsaScheme uses utility, since it determines the "quality" of a definition
+% IScheme uses utility, since it determines the "quality" of a definition
 % based on how many theorems it appears in?
-% IsaScheme uses novelty, using an equational rewrite system (Knuth-Bendix
+% IScheme uses novelty, using an equational rewrite system (Knuth-Bendix
 % completion) to remove redundancies
-% IsaScheme uses utility, since it focuses on the terms and definitions of a
+% IScheme uses utility, since it focuses on the terms and definitions of a
 % user's theory. However, don't all of them?
-% IsaScheme uses counterexample checking for empirical plausibility.
+% IScheme uses counterexample checking for empirical plausibility.
 
@@
 to some more recent tools in the remaining rows (based on our understanding of
-their function): \qspec{}~\cite{QuickSpec},
-Speculate~\cite{braquehais2017speculate},
-IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
-and IsaScheme~\cite{MontanoRivas2011}.
+their function): \quickspec{}~\cite{QuickSpec},
+\speculate{}~\cite{braquehais2017speculate},
+\isacosy{}~\cite{Johansson.Dixon.Bundy:conjecture-generation}
+and \isascheme{}~\cite{MontanoRivas2011}.
 
@@
 
-IsaCoSy ensures novelty using a constraint system to prevent special-cases being
-generated, whilst the others use post-hoc filters based on unification and term
-rewriting to achieve the same result. IsaCoSy is able to look for commonly
-desirable patterns, such as commutativity and associativity, which increases the
-utility of its output, whilst IsaScheme is completely based around such
-pattern-instantiation.
+\isacosy{} ensures novelty using a constraint system to prevent special-cases
+being generated, whilst the others use post-hoc filters based on unification and
+term rewriting to achieve the same result. \isacosy{} is able to look for
+commonly desirable patterns, such as commutativity and associativity, which
+increases the utility of its output, whilst \isascheme{} is completely based
+around such pattern-instantiation.
 
@@
 for one tool might not make sense for another, due to assumptions made about
-the algorithms' operation. For example, the novelty filters used by \qspec{}
-and Speculate are applied to the whole output set, guaranteeing that no
-special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
+the algorithms' operation. For example, the novelty filters used by \quickspec{}
+and \speculate{} are applied to the whole output set, guaranteeing that no
+special-cases will appear; yet we cannot assume this in the case of \isacosy{},
 since its constraint solver allows special cases if they're found \emph{before}

### 2019-06-02 PhDThesis/litreview_staticanalysis.tex
Headingy things
Thesis updates

@@
-\subsection{Automated Software Improvement}
+\section{Automated Software Improvement}
 
@@
 empirically by performing multiple runs, instantiating metavariables with
-synthetic data (\qcheck{} does this randomly, \textsc{SmallCheck} enumerates
+synthetic data (\quickcheck{} does this randomly, \smallcheck{} enumerates
 values, etc.).
@@
 when exploration is limited to equality properties (which is the case for
-\textsc{IsaCoSy} and \textsc{QuickSpec}). Quality metrics like efficiency (e.g.
+\isacosy{} and \quickspec{}). Quality metrics like efficiency (e.g.
 relative instruction count) can be considered a subset of the interestingness
@@
 discovery processes.
+
+% TODO: Evolutionary patches
+% TODO: Hutter search
+% TODO: Goedel machines

### 2019-06-02 PhDThesis/litreview_testing.tex
Headingy things
Thesis updates

@@
-\subsection{Property Checking}
+\section{Property Checking}
 
-Some prominent tools for mathematical theory exploration, like \qspec{}, were
-originally developed in the context of \emph{property checking} (although for
-\emph{discovering} software properties, rather than checking them), and only
+Some prominent tools for mathematical theory exploration, like \quickspec{},
+were originally developed in the context of \emph{property checking} (although
+for \emph{discovering} software properties, rather than checking them), and only
 later applied more directly to fields like formal mathematics (e.g. via the
 \hipster{} tool for Isabelle/HOL). In particular, the logical domain used by
-\qspec{} (inductive logic with higher-order functions) comes directly from its
-application to Haskell; and its search algorithm makes heavy use of the
-\qcheck{} property checker. The boundary between these two fields is hence
+\quickspec{} (inductive logic with higher-order functions) comes directly from
+its application to Haskell; and its search algorithm makes heavy use of the
+\quickcheck{} property checker. The boundary between these two fields is hence
 rather fuzzy, and cross-pollination of techniques is a compelling target for
@@
 
-\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
+\quickcheck{}~\cite{claessen2011quickcheck} popularised the use of
 property checking in Haskell, and has subsequently been ported to many other
-programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
+programming languages. \quickcheck{} generates input data \emph{randomly},
 either using explicit generators (via a function called \hs{forAll}), or by
@@
 understand \emph{why} a randomly generated value violated some property.
-\textsc{QuickCheck} tackles this in two ways: generators have access to a
+\quickcheck{} tackles this in two ways: generators have access to a
 ``size'' parameter, with the intention that this be used to decide how large the
@@
 
-The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
+The \smallcheck{} framework takes a different approach: \emph{enumerating}
 values (either using a user-provided enumeration function or using its
@@
 
-The \textsc{LazySmallCheck} framework is similar in spirit to
-\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
+The \lazysmallcheck{} framework is similar in spirit to
+\smallcheck{}, but utilises Haskell's lazy evaluation to avoid generating
 values up front~\cite{reich2013advances}. Instead, a partial value is generated
@@
 is an \hs{IO} action, pure properties which inspect such values will always
-be aborted, handing control back to \textsc{LazySmallCheck} which refines the
+be aborted, handing control back to \lazysmallcheck{} which refines the
 previous value and tries again. For example, a lazily generated \hs{Tree} may
@@
 
-\textsc{LazySmallCheck}'s automatic elaboration of free variables is similar in
+\lazysmallcheck{}'s automatic elaboration of free variables is similar in
 operation to the use of search procedures in logic programming languages like
-Kanren, Prolog (which has its own \textsc{QuickCheck} port) and Curry (a
+Kanren, Prolog (which has its own \quickcheck{} port) and Curry (a
 functional logic language, whose \textsc{EasyCheck} framework explicitly

### 2019-06-02 PhDThesis/ml4hs.tex
Thesis updates

@@
 We provide an implementation of our recurrent clustering algorithm in a tool
-called \textsc{ML4HS}, which consists of a loose collection of components shown
+called \mlforhs{}, which consists of a loose collection of components shown
 in Figure \ref{fig:ml4hs}. This arrangement makes it easy to swap out parts for
@@
   \begin{tikzpicture}[node distance=2cm]
-    \node [block] (hackage) {Hackage};
+    \node [block] (hackage) {\hackage{}};
 
@@
       \begin{tikzpicture}[node distance = 2cm, auto]
-        \node (mlspeclabel) {MLSpec};
-        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {QuickSpec};
+        \node (mlspeclabel) {\mlspec{}};
+        \node [block, anchor=north west, fill=white] at (mlspeclabel.south) (qs) {\quickspec{}};
       \end{tikzpicture}
@@
 reference, building our framework on top of GHC seems like a promising
-approach. Indeed, \hspec{} already invokes GHC's API to obtain the definitions
+approach. Indeed, \hipspec{} already invokes GHC's API to obtain the definitions
 of Haskell functions, in order to transform them into a form suitable for ATP
@@
 
-\subsection{\textsc{MLSpec}}
+\subsection{\mlspec{}}
 \label{sec:mlspec}
 
-We cannot supply these clusters as-is to \qspec{}, since it must be provided
-with a \emph{signature}. These are constructed by our \textsc{MLSpec} tool,
-using information from the \textsc{AstPlugin} database. Tasks performed by
-\textsc{MLSpec} include:
+We cannot supply these clusters as-is to \quickspec{}, since it must be provided
+with a \emph{signature}. These are constructed by our \mlspec{} tool,
+using information from the \astplugin{} database. Tasks performed by
+\mlspec{} include:
 
@@
     forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based
-  system like \qspec{} is unable to evaluate these expressions without
+  system like \quickspec{} is unable to evaluate these expressions without
   instantiating the variable \hs{t} to a specific type. Such an instantiation is
-  called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on
-  previous work in \qcheck{} by attempting to instantiate all type variables to
+  called \emph{monomorphising}, and in the case of \mlspec{} we build on
+  previous work in \quickcheck{} by attempting to instantiate all type variables to
   \hs{Integer}. We discard those cases where this is invalid, such as variable
@@
 
-\item{Variable definition}: Once a \qspec{} theory has been defined containing
+\item{Variable definition}: Once a \quickspec{} theory has been defined containing
   all of the given terms, we inspect the types it references and append three
   variables for each to the theory (enough to discover laws such as
-  associativity, but not too many to overflow the limit of \qspec{}'s exhaustive
+  associativity, but not too many to overflow the limit of \quickspec{}'s exhaustive
   search).
@@
   that all required packages and modules must be provided up-front, usually by
-  specification in a Cabal file. Since \textsc{MLSpec} builds signatures
+  specification in a Cabal file. Since \mlspec{} builds signatures
   \emph{dynamically}, depending on the cluster information it is given, we do
   not know what packages it may need. To work around this problem,
-  \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have
+  \mlspec{} invokes \quickspec{} for each cluster using a library we have
   built called \texttt{nix-eval}. This provides an \texttt{eval} function, like

### 2019-06-02 PhDThesis/related_tests.tex
Headingy things
Thesis updates

@@
 \subsection{Tests} \label{sec:tests}
+
+% TODO: This is not related work!
 
@@
 \emph{property checking}, which does give us useful data. Many Haskell property
-checkers exist, based on random testing (\textsc{QuickCheck}
-\cite{claessen2011quickcheck} and \textsc{SmartCheck}
-\cite{pike2014smartcheck}), enumeration (\textsc{SmallCheck}
-\cite{runciman2008smallcheck}), observation (\textsc{Lazy SmallCheck}
-\cite{reich2013advances}) and logic programming (\textsc{SparseCheck}
-\cite{sparsecheck}).
+checkers exist, based on random testing
+(\quickcheck{}~\cite{claessen2011quickcheck} and
+\smartcheck{}~\cite{pike2014smartcheck}), enumeration
+(\smallcheck{}~\cite{runciman2008smallcheck}), observation
+(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
+(\sparsecheck{}~\cite{sparsecheck}).
 
@@
 
-Here we consider the 10 most-downloaded packages from \textsc{Hackage} (as of
+Here we consider the 10 most-downloaded packages from \hackage{} (as of
 2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},

### 2019-06-03 PhDThesis/background_quickcheck.tex
Test quickcheck defs

@@
   where original 0 = 1
-        original n = n * naive (n - 1)
+        original n = n * original (n - 1)
 \end{haskell}
@@
 fact_base        = factorial 0 == factorial 1
+
+fact_increases, fact_div, fact_neg :: Int -> Bool  
 fact_increases n = factorial n <= factorial (n + 1)

### 2019-06-03 PhDThesis/clustering.tex
Add chapter

--- 2019-04-26 Bucketing/contributions.tex
@@
-\section{Contributions}
-\label{sec:contributions}
-
-\subsection{Recurrent Clustering}
+\chapter{Similarity and Clustering}
 \label{sec:recurrentclustering}
@@
 denote using the function $\phi$.
+
+\section{Core Syntax}\label{sec:core}
+
+The GHC Core language is based on \fc{}, and described in detail in
+\cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are
+mostly interested in the syntax of reducible expressions (representing Haskell
+of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below
+in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
+
+\begin{equation}\label{eq:coresyntax}
+  \begin{split}
+    expr\    \rightarrow\ & \CVar\ id                          \\
+                       |\ & \CLit\ literal                     \\
+                       |\ & \CApp\ expr\ expr                  \\
+                       |\ & \CLam\ \mathcal{L}\ expr           \\
+                       |\ & \CLet\ bind\ expr                  \\
+                       |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
+                       |\ & \CType                             \\
+    id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
+                       |\ & \CGlobal\      \mathcal{G}         \\
+                       |\ & \CConstructor\ \mathcal{D}         \\
+    literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
+                       |\ & \CLitStr\ \mathcal{S}              \\
+    alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
+    altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
+                       |\ & \CLitAlt\ literal                  \\
+                       |\ & \CDefault                          \\
+    bind\    \rightarrow\ & \CNonRec\ binder                   \\
+                       |\ & \CRec\ [binder]                    \\
+    binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
+  \end{split}
+\end{equation}
+Where:
+\begin{tabular}[t]{l @{ $=$ } l}
+  $\mathcal{S}$ & string literals    \\
+  $\mathcal{N}$ & numeric literals   \\
+  $\mathcal{L}$ & local identifiers  \\
+  $\mathcal{G}$ & global identifiers \\
+  $\mathcal{D}$ & constructor identifiers
+\end{tabular}
+
+The full Core language emitted by GHC (as of version 7.10.2, the latest at the
+time of writing) is translated automatically to this simplified form prior to
+recurrent clustering. Our major restriction is to ignore type-level entities
+(such as datatype definitions, explicit casts, and differences between
+types/kinds/coercions). Our implementation also handles several other forms of
+literal (machine words of various sizes, individual characters, etc.), but we
+omit them here for brevity as their treatment is similar to those of strings and
+numerals.
+
+\section{Rose Trees}\label{sec:rosetree}
+
+The $toTree$ function shown below transforms Core expressions, described in
+Appendix \ref{sec:core} to rose trees. We follow the presentation in
+\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is
+a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
+$T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the
+\emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where
+$n_T = 0$ are \emph{leaves}.
+
+The recursive definition is mostly routine; each repeated element (shown as
+$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
+$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
+have no internal structure for us to compare; they can only be compared based on
+their types, which we do not currently support. We also ignore values from
+$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
+and we conjecture that the effect of such ``magic values'' on clustering real
+code is low.
+
+%\begin{figure}
+  \begin{align*}\label{fig:totree}
+    toTree(e) &=
+    \begin{cases}
+      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
+      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
+      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
+      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
+      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
+      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
+      (\feature{\CType})                                                & \text{if $e = \CType$} \\
+      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
+      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
+      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
+      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
+      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
+      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
+      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
+      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
+      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
+      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
+      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
+      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
+    \end{cases}
+  \end{align*}
+
+%\end{figure}
 
@@
 
-Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$ are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match (others may match by coincidence, depending on $\phi$).
-These feature vectors might be deemed very dissimilar during clustering, despite the intuitive similarity of the expressions $X$ and $Y$ from which they derive.
+Here the features corresponding to the common sub-expression $\CLocal\ \id{y}$
+are misaligned, such that only $\frac{1}{3}$ of features are guaranteed to match
+(others may match by coincidence, depending on $\phi$). These feature vectors
+might be deemed very dissimilar during clustering, despite the intuitive
+similarity of the expressions $X$ and $Y$ from which they derive.
 
-If we were to align these features optimally, by padding the fourth column rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to match, making the similarity of the vectors more closely match our intuition and depend less on coincidence.
+If we were to align these features optimally, by padding the fourth column
+rather than the sixth, then $\frac{2}{3}$ of features would be guaranteed to
+match, making the similarity of the vectors more closely match our intuition and
+depend less on coincidence.
 
-The method we use to ``flatten'' expressions, described below, is a variation of breadth-first traversal which pads each level of nesting to a fixed size $c$ (for \emph{columns}).
-This doesn't guarantee alignment, but it does prevent mis-alignment from accumulating across different levels of nesting.
-Our method would align these features into the following vectors, if $c = 2$: \footnote{In fact, the constructor identifier $\id{C}$ would not appear in the vector, but this example is still accurate in terms of laying out the features as given.}
+The method we use to ``flatten'' expressions, described below, is a variation of
+breadth-first traversal which pads each level of nesting to a fixed size $c$
+(for \emph{columns}). This doesn't guarantee alignment, but it does prevent
+mis-alignment from accumulating across different levels of nesting. Our method
+would align these features into the following vectors, if $c = 2$: \footnote{In
+  fact, the constructor identifier $\id{C}$ would not appear in the vector, but
+  this example is still accurate in terms of laying out the features as given.}
 
@@
 
-Here $\frac{1}{2}$ of the original 6 features align, which is more than $breadthFirst$ but not optimal.
-Both vectors have also been padded by an extra 2 zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
+Here $\frac{1}{2}$ of the original 6 features align, which is more than
+$breadthFirst$ but not optimal. Both vectors have also been padded by an extra 2
+zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
 
-To perform this flattening we first transform the nested tokens of an expression into a \emph{rose tree} of features.
-Intuitively, these are simply an s-expression representation of the Core syntax, with the feature extraction function $\phi$ mapped over the labels.
-The full transformation is given in Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the result for the \hs{odd} function.
+To perform this flattening we first transform the nested tokens of an expression
+into a \emph{rose tree} of features. Intuitively, these are simply an
+s-expression representation of the Core syntax, with the feature extraction
+function $\phi$ mapped over the labels. The full transformation is given in
+Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the
+result for the \hs{odd} function.
 
@@
 
-\iffalse
-As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}.
-A valid topological ordering is given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding Core expressions to save space):
+As an example of this recurrent process, we can consider the Peano arithmetic
+functions from Figure \ref{fig:coreexample}. A valid topological ordering is
+given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
+Core expressions to save space):
 
@@
 concrete implementation, in particular to pin down values for the parameters
-such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.  \fi
+such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
+
+\section{Implementation}
+\label{sec:implementation}
+
+We provide an implementation of our recurrent clustering algorithm in a tool
+called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
+for the GHC compiler, which emits a serialised version of each Core definition
+as it is being compiled. This approach is more robust than, for example, parsing
+source files, since it avoids the complications of preprocessors and language
+extensions altering the syntax.
+
+A post-processing stage determines which Core definitions can be used by
+\quickspec{}, based on their type, visibility (whether they are encapsulated inside
+their module or visible to importers), etc. For each definition, we simply use
+the GHCi interpreter to check if calling \quickspec{} with that argument is
+well-typed. This information, along with type signatures, arity, etc. are stored
+alongside the Core definitions in JSON format.
+
+The definitions are then sorted topologically, based on the non-local
+identifiers appearing in their ASTs, and feature vectors are constructed using a
+Haskell implementation of the approach described in \S
+\ref{sec:recurrentclustering}. Since the features associated with each
+identifier may vary between iterations (as the clusters change), we leave the
+raw identifiers in the vector so their features can be extracted in the correct
+context.
+
+We implement Algorithm \ref{alg:recurrent} as a simple shell script, which
+invokes Weka for clustering and associates each Core definition with a cluster
+number. These numbers are used to finish the deferred feature extraction of
+identifiers, the resulting feature vectors are clustered, and the process is
+repeated until all SCCs have been processed.
+
+Once the recurrent clustering is complete, we generate a string of Haskell code
+for each of the final clusters, which will invoke \quickspec{} with a suitable
+signature. This involves:
+
+\begin{itemize}
+\item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall
+    t. t -> t -> Bool}, we must choose a concrete representation to use (in this
+  case for \hs{t}), in order to know which random generator to use. We take the
+  approach used in \quickcheck{} by attempting to instantiate all type variables to
+  \hs{Integer}. Any values where this is invalid, such as those with
+  incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where
+  \hs{Integer} does not implement \hs{IsString}) will not be included in the
+  signature (this is checked at the AST post-processing stage).
+
+\item{Qualifying}: All names are \emph{qualified} (prefixed by their module's
+  name), to avoid most ambiguity. There is still the possibility that multiple
+  packages will declare modules of the same name, although this is rare as it
+  causes problems for any Haskell programmer trying to use those modules. In
+  such cases the exploration process simply aborts.
+
+\item{Appending variables}: Once a \quickspec{} theory has been defined containing
+  all of the given terms, we inspect the types it references and append three
+  variables for each to the theory (enough to discover laws such as
+  associativity, but not too many to overflow the limit of \quickspec{}'s exhaustive
+  search).
+
+\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
+  that all required packages and modules must be provided up-front, usually by
+  specification in a Cabal file. Since \mlspec{} builds signatures
+  \emph{dynamically}, depending on the cluster information it is given, we do
+  not know what packages it may need. To work around this problem, we evaluate
+  these strings of Haskell using a custom library called \texttt{nix-eval}. This
+  uses the Nix package manager to obtain all of the required packages and make
+  them available to GHC, and is described further in
+  Appendix~\ref{sec:nix-eval}.
+
+\end{itemize}
+
+The equations resulting from evaluating these strings are collected and
+outputted as JSON values, to ease further processing (e.g. displaying in some
+form, sending to a theorem prover, etc.).
+
+\section{Evaluation}
+\label{sec:evaluation}
+
+% TODO: Move recurrent clustering to its own chapter, with this as its
+% evaluation?
+We have applied our recurrent clustering algorithm to several scenarios, with
+mixed results. A major difficulty in evaluating these clusters is that we have
+no ``ground truth'', i.e. there is no objectively correct way to compare
+expressions. Instead, we provide a qualitative overview of the more interesting
+characteristics.
+
+As a simple example, we clustered (Haskell equivalents of) the running examples
+used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
+\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
+implementations of several functions. We expect those with similar
+\emph{structure} to be clustered together, rather than those which implement the
+same function. The results are shown in Figure \ref{fig:haskellcluster}, where
+we can see the ``\hs{Tail}'' functions clearly distinguished, with little
+distinction between the tail recursive and na\"{\i}ve implementations.
+
+Next we tested whether these same functions would be clustered together when
+mixed with seemingly-unrelated functions, in this case 207 functions from
+Haskell's \hs{text} library. In fact, the \hs{helperFib} and \hs{fibTail}
+functions appeared together in a separate cluster from the rest. This was
+unexpected, with no obvious semantic connection between these two functions and
+the others in their cluster (although most are recursive, due to the nature of
+the \hs{text} library).
+
+\begin{figure}
+  \begin{common-lisp}
+(defun fact (n)
+  (if (zp n) 1 (* n (fact (- n 1)))))
+
+(defun helper-fact (n a)
+  (if (zp n) a (helper-fact (- n 1) (* a n))))
+
+(defun fact-tail (n)
+  (helper-fact n 1))
+
+(defun power (n)
+  (if (zp n) 1 (* 2 (power (- n 1)))))
+
+(defun helper-power (n a)
+  (if (zp n) a (helper-power (- n 1) (+ a a))))
+
+(defun power-tail (n)
+  (helper-power n 1))
+
+(defun fib (n)
+  (if (zp n)
+      0
+      (if (equal n 1)
+          1
+          (+ (fib (- n 1)) (fib (- n 2))))))
+
+(defun helper-fib (n j k)
+  (if (zp n)
+      j
+      (if (equal n 1)
+          k
+          (helper-fib (- n 1) k (+ j k)))))
+
+(defun fib-tail (n)
+  (helper-fib n 0 1))
+  \end{common-lisp}
+  \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
+  \label{fig:lisp}
+\end{figure}
+
+\begin{figure}
+  \begin{haskell}
+fact n = if n == 0
+            then 1
+            else n * fact (n - 1)
+
+helperFact n a = if n == 0
+                    then a
+                    else helperFact (n - 1) (a * n)
+
+factTail n = helperFact n 1
+
+power n = if n == 0
+             then 1
+             else 2 * power (n - 1)
+
+helperPower n a = if n == 0
+                     then a
+                     else helperPower (n - 1) (a + a)
+
+powerTail n = helperPower n 1
+
+fib n = if n == 0
+           then 0
+           else if n == 1
+                   then 1
+                   else fib (n - 1) + fib (n - 2)
+
+helperFib n j k = if n == 0
+                     then j
+                     else if n == 1
+                          then k
+                          else helperFib (n - 1) k (j + k)
+
+fibTail n = helperFib n 0 1
+  \end{haskell}
+  \caption{Haskell equivalents of the Common Lisp functions in Figure \ref{fig:lisp}.}
+  \label{fig:haskelllisp}
+\end{figure}
+
+\begin{figure}
+  \centering
+    \tikzstyle{block} = [rectangle, draw]
+    \tikzstyle{line}  = [draw, -latex']
+    \begin{tikzpicture}[node distance=3cm]
+      \node [block] (c1) {
+        \begin{tikzpicture}[node distance = 1cm]
+          \node [block]                    (factTail)  {\hs{factTail}};
+          \node [block, below of=factTail] (fibTail)   {\hs{fibTail}};
+          \node [block, below of=fibTail]  (powerTail) {\hs{powerTail}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=c1] (c2) {
+        \begin{tikzpicture}[node distance = 1cm]
+          \node [block]                (fact)       {\hs{fact}};
+          \node [block, below of=fact] (helperFact) {\hs{helperFact}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=c2] (c3) {
+        \begin{tikzpicture}[node distance = 1cm]
+          \node [block]                       (fib)         {\hs{fib}};
+          \node [block, below of=fib]         (helperFib)   {\hs{helperFib}};
+          \node [block, below of=helperFib]   (helperPower) {\hs{helperPower}};
+          \node [block, below of=helperPower] (power)       {\hs{power}};
+        \end{tikzpicture}
+      };
+  \end{tikzpicture}
+  \caption{Typical clusters for the functions in Figure \ref{fig:haskelllisp}.}
+  \label{fig:haskellcluster}
+\end{figure}
+
+We have also applied our recurrent clustering algorithm to a variety of the
+most-downloaded packages from \hackage{} (as of 2015-10-30), including
+\texttt{text} (as above), \texttt{pandoc}, \texttt{attoparsec},
+\texttt{scientific}, \texttt{yesod-core} and \texttt{blaze-html}. Whilst we
+expected functions with a similar purpose to appear together, such as the
+various reader and writer functions of \hs{pandoc}, there were always a few
+exceptions which became separate for reasons which are still unclear.
+
+When clustering the \hs{yesod} Web framework, the clustering did seem to match
+our intuitions, in particular since all 15 of Yesod's MIME type identifiers
+appeared in the same cluster.
+
+Whilst recurrent clustering has produced results which merit further
+investigation, the application to theory exploration has yet to be tested
+empirically. This is due to \quickspec{}'s use of \quickcheck{}'s \hs{Arbitrary} type
+class to generate random values for instantiating variables. Whilst we can
+automatically define \quickspec{} theories and invoke them with \hs{nix-eval}, not
+all types have \hs{Arbitrary} instances; those without cannot be given any
+variables in our signature, which severely limits the possible combinations
+which can be explored. In many cases, no variables can be included at all,
+leaving just equations involving constants. This has so far prevented us from
+measuring the direct impact on \quickspec{} performance, either directly by
+exploring the sub-sets identified through recurrent clustering, or indirectly by
+comparing the equations generated by a full brute-force search to our recurrent
+clusters: those equations relating terms from different clusters would not be
+discovered by our method. It is this ratio of equations found through brute
+force to those found after narrowing-down by clusters which is one of our key
+objectives to maximise at this stage; until we begin to pursue the
+\emph{interestingness} of the properties.
+
+The following less-serious problems were also encountered while applying
+\mlforhs{} to \hackage{} packages:
+
+\begin{itemize}
+\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
+  declarations to cluster. This is because they make all of their declarations
+  privately, e.g. in ``internal'' modules, then use separate modules to export
+  the public declarations. GHC's renaming phase makes all references to such
+  exports canonical, by pointing them to the private declarations. This forces
+  us to ignore such declarations, as \quickspec{} will not be able to access them.
+
+\item Since we do not support type-level entities, we ignore type
+  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
+  ``methods'') which occur in a type class instance. Instead of being clustered,
+  these result in references getting $f_{recursion}$ features. This is
+  especially noticable in libraries like \hs{scientific}, where only the
+  functions for constructing and destructing numbers in scientific notation are
+  clustered; all of the arithmetic is defined in type classes. One difficulty
+  with supporting methods is that their namespace in Core is disjoint from that
+  of regular Haskell identifiers: a transformation layer would be required,
+  along with explicit type annotations to avoid ambiguity.
+\end{itemize}
+
+It seems like this recurrent clustering method has promise, although it will
+require a more thorough exploration of the parameters to obtain more intuitively
+reliable results. These clusters can then be used in several ways to perform
+theory exploration; the most na\"{\i}ve way being to explore each cluster as
+\quickspec{} signature in its own right. \mlforhs{} already provides this
+functionality, although the lack of test generators severely limits what can be
+discovered.
+
+As alluded to previously, we also have the opportunity to reason by
+\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
+produce a general ``scheme'' from each equation we find (either through \quickspec{}
+or by data mining test suites); like Isabelle approaches have shown
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
+instantiated to a variety of similar values, in an attempt to find new theorems
+which are analogues of existing results from a different context. ``Mutating''
+existing theorem statements in such a way would also increase the chance of any
+result being considered interesting; since it's likely that the unmodified
+statement was deemed interesting, and the new result would not in general follow
+as a simple logical consequence.
+
+\subsection{Recurrent Clustering}
+\label{sec:clusteringexpressions}
+
+Our recurrent clustering algorithm is similar to that of ML4PG, as our
+transformation maps the elements of a syntax tree to distinct cells in a
+matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree
+elements: providing, for each level of the tree, the number of variables,
+nullary symbols, unary symbols, etc.
+
+Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it
+also ignores types. This is because obtaining the type of each component of a
+Core definition is more difficult than in ML4PG (which queries the current
+\textsc{Proof General} state), and is left for future work.
+
+The way we \emph{use} our clusters to inform theory exploration is actually more
+similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
+user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
+clusters are used to restrict the search space of a proof search, much like we
+restrict the scope of theory exploration.
+
+\section{Conclusion and Future Work}
+\label{sec:conclusion}
+
+Our use of clustering to pre-process \quickspec{} signatures has required many
+decisions and tradeoffs to be made. Hence our approach is just one possibility
+out of many alternatives which could be investigated to push this work further.
+
+The most obvious next step is to incorporate types. Types contain valuable
+information about an expression, and would allow us to distinguish between
+constructors. Since our algorithm closely follows that of ML4PG, which
+\emph{does} support types, the only barrier is the practical issue of
+propagating annotations through all of the Core definitions.
+
+More speculative directions include the use of \emph{learned} representations,
+rather than our hand-crafted features \cite{bengio2013representation}. This
+would provide an interesting comparison, as well as being more robust in the
+face of language evolution. Another intriguing possibility is to extend the
+recurrent nature of our algorithm to make use of the discovered properties
+during clustering; for example, by treating the discovered equations as rewrite
+rules to reduce the ASTs prior to feature extraction.

### 2019-06-05 PhDThesis/clustering.tex
Removed recurrent_clustering stuff into clustering

@@
 
-Rather than clustering Core expressions directly, we first perform \emph{feature
-  extraction} to transform them into numeric \emph{feature vectors}. We adapt
-the methodology of \emph{recurrent clustering} proposed in
-\cite{DBLP:journals/corr/HerasK14} \cite{heras2013proof}, which combines feature
-extraction and clustering into a single recursive algorithm, such that the
-feature vector of an expression depends on the feature vectors of those
-expressions it references. We suggest a new recurrent clustering and feature
-extraction algorithm for Haskell (shown in Algorithm \ref{alg:recurrent}) and
-compare its similarity and differences to those of ML4PG and ACL2(ml).
+To improve the performance of theory exploration tools, we need to avoid
+brute-force search in favour of a more informed strategy. This, in turn,
+requires more information about the terms in the given signature. One
+generally available source of information is a term's definition.
 
-We describe our algorithm in two stages: the first transforms the nested
-structure of expressions into a flat vector representation; the second converts
-the discrete symbols of Core syntax into features (real numbers), which we will
-denote using the function $\phi$.
+Unfortunately these definitions are over-complicated due to the relatively large
+syntax of Haskell, which includes many convenience features (``syntactic
+sugar'') and alternative ways to write essentially the same thing (e.g.
+\hs{let} vs \hs{where}, \hs{if} vs \hs{case}, \hs{f x = ...} vs
+\hs{f = \x -> ...}, etc.). This obscures the simplicity of the lambda calculus
+underlying the language, and more importantly it makes syntactic analysis less
+useful for our purposes by making distinctions between semantically identical
+terms (which will behave identically under theory exploration). For this reason
+we perform our analyses on the \emph{GHC Core} language instead.
 
-\section{Core Syntax}\label{sec:core}
+\section{GHC Core}\label{sec:core}
+\label{sec:ghccore}
 
-The GHC Core language is based on \fc{}, and described in detail in
-\cite[Appendix C]{sulzmann2007system}. For our machine learning purposes we are
-mostly interested in the syntax of reducible expressions (representing Haskell
-of the form \hs{f a b \dots = \dots}), and use the simplified syntax given below
-in BNF style ($[]$ and $(,)$ denote repetition and grouping, respectively):
-
-\begin{equation}\label{eq:coresyntax}
+\begin{figure}
+  \begin{equation*}
     \begin{split}
@@
     \end{split}
-\end{equation}
+  \end{equation*}
   Where:
@@
 
-The full Core language emitted by GHC (as of version 7.10.2, the latest at the
-time of writing) is translated automatically to this simplified form prior to
-recurrent clustering. Our major restriction is to ignore type-level entities
-(such as datatype definitions, explicit casts, and differences between
-types/kinds/coercions). Our implementation also handles several other forms of
-literal (machine words of various sizes, individual characters, etc.), but we
-omit them here for brevity as their treatment is similar to those of strings and
-numerals.
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
+    repetition and grouping, respectively. Other forms of literal, like machine
+    words, are treated in a similar way and have been omitted for brevity.}
+  \label{fig:coresyntax}
+\end{figure}
 
-\section{Rose Trees}\label{sec:rosetree}
+GHC Core is an intermediate language of the \textsc{GHC} compiler based on
+\fc{}, and described in detail in~\cite[Appendix C]{sulzmann2007system}. Core
+contains explicit type annotations and coercions, which we omit as they have no
+effect on runtime behaviour (they have no ``computational content''). The
+resulting sub-set of Core\footnote{As of GHC version 7.10.2.} is shown in
+Figure~\ref{fig:coresyntax}.
+
+\begin{figure}
+  \begin{subfigure}[haskell]{\textwidth}
+    \underline{\texttt{Haskell definitions}}
+    \begin{haskell}
+plus    Z  y = y
+plus (S x) y = S      (plus x y)
+
+mult    Z  y = Z
+mult (S x) y = plus y (mult x y)
+
+odd     Z  = False
+odd  (S n) = even n
+
+even    Z  = True
+even (S n) = odd n
+    \end{haskell}
+    \smallskip
+  \end{subfigure}
+  \begin{subfigure}[plus]{\textwidth}
+    \begin{small}
+      \underline{\texttt{plus}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       (Alt (DataAlt "Z") (Var (Local "y")))
+                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
+                                               (App (App (Var (Global "plus"))
+                                                         (Var (Local  "x")))
+                                                    (Var (Local "y"))))
+                                          "x")))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[mult]{\textwidth}
+    \begin{small}
+      \underline{\texttt{mult}}
+      \begin{verbatim}
+Lam "a" (Lam "y" (Case (Var (Local "a"))
+                       "b"
+                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
+                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
+                                                    (Var (Local  "y")))
+                                               (App (App (Var (Global "mult"))
+                                                         (Var (Local  "x")))
+                                                    (Var (Local  "y"))))
+                                          "x")))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[odd]{\textwidth}
+    \begin{small}
+      \underline{\texttt{odd}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              (Alt (DataAlt "Z") (Var (Constructor "False")))
+              (Alt (DataAlt "S") (App (Var (Global "even"))
+                                      (Var (Local  "n")))
+                                 "n"))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \begin{subfigure}[even]{\textwidth}
+    \begin{small}
+      \underline{\texttt{even}}
+      \begin{verbatim}
+Lam "a" (Case (Var (Local "a"))
+              "b"
+              (Alt (DataAlt "Z") (Var (Constructor "True")))
+              (Alt (DataAlt "S") (App (Var (Global "odd"))
+                                      (Var (Local  "n")))
+                                 "n"))
+      \end{verbatim}
+    \end{small}
+  \end{subfigure}
+  \caption{Example Haskell functions and their translations into the Core syntax
+    of Figure \ref{fig:coresyntax}. Notice the introduction of explicit
+    $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to
+    represent piecewise definitions. Fresh variables are chosen arbitrarily
+    as \texttt{"a"}, \texttt{"b"}, etc.}
+  \label{fig:coreexample}
+\end{figure}
+
+GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
+shows the Core representation of some simple functions. Although the Core is
+more verbose, we can see that similar structure in the Haskell definitions gives
+rise to similar structure in the Core; for example, the definitions of \hs{odd}
+and \hs{even} are identical in both languages, except for the particular
+identifiers used. It is this close correspondence which allows us to analyse
+Core expressions in place of their more complicated Haskell source.
+
+For simplicity we also avoid using constructors directly, choosing instead to
+wrap them with normal functions (e.g. \hs{s x = S x}, known as
+$\eta$-expansion), since this eliminates another distinction (constructors vs
+functions) and also reduces cross-language differences (e.g. Isabelle treats
+constructors in a different way to Haskell).
+
+\section{Feature Extraction}
+
+Raw syntax trees, even in Core form, are unsuitable for many forms of analysis
+due to their recursive structure, use of discrete labels and heavy reliance on
+references. These are unsuitable for many machine learning algorithms, which
+instead require a fixed-size vector of continuous numbers known as a
+\emph{feature vector}.
+
+The process of turning raw data into feature vectors is called \emph{feature
+  extraction}, and we adapt the methodology of \emph{recurrent clustering}
+proposed in~\cite{DBLP:journals/corr/HerasK14,heras2013proof}, since this
+handles recursive tree structures, resolves references and produces numeric
+vectors. We suggest a new recurrent clustering and feature extraction algorithm
+for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
+and differences to the existing approaches of ML4PG (Coq) and ACL2(ml) (ACL2).
+
+We describe our algorithm in two stages: the first transforms the nested
+structure of expressions into a flat vector representation; the second converts
+the discrete symbols of Core syntax into features (real numbers), which we will
+denote using the function $\phi$. Before introducing our algorithm, we briefly
+describe \emph{clustering}, and its use in \emph{recurrent clustering}.
+
+\subsection{Clustering}
+
+Clustering is an unsupervised machine learning task for grouping $n$ data points
+using a similarity metric. There are many variations on this theme, but in our
+case we make the following choices:
+
+\begin{itemize}
+\item For simplicity, we use the ``rule of thumb'' given in
+  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
+  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
+\item Data points will be $d$-dimensional feature vectors, as defined above.
+\item We will use euclidean distance (denoted $e$) as our similarity metric.
+\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
+  \cite{lloyd1982least}.
+\end{itemize}
+
+This is a standard setup, supported by off-the-shelf tools; in particular we use
+the \hPackage{k-means} Haskell package.
+
+Since k-means is iterative, we will use function notation to denote time steps,
+so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
+to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
+which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
+
+\begin{equation*}
+  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
+\end{equation*}
+
+Before k-means starts, we must choose \emph{seed} values for
+$\vect{m}^i(0)$. Many methods have been proposed for choosing these values
+\cite{arthur2007k}. For simplicity, we will choose values randomly from our
+data set $S$; this is known as the Forgy method.
+
+The elements of each cluster $C^i(t)$ are those data points closest to the mean
+value at the previous time step:
+
+\begin{equation*}
+  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
+\end{equation*}
+
+As $t$ increases, the clusters $C^i$ move from their initial location around the
+``seeds'', to converge on a local minimum of the ``within-cluster sum of squared
+error'' objective:
+
+\begin{equation} \label{eq:kmeansobjective}
+  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
+\end{equation}
+
+We chose k-means because it is a standard approach and we are more concerned
+with the application of feature extraction to Haskell and its use in theory
+exploration, rather than precise tuning of learning algorithms. There are many
+other clustering algorithms we could use, such as \emph{expectation
+  maximisation}~\footnote{In fact, k-means is very similar to
+  expectation-maximisation, as it alternates between an \emph{expectation step}
+  (finding the mean value of each cluster) and a \emph{maximisation step}
+  (assigning points to the cluster they're most similar to; or alternatively,
+  \emph{minimising} the distance of each point to the centre of its cluster, as
+  per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have
+shown little difference in their results; in effect, the quality of the features
+is the bottleneck to learning, so there is no reason to avoid a fast algorithm
+like k-means.
+
+In any case there are many conservative improvements to the standard k-means
+algorithm, which could be applied to our setup. For example, a more efficient
+approach like \emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make
+larger input sizes more practical to cluster, especially since recurrent
+clustering invokes k-means many times. The \emph{k-means++}
+approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
+select the ``seed'' values for the first timestep, and the \emph{x-means}
+algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use.
+
+\subsection{Recurrent Clustering}
+
+\emph{Recurrent clustering} has been implemented in the
+ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
+ACL2(ml) for ACL2~\cite{heras2013proof}. Whilst both transform syntax trees into
+matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements. In contrast,
+the matrices produced by ACL2(ml) \emph{summarise} information about the tree;
+for example, one column counts the number of variables appearing at each tree
+level, others count the number of function symbols which are nullary, unary,
+binary, etc. Whilst it may be interesting to contrast our current algorithm with
+an alternative based on that of ACL2(ml), it is unclear how such summaries could
+be extended to include types, which seems the next logical step for our
+approach. The ML4PG algorithm extends trivially, by using (term, type) pairs
+instead of just terms.
+
+The way we \emph{use} our clusters to inform theory exploration is actually more
+similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
+user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
+clusters are used to restrict the search space of a proof search, much like we
+restrict the scope of theory exploration.
+
+ACL2(ml) reasons by analogy: finding theorem statements which are similar to the
+current goal, and attempting to prove the goal in a similar way. In particular,
+the lemmas used to prove a theorem are mutated by substituting symbols for those
+which appear in the same cluster. For example, if \texttt{plus} and
+\texttt{multiply} are clustered together, and we are trying to prove a goal
+involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem
+involving \texttt{plus}. The lemmas used to prove that theorem will be mutated,
+for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an
+attempt to prove the goal.
+
+\iffalse
+% TODO: Move this to the end of the chapter?
+Whilst we do not currently reason by analogy, this is an interesting area for
+future work in theory exploration: given a set of theorems relating particular
+terms, we might form conjectures regarding similar terms found through
+clustering.
+\fi
+
+\subsection{Rose Trees}\label{sec:rosetree}
 
 The $toTree$ function shown below transforms Core expressions, described in
-Appendix \ref{sec:core} to rose trees. We follow the presentation in
-\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$ is
-a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
-$T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the
-\emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where
-$n_T = 0$ are \emph{leaves}.
+\S~\ref{sec:core}, to rose trees. We follow the presentation
+in~\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$
+is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
+$T_i$ are rose trees.
+
+$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
+
+$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
 
@@
 $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
-have no internal structure for us to compare; they can only be compared based on
-their types, which we do not currently support. We also ignore values from
+have no internal structure for us to compare. We also ignore values from
 $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
@@
 
-%\begin{figure}
+\begin{figure}
   \begin{align*}\label{fig:totree}
@@
   \end{align*}
-
-%\end{figure}
+\end{figure}
 

### 2019-06-05 PhDThesis/background_clustering.tex
Headingy things
Thesis updates
Background changes

@@
-Our approach to scaling up \qspec{} takes inspiration from two sources. The
+Our approach to scaling up \quickspec{} takes inspiration from two sources. The
 first is relevance filtering, which makes expensive algorithms used in theorem
 proving more practical by limiting the size of their inputs. We describe this
-approach in more details in \S \ref{sec:relevance}. Relevance filtering is a
+approach in more details in \S~\ref{sec:relevance}. Relevance filtering is a
 practical tool which has existing applications in software, such as the
@@
 brute force enumeration, such that related expressions are explored together.
-
-\subsection{K-Means}
-
-We use the Weka tool to perform k-means clustering
-\cite{Holmes.Donkin.Witten:1994}, since we are more concerned with the
-application of feature extraction to Haskell and its use in theory exploration,
-rather than precise tuning of learning algorithms. Since k-means is a standard
-method, there are many other implementations available. More interestingly,
-there are many other clustering algorithms we could use, such as
-\emph{expectation maximisation} \footnote{In fact, k-means is very similar to
-  expectation-maximisation, as it alternates between an \emph{expectation step}
-  (finding the mean value of each cluster) and a \emph{maximisation step}
-  (assigning points to the cluster they're most similar to; or alternatively,
-  \emph{minimising} the distance of each point to the centre of its cluster, as
-  per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have
-shown little difference in their results; in effect, the quality of our features
-is the bottleneck to learning, so there is no reason to avoid a fast algorithm
-like k-means.
-
-In any case there are many conservative improvements to the standard k-means
-algorithm, which could be applied to our setup. For example, a more efficient
-approach like \emph{yinyang k-means} \cite{conf/icml/DingZSMM15} could make
-larger input sizes more practical to cluster, especially since recurrent
-clustering invokes k-means many times. The \emph{k-means++} approach
-\cite{arthur2007k, bahmani2012scalable} can be used to more carefully select
-the ``seed'' values for the first timestep, and the \emph{x-means} algorithm
-\cite{pelleg2000x} is able to estimate how many clusters to use (our
-\emph{final} clusters should be tuned to maximise the performance of the
-subsequent theory exploration step, but x-means could still be useful in the
-recurrent clustering steps).
-
-\subsubsection{K-Means}
-\label{sec:kmeans}
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points
-using a similarity metric. There are many variations on this theme, but in our
-case we make the following choices:
-
-\begin{itemize}
-\item For simplicity, we use the ``rule of thumb'' given in
-  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
-  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
-\item Data points will be $d$-dimensional feature vectors, as defined above.
-\item We will use euclidean distance (denoted $e$) as our similarity metric.
-\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
-  \cite{lloyd1982least}.
-\end{itemize}
-
-This is a standard setup, supported by off-the-shelf tools; in particular we use
-the implementation provided by Weka \cite{Holmes.Donkin.Witten:1994}, due to
-its use by ML4PG, which makes our results more easily comparable.
-
-Since k-means is iterative, we will use function notation to denote time steps,
-so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
-to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
-which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation*}
-  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
-\end{equation*}
-
-Before k-means starts, we must choose \emph{seed} values for
-$\vect{m}^i(0)$. Many methods have been proposed for choosing these values
-\cite{arthur2007k}. For simplicity, we will choose values randomly from our
-data set $S$; this is known as the Forgy method.
-
-The elements of each cluster $C^i(t)$ are those data points closest to the mean
-value at the previous time step:
-
-\begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
-\end{equation*}
-
-As $t$ increases, the clusters $C^i$ move from their initial location around the
-``seeds'', to converge on a local minimum of the ``within-cluster sum of squared
-error'' objective:
-
-\begin{equation} \label{eq:kmeansobjective}
-  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation}

### 2019-06-05 PhDThesis/background_haskell.tex
Background changes

@@
 
-\begin{figure}
-  \begin{equation*}
-    \begin{split}
-      expr\    \rightarrow\ & \CVar\ id                          \\
-                         |\ & \CLit\ literal                     \\
-                         |\ & \CApp\ expr\ expr                  \\
-                         |\ & \CLam\ \mathcal{L}\ expr           \\
-                         |\ & \CLet\ bind\ expr                  \\
-                         |\ & \CCase\ expr\ \mathcal{L}\ [alt]   \\
-                         |\ & \CType                             \\
-      id\      \rightarrow\ & \CLocal\       \mathcal{L}         \\
-                         |\ & \CGlobal\      \mathcal{G}         \\
-                         |\ & \CConstructor\ \mathcal{D}         \\
-      literal\ \rightarrow\ & \CLitNum\ \mathcal{N}              \\
-                         |\ & \CLitStr\ \mathcal{S}              \\
-      alt\     \rightarrow\ & \CAlt\ altcon\ expr\ [\mathcal{L}] \\
-      altcon\  \rightarrow\ & \CDataAlt\ \mathcal{D}             \\
-                         |\ & \CLitAlt\ literal                  \\
-                         |\ & \CDefault                          \\
-      bind\    \rightarrow\ & \CNonRec\ binder                   \\
-                         |\ & \CRec\ [binder]                    \\
-      binder   \rightarrow\ & \CBind\ \mathcal{L}\ expr
-    \end{split}
-  \end{equation*}
-  Where:
-  \begin{tabular}[t]{l @{ $=$ } l}
-    $\mathcal{S}$ & string literals    \\
-    $\mathcal{N}$ & numeric literals   \\
-    $\mathcal{L}$ & local identifiers  \\
-    $\mathcal{G}$ & global identifiers \\
-    $\mathcal{D}$ & constructor identifiers
-  \end{tabular}
-
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
-    repetition and grouping, respectively.}
-  \label{fig:coresyntax}
-\end{figure}
-
 Further evidence of Haskell's suitability for theory exploration is given by the
@@
 to Haskell and invoking \hipspec{}~\cite{claessen2013automating}.
-
-\subsection{GHC Core}
-\label{sec:ghccore}
-
-Whilst our systems and experiments use normal Haskell code, for simplicity we
-perform some of our analyses on an intermediate representation of the
-\textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively
-large and complex syntax of Haskell proper. Core is based on \fc{}, which is
-described in detail in~\cite[Appendix C]{sulzmann2007system}.
-
-Core contains explicit type annotations and coercions, which we omit as they
-have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
-  of GHC version 7.10.2.} is shown in Figure~\ref{fig:coresyntax}; for brevity,
-we also omit several other forms of literal (machine words of various sizes,
-individual characters, etc.), since their treatment is similar to those of
-strings and numerals. We use quoted strings to denote names and literals,
-e.g. \hs{Local "foo"}, \hs{Global "bar"}, \hs{Constructor "Baz"}, \hs{LitStr
-  "quux"} and \hs{LitNum "42"}, and require only that they can be compared for
-equality.
-
-\begin{figure}
-  \begin{subfigure}[plus]{\textwidth}
-    \begin{small}
-      \underline{\texttt{plus}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Local "y")))
-                       (Alt (DataAlt "S") (App (Var (Constructor "S"))
-                                               (App (App (Var (Global "plus"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[mult]{\textwidth}
-    \begin{small}
-      \underline{\texttt{mult}}
-      \begin{verbatim}
-Lam "a" (Lam "y" (Case (Var (Local "a"))
-                       "b"
-                       (Alt (DataAlt "Z") (Var (Constructor "Z")))
-                       (Alt (DataAlt "S") (App (App (Var (Global "plus"))
-                                                    (Var (Local  "y")))
-                                               (App (App (Var (Global "mult"))
-                                                         (Var (Local  "x")))
-                                                    (Var (Local  "y"))))
-                                          "x")))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[odd]{\textwidth}
-    \begin{small}
-      \underline{\texttt{odd}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "False")))
-              (Alt (DataAlt "S") (App (Var (Global "even"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \begin{subfigure}[even]{\textwidth}
-    \begin{small}
-      \underline{\texttt{even}}
-      \begin{verbatim}
-Lam "a" (Case (Var (Local "a"))
-              "b"
-              (Alt (DataAlt "Z") (Var (Constructor "True")))
-              (Alt (DataAlt "S") (App (Var (Global "odd"))
-                                      (Var (Local  "n")))
-                                 "n"))
-      \end{verbatim}
-    \end{small}
-  \end{subfigure}
-  \caption{Translations of functions in Figure \ref{fig:haskellexample} into the
-    Core syntax of Figure \ref{fig:coresyntax}. Notice the introduction of
-    explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case}
-    to represent piecewise definitions. Fresh variables are chosen arbitrarily
-    as \texttt{"a"}, \texttt{"b"}, etc.}
-  \label{fig:coreexample}
-\end{figure}
-
-GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
-shows the Core representation of functions from Figure~\ref{fig:haskellexample}.
-Although the Core is more verbose, we can see that similar structure in the
-Haskell definitions gives rise to similar structure in the Core; for example,
-the definitions of \hs{odd} and \hs{even} are identical in both languages,
-except for the particular identifiers used. It is this close correspondence
-which allows us to analyse Core expressions in place of their more complicated
-Haskell source.
-
-Note that we exclude representations for type-level entities, including datatype
-definitions like that of \hs{Nat}. GHC can represent these, but we are mostly
-concerned with dynamic behaviour, which excludes types (they are \emph{erased}
-during compilation). For simplicity we also avoid using constructors directly,
-choosing instead to wrap them with normal functions (e.g. \hs{s x = S x}), since
-this makes our domain more symmetric and reduces cross-language differences.

### 2019-06-05 PhDThesis/bucketing.tex
Bucketing changes

@@
 all rely on enumerating combinations of definitions in some way, which can cause
-infeasible running times on larger inputs (based on results of the Theory
-Exploration Benchmark in \S~\ref{sec:benchmark}, this happens above a few dozen
+infeasible running times on larger inputs (based on experiments with our Theory
+Exploration Benchmark,%in \S~\ref{sec:benchmark}
+this happens above a few dozen
 definitions). To avoid this, users of these systems must carefully select only
@@
   \item Define the \emph{signature selection problem} for theory exploration.
-  \item Analyse the performance of theory exploration on the Theory Exploration
-    Benchmark.
+  \item Analyse system performance on the Theory Exploration Benchmark, to
+    demonstrate the effect of ``toxic'' definitions leading to timeouts.
   \item Analyse the effect of signature selection on the quantity of desirable
@@
 other hand, functions such as HTTP parsers and spell-checkers will not be
-related in many interesting ways; exploring their combinations is wasteful.
+related in many interesting ways, so exploring their combinations is wasteful.
 
-The signature selection problem is that of selecting sub-sets of a large
+The signature selection problem is that of choosing sub-sets of a large
 signature, such that:
@@
 \item Few subsets are selected
-\item Each sub-set is small
-\item As many of the total interesting relationships are still findable
+\item Each sub-set can be explored quickly
+\item Many of the interesting properties of the original signature can be
+  found in the union of properties of the subsets
 \end{itemize}
 
-\section{Clustering}
-\label{sec:clustering}
+There is a clear trade-off to be made between speed (the first two requirements)
+and ``quality'' (the final requirement): we can find all of the interesting
+properties of a signature if we explore the whole thing at once; whilst it
+is fastest to explore no definitions at all. In between are opportunities for a
+Pareto-optimal balance to be struck. Whilst the number of subsets can be
+determined simply by counting, the others require more thorough investigation.
 
-Our approach to scaling up \quickspec{} takes inspiration from two sources. The
-first is relevance filtering, which makes expensive algorithms used in theorem
-proving more practical by only considering clauses deemed ``relevant'' to the
-problem \cite{meng2009lightweight}.
+\section{Exploration Speed}
 
-Relevance is determined by comparing clauses to the target theorem, but theory
-exploration does not have such a distinguished term. Instead, we are interested
-in relationships between \emph{all} terms in a signature, and hence we need a
-different algorithm for considering the relevance of \emph{all terms} to
-\emph{all other terms}.
-
-A natural fit for this task is \emph{clustering}, which attempts to group
-similar inputs together in an unsupervised way. Based on their success in
-discovering relationships and patterns between expressions in Coq and ACL2 (in
-the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
-methods can fulfil the role of relevance filters for theory exploration:
-intelligently breaking up large signatures into smaller ones more amenable to
-brute force enumeration, such that related expressions are explored together.
-
-Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering,
-implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's
-algorithm \cite{lloyd1982least}, with randomly-selected input elements as the
-initial clusters. Rather than relying on the user to provide the number of
-clusters $k$, we use the ``rule of thumb'' given in
-\cite[pp. 365]{mardia1979multivariate} of clustering $n$ data points into
-$k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
-
-\section{Experiments}
-
-% TODO: This was taken from the Background section, so probably doesn't fit here
-% without some fiddling.
-% TODO: Describe all of our experimental setups:
-%
-%  - The use of the TEBenchmark methodology for running QSpec, including the
-%    sampling process
-%
-%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
-%    including how we picked the sizes
-%
-%  - The definition of HashSpec and its use as control; how we ran all of the
-%    same samples through it
-%
-%  - The use of optimal and pessimal oracles, defined via constraint
-%    satisfaction. How we it was only feasible to run these up to size 10 (11?)
-
-We follow the tradition of prior practitioners and use an existing corpus of
-theorems as a \emph{ground truth} against which to compare results. To analyse
-behaviour on large inputs (i.e. those for which exploration is currently
-infeasible), as required to study the effects of signature selection, we have
-used the Theory Exploration Benchmark of section~\ref{sec:tebenchmark}, which is
-based on the Tons of Inductive Problems (TIP) problem set for theorem
-provers~\cite{claessen2015tip}.
+The time taken to explore a subset of definitions from a signature can be easily
+measured, but only in hindsight. Choosing each subset must be done with only a
+\emph{prediction} of how long it will take to explore. Here we investigate the
+time taken by \quickspec{} on signatures chosen from our Theory Exploration
+Benchmark (containing definitions from Tons of Inductive Problems (TIP)
+benchmark~\cite{claessen2015tip}), and how this is affected by their content.
 
 \begin{figure}
+  \centering
   \scalebox{0.45}{\input{images/steppedall.pgf}}
@@
 
-Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{} running
-times, when given inputs containing different numbers of definitions. Many runs
+Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{}
+running times, when given inputs containing different numbers of definitions. Many runs
 finish quickly, with the remainder occupying a ``long tail'' which we cut off
@@
 
-The number of definitions in the input is linearly correlated with the amount of
-timeouts, except for the very smallest inputs (which are more constrained by the
-sampling procedure). One explanation for this is ``toxic'' definitions, whose
-presence in the input always leads to the exploration failing (with larger
-inputs being more likely to have sampled a toxic definition).
-
-\iffalse
 \begin{figure}
@@
 \end{figure}
-\fi
+
+Figure~\ref{fig:tailsize} shows that the number of definitions in the input is
+linearly correlated with the proportion of timeouts, except for the very
+smallest inputs (which are more constrained by the sampling procedure). One
+hypothesis to explain this is the existence of ``toxic'' definitions, whose
+presence in the input always leads to the exploration failing (with larger
+inputs being more likely to have sampled a toxic definition).
 
@@
   \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
-  \label{fig:proportions}
   \caption{Definitions, ordered by the ratio of successes to failures of the
@@
     without any toxic definitions.}
+  \label{fig:proportions}
 \end{figure}
@@
 
-Exploring each of these functions on its own does not require much memory, since
-Haskell generates the output lazily. However, comparing such large numbers for
-equality takes a lot of CPU time as the \texttt{S} constructors are successively
-unwrapped from each side, and this is why the timeout is reached. We confirmed
-this hypothesis by exploring with a custom data generator which only generates
-the values \texttt{Z}, \texttt{S Z} and \texttt{S (S Z)} (0, 1 and 2); this
-caused the exploration to finish quickly. Other interventions, like making the
-accumulator arguments strict (to prevent space leaks), did not prevent timeouts.
+Note that Haskell is lazily evaluated, which allows large structures like Peano
+numerals to be compared using little memory: the \texttt{S} constructors are
+successively unwrapped from each side, being calculated on-demand and
+immediately garbage collected in a tight loop. However, this still takes a lot
+of CPU time and hence causes timeouts. We confirmed this hypothesis by exploring
+with a custom data generator which only generates the values \texttt{Z},
+\texttt{S Z} and \texttt{S (S Z)} (0, 1 and 2), which caused the exploration to
+finish quickly in these cases. Other interventions, like making the accumulator
+arguments strict (to prevent space leaks), did not prevent timeouts.
 
-To assess the impact of these problematic definitions, we removed any samples
-containing them and repeated our analysis.
+To assess the impact of these toxic definitions, we removed any samples
+containing them and repeated our analysis: the results are shown on the right
+hand side of the previous figures. Whilst timeouts remain, they are
+substantially reduced, and completely eliminated for the smallest sample sizes.
+Some definitions never appeared in a failing signature.
+
+Our analysis of signature contents on theory exploration performance leads us to
+two recommendations for signature selection methods:
+
+\begin{itemize}
+\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
+  and/or taking many arguments (leading to a large number of combinations). In
+  particular, exposing the accumulator argument of tail-recursive functions can
+  make them harder to explore.
+\item Smaller sets are faster to explore than larger sets, but the
+\end{itemize}
+
+It also seems prudent for theory exploration tools to limit time and memory
+usage at the level of individual evaluations: these can abort evaluation of
+toxic definitions, whilst still being high enough to allow the majority of
+non-toxic definitions to be unhindered.
+
+\section{Maintaining Quality}
+
+Our final criterion concerns the ``quality'' of the selected subsets. This
+depends crucially on how we define whether a property is ``interesting''. We can
+use the same ground truth definition as our Theory Exploration Benchmark: a
+property is ``interesting'' if it appears in a particular corpus (in our case
+TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
+approach allows the criterion to be quantified: the quality of the selected
+subsets is proportion of the whole signature's interesting properties that also
+apply to the subsets.
+
+Since signature selection is largely independent of which tool we choose to
+explore the resulting subsets, we do not require that such exploration actually
+\emph{finds} all of the interesting properties; only that we have not
+\emph{prevented} them from being found. For example, we may be given a signature
+containing addition and multiplication functions and may consider commutativity
+and distributivity properties to be ``interesting''. Selecting a subset
+containing both would be high quality, since all of these properties are
+\emph{available} to find, even if our chosen tool doesn't spot them. In
+contrast, splitting addition and multiplication into \emph{separate} subsets is
+lower quality, since this prevents \emph{any} tool finding the distributivity
+property (since it involves both definitions).
+
+To understand the quality impact of limiting ourselves to such subsets, we
+applied four signature selection methods to signatures sampled from the Theory
+Exploration Benchmark. Signatures varied in size from a single definition up to
+100 definitions and 100 signatures of each size were sampled. Each method was
+run 19 times on each signature: splitting into between 1 and 20 subsets.
+
+All four methods place each definition in exactly one subset, to focus on the
+effect of splitting apart definitions. The tested methods are:
+
+\begin{itemize}
+\item \textbf{Optimal}: Uses knowledge of the ground truth to choose equal sized
+  subsets with the highest quality.
+\item \textbf{Pessimal}: Like optimal but chooses the lowest quality subsets.
+\item \textbf{Pseudorandom}: Assigns definitions to subsets pseudorandomly, used
+  as a control.
+\item \textbf{Clustering}: Subsets are chosen by a recurrent clustering
+  algorithm, based on their syntactic similarity.
+\end{itemize}
+
+The optimal and pessimal methods are unrealistic, since they require access to
+the ground truth (which only exists for benchmark problems), but set upper and
+lower bounds on the impact of separating a signature's definitions. These are
+implemented via constraint solving, whose exponential running time became
+infeasible for signatures containing more than 10 definitions. To avoid extreme
+solutions, such as all-but-one subsets being singletons, subset cardinality
+could only differ by at most one.
+
+The pseudorandom and clustering implementations are more realistic, since they
+do not use any information from the ground truth. Cardinality limits were also
+not imposed for these methods.
+
+% TODO: Insert graph of hash vs MLSpec on 1-100
+
+% TODO: Insert graph of 1-10 with all
+
+\iffalse
+% TODO
+\section{Clustering}
+\label{sec:clustering}
+
+Our approach to scaling up \quickspec{} takes inspiration from two sources. The
+first is relevance filtering, which makes expensive algorithms used in theorem
+proving more practical by only considering clauses deemed ``relevant'' to the
+problem~\cite{meng2009lightweight}.
+
+Relevance is determined by comparing clauses to the target theorem, but theory
+exploration does not have such a distinguished term. Instead, we are interested
+in relationships between \emph{all} terms in a signature, and hence we need a
+different algorithm for considering the relevance of \emph{all terms} to
+\emph{all other terms}.
+
+A natural fit for this task is \emph{clustering}, which attempts to group
+similar inputs together in an unsupervised way. Based on their success in
+discovering relationships and patterns between expressions in Coq and ACL2 (in
+the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
+methods can fulfil the role of relevance filters for theory exploration:
+intelligently breaking up large signatures into smaller ones more amenable to
+brute force enumeration, such that related expressions are explored together.
+
+Due to its use by ML4PG and ACL2(ml), we use \emph{k-means} clustering,
+implemented in the Weka tool \cite{Holmes.Donkin.Witten:1994} by Lloyd's
+algorithm \cite{lloyd1982least}, with randomly-selected input elements as the
+initial clusters. Rather than relying on the user to provide the number of
+clusters $k$, we use the ``rule of thumb'' given in
+\cite[pp. 365]{mardia1979multivariate} of clustering $n$ data points into
+$k = \lceil \sqrt{\frac{n}{2}} \rceil$ clusters.
+\fi
+
+%\section{Experiments}
+
+% TODO: This was taken from the Background section, so probably doesn't fit here
+% without some fiddling.
+% TODO: Describe all of our experimental setups:
+%
+%  - The use of the TEBenchmark methodology for running QSpec, including the
+%    sampling process
+%
+%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
+%    including how we picked the sizes
+%
+%  - The definition of HashSpec and its use as control; how we ran all of the
+%    same samples through it
+%
+%  - The use of optimal and pessimal oracles, defined via constraint
+%    satisfaction. How we it was only feasible to run these up to size 10 (11?)
+
+%We follow the tradition of prior practitioners and use an existing corpus of
+%theorems as a \emph{ground truth} against which to compare results. To analyse
+%behaviour on large inputs (i.e. those for which exploration is currently
+%infeasible), as required to study the effects of signature selection, we have
+%used the Theory Exploration Benchmark of section~\ref{sec:tebenchmark}, which is
+%based on the Tons of Inductive Problems (TIP) problem set for theorem
+%provers~\cite{claessen2015tip}.

### 2019-06-05 PhDThesis/support-for-benchmarks/mkBenchmarkGraphs.py
Suppress warnings properly

@@ -237,6 +237,8 @@ del(data)
 
 msg('Setting up plots')
 
+import warnings
+warnings.filterwarnings('ignore', module='matplotlib')
 import matplotlib as mpl
 import numpy      as np
 

### 2019-06-05 PhDThesis/support-for-bucketing/boundsData.py
Bounds graphs

@@ -1,0 +1,176 @@
+#!/usr/bin/env python3
+import json
+import sys
+
+msg = lambda x: (sys.stderr.write(repr(x)), sys.stderr.flush(), None)[-1]
+
+# Input format:
+#   {
+#     size: {
+#       rep: {
+#         "proportions": {
+#           bucketCount: {
+#             "min"/"max": {
+#               "proportion": number
+#   ...
+data = json.load(sys.stdin)
+
+# Average over the reps
+average  = lambda l: sum(l) / len(l)
+averaged = {
+    size: {
+        bucketCount: {
+            bound: average([
+                rep["proportions"][bucketCount][bound]["proportion"] \
+                for rep in reps.values()
+            ]) for bound in ["min", "max"]
+        } for bucketCount in reps["1"]["proportions"].keys()  # "1" is arbitrary
+    } for (size,reps) in data.items()
+}
+
+# Lines for our chart
+def getLines(bound):
+    return {
+        bucketCount: [
+            averaged[size][bucketCount][bound]           \
+            for size in sorted(averaged.keys(), key=int) \
+            if bucketCount in averaged[size]
+        ] for bucketCount in averaged["10"].keys()
+    }
+minima = getLines("min")
+maxima = getLines("max")
+del getLines
+
+# Show hash line for context
+import os
+hashprops = None
+with open(os.getenv('proportionsTsv'), 'r') as f:
+    # method, size, bucketCount, mean, stddev
+    rows = [line.strip().split('	') for line in f.readlines() \
+                                           if line.startswith('hashed')]
+
+    sizes        = sorted(list(set([row[1] for row in rows])), key=int)
+    bucketCounts = sorted(list(set([row[2] for row in rows])), key=int)
+
+    hashprops = {
+        bucketCount: [
+            float([row[3] for row in rows if row[1] == size and         \
+                                             row[2] == bucketCount][0]) \
+            for size in sizes
+        ] for bucketCount in bucketCounts
+    }
+
+# Based on https://python-graph-gallery.com/125-small-multiples-for-line-chart
+import warnings
+warnings.filterwarnings('ignore', module='matplotlib')
+import matplotlib.pyplot as plt
+import numpy             as np
+import pandas            as pd
+
+# Inspired by https://stackoverflow.com/a/29107972/884682
+def suplabels(x=None, y=None):
+    ''' Add super ylabel or xlabel to the figure
[…]

### 2019-06-14 PhDThesis/bucketing.tex
Bucketing changes in thesis

@@
 Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{}
-running times, when given inputs containing different numbers of definitions. Many runs
-finish quickly, with the remainder occupying a ``long tail'' which we cut off
-after 300 seconds\footnote{Chosen based on preliminary experiments, which showed
-  little difference in survival between 300 seconds and 1 hour}.
+running times, when given inputs containing different numbers of definitions.
+Many runs finish quickly, with the remainder occupying a ``long tail'' which
+didn't finish within our 300 second timeout\footnote{Chosen based on preliminary
+  experiments, which showed little difference in survival between 300 seconds
+  and 1 hour}.
 
@@
 
-Figure~\ref{fig:tailsize} shows that the number of definitions in the input is
-linearly correlated with the proportion of timeouts, except for the very
-smallest inputs (which are more constrained by the sampling procedure). One
+The height of these ``tails'' is linearly correlated with the number of
+definitions in the signature, as shown in Figure~\ref{fig:tailsize}. One
 hypothesis to explain this is the existence of ``toxic'' definitions, whose
@@
 
-This does appear to be the case, as Figure~\ref{fig:proportions} shows that five
-definitions appeared \emph{only} in failing inputs; these are named
-\texttt{mul3}, \texttt{mul3acc}, \texttt{mult2}, \texttt{op} and \texttt{qexp};
-their definitions appear in Figure~\ref{fig:faildefs}. All of these are
-functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
-exploration to time out by either exhausting the RAM with too many expressions
-to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
+To investigate the plausibility of such definitions, the ratio of successful and
+timed-out runs is plotted for each name in Figure~\ref{fig:proportions}; showing
+that five definitions appeared \emph{only} in failing inputs, and seem likely to
+be ``toxic''. These definitions are named \texttt{mul3}, \texttt{mul3acc},
+\texttt{mult2}, \texttt{op} and \texttt{qexp}; their definitions appear in
+Figure~\ref{fig:faildefs}. All of these are functions of Peano-encoded natural
+numbers (\texttt{Nat}), and they cause exploration to time out by either
+exhausting the RAM with too many expressions to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
 or by generating such deeply-nested outputs that comparing them takes too long.
@@
   and/or taking many arguments (leading to a large number of combinations). In
-  particular, exposing the accumulator argument of tail-recursive functions can
-  make them harder to explore.
-\item Smaller sets are faster to explore than larger sets, but the
+  particular, tail-recursive functions which expose their accumulator argument
+  are harder to explore.
+\item Smaller signatures are faster to explore, so should be preferred if
+  possible.
 \end{itemize}
@@
 It also seems prudent for theory exploration tools to limit time and memory
-usage at the level of individual evaluations: these can abort evaluation of
-toxic definitions, whilst still being high enough to allow the majority of
-non-toxic definitions to be unhindered.
+usage, both at the global level (to abort if it becomes clear we are in the
+``long tail'') and at the level of individual evaluations. In particular such
+fine-grained limits can abort evaluation of toxic definitions, whilst still
+being lenient enough to allow the majority of non-toxic definitions to be
+explored unhindered.
 
@@
 approach allows the criterion to be quantified: the quality of the selected
-subsets is proportion of the whole signature's interesting properties that also
-apply to the subsets.
+subsets is the proportion of the whole signature's interesting properties that
+also apply to the subsets taken individually.
 
@@
 containing addition and multiplication functions and may consider commutativity
-and distributivity properties to be ``interesting''. Selecting a subset
-containing both would be high quality, since all of these properties are
+and distributivity properties to be ``interesting''. A subset containing both
+would be a high quality selection, since all of these properties are
 \emph{available} to find, even if our chosen tool doesn't spot them. In
 contrast, splitting addition and multiplication into \emph{separate} subsets is
-lower quality, since this prevents \emph{any} tool finding the distributivity
+lower quality, since this stops \emph{any} tool from finding the distributivity
 property (since it involves both definitions).
@@
 Exploration Benchmark. Signatures varied in size from a single definition up to
-100 definitions and 100 signatures of each size were sampled. Each method was
-run 19 times on each signature: splitting into between 1 and 20 subsets.
+100; for each size, we sampled 100 signatures.
 
-All four methods place each definition in exactly one subset, to focus on the
-effect of splitting apart definitions. The tested methods are:
+All four methods were forced to place each definition in exactly one subset, to
+focus on the effect of splitting apart definitions rather than e.g. duplicating
+definitions across several subsets or dropping some entirely. The tested methods
+are:
 
@@
 infeasible for signatures containing more than 10 definitions. To avoid extreme
-solutions, such as all-but-one subsets being singletons, subset cardinality
-could only differ by at most one.
+solutions, such as placing all definitions in one subset and leaving the others
+empty, we required that all of the selected subsets are non-empty and their
+cardinality differs by at most one. As a consequence, these methods cannot
+divide signatures of size $N$ into more than $N$ subsets (since there are not
+enough definitions to populate every subset without duplicating).
 
 The pseudorandom and clustering implementations are more realistic, since they
-do not use any information from the ground truth. Cardinality limits were also
-not imposed for these methods.
+do not use any information from the ground truth. These methods may also leave
+some of their subsets empty (since, unlike the optimal and pessimal methods,
+they are less prone to extremes).
 
-% TODO: Insert graph of hash vs MLSpec on 1-100
+\begin{figure}
+  % \input{images/bounds.pgf}
+  \includegraphics[width=\textwidth]{images/bounds.png}
+  \caption{The average quality (proportion of ground-truth properties available)
+    when splitting signatures sampled using the Theory Exploration Benchmark
+    into subsets. Each plot shows division into a different number of subsets.
+    The shaded area is bounded by the optimal and pessimal methods, which are
+    can only produce $N$ subsets for signatures of size $\geq N$. Grey lines
+    show pseudorandom selection, for comparison; these extend into smaller
+    signature sizes since empty subsets are permitted.}
+  \label{fig:bounds}
+\end{figure}
 
-% TODO: Insert graph of 1-10 with all
+Figure~\ref{fig:bounds} shows the impact of selection (specifically, dividing
+definitions between subsets, with no skipping or duplication) on the proportion
+of interesting properties that remain available for subsequent theory
+exploration tools to discover. The shaded region shows the difference between
+(average) quality, with optimal and pessimal selections starting out equal (when
+dividing signatures of $N$ definitions into $N$ subsets, where there is no
+choice other than putting each definition on its own) but quickly diverging as
+choices become available. Note that the optimal and pessimal methods are
+constrained to avoid empty or unequally-distributed sets, so they cannot divide
+up a signature of size $N$ into more than $N$ subsets. Pseudorandom selection is
+shown as a grey line, and although the cardinality of its subsets is not
+constrained, it nevertheless appears consistently between the bounds.
+
+Two trends are visible in these charts: firstly the bounded region trends
+upwards as the signature size increases, which is promising for the application
+of signature selection to break apart large signatures. Secondly the upper bound
+reduces as the number of subsets increases: this is to be expected as more
+definitions must be separated to produce the extra sets, but it demonstrates the
+tradeoff between quality and subset size. The distance between the pseudorandom
+and optimal results shows the potential gains to be made by smarter signature
+selection methods.
+
+\begin{figure}
+  % \input{images/bucketing}
+  \includegraphics[width=\textwidth]{images/bucketing.png}
+  \caption{Comparison of subset quality when selected by recurrent clustering
+    (solid lines) versus pseudorandom selection (dashed lines). Darker lines
+    show division into a greater number of subsets (between 1 and 20). The solid
+    and dashed lines follow each other closely, indicating that clustering
+    performs no better than random at signature selection.}
+  \label{fig:bucketing}
+\end{figure}
+
+The signature selection problem is superficially similar to the
+\emph{clustering} problem in machine learning: grouping data points based on
+their similarity across multiple dimensions. To investigate this potential
+relationship we created a signature selection method based on a \emph{recurrent
+  clustering} algorithm for Haskell code: this flattens syntax trees into
+vectors, substitutes fixed numbers for keywords and recurses to determine
+suitable numbers for referenced expressions.
+
+The results of this method are shown in Figure~\ref{fig:bucketing}, along with
+the pseudorandom method as a control. The results of clustering turn out to be
+no better than random; in fact the curves show surprisingly similar patterns.
 

### 2019-06-14 PhDThesis/support-for-benchmarks/mkBenchmarkGraphs.py
Import textWidth, to avoid rebuilding
textWidth was a derivation, which depended on the contents of the .tex
files. Editing the .tex files would cause textWidth to rebuild, which would
cause everything that used textWidth to rebuild (e.g. our graphs).

Now we import the textWidth derivation, which gives us a string. Changing the
contents of the .tex files will cause that string to be recalculated, but the
things which depend on it (e.g. our graphs) will not get rebuilt if the string's
value hasn't changed (which it probably won't).

@@ -243,8 +243,7 @@ import matplotlib as mpl
 import numpy      as np
 
 def figSize(widthFraction, height=None):
-    with open(os.getenv('textWidth'), 'r') as textWidthFile:
-        textWidthPt = float(textWidthFile.read())
+    textWidthPt = float(os.getenv('textWidth'))
     ptToInch    = 1.0 / 72.27
     textWidthIn = textWidthPt * ptToInch
     goldMean    = (np.sqrt(5.0)-1.0) / 2.0

### 2019-06-18 PhDThesis/clustering.tex
Lots of clustering changes

@@
 requires more information about the terms in the given signature. One
-generally available source of information is a term's definition.
+generally available source of information is the syntax of a term's definition.
 
@@
 syntax of Haskell, which includes many convenience features (``syntactic
-sugar'') and alternative ways to write essentially the same thing (e.g.
-\hs{let} vs \hs{where}, \hs{if} vs \hs{case}, \hs{f x = ...} vs
-\hs{f = \x -> ...}, etc.). This obscures the simplicity of the lambda calculus
-underlying the language, and more importantly it makes syntactic analysis less
-useful for our purposes by making distinctions between semantically identical
-terms (which will behave identically under theory exploration). For this reason
-we perform our analyses on the \emph{GHC Core} language instead.
+sugar'') and alternative ways to express the same operation (e.g. binding using
+\hs{let} or \hs{where}, pattern matching with \hs{if} or \hs{case}, function
+abstraction via \hs{f x = ...} or \hs{f = \x -> ...}, etc.). This obscures the
+simplicity of the lambda calculus underlying the language, and more importantly
+it makes syntactic analysis less useful for our purposes by drawing distinctions
+between semantically identical terms. Since theory exploration is concerned only
+with semantics (e.g. telling us how code actually behaves), we perform our
+analyses on the syntax of the \emph{GHC Core} language instead.
 
@@
 GHC Core is an intermediate language of the \textsc{GHC} compiler based on
-\fc{}, and described in detail in~\cite[Appendix C]{sulzmann2007system}. Core
-contains explicit type annotations and coercions, which we omit as they have no
-effect on runtime behaviour (they have no ``computational content''). The
-resulting sub-set of Core\footnote{As of GHC version 7.10.2.} is shown in
+\fc{}, and described in detail in~\cite[Appendix C]{sulzmann2007system}. We
+ignore type annotations and coercions, since these have no effect on the
+semantics (they have no ``computational content''). The resulting sub-set of
+Core\footnote{As of GHC version 7.10.2.} is shown in
 Figure~\ref{fig:coresyntax}.
@@
 GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
-shows the Core representation of some simple functions. Although the Core is
-more verbose, we can see that similar structure in the Haskell definitions gives
-rise to similar structure in the Core; for example, the definitions of \hs{odd}
-and \hs{even} are identical in both languages, except for the particular
-identifiers used. It is this close correspondence which allows us to analyse
-Core expressions in place of their more complicated Haskell source.
-
-For simplicity we also avoid using constructors directly, choosing instead to
-wrap them with normal functions (e.g. \hs{s x = S x}, known as
-$\eta$-expansion), since this eliminates another distinction (constructors vs
-functions) and also reduces cross-language differences (e.g. Isabelle treats
-constructors in a different way to Haskell).
+shows the Core representation of some simple functions. Although Core is more
+verbose, it has fewer redundant features than the full Haskell syntax: for
+example all pattern-matching is represented with \texttt{Case}, all function
+abstraction uses \texttt{Lam}(bda), and so on. We can see that similar structure
+in the Haskell definitions gives rise to similar structure in the Core, such as
+the definitions of \hs{odd} and \hs{even} being identical except for the
+identifiers. It is this close correspondence which allows us to analyse Core
+expressions in place of their more complicated Haskell source.
+
+As a further simplification, we also replace data constructors in definitions
+with normal functions, e.g. replacing occurrences of a constructor
+\hs{S :: Nat -> Nat} with a function \hs{s} defined as \hs{s x = S x} (this is
+known as the $\eta$-expansion of \hs{S}). We do this to remove more distinctions
+from the syntax, and also to reduce cross-language differences (in particular,
+the Isabelle language used by the \isascheme{}, \isacosy{} and \hipster{} theory
+exploration systems treats constructors in a different way to Haskell, so having
+constructors in our data would complicate comparisons).
 
@@
 
-Raw syntax trees, even in Core form, are unsuitable for many forms of analysis
-due to their recursive structure, use of discrete labels and heavy reliance on
-references. These are unsuitable for many machine learning algorithms, which
-instead require a fixed-size vector of continuous numbers known as a
+Raw syntax trees, even in Core form, are difficult to analyse due to their
+recursive structure, use of discrete labels and heavy reliance on references.
+This makes them unsuitable for many machine learning algorithms, which instead
+require a fixed-size vector of continuous numbers known as a
 \emph{feature vector}.
@@
 handles recursive tree structures, resolves references and produces numeric
-vectors. We suggest a new recurrent clustering and feature extraction algorithm
+values. We suggest a new recurrent clustering and feature extraction algorithm
 for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
-and differences to the existing approaches of ML4PG (Coq) and ACL2(ml) (ACL2).
+and differences to the existing approaches of \textsc{ML4PG} and
+\textsc{ACL2(ml)}.
 
@@
 
-This is a standard setup, supported by off-the-shelf tools; in particular we use
-the \hPackage{k-means} Haskell package.
+We choose k-means because it is a standard approach and we are more concerned
+with the application of feature extraction to Haskell and its use in theory
+exploration, rather than precise tuning of learning algorithms. There are many
+other clustering algorithms we could use, such as \emph{expectation
+  maximisation}, but experiments with \textsc{ML4PG} have shown little
+difference in their results; it seems that the quality of the features is the
+bottleneck to learning, so there is no reason to avoid a fast algorithm like
+k-means. We use a Haskell implementation provided by the package
+\hPackage{k-means}.
 
@@
 \begin{equation*}
-  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|} \quad \text{for $t > 0$}
+  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|}
+    \quad \text{for $t > 0$}
 \end{equation*}
@@
 \begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x}, \vect{m}^j(t-1)) \} \quad \text{for $t > 0$}
+  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x},
+                                                        \vect{m}^j(t-1)) \}
+    \quad \text{for $t > 0$}
 \end{equation*}
@@
 
-We chose k-means because it is a standard approach and we are more concerned
-with the application of feature extraction to Haskell and its use in theory
-exploration, rather than precise tuning of learning algorithms. There are many
-other clustering algorithms we could use, such as \emph{expectation
-  maximisation}~\footnote{In fact, k-means is very similar to
-  expectation-maximisation, as it alternates between an \emph{expectation step}
-  (finding the mean value of each cluster) and a \emph{maximisation step}
-  (assigning points to the cluster they're most similar to; or alternatively,
-  \emph{minimising} the distance of each point to the centre of its cluster, as
-  per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have
-shown little difference in their results; in effect, the quality of the features
-is the bottleneck to learning, so there is no reason to avoid a fast algorithm
-like k-means.
-
-In any case there are many conservative improvements to the standard k-means
-algorithm, which could be applied to our setup. For example, a more efficient
-approach like \emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make
-larger input sizes more practical to cluster, especially since recurrent
-clustering invokes k-means many times. The \emph{k-means++}
+Many conservative improvements can be made to the standard k-means algorithm
+described above. For example, a more efficient approach like
+\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
+more practical to cluster; the \emph{k-means++}
 approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
-select the ``seed'' values for the first timestep, and the \emph{x-means}
-algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use.
-
-\subsection{Recurrent Clustering}
+select the ``seed'' values for the first timestep; and the \emph{x-means}
+algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
+(instead of relying on our ``rule of thumb''). However, like the choice of
+k-means versus other clustering algorithms, these algorithmic improvements would
+need higher quality data than we are capable of producing in these early
+experiments.
+
+\subsubsection{Recurrent Clustering}
 
@@
 ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
-ACL2(ml) for ACL2~\cite{heras2013proof}. Whilst both transform syntax trees into
-matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements. In contrast,
-the matrices produced by ACL2(ml) \emph{summarise} information about the tree;
-for example, one column counts the number of variables appearing at each tree
-level, others count the number of function symbols which are nullary, unary,
-binary, etc. Whilst it may be interesting to contrast our current algorithm with
-an alternative based on that of ACL2(ml), it is unclear how such summaries could
-be extended to include types, which seems the next logical step for our
-approach. The ML4PG algorithm extends trivially, by using (term, type) pairs
-instead of just terms.
-
-The way we \emph{use} our clusters to inform theory exploration is actually more
-similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
-user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
-clusters are used to restrict the search space of a proof search, much like we
-restrict the scope of theory exploration.
-
-ACL2(ml) reasons by analogy: finding theorem statements which are similar to the
-current goal, and attempting to prove the goal in a similar way. In particular,
-the lemmas used to prove a theorem are mutated by substituting symbols for those
-which appear in the same cluster. For example, if \texttt{plus} and
-\texttt{multiply} are clustered together, and we are trying to prove a goal
-involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem
-involving \texttt{plus}. The lemmas used to prove that theorem will be mutated,
-for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an
-attempt to prove the goal.
-
-\iffalse
-% TODO: Move this to the end of the chapter?
-Whilst we do not currently reason by analogy, this is an interesting area for
-future work in theory exploration: given a set of theorems relating particular
-terms, we might form conjectures regarding similar terms found through
-clustering.
-\fi
-
-\subsection{Rose Trees}\label{sec:rosetree}
-
-The $toTree$ function shown below transforms Core expressions, described in
-\S~\ref{sec:core}, to rose trees. We follow the presentation
-in~\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$
-is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
-$T_i$ are rose trees.
-
-$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
-
-$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
-
-The recursive definition is mostly routine; each repeated element (shown as
-$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
-$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
-have no internal structure for us to compare. We also ignore values from
-$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
-and we conjecture that the effect of such ``magic values'' on clustering real
-code is low.
-
-\begin{figure}
-  \begin{align*}\label{fig:totree}
-    toTree(e) &=
-    \begin{cases}
-      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
-      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
-      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
-      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
-      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
-      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
-      (\feature{\CType})                                                & \text{if $e = \CType$} \\
-      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
-      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
-      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
-      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
-      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
-      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
-      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
-      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
-      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
-      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
-      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
-      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
-    \end{cases}
-  \end{align*}
-\end{figure}
-
-\subsubsection{Expressions to Vectors}
+\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
+syntax trees into fixed-size matrices, then concatenate the rows of these
+matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
+\emph{summarising} information about the tree; for example, one column counts
+the number of variables appearing at each tree level, others count the number of
+function symbols which are nullary, unary, binary, etc. In contrast, the
+algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
+arranging them in the matrix based on their position in the syntax tree. It is
+this latter algorithm that we adapt for analysing GHC Core expressions.
+
+The key to recurrent clustering is its treatment of identifiers. From a purely
+symbolic perspective, they are atomic: two identifiers can either be identical
+or distinct. Consider the \hs{odd} and \hs{even} functions in
+Figure~\ref{fig:coreexample}, which differ only in their choice of identifiers.
+These functions are clearly different (opposite, in fact), so their identifiers
+should play a role in determining their similarity. Yet these distinctions, e.g.
+between \hs{True} and \hs{False}, seem intuitively small relative to the
+distinction between, for example, \hs{isEmpty} and \hs{isValidEmailAddress}.
+
+Recurrent clustering solves this dichotomy by introducing a finer granularity to
+the treatment of identifiers, by first clustering the definitions that they
+refer to. If two definitions appear in the same cluster, their identifiers will
+be denoted by similar numbers during subsequent feature extraction; if they
+appear in different clusters, their numbers will differ more during feature
+extraction. It is this use of clustering to determine features that makes the
+approach \emph{recurrecnt}.
+
+\subsection{Expressions to Vectors}
 \label{sec:expressionstovectors}
 
-Our recurrent clustering algorithm is based on k-means clustering, which
-considers the elements of a feature vector to be \emph{orthogonal}. Hence we
-must ensure that similar expressions not only give rise to similar numerical
-values, but crucially that these values appear \emph{at the same position} in
-the feature vectors. Since different patterns of nesting can alter the ``shape''
-of expressions, simple traversals (breadth-first, depth-first, post-order, etc.)
+The first step of our feature extraction algorithm is to convert the tree
+structure of Core expressions to a flat vector. Since k-means clustering
+considers the elements of a feature vector to be \emph{orthogonal}, we must
+ensure that similar expressions not only give rise to similar numerical values,
+but crucially that these values appear \emph{at the same position} in the
+feature vectors. Different patterns of nesting can alter the ``shape'' of
+expressions, so simple traversals (breadth-first, depth-first, post-order, etc.)
 may cause features from equivalent sub-expressions to be mis-aligned. For
-example, consider the following expressions, which represent pattern-match
+example, consider the following Core expressions, which represent pattern-match
 clauses with different patterns but the same body (\hs{\vlocal{y}}):
@@
 If we traverse these expressions in breadth-first order, converting each token
-to a feature using $\phi$ and padding to the same length with $0$, we would get
-the following feature vectors:
+to a feature using some function $\phi$ (defined later) and padding to the same
+length with $0$, we would get the following feature vectors:
 
@@
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      breadthFirst(X)\ &=\ (\feature{\CAlt},\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\ &\feature{\CLocal},\ &\feature{\id{y}} &) \\
-      breadthFirst(Y)\ &=\ (\feature{\CAlt},\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &)
+      breadthFirst(X)\ &=\ (\feature{\CAlt},\
+                           &\feature{\CDataAlt},\
+                           &\feature{\CVar},\
+                           &\feature{\id{C}},\
+                           &\feature{\CLocal},\
+                           &\feature{\id{y}} &) \\
+      breadthFirst(Y)\ &=\ (\feature{\CAlt},\
+                           &\feature{\CDefault},\
+                           &\feature{\CVar},\
+                           &\feature{\CLocal},\
+                           &\feature{\id{y}},\
+                           &0 &)
     \end{array}
@@
 mis-alignment from accumulating across different levels of nesting. Our method
-would align these features into the following vectors, if $c = 2$: \footnote{In
-  fact, the constructor identifier $\id{C}$ would not appear in the vector, but
-  this example is still accurate in terms of laying out the features as given.}
+would align these features into the following vectors, if $c = 2$:
 
@@
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
-      featureVec(X)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDataAlt},\ &\feature{\CVar},\ &\feature{\id{C}},\  &\feature{\CLocal},\ &\feature{\id{y}},\ &0 &) \\
-      featureVec(Y)\ &=\ (\feature{\CAlt},\ &0,\ &\feature{\CDefault},\ &\feature{\CVar},\ &\feature{\CLocal},\ &0,\                 &\feature{\id{y}},\ &0 &)
+      featureVec(X)\ &=\ (\feature{\CAlt},\
+                         &0,\
+                         &\feature{\CDataAlt},\
+                         &\feature{\CVar},\
+                         &\feature{\id{C}},\
+                         &\feature{\CLocal},\
+                         &\feature{\id{y}},\
+                         &0 &) \\
+      featureVec(Y)\ &=\ (\feature{\CAlt},\
+                         &0,\
+                         &\feature{\CDefault},\
+                         &\feature{\CVar},\
+                         &\feature{\CLocal},\
+                         &0,\
+                         &\feature{\id{y}},\
+                         &0 &)
     \end{array}
@@
 zeros compared to $breadthFirst$; raising their alignment to $\frac{5}{8}$.
-
-To perform this flattening we first transform the nested tokens of an expression
-into a \emph{rose tree} of features. Intuitively, these are simply an
-s-expression representation of the Core syntax, with the feature extraction
-function $\phi$ mapped over the labels. The full transformation is given in
-Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the
-result for the \hs{odd} function.
 
@@
 
+To perform this flattening we first transform the nested tokens of an expression
+into a \emph{rose tree} of features, via the function $toTree$ defined below.
+Intuitively, rose trees are simply an s-expression representation of the Core
+syntax, with the feature extraction function $\phi$ mapped over the labels:
+Figure~\ref{fig:rosetreeexample} shows the result of transforming the \hs{odd}
+function from Figure~\ref{fig:coreexample}.
+
+We follow the presentation in~\cite{blundell2012bayesian} and define rose trees
+recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$,
+where $f \in \mathbb{R}$ and $T_i$ are rose trees.
+
+$T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$.
+
+$n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}.
+
+The recursive definition of $toTree$ is mostly routine; each repeated element
+(shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$
+we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since
+constructors have no internal structure for us to compare. We also ignore values
+from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of
+$\phi$, and we conjecture that the effect of such ``magic values'' on clustering
+real code is low.
+
+\begin{align*}\label{fig:totree}
+  toTree(e) &=
+  \begin{cases}
+    (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
+    (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
+    (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
+    (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
+    (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
+    (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
+    (\feature{\CType})                                                & \text{if $e = \CType$} \\
+    (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
+    (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
+    (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
+    (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
+    (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
+    (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
+    (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
+    (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
+    (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
+    (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
+    (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
+    (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
+  \end{cases}
+\end{align*}
+
 \begin{figure}
@@
 
-These rose trees are then turned into matrices, as shown in Figure
-\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at
-depth $i$ in the rose tree, read left-to-right, followed by any required
+These rose trees are then turned into matrices, as shown in
+Figure~\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features
+at depth $i$ in the rose tree, read left-to-right, followed by any required
 padding. These matrices are then either truncated, or padded with rows (on the
@@
   Finally, matrices are turned into vectors by concatenating the rows from top
-  to bottom, hence Figure \ref{fig:matrixexample} will produce a vector
+  to bottom, hence Figure~\ref{fig:matrixexample} will produce a vector
   beginning
@@
 $\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
-assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14}
+assign a value to each, in a similar way to~\cite{DBLP:journals/corr/HerasK14}
 in Coq. We use a constant $\alpha$ to separate these values from those of other
-tokens (e.g. identifiers), but the order is essentially arbitrary: \footnote{In
-  \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix}
-  and \coq{cofix} are grouped together to reduce redundancy; we do not group
-  tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and
-  \CGlobal.}
+tokens (e.g. identifiers), but the order is essentially
+arbitrary:~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
+  tokens like \coq{fix} and \coq{cofix} are grouped together to reduce
+  redundancy; we do not group tokens, but we do put ``similar'' tokens close
+  together, such as \CLocal\ and \CGlobal.}
 
@@
 identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
-  index} of the identifier \cite{de1972lambda}, denoted $i_l$:
+  index} of the identifier~\cite{de1972lambda}, denoted $i_l$:
 
@@
 
-Instead, we use the approach taken in the latest versions of ML4PG and encode
-global identifiers \emph{indirectly}, by looking up the expressions which they
-\emph{reference}:
+Instead, we use the approach taken nu ML4PG and encode global identifiers
+\emph{indirectly}, by looking up the expressions which they \emph{reference}:
 
@@
 this way, we can avoid looking up expressions which haven't been clustered
-yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth}
+yet. To handle mutual recursion we use Tarjan's algorithm~\cite{tarjan1972depth}
 to produce a sorted list of \emph{strongly connected components} (SCCs), where
@@
 top-level expression as they are calculated, our algorithm can be computed
-\emph{iteratively} rather than recursively, as shown in Algorithm
-\ref{alg:recurrent}.
+\emph{iteratively} rather than recursively, as shown in
+Algorithm~\ref{alg:recurrent}.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[node distance=2cm]
+      \tikzset{vertex/.style = {minimum size=1.5em}}
+      \tikzset{edge/.style = {->,> = latex'}}
+
+      \node[vertex]                at (0.5\textwidth, 3)   (z)     {\hs{Z}};
+      \node[vertex]                at (0.5\textwidth, 0)   (s)     {\hs{S}};
+      \node[vertex]                at (0,             1.5) (plus)  {\hs{plus}};
+      \node[vertex, anchor=east]   at (\textwidth,    1.5) (false) {\hs{False}};
+      \node[vertex, right of=plus]                         (mult)  {\hs{mult}};
+      \node[vertex, left of=false]                         (odd)   {\hs{odd}};
+      \node[vertex, left of=odd]                           (even)  {\hs{even}};
+      \node[vertex, left of=even]                          (true)  {\hs{True}};
+
+      \draw[edge, dashed]     (z)     to (plus);
+      \draw[edge, dashed]     (z)     to (mult);
+      \draw[edge, dashed]     (z)     to (odd);
+      \draw[edge, dashed]     (z)     to (even);
+      \draw[edge, dashed]     (s)     to (plus);
+      \draw[edge, dashed]     (s)     to (mult);
+      \draw[edge, dashed]     (s)     to (odd);
+      \draw[edge, dashed]     (s)     to (even);
+      \draw[edge, dashed]     (false) to (odd);
+      \draw[edge, dashed]     (true)  to (even);
+      \draw[edge]             (plus)  to (mult);
+      \draw[edge]             (even)  to (odd);
+      \draw[edge]             (odd)   to (even);
+      \draw[edge, loop above] (plus)  to (plus);
+      \draw[edge, loop above] (mult)  to (mult);
+    \end{tikzpicture}
+
+    \caption{Dependency graph for Figure \ref{fig:coreexample}. Loops indicate
+      recursive functions, double arrows indicate mutual recursion. Dashed lines
+      show references to data constructors, which we do not consider in this
+      work.}
+    \label{fig:dagexample}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \centering
+    \tikzstyle{block} = [rectangle, draw]
+    \tikzstyle{line}  = [draw, -latex']
+    \begin{tikzpicture}[node distance=2cm]
+      \node [block] (plusset) {
+        \begin{tikzpicture}[node distance = 2cm]
+          \node [block] (plus) {\hs{plus}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=plusset] (parityset) {
+        \begin{tikzpicture}[node distance = 1cm]
+          \node [block]               (odd)  {\hs{odd}};
+          \node [block, below of=odd] (even) {\hs{even}};
+        \end{tikzpicture}
+      };
+      \node [block, right of=parityset] (multset) {
+        \begin{tikzpicture}[node distance = 2cm]
+          \node [block] (mult) {\hs{mult}};
+        \end{tikzpicture}
+      };
+
+      \path [line] (plusset)   -- (parityset);
+      \path [line] (parityset) -- (multset);
+    \end{tikzpicture}
+    \caption{One possible topological sorting of simply connected components for
+      Figure \ref{fig:dagexample} (ignoring constructors). \hs{even} and
+      \hs{odd} are mutually recursive, neither can appear before the other, so
+      they are grouped into one component and handled concurrently by Algorithm
+      \ref{alg:recurrent}.}
+    \label{fig:sccexample}
+  \end{subfigure}
+  \caption{Sorting functions from Figure \ref{fig:coreexample} into dependency order.}
+  \label{fig:dependencyexample}
+\end{figure}
 
 As an example of this recurrent process, we can consider the Peano arithmetic
-functions from Figure \ref{fig:coreexample}. A valid topological ordering is
-given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
+functions from Figure~\ref{fig:coreexample}. A valid topological ordering is
+given in Figure~\ref{fig:sccexample}, which can be our value for $d$ (eliding
 Core expressions to save space):
@@
 
-We can then trace the execution of Algorithm \ref{alg:recurrent} as follows:
+We can then trace the execution of Algorithm~\ref{alg:recurrent} as follows:
 
@@
 discarded by $toTree$) and recursive calls (which are replaced by
-$f_{recursion}$). A more extensive investigation of these features requires a
-concrete implementation, in particular to pin down values for the parameters
-such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.
+$f_{recursion}$).
 
@@
 called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
-for the GHC compiler, which emits a serialised version of each Core definition
-as it is being compiled. This approach is more robust than, for example, parsing
-source files, since it avoids the complications of preprocessors and language
-extensions altering the syntax.
+for the GHC compiler~\footnote{Available at
+  \url{https://github.com/warbo/ast-plugin}}, which emits a serialised version
+of each Core definition as it is being compiled. This approach is more robust
+than, for example, parsing source files, since it avoids the complications of
+preprocessors and language extensions altering the syntax.
 

### 2019-10-04 PhDThesis/background_theoryexploration.tex
Fold feedback into tex

@@
 (Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
-of definitions in some formal system (for example a programming language) and
+of definitions \iffalse TODO: Alison Some tools generate their own definitions
+\fi in some formal system (for example a programming language) and
 automatically generating a set of formal statements (properties) involving
@@
 sending conjectures to an automated theorem prover, or by having the generating
-procedure proceed in logically sound steps). These statements must also be
+procedure proceed in logically sound steps from a set of axioms). These
+statements should also be
 ``interesting'' in some way, to rule out unhelpful trivialities such as
@@
 The choice of appropriate methods can differ according to the underlying logic,
-etc. that is in use, so in this work we ground ourselves by applying TE to
+etc. that is in use: in this work we ground ourselves by applying TE to
 discovering properties of pure functional software libraries, mostly in the
 Haskell programming language. This also gives our results immediate utility to
-software engineering. We ignore the issue of \emph{proving} such conjectures,
-deferring that task to existing tools (such as \hipspec{}).
+software engineering. We focus on the generation of interesting conjectures; to
+\emph{prove} such conjectures, we defer to existing tools (such as \hip{} and
+\hipspec{}).
 
@@
 enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
-conjectures equations about Haskell code. We have used version 1 of
-\quickspec{} in our experiments, due to its availability, stability and tooling
-integration; hence references to \quickspec{} are to version 1 unless stated
-otherwise. At the time of writing there is a \quickspec{} version 2 available,
-which has a much improved generation procedure which can be significantly faster
-than its predecessor. Whilst \quickspec{} 2 has advanced the state of the art in
-theory exploration, our preliminary experience has found that it still suffers
-the scaling issues we identify in this work\footnote{We encountered memory
-  exhaustion when trying to run some of the examples included with the
-  \quickspec{} 2 source code; let alone auto-generated signatures.} Since
-\quickspec{} 2 has less integration with the tooling we have used, we leave a
-more thorough analysis of it (and related systems such as \speculate{}) for
-future work, pending the necessary infrastructure changes that would require.
+conjectures equations about Haskell code. We have used version 1 of \quickspec{}
+in our experiments, due to its availability, stability and tooling integration;
+hence references to \quickspec{} are to version 1 unless stated otherwise. At
+the time of writing \iffalse TODO: Alison: Include date \fi there is a
+\quickspec{} version 2 available, which has a much improved generation procedure
+which can be significantly faster than its predecessor. Whilst \quickspec{} 2
+has advanced the state of the art in theory exploration, our preliminary
+experience has found that it still suffers the scaling issues we identify in
+this work\footnote{Not only did we experience failures with our own experiments,
+  we also suffered memory exhaustion when trying to run some of the examples
+  included with the \quickspec{} 2 source code.} Since \quickspec{} 2 has less
+integration with the tooling we have used, we leave a more thorough analysis of
+it (and related systems such as \speculate{}) for future work, pending the
+necessary infrastructure changes that would require.
 

### 2019-10-04 PhDThesis/bucketing.tex
Incorporate feedback

@@
 all rely on enumerating combinations of definitions in some way, which can cause
-infeasible running times on larger inputs (based on experiments with our Theory
-Exploration Benchmark,%in \S~\ref{sec:benchmark}
-this happens above a few dozen
-definitions). To avoid this, users of these systems must carefully select only
-small subsets of their definitions to explore at a time.
+infeasible running times for input signatures with more than around a dozen
+definitions (based on experiments with our Theory Exploration Benchmark, in
+\S~\ref{sec:benchmark}. To avoid this, users of these systems must carefully
+select only small subsets of their definitions to explore at a time.
 
@@
 
+\iffalse TODO: Alison: This is very clear and good. Each chapter should have
+this at the start \fi
 In this chapter we will:
@@
 \end{enumerate}
+
+\iffalse
+TODO: Alison: Are you aiming to produce a set of recommendation for TE
+researchers to follow here?
+
+My thoughts: Not really. We are identifying an open problem (our attempt at a
+solution turns out not to work :( )
+\fi
 
@@
   \centering
-  \scalebox{0.45}{\input{images/steppedall.pgf}}
-  \scalebox{0.45}{\input{images/steppednontoxic.pgf}}
+  \scalebox{0.5}{\input{images/steppedall.pgf}}
+  \scalebox{0.5}{\input{images/timeoutsall.pgf}}
+  \scalebox{0.5}{\input{images/steppednontoxic.pgf}}
+  \scalebox{0.5}{\input{images/timeoutsnontoxic.pgf}}
   \caption{Kaplan-Meier survival plot for running \quickspec{} on inputs
@@
     denotes running time, which we cut short after 300 seconds. The height of
-    each line shows the proportion of \quickspec{} runs which were still going at
-    that time (lower is better). First plot is for all TIP definitions, second
-    plot removes runs given ``toxic'' definitions.}
+    each line shows the proportion of \quickspec{} runs which were still going
+    at that time (lower is better). First plot is for all TIP definitions,
+    second plot removes runs given ``toxic'' definitions.
+    \iffalse
+    TODO: Alison: The key needs further explanation (either here or in the caption)
+    TODO: Alison: Label the y axis (on both)
+    \fi}
   \label{fig:survival}
@@
 
+\iffalse
 \begin{figure}
@@
   \caption{Definitions, ordered by the ratio of successes to failures of the
-    runs they appeared in. First graph contains all TIP definitions, showing
-    ``toxic'' definitions which always failed. Second graph only contains runs
-    without any toxic definitions.}
+    runs they appeared in. The first graph contains all TIP definitions, showing
+    ``toxic'' definitions which always failed. The second graph only contains
+    runs without any toxic definitions.
+    \iffalse
+    TODO: Alison: ? (ME: The green/red bar graphs appear as empty white boxes in
+    the PDF Alison sent over. Check how it's rendered!)
+    TODO: The x-axis ticks are crap (all overlapping); turn them off, put a
+    single label like ``Definitions (ordered by failure rate)''
+    \fi}
   \label{fig:proportions}
@@
 
-To investigate the plausibility of such definitions, the ratio of successful and
-timed-out runs is plotted for each name in Figure~\ref{fig:proportions}; showing
+To investigate the plausibility of such definitions, we plotted the ratio of
+successful and timed-out runs for each name in Figure~\ref{fig:proportions}; showing
 that five definitions appeared \emph{only} in failing inputs, and seem likely to
@@
 \item Avoid ``toxic'' definitions: those producing exponentially large outputs,
-  and/or taking many arguments (leading to a large number of combinations). In
+  and/or taking many arguments (leading to a large number of combinations)
+  \iffalse TODO: Alison: Is it always obvious which are toxic? \fi. In
   particular, tail-recursive functions which expose their accumulator argument
-  are harder to explore.
+  are harder to explore. \iffalse TODO: Alison: These should be in the
+  contributions and further recommendations based on your other findings in this
+  chapter \fi
 \item Smaller signatures are faster to explore, so should be preferred if
@@
 
-Our final criterion concerns the ``quality'' of the selected subsets. This
-depends crucially on how we define whether a property is ``interesting''. We can
-use the same ground truth definition as our Theory Exploration Benchmark: a
-property is ``interesting'' if it appears in a particular corpus (in our case
-TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
-approach allows the criterion to be quantified: the quality of the selected
-subsets is the proportion of the whole signature's interesting properties that
-also apply to the subsets taken individually.
+Our final criterion \iffalse TODO: Alison: This looks odd since you haven't
+mentioned criteria before \fi concerns the ``quality'' of the selected
+subsets. This depends crucially on how we define whether a property is
+``interesting''. We can use the same ground truth definition as our Theory
+Exploration Benchmark: a property is ``interesting'' if it appears in a
+particular corpus (in our case TIP~\cite{claessen2015tip}), and
+``uninteresting'' if not. Whilst na\"ive, this approach allows the criterion to
+be quantified: the quality of the selected subsets is the proportion of the
+whole signature's interesting properties that also apply to the subsets taken
+individually.
 
@@
     into subsets. Each plot shows division into a different number of subsets.
-    The shaded area is bounded by the optimal and pessimal methods, which are
-    can only produce $N$ subsets for signatures of size $\geq N$. Grey lines
-    show pseudorandom selection, for comparison; these extend into smaller
-    signature sizes since empty subsets are permitted.}
+    The shaded area is bounded by the optimal and pessimal methods, which can
+    only produce $N$ subsets for signatures of size $\geq N$. Grey lines show
+    pseudorandom selection, for comparison; these extend into smaller signature
+    sizes since empty subsets are permitted.}
   \label{fig:bounds}
@@
 no better than random; in fact the curves show surprisingly similar patterns.
+\iffalse
+TODO: Alison: Does this translate into a recommendation (e.g. not to
+use clustering)? There should be some discussion of this - if not here then
+point to another chapter where you discuss these results
+
+TODO: Alison: Needs a summary section
+\fi
 

### 2019-10-04 PhDThesis/clustering.tex
Incorporate feedback

@@
 
+% TODO: Say what's in this chapter
+In this chapter we consider how the contents of a given signature can be
+analysed, such that subsequent exploration can be performed more efficiently. We
+frame this analysis as a task of \emph{unsupervised machine learning}, and
+describe how well-known \emph{clustering algorithms} can utilise the syntax of
+definitions to determine their similarity. This requires a detailed description
+of Haskell definition syntax, in particular of the ``Core'' intermediate
+language of the GHC compiler. We identify the difficulty posed by references,
+and how these can be made less opaque through the use of \emph{recurrent
+  clustering}.
+
+\section{Syntax}
+
 To improve the performance of theory exploration tools, we need to avoid
-brute-force search in favour of a more informed strategy. This, in turn,
-requires more information about the terms in the given signature. One
+brute-force search in favour of a more informed strategy.
+\iffalse
+TODO: Alison: This makes it sound as though current TE tools all use brute force
+(whereas all the tools I know already use informed strategies).
+MY THOUGHTS: ``more informed'' is a bit of weasel-wording from me; implying that
+they're not informed and that we're making them informed; whereas they're
+actually already somewhat informed, and we're trying to add to that. I agree
+that this weasel-wording should be replaced!
+\fi
+This, in turn,
+requires more information \iffalse TODO: Alison: More than what? \fi about the
+terms in the given signature. One
 generally available source of information is the syntax of a term's definition.
@@
 between semantically identical terms. Since theory exploration is concerned only
-with semantics (e.g. telling us how code actually behaves), we perform our
-analyses on the syntax of the \emph{GHC Core} language instead.
-
-\section{GHC Core}\label{sec:core}
+with semantics (telling us how code actually behaves), we perform our analyses
+on the simpler syntax of the \emph{GHC Core} language instead.
+
+\subsection{GHC Core}\label{sec:core}
 \label{sec:ghccore}
@@
   \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
-    repetition and grouping, respectively. Other forms of literal, like machine
+    repetition and grouping, respectively \iffalse TODO: Alison: Include examples? \fi. Other forms of literal, like machine
     words, are treated in a similar way and have been omitted for brevity.}
@@
     \underline{\texttt{Haskell definitions}}
+    \iffalse
+    TODO: Alison: Big space
+    MY THOUGHTS: I wanted to align things more neatly...
+    \fi
     \begin{haskell}
@@
   \caption{Example Haskell functions and their translations into the Core syntax
-    of Figure \ref{fig:coresyntax}. Notice the introduction of explicit
+    of Figure~\ref{fig:coresyntax}. Notice the introduction of explicit
     $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to
@@
 
-As a further simplification, we also replace data constructors in definitions
-with normal functions, e.g. replacing occurrences of a constructor
-\hs{S :: Nat -> Nat} with a function \hs{s} defined as \hs{s x = S x} (this is
-known as the $\eta$-expansion of \hs{S}). We do this to remove more distinctions
-from the syntax, and also to reduce cross-language differences (in particular,
-the Isabelle language used by the \isascheme{}, \isacosy{} and \hipster{} theory
-exploration systems treats constructors in a different way to Haskell, so having
-constructors in our data would complicate comparisons).
+Note that in test data such as our Theory Exploration Benchmark, data
+constructors in definitions are replaced with normal functions, e.g. replacing
+occurrences of a constructor \hs{S :: Nat -> Nat} with a function \hs{s} defined
+as \hs{s x = S x} (this is known as the $\eta$-expansion of \hs{S}). This
+removes even more distinctions from the syntax, and also reduces
+cross-language differences (in particular, the Isabelle language used by the
+\isascheme{}, \isacosy{} and \hipster{} theory exploration systems treats
+constructors in a different way to Haskell, so removing constructors from our
+data simplifies cross-language comparisons).
 
 \section{Feature Extraction}
+\label{sec:featureextraction}
 
@@
 values. We suggest a new recurrent clustering and feature extraction algorithm
-for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
+for Haskell (shown in Algorithm~\ref{alg:recurrent}) and compare its similarity
 and differences to the existing approaches of \textsc{ML4PG} and
@@
 the discrete symbols of Core syntax into features (real numbers), which we will
-denote using the function $\phi$. Before introducing our algorithm, we briefly
-describe \emph{clustering}, and its use in \emph{recurrent clustering}.
-
-\subsection{Clustering}
-
-Clustering is an unsupervised machine learning task for grouping $n$ data points
-using a similarity metric. There are many variations on this theme, but in our
-case we make the following choices:
-
-\begin{itemize}
-\item For simplicity, we use the ``rule of thumb'' given in
-  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
-  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
-\item Data points will be $d$-dimensional feature vectors, as defined above.
-\item We will use euclidean distance (denoted $e$) as our similarity metric.
-\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
-  \cite{lloyd1982least}.
-\end{itemize}
-
-We choose k-means because it is a standard approach and we are more concerned
-with the application of feature extraction to Haskell and its use in theory
-exploration, rather than precise tuning of learning algorithms. There are many
-other clustering algorithms we could use, such as \emph{expectation
-  maximisation}, but experiments with \textsc{ML4PG} have shown little
-difference in their results; it seems that the quality of the features is the
-bottleneck to learning, so there is no reason to avoid a fast algorithm like
-k-means. We use a Haskell implementation provided by the package
-\hPackage{k-means}.
-
-Since k-means is iterative, we will use function notation to denote time steps,
-so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
-to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
-which we denote as $\vect{m}^1$ to $\vect{m}^k$, hence:
-
-\begin{equation*}
-  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|}
-    \quad \text{for $t > 0$}
-\end{equation*}
-
-Before k-means starts, we must choose \emph{seed} values for
-$\vect{m}^i(0)$. Many methods have been proposed for choosing these values
-\cite{arthur2007k}. For simplicity, we will choose values randomly from our
-data set $S$; this is known as the Forgy method.
-
-The elements of each cluster $C^i(t)$ are those data points closest to the mean
-value at the previous time step:
-
-\begin{equation*}
-  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x},
-                                                        \vect{m}^j(t-1)) \}
-    \quad \text{for $t > 0$}
-\end{equation*}
-
-As $t$ increases, the clusters $C^i$ move from their initial location around the
-``seeds'', to converge on a local minimum of the ``within-cluster sum of squared
-error'' objective:
-
-\begin{equation} \label{eq:kmeansobjective}
-  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
-\end{equation}
-
-Many conservative improvements can be made to the standard k-means algorithm
-described above. For example, a more efficient approach like
-\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
-more practical to cluster; the \emph{k-means++}
-approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
-select the ``seed'' values for the first timestep; and the \emph{x-means}
-algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
-(instead of relying on our ``rule of thumb''). However, like the choice of
-k-means versus other clustering algorithms, these algorithmic improvements would
-need higher quality data than we are capable of producing in these early
-experiments.
-
-\subsubsection{Recurrent Clustering}
-
-\emph{Recurrent clustering} has been implemented in the
-ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
-\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
-syntax trees into fixed-size matrices, then concatenate the rows of these
-matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
-\emph{summarising} information about the tree; for example, one column counts
-the number of variables appearing at each tree level, others count the number of
-function symbols which are nullary, unary, binary, etc. In contrast, the
-algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
-arranging them in the matrix based on their position in the syntax tree. It is
-this latter algorithm that we adapt for analysing GHC Core expressions.
-
-The key to recurrent clustering is its treatment of identifiers. From a purely
-symbolic perspective, they are atomic: two identifiers can either be identical
-or distinct. Consider the \hs{odd} and \hs{even} functions in
-Figure~\ref{fig:coreexample}, which differ only in their choice of identifiers.
-These functions are clearly different (opposite, in fact), so their identifiers
-should play a role in determining their similarity. Yet these distinctions, e.g.
-between \hs{True} and \hs{False}, seem intuitively small relative to the
-distinction between, for example, \hs{isEmpty} and \hs{isValidEmailAddress}.
-
-Recurrent clustering solves this dichotomy by introducing a finer granularity to
-the treatment of identifiers, by first clustering the definitions that they
-refer to. If two definitions appear in the same cluster, their identifiers will
-be denoted by similar numbers during subsequent feature extraction; if they
-appear in different clusters, their numbers will differ more during feature
-extraction. It is this use of clustering to determine features that makes the
-approach \emph{recurrecnt}.
-
-\subsection{Expressions to Vectors}
+denote using the function $\phi$.
+
+\subsubsection{Expressions to Vectors}
 \label{sec:expressionstovectors}
@@
 example, consider the following Core expressions, which represent pattern-match
-clauses with different patterns but the same body (\hs{\vlocal{y}}):
-
-\begin{equation}\label{eq:xy}
+clauses with different patterns but the same body (\vlocal{y}):
+
+\begin{equation*}\label{eq:xy}
   \begin{array}{r@{}l@{}l@{}}
@@
   \end{array}
-\end{equation}
+\end{equation*}
 
@@
 \begin{small}
-  \begin{equation}
+  \begin{equation*}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
       breadthFirst(X)\ &=\ (\feature{\CAlt},\
-                           &\feature{\CDataAlt},\
-                           &\feature{\CVar},\
-                           &\feature{\id{C}},\
-                           &\feature{\CLocal},\
-                           &\feature{\id{y}} &) \\
+                           \ &\feature{\CDataAlt},\
+                           \ &\feature{\CVar},\
+                           \ &\feature{\id{C}},\
+                           \ &\feature{\CLocal},\
+                           \ &\feature{\id{y}} &) \\
       breadthFirst(Y)\ &=\ (\feature{\CAlt},\
-                           &\feature{\CDefault},\
-                           &\feature{\CVar},\
-                           &\feature{\CLocal},\
-                           &\feature{\id{y}},\
-                           &0 &)
+                           \ &\feature{\CDefault},\
+                           \ &\feature{\CVar},\
+                           \ &\feature{\CLocal},\
+                           \ &\feature{\id{y}},\
+                           \ &0 &)
     \end{array}
-  \end{equation}
+  \end{equation*}
 \end{small}
@@
 \begin{small}
-  \begin{equation}\label{eq:flattened}
+  \begin{equation*}\label{eq:flattened}
     \begin{array}{r@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l@{}l}
@@
     \end{array}
-  \end{equation}
+  \end{equation*}
 \end{small}
@@
   \caption[Rose tree for odd]{\label{fig:rosetreeexample} Rose tree for the
-    expression \hs{odd} from Figure \ref{fig:coreexample}. Each (sub-) rose tree
+    expression \hs{odd} from Figure~\ref{fig:coreexample}. Each (sub-) rose tree
     is rendered with its feature at the node and sub-trees beneath.}
@@
     \end{equation*}
-    \caption{Matrix generated from Figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
+    \caption{Matrix generated from Figure~\ref{fig:rosetreeexample}, padded to 6 columns. Each level of nesting in the tree corresponds to a row in the matrix.}
     \label{fig:matrixexample}
@@
 tokens (e.g. identifiers), but the order is essentially
-arbitrary:~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
+arbitrary~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
   tokens like \coq{fix} and \coq{cofix} are grouped together to reduce
   redundancy; we do not group tokens, but we do put ``similar'' tokens close
-  together, such as \CLocal\ and \CGlobal.}
+  together, such as \CLocal\ and \CGlobal.}:
 
@@
 
-Instead, we use the approach taken nu ML4PG and encode global identifiers
+Instead, we use the approach taken by ML4PG and encode global identifiers
 \emph{indirectly}, by looking up the expressions which they \emph{reference}:
@@
     \begin{cases}
-      i + 3 \alpha \quad & \text{if $g \in C_i$} \\
+      i + 3 \alpha \quad & \text{if $g \in C^i$} \\
       f_{recursion}         & \text{otherwise}
@@
 
-Where $C_i$ are our clusters (in some arbitrary order). This is where the
-recurrent nature of the algorithm appears: to determine the contents of $C_i$,
-we must perform k-means clustering \emph{during} feature extraction; yet that
-clustering step, in turn, requires that we perform feature extraction.
+Here $f_{recursion}$ is a constant value which is used for recursive calls, like
+those in the \hs{plus} and \hs{mult} definitions above. It will also appear in
+cases of mutual recursion, like the call to \hs{odd} within the definition of
+\hs{even} and \emph{vice versa}.
+
+$C$ contains the results (in some arbitrary order) of a \emph{clustering
+  algorithm} (described below in \S~\ref{sec:clustering}) applied to all of
+those definitions which are referenced non-recursively by this one. When we
+encounter such a reference, we look up which cluster that definition appeared
+in (denoted $C^i$), and use that index ($i$) as our feature value (using
+$\alpha$ to offset from other features as before).
+
+Defining the \emph{input} (feature vectors) for one definition in terms of the
+\emph{output} (clusters) of some other definitions gives rise to a
+\emph{recurrent clustering} algorithm, and blurs the boundary between feature
+extraction and clustering, which are normally distinct steps.
+
+\section{Clustering}
+\label{sec:clustering}
+
+Clustering is an unsupervised machine learning task for dividing $n$ data points
+into $k$ groups (``clusters'') such that points deemed similar a given metric
+tend to be grouped together. There are many variations on this theme, but in our
+case we make the following choices:
+
+\begin{itemize}
+\item For simplicity, we use the ``rule of thumb'' given in
+  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
+  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
+\item Data points will be $d$-dimensional feature vectors, as defined above in
+  \S~\ref{sec:featureextraction}.
+\item We will use euclidean distance (denoted $e$) as our similarity metric.
+\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
+  \cite{lloyd1982least}.
+\end{itemize}
+
+\iffalse
+TODO: Alison: It would be good to have justification of the previous
+items as well.
+\fi
+
+We choose k-means because it is a standard approach \iffalse TODO: Alison: Is
+there a reference for this? \fi. There are many other clustering algorithms we
+could use, such as \emph{expectation maximisation}, but experiments with
+\textsc{ML4PG} have shown little difference in their results; \iffalse TODO:
+Alison: Whose experiments? Yours (include something in appendix?) Someone else's
+(reference)? MY THOUGHTS: Conversation with Katya; maybe check ML4PG papers, or
+omit \fi it seems that \iffalse TODO: Alison: a bit informal and vague \fi the
+quality of the features is the bottleneck to learning, so there is no reason to
+avoid a fast algorithm like k-means. We use a Haskell implementation provided by
+the package \hPackage{k-means}.
+
+Since k-means is iterative, we will use function notation to denote time steps,
+so $x(t)$ denotes the value of $x$ at time $t$. \iffalse TODO: Alison: Pick a
+tense and stick to it. I'd suggest the present tense throughout \fi We denote
+the clusters as $C^1$
+to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
+which we denote as $\vect{m}^1$ to $\vect{m}^k$. The components of these vectors
+are treated as orthogonal, hence the $j$th component of $\vect{m}^i$ is the
+average of the $j$th components of all vectors in cluster $C^i$:
+
+\begin{equation*}
+  m^i_j(t) = \mean{C^i_j}(t) = \frac{\sum_{\vect{x} \in C^i(t)} x_j}{|C^i(t)|}
+    \quad \text{for $t > 0$}
+\end{equation*}
+
+Before k-means starts, we must choose \emph{seed} values for
+$\vect{m}^i(0)$. Many methods have been proposed for choosing these
+values~\cite{arthur2007k}. For simplicity, we will choose values randomly from
+our data set $S$; this is known as the Forgy method. \iffalse TODO: Alison: Any
+stronger justification? Why are the other methods not appropriate? \fi
+
+The elements of each cluster $C^i(t)$ are those data points closest to the mean
+value at the previous time step $\vect{m}^i(t-1)$:
+
+\begin{equation*}
+  C^i(t) = \{ \vect{x} \in S \mid i = \argmin\limits_j e(\vect{x},
+                                                        \vect{m}^j(t-1)) \}
+    \quad \text{for $t > 0$}
+\end{equation*}
+
+As $t$ increases, the clusters $C^i$ move from their initial location around the
+``seeds'', to converge on a local minimum of the ``within-cluster sum of squared
+error'' objective:
+
+\begin{equation} \label{eq:kmeansobjective}
+  \argmin\limits_C \sum_{i=1}^k \sum_{\vect{x} \in C^i} e(\vect{x}, \vect{m}^i)^2
+\end{equation}
+
+Many conservative improvements can be made to the standard k-means algorithm
+described above. For example, a more efficient approach like
+\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
+more practical to cluster; the \emph{k-means++}
+approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
+select the ``seed'' values for the first timestep; and the \emph{x-means}
+algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
+(instead of relying on our ``rule of thumb''). However, like the choice of
+k-means versus other clustering algorithms, the benefits of more, higher-quality
+data seem greater than those offered by such algorithmic tweaking. \iffalse
+TODO: Also describe better reasons for avoiding these improvements, i.e. they
+are algorithmically complicated and require multiple passes over the input data
+\fi
+
+\subsubsection{Recurrent Clustering}
+
+\emph{Recurrent clustering} has been implemented in the
+ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
+\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
+syntax trees into fixed-size matrices, then concatenate the rows of these
+matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
+\emph{summarising} information about the tree; for example, one column counts
+the number of variables appearing at each tree level, others count the number of
+function symbols which are nullary, unary, binary, etc. In contrast, the
+algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
+arranging them in the matrix based on their position in the syntax tree. It is
+this latter algorithm that we have adapted for analysing GHC Core expressions.
+
+The key to recurrent clustering is its treatment of identifiers. From a purely
+symbolic perspective, they are atomic: two identifiers can either be identical
+or distinct. Consider the \hs{odd} and \hs{even} functions in
+Figure~\ref{fig:coreexample}, which differ only in their choice of identifiers.
+These functions are clearly different (opposite, in fact), so their identifiers
+should play a role in determining their similarity. Yet these distinctions, e.g.
+between \hs{True} and \hs{False}, seem intuitively small relative to the
+distinction between, for example, \hs{isEmpty} and \hs{isValidEmailAddress}.
+
+Recurrent clustering solves this dichotomy by introducing a finer granularity to
+the treatment of identifiers, by first clustering the definitions that they
+refer to. If two definitions appear in the same cluster, references to those
+definitions will be denoted by similar numbers during subsequent feature
+extraction; if they appear in different clusters, their numbers will differ more
+during feature extraction. It is this use of clustering to determine features
+that makes the approach \emph{recurrent}.
 
 For this recursive process to be well-founded, we perform a topological sort on
-declarations based on their dependencies (the expressions they reference). In
-this way, we can avoid looking up expressions which haven't been clustered
+definitions based on their dependencies (the definitions they reference). In
+this way, we can avoid looking up definitions which haven't been clustered
 yet. To handle mutual recursion we use Tarjan's algorithm~\cite{tarjan1972depth}
@@
 cannot be found amongst those clustered so far, it must appear in the same SCC
-as the expression we are processing; hence we give that identifier the constant
-feature value $f_{recursion}$.
+as the expression we are processing; hence we give that identifier a constant
+feature value denoted as $f_{recursion}$.
 
@@
 
-    \caption{Dependency graph for Figure \ref{fig:coreexample}. Loops indicate
+    \caption{Dependency graph for Figure~\ref{fig:coreexample}. Loops indicate
       recursive functions, double arrows indicate mutual recursion. Dashed lines
-      show references to data constructors, which we do not consider in this
-      work.}
+      show references to data constructors, which we ignore in this work
+      (replacing such references with $\eta$-expanded alternatives).}
     \label{fig:dagexample}
@@
     \caption{One possible topological sorting of simply connected components for
-      Figure \ref{fig:dagexample} (ignoring constructors). \hs{even} and
+      Figure~\ref{fig:dagexample} (ignoring constructors). \hs{even} and
       \hs{odd} are mutually recursive, neither can appear before the other, so
-      they are grouped into one component and handled concurrently by Algorithm
-      \ref{alg:recurrent}.}
+      they are grouped into one component and handled concurrently by
+      Algorithm~\ref{alg:recurrent}.}
     \label{fig:sccexample}
   \end{subfigure}
-  \caption{Sorting functions from Figure \ref{fig:coreexample} into dependency order.}
+  \caption{Sorting functions from Figure~\ref{fig:coreexample} into dependency order.}
   \label{fig:dependencyexample}
@@
 cluster, since they only differ in their choice of constructor names (which are
-discarded by $toTree$) and recursive calls (which are replaced by
+discarded by $toTree$) and recursive calls (which are both replaced by
 $f_{recursion}$).
@@
 We provide an implementation of our recurrent clustering algorithm in a tool
-called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
-for the GHC compiler~\footnote{Available at
+called \mlforhs{}~\footnote{Available at
+  \url{https://github.com/warbo/ml4hsfe}}. We obtain Core ASTs from Haskell
+packages using a plugin for the GHC compiler~\footnote{Available at
   \url{https://github.com/warbo/ast-plugin}}, which emits a serialised version
@@
 
-A post-processing stage determines which Core definitions can be used by
-\quickspec{}, based on their type, visibility (whether they are encapsulated inside
-their module or visible to importers), etc. For each definition, we simply use
-the GHCi interpreter to check if calling \quickspec{} with that argument is
-well-typed. This information, along with type signatures, arity, etc. are stored
+A post-processing stage determines which Core definitions are suitable for
+exploration, based on their visibility (whether they are encapsulated inside
+their module or visible to importers), whether a data generator is available,
+etc. This information, along with type signatures, arity, etc. are stored
 alongside the Core definitions in JSON format.
@@
 identifiers appearing in their ASTs, and feature vectors are constructed using a
-Haskell implementation of the approach described in \S
-\ref{sec:recurrentclustering}. Since the features associated with each
+Haskell implementation of the approach described in
+\S~\ref{sec:featureextraction}. Since the features associated with each
 identifier may vary between iterations (as the clusters change), we leave the
@@
 
-We implement Algorithm \ref{alg:recurrent} as a simple shell script, which
-invokes Weka for clustering and associates each Core definition with a cluster
+We implement Algorithm~\ref{alg:recurrent} as a Haskell executable, which
+performs the clustering and associates each Core definition with a cluster
 number. These numbers are used to finish the deferred feature extraction of
 identifiers, the resulting feature vectors are clustered, and the process is
-repeated until all SCCs have been processed.
-
-Once the recurrent clustering is complete, we generate a string of Haskell code
-for each of the final clusters, which will invoke \quickspec{} with a suitable
-signature. This involves:
+repeated until all SCCs have been processed. Once the recurrent clustering is
+complete, we output a JSON array of the clusters (serialised in an arbitrary
+order).
+
+\iffalse
+% TODO: This seems less appropriate for this chapter, more suitable for
+% Signature Selection
+A separate application called \mlspec{} can read in this JSON, generate
+appropriate Haskell code to invoke \quickspec{} on each cluster individually,
+and evaluate that code with all appropriate dependencies installed. This
+decoupling allows our feature extraction algorithm to be used for other
+purposes, and allows alternative clustering involves:
 
@@
 form, sending to a theorem prover, etc.).
+\fi
 
@@
 
-% TODO: Move recurrent clustering to its own chapter, with this as its
-% evaluation?
+\iffalse
+TODO: Alison: Next chapter? It's good to lead into the next chapter at the end
+of the previous ones.
+\fi
+
 We have applied our recurrent clustering algorithm to several scenarios, with
@@
 no ``ground truth'', i.e. there is no objectively correct way to compare
-expressions. Instead, we provide a qualitative overview of the more interesting
+definitions. Instead, we provide a qualitative overview of the more interesting
 characteristics.
@@
 As a simple example, we clustered (Haskell equivalents of) the running examples
-used to present ACL2(ml) \cite{heras2013proof}, shown in Figure
-\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
+used to present \textsc{ACL2(ml)}~\cite{heras2013proof}, shown in
+Figure~\ref{fig:lisp}. These include tail-recursive and non-tail-recursive
 implementations of several functions. We expect those with similar
 \emph{structure} to be clustered together, rather than those which implement the
-same function. The results are shown in Figure \ref{fig:haskellcluster}, where
+same function. The results are shown in Figure~\ref{fig:haskellcluster}, where
 we can see the ``\hs{Tail}'' functions clearly distinguished, with little
@@
   \end{common-lisp}
-  \caption{Common Lisp functions, both tail-recursive and non-tail-recursive.}
+  \caption{Common Lisp functions used with \textsc{ACL2(ml)}, both
+    tail-recursive and non-tail-recursive.}
   \label{fig:lisp}
@@
 
+\iffalse
+Since we are working in the domain of Haskell programs, an abundant source of
+statements are available in the form of \emph{tests}. For manually-written
+tests, the effort required to write them implies that they must be of some
+interest to their author. Many forms of test are \emph{not} suitable for our
+purposes, such as \emph{unit tests} which are always trivially provable by
+$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
+the external environment. One reason to study Haskell is its widespread use of
+\emph{property checking}, which does give us useful data. Many Haskell property
+checkers exist, based on random testing
+(\quickcheck{}~\cite{claessen2011quickcheck} and
+\smartcheck{}~\cite{pike2014smartcheck}), enumeration
+(\smallcheck{}~\cite{runciman2008smallcheck}), observation
+(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
+(\sparsecheck{}~\cite{sparsecheck}).
+
+Thankfully the major differences between these systems are in the way they
+instantiate test arguments; their representations of properties are largely the
+same (modulo renaming of functions and types).
+
+Here we consider the 10 most-downloaded packages from \hackage{} (as of
+2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
+\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
+\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
+\texttt{blaze-html}.
+
+% TODO: Analysis
+\fi
+
+\iffalse
 Whilst recurrent clustering has produced results which merit further
@@
 \end{itemize}
-
-It seems like this recurrent clustering method has promise, although it will
+\fi
+
+It seems like this recurrent clustering method has promise\iffalse TODO: Alison: Informal, vague \fi, although it will
 require a more thorough exploration of the parameters to obtain more intuitively
@@
 theory exploration; the most na\"{\i}ve way being to explore each cluster as
-\quickspec{} signature in its own right. \mlforhs{} already provides this
-functionality, although the lack of test generators severely limits what can be
-discovered.
-
+\quickspec{} signature in its own right, which we investigate as a potential
+solution to the Signature Selection problem.
+
+\iffalse
+TODO: Alison: Next chapter? It's good to lead into the next chapter at the end of the previous ones.
+\fi
+
+\iffalse
 As alluded to previously, we also have the opportunity to reason by
@@
 as a simple logical consequence.
-
+\fi
+
+\iffalse
 \subsection{Recurrent Clustering}
@@
 rules to reduce the ASTs prior to feature extraction.
+\fi
+W

### 2019-10-04 PhDThesis/related_formalstats.tex
Fix listings
Incorporate feedback

@@
-\subsection{Statistics of Formal Systems}
+\section{Statistics of Formal Systems}
+
+\iffalse
+TODO: Alison: I'd like to revisit this chapter once I see the whole
+thesis. At the moment it feels in parts more like a discussion of your work,
+than of related work. It might help to read other people's related work sections
+to get a better idea of what to include here. I think it'll be more obvious too,
+once we have the intro, etc. I suspect you'll need to do a fair bit more work on
+this one.
+\fi
 
@@
 of formal systems. This problem has been tackled from various directions for a
-variety of reasons; here we summarise those contributions which seem of
+variety of reasons; here we summarise those contributions which are of
 particular importance for theory exploration.
 
-\subsubsection{Clustering}
+\iffalse
+\subsection{Clustering}
 \label{sec:clustering}
@@
 \cite{heras2013proof}
+\fi
 
-\subsubsection{Probability of Sentences}
+\subsection{Probability of Statements}
 
-The most important property of a logical formula is its truth value. Although we
-may be able to determine some truth values exactly, e.g. using decision or
-semi-decision procedures, it may be more efficient to \emph{approximate} truth
-values. One straightforward extension of truth values is \emph{probabilities},
-where we can assign probability $1$ to formulae which are known to be true, $0$
-to formulae known to be false, and intermediate values to those which we do not
-yet know.
+An important property of a logical formula (in general, and in particular with
+regards to interestingness) is its truth value. Although we may be able to
+determine some truth values exactly, e.g. via deduction, it may only be possible
+(or more efficient) to \emph{approximate} truth values. One straightforward
+approximation is to adopt \emph{probabilities}, where we assign probability $1$
+to formulae which are known to be true (e.g. axioms), $0$ to formulae known to
+be false, and intermediate values to those which are not known.
 
-% TODO
-\cite{Hutter.Lloyd.Ng.ea:2013}
+This can be useful for ``knowledge bases'' of non-logical axioms (e.g. the
+observed colours of Hempel's ravens~\cite{10.1093/mind/LIV.214.97} \iffalse
+TODO: Alison: Maybe a bit of explanation here \fi), and allows the probability of
+universal statements to be inferred from non-exhaustive
+cases~\cite{Hutter.Lloyd.Ng.ea:2013}. However, inferring consistent
+probabilities requires \emph{logical omniscience}: knowledge of all the logical
+consequences of the chosen axioms. This is unfortunate, since it is precisely
+what we are trying to avoid by approximating with probabilities.
 
-\subsubsection{Interestingness in Concept Formation}
+Avoiding this requirement for logical omniscience has given rise to the field of
+\emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
+subject to some computational constraints. For example, the ``logical inductor''
+of~\cite{garrabrant2016logical} avoids assigning particular probabilities to
+formal statements, and instead produces a \emph{sequence} of successive
+approximate probabilities, each computable with finite (though
+doubly-exponential!) resources. These approximations are only guaranteed to be
+consistent in the limit.
+
+The probability of a statement can affect its interestingness in a few ways.
+Statements whose probability is very close to $0$ or $1$ may be considered
+less interesting, since they are essentially known (e.g. easily derivable or
+refutable via known facts). Those closer to $\frac{1}{2}$ may be considered more
+interesting; although in a bounded rationality context this may simply indicate
+the requirement for some routine (but lengthy) computation: for example stating
+that the $n$th bit of $\pi$ is a zero, for some large $n$.
+
+\iffalse
+\subsection{Interestingness in Concept Formation}
 \label{sec:conceptformation}
@@
 % Eurisko, AM, etc.?
+\fi
 
-\subsubsection{Learning From Structured Data}
+\subsection{Learning From Structured Data}
 
@@
 
-\paragraph{Feature Extraction}\label{sec:featureextraction}
+\subsubsection{Feature Extraction}\label{sec:featureextraction}
 
@@
 
-\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
+\begin{haskell}
 data Maybe a = Nothing | Just a
@@
 data Either a b = Left a | Right b
-\end{lstlisting}
+\end{haskell}
 
@@
   \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
+  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a
+    \hs{b}.
 \end{itemize}
@@
 
-\paragraph{Truncation and Padding}
+\subsubsection{Truncation and Padding}
 
@@
 
-\paragraph{Dimension Reduction}
+\subsubsection{Dimension Reduction}
 
@@
 points in high-dimensional spaces, which we want to project into a
-lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit
-vectors).
+lower-dimensional space.
 
 Truncation is a trivial dimension reduction technique: take the first $N$
-coordinates (bits). More sophisticated projection functions consider the
+coordinates. More sophisticated projection functions consider the
 \emph{distribution} of the points, and project with the hyperplane which
@@
 There are many techniques to find these hyperplanes, such as \emph{principle
-  component analysis} (PCA) and \emph{autoencoding}; however, since these
+  component analysis} (PCA) and \emph{autoencoding}. If the data are labelled,
+these classes can also be taken into account during dimension
+reduction~\cite{Oveisi.Oveisi.Erfanian.ea:2012}. However, since these
 techniques are effectively ML algorithms in their own right, they suffer some of
@@
 
-\paragraph{Sequences}
+\subsubsection{Sequences}
 
-To handle input of \emph{variable} size, most attention has been given to the
-handling of \emph{sequences}. This is a lossless approach, which splits the
-input into fixed-size \emph{chunks}, which are fed into an appropriate ML
-algorithm one at a time. The sequence is terminated by a sentinel; an
-``end-of-sequence'' marker which, by construction, is distinguishable from the
-data chunks. This technique allows us to trade \emph{space} (the size of our
-input) for \emph{time} (the number of chunks in a sequence).
+To handle input of \emph{variable} size, research attention has been given to
+the handling of \emph{sequences}. This is a lossless approach, which splits the
+input into fixed-size \emph{chunks} (e.g. splitting text into a sequence of
+characters), which are fed into an appropriate ML algorithm one at a time. The
+sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by
+construction, is distinguishable from the data chunks. This technique allows us
+to trade \emph{space} (the size of our input) for \emph{time} (the number of
+chunks in a sequence).
 
@@
 
-\paragraph{Recursive Structure}
+An alternative to introducing recurrent connections, is the use of
+\emph{attention}. % TODO: cite
+In a (typically feed-forward) ANN with attention, the whole input is provided to
+the network, which produces some output as well as some ``attention''
+information; the network is run again, but with the input adjusted using this
+attention information (increasing and decreasing some of the input values); more
+output is produced, along with new attention information, and so on. Such
+networks do not need to ``remember'' long-distance relationships, whether by
+recurrent connections or gated nodes; they instead use the same input over and
+over again, but use their ``attention'' to emphasise those parts which should be
+relevant to future iterations.
+
+\subsubsection{Recursive Structure}
 
@@
 
-The application of \emph{kernel methods} to structured information is discussed
-in \cite{Gartner2003}, where the input data (including sequences, trees and
-graphs) are represented using \emph{generative models}, such as hidden Markov
-models, of a fixed size.
-
 \emph{Kernel methods} have also been applied to structured information, for
@@
 clustering alone.
-
-% TODO
-\cite{Gartner2003}
-\cite{Oveisi.Oveisi.Erfanian.ea:2012}
-\cite{bakir2007predicting}
-\cite{conf/ijcai/Plate91}
-\cite{goller1996learning}
-\cite{kwasny1995tail}
-\cite{pollack1990recursive}
-\cite{zanzotto2012distributed}

### 2019-10-16 PhDThesis/litreview_staticanalysis.tex
Bulk out static analysis

@@
 the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
-Linters are characterised by parsing the given code and looking for occurrences
-of certain syntax patterns, for example definitions which are never used, or
-``no-op'' constructs (e.g. \texttt{(x~<~0)} when \texttt{x} is of
-unsigned/non-negative type, making this comparison always false).
+Linters are standalone tools, characterised by parsing the given code and
+looking for occurrences of certain syntax patterns; for example definitions
+which are never used, ``no-op'' constructs (e.g. \texttt{if~(x~<~0)} when
+\texttt{x} is of unsigned/non-negative type) and expressions which are valid but
+usually used erroneously (e.g. \texttt{if (x = 1)} which performs assignment
+rather than comparison).
 
@@
 perform an open-ended search through the space of all programs to find the
-optimal implementation according to some quality measure or cost model (often a
-branch- and effect-free subset of machine code is used, and measured by counting
-instructions or cycles).  Superoptimisers need some method to check whether a
-synthesised program implements the required specification, such as a test suite
-and/or a theorem prover.
+optimal implementation according to some quality measure or cost model.
 
-Proving correctness of machine code programs is helped by the uniform, finite
-nature of their domain (e.g. a fixed set of instructions, operating on a fixed
+Supercompilers are often restricted to a branch- and effect-free subset of
+machine code, where cost can be reasonably estimated by counting instructions or
+cycles. Higher-level languages require more elaborate estimation methods, such
+as cost semantics~\cite{danner2015denotational}: such models will still contain
+many parameters, some of which may be determined by empirical measurement,
+whilst others may depend on characteristics of the input data. Despite this
+drawback, pareto-optimal optimisations can still be identified, and either
+presented to the user or tested against a realistic data set (known as
+\emph{profile-guided optimisation}~\cite{TODO}).
+
+Superoptimisers also need some method to check whether a synthesised program
+implements the required specification, such as a test suite and/or a theorem
+prover. Proving correctness of machine code fragments is helped by the uniform,
+finite nature of their domain (a fixed set of instructions, operating on a fixed
 set of fixed-size registers). Higher level languages like Haskell are more
 difficult to reason about and measure, since their behaviour and efficiency may
-depend on unknown runtime parameters. For example, a higher-order function $f$
-accepting another function $g$ as an argument may choose to memoise calls to $g$
-in order to avoid evaluating identical calls multiple times; however, storing
-previous results requires more memory, and each input must be looked up in the
-memo table to see if it's been used before. Whether this is an optimisation or
-pessimisation depends on the runtime data at each call site, including the
-function argument (whether it's cheaper than performing a comparison) and the
-data it will be applied to (whether it actually contains any repeated values).
+depend on unknown runtime parameters. In particular, optimising higher-order
+functions is difficult in isolation, since many performance tradeoffs cannot be
+decided without more knowledge of the functions involved. For example, a cache
+of previous return values (known as a \emph{memo table}) can avoid re-computing
+repeated calls to a function; but whether this actually saves time, or is worth
+the memory penalty, depends on how long that function takes to calculate; and
+how often it will be called with same arguments.
+
 In addition to this uncertainty, each function's input domain may be infinite
@@
 potentially undecidable.
-
-We can reduce the amount of unknown parameters in a concrete program by
-propagating statically-known values, for example using
-\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
-dramatically increase code size, as each function may be duplicated into
-multiple specialised instances. Related techniques, like \emph{symbolic
-  execution} (a form of \emph{abstract interpretation}), allow approximate
-execution of code, where unknown runtime data is represented as opaque
-metavariables. Property- and model-checkers are similar, except they work
-empirically by performing multiple runs, instantiating metavariables with
-synthetic data (\quickcheck{} does this randomly, \smallcheck{} enumerates
-values, etc.).
-
-Efficiency can be calculated using techniques like cost
-semantics~\cite{danner2015denotational}, with constant factors determined
-through empirical testing. This cannot solve the problem of optimality depending
-on the characteristics of runtime data, but it can expose these dependencies as
-parameters allowing the user to select between pareto-optimal strategies.
 
@@
 
-The open-ended search process of such superoptimisers is similar to application
-of mathematical theory exploration to software property discovery, especially
-when exploration is limited to equality properties (which is the case for
-\isacosy{} and \quickspec{}). Quality metrics like efficiency (e.g.
-relative instruction count) can be considered a subset of the interestingness
-criteria of theory exploration. The \emph{aims} of these analyses are in
-opposition, since theory exploration \emph{builds up} structure and abstraction
-latent in a set of definitions, whilst superoptimisation \emph{tears down} such
-layers to find minimal implementations; yet the practical necessity of
-superoptimising only small snippets at a time (e.g. to find ``peephole
-optimisations''~\cite{Bansal.Aiken:2006}) results in remarkably similar
-discovery processes.
+Rather than optimising individual units of code in isolation, we can instead
+work ``top down'' and perform \emph{whole program optimisation}.\iffalse TODO: Cite MLTon, Stalin,
+maybe others \fi In particular, for every function we may want to optimise, we
+know \emph{all} call sites of that function within the program, and hence avoid
+the need for a general solution. Statically-known values, including arguments to
+(higher-order) functions, can be propagated through the code using techniques
+like \emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
+expressions can be ahead of time; however this can dramatically increase code
+size, as each function may be duplicated into multiple specialised instances.
+
+Related techniques, like \emph{symbolic execution} (a form of \emph{abstract
+  interpretation}), allow information to be statically deduced by tracing the
+behaviour of code using an alternative execution model, where unknown runtime
+data are represented as opaque metavariables. Property- and model-checkers are
+similar, except they work dynamically by performing multiple runs, instantiating
+metavariables with exhaustive or synthetic data (\quickcheck{} does this
+randomly, \smallcheck{} enumerates values, etc.).
+
+The most common code-improvement technique performed \emph{during execution} is
+Just In Time (JIT) compilation~\iffalse TODO: Cite \fi, which allows an
+interpreter to compile sections of code into a faster form (usually machine
+code). Since this compilation takes time, it is usually reserved for sections
+which are executed often enough to exceed some predefined threshold (e.g. 100).
+By focusing on such ``hot paths'', the up-front costs of compiling everything
+ahead of time are avoided; although compiled forms are kept only in memory,
+getting recompiled as needed on each program run. Compiling at runtime also
+allows specialisation with dynamically-known information; for example, inlining
+specific implementations of a procedure rather than performing dynamic dispatch
+(this is the problem higher-order functions discussed above). To avoid altering
+semantics, such dynamic specialisations must be accompanied by a conservative
+validity check (e.g. that the given arguments should use the procedure
+implementation we have compiled-in), falling back to the interpreter upon
+failure.
+
+A more speculative use of optimisation during runtime has been proposed by
+Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
+algorithm begins by directly executing the given (unoptimised) program $p$ on
+the relevant input $x$ (this may be provided up front, or piecemeal throughout
+execution, e.g. as a stream of user interaction events). The variables $p_{fast}$
+and $t_{fast}$ are initialised to $p$ and $\infinity$ respectively, to indicate
+that $p$ is the fastest implementation we have so far, and we don't yet have an
+upper bound on its running time. Concurrently with the execution of $p$, two
+other processes are also started:
+
+\begin{itemize}
+\item Process A searches for proofs (in a pre-specified formal system) that some
+  program $p\prime$ is both semantically equivalent to $p$ and has a running
+  time bounded by some other program $t$, i.e. for all inputs $i$, the
+  time taken to compute $p\prime(i)$ is bounded by $t(i)$. A list is built up of
+  the $(p\prime, t)$ pairs found so far.
+\item Process B computes $t(x)$ for all of the bounds $t$ found so far by A.
+  Computations are interleaved via a schedule similar to Levin's universal
+  search algorithm~\iffalse TODO: Cite \fi, to prevent getting stuck on
+  non-halting programs. Whenever a calculation of some $t(x)$ finishes, it is
+  compared to $t_{fast}$: if smaller, $t_{fast}$ is set to $t(x)$, and $p_{fast}$ is
+  set to $p\prime$ (the program paired with this $t$).
+\end{itemize}
+
+The execution of $p_{fast}(x)$ (initially $p(x)$) is restarted at
+exponentially-spaced intervals: if the value of $p_{fast}$ gets updated,
+execution will switch to the new, provably-faster program at the next restart.
+Although wasteful, such restarting will only slow down execution by at most a
+factor of 2; whilst switching to an updated $p_{fast}$ may bring arbitrarily large
+speed-ups (depending on the problem). Since the proof search (process A) is
+independent of the input $x$, it only contributes a constant factor to the
+running time; although in the general case this factor will be exponential in
+the length of the proof.
+
+Hutter's algorithm is mostly of theoretical interest, being described as
+``another $O()$ warning showing, how important [to running time] factors, and
+even subdominant additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite
+this caveat, the general framework is still useful to compare with the code
+improvement techniques described above. In particular, the use of open-ended
+search (process A) is similar to that of superoptimisation; although the former
+enumerates equivalence proofs and extracts programs from them, whilst the latter
+enumerates programs and attempts to prove (or approximately determine)
+equivalence. Likewise the speed estimation of process B is analogous to those
+employed by superoptimisation. JIT compilation can also be achieved within this
+framework, for example by including our desired optimisations in the axioms of
+the formal system (making proof search trivial), biasing the search algorithm to
+consider applications of these optimisations first, and including a simple cost
+model for the language. Switching to faster programs can also be made
+incremental, e.g. allowing (stateless) functions to be replaced individually
+without having to restart execution, just like in a JIT-enabled interpreter.
+
+The open-ended search through program and proof space performed by
+superoptimisation and Hutter's algorithm are similar to that of Theory
+Exploration (e.g. the discovery of equivalent expressions by \isacosy{} and
+\quickspec{}). One major difference is in \emph{end result} of these techniques:
+the latter tends to \emph{build up} structure and abstraction latent in a set of
+definitions, whilst the former tend to \emph{tear down} such layers to find
+minimal implementations.
+
+Another seeming contrast is that Theory Exploration has no ``target'', unlike
+optimisers. However, the practical necessity of limiting superoptimisers to only
+small snippets of code at a time has resulted in applications like finding
+``peephole optimisations''~\cite{Bansal.Aiken:2006}, whose scattershot nature is
+more akin to Theory Exploration than traditional compiler optimisation.
+
+In any case, efficiency (whether instruction count, small upper bounds, cost
+semantics, etc.) can be considered part of the interestingness criteria of
+Theory Exploration, and there are certainly opportunities to cross-pollinate
+techniques between these fields.
 
 % TODO: Evolutionary patches
-% TODO: Hutter search
 % TODO: Goedel machines

### 2019-10-16 PhDThesis/clustering.tex
TODOs

@@
 
-% TODO: Say what's in this chapter
+\iffalse TODO: Alison: Each chapter should have a description at the start (see
+bucketing ``in this chapter'') \fi
+
 In this chapter we consider how the contents of a given signature can be
@@
 
-To improve the performance of theory exploration tools, we need to avoid
-brute-force search in favour of a more informed strategy.
-\iffalse
-TODO: Alison: This makes it sound as though current TE tools all use brute force
-(whereas all the tools I know already use informed strategies).
-MY THOUGHTS: ``more informed'' is a bit of weasel-wording from me; implying that
-they're not informed and that we're making them informed; whereas they're
-actually already somewhat informed, and we're trying to add to that. I agree
-that this weasel-wording should be replaced!
-\fi
-This, in turn,
-requires more information \iffalse TODO: Alison: More than what? \fi about the
-terms in the given signature. One
-generally available source of information is the syntax of a term's definition.
+To improve the performance of theory exploration tools, we need to provide their
+search strategies with more than just type information about the terms in the
+given signature, rather than treating them as ``black boxes''. One generally
+available source of information is the syntax of a term's definition.
 
@@
 
-  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote
-    repetition and grouping, respectively \iffalse TODO: Alison: Include examples? \fi. Other forms of literal, like machine
-    words, are treated in a similar way and have been omitted for brevity.}
+  \caption{Simplified syntax of GHC Core in BNF style. $[]$ denotes repetition
+    (zero or more occurrences). Other forms of literal, like machine words, are
+    treated in a similar way and have been omitted for brevity.}
   \label{fig:coresyntax}
@@
     \underline{\texttt{Haskell definitions}}
-    \iffalse
-    TODO: Alison: Big space
-    MY THOUGHTS: I wanted to align things more neatly...
-    \fi
     \begin{haskell}

### 2019-10-17 PhDThesis/litreview_staticanalysis.tex
More static analysis

@@
 Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
-algorithm begins by directly executing the given (unoptimised) program $p$ on
-the relevant input $x$ (this may be provided up front, or piecemeal throughout
-execution, e.g. as a stream of user interaction events). The variables $p_{fast}$
-and $t_{fast}$ are initialised to $p$ and $\infinity$ respectively, to indicate
-that $p$ is the fastest implementation we have so far, and we don't yet have an
-upper bound on its running time. Concurrently with the execution of $p$, two
-other processes are also started:
+algorithm (also referred \hsearch{}) begins by directly executing the given
+(unoptimised) program $p$ on the relevant input $x$ (this may be provided up
+front, or piecemeal throughout execution, e.g. as a stream of user interaction
+events). The variables $p_{fast}$ and $t_{fast}$ are initialised to $p$ and
+$\infinity$ respectively, to indicate that $p$ is the fastest implementation we
+have so far, and we don't yet have an upper bound on its running
+time. Concurrently with the execution of $p$, two other processes are also
+started:
 
@@
 
-Hutter's algorithm is mostly of theoretical interest, being described as
-``another $O()$ warning showing, how important [to running time] factors, and
-even subdominant additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite
-this caveat, the general framework is still useful to compare with the code
-improvement techniques described above. In particular, the use of open-ended
-search (process A) is similar to that of superoptimisation; although the former
-enumerates equivalence proofs and extracts programs from them, whilst the latter
-enumerates programs and attempts to prove (or approximately determine)
+\hsearch{} is mostly of theoretical interest, being described as ``another $O()$
+warning showing, how important [to running time] factors, and even subdominant
+additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite this caveat, the
+general framework is still useful to compare with the code improvement
+techniques described above. In particular, the use of open-ended search (process
+A) is similar to that of superoptimisation; although the former enumerates
+equivalence proofs and extracts programs from them, whilst the latter enumerates
+programs and attempts to prove (or approximately determine)
 equivalence. Likewise the speed estimation of process B is analogous to those
@@
 
-The open-ended search through program and proof space performed by
-superoptimisation and Hutter's algorithm are similar to that of Theory
-Exploration (e.g. the discovery of equivalent expressions by \isacosy{} and
-\quickspec{}). One major difference is in \emph{end result} of these techniques:
-the latter tends to \emph{build up} structure and abstraction latent in a set of
-definitions, whilst the former tend to \emph{tear down} such layers to find
-minimal implementations.
+The framework of the \goedelmachine{}~\cite{Schmidhuber:05icann} goes even
+further than \hsearch{}: rather than limiting optimisations to just the given
+program, it also searches for provable optimisations to \emph{itself}. In
+principle, this allows the search algorithm, execution schedule, etc. to all be
+optimised, replaced or even removed entirely, as long as a proof is found that
+it would be beneficial to do so. The \goedelmachine{} concept is framed in the
+language of reinforcement learning, with the goal of maximising``reward'' from
+some ``utility function'', but it is easily to recast in terms of improving
+efficiency: for example, choosing a utility function which provides a one-off
+reward of $\frac{1}{t}$ if a result provably-equivalent to the original
+program's is returned within $t$ steps; and 0 reward otherwise. The ability to
+provide a utility function separate from the initial implementation is also
+useful for specifying the desired task more declaratively: for example, if it is
+acceptable to deviate from the behaviour of the initial implementation, as long
+as some other criteria (given by the utility function) are satisfied. A relevant
+example might be altering a search algorithm to find more results faster, even
+if those aren't the same results that the original algorithm would have found.
 
-Another seeming contrast is that Theory Exploration has no ``target'', unlike
-optimisers. However, the practical necessity of limiting superoptimisers to only
-small snippets of code at a time has resulted in applications like finding
-``peephole optimisations''~\cite{Bansal.Aiken:2006}, whose scattershot nature is
-more akin to Theory Exploration than traditional compiler optimisation.
+The open-ended search through program and proof space performed by algorithms
+like superoptimisation, \hsearch{} and the \goedelmachine{} are similar to that
+of Theory Exploration (e.g. the discovery of equivalent expressions by
+\isacosy{} and \quickspec{}). One major difference is in \emph{end result} of
+these techniques: the latter tends to \emph{build up} structure and abstraction
+latent in a set of definitions, whilst the former tend to \emph{tear down} such
+layers to find direct, minimalistic implementations.
 
-In any case, efficiency (whether instruction count, small upper bounds, cost
-semantics, etc.) can be considered part of the interestingness criteria of
-Theory Exploration, and there are certainly opportunities to cross-pollinate
-techniques between these fields.
+Another seeming contrast is that Theory Exploration has no specific ``target'',
+unlike optimisers: it produces any properties found to match its
+``interestingness'' criteria. However, the practical necessity of limiting
+superoptimisers to only small snippets of code at a time has resulted in
+applications like finding ``peephole optimisations''~\cite{Bansal.Aiken:2006},
+whose scattershot nature is more akin to Theory Exploration than traditional
+compiler optimisation. Likewise, the improvement in efficiency of an
+optimisation (however that is measured or justified) contributes to its
+interestingness, and hence suitability as an outputs for Theory Exploration
+systems (especially in the use case of a programming assistant).
 
 % TODO: Evolutionary patches
-% TODO: Goedel machines

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Add genetic patches

@@
 
-At the other end of the code improvement spectrum we find
-\emph{superoptimisers}~\cite{massalin1987superoptimizer}, which can take an
-arbitrarily long time to find improvements, but whose deductive power is only
-limited by the undecidability of the source language. Superoptimisers, such as
-\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
+To find project-specific improvements we need some description or specification
+of how our program \emph{should} behave, to judge proposed improvements against.
+Few projects have a complete formal specification, due to the prohibitive cost
+of creating, validating, verifying and maintaining such artefacts. Partial
+specifications, in the form of automated test suites, are more common. These
+have been used to automatically patch errors in code (i.e. altering code until
+its test suite passes), using methods from evolutionary
+computation~\cite{Forrest.Nguyen.Weimer.ea:2009,weimer2009automatically,Schulte.Forrest.Weimer:2010}.
+
+Optimisations treat \empg{the code itself} as a specification: formally
+describing the desired behaviour, but not prescribing the exact steps that
+should be taken to implement it. This allows different implementations to be
+used as long as the result is equivalent\footnote{Optimisation assumes that
+  extra efficiency is always desirable. In some areas that is not the case, such
+  as security-sensitive code, where speed ups may expose sensitive information
+  to timings attacks~\cite{kocher1996timing}.}. Taking this to the extreme gives
+rise to \emph{superoptimisation}~\cite{massalin1987superoptimizer}, which
+performs an open-ended search through the space of all programs to find those
+equivalent to the given code, which are also optimal according to some quality
+measure or cost model. Superoptimisers, such a
+s \textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
 Assistant'', developed to accompany \emph{Hacker's
-  Delight}~\cite{warren2013hacker}) treat the given code as a specification, and
-perform an open-ended search through the space of all programs to find the
-optimal implementation according to some quality measure or cost model.
+  Delight}~\cite{warren2013hacker}), can take an arbitrarily long time to find
+improvements, but their optimisation power is limited only by the undecidability
+of the source language (optimisations which cannot be proven equivalent are
+necessarily ignored, like the uncomputable improvements demonstrated by
+Blum~\cite{blum1967machine}).
 

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Oops, optimise not compile

@@
 
-Supercompilers are often restricted to a branch- and effect-free subset of
+Superoptimisers are often restricted to a branch- and effect-free subset of
 machine code, where cost can be reasonably estimated by counting instructions or

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Typo

@@
 
-Optimisations treat \empg{the code itself} as a specification: formally
+Optimisations treat \emph{the code itself} as a specification: formally
 describing the desired behaviour, but not prescribing the exact steps that

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Citations

@@
 presented to the user or tested against a realistic data set (known as
-\emph{profile-guided optimisation}~\cite{TODO}).
+\emph{profile-guided optimisation}~\cite{hubicka2005profile}).
 
@@
 Rather than optimising individual units of code in isolation, we can instead
-work ``top down'' and perform \emph{whole program optimisation}.\iffalse TODO: Cite MLTon, Stalin,
-maybe others \fi In particular, for every function we may want to optimise, we
-know \emph{all} call sites of that function within the program, and hence avoid
-the need for a general solution. Statically-known values, including arguments to
-(higher-order) functions, can be propagated through the code using techniques
-like \emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
+work ``top down'' and perform \emph{whole program
+  optimisation}~\cite{weeks2006whole,siskind1999flow}. In particular, for every
+function we may want to optimise, we know \emph{all} call sites of that function
+within the program, and hence avoid the need for a general solution.
+Statically-known values, including arguments to (higher-order) functions, can be
+propagated through the code using techniques like
+\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
 expressions can be ahead of time; however this can dramatically increase code
@@
 warning showing, how important [to running time] factors, and even subdominant
-additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite this caveat, the
+additive terms, are.''~\cite[p. 4]{HUTTER:2002}. Despite this caveat, the
 general framework is still useful to compare with the code improvement

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Fix tex

@@
 events). The variables $p_{fast}$ and $t_{fast}$ are initialised to $p$ and
-$\infinity$ respectively, to indicate that $p$ is the fastest implementation we
+$\infty$ respectively, to indicate that $p$ is the fastest implementation we
 have so far, and we don't yet have an upper bound on its running

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
Remove fixed TODO
Add citations

@@
 The most common code-improvement technique performed \emph{during execution} is
-Just In Time (JIT) compilation~\iffalse TODO: Cite \fi, which allows an
+Just In Time (JIT) compilation~\cite{aycock2003brief}, which allows an
 interpreter to compile sections of code into a faster form (usually machine
@@
 A more speculative use of optimisation during runtime has been proposed by
-Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
-algorithm (also referred \hsearch{}) begins by directly executing the given
-(unoptimised) program $p$ on the relevant input $x$ (this may be provided up
-front, or piecemeal throughout execution, e.g. as a stream of user interaction
-events). The variables $p_{fast}$ and $t_{fast}$ are initialised to $p$ and
-$\infty$ respectively, to indicate that $p$ is the fastest implementation we
+Hutter~\cite{HUTTER:2002}. As in a JIT-enabled interpreter, Hutter's
+algorithm (which we refer to as \hsearch{}) begins by directly executing the
+given (unoptimised) program $p$ on the relevant input $x$ (this may be provided
+up front, or piecemeal throughout execution, e.g. as a stream of user
+interaction events). The variables $p_{fast}$ and $t_{fast}$ are initialised to $p$
+and $\infty$ respectively, to indicate that $p$ is the fastest implementation we
 have so far, and we don't yet have an upper bound on its running
@@
   Computations are interleaved via a schedule similar to Levin's universal
-  search algorithm~\iffalse TODO: Cite \fi, to prevent getting stuck on
+  search algorithm~\cite{levin1984randomness,ming}, to prevent getting stuck on
   non-halting programs. Whenever a calculation of some $t(x)$ finishes, it is
@@
 systems (especially in the use case of a programming assistant).
-
-% TODO: Evolutionary patches

### 2019-10-18 PhDThesis/clustering.tex
Fixed todos

@@
 \begin{itemize}
-\item For simplicity, we use the ``rule of thumb'' given in
-  \cite[pp. 365]{mardia1979multivariate} to fix the number of clusters at
-  $k = \lceil \sqrt{\frac{n}{2}} \rceil$.
+\item We will use \emph{k-means} clustering~\cite{macqueen1967some}, implemented
+  by Lloyd's algorithm~\cite{lloyd1982least}\footnote{We use a Haskell
+    implementation provided by the package \hPackage{k-means}.}, due to its
+  simplicity, fast convergence~\cite{bottou1994convergence} and use in prior
+  work on recurrent clustering~\cite{journals/corr/abs-1212-3618}.
+\item The number of clusters $k$ must be chosen (or, if we were to use a
+  hierarchical clustering algorithm, a level to ``cut off'' the hierarchy).
+  There is no ``correct'' answer, although algorithms like
+  \emph{x-means}~\cite{pelleg2000x} can estimate this during clustering. We
+  stick to a standard k-means algorithm for the reasons given above, and use
+  the common ``rule of thumb''~\cite[pp. 365]{mardia1979multivariate} of setting
+  $k = \lceil \sqrt{\frac{n}{2}} \rceil$
 \item Data points will be $d$-dimensional feature vectors, as defined above in
-  \S~\ref{sec:featureextraction}.
-\item We will use euclidean distance (denoted $e$) as our similarity metric.
-\item We will use \emph{k-means} clustering, implemented by Lloyd's algorithm
-  \cite{lloyd1982least}.
+  \S~\ref{sec:featureextraction}. This is the standard input format for such
+  machine learning tasks.
+\item We will use the standard euclidean distance (denoted $e$) as the
+  similarity metric between feature vectors. In principle we could construct a
+  task-specific distance function to better determine similarity, but that would
+  only shift the problem, since it is precisely the task that we are attempting
+  to solve through the use of (recurrent) clustering.
 \end{itemize}
 
-\iffalse
-TODO: Alison: It would be good to have justification of the previous
-items as well.
-\fi
-
-We choose k-means because it is a standard approach \iffalse TODO: Alison: Is
-there a reference for this? \fi. There are many other clustering algorithms we
-could use, such as \emph{expectation maximisation}, but experiments with
-\textsc{ML4PG} have shown little difference in their results; \iffalse TODO:
-Alison: Whose experiments? Yours (include something in appendix?) Someone else's
-(reference)? MY THOUGHTS: Conversation with Katya; maybe check ML4PG papers, or
-omit \fi it seems that \iffalse TODO: Alison: a bit informal and vague \fi the
-quality of the features is the bottleneck to learning, so there is no reason to
-avoid a fast algorithm like k-means. We use a Haskell implementation provided by
-the package \hPackage{k-means}.
-
-Since k-means is iterative, we will use function notation to denote time steps,
-so $x(t)$ denotes the value of $x$ at time $t$. \iffalse TODO: Alison: Pick a
-tense and stick to it. I'd suggest the present tense throughout \fi We denote
-the clusters as $C^1$
+Since k-means is iterative, we use function notation to denote time steps, so
+$x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$
 to $C^k$. As the name suggests, k-means uses the mean value of each cluster,
@@
 $\vect{m}^i(0)$. Many methods have been proposed for choosing these
-values~\cite{arthur2007k}. For simplicity, we will choose values randomly from
-our data set $S$; this is known as the Forgy method. \iffalse TODO: Alison: Any
-stronger justification? Why are the other methods not appropriate? \fi
+values~\cite{arthur2007k}. It is common to choose values randomly from
+our data set $S$ (known as the Forgy method), since this prevents having to
+synthesise (potentially unrealistic) data.
 
@@
 Many conservative improvements can be made to the standard k-means algorithm
-described above. For example, a more efficient approach like
-\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
-more practical to cluster; the \emph{k-means++}
-approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
-select the ``seed'' values for the first timestep; and the \emph{x-means}
-algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
-(instead of relying on our ``rule of thumb''). However, like the choice of
-k-means versus other clustering algorithms, the benefits of more, higher-quality
-data seem greater than those offered by such algorithmic tweaking. \iffalse
-TODO: Also describe better reasons for avoiding these improvements, i.e. they
-are algorithmically complicated and require multiple passes over the input data
-\fi
+described above, such as the more efficient \emph{yinyang
+  k-means}~\cite{conf/icml/DingZSMM15}, or the \emph{k-means++}
+approach~\cite{arthur2007k, bahmani2012scalable} to more carefully selecting the
+``seed'' values for the first timestep. Adding such complexities does not appear
+necessary for the moment, as clustering data sets like our Theory Exploration
+Benchmark only takes seconds.
 

### 2019-10-18 PhDThesis/litreview_testing.tex
Wording

@@
 \quickspec{} (inductive logic with higher-order functions) comes directly from
-its application to Haskell; and its search algorithm makes heavy use of the
+its application to Haskell; its search algorithm also makes heavy use of the
 \quickcheck{} property checker. The boundary between these two fields is hence
@@
 ``property''), usually involving components of a software system and, crucially,
-may contain free variables. \emph{Unit testing} is a common method of software
-quality assurance (outside of specialised domains like those which are
+which may contain free variables. \emph{Unit testing} is a common method of
+software quality assurance (outside of specialised domains like those which are
 life-critical), which can be characterised as proving \emph{existential}

### 2019-10-18 PhDThesis/litreview.tex
Thesis changes (e.g. lit review)
More content
Bunch more changes to thesis
Disable some bits of litreview for now
Headingy things
TODOs
Add intro

--- 2018-05-17 PhDThesis/litreview.tex
@@
-\section{Literature Review}
+\chapter{Literature Review}
 
-\input{litreview_mathhistory.tex}
+In this chapter we place existing work on Theory Exploration in context, survey
+related fields and those which we draw our own contributions and inspirations
+from; including Machine Learning, tools for identifying and applying
+improvements to software, premise selection in (automated and interactive)
+theorem proving, and measures for determining the ``interestingness'' of a
+statement or conjecture.
+
+\input{litreview_automatedreasoning.tex}
+\input{litreview_staticanalysis.tex}
+\input{litreview_testing.tex}
+\input{litreview_premiseselection.tex}
+\input{litreview_machinelearning.tex}
 \input{litreview_interestingness.tex}
-\input{litreview_theoryformation.tex}
-\input{litreview_theoryexploration.tex}
-\input{litreview_premiseselection.tex}

### 2019-10-18 PhDThesis/litreview_machinelearning.tex
Bunch more changes to thesis
Headingy things
More literview

@@
-\subsection{Machine Learning Applications in Mathematics, Logic and Programming Language Theory}
+\section{Machine Learning Applications in Mathematics, Logic and Programming
+  Language Theory}
 
-Machine learning techniques, used for finding statistical patterns and numerical
-correlations in arbitrary data, are characterised by their statistical pattern-finding
+Machine Learning (ML) applies statistical techniques to find patterns in data.
+ML algorithms are numerical in nature, and often rely on assumptions such as
+continuity and differentiability. In contrast, the worlds of mathematical
+formulae, formal logic and computer programs are usually discrete and symbolic.
+This makes application of ML to these domains more difficult than, say, image
+processing.
 
-AutoTuning
+One application of ML in these domains is to analyse real-world measurements
+parameterised by some discrete entity, such as time or power used to execute
+computer programs. Since flipping a single bit can cause arbitrarily large
+changes in program behaviour (if it is the scrutinee of a conditional, for
+example), such programs must be very similar for statistical patterns to emerge
+in their executions. In the case of \emph{auto-tuning}, many programs variants
+are measured, which all perform the same task in slightly different ways (e.g.
+using different compiler optimisations, or dividing work differently between
+hardware components); in this case the goal is to learn a predictive model of
+which variant performs best for a given input~\cite{ganapathi2009case}.
 
-DeepTune analyses raw program text to predict the best options for an optimising
-compiler. FIXME
+Whilst performance measurements are simple numeric values, the symbolic values
+need some conversion to become amenable to learning. This is usually performed
+by a \emph{feature extraction} step, which takes various measurements (e.g.
+counting occurrences of some element, assigning indices to different constants,
+etc.) and packs them into a vector for subsequent analysis. More recently, so
+called \emph{deep learning} systems have out performed those with manually
+chosen features. Deep learning doesn't actually avoid feature extraction, but
+its use of many ``layers'' of processing between the input and output makes the
+selection of those features less important. For example, the feature extraction
+in~\cite{cummins2017synthesizing} represents computer programs (after some
+normalisation to a consistent layout and naming scheme) as a sequence of
+individual characters, with each ASCII character replaced by a different number.
+Yet even this crude method is enough for the deep learning system that consumed
+these sequences to infer structure relevant for deciding whether it was better
+suited to running on a CPU or GPU. The \textsc{DeepTune} system encodes slightly
+more domain knowledge by tokenising language keywords, but is otherwise
+similar~\cite{cummins2017end}.
 
-Fuzzing AFL DeepSmith
+\emph{Generative models} are another useful application of ML techniques to
+symbolic systems. The \textsc{DeepSmith} system analyses a corpus of source code
+to build a model of the programming language, which can then generate realistic
+fragments of code, suitable for fuzz-testing compilers for that
+language~\cite{cummins2017deepsmith}.

### 2019-10-18 PhDThesis/litreview_premiseselection.tex
Headingy things
More literview

@@
-\subsection{Relevance Filtering}
+\section{Relevance Filtering}
 \label{sec:relevance}
 
-\emph{Relevance filtering} takes a specified ``target'' (such as the current
-goal of a theorem prover), a large list of inputs (e.g. existing lemmas from a
-theory) and produces a sub-set of those inputs which are predicted to be useful
-or relevant to the target. The example of choosing a sub-set of lemmas to help
-prove a theorem is known as \emph{premise
-  selection}~\cite{kuhlwein2012overview}.
+In the context of theorem proving, \emph{relevance filtering} takes a specified
+``target'' (such as the current goal of a theorem prover), a large list of
+inputs (e.g. existing lemmas from a theory) and produces a sub-set of those
+inputs which are predicted to be useful or relevant to the target. The example
+of choosing a sub-set of lemmas to help prove a theorem is known as
+\emph{premise selection}~\cite{kuhlwein2012overview}.
 

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
More literview

@@
 
-Software for analysing and improving code can be roughly divided into two
-categories. Some contribute to the running of a program, such as (optimising)
-compilers, linkers and interpreters. These must produce correct, predictable,
+One obvious application of software properties discovered through Theory
+Exploration is to improve that software, for example by \emph{refactoring} (e.g.
+simplifying definitions) or \emph{optimising} (e.g. replacing a slow calculation
+with an equivalent fast one). Existing tools for analysing and improving code
+can be roughly divided into two categories. Some contribute to the running of a
+program, such as (optimising) compilers, linkers and
+interpreters\footnote{Optimisation has traditionally been the domain of
+  compilers, but the recent popularity of \emph{link time optimisation} and
+  \emph{just-in-time compilation} have brought the latter tools firmly under the
+  code improvement umbrella.}. These must produce correct, predictable,
 desirable results in a timely manner. For example, compiler optimisations should
@@
 refactorings~\cite{robison2001impact}. Optimisations in ahead-of-time (AOT)
-compilers must be fast enough to run during every build, and should have a large
-enough impact on the generated code's performance that they
+compilers/linkers must be fast enough to run during every build, and should have
+a large enough impact on the generated code's performance that they
 ``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
@@
 its test suite passes), using methods from evolutionary
-computation~\cite{Forrest.Nguyen.Weimer.ea:2009,weimer2009automatically,Schulte.Forrest.Weimer:2010}.
+computation~\cite{Forrest.Nguyen.Weimer.ea:2009, weimer2009automatically,
+  Schulte.Forrest.Weimer:2010}. Such methods are necessarily standalone, since
+making such arbitrary changes to a program's behaviour without a full
+specification or user supervision has the potential to break the program in
+untested ways.
 
@@
 equivalent to the given code, which are also optimal according to some quality
-measure or cost model. Superoptimisers, such a
-s \textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
+measure or cost model. Superoptimisers, such as
+\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
 Assistant'', developed to accompany \emph{Hacker's
@@
 of the source language (optimisations which cannot be proven equivalent are
-necessarily ignored, like the uncomputable improvements demonstrated by
+necessarily ignored, like the uncomputable speedups demonstrated by
 Blum~\cite{blum1967machine}).
@@
 Superoptimisers also need some method to check whether a synthesised program
-implements the required specification, such as a test suite and/or a theorem
-prover. Proving correctness of machine code fragments is helped by the uniform,
-finite nature of their domain (a fixed set of instructions, operating on a fixed
-set of fixed-size registers). Higher level languages like Haskell are more
-difficult to reason about and measure, since their behaviour and efficiency may
-depend on unknown runtime parameters. In particular, optimising higher-order
-functions is difficult in isolation, since many performance tradeoffs cannot be
-decided without more knowledge of the functions involved. For example, a cache
-of previous return values (known as a \emph{memo table}) can avoid re-computing
-repeated calls to a function; but whether this actually saves time, or is worth
-the memory penalty, depends on how long that function takes to calculate; and
-how often it will be called with same arguments.
+is equivalent to that given: either partially, like a test suite, or robustly
+via a theorem prover. Proving correctness of machine code fragments is helped by
+the uniform, finite nature of their domain (a fixed set of instructions,
+operating on a fixed set of fixed-size registers). Higher level languages like
+Haskell are more difficult to reason about and measure, since their behaviour
+and efficiency may depend on unknown runtime parameters. In particular,
+optimising higher-order functions is difficult in isolation, since many
+performance tradeoffs cannot be decided without more knowledge of the functions
+involved. For example, introducing a cache of previous return values (known as a
+\emph{memo table}) can avoid re-computing repeated calls to a function; but
+whether this actually saves time, or is worth the memory penalty, depends on how
+long that function takes to calculate and how often it will be called with same
+arguments.
 
-In addition to this uncertainty, each function's input domain may be infinite
-(such as the natural numbers), and programs may use abstract interfaces and
-datatypes with a variety of possible implementations (such as lists or key/value
-mappings). All these factors makes effective testing more difficult, and proofs
-potentially undecidable.
+In addition to this uncertainty, high level languages allow the input domain of
+a function to be infinite (such as the natural numbers or lists), and programs
+may use abstract interfaces and datatypes which either hide characteristics
+required for establishing equivalence or cost, or may permit \emph{multiple}
+implementations with varying characteristics. All these factors make effective
+testing more difficult, and proofs potentially undecidable.
 
@@
 \emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
-expressions can be ahead of time; however this can dramatically increase code
-size, as each function may be duplicated into multiple specialised instances.
+expressions can be computed ahead of time; however this can dramatically
+increase code size, as each function may be duplicated into multiple specialised
+instances. Link-time optimisation permits a limited form of inter-procedural
+optimisation (e.g. inlining function definitions which were abstract black-boxes
+during compilation), but generally avoids the more invasive optimisations found
+in compilers.
 
-Related techniques, like \emph{symbolic execution} (a form of \emph{abstract
-  interpretation}), allow information to be statically deduced by tracing the
-behaviour of code using an alternative execution model, where unknown runtime
-data are represented as opaque metavariables. Property- and model-checkers are
-similar, except they work dynamically by performing multiple runs, instantiating
-metavariables with exhaustive or synthetic data (\quickcheck{} does this
-randomly, \smallcheck{} enumerates values, etc.).
+Other methods to determine function usage and behaviour include \emph{symbolic
+  execution} (a form of \emph{abstract
+  interpretation}), which allow information to be statically deduced by tracing
+the behaviour of code using an alternative execution model, where unknown
+runtime data are represented as opaque metavariables. Property- and
+model-checkers are similar, except they work dynamically by performing multiple
+runs, instantiating metavariables with exhaustive or synthetic data
+(\quickcheck{} does this randomly, \smallcheck{} enumerates values, etc.).
 
 The most common code-improvement technique performed \emph{during execution} is
-Just In Time (JIT) compilation~\cite{aycock2003brief}, which allows an
+just-in-time (JIT) compilation~\cite{aycock2003brief}, which allows an
 interpreter to compile sections of code into a faster form (usually machine
@@
 \item Process A searches for proofs (in a pre-specified formal system) that some
-  program $p\prime$ is both semantically equivalent to $p$ and has a running
+  program $p'$ is both semantically equivalent to $p$ and has a running
   time bounded by some other program $t$, i.e. for all inputs $i$, the
-  time taken to compute $p\prime(i)$ is bounded by $t(i)$. A list is built up of
-  the $(p\prime, t)$ pairs found so far.
+  time taken to compute $p'(i)$ is bounded by $t(i)$. A list is built up of
+  the $(p', t)$ pairs found so far.
 \item Process B computes $t(x)$ for all of the bounds $t$ found so far by A.
@@
   compared to $t_{fast}$: if smaller, $t_{fast}$ is set to $t(x)$, and $p_{fast}$ is
-  set to $p\prime$ (the program paired with this $t$).
+  set to $p'$ (the program paired with this $t$).
 \end{itemize}

### 2019-10-22 PhDThesis/litreview_machinelearning.tex
Added some ML

@@
 This makes application of ML to these domains more difficult than, say, image
-processing.
+processing. This difficulty is clear from attempts to train Artificial Neural
+Networks to evaluate snippets of program code~\cite{zaremba2014learning}: even
+with heavily restricted inputs (e.g. short integer calculations, with little
+nesting, requiring only linear time and memory use, and evaluating with
+``teacher forcing'' to prevent errors accumulating) the accuracy quickly drops
+with program length. Coupled with the success of \emph{deep learning} (efficient
+back-propagation/credit-assignment over many differentiable processing steps)
+compared to prior hand-engineered AI solutions, it is apparent that symbolic
+and statistical methods are \emph{complementary}; hence the need to identify
+relevant sub-tasks for each technique when combined into some larger system.
 
-One application of ML in these domains is to analyse real-world measurements
-parameterised by some discrete entity, such as time or power used to execute
+One way to incorporate ML into a traditionally symbolic domain is to analyse
+real-world measurements of a system, such as time or power taken to execute
 computer programs. Since flipping a single bit can cause arbitrarily large
@@
 Whilst performance measurements are simple numeric values, the symbolic values
-need some conversion to become amenable to learning. This is usually performed
-by a \emph{feature extraction} step, which takes various measurements (e.g.
-counting occurrences of some element, assigning indices to different constants,
-etc.) and packs them into a vector for subsequent analysis. More recently, so
-called \emph{deep learning} systems have out performed those with manually
-chosen features. Deep learning doesn't actually avoid feature extraction, but
-its use of many ``layers'' of processing between the input and output makes the
-selection of those features less important. For example, the feature extraction
-in~\cite{cummins2017synthesizing} represents computer programs (after some
-normalisation to a consistent layout and naming scheme) as a sequence of
-individual characters, with each ASCII character replaced by a different number.
-Yet even this crude method is enough for the deep learning system that consumed
-these sequences to infer structure relevant for deciding whether it was better
-suited to running on a CPU or GPU. The \textsc{DeepTune} system encodes slightly
-more domain knowledge by tokenising language keywords, but is otherwise
+still need some conversion to become amenable to learning. This is usually
+performed by a \emph{feature extraction} step, which takes various measurements
+(e.g.  counting occurrences of some element, assigning indices to different
+constants, etc.) and packs them into a vector for subsequent analysis. One major
+advantage of deep learning systems is their ability to find structure in rather
+crude representations: the first few layers effectively perform the job of
+feature extraction, without the need for hand-crafting task-specific extractors.
+For example, the feature extraction in~\cite{cummins2017synthesizing} represents
+computer programs (after some normalisation to a consistent layout and naming
+scheme) as a sequence of individual characters, with each ASCII character
+replaced by a different number. Yet even this na\"ive method is enough for the
+deep neural network that consumes these sequences to infer structure relevant
+for deciding whether the code is better suited to running on a CPU or GPU. The
+\textsc{DeepTune} system encodes slightly more domain knowledge by tokenising
+language keywords, rather than requiring them to be learned, but is otherwise
 similar~\cite{cummins2017end}.
 
-\emph{Generative models} are another useful application of ML techniques to
-symbolic systems. The \textsc{DeepSmith} system analyses a corpus of source code
-to build a model of the programming language, which can then generate realistic
-fragments of code, suitable for fuzz-testing compilers for that
-language~\cite{cummins2017deepsmith}.
+Another useful class of ML systems are \emph{generative models}, which are
+reversible. The \textsc{DeepSmith} system~\cite{cummins2017deepsmith} turns a
+corpus of source code into a stream of (numeric) tokens, feeds them into a deep
+neural network and learns to predict which token will appear next in the stream.
+Once trained, realistic fragments of code can be generated by providing an
+initial token as a ``seed'', then sampling from the predicted distribution
+(similar to using a Markov Chain). Generated programs are suitable for e.g.
+fuzz-testing compilers. Such systems are relevant to Theory Exploration, since
+those systems can be viewed as a generative model of interesting statements. One
+difficulty which makes it far from straightforward to implement a Theory
+Exploration system along the same lines as \textsc{DeepSmith} is the lack of
+ground-truth data from which to discern what is ``interesting'' and what is not
+(our Theory Exploration Benchmark %TODO Section ref
+provides enough for evaluation purposes, but learning (and subsequently
+validating) would require orders of magnitude more). Another is the
+ever-changing nature of what is interesting: sampling statements from a static
+distribution would eventually run out of novelty and surprisingness, at the
+least. There is, however, promise in using generative models to produce
+realistic input to some separate processing step, e.g. to filter or otherwise
+transform them to produce only interesting results.

### 2019-10-22 PhDThesis/bucketing.tex
TODOs
Fixes

@@
 
-\iffalse TODO: Alison: This is very clear and good. Each chapter should have
-this at the start \fi
 In this chapter we will:
@@
 \item Define the \emph{signature selection problem} for theory exploration.
-  \item Analyse system performance on the Theory Exploration Benchmark, to
-    demonstrate the effect of ``toxic'' definitions leading to timeouts.
 \item Analyse the effect of signature selection on the quantity of desirable
-    properties attainable by theory exploration on this problem set.
+  properties attainable by theory exploration on the Theory Exploration
+  Benchmark problem set.
+\item Identify the effect of ``toxic'' definitions on system performance, as
+  measured by the Theory Exploration Benchmark.
 \item Demonstrate how the signature selection problem is distinct from the
@@
   algorithms cannot be directly applied to signature selection.
+\item Propose recommendations to take signature selection and toxic definitions
+  into account.
 \end{enumerate}
-
-\iffalse
-TODO: Alison: Are you aiming to produce a set of recommendation for TE
-researchers to follow here?
-
-My thoughts: Not really. We are identifying an open problem (our attempt at a
-solution turns out not to work :( )
-\fi
 

### 2019-10-22 PhDThesis/litreview_testing.tex
Fixes

@@
 choosing a default generator for types implementing its \hs{Arbitrary}
-typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
-\cite{TODO} \fi).
+typeclass (the recent \texttt{hedgehog} package works in a similar way).
 

### 2019-10-22 PhDThesis/support-for-benchmarks/mkBenchmarkGraphs.py
Make textWidth available to other graphs

@@ -241,12 +241,13 @@ import warnings
 warnings.filterwarnings('ignore', module='matplotlib')
 import matplotlib as mpl
 import numpy      as np
+from math import sqrt
 
 def figSize(widthFraction, height=None):
     textWidthPt = float(os.getenv('textWidth'))
     ptToInch    = 1.0 / 72.27
     textWidthIn = textWidthPt * ptToInch
-    goldMean    = (np.sqrt(5.0)-1.0) / 2.0
+    goldMean    = (sqrt(5.0)-1.0) / 2.0
     calcWidth   = widthFraction * textWidthIn
     calcHeight  = textWidthIn * ((goldMean * widthFraction) \
                                  if height is None else height)
@@ -441,8 +442,8 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
 
         # Analytic values, calculated from our model
         anVar     = (p * (1.0 - p)) / count
-        anStddev  = math.sqrt(anVar)
-        anStderr  = anStddev / math.sqrt(count)
+        anStddev  = sqrt(anVar)
+        anStderr  = anStddev / sqrt(count)
 
         # Sample variability, calculated from our data.
         if len(totals) == 1:
@@ -465,7 +466,7 @@ def aggProp(system, sizes=None, agg=None, key=None, total=None):
             denom   = float(len(totals)) - 1.0
             sVar    = (sum([((float(correct) / tot) - p)**2 \
                             for tot, correct in zip(totals, corrects)])) / denom
-            sStddev = math.sqrt(sVar)
+            sStddev = sqrt(sVar)
 
         return {
             'bailed out'        : False,

### 2019-10-22 PhDThesis/support-for-bucketing/boundsData.py

@@ -71,11 +71,12 @@ warnings.filterwarnings('ignore', module='matplotlib')
 # Taken from http://bkanuka.com/articles/native-latex-plots
 import numpy      as np
 import matplotlib as mpl
+from math import sqrt
 def figSize(widthFraction, height=None):
     textWidthPt = float(os.getenv('textWidth'))
     ptToInch    = 1.0 / 72.27
     textWidthIn = textWidthPt * ptToInch
-    goldMean    = (np.sqrt(5.0)-1.0) / 2.0
+    goldMean    = (sqrt(5.0)-1.0) / 2.0
     calcWidth   = widthFraction * textWidthIn
     calcHeight  = textWidthIn * ((goldMean * widthFraction) \
                                  if height is None else height)

### 2019-10-22 PhDThesis/support-for-bucketing/survivalGraph.py

@@ -4,18 +4,33 @@ from os     import getenv
 from pandas import read_csv
 with open(getenv('csv'), 'r') as f:
     times = read_csv(f, header=0, index_col=0)
-
-label = getenv('label')
-
-del(getenv)
 del(read_csv)
+
+label       = getenv('label')
+textWidthPt = float(getenv('textWidth'))
+del(getenv)
+
+from math import sqrt
+
+def figSize(widthFraction, height=None):
+    ptToInch    = 1.0 / 72.27
+    textWidthIn = textWidthPt * ptToInch
+    goldMean    = (sqrt(5.0)-1.0) / 2.0
+    calcWidth   = widthFraction * textWidthIn
+    calcHeight  = textWidthIn * ((goldMean * widthFraction) \
+                                 if height is None else height)
+    return (calcWidth, calcHeight)
 
 from lifelines import KaplanMeierFitter
 import matplotlib.pyplot as plt
 
-def save(name, axes):
-    axes.get_figure().savefig(name + label + '.pdf')
-    axes.get_figure().savefig(name + label + '.pgf')
+def save(name, axes, size=None):
+    fig = axes.get_figure()
+    if size is not None:
+        (w, h) = size
+        fig.set_size_inches(figSize(w, h))
+    fig.savefig(name + label + '.pdf')
+    fig.savefig(name + label + '.pgf')
 
 axes = plt.subplot(111, label="stepped")
 kmf  = KaplanMeierFitter()

### 2019-10-22 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -1,23 +1,38 @@
 #!/usr/bin/env python3
-from os  import getenv
-from csv import DictReader
+from csv  import DictReader
+from math import sqrt
+from os   import getenv
 
-label = getenv('label')
+label       = getenv('label')
+textWidthPt = float(getenv('textWidth'))
 
 def msg(x):
     from sys import stderr
     stderr.write(repr(x))
     stderr.flush()
 
-def save(name, axes):
-    axes.get_figure().savefig(name + label + '.pdf')
-    axes.get_figure().savefig(name + label + '.pgf')
+def figSize(widthFraction, height=None):
+    ptToInch    = 1.0 / 72.27
+    textWidthIn = textWidthPt * ptToInch
+    goldMean    = (sqrt(5.0)-1.0) / 2.0
+    calcWidth   = widthFraction * textWidthIn
+    calcHeight  = textWidthIn * ((goldMean * widthFraction) \
+                                 if height is None else height)
+    return (calcWidth, calcHeight)
+
+def save(name, axes, size=None):
+    fig = axes.get_figure()
+    if size is not None:
+        (w, h) = size
+        fig.set_size_inches(figSize(w, h))
+    fig.savefig(name + label + '.pdf')
+    fig.savefig(name + label + '.pgf')
 
 with open(getenv('csv'), 'r') as f:
     times = DictReader(f)
     rows  = [row for row in times]
 
-del(getenv  )
+del(getenv)
 del(DictReader)
 
 counts = {}

### 2019-10-22 PhDThesis/support-for-bucketing/survivalGraph.py
Adjust scale of graphs
We want the two survival plots in a column next to the two failure lines. This
requires that each graph be at most 50% of the width.

We can set the width (from which the height is calculated via the golden ratio)
in the Python code. The larger this is, the more spread-out the graph contents
are.

We can use a scalebox in LaTeX to grow/shrink the graphs. This doesn't affect
the graph's layout, but it does make the whole thing smaller.

We want the width-proportion (in Python) * scale factor (in LaTeX) to be at most
0.5. This version uses 0.75 in Python and 0.66 in LaTeX

@@ -39,7 +39,7 @@ for s, grouped_df in times.groupby('size'):
         continue
     kmf.fit(grouped_df['time'], grouped_df['success'], label=str(s))
     axes = kmf.plot(ax=axes)
-save('stepped', axes)
+save('stepped', axes, size=(0.75, None))
 
 def crossValidate(name, fitter):
     from lifelines.utils import k_fold_cross_validation

### 2019-10-22 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -75,5 +75,5 @@ plt.xticks(xs)
 plt.xlabel('size')
 plt.ylabel('failure proportion')
 plt.title('Failures as size grows')
-save('timeouts', plt.gca())
+save('timeouts', plt.gca(), size=(0.75, None))
 #plt.savefig('timeouts.pdf')

### 2019-10-22 PhDThesis/support-for-bucketing/boundsData.py
Add TODO to hacked sizes

@@ -80,7 +80,7 @@ def figSize(widthFraction, height=None):
     calcWidth   = widthFraction * textWidthIn
     calcHeight  = textWidthIn * ((goldMean * widthFraction) \
                                  if height is None else height)
-    return (10, 10) #(calcWidth, calcHeight)
+    return (10, 10) # TODO: (calcWidth, calcHeight)
 
 mpl.rcParams.update({
     # 'pgf.texsystem' : 'pdflatex',

### 2019-10-22 PhDThesis/support-for-bucketing/survivalGraph.py
Add labels

@@ -39,6 +39,8 @@ for s, grouped_df in times.groupby('size'):
         continue
     kmf.fit(grouped_df['time'], grouped_df['success'], label=str(s))
     axes = kmf.plot(ax=axes)
+axes.set_xlabel('Time (seconds)')
+axes.set_ylabel('Proportion still running')
 save('stepped', axes, size=(0.75, None))
 
 def crossValidate(name, fitter):

### 2019-10-23 PhDThesis/support-for-bucketing/survivalGraph.py
Make graph layout "tight"

@@ -29,8 +29,8 @@ def save(name, axes, size=None):
     if size is not None:
         (w, h) = size
         fig.set_size_inches(figSize(w, h))
-    fig.savefig(name + label + '.pdf')
-    fig.savefig(name + label + '.pgf')
+    fig.savefig(name + label + '.pdf', bbox_inches='tight', pad_inches=0.0)
+    fig.savefig(name + label + '.pgf', bbox_inches='tight', pad_inches=0.0)
 
 axes = plt.subplot(111, label="stepped")
 kmf  = KaplanMeierFitter()

### 2019-10-23 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -25,8 +25,8 @@ def save(name, axes, size=None):
     if size is not None:
         (w, h) = size
         fig.set_size_inches(figSize(w, h))
-    fig.savefig(name + label + '.pdf')
-    fig.savefig(name + label + '.pgf')
+    fig.savefig(name + label + '.pdf', bbox_inches='tight', pad_inches=0.0)
+    fig.savefig(name + label + '.pgf', bbox_inches='tight', pad_inches=0.0)
 
 with open(getenv('csv'), 'r') as f:
     times = DictReader(f)

### 2019-10-23 PhDThesis/support-for-bucketing/survivalGraph.py
Fiddle graph size and make square
Golden mean is nice, but things were a bit too squashed vertically (e.g. the
legend on survival plots overlapping the lines)

@@ -41,7 +41,7 @@ for s, grouped_df in times.groupby('size'):
     axes = kmf.plot(ax=axes)
 axes.set_xlabel('Time (seconds)')
 axes.set_ylabel('Proportion still running')
-save('stepped', axes, size=(0.75, None))
+save('stepped', axes, size=(0.72, 0.72))
 
 def crossValidate(name, fitter):
     from lifelines.utils import k_fold_cross_validation

### 2019-10-23 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -75,5 +75,5 @@ plt.xticks(xs)
 plt.xlabel('size')
 plt.ylabel('failure proportion')
 plt.title('Failures as size grows')
-save('timeouts', plt.gca(), size=(0.75, None))
+save('timeouts', plt.gca(), size=(0.72, 0.72))
 #plt.savefig('timeouts.pdf')

### 2019-10-23 PhDThesis/bucketing.tex
Adjust scale of graphs
We want the two survival plots in a column next to the two failure lines. This
requires that each graph be at most 50% of the width.

We can set the width (from which the height is calculated via the golden ratio)
in the Python code. The larger this is, the more spread-out the graph contents
are.

We can use a scalebox in LaTeX to grow/shrink the graphs. This doesn't affect
the graph's layout, but it does make the whole thing smaller.

We want the width-proportion (in Python) * scale factor (in LaTeX) to be at most
0.5. This version uses 0.75 in Python and 0.66 in LaTeX

@@
   \centering
-  \scalebox{0.5}{\input{images/steppedall.pgf}}
-  \scalebox{0.5}{\input{images/timeoutsall.pgf}}
-  \scalebox{0.5}{\input{images/steppednontoxic.pgf}}
-  \scalebox{0.5}{\input{images/timeoutsnontoxic.pgf}}
-  \caption{Kaplan-Meier survival plot for running \quickspec{} on inputs
-    containing various numbers of definitions, sampled from TIP. The x axis
-    denotes running time, which we cut short after 300 seconds. The height of
-    each line shows the proportion of \quickspec{} runs which were still going
-    at that time (lower is better). First plot is for all TIP definitions,
-    second plot removes runs given ``toxic'' definitions.
-    \iffalse
-    TODO: Alison: The key needs further explanation (either here or in the caption)
-    TODO: Alison: Label the y axis (on both)
-    \fi}
+  \scalebox{0.69}{\input{images/steppedall.pgf}}
+  \scalebox{0.69}{\input{images/timeoutsall.pgf}}
+  \scalebox{0.69}{\input{images/steppednontoxic.pgf}}
+  \scalebox{0.69}{\input{images/timeoutsnontoxic.pgf}}
+
+  % NOTE: The list-of-figures description given as an optional argument to
+  % \caption is needed to allow multiple paragraphs without error.
+  \caption[Survival plots for \quickspec{} running on different sized
+  signatures.]{Running times when using \quickspec{} to explore signatures
+    sampled from the Theory Exploration Benchmark, The top row includes all
+    sampled signatures, whilst the second row excludes those containing ``toxic''
+    definitions.
+
+    \textbf{Left} Kaplan-Meier survival plots for exploring signatures sampled
+    from our TEB. Time runs horizontally, which we cut short after 300 seconds.
+    Each line tracks a population of signatures of a given size (shown in the
+    legend), as they're explored using \quickspec{}. The height of each line
+    shows the proportion of processes still running after that amount of time
+    (lower is better). Shaded regions denote 95\% confidence intervals.
+
+    \textbf{Right} Proportion of each population which timed out after 300
+    seconds, for signatures of different size. This corresponds to a slice
+    through the corresponding survival plot at the 300 second mark (if all of
+    the measured populations were included). Lines of best-fit use linear
+    regression, providing an estimate of the effect that ``toxic'' definitions
+    have on failure (timeout) rate.}
+    % TODO Put line slopes and error on graph
+    % TODO Bottom row should use a second set of samples (chosen to avoid toxic
+    % definitions), rather than just filtering down
   \label{fig:survival}

### 2019-10-23 PhDThesis/support-for-bucketing/survivalGraph.py
Force legend into upper right

@@ -41,6 +41,7 @@ for s, grouped_df in times.groupby('size'):
     axes = kmf.plot(ax=axes)
 axes.set_xlabel('Time (seconds)')
 axes.set_ylabel('Proportion still running')
+axes.legend(loc='upper right')
 save('stepped', axes, size=(0.72, 0.72))
 
 def crossValidate(name, fitter):

### 2019-10-23 PhDThesis/support-for-bucketing/timeoutGraph.py
Remove title

@@ -74,6 +74,4 @@ plt.ylim((0,1 ))
 plt.xticks(xs)
 plt.xlabel('size')
 plt.ylabel('failure proportion')
-plt.title('Failures as size grows')
 save('timeouts', plt.gca(), size=(0.72, 0.72))
-#plt.savefig('timeouts.pdf')

### 2019-10-23 PhDThesis/support-for-bucketing/survivalGraph.py
More evenly-spaced ticks

@@ -42,6 +42,7 @@ for s, grouped_df in times.groupby('size'):
 axes.set_xlabel('Time (seconds)')
 axes.set_ylabel('Proportion still running')
 axes.legend(loc='upper right')
+plt.yticks([0.0, 0.25, 0.5, 0.75, 1.0])
 save('stepped', axes, size=(0.72, 0.72))
 
 def crossValidate(name, fitter):

### 2019-10-23 PhDThesis/support-for-bucketing/timeoutGraph.py

@@ -69,9 +69,12 @@ ys = [f(x) for x in xs         ]
 
 ax = plt.plot(xs, ys)
 
+even = lambda x: x % 2 == 0
+
 plt.xlim((1,20))
 plt.ylim((0,1 ))
-plt.xticks(xs)
-plt.xlabel('size')
-plt.ylabel('failure proportion')
+plt.yticks([0.0, 0.25, 0.5, 0.75, 1.0])
+plt.xticks(list(filter(even, xs)))  # Reduce clutter
+plt.xlabel('Signature size')
+plt.ylabel('Failure proportion')
 save('timeouts', plt.gca(), size=(0.72, 0.72))

### 2019-10-25 PhDThesis/background_theoryexploration.tex
Add and rejiggle content

@@
 (Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
-of definitions \iffalse TODO: Alison Some tools generate their own definitions
-\fi in some formal system (for example a programming language) and
+of definitions in some formal system (for example a programming language) and
 automatically generating a set of formal statements (properties) involving
-those definitions. These may be conjectures or theorems (proven either by
-sending conjectures to an automated theorem prover, or by having the generating
-procedure proceed in logically sound steps from a set of axioms). These
-statements should also be
-``interesting'' in some way, to rule out unhelpful trivialities such as
-iterating a pattern over and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x},
-\dots).
-
-\iffalse
-It is beyond the scope of this paper
-to define what makes a mathematical statement ``interesting'', but this problem
-has been tackled extensively in the literature~\cite{colton2000notion}.
-\fi
-% END TOTO
+those definitions\footnote{Some tools, such as \isascheme{}, also generate their
+  own definitions. This work does not focus on such functionality, treating it
+  instead under the umbrella of \emph{Concept Formation}.}. These may be
+conjectures or theorems (proven either by sending conjectures to an automated
+theorem prover, or by having the generating procedure proceed in logically sound
+steps from a set of axioms). These statements should also be ``interesting'' in
+some way, to rule out unhelpful trivialities such as iterating a pattern over
+and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x}, \dots). We defer a more
+thorough treatment of what counts as ``interesting'' to
+\S~\ref{sec:interestingness}.
 
@@
 in our experiments, due to its availability, stability and tooling integration;
-hence references to \quickspec{} are to version 1 unless stated otherwise. At
-the time of writing \iffalse TODO: Alison: Include date \fi there is a
-\quickspec{} version 2 available, which has a much improved generation procedure
-which can be significantly faster than its predecessor. Whilst \quickspec{} 2
-has advanced the state of the art in theory exploration, our preliminary
-experience has found that it still suffers the scaling issues we identify in
-this work\footnote{Not only did we experience failures with our own experiments,
-  we also suffered memory exhaustion when trying to run some of the examples
-  included with the \quickspec{} 2 source code.} Since \quickspec{} 2 has less
-integration with the tooling we have used, we leave a more thorough analysis of
-it (and related systems such as \speculate{}) for future work, pending the
-necessary infrastructure changes that would require.
+hence references to \quickspec{} are to version 1 unless stated otherwise. As of
+2019 there is a \quickspec{} version 2 available, which has a much improved
+generation procedure which can be significantly faster than its predecessor.
+Whilst \quickspec{} 2 has advanced the state of the art in theory exploration,
+our preliminary experience has found that it still suffers the scaling issues we
+identify in this work\footnote{Not only did we experience failures with our own
+  experiments, we also suffered memory exhaustion when trying to run some of the
+  examples included with the \quickspec{} 2 source code.} Since \quickspec{} 2
+has less integration with the tooling we have used, we leave a more thorough
+analysis of it (and related systems such as \speculate{}) for future work,
+pending the necessary infrastructure changes that would require.
 
@@
 
-\quickspec{} (version 1) is written in the Haskell programming language, and
-works with definitions which are also written in Haskell. A thorough description
-of Haskell, including its suitability for theory exploration, can be found
-in \S~\ref{sec:haskell}. For illustrative purposes, some simple Haskell
-definitions are shown in Figure~\ref{fig:haskellteexample}. \quickspec{} uses
-the following procedure to generate conjectures, which are both plausible (due
-to testing on many examples) and potentially interesting (due to being mutually irreducible):
+\quickspec{} is written in the Haskell programming language, and works with
+definitions which are also written in Haskell. A thorough description of
+Haskell, including its suitability for theory exploration, can be found in
+\S~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
+are shown in Figure~\ref{fig:haskellteexample}. \quickspec{} uses the following
+procedure to generate conjectures, which are both plausible (due to testing on
+many examples) and potentially interesting (due to being mutually irreducible):
 
@@
 
-The elements of $terms_{\hs{Nat}}$ will be grouped into five classes, one each
-for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and
-\hs{Bool}. As the variables \hs{a}, \hs{b} and \hs{c} are instantiated to
+The elements of $terms_{\hs{Nat}}$ will be grouped into five equivalence classes,
+one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool}
+and \hs{Bool}. As the variables \hs{a}, \hs{b} and \hs{c} are instantiated to
 various randomly-generated numbers, the differences between these terms will
-be discovered and the equivalence classes will be divided, until eventually
-the equations in Figure \ref{fig:qspecresult} are conjectured.
+be discovered. For example, the \hs{Nat} class will initially contain (among
+other things) the terms \hs{Z}, \hs{x} and \hs{plus x Z}; this will be split
+apart once \hs{x} is tested with a non-zero value, since the latter terms will
+not equal \hs{Z}. Since those latter terms will always be equal, for any value
+of \hs{x}, they will never be split apart. Eventually (after removing
+redundancies) the equations in Figure~\ref{fig:qspecresult} are conjectured.
+These may be output to the user for perusal, or for another system to process
+like the \hipspec{} automated theorem prover.
+
+Note that we must be able to generate values of a type in order to include
+variables of that type. Also, only equivalence classes whose element type can be
+compared are split and conjectured about in this way\footnote{Both random data
+  generators and comparison functions can be derived mechanically from algebraic
+  datatype declarations; hence why they are omitted from our example. Note that
+  care must be taken when generating recursive structures, as na\"ive recursion
+  may produce values whose expected size exceeds the machine's memory.}; in
+particular, functions (like \hs{odd} and \hs{even}) are first-class values, so
+they will be put in an equivalence class, but they will not be subject to
+testing and comparison since Haskell has no generic notion of function
+equality~\footnote{In fact \quickspec{} compares expressions using a total
+  ordering rather than equality; this is even more restrictive than requiring a
+  notion of equality, but reduces the number of required comparisons}.
+\quickspec{} can hence conjecture that two functions are pointwise-equal (e.g.
+\hs{f x == g x}) but it cannot conjecture that the functions themselves are
+equal (e.g. \hs{f == g}).
 
@@
 are closely related, and hence obey many algebraic laws; arbitrary definitions
-from a typical software library or proof development are unlikely to have be
-related so strongly.
+from a typical software library or proof development are unlikely to be related
+so strongly.
 
@@
 and the consequences for testing and QA in general.
-\fi
-
-\iffalse
-% TODO: Work this into the example
-
-As an example, giving the definitions from Figure~\ref{fig:haskellteexample} to
-\quickspec{}, along with suitable comparison functions, random data generators and
-variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
-class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
-\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
-(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
-\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
-them up, and repeat until no more splits are found. Equations are then generated
-between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
-\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
-to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
-instance of the more general \hs{b == not (not b)}). The remaining equations
-(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
-output, either for the user to peruse or for another system to process, like the
-\hipspec{} automated theorem prover.
-
-Note that we must be able to generate values of a type in order to include
-variables of that type. Also, only equivalence classes whose element type can be
-compared are split and conjectured about in this way; in particular, functions
-(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
-equivalence class, but they will not be subject to testing and comparison since
-Haskell has no generic notion of function equality~\footnote{In
-  fact \quickspec{} compares expressions using a total ordering rather than equality;
-  this is even more restrictive than requiring a notion of equality, but reduces
-  the number of required comparisons}. \quickspec{} can hence conjecture that two
-functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
-that the functions themselves are equal (e.g. \hs{f == g}).
 \fi
@@
 % Communication: Latex, Wikis, APIs, communicating with aliens
-
-% \section{Exploration in Theorem Proving}
-% \label{sec:examples}
-
-% Before exploring abstract definitions of interestingness, we can first consider
-% some scenarios which arise during formal proof where we are forced to generate
-% conjectures. An analysis of these situations, and the subsequent theorems they
-% produce, will contribute towards an empirical justification for what is
-% interesting (at least from a utilitarian point of view) and inform our later
-% exploration of the literature.
-
-% \subsection{Generalisation}
-
-% \providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
-
-% When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
-% which $S$ is a special case. Although it seems counterintuitive, a generalised
-% statement can sometimes be \emph{easier} to prove than the original. This arises
-% often in inductive proofs, since the specific obligations which arise in the
-% proof may be incompatible with the available inductive hypotheses.
-
-% However, we cannot blindly generalise \emph{all} obligations we encounter, since
-% \emph{over-generalising} results in obligations which are so strong that they
-% are unprovable, or even false. We must therefore rely on heuristics to guide the
-% generation of generalised conjectures, and hence perform a kind of exploration.
-
-% An informative example is given by Boyer and Moore of the associativity of
-% multiplication in ACL2 \cite{boyer1983proof}:
-
-% $$(x * y) * z = x * (y * z)$$
-
-% During the course of the proof, the following obligation arises:
-
-% \begin{equation}
-%   \tag{conc3}
-%   (y + (x * y)) * z = (y * z) + ((x * y) * z)
-%   \label{eq:conc3}
-% \end{equation}
-
-% ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
-% sub-term $x * y$ with a fresh variable $w$:
-
-% \begin{equation}
-%   \tag{conc4}
-%   (y + w) * z = (y * z) + (w * z)
-%   \label{eq:conc4}
-% \end{equation}
-
-% This generalised form is clearly the distributivity law for multiplication and
-% addition, which can be proved separately to the original goal of
-% associativity. It would not be controversial to claim that this distributivity
-% law is interesting in its own right (relative to associativity, at least), in
-% addition to its usefulness in making this proof go through.
-
-% % TODO: Describe the ACL2 heuristics
-
-% Generalisation also occurs frequently when reasoning about \emph{tail-recursive}
-% definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can
-%   be executed in constant space using a loop, whereas recursion in non-tail
-%   positions may require a growing number of stack frames or nested closures. See
-%   \S \ref{sec:auxiliarylemmas} for example definitions of each type.}
-
-% \subsection{Analogy}
-
-% One way to characterise the interestingness of a statement is by \emph{analogy}
-% to existing interesting statements. By finding lemmas analogous to those of a
-% different theory, we may be able to re-use tactics and other forms of
-% meta-programming across both.
-
-% Existing theory exploration systems have been successfully applied to this
-% problem, however the use of pure exploration misses opportunities to
-% \emph{focus} the search, since we know which lemmas are used in those theories
-% where a technique succeeded. If we can find an analogy to map from such solved
-% problems to our unsolved goal, we can infer the approximate form of the lemmas
-% we require, and target these specifically.
-
-% The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
-% to solving a goal $G$ by making analogies via unsupervised clustering
-% \cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are used in two
-% ways:
-
-% \begin{itemize}
-% \item First, we use the cluster $C_G$ containing $G$ to identify analogous
-%   theorems.
-
-% \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
-%   $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
-%   to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
-%   the cluster $C_s$ containing $s$.
-% \end{itemize}
-
-% The running examples for demonstrating \textsc{ACL2(ml)} are equivalence
-% theorems for tail-recursive and non-tail-recursive calculations, as well as the
-% effect of repeating certain list operations:
-
-% \begin{itemize}
-
-%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
-
-%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
-
-%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
-
-%   \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
-
-%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
-% \end{itemize}
-
-% \subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
-
-% One consideration when generating conjectures is the difference between
-% theorems, lemmas, corollaries, etc. From a logical point of view, these are all
-% equivalent, and hence most proof assistants do not distinguish between
-% them. However, their \emph{intention} may be different: in a sense, theorems are
-% the interesting results; whilst lemmas are useful results, required for proving
-% the theorems.
-
-% Some systems, like Coq, allow users to \emph{label} each statement as being a
-% \coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
-% the same. This shows us immediately that lemmas outnumber theorems; in the Coq
-% standard library there are over five times as many lemmas as theorems
-% \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when
-%   excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of
-%   \coq{Lemma} in its \texttt{theories/} directory.}.
-
-% % TODO: Analyse them
-
-% % TODO: Theory exploration as lemma generation; give example from a HipSpec paper
-
-% We can find a need for auxiliary lemmas, once again, in the context of
-% tail-recursive functions. Consider proving the (pointwise) equality of the
-% following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
-
-% \begin{coqblock}
-% Inductive Nat : Set := Z : Nat
-%                      | S : Nat -> Nat.
-
-% Fixpoint plus      (n m : Nat) := match n with
-%                                       | Z    => m
-%                                       | S n' => S (plus n' m)
-%                                   end.
-
-% Fixpoint plus_tail (n m : Nat) := match n with
-%                                       | Z    => m
-%                                       | S n' => plus_tail n' (S m)
-%                                   end.
-% \end{coqblock}
-
-% \begin{haskell}
-% Haskell equivalent:
-
-% plus :: Nat -> Nat -> Nat
-% plus      n  Z    = n
-% plus      n (S m) = S (plus n m)
-
-% plus_tail :: Nat -> Nat -> Nat
-% plus_tail n  Z    = n
-% plus_tail n (S m) = plus_tail (S n) m
-% \end{haskell}
-
-% Both of these functions implement addition, but the \coq{plus_tail} variant is
-% tail-recursive. However, if we want to \emph{prove} that the definitions are
-% (pointwise) equal, we run into difficulties. In particular, when the inductive
-% step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
-% reasonable), we cannot make this go through using another inductive argument.
-
-% \begin{coqblock}
-% (* Solve equalities by beta-normalising both sides *)
-% Ltac triv := try (simpl; reflexivity).
-
-% (* Prove equivalence of plus and plus_tail *)
-% Theorem equiv : forall n m, plus n m = plus_tail n m.
-%   induction n; triv. (* Base case is trivial *)
-
-%   (* Inductive case: plus (S n) m = plus_tail (S n) m *)
-%   intro m.
-
-%   (* Beta-reduce the right-hand-side (justification is trivial) *)
-%   replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
-
-%   (* Use induction hypothesis to replace plus_tail with plus *)
-%   rewrite <- (IHn (S m)).
-% \end{coqblock}
-
-% Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
-% the equation we need:
-
-% \begin{coqblock}
-% IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
-% \end{coqblock}
-
-% Yet we cannot provide it with the argument it needs, as our original induction
-% hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):
-
-% \begin{coqblock}
-% IHn : forall x, plus (S n') x = plus_tail (S n') x
-% \end{coqblock}
-
-% We are forced to abandon the proof, despite such a reasonable-looking
-% intermediate goal.
-
-% In fact, if we attempt to prove that goal \emph{separately}, we can use a
-% straightforward argument by induction; even though it is actually
-% \emph{stronger} due to the absence of the \coq{IHn} assumption. Using this
-% separate result as a lemma, the pointwise equality is proven easily.
-
-% \begin{coqblock}
-% Lemma gen n m : plus (S n) m = plus n (S m).
-%   induction n; triv. (* Base case is trivial *)
-
-%   (* Move all S constructors outside *)
-%   simpl. rewrite <- IHn. simpl.
-
-%   (* Trivial *)
-%   reflexivity.
-% Defined.
-% \end{coqblock}
-
-% \begin{coqblock}
-%   rewrite (gen n m).
-%   reflexivity.
-% Defined.
-% \end{coqblock}

### 2019-10-25 PhDThesis/litreview_interestingness.tex
Fix reference
Add and rejiggle content

@@
-\subsection{Interestingness}
+\section{Interestingness}
+\label{sec:interestingness}
 
-A major challenge when generating conjectures is choosing how to narrow down the
-resulting set to only those deemed ``interesting'', since this is an imprecise
+A major challenge when generating conjectures is choosing how to focus attention
+on only those deemed ``interesting'' (either by avoiding uninteresting search
+paths, and/or by filtering down the eventual output), since this is an imprecise
 term with many different interpretations. For example, all existing approaches
 agree that simple tautologies are ``uninteresting'', but differ when it comes to
-more complex statements.  Colton \etal{} surveyed tools for theory formation
-and exploration, and their associated notions of ``interestingness'' for
+more complex statements.  Colton~\etal{} surveyed tools for theory formation
+and exploration, and their associated notions of ``interestingness'', for
 concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
@@
 
-% FIXME: This is original work, so should be in a different section
+% FIXME: This is original work, so should maybe be in a different section
 \begin{table}
@@
 in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
-based on this survey (the latter is Colton's own tool). We extend this analysis
-to some more recent tools in the remaining rows (based on our understanding of
-their function): \quickspec{}~\cite{QuickSpec},
+based on this survey (the latter is Colton's own tool). The remaining rows are
+our extension of this analysis to the (higher-order) tools we have focused on
+(based on our understanding of their function): \quickspec{}~\cite{QuickSpec},
 \speculate{}~\cite{braquehais2017speculate},
 \isacosy{}~\cite{Johansson.Dixon.Bundy:conjecture-generation}
-and \isascheme{}~\cite{MontanoRivas2011}.
+and \isascheme{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}.
 
@@
 
+\subsection{Exploration in Theorem Proving}
+\label{sec:examples}
+
+Conjecture generation (automated or manual) is crucial when formally proving
+statements, such verifying software against its specification. This provides a
+concrete, empirical setting to determine what makes statements interesting (at
+least from a utilitarian point of view).
+
+\paragraph{Generalisation}
+
+\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}
+
+When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
+which $S$ is a special case. Although it seems counterintuitive, a generalised
+statement can sometimes be \emph{easier} to prove than the original. This arises
+often in inductive proofs (e.g. when reasoning about tail-recursive
+definitions~\cite{kapur2003automatic}), since the specific obligations which
+arise in the proof may be incompatible with the available inductive hypotheses.
+An informative example is given by Boyer and Moore of the associativity of
+multiplication in ACL2~\cite{boyer1983proof}:
+
+$$(x * y) * z = x * (y * z)$$
+
+During the course of the proof, the following obligation arises:
+
+\begin{equation}
+  \tag{conc3}
+  (y + (x * y)) * z = (y * z) + ((x * y) * z)
+  \label{eq:conc3}
+\end{equation}
+
+ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
+sub-term $x * y$ with a fresh variable $w$:
+
+\begin{equation}
+  \tag{conc4}
+  (y + w) * z = (y * z) + (w * z)
+  \label{eq:conc4}
+\end{equation}
+
+This generalised form is clearly the distributivity law for multiplication and
+addition, which can be proved separately to the original goal of
+associativity. It would not be controversial to claim that this distributivity
+law is interesting in its own right (relative to associativity, at least), in
+addition to its usefulness in making this proof go through.
+
+% TODO: Describe the ACL2 heuristics
+
+Generality hence contributes to a conjecture's \emph{utility}, and also (by
+definition) to its \emph{applicability}. It is still important to balance
+against other concerns, since blindly generalising \emph{all} statements of
+proof obligations we come across would result in conjectures so strong that they
+are unprovable, or even false; which would make them much less interesting.
+
+\paragraph{Analogy}
+
+The interestingness of a statement can also be characterised by \emph{analogy}
+to existing interesting statements. By finding lemmas analogous to those of a
+different theory, we may be able to re-use proof tactics and other forms of
+meta-programming across both.
+
+Existing theory exploration systems have been successfully applied to this
+problem, however the use of pure exploration misses opportunities to
+\emph{focus} the search, since we know which lemmas are used in those theories
+where a technique succeeded. If we can find an analogy to map from such solved
+problems to a currently-unproven goal, we can infer the approximate form of the
+lemmas we require, and target these specifically.
+
+The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
+to solving a goal $G$ by making analogies via unsupervised
+clustering~\cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are
+used in two ways:
+
+\begin{itemize}
+\item First, we use the cluster $C_G$ containing $G$ to identify analogous
+  theorems.
+
+\item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
+  $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
+  to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
+  the cluster $C_s$ containing $s$.
+\end{itemize}
+
+The interestingness of such generated lemmas derives from both the existence of
+the original lemma (which is presumably interesting enough to prove and
+subsequently explore), that of the target theorem it may help to prove, and the
+novelty provided by reframing existing concepts in a new domain. The running
+examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
+tail-recursive and non-tail-recursive calculations\footnote{A tail-recursive
+  function can be executed in constant space using a loop, whereas recursion in
+  non-tail positions may require a growing number of stack frames or nested
+  closures.}), as well as the effect of repeating certain list operations:
+
+\begin{itemize}
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
+
+  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
+
+  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
+
+  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
+
+\end{itemize}
+
+\paragraph{Auxiliary Lemmas} \label{sec:auxiliarylemmas}
+
+One consideration when generating conjectures is the difference between
+theorems, lemmas, corollaries, etc. From a logical point of view, these are all
+equivalent, and hence most proof assistants do not distinguish between
+them. However, their \emph{intention} may be different: in a sense, theorems are
+the interesting results; whilst lemmas are useful results, required for proving
+the theorems. Corollaries are consequences and special-cases of theorems, which
+are interesting enough to state separately, but not as much as a theorem.
+
+Some systems, like Coq, allow users to \emph{label} each statement as being a
+\coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
+the same. This shows us immediately that lemmas outnumber theorems; in the Coq
+standard library there are over five times as many lemmas as theorems
+\footnote{Coq version \texttt{8.4pl6} contains (excluding comments) 1492
+  occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/}
+  directory.}.
+
+% TODO: Analyse them
+
+% TODO: Theory exploration as lemma generation; give example from a HipSpec paper
+
+We can find a need for auxiliary lemmas, once again, in the context of
+tail-recursive functions. Consider proving the (pointwise) equality of the
+following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:
+
+\begin{coqblock}
+Inductive Nat : Set := Z : Nat
+                     | S : Nat -> Nat.
+
+Fixpoint plus      (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => S (plus n' m)
+                                  end.
+
+Fixpoint plus_tail (n m : Nat) := match n with
+                                      | Z    => m
+                                      | S n' => plus_tail n' (S m)
+                                  end.
+\end{coqblock}
+
+These definitions are equivalent to the following Haskell:
+
+\begin{haskell}
+plus :: Nat -> Nat -> Nat
+plus      n  Z    = n
+plus      n (S m) = S (plus n m)
+
+plus_tail :: Nat -> Nat -> Nat
+plus_tail n  Z    = n
+plus_tail n (S m) = plus_tail (S n) m
+\end{haskell}
+
+Both of these functions implement addition, but the \coq{plus_tail} variant is
+tail-recursive. However, if we want to \emph{prove} that the definitions are
+(pointwise) equal, we run into difficulties. In particular, when the inductive
+step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
+reasonable), we cannot make this go through using another inductive argument.
+
+\begin{coqblock}
+(* Solve equalities by beta-normalising both sides *)
+Ltac triv := try (simpl; reflexivity).
+
+(* Prove equivalence of plus and plus_tail *)
+Theorem equiv : forall n m, plus n m = plus_tail n m.
+  induction n; triv. (* Base case is trivial *)
+
+  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
+  intro m.
+
+  (* Beta-reduce the right-hand-side (justification is trivial) *)
+  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.
+
+  (* Use induction hypothesis to replace plus_tail with plus *)
+  rewrite <- (IHn (S m)).
+\end{coqblock}
+
+Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
+the equation we need:
+
+\begin{coqblock}
+IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
+\end{coqblock}
+
+Yet we cannot provide it with the argument it needs, as our original induction
+hypothesis is \emph{too specific} (it has too many \coq{S} constructors):
+
+\begin{coqblock}
+IHn : forall x, plus (S n') x = plus_tail (S n') x
+\end{coqblock}
+
+We are forced to abandon the proof, despite such a reasonable-looking
+intermediate goal.
+
+In fact, if we attempt to prove that goal \emph{separately}, we can use a
+straightforward argument by induction; even though it is actually
+\emph{stronger} due to the absence of the \coq{IHn} assumption:
+
+\begin{coqblock}
+Lemma gen n m : plus (S n) m = plus n (S m).
+  induction n; triv. (* Base case is trivial *)
+
+  (* Move all S constructors outside *)
+  simpl. rewrite <- IHn. simpl.
+
+  (* Trivial *)
+  reflexivity.
+Defined.
+\end{coqblock}
+
+Using this separate result as a lemma, the orignal pointwise equality is proven
+easily:
+
+\begin{coqblock}
+  rewrite (gen n m).
+  reflexivity.
+Defined.
+\end{coqblock}
+
+Hence demonstrating the utility, and therefore interestingness, of such
+auxilliary lemmas.
+
 % TODO Include Colton/Bundy et al

### 2019-10-25 PhDThesis/litreview_testing.tex
Rejig

@@
 randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
-the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
-\hs{Tree} example this might check \hs{Leaf}, then \hs{Node Leaf
-  Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
-order of ``size'' ensures that counterexamples are always small, that tests are
-not repeated with the same values and that memory is not exhausted (since the
-stopping condition will trigger long before the enumeration reaches such large
-values). The disadvantage is that large inputs (which exercise properties,
-especially recursive ones, the most) may occur so late in the enumeration that
-testing never reaches them.
+the \texttt{testing-feat} package). In the \hs{Tree} example this might check
+\hs{Leaf}, then \hs{Node Leaf Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and
+so on. Enumerating in order of ``size'' ensures that counterexamples are always
+small, that tests are not repeated with the same values and that memory is not
+exhausted (since the stopping condition will trigger long before the enumeration
+reaches such large values). The disadvantage is that large inputs (which
+exercise properties, especially recursive ones, the most) may occur so late in
+the enumeration that testing never reaches them.
 

### 2019-10-25 PhDThesis/related_tests.tex
Rejig

@@
-\subsection{Tests} \label{sec:tests}
+\section{Tests} \label{sec:tests}
 
 % TODO: This is not related work!
-
-Since we are working in the domain of Haskell programs, an abundant source of
-statements are available in the form of \emph{tests}. For manually-written
-tests, the effort required to write them implies that they must be of some
-interest to their author. Many forms of test are \emph{not} suitable for our
-purposes, such as \emph{unit tests} which are always trivially provable by
-$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
-the external environment. One reason to study Haskell is its widespread use of
-\emph{property checking}, which does give us useful data. Many Haskell property
-checkers exist, based on random testing
-(\quickcheck{}~\cite{claessen2011quickcheck} and
-\smartcheck{}~\cite{pike2014smartcheck}), enumeration
-(\smallcheck{}~\cite{runciman2008smallcheck}), observation
-(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
-(\sparsecheck{}~\cite{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they
-instantiate test arguments; their representations of properties are largely the
-same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \hackage{} (as of
-2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
-\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
-\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
-\texttt{blaze-html}.
-
-% TODO: Analysis

### 2019-10-25 PhDThesis/bucketing.tex
Wording
Bucketing content

@@
 definitions (based on experiments with our Theory Exploration Benchmark, in
-\S~\ref{sec:benchmark}. To avoid this, users of these systems must carefully
+\S~\ref{sec:benchmark}). To avoid this, users of these systems must carefully
 select only small subsets of their definitions to explore at a time.
@@
 exploration, and we analyse its effect on the running time of theory exploration
-tools on the Theory Exploration Benchmark, and on the quality of the statements
-they are able to generate. An automated solution to this signature selection
-problem would accept a large number of definitions and efficiently select
-sub-sets which are both small enough to reasonably explore, whilst still
+tools on our Theory Exploration Benchmark (TEB), and on the quality of the
+statements they are able to generate. An automated solution to this signature
+selection problem would accept a large number of definitions and efficiently
+select sub-sets which are both small enough to reasonably explore, whilst still
 enabling interesting properties to be discovered.
@@
 \item Analyse the effect of signature selection on the quantity of desirable
-  properties attainable by theory exploration on the Theory Exploration
-  Benchmark problem set.
+  properties attainable by theory exploration on the TEB problem set.
 \item Identify the effect of ``toxic'' definitions on system performance, as
-  measured by the Theory Exploration Benchmark.
+  measured by the TEB.
+\item Recommend improvements for theory exploration systems, to avoid or
+  mitigate the problem of toxic definitions.
 \item Demonstrate how the signature selection problem is distinct from the
@@
   algorithms cannot be directly applied to signature selection.
-\item Propose recommendations to take signature selection and toxic definitions
-  into account.
 \end{enumerate}
@@
     through the corresponding survival plot at the 300 second mark (if all of
-    the measured populations were included). Lines of best-fit use linear
-    regression, providing an estimate of the effect that ``toxic'' definitions
-    have on failure (timeout) rate.}
+    the measured populations were included). Lines of best-fit use least-squares
+    linear regression, indicating the effect that ``toxic'' definitions have on
+    failure (timeout) rate.}
     % TODO Put line slopes and error on graph
@@
 
-\iffalse
-\begin{figure}
-  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
-  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
-  \caption{Proportion of samples which timed out per size, with least-squares
-    linear regression. First plot is for all TIP definitions, second removes
-    runs given ``toxic'' definitions.}
-  \label{fig:tailsize}
-\end{figure}
-\fi
-
 The height of these ``tails'' is linearly correlated with the number of
@@
 \begin{figure}
-  \scalebox{0.45}{\input{images/proportionsall.pgf}}
-  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
-  \caption{Definitions, ordered by the ratio of successes to failures of the
-    runs they appeared in. The first graph contains all TIP definitions, showing
+  \scalebox{0.69}{\input{images/proportionsall.pgf}}
+  \scalebox{0.69}{\input{images/proportionsnontoxic.pgf}}
+  \caption[]{Each bar is a definition from our Theory Exploration Benchmark,
+    showing the proportion of signatures they appeared in which were
+    successfully explored (green) compared to those which timed out (red).
+    Definitions are sorted from most successful to least. The left graph shows
+    all \input{text/proportions_graph_count_all.tex} TEB definitions explored, with five ``toxic'' definitions which always
+    failed. with s ratio of successes to failures of the runs they appeared in. The first graph contains all TIP definitions, showing
     ``toxic'' definitions which always failed. The second graph only contains
-    runs without any toxic definitions.
-    \iffalse
-    TODO: Alison: ? (ME: The green/red bar graphs appear as empty white boxes in
-    the PDF Alison sent over. Check how it's rendered!)
-    TODO: The x-axis ticks are crap (all overlapping); turn them off, put a
-    single label like ``Definitions (ordered by failure rate)''
-    \fi}
+    runs without any toxic definitions.}
   \label{fig:proportions}
@@
 \begin{itemize}
-\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
-  and/or taking many arguments (leading to a large number of combinations)
-  \iffalse TODO: Alison: Is it always obvious which are toxic? \fi. In
-  particular, tail-recursive functions which expose their accumulator argument
-  are harder to explore. \iffalse TODO: Alison: These should be in the
+\item Avoid, or de-prioritise, exploring definitions which appear ``toxic''.
+  We have not yet identified an \emph{a priori} toxicity test, but contributing
+  factors include the use of large/infinite types (such as numbers and lists),
+  recursion across multiple arguments and matching input/output types (i.e.
+  \emph{endomorphisms}, which can be combined in exponentially-many ways).
+  In particular, exposing the accumulator argument of a tail-recursive function
+  makes it harder to explore. \iffalse TODO: Alison: These should be in the
   contributions and further recommendations based on your other findings in this
@@
 
-It also seems prudent for theory exploration tools to limit time and memory
-usage, both at the global level (to abort if it becomes clear we are in the
-``long tail'') and at the level of individual evaluations. In particular such
-fine-grained limits can abort evaluation of toxic definitions, whilst still
-being lenient enough to allow the majority of non-toxic definitions to be
-explored unhindered.
+We also recommend that theory exploration systems allow limits to be placed on
+their time and memory usage: both at the global level, to abort if it becomes
+clear we are in the ``long tail'' (e.g. if the discovery rate approaches zero);
+and at the level of individual evaluations. The latter is more invasive, but
+provides robustness against the presence of toxic definitions: their evaluation
+will still drain resources for no benefit, but other (non-toxic) definitions
+will continue to be explored, which would not be the case with a global abort.
+
+More elaborate alternatives could also be pursued, like dynamic scheduling to
+prioritise exploring definitions with the highest discovery rate.
 
@@
 
-Our final criterion \iffalse TODO: Alison: This looks odd since you haven't
-mentioned criteria before \fi concerns the ``quality'' of the selected
-subsets. This depends crucially on how we define whether a property is
-``interesting''. We can use the same ground truth definition as our Theory
-Exploration Benchmark: a property is ``interesting'' if it appears in a
-particular corpus (in our case TIP~\cite{claessen2015tip}), and
-``uninteresting'' if not. Whilst na\"ive, this approach allows the criterion to
-be quantified: the quality of the selected subsets is the proportion of the
+Regardless of how fast a signature is to explore, we must also be concerned with
+the ``quality'' of the output when selecting its contents. This depends
+crucially on how we define whether a property is ``interesting''. We can use the
+same ground truth definition as our Theory Exploration Benchmark: a property is
+``interesting'' if it appears in a particular corpus (in our case
+TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
+approach is quantified and measurable: the quality of the subset of definitions
+chosen to explore is the proportion of the
 whole signature's interesting properties that also apply to the subsets taken
@@
 contrast, splitting addition and multiplication into \emph{separate} subsets is
-lower quality, since this stops \emph{any} tool from finding the distributivity
+lower quality, as this prevents \emph{any} tool finding the distributivity
 property (since it involves both definitions).
@@
 To understand the quality impact of limiting ourselves to such subsets, we
-applied four signature selection methods to signatures sampled from the Theory
-Exploration Benchmark. Signatures varied in size from a single definition up to
-100; for each size, we sampled 100 signatures.
+devised four signature selection methods and applied them to signatures sampled
+from TEB. We varied the size of signatures from a single definition up to 100;
+for each size, we sampled 100 signatures.
 
@@
 focus on the effect of splitting apart definitions rather than e.g. duplicating
-definitions across several subsets or dropping some entirely. The tested methods
-are:
+definitions across several subsets or dropping some entirely (practical
+implementations will not have these constraints, but it is important here to
+avoid gaming our evaluation). The tested methods are:
 
@@
 exploration tools to discover. The shaded region shows the difference between
-(average) quality, with optimal and pessimal selections starting out equal (when
-dividing signatures of $N$ definitions into $N$ subsets, where there is no
-choice other than putting each definition on its own) but quickly diverging as
-choices become available. Note that the optimal and pessimal methods are
-constrained to avoid empty or unequally-distributed sets, so they cannot divide
-up a signature of size $N$ into more than $N$ subsets. Pseudorandom selection is
-shown as a grey line, and although the cardinality of its subsets is not
-constrained, it nevertheless appears consistently between the bounds.
+the average quality of optimal and pessimal methods. These always start equal
+(when dividing signatures of $N$ definitions into $N$ subsets, which can only be
+done in one way) but quickly diverging as choices become available. The average
+quality of pseudorandom selection is shown as a grey line, and although the
+cardinality of its subsets is not constrained, it nevertheless appears
+consistently between the bounds.
 

### 2019-11-19 PhDThesis/related_formalstats.tex
Edits

@@
 An important property of a logical formula (in general, and in particular with
-regards to interestingness) is its truth value. Although we may be able to
-determine some truth values exactly, e.g. via deduction, it may only be possible
-(or more efficient) to \emph{approximate} truth values. One straightforward
-approximation is to adopt \emph{probabilities}, where we assign probability $1$
-to formulae which are known to be true (e.g. axioms), $0$ to formulae known to
-be false, and intermediate values to those which are not known.
+regards to interestingness) is its truth value. Although we may sometimes be
+able to determine truth values exactly, e.g. via deduction, for many formulae it
+may only be possible (or practical) to \emph{approximate} their truth value. One
+straightforward approximation is to adopt \emph{probabilities}, where we assign
+probability $1$ to formulae which are known to be true (e.g. axioms), $0$ to
+formulae known to be false, and intermediate values to those which are not
+known~\cite{Hutter.Lloyd.Ng.ea:2013}. For example, the truth value of universal
+statements (e.g. ``all ravens are black'') cannot be deduced from non-exhaustive
+examples (e.g. real-world observations of raven colours), but various induction
+methods may infer an (approximate) truth value~\cite{10.1093/mind/LIV.214.97}.
 
-This can be useful for ``knowledge bases'' of non-logical axioms (e.g. the
-observed colours of Hempel's ravens~\cite{10.1093/mind/LIV.214.97} \iffalse
-TODO: Alison: Maybe a bit of explanation here \fi), and allows the probability of
-universal statements to be inferred from non-exhaustive
-cases~\cite{Hutter.Lloyd.Ng.ea:2013}. However, inferring consistent
-probabilities requires \emph{logical omniscience}: knowledge of all the logical
-consequences of the chosen axioms. This is unfortunate, since it is precisely
-what we are trying to avoid by approximating with probabilities.
-
-Avoiding this requirement for logical omniscience has given rise to the field of
-\emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
+One difficulty with approximate truth values is maintaining consistency without
+requiring \emph{logical omniscience} (knowledge of all the logical consequences
+of the chosen axioms). Avoiding logical omniscience has given rise to the field
+of \emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
 subject to some computational constraints. For example, the ``logical inductor''
@@
 formal statements, and instead produces a \emph{sequence} of successive
-approximate probabilities, each computable with finite (though
-doubly-exponential!) resources. These approximations are only guaranteed to be
-consistent in the limit.
+approximate probabilities, each computable with finite resources (albeit very
+inefficiently), and guaranteed to be \emph{eventually consistent}.
 
@@
 that the $n$th bit of $\pi$ is a zero, for some large $n$.
+
+To avoid such trivially-easy and infeasibly-difficult statements (which we may
+consider uninteresting) we can instead consider the \emph{variance}
+(disagreement) between \emph{multiple} approximate truth values (similar to the
+problem solvers of~\cite{fernando2013design1, fernando2013design2}). Those
+statements which are assigned similar truth values by a variety of methods may
+be considered less interesting, as they are either easy enough to infer by all
+of our methods, or too hard for any of them (the stability of a single method's
+prediction over time would serve a similar role). One benefit of this
+application is that the approximation methods do not have to be particularly
+capable; in fact choosing ``superhuman'' algorithms would be \emph{less} useful
+in this context, since they would agree on the triviality of many statements we
+might consider interesting.
 

### 2019-11-24 PhDThesis/clustering.tex
More citations

@@
   \end{haskell}
-  \caption{Haskell equivalents of the Common Lisp functions in Figure \ref{fig:lisp}.}
-  \label{fig:haskelllisp}
+  \caption{Haskell equivalents of the Common Lisp functions in
+    Figure~\ref{fig:clusterlisp}.}
+  \label{fig:clusterhaskelllisp}
 \end{figure}
@@
   \end{tikzpicture}
-  \caption{Typical clusters for the functions in Figure \ref{fig:haskelllisp}.}
+  \caption{Typical clusters for the functions in
+    Figure~\ref{fig:clusterhaskelllisp}.}
   \label{fig:haskellcluster}

### 2019-11-24 PhDThesis/futurework.tex
Remove dupe label
More citations

@@
 \subsubsection{Interestingness}
-\label{sec:interestingness}
 
@@
 Since tools like \hipspec{} already call out to third-party automated theorem
-provers, and indeed our own \mlforhs{} system uses the external Weka program, there
-may also be merit in using external concept formation or conjecture generation
-tools (or reimplementations of their ideas), in order to build up more structure
-on top of that provided by the code we are exploring. For example, the
-approaches taken by AM \cite{lenat1977automated, lenat1979automated}, Graffiti
-\cite{delavina2005some, delavina2005graffiti} and HR
-\cite{colton1999automatic, colton2000agent} could be used alongside those of
+provers, there may also be merit in using external concept formation or
+conjecture generation tools (or reimplementations of their ideas), in order to
+build up more structure on top of that provided by the code we are exploring.
+For example, the approaches taken by
+AM~\cite{lenat1977automated, lenat1979automated},
+Graffiti~\cite{delavina2005some, delavina2005graffiti} and
+HR~\cite{colton1999automatic, colton2000agent} could be used alongside those of
 \quickspec{}.

### 2019-11-24 PhDThesis/litreview_machinelearning.tex
More citations

@@
-\section{Machine Learning Applications in Mathematics, Logic and Programming
-  Language Theory}
+\section{Machine Learning Applications to Formal Systems}
 
@@
 ML algorithms are numerical in nature, and often rely on assumptions such as
-continuity and differentiability. In contrast, the worlds of mathematical
-formulae, formal logic and computer programs are usually discrete and symbolic.
-This makes application of ML to these domains more difficult than, say, image
-processing. This difficulty is clear from attempts to train Artificial Neural
-Networks to evaluate snippets of program code~\cite{zaremba2014learning}: even
-with heavily restricted inputs (e.g. short integer calculations, with little
-nesting, requiring only linear time and memory use, and evaluating with
+continuity and differentiability. In contrast, formal systems such as
+mathematical formulae, formal logic and computer programs are usually discrete
+and symbolic. This makes application of ML to these domains more difficult than,
+say, image processing, which has a natural representation as numeric pixel
+values. This difficulty is clear from attempts to train Artificial Neural
+Networks (ANNs) to evaluate snippets of program code~\cite{zaremba2014learning}:
+even with heavily restricted inputs (e.g. short integer calculations, with
+little nesting, requiring only linear time and memory use, and evaluating with
 ``teacher forcing'' to prevent errors accumulating) the accuracy quickly drops

### 2019-11-24 PhDThesis/background_machinelearning.tex
Some rearranging and expansion

@@
 (learned compressor/decompressor pairs, rewarded by how well they preserve data
-through a round-trip) and adversarial or co-evolution tasks (coupled systems
-where each is rewarded by exploiting mis-prediction in the others).
+through a ``round trip'' compression-then-decompression) and adversarial or
+co-evolution tasks (coupled systems where each is rewarded by exploiting
+mis-prediction in the others).
+
+We focus on the application of statistical machine learning algorithms to
+\emph{languages}, such as logical statements or programming languages, This is
+made particularly difficult due to the ubiquity of recursive structures, from
+inductive datatypes, to recursive function definitions; from theorem statements,
+to proof objects. Such nested structures may extend to arbitrary depth, which
+makes them difficult to represent in the fixed numerical terms expected by most
+machine learning algorithms.
 
@@
 
-One major difficulty when applying statistical machine learning algorithms to
-\emph{languages}, such as Haskell, is the appearance of recursive
-structures. This can lead to nested expressions of arbitrary depth, which are
-difficult to compare in numerical ways. One common approach to this problem is
+\emph{Feature extraction} is a common pre-processing step for machine learning
+(ML). Rather than feeding ``raw'' data straight into our ML algorithm, we only
+learn a sample of \emph{relevant} details, known as \emph{features}. This has
+two benefits:
+
+\begin{itemize}
+\item \emph{Feature vectors} (ordered sets of features) are chosen to be more
+  compact than the data they're extracted from: feature extraction is
+  \emph{lossy compression}. This reduces the size of the ML problem, improving
+  efficiency (e.g. running time).
+\item We avoid learning irrelevant details such as the encoding system used,
+  improving \emph{data} efficiency (the number of samples required to spot a
+  pattern).
+\item Feature vectors are designed to have a fixed size, i.e. they will all have
+  length (or \emph{dimension}) $d$. Many machine learning algorithms only work
+  with inputs of a uniform size; feature extraction allows us to use these
+  algorithms in domains where the size of each input is not known, may vary or
+  may even be unbounded. For example, element-wise comparison of feature vectors
+  is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions
+  this is not so straightforward, as their nesting may give rise to very
+  different shapes.
+\item Unlike our expressions, which are discrete, we can continuously transform
+  one feature vector into another. This allows operations like averaging,
+  interpolation and gradient \emph{gradient descent} to be used, which are not
+  suitable for discrete terms.
+\end{itemize}
+
+As an example, say we want to learn relationships between the following program
+fragments:
+
+\begin{haskell}
+data Maybe a = Nothing | Just a
+
+data Either a b = Left a | Right b
+\end{haskell}
+
+We might hope our algorithm discovers relationships like:
+
+\begin{itemize}
+  \item Both are valid Haskell code.
+  \item Both describe datatypes.
+  \item Both datatypes have two constructors.
+  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a
+      = Either () a} and \hs{Nothing = Left ()}).
+  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to
+    \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
+  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
+  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a
+    \hs{b}.
+\end{itemize}
+
+However, this is too optimistic. Without our domain-knowledge of Haskell, an ML
+algorithm cannot impose any structure on these fragments, and will treat them as
+strings of bits. Our high-level hopes are obscured by low-level details: the
+desirable patterns of Haskell types are mixed with undesirable patterns of ASCII
+bytes, of letter frequency in English words, and so on.
+
+In theory we could throw more computing resources and data at a problem (this is
+precisely what \emph{deep learning} techniques do), but available hardware and
+corpora are always limited. Instead, feature extraction lets us narrow the ML
+problem to what we, with our domain knowlege, consider important.
+
+There is no \emph{fundamental} difference between raw representations and
+features: the identity function is a valid feature extractor. Likewise, there is
+no crisp distinction between feature extraction and machine learning: a
+sufficiently-powerful learner doesn't require feature extraction, and a
+sufficiently-powerful feature extractor doesn't require any learning!
+\footnote{Consider a classification problem, to assign a label $l \in L$ to each
+  input. If we only extract a single feature $f \in L$, we have solved the
+  classification problem without using a separate learning step.}
+
+Rather, the terms are distinguished for purely \emph{practical} reasons: by
+separating feature extraction from learning, we can distinguish straightforward,
+fast data transformation (feature extraction) from complex, slow statistical
+analysis (learning). This allows for modularity, separation of concerns, and in
+particular allows ``off-the-shelf'' ML to be re-used across a variety of
+different domains. Even with no domain knowledge, feature extraction can still
+be used to improve efficiency by compressing the input (for example using
+\emph{autoencoding}).
+
+\subsubsection{Truncation and Padding}
+
+The simplest way to limit the size of recursive structures is to truncate
+anything larger than a particular size (and pad anything smaller). This is the
+approach taken by ML4PG~\cite{journals/corr/abs-1302-6421}, which limits itself
+to trees with at most 10 levels and 10 elements per level; each tree is
+converted to a $30 \times 10$ matrix (3 values per tree node) and learning
+takes place on these normalised representations.
+
+Truncation is unsatisfactory in the way it balances \emph{data} efficiency with
+\emph{time} efficiency. Specifically, truncation works best when the input data
+contains no redundancy and is arranged with the most significant data first (in
+a sense, it is ``big-endian''). The less these assumptions hold, the less we can
+truncate. Since many ML algorithms scale poorly with input size, it is important
+to truncate aggressively, but there may be useful information deep in the trees
+which would be discarded.
+
+\subsubsection{Dimension Reduction}
+
+A more sophisticated approach to the problem of reducing input size is to view
+it as a \emph{dimension reduction} technique: our inputs can be modelled as
+points in a high-dimensional space, which we want to project into a
+lower-dimensional space.
+
+Truncation is a trivial dimension reduction technique: take the first $N$
+coordinates. More sophisticated projection functions, such as Principle
+Component Analysis (PCA) consider the \emph{distribution} of the points, and
+project with the hyperplane which preserves as much of the variance as possible
+(or, equivalently, reduces the \emph{mutual information} between the points).
+
+Dimension reduction can also be approached as an unsupervised ML task (even if
+the data are labelled for a supervised
+task~\cite{Oveisi.Oveisi.Erfanian.ea:2012}). One popular approach is the
+\emph{autoencoder}, which transforms training data into smaller feature vectors
+and back again, whilst trying to minimise reconstruction error. Such compressed
+representations can be used as feature vectors for subsequent learning
+tasks. More semantically-meaningful features can be found using
+\emph{variational autoencoding}~\cite{kingma2013auto}, which add a stochastic
+sampling step before decoding, forcing similar data to produce similar vectors.
+
+Since these techniques are effectively ML algorithms in their own right, they
+suffer some of the same difficulties we encounter with recursive expressions:
+
+\begin{itemize}
+  \item They operate \emph{offline}, requiring all input points up-front
+  \item All input points must have the same dimensionality
+\end{itemize}
+
+Sophisticated dimension reduction is still useful for \emph{compressing} large,
+redundant features into smaller, information-dense representations, and as such
+provides a good complement to more flexible but less informative techniques. For
+example, we might truncate (and pad) tree structures into large, sparse vectors
+of a fixed size; then use PCA to pick only a few of the most informative
+dimensions to constitute our resulting feature vectors.
+
+The requirement for offline ``batch'' processing is more difficult to overcome,
+since any learning performed for feature extraction will interfere with the
+core learning algorithm that's consuming these features (this is why deep
+learning is often done greedily).
+
+\subsubsection{Distributed Representations}
+
+A \emph{distributed representation} does not isolate each semantic feature (for
+example, presence of absence of some lexeme) to separate, independent elements
+of a feature vector. Instead, features are represented as \emph{patterns} across
+multiple elements: each feature uses many elements and each element contributes
+to many features~\cite{hinton1984distributed}. Such encodings provide simple
+mechanisms to represent \emph{combinations} of features (combining their
+patterns element-wise), \emph{partial} features (presence of the pattern across
+a subset of its elements) and \emph{generalisation} (patterns which are present
+as sub-patterns of multiple features).
+
+The compressed representations learned by autoencoders are distributed
+representations, and are usually continuous. A discrete alternative is to use
+\emph{hashing}, which can also be learned (e.g. using \emph{random
+forests}~\cite{vens2011random}).
+
+Particularly relevant to our work is the ability for distributed representations
+to approximate recursive data such as syntax trees. This is achieved by defining
+(or learning) functions to \emph{combine} the representations of parts into
+those of composite objects, to \emph{test} whether or not a given feature vector
+represents a composite object, and to \emph{project} composite representations
+back into those of their constituents. For example, if $c$ and $d$ are the
+compression and decompression routines of an auto-encoder for non-recursive
+data, we can define an auto-encoder ($c'$ and $d'$) for binary trees in terms
+of a combining function $\bigoplus$, a predicate $\texttt{node?}$ and projection
+functions $\pi_{left}}$ and $\pi_{right}}$ as follows:
+
+\begin{equation*}
+  c'(x) = \begin{cases}
+            c'(a) \bigoplus c'(b) & \text{if $x = \texttt{node}(a, b)$}
+            c(x)                  & \text{otherwise}
+          \end{cases}
+
+  d'(x) = \begin{cases}
+            \texttt{node}(d'(\pi_{left }(x)),
+                          d'(\pi_{right}(x))) & \text{if \texttt(node?)(x)}
+            d(x)                            & \text{otherwise}
+          \end{cases}
+\end{equation*}
+
+With a suitable distance function between trees (e.g. edit distance), we can
+learn $c$, $d$, \texttt{node?} and $\bigoplus$ together in an unsupervised
+manner, by minimising the round trip error on a training corpus. The resulting
+$c'$ function can then directly encode recursive data to fixed-size feature
+vector. Analogous constructions can be applied to different recursive
+structures.
+
+\subsection{Sequences}
+
+To handle input of variable size, such as natural or formal language, research
+attention has been given to the handling of \emph{sequences}. This is a lossless
+approach, which splits the input into fixed-size \emph{chunks} (e.g. splitting
+text into a sequence of characters), which are fed into an appropriate ML
+algorithm one at a time. The sequence is terminated by a sentinel; an
+``end-of-sequence'' marker which, by construction, is distinguishable from the
+data chunks. Tradeoffs can be made between chunk size and sequence length; for
+example, treating code as a sequence of
+characters~\cite{cummins2017synthesizing} or lexer tokens~\cite{cummins2017end}.
+
+Not all ML algorithms can be adapted to accept sequences. One notable approach
+is to use \emph{recurrent (artificial) neural networks} (RNNs), which allow
+arbitrary connections between nodes, including cycles. Compared to
+\emph{feed-forward} (FF) neural networks, which are acyclic, the \emph{future
+  output} of a RNN may depend arbitrarily on its \emph{past inputs} (in fact,
+RNNs are universal computers).
+
+The main problem with RNNs, compared to the more widely-used FF approach, is the
+difficulty of training them. If we extend the standard backpropagation algorithm
+to handle cycles, we get the \emph{backpropagation through time}
+algorithm~\cite{werbos1990backpropagation}. However, this suffers a problem
+known as the \emph{vanishing gradient}: error values decay exponentially as they
+propagate back through the cycles, which prevents effective learning of delayed
+dependencies, undermining the main advantage of RNNs. The vanishing gradient
+problem is the subject of current research, with countermeasures including
+\emph{neuroevolution} (using evolutionary computation techniques to complement
+or replace backpropagation) and \emph{gated neurons} (which are selectively
+updated, and hence resilient to spurious overwriting; the most common example
+being \emph{long short-term memory} (LSTM)~\cite{hochreiter1997long}).
+
+An alternative to introducing recurrent connections is the use of
+\emph{attention}~\cite{vaswani2017attention}. In a (typically FF) neural network
+with attention, the network is given a chunk of the input and, in addition to
+its usual output, also produces some ``attention'' information specifying which
+chunks of the input should be provided next. The network is run again, but with
+the input adjusted using this attention information; more output is produced,
+along with new attention information, and so on. Such networks do not need to
+``remember'' long-distance relationships, whether by recurrent connections or
+gated nodes; they instead use the same input over and over again, but use their
+``attention'' to emphasise those parts which are relevant to upcoming
+output and hence should be taken into account when computing it.
+
+\subsubsection{Recursive Structure}
+
+Using sequences to represent recursive data is problematic: if we want our
+learning algorithm to exploit structure (such as the depth of a token), it will
+need to discover how to parse the sequences for itself, which is wasteful since
+we already have a parsed representation to begin with. The
+\emph{back-propagation through structure} approach~\cite{goller1996learning} is
+a more direct solution to this problem, using FF neural networks to learn
+recursive distributed representations~\cite{pollack1990recursive} which
+correspond to the recursive structure of the inputs. Such distributed
+representations can also be used for sequences, which we can use to encode
+sub-trees when the branching factor of nodes is not
+uniform~\cite{kwasny1995tail}. Recursive structure can also be gated inside
+LSTM cells~\cite{zhu2015long}.
+
+A simpler alternative for generating recursive distributed representations is to
+use circular convolution \cite{conf/ijcai/Plate91}. Although promising results
+are shown for its use in \emph{distributed tree kernels}
+\cite{zanzotto2012distributed}, our preliminary experiments in applying
+circular convolution to functional programming expressions found most of the
+information to be lost in the process; presumably as the expressions are too
+small.
+
+\emph{Kernel methods} have also been applied to structured information, for
+example in \cite{Gartner2003} the input data (including sequences, trees and
+graphs) are represented using \emph{generative models}, such as hidden Markov
+models, of a fixed size suitable for learning. Many more applications of kernel
+methods to structured domains are given in \cite{bakir2007predicting}, which
+could be used to learn more subtle relations between expressions than recurrent
+clustering alone.
+
+ One common approach to this problem is
 to represent such structures as \emph{sequences}. \emph{Recurrent neural
@@
 
-\subsubsection{Distributed Representations}
-
-A \emph{distributed representation} does not isolate each semantic feature (for
-example, presence of absence of some lexeme) to separate, independent
-``computing elements'' (e.g. memory locations, such as the elements of a feature
-vector). Instead, features are represented as \emph{patterns} across multiple
-elements: each feature uses many elements and each element contributes to many
-features~\cite{hinton1984distributed}. Such encodings provide simple mechanisms
-to represent \emph{combinations} of features (combining their patterns
-element-wise), \emph{partial} features (presence of the pattern across a subset
-of its elements) and \emph{generalisation} (patterns which are present as
-sub-patterns of multiple features).
-
-% Distributed representations may be learned from data, for example by
-% \emph{auto-encoding}; or they may be calculated ``from whole cloth'' using
-% techniques like hashing. We are particularly interested in distributed
-% representations for storing structured data (like syntax trees) in fixed-length
-% feature vectors.
-
 \input{background_clustering.tex}

### 2019-11-24 PhDThesis/benchmark.tex
TODOs
Some rearranging and expansion

@@
 \label{sec:benchmark}
+
+\iffalse TODO: Alison: Each chapter should have a description at the start (see
+bucketing ``in this chapter'') \fi
 
@@
 conjecturing equations involving Haskell functions, described in more detail in
-$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark \quickspec{}, we
-need to automate some of the decisions which are normally left up to the user:
+$\S$\ref{sec:theoryexploration}. In order to thoroughly benchmark \quickspec{},
+we need to automate some of the decisions which are normally left up to the
+user:
 
@@
     is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses show runs which timed out, which occurred for every size. Each size
+    circles show runs which timed out, which occurred for every size. Each size
     has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
@@
     around $0.1$ for size 20, whilst mean recall remains roughly flat between
-    $0.2$ and $0.5$"}
+    $0.2$ and $0.5$}
   \label{fig:quickspec_precRec}
@@
     is a successful run (spread out horizontally to reduce overlaps). Red
-    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
+    circles are failed runs, caused by timeouts or out-of-memory. All runs of
     size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31

### 2019-11-24 PhDThesis/bucketing.tex
More citations
Some rearranging and expansion

@@
   Haskell packages\footnote{\hPackage{nix-eval}}.
-\item Translation of the signature selection problem to the constrain-solving domain, and
-  executable oracles for optimal (and pessimal) signature selection, when the desired
-  properties are already known\footnote{Part of
+\item Translation of the signature selection problem to the constrain-solving
+  domain, and executable oracles for optimal (and pessimal) signature selection,
+  when the desired properties are already known\footnote{Part of
     https://github.com/warbo/bucketing-algorithms}.
@@
   In particular, exposing the accumulator argument of a tail-recursive function
-  makes it harder to explore. \iffalse TODO: Alison: These should be in the
-  contributions and further recommendations based on your other findings in this
-  chapter \fi
+  makes it harder to explore.
 \item Smaller signatures are faster to explore, so should be preferred if
@@
 their similarity across multiple dimensions. To investigate this potential
-relationship we created a signature selection method based on a \emph{recurrent
-  clustering} algorithm for Haskell code: this flattens syntax trees into
-vectors, substitutes fixed numbers for keywords and recurses to determine
-suitable numbers for referenced expressions.
+relationship we created a signature selection method using the \emph{recurrent
+  clustering} algorithm for Haskell code described in
+\S~\ref{sec:recurrentclustering}: this flattens syntax trees into vectors,
+substitutes fixed numbers for keywords and recurses to determine suitable
+numbers for referenced expressions.
 

### 2019-11-24 PhDThesis/clustering.tex
Some rearranging and expansion

@@
 \iffalse
+Here we consider the 10 most-downloaded packages from \hackage{} (as of
+2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
+\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
+\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
+\texttt{blaze-html}.
+
+% TODO: Analysis
+\fi
+
+\iffalse
 TODO: Alison: Next chapter? It's good to lead into the next chapter at the end
@@
 our intuitions, in particular since all 15 of Yesod's MIME type identifiers
-appeared in the same cluster.
-
-\iffalse
-Since we are working in the domain of Haskell programs, an abundant source of
-statements are available in the form of \emph{tests}. For manually-written
-tests, the effort required to write them implies that they must be of some
-interest to their author. Many forms of test are \emph{not} suitable for our
-purposes, such as \emph{unit tests} which are always trivially provable by
-$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
-the external environment. One reason to study Haskell is its widespread use of
-\emph{property checking}, which does give us useful data. Many Haskell property
-checkers exist, based on random testing
-(\quickcheck{}~\cite{claessen2011quickcheck} and
-\smartcheck{}~\cite{pike2014smartcheck}), enumeration
-(\smallcheck{}~\cite{runciman2008smallcheck}), observation
-(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
-(\sparsecheck{}~\cite{sparsecheck}).
-
-Thankfully the major differences between these systems are in the way they
-instantiate test arguments; their representations of properties are largely the
-same (modulo renaming of functions and types).
-
-Here we consider the 10 most-downloaded packages from \hackage{} (as of
-2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
-\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
-\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
-\texttt{blaze-html}.
-
-% TODO: Analysis
-\fi
+appeared in the same cluster. Although promising, a more thorough exploration of
+the parameters would be needed to obtain more intuitively reliable results (at
+the risk of over-fitting).
 
@@
 
-It seems like this recurrent clustering method has promise\iffalse TODO: Alison: Informal, vague \fi, although it will
-require a more thorough exploration of the parameters to obtain more intuitively
-reliable results. These clusters can then be used in several ways to perform
-theory exploration; the most na\"{\i}ve way being to explore each cluster as
-\quickspec{} signature in its own right, which we investigate as a potential
-solution to the Signature Selection problem.
+These clusters can then be used in several ways to perform theory exploration;
+the most na\"{\i}ve way being to explore each cluster as \quickspec{} signature
+in its own right, which we investigate as a potential solution to the Signature
+Selection problem in \S~\ref{sec:signatureselection}.
 

### 2019-11-24 PhDThesis/futurework.tex
Some rearranging and expansion

@@
 
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.75]
-      % The image, for reference
-      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
-
-      % Axes
-      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
-      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
-      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
-
-      % Curve. The numbers come from tracing over wundt.png
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=180] (6.5,  8.2)
-           to[out=0,   in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-
-      % This version is closer, but a little jagged
-      \iffalse
-      \draw[black,ultra thick] (3, 4)
-           to[out=0,   in=240] (5.05, 6)
-           to[out=60,  in=225] (6,    8)
-           to[out=45,  in=180] (6.5,  8.2)
-           to[out=0,   in=135] (7.2,  8)
-           to[out=-45, in=105] (8.1,  6)
-           to[out=-75, in=120] (8.8,  4)
-           to[out=-60, in=150] (11.4, 0.8);
-      \fi
-  \end{tikzpicture}
-
-  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}.
-The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}
-
-  \label{fig:wundt}
-\end{figure}
-
-\emph{Artificial curiosity} (AC) describes active learning systems which are
-rewarded based on how interesting the input or data they discover is
-\cite{schmidhuber2006developmental}. Although framed in the context of
-\emph{reinforcement learning}, this reliance on interest is clearly relevant to
-our theory exploration setting.
-
-As an unsupervised learning task, AC has no access to labels or meanings
-associated with its input; the only features it can learn are the structure and
-relationships inherent in the data, in a similar way to our recurrent clustering
-algorithm. The unifying principle of AC methods is to force systems away from
-inputs which are not amenable to learning; either because they are so familiar
-that there is nothing left to learn, or so unfamiliar that they are
-unintelligible. The resulting behaviour is characterised by the \emph{Wundt
-  curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures
-  avoid negative values for simplicity, in which cases we replace all negative
-  points on the curve with zero.}, which has been used in psychology to explain
-human aesthetics and preferences \cite{berlyne1970novelty}. This same behaviour
-may be applicable to the theorems produced by a theory exploration system.
-
-We can divide AC approaches into two groups: those which make \emph{explicit}
-use of interestingness, learning from signals which follow a Wundt curve; whilst
-\emph{implicit} approaches modify the \emph{output} of their learning
-algorithm(s), to engineer the overall system behaviour to follow a Wundt curve
-as an emergent property.
-
-A framework encompassing many examples of the explicit approach is given in
-\cite{oudeyer2007intrinsic} in the context of reinforcement learning; for
-comparison, many similar measures are surveyed in a data mining context in
-\cite{geng2006interestingness}. Many more reinforcement learning examples can
-be found in \cite{Kaplan2006, Lipson2007, Luciw2011, Macedo2000,
-  Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991,
-  oudeyer2004intelligent}; whilst more general descriptions are given in
-\cite{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be
-more amenable to our theory exploration setting.
-
-Many of these reward signals are based on information theory, with a prominent
-example being \emph{compression progress}: given a compressed representation of
-our previous observations, the ``progress'' is the space saved if we include the
-current observation. Observations which are incompressible or trivially
-compressible don't save any space, whilst observations which provide new
-insights into the structure of past experience can provide a space saving when
-compressed together. This seems particularly relevant for our problem of
-identifying interesting theorems: those new theorems (``observations'') which
-shorten the proofs of previously discovered theorems may be more general, more
-powerful and therefore more \emph{interesting} and hence worth keeping. In fact
-this is very similar to \quickspec{}'s interestingness criterion.
-
-Another example of explicit artificial curiosity is given in
-\cite{Hester.Stone:2012}, where world states which cause \emph{disagreement}
-among a population of decision trees (a \emph{random forest}
-\cite{randomforests}) are considered interesting. Since the models make
-stochastic predictions, the disagreement follows a Wundt curve as the complexity
-of state transitions increases: for parts of the state space which have been
-fully learned, the models will agree on accurate predictions; for parts which
-are unlearnable, the models cannot infer any structure, and will converge to
-reporting the same average value. Whilst the latter predictions may not be
-\emph{accurate}, they will be \emph{in agreement}, hence pushing down the
-interestingness of states which are too complex.
-
-Many examples of the implicit case are based on \emph{coevolution}: rewarding
-one part of the system for exploiting another part, and vice versa. In
-\cite{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on
-the outcome of actions, and the winner is rewarded at the expense of the
-loser. Due to the risk involved, each algorithm will only bet when it is
-confident in its prediction, and bets will only be actioned when each algorithm
-is confident in a \emph{different} outcome. The overall behaviour of this system
-is therefore similar to the explicit measure of disagreement used in the random
-forest example. In terms of theory exploration, such a scheme could be used to
-find theorems which are \emph{non-obvious}, and hence informative in some way to
-the user.
-
-Another case is the ``darwinian brain'' of Fernando et
-al. \cite{fernando2013design1, fernando2013design2}. This coevolves a
-population of problem generators and problem solvers, rewarding the solvers
-based on their speed, and rewarding the generators based on the \emph{variance}
-of the solvers' speed. This avoids trivial problems (which all solvers can
-quickly overcome) and complex problems (which no solver can manage), and focuses
-on those with the most possibility for learning. It is easy to imagine such a
-general architecture being populated by conjecture generators and theorem
-provers to form a theory exploration system.
-
 \paragraph{Evolutionary Computation} \leavevmode \newline

### 2019-11-24 PhDThesis/litreview_interestingness.tex
Some rearranging and expansion

@@
 often in inductive proofs (e.g. when reasoning about tail-recursive
-definitions~\cite{kapur2003automatic}), since the specific obligations which
+definitions~\cite{kapur2003automatic}\footnote{Tail-recursive functions are
+  desirable due to their constant space usage, unlike recursion in non-tail
+  positions which requires a growing number of stack frames or nested closures.
+  Reasoning about non-tail recursive definitions is often simpler, since they
+  don't expose accumulator arguments, so equivalence ), since the specific obligations which
 arise in the proof may be incompatible with the available inductive hypotheses.
@@
 definition) to its \emph{applicability}. It is still important to balance
-against other concerns, since blindly generalising \emph{all} statements of
+against other concerns, since blindly generalising \emph{all} statements or
 proof obligations we come across would result in conjectures so strong that they
@@
 different theory, we may be able to re-use proof tactics and other forms of
-meta-programming across both.
-
-Existing theory exploration systems have been successfully applied to this
-problem, however the use of pure exploration misses opportunities to
-\emph{focus} the search, since we know which lemmas are used in those theories
-where a technique succeeded. If we can find an analogy to map from such solved
-problems to a currently-unproven goal, we can infer the approximate form of the
-lemmas we require, and target these specifically.
+meta-programming across both. In this case an analogy would map from a solved
+problem to a currently-unproven goal, and applying this map to the lemmas
+appearing in the existing proof allows us to infer the approximate form of the
+lemmas required for the new theorem.
 
@@
 examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
-tail-recursive and non-tail-recursive calculations\footnote{A tail-recursive
-  function can be executed in constant space using a loop, whereas recursion in
-  non-tail positions may require a growing number of stack frames or nested
-  closures.}), as well as the effect of repeating certain list operations:
+tail-recursive and their non-tail-recursive counterparts, as well as the effect
+of repeating certain list operations:
 
@@
 
-% TODO Include something about intrinsic motivation, compression progress,
+\subsection{Tests}
+
+Since we are working in the domain of Haskell programs, an abundant source of
+statements are available in the form of \emph{tests}. For manually-written
+tests, the effort required to write them implies that they must be of some
+interest to their author. Many forms of test are \emph{not} suitable for our
+purposes, such as \emph{unit tests} which are always trivially provable by
+$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
+the external environment. One reason to study Haskell is its widespread use of
+\emph{property checking}, which does give us useful data in the form of
+universally quantified statements. Many Haskell property checkers exist, based
+on random testing (\quickcheck{}~\cite{claessen2011quickcheck} and
+\smartcheck{}~\cite{pike2014smartcheck}), enumeration
+(\smallcheck{}~\cite{runciman2008smallcheck}), observation
+(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
+(\sparsecheck{}~\cite{sparsecheck}).
+
+Thankfully the major differences between these systems are in the way they
+instantiate test arguments; their representations of properties are largely the
+same (modulo renaming of functions and types): as Haskell functions with free
+variables encoded as function arguments.
+
+\subsection{Artificial Curiosity} \label{sec:intrinsic}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[scale=0.75]
+      % The image, for reference
+      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};
+
+      % Axes
+      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
+      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
+      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label
+
+      % Curve. The numbers come from tracing over wundt.png
+      \draw[black,ultra thick] (3, 4)
+           to[out=0,   in=240] (5.05, 6)
+           to[out=60,  in=180] (6.5,  8.2)
+           to[out=0,   in=105] (8.1,  6)
+           to[out=-75, in=120] (8.8,  4)
+           to[out=-60, in=150] (11.4, 0.8);
+
+      % This version is closer, but a little jagged
+      \iffalse
+      \draw[black,ultra thick] (3, 4)
+           to[out=0,   in=240] (5.05, 6)
+           to[out=60,  in=225] (6,    8)
+           to[out=45,  in=180] (6.5,  8.2)
+           to[out=0,   in=135] (7.2,  8)
+           to[out=-45, in=105] (8.1,  6)
+           to[out=-75, in=120] (8.8,  4)
+           to[out=-60, in=150] (11.4, 0.8);
+      \fi
+  \end{tikzpicture}
+
+  \caption{The Wundt curve, reproduced from~\cite{berlyne1970novelty}. The axes
+    ``hedonic value'' and ``arousal potential'' are described as covering
+    \textquote{reward value\dots preference or pleasure}, and \textquote{all the
+      stimulus properties that tend to raise arousal, including novelty and
+      complexity}, respectively.}
+  \label{fig:wundt}
+\end{figure}
+
+Another perspective on what is ``interesting'' comes from the artificial
+intelligence tasks of optimisation and reinforcement learning. Common pitfalls
+for such systems are highly non-convex problems (trapping optimisers in local
+minima), and sparse information (such as flat gradients; or highly delayed
+rewards, like a single win/lose reward at the end of a game). To mitigate these problems, various forms of internally-generated (or \emph{intrinsic}) rewards
+can be incorporated into a system. Such schemes provide reward for interesting
+discoveries, which encourages exploration in lieu of external direction, and are
+known as \emph{Artificial Curiosity}~\cite{schmidhuber2006developmental} (AC).
+
+The unifying principle of AC methods is to force systems away from data which
+are not amenable to learning; either because they are so familiar that there is
+nothing left to learn, or so unfamiliar that they are unintelligible. The
+resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures
+  avoid negative values for simplicity, in which cases we replace all negative
+  points on the curve with zero.}, which has been used in psychology to explain
+human aesthetics and preferences \cite{berlyne1970novelty}. This same behaviour
+may be applicable to the theorems produced by a theory exploration system.
+
+We can divide AC approaches into two groups: those which make \emph{explicit}
+use of interestingness, learning from signals which follow a Wundt curve; whilst
+\emph{implicit} approaches modify the \emph{output} of their learning
+algorithm(s), to engineer the overall system behaviour to follow a Wundt curve
+as an emergent property.
+
+A framework encompassing many examples of the explicit approach is given
+in~\cite{oudeyer2007intrinsic} in the context of reinforcement learning; for
+comparison, many similar measures are surveyed in a data mining context
+in~\cite{geng2006interestingness}. Many more reinforcement learning examples can
+be found in~\cite{Kaplan2006, Lipson2007, Luciw2011, Macedo2000,
+  Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991,
+  oudeyer2004intelligent}; whilst more general descriptions are given
+in~\cite{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may
+be more amenable to Theory Exploration setting.
+
+Many of these reward signals are based on information theory, with a prominent
+example being \emph{compression progress}: given a compressed representation of
+our previous observations, the ``progress'' is the space saved if we include the
+current observation. Observations which are incompressible or trivially
+compressible don't save any space, whilst observations which provide new
+insights into the structure of past experience can provide a space saving when
+compressed together. This seems particularly relevant for identifying
+interesting theorems: those new theorems (``observations'') which shorten the
+proofs of previously discovered theorems may be more general, more powerful and
+therefore more \emph{interesting}. In fact this is very similar to
+\quickspec{}'s interestingness criterion.
+
+Another example of explicit Artificial Curiosity is given
+in~\cite{Hester.Stone:2012}, where world states which cause \emph{disagreement}
+among a population of decision trees (a \emph{random
+  forest}~\cite{randomforests}) are considered interesting. Since the models
+make stochastic predictions, the disagreement follows a Wundt curve as the
+complexity of state transitions increases: for parts of the state space which
+have been fully learned, the models will agree on accurate predictions; for
+parts which are unlearnable, the models cannot infer any structure, and will
+converge to reporting the same average value. Whilst the latter predictions may
+not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the
+interestingness of states which are too complex.
+
+Many examples of the implicit case are based on \emph{coevolution}: rewarding
+one part of the system for exploiting another part, and vice versa. In~\cite{Schmidhuber1999} a pair of learning algorithms place virtual ``bets''
+on the outcome of actions, and the winner is rewarded at the expense of the
+loser. Due to the risk involved, each algorithm will only bet when it is
+confident in its prediction, and bets will only be actioned when each algorithm
+is confident in a \emph{different} outcome. The overall behaviour of this system
+is therefore similar to the explicit measure of disagreement used in the random
+forest example.
+
+More recent work on \emph{Generative Adversarial Networks} (GANs) trains a
+``discriminator'' program to distinguish real training examples (usually images)
+from synthetic data made by a ``generator'' program. Both programs are
+differentiable (usually some type of Artificial Neural Network), so the coupling
+between the generator's output and the discriminator's input can be used not
+only to feed synthetic data ``forwards'', but also to back-propagate the
+discriminator's errors (which it is trying to minimise) into the generator
+(which then tries to maximise, exploiting the discriminator by synthesising more
+realistic data).
+
+Another example of implicit AC is the ``darwinian brain'' of Fernando et
+al.~\cite{fernando2013design1, fernando2013design2}. This coevolves a
+population of problem generators and problem solvers, rewarding the solvers
+based on their speed, and rewarding the generators based on the \emph{variance}
+of the solvers' speed. This avoids trivial problems (which all solvers can
+quickly overcome) and complex problems (which no solver can manage), and focuses
+on those with the most possibility for learning. The role of problem generator
+is similar to that of Theory Exploration tools, with solvers being Automated
+Theorem Provers.
+
+% TODO Include something about
 % variance among experts (decision tree stuff as well as more established
 % splitting functions stuff)
+
+% TODO: Also mention the intrinsic reward idea about easily accessible regions
+% (e.g. balancing a pole because it's easier to get from there to anywhere else)
 

### 2019-11-24 PhDThesis/related_formalstats.tex
Some rearranging and expansion

@@
 \fi
-
-\subsection{Learning From Structured Data}
-
-One major difficulty with formal mathematics as a domain in which to apply
-statistical machine learning is the use of \emph{structure} to encode
-information in objects. In particular, \emph{trees} appear in many places: from
-inductive datatypes, to recursive function definitions; from theorem statements,
-to proof objects. Such nested structures may extend to arbitrary depth, which
-makes them difficult to represent with a fixed number of features, as is
-expected by most machine learning algorithms. Here we review a selection of
-solutions to this problem, and compare their distinguishing properties.
-
-\subsubsection{Feature Extraction}\label{sec:featureextraction}
-
-\emph{Feature extraction} is a common pre-processing step for machine learning
-(ML). Rather than feeding ``raw'' data straight into our ML algorithm, we only
-learn a sample of \emph{relevant} details, known as \emph{features}. This has
-two benefits:
-
-\begin{itemize}
-\item \emph{Feature vectors} (ordered sets of features) are chosen to be more
-  compact than the data they're extracted from: feature extraction is
-  \emph{lossy compression}. This reduces the size of the ML problem, improving
-  efficiency (e.g. running time).
-\item We avoid learning irrelevant details such as the encoding system used,
-  improving \emph{data} efficiency (the number of samples required to spot a
-  pattern).
-\item All of our feature vectors will be the same size, i.e. they will all have
-  length (or \emph{dimension}) $d$. Many machine learning algorithms only work
-  with inputs of a uniform size; feature extraction allows us to use these
-  algorithms in domains where the size of each input is not known, may vary or
-  may even be unbounded. For example, element-wise comparison of feature vectors
-  is trivial (compare the $i$th elements for $1 \leq i \leq d$); for expressions
-  this is not so straightforward, as their nesting may give rise to very
-  different shapes.
-\item Unlike our expressions, which are discrete, we can continuously transform
-  one feature vector into another. This enables many powerful machine learning
-  algorithms to be used, such as those based on \emph{gradient descent} or, in
-  our case, arithmetic means.
-\item Feature vectors can be chosen to represent the relevant information in a
-  more compressed form than the raw data; for example, we might replace verbose,
-  descriptive identifiers with sequential numbers. This reduces the input size
-  of the machine learning problem, improving efficiency.
-\end{itemize}
-
-Another benefit of feature extraction is to \emph{normalise} the input data to a
-fixed-size representation. Many ML algorithms only work with fixed-size inputs;
-for example, the popular \emph{backpropagation} \cite{Russell:2003:AIM:773294}
-algorithm works on models with \emph{fixed} topology (e.g. \emph{artificial
-  neural networks} with fixed connections between nodes). This requires some
-form of pre-processing in domains where the size of each input is not known, may
-vary or may even be unbounded.
-
-For example, in the case of \emph{online} learning we must make
-predictions/decisions before seeing all of the inputs. Unbounded input appears
-in domains such as programming and theorem proving, where individual term may be
-trees of unbounded depth. In these situations we need a mapping from arbitrary
-inputs to a fixed representation which is amenable to learning.
-
-As an example, say we want to learn relationships between the following program
-fragments:
-
-\begin{haskell}
-data Maybe a = Nothing | Just a
-
-data Either a b = Left a | Right b
-\end{haskell}
-
-We might hope our algorithm discovers relationships like:
-
-\begin{itemize}
-  \item Both are valid Haskell code.
-  \item Both describe datatypes.
-  \item Both datatypes have two constructors.
-  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a
-      = Either () a} and \hs{Nothing = Left ()}).
-  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to
-    \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
-  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
-  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a
-    \hs{b}.
-\end{itemize}
-
-However, this is too optimistic. Without our domain-knowledge of Haskell, an ML
-algorithm cannot impose any structure on these fragments, and will treat them as
-strings of bits. Our high-level hopes are obscured by low-level details: the
-desirable patterns of Haskell types are mixed with undesirable patterns of ASCII
-bytes, of letter frequency in English words, and so on.
-
-In theory we could throw more computing resources and data at a problem, but
-available hardware and corpora are always limited. Instead, feature extraction
-lets us narrow the ML problem to what we, with our domain knowlege, consider
-important.
-
-There is no \emph{fundamental} difference between raw representations and
-features: the identity function is a valid feature extractor. Likewise, there is
-no crisp distinction between feature extraction and machine learning: a
-sufficiently-powerful learner doesn't require feature extraction, and a
-sufficiently-powerful feature extractor doesn't require any learning!
-\footnote{Consider a classification problem, to assign a label $l \in L$ to each
-  input. If we only extract a single feature $f \in L$, we have solved the
-  classification problem without using a separate learning step.}
-
-Rather, the terms are distinguished for purely \emph{practical} reasons: by
-separating feature extraction from learning, we can distinguish straightforward,
-fast data transformation (feature extraction) from complex, slow statistical
-analysis (learning). This allows for modularity, separation of concerns, and in
-particular allows ``off-the-shelf'' ML to be re-used across a variety of
-different domains.
-
-Even if we have no domain knowledge, we can still use a feature extraction phase
-to improve efficiency: first we learn a compact representation for our data, for
-example using \emph{autoencoding}; then we use that encoder as a feature
-extractor for our main learning task. This stacking of one learning algorithm on
-top of another, especially with greedy learning of each layer, has lead to the
-recent trend of \emph{deep learning}.
-
-\subsubsection{Truncation and Padding}
-
-The simplest way to limit the size of our inputs is to truncate anything larger
-than a particular size (and pad anything smaller). This is the approach taken by
-ML4PG \cite{journals/corr/abs-1302-6421}, which limits itself to trees with at
-most 10 levels and 10 elements per level; each tree is converted to a
-$30 \times 10$ matrix (3 values per tree node) and learning takes place on these
-normalised representations.
-
-Truncation is unsatisfactory in the way it balances \emph{data} efficiency with
-\emph{time} efficiency. Specifically, truncation works best when the input data
-contains no redundancy and is arranged with the most significant data first (in
-a sense, it is ``big-endian''). The less these assumptions hold, the less we can
-truncate. Since many ML algorithms scale poorly with input size, we would prefer
-to eliminate the redundancy using a more aggressive algorithm, to keep the
-resulting feature size as low as possible.
-
-\subsubsection{Dimension Reduction}
-
-A more sophisticated approach to the problem of reducing input size is to view
-it as a \emph{dimension reduction} technique: our inputs can be modelled as
-points in high-dimensional spaces, which we want to project into a
-lower-dimensional space.
-
-Truncation is a trivial dimension reduction technique: take the first $N$
-coordinates. More sophisticated projection functions consider the
-\emph{distribution} of the points, and project with the hyperplane which
-preserves as much of the variance as possible (or, equivalently, reduces the
-\emph{mutual information} between the points).
-
-There are many techniques to find these hyperplanes, such as \emph{principle
-  component analysis} (PCA) and \emph{autoencoding}. If the data are labelled,
-these classes can also be taken into account during dimension
-reduction~\cite{Oveisi.Oveisi.Erfanian.ea:2012}. However, since these
-techniques are effectively ML algorithms in their own right, they suffer some of
-the same constraints we're trying to avoid:
-
-\begin{itemize}
-  \item They operate \emph{offline}, requiring all input points up-front
-  \item All input points must have the same dimensionality
-\end{itemize}
-
-In particular, the second constraint is precisely what we're trying to
-avoid. Sophisticated dimension reduction is still useful for \emph{compressing}
-large, redundant features into smaller, information-dense representations, and
-as such provides a good complement to truncation.
-
-The requirement for offline ``batch'' processing is more difficult to overcome,
-since any learning we perform for feature extraction will interfere with the
-core learning algorithm that's consuming these features (this is why deep
-learning is often done greedily).
-
-\subsubsection{Sequences}
-
-To handle input of \emph{variable} size, research attention has been given to
-the handling of \emph{sequences}. This is a lossless approach, which splits the
-input into fixed-size \emph{chunks} (e.g. splitting text into a sequence of
-characters), which are fed into an appropriate ML algorithm one at a time. The
-sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by
-construction, is distinguishable from the data chunks. This technique allows us
-to trade \emph{space} (the size of our input) for \emph{time} (the number of
-chunks in a sequence).
-
-Not all ML algorithms can be adapted to accept sequences. One notable approach
-is to use \emph{recurrent ANNs} (RANNs), which allow arbitrary connections
-between nodes, including cycles. Compared to \emph{feed-forward} ANNs (FFANNs),
-which are acyclic, the \emph{future output} of a RANN may depend arbitrarily on
-its \emph{past inputs} (in fact, RANNs are universal computers).
-
-The main problem with RANNs, compared to the more widely-used FFANNs, is the
-difficulty of training them. If we extend the standard backpropagation algorithm
-to handle cycles, we get the \emph{backpropagation through time} algorithm
-\cite{werbos1990backpropagation}. However, this suffers a problem known as the
-\emph{vanishing gradient}: error values decay exponentially as they propagate
-back through the cycles, which prevents effective learning of delayed
-dependencies, undermining the main advantage of RANNs. The vanishing gradient
-problem is the subject of current research, with countermeasures including
-\emph{neuroevolution} (using evolutionary computation techniques to train an
-ANN) and \emph{long short-term memory} (LSTM; introducing a few special,
-untrainable nodes to persist values for long time periods
-\cite{hochreiter1997long}).
-
-An alternative to introducing recurrent connections, is the use of
-\emph{attention}. % TODO: cite
-In a (typically feed-forward) ANN with attention, the whole input is provided to
-the network, which produces some output as well as some ``attention''
-information; the network is run again, but with the input adjusted using this
-attention information (increasing and decreasing some of the input values); more
-output is produced, along with new attention information, and so on. Such
-networks do not need to ``remember'' long-distance relationships, whether by
-recurrent connections or gated nodes; they instead use the same input over and
-over again, but use their ``attention'' to emphasise those parts which should be
-relevant to future iterations.
-
-\subsubsection{Recursive Structure}
-
-Recursive structures, like trees and lists, have \emph{fractal} dimension:
-adding layers to a recursive structure gives us more \emph{fine-grained}
-features, rather than \emph{orthogonal} features. For data mining context-free
-languages (e.g. those of programming and theorem-proving systems), we will
-mainly be concerned with tree and graph structures of variable size.
-
-Using sequences to represent recursive structures is also problematic: if we
-want our learning algorithm to exploit structure (such as the depth of a token),
-it will have to discover how to parse the sequences for itself, which seems
-wasteful. The \emph{back-propagation through structure} approach
-\cite{goller1996learning} is a more direct solution to this problem, using a
-feed-forward NN to learn recursive distributed representations
-\cite{pollack1990recursive} which correspond to the recursive structure of the
-inputs. Such distributed representations can also be used for sequences, which
-we can use to encode sub-trees when the branching factor of nodes is not uniform
-\cite{kwasny1995tail}. More recent work has investigated storing recursive
-structures inside LSTM cells \cite{zhu2015long}.
-
-A simpler alternative for generating recursive distributed representations is to
-use circular convolution \cite{conf/ijcai/Plate91}. Although promising results
-are shown for its use in \emph{distributed tree kernels}
-\cite{zanzotto2012distributed}, our preliminary experiments in applying
-circular convolution to functional programming expressions found most of the
-information to be lost in the process; presumably as the expressions are too
-small.
-
-\emph{Kernel methods} have also been applied to structured information, for
-example in \cite{Gartner2003} the input data (including sequences, trees and
-graphs) are represented using \emph{generative models}, such as hidden Markov
-models, of a fixed size suitable for learning. Many more applications of kernel
-methods to structured domains are given in \cite{bakir2007predicting}, which
-could be used to learn more subtle relations between expressions than recurrent
-clustering alone.

### 2019-11-29 PhDThesis/bucketing.tex
Include better bucketing graph

@@
 The pseudorandom and clustering implementations are more realistic, since they
-do not use any information from the ground truth. These methods may also leave
-some of their subsets empty (since, unlike the optimal and pessimal methods,
-they are less prone to extremes).
+do not use any information from the ground truth. We also permit these methods
+to leave some of their subsets empty (since, unlike the optimal and pessimal
+methods, they are less prone to extremes).
 
@@
 \begin{figure}
-  % \input{images/bucketing}
-  \includegraphics[width=\textwidth]{images/bucketing.png}
-  \caption{Comparison of subset quality when selected by recurrent clustering
-    (solid lines) versus pseudorandom selection (dashed lines). Darker lines
-    show division into a greater number of subsets (between 1 and 20). The solid
-    and dashed lines follow each other closely, indicating that clustering
-    performs no better than random at signature selection.}
+  % \input{images/bucketed}
+  \includegraphics[width=\textwidth]{images/bucketed.png}
+  \caption{Quality evaluation of bucketing via recurrent clustering. Each plot
+    uses a fixed number of buckets (5, 10, 15 or 20) and shows the number of
+    extra theorems available when clusters are used as buckets, compared to a
+    pseudorandom control. The difference is not statistically significant. There
+    is a slight upward trend for larger sizes, but this is relatively small
+    compared to the corresponding increase in total theorems available.}
   \label{fig:bucketing}
@@
 substitutes fixed numbers for keywords and recurses to determine suitable
-numbers for referenced expressions.
+numbers for referenced expressions. These vectors are clustered based on their
+euclidean distance, using a \emph{k-means} algorithm, and these clusters are
+chosen as our buckets.
 
-The results of this method are shown in Figure~\ref{fig:bucketing}, along with
-the pseudorandom method as a control. The results of clustering turn out to be
-no better than random; in fact the curves show surprisingly similar patterns.
+The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
+the difference between bucketing based on this clustering approach, compared to
+simply choosing buckets pseudorandomly (as a control). The results of clustering
+are statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
+  test finds no significant difference for most of the plotted points; whilst
+  not a rigorous comparison (hence its ommission), this is nevertheless a clear
+  indication of their similarity.}
+
 \iffalse

### 2019-11-29 PhDThesis/support-for-bucketing/clustering_vs_random/grapher.py
Avoid conflicting filename

@@ -142,6 +142,6 @@ suplabels(x='Sample size', y='Extra theorems (vs random)')
 
 # Saving
 
-plt.savefig('bucketing.png', bbox_inches='tight', pad_inches=0.0)
-plt.savefig('bucketing.pdf', bbox_inches='tight', pad_inches=0.0)
-plt.savefig('bucketing.pgf', bbox_inches='tight', pad_inches=0.0)
+plt.savefig('bucketed.png', bbox_inches='tight', pad_inches=0.0)
+plt.savefig('bucketed.pdf', bbox_inches='tight', pad_inches=0.0)
+plt.savefig('bucketed.pgf', bbox_inches='tight', pad_inches=0.0)

### 2019-11-29 PhDThesis/litreview_interestingness.tex
Shrink table to fit
Fix unclosed footnote

@@
   \centering
+  \scalebox{0.9}{
     \begin{tabular}{ |l|l|c|c|c|c|c|c| }
@@
     \end{tabular}
+  }
   \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
@@
   positions which requires a growing number of stack frames or nested closures.
-  Reasoning about non-tail recursive definitions is often simpler, since they
-  don't expose accumulator arguments, so equivalence ), since the specific obligations which
-arise in the proof may be incompatible with the available inductive hypotheses.
-An informative example is given by Boyer and Moore of the associativity of
-multiplication in ACL2~\cite{boyer1983proof}:
+  The presence of accumulator arguments makes reasoning harder, so it is common
+  to use the direct implementation for proofs, then transport them over to the
+  tail-recursive definition via an equivalence relation.}), since the specific
+obligations which arise in the proof may be incompatible with the available
+inductive hypotheses. An informative example is given by Boyer and Moore of the
+associativity of multiplication in \textsc{ACL2}~\cite{boyer1983proof}:
 
@@
 examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
-tail-recursive and their non-tail-recursive counterparts, as well as the effect
-of repeating certain list operations:
+tail-recursive definitions with their direct (non-tail-recursive) counterparts,
+as well as the effect of repeating certain list operations:
 
 \begin{itemize}
+\item
+  $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) =
+  \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural
+  number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and
+  non-tail-recursive implementations of factorial, respectively.
 
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.
+\item
+  $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) =
+  \texttt{power}(n)$, where \texttt{power-tail} and \texttt{power} calculate
+  powers of 2.
 
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.
+\item
+  $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) =
+  \texttt{fib}(n)$, where \texttt{fib-tail} and \texttt{fib} calculate fibonacci
+  numbers.
 
-  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.
+\item
+  $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x))
+  = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp}
+  and list-sorting function \texttt{sort}.
 
-  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.
+\item
+  $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) =
+  x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list
+  structure and \texttt{rev} is list reversal.
 
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.
-
-  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).
-
+\item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$,
+  where \texttt{int} is the intersection of lists (i.e. a list of elements
+  common to each).
 \end{itemize}
@@
 plus :: Nat -> Nat -> Nat
-plus      n  Z    = n
-plus      n (S m) = S (plus n m)
+plus      Z      m = m
+plus      (S n') m = S (plus n' m)
 
 plus_tail :: Nat -> Nat -> Nat
-plus_tail n  Z    = n
-plus_tail n (S m) = plus_tail (S n) m
+plus_tail Z      m = m
+plus_tail (S n') m = plus_tail n' (S m)
 \end{haskell}

### 2019-11-29 PhDThesis/bucketing.tex
Finish off bucketing

@@
 \item Define the \emph{signature selection problem} for theory exploration.
-\item Analyse the effect of signature selection on the quantity of desirable
-  properties attainable by theory exploration on the TEB problem set.
 \item Identify the effect of ``toxic'' definitions on system performance, as
   measured by the TEB.
-\item Recommend improvements for theory exploration systems, to avoid or
-  mitigate the problem of toxic definitions.
+\item Recommend improvements for theory exploration systems, such as
+  resource-bounding of individual evaluations, to avoid or mitigate the
+  problem of toxic definitions.
+\item Analyse the effect of signature selection on the quality of discovered
+  results, by measuring how many desirable properties of a benchmark problem
+  (TEB) remain discoverable after splitting into multiple small signatures.
 \item Demonstrate how the signature selection problem is distinct from the
@@
   \scalebox{0.69}{\input{images/proportionsnontoxic.pgf}}
-  \caption[]{Each bar is a definition from our Theory Exploration Benchmark,
-    showing the proportion of signatures they appeared in which were
-    successfully explored (green) compared to those which timed out (red).
+  \caption[]{Each vertical bar is a definition from our Theory Exploration
+    Benchmark (TEB), showing the proportion of signatures they appeared in which
+    were successfully explored (green) compared to those which timed out (red).
     Definitions are sorted from most successful to least. The left graph shows
-    all \input{text/proportions_graph_count_all.tex} TEB definitions explored, with five ``toxic'' definitions which always
-    failed. with s ratio of successes to failures of the runs they appeared in. The first graph contains all TIP definitions, showing
-    ``toxic'' definitions which always failed. The second graph only contains
-    runs without any toxic definitions.}
+    all \input{text/proportions_graph_count_all.tex} TEB definitions explored,
+    where five ``toxic'' definitions always failed. In comparison, the right
+    graph ignores signatures containing those five definitions; the much higher
+    proportion of success highlights the disproportionate impact a few toxic
+    definitions can have.}
   \label{fig:proportions}
@@
   \caption{The average quality (proportion of ground-truth properties available)
-    when splitting signatures sampled using the Theory Exploration Benchmark
-    into subsets. Each plot shows division into a different number of subsets.
-    The shaded area is bounded by the optimal and pessimal methods, which can
-    only produce $N$ subsets for signatures of size $\geq N$. Grey lines show
-    pseudorandom selection, for comparison; these extend into smaller signature
-    sizes since empty subsets are permitted.}
+    when selecting subsets of definitions sampled from the Theory Exploration
+    Benchmark. Each plot shows division into a different number of subsets.
+    The shaded area is bounded by the optimal and pessimal methods, whose
+    constraints prevent selecting $N$ signatures if there are fewer than $N$
+    definitions to pick from. Grey lines show pseudorandom selection for
+    comparison; these are unconstrained, so extend into smaller signature
+    sizes.}
   \label{fig:bounds}
@@
   \includegraphics[width=\textwidth]{images/bucketed.png}
-  \caption{Quality evaluation of bucketing via recurrent clustering. Each plot
-    uses a fixed number of buckets (5, 10, 15 or 20) and shows the number of
-    extra theorems available when clusters are used as buckets, compared to a
-    pseudorandom control. The difference is not statistically significant. There
-    is a slight upward trend for larger sizes, but this is relatively small
-    compared to the corresponding increase in total theorems available.}
+  \caption{Evaluating the quality of selecting signatures using recurrent
+    clustering. Each plot uses a fixed number of subsets (5, 10, 15 or 20) and
+    shows the number of extra theorems available when clusters are used as these
+    subsets, compared to a pseudorandom control. The difference is not
+    statistically significant. There is a slight upward trend for larger sizes,
+    but this is relatively small compared to the corresponding increase in total
+    theorems available.}
   \label{fig:bucketing}
@@
 euclidean distance, using a \emph{k-means} algorithm, and these clusters are
-chosen as our buckets.
+chosen as the subsets to explore separately.
 
 The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
-the difference between bucketing based on this clustering approach, compared to
-simply choosing buckets pseudorandomly (as a control). The results of clustering
-are statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
+the difference between selecting clusters of definitions, compared to simply
+choosing pseudorandomly (as a control). The results of clustering are
+statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
   test finds no significant difference for most of the plotted points; whilst
   not a rigorous comparison (hence its ommission), this is nevertheless a clear
-  indication of their similarity.}
+  indication of their similarity.} Despite this, two trends are apparent: a
+slight increase in clustering's (relative) quality as the sample size increases,
+and a decrease in variance as the number of subsets is increased. The variance
+may be explained by the larger number of permutations available when using more
+subsets, which decreases the chance of finding an outlier (either good \emph{or}
+bad). The absolute increase in theorems found by clustering for large sizes, if
+real, is dwarfed by the corresponding increase in the ground truth (the total
+number of theorems available, without selecting subsets); the quality difference
+between these selection methods as a \emph{proportion} of the ground truth
+actually \emph{decreases} for larger sizes.
 
-\iffalse
-TODO: Alison: Does this translate into a recommendation (e.g. not to
-use clustering)? There should be some discussion of this - if not here then
-point to another chapter where you discuss these results
+\section{Summary}
 
-TODO: Alison: Needs a summary section
-\fi
+We have identified the \emph{signature selection problem} for theory
+exploration, which narrows the available definitions (e.g. from a software
+library) down to a subset fast enough to explore. This must currently be solved
+by the user, which imposes a knowledge burden (reducing TE's utility as an aid
+to learning) and limits how novel or surprising the discovered properties can
+be.
+
+We have demonstrated the effect of signature size (and hence the importance of
+selection) on the time taken to explore, in particular noting the ``heavy tail''
+of explorations timing out, which increases linearly with signature size. We
+have attributed this to ``toxic'' definitions, whose presence in a signature
+will exhaust the allocated time and/or memory. To mitigate this problem, and
+hence reduce the severity of signature selection, we recommend a more
+fine-grained approach to resource management and/or scheduling: by limiting the
+resources of individual tests, evaluation of toxic definitions can be terminated
+without having to abort the whole exploration run.
+
+We have shown the impact of signature selection on the quality of discoverable
+properties, by sampling sets of definitions from the Theory Exploration
+Benchmark and comparing how many of their associated theorems become impossible
+to discover (due to their definitions being separated) as these sets are
+subdivided by different selection methods. Optimal selection (calculated with
+knowledge of the benchmark theorems) retains a large proportion of the desired
+output, showing the potential to maintain high quality. Pseudorandom selection
+is found to be closer to pessimal, and selecting based on similarity (determined
+via an unsupervised machine learning algorithm) performs no different to random.
+
+Automated signature selection is hence an important problem, requiring further
+research to establish if, and how, the potential benefits can be reaped.
 

### 2019-11-29 PhDThesis/background_machinelearning.tex
Fix recursive ML equations

@@
 of a combining function $\bigoplus$, a predicate $\texttt{node?}$ and projection
-functions $\pi_{left}}$ and $\pi_{right}}$ as follows:
+functions $\pi_{left}$ and $\pi_{right}$ as follows:
 
-\begin{equation*}
-  c'(x) = \begin{cases}
-            c'(a) \bigoplus c'(b) & \text{if $x = \texttt{node}(a, b)$}
+\begin{align*}
+  c'(x) &= \begin{cases}
+             c'(a) \bigoplus c'(b) & \text{if $x = \texttt{node}(a, b)$} \\
              c(x)                  & \text{otherwise}
-          \end{cases}
-
-  d'(x) = \begin{cases}
+           \end{cases} \\
+  d'(x) &= \begin{cases}
              \texttt{node}(d'(\pi_{left }(x)),
-                          d'(\pi_{right}(x))) & \text{if \texttt(node?)(x)}
+                           d'(\pi_{right}(x))) & \text{if \texttt{node?}(x)} \\
              d(x)                              & \text{otherwise}
            \end{cases}
-\end{equation*}
+\end{align*}
 

### 2019-11-29 PhDThesis/bucketing.tex
Clustering text improvements

@@
   test finds no significant difference for most of the plotted points; whilst
-  not a rigorous comparison (hence its ommission), this is nevertheless a clear
-  indication of their similarity.} Despite this, two trends are apparent: a
-slight increase in clustering's (relative) quality as the sample size increases,
-and a decrease in variance as the number of subsets is increased. The variance
-may be explained by the larger number of permutations available when using more
-subsets, which decreases the chance of finding an outlier (either good \emph{or}
-bad). The absolute increase in theorems found by clustering for large sizes, if
-real, is dwarfed by the corresponding increase in the ground truth (the total
-number of theorems available, without selecting subsets); the quality difference
-between these selection methods as a \emph{proportion} of the ground truth
-actually \emph{decreases} for larger sizes.
+  not a rigorous overall comparison (hence its ommission), this is nevertheless
+  a clear indication of their similarity.} Despite this, two trends are
+apparent: a slight increase in clustering's (relative) quality as the sample
+size increases, and a decrease in variance as the number of subsets is
+increased. The variance may be explained by the larger number of permutations
+available when using more subsets, which decreases the chance of finding an
+outlier (either good \emph{or} bad). The absolute increase in theorems found by
+clustering for large sizes, if real, is dwarfed by the corresponding increase in
+the ground truth (the total number of theorems available, without selecting
+subsets); the quality difference between these selection methods as a
+\emph{proportion} of the ground truth actually \emph{decreases} for larger
+sizes.
 
@@
 will exhaust the allocated time and/or memory. To mitigate this problem, and
-hence reduce the severity of signature selection, we recommend a more
+hence the severity of signature selection's pruning, we recommend a more
 fine-grained approach to resource management and/or scheduling: by limiting the
@@
 
-We have shown the impact of signature selection on the quality of discoverable
-properties, by sampling sets of definitions from the Theory Exploration
+We have shown the impact of signature selection on the quality of theory
+exploration output, by sampling sets of definitions from the Theory Exploration
 Benchmark and comparing how many of their associated theorems become impossible
@@
 is found to be closer to pessimal, and selecting based on similarity (determined
-via an unsupervised machine learning algorithm) performs no different to random.
+via an unsupervised machine learning algorithm) performs no different than
+random.
 
-Automated signature selection is hence an important problem, requiring further
-research to establish if, and how, the potential benefits can be reaped.
+Automated signature selection is hence an important unsolved problem, requiring
+further research to establish if, and how, the potential benefits shown by our
+oracle-assisted algorithms can be reaped.
 
@@
 \fi
-
-%\section{Experiments}
-
-% TODO: This was taken from the Background section, so probably doesn't fit here
-% without some fiddling.
-% TODO: Describe all of our experimental setups:
-%
-%  - The use of the TEBenchmark methodology for running QSpec, including the
-%    sampling process
-%
-%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
-%    including how we picked the sizes
-%
-%  - The definition of HashSpec and its use as control; how we ran all of the
-%    same samples through it
-%
-%  - The use of optimal and pessimal oracles, defined via constraint
-%    satisfaction. How we it was only feasible to run these up to size 10 (11?)
-
-%We follow the tradition of prior practitioners and use an existing corpus of
-%theorems as a \emph{ground truth} against which to compare results. To analyse
-%behaviour on large inputs (i.e. those for which exploration is currently
-%infeasible), as required to study the effects of signature selection, we have
-%used the Theory Exploration Benchmark of section~\ref{sec:tebenchmark}, which is
-%based on the Tons of Inductive Problems (TIP) problem set for theorem
-%provers~\cite{claessen2015tip}.

### 2019-11-29 PhDThesis/clustering.tex
Clustering text improvements

@@
 
-\iffalse TODO: Alison: Each chapter should have a description at the start (see
-bucketing ``in this chapter'') \fi
-
 In this chapter we consider how the contents of a given signature can be
-analysed, such that subsequent exploration can be performed more efficiently. We
-frame this analysis as a task of \emph{unsupervised machine learning}, and
-describe how well-known \emph{clustering algorithms} can utilise the syntax of
-definitions to determine their similarity. This requires a detailed description
-of Haskell definition syntax, in particular of the ``Core'' intermediate
-language of the GHC compiler. We identify the difficulty posed by references,
-and how these can be made less opaque through the use of \emph{recurrent
-  clustering}.
+analysed syntactically, to aid theory exploration systems (either by improving
+efficiency, or finding more interesting properties). Inspired by the approaches
+of \aclml{} and \mlforpg{}, we frame this analysis as a task of
+\emph{unsupervised machine learning}, and describe how well-known
+\emph{clustering algorithms} can utilise the syntax of definitions to determine
+their similarity. This requires a detailed description of Haskell syntax, in
+particular we use the ``Core'' intermediate language of the GHC compiler. We
+identify the difficulty posed by references, and how these can be made less
+opaque through the use of \emph{recurrent clustering}.
+
+Our main contribution is a recurrent clustering method to determine the
+syntactic similarity of Haskell definitions, which is transitive across
+references (e.g. the similarity between definitions \hs{A} and \hs{B}, which
+reference definitions \hs{C} and \hs{D} respectively, is correlated to the
+similarity between \hs{C} and \hs{D}). We also provide an iterative
+algorithm, which is more efficient than a straightforward (na\"ive)
+implementation of the defined recurrence relation. We have made a prototype of
+our system available, including a plugin for the \textsc{GHC} compiler to
+enable end-to-end automation, taking a standard Haskell package and producing
+clusters of definitions; these may subsequently be explored, e.g. via
+\quickspec{}, which we investigate further in \S~\ref{sec:signatureselection}.
+
+Without an independent ``gold standard'' to compare our results against, we
+instead perform a qualitative analysis; using (translations of) examples from
+\aclml{} and real Haskell packages.
 
@@
 \fi
-W
+
+% FIXME

### 2019-11-29 PhDThesis/clustering.tex
Incorporate a bunch of commented-out content into the text

@@
 handles recursive tree structures, resolves references and produces numeric
-values. We suggest a new recurrent clustering and feature extraction algorithm
-for Haskell (shown in Algorithm~\ref{alg:recurrent}) and compare its similarity
-and differences to the existing approaches of \textsc{ML4PG} and
-\textsc{ACL2(ml)}.
+values. There are two recurrent clustering approaches in the literature: those
+of \mlforpg{} and \aclml{}. The former produces features which \emph{summarise}
+the elements of a syntax tree: providing, for each level of the tree, the number
+of variables, nullary symbols, unary symbols, etc. In contrast, that of
+\mlforpg{} maps the elements of a syntax tree to distinct cells in a matrix,
+then flattens the matrix into a vector.
+
+Our approach, shown in Algorithm~\ref{alg:recurrent}, follows that of
+\mlforpg{}, but can additionally handle mutually-recursive definitions. We
+ignore types, since obtaining the type of each component of a Core definition is
+more difficult than in Coq (\mlforpg{} queries the current
+\textsc{Proof General} state inside Emacs), and is left for future work.
 
@@
 
-\iffalse
-% TODO: This seems less appropriate for this chapter, more suitable for
-% Signature Selection
-A separate application called \mlspec{} can read in this JSON, generate
-appropriate Haskell code to invoke \quickspec{} on each cluster individually,
-and evaluate that code with all appropriate dependencies installed. This
-decoupling allows our feature extraction algorithm to be used for other
-purposes, and allows alternative clustering involves:
-
-\begin{itemize}
-\item{Monomorphising}: If a value has polymorphic type, e.g. \hs{equal :: forall
-    t. t -> t -> Bool}, we must choose a concrete representation to use (in this
-  case for \hs{t}), in order to know which random generator to use. We take the
-  approach used in \quickcheck{} by attempting to instantiate all type variables to
-  \hs{Integer}. Any values where this is invalid, such as those with
-  incompatible class constraints (e.g. \hs{forall t. IsString t => t}, where
-  \hs{Integer} does not implement \hs{IsString}) will not be included in the
-  signature (this is checked at the AST post-processing stage).
-
-\item{Qualifying}: All names are \emph{qualified} (prefixed by their module's
-  name), to avoid most ambiguity. There is still the possibility that multiple
-  packages will declare modules of the same name, although this is rare as it
-  causes problems for any Haskell programmer trying to use those modules. In
-  such cases the exploration process simply aborts.
-
-\item{Appending variables}: Once a \quickspec{} theory has been defined containing
-  all of the given terms, we inspect the types it references and append three
-  variables for each to the theory (enough to discover laws such as
-  associativity, but not too many to overflow the limit of \quickspec{}'s exhaustive
-  search).
-
-\item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is
-  that all required packages and modules must be provided up-front, usually by
-  specification in a Cabal file. Since \mlspec{} builds signatures
-  \emph{dynamically}, depending on the cluster information it is given, we do
-  not know what packages it may need. To work around this problem, we evaluate
-  these strings of Haskell using a custom library called \texttt{nix-eval}. This
-  uses the Nix package manager to obtain all of the required packages and make
-  them available to GHC, and is described further in
-  Appendix~\ref{sec:nix-eval}.
-
-\end{itemize}
-
-The equations resulting from evaluating these strings are collected and
-outputted as JSON values, to ease further processing (e.g. displaying in some
-form, sending to a theorem prover, etc.).
-\fi
+These clusters can be fed into an augmented version of \quickspec{}, which makes
+decisions such as monomorphising types, choosing variables, etc. and is
+described further in \S~\ref{sec:implementation}. The utility of splitting up
+large signatures into smaller parts like this is the ``signature selection
+problem'', which we investigate in \S~\ref{sec:signatureselection}.
 
@@
 \label{sec:evaluation}
-
-\iffalse
-Here we consider the 10 most-downloaded packages from \hackage{} (as of
-2015-10-30) which have property tests; these are \texttt{warp}, \texttt{aeson},
-\texttt{text}, \texttt{lens}, \texttt{conduit}, \texttt{pandoc},
-\texttt{attoparsec}, \texttt{scientific}, \texttt{yesod-core} and
-\texttt{blaze-html}.
-
-% TODO: Analysis
-\fi
-
-\iffalse
-TODO: Alison: Next chapter? It's good to lead into the next chapter at the end
-of the previous ones.
-\fi
 
@@
 
-\iffalse
-Whilst recurrent clustering has produced results which merit further
-investigation, the application to theory exploration has yet to be tested
-empirically. This is due to \quickspec{}'s use of \quickcheck{}'s \hs{Arbitrary} type
-class to generate random values for instantiating variables. Whilst we can
-automatically define \quickspec{} theories and invoke them with \hs{nix-eval}, not
-all types have \hs{Arbitrary} instances; those without cannot be given any
-variables in our signature, which severely limits the possible combinations
-which can be explored. In many cases, no variables can be included at all,
-leaving just equations involving constants. This has so far prevented us from
-measuring the direct impact on \quickspec{} performance, either directly by
-exploring the sub-sets identified through recurrent clustering, or indirectly by
-comparing the equations generated by a full brute-force search to our recurrent
-clusters: those equations relating terms from different clusters would not be
-discovered by our method. It is this ratio of equations found through brute
-force to those found after narrowing-down by clusters which is one of our key
-objectives to maximise at this stage; until we begin to pursue the
-\emph{interestingness} of the properties.
-
-The following less-serious problems were also encountered while applying
-\mlforhs{} to \hackage{} packages:
-
-\begin{itemize}
-\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
-  declarations to cluster. This is because they make all of their declarations
-  privately, e.g. in ``internal'' modules, then use separate modules to export
-  the public declarations. GHC's renaming phase makes all references to such
-  exports canonical, by pointing them to the private declarations. This forces
-  us to ignore such declarations, as \quickspec{} will not be able to access them.
-
-\item Since we do not support type-level entities, we ignore type
-  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
-  ``methods'') which occur in a type class instance. Instead of being clustered,
-  these result in references getting $f_{recursion}$ features. This is
-  especially noticable in libraries like \hs{scientific}, where only the
-  functions for constructing and destructing numbers in scientific notation are
-  clustered; all of the arithmetic is defined in type classes. One difficulty
-  with supporting methods is that their namespace in Core is disjoint from that
-  of regular Haskell identifiers: a transformation layer would be required,
-  along with explicit type annotations to avoid ambiguity.
-\end{itemize}
-\fi
-
-These clusters can then be used in several ways to perform theory exploration;
-the most na\"{\i}ve way being to explore each cluster as \quickspec{} signature
-in its own right, which we investigate as a potential solution to the Signature
-Selection problem in \S~\ref{sec:signatureselection}.
-
-\iffalse
-TODO: Alison: Next chapter? It's good to lead into the next chapter at the end of the previous ones.
-\fi
-
-\iffalse
-As alluded to previously, we also have the opportunity to reason by
-\emph{analogy}. Similar to the work on ACL2(ml) \cite{heras2013proof}, we could
-produce a general ``scheme'' from each equation we find (either through \quickspec{}
-or by data mining test suites); like Isabelle approaches have shown
-\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
-instantiated to a variety of similar values, in an attempt to find new theorems
-which are analogues of existing results from a different context. ``Mutating''
-existing theorem statements in such a way would also increase the chance of any
-result being considered interesting; since it's likely that the unmodified
-statement was deemed interesting, and the new result would not in general follow
-as a simple logical consequence.
-\fi
-
-\iffalse
-\subsection{Recurrent Clustering}
-\label{sec:clusteringexpressions}
-
-Our recurrent clustering algorithm is similar to that of ML4PG, as our
-transformation maps the elements of a syntax tree to distinct cells in a
-matrix. In contrast, the matrices produced by ACL2(ml) \emph{summarise} the tree
-elements: providing, for each level of the tree, the number of variables,
-nullary symbols, unary symbols, etc.
-
-Unlike ML4PG, our algorithm can handle mutually-recursive definitions, but it
-also ignores types. This is because obtaining the type of each component of a
-Core definition is more difficult than in ML4PG (which queries the current
-\textsc{Proof General} state), and is left for future work.
-
-The way we \emph{use} our clusters to inform theory exploration is actually more
-similar to that of ACL2(ml) than ML4PG. ML4PG can either present clusters to the
-user for inspection, or produce automata for recreating proofs. In ACL2(ml), the
-clusters are used to restrict the search space of a proof search, much like we
-restrict the scope of theory exploration.
-
 \section{Conclusion and Future Work}
@@
 information about an expression, and would allow us to distinguish between
-constructors. Since our algorithm closely follows that of ML4PG, which
+constructors. Since our algorithm closely follows that of \mlforpg{}, which
 \emph{does} support types, the only barrier is the practical issue of
@@
 More speculative directions include the use of \emph{learned} representations,
-rather than our hand-crafted features \cite{bengio2013representation}. This
+rather than our hand-crafted features~\cite{bengio2013representation}. This
 would provide an interesting comparison, as well as being more robust in the
@@
 during clustering; for example, by treating the discovered equations as rewrite
-rules to reduce the ASTs prior to feature extraction.
-\fi
-
-% FIXME
+rules to reduce the ASTs prior to feature extraction. Version 2 of \quickspec{}
+already constructs such rewrite rules (via the Knuth-Bendix completion
+algorithm) to determine whether a generated expression reduces to an existing
+equivalence class.
+
+Similarity measures like ours have other applications, such as the opportunity
+to reason by \emph{analogy}. Similar to the work on
+\aclml{}~\cite{heras2013proof}, we could produce a general ``scheme'' from each
+equation we find (either through \quickspec{}, or by other means like data
+mining benchmarks and test suites); like Isabelle approaches have shown
+\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, these schemes could then be
+instantiated to a variety of ``similar'' values, in an attempt to find new
+theorems which are analogues of existing results from a different context.
+Such ``mutations'' of existing theorem statements increases the chance of any
+result being considered interesting; since it's likely that the unmodified
+statement was deemed interesting, and the new result would not in general follow
+as a simple logical consequence.

### 2019-11-29 PhDThesis/ml4hs.tex
Incorporate a bunch of commented-out content into the text

@@
 \end{itemize}
+
+\subsection{Difficulties}
+
+The application to theory exploration to real-world Haskell packages is
+complicated due to \quickspec{}'s use of \quickcheck{}'s \hs{Arbitrary} type
+class to generate random values for instantiating variables. Whilst we can
+automatically define \quickspec{} theories and invoke them with \hs{nix-eval},
+not all types have \hs{Arbitrary} instances; those without cannot be given any
+variables in our signature, which severely limits the possible combinations
+which can be explored. In many cases, no variables can be included at all,
+leaving just equations involving constants. This has so far prevented us from
+measuring the direct impact on \quickspec{} performance, either directly by
+exploring the sub-sets identified through recurrent clustering, or indirectly by
+comparing the equations generated by a full brute-force search to our recurrent
+clusters: those equations relating terms from different clusters would not be
+discovered by our method. It is this ratio of equations found through brute
+force to those found after narrowing-down by clusters which is one of our key
+objectives to maximise at this stage; until we begin to pursue the
+\emph{interestingness} of the properties.
+
+The following less-serious problems were also encountered while applying
+\mlforhs{} to \hackage{} packages:
+
+\begin{itemize}
+\item Some packages, such as \texttt{warp} and \texttt{conduits}, get no
+  declarations to cluster. This is because they make all of their declarations
+  privately, e.g. in ``internal'' modules, then use separate modules to export
+  the public declarations. GHC's renaming phase makes all references to such
+  exports canonical, by pointing them to the private declarations. This forces
+  us to ignore such declarations, as \quickspec{} will not be able to access them.
+
+\item Since we do not support type-level entities, we ignore type
+  classes. Unfortunately, this also means ignoring any value-level bindings (AKA
+  ``methods'') which occur in a type class instance. Instead of being clustered,
+  these result in references getting $f_{recursion}$ features. This is
+  especially noticable in libraries like \hs{scientific}, where only the
+  functions for constructing and destructing numbers in scientific notation are
+  clustered; all of the arithmetic is defined in type classes. One difficulty
+  with supporting methods is that their namespace in Core is disjoint from that
+  of regular Haskell identifiers: a transformation layer would be required,
+  along with explicit type annotations to avoid ambiguity.
+\end{itemize}

### 2019-11-29 PhDThesis/background.tex
Thesis bits
Headingy things
Background changes
TODOs
Remove some duplicate text
A bunch of cleanup

@@
-\section{Background}
+\chapter{Background}
+
+% Background should cover approaches to automated reasoning about software, and
+% formal systems in general. We can highlight how the language/representation
+% imposes invariants and guarantees on what we can know, and how we can extend
+% those guarantees with features like purity (prevents behaviour depending on
+% ambient environment/state), lazy evaluation (prevents problems with
+% non-termination), sound type systems (limits the space of possible values for
+% each expression), etc. and therefore how Haskell is a natural vehicle for us to
+% use. Giving background on Haskell naturally leads to descendants like Agda and
+% Idris, which are also interactive theorem provers; which lets us discuss
+% Isabelle, Coq, HOL, etc., how they're difficult for existing reasoning
+% algorithms to handle (e.g. resolution, which works well in first-order logics),
+% approaches to helping
+
+This chapter provides background on the fields we investigate, and the
+techniques we apply. It includes descriptions of the proof- and
+programming-languages we use, and justifications for their use in our work. A
+very short summary of relevant Machine Learning topics is also given, to provide
+context for the techniques we use (namely unsupervised clustering) and the
+potential alternatives and future directions that are most applicable.
 
@@
 \input{background_theoryexploration.tex}
-\input{background_conjecturegeneration.tex}
+
+% FIXME: This seems to be repetitive of Theory Exploration?
+% \input{background_conjecturegeneration.tex}
+
 \input{background_machinelearning.tex}
-\input{background_recurrentclustering.tex}

### 2019-11-29 PhDThesis/background_machinelearning.tex
Remove some duplicate text
A bunch of cleanup

@@
 learn a sample of \emph{relevant} details, known as \emph{features}. This has
-two benefits:
+a few benefits:
 
@@
 
- One common approach to this problem is
-to represent such structures as \emph{sequences}. \emph{Recurrent neural
-  networks} (RNNs) are a popular choice for processing sequences, especially
-when combined with mechanisms such as \emph{long short-term memory} (LSTM) for
-preserving information across long sequences \cite{hochreiter1997long}. Such
-systems have been used, for example, to parse and execute computer programs
-\cite{zaremba2014learning}. However, learning to parse sequences seems
-inefficient considering that we already have correctly-parsed ASTs.
-
-Whilst neural networks have been applied directly to recursive structures
-\cite{goller1996learning}, including using LSTM \cite{zhu2015long}, a more
-popular approach is to use \emph{kernel methods}
-\cite{bakir2007predicting}. These are promising as a more principled alternative
-to our current hand-crafted translation of ASTs to vectors.
-
-\input{background_clustering.tex}
+% FIXME: Include clustering

### 2019-11-29 PhDThesis/background_theoryexploration.tex
A bunch of cleanup

@@
 
-\iffalse
-\quickspec{} (and \hipspec{}) is also compatible with Haskell's existing testing
+\subsection{Related Fields}
+
+\quickspec{} (and \hipspec{}) grew out of software engineering tools like
+\quickcheck{}, and they are in fact compatible with Haskell's existing testing
 infrastructure, such that an invocation of \texttt{cabal test} can run these
@@
 
-In fact, there are similarities between the way a TE system like \quickspec{} can
-generalise from checking \emph{particular} properties to \emph{inventing} new
+There are similarities between the way a TE system like \quickspec{} can
+generalise from checking \emph{particular} properties to \emph{generating} new
 ones, and the way counterexample finders like \quickcheck{} can generalise from
-testing \emph{particular} expressions to \emph{inventing} expressions to
-test. One of our aims is to understand the implications of this generalisation,
-the lessons that each can learn from the other's approach to term generation,
-and the consequences for testing and QA in general.
-\fi
+testing \emph{particular} expressions to generating expressions to test.
 
-% TODO
-%%%%% THE BELOW SEEMS TO BE INTERESTINGNESS?
+\emph{Experimental mathematics} is related to theory exploration, in that both
+seek to use empirical methods to discover high-level properties of
+well-specified formal systems. The theory is itself may be a scientific model,
+in which case theory exploration becomes part of the scientific method; in
+particular, finding ``testable predictions'' (e.g. equations relating measurable
+quantities, which can be experimentally tested) is a special case of theory
+exploration's search for ``interesting conjectures''.
 
-%\begin{description}
-%\item{Interestingness} Various alternative interestingness criteria have been
-%  proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing
-%  the criteria may be useful, for example to distinguish useful relationships
-%  from incidental coincidences; or to prevent surprising, insightful equations
-%  from being discarded because they can be simplified.
-%\end{description}
-
-% Theory exploration is similar to \emph{experimental mathematics}
-
-% - Relation to Science
-%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
-% - Relation to AI tasks: exploring surroundings, etc.
-
-% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.
-
-% Theory formation: Alison? Others.
-% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
-% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
-% Communication: Latex, Wikis, APIs, communicating with aliens
+\emph{Theory formation} and \emph{concept formation} go beyond mere exploration;
+inventing new definitions, which can give rise to subsequent properties. This
+gives rise to a feedback loop: concepts are ``interesting'' if they have
+``interesting'' properties, whilst properties are ``interesting'' due to the
+concepts they involve. This gives rise to some deep philosophical questions
+about the nature of mathematics, but is so open-ended that it can be difficult
+to apply to day-to-day tasks; hence why we generally avoid invention systems
+in this work.

### 2019-11-29 PhDThesis/bucketing.tex
A bunch of cleanup

@@
 \fi
+
+\iffalse
+Our approach to scaling up \quickspec{} takes inspiration from two sources. The
+first is relevance filtering, which makes expensive algorithms used in theorem
+proving more practical by limiting the size of their inputs. We describe this
+approach in more details in \S~\ref{sec:relevance}. Relevance filtering is a
+practical tool which has existing applications in software, such as the
+\emph{Sledgehammer} component of the Isabelle/HOL theorem prover.
+
+Despite the idea's promise, we cannot simply invoke existing relevance filter
+algorithms in our theory exploration setting. The reason is that relevance
+filtering is a supervised learning method, i.e. it would require a distinguished
+expression to compare everything against. Theory exploration does not have such
+a distinguished expression; instead, we are interested in relationships between
+\emph{any} terms generated from a signature, and hence we must consider the
+relevance of \emph{all terms} to \emph{all other terms}.
+
+A natural fit for this task is \emph{clustering}, which attempts to group
+similar inputs together in an unsupervised way. Based on their success in
+discovering relationships and patterns between expressions in Coq and ACL2 (in
+the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering
+methods can fulfil the role of relevance filters for theory exploration:
+intelligently breaking up large signatures into smaller ones more amenable to
+brute force enumeration, such that related expressions are explored together.
+\fi

### 2019-11-29 PhDThesis/litreview_automatedreasoning.tex
A bunch of cleanup

@@
 
-The successor of AM is Eurisko, a discovery system intended to be useful in more
-general domains than just mathematical theory formation. Despite claiming some
-early successes % FIXME Traveller TCS
-work on Eurisko was put on hold due to a lack of machine-readable data about
-real world systems for it to work with % FIXME Find something quotable
-. This lead to the Cyc project, in an attempt to encode such knowledge in a
-semantically rich and consistent form. Cyc is an ongoing project, with parts of
-its database made available via OpenCyc.
+The successor of AM is Eurisko~\cite{lenat83}, a discovery system intended to be
+useful in more general domains than just mathematical theory formation. Despite
+claiming some early successes (such as winning the ``Traveller TCS'' game) work
+on Eurisko was put on hold in favour of the Cyc project~\cite{lenat1995cyc}: an
+attempt to encode real-world knowledge in a semantically rich and consistent
+form, for consumption by Eurisko-like algorithms. Cyc is an ongoing project,
+with parts of its database made available via OpenCyc.
 

### 2019-11-29 PhDThesis/related_formalstats.tex
A bunch of cleanup

@@
 
-\iffalse
 \subsection{Clustering}
@@
 
-% TODO: ML4PG
-\cite{journals/corr/abs-1212-3618}
-% TODO: ACL2(ml) (also examples section)
-\cite{heras2013proof}
-\fi
+The most relevant related work for our recurrent clustering approach is
+\mlforpg{}~\cite{journals/corr/abs-1212-3618}. This clusters expressions in the
+Coq proof assistant, for presenting to the user (e.g. if they are stuck on a
+proof, it may be useful to read ``similar'' proofs). Like our approach, this
+transforms syntax trees into a matrix of values, which is then flattened into a
+vector. We use the same idea of clustering referred-to definitions in order to
+attain transitive similarity; however, our approach must also handle mutual
+recursion (which is much more restricted in Coq, and hence unsupported by
+\mlforpg{}).
+
+The other prior work for recurrent clustering is\aclml{}~\cite{heras2013proof},
+which is similar to \mlforpg{} but applies to ACL2 (Common Lisp) and is hence
+untyped. The algorithm also differs, since it \emph{summarises} trees, rather
+than directly encoding them to vector elements.
 

### 2019-12-02 PhDThesis/benchmark.tex
A bunch of cleanup
Fix diagram spacing

@@
 \label{sec:benchmark}
-
-\iffalse TODO: Alison: Each chapter should have a description at the start (see
-bucketing ``in this chapter'') \fi
 
@@
   \centering
-  \begin{tikzpicture}[node distance=1cm]
+  \scalebox{0.8}{\begin{tikzpicture}[node distance=1cm]
 
@@
           \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=1.3cm]
        (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
-         node [black, midway, right, xshift=0.3cm]
+         node [black, midway, right, xshift=0.5cm]
               {Preparation $\S$\ref{section:prep}};
@@
           \p{rec} = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=1.3cm]
        (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
-         node [black, midway, right, xshift=0.3cm]
+         node [black, midway, right, xshift=0.5cm]
               {Sampling $\S$\ref{section:sampling}};
@@
           \p{rec}  = (rec.south east)
-       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
+       in [decorate,decoration={brace, amplitude=10pt}, xshift=1.3cm]
        (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
-         node [black, midway, right, xshift=0.3cm]
+         node [black, midway, right, xshift=0.5cm]
               {Evaluation $\S$\ref{section:evaluation}};
 
-  \end{tikzpicture}
+  \end{tikzpicture}}
   \caption[]{High-level view of the benchmarking methodology described in
